### 2019-09-15 pldi/draft.tex
Add PLDI template

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title{Distribution-oblivious Programming}         %% [Short Title] is optional;
% \titlenote{with title note}             %% \titlenote is optional;
%                                         %% can be repeated if necessary;
%                                         %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\sys}{{\scshape Dish}\xspace}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{../../bib/bib}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-09-16 pldi/draft.tex
Import modules
Dump definitions from other papers
Add code styling

@@
 %\documentclass[sigplan]{acmart}\settopmatter{}
+
+\usepackage{enumitem}
+\usepackage{booktabs}
+\usepackage{amssymb}
+\usepackage{soul}
+\usepackage{xspace}
+\usepackage{color}
+\usepackage{xcolor}
+\usepackage{upquote}
+\usepackage{listings}
+\usepackage{amsmath}
+
+\captionsetup[figure]{font=footnotesize,name={Fig.},labelfont={bf, footnotesize}}
+\captionsetup[table]{font=footnotesize,name={Tab.},labelfont={bf, footnotesize}, skip=2pt, aboveskip=2pt}
+\captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
+
+\newcommand{\eg}{{\em e.g.}, }
+\newcommand{\ie}{{\em i.e.}, }
+\newcommand{\etc}{{\em etc.}\xspace}
+\newcommand{\vs}{{\em vs.} }
+\newcommand{\cmpn}{compartmentalization} 
+\newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
+\newcommand{\ttt}[1]{\texttt{\small #1}}
+\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
+\newcommand{\spol}[1]{\scriptsize{\sc#1}}
+\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
+\newcommand{\rf}[1]{\ref{#1}}
+\newcommand{\wka}{\ttt{a\textsubscript{1}}}
+\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}
+
+% For comments
+\newcommand{\eat}[1]{}
+\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
+\newcommand{\todo}[1]{\hl{#1}\xspace}
+\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
+\newcommand{\review}[1]{{\color{red}#1}}
+
+\definecolor{editorGray}{rgb}{0.95, 0.95, 0.95}
+\definecolor{editorOcher}{rgb}{1, 0.5, 0} % #FF7F00 -> rgb(239, 169, 0)
+\definecolor{editorGreen}{rgb}{0, 0.5, 0} % #007C00 -> rgb(0, 124, 0)
+
+\definecolor{cdb}{rgb}{0.37, 0.62, 0.63} % cadet blue
+
+\lstdefinelanguage{sh}{
+  morekeywords={for, in, do, done, \|},
+  keywordstyle=\color{purple}\ttfamily,
+  % ndkeywords={curl, grep, wget, awk, xargs, find, nc, mdc, gunzip, cut, sort, head, join},
+  ndkeywordstyle=\color{black}\ttfamily\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  comment=[l]{\#},
+  commentstyle=\color{lightgray},
+% morecomment=[s]{/\\*\\*, \\*/},
+  stringstyle=\color{darkgray}\ttfamily,
+  morestring=[b]',
+  morestring=[b]",
+% numbersep=1pt,
+% numberstyle=\footnotesize\bf\color{gray},   % the style that is used for the line-numbers
+  abovecaptionskip=0pt,
+  aboveskip=0pt,
+  belowcaptionskip=0pt,
+  belowskip=0pt,
+  frame=none                     % adds a frame around the code
+% moredelim=[s][\color{gray}]{c:}{>},
+% moredelim=[s][\color{orange}]{/*}{/}
+}
+
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
+  basicstyle=\small\ttfamily,  % the size of the fonts that are used for the code
+  upquote=true,
+  captionpos=b,                    % sets the caption-position to bottom
+% frame=B,                    % adds a frame around the code
+  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=2pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{gray},   % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  framerule=0pt,
+	xleftmargin=0pt,
+	xrightmargin=0pt,
+	breakindent=0pt,
+  aboveskip=0pt,
+  framesep=0pt,
+  abovecaptionskip=0pt,
+  aboveskip=0pt,
+  belowcaptionskip=0pt,
+  belowskip=0pt,
+  frame=none,
+  framexbottommargin=0pt,
+  resetmargins=true
+}
 

### 2019-09-16 pldi/draft.tex
Add notes on intro

@@
 
+%% Abstract
+%% Note: \begin{abstract}...\end{abstract} environment must come
+%% before \maketitle command
+\begin{abstract}
+  Distributed systems offer notable benefits over centralized systems.
+  Reaping these benefits, however, requires developers to develop and deploy
+  software in a manner that makes distribution explicit---a process
+  that is quite different from how they compose software for centralized
+  systems.
+% 
+  \emph{Light-touch distribution} is a new approach for scaling out distribution-oblivious shell scripts.
+  It leverages the fact that 
+% introduced as a drop-in replacement of a language runtime's module system, 
+  Our \sys prototype shows substantial speedups over 
+  as well as the ability to operate on large input sizes, all without
+  any developer input.
+\end{abstract}
+
 %% 2012 ACM Computing Classification System (CSS) concepts
@@
 
+\section{Introduction}
+
+Distributed systems offer significant benefits over their centralized counterparts.
+By parallelizing computation they can  
+can speed up computations, mitigate resource-exhaustion attacks, improve fault-tolerance, and balance load during spikes.
+% offer notable benefits over their centralized counterparts:
+%   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
+Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
+The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
+
+Despite these benefits, the development and deployment of distributed software remains different from and significantly more difficult than its centralized counterpart.
+Whereas anyone can stitch together a Bash script to compute on a single machine, 
+  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
+   scaling out to multiple computers requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{maas2014holistic, schwarzkopf2013dios, andromeda:hotos:2015, jonas2017occupy, varghese2018next, nefele:18}.
+Consider the following Unix pipeline featured in Bentley's Programming Pearls column, slightly modified to work on today's Unix.
+
+% \begin{lstlisting}[language=sh,float=h,numbers=none]
+% cat doc.ms |                   # print input file
+% groff -t -e -mandoc -Tascii |  # remove formatting
+% col -bx |                      # remove backspaces 
+% tr A-Z a-z |                   # convert to lower case
+% tr -d '[:punct:]' |            # remove punctuation
+% sort |                         # put words in alphabetical order
+% unique |                       # remove duplicate words
+% comm -13 /usr/share/dict -     # report words not in dictionary 
+% \end{lstlisting}
+
+\begin{lstlisting}[language=sh,float=h,numbers=none]
+ cat *.txt |
+ tr -cs A-Za-z`\n` |
+ tr A-Z a-z |
+ sort |
+ uniq -c |
+ sort -rn |
+ head 5 > output.txt
+\end{lstlisting}
+
+This pipeline takes an input file (line 1), removes
+
 %% Acknowledgments

### 2019-09-17 pldi/draft.tex
Update preamble
Minor notes in intro

@@
 %% Title information
-\title{Distribution-oblivious Programming}         %% [Short Title] is optional;
+\title{Distribution-oblivious Programming with Dish}         %% [Short Title] is optional;
 % \titlenote{with title note}             %% \titlenote is optional;
@@
 \newcommand{\sys}{{\scshape Dish}\xspace}
+\newcommand{\unix}{{\scshape Unix}\xspace}
+
+\setlist{noitemsep,leftmargin=10pt,topsep=2pt,parsep=2pt,partopsep=2pt}
 
@@
 
+% 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
+% They are an easy way to , because they combine a set of assumptions that work well with Unix
+% They work great on a single machine, but are difficult to scale out
+% Could we fully automate distribution? 
+% The key insight is that pipelines already express a domain-specific
+% computation that is easily amenable to distribution.
+%
+
 Distributed systems offer significant benefits over their centralized counterparts.
-By parallelizing computation they can  
-can speed up computations, mitigate resource-exhaustion attacks, improve fault-tolerance, and balance load during spikes.
+For example, they can speed up computations and can process amounts that would not fit into any single machine.
 % offer notable benefits over their centralized counterparts:
 %   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
-Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
-The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
+% Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
+% The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
 
-Despite these benefits, the development and deployment of distributed software remains different from and significantly more difficult than its centralized counterpart.
-Whereas anyone can stitch together a Bash script to compute on a single machine, 
+Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
+Whereas anyone can stitch together a Bash script to compute on a single computer, 
   % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-   scaling out to multiple computers requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{maas2014holistic, schwarzkopf2013dios, andromeda:hotos:2015, jonas2017occupy, varghese2018next, nefele:18}.
-Consider the following Unix pipeline featured in Bentley's Programming Pearls column, slightly modified to work on today's Unix.
+   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{maas2014holistic, schwarzkopf2013dios, andromeda:hotos:2015, jonas2017occupy, varghese2018next, nefele:18}.
+
+To understand this sharp contrast, consider calculating frequencies of terms within a set of input files.
+For a developer with \emph{one} computer, this is achievable with a short \unix pipeline:
 
@@
 % \end{lstlisting}
-
+% greece; stops
+% AWS policy specification
 \begin{lstlisting}[language=sh,float=h,numbers=none]
- cat *.txt |
- tr -cs A-Za-z`\n` |
- tr A-Z a-z |
- sort |
- uniq -c |
- sort -rn |
- head 5 > output.txt
+ cat * | tr -cs A-Za-z`\n` | tr A-Z a-z |
+   sort | uniq -c | sort -rn | head 5 > out
 \end{lstlisting}
 
-This pipeline takes an input file (line 1), removes
+Combining several features, Unix makes small tasks easy to express;
+  composition allows the composition of general primitives
+Key elements of \unix are the ability to stitch together programs written in different languages, the abstraction of a file system as a set of resident streams, as well as naming---the ability to resolve command, file names within a global context.
+Under the hood, the \unix kernel buffers results, concurrency control
 
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
+% Key insight 
+% % The garden-hose philosophy is similar to Haskell 
+% This pipeline takes an input file (line 1), removes 
+% * do not need to use specialized frameworks for composition---instead, as long as they conform to the shell, individual primitives can be written in any programming language.
+% * do not need to rewrite ---
+
+Unfortunately, 
+A developer with many computers has to choose between three options, all of which require significant manual effort.
+The most popular is to leverage domain-specific ; this requires rewriting portions of the program in a new language and use operations
+
+% If developers have already expressed their computation as a Unix pipeline, they should not have to manually rewrite the program in other environment---\eg Hadoop, Spark---to exploit distribution.
+For simple, one-off pipelines that take a few minutes to compose this cost---order of magnitude higher than---becomes unjustifiable.
+For many legacy pipelines  
+Could the generation and execution of a distributing version of the pipeline be automated?
+
+The key insight behind our work is that Unix pipelines already provide most of the information for distributing a computation.
+Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations on streams.
+Primitives 
+\sys enables \emph{distribution-oblivious programming}: 
+  multi-order of magnitude speedups with correctness guarantees and without any developer effort.
+
+The a series of transformations that 
+Second, we are using describing composition . For this, we are restricting the study to a subset of the POSIX shell--
+
+Local operations and identifiers have to be carefully translated to distributed ones, a translation that depends on the types of the operations and associated identifiers.
+
+The semantics is enough to capture powerful features such as stream branching and feedback constructs found in practice and in the system's evaluation.
+
+The paper is structured as follows.
+We first start with a background section outlining pipeline concepts.
+Sections \ref{profiling}--\ref{recipes} highlight our key contributions:
+\begin{itemize}
+
+  \item
+  \S\ref{overview} overviews the system and introduces the notion of command classes.
+	introduces \emph{load-detection transformations} that collect windowed statistics about load at each module boundary.
+  Control-plane coordination with a global view of load and available resources helps decide when to initiate scale-out of a bottlenecked module.
+
+  \item
+  \S\ref{core} presents a set of parametrizable \emph{distribution transformations} that transparently scale modules out.
+  These transformations can create module replicas, hook communication channels among them, schedule requests, and forward side-effects such as mutation and collection of memory.
+
+  \item
+  \S\ref{recipes} outlines \emph{distribution recipes}, lightweight annotations that guide the semantics of the resulting distribution.
+  They offer significant flexibility by\eat{ generating configurations that} parametrizing transformations, including tuning state management, replication consistency, event propagation, and colocation preferences.
+\end{itemize}
+
+\noindent
+\sys's evaluation~\sx{eval} uses a combination of one-liner micro-benchmarks that highlight certain features and three multi-line programs---a web crawling and indexing engine, a production genomics pipeline, and a climate data processing.
+We close with a discussion of related prior work~\sx{related} and possible future directions~\sx{discussion}.
+
+\section{Background}
+
+\begin{acks}
+  % Dumping people so that we don't forget
+  % 
   This material is based upon work supported by the
@@
 \appendix
-\section{Appendix}
+\section{Scripts used in the evaluation}
 
-Text of appendix \ldots
+This appendix contains the source code of the scripts used in the evaluation of
+the \sys. They are part of the codebase (released as open source with the camera
+ready), and are provided here only to aid the reviewers.
 

### 2019-09-21 pldi/draft.tex
Make bib local

@@
   % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{maas2014holistic, schwarzkopf2013dios, andromeda:hotos:2015, jonas2017occupy, varghese2018next, nefele:18}.
+   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:2015, pywren:17, futuredata:18, nefele:18}.
 
 To understand this sharp contrast, consider calculating frequencies of terms within a set of input files.
-For a developer with \emph{one} computer, this is achievable with a short \unix pipeline:
+For a developer with \emph{one} computer, this is achievable with a short \unix
+pipeline~\cite{bentley1986literate}:
 
@@
 %% Bibliography
-\bibliography{../../bib/bib}
+\bibliography{./bib}
 

### 2019-09-21 pldi/draft.tex
Start examples

@@
 % AWS policy specification
-\begin{lstlisting}[language=sh,float=h,numbers=none]
- cat * | tr -cs A-Za-z`\n` | tr A-Z a-z |
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat * | tr -cs A-Za-z`\n` | tr A-Z a-z |      (($$p_1$$))
    sort | uniq -c | sort -rn | head 5 > out
@@
 
+This section breaks down $p_1$ to presents important background on and challenges
+of shell scripting.
+
+\subsection{Unix Pipelines, Informally}
+
+Shell scripting
+
+\heading{Basics}
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ ls -l | wc -l                                                  (($$p_2$$))
+\end{lstlisting}
+
+\heading{Generation}
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat * | wc -l > out.txt                                          (($$p_2$$))
+\end{lstlisting}
+
+\heading{Identifiers}
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat /proc/ >                                                   (($$p_2$$))
+\end{lstlisting}
+
+\heading{Stream Manipulation}
+Explicit Redirection
+Implicit splitting and merging 
+This can be done through the file-system operators
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat /proc/ >                                                   (($$p_2$$))
+\end{lstlisting}
+
+\heading{Grouping}
+Composition
+
+The shell provides several ways of grouping commands and manipulating their
+overall input and output channels
+
+{ echo 1; echo 2; } > out.txt
+
+\subsection{Comnands}
+
+% Sets vs. Classes
+
+Distributability classes, based on how
+
+\heading{Preliminaries}
+
+\heading{POSIX, GNU Core-utils, and beyond}
+
+\heading{Stateless}
+
+\heading{Pure}
+
+\heading{File-system}
+
+\heading{Side-effectful}
+
+\heading{Irreversably Side-effectful}
+
+\subsection{The Shell Language}
+
+\heading{Implicits}
+Expansion
+Implicits
+Context
+
+\heading{Language Constructs}
+
+Even without these, the Unix pipeline language is powerful enough to capture complex pipelines.
+
+We can encode powerful
+
+\section{Dish Overview}
+
+\section{Related Work}
+
 \begin{acks}

### 2019-09-21 pldi/draft.tex
Update intro

@@
 \section{Background}
+\label{bg}
 
@@
 \subsection{Unix Pipelines, Informally}
+\label{bg:pipelines}
 
-Shell scripting
+Pipelines are a mechanism for program composition.
+Individual programs, specified as commands, conform to a common but general interface. 
+Broadly, this interface commands allows commands to do a combination of the following:
+  accept input from their input stream (stdin),
+  generate output to their output stream (stdout)
+Input and output streams are standard text streams---lines of text separated by the newline character.
+Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
+% This concept gave rise to the toolbox philosophy: 
+%   programs that do one thing and do it well, that work together, and that handle text streams as a universal interface.
 
 \heading{Basics}
+
+The standard shell syntax for anonymous pipes is to list multiple commands, separated by vertical bars ("pipes" in common Unix verbiage).
+that are not interpreted by the shell, but are local to each command.
+
+Additional details augment this simple interface.
+Commands can be specified with options arguments
 
@@
 \heading{Generation}
+A central concept in pipelines---and Unix, in general---is the file system, which allows interfacing with persistent data.
+There are many commands that take as input a file identifier 
+Reading a file or a set of files generates a stream; 
+  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
+
+Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
+The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
+
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat * | wc -l > out.txt                                          (($$p_2$$))
+ cat * | wc -l > out.txt                 (($$p_3$$))
 \end{lstlisting}
@@
 \heading{Identifiers}
+The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).
+as such, different identifiers many not resolve to actual files, but rather in-memory or OS-internal structures.
+
+Unix already specifies a few classes of files:
+
+Here are a few special files:
+  \ttt{proc} called
+
+Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
+Similarly, open files and other ephemeral identifiers
+
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                                                   (($$p_2$$))
+ cat /proc/ >                            (($$p_4$$))
 \end{lstlisting}
@@
 \heading{Stream Manipulation}
+A small but powerful set of primitives are concerned with stream manipulation.
+First, as described above, several commands can be redirected
+Second,
+
+\ttt{xargs}
+\ttt{tee}
+
+In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
+
 Explicit Redirection
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                                                   (($$p_2$$))
+ cat /proc/ >                            (($$p_5$$))
 \end{lstlisting}
 
-\heading{Grouping}
+\heading{Composition}
 Composition
@@
 \subsection{Comnands}
+\label{bg:cmd}
 
@@
 \subsection{The Shell Language}
+\label{bg:shell}
 

### 2019-09-22 pldi/draft.tex
Notes on streams

@@
 Individual programs, specified as commands, conform to a common but general interface. 
-Broadly, this interface commands allows commands to do a combination of the following:
-  accept input from their input stream (stdin),
-  generate output to their output stream (stdout)
-Input and output streams are standard text streams---lines of text separated by the newline character.
+For now, this interface allows a combination of the following:
+  accepting input from the standard input stream (stdin),
+  generating output to the standard output stream (stdout), and
+  generating errors or diagnostics to the standard error stream (stderr)
+Standard streams are streams of characters with a few special characters---for example, EOF is a special end-of-file character signaling the end of the input stream.
+% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.
+
 Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-% This concept gave rise to the toolbox philosophy: 
-%   programs that do one thing and do it well, that work together, and that handle text streams as a universal interface.
+Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
+  programs do one thing well, they work together, and they manipulate text streams as their universal interface.
+The next few paragraphs describe the details.
 
 \heading{Basics}
+The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).
 
-The standard shell syntax for anonymous pipes is to list multiple commands, separated by vertical bars ("pipes" in common Unix verbiage).
 that are not interpreted by the shell, but are local to each command.

### 2019-09-22 pldi/draft.tex
Notes on POSIX file types

@@
 
+At a high level, POSIX defines several file types:
+(-) regular files, storing lines of text or binary data,
+(d) directories, grouping together multiple other files,
+(l) symbolic links, pointing to other files,
+(p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
+(s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
+(b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
+
+Most of these identifiers are manipulated by file-system commands---\eg \ttt{mkdir} for directories and \ttt{mkfifo} for FIFOs.
+FIFOs and DSes look like files but are streams.
+
+Special files are particularly interesting because they are resolved to OS structures.
+One set is pseudo-devices---device files that simply;
+For example /dev/urandom returns a stream of random characters and /dev/null accepts and discards all input written to it.
+Another set is proc filesystem (procfs), a special filesystem in Unix-like operating systems that presents information about processes and other system information.
+For example, /proc/PID/cmdline contains command that originally started the process and /proc/PID/cwd contains a symlink to the current working directory of the process.
+
+sysfs is a pseudo file system provided by the Linux kernel that exports information about various kernel subsystems, hardware devices, and
+
+/modules, one of the most important files in /proc, containing a list of the kernel modules currently loaded . It gives some indication (not always entirely correct) of dependencies.
+/cpuinfo
+
+symbolic links, FIFO specials, 
+device files
+block special, character special, 
+and domain sockets.
+
+There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
+These correspond to \ttt{/dev/fd/0}.
+% These streams are created by the C library (glibc) at the start of program execution,
+and new streams can be created to connect to files, sockets, pipes,
+
 Unix already specifies a few classes of files:
@@
 \heading{Composition}
+anonymous pipes
+
 Composition

### 2019-10-02 pldi/draft.tex
Abstract placeholder

@@
 \begin{abstract}
-  Distributed systems offer notable benefits over centralized systems.
-  Reaping these benefits, however, requires developers to develop and deploy
-  software in a manner that makes distribution explicit---a process
-  that is quite different from how they compose software for centralized
-  systems.
+  \nv{placeholder}
+  Distributed systems offer notable benefits over centralized ones.
+  Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed framework interfaces, or distribution annotations.
 % 
-  \emph{Light-touch distribution} is a new approach for scaling out distribution-oblivious shell scripts.
-  It leverages the fact that 
-% introduced as a drop-in replacement of a language runtime's module system, 
-  Our \sys prototype shows substantial speedups over 
-  as well as the ability to operate on large input sizes, all without
-  any developer input.
+%   \emph{Distribution-oblivious programming} is a new approach for..?
+  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. %  with regards to the sequential program.
+  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known distributability classes.
+  \sys's insight is that shell pipelines already express stream computations that can automatically distributed.
+  To achieve this, \sys
+% a series of techniques for automatically 
+decomposes primitives into distributability classes,
+identifies high-distributability stages,
+applies rewriting rules for largest possible subprograms,
+and orchestrates the distributed program.
+Its runtime component provides orchestration and planning support during the execution of the program.
+% 
+% Key challenges include 
+% surprisingly expressive, 
+% maximally distributable subprograms
+% 
+Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
 \end{abstract}

### 2019-10-02 pldi/draft.tex
Update intro

@@
 
-Distributed systems offer significant benefits over their centralized counterparts.
-For example, they can speed up computations and can process amounts that would not fit into any single machine.
+Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or process data that would not fit into any single machine.
 % offer notable benefits over their centralized counterparts:
@@
 % The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
-
 Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
-Whereas anyone can stitch together a Bash script to compute on a single computer, 
+Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
   % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:2015, pywren:17, futuredata:18, nefele:18}.
+   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
 
@@
 \heading{Identifiers}
+
 The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).

### 2019-10-02 pldi/draft.tex
Update intro

@@
 
-To understand this sharp contrast, consider calculating frequencies of terms within a set of input files.
-For a developer with \emph{one} computer, this is achievable with a short \unix
-pipeline~\cite{bentley1986literate}:
+To understand this contrast, consider calculating term frequencies over a set of input files.
+For a user with \emph{one} computer, this is achievable with a short \unix pipeline~\cite{bentley1986literate}:
 
@@
 % \end{lstlisting}
-% greece; stops
-% AWS policy specification
+
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z`\n` | tr A-Z a-z | (($$p_1$$))
+cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
   sort | uniq -c | sort -rn | head 5 > out
@@
 
-Combining several features, Unix makes small tasks easy to express;
-  composition allows the composition of general primitives
-Key elements of \unix are the ability to stitch together programs written in different languages, the abstraction of a file system as a set of resident streams, as well as naming---the ability to resolve command, file names within a global context.
-Under the hood, the \unix kernel buffers results, concurrency control
+Combining several features, Unix makes small tasks easy to express:
+  $p_1$ creates a character stream, breaks it down to words, transliterates them lower-case, sorts them to group duplicates, counts and replaces duplicates, sorts them numerically and in reverse order, picks the top five results and writes them to a file \ttt{out}.
+% composition allows the composition of general primitives
+Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
+Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
 
@@
 
-Unfortunately, 
-A developer with many computers has to choose between three options, all of which require significant manual effort.
-The most popular is to leverage domain-specific ; this requires rewriting portions of the program in a new language and use operations
+%      dish      ignis, mozart    hadoop, spark, ...      erlang, elixir
+%       \/           \   /             \     /                \   /
+% auto <-------------------------------------------------------------> manual
+%                      [rewrite (not write) cost]
+
+For a user with many computers, however, rewriting this pipeline to exploit distribution requires a significant cost in terms of manual effort.
+% The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
+Such cost is at best moderate when pipelines can be expressed in domain-specific frameworks---for example, Hadoop and Spark provide a few purely functional primitives that 
+As long as the program fits into the model this is only moderately difficult;
+For simple, one-off pipelines that take a few minutes to compose this cost---order of magnitude higher than---becomes unjustifiable.
+
+More realistic programs, such as the pipelines presented in this paper, require distributed programming language or a distributed operating system---\eg Plan9's \ttt{rc} shell.
+For many legacy pipelines  
+Could the generation and execution of $p_1$'s distributing version be fully and correctly automated?
+
+ has to choose between three options, all of which require significant manual effort.
+The most invasive is to rewrite the entire program in a distributed programming language
+The most popular is to leverage
+
+Use a distributed programming language, even higher rewrites
+Or use some form of annotations
+
+ this requires rewriting portions of the program in a new language and use operations
+
+In the simplest case, annotations
+
+Often, they only focus on a few parts of the system---for example, upgrading to a distributed storage layer.
+More rarely, companies rewrite entire systems (\eg Twitter's Ruby-to-Scala rewrite~\cite{twitter}), a process that is notoriously difficult under schedule constraints and competitive pressures~\cite{rewrite1, rewrite2}.
+The manual effort is expensive, and can introduce new bugs, cascading changes, or regressions of previously fixed performance issues, especially since software today makes extensive use of third-party modules~\cite{breakapp:plos:2017}.
 
 % If developers have already expressed their computation as a Unix pipeline, they should not have to manually rewrite the program in other environment---\eg Hadoop, Spark---to exploit distribution.
-For simple, one-off pipelines that take a few minutes to compose this cost---order of magnitude higher than---becomes unjustifiable.
-For many legacy pipelines  
-Could the generation and execution of a distributing version of the pipeline be automated?
 
-The key insight behind our work is that Unix pipelines already provide most of the information for distributing a computation.
-Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations on streams.
+The key insight behind our work is that \unix pipelines already provide most of the information for distributing a computation.
+Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
 Primitives 

### 2019-10-03 pldi/draft.tex
Update intro

@@
 
-To understand this contrast, consider calculating term frequencies over a set of input files.
-For a user with \emph{one} computer, this is achievable with a short \unix pipeline~\cite{bentley1986literate}:
+To understand this sharp contrast, consider calculating term frequencies over a set of input files:
 
@@
 
-Combining several features, Unix makes small tasks easy to express:
+Combining several features, \unix makes small tasks easy to express:
   $p_1$ creates a character stream, breaks it down to words, transliterates them lower-case, sorts them to group duplicates, counts and replaces duplicates, sorts them numerically and in reverse order, picks the top five results and writes them to a file \ttt{out}.
+For a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
-Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
-Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
 
@@
 
-For a user with many computers, however, rewriting this pipeline to exploit distribution requires a significant cost in terms of manual effort.
+For a user with many computers and larger inputs, however, developing $p_1$'s distributed equivalent requires a significant effort.
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.

### 2019-10-03 pldi/draft.tex
Update intro

@@
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-Such cost is at best moderate when pipelines can be expressed in domain-specific frameworks---for example, Hadoop and Spark provide a few purely functional primitives that 
-As long as the program fits into the model this is only moderately difficult;
-For simple, one-off pipelines that take a few minutes to compose this cost---order of magnitude higher than---becomes unjustifiable.
+In simple cases comprised of pure functions, has to be rewritten in the language and interfaces exposed by the chosen big-data framework or DSL.
+This cost is unjustifiable for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language or a distributed operating system---\eg Plan9's \ttt{rc} shell.
+In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality---
+Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
 
-More realistic programs, such as the pipelines presented in this paper, require distributed programming language or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-For many legacy pipelines  
-Could the generation and execution of $p_1$'s distributing version be fully and correctly automated?
-
- has to choose between three options, all of which require significant manual effort.
-The most invasive is to rewrite the entire program in a distributed programming language
-The most popular is to leverage
-
-Use a distributed programming language, even higher rewrites
-Or use some form of annotations
-
- this requires rewriting portions of the program in a new language and use operations
-
-In the simplest case, annotations
-
-Often, they only focus on a few parts of the system---for example, upgrading to a distributed storage layer.
-More rarely, companies rewrite entire systems (\eg Twitter's Ruby-to-Scala rewrite~\cite{twitter}), a process that is notoriously difficult under schedule constraints and competitive pressures~\cite{rewrite1, rewrite2}.
-The manual effort is expensive, and can introduce new bugs, cascading changes, or regressions of previously fixed performance issues, especially since software today makes extensive use of third-party modules~\cite{breakapp:plos:2017}.
+% Some (parts of) pipelines require only moderate effort, as they are 
+% For some pipelines, this effort is merely moderate, as 
+%  can be expressed in domain-specific frameworks---for example, Hadoop and Spark provide a few purely functional primitives that 
+%   ---still requiring rewriting the program in a different (\eg Scala for Spark).
+% 
+% This effort is moderate if the program expressed by these pipelines fits into a domain-specific framework.
+% At times, such pipelines
+% As long as the program fits into the model this is only moderately difficult;
+% 
+% For many legacy pipelines  
+%  has to choose between three options, all of which require significant manual effort.
+% The most invasive is to rewrite the entire program in a distributed programming language
+% The most popular is to leverage 
+% 
+% Use a distributed programming language, even higher rewrites
+% Or use some form of annotations 
+% 
+%  this requires rewriting portions of the program in a new language and use operations
+% 
+% 
+% In the simplest case, annotations
+% 
+% Often, they only focus on a few parts of the system---for example, upgrading to a distributed storage layer.
+% More rarely, companies rewrite entire systems (\eg Twitter's Ruby-to-Scala rewrite~\cite{twitter}), a process that is notoriously difficult under schedule constraints and competitive pressures~\cite{rewrite1, rewrite2}.
+%  especially since software today makes extensive use of third-party modules~\cite{breakapp:plos:2017}.
+%
 

### 2019-10-03 pldi/draft.tex
Update intro

@@
 The key insight behind our work is that \unix pipelines already provide most of the information for distributing a computation.
+Commands
+
 Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
+Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
+Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
+
 Primitives 

### 2019-10-09 pldi/draft.tex
Update intro

--- 2019-10-03 pldi/draft.tex
@@
 
-Combining several features, \unix makes small tasks easy to express:
-  $p_1$ creates a character stream, breaks it down to words, transliterates them lower-case, sorts them to group duplicates, counts and replaces duplicates, sorts them numerically and in reverse order, picks the top five results and writes them to a file \ttt{out}.
-For a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+\noindent
+Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, summarizes them with a counter, sorts them in descending order, picks the top five results and writes them to a file \ttt{out}.
+Combining several features, \unix makes small tasks easy to express;
+  for a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
@@
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-In simple cases comprised of pure functions, has to be rewritten in the language and interfaces exposed by the chosen big-data framework or DSL.
-This cost is unjustifiable for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality---
+For simple pipelines that capture in restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
+%  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
+In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality.
 Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
@@
 
-The key insight behind our work is that \unix pipelines already provide most of the information for distributing a computation.
-Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
-Primitives 
-\sys enables \emph{distribution-oblivious programming}: 
-  multi-order of magnitude speedups with correctness guarantees and without any developer effort.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required for distributing a computation.
+\sys builds on this insight with a careful study of the scalability properties of shell primitives and its commands, paired with a novel rewriting scheme that identifies maximal pipeline sub-expressions that are candidates for scale-out.
+We term the result \emph{distribution-oblivious programming}---a series of techniques that fully automates program distribution while provably maintaining correct (non-distributed) semantics.
 
-The a series of transformations that 
-Second, we are using describing composition . For this, we are restricting the study to a subset of the POSIX shell--
+Key benefits include the ability to scale legacy computations onto
+\sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
+Most importantly, \sys provides an architectural lesson:
+  the current design shows that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 
-Local operations and identifiers have to be carefully translated to distributed ones, a translation that depends on the types of the operations and associated identifiers.
+% % Make Unix benefits explicit?
+% Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
+% Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
+% Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
 
-The semantics is enough to capture powerful features such as stream branching and feedback constructs found in practice and in the system's evaluation.
+% Primitives 
+% \sys enables \emph{distribution-oblivious programming}: 
+%   multi-order of magnitude speedups with correctness guarantees and without any developer effort.
+% 
+% The a series of transformations that 
+% Second, we are using describing composition . For this, we are restricting the study to a subset of the POSIX shell--
+% 
+% Local operations and identifiers have to be carefully translated to distributed ones, a translation that depends on the types of the operations and associated identifiers.
+% 
+% The semantics is enough to capture powerful features such as stream branching and feedback constructs found in practice and in the system's evaluation.
 
 The paper is structured as follows.
-We first start with a background section outlining pipeline concepts.
-Sections \ref{profiling}--\ref{recipes} highlight our key contributions:
+We first start with a background section outlining pipeline concepts~\sx{bg}.
+Sections \ref{classes}--\ref{recipes} highlight our key contributions:\footnote{
+  This is not precise, we'll need to wait until it matures a bit.
+}
 \begin{itemize}
@@
   \item
-  \S\ref{overview} overviews the system and introduces the notion of command classes.
-	introduces \emph{load-detection transformations} that collect windowed statistics about load at each module boundary.
-  Control-plane coordination with a global view of load and available resources helps decide when to initiate scale-out of a bottlenecked module.
+  \S\ref{classes} overviews \sys and introduces several scalability and distributability classes.
 
   \item
-  \S\ref{core} presents a set of parametrizable \emph{distribution transformations} that transparently scale modules out.
-  These transformations can create module replicas, hook communication channels among them, schedule requests, and forward side-effects such as mutation and collection of memory.
+  \S\ref{rewrite} presents a set of rewriting transformations for identifying maximal pipeline sub-expressions, candidates for scale-out.
 
   \item
-  \S\ref{recipes} outlines \emph{distribution recipes}, lightweight annotations that guide the semantics of the resulting distribution.
-  They offer significant flexibility by\eat{ generating configurations that} parametrizing transformations, including tuning state management, replication consistency, event propagation, and colocation preferences.
+  \S\ref{other} details other concerns, such as distributed state management and extensibility.
 \end{itemize}
@@
 \noindent
-\sys's evaluation~\sx{eval} uses a combination of one-liner micro-benchmarks that highlight certain features and three multi-line programs---a web crawling and indexing engine, a production genomics pipeline, and a climate data processing.
+\sys's evaluation~\sx{eval} uses a combination of one-liner micro-benchmarks that highlight certain features and three multi-line macro-benchmarks---a web crawling and indexing engine, a production genomics pipeline, and a large-scale weather analysis.
 We close with a discussion of related prior work~\sx{related} and possible future directions~\sx{discussion}.

### 2019-11-18 pldi/draft.tex
Add spectrum figure
Add overview figure
Update intro

@@
 
-Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or process data that would not fit into any single machine.
+Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
 % offer notable benefits over their centralized counterparts:
@@
 
-To understand this sharp contrast, consider calculating term frequencies over a set of input files:
+To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
 
@@
 \noindent
-Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, summarizes them with a counter, sorts them in descending order, picks the top five results and writes them to a file \ttt{out}.
+Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
 Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
+\caption{
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
+}
+\label{fig:spectrum}
+\end{figure}
+
 %      dish      ignis, mozart    hadoop, spark, ...      erlang, elixir
@@
 
-For a user with many computers and larger inputs, however, developing $p_1$'s distributed equivalent requires a significant effort.
+Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
+% For a user with many computers and larger inputs, 
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines that capture in restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
+For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
 %  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality.
+In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
 Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
@@
 
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required for distributing a computation.
-\sys builds on this insight with a careful study of the scalability properties of shell primitives and its commands, paired with a novel rewriting scheme that identifies maximal pipeline sub-expressions that are candidates for scale-out.
-We term the result \emph{distribution-oblivious programming}---a series of techniques that fully automates program distribution while provably maintaining correct (non-distributed) semantics.
+To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
+\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
 
-Key benefits include the ability to scale legacy computations onto
-\sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-Most importantly, \sys provides an architectural lesson:
-  the current design shows that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
+% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
+First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}, SLURM, calls to \ttt{parallel}, or any manual rewriting.
+Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt t}  \#threads, \ttt{-t}, \ttt{-p} \etc
+Most importantly, \sys provides an architectural lesson for system designers---namely that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 
@@
 
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+\caption{
+  \textbf{Applying \sys to $p_1$.}
+}
+\label{fig:example}
+\end{figure*}
+
 \section{Background}
@@
 Commands can be specified with options arguments
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
+\label{fig:example}
+\end{figure}
 

### 2019-11-18 pldi/draft.tex
Teensy update

@@
 % \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}, SLURM, calls to \ttt{parallel}, or any manual rewriting.
-Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt t}  \#threads, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  \#threads, \ttt{-t}, \ttt{-p} \etc
+Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 

### 2019-11-18 pldi/draft.tex
Teensy updates in intro

@@
 First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  \#threads, \ttt{-t}, \ttt{-p} \etc
+Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
 Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+% Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
 
@@
 The paper is structured as follows.
-We first start with a background section outlining pipeline concepts~\sx{bg}.
-Sections \ref{classes}--\ref{recipes} highlight our key contributions:\footnote{
+It first starts with a background section outlining pipeline concepts and overviewing \sys~\sx{bg}.
+Sections \ref{classes}--\ref{recipes} highlight key contributions:\footnote{
   This is not precise, we'll need to wait until it matures a bit.
@@
 \noindent
-\sys's evaluation~\sx{eval} uses a combination of one-liner micro-benchmarks that highlight certain features and three multi-line macro-benchmarks---a web crawling and indexing engine, a production genomics pipeline, and a large-scale weather analysis.
-We close with a discussion of related prior work~\sx{related} and possible future directions~\sx{discussion}.
+\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
+After a comparison with related prior work~\sx{related}, the paper closes with a discussion of related work and possible future directions~\sx{discussion}.
 

### 2019-11-19 pldi/draft.tex
Draft content for intermediate sections, to estimate budget and feasibility

--- 2019-11-18 pldi/draft.tex
@@
 \usepackage{amsmath}
+\usepackage{cleveref}
 
@@
 \newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
+\newcommand{\kk}[1]{[{\color{magenta}#1 --- KK}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 
-Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
+Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or process data that would not fit into any single machine.
 % offer notable benefits over their centralized counterparts:
@@
 
-To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
+To understand this sharp contrast, consider calculating term frequencies over a set of input files:
 
@@
 \noindent
-Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
+Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, summarizes them with a counter, sorts them in descending order, picks the top five results and writes them to a file \ttt{out}.
 Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+  for a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
@@
 
-Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
-% For a user with many computers and larger inputs, 
+For a user with many computers and larger inputs, however, developing $p_1$'s distributed equivalent requires a significant effort.
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
+For simple pipelines that capture in restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
 %  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
+In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality.
 Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
@@
 
-To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required for distributing a computation.
+\sys builds on this insight with a careful study of the scalability properties of shell primitives and its commands, paired with a novel rewriting scheme that identifies maximal pipeline sub-expressions that are candidates for scale-out.
+We term the result \emph{distribution-oblivious programming}---a series of techniques that fully automates program distribution while provably maintaining correct (non-distributed) semantics.
 
-The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
-% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}, SLURM, calls to \ttt{parallel}, or any manual rewriting.
-Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt t}  \#threads, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+Key benefits include the ability to scale legacy computations onto
+\sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
+Most importantly, \sys provides an architectural lesson:
+  the current design shows that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 
@@
 \heading{Identifiers}
+
+\kk{We have to mention character vs line based parallelism somewhere,
+  and that we choose to see lines as the data quantum}
 
@@
 
+\section{Extensibility}
+
+ULTRA DRAFTY:
+
+\kk{Text point: A problem that is specific to the shell is
+  extensibility.}
+
+The major characteristic that makes the shell popular and very useful
+is its extensibility, as also mentioned in \Cref{bg}. Commands change,
+get extended, and each user continuously installs more commands to use
+in their system.
+
+\kk{Text point: This restricts the space of solutions for distribution
+  of shell scripts}
+
+This means that any one-time solution is not satisfactory, since it
+would very easily become obsolete, or it would only handle a limited
+amount of shell scripts.
+
+\kk{Text point: The characteristics of a good solution}
+
+A solution that applies to shell programms should thus be extensible,
+in the sense that adding or changing a command, doesn't invalidate the
+assumptions that the tool makes. In our case, the information about
+the command categories. A developer of a new command, should be able
+to categorize it accordingly, and succinctly describe how the system
+can distribute a command.
+
+\kk{Solution: A lightweight language that can be used by a command
+  developer to categorize it for distribution.}
+
+We address this issue by designing a lightweight language whose
+purpose is to categorize a command, describes how it reads its input
+(meaning in what order), and how to distribute it (in case it is
+pure). \kk{I am not sure if the order in which a command reads its
+  input will be part of the language since many commands can be made
+  to read from one place (like stdin or their first argument). Even in
+  that case though, where the command only reads from one argument, we
+  have to make sure that there is a way for the user to pinpoint which
+  argument does a command read from.}
+
+\subsection{Categorization Language}
+
+\kk{Text point: Introduce problem: Most shell commands, belong to
+  different classes depending on their flags, arguments}
+
+Consider the command \texttt{cat} \kk{Maybe that is not he best
+  example. TODO: Find a good example of a simple command that is in
+  different classes depending on the arguments.}. When used without
+any flag it is stateless, etc...
+
+\kk{Text point: Our language must be able to categorize commands to
+  different classes according to their arguments.}
+
+\kk{We should have a crisp point about why we have this categorization
+  language, and why don't we just allow someone to write a function in
+  python for each command, that given the command and its arguments,
+  returns whether the command category. Possible arguments include,
+  the fact that this language is simpler to use, especially by non
+  experts that just run some script but have installed some commands
+  that are not ``supported''. Another possible argument is that we
+  could be able to reason about the constructs of the categorization
+  lagnauge, and that they will be easier to read and
+  understand. Another argument is that since these categorizations
+  should be shareable among users, it would be bad to execute
+  arbitrary python code, so using this language, expressivity is
+  limited. All of these arguments are a little bit weak though. Niko,
+  what do you think?}
+
+Sketch of the language:
+
+Some kind of lightweight annotation configuration language (like
+yaml), that contains a record for each command.
+
+A record contains a sequence of predicates => categories. Each
+predicate is a combination of not, or, and and the atomic predicates
+are the presence of a flag, or an equality that checks that the value
+of a flag is equal to something (because flags can also have
+values). The final predicate can be left empty, and it is trivially true.
+
+Interpreter:
+
+Given a command and the set of its arguments, the intepreter returns
+the command category. In does so by returning the category of the first
+predicate that matches.
+
+\kk{Would it be meaningful to give the syntaxx of the language
+  formally?}
+
+\kk{If we talk about the language it would be good to give some
+  statistics about how many commands (out of the ones that we have
+  classified) can be represented with one, two clauses etc. If we have
+  a different solution (like python function that given a command and
+  its arguments returns the category) then we could talk about how
+  many lines of code it took to categorize all the commands that we
+  did.}
+
+As described in distributability\ref{}, the abstract notion of a
+program that we have is programs with one input and one output
+stream. The command developer must be able to express the input
+argument of a commmand (or if it is the stdin).
+
+\kk{Question: Do we need to handle all commands (even the ones that
+  could have their input split in several different arguments? or can
+  we somehow easily make these ones just take their input from one
+  place (e.g. by adding a cat before them).}
+
+\kk{Possible solutions for this (if we have already turned commands
+  with more that one input to commands with one input): Have a
+  positional reference to the argument that is input (or ``stdin'' if
+  it is the stdin). Example: the last argument [-1] of grep is its
+  input.}
+
+\kk{It seems that even the arguments might depend on flags (e.g. grep
+  checks the current directory if asked recursively, or the stdin
+  otherwise), so maybe we should again have a sequence of predicates}
+
+\kk{Note: Since a categorization language will probably not be
+  complete (especially the one showing the input arguments) there
+  should be a backup mechanism, when the language is not expressive
+  enough, like a python function that categorizes the command, and
+  returns its input argument if it is stateless.}
+
+\subsection{Pure Commands}
+
+Even though stateless commands are a big percentage of all of them,
+there are a lot of useful and popular pure commands.
+
+As mentioned in distributability, pure commands can be broken down
+into different categories. We might be able to get ideas about this
+from this paper:
+\url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
+continuation in PLDI 2019.
+
+Can we find a solution for the commands in coreutils?
+
+\section{Front End}
+
+We parse a shell script using a posix compliant parser that returns
+the ast.
+
+\kk{Text point: The AST Is not a very good data structure to
+  manipulate for distribution, so we translate it to a kind of a
+  dataflow graph.}
+
+\subsection{Distributable Subtrees}
+
+\kk{Text point: Problem: Not all parts of a shell script are
+  distributable, so we cannot arbitrarily distribute}
+
+Even though commands are split in categories, and several pipelines
+are distributable, not all parts of a shell script are stateless and
+several of them have effects that could affect the behaviour of a
+program after distribution.
+
+\kk{Solution: We have found that distributing subtrees of the shell
+  AST is safe. Can we have a semi-formal argument about that?}
+
+Mention all the different constructs that our front end handles, and
+why it is safe to add them to the distribution pipeline.
+
+\subsection{Intermediate Representation/Dataflow graph}
+
+Mention its constructs, and its semantics, and why it is a bit
+different that standard dataflow graphs (where usually incoming edges
+are either consumed with some arbitrary interleaving, or pair by
+pair).
+
+\kk{Text point: The main shell abstraction is files, which are batches
+  or streams of characters (or lines) where the order matters. So we
+  want a representation that captures order of inputs and outputs (and
+  ours does because its node in the graph has a sequence of inputs.}
+
+Explain that the intermediate representation represents workers that
+communicate through files, and that it captures parallelism both in
+the operator level, as well as data parallelism.
+
+\section{Distribution Planner}
+
+\kk{Maybe this should be merged with the one above, especially if it
+  talks about how distribution happens on the intermediate
+  representation.}
+
+\subsection{Scaling up the distributed computation/Vertically expanding the graph}
+
+The planner acts on the dataflow graph, by first expanding it ``as
+much as possible'' to expose parallelization opportunities. It uses
+knowledge about the inputs of each stateless and pure command.
+
+\kk{Explain the algorithm and how it pushes parallelization through
+  the graph, while preserving correctness.}
+
+\kk{Give an example}
+
+\subsection{Mapping operators to nodes}
+
+Explain the simple algorithm that we used to minimize intra node
+communication, that tries to map contiguous parts of the graph to the
+same node. We could do this by having an algorithm that minimizes the
+number of cuts or something.
+
+We don't try to make a proper planner, as there is a lot of work on
+operator placement etc. that we can borrow from.
+
+\subsection{Just in time planning}
+
+Important point: Shell is extremely dynamic. Because of this, a
+distributor cannot decide how to distribute a subexpression
+statically, by just reading the script, (as in many other systems like
+MapReduce, Spark, etc), since most of the information is not there
+before the script starts executing. Environment variables,
+unexpanded(unevaluated) strings \kk{Make sure that terminology is
+  consistent with Greenberg}, could all contain information that is
+valuable to make the distribution plan \kk{give an example}.
+
+Future work on this: calling the shell shtepper to only partially
+evaluate strings, just expand arguments, and not do any significant
+computation.
+
+\section{Limitations}
+
+\kk{This is not really a section on its own, but rather part of something else.}
+
+\begin{itemize}
+\item Fault Tolerance
+\item Distributed Planning: The planner at the moment is naive, and
+  there is a huge literature on operator placement of distributed
+  dataflow graphs that can be used to improve it.
+\item Support of shell constructs: At the moment we have been
+  conservative in the shell subsets that we handle, so that we don't
+  introduce any unsafe parallelism. The point of this work was not to
+  be able to handle a complete distributable subset of the shell, but
+  rather a significantly big part of it, to demonstrate that there can
+  be performance benefit from it.
+\item Support distributing different classes: At the moment we are
+  only dealing with stateless and pure commands, that is the ones that
+  are most popular but also can offer the biggest benefits from
+  parallelizing them. However, it would be interesting to explore the
+  distributability of the file system commands, by using a distributed
+  file system. THis would also require considerations about
+  replication, consistency, and failures.
+\end{itemize}
+
 \section{Related Work}

### 2019-11-20 pldi/draft.tex
Update intro
Update background

@@
 \section{Introduction}
+\label{intro}
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
+\includegraphics[width=0.49\textwidth]{./figs/dish_schematic.pdf}
 \caption{
-  \textbf{Cost of Manual Effort.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-A more complete picture is presented in the related-work section~\sx{rt}.
+  \textbf{\sys high-level schematic.}
+	\sys XYZ
 }
-\label{fig:spectrum}
+\label{fig:schematic}
 \end{figure}
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
-\caption{
-  \textbf{Applying \sys to $p_1$.}
-}
-\label{fig:example}
-\end{figure*}
+% \begin{figure*}[t]
+% \centering
+% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+% \caption{
+%   \textbf{Applying \sys to $p_1$.}
+% }
+% \label{fig:example}
+% \end{figure*}
+%
 
@@
 
-Pipelines are a mechanism for program composition.
-Individual programs, specified as commands, conform to a common but general interface. 
-For now, this interface allows a combination of the following:
+A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
+Behind the scenes, \unix facilitates program scheduling and communication synchronization.
+Intermediate programs run concurrently consuming input and producing output;
+  they can be developed in any programming language, as long as they conform to a common interface.
+This interface allows any combination of the following:
   accepting input from the standard input stream (stdin),
@@
   generating errors or diagnostics to the standard error stream (stderr)
-Standard streams are streams of characters with a few special characters---for example, EOF is a special end-of-file character signaling the end of the input stream.
+These standard streams are lines of characters, with a few special characters---for example, EOF is a special end-of-file character signaling the end of a stream.
+Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
+% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}
+
+Along with their implicit context---these standard streams and environment variables---commands can express
+Command flags (options)
+
+% TODO: Talk about security?
+Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
+  files are simply persistent streams that can be manipulated by commands or pipelines.
+Reading a file or a set of files generates a stream; 
+  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
+Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
+\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
+As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
+
+% At a high level, POSIX defines several file types:
+% (-) regular files, storing lines of text or binary data,
+% (d) directories, grouping together multiple other files,
+% (l) symbolic links, pointing to other files,
+% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
+% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
+% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
+
+Special files are particularly interesting because they are resolved to OS structures.
+One example is the proc filesystem (procfs), a special file-system that presents information about processes and other system information.
+This file-system enables a form of programmatic introspection, in which commands can inspect their directory or the flags passed to them.
+% Here are a few examples:
+% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
+% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
+% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
+Similarly, \unix exposes file-system handles for various process-related information, include identifiers for ephmeral streams---\ie ones 
+OS-internal structures.
+
+There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
+These correspond to \ttt{/dev/fd/0}.
+% These streams are created by the C library (glibc) at the start of program execution,
+and new streams can be created to connect to files, sockets, pipes,
+
+This isomorphism is bidirectional---
+
+---and a key enabler of the runtime rebinding achieved during our transformations.
+between file-system handles and ephmeral streams
+
+Unix already specifies a few classes of files:
+
+Here are a few special files:
+  \ttt{proc} called
+
+Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
+Similarly, open files and other ephemeral identifiers
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat /proc/ >                            (($$p_4$$))
+\end{lstlisting}
+
 % For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.
 
-Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
-  programs do one thing well, they work together, and they manipulate text streams as their universal interface.
-The next few paragraphs describe the details.
+% Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
+%   programs do one thing well, they work together, and they manipulate text streams as their universal interface.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+\caption{
+  \textbf{\sys transformation overview.}
+	\sys XYZ
+}
+\label{fig:overview}
+\end{figure}
 
@@
 \heading{Generation}
-A central concept in pipelines---and Unix, in general---is the file system, which allows interfacing with persistent data.
-There are many commands that take as input a file identifier 
-Reading a file or a set of files generates a stream; 
-  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-
 Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
@@
 \heading{Identifiers}
-
-\kk{We have to mention character vs line based parallelism somewhere,
-  and that we choose to see lines as the data quantum}
-
-The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).
-as such, different identifiers many not resolve to actual files, but rather in-memory or OS-internal structures.
-
-At a high level, POSIX defines several file types:
-(-) regular files, storing lines of text or binary data,
-(d) directories, grouping together multiple other files,
-(l) symbolic links, pointing to other files,
-(p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
-(s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
-(b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-
-Most of these identifiers are manipulated by file-system commands---\eg \ttt{mkdir} for directories and \ttt{mkfifo} for FIFOs.
-FIFOs and DSes look like files but are streams.
-
-Special files are particularly interesting because they are resolved to OS structures.
-One set is pseudo-devices---device files that simply;
-For example /dev/urandom returns a stream of random characters and /dev/null accepts and discards all input written to it.
-Another set is proc filesystem (procfs), a special filesystem in Unix-like operating systems that presents information about processes and other system information.
-For example, /proc/PID/cmdline contains command that originally started the process and /proc/PID/cwd contains a symlink to the current working directory of the process.
-
-sysfs is a pseudo file system provided by the Linux kernel that exports information about various kernel subsystems, hardware devices, and
-
-/modules, one of the most important files in /proc, containing a list of the kernel modules currently loaded . It gives some indication (not always entirely correct) of dependencies.
-/cpuinfo
-
-symbolic links, FIFO specials, 
-device files
-block special, character special, 
-and domain sockets.
-
-There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
-These correspond to \ttt{/dev/fd/0}.
-% These streams are created by the C library (glibc) at the start of program execution,
-and new streams can be created to connect to files, sockets, pipes,
-
-Unix already specifies a few classes of files:
-
-Here are a few special files:
-  \ttt{proc} called
-
-Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
-Similarly, open files and other ephemeral identifiers
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                            (($$p_4$$))
-\end{lstlisting}
 
@@
 \section{Related Work}
+\label{related}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
+\caption{
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
+}
+\label{fig:spectrum}
+\end{figure}
+
+\section{Conclusion}
+\label{discussion}
 

### 2019-11-20 pldi/draft.tex
Add a comment about a formal argument in the scaleup section
Minor changes to text

@@
 
-Along with their implicit context---these standard streams and environment variables---commands can express
-Command flags (options)
+The operation of commands can be tuned by two mechanisms, namely command-line options and environment variables.
+Options control the execution of a program---\eg changing the \ttt{sort}ing order.
+The shell does not have any visibility into these options; 
+  their parsing and evaluation is left entirely up to each command.
+Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
+They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
 
-% TODO: Talk about security?
+A small but powerful set of features is concerned with stream manipulation.
+Some of these features are built into the shell;
+for example, given a file identifier, the stream redirection operator can direct stdout or stderr to the file ---of course, this can be a FIFO that is concurrently used as input by a different command.
+More interestingly, some commands allow splitting a single stream into multiple streams while others allow merging streams together (either one after the other, or fusing (zipping) them column-wise into a single stream.
+
+First, as described above, several commands can be redirected
+Second,
+
+\ttt{xargs}
+\ttt{tee}
+
+In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
+
+Explicit Redirection
+Implicit splitting and merging 
+This can be done through the file-system operators
+
+% TODO: Talk about security primitives?
 Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
@@
 Special files are particularly interesting because they are resolved to OS structures.
-One example is the proc filesystem (procfs), a special file-system that presents information about processes and other system information.
-This file-system enables a form of programmatic introspection, in which commands can inspect their directory or the flags passed to them.
+One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
+  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
 % Here are a few examples:
@@
 % (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
-Similarly, \unix exposes file-system handles for various process-related information, include identifiers for ephmeral streams---\ie ones 
-OS-internal structures.
+Similarly, \unix exposes file-system handles for various ephmeral structures, including standard streams.
 
-There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
-These correspond to \ttt{/dev/fd/0}.
+The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines---for example:
+
 % These streams are created by the C library (glibc) at the start of program execution,
-and new streams can be created to connect to files, sockets, pipes,
+% and new streams can be created to connect to files, sockets, pipes,
 
-This isomorphism is bidirectional---
+This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
 
----and a key enabler of the runtime rebinding achieved during our transformations.
-between file-system handles and ephmeral streams
+Apart from pipes, the \unix shell features several  other forms of program composition.
+Command substitution, enclosed in \ttt{\$()}, replaces the contents of the 
+Code block redirects { - }, 
+grouping command 
+single redirection to redirect the standard output of both commands in the
+this is analogous to an anonymous function, as it can be assigned to a variable and be manipulated as a subscript.
+A subshell enclosed in \ttt{()} is similar , but executes the block in a new sub-shell---avoiding polluting environment. % but incurring higher overhead.
 
-Unix already specifies a few classes of files:
-
-Here are a few special files:
-  \ttt{proc} called
-
-Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
-Similarly, open files and other ephemeral identifiers
+% \begin{lstlisting}[
+%   language=es,
+%   mathescape,
+%   float=t,
+%   belowskip=-5mm,
+%   label=core-pl,
+%   upquote=true,
+%   caption={
+%     \textbf{The core language of the POSIX shell.}
+%     The language used by ; file identifiers and expansion are not shown.
+%   }]
+% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
+% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
+% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
+% \end{lstlisting}
 
@@
 \heading{Identifiers}
-
-\heading{Stream Manipulation}
-A small but powerful set of primitives are concerned with stream manipulation.
-First, as described above, several commands can be redirected
-Second,
-
-\ttt{xargs}
-\ttt{tee}
-
-In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
-
-Explicit Redirection
-Implicit splitting and merging 
-This can be done through the file-system operators
 

### 2019-11-20 pldi/draft.tex
Massive rewrite of background section

@@
 \usepackage{cleveref}
+\usepackage{wrapfig}
 
@@
 
-\subsection{Unix Pipelines, Informally}
+\subsection{Unix Composition, Informally}
 \label{bg:pipelines}
 
-A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
-Behind the scenes, \unix facilitates program scheduling and communication synchronization.
-Intermediate programs run concurrently consuming input and producing output;
-  they can be developed in any programming language, as long as they conform to a common interface.
-This interface allows any combination of the following:
-  accepting input from the standard input stream (stdin),
-  generating output to the standard output stream (stdout), and
-  generating errors or diagnostics to the standard error stream (stderr)
-These standard streams are lines of characters, with a few special characters---for example, EOF is a special end-of-file character signaling the end of a stream.
-Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
+% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
+\unix provides several ways for succinct program composition.
+Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
+These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
+Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
+\begin{wrapfigure}{r}{0.06\textwidth}
+  \vspace{-12pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
+  % \end{center}
+  \vspace{-30pt}
+\end{wrapfigure}
+This interface exposes a stream of contiguous character lines.
+A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
+As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
+% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
+% generating errors or diagnostics to the standard 
+% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
 % TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}
 
-The operation of commands can be tuned by two mechanisms, namely command-line options and environment variables.
+Commands can be tuned by two mechanisms, namely command-line options and environment variables.
 Options control the execution of a program---\eg changing the \ttt{sort}ing order.
@@
 
-A small but powerful set of features is concerned with stream manipulation.
-Some of these features are built into the shell;
-for example, given a file identifier, the stream redirection operator can direct stdout or stderr to the file ---of course, this can be a FIFO that is concurrently used as input by a different command.
-More interestingly, some commands allow splitting a single stream into multiple streams while others allow merging streams together (either one after the other, or fusing (zipping) them column-wise into a single stream.
-
-First, as described above, several commands can be redirected
-Second,
-
-\ttt{xargs}
-\ttt{tee}
-
-In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
-
-Explicit Redirection
-Implicit splitting and merging 
-This can be done through the file-system operators
-
 % TODO: Talk about security primitives?
@@
   files are simply persistent streams that can be manipulated by commands or pipelines.
+\begin{wrapfigure}{r}{0.15\textwidth}
+  \vspace{-15pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
+  % \end{center}
+  \vspace{-25pt}
+\end{wrapfigure}
 Reading a file or a set of files generates a stream; 
@@
 % (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
-Similarly, \unix exposes file-system handles for various ephmeral structures, including standard streams.
+Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
+This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
 
-The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines---for example:
+A small but powerful set of features is concerned with stream manipulation.
+Some of these features are built into the shell;
+  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
+\begin{wrapfigure}{r}{0.15\textwidth}
+  \vspace{-15pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
+  % \end{center}
+  \vspace{-25pt}
+\end{wrapfigure}
+As described before, the file can be a FIFO dubbing as input to another command.
+More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
+  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
+% TODO: talk about \ttt{xargs}?
 
+% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators
+
+The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
+The most interesting feature of these pipelines is the presence of cycles:
+  a pipeline segment processes some input, part of which is extracting the next input to process.
+One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
 % These streams are created by the C library (glibc) at the start of program execution,
@@
 
-This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
-
-Apart from pipes, the \unix shell features several  other forms of program composition.
-Command substitution, enclosed in \ttt{\$()}, replaces the contents of the 
-Code block redirects { - }, 
-grouping command 
-single redirection to redirect the standard output of both commands in the
-this is analogous to an anonymous function, as it can be assigned to a variable and be manipulated as a subscript.
-A subshell enclosed in \ttt{()} is similar , but executes the block in a new sub-shell---avoiding polluting environment. % but incurring higher overhead.
+Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
+Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
+Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
+  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
+Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
 
@@
 
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                            (($$p_4$$))
-\end{lstlisting}
-
-% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.
-
-% Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
-%   programs do one thing well, they work together, and they manipulate text streams as their universal interface.
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat /proc/ >                            (($$p_4$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  ls -l | wc -l                           (($$p_2$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat * | wc -l > out.txt                 (($$p_3$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat /proc/ >                            (($$p_5$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+% { echo 1; echo 2; } > out.txt
+% \end{lstlisting}
 
@@
 
-\heading{Basics}
-The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).
+% \heading{Generation}
+% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
+% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
 
-that are not interpreted by the shell, but are local to each command.
-
-Additional details augment this simple interface.
-Commands can be specified with options arguments
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
-\label{fig:example}
-\end{figure}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- ls -l | wc -l                           (($$p_2$$))
-\end{lstlisting}
-
-\heading{Generation}
-Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
-The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat * | wc -l > out.txt                 (($$p_3$$))
-\end{lstlisting}
-
-\heading{Identifiers}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                            (($$p_5$$))
-\end{lstlisting}
-
-\heading{Composition}
-anonymous pipes
-
-Composition
-
-The shell provides several ways of grouping commands and manipulating their
-overall input and output channels
-
-{ echo 1; echo 2; } > out.txt
+\subsection{Overview}
 

### 2019-11-20 pldi/draft.tex
Update overview

@@
 
+As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
+Intuitively
+ (i) the parallel-composition operator exposes task parallelism,
+ (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+ (iii) the pipe operator enables both task and data parallelism, and
+ (iv)  other composition operators can be viewed as barriers.
+To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.
+
+Understanding the primitives of the shell is not enough;
+  one has to understand the characteristics of commands used to compose larger programs.
+This challenge has two parts
+One, characterize the commands already in the shell---that is, built-ins .
+This study identifies a small set of well-understood distributability classes and is already useful enough to allow the composition of practical pipelines.
+The second part is to use these classes to 
+Using this language -> extensions
+A command developer can only annotate the default behavior;
+  if a new flag is added a few years later, and only if that flag alters the command's class, only then a new annotation is needed.
+
+Given a set of , \sys's has to analyze the given pipeline and identify candidate subexpressions for distribution.
+\sys addresses this challenge with a dataflow analysis that identifies longest subexpressions.
+This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc
+
+Given a distributed pipeline with runtime instrumentation---two challenges.
+First, wrapping 
+  just-in-time wrappers for all the, generated by the distributability specification outlined earlier.
+The reasoning behind thin wrappers is that they should not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
+Second, fill-in planing holes left by the rewriting engine.
+This is achieved with \emph{planner activators}, commands yield (passing their arguments) to the scheduler.
+From the point of view of the shell, activators are higher-order commands a la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding.
+
+The next few sections , and discuss other challenges such as the synchronization of environment variables, distributed file operation, \etc.
+
+\section{Distributability Classes}
+
 \subsection{Comnands}
@@
 \section{Extensibility}
+
+Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
+% ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
+%    | -h :: ( Sless ) => { Out }                              -- if needed
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+(bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
+  |-h :: (Sless)($$\Rightarrow$$){Out}                              
+\end{lstlisting}
+
+It states that the command defaults to the Pure class.
+It's operation can be thought as a transformation from input streams to output streams.
+It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
+% TODO: check actual man page for how to specify files
+  it writes to the output and error stream rather than a file.
+The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
+  its output is a constant function.
 

### 2019-11-20 pldi/draft.tex
Update overview

--- 2019-11-20 pldi/draft.tex
@@
 
+As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
+Intuitively,
+ (i) the parallel-composition operator exposes task parallelism,
+ (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+ (iii) the pipe operator enables both task and data parallelism, and
+ (iv)  other composition operators can be viewed as barrier candidates.
+To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.
+
+Understanding the shell's primitives is not enough;
+  \sys has to understand the distribution characteristics of individual commands used to compose larger programs.
+This challenge can be broken down into two parts.
+First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
+One example is the set of commands defined by the \textsc{POSIX} standard~\cite{}.
+\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
+This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.
+
+The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
+Commands are arbitrary programs, but still conform to the shell's pipeline interface.
+To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
+Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
+A command developer can only annotate the default behavior;
+  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
+
+Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
+\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying maximal distributability sub-expressions.
+It converts the AST to a dataflow graph, and runs...
+This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc
+
+Finally, given a distributed pipeline with runtime \todo{annotations}, \sys needs to schedule and execute it.
+This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
+First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
+  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
+The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
+
+Second, \sys needs to fill-in planing details left blank by the rewriting pass.
+This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
+From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.
+
+The next few sections discuss the details.
+They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
+
+\section{Distributability Classes}
+
 \subsection{Comnands}
@@
 
-Distributability classes, based on how
+\begin{table}[h]
+\center
+\footnotesize
+\setlength\tabcolsep{3pt}
+\caption{
+  \footnotesize{
+    \textbf{Distributability Classes}.
+    Broadly, \unix commands can be broken down into to five classses.
+  }
+}
+\begin{tabular}{l @{\extracolsep{\fill}} lll}
+\toprule
+Class                           & Example Commands        & POSIX            & Core-utils\\
+\midrule
+Stateless                       &                         &                  &           \\
+Pure                            &                         &                  &           \\
+(Distr.) File-System            &                         &                  &           \\
+Side-effectful                  &                         &                  &           \\
+Irreversibly Side-effectful     &                         &                  &           \\
+% \midrule
+% Shell PL Constructs             &                         &                  \\
+% \etc                            &                         &                  \\
+\bottomrule
+\end{tabular}
+\label{tab:classes}
+\end{table}
+
+Define distributability 
+Explain distributability classes, based on how
 

### 2019-11-20 pldi/draft.tex
Progress on the intermediate representation section, especially the part about the graph nodes

--- 2019-11-20 pldi/draft.tex
@@
 \usepackage{cleveref}
-\usepackage{wrapfig}
 
@@
 \newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
-\newcommand{\kk}[1]{[{\color{magenta}#1 --- KK}]}
+\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 \section{Introduction}
-\label{intro}
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_schematic.pdf}
+\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
 \caption{
-  \textbf{\sys high-level schematic.}
-	\sys XYZ
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
 }
-\label{fig:schematic}
+\label{fig:spectrum}
 \end{figure}
@@
 
-% \begin{figure*}[t]
-% \centering
-% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
-% \caption{
-%   \textbf{Applying \sys to $p_1$.}
-% }
-% \label{fig:example}
-% \end{figure*}
-%
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+\caption{
+  \textbf{Applying \sys to $p_1$.}
+}
+\label{fig:example}
+\end{figure*}
 
@@
 
-\subsection{Unix Composition, Informally}
+\subsection{Unix Pipelines, Informally}
 \label{bg:pipelines}
 
-% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
-\unix provides several ways for succinct program composition.
-Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
-These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
-Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
-\begin{wrapfigure}{r}{0.06\textwidth}
-  \vspace{-12pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
-  % \end{center}
-  \vspace{-30pt}
-\end{wrapfigure}
-This interface exposes a stream of contiguous character lines.
-A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
-As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
-% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
-% generating errors or diagnostics to the standard 
-% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}
+Pipelines are a mechanism for program composition.
+Individual programs, specified as commands, conform to a common but general interface. 
+For now, this interface allows a combination of the following:
+  accepting input from the standard input stream (stdin),
+  generating output to the standard output stream (stdout), and
+  generating errors or diagnostics to the standard error stream (stderr)
+Standard streams are streams of characters with a few special characters---for example, EOF is a special end-of-file character signaling the end of the input stream.
+% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.
 
-Commands can be tuned by two mechanisms, namely command-line options and environment variables.
-Options control the execution of a program---\eg changing the \ttt{sort}ing order.
-The shell does not have any visibility into these options; 
-  their parsing and evaluation is left entirely up to each command.
-Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
-They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
+Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
+Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
+  programs do one thing well, they work together, and they manipulate text streams as their universal interface.
+The next few paragraphs describe the details.
 
-% TODO: Talk about security primitives?
-Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
-  files are simply persistent streams that can be manipulated by commands or pipelines.
-\begin{wrapfigure}{r}{0.15\textwidth}
-  \vspace{-15pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
-  % \end{center}
-  \vspace{-25pt}
-\end{wrapfigure}
-Reading a file or a set of files generates a stream; 
-  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
-\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
-As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
+\heading{Basics}
+The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).
 
-% At a high level, POSIX defines several file types:
-% (-) regular files, storing lines of text or binary data,
-% (d) directories, grouping together multiple other files,
-% (l) symbolic links, pointing to other files,
-% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
-% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
-% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
+that are not interpreted by the shell, but are local to each command.
 
-Special files are particularly interesting because they are resolved to OS structures.
-One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
-  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
-% Here are a few examples:
-% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
-% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
-% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
-Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
-This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
-
-A small but powerful set of features is concerned with stream manipulation.
-Some of these features are built into the shell;
-  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
-\begin{wrapfigure}{r}{0.15\textwidth}
-  \vspace{-15pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
-  % \end{center}
-  \vspace{-25pt}
-\end{wrapfigure}
-As described before, the file can be a FIFO dubbing as input to another command.
-More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
-  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
-% TODO: talk about \ttt{xargs}?
-
-% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators
-
-The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
-The most interesting feature of these pipelines is the presence of cycles:
-  a pipeline segment processes some input, part of which is extracting the next input to process.
-One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
-% These streams are created by the C library (glibc) at the start of program execution,
-% and new streams can be created to connect to files, sockets, pipes,
-
-Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
-Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
-Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
-  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
-Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
-
-% \begin{lstlisting}[
-%   language=es,
-%   mathescape,
-%   float=t,
-%   belowskip=-5mm,
-%   label=core-pl,
-%   upquote=true,
-%   caption={
-%     \textbf{The core language of the POSIX shell.}
-%     The language used by ; file identifiers and expansion are not shown.
-%   }]
-% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
-% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
-% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
-% \end{lstlisting}
-
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_4$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  ls -l | wc -l                           (($$p_2$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat * | wc -l > out.txt                 (($$p_3$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_5$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-% { echo 1; echo 2; } > out.txt
-% \end{lstlisting}
+Additional details augment this simple interface.
+Commands can be specified with options arguments
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
-\caption{
-  \textbf{\sys transformation overview.}
-	\sys XYZ
-}
-\label{fig:overview}
+\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
+\label{fig:example}
 \end{figure}
 
-% \heading{Generation}
-% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
-% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ ls -l | wc -l                           (($$p_2$$))
+\end{lstlisting}
 
-\subsection{Overview}
+\heading{Generation}
+A central concept in pipelines---and Unix, in general---is the file system, which allows interfacing with persistent data.
+There are many commands that take as input a file identifier 
+Reading a file or a set of files generates a stream; 
+  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
+
+Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
+The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat * | wc -l > out.txt                 (($$p_3$$))
+\end{lstlisting}
+
+\heading{Identifiers}
+
+\kk{We have to mention character vs line based parallelism somewhere,
+  and that we choose to see lines as the data quantum}
+
+The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).
+as such, different identifiers many not resolve to actual files, but rather in-memory or OS-internal structures.
+
+At a high level, POSIX defines several file types:
+(-) regular files, storing lines of text or binary data,
+(d) directories, grouping together multiple other files,
+(l) symbolic links, pointing to other files,
+(p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
+(s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
+(b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
+
+Most of these identifiers are manipulated by file-system commands---\eg \ttt{mkdir} for directories and \ttt{mkfifo} for FIFOs.
+FIFOs and DSes look like files but are streams.
+
+Special files are particularly interesting because they are resolved to OS structures.
+One set is pseudo-devices---device files that simply;
+For example /dev/urandom returns a stream of random characters and /dev/null accepts and discards all input written to it.
+Another set is proc filesystem (procfs), a special filesystem in Unix-like operating systems that presents information about processes and other system information.
+For example, /proc/PID/cmdline contains command that originally started the process and /proc/PID/cwd contains a symlink to the current working directory of the process.
+
+sysfs is a pseudo file system provided by the Linux kernel that exports information about various kernel subsystems, hardware devices, and
+
+/modules, one of the most important files in /proc, containing a list of the kernel modules currently loaded . It gives some indication (not always entirely correct) of dependencies.
+/cpuinfo
+
+symbolic links, FIFO specials, 
+device files
+block special, character special, 
+and domain sockets.
+
+There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
+These correspond to \ttt{/dev/fd/0}.
+% These streams are created by the C library (glibc) at the start of program execution,
+and new streams can be created to connect to files, sockets, pipes,
+
+Unix already specifies a few classes of files:
+
+Here are a few special files:
+  \ttt{proc} called
+
+Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
+Similarly, open files and other ephemeral identifiers
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat /proc/ >                            (($$p_4$$))
+\end{lstlisting}
+
+\heading{Stream Manipulation}
+A small but powerful set of primitives are concerned with stream manipulation.
+First, as described above, several commands can be redirected
+Second,
+
+\ttt{xargs}
+\ttt{tee}
+
+In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
+
+Explicit Redirection
+Implicit splitting and merging 
+This can be done through the file-system operators
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat /proc/ >                            (($$p_5$$))
+\end{lstlisting}
+
+\heading{Composition}
+anonymous pipes
+
+Composition
+
+The shell provides several ways of grouping commands and manipulating their
+overall input and output channels
+
+{ echo 1; echo 2; } > out.txt
 
@@
 
+\section{Intermediate Representation/Dataflow graph}
+
+In order to effectively apply optimizations and distribute a shell
+script, it first has to be translated to a manipulable
+representation. As mentioned in \ref{}, the two main abstractions of
+the shell, are files that contain data, and commands that communicate
+through these files. We propose a slight variation of the standard
+dataflow graph, that is commonly found in stream processing systems,
+as the representation of a distributed shell computation. The dataflow
+graph nodes represent commands, and the edges represent files.
+
+\kk{Before that I have to mention that the basic data type $D$ is a
+  line of characters.}
+
+\paragraph{Nodes - Commands}
+
+Nodes $f$ of the graph represent functions from one (possibly empty)
+input stream to an output stream $f : D^* \rightarrow D^*$.  This
+representation captures the majority of shell commands. The nodes $f$
+are monotonic, meaning that they cannot retract their output. More
+formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
+that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
+omitted from now on. An example of a node is the following command
+\kk{Does it need ref? Also maybe this example is too trivial and can
+  be removed.}:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ grep ``foo''
+\end{lstlisting}
+
+\noindent
+This command takes a stream of lines from its standard input and
+returns a stream of lines on its standard output. An important
+observation though is that commands in the shell might read their
+input stream from a sequence of different sources. An illustrative
+example is the command \kk{Inline listing}cat which reads its input
+from a combination of files and its standard input.
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat x - y
+\end{lstlisting}
+
+\kk{Should I reference that this example was takes from the cat
+  manpage?} The example above reads file \texttt{x} until it
+encounters the end of file, then reads from \texttt{stdin}, and then
+reads from \texttt{y}. In order to handle this particularity, nodes in
+the dataflow graph can have many incoming edges, that are clearly
+ordered in a sequence. Note that this comes in contrast with standard
+dataflow graph representations where multiple incoming edges either
+mean zipping the input to pairs, or arbitraliy interleaving the
+different inputs.
+
+\kk{A note about nodes and commands that can write and read from more
+  files that the ones in their streams}
+
+Note that nodes can read from more files than just the files of their
+stream (or respectively write to more files that their output stream),
+however these files are static (e.g. used for configuration or logging
+\kk{Would an example here help}) and do not represent streams of data,
+and therefore are not considered part of the dataflow graph. Because
+of the assumption in section Front End \ref{}, that each file exists
+only once in the dataflow graph, we can safely assume that reading and
+writing to these static files does not interfere with the distributed
+implementation, and doesn't alter the behaviour of the program.
+
+\paragraph{Edges - Files}
+
+\kk{Maybe edges has to be moved before nodes, so that we can define
+  the line data type}
+
+Edges in the dataflow graph represent files, the basic data
+abstraction of the shell. They are used as communication channels
+between nodes in the graph, and as the input or output of the whole
+graph. Edges are represented as possibly unbounded streams of type
+$D*$, where $D$ represents a line, that is a sequence of characters
+followed by \verb|\n| \kk{or EOF?}.
+
+\kk{The following paragraph falls in the trap of just explaining
+  without motivating}
+
+Edges could either refer to a specific named file in the file system,
+or just be FIFO pipes that used for interprocess communication \kk{Are
+  there any other file types, such as URLs, ...?}. Edges that refer to
+real files, contain their exact number of lines, while edges that are
+fifo pipes can have an upper bound on the number of lines that they
+will transfer through them, thus being bounded in size.
+
+\kk{Depending on which of the two paragraphs goes last (nodes/edges),
+  I have to explain there that the last incoming edge to a node must
+  be unbounded (if it doesn't point to a resource).}
+
+\kk{Text point: The main shell abstraction is files, which are
+  sequences of lines, where the order matters. So we want a
+  representation that captures order of inputs and outputs (and ours
+  does because its node in the graph has a sequence of inputs.}
+
+\paragraph{Merge and Split}
+
+\kk{Mergers and splitters, and how they are implemented using shell constructs (cat and head)}
+
+As we described above, a node could have many incoming edges
+
+\kk{Section notes:}
+
+\kk{Files have an associated number of lines}
+
+\kk{Explain how the assumptions made in the front end give as correct
+  distribution}
+
 \section{Front End}
+
+\kk{Maybe this becomes a subsection either in the intermediate
+  representation or in the planner.}
 
@@
 \kk{Text point: Problem: Not all parts of a shell script are
-  distributable, so we cannot arbitrarily distribute}
+  distributable, so we cannot arbitrarily distribute. Here we have to
+  make our assumptions clear}
 
@@
 
-\subsection{Intermediate Representation/Dataflow graph}
+\kk{The same file exists once in the substree. This is a conservative
+  assumption that ensures that no two commands write or read to the
+  same file concurrently.}
 
-Mention its constructs, and its semantics, and why it is a bit
-different that standard dataflow graphs (where usually incoming edges
-are either consumed with some arbitrary interleaving, or pair by
-pair).
-
-\kk{Text point: The main shell abstraction is files, which are batches
-  or streams of characters (or lines) where the order matters. So we
-  want a representation that captures order of inputs and outputs (and
-  ours does because its node in the graph has a sequence of inputs.}
-
-Explain that the intermediate representation represents workers that
-communicate through files, and that it captures parallelism both in
-the operator level, as well as data parallelism.
+\kk{How can we make sure that nodes in the IR are deterministic. Do we
+  make some assumption that enforces this?}
 
@@
 
-\kk{Maybe this should be merged with the one above, especially if it
-  talks about how distribution happens on the intermediate
-  representation.}
+\kk{Mention that different nodes can be safely executed independently,
+  and that this exposes operator level parallelism}
 
-\subsection{Scaling up the distributed computation/Vertically expanding the graph}
+\subsection{Scaling up the distributed computation/Vertically expanding the graph/Graph Transformations/ Equations}
 
@@
 \kk{Give an example}
+
+\kk{Can we have a semi-formal argument here?}
+
+\kk{Explain how our graph transformations expose data parallelism as
+  operator level parallelism.}
+
+\kk{Different transformations to explain: Stateless, Pure, File
+  splits, XARGS. Mention as much as possible formally using the
+  definition of a node.}
+
+The following equation holds for any stateless command $f$.
+
+\[
+f(x.x') = f(x).f(x')
+\]
+
+This is the basis of all optimizations that the graph does, and allows
+for correct data parallelism.
 
@@
   replication, consistency, and failures.
+\item No cycles
 \end{itemize}
@@
 \section{Related Work}
-\label{related}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
-\caption{
-  \textbf{Cost of Manual Effort.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-A more complete picture is presented in the related-work section~\sx{rt}.
-}
-\label{fig:spectrum}
-\end{figure}
-
-\section{Conclusion}
-\label{discussion}
 

### 2019-11-20 pldi/draft.tex
Complete a first draft of Section 5

@@
 
+\kk{Try to be more formal in the section, by using the notation that
+  is defined in the nodes paragraph.}
+
 In order to effectively apply optimizations and distribute a shell
@@
 dataflow graph, that is commonly found in stream processing systems,
-as the representation of a distributed shell computation. The dataflow
-graph nodes represent commands, and the edges represent files.
+as the representation of a distributed shell computation. The main
+benefit of the dataflow graph model is that it clearly exposes
+operator level parallelism, as different nodes can be independent
+processing workers. In our case each node represents a command, and
+each edge represents a file.
 
 \kk{Before that I have to mention that the basic data type $D$ is a
-  line of characters.}
+  line of characters. If we move the edges section before the nodes
+  that will be fine.}
 
@@
 $D*$, where $D$ represents a line, that is a sequence of characters
-followed by \verb|\n| \kk{or EOF?}.
-
-\kk{The following paragraph falls in the trap of just explaining
-  without motivating}
-
-Edges could either refer to a specific named file in the file system,
-or just be FIFO pipes that used for interprocess communication \kk{Are
-  there any other file types, such as URLs, ...?}. Edges that refer to
-real files, contain their exact number of lines, while edges that are
-fifo pipes can have an upper bound on the number of lines that they
-will transfer through them, thus being bounded in size.
-
-\kk{Depending on which of the two paragraphs goes last (nodes/edges),
-  I have to explain there that the last incoming edge to a node must
-  be unbounded (if it doesn't point to a resource).}
-
-\kk{Text point: The main shell abstraction is files, which are
-  sequences of lines, where the order matters. So we want a
-  representation that captures order of inputs and outputs (and ours
-  does because its node in the graph has a sequence of inputs.}
+followed by \verb|\n| \kk{or EOF?}. They could either refer to a
+specific named file in the file system, or just be FIFO pipes that
+used for interprocess communication \kk{Are there any other file
+  types, such as URLs, ...?}.
+
+\kk{Show an example pipe and its graph}.
+
+In order to expose operator parallelism with the output of a command,
+files that don't point to a resource can have an upper bound on the
+lines that they can tranfer. The following example illustrates why
+this is essential.
+
+\kk{An example of a command that has two outputs and one goes to one
+  command, while another output goes to a different one. Maybe only
+  show the graph here and not the command itself.}
+
+On the other hand, some forms of data parallelism can be exposed when
+knowing the size of the input files. As mentioned in \ref{} some pure
+commands (such as cat -n) only need line information to become
+stateless, and knowing the size of a file could allow the system to
+split it in different chunks that can be processed independently. To
+account for that, edges that refer to an input resource contain the
+number of lines of the file that they refer to. \kk{At the moment this
+  is not supported by our real system. The whole paragraph above can
+  go if we decide that it is not a good idea to talk about it now.}
+
+Finally, the edges that don't refer to a resource and do not start
+from a node in the graph represent the graph input, while the edges
+that don't refer to a resource and do not point to a node in the graph
+are its outputs.
+
+\kk{(Maybe) Depending on which of the two paragraphs goes last
+  (nodes/edges), I have to explain there that the last incoming edge
+  to a node must be unbounded (if it doesn't point to a resource). I
+  am actually not sure whether I should include this here.}
+
+\kk{(Maybe) It might be beneficial to mention that many inputs do not
+  represent the same thing as many outputs. These two notions are not
+  symmetric.}
 
@@
 
-\kk{Mergers and splitters, and how they are implemented using shell constructs (cat and head)}
-
-As we described above, a node could have many incoming edges
-
-\kk{Section notes:}
-
-\kk{Files have an associated number of lines}
-
-\kk{Explain how the assumptions made in the front end give as correct
-  distribution}
+\kk{Mergers and splitters, and how they are implemented using shell
+  constructs (cat and head). This subsection might be a better fit in
+  some other section, maybe something related to the planner or the
+  implementation, as it is a very elegant way to implement splitters
+  and mergers while staying completely inside the shell programming
+  model.}
+
+It is common in the dataflow graph for a node to have many incoming
+edges, or many outgoing edges (when all but the last outgoing edges
+are bounded). However some nodes, might read their input from one file
+(e.g. stdin). In this case, the distributed implementation needs to
+concatenate input files, and split output files. Here is an example:
+
+\kk{Show an example graph! of cat | tr | sort. The command in the
+  middle must only take input from one file (maybe just stdin).}
+
+\kk{This transition needs to be stronger} Luckily, the shell offers
+primitives for file manipulations that can be used to implement a
+merger and a splitter. More concretely a merger can be implemented by
+a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
+and an output file \verb|out|, a merger can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ cat $in1 $in2 ... > $out
+\end{lstlisting}
+
+\noindent
+On the other hand, given an input file \verb|in| that is a FIFO pipe,
+and two output files \verb|out1, out2| the first of which is bounded
+to \verb|N| lines, a 2-splitter can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ head -n $N $in > $out1 ; cat $in > $out2
+\end{lstlisting}
+
+\noindent
+Using the 2-splitter as a building block, a pipe can be split to
+arbitrarily many different pipes.
+
+\kk{Do I also need to include a splitter for normal files?}
+
+\kk{Intermediate Representation Section notes:}
+
+\kk{Explain how the assumptions made in the front end give as
+  correct/safe operator level parallelism. Maybe this must go in the
+  Front-end section.}
 

### 2019-11-20 pldi/draft.tex
Add a very drafty draft of the frond end section

@@
 
+\kk{The title is too abstract and doesn't really hint that we
+  distribute shell code. Also we could consider capitalizing the name
+  as DiSh?}
+
 % 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
@@
 \section{Extensibility}
+\label{ext}
 
@@
 \section{Intermediate Representation/Dataflow graph}
+\label{ir}
 
@@
 
-\kk{Maybe this becomes a subsection either in the intermediate
-  representation or in the planner.}
-
-We parse a shell script using a posix compliant parser that returns
-the ast.
-
-\kk{Text point: The AST Is not a very good data structure to
-  manipulate for distribution, so we translate it to a kind of a
-  dataflow graph.}
+\kk{It might make sense to merge this section with the one below with
+  the name: From Shell to IR and back}
+
+As mentioned in \Cref{ir}, in order to generate the distributed
+implementation of a shell script, \texttt{Dish} first translates it to
+a dataflow graph. However, there are several components in shell
+programs that cannot be arbitrarily parallelized without affecting the
+program behaviour.
+
+\kk{Give examples of writing to environment variables, producing
+  irreversible side effects, the ; and \&\& operators.}
+
+In order to address this issue, we introduce the notion of
+distributable AST subtrees, and we design an analysis that given the
+shell script Abstract Syntax Tree identifies distributable AST
+subtrees and translates each one of them to a dataflow graph. The AST
+of the shell script is produced using a POSIX compliant parser
+\kk{Reference libdash github}.
 
@@
 
-\kk{Text point: Problem: Not all parts of a shell script are
-  distributable, so we cannot arbitrarily distribute. Here we have to
-  make our assumptions clear}
-
-Even though commands are split in categories, and several pipelines
-are distributable, not all parts of a shell script are stateless and
-several of them have effects that could affect the behaviour of a
-program after distribution.
-
-\kk{Solution: We have found that distributing subtrees of the shell
-  AST is safe. Can we have a semi-formal argument about that?}
-
-Mention all the different constructs that our front end handles, and
-why it is safe to add them to the distribution pipeline.
-
-\kk{The same file exists once in the substree. This is a conservative
-  assumption that ensures that no two commands write or read to the
-  same file concurrently.}
-
-\kk{How can we make sure that nodes in the IR are deterministic. Do we
-  make some assumption that enforces this?}
+\kk{The goal of this section is to convince the reader ``intuitively''
+  that we choose safe parts of a shell program to distribute. There is
+  no formal argument, but one should be able to intuitively believe
+  us.}
+
+Consider the following example:
+
+\kk{Give an example of a shell script that cannot be distributed
+  completely, Possibly involving ; and writing to a variable.}
+
+\kk{Intuitively explain what could be distributed and what not.}
+
+\kk{Our notion of distributable subtrees has two dimensions. First,
+  the state of the shell should be an invariant in a distributable
+  subtree. We cannot know this completely without evaluating, so we do
+  a conservative analysis. The second dimension is that there exists
+  at least one component that can be abstracted as a node of the
+  dataflow graph. This can be anything that includes a command.}
+
+%% Even though commands are split in categories, and several pipelines
+%% are distributable, not all parts of a shell script are stateless and
+%% several of them have effects that could affect the behaviour of a
+%% program after distribution.
+
+\kk{All of our assumptions and study are based on the POSIX
+  standard. Reference?}
+
+\kk{Not checked Assumption. A command doesn't write/read to arbitrary
+  files that are not mentioned in its arguments. This is necessary to
+  enforce that no two commands concurrently read/write from the same
+  file.}
+
+\kk{Not checked Assumption: No signals, traps, and sets at the
+  moment. At the moment we don't have a way to handle signals and
+  traps.}
+
+\kk{Partially checked Assumption. The same file exists once in the
+  substree. This is a conservative assumption that ensures that no two
+  commands write or read to the same file concurrently. We kind of
+  check it but not completely. We check that the expanded evaluated
+  files are not the same, but one could have part of a file in a
+  variable etc.}
+
+\kk{Checked assumption: How can we make sure that nodes in the IR are
+  deterministic. Do we make some assumption that enforces this? I
+  guess that commands that are nodes in the IR are all deterministic?
+  That is stateless and pure commands are deterministic.}
+
+\paragraph{analysis}
+
+The analysis is done in a DFS fashion on the AST. Distributable
+subtrees grow from the bottom up, until a construct that doesn't
+``allow'' them to go through is encountered.
+
+%% Mention all the different constructs that our front end handles, and
+%% intuitively mention why it is ``safe'' to add them to the distribution
+%% pipeline.
+
+Subtrees don't cross the following:
+
+\begin{itemize}
+\item Sequence, And,  or
+\item Function definitions
+\item Command Substitutions
+\item control flow: while, for, if, case: While these constructs can
+  be parallelized and there has been research on it, it is orthogonal
+  to our work, and can be incorporated as future work.
+\end{itemize}
+
+On the other hand: A distributable subtree starts whenever there is a
+pipe, or asynchronous list. If the subcomponents of the pipe or
+asynchronous list were already graphs, then we connect them to the
+bigger graph \kk{explain how we do pipe and asynchronous list}, and if
+they were not graphs, we make each subcomponent a node in the subtree,
+event though they are not commands. These nodes do not belong to any
+category, but still have stdin and stdout, and can be executed
+concurrently (as long as they satisfy the checked assumptions above),
+since they don't introduce new behaviours, they would still be
+executed concurrently when ran in the standard shell, producing
+possibly non-deterministic results with different interleavings.
+
+\kk{Explain how we merge two graphs when seeing a pipe}
+
+Whenever we see a subtree of the AST that is a command, we check the
+files that it writes to. If any of its arguments is a command
+substitution or contains anything that is evaluated\kk{make this more
+  precise}, its category is marked as sequential\kk{name?}. Otherwise
+we check its its category and its input output arguments (using the
+methods described in \Cref{ext}). The command is then initialized as
+the node of a singular graph, with the inputs and outputs of the graph
+being the ones of the node.
+
+\kk{Maybe mention the recursive case. I am not sure if that should be
+  here, or in the distribution planner. In any case, it should be
+  after we say that an IR is replaced with a call to the planner with
+  the IR saved in a file. The example is a command in a pipeline whose
+  first argument is a command substitution with a pipeline. Executing
+  the outside dataflow graph, would lead to the command to be executed
+  as part of this shell. However its first argument would have a call
+  to the planner itself, leading to another distributed implementation
+  running in parallel. The rest of the pipeline would wait for this to
+  end before it continues. Assuming that the files that both of these
+  graphs use are different, and that the above assumptions hold, there
+  should be no interference between the two graphs and they should be
+  able to run in parallel}
 
@@
 evaluate strings, just expand arguments, and not do any significant
-computation.
+computation. Even better, we could be calling the shtepper just after
+having parsed the ast, and decide on maximal distributable subtrees as
+late as possible in the process. This would be awesome.
+
+\section{Implementation}
+
+\kk{We should make sure to measure the time that it takes for the
+  whole process to run (parsing, translating, optimizing, planning)
+  together with the scripts. This will probably be negligible, but we
+  still have to mention it. }
+
+\kk{Also, we have to have one-two handcrafted examples that have more
+  than one pipeline, to show the execution time end to end, as this
+  cannot be shown with the one-pipeline examples since they only have
+  one graph, so no back and forth between shell and dish.}
+
+\kk{Furthermore it would be great to have one distributed example that
+  runs in more than one node (if we have time)}
+
+\kk{Also we have to make sure to measure with and without the final
+  cat of output file, so that we show that this is also negligible.}
+
+\kk{We have to make sure that we repeat that the results are the same
+  in all tests that we run. Corectness, blah blah ...}
 
@@
   replication, consistency, and failures.
-\item No cycles
+\item No cycles (multiple commands writing and reading from the same file)
+\item Formal proof that the analysis indeed returns distributable
+  subtrees. Previously there was no formal semantics for shell, but
+  recent work (cite greenberg) allows us to make a formal proof that
+  the analysis is indeed sound.
 \end{itemize}

### 2019-11-21 pldi/draft.tex
A draft on the distribution planner, graph transformations

@@
 
-Can we find a solution for the commands in coreutils?
+\kk{Can we find a solution for the commands in coreutils?}
+
+\kk{Important TODO: Find at least one more command that can be
+  executed like sort. Does wc work? If there is no pure command that
+  can be easily parallelized other than sort, this story is not very
+  nice.}
 
@@
 \label{ir}
+
+\kk{Inline listing?}
 
@@
 benefit of the dataflow graph model is that it clearly exposes
-operator level parallelism, as different nodes can be independent
+operator level parallelism \kk{Is this a common term? Is there any
+  definitive citation for it?}, as different nodes can be independent
 processing workers. In our case each node represents a command, and
@@
 
-\kk{Partially checked Assumption. The same file exists once in the
-  substree. This is a conservative assumption that ensures that no two
-  commands write or read to the same file concurrently. We kind of
-  check it but not completely. We check that the expanded evaluated
-  files are not the same, but one could have part of a file in a
-  variable etc.}
+\kk{Partially checked Assumption. Only one reader and one writer to
+  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
+  file exists once in the substree. This is a conservative assumption
+  that ensures that no two commands write or read to the same file
+  concurrently. We kind of check it but not completely. We check that
+  the expanded evaluated files are not the same, but one could have
+  part of a file in a variable etc.}
 
@@
 
-\kk{Mention that different nodes can be safely executed independently,
-  and that this exposes operator level parallelism}
-
-\subsection{Scaling up the distributed computation/Vertically expanding the graph/Graph Transformations/ Equations}
-
-The planner acts on the dataflow graph, by first expanding it ``as
-much as possible'' to expose parallelization opportunities. It uses
-knowledge about the inputs of each stateless and pure command.
-
-\kk{Explain the algorithm and how it pushes parallelization through
-  the graph, while preserving correctness.}
-
-\kk{Give an example}
-
-\kk{Can we have a semi-formal argument here?}
-
-\kk{Explain how our graph transformations expose data parallelism as
-  operator level parallelism.}
-
-\kk{Different transformations to explain: Stateless, Pure, File
-  splits, XARGS. Mention as much as possible formally using the
-  definition of a node.}
-
-The following equation holds for any stateless command $f$.
+\kk{I think that there is not enough content for the planner section
+  and it should probably just be called graph transformations. Then we
+  could have a small subsection at the end about how the planner
+  applies transformations iteratively starting from the source of the
+  graph and pushing them until the end.}
+
+As mentioned in \Cref{ir}, the dataflow graph exposes operator level
+parallelism. Different nodes in the graph can be executed
+concurrently, communicating through the edges. Operator level
+parallelism is already exposed by the shell, by using pipes \texttt{|}
+and the background operator \texttt{\&}. Unfortunately, simple
+operator level parallelism leaves a lot to be desired, since scaling a
+computation requires manually introducing more operators.
+
+In order to adequately utilize the available computational resources,
+a distributed implementation should exploit data parallelism. Data
+parallelism is based on the observation that some operations produce
+the same result when performed independently on subsets of the input
+data, merging the results afterwards. Data level parallelism is the
+basic optimization that is used by popular distributed processing
+systems such as MapReduce\kk{cite} and Apache Spark\kk{cite}.
+
+\subsection{Graph transformations}
+
+In our case, this is achieved by iteratively applying graph
+transformations that preserve correctness, while exposing
+parallelization opportunities. We illustrate that with a motivating
+example; consider the following program and its dataflow graph.
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ grep ``foo'' f1 f2 // Add its graph
+\end{lstlisting}
+
+\noindent
+In this example, \texttt{grep} is applied to files f1 and f2, and its
+result is sent to standard output. However, the result would be the
+same if we concatenated the results of \texttt{grep ``foo'' f1} and
+\texttt{grep ``foo'' f2}. This intuition can be generalized to all
+stateless commands, as they are all essentially ``linear'' with
+respect to concatenation and the following equation holds.
 
 \[
-f(x.x') = f(x).f(x')
+\forall f, \mathrm{stateless}(f) \implies f(x.x') = f(x).f(x')
 \]
 
-This is the basis of all optimizations that the graph does, and allows
-for correct data parallelism.
+\kk{Complete this story with pure, after we have written the section
+  about command categories.}
+
+\subsection{Planner Algorithm}
+
+The \texttt{Dish} planner extends the transformations that were
+described above with edge splits in order to allow for the above
+transformations to be performed. More precisely, it splits input files
+and graph edges into several chunks so that they can be processed in
+parallel by the next levels.
+
+\kk{Explain the greedy algorithm that our planner performs. It splits
+  input files, and then pushes parallelization through the graph,
+  starting from the source nodes.}
+
+\kk{Maybe give an example}
 
 \subsection{Mapping operators to nodes}
+
+\kk{I think we should actually remove this completely.}
 
@@
 \subsection{Just in time planning}
+
+\kk{This whole section for just in itme planning, is also relevant for
+  the front end. If the front end is executed just in time, it would
+  have more information, and maybe it could make a sound analysis.}
 
@@
 
-\section{Implementation}
+\section{Evaluation}
 
@@
   in all tests that we run. Corectness, blah blah ...}
+
+\kk{We should have at least some pipelines that have a bad pure
+  command in them.}
 
@@
 
+\kk{We need to cite DGSH}
+
 \begin{acks}

### 2019-11-21 evaluation/minimal_grep_100_distr.sh
Add a minimal example with grep in the evaluation

@@ -1,0 +1,1301 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file129"
+mkfifo "#file129"
+rm -f "#file155"
+mkfifo "#file155"
+rm -f "#file112"
+mkfifo "#file112"
+rm -f "#file318"
+mkfifo "#file318"
+rm -f "#file275"
+mkfifo "#file275"
+rm -f "#file398"
+mkfifo "#file398"
+rm -f "#file107"
+mkfifo "#file107"
+rm -f "#file325"
+mkfifo "#file325"
+rm -f "#file150"
+mkfifo "#file150"
+rm -f "#file405"
+mkfifo "#file405"
+rm -f "#file265"
+mkfifo "#file265"
+rm -f "#file289"
+mkfifo "#file289"
+rm -f "#file277"
+mkfifo "#file277"
+rm -f "#file360"
+mkfifo "#file360"
+rm -f "#file142"
+mkfifo "#file142"
+rm -f "#file223"
+mkfifo "#file223"
+rm -f "#file343"
+mkfifo "#file343"
+rm -f "#file378"
+mkfifo "#file378"
+rm -f "#file249"
+mkfifo "#file249"
+rm -f "#file381"
+mkfifo "#file381"
+rm -f "#file118"
+mkfifo "#file118"
+rm -f "#file236"
+mkfifo "#file236"
+rm -f "#file314"
+mkfifo "#file314"
+rm -f "#file302"
+mkfifo "#file302"
+rm -f "#file231"
+mkfifo "#file231"
+rm -f "#file140"
+mkfifo "#file140"
+rm -f "#file156"
+mkfifo "#file156"
+rm -f "#file261"
+mkfifo "#file261"
+rm -f "#file263"
+mkfifo "#file263"
+rm -f "#file320"
+mkfifo "#file320"
+rm -f "#file373"
+mkfifo "#file373"
+rm -f "#file247"
+mkfifo "#file247"
+rm -f "#file244"
+mkfifo "#file244"
+rm -f "#file307"
+mkfifo "#file307"
+rm -f "#file180"
+mkfifo "#file180"
+rm -f "#file400"
+mkfifo "#file400"
+rm -f "#file355"
+mkfifo "#file355"
+rm -f "#file134"
+mkfifo "#file134"
+rm -f "#file256"
+mkfifo "#file256"
+rm -f "#file184"
+mkfifo "#file184"
+rm -f "#file207"
+mkfifo "#file207"
[…]

### 2019-11-21 evaluation/minimal_grep_2_seq.sh

@@ -1,0 +1,1 @@
+cat $IN $IN | tr A-Z a-z | grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4'

### 2019-11-21 evaluation/minimal_grep_4_distr.sh

@@ -1,0 +1,53 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file17"
+mkfifo "#file17"
+rm -f "#file18"
+mkfifo "#file18"
+rm -f "#file11"
+mkfifo "#file11"
+rm -f "#file14"
+mkfifo "#file14"
+rm -f "#file16"
+mkfifo "#file16"
+rm -f "#file12"
+mkfifo "#file12"
+rm -f "#file13"
+mkfifo "#file13"
+rm -f "#file20"
+mkfifo "#file20"
+rm -f "#file22"
+mkfifo "#file22"
+rm -f "#file15"
+mkfifo "#file15"
+rm -f "#file21"
+mkfifo "#file21"
+rm -f "#file19"
+mkfifo "#file19"
+cat $IN > "#file11" &
+cat $IN > "#file12" &
+cat $IN > "#file13" &
+cat $IN > "#file14" &
+cat "#file11" | tr A-Z a-z > "#file15" &
+cat "#file12" | tr A-Z a-z > "#file16" &
+cat "#file13" | tr A-Z a-z > "#file17" &
+cat "#file14" | tr A-Z a-z > "#file18" &
+cat "#file15" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file19" &
+cat "#file16" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file20" &
+cat "#file17" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file21" &
+cat "#file18" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file22" &
+cat "#file19" > /tmp/dish_output//0
+cat "#file20" > /tmp/dish_output//1
+cat "#file21" > /tmp/dish_output//2
+cat "#file22" > /tmp/dish_output//3
+rm -f "#file17"
+rm -f "#file18"
+rm -f "#file11"
+rm -f "#file14"
+rm -f "#file16"
+rm -f "#file12"
+rm -f "#file13"
+rm -f "#file20"
+rm -f "#file22"
+rm -f "#file15"
+rm -f "#file21"
+rm -f "#file19"

### 2019-11-21 evaluation/minimal_sort_10_distr.sh

@@ -1,66 +1,66 @@
 mkdir -p /tmp/dish_output/
-rm -f "#file31"
-mkfifo "#file31"
-rm -f "#file40"
-mkfifo "#file40"
-rm -f "#file34"
-mkfifo "#file34"
-rm -f "#file23"
-mkfifo "#file23"
-rm -f "#file35"
-mkfifo "#file35"
+rm -f "#file22"
+mkfifo "#file22"
+rm -f "#file30"
+mkfifo "#file30"
+rm -f "#file45"
+mkfifo "#file45"
+rm -f "#file28"
+mkfifo "#file28"
+rm -f "#file29"
+mkfifo "#file29"
+rm -f "#file39"
+mkfifo "#file39"
 rm -f "#file36"
 mkfifo "#file36"
-rm -f "#file44"
-mkfifo "#file44"
+rm -f "#file31"
+mkfifo "#file31"
+rm -f "#file25"
+mkfifo "#file25"
 rm -f "#file41"
 mkfifo "#file41"
 rm -f "#file21"
 mkfifo "#file21"
-rm -f "#file32"
-mkfifo "#file32"
-rm -f "#file39"
-mkfifo "#file39"
-rm -f "#file38"
-mkfifo "#file38"
-rm -f "#file45"
-mkfifo "#file45"
-rm -f "#file27"
-mkfifo "#file27"
 rm -f "#file43"
 mkfifo "#file43"
-rm -f "#file30"
-mkfifo "#file30"
-rm -f "#file28"
-mkfifo "#file28"
-rm -f "#file19"
-mkfifo "#file19"
-rm -f "#file42"
-mkfifo "#file42"
-rm -f "#file24"
-mkfifo "#file24"
-rm -f "#file16"
-mkfifo "#file16"
-rm -f "#file20"
-mkfifo "#file20"
-rm -f "#file18"
-mkfifo "#file18"
-rm -f "#file22"
-mkfifo "#file22"
-rm -f "#file33"
-mkfifo "#file33"
+rm -f "#file44"
+mkfifo "#file44"
 rm -f "#file26"
 mkfifo "#file26"
-rm -f "#file25"
-mkfifo "#file25"
-rm -f "#file37"
-mkfifo "#file37"
-rm -f "#file29"
-mkfifo "#file29"
+rm -f "#file35"
+mkfifo "#file35"
+rm -f "#file20"
+mkfifo "#file20"
+rm -f "#file38"
+mkfifo "#file38"
 rm -f "#file46"
[…]

### 2019-11-21 evaluation/minimal_sort_2_distr.sh

@@ -1,18 +1,18 @@
 mkdir -p /tmp/dish_output/
 rm -f "#file10"
 mkfifo "#file10"
-rm -f "#file9"
-mkfifo "#file9"
 rm -f "#file12"
 mkfifo "#file12"
-rm -f "#file14"
-mkfifo "#file14"
 rm -f "#file13"
 mkfifo "#file13"
-rm -f "#file11"
-mkfifo "#file11"
 rm -f "#file8"
 mkfifo "#file8"
+rm -f "#file11"
+mkfifo "#file11"
+rm -f "#file9"
+mkfifo "#file9"
+rm -f "#file14"
+mkfifo "#file14"
 cat $IN > "#file9" &
 cat $IN > "#file10" &
 cat "#file9" | tr A-Z a-z > "#file11" &
@@ -22,9 +22,9 @@ cat "#file12" | sort  > "#file14" &
 sort -m "#file13" "#file14" > "#file8" &
 cat "#file8" > /tmp/dish_output//0
 rm -f "#file10"
-rm -f "#file9"
 rm -f "#file12"
-rm -f "#file14"
 rm -f "#file13"
-rm -f "#file11"
 rm -f "#file8"
+rm -f "#file11"
+rm -f "#file9"
+rm -f "#file14"

### 2019-11-21 evaluation/minimal_sort_50_distr.sh

@@ -1,306 +1,306 @@
 mkdir -p /tmp/dish_output/
-rm -f "#file105"
-mkfifo "#file105"
-rm -f "#file190"
-mkfifo "#file190"
-rm -f "#file118"
-mkfifo "#file118"
-rm -f "#file206"
-mkfifo "#file206"
-rm -f "#file151"
-mkfifo "#file151"
-rm -f "#file115"
-mkfifo "#file115"
-rm -f "#file170"
-mkfifo "#file170"
 rm -f "#file155"
 mkfifo "#file155"
-rm -f "#file58"
-mkfifo "#file58"
-rm -f "#file168"
-mkfifo "#file168"
-rm -f "#file182"
-mkfifo "#file182"
-rm -f "#file85"
-mkfifo "#file85"
-rm -f "#file94"
-mkfifo "#file94"
-rm -f "#file187"
-mkfifo "#file187"
-rm -f "#file162"
-mkfifo "#file162"
-rm -f "#file154"
-mkfifo "#file154"
-rm -f "#file140"
-mkfifo "#file140"
-rm -f "#file107"
-mkfifo "#file107"
-rm -f "#file156"
-mkfifo "#file156"
-rm -f "#file138"
-mkfifo "#file138"
-rm -f "#file174"
-mkfifo "#file174"
-rm -f "#file161"
-mkfifo "#file161"
-rm -f "#file172"
-mkfifo "#file172"
-rm -f "#file177"
-mkfifo "#file177"
-rm -f "#file119"
-mkfifo "#file119"
-rm -f "#file77"
-mkfifo "#file77"
-rm -f "#file109"
-mkfifo "#file109"
-rm -f "#file200"
-mkfifo "#file200"
-rm -f "#file149"
-mkfifo "#file149"
-rm -f "#file202"
-mkfifo "#file202"
-rm -f "#file191"
-mkfifo "#file191"
-rm -f "#file199"
-mkfifo "#file199"
-rm -f "#file132"
-mkfifo "#file132"
-rm -f "#file186"
-mkfifo "#file186"
-rm -f "#file72"
-mkfifo "#file72"
-rm -f "#file124"
-mkfifo "#file124"
-rm -f "#file145"
-mkfifo "#file145"
-rm -f "#file117"
-mkfifo "#file117"
-rm -f "#file192"
-mkfifo "#file192"
-rm -f "#file181"
-mkfifo "#file181"
-rm -f "#file96"
-mkfifo "#file96"
[…]

### 2019-11-21 front_end/translate_evaluation_scripts.sh

@@ -9,3 +9,11 @@ distr_output_dir=/tmp/dish_output/
 ./translate_script.sh "../evaluation/minimal_sort_20" $distr_output_dir
 ./translate_script.sh "../evaluation/minimal_sort_50" $distr_output_dir
 ./translate_script.sh "../evaluation/minimal_sort_100" $distr_output_dir
+
+./translate_script.sh "../evaluation/minimal_grep_2" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_4" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_10" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_20" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_50" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_100" $distr_output_dir
+./translate_script.sh "../evaluation/minimal_grep_200" $distr_output_dir

### 2019-11-21 evaluation/minimal_grep_10_distr.sh
Add a minimal example with grep in the evaluation

@@ -1,0 +1,131 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file19"
+mkfifo "#file19"
+rm -f "#file40"
+mkfifo "#file40"
+rm -f "#file30"
+mkfifo "#file30"
+rm -f "#file41"
+mkfifo "#file41"
+rm -f "#file36"
+mkfifo "#file36"
+rm -f "#file32"
+mkfifo "#file32"
+rm -f "#file18"
+mkfifo "#file18"
+rm -f "#file29"
+mkfifo "#file29"
+rm -f "#file23"
+mkfifo "#file23"
+rm -f "#file42"
+mkfifo "#file42"
+rm -f "#file46"
+mkfifo "#file46"
+rm -f "#file31"
+mkfifo "#file31"
+rm -f "#file37"
+mkfifo "#file37"
+rm -f "#file20"
+mkfifo "#file20"
+rm -f "#file21"
+mkfifo "#file21"
+rm -f "#file43"
+mkfifo "#file43"
+rm -f "#file25"
+mkfifo "#file25"
+rm -f "#file35"
+mkfifo "#file35"
+rm -f "#file33"
+mkfifo "#file33"
+rm -f "#file39"
+mkfifo "#file39"
+rm -f "#file34"
+mkfifo "#file34"
+rm -f "#file26"
+mkfifo "#file26"
+rm -f "#file24"
+mkfifo "#file24"
+rm -f "#file44"
+mkfifo "#file44"
+rm -f "#file38"
+mkfifo "#file38"
+rm -f "#file45"
+mkfifo "#file45"
+rm -f "#file17"
+mkfifo "#file17"
+rm -f "#file27"
+mkfifo "#file27"
+rm -f "#file28"
+mkfifo "#file28"
+rm -f "#file22"
+mkfifo "#file22"
+cat $IN > "#file17" &
+cat $IN > "#file18" &
+cat $IN > "#file19" &
+cat $IN > "#file20" &
+cat $IN > "#file21" &
+cat $IN > "#file22" &
+cat $IN > "#file23" &
+cat $IN > "#file24" &
+cat $IN > "#file25" &
+cat $IN > "#file26" &
+cat "#file17" | tr A-Z a-z > "#file27" &
+cat "#file18" | tr A-Z a-z > "#file28" &
+cat "#file19" | tr A-Z a-z > "#file29" &
[…]

### 2019-11-21 evaluation/minimal_grep_10_seq.sh

@@ -1,0 +1,1 @@
+cat $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN | tr A-Z a-z | grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4'

### 2019-11-21 evaluation/minimal_grep_200_distr.sh

@@ -1,0 +1,2601 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file407"
+mkfifo "#file407"
+rm -f "#file327"
+mkfifo "#file327"
+rm -f "#file252"
+mkfifo "#file252"
+rm -f "#file515"
+mkfifo "#file515"
+rm -f "#file791"
+mkfifo "#file791"
+rm -f "#file207"
+mkfifo "#file207"
+rm -f "#file734"
+mkfifo "#file734"
+rm -f "#file242"
+mkfifo "#file242"
+rm -f "#file241"
+mkfifo "#file241"
+rm -f "#file719"
+mkfifo "#file719"
+rm -f "#file796"
+mkfifo "#file796"
+rm -f "#file706"
+mkfifo "#file706"
+rm -f "#file697"
+mkfifo "#file697"
+rm -f "#file696"
+mkfifo "#file696"
+rm -f "#file448"
+mkfifo "#file448"
+rm -f "#file672"
+mkfifo "#file672"
+rm -f "#file506"
+mkfifo "#file506"
+rm -f "#file351"
+mkfifo "#file351"
+rm -f "#file622"
+mkfifo "#file622"
+rm -f "#file739"
+mkfifo "#file739"
+rm -f "#file332"
+mkfifo "#file332"
+rm -f "#file729"
+mkfifo "#file729"
+rm -f "#file338"
+mkfifo "#file338"
+rm -f "#file742"
+mkfifo "#file742"
+rm -f "#file789"
+mkfifo "#file789"
+rm -f "#file382"
+mkfifo "#file382"
+rm -f "#file649"
+mkfifo "#file649"
+rm -f "#file250"
+mkfifo "#file250"
+rm -f "#file278"
+mkfifo "#file278"
+rm -f "#file477"
+mkfifo "#file477"
+rm -f "#file219"
+mkfifo "#file219"
+rm -f "#file512"
+mkfifo "#file512"
+rm -f "#file347"
+mkfifo "#file347"
+rm -f "#file574"
+mkfifo "#file574"
+rm -f "#file750"
+mkfifo "#file750"
+rm -f "#file235"
+mkfifo "#file235"
+rm -f "#file430"
+mkfifo "#file430"
+rm -f "#file240"
+mkfifo "#file240"
+rm -f "#file555"
+mkfifo "#file555"
+rm -f "#file580"
+mkfifo "#file580"
+rm -f "#file401"
[…]

### 2019-11-21 evaluation/minimal_grep_200_seq.sh

@@ -1,0 +1,1 @@
+cat $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN | tr A-Z a-z | grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4'

### 2019-11-21 evaluation/minimal_grep_20_distr.sh

@@ -1,0 +1,261 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file59"
+mkfifo "#file59"
+rm -f "#file63"
+mkfifo "#file63"
+rm -f "#file53"
+mkfifo "#file53"
+rm -f "#file76"
+mkfifo "#file76"
+rm -f "#file35"
+mkfifo "#file35"
+rm -f "#file47"
+mkfifo "#file47"
+rm -f "#file30"
+mkfifo "#file30"
+rm -f "#file66"
+mkfifo "#file66"
+rm -f "#file75"
+mkfifo "#file75"
+rm -f "#file51"
+mkfifo "#file51"
+rm -f "#file54"
+mkfifo "#file54"
+rm -f "#file82"
+mkfifo "#file82"
+rm -f "#file84"
+mkfifo "#file84"
+rm -f "#file36"
+mkfifo "#file36"
+rm -f "#file39"
+mkfifo "#file39"
+rm -f "#file61"
+mkfifo "#file61"
+rm -f "#file46"
+mkfifo "#file46"
+rm -f "#file32"
+mkfifo "#file32"
+rm -f "#file27"
+mkfifo "#file27"
+rm -f "#file50"
+mkfifo "#file50"
+rm -f "#file69"
+mkfifo "#file69"
+rm -f "#file64"
+mkfifo "#file64"
+rm -f "#file29"
+mkfifo "#file29"
+rm -f "#file31"
+mkfifo "#file31"
+rm -f "#file38"
+mkfifo "#file38"
+rm -f "#file73"
+mkfifo "#file73"
+rm -f "#file28"
+mkfifo "#file28"
+rm -f "#file85"
+mkfifo "#file85"
+rm -f "#file40"
+mkfifo "#file40"
+rm -f "#file34"
+mkfifo "#file34"
+rm -f "#file65"
+mkfifo "#file65"
+rm -f "#file43"
+mkfifo "#file43"
+rm -f "#file48"
+mkfifo "#file48"
+rm -f "#file49"
+mkfifo "#file49"
+rm -f "#file77"
+mkfifo "#file77"
+rm -f "#file79"
+mkfifo "#file79"
+rm -f "#file74"
+mkfifo "#file74"
+rm -f "#file41"
+mkfifo "#file41"
+rm -f "#file58"
+mkfifo "#file58"
+rm -f "#file70"
+mkfifo "#file70"
+rm -f "#file56"
+mkfifo "#file56"
[…]

### 2019-11-21 evaluation/minimal_grep_2_distr.sh

@@ -1,0 +1,27 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file14"
+mkfifo "#file14"
+rm -f "#file13"
+mkfifo "#file13"
+rm -f "#file11"
+mkfifo "#file11"
+rm -f "#file12"
+mkfifo "#file12"
+rm -f "#file9"
+mkfifo "#file9"
+rm -f "#file10"
+mkfifo "#file10"
+cat $IN > "#file9" &
+cat $IN > "#file10" &
+cat "#file9" | tr A-Z a-z > "#file11" &
+cat "#file10" | tr A-Z a-z > "#file12" &
+cat "#file11" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file13" &
+cat "#file12" | grep "\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4" > "#file14" &
+cat "#file13" > /tmp/dish_output//0
+cat "#file14" > /tmp/dish_output//1
+rm -f "#file14"
+rm -f "#file13"
+rm -f "#file11"
+rm -f "#file12"
+rm -f "#file9"
+rm -f "#file10"

### 2019-11-21 evaluation/minimal_grep_50_distr.sh

@@ -1,0 +1,651 @@
+mkdir -p /tmp/dish_output/
+rm -f "#file174"
+mkfifo "#file174"
+rm -f "#file95"
+mkfifo "#file95"
+rm -f "#file189"
+mkfifo "#file189"
+rm -f "#file104"
+mkfifo "#file104"
+rm -f "#file164"
+mkfifo "#file164"
+rm -f "#file78"
+mkfifo "#file78"
+rm -f "#file75"
+mkfifo "#file75"
+rm -f "#file151"
+mkfifo "#file151"
+rm -f "#file60"
+mkfifo "#file60"
+rm -f "#file177"
+mkfifo "#file177"
+rm -f "#file132"
+mkfifo "#file132"
+rm -f "#file100"
+mkfifo "#file100"
+rm -f "#file146"
+mkfifo "#file146"
+rm -f "#file172"
+mkfifo "#file172"
+rm -f "#file134"
+mkfifo "#file134"
+rm -f "#file195"
+mkfifo "#file195"
+rm -f "#file126"
+mkfifo "#file126"
+rm -f "#file119"
+mkfifo "#file119"
+rm -f "#file82"
+mkfifo "#file82"
+rm -f "#file58"
+mkfifo "#file58"
+rm -f "#file178"
+mkfifo "#file178"
+rm -f "#file190"
+mkfifo "#file190"
+rm -f "#file192"
+mkfifo "#file192"
+rm -f "#file162"
+mkfifo "#file162"
+rm -f "#file202"
+mkfifo "#file202"
+rm -f "#file166"
+mkfifo "#file166"
+rm -f "#file69"
+mkfifo "#file69"
+rm -f "#file187"
+mkfifo "#file187"
+rm -f "#file99"
+mkfifo "#file99"
+rm -f "#file89"
+mkfifo "#file89"
+rm -f "#file197"
+mkfifo "#file197"
+rm -f "#file193"
+mkfifo "#file193"
+rm -f "#file199"
+mkfifo "#file199"
+rm -f "#file115"
+mkfifo "#file115"
+rm -f "#file147"
+mkfifo "#file147"
+rm -f "#file116"
+mkfifo "#file116"
+rm -f "#file80"
+mkfifo "#file80"
+rm -f "#file143"
+mkfifo "#file143"
+rm -f "#file191"
+mkfifo "#file191"
+rm -f "#file144"
+mkfifo "#file144"
[…]

### 2019-11-21 evaluation/minimal_sort_100_distr.sh

@@ -1,606 +1,606 @@
 mkdir -p /tmp/dish_output/
-rm -f "#file134"
-mkfifo "#file134"
-rm -f "#file175"
-mkfifo "#file175"
-rm -f "#file306"
-mkfifo "#file306"
-rm -f "#file205"
-mkfifo "#file205"
-rm -f "#file269"
-mkfifo "#file269"
-rm -f "#file300"
-mkfifo "#file300"
-rm -f "#file391"
-mkfifo "#file391"
-rm -f "#file147"
-mkfifo "#file147"
-rm -f "#file397"
-mkfifo "#file397"
-rm -f "#file385"
-mkfifo "#file385"
-rm -f "#file151"
-mkfifo "#file151"
-rm -f "#file160"
-mkfifo "#file160"
-rm -f "#file162"
-mkfifo "#file162"
-rm -f "#file188"
-mkfifo "#file188"
-rm -f "#file289"
-mkfifo "#file289"
-rm -f "#file143"
-mkfifo "#file143"
-rm -f "#file390"
-mkfifo "#file390"
-rm -f "#file401"
-mkfifo "#file401"
-rm -f "#file271"
-mkfifo "#file271"
-rm -f "#file303"
-mkfifo "#file303"
-rm -f "#file287"
-mkfifo "#file287"
-rm -f "#file380"
-mkfifo "#file380"
-rm -f "#file106"
-mkfifo "#file106"
-rm -f "#file366"
-mkfifo "#file366"
-rm -f "#file200"
-mkfifo "#file200"
-rm -f "#file193"
-mkfifo "#file193"
-rm -f "#file264"
-mkfifo "#file264"
-rm -f "#file313"
-mkfifo "#file313"
-rm -f "#file217"
-mkfifo "#file217"
-rm -f "#file275"
-mkfifo "#file275"
-rm -f "#file404"
-mkfifo "#file404"
-rm -f "#file358"
-mkfifo "#file358"
-rm -f "#file362"
-mkfifo "#file362"
-rm -f "#file166"
-mkfifo "#file166"
-rm -f "#file285"
-mkfifo "#file285"
-rm -f "#file152"
-mkfifo "#file152"
-rm -f "#file310"
-mkfifo "#file310"
-rm -f "#file383"
-mkfifo "#file383"
-rm -f "#file168"
-mkfifo "#file168"
-rm -f "#file250"
-mkfifo "#file250"
-rm -f "#file181"
[…]

### 2019-11-21 evaluation/minimal_sort_20_distr.sh

@@ -1,126 +1,126 @@
 mkdir -p /tmp/dish_output/
-rm -f "#file73"
-mkfifo "#file73"
-rm -f "#file71"
-mkfifo "#file71"
-rm -f "#file48"
-mkfifo "#file48"
-rm -f "#file36"
-mkfifo "#file36"
-rm -f "#file69"
-mkfifo "#file69"
-rm -f "#file74"
-mkfifo "#file74"
-rm -f "#file77"
-mkfifo "#file77"
-rm -f "#file56"
-mkfifo "#file56"
-rm -f "#file64"
-mkfifo "#file64"
-rm -f "#file65"
-mkfifo "#file65"
-rm -f "#file43"
-mkfifo "#file43"
-rm -f "#file47"
-mkfifo "#file47"
-rm -f "#file39"
-mkfifo "#file39"
-rm -f "#file31"
-mkfifo "#file31"
-rm -f "#file34"
-mkfifo "#file34"
-rm -f "#file44"
-mkfifo "#file44"
-rm -f "#file68"
-mkfifo "#file68"
-rm -f "#file75"
-mkfifo "#file75"
-rm -f "#file27"
-mkfifo "#file27"
-rm -f "#file40"
-mkfifo "#file40"
-rm -f "#file33"
-mkfifo "#file33"
-rm -f "#file42"
-mkfifo "#file42"
-rm -f "#file55"
-mkfifo "#file55"
-rm -f "#file86"
-mkfifo "#file86"
-rm -f "#file70"
-mkfifo "#file70"
-rm -f "#file41"
-mkfifo "#file41"
-rm -f "#file45"
-mkfifo "#file45"
-rm -f "#file61"
-mkfifo "#file61"
-rm -f "#file83"
-mkfifo "#file83"
-rm -f "#file46"
-mkfifo "#file46"
-rm -f "#file78"
-mkfifo "#file78"
-rm -f "#file28"
-mkfifo "#file28"
 rm -f "#file60"
 mkfifo "#file60"
-rm -f "#file79"
-mkfifo "#file79"
-rm -f "#file57"
-mkfifo "#file57"
-rm -f "#file26"
-mkfifo "#file26"
-rm -f "#file67"
-mkfifo "#file67"
-rm -f "#file54"
-mkfifo "#file54"
-rm -f "#file38"
-mkfifo "#file38"
-rm -f "#file53"
-mkfifo "#file53"
-rm -f "#file35"
-mkfifo "#file35"
[…]

### 2019-11-21 evaluation/minimal_sort_4_distr.sh

@@ -1,30 +1,30 @@
 mkdir -p /tmp/dish_output/
-rm -f "#file13"
-mkfifo "#file13"
-rm -f "#file14"
-mkfifo "#file14"
-rm -f "#file20"
-mkfifo "#file20"
-rm -f "#file11"
-mkfifo "#file11"
-rm -f "#file12"
-mkfifo "#file12"
-rm -f "#file10"
-mkfifo "#file10"
 rm -f "#file18"
 mkfifo "#file18"
-rm -f "#file17"
-mkfifo "#file17"
-rm -f "#file22"
-mkfifo "#file22"
-rm -f "#file16"
-mkfifo "#file16"
 rm -f "#file15"
 mkfifo "#file15"
+rm -f "#file12"
+mkfifo "#file12"
+rm -f "#file14"
+mkfifo "#file14"
+rm -f "#file17"
+mkfifo "#file17"
 rm -f "#file21"
 mkfifo "#file21"
 rm -f "#file19"
 mkfifo "#file19"
+rm -f "#file11"
+mkfifo "#file11"
+rm -f "#file13"
+mkfifo "#file13"
+rm -f "#file20"
+mkfifo "#file20"
+rm -f "#file16"
+mkfifo "#file16"
+rm -f "#file22"
+mkfifo "#file22"
+rm -f "#file10"
+mkfifo "#file10"
 cat $IN > "#file11" &
 cat $IN > "#file12" &
 cat $IN > "#file13" &
@@ -39,16 +39,16 @@ cat "#file17" | sort  > "#file21" &
 cat "#file18" | sort  > "#file22" &
 sort -m "#file19" "#file20" "#file21" "#file22" > "#file10" &
 cat "#file10" > /tmp/dish_output//0
-rm -f "#file13"
-rm -f "#file14"
-rm -f "#file20"
-rm -f "#file11"
-rm -f "#file12"
-rm -f "#file10"
 rm -f "#file18"
-rm -f "#file17"
-rm -f "#file22"
-rm -f "#file16"
 rm -f "#file15"
+rm -f "#file12"
+rm -f "#file14"
+rm -f "#file17"
 rm -f "#file21"
 rm -f "#file19"
+rm -f "#file11"
+rm -f "#file13"
+rm -f "#file20"
+rm -f "#file16"
+rm -f "#file22"
+rm -f "#file10"

### 2019-11-21 pldi/draft.tex
Remove some obsolete comments
Checkpoint related work to pull

--- 2019-11-20 pldi/draft.tex
@@
 \newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
-\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
+\newcommand{\kk}[1]{[{\color{magenta}#1 --- KK}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 
-\begin{figure*}[t]
+% \begin{figure*}[t]
+% \centering
+% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+% \caption{
+%   \textbf{Applying \sys to $p_1$.}
+% }
+% \label{fig:example}
+% \end{figure*}
+%
+
+\section{Background}
+\label{bg}
+
+This section breaks down $p_1$ to presents important background on and challenges
+of shell scripting.
+
+\subsection{Unix Composition, Informally}
+\label{bg:pipelines}
+
+% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
+\unix provides several ways for succinct program composition.
+Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
+These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
+Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
+\begin{wrapfigure}{r}{0.06\textwidth}
+  \vspace{-12pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
+  % \end{center}
+  \vspace{-30pt}
+\end{wrapfigure}
+This interface exposes a stream of contiguous character lines.
+A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
+As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
+% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
+% generating errors or diagnostics to the standard 
+% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
+% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}
+
+Commands can be tuned by two mechanisms, namely command-line options and environment variables.
+Options control the execution of a program---\eg changing the \ttt{sort}ing order.
+The shell does not have any visibility into these options; 
+  their parsing and evaluation is left entirely up to each command.
+Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
+They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
+
+% TODO: Talk about security primitives?
+Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
+  files are simply persistent streams that can be manipulated by commands or pipelines.
+\begin{wrapfigure}{r}{0.15\textwidth}
+  \vspace{-15pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
+  % \end{center}
+  \vspace{-25pt}
+\end{wrapfigure}
+Reading a file or a set of files generates a stream; 
+  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
+Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
+\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
+As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
+
+% At a high level, POSIX defines several file types:
+% (-) regular files, storing lines of text or binary data,
+% (d) directories, grouping together multiple other files,
+% (l) symbolic links, pointing to other files,
+% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
+% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
+% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
+
+Special files are particularly interesting because they are resolved to OS structures.
+One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
+  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
+% Here are a few examples:
+% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
+% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
+% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
+Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
+This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
+
+A small but powerful set of features is concerned with stream manipulation.
+Some of these features are built into the shell;
+  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
+\begin{wrapfigure}{r}{0.15\textwidth}
+  \vspace{-15pt}
+  % \begin{center}
+    % \includegraphics[width=0.48\textwidth]{gull}
+    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
+  % \end{center}
+  \vspace{-25pt}
+\end{wrapfigure}
+As described before, the file can be a FIFO dubbing as input to another command.
+More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
+  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
+% TODO: talk about \ttt{xargs}?
+
+% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators
+
+The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
+The most interesting feature of these pipelines is the presence of cycles:
+  a pipeline segment processes some input, part of which is extracting the next input to process.
+One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
+% These streams are created by the C library (glibc) at the start of program execution,
+% and new streams can be created to connect to files, sockets, pipes,
+
+Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
+Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
+Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
+  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
+Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
+
+% \begin{lstlisting}[
+%   language=es,
+%   mathescape,
+%   float=t,
+%   belowskip=-5mm,
+%   label=core-pl,
+%   upquote=true,
+%   caption={
+%     \textbf{The core language of the POSIX shell.}
+%     The language used by ; file identifiers and expansion are not shown.
+%   }]
+% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
+% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
+% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
+% \end{lstlisting}
+
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat /proc/ >                            (($$p_4$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  ls -l | wc -l                           (($$p_2$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat * | wc -l > out.txt                 (($$p_3$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%  cat /proc/ >                            (($$p_5$$))
+% \end{lstlisting}
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+% { echo 1; echo 2; } > out.txt
+% \end{lstlisting}
+
+\begin{figure}[t]
 \centering
@@
 \caption{
-  \textbf{Applying \sys to $p_1$.}
+  \textbf{\sys transformation overview.}
+	\sys XYZ
 }
-\label{fig:example}
-\end{figure*}
-
-\section{Background}
-\label{bg}
-
-This section breaks down $p_1$ to presents important background on and challenges
-of shell scripting.
-
-\subsection{Unix Pipelines, Informally}
-\label{bg:pipelines}
-
-Pipelines are a mechanism for program composition.
-Individual programs, specified as commands, conform to a common but general interface. 
-For now, this interface allows a combination of the following:
-  accepting input from the standard input stream (stdin),
-  generating output to the standard output stream (stdout), and
-  generating errors or diagnostics to the standard error stream (stderr)
-Standard streams are streams of characters with a few special characters---for example, EOF is a special end-of-file character signaling the end of the input stream.
-% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.
-
-Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
-  programs do one thing well, they work together, and they manipulate text streams as their universal interface.
-The next few paragraphs describe the details.
-
-\heading{Basics}
-The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).
-
-that are not interpreted by the shell, but are local to each command.
-
-Additional details augment this simple interface.
-Commands can be specified with options arguments
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
-\label{fig:example}
+\label{fig:overview}
 \end{figure}
 
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- ls -l | wc -l                           (($$p_2$$))
-\end{lstlisting}
-
-\heading{Generation}
-A central concept in pipelines---and Unix, in general---is the file system, which allows interfacing with persistent data.
-There are many commands that take as input a file identifier 
-Reading a file or a set of files generates a stream; 
-  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-
-Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
-The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat * | wc -l > out.txt                 (($$p_3$$))
-\end{lstlisting}
-
-\heading{Identifiers}
-
-\kk{We have to mention character vs line based parallelism somewhere,
-  and that we choose to see lines as the data quantum}
-
-The file-system is a common abstraction layer used for many other operations than just serving files (this will be become clearer when talking about file-system commands in \S\ref{bg:cmd}).
-as such, different identifiers many not resolve to actual files, but rather in-memory or OS-internal structures.
-
-At a high level, POSIX defines several file types:
-(-) regular files, storing lines of text or binary data,
-(d) directories, grouping together multiple other files,
-(l) symbolic links, pointing to other files,
-(p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
-(s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
-(b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-
-Most of these identifiers are manipulated by file-system commands---\eg \ttt{mkdir} for directories and \ttt{mkfifo} for FIFOs.
-FIFOs and DSes look like files but are streams.
-
-Special files are particularly interesting because they are resolved to OS structures.
-One set is pseudo-devices---device files that simply;
-For example /dev/urandom returns a stream of random characters and /dev/null accepts and discards all input written to it.
-Another set is proc filesystem (procfs), a special filesystem in Unix-like operating systems that presents information about processes and other system information.
-For example, /proc/PID/cmdline contains command that originally started the process and /proc/PID/cwd contains a symlink to the current working directory of the process.
-
-sysfs is a pseudo file system provided by the Linux kernel that exports information about various kernel subsystems, hardware devices, and
-
-/modules, one of the most important files in /proc, containing a list of the kernel modules currently loaded . It gives some indication (not always entirely correct) of dependencies.
-/cpuinfo
-
-symbolic links, FIFO specials, 
-device files
-block special, character special, 
-and domain sockets.
-
-There are 3 standard streams (stdin, stdout, stderr) created automatically for most Unix programs.
-These correspond to \ttt{/dev/fd/0}.
-% These streams are created by the C library (glibc) at the start of program execution,
-and new streams can be created to connect to files, sockets, pipes,
-
-Unix already specifies a few classes of files:
-
-Here are a few special files:
-  \ttt{proc} called
-
-Technically, anonymous pipes show up as file identifiers in a hidden filesystem.
-Similarly, open files and other ephemeral identifiers
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                            (($$p_4$$))
-\end{lstlisting}
-
-\heading{Stream Manipulation}
-A small but powerful set of primitives are concerned with stream manipulation.
-First, as described above, several commands can be redirected
-Second,
-
-\ttt{xargs}
-\ttt{tee}
-
-In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?
-
-Explicit Redirection
-Implicit splitting and merging 
-This can be done through the file-system operators
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat /proc/ >                            (($$p_5$$))
-\end{lstlisting}
-
-\heading{Composition}
-anonymous pipes
-
-Composition
-
-The shell provides several ways of grouping commands and manipulating their
-overall input and output channels
-
-{ echo 1; echo 2; } > out.txt
+% \heading{Generation}
+% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
+% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
+
+\subsection{Overview}
+
+As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
+Intuitively,
+ (i) the parallel-composition operator exposes task parallelism,
+ (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+ (iii) the pipe operator enables both task and data parallelism, and
+ (iv)  other composition operators can be viewed as barrier candidates.
+To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.
+
+Understanding the shell's primitives is not enough;
+  \sys has to understand the distribution characteristics of individual commands used to compose larger programs.
+This challenge can be broken down into two parts.
+First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
+One example is the set of commands defined by the \textsc{POSIX} standard~\cite{}.
+\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
+This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.
+
+The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
+Commands are arbitrary programs, but still conform to the shell's pipeline interface.
+To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
+Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
+A command developer can only annotate the default behavior;
+  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
+
+Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
+\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying maximal distributability sub-expressions.
+It converts the AST to a dataflow graph, and runs...
+This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc
+
+Finally, given a distributed pipeline with runtime \todo{annotations}, \sys needs to schedule and execute it.
+This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
+First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
+  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
+The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
+
+Second, \sys needs to fill-in planing details left blank by the rewriting pass.
+This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
+From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.
+
+The next few sections discuss the details.
+They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
+
+\section{Distributability Classes}
 
@@
 
-\section{Intermediate Representation/Dataflow graph}
-
-\kk{Try to be more formal in the section, by using the notation that
-  is defined in the nodes paragraph.}
-
-In order to effectively apply optimizations and distribute a shell
-script, it first has to be translated to a manipulable
-representation. As mentioned in \ref{}, the two main abstractions of
-the shell, are files that contain data, and commands that communicate
-through these files. We propose a slight variation of the standard
-dataflow graph, that is commonly found in stream processing systems,
-as the representation of a distributed shell computation. The main
-benefit of the dataflow graph model is that it clearly exposes
-operator level parallelism, as different nodes can be independent
-processing workers. In our case each node represents a command, and
-each edge represents a file.
-
-\kk{Before that I have to mention that the basic data type $D$ is a
-  line of characters. If we move the edges section before the nodes
-  that will be fine.}
-
-\paragraph{Nodes - Commands}
-
-Nodes $f$ of the graph represent functions from one (possibly empty)
-input stream to an output stream $f : D^* \rightarrow D^*$.  This
-representation captures the majority of shell commands. The nodes $f$
-are monotonic, meaning that they cannot retract their output. More
-formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
-that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
-omitted from now on. An example of a node is the following command
-\kk{Does it need ref? Also maybe this example is too trivial and can
-  be removed.}:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- grep ``foo''
-\end{lstlisting}
-
-\noindent
-This command takes a stream of lines from its standard input and
-returns a stream of lines on its standard output. An important
-observation though is that commands in the shell might read their
-input stream from a sequence of different sources. An illustrative
-example is the command \kk{Inline listing}cat which reads its input
-from a combination of files and its standard input.
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat x - y
-\end{lstlisting}
-
-\kk{Should I reference that this example was takes from the cat
-  manpage?} The example above reads file \texttt{x} until it
-encounters the end of file, then reads from \texttt{stdin}, and then
-reads from \texttt{y}. In order to handle this particularity, nodes in
-the dataflow graph can have many incoming edges, that are clearly
-ordered in a sequence. Note that this comes in contrast with standard
-dataflow graph representations where multiple incoming edges either
-mean zipping the input to pairs, or arbitraliy interleaving the
-different inputs.
-
-\kk{A note about nodes and commands that can write and read from more
-  files that the ones in their streams}
-
-Note that nodes can read from more files than just the files of their
-stream (or respectively write to more files that their output stream),
-however these files are static (e.g. used for configuration or logging
-\kk{Would an example here help}) and do not represent streams of data,
-and therefore are not considered part of the dataflow graph. Because
-of the assumption in section Front End \ref{}, that each file exists
-only once in the dataflow graph, we can safely assume that reading and
-writing to these static files does not interfere with the distributed
-implementation, and doesn't alter the behaviour of the program.
-
-\paragraph{Edges - Files}
-
-\kk{Maybe edges has to be moved before nodes, so that we can define
-  the line data type}
-
-Edges in the dataflow graph represent files, the basic data
-abstraction of the shell. They are used as communication channels
-between nodes in the graph, and as the input or output of the whole
-graph. Edges are represented as possibly unbounded streams of type
-$D*$, where $D$ represents a line, that is a sequence of characters
-followed by \verb|\n| \kk{or EOF?}. They could either refer to a
-specific named file in the file system, or just be FIFO pipes that
-used for interprocess communication \kk{Are there any other file
-  types, such as URLs, ...?}.
-
-\kk{Show an example pipe and its graph}.
-
-In order to expose operator parallelism with the output of a command,
-files that don't point to a resource can have an upper bound on the
-lines that they can tranfer. The following example illustrates why
-this is essential.
-
-\kk{An example of a command that has two outputs and one goes to one
-  command, while another output goes to a different one. Maybe only
-  show the graph here and not the command itself.}
-
-On the other hand, some forms of data parallelism can be exposed when
-knowing the size of the input files. As mentioned in \ref{} some pure
-commands (such as cat -n) only need line information to become
-stateless, and knowing the size of a file could allow the system to
-split it in different chunks that can be processed independently. To
-account for that, edges that refer to an input resource contain the
-number of lines of the file that they refer to. \kk{At the moment this
-  is not supported by our real system. The whole paragraph above can
-  go if we decide that it is not a good idea to talk about it now.}
-
-Finally, the edges that don't refer to a resource and do not start
-from a node in the graph represent the graph input, while the edges
-that don't refer to a resource and do not point to a node in the graph
-are its outputs.
-
-\kk{(Maybe) Depending on which of the two paragraphs goes last
-  (nodes/edges), I have to explain there that the last incoming edge
-  to a node must be unbounded (if it doesn't point to a resource). I
-  am actually not sure whether I should include this here.}
-
-\kk{(Maybe) It might be beneficial to mention that many inputs do not
-  represent the same thing as many outputs. These two notions are not
-  symmetric.}
-
-\paragraph{Merge and Split}
-
-\kk{Mergers and splitters, and how they are implemented using shell
-  constructs (cat and head). This subsection might be a better fit in
-  some other section, maybe something related to the planner or the
-  implementation, as it is a very elegant way to implement splitters
-  and mergers while staying completely inside the shell programming
-  model.}
-
-It is common in the dataflow graph for a node to have many incoming
-edges, or many outgoing edges (when all but the last outgoing edges
-are bounded). However some nodes, might read their input from one file
-(e.g. stdin). In this case, the distributed implementation needs to
-concatenate input files, and split output files. Here is an example:
-
-\kk{Show an example graph! of cat | tr | sort. The command in the
-  middle must only take input from one file (maybe just stdin).}
-
-\kk{This transition needs to be stronger} Luckily, the shell offers
-primitives for file manipulations that can be used to implement a
-merger and a splitter. More concretely a merger can be implemented by
-a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
-and an output file \verb|out|, a merger can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
- cat $in1 $in2 ... > $out
-\end{lstlisting}
-
-\noindent
-On the other hand, given an input file \verb|in| that is a FIFO pipe,
-and two output files \verb|out1, out2| the first of which is bounded
-to \verb|N| lines, a 2-splitter can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
- head -n $N $in > $out1 ; cat $in > $out2
-\end{lstlisting}
-
-\noindent
-Using the 2-splitter as a building block, a pipe can be split to
-arbitrarily many different pipes.
-
-\kk{Do I also need to include a splitter for normal files?}
-
-\kk{Intermediate Representation Section notes:}
-
-\kk{Explain how the assumptions made in the front end give as
-  correct/safe operator level parallelism. Maybe this must go in the
-  Front-end section.}
-
 \section{Front End}
-
-\kk{Maybe this becomes a subsection either in the intermediate
-  representation or in the planner.}
 
@@
 \kk{Text point: Problem: Not all parts of a shell script are
-  distributable, so we cannot arbitrarily distribute. Here we have to
-  make our assumptions clear}
+  distributable, so we cannot arbitrarily distribute}
 
@@
 
-\kk{The same file exists once in the substree. This is a conservative
-  assumption that ensures that no two commands write or read to the
-  same file concurrently.}
-
-\kk{How can we make sure that nodes in the IR are deterministic. Do we
-  make some assumption that enforces this?}
+\subsection{Intermediate Representation/Dataflow graph}
+
+Mention its constructs, and its semantics, and why it is a bit
+different that standard dataflow graphs (where usually incoming edges
+are either consumed with some arbitrary interleaving, or pair by
+pair).
+
+\kk{Text point: The main shell abstraction is files, which are batches
+  or streams of characters (or lines) where the order matters. So we
+  want a representation that captures order of inputs and outputs (and
+  ours does because its node in the graph has a sequence of inputs.}
+
+Explain that the intermediate representation represents workers that
+communicate through files, and that it captures parallelism both in
+the operator level, as well as data parallelism.
 
@@
 
-\kk{Mention that different nodes can be safely executed independently,
-  and that this exposes operator level parallelism}
-
-\subsection{Scaling up the distributed computation/Vertically expanding the graph/Graph Transformations/ Equations}
+\kk{Maybe this should be merged with the one above, especially if it
+  talks about how distribution happens on the intermediate
+  representation.}
+
+\subsection{Scaling up the distributed computation/Vertically expanding the graph}
 
@@
 \kk{Give an example}
-
-\kk{Can we have a semi-formal argument here?}
-
-\kk{Explain how our graph transformations expose data parallelism as
-  operator level parallelism.}
-
-\kk{Different transformations to explain: Stateless, Pure, File
-  splits, XARGS. Mention as much as possible formally using the
-  definition of a node.}
-
-The following equation holds for any stateless command $f$.
-
-\[
-f(x.x') = f(x).f(x')
-\]
-
-This is the basis of all optimizations that the graph does, and allows
-for correct data parallelism.
 
@@
   replication, consistency, and failures.
-\item No cycles
 \end{itemize}
@@
 \section{Related Work}
+\label{related}
+
+At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
+\caption{
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
+}
+\label{fig:spectrum}
+\end{figure}
+
+% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
+\heading{Automated Parallelization and Distribution}
+There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
+
+% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
+% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
+% While they Moreover, these systems take advantage of functional purity;
+%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
+A plethora of systems assist in the construction of distributed software.
+Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
+% simplify many of the problems of distribution and
+provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
+However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
+\TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
+
+\heading{Annotation-based Distribution}
+Ignis and Mozart
+
+\heading{DSLs for Distribution}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+simplify certain patterns, 
+but do not offer the flexibility of a full-fledged environment.
+% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
+Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
+
+More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
+These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
+%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
+Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
+%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
+
+\section{Conclusion}
+\label{discussion}
 

### 2019-11-21 pldi/draft.tex

@@
 
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required for distributing a computation.
-\sys builds on this insight with a careful study of the scalability properties of shell primitives and its commands, paired with a novel rewriting scheme that identifies maximal pipeline sub-expressions that are candidates for scale-out.
-We term the result \emph{distribution-oblivious programming}---a series of techniques that fully automates program distribution while provably maintaining correct (non-distributed) semantics.
-
-Key benefits include the ability to scale legacy computations onto
-\sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-Most importantly, \sys provides an architectural lesson:
-  the current design shows that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
+\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
+
+The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
+% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
+First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
+Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+% Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
 

### 2019-11-21 pldi/draft.tex

@@
 
-Can we find a solution for the commands in coreutils?
+\kk{Can we find a solution for the commands in coreutils?}
+
+\kk{Important TODO: Find at least one more command that can be
+  executed like sort. Does wc work? If there is no pure command that
+  can be easily parallelized other than sort, this story is not very
+  nice.}
+
+\section{Intermediate Representation/Dataflow graph}
+\label{ir}
+
+\kk{Inline listing?}
+
+\kk{Try to be more formal in the section, by using the notation that
+  is defined in the nodes paragraph.}
+
+In order to effectively apply optimizations and distribute a shell
+script, it first has to be translated to a manipulable
+representation. As mentioned in \ref{}, the two main abstractions of
+the shell, are files that contain data, and commands that communicate
+through these files. We propose a slight variation of the standard
+dataflow graph, that is commonly found in stream processing systems,
+as the representation of a distributed shell computation. The main
+benefit of the dataflow graph model is that it clearly exposes
+operator level parallelism \kk{Is this a common term? Is there any
+  definitive citation for it?}, as different nodes can be independent
+processing workers. In our case each node represents a command, and
+each edge represents a file.
+
+\kk{Before that I have to mention that the basic data type $D$ is a
+  line of characters. If we move the edges section before the nodes
+  that will be fine.}
+
+\paragraph{Nodes - Commands}
+
+Nodes $f$ of the graph represent functions from one (possibly empty)
+input stream to an output stream $f : D^* \rightarrow D^*$.  This
+representation captures the majority of shell commands. The nodes $f$
+are monotonic, meaning that they cannot retract their output. More
+formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
+that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
+omitted from now on. An example of a node is the following command
+\kk{Does it need ref? Also maybe this example is too trivial and can
+  be removed.}:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ grep ``foo''
+\end{lstlisting}
+
+\noindent
+This command takes a stream of lines from its standard input and
+returns a stream of lines on its standard output. An important
+observation though is that commands in the shell might read their
+input stream from a sequence of different sources. An illustrative
+example is the command \kk{Inline listing}cat which reads its input
+from a combination of files and its standard input.
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat x - y
+\end{lstlisting}
+
+\kk{Should I reference that this example was takes from the cat
+  manpage?} The example above reads file \texttt{x} until it
+encounters the end of file, then reads from \texttt{stdin}, and then
+reads from \texttt{y}. In order to handle this particularity, nodes in
+the dataflow graph can have many incoming edges, that are clearly
+ordered in a sequence. Note that this comes in contrast with standard
+dataflow graph representations where multiple incoming edges either
+mean zipping the input to pairs, or arbitraliy interleaving the
+different inputs.
+
+\kk{A note about nodes and commands that can write and read from more
+  files that the ones in their streams}
+
+Note that nodes can read from more files than just the files of their
+stream (or respectively write to more files that their output stream),
+however these files are static (e.g. used for configuration or logging
+\kk{Would an example here help}) and do not represent streams of data,
+and therefore are not considered part of the dataflow graph. Because
+of the assumption in section Front End \ref{}, that each file exists
+only once in the dataflow graph, we can safely assume that reading and
+writing to these static files does not interfere with the distributed
+implementation, and doesn't alter the behaviour of the program.
+
+\paragraph{Edges - Files}
+
+\kk{Maybe edges has to be moved before nodes, so that we can define
+  the line data type}
+
+Edges in the dataflow graph represent files, the basic data
+abstraction of the shell. They are used as communication channels
+between nodes in the graph, and as the input or output of the whole
+graph. Edges are represented as possibly unbounded streams of type
+$D*$, where $D$ represents a line, that is a sequence of characters
+followed by \verb|\n| \kk{or EOF?}. They could either refer to a
+specific named file in the file system, or just be FIFO pipes that
+used for interprocess communication \kk{Are there any other file
+  types, such as URLs, ...?}.
+
+\kk{Show an example pipe and its graph}.
+
+In order to expose operator parallelism with the output of a command,
+files that don't point to a resource can have an upper bound on the
+lines that they can tranfer. The following example illustrates why
+this is essential.
+
+\kk{An example of a command that has two outputs and one goes to one
+  command, while another output goes to a different one. Maybe only
+  show the graph here and not the command itself.}
+
+On the other hand, some forms of data parallelism can be exposed when
+knowing the size of the input files. As mentioned in \ref{} some pure
+commands (such as cat -n) only need line information to become
+stateless, and knowing the size of a file could allow the system to
+split it in different chunks that can be processed independently. To
+account for that, edges that refer to an input resource contain the
+number of lines of the file that they refer to. \kk{At the moment this
+  is not supported by our real system. The whole paragraph above can
+  go if we decide that it is not a good idea to talk about it now.}
+
+Finally, the edges that don't refer to a resource and do not start
+from a node in the graph represent the graph input, while the edges
+that don't refer to a resource and do not point to a node in the graph
+are its outputs.
+
+\kk{(Maybe) Depending on which of the two paragraphs goes last
+  (nodes/edges), I have to explain there that the last incoming edge
+  to a node must be unbounded (if it doesn't point to a resource). I
+  am actually not sure whether I should include this here.}
+
+\kk{(Maybe) It might be beneficial to mention that many inputs do not
+  represent the same thing as many outputs. These two notions are not
+  symmetric.}
+
+\paragraph{Merge and Split}
+
+\kk{Mergers and splitters, and how they are implemented using shell
+  constructs (cat and head). This subsection might be a better fit in
+  some other section, maybe something related to the planner or the
+  implementation, as it is a very elegant way to implement splitters
+  and mergers while staying completely inside the shell programming
+  model.}
+
+It is common in the dataflow graph for a node to have many incoming
+edges, or many outgoing edges (when all but the last outgoing edges
+are bounded). However some nodes, might read their input from one file
+(e.g. stdin). In this case, the distributed implementation needs to
+concatenate input files, and split output files. Here is an example:
+
+\kk{Show an example graph! of cat | tr | sort. The command in the
+  middle must only take input from one file (maybe just stdin).}
+
+\kk{This transition needs to be stronger} Luckily, the shell offers
+primitives for file manipulations that can be used to implement a
+merger and a splitter. More concretely a merger can be implemented by
+a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
+and an output file \verb|out|, a merger can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ cat $in1 $in2 ... > $out
+\end{lstlisting}
+
+\noindent
+On the other hand, given an input file \verb|in| that is a FIFO pipe,
+and two output files \verb|out1, out2| the first of which is bounded
+to \verb|N| lines, a 2-splitter can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ head -n $N $in > $out1 ; cat $in > $out2
+\end{lstlisting}
+
+\noindent
+Using the 2-splitter as a building block, a pipe can be split to
+arbitrarily many different pipes.
+
+\kk{Do I also need to include a splitter for normal files?}
 
@@
 
-We parse a shell script using a posix compliant parser that returns
-the ast.
-
-\kk{Text point: The AST Is not a very good data structure to
-  manipulate for distribution, so we translate it to a kind of a
-  dataflow graph.}
+\kk{It might make sense to merge this section with the one below with
+  the name: From Shell to IR and back}
+
+As mentioned in \Cref{ir}, in order to generate the distributed
+implementation of a shell script, \texttt{Dish} first translates it to
+a dataflow graph. However, there are several components in shell
+programs that cannot be arbitrarily parallelized without affecting the
+program behaviour.
+
+\kk{Give examples of writing to environment variables, producing
+  irreversible side effects, the ; and \&\& operators.}
+
+In order to address this issue, we introduce the notion of
+distributable AST subtrees, and we design an analysis that given the
+shell script Abstract Syntax Tree identifies distributable AST
+subtrees and translates each one of them to a dataflow graph. The AST
+of the shell script is produced using a POSIX compliant parser
+\kk{Reference libdash github}.
 
@@
 
-\kk{Text point: Problem: Not all parts of a shell script are
-  distributable, so we cannot arbitrarily distribute}
-
-Even though commands are split in categories, and several pipelines
-are distributable, not all parts of a shell script are stateless and
-several of them have effects that could affect the behaviour of a
-program after distribution.
-
-\kk{Solution: We have found that distributing subtrees of the shell
-  AST is safe. Can we have a semi-formal argument about that?}
-
-Mention all the different constructs that our front end handles, and
-why it is safe to add them to the distribution pipeline.
-
-\subsection{Intermediate Representation/Dataflow graph}
-
-Mention its constructs, and its semantics, and why it is a bit
-different that standard dataflow graphs (where usually incoming edges
-are either consumed with some arbitrary interleaving, or pair by
-pair).
-
-\kk{Text point: The main shell abstraction is files, which are batches
-  or streams of characters (or lines) where the order matters. So we
-  want a representation that captures order of inputs and outputs (and
-  ours does because its node in the graph has a sequence of inputs.}
-
-Explain that the intermediate representation represents workers that
-communicate through files, and that it captures parallelism both in
-the operator level, as well as data parallelism.
+\kk{The goal of this section is to convince the reader ``intuitively''
+  that we choose safe parts of a shell program to distribute. There is
+  no formal argument, but one should be able to intuitively believe
+  us.}
+
+Consider the following example:
+
+\kk{Give an example of a shell script that cannot be distributed
+  completely, Possibly involving ; and writing to a variable.}
+
+\kk{Intuitively explain what could be distributed and what not.}
+
+\kk{Our notion of distributable subtrees has two dimensions. First,
+  the state of the shell should be an invariant in a distributable
+  subtree. We cannot know this completely without evaluating, so we do
+  a conservative analysis. The second dimension is that there exists
+  at least one component that can be abstracted as a node of the
+  dataflow graph. This can be anything that includes a command.}
+
+%% Even though commands are split in categories, and several pipelines
+%% are distributable, not all parts of a shell script are stateless and
+%% several of them have effects that could affect the behaviour of a
+%% program after distribution.
+
+\kk{All of our assumptions and study are based on the POSIX
+  standard. Reference?}
+
+\kk{Not checked Assumption. A command doesn't write/read to arbitrary
+  files that are not mentioned in its arguments. This is necessary to
+  enforce that no two commands concurrently read/write from the same
+  file.}
+
+\kk{Not checked Assumption: No signals, traps, and sets at the
+  moment. At the moment we don't have a way to handle signals and
+  traps.}
+
+\kk{Partially checked Assumption. Only one reader and one writer to
+  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
+  file exists once in the substree. This is a conservative assumption
+  that ensures that no two commands write or read to the same file
+  concurrently. We kind of check it but not completely. We check that
+  the expanded evaluated files are not the same, but one could have
+  part of a file in a variable etc.}
+
+\kk{Checked assumption: How can we make sure that nodes in the IR are
+  deterministic. Do we make some assumption that enforces this? I
+  guess that commands that are nodes in the IR are all deterministic?
+  That is stateless and pure commands are deterministic.}
+
+\paragraph{analysis}
+
+The analysis is done in a DFS fashion on the AST. Distributable
+subtrees grow from the bottom up, until a construct that doesn't
+``allow'' them to go through is encountered.
+
+%% Mention all the different constructs that our front end handles, and
+%% intuitively mention why it is ``safe'' to add them to the distribution
+%% pipeline.
+
+Subtrees don't cross the following:
+
+\begin{itemize}
+\item Sequence, And,  or
+\item Function definitions
+\item Command Substitutions
+\item control flow: while, for, if, case: While these constructs can
+  be parallelized and there has been research on it, it is orthogonal
+  to our work, and can be incorporated as future work.
+\end{itemize}
+
+On the other hand: A distributable subtree starts whenever there is a
+pipe, or asynchronous list. If the subcomponents of the pipe or
+asynchronous list were already graphs, then we connect them to the
+bigger graph \kk{explain how we do pipe and asynchronous list}, and if
+they were not graphs, we make each subcomponent a node in the subtree,
+event though they are not commands. These nodes do not belong to any
+category, but still have stdin and stdout, and can be executed
+concurrently (as long as they satisfy the checked assumptions above),
+since they don't introduce new behaviours, they would still be
+executed concurrently when ran in the standard shell, producing
+possibly non-deterministic results with different interleavings.
+
+\kk{Explain how we merge two graphs when seeing a pipe}
+
+Whenever we see a subtree of the AST that is a command, we check the
+files that it writes to. If any of its arguments is a command
+substitution or contains anything that is evaluated\kk{make this more
+  precise}, its category is marked as sequential\kk{name?}. Otherwise
+we check its its category and its input output arguments (using the
+methods described in \Cref{ext}). The command is then initialized as
+the node of a singular graph, with the inputs and outputs of the graph
+being the ones of the node.
+
+\kk{Maybe mention the recursive case. I am not sure if that should be
+  here, or in the distribution planner. In any case, it should be
+  after we say that an IR is replaced with a call to the planner with
+  the IR saved in a file. The example is a command in a pipeline whose
+  first argument is a command substitution with a pipeline. Executing
+  the outside dataflow graph, would lead to the command to be executed
+  as part of this shell. However its first argument would have a call
+  to the planner itself, leading to another distributed implementation
+  running in parallel. The rest of the pipeline would wait for this to
+  end before it continues. Assuming that the files that both of these
+  graphs use are different, and that the above assumptions hold, there
+  should be no interference between the two graphs and they should be
+  able to run in parallel}
 
@@
 
-\kk{Maybe this should be merged with the one above, especially if it
-  talks about how distribution happens on the intermediate
-  representation.}
-
-\subsection{Scaling up the distributed computation/Vertically expanding the graph}
-
-The planner acts on the dataflow graph, by first expanding it ``as
-much as possible'' to expose parallelization opportunities. It uses
-knowledge about the inputs of each stateless and pure command.
-
-\kk{Explain the algorithm and how it pushes parallelization through
-  the graph, while preserving correctness.}
-
-\kk{Give an example}
+\kk{I think that there is not enough content for the planner section
+  and it should probably just be called graph transformations. Then we
+  could have a small subsection at the end about how the planner
+  applies transformations iteratively starting from the source of the
+  graph and pushing them until the end.}
+
+As mentioned in \Cref{ir}, the dataflow graph exposes operator level
+parallelism. Different nodes in the graph can be executed
+concurrently, communicating through the edges. Operator level
+parallelism is already exposed by the shell, by using pipes \texttt{|}
+and the background operator \texttt{\&}. Unfortunately, simple
+operator level parallelism leaves a lot to be desired, since scaling a
+computation requires manually introducing more operators.
+
+In order to adequately utilize the available computational resources,
+a distributed implementation should exploit data parallelism. Data
+parallelism is based on the observation that some operations produce
+the same result when performed independently on subsets of the input
+data, merging the results afterwards. Data level parallelism is the
+basic optimization that is used by popular distributed processing
+systems such as MapReduce\kk{cite} and Apache Spark\kk{cite}.
+
+\subsection{Graph transformations}
+
+In our case, this is achieved by iteratively applying graph
+transformations that preserve correctness, while exposing
+parallelization opportunities. We illustrate that with a motivating
+example; consider the following program and its dataflow graph.
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ grep ``foo'' f1 f2 // Add its graph
+\end{lstlisting}
+
+\noindent
+In this example, \texttt{grep} is applied to files f1 and f2, and its
+result is sent to standard output. However, the result would be the
+same if we concatenated the results of \texttt{grep ``foo'' f1} and
+\texttt{grep ``foo'' f2}. This intuition can be generalized to all
+stateless commands, as they are all essentially ``linear'' with
+respect to concatenation and the following equation holds.
+
+\[
+\forall f, \mathrm{stateless}(f) \implies f(x.x') = f(x).f(x')
+\]
+
+\kk{Complete this story with pure, after we have written the section
+  about command categories.}
+
+\subsection{Planner Algorithm}
+
+The \texttt{Dish} planner extends the transformations that were
+described above with edge splits in order to allow for the above
+transformations to be performed. More precisely, it splits input files
+and graph edges into several chunks so that they can be processed in
+parallel by the next levels.
+
+\kk{Explain the greedy algorithm that our planner performs. It splits
+  input files, and then pushes parallelization through the graph,
+  starting from the source nodes.}
+
+\kk{Maybe give an example}
 
 \subsection{Mapping operators to nodes}
+
+\kk{I think we should actually remove this completely.}
 
@@
 evaluate strings, just expand arguments, and not do any significant
-computation.
+computation. Even better, we could be calling the shtepper just after
+having parsed the ast, and decide on maximal distributable subtrees as
+late as possible in the process. This would be awesome.
+
+\section{Evaluation}
+
+\kk{We should make sure to measure the time that it takes for the
+  whole process to run (parsing, translating, optimizing, planning)
+  together with the scripts. This will probably be negligible, but we
+  still have to mention it. }
+
+\kk{Also, we have to have one-two handcrafted examples that have more
+  than one pipeline, to show the execution time end to end, as this
+  cannot be shown with the one-pipeline examples since they only have
+  one graph, so no back and forth between shell and dish.}
+
+\kk{Furthermore it would be great to have one distributed example that
+  runs in more than one node (if we have time)}
+
+\kk{Also we have to make sure to measure with and without the final
+  cat of output file, so that we show that this is also negligible.}
+
+\kk{We have to make sure that we repeat that the results are the same
+  in all tests that we run. Corectness, blah blah ...}
+
+\kk{We should have at least some pipelines that have a bad pure
+  command in them.}
 
@@
   replication, consistency, and failures.
+\item No cycles (multiple commands writing and reading from the same file)
+\item Formal proof that the analysis indeed returns distributable
+  subtrees. Previously there was no formal semantics for shell, but
+  recent work (cite greenberg) allows us to make a formal proof that
+  the analysis is indeed sound.
 \end{itemize}

### 2019-11-21 pldi/draft.tex
Some comments in intro+overview

@@
 To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation\kk{I am not sure whether stream processing provides information for distributing a computation on its own.
+Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
+\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies \todo{maximal} sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams \kk{Isn't uniq pure?
+I am not sure how we can handle it}, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
 
@@
 The shell does not have any visibility into these options; 
-  their parsing and evaluation is left entirely up to each command.
+  their parsing and evaluation is left entirely up to each command \kk{A minor point: the shell does the expansion, and then hands it of to the command (I am saying that because evaluation might be confounded with expansion so we might want to clarify that)}.
 Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
@@
 Intuitively,
- (i) the parallel-composition operator exposes task parallelism,
+ (i) the parallel-composition operator exposes \todo{task} parallelism \kk{I have named that operator level parallelism, which one do we prefer?},
  (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator enables both task and data parallelism, and
+ (iii) the pipe operator enables both task and data parallelism, and \kk{How does it enable data parallelism?}
  (iv)  other composition operators can be viewed as barrier candidates.
@@
 Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
-\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying maximal distributability sub-expressions.
+\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying \todo{maximal} distributable sub-expressions.
 It converts the AST to a dataflow graph, and runs...
@@
 
-Finally, given a distributed pipeline with runtime \todo{annotations}, \sys needs to schedule and execute it.
+\kk{On the dataflow graph, Dish iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.}
+
+Finally, given a distributed pipeline with runtime \todo{annotations}\kk{the runtime just gets a dataflow graph}, \sys needs to schedule and execute it.
 This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
@@
 From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.
+\kk{I am not exactly sure I understand that :) }
 

### 2019-11-21 pldi/draft.tex
Add notes in related work} git push

@@
 
+\kk{Rename this section to panner activators}
+
 \kk{This whole section for just in itme planning, is also relevant for
@@
 
+\kk{On our dataflow graph model: Our data flow graph model is
+  different than most distributed batch and stream processing systems,
+  since the shell is a hybrid between the two. Batch processing
+  systems have no notion of data order, and stream processing systems
+  usually have order, but none of these systems supports this notion
+  of reading from different sources in sequence. Our dataflow graph
+  model supports this by including information about incoming edge
+  order. Another benefit of our dataflow model, is that together with
+  the equations that hold on stateless and pure commands, data
+  parallelism is exposed in the graph as task parallelism. This allows
+  the unoptimized and optimized dataflow graph to have the same
+  unified representation, and doesn't lead to an ad-hoc optimization
+  of the dataflow graph as done in the other systems (the last bit is
+  too aggressive). }
+
 At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
@@
 \heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
 simplify certain patterns, 
@@
 
+\kk{We need to cite DGSH}
+
 \section{Conclusion}
 \label{discussion}
-
-\kk{We need to cite DGSH}
 

### 2019-11-21 pldi/draft.tex
A syntax for the DSL, and some progress on Front end

@@
 \usepackage{wrapfig}
+\usepackage{syntax}
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_schematic.pdf}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
 \caption{
@@
 % \centering
-% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
 % \caption{
@@
     % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
+    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
   % \end{center}
@@
     % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
+    \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
   % \end{center}
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
 \caption{
@@
 
+\kk{New sketch of the language}
+
+We assume that commands satisfy the standard that options and flags
+come first, and then file arguments. The language contains a record
+for each command name, together with a sequence of predicates that map
+to command categories and designation of input output.
+
+%% Can be tricked according to this:
+%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
+
+\begin{grammar}
+  <flag> ::= `-' <string>
+
+  <category> ::= `stateless' | `pure' | ...
+
+  <arg> ::= `arg[' <int> `]'
+
+  <input> ::= `stdin' | <arg>
+
+  <output> ::= `stdout' | <arg>
+
+  <flag-pred> ::= <flag>
+  \alt `value' <flag> = <string>
+  \alt `!' <flag-pred>
+  \alt <flag-pred> `or' <flag-pred>
+  \alt <flag-pred> `and' <flag-pred>
+
+  <assignment> ::= `(' <category>, <input>, <output> `)'
+
+  <predicate> ::= <flag-pred> `=>' <assignment>
+
+  <pred-list> ::= `|' <predicate> <pred-list>
+  \alt `|' `otherwise' `=>' <assignment>
+
+  <command> ::= <name> `:' <pred-list>
+
+\end{grammar}
+
+\kk{We don't need to designate the stderror, because we assume that it
+  is never the main output of a command and it will never be used by
+  the input of another command in the pipeline. If someone indeed
+  wants to do this, they can just use a redirect I think to get around
+  it.}
+
+\kk{Give an example of a flag that has value and how this could alter
+  category or input etc...}
+
 Sketch of the language:
@@
 
-\paragraph{Merge and Split}
-
-\kk{Mergers and splitters, and how they are implemented using shell
-  constructs (cat and head). This subsection might be a better fit in
-  some other section, maybe something related to the planner or the
-  implementation, as it is a very elegant way to implement splitters
-  and mergers while staying completely inside the shell programming
-  model.}
-
-It is common in the dataflow graph for a node to have many incoming
-edges, or many outgoing edges (when all but the last outgoing edges
-are bounded). However some nodes, might read their input from one file
-(e.g. stdin). In this case, the distributed implementation needs to
-concatenate input files, and split output files. Here is an example:
-
-\kk{Show an example graph! of cat | tr | sort. The command in the
-  middle must only take input from one file (maybe just stdin).}
-
-\kk{This transition needs to be stronger} Luckily, the shell offers
-primitives for file manipulations that can be used to implement a
-merger and a splitter. More concretely a merger can be implemented by
-a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
-and an output file \verb|out|, a merger can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
- cat $in1 $in2 ... > $out
-\end{lstlisting}
-
-\noindent
-On the other hand, given an input file \verb|in| that is a FIFO pipe,
-and two output files \verb|out1, out2| the first of which is bounded
-to \verb|N| lines, a 2-splitter can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
- head -n $N $in > $out1 ; cat $in > $out2
-\end{lstlisting}
-
-\noindent
-Using the 2-splitter as a building block, a pipe can be split to
-arbitrarily many different pipes.
-
-\kk{Do I also need to include a splitter for normal files?}
-
 \section{Front End}
@@
 programs that cannot be arbitrarily parallelized without affecting the
-program behaviour.
-
-\kk{Give examples of writing to environment variables, producing
-  irreversible side effects, the ; and \&\& operators.}
+program behaviour, e.g. commands producing an irreversible effect, or
+commands connected with a \texttt{\&\&} operator. In addition, the
+fact that the shell is \todo{extremely dynamic}, further increases the
+challenge of correctly identifying what parts can be distributed.
+
+%% \kk{Give examples of writing to environment variables, producing
+%%   irreversible side effects, the ; and \&\& operators.}
 
 In order to address this issue, we introduce the notion of
-distributable AST subtrees, and we design an analysis that given the
-shell script Abstract Syntax Tree identifies distributable AST
-subtrees and translates each one of them to a dataflow graph. The AST
-of the shell script is produced using a POSIX compliant parser
-\kk{Reference libdash github}.
-
-\subsection{Distributable Subtrees}
-
-\kk{The goal of this section is to convince the reader ``intuitively''
-  that we choose safe parts of a shell program to distribute. There is
-  no formal argument, but one should be able to intuitively believe
-  us.}
+distributable program regions, and we design an analysis that given the
+abstract syntax tree of a shell script identifies distributable
+regions and translates each of them to a dataflow graph. The
+AST of the shell script is produced using a POSIX compliant parser
+\kk{Reference libdash github. This reference should probably go to
+  evaluation}.
+
+\subsection{Distributable Regions}
+
+\kk{I wanted this to be purely about the regions, and a next
+  subsection be about the analysis, but the separation line is very
+  blurry, so they are now in the same subsection.}
+
+\kk{The goal of this subsection is to explain that Dish does a safe
+  but effective analysis to identify which parts of the scripts to
+  parallelize.}
 
@@
 
-\kk{Give an example of a shell script that cannot be distributed
-  completely, Possibly involving ; and writing to a variable.}
-
-\kk{Intuitively explain what could be distributed and what not.}
-
-\kk{Our notion of distributable subtrees has two dimensions. First,
-  the state of the shell should be an invariant in a distributable
-  subtree. We cannot know this completely without evaluating, so we do
-  a conservative analysis. The second dimension is that there exists
-  at least one component that can be abstracted as a node of the
-  dataflow graph. This can be anything that includes a command.}
-
-%% Even though commands are split in categories, and several pipelines
-%% are distributable, not all parts of a shell script are stateless and
-%% several of them have effects that could affect the behaviour of a
-%% program after distribution.
-
-\kk{All of our assumptions and study are based on the POSIX
-  standard. Reference?}
-
-\kk{Not checked Assumption. A command doesn't write/read to arbitrary
-  files that are not mentioned in its arguments. This is necessary to
-  enforce that no two commands concurrently read/write from the same
-  file.}
-
-\kk{Not checked Assumption: No signals, traps, and sets at the
-  moment. At the moment we don't have a way to handle signals and
-  traps.}
-
-\kk{Partially checked Assumption. Only one reader and one writer to
-  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
-  file exists once in the substree. This is a conservative assumption
-  that ensures that no two commands write or read to the same file
-  concurrently. We kind of check it but not completely. We check that
-  the expanded evaluated files are not the same, but one could have
-  part of a file in a variable etc.}
-
-\kk{Checked assumption: How can we make sure that nodes in the IR are
-  deterministic. Do we make some assumption that enforces this? I
-  guess that commands that are nodes in the IR are all deterministic?
-  That is stateless and pure commands are deterministic.}
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+  cat f1 f2 | grep ``foo'' > f3 &&
+  sort f3 #Replace
+\end{lstlisting}
+
+Notice that the structure of a shell program clearly identifies
+components that can be executed concurrently, and barriers that
+enforce synchronization. More precisely, \texttt{cat f1 f2} and
+\texttt{grep ``foo''} already execute concurrently in the standard
+shell, while \texttt{sort f3} waits for their completion before being
+executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and \texttt{sort
+  f3} are maximal distributable regions, in the sense that they
+cannot be extended beyond \texttt{\&\&}.
+
+Concretely, a distributable region cannot permeate through the
+following shell constructs: ;, \&\&, \texttt{||}, function
+definitions, command substitutions, and control flow primitives. On
+the other hand, the pipe operator \texttt{|} and the parallel
+composition operator \& enable parallelization, and extend
+distributable regions. Essentially, distributed regions correspond to
+subsections of the program that would be executed by different shell
+processes.
+
+\kk{I don't know whether I should mention the following: While these
+  control flow constructs can be parallelized and there has been
+  research on it, it is orthogonal to our work, and can be
+  incorporated as future work.}
+
+Dish performs a depth first search pass on the AST of the given shell
+program, extending the distributable regions from the bottom
+up, translating their \todo{independent} components to dataflow graph
+nodes, until it reaches a barrier construct. Note that barrier
+constructs can be contained in a distributable subexpression, as a
+\todo{whole independent component}. An illustrative example follows:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+  (sort f1 > g1 && sort f2 > g2) &
+  sort -m g1 g2
+\end{lstlisting}
+
+In the above example\kk{Can we make example not floating?}, while
+\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
+  g2} are all in different distributable regions. The dataflow graph
+that corresponds to the \todo{large} region, has two nodes,
+\texttt{sort f1 > g1 \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.
+
+The analysis algorithm also finds the category of each node that it
+adds to the dataflow graph, using the rules described in \ref{}. Some
+interesting cases of the analysis algorithm are shown in pseudocode
+below:
+
+\begin{lstlisting}[language=python, float=h]
+  def analysis(node):
+    ...
+    if(node.name == 'Pipe'):
+      return pipe_graphs([analysis(child)
+        for child in node.children])
+    elif(node.name == 'And'):
+      node.children = [analysis(child)
+        for child in node.children]
+      return node
+    elif(node.name == 'Command'):
+      if(not safe(node)):
+        return make_ir_node(node, 'unsafe')
+      else:
+        return make_ir_node(node,
+          find_category(node))
+    ...
+\end{lstlisting}
+
+\kk{Explain how the three above cases work, mentioning what does
+  pipe-graphs and safe do. The important point to get across is to
+  explain why we do the ``conservative checks'' in safe, and how they
+  preserve safety.}
 
 \paragraph{analysis}
-
-The analysis is done in a DFS fashion on the AST. Distributable
-subtrees grow from the bottom up, until a construct that doesn't
-``allow'' them to go through is encountered.
-
-%% Mention all the different constructs that our front end handles, and
-%% intuitively mention why it is ``safe'' to add them to the distribution
-%% pipeline.
-
-Subtrees don't cross the following:
-
-\begin{itemize}
-\item Sequence, And,  or
-\item Function definitions
-\item Command Substitutions
-\item control flow: while, for, if, case: While these constructs can
-  be parallelized and there has been research on it, it is orthogonal
-  to our work, and can be incorporated as future work.
-\end{itemize}
 
@@
 
-\kk{Maybe mention the recursive case. I am not sure if that should be
-  here, or in the distribution planner. In any case, it should be
-  after we say that an IR is replaced with a call to the planner with
-  the IR saved in a file. The example is a command in a pipeline whose
-  first argument is a command substitution with a pipeline. Executing
-  the outside dataflow graph, would lead to the command to be executed
-  as part of this shell. However its first argument would have a call
-  to the planner itself, leading to another distributed implementation
-  running in parallel. The rest of the pipeline would wait for this to
-  end before it continues. Assuming that the files that both of these
-  graphs use are different, and that the above assumptions hold, there
-  should be no interference between the two graphs and they should be
-  able to run in parallel}
+\paragraph{Assumptions}
+
+\kk{Rephrase that paragraph, to mention the assumtpions that the front
+  end makes, and why they are not so unreasonable}
+
+There are some assumptions that we make in order to preserve the
+correctness of the computation under distributability optimizations.
+
+\kk{Our notion of distributable subtrees has two dimensions. First,
+  the state of the shell should be an invariant in a distributable
+  subtree. We cannot know this completely without evaluating, so we do
+  a conservative analysis. The second dimension is that there exists
+  at least one component that can be abstracted as a node of the
+  dataflow graph. This can be anything that includes a command.}
+
+%% Even though commands are split in categories, and several pipelines
+%% are distributable, not all parts of a shell script are stateless and
+%% several of them have effects that could affect the behaviour of a
+%% program after distribution.
+
+\kk{All of our assumptions and study are based on the POSIX
+  standard. Reference?}
+
+\kk{Not checked Assumption. A command doesn't write/read to arbitrary
+  files that are not mentioned in its arguments. This is necessary to
+  enforce that no two commands concurrently read/write from the same
+  file.}
+
+\kk{Not checked Assumption: No signals, traps, and sets at the
+  moment. At the moment we don't have a way to handle signals and
+  traps.}
+
+\kk{Partially checked Assumption. Only one reader and one writer to
+  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
+  file exists once in the substree. This is a conservative assumption
+  that ensures that no two commands write or read to the same file
+  concurrently. We kind of check it but not completely. We check that
+  the expanded evaluated files are not the same, but one could have
+  part of a file in a variable etc.}
+
+\kk{Checked assumption: How can we make sure that nodes in the IR are
+  deterministic. Do we make some assumption that enforces this? I
+  guess that commands that are nodes in the IR are all deterministic?
+  That is stateless and pure commands are deterministic.}
 
@@
 
+\kk{This section could be moved to the intermediate representation,
+  leaving a small section about the iterative process of the planner
+  in the implementation/runtime section}
+
 In our case, this is achieved by iteratively applying graph
@@
   about command categories.}
+
+\kk{Mention the assumptions that need to hold for the graph
+  transformations to be valid. Commands must be deterministic, they
+  must not do any other side effect (such as writing to other files,
+  sending signals, etc). However, these assumptions must already be
+  checked when the developer designates the categories.}
 
@@
 late as possible in the process. This would be awesome.
+
+\paragraph{Merge and Split}
+
+\kk{Mergers and splitters, and how they are implemented using shell
+  constructs (cat and head). This subsection might be a better fit in
+  some other section, maybe something related to the planner or the
+  implementation, as it is a very elegant way to implement splitters
+  and mergers while staying completely inside the shell programming
+  model.}
+
+It is common in the dataflow graph for a node to have many incoming
+edges, or many outgoing edges (when all but the last outgoing edges
+are bounded). However some nodes, might read their input from one file
+(e.g. stdin). In this case, the distributed implementation needs to
+concatenate input files, and split output files. Here is an example:
+
+\kk{Show an example graph! of cat | tr | sort. The command in the
+  middle must only take input from one file (maybe just stdin).}
+
+\kk{This transition needs to be stronger} Luckily, the shell offers
+primitives for file manipulations that can be used to implement a
+merger and a splitter. More concretely a merger can be implemented by
+a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
+and an output file \verb|out|, a merger can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ cat $in1 $in2 ... > $out
+\end{lstlisting}
+
+\noindent
+On the other hand, given an input file \verb|in| that is a FIFO pipe,
+and two output files \verb|out1, out2| the first of which is bounded
+to \verb|N| lines, a 2-splitter can be implemented as:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+ head -n $N $in > $out1 ; cat $in > $out2
+\end{lstlisting}
+
+\noindent
+Using the 2-splitter as a building block, a pipe can be split to
+arbitrarily many different pipes.
+
+\kk{Do I also need to include a splitter for normal files?}
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
 \caption{

### 2019-11-21 pldi/draft.tex
A draft of the front end

@@
 
+\kk{The writing in this section is wandering around without a clear
+  purpose :( }
+
 \kk{It might make sense to merge this section with the one below with
@@
 In order to address this issue, we introduce the notion of
-distributable program regions, and we design an analysis that given the
-abstract syntax tree of a shell script identifies distributable
-regions and translates each of them to a dataflow graph. The
-AST of the shell script is produced using a POSIX compliant parser
-\kk{Reference libdash github. This reference should probably go to
-  evaluation}.
+distributable program regions, and we design a translation pass that
+given the abstract syntax tree of a shell script identifies
+distributable regions and translates each of them to a dataflow
+graph. The AST of the shell script is produced using a POSIX compliant
+parser \kk{Reference libdash github. This las sentence should probably
+  go to evaluation}.
 
@@
 
-\kk{I wanted this to be purely about the regions, and a next
-  subsection be about the analysis, but the separation line is very
-  blurry, so they are now in the same subsection.}
+%% \kk{I wanted this to be purely about the regions, and a next
+%%   subsection be about the analysis, but the separation line is very
+%%   blurry, so they are now in the same subsection.}
 
@@
   cat f1 f2 | grep ``foo'' > f3 &&
-  sort f3 #Replace
+  sort f3 # Replace with sth better
 \end{lstlisting}
@@
 enforce synchronization. More precisely, \texttt{cat f1 f2} and
-\texttt{grep ``foo''} already execute concurrently in the standard
-shell, while \texttt{sort f3} waits for their completion before being
-executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and \texttt{sort
-  f3} are maximal distributable regions, in the sense that they
-cannot be extended beyond \texttt{\&\&}.
+\texttt{grep ``foo''} would already execute concurrently in the
+standard shell, while \texttt{sort f3} waits for their completion
+before being executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and
+\texttt{sort f3} are maximal distributable regions, in the sense that
+they cannot be extended beyond \texttt{\&\&}.
 
 Concretely, a distributable region cannot permeate through the
-following shell constructs: ;, \&\&, \texttt{||}, function
+following barrier constructs: ;, \&\&, \texttt{||}, function
 definitions, command substitutions, and control flow primitives. On
@@
 
+\subsection{Translation Pass}
+
 Dish performs a depth first search pass on the AST of the given shell
-program, extending the distributable regions from the bottom
-up, translating their \todo{independent} components to dataflow graph
-nodes, until it reaches a barrier construct. Note that barrier
-constructs can be contained in a distributable subexpression, as a
-\todo{whole independent component}. An illustrative example follows:
+program, extending the distributable regions from the bottom up,
+translating their independent components to dataflow graph
+nodes, until a barrier construct is reached. Note that barrier
+constructs can be contained in a dataflow graph, as long as they don't
+extend beyong a single node. An illustrative example follows:
 
@@
 \texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
-  g2} are all in different distributable regions. The dataflow graph
-that corresponds to the \todo{large} region, has two nodes,
-\texttt{sort f1 > g1 \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.
-
-The analysis algorithm also finds the category of each node that it
-adds to the dataflow graph, using the rules described in \ref{}. Some
+  g2} are all in different dataflow graphs. The dataflow graph that
+corresponds to the whole program, has two nodes, \texttt{sort f1 > g1
+  \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.
+
+In order to identify possible opportunities for parallelism, the
+translation pass also identifies the category of each node, together
+with its input and output, using the rules described in \ref{}. Some
 interesting cases of the analysis algorithm are shown in pseudocode
@@
 \begin{lstlisting}[language=python, float=h]
-  def analysis(node):
+  def translate(node):
     ...
-    if(node.name == 'Pipe'):
-      return pipe_graphs([analysis(child)
+    elif(node.name == 'Pipe'):
+      return pipe_graphs([translate(child)
         for child in node.children])
     elif(node.name == 'And'):
-      node.children = [analysis(child)
+      node.children = [translate(child)
         for child in node.children]
@@
 
-\kk{Explain how the three above cases work, mentioning what does
-  pipe-graphs and safe do. The important point to get across is to
-  explain why we do the ``conservative checks'' in safe, and how they
-  preserve safety.}
-
-\paragraph{analysis}
-
-On the other hand: A distributable subtree starts whenever there is a
-pipe, or asynchronous list. If the subcomponents of the pipe or
-asynchronous list were already graphs, then we connect them to the
-bigger graph \kk{explain how we do pipe and asynchronous list}, and if
-they were not graphs, we make each subcomponent a node in the subtree,
-event though they are not commands. These nodes do not belong to any
-category, but still have stdin and stdout, and can be executed
-concurrently (as long as they satisfy the checked assumptions above),
-since they don't introduce new behaviours, they would still be
-executed concurrently when ran in the standard shell, producing
-possibly non-deterministic results with different interleavings.
-
-\kk{Explain how we merge two graphs when seeing a pipe}
-
-Whenever we see a subtree of the AST that is a command, we check the
-files that it writes to. If any of its arguments is a command
-substitution or contains anything that is evaluated\kk{make this more
-  precise}, its category is marked as sequential\kk{name?}. Otherwise
-we check its its category and its input output arguments (using the
-methods described in \Cref{ext}). The command is then initialized as
-the node of a singular graph, with the inputs and outputs of the graph
-being the ones of the node.
+%% \kk{Explain how the three above cases work, mentioning what does
+%%   pipe-graphs and safe do. The important point to get across is to
+%%   explain why we do the ``conservative checks'' in safe, and how they
+%%   preserve safety.}
+
+Whenever the algorithm encounters a pipe node, it recursively
+translates its subcomponents, and then merges them in a dataflow graph
+by connecting the output of the first with the input of the second,
+\etc. Before the different dataflow graphs are connected, the
+algorithm checks that at most one node in the graph writes at every
+file, and at most one node reads from a file that another node writes
+to. This is important, in order to avoid inconsistent behaviour after
+distributing nodes in the dataflow graph \kk{Does this need an
+  example?}. Whenever the algorithm encounter an \texttt{\&\&} node,
+it recursively translates its subcomponents (similarly to the pipe
+node), but then doesn't merge them in a large dataflow graph. Finally,
+whenever the algorithm encounters a command, it first checks whether
+it is safe to possibly parallelize this command. A command is
+considered to be safe, if it does not make any modification to the
+shell state (e.g. changing the current directory, modifying
+environment variables), and if it does not read from a
+\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
+  unsafe command can still be safely added to the dataflow graph as a
+  single node, since this doesn't introduce any new
+  behaviours. However it cannot be parallelized further than that.} If
+the command is considered safe, its category is identified and a node
+containing that command is created and added to a singleton dataflow
+graph.
+
+It is important to note that due to the highly dynamic nature of the
+shell, the safety properties that were mentioned above cannot be
+soundly checked statically. For example in the first case, two
+unexpaned strings could refer to the same file, but the translation
+pass cannot infer that. Hence, in both cases, the checks were
+approximated using a \todo{best-effort} method. \kk{Maybe mention that
+  as future work, we want to tighly integrate with the shell shtepper
+  that will allow to maximally expand everything before making this
+  decision.}
 
@@
 
-\kk{Rephrase that paragraph, to mention the assumtpions that the front
-  end makes, and why they are not so unreasonable}
-
-There are some assumptions that we make in order to preserve the
-correctness of the computation under distributability optimizations.
-
-\kk{Our notion of distributable subtrees has two dimensions. First,
-  the state of the shell should be an invariant in a distributable
-  subtree. We cannot know this completely without evaluating, so we do
-  a conservative analysis. The second dimension is that there exists
-  at least one component that can be abstracted as a node of the
-  dataflow graph. This can be anything that includes a command.}
-
-%% Even though commands are split in categories, and several pipelines
-%% are distributable, not all parts of a shell script are stateless and
-%% several of them have effects that could affect the behaviour of a
-%% program after distribution.
-
-\kk{All of our assumptions and study are based on the POSIX
-  standard. Reference?}
-
-\kk{Not checked Assumption. A command doesn't write/read to arbitrary
-  files that are not mentioned in its arguments. This is necessary to
-  enforce that no two commands concurrently read/write from the same
-  file.}
-
-\kk{Not checked Assumption: No signals, traps, and sets at the
-  moment. At the moment we don't have a way to handle signals and
-  traps.}
-
-\kk{Partially checked Assumption. Only one reader and one writer to
-  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
-  file exists once in the substree. This is a conservative assumption
-  that ensures that no two commands write or read to the same file
-  concurrently. We kind of check it but not completely. We check that
-  the expanded evaluated files are not the same, but one could have
-  part of a file in a variable etc.}
-
-\kk{Checked assumption: How can we make sure that nodes in the IR are
-  deterministic. Do we make some assumption that enforces this? I
-  guess that commands that are nodes in the IR are all deterministic?
-  That is stateless and pure commands are deterministic.}
+\kk{This paragraph is not extremely necessary. Maybe it can be moved
+  to limitations?}
+
+There are two additional assumptions that have to be satisfied in
+order for the distributed implementation that is generated by Dish to
+have equivalent behaviour with the original implementation. First of
+all, there should be no external signals to the commands while they
+execute. At the moment Dish assumes no faults, and that communication
+always succeeds. This assumption can be lifted by extending Dish with
+a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
+command reads and writes to arbitrary files that are not mentioned in
+their arguments. This is a necessary assumption, since commands are
+considered as black boxes and the only information that can be
+inferred about them is through their arguments.
+
+\kk{The following assumption should probably be mentioned in the
+  categories, since a command that is non-deterministic should not be
+  categorized to be neither stateless or pure. Assumption: How can we
+  make sure that nodes in the IR are deterministic. Do we make some
+  assumption that enforces this? I guess that commands that are nodes
+  in the IR are all deterministic?  That is stateless and pure
+  commands are deterministic.}
+
+%% \kk{All of our assumptions and study are based on the POSIX
+%%   standard. Reference?}
 

### 2019-11-22 pldi/draft.tex
Notes on distributability

--- 2019-11-21 pldi/draft.tex
@@
 \newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
+\newcommand{\tti}[1]{\texttt{\scriptsize #1}}
 \newcommand{\spol}[1]{\scriptsize{\sc#1}}
@@
 \newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}
+
+\newcommand{\cn}[1]{\mbox{\textcircled{\footnotesize #1}}}
+\newcommand{\tcn}[1]{\mbox{\textcircled{\scriptsize #1}}}
+
+\newcommand{\pur}{\cn{\textsc{P}}}
+\newcommand{\sta}{\cn{\textsc{S}}}
+\newcommand{\dfs}{\cn{\textsc{F}}}
+\newcommand{\sid}{\cn{\textsc{E}}}
+\newcommand{\irr}{\cn{\textsc{I}}}
+
+\newcommand{\tpur}{\tcn{\textsc{P}}}
+\newcommand{\tsta}{\tcn{\textsc{S}}}
+\newcommand{\tdfs}{\tcn{\textsc{F}}}
+\newcommand{\tsid}{\tcn{\textsc{E}}}
+\newcommand{\tirr}{\tcn{\textsc{I}}}
 
@@
 To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation\kk{I am not sure whether stream processing provides information for distributing a computation on its own.
-Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies \todo{maximal} sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams \kk{Isn't uniq pure?
-I am not sure how we can handle it}, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
+\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
 
@@
 % Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}
 
@@
 The shell does not have any visibility into these options; 
-  their parsing and evaluation is left entirely up to each command \kk{A minor point: the shell does the expansion, and then hands it of to the command (I am saying that because evaluation might be confounded with expansion so we might want to clarify that)}.
+  their parsing and evaluation is left entirely up to each command.
 Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
@@
 Intuitively,
- (i) the parallel-composition operator exposes \todo{task} parallelism \kk{I have named that operator level parallelism, which one do we prefer?},
+ (i) the parallel-composition operator exposes task parallelism,
  (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator enables both task and data parallelism, and \kk{How does it enable data parallelism?}
+ (iii) the pipe operator enables both task and data parallelism, and
  (iv)  other composition operators can be viewed as barrier candidates.
@@
 Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
-\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying \todo{maximal} distributable sub-expressions.
+\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying maximal distributability sub-expressions.
 It converts the AST to a dataflow graph, and runs...
@@
 
-\kk{On the dataflow graph, Dish iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.}
-
-Finally, given a distributed pipeline with runtime \todo{annotations}\kk{the runtime just gets a dataflow graph}, \sys needs to schedule and execute it.
+Finally, given a distributed pipeline with runtime \todo{annotations}, \sys needs to schedule and execute it.
 This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
@@
 From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.
-\kk{I am not exactly sure I understand that :) }
 
@@
 \section{Distributability Classes}
-
-\subsection{Comnands}
-\label{bg:cmd}
-
-% Sets vs. Classes
-
-\begin{table}[h]
+\label{distributability}
+
+To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of individual commands involved in them.
+This understanding is encoded in a command's \emph{distributability} class.
+Put informally, a distributability class representing the level of synchronization required by copies of a command executing concurrently.
+The class captures characteristics that are imprortant for distribution, rather than all the observable behavior of a command.
+\sys leans towards a few, coarse classes rather than many, detailed ones---after all, these classes should be easily understood by command developers.
+% Make the connections that this is like a type, that doesn't capture everything, but is good-enough
+
+The section starts with a distributability study of standard libraries commands, such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+Building on this study, it presents a DSL for describing a command's distributability characteristics~\sx{ext}.
+In return, \sys uses this DSL to tag commands in the standard libraries and generate the transformations presented in later sections.
+
+\subsection{Distributability of Standard Libraries}
+\label{cmd}
+
+% nv TODO: this is a hierarchy
+Broadly, shell commands can be split into five major classes with respect to their distribution characteristics, depending on how they interact with state.
+Tab.~\ref{tab:classes} summarizes the discussion;
+These classes are listed (and presented later) in ascending difficulty of distribution; later classes require further effort.
+In this order, each class can be thought as a subset of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+
+\begin{table}[t]
 \center
@@
     \textbf{Distributability Classes}.
-    Broadly, \unix commands can be broken down into to five classses.
+    Broadly, \unix commands can be broken down into to five classes.
   }
 }
-\begin{tabular}{l @{\extracolsep{\fill}} lll}
+\begin{tabular}{l @{\extracolsep{\fill}} llll}
 \toprule
-Class                           & Example Commands        & POSIX            & Core-utils\\
+Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
 \midrule
-Stateless                       &                         &                  &           \\
-Pure                            &                         &                  &           \\
-(Distr.) File-System            &                         &                  &           \\
-Side-effectful                  &                         &                  &           \\
-Irreversibly Side-effectful     &                         &                  &           \\
+Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
+Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
+(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
+Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
+Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{restart}     &  5  (0.4\%)      &           \\  % 5
 % \midrule
@@
 
-Define distributability 
-Explain distributability classes, based on how
-
-\heading{Preliminaries}
-
-\heading{POSIX, GNU Core-utils, and beyond}
-
-\heading{Stateless}
-
-\heading{Pure}
-
-\heading{File-system}
-
-\heading{Side-effectful}
-
-\heading{Irreversably Side-effectful}
+% \heading{Preliminaries}
+% \heading{POSIX, GNU Core-utils, and beyond}
+
+\heading{Stateless Commands}
+The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
+These are commands expressible as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
+At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
+Workloads that use only stateless commands are trivial to parallelize:
+  they do not require any synchronization to maintain correctness,
+  they do not need care planning on where to split.
+ % map :: (a -> b) -> [a] -> [b]
+
+Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time);
+  this could potentially allow further parallelization within a single line element.
+However, exactly because of their local, commands operating on individual characters are not particularly expensive computationally.
+
+\heading{Pure Commands}
+The second class contains pure commands \pur.
+These commands respect functional purity, returning the same results for the same inputs;
+  however, they need to maintain internal state across the entire invocation.
+
+Pure commands often cannot operate in an online fashion, but need to block until the end of the stream.
+A typical example of this is \ttt{sort}, a command that cannot start before the last line of input has been consumed.
+
+Parallelizing these commands is slightly more challenging and depends on the command.
+Some commands have known divide-and-conquer equivalents.
+
+Pure commands~ones that maintain 
+Distribution:
+
+\heading{File-system Commands}
+
+This is the largest class, as the file (and file-system) abstraction is so central to \unix.
+In fact, \unix proxies several file-unrelated operators through the file-system---for example, security in \unix is grafted upon the file-system:
+
+(and all associated commands) are always
+And most of these commands seem control-plane operations that operate on identifiers~\sx{bg}---for example, renaming a file or changing group ownership.
+% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
+These are mostly distributable
+
+Distribution:
+
+\heading{Side-effectful Commands}
+Another large class of commands are ones that have side-effects to memory.
+Because of timesharing and multi-tasking (many users can operate multiple terminals on single \unix machine), a set of commands are devoted to accessing variables---environment variables, shell parsing flags,
+The different formats have multiplied these commands
+
+Thankfully, the vast majority of these commands (a ratio of 21:4) only \emph{read} values;
+Even better,  usually read values that are not altered by user---\eg \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information.
+Even if \sid commands are protected with transactions
+these benefits are even greater for read-only or read-mostly accesses.
+
+\heading{Irreversibly Side-effectful Commands}
 
@@
 We can encode powerful
-
-\section{Dish Overview}
 
@@
 \subsection{Just in time planning}
-
-\kk{Rename this section to panner activators}
 
@@
 
-\kk{On our dataflow graph model: Our data flow graph model is
-  different than most distributed batch and stream processing systems,
-  since the shell is a hybrid between the two. Batch processing
-  systems have no notion of data order, and stream processing systems
-  usually have order, but none of these systems supports this notion
-  of reading from different sources in sequence. Our dataflow graph
-  model supports this by including information about incoming edge
-  order. Another benefit of our dataflow model, is that together with
-  the equations that hold on stateless and pure commands, data
-  parallelism is exposed in the graph as task parallelism. This allows
-  the unoptimized and optimized dataflow graph to have the same
-  unified representation, and doesn't lead to an ad-hoc optimization
-  of the dataflow graph as done in the other systems (the last bit is
-  too aggressive). }
-
 At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
@@
 \heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
 simplify certain patterns, 
@@
 
-\kk{We need to cite DGSH}
-
 \section{Conclusion}
 \label{discussion}
+
+\kk{We need to cite DGSH}
 

### 2019-11-22 pldi/draft.tex
Update acmart
Cleanup

@@
 %\documentclass[sigplan]{acmart}\settopmatter{}
+
+\acmSubmissionID{248}
 
@@
 \begin{abstract}
-  \nv{placeholder}
   Distributed systems offer notable benefits over centralized ones.
-  Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed framework interfaces, or distribution annotations.
-% 
-%   \emph{Distribution-oblivious programming} is a new approach for..?
-  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. %  with regards to the sequential program.
-  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known distributability classes.
-  \sys's insight is that shell pipelines already express stream computations that can automatically distributed.
+  Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed frameworks, or source-level annotations.
+  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
+  \sys's insight is that shell pipelines already express stream computations that can be automatically distributed.
   To achieve this, \sys
-% a series of techniques for automatically 
 decomposes primitives into distributability classes,
@@
 Its runtime component provides orchestration and planning support during the execution of the program.
+Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
+\end{abstract}
+
+% These are from the abstract:
+%  with regards to the sequential program.
+  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known distributability classes.
+% a series of techniques for automatically 
 % 
@@
 %
-Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
-\end{abstract}
 
@@
 After a comparison with related prior work~\sx{related}, the paper closes with a discussion of related work and possible future directions~\sx{discussion}.
+
+% Additional material, in anonymized form, is available at~\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
 
@@
 
-\subsection{The Shell Language}
-\label{bg:shell}
-
-\heading{Implicits}
-Expansion
-Implicits
-Context
-
-\heading{Language Constructs}
-
-Even without these, the Unix pipeline language is powerful enough to capture complex pipelines.
-
-We can encode powerful
-
 \section{Extensibility}

### 2019-11-22 evaluation/minimal_sort_100_env.sh
Results

@@ -1,1 +1,1 @@
-IN=../scripts/input/i10M.txt
+IN=../scripts/input/i100M.txt

### 2019-11-22 evaluation/minimal_sort_4_env.sh

@@ -1,1 +1,1 @@
-IN=../scripts/input/i1G.txt
+IN=../scripts/input/i100M.txt

### 2019-11-22 evaluation/topn_2_env.sh

@@ -1,2 +1,2 @@
 N=1000
-IN=../scripts/input/i1M.txt
+IN=../scripts/input/i10M.txt

### 2019-11-22 evaluation/wf_2_env.sh

@@ -1,1 +1,1 @@
-IN=../scripts/input/i1M.txt
+IN=../scripts/input/i10M.txt

### 2019-11-22 pldi/draft.tex

--- 2019-11-21 pldi/draft.tex
@@
 
-\section{Intermediate Representation/Dataflow graph}
+\section{Dataflow Graph Model}
 \label{ir}
 
-\kk{Inline listing?}
-
-\kk{Try to be more formal in the section, by using the notation that
-  is defined in the nodes paragraph.}
+\TODO{Replace texttt with inline listings?}
 
@@
 through these files. We propose a slight variation of the standard
-dataflow graph, that is commonly found in stream processing systems,
-as the representation of a distributed shell computation. The main
-benefit of the dataflow graph model is that it clearly exposes
-operator level parallelism \kk{Is this a common term? Is there any
-  definitive citation for it?}, as different nodes can be independent
-processing workers. In our case each node represents a command, and
-each edge represents a file.
-
-\kk{Before that I have to mention that the basic data type $D$ is a
-  line of characters. If we move the edges section before the nodes
-  that will be fine.}
+dataflow graph model, that is commonly found in stream processing
+systems, as the representation of a distributed shell computation. The
+main benefit of the dataflow graph model is that it clearly exposes
+task parallelism, as different nodes can be independent processing
+workers. In our case each node represents a command, and each edge
+represents a file.
+
+\subsection{Graph Components}
 
@@
 Nodes $f$ of the graph represent functions from one (possibly empty)
-input stream to an output stream $f : D^* \rightarrow D^*$.  This
-representation captures the majority of shell commands. The nodes $f$
-are monotonic, meaning that they cannot retract their output. More
-formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
-that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
-omitted from now on. An example of a node is the following command
-\kk{Does it need ref? Also maybe this example is too trivial and can
-  be removed.}:
+input stream to an output stream $f : D^* \rightarrow D^*$. $D$
+represents the basic data type of a line of characters. This
+representation captures the majority of shell commands. The nodes are
+monotonic, meaning that they cannot retract their output. More
+formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
+$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
+and will be omitted when obvious. An example of a node is the
+following command \kk{Does it need ref? Also maybe this example is too
+  trivial and can be removed.}:
 
@@
 returns a stream of lines on its standard output. An important
-observation though is that commands in the shell might read their
-input stream from a sequence of different sources. An illustrative
-example is the command \kk{Inline listing}cat which reads its input
-from a combination of files and its standard input.
+observation is that commands in the shell might read their input
+stream from a sequence of different sources. An illustrative example
+is the command \texttt{cat} which reads its input from a combination
+of files and its standard input.
 
@@
 
-\kk{Should I reference that this example was takes from the cat
-  manpage?} The example above reads file \texttt{x} until it
-encounters the end of file, then reads from \texttt{stdin}, and then
-reads from \texttt{y}. In order to handle this particularity, nodes in
-the dataflow graph can have many incoming edges, that are clearly
-ordered in a sequence. Note that this comes in contrast with standard
-dataflow graph representations where multiple incoming edges either
-mean zipping the input to pairs, or arbitraliy interleaving the
-different inputs.
-
-\kk{A note about nodes and commands that can write and read from more
-  files that the ones in their streams}
+The example above reads file \texttt{x} until it encounters the EOF
+character, then reads from \texttt{stdin}, and then reads from
+\texttt{y}. In order to handle this particularity, nodes in the
+dataflow graph can have many incoming edges, that are ordered in a
+sequence. Note that this does not correspond to standard dataflow
+graph representations where the semantics of multiple incoming edges
+is either zipping the inputs in pairs, or arbitraliy interleaving them.
+
+\TODO{Is the following paragraph best suited here? It describes an
+  assumption that we make about commands, to guarantee the validity of
+  our transformations. The assumption is needed due to the limitation
+  of the programming model of one input one output per command.}
 
@@
 \kk{Would an example here help}) and do not represent streams of data,
-and therefore are not considered part of the dataflow graph. Because
-of the assumption in section Front End \ref{}, that each file exists
-only once in the dataflow graph, we can safely assume that reading and
-writing to these static files does not interfere with the distributed
-implementation, and doesn't alter the behaviour of the program.
+and therefore are not considered part of the dataflow graph. We assume
+that these static files can only be accessed by one node, thus
+ensuring that they do not interfere with the distributed
+implementation.
+%% Because of the assumption in section Front End \ref{}, that each file
+%% exists only once in the dataflow graph, we can safely assume that
+%% reading and writing to these static files does not interfere with the
+%% distributed implementation, and doesn't alter the behaviour of the
+%% program.
 
 \paragraph{Edges - Files}
-
-\kk{Maybe edges has to be moved before nodes, so that we can define
-  the line data type}
 
@@
 graph. Edges are represented as possibly unbounded streams of type
-$D*$, where $D$ represents a line, that is a sequence of characters
-followed by \verb|\n| \kk{or EOF?}. They could either refer to a
-specific named file in the file system, or just be FIFO pipes that
-used for interprocess communication \kk{Are there any other file
-  types, such as URLs, ...?}.
-
-\kk{Show an example pipe and its graph}.
-
-In order to expose operator parallelism with the output of a command,
+$D*$. They could either refer to a specific named file in the file
+system, or just be FIFO pipes that used for interprocess communication
+\TODO{Are there any other file types, such as URLs, ...?}.
+
+\TODO{Show an example pipe and its graph}.
+
+In order to expose data parallelism from the output of a command,
 files that don't point to a resource can have an upper bound on the
@@
 
-\kk{An example of a command that has two outputs and one goes to one
-  command, while another output goes to a different one. Maybe only
-  show the graph here and not the command itself.}
+\TODO{Add an example of a command that has two outputs and one goes to
+  one command, while another output goes to a different one. Maybe
+  only show the graph here and not the command itself.}
 
@@
 account for that, edges that refer to an input resource contain the
-number of lines of the file that they refer to. \kk{At the moment this
-  is not supported by our real system. The whole paragraph above can
-  go if we decide that it is not a good idea to talk about it now.}
+number of lines of the file that they refer to.
 
@@
 
-\kk{(Maybe) Depending on which of the two paragraphs goes last
-  (nodes/edges), I have to explain there that the last incoming edge
-  to a node must be unbounded (if it doesn't point to a resource). I
-  am actually not sure whether I should include this here.}
-
 \kk{(Maybe) It might be beneficial to mention that many inputs do not
@@
   symmetric.}
+
+\subsection{Graph transformations}
+\label{ir:transformations}
+
+As mentioned above, the dataflow graph exposes task
+parallelism. Different nodes in the graph can be executed
+concurrently, communicating through the edges. Task level parallelism
+is already exposed by the shell, by using pipes \texttt{|} and the
+background operator \texttt{\&}. Unfortunately, simple task
+parallelism leaves a lot to be desired, since scaling a computation
+requires manually introducing more operators.
+
+In order to adequately utilize the available computational resources,
+a distributed implementation should exploit data parallelism. Data
+parallelism is based on the observation that some operations produce
+the same result when performed independently on subsets of the input
+data, merging the results afterwards. This is the basic optimization
+that is used by popular distributed processing systems such as
+MapReduce\kk{cite} and Apache Spark\kk{cite}. \todo{Unfortunately, the
+  dataflow graph model of most of these systems is not expressive
+  enough to represent data parallelism, so the distributed generations
+  that they generate, can not be represented in their dataflow model.}
+\kk{This last sentence is extremely important and we should make it
+  shine.}
+
+In contrast, our dataflow model inherently supports the data
+parallelism that is found in shell programs. More precisely, data
+parallelism can be easily exposed as task parallelism in our model. A
+very important consequence of this is that we can iteratively apply
+optimizations that expose data parallelism, or even revert them when
+the resulting dataflow graph is too wide. This flexibility enables
+unified reasoning and implementation of the unoptimized and optimized
+version of the dataflow graph. \kk{This is exciting, and maybe it
+  should be moved up (or rephrased in the introduction)}
+
+Before describing the different types of parallelization
+optimizations, we formalize our intuition about stateless and pure
+commands. Recall the example \ref{} from section Categories
+\ref{}. This example shows the stateless command x being applied to
+...\kk{Complete this after we have an example}. Formally, a commmand
+$f$ is stateless if it is ``linear'' with respect to concatenation,
+i.e. the following equation holds:
+
+\[
+\forall x, x', f(x.x') = f(x).f(x')
+\]
+
+Similarly, every pure command $f$ that can be written as a map $m$ and a
+reduce $r$ step satisfies the following:
+
+\[
+\forall x, x', f(x.x') = r(m(x).m(x'))
+\]
+
+Based on these equations, we can define a maximal node split operation
+on a stateless node $v$ with $n$ input ordered edges and an output
+edge to node $v'$, to replace $v$ with $n$ new nodes, giving each one
+of them one of the incoming edges of $v$, and routing them in order to
+point to $v'$. The maximal node split operation can similarly extended
+for pure nodes.
+
+\kk{If there is time I can work out a formal definition and a proof
+  sketch why this transformation preserves the output of the dataflow
+  graph}.
+
+%% \begin{definition}
+%% Given a dataflow graph $G = (V, E, O)$, where $V$ is a set of nodes
+%% representing commands, $E$ is a set of edges representing files, and
+%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
+%% edges. $v_{out}$ represents a concatenation of all the outputs of the
+%% dataflow graph. We represent the total order for a node $v$ as
+%% $<_v$. Given a node a node $v$
+%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
+%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
+%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
+%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
+%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
+%% \end{definition}
+
+%% \begin{lemma}
+%% tofill
+%% \end{lemma}
+
+In addition to node splits, we can also perform edge splits in order
+to increase the number of incoming edges for some nodes to expose
+possible data parallelism. If the edge refers to a real file, then the
+edge split represents splitting the real file in chunks, while in
+cases where the edge connects two nodes, it is split to several edges
+with a bounded number of lines. Note that the final output edge of a
+valid node in the graph must always be unbounded if it does not refer
+to a file.
+
+\kk{Remember to mention the assumptions that need to hold for the
+  graph transformations to be valid in the Command categories
+  section. Commands must be deterministic, they must not do any other
+  side effect (such as writing to other files, sending signals,
+  etc). However, these assumptions must already be checked when the
+  developer designates the categories.}
 
@@
 
-\kk{The following assumption should probably be mentioned in the
-  categories, since a command that is non-deterministic should not be
-  categorized to be neither stateless or pure. Assumption: How can we
-  make sure that nodes in the IR are deterministic. Do we make some
-  assumption that enforces this? I guess that commands that are nodes
-  in the IR are all deterministic?  That is stateless and pure
-  commands are deterministic.}
-
-%% \kk{All of our assumptions and study are based on the POSIX
-%%   standard. Reference?}
-
-\section{Distribution Planner}
-
-\kk{I think that there is not enough content for the planner section
-  and it should probably just be called graph transformations. Then we
-  could have a small subsection at the end about how the planner
-  applies transformations iteratively starting from the source of the
-  graph and pushing them until the end.}
-
-As mentioned in \Cref{ir}, the dataflow graph exposes operator level
-parallelism. Different nodes in the graph can be executed
-concurrently, communicating through the edges. Operator level
-parallelism is already exposed by the shell, by using pipes \texttt{|}
-and the background operator \texttt{\&}. Unfortunately, simple
-operator level parallelism leaves a lot to be desired, since scaling a
-computation requires manually introducing more operators.
-
-In order to adequately utilize the available computational resources,
-a distributed implementation should exploit data parallelism. Data
-parallelism is based on the observation that some operations produce
-the same result when performed independently on subsets of the input
-data, merging the results afterwards. Data level parallelism is the
-basic optimization that is used by popular distributed processing
-systems such as MapReduce\kk{cite} and Apache Spark\kk{cite}.
-
-\subsection{Graph transformations}
-
-\kk{This section could be moved to the intermediate representation,
-  leaving a small section about the iterative process of the planner
-  in the implementation/runtime section}
-
-In our case, this is achieved by iteratively applying graph
-transformations that preserve correctness, while exposing
-parallelization opportunities. We illustrate that with a motivating
-example; consider the following program and its dataflow graph.
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- grep ``foo'' f1 f2 // Add its graph
-\end{lstlisting}
-
-\noindent
-In this example, \texttt{grep} is applied to files f1 and f2, and its
-result is sent to standard output. However, the result would be the
-same if we concatenated the results of \texttt{grep ``foo'' f1} and
-\texttt{grep ``foo'' f2}. This intuition can be generalized to all
-stateless commands, as they are all essentially ``linear'' with
-respect to concatenation and the following equation holds.
-
-\[
-\forall f, \mathrm{stateless}(f) \implies f(x.x') = f(x).f(x')
-\]
-
-\kk{Complete this story with pure, after we have written the section
-  about command categories.}
-
-\kk{Mention the assumptions that need to hold for the graph
-  transformations to be valid. Commands must be deterministic, they
-  must not do any other side effect (such as writing to other files,
-  sending signals, etc). However, these assumptions must already be
-  checked when the developer designates the categories.}
-
-\subsection{Planner Algorithm}
-
-The \texttt{Dish} planner extends the transformations that were
-described above with edge splits in order to allow for the above
-transformations to be performed. More precisely, it splits input files
-and graph edges into several chunks so that they can be processed in
-parallel by the next levels.
-
-\kk{Explain the greedy algorithm that our planner performs. It splits
-  input files, and then pushes parallelization through the graph,
-  starting from the source nodes.}
+\section{Dish Implementation}
+
+\kk{I think that this section should be merged with the front end}
+
+\paragraph{Dish Optimizer}
+
+The \texttt{Dish} optimizer is given a dataflow graph representing a
+shell program and outputs an optimized dataflow graph, where data
+parallelism opportunities have been exploited. More concretely, the
+optimizer starts by splitting input files in several chunks, given the
+desirable output dataflow graph width. It then starts from the source
+nodes of the graph, iteratively performing the graph transformations
+that were described in \Cref{ir:transformations}. If the incoming
+edges of the node are less than the desirable graph width, it performs
+edge splits, and then it performs maximal node splits. The final graph
+can then be implemented by spawning a process for each node, and
+redirecting the outputs according to the graph edges.
 
@@
 
-\subsection{Mapping operators to nodes}
-
-\kk{I think we should actually remove this completely.}
-
-Explain the simple algorithm that we used to minimize intra node
-communication, that tries to map contiguous parts of the graph to the
-same node. We could do this by having an algorithm that minimizes the
-number of cuts or something.
-
-We don't try to make a proper planner, as there is a lot of work on
-operator placement etc. that we can borrow from.
+%% \subsection{Mapping operators to nodes}
+
+%% \kk{I don't think we should mention this.}
+
+%% Explain the simple algorithm that we used to minimize intra node
+%% communication, that tries to map contiguous parts of the graph to the
+%% same node. We could do this by having an algorithm that minimizes the
+%% number of cuts or something.
+
+%% We don't try to make a proper planner, as there is a lot of work on
+%% operator placement etc. that we can borrow from.
 
@@
 
-\kk{Rename this section to panner activators}
+\TODO{Rename this section to planner activators and write it after I
+  discuss with Nikos.}
 
@@
 
-\paragraph{Merge and Split}
-
-\kk{Mergers and splitters, and how they are implemented using shell
-  constructs (cat and head). This subsection might be a better fit in
-  some other section, maybe something related to the planner or the
-  implementation, as it is a very elegant way to implement splitters
-  and mergers while staying completely inside the shell programming
-  model.}
+\subsection{Merge and Split Implementation}
+
+\TODO{Niko, could you make this a bit stronger?}
 
@@
 edges, or many outgoing edges (when all but the last outgoing edges
-are bounded). However some nodes, might read their input from one file
-(e.g. stdin). In this case, the distributed implementation needs to
-concatenate input files, and split output files. Here is an example:
+are bounded). However some nodes, might only read their input from one
+source (e.g. stdin). In this case, the distributed implementation
+needs to concatenate input files (or conversely split output
+files). Here is an example:
 
@@
 
-\kk{This transition needs to be stronger} Luckily, the shell offers
-primitives for file manipulations that can be used to implement a
-merger and a splitter. More concretely a merger can be implemented by
-a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
-and an output file \verb|out|, a merger can be implemented as:
+We address this issue using primitive shell constructs for file
+manipulation. More concretely a merger can be implemented by a simple
+\verb|cat|. Given a set of input files \verb|in1, in2, ...| and an
+output file \verb|out|, a merger can be implemented as:
 
@@
 \noindent
+\TODO{Change that with the new splitter}
 On the other hand, given an input file \verb|in| that is a FIFO pipe,
@@
 arbitrarily many different pipes.
-
-\kk{Do I also need to include a splitter for normal files?}
 

### 2019-11-22 pldi/draft.tex
Updates on commands

--- 2019-11-22 pldi/draft.tex
@@
 
-\newcommand{\pur}{\cn{\textsc{P}}}
-\newcommand{\sta}{\cn{\textsc{S}}}
-\newcommand{\dfs}{\cn{\textsc{F}}}
-\newcommand{\sid}{\cn{\textsc{E}}}
-\newcommand{\irr}{\cn{\textsc{I}}}
-
-\newcommand{\tpur}{\tcn{\textsc{P}}}
-\newcommand{\tsta}{\tcn{\textsc{S}}}
-\newcommand{\tdfs}{\tcn{\textsc{F}}}
-\newcommand{\tsid}{\tcn{\textsc{E}}}
-\newcommand{\tirr}{\tcn{\textsc{I}}}
+\newcommand{\pur}{\cn{\textsc{P}}\xspace}
+\newcommand{\sta}{\cn{\textsc{S}}\xspace}
+\newcommand{\dfs}{\cn{\textsc{F}}\xspace}
+\newcommand{\sid}{\cn{\textsc{E}}\xspace}
+\newcommand{\irr}{\cn{\textsc{I}}\xspace}
+
+\newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
+\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
+\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
+\newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
+\newcommand{\tirr}{\tcn{\textsc{I}}\xspace}
 
@@
 %% Title information
-\title{Distribution-oblivious Programming with Dish}         %% [Short Title] is optional;
+\title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
 % \titlenote{with title note}             %% \titlenote is optional;
@@
 
-\kk{The title is too abstract and doesn't really hint that we
-  distribute shell code. Also we could consider capitalizing the name
-  as DiSh?}
-
 % 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
@@
 
-To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of individual commands involved in them.
+To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
 This understanding is encoded in a command's \emph{distributability} class.
-Put informally, a distributability class representing the level of synchronization required by copies of a command executing concurrently.
-The class captures characteristics that are imprortant for distribution, rather than all the observable behavior of a command.
-\sys leans towards a few, coarse classes rather than many, detailed ones---after all, these classes should be easily understood by command developers.
+Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
+It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
+\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
 % Make the connections that this is like a type, that doesn't capture everything, but is good-enough
 
-The section starts with a distributability study of standard libraries commands, such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
-Building on this study, it presents a DSL for describing a command's distributability characteristics~\sx{ext}.
-In return, \sys uses this DSL to tag commands in the standard libraries and generate the transformations presented in later sections.
+This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
+In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers presented in later sections.
 
@@
 % nv TODO: this is a hierarchy
-Broadly, shell commands can be split into five major classes with respect to their distribution characteristics, depending on how they interact with state.
-Tab.~\ref{tab:classes} summarizes the discussion;
-These classes are listed (and presented later) in ascending difficulty of distribution; later classes require further effort.
-In this order, each class can be thought as a subset of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
+These classes are ordered in ascending difficulty of distribution:
+  later classes require more effort.
+In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
 
@@
 Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
-Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{restart}     &  5  (0.4\%)      &           \\  % 5
+Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)      &           \\  % 5
 % \midrule
@@
 The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
-These are commands expressible as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
+These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
 At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
 Workloads that use only stateless commands are trivial to parallelize:
-  they do not require any synchronization to maintain correctness,
-  they do not need care planning on where to split.
+  they do not require any synchronization to maintain correctness, nor caution about where to split.
  % map :: (a -> b) -> [a] -> [b]
 
-Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time);
-  this could potentially allow further parallelization within a single line element.
-However, exactly because of their local, commands operating on individual characters are not particularly expensive computationally.
+Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
+This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
+However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.
 
@@
 The second class contains pure commands \pur.
-These commands respect functional purity, returning the same results for the same inputs;
-  however, they need to maintain internal state across the entire invocation.
-
-Pure commands often cannot operate in an online fashion, but need to block until the end of the stream.
-A typical example of this is \ttt{sort}, a command that cannot start before the last line of input has been consumed.
-
-Parallelizing these commands is slightly more challenging and depends on the command.
-Some commands have known divide-and-conquer equivalents.
-
-Pure commands~ones that maintain 
-Distribution:
+These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
+The details of this state and its propagation during piecewise element processing affects their distributability characteristics.
+
+Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
+Others, such as \ttt{sha1sum}, maintain more complex state that is not applied commutatively.
+
+Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
+Non-streaming constraints affect task parallelism, but not data parallelism:
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques.
 
 \heading{File-system Commands}
-
-This is the largest class, as the file (and file-system) abstraction is so central to \unix.
-In fact, \unix proxies several file-unrelated operators through the file-system---for example, security in \unix is grafted upon the file-system:
-
-(and all associated commands) are always
-And most of these commands seem control-plane operations that operate on identifiers~\sx{bg}---for example, renaming a file or changing group ownership.
+The third class contains commands that access the file-system~\dfs.
+This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
+In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.
+
 % TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
-These are mostly distributable
-
-Distribution:
+Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
+Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
+The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
+% these operate by changing an inode
+Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
+Operations using different identifiers pose no problems;
+  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
+While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).
+
+% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.
 
 \heading{Side-effectful Commands}
-Another large class of commands are ones that have side-effects to memory.
-Because of timesharing and multi-tasking (many users can operate multiple terminals on single \unix machine), a set of commands are devoted to accessing variables---environment variables, shell parsing flags,
-The different formats have multiplied these commands
-
-Thankfully, the vast majority of these commands (a ratio of 21:4) only \emph{read} values;
-Even better,  usually read values that are not altered by user---\eg \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information.
+Another large class of commands are ones that have side-effects to memory~\sid.
+Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.
+
+A command in \sid will require some form of concurrency control to distribute property.
+However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.
+
+The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
+For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
 Even if \sid commands are protected with transactions
@@
 \heading{Irreversibly Side-effectful Commands}
+Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
+Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
 

### 2019-11-22 pldi/draft.tex
Add intro to eval

@@
 \section{Evaluation}
+\label{eval}
 
@@
   command in them.}
+
+At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
+To achieve this, we use a combination of micro- and macro-benchmarks, most of which were collected from prior literature and online repositories.
+Micro-benchmarks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write (usually interactively);
+  such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
+These pipelines are taken from real, often striking~\cite{}, use cases and have the benefit of highlighting stress on a handful of stages, precisely due to their small size.
+
+Three complex macro-benchmarks (\S\ref{macro1}--\ref{macro3}) involve third-party commands and handle realistic workloads by today's standards.
+These scripts are about an order of magnitude larger than the micro-benchmarks, but this perceived lack of complexity is deceiving:
+  as shown below, if written in conventional languages, these pipelines would correspond to programs on the order of hundreds of lines of code.
+
+Several results are worth highlighting.
+First and foremost, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
+\sys's transformation and planning phase take around XX--XXms, negligible (XX\%) for short-running pipelines and virtually non-existent for long-running ones.
+As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
+Finally, \sys preserves the productivity of the shell:
+  in one case, the equivalent program for a small part of the pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop;
+  in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
+
+Experiments were run on a \todo{network of five workstations} connected by 1Gbps links: 
+ one large machine (\wka) with 512GB of memory and 128 2.1GHz Intel Xeon E5-2683 cores, and 
+ \todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
+For our software setup, we use SMP Debian 4.9.144-3.1 (2019-02-19), GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-release (x86\_64-pc-linux-gnu), Python 3.7.3, and OCaml 4.05.0.
+Generally, no special configuration was made in hardware or software beyond disabling hyper-threading. % ---specifically, the network protocol stack was left unoptimized.
+More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
 

### 2019-11-22 pldi/draft.tex

--- 2019-11-22 pldi/draft.tex
@@
 \newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}
+
+\newcommand{\cn}[1]{\mbox{\textcircled{\footnotesize #1}}}
+\newcommand{\tcn}[1]{\mbox{\textcircled{\scriptsize #1}}}
+
+\newcommand{\pur}{\cn{\textsc{P}}\xspace}
+\newcommand{\sta}{\cn{\textsc{S}}\xspace}
+\newcommand{\dfs}{\cn{\textsc{F}}\xspace}
+\newcommand{\sid}{\cn{\textsc{E}}\xspace}
+\newcommand{\irr}{\cn{\textsc{I}}\xspace}
+
+\newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
+\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
+\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
+\newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
+\newcommand{\tirr}{\tcn{\textsc{I}}\xspace}
 
@@
 \label{intro}
-
-\kk{The title is too abstract and doesn't really hint that we
-  distribute shell code. Also we could consider capitalizing the name
-  as DiSh?}
 
@@
 
+% Additional material, in anonymized form, is available at~\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+
 % \begin{figure*}[t]
@@
 \section{Distributability Classes}
-
-\subsection{Comnands}
-\label{bg:cmd}
-
-% Sets vs. Classes
-
-\begin{table}[h]
+\label{distributability}
+
+To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
+This understanding is encoded in a command's \emph{distributability} class.
+Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
+It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
+\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
+% Make the connections that this is like a type, that doesn't capture everything, but is good-enough
+
+This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
+In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers presented in later sections.
+
+\subsection{Distributability of Standard Libraries}
+\label{cmd}
+
+% nv TODO: this is a hierarchy
+Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
+These classes are ordered in ascending difficulty of distribution:
+  later classes require more effort.
+In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+
+\begin{table}[t]
 \center
@@
     \textbf{Distributability Classes}.
-    Broadly, \unix commands can be broken down into to five classses.
+    Broadly, \unix commands can be broken down into to five classes.
   }
 }
-\begin{tabular}{l @{\extracolsep{\fill}} lll}
+\begin{tabular}{l @{\extracolsep{\fill}} llll}
 \toprule
-Class                           & Example Commands        & POSIX            & Core-utils\\
+Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
 \midrule
-Stateless                       &                         &                  &           \\
-Pure                            &                         &                  &           \\
-(Distr.) File-System            &                         &                  &           \\
-Side-effectful                  &                         &                  &           \\
-Irreversibly Side-effectful     &                         &                  &           \\
+Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
+Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
+(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
+Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
+Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)      &           \\  % 5
 % \midrule
@@
 
-Define distributability 
-Explain distributability classes, based on how
-
-\heading{Preliminaries}
-
-\heading{POSIX, GNU Core-utils, and beyond}
-
-\heading{Stateless}
-
-\heading{Pure}
-
-\heading{File-system}
-
-\heading{Side-effectful}
-
-\heading{Irreversably Side-effectful}
-
-\subsection{The Shell Language}
-\label{bg:shell}
-
-\heading{Implicits}
-Expansion
-Implicits
-Context
-
-\heading{Language Constructs}
-
-Even without these, the Unix pipeline language is powerful enough to capture complex pipelines.
-
-We can encode powerful
-
-\section{Dish Overview}
+% \heading{Preliminaries}
+% \heading{POSIX, GNU Core-utils, and beyond}
+
+\heading{Stateless Commands}
+The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
+These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
+At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
+Workloads that use only stateless commands are trivial to parallelize:
+  they do not require any synchronization to maintain correctness, nor caution about where to split.
+ % map :: (a -> b) -> [a] -> [b]
+
+Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
+This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
+However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.
+
+\heading{Pure Commands}
+The second class contains pure commands \pur.
+These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
+The details of this state and its propagation during piecewise element processing affects their distributability characteristics.
+
+Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
+Others, such as \ttt{sha1sum}, maintain more complex state that is not applied commutatively.
+
+Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
+Non-streaming constraints affect task parallelism, but not data parallelism:
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques.
+
+\heading{File-system Commands}
+The third class contains commands that access the file-system~\dfs.
+This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
+In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.
+
+% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
+Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
+Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
+The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
+% these operate by changing an inode
+Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
+Operations using different identifiers pose no problems;
+  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
+While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).
+
+% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.
+
+\heading{Side-effectful Commands}
+Another large class of commands are ones that have side-effects to memory~\sid.
+Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.
+
+A command in \sid will require some form of concurrency control to distribute property.
+However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.
+
+The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
+For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
+Even if \sid commands are protected with transactions
+  these benefits are even greater for read-only or read-mostly accesses.
+
+\heading{Irreversibly Side-effectful Commands}
+Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
+Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
 

### 2019-11-22 pldi/draft.tex
Limitations

@@
 
-\paragraph{Assumptions}
-
-\kk{This paragraph is not extremely necessary. Maybe it can be moved
-  to limitations?}
-
-There are two additional assumptions that have to be satisfied in
-order for the distributed implementation that is generated by Dish to
-have equivalent behaviour with the original implementation. First of
-all, there should be no external signals to the commands while they
-execute. At the moment Dish assumes no faults, and that communication
-always succeeds. This assumption can be lifted by extending Dish with
-a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
-command reads and writes to arbitrary files that are not mentioned in
-their arguments. This is a necessary assumption, since commands are
-considered as black boxes and the only information that can be
-inferred about them is through their arguments.
-
 \section{Dish Implementation}
@@
 More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
-
-\section{Limitations}
-
-\kk{This is not really a section on its own, but rather part of something else.}
-
-\begin{itemize}
-\item Fault Tolerance
-\item Distributed Planning: The planner at the moment is naive, and
-  there is a huge literature on operator placement of distributed
-  dataflow graphs that can be used to improve it.
-\item Support of shell constructs: At the moment we have been
-  conservative in the shell subsets that we handle, so that we don't
-  introduce any unsafe parallelism. The point of this work was not to
-  be able to handle a complete distributable subset of the shell, but
-  rather a significantly big part of it, to demonstrate that there can
-  be performance benefit from it.
-\item Support distributing different classes: At the moment we are
-  only dealing with stateless and pure commands, that is the ones that
-  are most popular but also can offer the biggest benefits from
-  parallelizing them. However, it would be interesting to explore the
-  distributability of the file system commands, by using a distributed
-  file system. THis would also require considerations about
-  replication, consistency, and failures.
-\item No cycles (multiple commands writing and reading from the same file)
-\item Formal proof that the analysis indeed returns distributable
-  subtrees. Previously there was no formal semantics for shell, but
-  recent work (cite greenberg) allows us to make a formal proof that
-  the analysis is indeed sound.
-\end{itemize}
 
@@
 
-\kk{We need to cite DGSH}
-
-\section{Conclusion}
+\heading{Concurrent and Distributed Shells}
+Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18}.
+However, their work is 
+dgsh~\cite{dagsh:17}.
+
+\section{Discussion}
 \label{discussion}
+
+This section reflects on \sys's design and implementation, touches on future work, and closes the paper.
+
+\heading{Limitations}
+Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct--- despite \sys being a research prototype.
+
+\sys does not adequately handle failures or network partitions in the general sense, which would include re-scheduling passes that were executing on failed replicas.
+Prior work on fault-tolerant distributed stream processing~\cite{} can be of significant aid here.
+
+The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
+Prior work on operator placement of distributed dataflow graphs~\cite{} can be used to improve it.
+
+\sys is conservative in the shell subsets that it handles, so that it doesn't introduce any unsafe parallelism.
+The point of this work was not to be able to handle a complete distributable
+subset of the shell, but rather use significant part of it to demonstrate
+performance benefits.
+
+\heading{Future Work}
+% \label{limitation}
+% \item No cycles (multiple commands writing and reading from the same file)
+There are several worthwhile directions for future work.
+One direction would be a formal study of \sys and specifically a proof that the analysis~\sx{dataflow} does indeed return distributable subtrees.
+Recent work on formalizing the semantics of the POSIX shell~\cite{smoosh:20} make such an attempt possible.
+
+Another direction is to extend \sys to handle other classes outside \sta and \pur.
+These two classes enjoy a certain in quick, one-off pipelines and are the easiest to distribute, but 
+it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).
+This direction would naturally introduce considerations about replication, consistency, and failure recovery.
+
+% There are two additional assumptions that have to be satisfied in
+% order for the distributed implementation that is generated by Dish to
+% have equivalent behaviour with the original implementation. First of
+% all, there should be no external signals to the commands while they
+% execute. At the moment Dish assumes no faults, and that communication
+% always succeeds. This assumption can be lifted by extending Dish with
+% a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
+% command reads and writes to arbitrary files that are not mentioned in
+% their arguments. This is a necessary assumption, since commands are
+% considered as black boxes and the only information that can be
+% inferred about them is through their arguments.
+
+\heading{Conclusion}
+This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
+\sys's insight is that shell pipelines already express stream computations that can be automatically distributed.
+To achieve this, \sys
+  decomposes primitives into distributability classes,
+  identifies high-distributability stages,
+  applies rewriting rules for largest possible subprograms,
+  and orchestrates the execution of the distributed program.
+% Its runtime component provides orchestration and planning support during the execution of the program.
+Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
 

### 2019-11-22 pldi/draft.tex
Small updates

@@
   This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
-  \sys's insight is that shell pipelines already express stream computations that can be automatically distributed.
-  To achieve this, \sys
+  Its insight is that pipelines already express stream computations that can be automatically distributed.
+  To distribute them, \sys
     decomposes primitives into distributability classes,
@@
     applies rewriting rules for largest possible subprograms,
-and orchestrates the distributed program.
-Its runtime component provides orchestration and planning support during the execution of the program.
+    and orchestrates the execution of the distributed program.
 Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
@@
 
+% Its runtime component provides orchestration and planning support during the execution of the program.
 % These are from the abstract:

### 2019-11-22 scripts/max-temp/p2.sh
Break down scripts

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+cat p1.txt |
+  grep gz |
+# head -n 1 | # <-- remove this live for full scale
+  tr -s ' ' |
+  cut -d ' ' -f9 > p2.txt
+
+

### 2019-11-22 pldi/draft.tex

@@
 (meaning in what order), and how to distribute it (in case it is
-pure). \kk{I am not sure if the order in which a command reads its
-  input will be part of the language since many commands can be made
-  to read from one place (like stdin or their first argument). Even in
-  that case though, where the command only reads from one argument, we
-  have to make sure that there is a way for the user to pinpoint which
-  argument does a command read from.}
+pure).
+
+%% \kk{I am not sure if the order in which a command reads its
+%%   input will be part of the language since many commands can be made
+%%   to read from one place (like stdin or their first argument). Even in
+%%   that case though, where the command only reads from one argument, we
+%%   have to make sure that there is a way for the user to pinpoint which
+%%   argument does a command read from.}
 
@@
 
-Consider the command \texttt{cat} \kk{Maybe that is not he best
+Consider the command \ttt{cat} \kk{Maybe that is not he best
   example. TODO: Find a good example of a simple command that is in
@@
 
-%% Can be tricked according to this:
+A record contains a sequence of predicates => categories. Each
+predicate is a combination of not, or, and and the atomic predicates
+are the presence of a flag, or an equality that checks that the value
+of a flag is equal to something (because flags can also have
+values). The final predicate can be left empty, and it is trivially true.
+
+Pythonic range for several inputs/outputs
+
+%% Can be tweaked according to this:
 %% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
@@
 
-Sketch of the language:
-
-Some kind of lightweight annotation configuration language (like
-yaml), that contains a record for each command.
-
-A record contains a sequence of predicates => categories. Each
-predicate is a combination of not, or, and and the atomic predicates
-are the presence of a flag, or an equality that checks that the value
-of a flag is equal to something (because flags can also have
-values). The final predicate can be left empty, and it is trivially true.
-
 Interpreter:
@@
 predicate that matches.
-
-\kk{Would it be meaningful to give the syntaxx of the language
-  formally?}
 
@@
 
-\kk{Question: Do we need to handle all commands (even the ones that
-  could have their input split in several different arguments? or can
-  we somehow easily make these ones just take their input from one
-  place (e.g. by adding a cat before them).}
-
-\kk{Possible solutions for this (if we have already turned commands
-  with more that one input to commands with one input): Have a
-  positional reference to the argument that is input (or ``stdin'' if
-  it is the stdin). Example: the last argument [-1] of grep is its
-  input.}
-
-\kk{It seems that even the arguments might depend on flags (e.g. grep
-  checks the current directory if asked recursively, or the stdin
-  otherwise), so maybe we should again have a sequence of predicates}
-
 \kk{Note: Since a categorization language will probably not be
@@
 \label{ir}
-
-\TODO{Replace texttt with inline listings?}
 
@@
 stream from a sequence of different sources. An illustrative example
-is the command \texttt{cat} which reads its input from a combination
+is the command \ttt{cat} which reads its input from a combination
 of files and its standard input.
@@
 
-The example above reads file \texttt{x} until it encounters the EOF
-character, then reads from \texttt{stdin}, and then reads from
-\texttt{y}. In order to handle this particularity, nodes in the
+The example above reads file \ttt{x} until it encounters the EOF
+character, then reads from \ttt{stdin}, and then reads from
+\ttt{y}. In order to handle this particularity, nodes in the
 dataflow graph can have many incoming edges, that are ordered in a
@@
 concurrently, communicating through the edges. Task level parallelism
-is already exposed by the shell, by using pipes \texttt{|} and the
-background operator \texttt{\&}. Unfortunately, simple task
+is already exposed by the shell, by using pipes \ttt{|} and the
+background operator \ttt{\&}. Unfortunately, simple task
 parallelism leaves a lot to be desired, since scaling a computation
@@
 As mentioned in \Cref{ir}, in order to generate the distributed
-implementation of a shell script, \texttt{Dish} first translates it to
+implementation of a shell script, \ttt{Dish} first translates it to
 a dataflow graph. However, there are several components in shell
@@
 program behaviour, e.g. commands producing an irreversible effect, or
-commands connected with a \texttt{\&\&} operator. In addition, the
+commands connected with a \ttt{\&\&} operator. In addition, the
 fact that the shell is \todo{extremely dynamic}, further increases the
@@
 components that can be executed concurrently, and barriers that
-enforce synchronization. More precisely, \texttt{cat f1 f2} and
-\texttt{grep ``foo''} would already execute concurrently in the
-standard shell, while \texttt{sort f3} waits for their completion
-before being executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and
-\texttt{sort f3} are maximal distributable regions, in the sense that
-they cannot be extended beyond \texttt{\&\&}.
+enforce synchronization. More precisely, \ttt{cat f1 f2} and
+\ttt{grep ``foo''} would already execute concurrently in the
+standard shell, while \ttt{sort f3} waits for their completion
+before being executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and
+\ttt{sort f3} are maximal distributable regions, in the sense that
+they cannot be extended beyond \ttt{\&\&}.
 
 Concretely, a distributable region cannot permeate through the
-following barrier constructs: ;, \&\&, \texttt{||}, function
+following barrier constructs: ;, \&\&, \ttt{||}, function
 definitions, command substitutions, and control flow primitives. On
-the other hand, the pipe operator \texttt{|} and the parallel
+the other hand, the pipe operator \ttt{|} and the parallel
 composition operator \& enable parallelization, and extend
@@
 In the above example\kk{Can we make example not floating?}, while
-\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
+\ttt{sort f1 > g1}, \ttt{sort f2 > g2}, and \ttt{sort -m g1
   g2} are all in different dataflow graphs. The dataflow graph that
-corresponds to the whole program, has two nodes, \texttt{sort f1 > g1
-  \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.
+corresponds to the whole program, has two nodes, \ttt{sort f1 > g1
+  \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.
 
@@
 distributing nodes in the dataflow graph \kk{Does this need an
-  example?}. Whenever the algorithm encounter an \texttt{\&\&} node,
+  example?}. Whenever the algorithm encounter an \ttt{\&\&} node,
 it recursively translates its subcomponents (similarly to the pipe
@@
 
-The \texttt{Dish} optimizer is given a dataflow graph representing a
+The \ttt{Dish} optimizer is given a dataflow graph representing a
 shell program and outputs an optimized dataflow graph, where data

### 2019-11-22 pldi/draft.tex
Write extensibility subsection

@@
 \newcommand{\review}[1]{{\color{red}#1}}
+\newcommand{\tr}[1]{} %% Text and comments for technical report
 
@@
   \textbf{\sys transformation overview.}
-	\sys XYZ
+	\sys XYZ \todo{This is slightly cropped?}
 }
@@
 
-\section{Extensibility}
+\subsection{Extensibility}
 \label{ext}
 
-Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
-% ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
-%    | -h :: ( Sless ) => { Out }                              -- if needed
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-(bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
-  |-h :: (Sless)($$\Rightarrow$$){Out}                              
-\end{lstlisting}
-
-It states that the command defaults to the Pure class.
-It's operation can be thought as a transformation from input streams to output streams.
-It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
-% TODO: check actual man page for how to specify files
-  it writes to the output and error stream rather than a file.
-The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
-  its output is a constant function.
-
-ULTRA DRAFTY:
-
-\kk{Text point: A problem that is specific to the shell is
-  extensibility.}
-
-The major characteristic that makes the shell popular and very useful
-is its extensibility, as also mentioned in \Cref{bg}. Commands change,
-get extended, and each user continuously installs more commands to use
-in their system.
-
-\kk{Text point: This restricts the space of solutions for distribution
-  of shell scripts}
-
-This means that any one-time solution is not satisfactory, since it
-would very easily become obsolete, or it would only handle a limited
-amount of shell scripts.
-
-\kk{Text point: The characteristics of a good solution}
+%% Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
+%% % ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
+%% %    | -h :: ( Sless ) => { Out }                              -- if needed
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%% (bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
+%%   |-h :: (Sless)($$\Rightarrow$$){Out}                              
+%% \end{lstlisting}
+
+%% It states that the command defaults to the Pure class.
+%% It's operation can be thought as a transformation from input streams to output streams.
+%% It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
+%% % TODO: check actual man page for how to specify files
+%%   it writes to the output and error stream rather than a file.
+%% The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
+%%   its output is a constant function.
+
+One of the major characteristics of the shell is its extensibility, as
+also mentioned in \Cref{bg}. Existing commands are modified and
+extended over time and users are free to install custom commands from
+a wide variety of sources.
+%% \kk{Text point: This restricts the space of solutions for
+%%   distribution of shell scripts}
+This means that any one-time solution for categorizing shell commands
+is not satisfactory, since it would very quickly become obsolete,
+being able to support a very limited amount of shell scripts.
+
+%% \kk{Text point: The characteristics of a good solution}
 
@@
 in the sense that adding or changing a command, doesn't invalidate the
-assumptions that the tool makes. In our case, the information about
-the command categories. A developer of a new command, should be able
-to categorize it accordingly, and succinctly describe how the system
-can distribute a command.
-
-\kk{Solution: A lightweight language that can be used by a command
-  developer to categorize it for distribution.}
+existing information on command categories. A developer of a new
+command, should be able to categorize it accordingly, and succinctly
+express the information that the system needs in order to distribute
+it.
+
+%% \kk{Solution: A lightweight language that can be used by a command
+%%   developer to categorize it for distribution.}
+
+\TODO{Make sure that before this point, we mention that our notion of
+  command has one input and one output. Concretely: the abstract
+  notion of a program that we have is programs with one input and one
+  output stream. The command developer must be able to express the
+  input argument of a commmand (or if it is the stdin).  }
 
 We address this issue by designing a lightweight language whose
-purpose is to categorize a command, describes how it reads its input
-(meaning in what order), and how to distribute it (in case it is
-pure).
-
-%% \kk{I am not sure if the order in which a command reads its
-%%   input will be part of the language since many commands can be made
-%%   to read from one place (like stdin or their first argument). Even in
-%%   that case though, where the command only reads from one argument, we
-%%   have to make sure that there is a way for the user to pinpoint which
-%%   argument does a command read from.}
-
-\subsection{Categorization Language}
-
-\kk{Text point: Introduce problem: Most shell commands, belong to
-  different classes depending on their flags, arguments}
-
-Consider the command \ttt{cat} \kk{Maybe that is not he best
-  example. TODO: Find a good example of a simple command that is in
-  different classes depending on the arguments.}. When used without
-any flag it is stateless, etc...
-
-\kk{Text point: Our language must be able to categorize commands to
-  different classes according to their arguments.}
-
-\kk{We should have a crisp point about why we have this categorization
+purpose is to categorize a command and describe the sequence of
+sources from which it reads its input.
+
+%% \subsection{Categorization Language}
+
+%% \kk{Text point: Introduce problem: Most shell commands, belong to
+%%   different classes depending on their flags, arguments}
+
+\TODO{Mention the point of these two paragraphs in the intro?}
+
+Consider the command \ttt{comm}. It takes as input two files and
+implements some form of join. When invoked without any option, it
+produces three-column output. The first column contains lines that
+unique to the first file, the second column contains lines that are
+unique to the second file, and the last column contains lines that
+exist in both files. In the general case \ttt{comm} is pure, since it
+needs to wait until it reads both files completely before it can
+output its results.
+
+However, \ttt{comm} can also be invoked using some options that
+supress the output of any combination of columns. More precisely, the
+option \ttt{-1} suppresses the first column of the output. Similarly,
+\ttt{-2} and \ttt{-3} supress the second and third column. Using both
+\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command becomes stateless
+regarding its first (resp. the second) file argument.
+
+In order to address this, we have designed a categorization language
+that is flexible enough to identify commands to different categories
+given their options. The complete language syntax can be seen in
+\Cref{fig:dsl}.
+
+%% Can be tweaked according to this:
+%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
+
+\begin{figure}
+  \centering
+  \begin{grammar}
+    <option> ::= `-' <string>
+
+    <category> ::= `stateless' | `pure' | ...
+
+    <maybe-int> ::= | <int>
+
+    <arg> ::= `args[' <int> `]'
+
+    <args> ::= <arg>
+    \alt `args[' <maybe-int> `:' <maybe-int> `]'
+
+    <input> ::= `stdin' | <args>
+
+    <inputs> ::= <input>
+    \alt <input> `,' <inputs>
+
+    <output> ::= `stdout' | <arg>
+
+    <option-pred> ::= <option>
+    \alt `value' <option> = <string>
+    \alt `not' <option-pred>
+    \alt <option-pred> `or' <option-pred>
+    \alt <option-pred> `and' <option-pred>
+
+    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
+
+    <predicate> ::= <option-pred> `=>' <assignment>
+
+    <pred-list> ::= `|' <predicate> <pred-list>
+    \alt `|' `otherwise' `=>' <assignment>
+
+    <command> ::= <name> `\{' <pred-list> `\}'
+
+    <command-list> ::= <command>
+    \alt <command> <command-list>
+  \end{grammar}
+  \caption{\todo{Categorization language}}
+  \label{fig:dsl}
+\end{figure}
+
+We assume that the arguments of a command can always be separated to
+options and file arguments (checking if they start with a
+\ttt{-}~\footnote{Note that the single character \ttt{-} is not an
+  option as it is most often used to designate that a command reads
+  from its standard input}). A file written in the language contains a
+record containing a mapping from predicates to assignments for each
+command. Predicates can be simple boolean formulas on the existence
+and the value of all options. Each assignment contains the command
+category, the sequence of its input sources, and its output
+source. Inputs to commands are a sequence of file arguments and its
+standard input. Similarly, the output is either its standard output or
+a file argument. An example record for comm can be seen below:
+
+\begin{lstlisting}[float=h, numbers=none]
+  comm {
+    | -1 and -3 =>
+      (stateless, [args[1]], stdout)
+    | -2 and -3 =>
+      (stateless, [args[0]], stdout)
+    | otherwise =>
+      (pure, [args[0]], stdout)
+  }
+\end{lstlisting}
+
+The semantics of the language is straightforward: given a command, the
+interpreter collects its options and then returns the assignment of
+the first predicate that is satisfied. The final predicate is always
+satisfied. For example the following two invocations of comm would be
+classified as follows:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ comm -13 f1 f2 # => (stateless, f2, stdout)
+ comm f1 f2     # => (pure, f1, stdout)
+\end{lstlisting}
+
+\tr{We don't need to designate the stderror, because we assume that it
+  is never the main output of a command and it will never be used by
+  the input of another command in the pipeline. If someone indeed
+  wants to do this, they can just use a redirect I think to get around
+  it.}
+
+\tr{If we talk about the language it would be good to give some
+  statistics about how many commands (out of the ones that we have
+  classified) can be represented with one, two clauses etc. If we have
+  a different solution (like python function that given a command and
+  its arguments returns the category) then we could talk about how
+  many lines of code it took to categorize all the commands that we
+  did.}
+
+\tr{Note: Since a categorization language will probably not be
+  complete (especially the one showing the input arguments) there
+  should be a backup mechanism, when the language is not expressive
+  enough, like a python function that categorizes the command, and
+  returns its input argument if it is stateless.}
+
+\tr{(Maybe) We should have a crisp point about why we have this categorization
   language, and why don't we just allow someone to write a function in
@@
 
-\kk{New sketch of the language}
-
-We assume that commands satisfy the standard that options and flags
-come first, and then file arguments. The language contains a record
-for each command name, together with a sequence of predicates that map
-to command categories and designation of input output.
-
-A record contains a sequence of predicates => categories. Each
-predicate is a combination of not, or, and and the atomic predicates
-are the presence of a flag, or an equality that checks that the value
-of a flag is equal to something (because flags can also have
-values). The final predicate can be left empty, and it is trivially true.
-
-Pythonic range for several inputs/outputs
-
-%% Can be tweaked according to this:
-%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
-
-\begin{grammar}
-  <flag> ::= `-' <string>
-
-  <category> ::= `stateless' | `pure' | ...
-
-  <arg> ::= `arg[' <int> `]'
-
-  <input> ::= `stdin' | <arg>
-
-  <output> ::= `stdout' | <arg>
-
-  <flag-pred> ::= <flag>
-  \alt `value' <flag> = <string>
-  \alt `!' <flag-pred>
-  \alt <flag-pred> `or' <flag-pred>
-  \alt <flag-pred> `and' <flag-pred>
-
-  <assignment> ::= `(' <category>, <input>, <output> `)'
-
-  <predicate> ::= <flag-pred> `=>' <assignment>
-
-  <pred-list> ::= `|' <predicate> <pred-list>
-  \alt `|' `otherwise' `=>' <assignment>
-
-  <command> ::= <name> `:' <pred-list>
-
-\end{grammar}
-
-\kk{We don't need to designate the stderror, because we assume that it
-  is never the main output of a command and it will never be used by
-  the input of another command in the pipeline. If someone indeed
-  wants to do this, they can just use a redirect I think to get around
-  it.}
-
-\kk{Give an example of a flag that has value and how this could alter
-  category or input etc...}
-
-Interpreter:
-
-Given a command and the set of its arguments, the intepreter returns
-the command category. In does so by returning the category of the first
-predicate that matches.
-
-\kk{If we talk about the language it would be good to give some
-  statistics about how many commands (out of the ones that we have
-  classified) can be represented with one, two clauses etc. If we have
-  a different solution (like python function that given a command and
-  its arguments returns the category) then we could talk about how
-  many lines of code it took to categorize all the commands that we
-  did.}
-
-As described in distributability\ref{}, the abstract notion of a
-program that we have is programs with one input and one output
-stream. The command developer must be able to express the input
-argument of a commmand (or if it is the stdin).
-
-\kk{Note: Since a categorization language will probably not be
-  complete (especially the one showing the input arguments) there
-  should be a backup mechanism, when the language is not expressive
-  enough, like a python function that categorizes the command, and
-  returns its input argument if it is stateless.}
-
-\subsection{Pure Commands}
+\paragraph{Pure Commands}
+
+\TODO{Complete this, giving example of sort and wc}
 
@@
 
-\kk{Can we find a solution for the commands in coreutils?}
-
-\kk{Important TODO: Find at least one more command that can be
-  executed like sort. Does wc work? If there is no pure command that
-  can be easily parallelized other than sort, this story is not very
-  nice.}
+\tr{Can we find a solution for the commands in coreutils?}
 
@@
 
+\TODO{We have to also cite the paper by Konstantinos, Caleb, Rajeev,
+  etc since they also try to do semantics preserving optimizations on
+  dataflow graphs}
+
 At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).

### 2019-11-22 pldi/draft.tex
Selective rewriting

--- 2019-11-22 pldi/draft.tex
@@
 % \end{lstlisting}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+\smallskip
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
 cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
@@
 \end{lstlisting}
-
-\noindent
+\smallskip
+% \noindent
 Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
 \caption{
-  \textbf{\sys high-level schematic.}
-	\sys XYZ
+  \textbf{High-level schematic.}
+  \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
+  execution (right).
 }
+\vspace{-15pt}
 \label{fig:schematic}
 \end{figure}
-
-%      dish      ignis, mozart    hadoop, spark, ...      erlang, elixir
-%       \/           \   /             \     /                \   /
-% auto <-------------------------------------------------------------> manual
-%                      [rewrite (not write) cost]
 
@@
 
-To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation\kk{I am not sure whether stream processing provides information for distributing a computation on its own.
+To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
+The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation
+% FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
 Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
 \sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies \todo{maximal} sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams \kk{Isn't uniq pure?
-I am not sure how we can handle it}, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.
+A \sys-enabled $p_1$ will run \ttt{tr}s in parallel streams,
+  use a mostly-parallel tree of \ttt{sort}s,
+  run \ttt{uniq} mostly in parallel,
+  \ttt{head} in one of the streams,
+  and (trivially) merge streams in \ttt{out}.
 
@@
 It first starts with a background section outlining pipeline concepts and overviewing \sys~\sx{bg}.
-Sections \ref{classes}--\ref{recipes} highlight key contributions:\footnote{
-  This is not precise, we'll need to wait until it matures a bit.
-}
+Sections \ref{distributability}--\ref{recipes} highlight key contributions:
 \begin{itemize}
@@
   \item
-  \S\ref{classes} overviews \sys and introduces several scalability and distributability classes.
+  \S\ref{distributability} overviews \sys and introduces several scalability and distributability classes.
 
   \item
-  \S\ref{rewrite} presents a set of rewriting transformations for identifying maximal pipeline sub-expressions, candidates for scale-out.
+  \S\ref{ir} presents a set of rewriting transformations for identifying maximal pipeline sub-expressions, candidates for scale-out.
 
   \item
-  \S\ref{other} details other concerns, such as distributed state management and extensibility.
+  \S\ref{frontend} details other concerns, such as distributed state management and extensibility.
 \end{itemize}
@@
 \noindent
-\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
+\sys's implementation~\sx{impl} evaluation~\sx{eval} show significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
 After a comparison with related prior work~\sx{related}, the paper closes with a discussion of related work and possible future directions~\sx{discussion}.
-
-% Additional material, in anonymized form, is available at~\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+% \bigskip
+% \begin{quote}
+% \footnotesize
+Additional material, in anonymized form, is available at
+\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+% \end{quote}
 
@@
 
-\section{Background}
+\section{Background and Overview}
 \label{bg}
 
-This section breaks down $p_1$ to presents important background on and challenges
-of shell scripting.
+This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii)) a high-level overview of \sys's design and implementation.
 
@@
 Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
-\begin{wrapfigure}{r}{0.06\textwidth}
-  \vspace{-12pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
-  % \end{center}
-  \vspace{-30pt}
-\end{wrapfigure}
+%\begin{wrapfigure}{r}{0.06\textwidth}
+%  \vspace{-12pt}
+%  % \begin{center}
+%    % \includegraphics[width=0.48\textwidth]{gull}
+%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
+%  % \end{center}
+%  \vspace{-30pt}
+%\end{wrapfigure}
 This interface exposes a stream of contiguous character lines.
@@
 % Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-
-Commands can be tuned by two mechanisms, namely command-line options and environment variables.
+\smallskip
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+  ls -l | wc -l                          [($$p_2$$)]
+\end{lstlisting}
+\smallskip
+Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
 Options control the execution of a program---\eg changing the \ttt{sort}ing order.
 The shell does not have any visibility into these options; 
-  their parsing and evaluation is left entirely up to each command \kk{A minor point: the shell does the expansion, and then hands it of to the command (I am saying that because evaluation might be confounded with expansion so we might want to clarify that)}.
+  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
 Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
@@
   files are simply persistent streams that can be manipulated by commands or pipelines.
-\begin{wrapfigure}{r}{0.15\textwidth}
-  \vspace{-15pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-  % \end{center}
-  \vspace{-25pt}
-\end{wrapfigure}
+% \begin{wrapfigure}{r}{0.15\textwidth}
+%   \vspace{-15pt}
+%   % \begin{center}
+%     % \includegraphics[width=0.48\textwidth]{gull}
+%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
+%   % \end{center}
+%   \vspace{-25pt}
+% \end{wrapfigure}
 Reading a file or a set of files generates a stream; 
   similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
+Files are named, in the sense that they correspond to a global identifier---for example, \ttt{/x/y} corresponds to a persistent stream.
 \unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
@@
 % (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-
+\smallskip
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
+\end{lstlisting}
+\smallskip
+
+\noindent
 Special files are particularly interesting because they are resolved to OS structures.
@@
   for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
-\begin{wrapfigure}{r}{0.15\textwidth}
-  \vspace{-15pt}
-  % \begin{center}
-    % \includegraphics[width=0.48\textwidth]{gull}
-    \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-  % \end{center}
-  \vspace{-25pt}
-\end{wrapfigure}
+% \begin{wrapfigure}{r}{0.15\textwidth}
+%   \vspace{-15pt}
+%   % \begin{center}
+%     % \includegraphics[width=0.48\textwidth]{gull}
+%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
+%   % \end{center}
+%   \vspace{-25pt}
+% \end{wrapfigure}
 As described before, the file can be a FIFO dubbing as input to another command.
@@
 
+\smallskip
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
+\end{lstlisting}
+\smallskip
+
+\noindent
 % Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators 
-
-The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
+The combination of ephemeral file-system identifiers and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
 The most interesting feature of these pipelines is the presence of cycles:
@@
 
+\smallskip
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
+\end{lstlisting}
+\smallskip
+
+\noindent
 Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
@@
 Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
+
+\unix pipelines are surprisingly powerful, a feature that \sys exploits by re-targeting shell scripts during its rewriting passes that target shell.
 
@@
 % \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  ls -l | wc -l                           (($$p_2$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat * | wc -l > out.txt                 (($$p_3$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 %  cat /proc/ >                            (($$p_5$$))
@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
 \caption{
   \textbf{\sys transformation overview.}
-	\sys XYZ
+  Given a shell pipeline, \sys leverages command distributability to identify
+  high-distributability stages, rewrite them, and orchestrate their execution.
 }
@@
 
-\subsection{Overview}
+\subsection{\sys Design Overview}
 
@@
 Intuitively,
- (i) the parallel-composition operator exposes \todo{task} parallelism \kk{I have named that operator level parallelism, which one do we prefer?},
+ (i) the parallel-composition operator exposes \todo{task} parallelism
+% FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
  (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator enables both task and data parallelism, and \kk{How does it enable data parallelism?}
+ (iii) the pipe operator is candidate for both task and data parallelism, and
+% \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
  (iv)  other composition operators can be viewed as barrier candidates.
@@
 
-Understanding the shell's primitives is not enough;
-  \sys has to understand the distribution characteristics of individual commands used to compose larger programs.
+% However, understanding the shell's primitives is not enough
+\sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
 This challenge can be broken down into two parts.
@@
 
-Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
+Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
 \sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying \todo{maximal} distributable sub-expressions.
-It converts the AST to a dataflow graph, and runs...
-This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc
-
-\kk{On the dataflow graph, Dish iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.}
-
-Finally, given a distributed pipeline with runtime \todo{annotations}\kk{the runtime just gets a dataflow graph}, \sys needs to schedule and execute it.
-This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
+It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
+These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc
+
+Finally, given a distributed pipeline as a dataflow graph, \sys needs to schedule and execute it.
+To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
 First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
@@
 This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
-From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.
-\kk{I am not exactly sure I understand that :) }
-
-The next few sections discuss the details.
+From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
+Planner activators achieves the latest possible binding, when critical runtime information is available.
+% FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}
+
+The next few sections discuss \sys's design and implementation details.
 They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
@@
 Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
-In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers presented in later sections.
+In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.
 
@@
 }
-\begin{tabular}{l @{\extracolsep{\fill}} llll}
+\begin{tabular}{l @{\extracolsep{\fill}} lll}
 \toprule
-Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
+Class                           &  Key    & Example Commands                            & Coreutils       \\ % & POSIX       \\
 \midrule
-Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
-Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
-(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
-Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
-Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)      &           \\  % 5
+Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)    \\ %  &           \\  % 22
+Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)    \\ %  &           \\  % 21
+(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)    \\ %  &           \\  % 27
+Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)      \\ %  &           \\  % 25
+Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     \\ %  &           \\  % 5
 % \midrule
@@
 
-\section{Extensibility}
+\heading{Discussion}
+As described earlier~\sx{bg}, shell commands change their operation based on command-line options.
+These options can actually change the class in which a command belongs.
+Consider two examples:
+  (i) \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter;
+  (ii) \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
+% TODO: closure?
+
+\subsection{Extensibility}
 \label{ext}
@@
 optimizations, we formalize our intuition about stateless and pure
-commands. Recall the example \ref{} from section Categories
+commands.
+\nv{TODOs
+  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me);
+  (ii) try to add a few paragraphs / subsections on parallelization optimizations that make use of these formalisms (to show their value);
+  (iii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
+}
+ Recall the example \ref{} from section Categories
 \ref{}. This example shows the stateless command x being applied to
@@
 \section{Front End}
+\label{frontend}
 
@@
 
-\section{Dish Implementation}
-
+\section{Implementation}
+\label{impl}
 \kk{I think that this section should be merged with the front end}

### 2019-11-22 pldi/draft.tex

--- 2019-11-22 pldi/draft.tex
@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
 \caption{
   \textbf{\sys transformation overview.}
-	\sys XYZ \todo{This is slightly cropped?}
+  Given a shell pipeline, \sys leverages command distributability to identify
+  high-distributability stages, rewrite them, and orchestrate their execution.
 }
@@
 Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
+
+\heading{Discussion}
+As described earlier~\sx{bg}, shell commands change their operation based on command-line options.
+These options can actually change the class in which a command belongs.
+Consider two examples:
+  (i) \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter;
+  (ii) \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
+% TODO: closure?
 

### 2019-11-22 pldi/draft.tex
Pass over extensibility

@@
 }
+\vspace{-15pt}
 \label{fig:overview}
@@
 Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying \todo{maximal} distributable sub-expressions.
+\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying \todo{maximal} distributable sub-expressions.
 It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
@@
 
-Finally, given a distributed pipeline as a dataflow graph, \sys needs to schedule and execute it.
+Finally, \sys needs to schedule and execute the distributed dataflow graph.
 To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
@@
 \heading{Discussion}
-As described earlier~\sx{bg}, shell commands change their operation based on command-line options.
-These options can actually change the class in which a command belongs.
-Consider two examples:
-  (i) \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter;
-  (ii) \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
-% TODO: closure?
+There are several technical details about \sys's design decisions worth noting.
+
+In terms of inputs, \sys supports commands with multiple input (resp. output) streams, by prepending (resp. appending) shell built-ins that merge input (resp. split output) streams~\sx{impl}.
+Thus, a command can be modelled as having one input and one output stream, a representation that is used throughout \sys including the extensibility DSL~\sx{ext}.
+
+Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
+Other commands are treated as  non-distributable for now~\sx{conclusion};
+  by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.
+Even with only these two classes, however, \sys supports highly useful pipelines that are more expressive than ones in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
 
@@
 
-One of the major characteristics of the shell is its extensibility, as
-also mentioned in \Cref{bg}. Existing commands are modified and
-extended over time and users are free to install custom commands from
-a wide variety of sources.
+% FIXME: \nv{TODO: remember to bubble up(me)}
+An important characteristic of the \unix shell is its language-agnostic extensibility~\sx{bg}.
+Users composing pipelines are free to install custom commands available from various sources.
+Being general programs, commands are developed in a variety of languages and are extended over long periods of time.
 %% \kk{Text point: This restricts the space of solutions for
 %%   distribution of shell scripts}
-This means that any one-time solution for categorizing shell commands
-is not satisfactory, since it would very quickly become obsolete,
-being able to support a very limited amount of shell scripts.
-
+As a result, any static characterization of shell commands is unsatisfactory as it would quickly become obsolete, supporting only a small subset of commands.
+This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
 %% \kk{Text point: The characteristics of a good solution}
 
-A solution that applies to shell programms should thus be extensible,
-in the sense that adding or changing a command, doesn't invalidate the
-existing information on command categories. A developer of a new
-command, should be able to categorize it accordingly, and succinctly
-express the information that the system needs in order to distribute
-it.
-
-%% \kk{Solution: A lightweight language that can be used by a command
-%%   developer to categorize it for distribution.}
-
-\TODO{Make sure that before this point, we mention that our notion of
-  command has one input and one output. Concretely: the abstract
-  notion of a program that we have is programs with one input and one
-  output stream. The command developer must be able to express the
-  input argument of a commmand (or if it is the stdin).  }
-
-We address this issue by designing a lightweight language whose
-purpose is to categorize a command and describe the sequence of
-sources from which it reads its input.
-
-%% \subsection{Categorization Language}
-
-%% \kk{Text point: Introduce problem: Most shell commands, belong to
-%%   different classes depending on their flags, arguments}
-
-\TODO{Mention the point of these two paragraphs in the intro?}
-
+To address additions and changes to the set of commands available to the shell, \sys provides command developers lightweight, domain-specific language (DSL).
+The DSL allows developers to express semantic information about a command's distributanility that is critical to \sys.
+This DSL can be used by the developers of new commands succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
+It is also used by developers that work on existing commands to extend or maintain them, to express additions or changes to the command's semantics.
+
+\heading{Command-line Options}
+As described earlier~\sx{bg}, command-line options and environment variables can affect the operation of a command.
+Options are a particularly popular way to control a command, 
+  directly affecting a command's distributability classification.
+That is, commands are assigned to default classes, and commands can change the class in which a command belongs.
+
+As simple examples, consider \ttt{cat} and \ttt{sort}.
+By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
+Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta.
+% TODO: closure?
+
+\heading{Command Inputs and Outputs}
 Consider the command \ttt{comm}. It takes as input two files and
@@
 However, \ttt{comm} can also be invoked using some options that
-supress the output of any combination of columns. More precisely, the
+suppress the output of any combination of columns. More precisely, the
 option \ttt{-1} suppresses the first column of the output. Similarly,
@@
 regarding its first (resp. the second) file argument.
+
+% \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
 
@@
   \end{grammar}
-  \caption{\todo{Categorization language}}
+  \caption{
+\textbf{Distributability description language.}
+\todo{The DSL attempts to capture important elements regarding the distributability aspects of various commands.}
+\vspace{-15pt}
+}
+
   \label{fig:dsl}

### 2019-11-22 pldi/draft.tex
Massive cleanup

@@
   \item
-  \S\ref{frontend} details other concerns, such as distributed state management and extensibility.
+  \S\ref{impl} details other concerns, such as distributed state management and extensibility.
 \end{itemize}
@@
 \noindent
-\sys's implementation~\sx{impl} evaluation~\sx{eval} show significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
+\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
 After a comparison with related prior work~\sx{related}, the paper closes with a discussion of related work and possible future directions~\sx{discussion}.
@@
 
+%% Can be tweaked according to this:
+%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
+
+\begin{figure}
+  \centering
+  \begin{grammar}
+    <option> ::= `-' <string>
+
+    <category> ::= `stateless' | `pure' | ...
+
+    <maybe-int> ::= | <int>
+
+    <arg> ::= `args[' <int> `]'
+
+    <args> ::= <arg>
+    \alt `args[' <maybe-int> `:' <maybe-int> `]'
+
+    <input> ::= `stdin' | <args>
+
+    <inputs> ::= <input>
+    \alt <input> `,' <inputs>
+
+    <output> ::= `stdout' | <arg>
+
+    <option-pred> ::= <option>
+    \alt `value' <option> = <string>
+    \alt `not' <option-pred>
+    \alt <option-pred> `or' <option-pred>
+    \alt <option-pred> `and' <option-pred>
+
+    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
+
+    <predicate> ::= <option-pred> `=>' <assignment>
+
+    <pred-list> ::= `|' <predicate> <pred-list>
+    \alt `|' `otherwise' `=>' <assignment>
+
+    <command> ::= <name> `\{' <pred-list> `\}'
+
+    <command-list> ::= <command>
+    \alt <command> <command-list>
+  \end{grammar}
+  \caption{
+\textbf{Distributability description language.}
+\todo{The DSL attempts to capture important elements regarding the distributability aspects of various commands.}
+\vspace{-20pt}
+}
+
+  \label{fig:dsl}
+\end{figure}
+
 \heading{Command Inputs and Outputs}
@@
 
-In order to address this, we have designed a categorization language
-that is flexible enough to identify commands to different categories
-given their options. The complete language syntax can be seen in
-\Cref{fig:dsl}.
-
-%% Can be tweaked according to this:
-%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
-
-\begin{figure}
-  \centering
-  \begin{grammar}
-    <option> ::= `-' <string>
-
-    <category> ::= `stateless' | `pure' | ...
-
-    <maybe-int> ::= | <int>
-
-    <arg> ::= `args[' <int> `]'
-
-    <args> ::= <arg>
-    \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-    <input> ::= `stdin' | <args>
-
-    <inputs> ::= <input>
-    \alt <input> `,' <inputs>
-
-    <output> ::= `stdout' | <arg>
-
-    <option-pred> ::= <option>
-    \alt `value' <option> = <string>
-    \alt `not' <option-pred>
-    \alt <option-pred> `or' <option-pred>
-    \alt <option-pred> `and' <option-pred>
-
-    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
-
-    <predicate> ::= <option-pred> `=>' <assignment>
-
-    <pred-list> ::= `|' <predicate> <pred-list>
-    \alt `|' `otherwise' `=>' <assignment>
-
-    <command> ::= <name> `\{' <pred-list> `\}'
-
-    <command-list> ::= <command>
-    \alt <command> <command-list>
-  \end{grammar}
-  \caption{
-\textbf{Distributability description language.}
-\todo{The DSL attempts to capture important elements regarding the distributability aspects of various commands.}
-\vspace{-15pt}
-}
-
-  \label{fig:dsl}
-\end{figure}
-
-We assume that the arguments of a command can always be separated to
+\heading{Distributability DSL}
+\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is designed to be able to classify a commands into multiple categories given their options.
+
+It assumes that the arguments of a command can always be separated to
 options and file arguments (checking if they start with a
@@
 
-As mentioned in distributability, pure commands can be broken down
-into different categories. We might be able to get ideas about this
-from this paper:
-\url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
-continuation in PLDI 2019.
-
+% As mentioned in distributability, pure commands can be broken down
+% into different categories. We might be able to get ideas about this
+% from this paper:
+% \url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
+% continuation in PLDI 2019.
+% FIXME: \nv{Cite this paper?}
 \tr{Can we find a solution for the commands in coreutils?}
@@
 
-\section{Front End}
-\label{frontend}
-
-\kk{The writing in this section is wandering around without a clear
-  purpose :( }
-
-\kk{It might make sense to merge this section with the one below with
-  the name: From Shell to IR and back}
+\section{Implementation}
+\label{impl}
+% \kk{The writing in this section is wandering around without a clear
+%   purpose :( }
+% 
+% \kk{It might make sense to merge this section with the one below with
+%   the name: From Shell to IR and back}
 
@@
 
-\section{Implementation}
-\label{impl}
-\kk{I think that this section should be merged with the front end}
-
-\paragraph{Dish Optimizer}
+\subsection{Dish Optimizer}
 
@@
 
-\kk{Maybe give an example}
+\nv{Give example, extend}
 
@@
 
-\subsection{Just in time planning}
-
-\TODO{Rename this section to planner activators and write it after I
-  discuss with Nikos.}
-
-\kk{This whole section for just in itme planning, is also relevant for
-  the front end. If the front end is executed just in time, it would
-  have more information, and maybe it could make a sound analysis.}
+\subsection{Planning Activators and Just-in-Time Planning}
+
+\nv{Nikos working on this and next}
+
+% \kk{This whole section for just in itme planning, is also relevant for
+%   the front end. If the front end is executed just in time, it would
+%   have more information, and maybe it could make a sound analysis.}
 
@@
 
-\TODO{Niko, could you make this a bit stronger?}
-
 It is common in the dataflow graph for a node to have many incoming
@@
 
-\kk{We should make sure to measure the time that it takes for the
-  whole process to run (parsing, translating, optimizing, planning)
-  together with the scripts. This will probably be negligible, but we
-  still have to mention it. }
-
-\kk{Also, we have to have one-two handcrafted examples that have more
-  than one pipeline, to show the execution time end to end, as this
-  cannot be shown with the one-pipeline examples since they only have
-  one graph, so no back and forth between shell and dish.}
-
-\kk{Furthermore it would be great to have one distributed example that
-  runs in more than one node (if we have time)}
-
-\kk{Also we have to make sure to measure with and without the final
-  cat of output file, so that we show that this is also negligible.}
-
-\kk{We have to make sure that we repeat that the results are the same
-  in all tests that we run. Corectness, blah blah ...}
-
-\kk{We should have at least some pipelines that have a bad pure
-  command in them.}
+% FIXME
+% \kk{We should make sure to measure the time that it takes for the
+%   whole process to run (parsing, translating, optimizing, planning)
+%   together with the scripts. This will probably be negligible, but we
+%   still have to mention it. }
+% 
+% \kk{Also, we have to have one-two handcrafted examples that have more
+%   than one pipeline, to show the execution time end to end, as this
+%   cannot be shown with the one-pipeline examples since they only have
+%   one graph, so no back and forth between shell and dish.}
+% 
+% \kk{Furthermore it would be great to have one distributed example that
+%   runs in more than one node (if we have time)}
+% 
+% \kk{Also we have to make sure to measure with and without the final
+%   cat of output file, so that we show that this is also negligible.}
+% 
+% \kk{We have to make sure that we repeat that the results are the same
+%   in all tests that we run. Corectness, blah blah ...}
+% 
+% \kk{We should have at least some pipelines that have a bad pure
+%   command in them.}
 
@@
 More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
+
+\subsection{Microbenchmarks: Shell One-liners}
+
+\subsection{Macro-benchmark: Weather Analysis}
+
+\subsection{Macro-benchmark: Web Crawling}
+
+\subsection{Macro-benchmark: Bioinformatics Transform}
 

### 2019-11-22 pldi/draft.tex
A cleaning pass on 3.2

@@
 
-To address additions and changes to the set of commands available to the shell, \sys provides command developers lightweight, domain-specific language (DSL).
+To address additions and changes to the set of commands available to the shell, \sys provides command developers a lightweight, domain-specific language (DSL).
 The DSL allows developers to express semantic information about a command's distributanility that is critical to \sys.
-This DSL can be used by the developers of new commands succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
-It is also used by developers that work on existing commands to extend or maintain them, to express additions or changes to the command's semantics.
+This DSL can be used by the developers of new commands to succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
+It could also be used by developers that extend or maintain existing commands, to express additions or changes to the command's implementation or interface.
 
@@
   directly affecting a command's distributability classification.
-That is, commands are assigned to default classes, and commands can change the class in which a command belongs.
+That is, commands are assigned to default classes, and options can change the class in which they belong.
 
@@
 \textbf{Distributability description language.}
-\todo{The DSL attempts to capture important elements regarding the distributability aspects of various commands.}
+The DSL captures important information regarding the distributability of a command.
 \vspace{-20pt}
@@
 \heading{Command Inputs and Outputs}
+%
 Consider the command \ttt{comm}. It takes as input two files and
-implements some form of join. When invoked without any option, it
+performs a join operation. When invoked without any option, it
 produces three-column output. The first column contains lines that
@@
 unique to the second file, and the last column contains lines that
-exist in both files. In the general case \ttt{comm} is pure, since it
-needs to wait until it reads both files completely before it can
+exist in both files. In the general case \ttt{comm} is in \pur, since
+it needs to wait until it reads both files completely before it can
 output its results.
@@
 \ttt{-2} and \ttt{-3} supress the second and third column. Using both
-\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command becomes stateless
+\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command becomes \sta
 regarding its first (resp. the second) file argument.
@@
 \heading{Distributability DSL}
-\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is designed to be able to classify a commands into multiple categories given their options.
+%
+\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is
+designed to be able to classify a commands into multiple categories
+given their options.
 
@@
 command. Predicates can be simple boolean formulas on the existence
-and the value of all options. Each assignment contains the command
+and the value of any option. Each assignment contains the command
 category, the sequence of its input sources, and its output

### 2019-11-22 pldi/draft.tex
Clean up and pass on 4

@@
 Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
-Others, such as \ttt{sha1sum}, maintain more complex state that is not applied commutatively.
+Others, such as \ttt{sha1sum}, maintain more complex state that has to be updated in a sequential fashion.
 
@@
 Non-streaming constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques.
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques, that is implementing it as a map and a reduce function.
 
@@
 that is used by popular distributed processing systems such as
-MapReduce\kk{cite} and Apache Spark\kk{cite}. \todo{Unfortunately, the
-  dataflow graph model of most of these systems is not expressive
-  enough to represent data parallelism, so the distributed generations
-  that they generate, can not be represented in their dataflow model.}
-\kk{This last sentence is extremely important and we should make it
-  shine.}
+MapReduce\kk{cite} and Apache Spark\kk{cite}.Unfortunately, the
+dataflow graph model of most of these systems is not expressive enough
+to represent data parallelism, so the distributed implementations that
+they perform do not preserve the semantics of their dataflow model.
 
@@
   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me);
-  (ii) try to add a few paragraphs / subsections on parallelization optimizations that make use of these formalisms (to show their value);
-  (iii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
+  (ii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
 }
@@
 
-Based on these equations, we can define a maximal node split operation
-on a stateless node $v$ with $n$ input ordered edges and an output
-edge to node $v'$, to replace $v$ with $n$ new nodes, giving each one
-of them one of the incoming edges of $v$, and routing them in order to
-point to $v'$. The maximal node split operation can similarly extended
-for pure nodes.
-
-\kk{If there is time I can work out a formal definition and a proof
+\heading{Parallelization Transformations}
+%
+Based on these equations, we can define a maximal \emph{node split
+  operation} on a stateless node $v$ with $n$ input ordered edges and
+an output edge to node $v'$, to replace $v$ with $n$ new nodes, giving
+each one of them one of the incoming edges of $v$, and routing them in
+order to point to $v'$. Since each incoming edge represents a stream
+of data $x_i : D^*$, and the only behaviour of a dataflow graph is its
+output, this optimization $ v(x_1.x_2...x_n) \Rightarrow
+v(x_1).v(x_2)...v(x_n)$ can be shown to preserve the behaviour of the
+graph. Furthermore, it can be straightforwardly extended to pure
+commands that can be implemented by a map-reduce pair $(m, r)$ as $
+v(x_1.x_2...x_n) \Rightarrow r(m(x_1).m(x_2)...m(x_n))$.
+
+\tr{Remember to mention the assumptions that need to hold for the
+  graph transformations to be valid in the Command categories
+  section. Commands must be deterministic, they must not do any other
+  side effect (such as writing to other files, sending signals,
+  etc). However, these assumptions must already be checked when the
+  developer designates the categories.}
+
+\tr{If there is time I can work out a formal definition and a proof
   sketch why this transformation preserves the output of the dataflow
-  graph}.
+  graph.}
 
@@
 to a file.
-
-\kk{Remember to mention the assumptions that need to hold for the
-  graph transformations to be valid in the Command categories
-  section. Commands must be deterministic, they must not do any other
-  side effect (such as writing to other files, sending signals,
-  etc). However, these assumptions must already be checked when the
-  developer designates the categories.}
 

### 2019-11-22 pldi/draft.tex
Updates

--- 2019-11-22 pldi/draft.tex
@@
 
-\kk{On our dataflow graph model: Our data flow graph model is
+At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
+\caption{
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
+}
+\label{fig:spectrum}
+\end{figure}
+
+% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
+\heading{Automated Parallelization and Distribution}
+There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
+
+% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
+% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
+% While they Moreover, these systems take advantage of functional purity;
+%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
+A plethora of systems assist in the construction of distributed software.
+Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
+% simplify many of the problems of distribution and
+provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
+However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
+% TODO: \TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
+
+\heading{Annotation-based Distribution}
+Recent systems for light-touch distribution, Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, give developers the ability to express through annotations;
+
+Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
+This is different
+
+Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
+\sys is designed to be used, so it is somewhat more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared toward the deveopers of these components---not the users composing them.
+As such..
+Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
+The side-effects present in the shell dwarf those of functional languages
+The shell is also more general.
+
+\heading{DSLs for Distribution}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+simplify certain patterns, 
+but do not offer the flexibility of a full-fledged environment.
+% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
+Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
+
+More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
+These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
+%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
+Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
+%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
+
+\heading{Concurrent and Distributed Shells}
+\nv{NV}
+Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18};
+Their work does not automatically parallelize pipelines, but rather argues that the shell language is itself an good DSL for concurrency;
+This argument is different from \sys, 
+an argument complementary to \sys's which automatically (and correcctly) parallelizes pipelines even when developers have---hence distribution-\emph{oblivious} shell scripting.
+
+Several shells add support for non-linear pipe topologies;
+  this 
+2dsh~\cite{}
+rc~\cite{}
+gsh~\cite{}
+dgsh~\cite{dagsh:17}.
+
+Commands for explicitly specifying parallelism such \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+These commands require the user to express the program explicitely, work only with embarrasingly parallel (and short) programs;
+  \sys and scales out to large pipelines
+Often, these commands provide options to support very special cases\footnote{
+---a stark contradiction to the celebrated \unix philosophy.
+For example, \tti{parallel}, an official GNU project, flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs that
+with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
+
+\heading{Dataflow Graph Model}
+
+\nv{needs KM}
+On our dataflow graph model: Our data flow graph model is
   different than most distributed batch and stream processing systems,
@@
   of the dataflow graph as done in the other systems (the last bit is
-  too aggressive). }
+  too aggressive).
 
@@
   dataflow graphs}
-
-At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
-\caption{
-  \textbf{Cost of Manual Effort.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-A more complete picture is presented in the related-work section~\sx{rt}.
-}
-\label{fig:spectrum}
-\end{figure}
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Automated Parallelization and Distribution}
-There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
-
-% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
-% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
-% While they Moreover, these systems take advantage of functional purity;
-%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-% simplify many of the problems of distribution and
-provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
-However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
-\TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
-
-\heading{Annotation-based Distribution}
-Ignis and Mozart
-
-\heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
-simplify certain patterns, 
-but do not offer the flexibility of a full-fledged environment.
-% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
-Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
-%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
-Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
-
-\heading{Concurrent and Distributed Shells}
-Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18}.
-However, their work is 
-dgsh~\cite{dagsh:17}.
 

### 2019-11-22 pldi/draft.tex
Fix make

@@
 Often, these commands provide options to support very special cases\footnote{
----a stark contradiction to the celebrated \unix philosophy.
+---a stark contradiction to the celebrated \unix philosophy.}
 For example, \tti{parallel}, an official GNU project, flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs that

### 2019-11-22 pldi/draft.tex
Fix

--- 2019-11-22 pldi/draft.tex
@@
   \sys and scales out to large pipelines
-Often, these commands provide options to support very special cases\footnote{
+Often, these commands provide options to support very special cases
 ---a stark contradiction to the celebrated \unix philosophy.

### 2019-11-22 pldi/draft.tex
Add figs

@@
   later classes require more effort.
-In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+In this order, each class can be thought as a subset \kk{I am not sure this holds for all classes.
+I would propose saying that some classes are subsets, e.g....} of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
 
@@
 the first predicate that is satisfied. The final predicate is always
-satisfied. For example the following two invocations of comm would be
-classified as follows:
+satisfied. For example the following two invocations of \ttt{comm}
+would be classified as follows:
 
@@
 
-\paragraph{Pure Commands}
-
-\TODO{Complete this, giving example of sort and wc}
-
-Even though stateless commands are a big percentage of all of them,
-there are a lot of useful and popular pure commands.
+\heading{Pure Commands}
+%
+Together with the commands in \sta, that are always easy to
+parallelize, there also exists a subclass of commands in \pur that can
+be parallelized by separating them into a map and a reduce
+operation. In order to allow command developers to extend \sys with
+information about how to parallelize new pure commands, \sys offers an
+interface to specify how to distribute a pure command.
+
+\TODO{I am not sure a general interface is so easy to design. It needs
+  more though. It might be beneficial to just talk about sort and wc
+  here and how we implemented them and nothing more. Or maybe this
+  could then go to the implementation? Or maybe say that one can write
+  a python function that given a node of the graph, transforms it into
+  many. I am not sure what is best...}
 
@@
 
-On the other hand, some forms of data parallelism can be exposed when
-knowing the size of the input files. As mentioned in \ref{} some pure
-commands (such as cat -n) only need line information to become
-stateless, and knowing the size of a file could allow the system to
-split it in different chunks that can be processed independently. To
-account for that, edges that refer to an input resource contain the
-number of lines of the file that they refer to.
+\tr{On the other hand, some forms of data parallelism can be exposed
+  when knowing the size of the input files. As mentioned in \ref{}
+  some pure commands (such as cat -n) only need line information to
+  become stateless, and knowing the size of a file could allow the
+  system to split it in different chunks that can be processed
+  independently. To account for that, edges that refer to an input
+  resource contain the number of lines of the file that they refer
+  to.}
 
@@
 
+\begin{figure*}[t]
+    \centering
+    %% \subcaptionbox{\label{eval:minimal_grep}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+    %% }
+    %% \subcaptionbox{\label{eval:grep}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
+    %% }
+    %% \subcaptionbox{\label{eval:minimal_sort}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
+    %% }
+    %% \subcaptionbox{\label{eval:topn}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
+    %% }
+    %% \subcaptionbox{\label{eval:wf}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
+    %% }
+    %% \subcaptionbox{\label{eval:spell}}{
+    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+    %% }
+    %% \caption{a}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:minimal_grep}
+    \end{subfigure}%
+    ~
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:grep}
+    \end{subfigure}%
+    ~
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:minimal_sort}
+    \end{subfigure}%
+
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:topn}
+    \end{subfigure}%
+    ~
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:wf}
+    \end{subfigure}%
+    ~
+    \begin{subfigure}[b]{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:spell}
+    \end{subfigure}%
+    \caption{todo}
+    \label{fig:microbenchmarks}
+\end{figure*}
+
 \subsection{Macro-benchmark: Weather Analysis}

### 2019-11-22 pldi/draft.tex
Pass on 4

@@
 script, it first has to be translated to a manipulable
-representation. As mentioned in \ref{}, the two main abstractions of
-the shell, are files that contain data, and commands that communicate
-through these files. We propose a slight variation of the standard
-dataflow graph model, that is commonly found in stream processing
-systems, as the representation of a distributed shell computation. The
-main benefit of the dataflow graph model is that it clearly exposes
-task parallelism, as different nodes can be independent processing
-workers. In our case each node represents a command, and each edge
-represents a file.
+representation. As mentioned in \Cref{bg:pipelines}, the two main
+abstractions of the shell, are files that contain data, and commands
+that communicate through these files. We propose a slight variation of
+the standard dataflow graph model, that is commonly found in stream
+processing systems, as the representation of a distributed shell
+computation. The main benefit of the dataflow graph model is that it
+clearly exposes task parallelism, as different nodes can be
+independent processing workers. In our case each node represents a
+command, and each edge represents a file.
 
@@
 and will be omitted when obvious. An example of a node is the
-following command \kk{Does it need ref? Also maybe this example is too
-  trivial and can be removed.}:
+following command:
 
@@
 system, or just be FIFO pipes that used for interprocess communication
-\TODO{Are there any other file types, such as URLs, ...?}.
+%% \TODO{Are there any other file types, such as URLs, ...?}.
 
@@
 
-\TODO{Add an example of a command that has two outputs and one goes to
-  one command, while another output goes to a different one. Maybe
-  only show the graph here and not the command itself.}
+\TODO{Add an example of a command that has two outputs, one of which
+  goes to one command, while another goes to a different
+  command. Maybe only show the graph here and not the command itself.}
 
@@
 unified reasoning and implementation of the unoptimized and optimized
-version of the dataflow graph. \kk{This is exciting, and maybe it
-  should be moved up (or rephrased in the introduction)}
+version of the dataflow graph. \kk{This is interesting, and should be
+  moved up (or rephrased in the introduction)}
 
 Before describing the different types of parallelization
-optimizations, we formalize our intuition about stateless and pure
-commands.
+optimizations, we formalize the intuition about stateless and pure
+commands that was described in \Cref{cmd}.
 \nv{TODOs
@@
 }
- Recall the example \ref{} from section Categories
-\ref{}. This example shows the stateless command x being applied to
-...\kk{Complete this after we have an example}. Formally, a commmand
-$f$ is stateless if it is ``linear'' with respect to concatenation,
-i.e. the following equation holds:
+%
+As mentioned there, stateless commands operate independently on
+individual elements of the stream (characters, lines, or files)
+without maintaining any state. Some examples include \ttt{grep},
+\ttt{basename}, and \ttt{tr}. Formally, a commmand $f$ is stateless if
+it is ``linear'' with respect to concatenation, i.e. the following
+equation holds:
 
@@
 
-Similarly, every pure command $f$ that can be written as a map $m$ and a
-reduce $r$ step satisfies the following:
+Similarly, it was mentioned that some pure commands (such as
+\ttt{sort} and \ttt{wc}) can be parallelized using divide-and-conquer
+techniques. More formally, there pure commands $f$ can be implemented
+as a map $m$ and a reduce $r$ function, satisfying the following:
 

### 2019-11-22 pldi/draft.tex
Pass of 5, 5.1, 5.2

@@
 \label{impl}
-% \kk{The writing in this section is wandering around without a clear
-%   purpose :( }
-% 
-% \kk{It might make sense to merge this section with the one below with
-%   the name: From Shell to IR and back}
 
@@
 
-In order to address this issue, we introduce the notion of
+In order to address this issue, we \todo{introduce} the notion of
 distributable program regions, and we design a translation pass that
@@
 graph. The AST of the shell script is produced using a POSIX compliant
-parser \kk{Reference libdash github. This las sentence should probably
-  go to evaluation}.
+parser~\cite{libdash}.
 
@@
 
-%% \kk{I wanted this to be purely about the regions, and a next
-%%   subsection be about the analysis, but the separation line is very
-%%   blurry, so they are now in the same subsection.}
-
-\kk{The goal of this subsection is to explain that Dish does a safe
-  but effective analysis to identify which parts of the scripts to
-  parallelize.}
+%% \kk{The goal of this subsection is to explain that Dish does a safe
+%%   but effective analysis to identify which parts of the scripts to
+%%   parallelize.}
 
@@
 
-Notice that the structure of a shell program clearly identifies
+\noindent
+Notice that the structure of a shell program can be used to identify
 components that can be executed concurrently, and barriers that
-enforce synchronization. More precisely, \ttt{cat f1 f2} and
-\ttt{grep ``foo''} would already execute concurrently in the
-standard shell, while \ttt{sort f3} waits for their completion
-before being executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and
-\ttt{sort f3} are maximal distributable regions, in the sense that
-they cannot be extended beyond \ttt{\&\&}.
-
-Concretely, a distributable region cannot permeate through the
-following barrier constructs: ;, \&\&, \ttt{||}, function
-definitions, command substitutions, and control flow primitives. On
-the other hand, the pipe operator \ttt{|} and the parallel
-composition operator \& enable parallelization, and extend
-distributable regions. Essentially, distributed regions correspond to
-subsections of the program that would be executed by different shell
-processes.
-
-\kk{I don't know whether I should mention the following: While these
+enforce synchronization. More precisely, \ttt{cat f1 f2} and \ttt{grep
+  ``foo''} would already execute concurrently in the standard shell,
+while \ttt{sort f3} would wait for their completion before being
+executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3}
+are maximal distributable regions, in the sense that they cannot be
+extended beyond \ttt{\&\&}.
+
+Concretely, a distributable region cannot permeate through barrier
+constructs (e.g. \ttt{;}, \ttt{\&\&}, and function definitions). On
+the other hand, the pipe operator \ttt{|} and the parallel composition
+operator \ttt{\&} enable parallelization, and extend distributable
+regions. Intuitively, distributed regions correspond to subsections of
+the program that would be allowed to execute concurrently by different
+processes in the POSIX standard~\cite{posix}.
+
+\tr{I don't know whether I should mention the following: While these
   control flow constructs can be parallelized and there has been
@@
 nodes, until a barrier construct is reached. Note that barrier
-constructs can be contained in a dataflow graph, as long as they don't
+constructs can exist in a dataflow graph, as long as they don't
 extend beyong a single node. An illustrative example follows:
@@
 
-In the above example\kk{Can we make example not floating?}, while
-\ttt{sort f1 > g1}, \ttt{sort f2 > g2}, and \ttt{sort -m g1
-  g2} are all in different dataflow graphs. The dataflow graph that
-corresponds to the whole program, has two nodes, \ttt{sort f1 > g1
-  \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.
+In the above example, while \ttt{sort f1 > g1}, \ttt{sort f2 > g2},
+and \ttt{sort -m g1 g2} are all in different dataflow graphs, the
+dataflow graph that corresponds to the whole program, has two nodes,
+\ttt{sort f1 > g1 \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.
 
@@
 translation pass also identifies the category of each node, together
-with its input and output, using the rules described in \ref{}. Some
-interesting cases of the analysis algorithm are shown in pseudocode
-below:
+with its input and output, using records written by developers using
+the language that was described in \Cref{ext}. Some interesting cases
+of the analysis algorithm are shown in pseudocode below:
 
@@
 
-%% \kk{Explain how the three above cases work, mentioning what does
-%%   pipe-graphs and safe do. The important point to get across is to
-%%   explain why we do the ``conservative checks'' in safe, and how they
-%%   preserve safety.}
-
 Whenever the algorithm encounters a pipe node, it recursively
@@
 to. This is important, in order to avoid inconsistent behaviour after
-distributing nodes in the dataflow graph \kk{Does this need an
-  example?}. Whenever the algorithm encounter an \ttt{\&\&} node,
+distributing nodes in the dataflow graph (i.e. by introducing to
+concurrent reads and writes to the same file when
+parallelizing). Whenever the algorithm encounter an \ttt{\&\&} node,
 it recursively translates its subcomponents (similarly to the pipe
-node), but then doesn't merge them in a large dataflow graph. Finally,
-whenever the algorithm encounters a command, it first checks whether
-it is safe to possibly parallelize this command. A command is
-considered to be safe, if it does not make any modification to the
-shell state (e.g. changing the current directory, modifying
-environment variables), and if it does not read from a
-\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
-  unsafe command can still be safely added to the dataflow graph as a
-  single node, since this doesn't introduce any new
-  behaviours. However it cannot be parallelized further than that.} If
-the command is considered safe, its category is identified and a node
-containing that command is created and added to a singleton dataflow
-graph.
-
-It is important to note that due to the highly dynamic nature of the
-shell, the safety properties that were mentioned above cannot be
+node), but then does not merge them in a large dataflow
+graph. Finally, whenever the algorithm encounters a command, it first
+checks whether it is safe to parallelize it. A command is considered
+to be safe, if it does not make any modification to the shell state
+(e.g. changing the current directory, modifying environment
+variables), and if it does not read from any \todo{read-once} file,
+like a fifo pipe. If the command is considered safe, its category is
+identified and a node containing that command is created and added to
+a singleton dataflow graph. If the command is considered unsafe, it is
+still added to the dataflow graph, but cannot be further parallelized
+by the optimizer.
+
+An important consideration is that due to the highly dynamic nature of
+the shell, the safety properties that are mentioned above cannot be
 soundly checked statically. For example in the first case, two
 unexpaned strings could refer to the same file, but the translation
-pass cannot infer that. Hence, in both cases, the checks were
-approximated using a \todo{best-effort} method. \kk{Maybe mention that
-  as future work, we want to tighly integrate with the shell shtepper
-  that will allow to maximally expand everything before making this
-  decision.}
+pass cannot infer that.
 
@@
 \heading{Limitations}
-Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct--- despite \sys being a research prototype.
+Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct\TODO{with respect to some assumptions}--- despite \sys being a research prototype.
 
@@
 
-\sys is conservative in the shell subsets that it handles, so that it doesn't introduce any unsafe parallelism.
-The point of this work was not to be able to handle a complete distributable
-subset of the shell, but rather use significant part of it to demonstrate
-performance benefits.
+\sys is conservative in the shell subsets that it handles, in order to
+not introduce unsafe behaviours. The point of this work was not to be
+able to handle a complete distributable subset of the shell, but
+rather use significant part of it to demonstrate performance benefits.
+
+Shell is highly dynamic, thus making it impossible to design sound
+meaningful analyses to determine distributable regions and maximal
+parallelization. In order to address this particularity, \sys could be
+tighly integrated with a posix compliant shell
+interpreter~\cite{smoosh:20} so that dynamic information can be
+acquired selectively to aid the analysis and optimization process.
 
@@
 % \item No cycles (multiple commands writing and reading from the same file)
-There are several worthwhile directions for future work.
-One direction would be a formal study of \sys and specifically a proof that the analysis~\sx{dataflow} does indeed return distributable subtrees.
-Recent work on formalizing the semantics of the POSIX shell~\cite{smoosh:20} make such an attempt possible.
+There are several worthwhile directions for future work.  One
+direction would be an extension of the distributability analysis of
+\sys, in order to guide it using dynamic information. This could
+enable a formal study of the analysis~\sx{impl} to show that it
+returns regions that are safe to distribute.  Recent work on
+formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
+such an attempt possible.
 

### 2019-11-22 pldi/draft.tex
Minor updates

--- 2019-11-22 pldi/draft.tex
@@
   later classes require more effort.
-In this order, each class can be thought as a subset \kk{I am not sure this holds for all classes.
-I would propose saying that some classes are subsets, e.g....} of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
 
@@
 the first predicate that is satisfied. The final predicate is always
-satisfied. For example the following two invocations of \ttt{comm}
-would be classified as follows:
+satisfied. For example the following two invocations of comm would be
+classified as follows:
 
@@
 
-\heading{Pure Commands}
-%
-Together with the commands in \sta, that are always easy to
-parallelize, there also exists a subclass of commands in \pur that can
-be parallelized by separating them into a map and a reduce
-operation. In order to allow command developers to extend \sys with
-information about how to parallelize new pure commands, \sys offers an
-interface to specify how to distribute a pure command.
-
-\TODO{I am not sure a general interface is so easy to design. It needs
-  more though. It might be beneficial to just talk about sort and wc
-  here and how we implemented them and nothing more. Or maybe this
-  could then go to the implementation? Or maybe say that one can write
-  a python function that given a node of the graph, transforms it into
-  many. I am not sure what is best...}
+\paragraph{Pure Commands}
+
+\TODO{Complete this, giving example of sort and wc}
+
+Even though stateless commands are a big percentage of all of them,
+there are a lot of useful and popular pure commands.
 
@@
 
-\tr{On the other hand, some forms of data parallelism can be exposed
-  when knowing the size of the input files. As mentioned in \ref{}
-  some pure commands (such as cat -n) only need line information to
-  become stateless, and knowing the size of a file could allow the
-  system to split it in different chunks that can be processed
-  independently. To account for that, edges that refer to an input
-  resource contain the number of lines of the file that they refer
-  to.}
+On the other hand, some forms of data parallelism can be exposed when
+knowing the size of the input files. As mentioned in \ref{} some pure
+commands (such as cat -n) only need line information to become
+stateless, and knowing the size of a file could allow the system to
+split it in different chunks that can be processed independently. To
+account for that, edges that refer to an input resource contain the
+number of lines of the file that they refer to.
 
@@
 
-\begin{figure*}[t]
-    \centering
-    %% \subcaptionbox{\label{eval:minimal_grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:minimal_sort}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:topn}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:wf}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:spell}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \caption{a}
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_sort}
-    \end{subfigure}%
-
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:topn}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:wf}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:spell}
-    \end{subfigure}%
-    \caption{todo}
-    \label{fig:microbenchmarks}
-\end{figure*}
+\begin{table}[t]
+\center
+\footnotesize
+% \setlength\tabcolsep{3pt}
+\caption{
+  \footnotesize{
+    \textbf{Summary of Micro-benchmarks}.
+    Six micro-benchmarks used to evaluate \sys.
+  }
+}
+\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} lll}
+\toprule
+  Script                 ~&~ Structure                  & Input Size    & Seq.  Time \\
+\midrule
+  \tti{min-grep}         ~&~$2\times$\tsta              &               &            \\
+  \tti{min-sort}         ~&~~\tpur                      &               &            \\
+  \tti{top-n}            ~&~~\tpur                      &               &            \\
+  \tti{wf}               ~&~~\tpur                      &               &            \\
+  \tti{spell}            ~&~~\tpur                      &               &            \\
+  \tti{diff}             ~&~~\tpur                      &               &            \\
+\bottomrule
+\end{tabular*}
+\label{tab:eval}
+\end{table}
+
+We first apply \sys onto a collection of micro-benchmarks
+
+The first benchmarks runs a grep over a series of complex, DFA
+(Fig.~\cite{})
 
 \subsection{Macro-benchmark: Weather Analysis}
+
+For \sys's first macro-benchmark, we use we
+from introductory chapter of Hadoop Definitive Guide~\cite[Chapter 2]{hadoop:15}
+This example is used as a data analysis Hadoop is well suited for, and combines the 
+There are several interesting aspects about this case.
+First, Hadoop is used 
+McIlroy would have probably expressed---our pipeline 
+14 stages long;
+
+The effort required to write this pipeline is astonishingly low:
+  it amounts to 240 characters 
+(Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.)
+
+% /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
+% hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
 
@@
 At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+
+% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
+\heading{Automated Parallelization and Distribution}
+There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
+
+% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
+% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
+% While they Moreover, these systems take advantage of functional purity;
+%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
+A plethora of systems assist in the construction of distributed software.
+Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
+% simplify many of the problems of distribution and
+provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
+However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
+% TODO: \TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
+
+\heading{Annotation-based Distribution}
+Recent systems for light-touch distribution, Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, give developers the ability to express through annotations;
+
+Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
+This is different
+
+Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
+\sys is designed to be used, so it is somewhat more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared toward the deveopers of these components---not the users composing them.
+As such..
+Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
+The side-effects present in the shell dwarf those of functional languages
+The shell is also more general.
 
@@
 \end{figure}
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Automated Parallelization and Distribution}
-There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
-
-% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
-% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
-% While they Moreover, these systems take advantage of functional purity;
-%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-% simplify many of the problems of distribution and
-provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
-However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
-% TODO: \TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
-
-\heading{Annotation-based Distribution}
-Recent systems for light-touch distribution, Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, give developers the ability to express through annotations;
-
-Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-This is different
-
-Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
-\sys is designed to be used, so it is somewhat more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared toward the deveopers of these components---not the users composing them.
-As such..
-Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
-The side-effects present in the shell dwarf those of functional languages
-The shell is also more general.
 

### 2019-11-22 evaluation/spell_10_seq.sh
Spell experiment

@@ -1,4 +1,4 @@
-cat $IN                     |  # other tricks
+cat $IN $IN $IN $IN $IN $IN $IN $IN $IN $IN                   |  # other tricks
 # groff -t -e -mandoc -Tascii |  # remove formatting commands
 col -bx |                      # remove backspaces / linefeeds
 tr -cs A-Za-z '\n' |

### 2019-11-22 pldi/draft.tex

--- 2019-11-22 pldi/draft.tex
@@
 \subsection{Microbenchmarks: Shell One-liners}
+
+\begin{table}[t]
+\center
+\footnotesize
+% \setlength\tabcolsep{3pt}
+\caption{
+  \footnotesize{
+    \textbf{Summary of Micro-benchmarks}.
+    Six micro-benchmarks used to evaluate \sys.
+  }
+}
+\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} lll}
+\toprule
+  Script                 ~&~ Structure                  & Input Size    & Seq.  Time \\
+\midrule
+  \tti{min-grep}         ~&~$2\times$\tsta              &               &            \\
+  \tti{min-sort}         ~&~~\tpur                      &               &            \\
+  \tti{top-n}            ~&~~\tpur                      &               &            \\
+  \tti{wf}               ~&~~\tpur                      &               &            \\
+  \tti{spell}            ~&~~\tpur                      &               &            \\
+  \tti{diff}             ~&~~\tpur                      &               &            \\
+\bottomrule
+\end{tabular*}
+\label{tab:eval}
+\end{table}
+
+We first apply \sys onto a collection of micro-benchmarks
+
+The first benchmarks runs a grep over a series of complex, DFA
+(Fig.~\cite{})
 
@@
 
+For \sys's first macro-benchmark, we use we
+from introductory chapter of Hadoop Definitive Guide~\cite[Chapter 2]{hadoop:15}
+This example is used as a data analysis Hadoop is well suited for, and combines the 
+There are several interesting aspects about this case.
+First, Hadoop is used 
+McIlroy would have probably expressed---our pipeline 
+14 stages long;
+
+The effort required to write this pipeline is astonishingly low:
+  it amounts to 240 characters 
+(Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.)
+
+% /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
+% hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
+
 \subsection{Macro-benchmark: Web Crawling}

### 2019-11-23 pldi/draft.tex
starting pass

--- 2019-11-22 pldi/draft.tex
@@
 \newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
+\newcommand{\km}[1]{[{\color{blue}#1 --- km}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 
+\km{Test}
+
 \subsection{Graph Components}
@@
 
-\begin{table}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of Micro-benchmarks}.
-    Six micro-benchmarks used to evaluate \sys.
-  }
-}
-\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} lll}
-\toprule
-  Script                 ~&~ Structure                  & Input Size    & Seq.  Time \\
-\midrule
-  \tti{min-grep}         ~&~$2\times$\tsta              &               &            \\
-  \tti{min-sort}         ~&~~\tpur                      &               &            \\
-  \tti{top-n}            ~&~~\tpur                      &               &            \\
-  \tti{wf}               ~&~~\tpur                      &               &            \\
-  \tti{spell}            ~&~~\tpur                      &               &            \\
-  \tti{diff}             ~&~~\tpur                      &               &            \\
-\bottomrule
-\end{tabular*}
-\label{tab:eval}
-\end{table}
-
-We first apply \sys onto a collection of micro-benchmarks
-
-The first benchmarks runs a grep over a series of complex, DFA
-(Fig.~\cite{})
-
 \subsection{Macro-benchmark: Weather Analysis}
-
-For \sys's first macro-benchmark, we use we
-from introductory chapter of Hadoop Definitive Guide~\cite[Chapter 2]{hadoop:15}
-This example is used as a data analysis Hadoop is well suited for, and combines the 
-There are several interesting aspects about this case.
-First, Hadoop is used 
-McIlroy would have probably expressed---our pipeline 
-14 stages long;
-
-The effort required to write this pipeline is astonishingly low:
-  it amounts to 240 characters 
-(Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.)
-
-% /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
-% hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
 
@@
 
-At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Automated Parallelization and Distribution}
-There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
-
-% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
-% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
-% While they Moreover, these systems take advantage of functional purity;
-%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-% simplify many of the problems of distribution and
-provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
-However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
-% TODO: \TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
-
-\heading{Annotation-based Distribution}
-Recent systems for light-touch distribution, Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, give developers the ability to express through annotations;
-
-Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-This is different
-
-Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
-\sys is designed to be used, so it is somewhat more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared toward the deveopers of these components---not the users composing them.
-As such..
-Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
-The side-effects present in the shell dwarf those of functional languages
-The shell is also more general.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
-\caption{
-  \textbf{Cost of Manual Effort.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-A more complete picture is presented in the related-work section~\sx{rt}.
-}
-\label{fig:spectrum}
-\end{figure}
-
-\heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
-simplify certain patterns, 
-but do not offer the flexibility of a full-fledged environment.
-% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
-Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
-%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
-Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
-
-\heading{Concurrent and Distributed Shells}
-\nv{NV}
-Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18};
-Their work does not automatically parallelize pipelines, but rather argues that the shell language is itself an good DSL for concurrency;
-This argument is different from \sys, 
-an argument complementary to \sys's which automatically (and correcctly) parallelizes pipelines even when developers have---hence distribution-\emph{oblivious} shell scripting.
-
-Several shells add support for non-linear pipe topologies;
-  this 
-2dsh~\cite{}
-rc~\cite{}
-gsh~\cite{}
-dgsh~\cite{dagsh:17}.
-
-Commands for explicitly specifying parallelism such \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-These commands require the user to express the program explicitely, work only with embarrasingly parallel (and short) programs;
-  \sys and scales out to large pipelines
-Often, these commands provide options to support very special cases
----a stark contradiction to the celebrated \unix philosophy.
-For example, \tti{parallel}, an official GNU project, flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs that
-with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
-
-\heading{Dataflow Graph Model}
-
-\nv{needs KM}
-On our dataflow graph model: Our data flow graph model is
+\kk{On our dataflow graph model: Our data flow graph model is
   different than most distributed batch and stream processing systems,
@@
   of the dataflow graph as done in the other systems (the last bit is
-  too aggressive).
+  too aggressive). }
 
@@
   dataflow graphs}
+
+At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
+\caption{
+  \textbf{Cost of Manual Effort.}
+	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+A more complete picture is presented in the related-work section~\sx{rt}.
+}
+\label{fig:spectrum}
+\end{figure}
+
+% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
+\heading{Automated Parallelization and Distribution}
+There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
+
+% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
+% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
+% While they Moreover, these systems take advantage of functional purity;
+%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
+A plethora of systems assist in the construction of distributed software.
+Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
+% simplify many of the problems of distribution and
+provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
+However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
+\TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
+
+\heading{Annotation-based Distribution}
+Ignis and Mozart
+
+\heading{DSLs for Distribution}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+simplify certain patterns, 
+but do not offer the flexibility of a full-fledged environment.
+% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
+Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
+
+More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
+These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
+%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
+Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
+%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
+
+\heading{Concurrent and Distributed Shells}
+Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18}.
+However, their work is 
+dgsh~\cite{dagsh:17}.
 

### 2019-11-23 evaluation/spell_2_distr.sh
Spell results

@@ -1,222 +1,66 @@
-mkdir -p /tmp/dish_output/
-rm -f "#file13"
-mkfifo "#file13"
-rm -f "#file43"
-mkfifo "#file43"
-rm -f "#file18"
-mkfifo "#file18"
-rm -f "#file85"
-mkfifo "#file85"
-rm -f "#file42"
-mkfifo "#file42"
-rm -f "#file67"
-mkfifo "#file67"
-rm -f "#file59"
-mkfifo "#file59"
-rm -f "#file47"
-mkfifo "#file47"
-rm -f "#file50"
-mkfifo "#file50"
-rm -f "#file65"
-mkfifo "#file65"
-rm -f "#file87"
-mkfifo "#file87"
-rm -f "#file20"
-mkfifo "#file20"
-rm -f "#file15"
-mkfifo "#file15"
-rm -f "#file31"
-mkfifo "#file31"
-rm -f "#file51"
-mkfifo "#file51"
-rm -f "#file26"
-mkfifo "#file26"
-rm -f "#file37"
-mkfifo "#file37"
-rm -f "#file55"
-mkfifo "#file55"
-rm -f "#file53"
-mkfifo "#file53"
-rm -f "#file57"
-mkfifo "#file57"
-rm -f "#file76"
-mkfifo "#file76"
-rm -f "#file71"
-mkfifo "#file71"
-rm -f "#file45"
-mkfifo "#file45"
-rm -f "#file21"
-mkfifo "#file21"
-rm -f "#file34"
-mkfifo "#file34"
-rm -f "#file78"
-mkfifo "#file78"
-rm -f "#file86"
-mkfifo "#file86"
+rm -rf 1
+mkdir -p 1
+mkdir -p /dev/shm/dish
+rm -f "#file30"
+mkfifo "#file30"
 rm -f "#file22"
 mkfifo "#file22"
-rm -f "#file83"
-mkfifo "#file83"
-rm -f "#file61"
-mkfifo "#file61"
-rm -f "#file32"
-mkfifo "#file32"
-rm -f "#file70"
-mkfifo "#file70"
-rm -f "#file60"
-mkfifo "#file60"
-rm -f "#file38"
-mkfifo "#file38"
-rm -f "#file63"
-mkfifo "#file63"
-rm -f "#file40"
-mkfifo "#file40"
-rm -f "#file39"
-mkfifo "#file39"
-rm -f "#file46"
-mkfifo "#file46"
[…]

### 2019-11-23 evaluation/spell_4_distr.sh

@@ -1,60 +1,60 @@
-rm -rf /tmp/distr_output
-mkdir -p /tmp/distr_output
+rm -rf 1
+mkdir -p 1
 mkdir -p /dev/shm/dish
-rm -f "#file24"
-mkfifo "#file24"
-rm -f "#file25"
-mkfifo "#file25"
-rm -f "#file43"
-mkfifo "#file43"
-rm -f "#file21"
-mkfifo "#file21"
-rm -f "#file30"
-mkfifo "#file30"
-rm -f "#file16"
-mkfifo "#file16"
-rm -f "#file39"
-mkfifo "#file39"
-rm -f "#file18"
-mkfifo "#file18"
-rm -f "#file35"
-mkfifo "#file35"
-rm -f "#file28"
-mkfifo "#file28"
-rm -f "#file38"
-mkfifo "#file38"
-rm -f "#file29"
-mkfifo "#file29"
-rm -f "#file23"
-mkfifo "#file23"
-rm -f "#file42"
-mkfifo "#file42"
-rm -f "#file26"
-mkfifo "#file26"
-rm -f "#file36"
-mkfifo "#file36"
-rm -f "#file22"
-mkfifo "#file22"
-rm -f "#file40"
-mkfifo "#file40"
-rm -f "#file32"
-mkfifo "#file32"
-rm -f "#file27"
-mkfifo "#file27"
 rm -f "#file37"
 mkfifo "#file37"
-rm -f "#file44"
-mkfifo "#file44"
+rm -f "#file38"
+mkfifo "#file38"
 rm -f "#file34"
 mkfifo "#file34"
-rm -f "#file20"
-mkfifo "#file20"
+rm -f "#file29"
+mkfifo "#file29"
+rm -f "#file21"
+mkfifo "#file21"
+rm -f "#file31"
+mkfifo "#file31"
+rm -f "#file22"
+mkfifo "#file22"
+rm -f "#file25"
+mkfifo "#file25"
+rm -f "#file35"
+mkfifo "#file35"
+rm -f "#file44"
+mkfifo "#file44"
+rm -f "#file16"
+mkfifo "#file16"
+rm -f "#file36"
+mkfifo "#file36"
+rm -f "#file28"
+mkfifo "#file28"
+rm -f "#file43"
+mkfifo "#file43"
 rm -f "#file33"
 mkfifo "#file33"
 rm -f "#file41"
 mkfifo "#file41"
-rm -f "#file31"
-mkfifo "#file31"
[…]

### 2019-11-23 evaluation/spell_50_distr.sh
Spell results

@@ -1,0 +1,1218 @@
+rm -rf 1
+mkdir -p 1
+mkdir -p /dev/shm/dish
+rm -f "#file278"
+mkfifo "#file278"
+rm -f "#file157"
+mkfifo "#file157"
+rm -f "#file163"
+mkfifo "#file163"
+rm -f "#file160"
+mkfifo "#file160"
+rm -f "#file165"
+mkfifo "#file165"
+rm -f "#file319"
+mkfifo "#file319"
+rm -f "#file332"
+mkfifo "#file332"
+rm -f "#file340"
+mkfifo "#file340"
+rm -f "#file352"
+mkfifo "#file352"
+rm -f "#file76"
+mkfifo "#file76"
+rm -f "#file343"
+mkfifo "#file343"
+rm -f "#file234"
+mkfifo "#file234"
+rm -f "#file79"
+mkfifo "#file79"
+rm -f "#file296"
+mkfifo "#file296"
+rm -f "#file94"
+mkfifo "#file94"
+rm -f "#file267"
+mkfifo "#file267"
+rm -f "#file184"
+mkfifo "#file184"
+rm -f "#file322"
+mkfifo "#file322"
+rm -f "#file142"
+mkfifo "#file142"
+rm -f "#file179"
+mkfifo "#file179"
+rm -f "#file96"
+mkfifo "#file96"
+rm -f "#file168"
+mkfifo "#file168"
+rm -f "#file290"
+mkfifo "#file290"
+rm -f "#file354"
+mkfifo "#file354"
+rm -f "#file304"
+mkfifo "#file304"
+rm -f "#file258"
+mkfifo "#file258"
+rm -f "#file240"
+mkfifo "#file240"
+rm -f "#file152"
+mkfifo "#file152"
+rm -f "#file326"
+mkfifo "#file326"
+rm -f "#file315"
+mkfifo "#file315"
+rm -f "#file159"
+mkfifo "#file159"
+rm -f "#file185"
+mkfifo "#file185"
+rm -f "#file84"
+mkfifo "#file84"
+rm -f "#file134"
+mkfifo "#file134"
+rm -f "#file124"
+mkfifo "#file124"
+rm -f "#file316"
+mkfifo "#file316"
+rm -f "#file92"
+mkfifo "#file92"
+rm -f "#file291"
+mkfifo "#file291"
+rm -f "#file215"
+mkfifo "#file215"
+rm -f "#file196"
+mkfifo "#file196"
[…]

### 2019-11-23 pldi/draft.tex
Checkpoint to revert

--- 2019-11-22 pldi/draft.tex
@@
   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-These pipelines are taken from real, often striking~\cite{}, use cases and have the benefit of highlighting stress on a handful of stages, precisely due to their small size.
+These pipelines are taken from real use cases (often from the literature~\cite{bentley1985spelling, bentley1985spelling}), and, precisely due to their small size, have the benefit of highlighting stress on a handful of stages.
 
@@
 We first apply \sys onto a collection of micro-benchmarks
+
+Spell program~\cite{bentley1985spelling}
+
+\ttt{wf}
+Word-count~\cite{bentley1986literate}
+
+\ttt{top-n}
 
@@
 \subsection{Macro-benchmark: Weather Analysis}
+\label{macro1}
 
@@
 
+The pipeline starts with a few stages in \sta  that generate URLs;
+  these URLs point to index files, each one of which contains pointers 
+The stream of index files is analyzed (4 stages) to extract the URLs of the compressed data files.
+It is then passed again through a download command
+After downloading and uncompressing these files, the stream is analyzed to extract the maximum temperature.
+
+A less obvious lesson from this case was that  
+First, large, complex pipelines enable significant freedom in how to express some stages.
+Several stages in the original program are compressed into a single \ttt{awk} script;
+  unfortunately, \ttt{awk} is too general to have a helpful distributability signature.
+For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
+  this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
+
+The Hadoop program 
+Hadoop 3.2.1
+openjdk 11.0.5
+ruby 2.5.5p157
+
+few different ways to write some phases, 
+  for example, 
+A large pipeline resulted in a couple of 
+An interesting lesson here was that there are multiple
+First, longer pipelines can be written in multiple ways
+It is worth noting
+These are
+
 % /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
@@
 
+\subsection{Macro-benchmark: Bioinformatics Transform}
+\label{macro3}
+
 \subsection{Macro-benchmark: Web Crawling}
-
-\subsection{Macro-benchmark: Bioinformatics Transform}
+\label{macro2}
 

### 2019-11-23 pldi/draft.tex
Selective replay of km's changes
notation
some references

@@
 \newcommand{\todo}[1]{\hl{#1}\xspace}
-\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
+\newcommand{\nv}[1]{[{\color{cyan}#1 --- nv}]}
 \newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
+\newcommand{\km}[1]{[{\color{blue}#1 --- km}]}
 \newcommand{\review}[1]{{\color{red}#1}}
 \newcommand{\tr}[1]{} %% Text and comments for technical report
+
+\newcommand{\kstar}{^{\textstyle *}}
 
@@
 Nodes $f$ of the graph represent functions from one (possibly empty)
-input stream to an output stream $f : D^* \rightarrow D^*$. $D$
+input stream to an output stream $f : D\kstar \to D\kstar$. $D$
 represents the basic data type of a line of characters. This
@@
 
+\cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but their correctness is established using informal arguments based on operational intution.
+
+The paper \cite{MSAIT2019} proposes a denotational semantic framework for stream processing where the data streams are seen as partial orders, and establishes the soundness of some common parallelizing transformations on dataflow graphs.
+
 \section{Discussion}

### 2019-11-23 pldi/draft.tex
tikz example for drawing dataflow graphs
fixed refs

@@
 \usepackage{syntax}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata,shapes.misc,shapes.geometric,positioning}
 
@@
 graph. Edges are represented as possibly unbounded streams of type
-$D*$. They could either refer to a specific named file in the file
+$D\kstar$. They could either refer to a specific named file in the file
 system, or just be FIFO pipes that used for interprocess communication
@@
 \TODO{Show an example pipe and its graph}.
+
+\begin{center}
+\small
+\begin{tikzpicture}[node distance=2.25cm, ->]
+\node (S) {};
+\node (A) [draw, right of=S, node distance=1.75cm] {node 1};
+\node (B) [draw, right of=A] {node 2};
+\node (C) [draw, right of=B] {node 3};
+\node (E) [right of=C, node distance=1.75cm] {};
+\path (S) edge node[above] {input} (A);
+\path (A) edge node[above] {stream} (B);
+\path (B) edge node[above] {stream} (C);
+\path (C) edge node[above] {output} (E);
+\end{tikzpicture}
+\end{center}
 
@@
 that is used by popular distributed processing systems such as
-MapReduce\kk{cite} and Apache Spark\kk{cite}.Unfortunately, the
+MapReduce \cite{DG2004MR} and Apache Spark\cite{spark:12}. Unfortunately, the
 dataflow graph model of most of these systems is not expressive enough
-to represent data parallelism, so the distributed implementations that
+to represent data parallelism \km{not sure this is true}, so the distributed implementations that
 they perform do not preserve the semantics of their dataflow model.

### 2019-11-23 pldi/draft.tex
Update evaluation

--- 2019-11-23 pldi/draft.tex
@@
 \newcommand{\todo}[1]{\hl{#1}\xspace}
-\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
+\newcommand{\nv}[1]{[{\color{cyan}#1 --- nv}]}
 \newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
+\newcommand{\km}[1]{[{\color{blue}#1 --- km}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 
-% Some (parts of) pipelines require only moderate effort, as they are 
-% For some pipelines, this effort is merely moderate, as 
-%  can be expressed in domain-specific frameworks---for example, Hadoop and Spark provide a few purely functional primitives that 
-%   ---still requiring rewriting the program in a different (\eg Scala for Spark).
-% 
-% This effort is moderate if the program expressed by these pipelines fits into a domain-specific framework.
-% At times, such pipelines
-% As long as the program fits into the model this is only moderately difficult;
-% 
-% For many legacy pipelines  
-%  has to choose between three options, all of which require significant manual effort.
-% The most invasive is to rewrite the entire program in a distributed programming language
-% The most popular is to leverage 
-% 
-% Use a distributed programming language, even higher rewrites
-% Or use some form of annotations 
-% 
-%  this requires rewriting portions of the program in a new language and use operations
-% 
-% 
-% In the simplest case, annotations
-% 
-% Often, they only focus on a few parts of the system---for example, upgrading to a distributed storage layer.
-% More rarely, companies rewrite entire systems (\eg Twitter's Ruby-to-Scala rewrite~\cite{twitter}), a process that is notoriously difficult under schedule constraints and competitive pressures~\cite{rewrite1, rewrite2}.
-%  especially since software today makes extensive use of third-party modules~\cite{breakapp:plos:2017}.
-%
-
-% If developers have already expressed their computation as a Unix pipeline, they should not have to manually rewrite the program in other environment---\eg Hadoop, Spark---to exploit distribution.
-
 To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
@@
 % Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
-
-% Primitives 
-% \sys enables \emph{distribution-oblivious programming}: 
-%   multi-order of magnitude speedups with correctness guarantees and without any developer effort.
-% 
-% The a series of transformations that 
-% Second, we are using describing composition . For this, we are restricting the study to a subset of the POSIX shell--
-% 
-% Local operations and identifiers have to be carefully translated to distributed ones, a translation that depends on the types of the operations and associated identifiers.
-% 
-% The semantics is enough to capture powerful features such as stream branching and feedback constructs found in practice and in the system's evaluation.
 
@@
 
+\begin{table*}[t]
+\center
+\footnotesize
+% \setlength\tabcolsep{3pt}
+\caption{
+  \footnotesize{
+    \textbf{Summary of Micro-benchmarks}.
+    The micro-benchmarks are small pipelines (\ie ones developed on-the-fly) drawn from various sources and applied to large datasets.
+  }
+}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllll}
+\toprule
+  Script                 ~&~ Structure                    & Input (MB)& Time (Seq)    & Script Size (LoC)             & Highlights                                        \\
+\midrule
+% https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
+  \tti{min-grep}         ~&~$3\times$\tsta                &           &               &   17\qquad 134\qquad  1304    & complex NFA regex                                 \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
+  \tti{min-sort}         ~&~~$\tsta, \tpur$               &           &               &   X \qquad 129\qquad  1209    & \tti{sort}ing                                     \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
+  \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ &           &               &   X \qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
+  \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ &           &               &   X \qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
+  \tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ &           &               &   X \qquad  X \qquad          & stream shifting and merging                       \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
+  \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ &           &               &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
+% https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
+  \tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ &           &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\
+\bottomrule
+\end{tabular*}
+\label{tab:eval}
+\end{table*}
+
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
@@
 
-\begin{table}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of Micro-benchmarks}.
-    Six micro-benchmarks used to evaluate \sys.
-  }
-}
-\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} lll}
-\toprule
-  Script                 ~&~ Structure                  & Input Size    & Seq.  Time \\
-\midrule
-  \tti{min-grep}         ~&~$2\times$\tsta              &               &            \\
-  \tti{min-sort}         ~&~~\tpur                      &               &            \\
-  \tti{top-n}            ~&~~\tpur                      &               &            \\
-  \tti{wf}               ~&~~\tpur                      &               &            \\
-  \tti{spell}            ~&~~\tpur                      &               &            \\
-  \tti{diff}             ~&~~\tpur                      &               &            \\
-\bottomrule
-\end{tabular*}
-\label{tab:eval}
-\end{table}
-
-We first apply \sys onto a collection of micro-benchmarks
+Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
+Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
+Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
+Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+
+The size of the script grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+The first number alone is interesting, as it captures \sys's output for the sequential script.
+As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
+% this is the size of 
+This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication---and FIFOs
+are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
+
+All pipelines are set to read and write from the filesystem; 
+  \sys 
+The centralized pipelines read files of various sizes from disk and write to disk.
 
 Spell program~\cite{bentley1985spelling}
+
+Cannot be represented using the Thompson NFA, backtracking
 
@@
 \heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
+At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
 simplify certain patterns, 

### 2019-11-23 pldi/draft.tex
dataflow refs
the prefix order (to define monotonicity)

--- 2019-11-23 pldi/draft.tex
@@
 \newcommand{\kstar}{^{\textstyle *}}
+\newcommand{\eps}{\varepsilon}
 
@@
 
-Nodes $f$ of the graph represent functions from one (possibly empty)
-input stream to an output stream $f : D\kstar \to D\kstar$. $D$
+For a set $D$, we write $D\kstar$ to denote the set of all finite
+words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
+or $xy$ to denote their concatenation. We write $\eps$ for the empty
+word.
+We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
+It is easy to see that $\leq$ is a partial order, and it is often referred to as the \emph{prefix order}.
+
+A node $f$ of the graph represents a function from one (possibly empty)
+input stream to an output stream $f : D\kstar \to D\kstar$, where $D$
 represents the basic data type of a line of characters. This
-representation captures the majority of shell commands. The nodes are
-monotonic, meaning that they cannot retract their output. More
-formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
-$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
-and will be omitted when obvious. An example of a node is the
-following command:
+representation captures the majority of shell commands. We require that
+the function $f$ is monotone w.r.t.\ the prefix order. This captures
+the idea that a node cannot retract output after it has been emitted.
+%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
+%concatenation and will be omitted when obvious.
+An example of a node is the following command:
 
@@
 
+\km{Some very early references on the dataflow model of computation (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN, KMacQ1977}}
+
 \cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but their correctness is established using informal arguments based on operational intution.

### 2019-11-23 pldi/draft.tex
question about model

@@
   our transformations. The assumption is needed due to the limitation
-  of the programming model of one input one output per command.}
+  of the programming model of one input one output per command.}pldi
 
@@
 that is used by popular distributed processing systems such as
-MapReduce \cite{DG2004MR} and Apache Spark\cite{spark:12}. Unfortunately, the
+MapReduce \cite{DG2004MR} and Apache Spark \cite{spark:12}. Unfortunately, the
 dataflow graph model of most of these systems is not expressive enough
@@
 \]
+
+\km{This is a nice characterization, which I think is true in most settings.
+There is a caveat though: You can have stateless programs that emit output when they start (before they consume the stream) and/or when they consume the EOF symbol.
+In this case I am not sure the equation above holds.}
 

### 2019-11-23 pldi/draft.tex
typo fix
tikz samples for data parallelism & shuffle

@@
   our transformations. The assumption is needed due to the limitation
-  of the programming model of one input one output per command.}pldi
+  of the programming model of one input one output per command.}
 
@@
   symmetric.}
+
+\begin{figure*}
+\centering
+\begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
+%
+\small
+%
+\node (M1) {};
+\node (M2) [draw, right of=M1, node distance=2cm] {Map};
+\node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
+\node (M4) [right of=M3, node distance=3cm] {};
+%
+\path (M1) edge node {input} (M2);
+\path (M2) edge node {stream} (M3);
+\path (M3) edge node {output} (M4);
+\end{tikzpicture}
+\end{figure*}
+
+\begin{figure*}
+\newcommand{\Split}{\text{Split}}
+\newcommand{\Merge}{\text{Merge}}
+\newcommand{\Map}{\text{Map}}
+\newcommand{\GroupBy}{\text{GroupBy}}
+\newcommand{\Reduce}{\text{Reduce}}
+\centering
+\begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
+%
+\footnotesize
+%
+\node (M1) {};
+\node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
+\node (M3) [right of=M2, node distance=1.5cm] {};
+\node (M4) [right of=M3, node distance=2.5cm] {};
+\node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
+\node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
+\node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
+\node (M8) [right of=M7, node distance=2cm] {};
+%
+\node (T3) [draw, above of=M3] {$\Map$};
+\node (T4) [draw, above of=M4] {$\Split_K$};
+\node (T5) [draw, above of=M5] {$\Merge$};
+\node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
+%
+\node (B3) [draw, below of=M3] {$\Map$};
+\node (B4) [draw, below of=M4] {$\Split_K$};
+\node (B5) [draw, below of=M5] {$\Merge$};
+\node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
+%
+\path (M1) edge node {$A$} (M2);
+\path (M2) edge[bend left=10] node {$A$} (T3);
+\path (M2) edge[bend right=10] node[swap] {$A$} (B3);
+\path (M5) edge node {$K_2 \times B$} (M6);
+\path (M6) edge (M7);
+\path (M7) edge node {$K \times B$} (M8);
+%
+\path (T3) edge node {$K \times B$} (T4);
+\path (T4) edge (T5);
+\path (T4) edge (M5);
+\path (T4.330) edge[bend right=10] (B5.170);
+\path (T5) edge node {$K_1 \times B$} (T6);
+\path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
+%
+\path (B3) edge node {$K \times B$} (B4);
+\path (B4) edge (B5);
+\path (B4) edge (M5);
+\path (B4.30) edge[bend left=10] (T5.190);
+\path (B5) edge node {$K_3 \times B$} (B6);
+\path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
+%
+\end{tikzpicture}
+\caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
+\label{fig:mapReduce}
+\end{figure*}
 

### 2019-11-23 pldi/draft.tex
More on eval

--- 2019-11-23 pldi/draft.tex
@@
 \usepackage{syntax}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,shapes.misc,shapes.geometric,positioning}
 
@@
 \newcommand{\tr}[1]{} %% Text and comments for technical report
-
-\newcommand{\kstar}{^{\textstyle *}}
-\newcommand{\eps}{\varepsilon}
 
@@
 
-For a set $D$, we write $D\kstar$ to denote the set of all finite
-words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
-or $xy$ to denote their concatenation. We write $\eps$ for the empty
-word.
-We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-It is easy to see that $\leq$ is a partial order, and it is often referred to as the \emph{prefix order}.
-
-A node $f$ of the graph represents a function from one (possibly empty)
-input stream to an output stream $f : D\kstar \to D\kstar$, where $D$
+Nodes $f$ of the graph represent functions from one (possibly empty)
+input stream to an output stream $f : D^* \rightarrow D^*$. $D$
 represents the basic data type of a line of characters. This
-representation captures the majority of shell commands. We require that
-the function $f$ is monotone w.r.t.\ the prefix order. This captures
-the idea that a node cannot retract output after it has been emitted.
-%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
-%concatenation and will be omitted when obvious.
-An example of a node is the following command:
+representation captures the majority of shell commands. The nodes are
+monotonic, meaning that they cannot retract their output. More
+formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
+$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
+and will be omitted when obvious. An example of a node is the
+following command:
 
@@
 graph. Edges are represented as possibly unbounded streams of type
-$D\kstar$. They could either refer to a specific named file in the file
+$D*$. They could either refer to a specific named file in the file
 system, or just be FIFO pipes that used for interprocess communication
@@
 \TODO{Show an example pipe and its graph}.
-
-\begin{center}
-\small
-\begin{tikzpicture}[node distance=2.25cm, ->]
-\node (S) {};
-\node (A) [draw, right of=S, node distance=1.75cm] {node 1};
-\node (B) [draw, right of=A] {node 2};
-\node (C) [draw, right of=B] {node 3};
-\node (E) [right of=C, node distance=1.75cm] {};
-\path (S) edge node[above] {input} (A);
-\path (A) edge node[above] {stream} (B);
-\path (B) edge node[above] {stream} (C);
-\path (C) edge node[above] {output} (E);
-\end{tikzpicture}
-\end{center}
 
@@
 that is used by popular distributed processing systems such as
-MapReduce \cite{DG2004MR} and Apache Spark\cite{spark:12}. Unfortunately, the
+MapReduce\kk{cite} and Apache Spark\kk{cite}.Unfortunately, the
 dataflow graph model of most of these systems is not expressive enough
-to represent data parallelism \km{not sure this is true}, so the distributed implementations that
+to represent data parallelism, so the distributed implementations that
 they perform do not preserve the semantics of their dataflow model.
@@
 
-Experiments were run on a \todo{network of five workstations} connected by 1Gbps links: 
- one large machine (\wka) with 512GB of memory and 128 2.1GHz Intel Xeon E5-2683 cores, and 
- \todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
+Experiments were run on a %\todo{network of five workstations} connected by 1Gbps links: 
+ machine with 512GB of memory and 128 2.1GHz Intel Xeon E5-2683 cores.
+%\todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
 For our software setup, we use SMP Debian 4.9.144-3.1 (2019-02-19), GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-release (x86\_64-pc-linux-gnu), Python 3.7.3, and OCaml 4.05.0.
 Generally, no special configuration was made in hardware or software beyond disabling hyper-threading. % ---specifically, the network protocol stack was left unoptimized.
+All pipelines are set to (initially) read from and (finally) write to the file-system (except as otherwise stated).
+Whenever \ttt{curl} is part of a pipeline, it fetches data from a different physical host on the same network connected by 1Gbps links.
 More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
@@
 
-\begin{table}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of Micro-benchmarks}.
-    Six micro-benchmarks used to evaluate \sys.
-  }
-}
-\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} lll}
-\toprule
-  Script                 ~&~ Structure                  & Input Size    & Seq.  Time \\
-\midrule
-  \tti{min-grep}         ~&~$2\times$\tsta              &               &            \\
-  \tti{min-sort}         ~&~~\tpur                      &               &            \\
-  \tti{top-n}            ~&~~\tpur                      &               &            \\
-  \tti{wf}               ~&~~\tpur                      &               &            \\
-  \tti{spell}            ~&~~\tpur                      &               &            \\
-  \tti{diff}             ~&~~\tpur                      &               &            \\
-\bottomrule
-\end{tabular*}
-\label{tab:eval}
-\end{table}
-
-We first apply \sys onto a collection of micro-benchmarks
-
-Spell program~\cite{bentley1985spelling}
-
-\ttt{wf}
-Word-count~\cite{bentley1986literate}
-
-\ttt{top-n}
-
-The first benchmarks runs a grep over a series of complex, DFA
-(Fig.~\cite{})
+Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
+Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
+% The centralized pipelines read files of various sizes from disk and write to disk.
+
+Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
+Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+
+The size of the script grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+The first number alone is interesting, as it captures \sys's output for the sequential script.
+As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
+% this is the size of 
+This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
+FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
+
+In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
+% a DFA-based regular expression matching phase;
+  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
+The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
+The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
+  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
+The \ttt{bi-gram} pipeline calculates n-grams of base two;
+  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
+\ttt{spell} is based on 
+spell program~\cite{bentley1985spelling} 
+The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
+  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
+
+Generally, the planning
 
@@
 
-\km{Some very early references on the dataflow model of computation (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN, KMacQ1977}}
-
-\cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but their correctness is established using informal arguments based on operational intution.
-
-The paper \cite{MSAIT2019} proposes a denotational semantic framework for stream processing where the data streams are seen as partial orders, and establishes the soundness of some common parallelizing transformations on dataflow graphs.
-
 \section{Discussion}

### 2019-11-23 pldi/draft.tex

@@
 \newcommand{\tr}[1]{} %% Text and comments for technical report
+
+\newcommand{\kstar}{^{\textstyle *}}
+\newcommand{\eps}{\varepsilon}
 
@@
 
-Nodes $f$ of the graph represent functions from one (possibly empty)
-input stream to an output stream $f : D^* \rightarrow D^*$. $D$
+For a set $D$, we write $D\kstar$ to denote the set of all finite
+words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
+or $xy$ to denote their concatenation. We write $\eps$ for the empty
+word.
+We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
+It is easy to see that $\leq$ is a partial order, and it is often referred to as the \emph{prefix order}.
+
+A node $f$ of the graph represents a function from one (possibly empty)
+input stream to an output stream $f : D\kstar \to D\kstar$, where $D$
 represents the basic data type of a line of characters. This
-representation captures the majority of shell commands. The nodes are
-monotonic, meaning that they cannot retract their output. More
-formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
-$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
-and will be omitted when obvious. An example of a node is the
-following command:
+representation captures the majority of shell commands. We require that
+the function $f$ is monotone w.r.t.\ the prefix order. This captures
+the idea that a node cannot retract output after it has been emitted.
+%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
+%concatenation and will be omitted when obvious.
+An example of a node is the following command:
 
@@
 
+\begin{figure*}
+\centering
+\begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
+%
+\small
+%
+\node (M1) {};
+\node (M2) [draw, right of=M1, node distance=2cm] {Map};
+\node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
+\node (M4) [right of=M3, node distance=3cm] {};
+%
+\path (M1) edge node {input} (M2);
+\path (M2) edge node {stream} (M3);
+\path (M3) edge node {output} (M4);
+\end{tikzpicture}
+\end{figure*}
+
+\begin{figure*}
+\newcommand{\Split}{\text{Split}}
+\newcommand{\Merge}{\text{Merge}}
+\newcommand{\Map}{\text{Map}}
+\newcommand{\GroupBy}{\text{GroupBy}}
+\newcommand{\Reduce}{\text{Reduce}}
+\centering
+\begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
+%
+\footnotesize
+%
+\node (M1) {};
+\node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
+\node (M3) [right of=M2, node distance=1.5cm] {};
+\node (M4) [right of=M3, node distance=2.5cm] {};
+\node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
+\node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
+\node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
+\node (M8) [right of=M7, node distance=2cm] {};
+%
+\node (T3) [draw, above of=M3] {$\Map$};
+\node (T4) [draw, above of=M4] {$\Split_K$};
+\node (T5) [draw, above of=M5] {$\Merge$};
+\node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
+%
+\node (B3) [draw, below of=M3] {$\Map$};
+\node (B4) [draw, below of=M4] {$\Split_K$};
+\node (B5) [draw, below of=M5] {$\Merge$};
+\node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
+%
+\path (M1) edge node {$A$} (M2);
+\path (M2) edge[bend left=10] node {$A$} (T3);
+\path (M2) edge[bend right=10] node[swap] {$A$} (B3);
+\path (M5) edge node {$K_2 \times B$} (M6);
+\path (M6) edge (M7);
+\path (M7) edge node {$K \times B$} (M8);
+%
+\path (T3) edge node {$K \times B$} (T4);
+\path (T4) edge (T5);
+\path (T4) edge (M5);
+\path (T4.330) edge[bend right=10] (B5.170);
+\path (T5) edge node {$K_1 \times B$} (T6);
+\path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
+%
+\path (B3) edge node {$K \times B$} (B4);
+\path (B4) edge (B5);
+\path (B4) edge (M5);
+\path (B4.30) edge[bend left=10] (T5.190);
+\path (B5) edge node {$K_3 \times B$} (B6);
+\path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
+%
+\end{tikzpicture}
+\caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
+\label{fig:mapReduce}
+\end{figure*}
+
 \subsection{Graph transformations}
@@
 that is used by popular distributed processing systems such as
-MapReduce\kk{cite} and Apache Spark\kk{cite}.Unfortunately, the
+MapReduce \cite{DG2004MR} and Apache Spark \cite{spark:12}. Unfortunately, the
 dataflow graph model of most of these systems is not expressive enough
-to represent data parallelism, so the distributed implementations that
+to represent data parallelism \km{not sure this is true}, so the distributed implementations that
 they perform do not preserve the semantics of their dataflow model.
@@
 
+\km{Some very early references on the dataflow model of computation (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN, KMacQ1977}}
+
+\cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but their correctness is established using informal arguments based on operational intution.
+
+The paper \cite{MSAIT2019} proposes a denotational semantic framework for stream processing where the data streams are seen as partial orders, and establishes the soundness of some common parallelizing transformations on dataflow graphs.
+
 \section{Discussion}

### 2019-11-23 pldi/draft.tex
Minor comment from discussion

@@
 
+\kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
+  na ginoun formally proven claims gia to equivalence preservation twn
+  optimizations etc.}
+
 \heading{Dataflow Graph Model}
@@
 
-\TODO{We have to also cite the paper by Konstantinos, Caleb, Rajeev,
-  etc since they also try to do semantics preserving optimizations on
-  dataflow graphs}
+%% \TODO{We have to also cite the paper by Konstantinos, Caleb, Rajeev,
+%%   etc since they also try to do semantics preserving optimizations on
+%%   dataflow graphs}
 

### 2019-11-23 front_end/execute_evaluation_scripts.sh
Minor change

@@ -1,5 +1,6 @@
 #!/bin/bash
 
+distr_output_dir=/tmp/distr_output
 
 ./execute_compile_evaluation_script.sh "minimal_sort_2" $distr_output_dir 1 0
 ./execute_compile_evaluation_script.sh "minimal_sort_4" $distr_output_dir 1 0

### 2019-11-23 front_end/gather_results.py
Update plots

@@ -26,7 +26,8 @@ experiments = ["minimal_grep",
                "topn",
                "wf",
                "grep",
-               "spell"]
+               "spell",
+               "shortest_scripts"]
 
 def get_experiment_files(experiment, results_dir):
     files = [f for f in os.listdir(results_dir) if f.startswith(experiment)]

### 2019-11-23 pldi/draft.tex
Update plots
Update eval

--- 2019-11-23 pldi/draft.tex
@@
 \subsection{Microbenchmarks: Shell One-liners}
-
-Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
-Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
-% The centralized pipelines read files of various sizes from disk and write to disk.
-
-Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
-Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
-
-The size of the script grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
-The first number alone is interesting, as it captures \sys's output for the sequential script.
-As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
-% this is the size of 
-This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
-
-In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
-% a DFA-based regular expression matching phase;
-  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
-The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
-The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
-  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-The \ttt{bi-gram} pipeline calculates n-grams of base two;
-  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
-\ttt{spell} is based on 
-spell program~\cite{bentley1985spelling} 
-The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
-  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
-
-Generally, the planning
 
@@
     \end{subfigure}%
-    \caption{todo}
+    \caption{
+      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--200).} 
+Three different configurations per benchmark:
+  (i) distributed, focusing only on the distributed execution,
+  (ii) +compile, adding \sys' pre-processing overheads, and
+  (iii) +cat, adding a final merger step.
+    }
+    \vspace{-15pt}
     \label{fig:microbenchmarks}
 \end{figure*}
+
+Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
+Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
+% The centralized pipelines read files of various sizes from disk and write to disk.
+
+Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
+Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+
+The size of a script (col.
+5) grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+The first number alone is interesting, as it captures \sys's output for the sequential script.
+As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
+% this is the size of 
+This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
+FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
+
+In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
+% a DFA-based regular expression matching phase;
+  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
+The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
+The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
+  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
+The \ttt{bi-gram} pipeline calculates n-grams of base two;
+  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
+The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
+  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
+
+% TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
+Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200).
+% we experiment with a $200\times$ level of parallelism to see potential overheads
+Each one of these plots reports on three different configurations:
+  (i) distributed, which captures only the runtime of the distributed computation,
+  (ii) compile + distributed, which adds the \sys' analysis and generation overhead, and
+  (iii) cat + compile + distributed, which adds a final merger  gathering the results at the end of the pipeline.
+The results show significant speedups, ranging between \todo{7--70$\times$}, depending on the distributability characteristics of individual pipelines.
+
+Interestingly, the overhead from \sys's analysis and generation phases remains collectively below 1s even for the high-scalability cases.
+We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice (\S\ref{macro1}--\ref{macro3}).
+We find the resulting times quite acceptable:
+  \todo{8.23s} for parsing the script, 
+  \todo{3.22s} for analysis, and 
+  \todo{X.XXs} for rewriting the results of the analysis.
 

### 2019-11-23 pldi/draft.tex

@@
         \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
+        \includegraphics[width=\textwidth]{\detokenize{./figs/spell_throughput_scaleup.pdf}}
         %% \caption{TODO}
@@
         \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+        \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
         %% \caption{TODO}
@@
 with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
+
+\kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
+  na ginoun formally proven claims gia to equivalence preservation twn
+  optimizations etc.}
 

### 2019-11-23 pldi/draft.tex
Add miniplots
clean up 4

--- 2019-11-23 pldi/draft.tex
@@
 To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation
+The key insight behind \sys is that a large part of the shell language can be encoded using a dataflow graph model, providing structure to distribute computations.
+%% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
 % FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
 Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies \todo{maximal} sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
 A \sys-enabled $p_1$ will run \ttt{tr}s in parallel streams,
@@
 
-The paper is structured as follows.
-It first starts with a background section outlining pipeline concepts and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{recipes} highlight key contributions:
+The paper is structured as follows.  It first starts with a background
+section outlining pipeline concepts and overviewing \sys~\sx{bg}.
+Sections \ref{distributability}--\ref{recipes} highlight key
+contributions: \TODO{Change the 1st and the 3rd to reflect the new
+  section layout. I already changed the second.}
 \begin{itemize}
@@
   \item
-  \S\ref{ir} presents a set of rewriting transformations for identifying maximal pipeline sub-expressions, candidates for scale-out.
+  \S\ref{ir} presents a dataflow graph model that can encode shell programs, and a set of parallelization transformations that preserve its semantics.
 
@@
 
-\heading{Pure Commands}
+%% \heading{Pure Commands}
 %
+\noindent
 Together with the commands in \sta, that are always easy to
@@
 operation. In order to allow command developers to extend \sys with
-information about how to parallelize new pure commands, \sys offers an
-interface to specify how to distribute a pure command.
-
-\TODO{I am not sure a general interface is so easy to design. It needs
+information about how to parallelize new pure commands, \sys offers a
+programmatic interface to specify how to distribute a pure command.
+
+\tr{I am not sure a general interface is so easy to design. It needs
   more though. It might be beneficial to just talk about sort and wc
@@
 
-\TODO{Is the following paragraph best suited here? It describes an
-  assumption that we make about commands, to guarantee the validity of
-  our transformations. The assumption is needed due to the limitation
-  of the programming model of one input one output per command.}
+%% \TODO{Is the following paragraph best suited here? It describes an
+%%   assumption that we make about commands, to guarantee the validity of
+%%   our transformations. The assumption is needed due to the limitation
+%%   of the programming model of one input one output per command.}
 
@@
 stream (or respectively write to more files that their output stream),
-however these files are static (e.g. used for configuration or logging
-\kk{Would an example here help}) and do not represent streams of data,
-and therefore are not considered part of the dataflow graph. We assume
-that these static files can only be accessed by one node, thus
-ensuring that they do not interfere with the distributed
-implementation.
+however these files are static (e.g. used for configuration or
+logging) and do not represent streams of data, and therefore are not
+considered part of the dataflow graph. We assume that these static
+files can only be accessed by one node, thus ensuring that they do not
+interfere with the distributed implementation.
 %% Because of the assumption in section Front End \ref{}, that each file
@@
 
-\TODO{Show an example pipe and its graph}.
-
-\begin{center}
-\small
-\begin{tikzpicture}[node distance=2.25cm, ->]
-\node (S) {};
-\node (A) [draw, right of=S, node distance=1.75cm] {node 1};
-\node (B) [draw, right of=A] {node 2};
-\node (C) [draw, right of=B] {node 3};
-\node (E) [right of=C, node distance=1.75cm] {};
-\path (S) edge node[above] {input} (A);
-\path (A) edge node[above] {stream} (B);
-\path (B) edge node[above] {stream} (C);
-\path (C) edge node[above] {output} (E);
-\end{tikzpicture}
-\end{center}
+%% \TODO{Show an example pipe and its graph}.
+
+%% \begin{center}
+%% \small
+%% \begin{tikzpicture}[node distance=2.25cm, ->]
+%% \node (S) {};
+%% \node (A) [draw, right of=S, node distance=1.75cm] {node 1};
+%% \node (B) [draw, right of=A] {node 2};
+%% \node (C) [draw, right of=B] {node 3};
+%% \node (E) [right of=C, node distance=1.75cm] {};
+%% \path (S) edge node[above] {input} (A);
+%% \path (A) edge node[above] {stream} (B);
+%% \path (B) edge node[above] {stream} (C);
+%% \path (C) edge node[above] {output} (E);
+%% \end{tikzpicture}
+%% \end{center}
 
@@
 
-\kk{(Maybe) It might be beneficial to mention that many inputs do not
-  represent the same thing as many outputs. These two notions are not
-  symmetric.}
-
-\begin{figure*}
-\centering
-\begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
-%
-\small
-%
-\node (M1) {};
-\node (M2) [draw, right of=M1, node distance=2cm] {Map};
-\node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
-\node (M4) [right of=M3, node distance=3cm] {};
-%
-\path (M1) edge node {input} (M2);
-\path (M2) edge node {stream} (M3);
-\path (M3) edge node {output} (M4);
-\end{tikzpicture}
-\end{figure*}
-
-\begin{figure*}
-\newcommand{\Split}{\text{Split}}
-\newcommand{\Merge}{\text{Merge}}
-\newcommand{\Map}{\text{Map}}
-\newcommand{\GroupBy}{\text{GroupBy}}
-\newcommand{\Reduce}{\text{Reduce}}
-\centering
-\begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
-%
-\footnotesize
-%
-\node (M1) {};
-\node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
-\node (M3) [right of=M2, node distance=1.5cm] {};
-\node (M4) [right of=M3, node distance=2.5cm] {};
-\node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
-\node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
-\node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
-\node (M8) [right of=M7, node distance=2cm] {};
-%
-\node (T3) [draw, above of=M3] {$\Map$};
-\node (T4) [draw, above of=M4] {$\Split_K$};
-\node (T5) [draw, above of=M5] {$\Merge$};
-\node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
-%
-\node (B3) [draw, below of=M3] {$\Map$};
-\node (B4) [draw, below of=M4] {$\Split_K$};
-\node (B5) [draw, below of=M5] {$\Merge$};
-\node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
-%
-\path (M1) edge node {$A$} (M2);
-\path (M2) edge[bend left=10] node {$A$} (T3);
-\path (M2) edge[bend right=10] node[swap] {$A$} (B3);
-\path (M5) edge node {$K_2 \times B$} (M6);
-\path (M6) edge (M7);
-\path (M7) edge node {$K \times B$} (M8);
-%
-\path (T3) edge node {$K \times B$} (T4);
-\path (T4) edge (T5);
-\path (T4) edge (M5);
-\path (T4.330) edge[bend right=10] (B5.170);
-\path (T5) edge node {$K_1 \times B$} (T6);
-\path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
-%
-\path (B3) edge node {$K \times B$} (B4);
-\path (B4) edge (B5);
-\path (B4) edge (M5);
-\path (B4.30) edge[bend left=10] (T5.190);
-\path (B5) edge node {$K_3 \times B$} (B6);
-\path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
-%
-\end{tikzpicture}
-\caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
-\label{fig:mapReduce}
-\end{figure*}
+%% \kk{(Maybe) It might be beneficial to mention that many inputs do not
+%%   represent the same thing as many outputs. These two notions are not
+%%   symmetric.}
+
+%% \begin{figure*}
+%% \centering
+%% \begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
+%% %
+%% \small
+%% %
+%% \node (M1) {};
+%% \node (M2) [draw, right of=M1, node distance=2cm] {Map};
+%% \node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
+%% \node (M4) [right of=M3, node distance=3cm] {};
+%% %
+%% \path (M1) edge node {input} (M2);
+%% \path (M2) edge node {stream} (M3);
+%% \path (M3) edge node {output} (M4);
+%% \end{tikzpicture}
+%% \end{figure*}
+
+%% \begin{figure*}
+%% \newcommand{\Split}{\text{Split}}
+%% \newcommand{\Merge}{\text{Merge}}
+%% \newcommand{\Map}{\text{Map}}
+%% \newcommand{\GroupBy}{\text{GroupBy}}
+%% \newcommand{\Reduce}{\text{Reduce}}
+%% \centering
+%% \begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
+%% %
+%% \footnotesize
+%% %
+%% \node (M1) {};
+%% \node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
+%% \node (M3) [right of=M2, node distance=1.5cm] {};
+%% \node (M4) [right of=M3, node distance=2.5cm] {};
+%% \node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
+%% \node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
+%% \node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
+%% \node (M8) [right of=M7, node distance=2cm] {};
+%% %
+%% \node (T3) [draw, above of=M3] {$\Map$};
+%% \node (T4) [draw, above of=M4] {$\Split_K$};
+%% \node (T5) [draw, above of=M5] {$\Merge$};
+%% \node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
+%% %
+%% \node (B3) [draw, below of=M3] {$\Map$};
+%% \node (B4) [draw, below of=M4] {$\Split_K$};
+%% \node (B5) [draw, below of=M5] {$\Merge$};
+%% \node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
+%% %
+%% \path (M1) edge node {$A$} (M2);
+%% \path (M2) edge[bend left=10] node {$A$} (T3);
+%% \path (M2) edge[bend right=10] node[swap] {$A$} (B3);
+%% \path (M5) edge node {$K_2 \times B$} (M6);
+%% \path (M6) edge (M7);
+%% \path (M7) edge node {$K \times B$} (M8);
+%% %
+%% \path (T3) edge node {$K \times B$} (T4);
+%% \path (T4) edge (T5);
+%% \path (T4) edge (M5);
+%% \path (T4.330) edge[bend right=10] (B5.170);
+%% \path (T5) edge node {$K_1 \times B$} (T6);
+%% \path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
+%% %
+%% \path (B3) edge node {$K \times B$} (B4);
+%% \path (B4) edge (B5);
+%% \path (B4) edge (M5);
+%% \path (B4.30) edge[bend left=10] (T5.190);
+%% \path (B5) edge node {$K_3 \times B$} (B6);
+%% \path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
+%% %
+%% \end{tikzpicture}
+%% \caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
+%% \label{fig:mapReduce}
+%% \end{figure*}
+
+% \begin{figure}[t]
+% \centering
+\includegraphics[width=0.25\textwidth]{\detokenize{./figs/dish_g1.pdf}}
+
+\includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_g2.pdf}}
+% \caption{
+%   \textbf{\sys transformation overview.}
+%   Given a shell pipeline, \sys leverages command distributability to identify
+%   high-distributability stages, rewrite them, and orchestrate their execution.
+% }
+% \vspace{-15pt}
+% \label{fig:overview}
+% \end{figure}
 
@@
 parallelism. Different nodes in the graph can be executed
-concurrently, communicating through the edges. Task level parallelism
-is already exposed by the shell, by using pipes \ttt{|} and the
-background operator \ttt{\&}. Unfortunately, simple task
-parallelism leaves a lot to be desired, since scaling a computation
-requires manually introducing more operators.
+concurrently, communicating through the edges.
+%% Task level parallelism is already exposed by the shell, by using
+%% pipes \ttt{|} and the background operator \ttt{\&}.
+Unfortunately, simple task parallelism leaves a lot to be desired,
+since scaling a computation requires manually introducing more
+operators.
 
@@
 the same result when performed independently on subsets of the input
-data, merging the results afterwards. This is the basic optimization
-that is used by popular distributed processing systems such as
-MapReduce \cite{DG2004MR} and Apache Spark \cite{spark:12}. Unfortunately, the
-dataflow graph model of most of these systems is not expressive enough
-to represent data parallelism \km{not sure this is true}, so the distributed implementations that
-they perform do not preserve the semantics of their dataflow model.
+data, merging the results afterwards. Exploiting data parallelism is
+one of the major reasons that popular distributed stream processing
+systems (such as MapReduce \cite{DG2004MR} and Apache Spark
+\cite{spark:12}) achieve significant performance. Unfortunately, the
+optimizations in many of these systems do not preserve the dataflow
+model semantics, blurring the lines between specification,
+optimization, and implementation.
 
@@
 optimizations that expose data parallelism, or even revert them when
-the resulting dataflow graph is too wide. This flexibility enables
-unified reasoning and implementation of the unoptimized and optimized
-version of the dataflow graph. \kk{This is interesting, and should be
-  moved up (or rephrased in the introduction)}
+the resulting dataflow graph is too wide. This flexibility unifies the
+unoptimized and optimized versions of a computation, enabling thinking
+about them with the same conceptual tools.
 
@@
 \nv{TODOs
-  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me);
-  (ii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
+  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
 }
@@
 
-\km{This is a nice characterization, which I think is true in most settings.
-There is a caveat though: You can have stateless programs that emit output when they start (before they consume the stream) and/or when they consume the EOF symbol.
-In this case I am not sure the equation above holds.}
+\tr{This is a nice characterization, which I think is true in most
+  settings. There is a caveat though: You can have stateless programs
+  that emit output when they start (before they consume the stream)
+  and/or when they consume the EOF symbol. In this case I am not sure
+  the equation above holds.}
 
@@
 commands connected with a \ttt{\&\&} operator. In addition, the
-fact that the shell is \todo{extremely dynamic}, further increases the
+fact that the shell is highly dynamic, further increases the
 challenge of correctly identifying what parts can be distributed.
@@
 
-In order to address this issue, we \todo{introduce} the notion of
+In order to address this issue, we introduce the notion of
 distributable program regions, and we design a translation pass that
@@
 (e.g. changing the current directory, modifying environment
-variables), and if it does not read from any \todo{read-once} file,
-like a fifo pipe. If the command is considered safe, its category is
+variables), and if it does not read from any read-once file
+(e.g. pipe). If the command is considered safe, its category is
 identified and a node containing that command is created and added to
@@
 \nv{Give example, extend}
+
+\kk{Mention sort, wc}
 

### 2019-11-23 pldi/draft.tex
Almost done with section 5

@@
 files that don't point to a resource can have an upper bound on the
-lines that they can tranfer. The following example illustrates why
-this is essential.
-
-\TODO{Add an example of a command that has two outputs, one of which
-  goes to one command, while another goes to a different
-  command. Maybe only show the graph here and not the command itself.}
+lines that they can tranfer. This can be useful to utilize avaliable
+parallelism after encountering a command that cannot be parallelized
+(e.g. \ttt{sha1sum}). The following example illustrates this scenario.
+
+\includegraphics[width=0.40\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
 
@@
 
-% \begin{figure}[t]
-% \centering
-\includegraphics[width=0.25\textwidth]{\detokenize{./figs/dish_g1.pdf}}
-
-\includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_g2.pdf}}
-% \caption{
-%   \textbf{\sys transformation overview.}
-%   Given a shell pipeline, \sys leverages command distributability to identify
-%   high-distributability stages, rewrite them, and orchestrate their execution.
-% }
-% \vspace{-15pt}
-% \label{fig:overview}
-% \end{figure}
-
 \subsection{Graph transformations}
@@
 can then be implemented by spawning a process for each node, and
-redirecting the outputs according to the graph edges.
-
-\nv{Give example, extend}
-
-\kk{Mention sort, wc}
+redirecting the outputs according to the graph edges. An example
+transformation can be seen in \Cref{fig:optimizer-transformation}.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.70\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
+\caption{
+  \textbf{Example graph transformation.}
+  The optimizer splits a stateless node to utilize
+  available data parallelism.
+}
+\vspace{-15pt}
+\label{fig:optimizer-transformation}
+\end{figure}
+
+The optimizer also applies custom transformations on several commands
+to further improve performance. Two notable examples are the
+transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
+that have more than one input are tranformed to \ttt{sort} commands on
+each input, followed by a \ttt{sort -m}, to produce the final sorted
+result. The transformation on occurences of \ttt{wc} that have more
+than input \ttt{i1, t2} is illustrated below as a shell script:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+  paste -d '+' \
+   <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
+   <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
+   bc | tr -s '\n' ' ' |
+   sed 's/^/   /' | sed 's/$/\ /'
+\end{lstlisting}
+
+The above script takes the outputs produced by two \ttt{wc} commands
+and aggregates them accordingly.
 
@@
 needs to concatenate input files (or conversely split output
-files). Here is an example:
-
-\kk{Show an example graph! of cat | tr | sort. The command in the
-  middle must only take input from one file (maybe just stdin).}
-
+files).
+%
+%% \tr{Show an example graph! of cat | tr | sort. The command in the
+%%   middle must only take input from one file (maybe just stdin).}
+%
 We address this issue using primitive shell constructs for file

### 2019-11-23 pldi/draft.tex

@@
 \noindent
-\TODO{Change that with the new splitter}
 On the other hand, given an input file \verb|in| that is a FIFO pipe,
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none]
- head -n $N $in > $out1 ; cat $in > $out2
+  tee >(head -n $N > $temp1;
+        dd of=/dev/null > /dev/null 2>&1 &
+        cat $temp1 > $out1) |
+       (tail -n +${N+1} > $out2;
+        dd of=/dev/null > /dev/null 2>&1)
 \end{lstlisting}
+ %% head -n $N $in > $out1 ; cat $in > $out2
 

### 2019-11-23 pldi/draft.tex

@@
 
-Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
-Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
-% The centralized pipelines read files of various sizes from disk and write to disk.
-
-Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
-Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
-
-The size of the script grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
-The first number alone is interesting, as it captures \sys's output for the sequential script.
-As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
-% this is the size of 
-This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
-
-In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
-% a DFA-based regular expression matching phase;
-  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
-The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
-The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
-  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-The \ttt{bi-gram} pipeline calculates n-grams of base two;
-  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
-\ttt{spell} is based on 
-spell program~\cite{bentley1985spelling} 
-The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
-  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
-
-Generally, the planning
-
 \begin{figure*}[t]
@@
     \end{subfigure}%
-    \caption{todo}
+    \caption{
+      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--200).} 
+      Three different configurations per benchmark:
+      (i) distributed, focusing only on the distributed execution,
+      (ii) +compile, adding \sys' pre-processing overheads, and
+      (iii) +cat, adding a final merger step.
+    }
+    \vspace{-15pt}
     \label{fig:microbenchmarks}
 \end{figure*}
+
+Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
+Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
+% The centralized pipelines read files of various sizes from disk and write to disk.
+
+Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
+Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+
+The size of a script (col.
+5) grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+The first number alone is interesting, as it captures \sys's output for the sequential script.
+As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
+% this is the size of 
+This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
+FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
+
+In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
+% a DFA-based regular expression matching phase;
+  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
+The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
+The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
+  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
+The \ttt{bi-gram} pipeline calculates n-grams of base two;
+  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
+The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
+  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
+
+% TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
+Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200).
+% we experiment with a $200\times$ level of parallelism to see potential overheads
+Each one of these plots reports on three different configurations:
+  (i) distributed, which captures only the runtime of the distributed computation,
+  (ii) compile + distributed, which adds the \sys' analysis and generation overhead, and
+  (iii) cat + compile + distributed, which adds a final merger  gathering the results at the end of the pipeline.
+The results show significant speedups, ranging between \todo{7--70$\times$}, depending on the distributability characteristics of individual pipelines.
+
+Interestingly, the overhead from \sys's analysis and generation phases remains collectively below 1s even for the high-scalability cases.
+We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice (\S\ref{macro1}--\ref{macro3}).
+We find the resulting times quite acceptable:
+  \todo{8.23s} for parsing the script, 
+  \todo{3.22s} for analysis, and 
+  \todo{X.XXs} for rewriting the results of the analysis.
 

### 2019-11-23 pldi/draft.tex
Weather pipeline execution times
Wrap figure

@@
 
-In order to expose data parallelism from the output of a command,
-files that don't point to a resource can have an upper bound on the
-lines that they can tranfer. This can be useful to utilize avaliable
+In order to expose data parallelism from the output of a command, files that
+don't point to a resource can have
+\begin{wrapfigure}[3]{r}{0.20\columnwidth}
+\vspace{-7pt}
+  \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
+\vspace{-20pt}
+\end{wrapfigure}
+an upper bound on the lines that they can transfer. This can be useful to utilize available
 parallelism after encountering a command that cannot be parallelized
-(e.g. \ttt{sha1sum}). The following example illustrates this scenario.
-
-\includegraphics[width=0.40\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
+(\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.
 

### 2019-11-23 pldi/draft.tex
Update plot names and add synthetic benchmark numbers

--- 2019-11-23 pldi/draft.tex
@@
 We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice (\S\ref{macro1}--\ref{macro3}).
-We find the resulting times quite acceptable:
-  \todo{8.23s} for parsing the script, 
-  \todo{3.22s} for analysis, and 
-  \todo{X.XXs} for rewriting the results of the analysis.
+We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and for optimizing.
+If we also push the limit of input arguments to 100, \sys takes 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has 100 thousand edges, in a file that is almost 11MB.
 
@@
 % hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
+
+% Dish p3 + p4 me platos pipeline 2 => real 4m30.554s (perilambanei download twn gz kai to processing)
+% Dish p3 + p4 me platos pipeline 10 => real 1m7.183s
+
+% Dish p3 me platos pipeline 2 => real	3m47.470s (perilambanei download twn gz kai save sto file system)
+% Dish p3 me platos pipeline 10 => real	1m38.945s
 

### 2019-11-23 front_end/gather_results.py
Update plot names and add synthetic benchmark numbers

@@ -26,8 +26,17 @@ experiments = ["minimal_grep",
                "topn",
                "wf",
                "grep",
-               "spell",
+               # "spell",
                "shortest_scripts"]
+
+pretty_names = {"minimal_grep" : "min-grep",
+                "minimal_sort" : "min-sort",
+                "wf" : "wf",
+                "topn" : "top-n",
+                "grep" : "grep",
+                "bigram" : "bi-gram",
+                "spell" : "spell",
+                "shortest_scripts" : "shortest-scripts"}
 
 def get_experiment_files(experiment, results_dir):
     files = [f for f in os.listdir(results_dir) if f.startswith(experiment)]
@@ -81,7 +90,7 @@ def collect_scaleup_times(experiment, results_dir):
     # plt.yscale("log")
     plt.xticks(all_scaleup_numbers[1:])
     plt.legend(loc='lower right')
-    plt.title(experiment)
+    plt.title(pretty_names[experiment])
 
 
     plt.tight_layout()

### 2019-11-23 pldi/draft.tex
Update table

@@
 % https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
-  \tti{min-grep}         ~&~$3\times$\tsta                &           &               &   17\qquad 134\qquad  1304    & complex NFA regex                                 \\
+  \tti{min-grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
-  \tti{min-sort}         ~&~~$\tsta, \tpur$               &           &               &   X \qquad 129\qquad  1209    & \tti{sort}ing                                     \\
+  \tti{min-sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ &           &               &   X \qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+  \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ & 1000      &   2m 11.18s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ &           &               &   X \qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+  \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ & 1000      &   6m 53.55s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
-  \tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ &           &               &   X \qquad  X \qquad          & stream shifting and merging                       \\
+  \tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
-  \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ &           &               &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
+  \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ & 1000      &   8m 13.42s   &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
-  \tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ &           &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\
+  \tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\
 \bottomrule
@@
 Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
-Script sizes report on the \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+Script sizes report on the \sys's rewritten output for three different distribution sizes---$2\times$, 10$\times$, and $100\times$.
 

### 2019-11-23 pldi/draft.tex
Proof reading

@@
   \item
-  \S\ref{distributability} overviews \sys and introduces several scalability and distributability classes.
+  \S\ref{distributability} describes several scalability and distributability classes and introduces the distributability characterization language.
 
@@
   \item
-  \S\ref{impl} details other concerns, such as distributed state management and extensibility.
+    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of shell programs.
+    %% details other concerns, such as distributed state management and extensibility. 
 \end{itemize}
@@
 Intuitively,
- (i) the parallel-composition operator exposes \todo{task} parallelism
+ (i) the parallel-composition operator exposes task parallelism
 % FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
@@
 First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
-One example is the set of commands defined by the \textsc{POSIX} standard~\cite{}.
+One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
 \sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
@@
 Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying \todo{maximal} distributable sub-expressions.
+\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying  distributable regions.
 It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
@@
 Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
-Other commands are treated as  non-distributable for now~\sx{conclusion};
+Other commands are treated as  non-distributable for now~\sx{discussion};
   by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.

### 2019-11-23 pldi/draft.tex
Related work

--- 2019-11-23 pldi/draft.tex
@@
 We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice (\S\ref{macro1}--\ref{macro3}).
-We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and for optimizing.
-If we also push the limit of input arguments to 100, \sys takes 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has 100 thousand edges, in a file that is almost 11MB.
+We find the resulting times quite acceptable:
+  \todo{8.23s} for parsing the script, 
+  \todo{3.22s} for analysis, and 
+  \todo{X.XXs} for rewriting the results of the analysis.
 
@@
 
-At a high-level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+At a high level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+
+\heading{Distributed Operating Systems and Shells}
+A plethora of systems assist in the construction of distributed software.
+Distributed operating systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
+hide many distribution problems, as long as a program uses the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
+% simplify many of the problems of distribution and
+
+Several packages expose commands for specifying parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}.
+%  or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+Different from \sys, their effectiveness is predicated upon explicit invocation and is limited to embarrassingly parallel (and short) programs.
+Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
+For example, \tti{parallel}, an official GNU project, contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
+\sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently.
+
+Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target distribution.
+Here too, however, developers are expected to manually rewrite scripts to exploit these primitives.
+
+Recently, the Smoosh authors have argued that certain shell features are useful for concurrency~\cite{smoosh:18}.
+Their argument is somewhat antithetical to \sys, % but rather that the shell language is itself an good DSL for concurrency;
+  which argues for automatically (and correcctly) parallelizing pipelines when users have not expressed parallelism explicitly---hence distribution-\emph{oblivious} shell scripting.
+
+\heading{Cluster Computing}
+Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation.
+By providing a few carefully-designed primitives, they 
+Due to the specialization of these primitives, however, these systems suffer from a lack of generality ;
+% and they still require developers to compose their work using the pipelines
+
+They also support a single class of distributability, arguing that is enough 
+by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
+Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
 
@@
 %   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{rashid1981accent, walker1983locus, ousterhout1988sprite, mullender1990amoeba, pike1990plan9, rozier1991overview, dorward1997inferno, barak1998mosix, schwarzkopf2013dios, sacha2013osprey} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-% simplify many of the problems of distribution and
-provide a significant amount of automation. % but require development in a new system or language.  While they take care of all the distribution minutiae, 
-However, they involve significant manual effort using the provided abstractions, which are strongly coupled with the underlying operating or runtime system.
-% TODO: \TODO{This has to be broken down into systems and PLs, coz systems are the most relevant/important class.}
-
 \heading{Annotation-based Distribution}
@@
 \caption{
-  \textbf{Cost of Manual Effort.}
+  \textbf{Manual--automated distribution spectrum.}
 	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-A more complete picture is presented in the related-work section~\sx{rt}.
 }
@@
 \end{figure}
-
-\heading{DSLs for Distribution}
-At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
-simplify certain patterns, 
-but do not offer the flexibility of a full-fledged environment.
-% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
-Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
 
@@
 %Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
-
-\heading{Concurrent and Distributed Shells}
-\nv{NV}
-Among other work, the Smoosh authors have argued that certain features of the shell are useful for concurrency~\cite{smoosh:18};
-Their work does not automatically parallelize pipelines, but rather argues that the shell language is itself an good DSL for concurrency;
-This argument is different from \sys, 
-an argument complementary to \sys's which automatically (and correcctly) parallelizes pipelines even when developers have---hence distribution-\emph{oblivious} shell scripting.
-
-Several shells add support for non-linear pipe topologies;
-  this 
-2dsh~\cite{}
-rc~\cite{}
-gsh~\cite{}
-dgsh~\cite{dagsh:17}.
-
-Commands for explicitly specifying parallelism such \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-These commands require the user to express the program explicitely, work only with embarrasingly parallel (and short) programs;
-  \sys and scales out to large pipelines
-Often, these commands provide options to support very special cases
----a stark contradiction to the celebrated \unix philosophy.
-For example, \tti{parallel}, an official GNU project, flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs that
-with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
 

### 2019-11-23 pldi/draft.tex
clean up in 6

--- 2019-11-23 pldi/draft.tex
@@
 % https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
-  \tti{min-grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
+  \tti{grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
-  \tti{min-sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
+  \tti{sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
 % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
@@
 % https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
-  \tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\
+  \tr{\tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\}
 % https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
@@
 % https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
-  \tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\
+  \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
+  \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & \todo{x}                          \\
 \bottomrule
@@
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use a combination of micro- and macro-benchmarks, most of which were collected from prior literature and online repositories.
-Micro-benchmarks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write (usually interactively);
+To achieve this, we use a combination of micro-benchmarks and one macro-benchmark, most of which were collected from prior literature and online repositories.
+Micro-benchmarks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-These pipelines are taken from real use cases (often from the literature~\cite{bentley1985spelling, bentley1985spelling}), and, precisely due to their small size, have the benefit of highlighting stress on a handful of stages.
-
-Three complex macro-benchmarks (\S\ref{macro1}--\ref{macro3}) involve third-party commands and handle realistic workloads by today's standards.
-These scripts are about an order of magnitude larger than the micro-benchmarks, but this perceived lack of complexity is deceiving:
-  as shown below, if written in conventional languages, these pipelines would correspond to programs on the order of hundreds of lines of code.
+These pipelines are taken from real use cases (often from the literature~\cite{bentley1985spelling}), and, precisely due to their small size, have the benefit of highlighting stress on a handful of stages.
+
+The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
+It is about an order of magnitude larger than the micro-benchmarks, but this perceived lack of complexity is deceiving:
+  as shown below, if written in conventional languages, this pipeline would correspond to a program on the order of hundreds of lines of code.
 
@@
 First and foremost, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around XX--XXms, negligible (XX\%) for short-running pipelines and virtually non-existent for long-running ones.
+\sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
 As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
 Finally, \sys preserves the productivity of the shell:
-  in one case, the equivalent program for a small part of the pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop;
-  in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
+  the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
+  %% in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
 
@@
 \subsection{Microbenchmarks: Shell One-liners}
+\label{micro}
 
@@
   they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-The \ttt{bi-gram} pipeline calculates n-grams of base two;
-  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.
+\tr{The \ttt{bi-gram} pipeline calculates n-grams of base two;
+  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
 The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
@@
   (iii) cat + compile + distributed, which adds a final merger  gathering the results at the end of the pipeline.
-The results show significant speedups, ranging between \todo{7--70$\times$}, depending on the distributability characteristics of individual pipelines.
+The results show significant speedups, ranging between 4--70$\times$, depending on the distributability characteristics of individual pipelines.
 

### 2019-11-23 pldi/draft.tex
Pass over related work

--- 2019-11-23 pldi/draft.tex
@@
 
-At a high level, techniques for automatically distributing fall under a spectrum that ranges from fully automated (but potentially sub-optimal) distribution to fully manual (but, ideally, optimal) distribution (Fig.~\ref{fig:spectrum}).
+At a high level, techniques for re-writing programs to exploit distribution fall under a wide spectrum that ranges from
+  fully manual (but, ideally, optimal) to 
+  fully automated (but potentially sub-optimal) distribution   (Fig.~\ref{fig:spectrum}).
 
@@
 Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \tti{parallel}, an official GNU project, contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user would achieve with \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
+For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
 \sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently.
@@
 Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target distribution.
-Here too, however, developers are expected to manually rewrite scripts to exploit these primitives.
+Here too, however, developers are expected to manually rewrite scripts to exploit these new primitives.
 
@@
 \heading{Cluster Computing}
-Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation.
-By providing a few carefully-designed primitives, they 
-Due to the specialization of these primitives, however, these systems suffer from a lack of generality ;
-% and they still require developers to compose their work using the pipelines
-
-They also support a single class of distributability, arguing that is enough 
-by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
-Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
+Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
+These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
+By targeting specific classes of computation (\emph{viz.} \sys's distributability), they render distributed computation efficiently.
+
+% Lack of generality, no formal guaratees, their setup and is too tedious 
+\sys chooses a more general approach:
+  it does not require users to set up a new framework for every new class of computation they need to execute, nor to rewrite different parts of their computation using the specific abstractions provided by each framework they use.
+It also takes a more principled approach: by developing the semantic model outlined earlier, it can guarantee that transformations are correct with respect to the sequential program.
+
+% % Although still a research prototype, \sys's setup and deployment is simpler.
+% It also does not require these specialized frameworks, manually (re)write , and 
+% Different from \sys, however, 
+% Due to the specialization of these primitives, however, these systems suffer from a lack of generality ;
+% % and they still require developers to compose their work using the pipelines
+% Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
+
+\heading{Annotation-based Distribution}
+Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
+They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
+Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
+For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
+% Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
+Unfortunately, a challenge with this approach is that users composing applications by using third-party libraries do not understand their distributability details well-enough to annotate them.
+
+\sys occupies a different point in the design space, by removing any work from composers and adding a small load to developers.
+% More specifically, it is designed to be used automatically by composers,
+Its DSL is designed to be more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared towards the developers of these components---not the users composing them.
+
+% It also provides a more principled approach than both of these systems.
+% Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
+% The side-effects present in the shell dwarf those of functional languages
 
 % Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Automated Parallelization and Distribution}
-There is a long history of automated parallelization starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+\heading{Low-level Parallelization}
+Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
 These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
+
+More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
+These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
+%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
+Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
+%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
 
@@
 %   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-\heading{Annotation-based Distribution}
-Recent systems for light-touch distribution, Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, give developers the ability to express through annotations;
-
-Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-This is different
-
-Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
-\sys is designed to be used, so it is somewhat more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared toward the deveopers of these components---not the users composing them.
-As such..
-Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
-The side-effects present in the shell dwarf those of functional languages
-The shell is also more general.
 
@@
 \end{figure}
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
-%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
-Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
 

### 2019-11-23 pldi/draft.tex
remove two subsections

--- 2019-11-23 pldi/draft.tex
@@
 \noindent
-\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and multi-line macro-benchmarks---a web crawling and indexing engine, a genomics pipeline, and a popular weather analysis script.
-After a comparison with related prior work~\sx{related}, the paper closes with a discussion of related work and possible future directions~\sx{discussion}.
+\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
+After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
 % \bigskip
@@
 
-\subsection{Macro-benchmark: Bioinformatics Transform}
-\label{macro3}
-
-\subsection{Macro-benchmark: Web Crawling}
-\label{macro2}
-
 \section{Related Work}

### 2019-11-23 pldi/draft.tex
Update

--- 2019-11-23 pldi/draft.tex
@@
 To achieve this, we use a combination of micro-benchmarks and one macro-benchmark, most of which were collected from prior literature and online repositories.
-Micro-benchmarks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
+Micro-bench\-marks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
@@
 
-\kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
-  na ginoun formally proven claims gia to equivalence preservation twn
-  optimizations etc.}
+% \kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
+%   na ginoun formally proven claims gia to equivalence preservation twn
+%   optimizations etc.}
 

### 2019-11-23 pldi/draft.tex
more

@@
   \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
-  \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & \todo{x}                          \\
+  \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & extensive file-system operation   \\
 \bottomrule
@@
 The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
-It is about an order of magnitude larger than the micro-benchmarks, but this perceived lack of complexity is deceiving:
-  as shown below, if written in conventional languages, this pipeline would correspond to a program on the order of hundreds of lines of code.
+It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
+  as shown if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
 
@@
 
-In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command.
+In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command;
 % a DFA-based regular expression matching phase;
@@
   they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-\tr{The \ttt{bi-gram} pipeline calculates n-grams of base two;
-  it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
+% \tr{The \ttt{bi-gram} pipeline calculates n-grams of base two;
+%   it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
 The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
   after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
+Finally, \ttt{shortest-scripts} is a pipeline~\cite[pg. 7]{taylor2004wicked}that identifies the 15 shortest shell scripts in the user's \ttt{PATH};
+   interestingly, it uses \ttt{file} utility and a higher-order \ttt{wc}.
 

### 2019-11-23 pldi/draft.tex
[Minor]

@@
 \heading{Limitations}
-Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct\TODO{with respect to some assumptions}--- despite \sys being a research prototype.
+Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
 
 \sys does not adequately handle failures or network partitions in the general sense, which would include re-scheduling passes that were executing on failed replicas.
-Prior work on fault-tolerant distributed stream processing~\cite{} can be of significant aid here.
+Prior work on fault-tolerant distributed stream processing can be of significant aid here.
 
 The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
-Prior work on operator placement of distributed dataflow graphs~\cite{} can be used to improve it.
+Prior work on operator placement of distributed dataflow graphs can be used to build a more sophisticated planner.
 
 \sys is conservative in the shell subsets that it handles, in order to
-not introduce unsafe behaviours. The point of this work was not to be
+avoid introducing unsafe behaviours. The point of this work was not to be
 able to handle a complete distributable subset of the shell, but
-rather use significant part of it to demonstrate performance benefits.
-
-Shell is highly dynamic, thus making it impossible to design sound
-meaningful analyses to determine distributable regions and maximal
-parallelization. In order to address this particularity, \sys could be
-tighly integrated with a posix compliant shell
-interpreter~\cite{smoosh:20} so that dynamic information can be
-acquired selectively to aid the analysis and optimization process.
+rather to use significant part of it to demonstrate performance benefits.
+
+% Shell is highly dynamic, thus making it impossible to design sound
+% meaningful analyses to determine distributable regions and maximal
+% parallelization. In order to address this particularity, \sys could be
+% tighly integrated with a posix compliant shell
+% interpreter~\cite{smoosh:20} so that dynamic information can be
+% acquired selectively to aid the analysis and optimization process.
 
@@
 Another direction is to extend \sys to handle other classes outside \sta and \pur.
-These two classes enjoy a certain in quick, one-off pipelines and are the easiest to distribute, but 
+These two classes enjoy a certain popularity with quick, one-off pipelines and are the easiest to distribute, but 
 it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).

### 2019-11-23 pldi/draft.tex
related work

--- 2019-11-23 pldi/draft.tex
@@
 
-\heading{Dataflow Graph Model}
-
-\nv{needs KM}
-On our dataflow graph model: Our data flow graph model is
-  different than most distributed batch and stream processing systems,
-  since the shell is a hybrid between the two. Batch processing
-  systems have no notion of data order, and stream processing systems
-  usually have order, but none of these systems supports this notion
-  of reading from different sources in sequence. Our dataflow graph
-  model supports this by including information about incoming edge
-  order. Another benefit of our dataflow model, is that together with
-  the equations that hold on stateless and pure commands, data
-  parallelism is exposed in the graph as task parallelism. This allows
-  the unoptimized and optimized dataflow graph to have the same
-  unified representation, and doesn't lead to an ad-hoc optimization
-  of the dataflow graph as done in the other systems (the last bit is
-  too aggressive).
-
-%% \TODO{We have to also cite the paper by Konstantinos, Caleb, Rajeev,
-%%   etc since they also try to do semantics preserving optimizations on
-%%   dataflow graphs}
-
-\km{Some very early references on the dataflow model of computation (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN, KMacQ1977}}
-
-\cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but their correctness is established using informal arguments based on operational intution.
-
-The paper \cite{MSAIT2019} proposes a denotational semantic framework for stream processing where the data streams are seen as partial orders, and establishes the soundness of some common parallelizing transformations on dataflow graphs.
+\heading{Correct Distribution of Dataflow Graphs}
+%
+The dataflow graph is a prevalent programming model in several areas
+of data processing, as it is used by several batch~\cite{mapreduce:08,
+  spark:12} and stream processing systems~\cite{murray2013naiad,
+  carbone2015flink}. Despite its popularity, it has been noted that
+most systems perform optimizations that do not preserve its semantics,
+introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
+  SHGW2015, MSAIT2019} is trying to address this issue, by restricting
+optimizations to only be performed in cases where correctness would be
+preserved. Our work draws inspiration from these attempts, as we are
+also interested in optimizations that exploit parallelization but
+preserve the semantics of the underlying model. However, our proposed
+dataflow model is different from these other works, as nodes in our
+model read from their inputs \emph{in sequence}. Due to this
+intricacy, previous work is not straightforwardly applicable.
+
+%% \cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but
+%% their correctness is established using informal arguments based on
+%% operational intution.
+
+%% The paper \cite{MSAIT2019} proposes a denotational semantic framework
+%% for stream processing where the data streams are seen as partial
+%% orders, and establishes the soundness of some common parallelizing
+%% transformations on dataflow graphs.
+
+%% \km{Some very early references on the dataflow model of computation
+%% (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN,
+%% KMacQ1977}}
 

### 2019-11-23 pldi/draft.tex
Update

--- 2019-11-23 pldi/draft.tex
@@
 one of the major reasons that popular distributed stream processing
-systems (such as MapReduce \cite{DG2004MR} and Apache Spark
+systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
 \cite{spark:12}) achieve significant performance. Unfortunately, the
@@
 
-\subsection{Planning Activators and Just-in-Time Planning}
-
-\nv{Nikos working on this and next}
-
-% \kk{This whole section for just in itme planning, is also relevant for
-%   the front end. If the front end is executed just in time, it would
-%   have more information, and maybe it could make a sound analysis.}
-
-Important point: Shell is extremely dynamic. Because of this, a
-distributor cannot decide how to distribute a subexpression
-statically, by just reading the script, (as in many other systems like
-MapReduce, Spark, etc), since most of the information is not there
-before the script starts executing. Environment variables,
-unexpanded(unevaluated) strings \kk{Make sure that terminology is
-  consistent with Greenberg}, could all contain information that is
-valuable to make the distribution plan \kk{give an example}.
-
-Future work on this: calling the shell shtepper to only partially
-evaluate strings, just expand arguments, and not do any significant
-computation. Even better, we could be calling the shtepper just after
-having parsed the ast, and decide on maximal distributable subtrees as
-late as possible in the process. This would be awesome.
-
+% \subsection{Planning Activators and Just-in-Time Planning}
+% 
+% \nv{Nikos working on this and next}
+% 
+% % \kk{This whole section for just in itme planning, is also relevant for
+% %   the front end. If the front end is executed just in time, it would
+% %   have more information, and maybe it could make a sound analysis.}
+% 
+% Important point: Shell is extremely dynamic. Because of this, a
+% distributor cannot decide how to distribute a subexpression
+% statically, by just reading the script, (as in many other systems like
+% MapReduce, Spark, etc), since most of the information is not there
+% before the script starts executing. Environment variables,
+% unexpanded(unevaluated) strings \kk{Make sure that terminology is
+%   consistent with Greenberg}, could all contain information that is
+% valuable to make the distribution plan \kk{give an example}.
+% 
+% 
+% Future work on this: calling the shell shtepper to only partially
+% evaluate strings, just expand arguments, and not do any significant
+% computation. Even better, we could be calling the shtepper just after
+% having parsed the ast, and decide on maximal distributable subtrees as
+% late as possible in the process. This would be awesome.
+% 
 \subsection{Merge and Split Implementation}
@@
 
-For \sys's first macro-benchmark, we use we
-from introductory chapter of Hadoop Definitive Guide~\cite[Chapter 2]{hadoop:15}
-This example is used as a data analysis Hadoop is well suited for, and combines the 
-There are several interesting aspects about this case.
-First, Hadoop is used 
-McIlroy would have probably expressed---our pipeline 
-14 stages long;
+To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory program of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
+At a high level,
+
+The original task can be thought of comprising three sub-tasks:
+  download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
+
+---only one of which is the focus of the chapter
+
+It is presented to demonstrate the benefits of Hadoop is well suited to tackle.
+
+The Java program required for this
 

### 2019-11-23 pldi/draft.tex
Remove appendix
Nits

@@
 
-To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
+To answer this question, this paper presents \sys, a shell variant that transforms traditional shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
 The key insight behind \sys is that a large part of the shell language can be encoded using a dataflow graph model, providing structure to distribute computations.
@@
 section outlining pipeline concepts and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{recipes} highlight key
-contributions: \TODO{Change the 1st and the 3rd to reflect the new
-  section layout. I already changed the second.}
+Sections \ref{distributability}--\ref{impl} highlight key
+contributions:
 \begin{itemize}
@@
 
-This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii)) a high-level overview of \sys's design and implementation.
+This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation.
 
@@
 commands that was described in \Cref{cmd}.
-\nv{TODOs
-  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
-}
+%% \nv{TODOs
+%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
+%% }
 %
@@
 
-%% Appendix
-\appendix
-\section{Scripts used in the evaluation}
-
-This appendix contains the source code of the scripts used in the evaluation of
-the \sys. They are part of the codebase (released as open source with the camera
-ready), and are provided here only to aid the reviewers.
+%% %% Appendix
+%% \appendix
+%% \section{Scripts used in the evaluation}
+
+%% This appendix contains the source code of the scripts used in the evaluation of
+%% the \sys. They are part of the codebase (released as open source with the camera
+%% ready), and are provided here only to aid the reviewers.
 

### 2019-11-23 pldi/draft.tex
Nits

@@
 
-This section reflects on \sys's design and implementation, touches on future work, and closes the paper.
+This section reflects on \sys's design and implementation, touches on future work, and concludes the paper.
 
@@
 
-\sys does not adequately handle failures or network partitions in the general sense, which would include re-scheduling passes that were executing on failed replicas.
+\sys does not adequately handle failures or network partitions in the general sense, which would require re-scheduling passes on failed replicas.
 Prior work on fault-tolerant distributed stream processing can be of significant aid here.
@@
 This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
-\sys's insight is that shell pipelines already express stream computations that can be automatically distributed.
-To achieve this, \sys
+\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
+To achieve its goal, \sys
   decomposes primitives into distributability classes,

### 2019-11-23 pldi/draft.tex
Update

--- 2019-11-23 pldi/draft.tex
@@
 To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory program of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
-At a high level,
-
-The original task can be thought of comprising three sub-tasks:
+At a high level, the goal is to process multi-year weather data from NCDC to identify the temperature over the past several years.
+The original task, as presented in the book, can be thought of comprising three sub-tasks:
   download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
-
----only one of which is the focus of the chapter
-
-It is presented to demonstrate the benefits of Hadoop is well suited to tackle.
-
-The Java program required for this
-
-The effort required to write this pipeline is astonishingly low:
-  it amounts to 240 characters 
-(Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.)
-
+Only the last of these tasks is the real focus of the chapter, demonstrate the benefits of using distributed analytics system such as Hadoop.
+
+For \sys, we consider the full pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
 The pipeline starts with a few stages in \sta  that generate URLs;
-  these URLs point to index files, each one of which contains pointers 
-The stream of index files is analyzed (4 stages) to extract the URLs of the compressed data files.
-It is then passed again through a download command
-After downloading and uncompressing these files, the stream is analyzed to extract the maximum temperature.
-
-A less obvious lesson from this case was that  
-First, large, complex pipelines enable significant freedom in how to express some stages.
-Several stages in the original program are compressed into a single \ttt{awk} script;
-  unfortunately, \ttt{awk} is too general to have a helpful distributability signature.
-For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
-  this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
-
-The Hadoop program 
-Hadoop 3.2.1
-openjdk 11.0.5
-ruby 2.5.5p157
-
-few different ways to write some phases, 
-  for example, 
-A large pipeline resulted in a couple of 
-An interesting lesson here was that there are multiple
-First, longer pipelines can be written in multiple ways
-It is worth noting
-These are
+  these URLs point to index files, each one of which contains pointers to compressed data files.
+The stream of index files is then analyzed (4 stages) to extract the URLs of the data files, passed again through the command for downloading data.
+After downloading and uncompressing these files, the stream is analyzed to extract the maximum temperature (another 4 stages).
+
+% The first interesting result 
+The effort required to write this pipeline is astonishingly low:\footnote{
+  Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.
+}
+  it amounts to 14 stages and, when expressed as a single line, only 240 characters.
+The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 117 lines of Java code---not accounting for code moving input and output in and out of HDFS.
+
+In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1:09 for moving the dataset to HDFS.
+For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
 
@@
 % Dish p3 me platos pipeline 10 => real	1m38.945s
+
+A less obvious take-away from this case was that large, complex pipelines enable significant freedom in how to express their stages.
+Several stages in the original shell program (taken from the book) are expressed as a single \ttt{awk} script;
+  unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
+A much simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
+  but might not perform as well in the centralized case.
+% For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
+%   this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
+
+% few different ways to write some phases, 
+%   for example, 
+% A large pipeline resulted in a couple of 
+% An interesting lesson here was that there are multiple
+% First, longer pipelines can be written in multiple ways
+% It is worth noting
+% These are
+%
 

### 2019-11-23 pldi/draft.tex
Nits

@@
 
-The paper is structured as follows.  It first starts with a background
+The paper is structured as follows.  It starts with a background
 section outlining pipeline concepts and overviewing \sys~\sx{bg}.
@@
 To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory program of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
-At a high level, the goal is to process multi-year weather data from NCDC to identify the temperature over the past several years.
+At a high level, the goal is to process multi-year weather data from NCDC to identify the maximum temperature over the past several years.
 The original task, as presented in the book, can be thought of comprising three sub-tasks:
@@
 
-A less obvious take-away from this case was that large, complex pipelines enable significant freedom in how to express their stages.
+A less obvious take-away from this case is that large, complex pipelines enable significant freedom in how to express their stages.
 Several stages in the original shell program (taken from the book) are expressed as a single \ttt{awk} script;

### 2019-11-24 pldi/draft.tex
Minor changes

@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
+cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
   sort | uniq -c | sort -rn | head 5 > out
@@
 \heading{Command-line Options}
-As described earlier~\sx{bg}, command-line options and environment variables can affect the operation of a command.
+As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
 Options are a particularly popular way to control a command, 
@@
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use a combination of micro-benchmarks and one macro-benchmark, most of which were collected from prior literature and online repositories.
+To achieve this, we use a combination of micro-bench\-marks and one macro-bench\-mark, most of which were collected from prior literature and online repositories.
 Micro-bench\-marks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
@@
 It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-  as shown if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
+  as shown, if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
 
@@
 First and foremost, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
+\sys's transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
 As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
@@
   it amounts to 14 stages and, when expressed as a single line, only 240 characters.
-The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 117 lines of Java code---not accounting for code moving input and output in and out of HDFS.
-
-In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1:09 for moving the dataset to HDFS.
+The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 137 lines of Java code---not accounting for code moving input and output in and out of HDFS.
+
+In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.
 For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
@@
   unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A much simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
-  but might not perform as well in the centralized case.
+A much simpler pipeline that builds on basic \unix primitives might be out-performed by specialized programs on a single node, but turns out to be  trivially scalable to multiple nodes.
 % For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;

### 2019-11-24 pldi/draft.tex
Fixes to intro

@@
   Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed frameworks, or source-level annotations.
-  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
+  This paper presents \sys, a shell variant that automatically and correctly scales out \unix shell pipelines. 
   Its insight is that pipelines already express stream computations that can be automatically distributed.
@@
     applies rewriting rules for largest possible subprograms,
-    and orchestrates the execution of the distributed program.
-Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
+    and orchestrates their execution.
+Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
 \end{abstract}
@@
    scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
-
+% 
 To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
@@
 Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+  for a user with \emph{one} computer and a small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
@@
 
-Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
+Unfortunately, developing $p_1$'s distributed equivalent requires significant effort.
 % For a user with many computers and larger inputs, 
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
 More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
@@
 
-To answer this question, this paper presents \sys, a shell variant that transforms traditional shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that a large part of the shell language can be encoded using a dataflow graph model, providing structure to distribute computations.
+To answer this question we develop \sys, a shell variant that transforms traditional \unix pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
+The key insight behind \sys is that a large part of the shell language can be encoded in a distributed dataflow graph model, providing structure to distribute computations.
 %% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
@@
 \sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr}s in parallel streams,
+A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
   use a mostly-parallel tree of \ttt{sort}s,
   run \ttt{uniq} mostly in parallel,
-  \ttt{head} in one of the streams,
+  execute \ttt{head} in one of the streams,
   and (trivially) merge streams in \ttt{out}.
@@
 First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+Second, developers of new shell commands can use a lightweight domain-specific language to express their distributability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
+Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 % Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
@@
 
-The paper is structured as follows.  It starts with a background
-section outlining pipeline concepts and overviewing \sys~\sx{bg}.
+The paper is structured as follows.
+It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
 Sections \ref{distributability}--\ref{impl} highlight key
@@
   \item
-  \S\ref{distributability} describes several scalability and distributability classes and introduces the distributability characterization language.
+  \S\ref{distributability} studies several distributability classes and introduces the distributability characterization language.
 
   \item
-  \S\ref{ir} presents a dataflow graph model that can encode shell programs, and a set of parallelization transformations that preserve its semantics.
+  \S\ref{ir} presents a dataflow graph model for encoding shell pipelines, and a set of parallelization transformations that preserve the semantics of the sequential program.
 
   \item
-    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of shell programs.
+    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of distributed shell programs.
     %% details other concerns, such as distributed state management and extensibility. 
@@
 \noindent
-\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
+\sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$) for parallel execution using a combination of real pipelines.
+%micro-bench\-marks %---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
 After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.

### 2019-11-24 pldi/draft.tex
Update background and overview

@@
 
-This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation.
+This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation~\sx{bg:overview}.
 
@@
 Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
-These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
-Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
+These programs produce output and consume input concurrently and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
+Intermediate programs, termed \emph{commands}, can also be developed in different programming languages, as long as they conform to a common interface.
 %\begin{wrapfigure}{r}{0.06\textwidth}
@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-  ls -l | wc -l                          [($$p_2$$)]
+  ls -l | wc -l                           ($$(p_2)$$)
 \end{lstlisting}
 \smallskip
-Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
+Commands can be tuned by two mechanisms: com\-mand-line options and environment variables.
 Options control the execution of a program---\eg changing the \ttt{sort}ing order.
 The shell does not have any visibility into these options; 
-  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
-Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
-They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
+  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
+Environment variables carry more general information about the surrounding system---\eg search paths, system defaults, parsing switches \etc 
+They are organized as a map from names to values, both of which are strings, and can be evaluated at any point in the program---\eg as commands, flags, or inputs.
+% a shell feature termed parameter \emph{expansion}.
 
 % TODO: Talk about security primitives?
-Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
-  files are simply persistent streams that can be manipulated by commands or pipelines.
+Streams are a key abstraction in \unix, and go hand-in-hand with another \unix abstraction: \emph{files}.
+Files are persistent streams that can be manipulated by individual commands or pipelines:
 % \begin{wrapfigure}{r}{0.15\textwidth}
@@
 % \end{wrapfigure}
-Reading a file or a set of files generates a stream; 
-  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-Files are named, in the sense that they correspond to a global identifier---for example, \ttt{/x/y} corresponds to a persistent stream.
-\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
+  reading a (set of) file(s) generates a stream; 
+  conversely, redirecting to a file identifier generates a file.
+% streams can be trivially redirected to the file system---by appending with the file identifier.
+Files are named, corresponding to global identifiers---\eg \ttt{/x/y} corresponds to a persistent stream at the location \ttt{/y} in \ttt{/x}.
+\unix exposes a few different types of files---\eg directories, symbolic links, and named pipes (or FIFOs). % , and special-character devices.
 As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
+  ls | sed 's/1/2/' | > out.t 2> err.t    ($$(p_3)$$)
 \end{lstlisting}
@@
 \noindent
-Special files are particularly interesting because they are resolved to OS structures.
+Special files are particularly interesting because they resolve to OS structures.
 One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
@@
 Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
-This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
+This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---enables the runtime wrapping and rebinding achieved during \sys's transformations.
+% TODO: compatibility
 
@@
 Some of these features are built into the shell;
-  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
+  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file---which itself can be a FIFO dubbing as input to another command.
 % \begin{wrapfigure}{r}{0.15\textwidth}
@@
 % \end{wrapfigure}
-As described before, the file can be a FIFO dubbing as input to another command.
 More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
+  cat * | tee >(wc -l > l.t) | zip > a.z   ($$(p_4)$$)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
+  read | { calc | square; } >/dev/fd/0     ($$(p_5)$$)
 \end{lstlisting}
@@
 
-\unix pipelines are surprisingly powerful, a feature that \sys exploits by re-targeting shell scripts during its rewriting passes that target shell.
+To conclude, \unix pipelines are surprisingly powerful.
+This power is exploited directly by \sys's rewriting passes which re-target the shell rather than an implementation in Java or Go.
 
@@
 \subsection{\sys Design Overview}
+\label{bg:overview}
 
@@
  (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator is candidate for both task and data parallelism, and
+ (iii) the pipe operator is a candidate for both task and data parallelism, and
 % \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
@@
 % However, understanding the shell's primitives is not enough
-\sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
+First, \sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
 This challenge can be broken down into two parts.
-First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
+First, understand standard commands part of the shell---that is, built-ins that come with any shell and support its core functions.
 One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
-\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
-This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.
-
-The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
-Commands are arbitrary programs, but still conform to the shell's pipeline interface.
-To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
-Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
-A command developer can only annotate the default behavior;
-  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
-
-Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying  distributable regions.
-It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
+\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes~\sx{cmd}.
+This group is already useful-enough to allow the composition of practical pipelines~\sx{micro}.
+
+The second part is understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power.
+New commands are arbitrary programs, but still need to conform to the shell's pipeline interface.
+To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
+This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the distributability of the commands they are developing.
+% We haven't introduced the notion of flags
+% A command developer can only annotate the default behavior;
+%   if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
+
+Given a pipeline of commands and their distributability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
+It addresses this challenge with a series of rewriting passes that perform dataflow analysis to identify distributable regions.
+It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
 These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc
 
-Finally, \sys needs to schedule and execute the distributed dataflow graph.
-To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
-First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
-  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
-The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
+To schedule and execute the resulting distributed dataflow graph, \sys needs to solve several challenges~\sx{impl}.
+First, it needs to wrap single-node commands with corrective runtime interposition based on their classes.
+This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the distributability specification described earlier.
+Thin wrappers avoid interference with legacy functionality, as \unix commands contain non-obvious logic and flags added over extended periods of time.
 
 Second, \sys needs to fill-in planing details left blank by the rewriting pass.
-This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
-From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
-Planner activators achieves the latest possible binding, when critical runtime information is available.
+This is achieved with \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
+They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is the empty character).
+They make clever use of shell features such as command expansion to achieve the latest possible binding, when critical runtime information is available.
+% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
+Planner activators
+
 % FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}
 
-The next few sections discuss \sys's design and implementation details.
+The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
 They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.

### 2019-11-25 pldi/draft.tex
Update section 3

@@
 Second, \sys needs to fill-in planing details left blank by the rewriting pass.
-This is achieved with \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
-They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is the empty character).
-They make clever use of shell features such as command expansion to achieve the latest possible binding, when critical runtime information is available.
+This is achieved by \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
+They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is empty, not affecting pipelines).
+They make clever use of shell features such as expansion to achieve the latest possible binding, when critical runtime information becomes available.
 % From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
-Planner activators
 
@@
 The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
-They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
+% They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
 
@@
 
-To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
+To have any hope of distributing pipelines, \sys first needs to understand the distribution characteristics of their individual commands.
 This understanding is encoded in a command's \emph{distributability} class.
-Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
-It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
-\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
+Informally, a distributability class represents the level of synchronization required by copies of a command that execute in parallel.
+It focuses on capturing characteristics that are important for distribution, rather than the entire the observable behavior of a command.
+\sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
 % Make the connections that this is like a type, that doesn't capture everything, but is good-enough
 
-This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
-Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
+This section starts by defining these classes, along with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+Building on this study, it develops a DSL that enables command classification by its developers~\sx{ext}.
 In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.
@@
   later classes require more effort.
-In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
 
@@
 The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
-These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
-At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
+These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
+At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
 Workloads that use only stateless commands are trivial to parallelize:
-  they do not require any synchronization to maintain correctness, nor caution about where to split.
+  they do not require any synchronization to maintain correctness, nor caution about where to be split.
  % map :: (a -> b) -> [a] -> [b]
 
-Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
-This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
-However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.
+Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters within a line, one at a time).
+This property allows further parallelization, a feature that might seem of limited use as these commands are computationally inexpensive, precisely due to their narrow focus.
+However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
 
@@
 The second class contains pure commands \pur.
-These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
-The details of this state and its propagation during piecewise element processing affects their distributability characteristics.
+These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
+The details of this state and its propagation during piecewise element processing affect their distributability characteristics.
 
@@
 
-Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
-Non-streaming constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques, that is implementing it as a map and a reduce function.
+Often these commands do not operate in an online fashion, but need to block until the end of a stream.
+A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
+Such constraints affect task parallelism, but not data parallelism:
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $reduce$ function.
 
@@
 This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
-In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.
+In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.
 
@@
 Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
-Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
-The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
+Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for all the commands in \dfs.
+The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---\eg \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
 % these operate by changing an inode
@@
 Operations using different identifiers pose no problems;
-  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
-While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).
+  operations sharing a single identifier might introduce problems---but these risks exist even in the non-distributed case.
+Such pathological cases are indeed possible, but are rare enough to not pose a concern---and have been, since \unix's inception~\cite[\S3.6]{unix}.
 
@@
 A command in \sid will require some form of concurrency control to distribute property.
-However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.
-
-The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
+However, because its side-effects stay within the distributed \unix environment, they can be manipulated using a transactional control mechanism---including optimistic and multi-versioning schemes that have the ability to roll-back side-effects.
+
+The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
 For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
-Even if \sid commands are protected with transactions
-  these benefits are even greater for read-only or read-mostly accesses.
+The benefits of a scalable transactional scheme are even greater when the workload features only or mostly read accesses.
 
 \heading{Irreversibly Side-effectful Commands}
-Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
-Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
+Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
+Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.
 
 \heading{Discussion}
-There are several technical details about \sys's design decisions worth noting.
+There are a few technical details about \sys's design decisions worth noting.
 
@@
 
-Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
-Other commands are treated as  non-distributable for now~\sx{discussion};
-  by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.
-Even with only these two classes, however, \sys supports highly useful pipelines that are more expressive than ones in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
+The focus for the rest paper will be commands in \sta and \pur.
+\sys currently treats other commands as non-distributable~\sx{discussion}.
+In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
+Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
 

### 2019-11-27 pldi/draft.tex
Polish technical report

--- 2019-11-24 pldi/draft.tex
@@
   Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed frameworks, or source-level annotations.
-  This paper presents \sys, a shell variant that automatically and correctly scales out \unix shell pipelines. 
+  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
   Its insight is that pipelines already express stream computations that can be automatically distributed.
@@
     applies rewriting rules for largest possible subprograms,
-    and orchestrates their execution.
-Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
+    and orchestrates the execution of the distributed program.
+Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
 \end{abstract}
@@
    scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
-% 
+
 To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
@@
 \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
+cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
   sort | uniq -c | sort -rn | head 5 > out
@@
 Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and a small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
 % composition allows the composition of general primitives
@@
 
-Unfortunately, developing $p_1$'s distributed equivalent requires significant effort.
+Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
 % For a user with many computers and larger inputs, 
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
 More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
@@
 
-To answer this question we develop \sys, a shell variant that transforms traditional \unix pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that a large part of the shell language can be encoded in a distributed dataflow graph model, providing structure to distribute computations.
+To answer this question, this paper presents \sys, a shell variant that transforms traditional shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
+The key insight behind \sys is that a large part of the shell language can be encoded using a dataflow graph model, providing structure to distribute computations.
 %% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
@@
 \sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
+A \sys-enabled $p_1$ will run \ttt{tr}s in parallel streams,
   use a mostly-parallel tree of \ttt{sort}s,
   run \ttt{uniq} mostly in parallel,
-  execute \ttt{head} in one of the streams,
+  \ttt{head} in one of the streams,
   and (trivially) merge streams in \ttt{out}.
@@
 First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Second, developers of new shell commands can use a lightweight domain-specific language to express their distributability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
+Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
 % Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
@@
 
-The paper is structured as follows.
-It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
+The paper is structured as follows.  It starts with a background
+section outlining pipeline concepts and overviewing \sys~\sx{bg}.
 Sections \ref{distributability}--\ref{impl} highlight key
@@
   \item
-  \S\ref{distributability} studies several distributability classes and introduces the distributability characterization language.
+  \S\ref{distributability} describes several scalability and distributability classes and introduces the distributability characterization language.
 
   \item
-  \S\ref{ir} presents a dataflow graph model for encoding shell pipelines, and a set of parallelization transformations that preserve the semantics of the sequential program.
+  \S\ref{ir} presents a dataflow graph model that can encode shell programs, and a set of parallelization transformations that preserve its semantics.
 
   \item
-    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of distributed shell programs.
+    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of shell programs.
     %% details other concerns, such as distributed state management and extensibility. 
@@
 \noindent
-\sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$) for parallel execution using a combination of real pipelines.
-%micro-bench\-marks %---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
+\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
 After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
@@
 \heading{Command-line Options}
-As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
+As described earlier~\sx{bg}, command-line options and environment variables can affect the operation of a command.
 Options are a particularly popular way to control a command, 
@@
 standard input. Similarly, the output is either its standard output or
-a file argument. An example record for comm can be seen below:
+a file argument. An example record for \ttt{comm} can be seen below:
 
@@
 
-\tr{We don't need to designate the stderror, because we assume that it
+\tr{\kk{We don't need to designate the stderror, because we assume that it
   is never the main output of a command and it will never be used by
@@
   wants to do this, they can just use a redirect I think to get around
-  it.}
-
-\tr{If we talk about the language it would be good to give some
+  it.}}
+
+\tr{\kk{If we talk about the language it would be good to give some
   statistics about how many commands (out of the ones that we have
@@
   many lines of code it took to categorize all the commands that we
-  did.}
-
-\tr{Note: Since a categorization language will probably not be
+  did.}}
+
+\tr{\kk{Note: Since a categorization language will probably not be
   complete (especially the one showing the input arguments) there
@@
   enough, like a python function that categorizes the command, and
-  returns its input argument if it is stateless.}
-
-\tr{(Maybe) We should have a crisp point about why we have this categorization
+  returns its input argument if it is stateless.}}
+
+\tr{\kk{(Maybe) We should have a crisp point about why we have this categorization
   language, and why don't we just allow someone to write a function in
@@
   limited. All of these arguments are a little bit weak though. Niko,
-  what do you think?}
+  what do you think?}}
 
@@
 \noindent
-Together with the commands in \sta, that are always easy to
-parallelize, there also exists a subclass of commands in \pur that can
-be parallelized by separating them into a map and a reduce
-operation. In order to allow command developers to extend \sys with
-information about how to parallelize new pure commands, \sys offers a
-programmatic interface to specify how to distribute a pure command.
-
-\tr{I am not sure a general interface is so easy to design. It needs
+In addition to the commands in \sta that can be parallelized in a
+straightforward way, there also exists a subclass of commands in \pur
+that can be parallelized using divide-and-conquer techniques (as
+mentioned in \cref{cmd}). In order to support the parallelization of
+arbitrary commands in this category, \sys allows command developers to
+provide a custom map and reduce step for these commands, where the map
+is in \sta and the reduce in \pur, and their composition produces the
+same output with the initial command. \sys can then replace occurences
+of these commands with their map and reduce parts, enabling their
+parallelization.
+
+\tr{\kk{I am not sure a general interface is so easy to design. It needs
   more though. It might be beneficial to just talk about sort and wc
@@
   a python function that given a node of the graph, transforms it into
-  many. I am not sure what is best...}
+  many. I am not sure what is best...}}
 
@@
 
+\kk{Move some variation of that to intro/overview:
+
+In the core of \sys there is a dataflow graph model that is inspired
+by the ones that are used in popular distributed stream processing
+systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
+\cite{spark:12}). These systems exploit data parallelism by
+parallelizing nodes of the dataflow graph. Unfortunately, the
+optimizations in many of these systems do not preserve the dataflow
+model semantics, blurring the lines between specification,
+optimization, and implementation. In contrast, our dataflow model
+inherently supports the data parallelism that is found in shell
+programs. More precisely, data parallelism can be easily exposed as
+task parallelism in our model. We exploit that by developing a set of
+parallelization-exposing optimizations that preserve the dataflow
+model semantics, and can be represented as graph transformations. The
+\sys optimizer then applies these transformations in an iterative way,
+optimizing the program to exploit the available computational
+resources.
+
+%% The standard dataflow graph exposes task parallelism. Different nodes
+%% in the graph can be executed concurrently, communicating through the
+%% edges.
+%% Task level parallelism is already exposed by the shell, by using
+%% pipes \ttt{|} and the background operator \ttt{\&}.
+%% Unfortunately, simple task parallelism leaves a lot to be desired,
+%% since scaling a computation requires manually introducing more
+%% operators.
+
+%% In order to adequately utilize the available computational resources,
+%% a distributed implementation should exploit data parallelism. Data
+%% parallelism is based on the observation that some operations produce
+%% the same result when performed independently on subsets of the input
+%% data, merging the results afterwards. Exploiting data parallelism is
+%% one of the major reasons that popular distributed stream processing
+%% systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
+%% \cite{spark:12}) achieve significant performance.
+
+%% A very important consequence of this is that we can iteratively apply
+%% optimizations that expose data parallelism, or even revert them when
+%% the resulting dataflow graph is too wide. This flexibility unifies the
+%% unoptimized and optimized versions of a computation, enabling thinking
+%% about them with the same conceptual tools.
+}
+
 In order to effectively apply optimizations and distribute a shell
-script, it first has to be translated to a manipulable
-representation. As mentioned in \Cref{bg:pipelines}, the two main
-abstractions of the shell, are files that contain data, and commands
-that communicate through these files. We propose a slight variation of
-the standard dataflow graph model, that is commonly found in stream
-processing systems, as the representation of a distributed shell
-computation. The main benefit of the dataflow graph model is that it
-clearly exposes task parallelism, as different nodes can be
-independent processing workers. In our case each node represents a
-command, and each edge represents a file.
+script, it first has to be translated to a manipulable representation.
+In \Cref{graph-components} we propose a slight variation of the
+standard dataflow graph model as a candidate intermediate
+representation. The main benefit of our model is that it exposes both
+task and data parallelism. In \Cref{ir:transformations} we define a
+set of optimizations that expose and exploit data parallelism of
+commands in \sta and \pur as semantics-preserving graph
+transformations.
 
 \subsection{Graph Components}
+\label{graph-components}
+
+As mentioned in \Cref{bg:pipelines}, the two main abstractions of the
+shell are files that contain data, and commands that communicate
+through these files. Based on that, we define nodes in our graph model
+to represent commands, and edges to represent files.
 
@@
 
-As mentioned above, the dataflow graph exposes task
-parallelism. Different nodes in the graph can be executed
-concurrently, communicating through the edges.
-%% Task level parallelism is already exposed by the shell, by using
-%% pipes \ttt{|} and the background operator \ttt{\&}.
-Unfortunately, simple task parallelism leaves a lot to be desired,
-since scaling a computation requires manually introducing more
-operators.
-
-In order to adequately utilize the available computational resources,
-a distributed implementation should exploit data parallelism. Data
-parallelism is based on the observation that some operations produce
-the same result when performed independently on subsets of the input
-data, merging the results afterwards. Exploiting data parallelism is
-one of the major reasons that popular distributed stream processing
-systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
-\cite{spark:12}) achieve significant performance. Unfortunately, the
-optimizations in many of these systems do not preserve the dataflow
-model semantics, blurring the lines between specification,
-optimization, and implementation.
-
-In contrast, our dataflow model inherently supports the data
-parallelism that is found in shell programs. More precisely, data
-parallelism can be easily exposed as task parallelism in our model. A
-very important consequence of this is that we can iteratively apply
-optimizations that expose data parallelism, or even revert them when
-the resulting dataflow graph is too wide. This flexibility unifies the
-unoptimized and optimized versions of a computation, enabling thinking
-about them with the same conceptual tools.
-
 Before describing the different types of parallelization
@@
 
-\tr{This is a nice characterization, which I think is true in most
-  settings. There is a caveat though: You can have stateless programs
-  that emit output when they start (before they consume the stream)
-  and/or when they consume the EOF symbol. In this case I am not sure
-  the equation above holds.}
-
-Similarly, it was mentioned that some pure commands (such as
-\ttt{sort} and \ttt{wc}) can be parallelized using divide-and-conquer
-techniques. More formally, there pure commands $f$ can be implemented
-as a map $m$ and a reduce $r$ function, satisfying the following:
+%% This is a nice characterization, which I think is true in most
+%% settings. There is a caveat though: You can have stateless programs
+%% that emit output when they start (before they consume the stream)
+%% and/or when they consume the EOF symbol. In this case I am not sure
+%% the equation above holds.
+
+\noindent
+This means that we get the same output by applying a command $f$ to a
+concatenation of two inputs $x, x'$, or by concatenating the result
+produced by the application of $f$ to these inputs
+separately. Similarly, it was mentioned that some pure commands (such
+as \ttt{sort} and \ttt{wc}) can be parallelized using
+divide-and-conquer techniques. More formally, there pure commands $f$
+can be implemented as a map $m$ and a reduce $r$ function, satisfying
+the following:
 
@@
 \]
+
+\noindent
+Similarly to the above, this means that we get the same output by
+applying $f$ to a concatenation of two inputs $x, x'$, or by applying
+the reduce function $r$ to the concatenation of the outputs produced
+by applying map $m$ to each of these inputs.
 
@@
 
+In this section, we describe challenges related to the implementation
+of \sys. More concretely, in \Cref{front-end} we describe how the \sys
+front-end translates a shell script to the dataflow graph that was
+defined in \Cref{ir}, and in \Cref{optimizer}, we describe how the
+\sys optimizer distributes and implements a given dataflow graph.
+
+\subsection{Front End}
+\label{front-end}
+
 As mentioned in \Cref{ir}, in order to generate the distributed
-implementation of a shell script, \ttt{Dish} first translates it to
-a dataflow graph. However, there are several components in shell
+implementation of a shell script, \ttt{Dish} first translates it to a
+dataflow graph. However, there are several components in shell
 programs that cannot be arbitrarily parallelized without affecting the
-program behaviour, e.g. commands producing an irreversible effect, or
-commands connected with a \ttt{\&\&} operator. In addition, the
-fact that the shell is highly dynamic, further increases the
-challenge of correctly identifying what parts can be distributed.
+program behaviour, e.g. commands connected with a \ttt{\&\&}
+operator. In addition, the fact that the shell is highly dynamic,
+further increases the challenge of correctly identifying what parts
+can be distributed, e.g. by not being able to know what file does an
+unexpanded command argument refer to.
 
@@
 
-\subsection{Distributable Regions}
-
+\heading{Distributable Regions}
+%
 %% \kk{The goal of this subsection is to explain that Dish does a safe
@@
 %%   parallelize.}
-
+%
 Consider the following example:
@@
 the program that would be allowed to execute concurrently by different
-processes in the POSIX standard~\cite{posix}.
+processes in the POSIX standard~\cite{posix}. Consider the following
+illustrative example:
 
@@
 
-\subsection{Translation Pass}
-
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+  (tr ... f1 > g1 && tr ... f2 >> g1) &
+  grep ``foo'' g1
+\end{lstlisting}
+
+In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
+  g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
+the dataflow graph that corresponds to the whole program, has two
+nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
+  g1}.
+
+\heading{Translation Pass}
+%
 Dish performs a depth first search pass on the AST of the given shell
 program, extending the distributable regions from the bottom up,
-translating their independent components to dataflow graph
-nodes, until a barrier construct is reached. Note that barrier
-constructs can exist in a dataflow graph, as long as they don't
-extend beyong a single node. An illustrative example follows:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  (sort f1 > g1 && sort f2 > g2) &
-  sort -m g1 g2
-\end{lstlisting}
-
-In the above example, while \ttt{sort f1 > g1}, \ttt{sort f2 > g2},
-and \ttt{sort -m g1 g2} are all in different dataflow graphs, the
-dataflow graph that corresponds to the whole program, has two nodes,
-\ttt{sort f1 > g1 \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.
-
-In order to identify possible opportunities for parallelism, the
-translation pass also identifies the category of each node, together
-with its input and output, using records written by developers using
-the language that was described in \Cref{ext}. Some interesting cases
-of the analysis algorithm are shown in pseudocode below:
+translating their independent components to dataflow graph nodes,
+until a barrier construct is reached. In order to identify possible
+opportunities for parallelism, the translation pass also identifies
+the category of each node, together with its input and output, using
+records written by developers using the language that was described in
+\Cref{ext}. Some interesting cases of the analysis algorithm are shown
+in pseudocode below:
 
@@
 
-\subsection{Dish Optimizer}
+\subsection{Optimizer}
+\label{optimizer}
 
@@
 %
-\subsection{Merge and Split Implementation}
-
+
+\heading{Merge and Split Implementation}
+%
 It is common in the dataflow graph for a node to have many incoming
@@
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use a combination of micro-bench\-marks and one macro-bench\-mark, most of which were collected from prior literature and online repositories.
+To achieve this, we use a combination of micro-benchmarks and one macro-benchmark, most of which were collected from prior literature and online repositories.
 Micro-bench\-marks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
@@
 It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-  as shown, if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
+  as shown if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
 
@@
 First and foremost, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
+\sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
 As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
@@
   it amounts to 14 stages and, when expressed as a single line, only 240 characters.
-The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 137 lines of Java code---not accounting for code moving input and output in and out of HDFS.
-
-In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.
+The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 117 lines of Java code---not accounting for code moving input and output in and out of HDFS.
+
+In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1:09 for moving the dataset to HDFS.
 For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
@@
   unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A much simpler pipeline that builds on basic \unix primitives might be out-performed by specialized programs on a single node, but turns out to be  trivially scalable to multiple nodes.
+A much simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
+  but might not perform as well in the centralized case.
 % For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;

### 2019-11-28 pldi/draft.tex
Update section 3

--- 2019-11-27 pldi/draft.tex
@@
 %%   distribution of shell scripts}
-As a result, any static characterization of shell commands is unsatisfactory as it would quickly become obsolete, supporting only a small subset of commands.
+As a result, any static characterization of shell commands is unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
 This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
@@
 
-To address additions and changes to the set of commands available to the shell, \sys provides command developers a lightweight, domain-specific language (DSL).
-The DSL allows developers to express semantic information about a command's distributanility that is critical to \sys.
-This DSL can be used by the developers of new commands to succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
-It could also be used by developers that extend or maintain existing commands, to express additions or changes to the command's implementation or interface.
+To address % additions and changes to the set of commands available to the shell,
+  this challenge, \sys arms command developers with a lightweight, domain-specific language (DSL).
+The DSL allows developers to express semantic information about a command's distributability that is critical to \sys.
+It can be used by the developers of new commands to succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
+It can also be used by maintainers extending or maintain existing commands, to express additions or changes to the command's implementation or interface.
 
@@
 By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
-Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta.
+Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta because it generates a static stream regardless of other input.
 % TODO: closure?
@@
 \heading{Command Inputs and Outputs}
-%
-Consider the command \ttt{comm}. It takes as input two files and
+A second concern captured by the DSL is the order of inputs and outputs.
+Commands generally read multiple input streams (and, at cases, write to multiple output streams).
+The invariants between inputs and outputs must be maintained over the distributed execution of the program.
+% but these maintain different invariants, enabling very different distributability characteristics.
+
+Consider the command \ttt{comm} that takes as input two files and
 performs a join operation. When invoked without any option, it
-produces three-column output. The first column contains lines that
-unique to the first file, the second column contains lines that are
+produces a three-column output. The first column contains lines that
+are unique to the first file, the second column contains lines that are
 unique to the second file, and the last column contains lines that
@@
 \ttt{-2} and \ttt{-3} supress the second and third column. Using both
-\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command becomes \sta
+\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command jumps to \sta
 regarding its first (resp. the second) file argument.
@@
 \sys's distributability description DSL (Fig.~\ref{fig:dsl}) is
-designed to be able to classify a commands into multiple categories
+designed to classify commands into multiple categories
 given their options.
@@
   option as it is most often used to designate that a command reads
-  from its standard input}). A file written in the language contains a
+  from its standard input}). A file written in the DSL contains a
 record containing a mapping from predicates to assignments for each
@@
 category, the sequence of its input sources, and its output
-source. Inputs to commands are a sequence of file arguments and its
+source. Inputs to commands are a sequence of file identifiers and its
 standard input. Similarly, the output is either its standard output or
-a file argument. An example record for \ttt{comm} can be seen below:
+a file identifier. An example record for \ttt{comm} can be seen below:
 
@@
 arbitrary commands in this category, \sys allows command developers to
-provide a custom map and reduce step for these commands, where the map
+provide custom map and reduce steps for them, where the map
 is in \sta and the reduce in \pur, and their composition produces the
-same output with the initial command. \sys can then replace occurences
+same output with the initial command. \sys can then replace occurrences
 of these commands with their map and reduce parts, enabling their

### 2019-11-28 pldi/draft.tex
Update graph components

@@
 % \end{lstlisting}
-\smallskip
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \smallskip
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
@@
 \end{lstlisting}
-\smallskip
+% \smallskip
 % \noindent
@@
 % Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-\smallskip
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \smallskip
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
   ls -l | wc -l                          [($$p_2$$)]
 \end{lstlisting}
-\smallskip
+% \smallskip
+
 Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
@@
 % (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-\smallskip
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \smallskip
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
   ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
 \end{lstlisting}
-\smallskip
+%\smallskip
 
@@
 
-\smallskip
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \smallskip
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
   cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
 \end{lstlisting}
-\smallskip
+% \smallskip
 
@@
 
-\smallskip
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \smallskip
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
   read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
 \end{lstlisting}
-\smallskip
+% \smallskip
 
@@
 
-\kk{Move some variation of that to intro/overview:
-
-In the core of \sys there is a dataflow graph model that is inspired
-by the ones that are used in popular distributed stream processing
-systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
-\cite{spark:12}). These systems exploit data parallelism by
-parallelizing nodes of the dataflow graph. Unfortunately, the
-optimizations in many of these systems do not preserve the dataflow
-model semantics, blurring the lines between specification,
-optimization, and implementation. In contrast, our dataflow model
-inherently supports the data parallelism that is found in shell
-programs. More precisely, data parallelism can be easily exposed as
-task parallelism in our model. We exploit that by developing a set of
-parallelization-exposing optimizations that preserve the dataflow
-model semantics, and can be represented as graph transformations. The
-\sys optimizer then applies these transformations in an iterative way,
-optimizing the program to exploit the available computational
-resources.
+The core of \sys is a dataflow graph model, inspired by the ones used in popular distributed stream processing systems~\sx{related}.
+% such as MapReduce \cite{mapreduce:08} and Apache Spark \cite{spark:12}).
+These systems introduce data parallelism by parallelizing nodes of the dataflow graph.
+Unfortunately, % their optimizations in many of these systems 
+ they often do not preserve the model's semantics, blurring the lines between specification, optimization, and implementation.
+In contrast, \sys's dataflow model inherently supports the data parallelism found in shell pipelines.
+More specifically, it develops a set of parallelization-exposing optimizations represented as semantics-preserving graph transformations, effectively exposing data parallelism as task parallelism.
+The \sys optimizer then applies these transformations in an iterative way, optimizing the program to exploit the available computational resources.
 
@@
 %% about them with the same conceptual tools.
-}
-
-In order to effectively apply optimizations and distribute a shell
-script, it first has to be translated to a manipulable representation.
-In \Cref{graph-components} we propose a slight variation of the
-standard dataflow graph model as a candidate intermediate
-representation. The main benefit of our model is that it exposes both
-task and data parallelism. In \Cref{ir:transformations} we define a
-set of optimizations that expose and exploit data parallelism of
-commands in \sta and \pur as semantics-preserving graph
-transformations.
+
+In order to effectively apply optimizations and distribute a shell script, \sys first translates it to a manipulable representation.
+\sys first introduces a variant of the standard dataflow graph model as a candidate intermediate representation~\sx{graph-components}.
+The main benefit of this model is that it exposes both task and data parallelism.
+It then defines a set of optimizations that expose and exploit the data parallelism of commands in \sta and \pur as semantics-preserving graph transformations~\sx{ir:transformations}.
 
@@
 
-As mentioned in \Cref{bg:pipelines}, the two main abstractions of the
-shell are files that contain data, and commands that communicate
-through these files. Based on that, we define nodes in our graph model
-to represent commands, and edges to represent files.
-
-\paragraph{Nodes - Commands}
+As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of the
+shell are (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii) commands that communicate
+through these files. Based on that, \sys's graph model defines nodes to represent commands and edges to represent files.
+
+\paragraph{Nodes---Commands}
 
@@
 We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-It is easy to see that $\leq$ is a partial order, and it is often referred to as the \emph{prefix order}.
+It is easy to see that $\leq$ is a partial order, often called the \emph{prefix order}.
 
@@
 This command takes a stream of lines from its standard input and
-returns a stream of lines on its standard output. An important
-observation is that commands in the shell might read their input
+returns a stream of lines on its standard output.
+
+An important
+observation is that commands might read their input
 stream from a sequence of different sources. An illustrative example
-is the command \ttt{cat} which reads its input from a combination
-of files and its standard input.
+is the command \ttt{cat}, which reads its input from a combination
+of files and its standard input:
 
@@
 
-The example above reads file \ttt{x} until it encounters the EOF
-character, then reads from \ttt{stdin}, and then reads from
-\ttt{y}. In order to handle this particularity, nodes in the
-dataflow graph can have many incoming edges, that are ordered in a
-sequence. Note that this does not correspond to standard dataflow
+\noindent
+In the example above, \ttt{cat} reads file \ttt{x} until it encounters the EOF
+character, then reads from \ttt{stdin} until in encounters EOF, and finally reads from
+\ttt{y}.
+
+In order to handle this particularity, nodes in the
+dataflow graph can have many incoming edges, that are \emph{ordered} in a
+sequence. This is different from standard dataflow
 graph representations where the semantics of multiple incoming edges
-is either zipping the inputs in pairs, or arbitraliy interleaving them.
+is either zipping the inputs into pairs, or interleaving them arbitrarily.
 
@@
 
-Note that nodes can read from more files than just the files of their
-stream (or respectively write to more files that their output stream),
-however these files are static (e.g. used for configuration or
+Nodes may read from more files than just the files of their
+input stream (resp. write to more files than their output stream).
+However, these files are static (\eg used for configuration or
 logging) and do not represent streams of data, and therefore are not
-considered part of the dataflow graph. We assume that these static
+considered part of the dataflow graph.
+For now, we assume that these static
 files can only be accessed by one node, thus ensuring that they do not
@@
 
-\paragraph{Edges - Files}
+\paragraph{Edges---Files}
 
@@
 $D\kstar$. They could either refer to a specific named file in the file
-system, or just be FIFO pipes that used for interprocess communication
+system, or just be FIFO pipes used for interprocess communication
 %% \TODO{Are there any other file types, such as URLs, ...?}.
@@
 In order to expose data parallelism from the output of a command, files that
-don't point to a resource can have
+do not point to a resource can have
 \begin{wrapfigure}[3]{r}{0.20\columnwidth}
@@
 \end{wrapfigure}
-an upper bound on the lines that they can transfer. This can be useful to utilize available
+an upper bound on the lines that they can transfer. This helps utilize available
 parallelism after encountering a command that cannot be parallelized
@@
 
-Finally, the edges that don't refer to a resource and do not start
-from a node in the graph represent the graph input, while the edges
-that don't refer to a resource and do not point to a node in the graph
-are its outputs.
+Finally, edges that do not refer to a resource and do not start
+from a node in the graph represent the graph input;
+edges that do not refer to a resource and do not point to a node in the graph are its outputs.
 

### 2019-11-28 pldi/draft.tex
Transformations

@@
 
-\subsection{Graph transformations}
+\subsection{Graph Transformations}
 \label{ir:transformations}
@@
 optimizations, we formalize the intuition about stateless and pure
-commands that was described in \Cref{cmd}.
+commands that was described earlier~\sx{cmd}.
 %% \nv{TODOs
@@
 %
-As mentioned there, stateless commands operate independently on
+
+As mentioned there, stateless commands such as \ttt{tr} operate independently on
 individual elements of the stream (characters, lines, or files)
-without maintaining any state. Some examples include \ttt{grep},
-\ttt{basename}, and \ttt{tr}. Formally, a commmand $f$ is stateless if
-it is ``linear'' with respect to concatenation, i.e. the following
-equation holds:
+without maintaining any state.
+% Some examples include \ttt{grep}, \ttt{basename}, and \ttt{tr}.
+Formally, a command $f$ is stateless if
+it is ``linear'' with respect to concatenation, \ie 
+% the following equation holds:
+satisfying the following equation:
 
@@
 \noindent
-This means that we get the same output by applying a command $f$ to a
-concatenation of two inputs $x, x'$, or by concatenating the result
+This means that one gets the same output by applying a command $f$ to a
+concatenation of two inputs $x, x'$ as by concatenating the result
 produced by the application of $f$ to these inputs
-separately. Similarly, it was mentioned that some pure commands (such
+separately.
+
+Similarly, it was mentioned that some pure commands (such
 as \ttt{sort} and \ttt{wc}) can be parallelized using
-divide-and-conquer techniques. More formally, there pure commands $f$
-can be implemented as a map $m$ and a reduce $r$ function, satisfying
-the following:
+divide-and-conquer techniques. More formally, these pure commands $f$
+can be implemented as a combination of functions map $m$ and reduce $r$, satisfying
+the following equation:
 
@@
 \noindent
-Similarly to the above, this means that we get the same output by
+Similarly to the earlier equation, this means that we get the same output by
 applying $f$ to a concatenation of two inputs $x, x'$, or by applying
@@
 In addition to node splits, we can also perform edge splits in order
-to increase the number of incoming edges for some nodes to expose
-possible data parallelism. If the edge refers to a real file, then the
-edge split represents splitting the real file in chunks, while in
-cases where the edge connects two nodes, it is split to several edges
+to increase the number of incoming edges for some nodes to further expose
+possible data parallelism. In cases where the edge refers to a real file, the
+edge split represents splitting the real file in chunks. In
+cases where the edge connects two nodes, the split effectively refers to several edges
 with a bounded number of lines. Note that the final output edge of a
@@
 
-In this section, we describe challenges related to the implementation
-of \sys. More concretely, in \Cref{front-end} we describe how the \sys
-front-end translates a shell script to the dataflow graph that was
-defined in \Cref{ir}, and in \Cref{optimizer}, we describe how the
-\sys optimizer distributes and implements a given dataflow graph.
+This section describes challenges related to the implementation of \sys. 
+It starts by describing how the \sys front-end translates a shell script to the dataflow graph model defined in \S\ref{ir}, by applying transformation on distributable regions~\sx{front-end}.
+It closes with the \sys planner, which  optimizes, distributes, and orchestrates the execution of a given dataflow graph~\sx{optimizer}.
 
@@
 
-As mentioned in \Cref{ir}, in order to generate the distributed
-implementation of a shell script, \ttt{Dish} first translates it to a
-dataflow graph. However, there are several components in shell
-programs that cannot be arbitrarily parallelized without affecting the
-program behaviour, e.g. commands connected with a \ttt{\&\&}
-operator. In addition, the fact that the shell is highly dynamic,
-further increases the challenge of correctly identifying what parts
-can be distributed, e.g. by not being able to know what file does an
-unexpanded command argument refer to.
+As mentioned in \Cref{ir}, in order to generate the distributed implementation of a shell script, \sys translates it to a dataflow graph.
+However, there are several components in shell programs that cannot be arbitrarily parallelized without affecting the program behaviour---\eg commands connected with a \ttt{\&\&} operator~\sx{bg}.
+The fact that the shell is highly dynamic compounds the challenge---\eg by not knowing what files does an unexpanded command argument refer to.
+% of correctly identifying what parts can be distributed, e.g.
 
@@
 
-In order to address this issue, we introduce the notion of
-distributable program regions, and we design a translation pass that
-given the abstract syntax tree of a shell script identifies
-distributable regions and translates each of them to a dataflow
-graph. The AST of the shell script is produced using a POSIX compliant
-parser~\cite{libdash}.
-
-\heading{Distributable Regions}
+To address this issue, we introduce the notion of distributable program regions and a translation pass for converting them to a dataflow graph.
+The translation pass identifies distributable regions and translates each of them to a dataflow graph, provided the abstract syntax tree (AST) of a shell script.
+The AST is produced by \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
+
+\paragraph{Distributable Regions}
 %
@@
 %
+Distributable regions are (ideally, maximal) program sub-expressions that can be parallelized safely with respect to their sequential execution.
+The structure of a shell program can be used to identify components that can be executed concurrently and barriers that are natural points for enforcing synchronization.
 Consider the following example:
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat f1 f2 | grep ``foo'' > f3 &&
-  sort f3 # Replace with sth better
+  cat f1 f2 | grep ``foo'' > f3 && sort f3 
 \end{lstlisting}
+% # Replace sort with sth better
 
 \noindent
-Notice that the structure of a shell program can be used to identify
-components that can be executed concurrently, and barriers that
-enforce synchronization. More precisely, \ttt{cat f1 f2} and \ttt{grep
-  ``foo''} would already execute concurrently in the standard shell,
-while \ttt{sort f3} would wait for their completion before being
-executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3}
-are maximal distributable regions, in the sense that they cannot be
-extended beyond \ttt{\&\&}.
-
-Concretely, a distributable region cannot permeate through barrier
-constructs (e.g. \ttt{;}, \ttt{\&\&}, and function definitions). On
-the other hand, the pipe operator \ttt{|} and the parallel composition
-operator \ttt{\&} enable parallelization, and extend distributable
-regions. Intuitively, distributed regions correspond to subsections of
-the program that would be allowed to execute concurrently by different
-processes in the POSIX standard~\cite{posix}. Consider the following
-illustrative example:
-
-\tr{I don't know whether I should mention the following: While these
-  control flow constructs can be parallelized and there has been
-  research on it, it is orthogonal to our work, and can be
-  incorporated as future work.}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  (tr ... f1 > g1 && tr ... f2 >> g1) &
-  grep ``foo'' g1
-\end{lstlisting}
-
-In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
-  g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
-the dataflow graph that corresponds to the whole program, has two
-nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
-  g1}.
-
-\heading{Translation Pass}
-%
-Dish performs a depth first search pass on the AST of the given shell
-program, extending the distributable regions from the bottom up,
-translating their independent components to dataflow graph nodes,
-until a barrier construct is reached. In order to identify possible
-opportunities for parallelism, the translation pass also identifies
-the category of each node, together with its input and output, using
-records written by developers using the language that was described in
-\Cref{ext}. Some interesting cases of the analysis algorithm are shown
-in pseudocode below:
+The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would already execute concurrently in the standard shell, while \ttt{sort f3} would wait for their completion before executing.
+Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3} are distributable regions---maximal sub-expressions that cannot be extended beyond \ttt{\&\&}.
+
+Intuitively, distributed regions correspond to subsections of the program that would be allowed to execute concurrently by different processes in the POSIX standard~\cite{posix}.
+The pipe operator \ttt{|} and the parallel composition operator \ttt{\&} enable parallelization, and extend distributable regions.
+Conversely, the \ttt{;} sequential composition operator, the \ttt{\&\&} logical operator, and function definitions represent barrier constructs that do not allow distributable regions to permeate through.
+
+% \tr{I don't know whether I should mention the following: While these
+%   control flow constructs can be parallelized and there has been
+%   research on it, it is orthogonal to our work, and can be
+%   incorporated as future work.}
+% 
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%   (tr ... f1 > g1 && tr ... f2 >> g1) &
+%   grep ``foo'' g1
+% \end{lstlisting}
+% 
+% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
+%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
+% the dataflow graph that corresponds to the whole program, has two
+% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
+%   g1}.
+%
+
+\paragraph{Translation Pass}
+\sys performs a depth-first search pass on the AST of the given shell program.
+During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
+To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
+To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.
+
+A few illustrative cases of the analysis algorithm are shown in pseudocode below:
 
@@
 
-Whenever the algorithm encounters a pipe node, it recursively
-translates its subcomponents, and then merges them in a dataflow graph
-by connecting the output of the first with the input of the second,
-\etc. Before the different dataflow graphs are connected, the
+Whenever the algorithm encounters a distributable node, it recursively
+translates its subcomponents.
+It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
+% what first and second? we didn't say anything about first and second
+Before the different dataflow graphs are connected, the
 algorithm checks that at most one node in the graph writes at every
-file, and at most one node reads from a file that another node writes
+file, and that at most one node reads from a file that another node writes
 to. This is important, in order to avoid inconsistent behaviour after
-distributing nodes in the dataflow graph (i.e. by introducing to
+distributing nodes in the dataflow graph (\ie by introducing to
 concurrent reads and writes to the same file when
-parallelizing). Whenever the algorithm encounter an \ttt{\&\&} node,
-it recursively translates its subcomponents (similarly to the pipe
-node), but then does not merge them in a large dataflow
-graph. Finally, whenever the algorithm encounters a command, it first
+parallelizing).
+
+Whenever the algorithm encounters a barrier node,
+it recursively translates its subcomponents, similarly to the distributable
+case.
+However, different from the distributable case, it does not merge them in a large dataflow graph.
+
+Finally, whenever the algorithm encounters a command, it first
 checks whether it is safe to parallelize it. A command is considered
-to be safe, if it does not make any modification to the shell state
-(e.g. changing the current directory, modifying environment
-variables), and if it does not read from any read-once file
-(e.g. pipe). If the command is considered safe, its category is
-identified and a node containing that command is created and added to
-a singleton dataflow graph. If the command is considered unsafe, it is
-still added to the dataflow graph, but cannot be further parallelized
-by the optimizer.
+safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+% why couldn't we replicate this?
+If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
+If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
 
@@
 the shell, the safety properties that are mentioned above cannot be
-soundly checked statically. For example in the first case, two
-unexpaned strings could refer to the same file, but the translation
-pass cannot infer that.
+soundly checked statically.
+For safety purposes, \sys takes a conservative approach where it does not parallelize nodes without complete information.
+For example, the translation pass cannot infer that two un-expand strings do not refer to the same file identifier, and as a result does not parallelize this sub-expression.
 
@@
 
-The \ttt{Dish} optimizer is given a dataflow graph representing a
-shell program and outputs an optimized dataflow graph, where data
-parallelism opportunities have been exploited. More concretely, the
+Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.
+More concretely, the
 optimizer starts by splitting input files in several chunks, given the
@@
 
-The optimizer also applies custom transformations on several commands
-to further improve performance. Two notable examples are the
+The optimizer also applies custom transformations on several commands in \pur
+to further improve their performance. Two notable examples are the
 transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
@@
 than input \ttt{i1, t2} is illustrated below as a shell script:
+% \nv{do not understand i1, t2}
 
@@
 It is common in the dataflow graph for a node to have many incoming
-edges, or many outgoing edges (when all but the last outgoing edges
+edges or many outgoing edges (when all but the last outgoing edges
 are bounded). However some nodes, might only read their input from one
@@
 We address this issue using primitive shell constructs for file
-manipulation. More concretely a merger can be implemented by a simple
-\verb|cat|. Given a set of input files \verb|in1, in2, ...| and an
-output file \verb|out|, a merger can be implemented as:
+manipulation. More concretely a merger can be implemented by means of a simple
+\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
+output file \ttt{out}, a merger can be implemented as:
 
@@
 \noindent
-On the other hand, given an input file \verb|in| that is a FIFO pipe,
-and two output files \verb|out1, out2| the first of which is bounded
-to \verb|N| lines, a 2-splitter can be implemented as:
+On the other hand, given an input file \ttt{in} that is a FIFO pipe,
+and two output files \ttt{out1, out2} the first of which is bounded
+to \ttt{|N} lines, a 2-splitter can be implemented as:
 

### 2019-11-28 pldi/draft.tex
Update eval and rt

@@
 }
-\vspace{-15pt}
+\vspace{-10pt}
 \label{fig:optimizer-transformation}
@@
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use a combination of micro-benchmarks and one macro-benchmark, most of which were collected from prior literature and online repositories.
-Micro-bench\-marks~\sx{micro} are one-off, simple one-line pipelines that take a few seconds to write;
+To achieve this, we use six micro-benchmarks and one macro-benchmark, collected primarily from prior literature.
+
+Micro-bench\-marks~\sx{micro} are one-off, simple one-liners that take a few seconds to write;
   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-These pipelines are taken from real use cases (often from the literature~\cite{bentley1985spelling}), and, precisely due to their small size, have the benefit of highlighting stress on a handful of stages.
+These pipelines are taken from real use cases~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked} and, precisely due to their small size, highlight stress on a handful of stages.
 
@@
 It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-  as shown if written in a conventional language, it correspond to a program on the order of hundreds of lines of code.
+  as shown, if written in a conventional language it would correspond to a program on the order of hundreds of lines of code.
 
 Several results are worth highlighting.
-First and foremost, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
+First, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
 \sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
@@
 Experiments were run on a %\todo{network of five workstations} connected by 1Gbps links: 
- machine with 512GB of memory and 128 2.1GHz Intel Xeon E5-2683 cores.
+ machine with 512GB of memory and 128 $\times$ 2.1GHz Intel Xeon E5-2683 cores.
 %\todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
-For our software setup, we use SMP Debian 4.9.144-3.1 (2019-02-19), GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-release (x86\_64-pc-linux-gnu), Python 3.7.3, and OCaml 4.05.0.
+For our software setup, we use Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
 Generally, no special configuration was made in hardware or software beyond disabling hyper-threading. % ---specifically, the network protocol stack was left unoptimized.
@@
       (ii) +compile, adding \sys' pre-processing overheads, and
-      (iii) +cat, adding a final merger step.
+      (iii) +merge, adding a final merger step.
     }
@@
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
-Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}).
+Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}), and structure (col. 2) summarizes the different classes of commands used in the script.
 % The centralized pipelines read files of various sizes from disk and write to disk.
-
-Input sizes and times (col. 2, 3) summarize the longest sequential execution (200 files in Fig.~\ref{fig:microbenchmarks}).
-Script sizes report on the \sys's rewritten output for three different distribution sizes---$2\times$, 10$\times$, and $100\times$.
-
-The size of a script (col.
-5) grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
+
+Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
+Size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
 The first number alone is interesting, as it captures \sys's output for the sequential script.
-As there is no parallelism, it can be viewed as the initial cost corresponding to using \sys.
+As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
 % this is the size of 
-This change in size does not necessarily translate to observed runtime, as most of this size gain comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-FIFOs are simply pipes made, the in-kernel buffering and synchronization mechanisms are the same, leading to non-observable changes in runtime performance.
+This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
+FIFOs are simply pipes made explicit;
+  the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
 
@@
 % a DFA-based regular expression matching phase;
-  while \ttt{grep} defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
+  while the \unix \ttt{grep} command defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
 The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
@@
 Finally, \ttt{shortest-scripts} is a pipeline~\cite[pg. 7]{taylor2004wicked}that identifies the 15 shortest shell scripts in the user's \ttt{PATH};
-   interestingly, it uses \ttt{file} utility and a higher-order \ttt{wc}.
+   interestingly, it uses the \ttt{file} utility and a higher-order \ttt{wc} via \ttt{xargs}.
 
 % TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200).
+Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200$\times$).
 % we experiment with a $200\times$ level of parallelism to see potential overheads
@@
   (i) distributed, which captures only the runtime of the distributed computation,
-  (ii) compile + distributed, which adds the \sys' analysis and generation overhead, and
-  (iii) cat + compile + distributed, which adds a final merger  gathering the results at the end of the pipeline.
+  (ii) +compile, which adds the \sys' analysis and generation overhead, and
+  (iii) +merge, which adds a final merger  gathering the results at the end of the pipeline.
+Parallelism of 200$\times$ was explicitly chosen to show the \sys-internal overheads on suboptimal configurations;
+  \sys itself would never attempt a factor that is above the number of available compute nodes.
 The results show significant speedups, ranging between 4--70$\times$, depending on the distributability characteristics of individual pipelines.
@@
 Interestingly, the overhead from \sys's analysis and generation phases remains collectively below 1s even for the high-scalability cases.
-We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice (\S\ref{macro1}--\ref{macro3}).
-We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and for optimizing.
-If we also push the limit of input arguments to 100, \sys takes 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has 100 thousand edges, in a file that is almost 11MB.
+We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice.
+% (\S\ref{macro1}--\ref{macro3}).
+We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and optimization.
+Pushing the input file arguments to 100, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
 
@@
 
-To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory program of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
+To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
 At a high level, the goal is to process multi-year weather data from NCDC to identify the maximum temperature over the past several years.
@@
   download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
-Only the last of these tasks is the real focus of the chapter, demonstrate the benefits of using distributed analytics system such as Hadoop.
-
+Only the last of these tasks is the real focus of the chapter, demonstrating the benefits of using distributed analytics system such as Hadoop.
 For \sys, we consider the full pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
+
 The pipeline starts with a few stages in \sta  that generate URLs;
   these URLs point to index files, each one of which contains pointers to compressed data files.
-The stream of index files is then analyzed (4 stages) to extract the URLs of the data files, passed again through the command for downloading data.
-After downloading and uncompressing these files, the stream is analyzed to extract the maximum temperature (another 4 stages).
+The stream of index files is then analyzed (4 stages) to extract the URLs of the data files, passed again through the command to download the data.
+After download and un-compression, the stream is analyzed to extract the maximum temperature (another 4 stages).
 
@@
   unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A much simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
+A simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
   but might not perform as well in the centralized case.
@@
   fully automated (but potentially sub-optimal) distribution   (Fig.~\ref{fig:spectrum}).
+A significant concern with approaches that require input from developers is correctness:
+  developers---often neither experts in distributed systems nor educated on the internals of the code they depend on---can easily break the semantics of centralized programs, especially when these depend on existing code (such as libraries or shell commands).
 
@@
 %  or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Different from \sys, their effectiveness is predicated upon explicit invocation and is limited to embarrassingly parallel (and short) programs.
+Different from \sys, their effectiveness is predicated upon explicit and careful invocation and is limited to embarrassingly parallel (and short) programs.
 Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
@@
 
-Recently, the Smoosh authors have argued that certain shell features are useful for concurrency~\cite{smoosh:18}.
+Recently, the Smoosh authors have argued that certain shell features are useful for making concurrency explicit~\cite{smoosh:18}.
 Their argument is somewhat antithetical to \sys, % but rather that the shell language is itself an good DSL for concurrency;
@@
 They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
-Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
+% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
 For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
@@
 }
+\vspace{-15pt}
 \label{fig:spectrum}
@@
 %
-The dataflow graph is a prevalent programming model in several areas
-of data processing, as it is used by several batch~\cite{mapreduce:08,
-  spark:12} and stream processing systems~\cite{murray2013naiad,
-  carbone2015flink}. Despite its popularity, it has been noted that
+The dataflow graph is a prevalent model in several areas
+of data processing, as it is used by several batch-~\cite{mapreduce:08,
+  spark:12} and stream-processing systems~\cite{murray2013naiad,
+  carbone2015flink}. Despite its popularity, 
 most systems perform optimizations that do not preserve its semantics,
 introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
-  SHGW2015, MSAIT2019} is trying to address this issue, by restricting
-optimizations to only be performed in cases where correctness would be
-preserved. Our work draws inspiration from these attempts, as we are
+  SHGW2015, MSAIT2019} attempts to address this issue by performing
+optimizations only in cases where correctness is
+preserved. \sys draws inspiration from these efforts, as we are
 also interested in optimizations that exploit parallelization but

### 2019-11-28 pldi/draft.tex

@@
 
-\acmSubmissionID{248}
+% \acmSubmissionID{248}
 
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
+cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
   sort | uniq -c | sort -rn | head 5 > out
@@
 }
-\vspace{-15pt}
+\vspace{-30pt}
 \label{fig:schematic}
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls -l | wc -l                          [($$p_2$$)]
+  ls -l | wc -l                          ($$(p_2)$$)
 \end{lstlisting}
@@
 
-Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
+Commands can be tuned by two mechanisms: com\-mand-line options and environment variables.
 Options control the execution of a program---\eg changing the \ttt{sort}ing order.
 The shell does not have any visibility into these options; 
-  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
-Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
-They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
+  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
+Environment variables carry more general information about the surrounding system---\eg search paths, system defaults, parsing switches \etc 
+They are organized as a map from names to values, both of which are strings, and can be evaluated at any point in the program---\eg as commands, flags, or inputs.
+% a shell feature termed parameter \emph{expansion}.
 
 % TODO: Talk about security primitives?
-Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
-  files are simply persistent streams that can be manipulated by commands or pipelines.
+Streams are a key abstraction in \unix, and go hand-in-hand with another \unix abstraction: \emph{files}.
+Files are persistent streams that can be manipulated by individual commands or pipelines:
 % \begin{wrapfigure}{r}{0.15\textwidth}
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
+  ls | sed 's/1/2/' | > out.t 2> err.t    ($$(p_3)$$)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
+  cat * | tee >(wc -l > l.t) | zip > a.z   ($$(p_4)$$)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
+  read | { calc | square; } >/dev/fd/0     ($$(p_5)$$)
 \end{lstlisting}
@@
 % However, understanding the shell's primitives is not enough
-\sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
+First, \sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
 This challenge can be broken down into two parts.
-First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
+First, understand standard commands part of the shell---that is, built-ins that come with any shell and support its core functions.
 One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
-\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
-This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.
-
-The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
-Commands are arbitrary programs, but still conform to the shell's pipeline interface.
-To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
-Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
-A command developer can only annotate the default behavior;
-  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
-
-Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying  distributable regions.
-It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
+\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes~\sx{cmd}.
+This group is already useful-enough to allow the composition of practical pipelines~\sx{micro}.
+
+The second part is understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power.
+New commands are arbitrary programs, but still need to conform to the shell's pipeline interface.
+To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
+This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the distributability of the commands they are developing.
+% We haven't introduced the notion of flags
+% A command developer can only annotate the default behavior;
+%   if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
+
+Given a pipeline of commands and their distributability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
+It addresses this challenge with a series of rewriting passes that perform dataflow analysis to identify distributable regions.
+It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
 These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc
 
-Finally, \sys needs to schedule and execute the distributed dataflow graph.
-To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
-First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
-  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
-The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
+To schedule and execute the resulting distributed dataflow graph, \sys needs to solve several challenges~\sx{impl}.
+First, it needs to wrap single-node commands with corrective runtime interposition based on their classes.
+This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the distributability specification described earlier.
+Thin wrappers avoid interference with legacy functionality, as \unix commands contain non-obvious logic and flags added over extended periods of time.
 
 Second, \sys needs to fill-in planing details left blank by the rewriting pass.
-This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
-From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
-Planner activators achieves the latest possible binding, when critical runtime information is available.
+This is achieved by \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
+They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is empty, not affecting pipelines).
+They make clever use of shell features such as expansion to achieve the latest possible binding, when critical runtime information becomes available.
+% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
+
 % FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}
 
-The next few sections discuss \sys's design and implementation details.
-They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
+The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
+% They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
 
@@
 
-Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
-Other commands are treated as  non-distributable for now~\sx{discussion};
-  by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.
-Even with only these two classes, however, \sys supports highly useful pipelines that are more expressive than ones in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
-
-\subsection{Extensibility}
+The focus for the rest paper will be commands in \sta and \pur.
+\sys currently treats other commands as non-distributable~\sx{discussion}.
+In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
+Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
+
+\subsection{Distributability Extensibility}
 \label{ext}
@@
 First, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
+\sys's transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
 As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
@@
   unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
-  but might not perform as well in the centralized case.
+A simpler pipeline that builds on built-in \unix primitives might be outperformed by specialized programs on a single node, but turns out to be  trivially scalable to multiple nodes.
 % For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;

### 2019-11-28 pldi/draft.tex
Convert heading to paragraph

@@
 }
-\vspace{-30pt}
+\vspace{-10pt}
 \label{fig:schematic}
@@
 
-% \heading{Generation}
+% \paragraph{Generation}
 % Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
@@
 
-% \heading{Preliminaries}
-% \heading{POSIX, GNU Core-utils, and beyond}
-
-\heading{Stateless Commands}
+% \paragraph{Preliminaries}
+% \paragraph{POSIX, GNU Core-utils, and beyond}
+
+\paragraph{Stateless Commands}
 The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
@@
 
-\heading{Pure Commands}
+\paragraph{Pure Commands}
 The second class contains pure commands \pur.
@@
 
-\heading{File-system Commands}
+\paragraph{File-system Commands}
 The third class contains commands that access the file-system~\dfs.
@@
 
-\heading{Side-effectful Commands}
+\paragraph{Side-effectful Commands}
 Another large class of commands are ones that have side-effects to memory~\sid.
@@
 
-\heading{Irreversibly Side-effectful Commands}
+\paragraph{Irreversibly Side-effectful Commands}
 Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
@@
 
-\heading{Discussion}
+\paragraph{Discussion}
 There are a few technical details about \sys's design decisions worth noting.
@@
 %%   distribution of shell scripts}
-As a result, any static characterization of shell commands is unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
+As a result, any static characterization of shell commands would bs unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
 This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
@@
 
-\heading{Command-line Options}
+\paragraph{Command-line Options}
 As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
@@
 
-\heading{Command Inputs and Outputs}
+\paragraph{Command Inputs and Outputs}
 A second concern captured by the DSL is the order of inputs and outputs.
@@
 
-\heading{Distributability DSL}
-%
-\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is
+\paragraph{Distributability DSL}
+The DSL for distributability description (Fig.~\ref{fig:dsl}) is
 designed to classify commands into multiple categories
@@
 
-%% \heading{Pure Commands}
+%% \paragraph{Pure Commands}
 %
@@
 
-\heading{Parallelization Transformations}
+\paragraph{Parallelization Transformations}
 %
@@
 \paragraph{Translation Pass}
-\sys performs a depth-first search pass on the AST of the given shell program.
+The \sys front-end performs a depth-first search pass on the AST of the given shell program.
 During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
@@
 
-\heading{Merge and Split Implementation}
+\paragraph{Merge and Split Implementation}
 %
@@
 
-\heading{Distributed Operating Systems and Shells}
+\paragraph{Distributed Operating Systems and Shells}
 A plethora of systems assist in the construction of distributed software.
@@
 
-\heading{Cluster Computing}
+\paragraph{Cluster Computing}
 Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
@@
 
-\heading{Annotation-based Distribution}
+\paragraph{Annotation-based Distribution}
 Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
@@
 % Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Low-level Parallelization}
+\paragraph{Low-level Parallelization}
 Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
@@
 
-\heading{Correct Distribution of Dataflow Graphs}
+\paragraph{Correct Distribution of Dataflow Graphs}
 %
@@
 
-\heading{Limitations}
+\paragraph{Limitations}
 Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
@@
 
-\heading{Future Work}
+\paragraph{Future Work}
 % \label{limitation}
@@
 
-\heading{Conclusion}
+\paragraph{Conclusion}
 This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 

### 2020-01-03 osdi/draft.tex
Add paper for dumping notes

--- 2019-11-28 pldi/draft.tex
@@
 \documentclass[sigplan,10pt,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\settopmatter{printfolios=false,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
 \begin{abstract}
-  Distributed systems offer notable benefits over centralized ones.
-  Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed frameworks, or source-level annotations.
-  This paper presents \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
-  Its insight is that pipelines already express stream computations that can be automatically distributed.
-  To distribute them, \sys
-    decomposes primitives into distributability classes,
-    identifies high-distributability stages,
-    applies rewriting rules for largest possible subprograms,
-    and orchestrates the execution of the distributed program.
-Experiments for complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
+  one two three
 \end{abstract}
-
-% Its runtime component provides orchestration and planning support during the execution of the program.
-% These are from the abstract:
-%  with regards to the sequential program.
-  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known distributability classes.
-% a series of techniques for automatically 
-% 
-% Key challenges include 
-% surprisingly expressive, 
-% maximally distributable subprograms
-%
 
@@
 
-%% Keywords
-%% comma separated list
 % \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
 
-\section{Introduction}
-\label{intro}
-
-% 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
-% They are an easy way to , because they combine a set of assumptions that work well with Unix
-% They work great on a single machine, but are difficult to scale out
-% Could we fully automate distribution? 
-% The key insight is that pipelines already express a domain-specific
-% computation that is easily amenable to distribution.
-%
-
-Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
-% offer notable benefits over their centralized counterparts:
-%   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
-% Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
-% The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
-Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
-Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
-  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
-
-To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
-
-% \begin{lstlisting}[language=sh,float=h,numbers=none]
-% cat doc.ms |                   # print input file
-% groff -t -e -mandoc -Tascii |  # remove formatting
-% col -bx |                      # remove backspaces 
-% tr A-Z a-z |                   # convert to lower case
-% tr -d '[:punct:]' |            # remove punctuation
-% sort |                         # put words in alphabetical order
-% unique |                       # remove duplicate words
-% comm -13 /usr/share/dict -     # report words not in dictionary 
-% \end{lstlisting}
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$[p_1]$$)
-  sort | uniq -c | sort -rn | head 5 > out
-\end{lstlisting}
-% \smallskip
-% \noindent
-Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
-Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
-% composition allows the composition of general primitives
-
-% Key insight 
-% % The garden-hose philosophy is similar to Haskell 
-% This pipeline takes an input file (line 1), removes 
-% * do not need to use specialized frameworks for composition---instead, as long as they conform to the shell, individual primitives can be written in any programming language.
-% * do not need to rewrite ---
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-\caption{
-  \textbf{High-level schematic.}
-  \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
-  execution (right).
-}
-\vspace{-15pt}
-\label{fig:schematic}
-\end{figure}
-
-Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
-% For a user with many computers and larger inputs, 
-% The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
-%  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
-Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
-
-To answer this question, this paper presents \sys, a shell variant that transforms traditional shell pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that a large part of the shell language can be encoded using a dataflow graph model, providing structure to distribute computations.
-%% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
-% FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
-Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr}s in parallel streams,
-  use a mostly-parallel tree of \ttt{sort}s,
-  run \ttt{uniq} mostly in parallel,
-  \ttt{head} in one of the streams,
-  and (trivially) merge streams in \ttt{out}.
-
-The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
-% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
-% Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
-
-% % Make Unix benefits explicit?
-% Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
-% Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
-% Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
-
-The paper is structured as follows.  It starts with a background
-section outlining pipeline concepts and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{impl} highlight key
-contributions:
-\begin{itemize}
-
-  \item
-  \S\ref{distributability} describes several scalability and distributability classes and introduces the distributability characterization language.
-
-  \item
-  \S\ref{ir} presents a dataflow graph model that can encode shell programs, and a set of parallelization transformations that preserve its semantics.
-
-  \item
-    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of shell programs.
-    %% details other concerns, such as distributed state management and extensibility. 
-\end{itemize}
-
-\noindent
-\sys's evaluation~\sx{eval} shows significant speedups on both parallel and distributed execution, using a combination of micro-bench\-marks---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
-After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
-% \bigskip
-% \begin{quote}
-% \footnotesize
-Additional material, in anonymized form, is available at
-\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
-% \end{quote}
-
-% \begin{figure*}[t]
-% \centering
-% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-% \caption{
-%   \textbf{Applying \sys to $p_1$.}
-% }
-% \label{fig:example}
-% \end{figure*}
-%
-
-\section{Background and Overview}
-\label{bg}
-
-This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation.
-
-\subsection{Unix Composition, Informally}
-\label{bg:pipelines}
-
-% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
-\unix provides several ways for succinct program composition.
-Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
-These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
-Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
-%\begin{wrapfigure}{r}{0.06\textwidth}
-%  \vspace{-12pt}
-%  % \begin{center}
-%    % \includegraphics[width=0.48\textwidth]{gull}
-%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
-%  % \end{center}
-%  \vspace{-30pt}
-%\end{wrapfigure}
-This interface exposes a stream of contiguous character lines.
-A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
-As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
-% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
-% generating errors or diagnostics to the standard 
-% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-% \smallskip
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls -l | wc -l                          [($$p_2$$)]
-\end{lstlisting}
-% \smallskip
-
-Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
-Options control the execution of a program---\eg changing the \ttt{sort}ing order.
-The shell does not have any visibility into these options; 
-  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
-Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
-They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.
-
-% TODO: Talk about security primitives?
-Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
-  files are simply persistent streams that can be manipulated by commands or pipelines.
-% \begin{wrapfigure}{r}{0.15\textwidth}
-%   \vspace{-15pt}
-%   % \begin{center}
-%     % \includegraphics[width=0.48\textwidth]{gull}
-%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-%   % \end{center}
-%   \vspace{-25pt}
-% \end{wrapfigure}
-Reading a file or a set of files generates a stream; 
-  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
-Files are named, in the sense that they correspond to a global identifier---for example, \ttt{/x/y} corresponds to a persistent stream.
-\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
-As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
-
-% At a high level, POSIX defines several file types:
-% (-) regular files, storing lines of text or binary data,
-% (d) directories, grouping together multiple other files,
-% (l) symbolic links, pointing to other files,
-% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
-% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
-% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-% \smallskip
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
-\end{lstlisting}
-%\smallskip
-
-\noindent
-Special files are particularly interesting because they are resolved to OS structures.
-One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
-  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
-% Here are a few examples:
-% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
-% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
-% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
-Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
-This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.
-
-A small but powerful set of features is concerned with stream manipulation.
-Some of these features are built into the shell;
-  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
-% \begin{wrapfigure}{r}{0.15\textwidth}
-%   \vspace{-15pt}
-%   % \begin{center}
-%     % \includegraphics[width=0.48\textwidth]{gull}
-%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-%   % \end{center}
-%   \vspace{-25pt}
-% \end{wrapfigure}
-As described before, the file can be a FIFO dubbing as input to another command.
-More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
-  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
-% TODO: talk about \ttt{xargs}?
-
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
-\end{lstlisting}
-% \smallskip
-
-\noindent
-% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators 
-The combination of ephemeral file-system identifiers and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
-The most interesting feature of these pipelines is the presence of cycles:
-  a pipeline segment processes some input, part of which is extracting the next input to process.
-One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
-% These streams are created by the C library (glibc) at the start of program execution,
-% and new streams can be created to connect to files, sockets, pipes,
-
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
-\end{lstlisting}
-% \smallskip
-
-\noindent
-Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
-Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
-Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
-  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
-Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
-
-\unix pipelines are surprisingly powerful, a feature that \sys exploits by re-targeting shell scripts during its rewriting passes that target shell.
-
-% \begin{lstlisting}[
-%   language=es,
-%   mathescape,
-%   float=t,
-%   belowskip=-5mm,
-%   label=core-pl,
-%   upquote=true,
-%   caption={
-%     \textbf{The core language of the POSIX shell.}
-%     The language used by ; file identifiers and expansion are not shown.
-%   }]
-% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
-% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
-% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
-% \end{lstlisting}
-
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_4$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_5$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-% { echo 1; echo 2; } > out.txt
-% \end{lstlisting}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
-\caption{
-  \textbf{\sys transformation overview.}
-  Given a shell pipeline, \sys leverages command distributability to identify
-  high-distributability stages, rewrite them, and orchestrate their execution.
-}
-\vspace{-15pt}
-\label{fig:overview}
-\end{figure}
-
-% \heading{Generation}
-% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
-% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
-
-\subsection{\sys Design Overview}
-
-As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
-Intuitively,
- (i) the parallel-composition operator exposes task parallelism
-% FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
- (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator is candidate for both task and data parallelism, and
-% \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
- (iv)  other composition operators can be viewed as barrier candidates.
-To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.
-
-% However, understanding the shell's primitives is not enough
-\sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
-This challenge can be broken down into two parts.
-First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
-One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
-\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
-This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.
-
-The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
-Commands are arbitrary programs, but still conform to the shell's pipeline interface.
-To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
-Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
-A command developer can only annotate the default behavior;
-  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
-
-Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying  distributable regions.
-It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
-These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc
-
-Finally, \sys needs to schedule and execute the distributed dataflow graph.
-To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
-First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
-  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
-The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.
-
-Second, \sys needs to fill-in planing details left blank by the rewriting pass.
-This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
-From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
-Planner activators achieves the latest possible binding, when critical runtime information is available.
-% FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}
-
-The next few sections discuss \sys's design and implementation details.
-They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
-
-\section{Distributability Classes}
-\label{distributability}
-
-To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
-This understanding is encoded in a command's \emph{distributability} class.
-Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
-It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
-\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
-% Make the connections that this is like a type, that doesn't capture everything, but is good-enough
-
-This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
-Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
-In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.
-
-\subsection{Distributability of Standard Libraries}
-\label{cmd}
-
-% nv TODO: this is a hierarchy
-Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
-These classes are ordered in ascending difficulty of distribution:
-  later classes require more effort.
-In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
-
-\begin{table}[t]
-\center
-\footnotesize
-\setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Distributability Classes}.
-    Broadly, \unix commands can be broken down into to five classes.
-  }
-}
-\begin{tabular}{l @{\extracolsep{\fill}} lll}
-\toprule
-Class                           &  Key    & Example Commands                            & Coreutils       \\ % & POSIX       \\
-\midrule
-Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)    \\ %  &           \\  % 22
-Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)    \\ %  &           \\  % 21
-(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)    \\ %  &           \\  % 27
-Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)      \\ %  &           \\  % 25
-Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     \\ %  &           \\  % 5
-% \midrule
-% Shell PL Constructs             &                         &                  \\
-% \etc                            &                         &                  \\
-\bottomrule
-\end{tabular}
-\label{tab:classes}
-\end{table}
-
-% \heading{Preliminaries}
-% \heading{POSIX, GNU Core-utils, and beyond}
-
-\heading{Stateless Commands}
-The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
-These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
-At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
-Workloads that use only stateless commands are trivial to parallelize:
-  they do not require any synchronization to maintain correctness, nor caution about where to split.
- % map :: (a -> b) -> [a] -> [b]
-
-Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
-This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
-However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.
-
-\heading{Pure Commands}
-The second class contains pure commands \pur.
-These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
-The details of this state and its propagation during piecewise element processing affects their distributability characteristics.
-
-Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
-Others, such as \ttt{sha1sum}, maintain more complex state that has to be updated in a sequential fashion.
-
-Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
-Non-streaming constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques, that is implementing it as a map and a reduce function.
-
-\heading{File-system Commands}
-The third class contains commands that access the file-system~\dfs.
-This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
-In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.
-
-% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
-Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
-Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
-The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
-% these operate by changing an inode
-Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
-Operations using different identifiers pose no problems;
-  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
-While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).
-
-% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.
-
-\heading{Side-effectful Commands}
-Another large class of commands are ones that have side-effects to memory~\sid.
-Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.
-
-A command in \sid will require some form of concurrency control to distribute property.
-However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.
-
-The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
-For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
-Even if \sid commands are protected with transactions
-  these benefits are even greater for read-only or read-mostly accesses.
-
-\heading{Irreversibly Side-effectful Commands}
-Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
-Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
-
-\heading{Discussion}
-There are several technical details about \sys's design decisions worth noting.
-
-In terms of inputs, \sys supports commands with multiple input (resp. output) streams, by prepending (resp. appending) shell built-ins that merge input (resp. split output) streams~\sx{impl}.
-Thus, a command can be modelled as having one input and one output stream, a representation that is used throughout \sys including the extensibility DSL~\sx{ext}.
-
-Wanting to present the full picture for completeness purposes, the focus for the rest paper will be commands in \sta and \pur.
-Other commands are treated as  non-distributable for now~\sx{discussion};
-  by building infrastructure to handle other classes, \sys would be able to (i) extract more parallelism and (ii) handle larger programs.
-Even with only these two classes, however, \sys supports highly useful pipelines that are more expressive than ones in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
-
-\subsection{Extensibility}
-\label{ext}
-
-%% Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
-%% % ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
-%% %    | -h :: ( Sless ) => { Out }                              -- if needed
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%% (bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
-%%   |-h :: (Sless)($$\Rightarrow$$){Out}                              
-%% \end{lstlisting}
-
-%% It states that the command defaults to the Pure class.
-%% It's operation can be thought as a transformation from input streams to output streams.
-%% It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
-%% % TODO: check actual man page for how to specify files
-%%   it writes to the output and error stream rather than a file.
-%% The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
-%%   its output is a constant function.
-
-% FIXME: \nv{TODO: remember to bubble up(me)}
-An important characteristic of the \unix shell is its language-agnostic extensibility~\sx{bg}.
-Users composing pipelines are free to install custom commands available from various sources.
-Being general programs, commands are developed in a variety of languages and are extended over long periods of time.
-%% \kk{Text point: This restricts the space of solutions for
-%%   distribution of shell scripts}
-As a result, any static characterization of shell commands is unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
-This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
-%% \kk{Text point: The characteristics of a good solution}
-
-To address % additions and changes to the set of commands available to the shell,
-  this challenge, \sys arms command developers with a lightweight, domain-specific language (DSL).
-The DSL allows developers to express semantic information about a command's distributability that is critical to \sys.
-It can be used by the developers of new commands to succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
-It can also be used by maintainers extending or maintain existing commands, to express additions or changes to the command's implementation or interface.
-
-\heading{Command-line Options}
-As described earlier~\sx{bg}, command-line options and environment variables can affect the operation of a command.
-Options are a particularly popular way to control a command, 
-  directly affecting a command's distributability classification.
-That is, commands are assigned to default classes, and options can change the class in which they belong.
-
-As simple examples, consider \ttt{cat} and \ttt{sort}.
-By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
-Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta because it generates a static stream regardless of other input.
-% TODO: closure?
-
-%% Can be tweaked according to this:
-%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
-
-\begin{figure}
-  \centering
-  \begin{grammar}
-    <option> ::= `-' <string>
-
-    <category> ::= `stateless' | `pure' | ...
-
-    <maybe-int> ::= | <int>
-
-    <arg> ::= `args[' <int> `]'
-
-    <args> ::= <arg>
-    \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-    <input> ::= `stdin' | <args>
-
-    <inputs> ::= <input>
-    \alt <input> `,' <inputs>
-
-    <output> ::= `stdout' | <arg>
-
-    <option-pred> ::= <option>
-    \alt `value' <option> = <string>
-    \alt `not' <option-pred>
-    \alt <option-pred> `or' <option-pred>
-    \alt <option-pred> `and' <option-pred>
-
-    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
-
-    <predicate> ::= <option-pred> `=>' <assignment>
-
-    <pred-list> ::= `|' <predicate> <pred-list>
-    \alt `|' `otherwise' `=>' <assignment>
-
-    <command> ::= <name> `\{' <pred-list> `\}'
-
-    <command-list> ::= <command>
-    \alt <command> <command-list>
-  \end{grammar}
-  \caption{
-\textbf{Distributability description language.}
-The DSL captures important information regarding the distributability of a command.
-\vspace{-20pt}
-}
-
-  \label{fig:dsl}
-\end{figure}
-
-\heading{Command Inputs and Outputs}
-A second concern captured by the DSL is the order of inputs and outputs.
-Commands generally read multiple input streams (and, at cases, write to multiple output streams).
-The invariants between inputs and outputs must be maintained over the distributed execution of the program.
-% but these maintain different invariants, enabling very different distributability characteristics.
-
-Consider the command \ttt{comm} that takes as input two files and
-performs a join operation. When invoked without any option, it
-produces a three-column output. The first column contains lines that
-are unique to the first file, the second column contains lines that are
-unique to the second file, and the last column contains lines that
-exist in both files. In the general case \ttt{comm} is in \pur, since
-it needs to wait until it reads both files completely before it can
-output its results.
-
-However, \ttt{comm} can also be invoked using some options that
-suppress the output of any combination of columns. More precisely, the
-option \ttt{-1} suppresses the first column of the output. Similarly,
-\ttt{-2} and \ttt{-3} supress the second and third column. Using both
-\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command jumps to \sta
-regarding its first (resp. the second) file argument.
-
-% \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
-
-\heading{Distributability DSL}
-%
-\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is
-designed to classify commands into multiple categories
-given their options.
-
-It assumes that the arguments of a command can always be separated to
-options and file arguments (checking if they start with a
-\ttt{-}~\footnote{Note that the single character \ttt{-} is not an
-  option as it is most often used to designate that a command reads
-  from its standard input}). A file written in the DSL contains a
-record containing a mapping from predicates to assignments for each
-command. Predicates can be simple boolean formulas on the existence
-and the value of any option. Each assignment contains the command
-category, the sequence of its input sources, and its output
-source. Inputs to commands are a sequence of file identifiers and its
-standard input. Similarly, the output is either its standard output or
-a file identifier. An example record for \ttt{comm} can be seen below:
-
-\begin{lstlisting}[float=h, numbers=none]
-  comm {
-    | -1 and -3 =>
-      (stateless, [args[1]], stdout)
-    | -2 and -3 =>
-      (stateless, [args[0]], stdout)
-    | otherwise =>
-      (pure, [args[0]], stdout)
-  }
-\end{lstlisting}
-
-The semantics of the language is straightforward: given a command, the
-interpreter collects its options and then returns the assignment of
-the first predicate that is satisfied. The final predicate is always
-satisfied. For example the following two invocations of \ttt{comm}
-would be classified as follows:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- comm -13 f1 f2 # => (stateless, f2, stdout)
- comm f1 f2     # => (pure, f1, stdout)
-\end{lstlisting}
-
-\tr{\kk{We don't need to designate the stderror, because we assume that it
-  is never the main output of a command and it will never be used by
-  the input of another command in the pipeline. If someone indeed
-  wants to do this, they can just use a redirect I think to get around
-  it.}}
-
-\tr{\kk{If we talk about the language it would be good to give some
-  statistics about how many commands (out of the ones that we have
-  classified) can be represented with one, two clauses etc. If we have
-  a different solution (like python function that given a command and
-  its arguments returns the category) then we could talk about how
-  many lines of code it took to categorize all the commands that we
-  did.}}
-
-\tr{\kk{Note: Since a categorization language will probably not be
-  complete (especially the one showing the input arguments) there
-  should be a backup mechanism, when the language is not expressive
-  enough, like a python function that categorizes the command, and
-  returns its input argument if it is stateless.}}
-
-\tr{\kk{(Maybe) We should have a crisp point about why we have this categorization
-  language, and why don't we just allow someone to write a function in
-  python for each command, that given the command and its arguments,
-  returns whether the command category. Possible arguments include,
-  the fact that this language is simpler to use, especially by non
-  experts that just run some script but have installed some commands
-  that are not ``supported''. Another possible argument is that we
-  could be able to reason about the constructs of the categorization
-  lagnauge, and that they will be easier to read and
-  understand. Another argument is that since these categorizations
-  should be shareable among users, it would be bad to execute
-  arbitrary python code, so using this language, expressivity is
-  limited. All of these arguments are a little bit weak though. Niko,
-  what do you think?}}
-
-%% \heading{Pure Commands}
-%
-\noindent
-In addition to the commands in \sta that can be parallelized in a
-straightforward way, there also exists a subclass of commands in \pur
-that can be parallelized using divide-and-conquer techniques (as
-mentioned in \cref{cmd}). In order to support the parallelization of
-arbitrary commands in this category, \sys allows command developers to
-provide custom map and reduce steps for them, where the map
-is in \sta and the reduce in \pur, and their composition produces the
-same output with the initial command. \sys can then replace occurrences
-of these commands with their map and reduce parts, enabling their
-parallelization.
-
-\tr{\kk{I am not sure a general interface is so easy to design. It needs
-  more though. It might be beneficial to just talk about sort and wc
-  here and how we implemented them and nothing more. Or maybe this
-  could then go to the implementation? Or maybe say that one can write
-  a python function that given a node of the graph, transforms it into
-  many. I am not sure what is best...}}
-
-% As mentioned in distributability, pure commands can be broken down
-% into different categories. We might be able to get ideas about this
-% from this paper:
-% \url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
-% continuation in PLDI 2019.
-% FIXME: \nv{Cite this paper?}
-\tr{Can we find a solution for the commands in coreutils?}
-
-\section{Dataflow Graph Model}
-\label{ir}
-
-The core of \sys is a dataflow graph model, inspired by the ones used in popular distributed stream processing systems~\sx{related}.
-% such as MapReduce \cite{mapreduce:08} and Apache Spark \cite{spark:12}).
-These systems introduce data parallelism by parallelizing nodes of the dataflow graph.
-Unfortunately, % their optimizations in many of these systems 
- they often do not preserve the model's semantics, blurring the lines between specification, optimization, and implementation.
-In contrast, \sys's dataflow model inherently supports the data parallelism found in shell pipelines.
-More specifically, it develops a set of parallelization-exposing optimizations represented as semantics-preserving graph transformations, effectively exposing data parallelism as task parallelism.
-The \sys optimizer then applies these transformations in an iterative way, optimizing the program to exploit the available computational resources.
-
-%% The standard dataflow graph exposes task parallelism. Different nodes
-%% in the graph can be executed concurrently, communicating through the
-%% edges.
-%% Task level parallelism is already exposed by the shell, by using
-%% pipes \ttt{|} and the background operator \ttt{\&}.
-%% Unfortunately, simple task parallelism leaves a lot to be desired,
-%% since scaling a computation requires manually introducing more
-%% operators.
-
-%% In order to adequately utilize the available computational resources,
-%% a distributed implementation should exploit data parallelism. Data
-%% parallelism is based on the observation that some operations produce
-%% the same result when performed independently on subsets of the input
-%% data, merging the results afterwards. Exploiting data parallelism is
-%% one of the major reasons that popular distributed stream processing
-%% systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
-%% \cite{spark:12}) achieve significant performance.
-
-%% A very important consequence of this is that we can iteratively apply
-%% optimizations that expose data parallelism, or even revert them when
-%% the resulting dataflow graph is too wide. This flexibility unifies the
-%% unoptimized and optimized versions of a computation, enabling thinking
-%% about them with the same conceptual tools.
-
-In order to effectively apply optimizations and distribute a shell script, \sys first translates it to a manipulable representation.
-\sys first introduces a variant of the standard dataflow graph model as a candidate intermediate representation~\sx{graph-components}.
-The main benefit of this model is that it exposes both task and data parallelism.
-It then defines a set of optimizations that expose and exploit the data parallelism of commands in \sta and \pur as semantics-preserving graph transformations~\sx{ir:transformations}.
-
-\subsection{Graph Components}
-\label{graph-components}
-
-As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of the
-shell are (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii) commands that communicate
-through these files. Based on that, \sys's graph model defines nodes to represent commands and edges to represent files.
-
-\paragraph{Nodes---Commands}
-
-For a set $D$, we write $D\kstar$ to denote the set of all finite
-words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
-or $xy$ to denote their concatenation. We write $\eps$ for the empty
-word.
-We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-It is easy to see that $\leq$ is a partial order, often called the \emph{prefix order}.
-
-A node $f$ of the graph represents a function from one (possibly empty)
-input stream to an output stream $f : D\kstar \to D\kstar$, where $D$
-represents the basic data type of a line of characters. This
-representation captures the majority of shell commands. We require that
-the function $f$ is monotone w.r.t.\ the prefix order. This captures
-the idea that a node cannot retract output after it has been emitted.
-%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
-%concatenation and will be omitted when obvious.
-An example of a node is the following command:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- grep ``foo''
-\end{lstlisting}
-
-\noindent
-This command takes a stream of lines from its standard input and
-returns a stream of lines on its standard output.
-
-An important
-observation is that commands might read their input
-stream from a sequence of different sources. An illustrative example
-is the command \ttt{cat}, which reads its input from a combination
-of files and its standard input:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat x - y
-\end{lstlisting}
-
-\noindent
-In the example above, \ttt{cat} reads file \ttt{x} until it encounters the EOF
-character, then reads from \ttt{stdin} until in encounters EOF, and finally reads from
-\ttt{y}.
-
-In order to handle this particularity, nodes in the
-dataflow graph can have many incoming edges, that are \emph{ordered} in a
-sequence. This is different from standard dataflow
-graph representations where the semantics of multiple incoming edges
-is either zipping the inputs into pairs, or interleaving them arbitrarily.
-
-%% \TODO{Is the following paragraph best suited here? It describes an
-%%   assumption that we make about commands, to guarantee the validity of
-%%   our transformations. The assumption is needed due to the limitation
-%%   of the programming model of one input one output per command.}
-
-Nodes may read from more files than just the files of their
-input stream (resp. write to more files than their output stream).
-However, these files are static (\eg used for configuration or
-logging) and do not represent streams of data, and therefore are not
-considered part of the dataflow graph.
-For now, we assume that these static
-files can only be accessed by one node, thus ensuring that they do not
-interfere with the distributed implementation.
-%% Because of the assumption in section Front End \ref{}, that each file
-%% exists only once in the dataflow graph, we can safely assume that
-%% reading and writing to these static files does not interfere with the
-%% distributed implementation, and doesn't alter the behaviour of the
-%% program.
-
-\paragraph{Edges---Files}
-
-Edges in the dataflow graph represent files, the basic data
-abstraction of the shell. They are used as communication channels
-between nodes in the graph, and as the input or output of the whole
-graph. Edges are represented as possibly unbounded streams of type
-$D\kstar$. They could either refer to a specific named file in the file
-system, or just be FIFO pipes used for interprocess communication
-%% \TODO{Are there any other file types, such as URLs, ...?}.
-
-%% \TODO{Show an example pipe and its graph}.
-
-%% \begin{center}
-%% \small
-%% \begin{tikzpicture}[node distance=2.25cm, ->]
-%% \node (S) {};
-%% \node (A) [draw, right of=S, node distance=1.75cm] {node 1};
-%% \node (B) [draw, right of=A] {node 2};
-%% \node (C) [draw, right of=B] {node 3};
-%% \node (E) [right of=C, node distance=1.75cm] {};
-%% \path (S) edge node[above] {input} (A);
-%% \path (A) edge node[above] {stream} (B);
-%% \path (B) edge node[above] {stream} (C);
-%% \path (C) edge node[above] {output} (E);
-%% \end{tikzpicture}
-%% \end{center}
-
-In order to expose data parallelism from the output of a command, files that
-do not point to a resource can have
-\begin{wrapfigure}[3]{r}{0.20\columnwidth}
-\vspace{-7pt}
-  \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
-\vspace{-20pt}
-\end{wrapfigure}
-an upper bound on the lines that they can transfer. This helps utilize available
-parallelism after encountering a command that cannot be parallelized
-(\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.
-
-\tr{On the other hand, some forms of data parallelism can be exposed
-  when knowing the size of the input files. As mentioned in \ref{}
-  some pure commands (such as cat -n) only need line information to
-  become stateless, and knowing the size of a file could allow the
-  system to split it in different chunks that can be processed
-  independently. To account for that, edges that refer to an input
-  resource contain the number of lines of the file that they refer
-  to.}
-
-Finally, edges that do not refer to a resource and do not start
-from a node in the graph represent the graph input;
-edges that do not refer to a resource and do not point to a node in the graph are its outputs.
-
-%% \kk{(Maybe) It might be beneficial to mention that many inputs do not
-%%   represent the same thing as many outputs. These two notions are not
-%%   symmetric.}
-
-%% \begin{figure*}
-%% \centering
-%% \begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
-%% %
-%% \small
-%% %
-%% \node (M1) {};
-%% \node (M2) [draw, right of=M1, node distance=2cm] {Map};
-%% \node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
-%% \node (M4) [right of=M3, node distance=3cm] {};
-%% %
-%% \path (M1) edge node {input} (M2);
-%% \path (M2) edge node {stream} (M3);
-%% \path (M3) edge node {output} (M4);
-%% \end{tikzpicture}
-%% \end{figure*}
-
-%% \begin{figure*}
-%% \newcommand{\Split}{\text{Split}}
-%% \newcommand{\Merge}{\text{Merge}}
-%% \newcommand{\Map}{\text{Map}}
-%% \newcommand{\GroupBy}{\text{GroupBy}}
-%% \newcommand{\Reduce}{\text{Reduce}}
-%% \centering
-%% \begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
-%% %
-%% \footnotesize
-%% %
-%% \node (M1) {};
-%% \node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
-%% \node (M3) [right of=M2, node distance=1.5cm] {};
-%% \node (M4) [right of=M3, node distance=2.5cm] {};
-%% \node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
-%% \node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
-%% \node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
-%% \node (M8) [right of=M7, node distance=2cm] {};
-%% %
-%% \node (T3) [draw, above of=M3] {$\Map$};
-%% \node (T4) [draw, above of=M4] {$\Split_K$};
-%% \node (T5) [draw, above of=M5] {$\Merge$};
-%% \node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
-%% %
-%% \node (B3) [draw, below of=M3] {$\Map$};
-%% \node (B4) [draw, below of=M4] {$\Split_K$};
-%% \node (B5) [draw, below of=M5] {$\Merge$};
-%% \node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
-%% %
-%% \path (M1) edge node {$A$} (M2);
-%% \path (M2) edge[bend left=10] node {$A$} (T3);
-%% \path (M2) edge[bend right=10] node[swap] {$A$} (B3);
-%% \path (M5) edge node {$K_2 \times B$} (M6);
-%% \path (M6) edge (M7);
-%% \path (M7) edge node {$K \times B$} (M8);
-%% %
-%% \path (T3) edge node {$K \times B$} (T4);
-%% \path (T4) edge (T5);
-%% \path (T4) edge (M5);
-%% \path (T4.330) edge[bend right=10] (B5.170);
-%% \path (T5) edge node {$K_1 \times B$} (T6);
-%% \path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
-%% %
-%% \path (B3) edge node {$K \times B$} (B4);
-%% \path (B4) edge (B5);
-%% \path (B4) edge (M5);
-%% \path (B4.30) edge[bend left=10] (T5.190);
-%% \path (B5) edge node {$K_3 \times B$} (B6);
-%% \path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
-%% %
-%% \end{tikzpicture}
-%% \caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
-%% \label{fig:mapReduce}
-%% \end{figure*}
-
-\subsection{Graph Transformations}
-\label{ir:transformations}
-
-Before describing the different types of parallelization
-optimizations, we formalize the intuition about stateless and pure
-commands that was described earlier~\sx{cmd}.
-%% \nv{TODOs
-%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
-%% }
-%
-
-As mentioned there, stateless commands such as \ttt{tr} operate independently on
-individual elements of the stream (characters, lines, or files)
-without maintaining any state.
-% Some examples include \ttt{grep}, \ttt{basename}, and \ttt{tr}.
-Formally, a command $f$ is stateless if
-it is ``linear'' with respect to concatenation, \ie 
-% the following equation holds:
-satisfying the following equation:
-
-\[
-\forall x, x', f(x.x') = f(x).f(x')
-\]
-
-%% This is a nice characterization, which I think is true in most
-%% settings. There is a caveat though: You can have stateless programs
-%% that emit output when they start (before they consume the stream)
-%% and/or when they consume the EOF symbol. In this case I am not sure
-%% the equation above holds.
-
-\noindent
-This means that one gets the same output by applying a command $f$ to a
-concatenation of two inputs $x, x'$ as by concatenating the result
-produced by the application of $f$ to these inputs
-separately.
-
-Similarly, it was mentioned that some pure commands (such
-as \ttt{sort} and \ttt{wc}) can be parallelized using
-divide-and-conquer techniques. More formally, these pure commands $f$
-can be implemented as a combination of functions map $m$ and reduce $r$, satisfying
-the following equation:
-
-\[
-\forall x, x', f(x.x') = r(m(x).m(x'))
-\]
-
-\noindent
-Similarly to the earlier equation, this means that we get the same output by
-applying $f$ to a concatenation of two inputs $x, x'$, or by applying
-the reduce function $r$ to the concatenation of the outputs produced
-by applying map $m$ to each of these inputs.
-
-\heading{Parallelization Transformations}
-%
-Based on these equations, we can define a maximal \emph{node split
-  operation} on a stateless node $v$ with $n$ input ordered edges and
-an output edge to node $v'$, to replace $v$ with $n$ new nodes, giving
-each one of them one of the incoming edges of $v$, and routing them in
-order to point to $v'$. Since each incoming edge represents a stream
-of data $x_i : D^*$, and the only behaviour of a dataflow graph is its
-output, this optimization $ v(x_1.x_2...x_n) \Rightarrow
-v(x_1).v(x_2)...v(x_n)$ can be shown to preserve the behaviour of the
-graph. Furthermore, it can be straightforwardly extended to pure
-commands that can be implemented by a map-reduce pair $(m, r)$ as $
-v(x_1.x_2...x_n) \Rightarrow r(m(x_1).m(x_2)...m(x_n))$.
-
-\tr{Remember to mention the assumptions that need to hold for the
-  graph transformations to be valid in the Command categories
-  section. Commands must be deterministic, they must not do any other
-  side effect (such as writing to other files, sending signals,
-  etc). However, these assumptions must already be checked when the
-  developer designates the categories.}
-
-\tr{If there is time I can work out a formal definition and a proof
-  sketch why this transformation preserves the output of the dataflow
-  graph.}
-
-%% \begin{definition}
-%% Given a dataflow graph $G = (V, E, O)$, where $V$ is a set of nodes
-%% representing commands, $E$ is a set of edges representing files, and
-%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
-%% edges. $v_{out}$ represents a concatenation of all the outputs of the
-%% dataflow graph. We represent the total order for a node $v$ as
-%% $<_v$. Given a node a node $v$
-%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
-%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
-%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
-%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
-%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
-%% \end{definition}
-
-%% \begin{lemma}
-%% tofill
-%% \end{lemma}
-
-In addition to node splits, we can also perform edge splits in order
-to increase the number of incoming edges for some nodes to further expose
-possible data parallelism. In cases where the edge refers to a real file, the
-edge split represents splitting the real file in chunks. In
-cases where the edge connects two nodes, the split effectively refers to several edges
-with a bounded number of lines. Note that the final output edge of a
-valid node in the graph must always be unbounded if it does not refer
-to a file.
-
-\section{Implementation}
-\label{impl}
-
-This section describes challenges related to the implementation of \sys. 
-It starts by describing how the \sys front-end translates a shell script to the dataflow graph model defined in \S\ref{ir}, by applying transformation on distributable regions~\sx{front-end}.
-It closes with the \sys planner, which  optimizes, distributes, and orchestrates the execution of a given dataflow graph~\sx{optimizer}.
-
-\subsection{Front End}
-\label{front-end}
-
-As mentioned in \Cref{ir}, in order to generate the distributed implementation of a shell script, \sys translates it to a dataflow graph.
-However, there are several components in shell programs that cannot be arbitrarily parallelized without affecting the program behaviour---\eg commands connected with a \ttt{\&\&} operator~\sx{bg}.
-The fact that the shell is highly dynamic compounds the challenge---\eg by not knowing what files does an unexpanded command argument refer to.
-% of correctly identifying what parts can be distributed, e.g.
-
-%% \kk{Give examples of writing to environment variables, producing
-%%   irreversible side effects, the ; and \&\& operators.}
-
-To address this issue, we introduce the notion of distributable program regions and a translation pass for converting them to a dataflow graph.
-The translation pass identifies distributable regions and translates each of them to a dataflow graph, provided the abstract syntax tree (AST) of a shell script.
-The AST is produced by \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
-
-\paragraph{Distributable Regions}
-%
-%% \kk{The goal of this subsection is to explain that Dish does a safe
-%%   but effective analysis to identify which parts of the scripts to
-%%   parallelize.}
-%
-Distributable regions are (ideally, maximal) program sub-expressions that can be parallelized safely with respect to their sequential execution.
-The structure of a shell program can be used to identify components that can be executed concurrently and barriers that are natural points for enforcing synchronization.
-Consider the following example:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat f1 f2 | grep ``foo'' > f3 && sort f3 
-\end{lstlisting}
-% # Replace sort with sth better
-
-\noindent
-The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would already execute concurrently in the standard shell, while \ttt{sort f3} would wait for their completion before executing.
-Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3} are distributable regions---maximal sub-expressions that cannot be extended beyond \ttt{\&\&}.
-
-Intuitively, distributed regions correspond to subsections of the program that would be allowed to execute concurrently by different processes in the POSIX standard~\cite{posix}.
-The pipe operator \ttt{|} and the parallel composition operator \ttt{\&} enable parallelization, and extend distributable regions.
-Conversely, the \ttt{;} sequential composition operator, the \ttt{\&\&} logical operator, and function definitions represent barrier constructs that do not allow distributable regions to permeate through.
-
-% \tr{I don't know whether I should mention the following: While these
-%   control flow constructs can be parallelized and there has been
-%   research on it, it is orthogonal to our work, and can be
-%   incorporated as future work.}
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%   (tr ... f1 > g1 && tr ... f2 >> g1) &
-%   grep ``foo'' g1
-% \end{lstlisting}
-% 
-% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
-%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
-% the dataflow graph that corresponds to the whole program, has two
-% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
-%   g1}.
-%
-
-\paragraph{Translation Pass}
-\sys performs a depth-first search pass on the AST of the given shell program.
-During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
-To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
-To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.
-
-A few illustrative cases of the analysis algorithm are shown in pseudocode below:
-
-\begin{lstlisting}[language=python, float=h]
-  def translate(node):
-    ...
-    elif(node.name == 'Pipe'):
-      return pipe_graphs([translate(child)
-        for child in node.children])
-    elif(node.name == 'And'):
-      node.children = [translate(child)
-        for child in node.children]
-      return node
-    elif(node.name == 'Command'):
-      if(not safe(node)):
-        return make_ir_node(node, 'unsafe')
-      else:
-        return make_ir_node(node,
-          find_category(node))
-    ...
-\end{lstlisting}
-
-Whenever the algorithm encounters a distributable node, it recursively
-translates its subcomponents.
-It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
-% what first and second? we didn't say anything about first and second
-Before the different dataflow graphs are connected, the
-algorithm checks that at most one node in the graph writes at every
-file, and that at most one node reads from a file that another node writes
-to. This is important, in order to avoid inconsistent behaviour after
-distributing nodes in the dataflow graph (\ie by introducing to
-concurrent reads and writes to the same file when
-parallelizing).
-
-Whenever the algorithm encounters a barrier node,
-it recursively translates its subcomponents, similarly to the distributable
-case.
-However, different from the distributable case, it does not merge them in a large dataflow graph.
-
-Finally, whenever the algorithm encounters a command, it first
-checks whether it is safe to parallelize it. A command is considered
-safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
-% why couldn't we replicate this?
-If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
-
-An important consideration is that due to the highly dynamic nature of
-the shell, the safety properties that are mentioned above cannot be
-soundly checked statically.
-For safety purposes, \sys takes a conservative approach where it does not parallelize nodes without complete information.
-For example, the translation pass cannot infer that two un-expand strings do not refer to the same file identifier, and as a result does not parallelize this sub-expression.
-
-\subsection{Optimizer}
-\label{optimizer}
-
-Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.
-More concretely, the
-optimizer starts by splitting input files in several chunks, given the
-desirable output dataflow graph width. It then starts from the source
-nodes of the graph, iteratively performing the graph transformations
-that were described in \Cref{ir:transformations}. If the incoming
-edges of the node are less than the desirable graph width, it performs
-edge splits, and then it performs maximal node splits. The final graph
-can then be implemented by spawning a process for each node, and
-redirecting the outputs according to the graph edges. An example
-transformation can be seen in \Cref{fig:optimizer-transformation}.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.70\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
-\caption{
-  \textbf{Example graph transformation.}
-  The optimizer splits a stateless node to utilize
-  available data parallelism.
-}
-\vspace{-10pt}
-\label{fig:optimizer-transformation}
-\end{figure}
-
-The optimizer also applies custom transformations on several commands in \pur
-to further improve their performance. Two notable examples are the
-transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
-that have more than one input are tranformed to \ttt{sort} commands on
-each input, followed by a \ttt{sort -m}, to produce the final sorted
-result. The transformation on occurences of \ttt{wc} that have more
-than input \ttt{i1, t2} is illustrated below as a shell script:
-% \nv{do not understand i1, t2}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  paste -d '+' \
-   <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
-   <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
-   bc | tr -s '\n' ' ' |
-   sed 's/^/   /' | sed 's/$/\ /'
-\end{lstlisting}
-
-The above script takes the outputs produced by two \ttt{wc} commands
-and aggregates them accordingly.
-
-%% \subsection{Mapping operators to nodes}
-
-%% \kk{I don't think we should mention this.}
-
-%% Explain the simple algorithm that we used to minimize intra node
-%% communication, that tries to map contiguous parts of the graph to the
-%% same node. We could do this by having an algorithm that minimizes the
-%% number of cuts or something.
-
-%% We don't try to make a proper planner, as there is a lot of work on
-%% operator placement etc. that we can borrow from.
-
-% \subsection{Planning Activators and Just-in-Time Planning}
-% 
-% \nv{Nikos working on this and next}
-% 
-% % \kk{This whole section for just in itme planning, is also relevant for
-% %   the front end. If the front end is executed just in time, it would
-% %   have more information, and maybe it could make a sound analysis.}
-% 
-% Important point: Shell is extremely dynamic. Because of this, a
-% distributor cannot decide how to distribute a subexpression
-% statically, by just reading the script, (as in many other systems like
-% MapReduce, Spark, etc), since most of the information is not there
-% before the script starts executing. Environment variables,
-% unexpanded(unevaluated) strings \kk{Make sure that terminology is
-%   consistent with Greenberg}, could all contain information that is
-% valuable to make the distribution plan \kk{give an example}.
-% 
-% 
-% Future work on this: calling the shell shtepper to only partially
-% evaluate strings, just expand arguments, and not do any significant
-% computation. Even better, we could be calling the shtepper just after
-% having parsed the ast, and decide on maximal distributable subtrees as
-% late as possible in the process. This would be awesome.
-%
-
-\heading{Merge and Split Implementation}
-%
-It is common in the dataflow graph for a node to have many incoming
-edges or many outgoing edges (when all but the last outgoing edges
-are bounded). However some nodes, might only read their input from one
-source (e.g. stdin). In this case, the distributed implementation
-needs to concatenate input files (or conversely split output
-files).
-%
-%% \tr{Show an example graph! of cat | tr | sort. The command in the
-%%   middle must only take input from one file (maybe just stdin).}
-%
-We address this issue using primitive shell constructs for file
-manipulation. More concretely a merger can be implemented by means of a simple
-\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
-output file \ttt{out}, a merger can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
- cat $in1 $in2 ... > $out
-\end{lstlisting}
-
-\noindent
-On the other hand, given an input file \ttt{in} that is a FIFO pipe,
-and two output files \ttt{out1, out2} the first of which is bounded
-to \ttt{|N} lines, a 2-splitter can be implemented as:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
-  tee >(head -n $N > $temp1;
-        dd of=/dev/null > /dev/null 2>&1 &
-        cat $temp1 > $out1) |
-       (tail -n +${N+1} > $out2;
-        dd of=/dev/null > /dev/null 2>&1)
-\end{lstlisting}
- %% head -n $N $in > $out1 ; cat $in > $out2
-
-\noindent
-Using the 2-splitter as a building block, a pipe can be split to
-arbitrarily many different pipes.
-
-\section{Evaluation}
-\label{eval}
-
-% FIXME
-% \kk{We should make sure to measure the time that it takes for the
-%   whole process to run (parsing, translating, optimizing, planning)
-%   together with the scripts. This will probably be negligible, but we
-%   still have to mention it. }
-% 
-% \kk{Also, we have to have one-two handcrafted examples that have more
-%   than one pipeline, to show the execution time end to end, as this
-%   cannot be shown with the one-pipeline examples since they only have
-%   one graph, so no back and forth between shell and dish.}
-% 
-% \kk{Furthermore it would be great to have one distributed example that
-%   runs in more than one node (if we have time)}
-% 
-% \kk{Also we have to make sure to measure with and without the final
-%   cat of output file, so that we show that this is also negligible.}
-% 
-% \kk{We have to make sure that we repeat that the results are the same
-%   in all tests that we run. Corectness, blah blah ...}
-% 
-% \kk{We should have at least some pipelines that have a bad pure
-%   command in them.}
-
-\begin{table*}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of Micro-benchmarks}.
-    The micro-benchmarks are small pipelines (\ie ones developed on-the-fly) drawn from various sources and applied to large datasets.
-  }
-}
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllll}
-\toprule
-  Script                 ~&~ Structure                    & Input (MB)& Time (Seq)    & Script Size (LoC)             & Highlights                                        \\
-\midrule
-% https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
-  \tti{grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
-  \tti{sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ & 1000      &   2m 11.18s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ & 1000      &   6m 53.55s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
-  \tr{\tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\}
-% https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
-  \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ & 1000      &   8m 13.42s   &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
-  \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
-  \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & extensive file-system operation   \\
-\bottomrule
-\end{tabular*}
-\label{tab:eval}
-\end{table*}
-
-At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use six micro-benchmarks and one macro-benchmark, collected primarily from prior literature.
-
-Micro-bench\-marks~\sx{micro} are one-off, simple one-liners that take a few seconds to write;
-  such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-These pipelines are taken from real use cases~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked} and, precisely due to their small size, highlight stress on a handful of stages.
-
-The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
-It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-  as shown, if written in a conventional language it would correspond to a program on the order of hundreds of lines of code.
-
-Several results are worth highlighting.
-First, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around 50--300ms, negligible ($<$5\%) for short-running pipelines and virtually non-existent for long-running ones.
-As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
-Finally, \sys preserves the productivity of the shell:
-  the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
-  %% in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
-
-Experiments were run on a %\todo{network of five workstations} connected by 1Gbps links: 
- machine with 512GB of memory and 128 $\times$ 2.1GHz Intel Xeon E5-2683 cores.
-%\todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
-For our software setup, we use Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
-Generally, no special configuration was made in hardware or software beyond disabling hyper-threading. % ---specifically, the network protocol stack was left unoptimized.
-All pipelines are set to (initially) read from and (finally) write to the file-system (except as otherwise stated).
-Whenever \ttt{curl} is part of a pipeline, it fetches data from a different physical host on the same network connected by 1Gbps links.
-More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
-
-\subsection{Microbenchmarks: Shell One-liners}
-\label{micro}
-
-\begin{figure*}[t]
-    \centering
-    %% \subcaptionbox{\label{eval:minimal_grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:minimal_sort}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:topn}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:wf}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:spell}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \caption{a}
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/spell_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_sort}
-    \end{subfigure}%
-
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:topn}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:wf}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.32\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:spell}
-    \end{subfigure}%
-    \caption{
-      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--200).} 
-      Three different configurations per benchmark:
-      (i) distributed, focusing only on the distributed execution,
-      (ii) +compile, adding \sys' pre-processing overheads, and
-      (iii) +merge, adding a final merger step.
-    }
-    \vspace{-15pt}
-    \label{fig:microbenchmarks}
-\end{figure*}
-
-Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
-Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}), and structure (col. 2) summarizes the different classes of commands used in the script.
-% The centralized pipelines read files of various sizes from disk and write to disk.
-Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
-
-Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
-Size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
-The first number alone is interesting, as it captures \sys's output for the sequential script.
-As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
-% this is the size of 
-This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-FIFOs are simply pipes made explicit;
-  the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
-
-In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command;
-% a DFA-based regular expression matching phase;
-  while the \unix \ttt{grep} command defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
-The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
-The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
-  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-% \tr{The \ttt{bi-gram} pipeline calculates n-grams of base two;
-%   it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
-The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
-  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
-Finally, \ttt{shortest-scripts} is a pipeline~\cite[pg. 7]{taylor2004wicked}that identifies the 15 shortest shell scripts in the user's \ttt{PATH};
-   interestingly, it uses the \ttt{file} utility and a higher-order \ttt{wc} via \ttt{xargs}.
-
-% TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200$\times$).
-% we experiment with a $200\times$ level of parallelism to see potential overheads
-Each one of these plots reports on three different configurations:
-  (i) distributed, which captures only the runtime of the distributed computation,
-  (ii) +compile, which adds the \sys' analysis and generation overhead, and
-  (iii) +merge, which adds a final merger  gathering the results at the end of the pipeline.
-Parallelism of 200$\times$ was explicitly chosen to show the \sys-internal overheads on suboptimal configurations;
-  \sys itself would never attempt a factor that is above the number of available compute nodes.
-The results show significant speedups, ranging between 4--70$\times$, depending on the distributability characteristics of individual pipelines.
-
-Interestingly, the overhead from \sys's analysis and generation phases remains collectively below 1s even for the high-scalability cases.
-We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice.
-% (\S\ref{macro1}--\ref{macro3}).
-We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and optimization.
-Pushing the input file arguments to 100, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
-
-\subsection{Macro-benchmark: Weather Analysis}
-\label{macro1}
-
-To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
-At a high level, the goal is to process multi-year weather data from NCDC to identify the maximum temperature over the past several years.
-The original task, as presented in the book, can be thought of comprising three sub-tasks:
-  download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
-Only the last of these tasks is the real focus of the chapter, demonstrating the benefits of using distributed analytics system such as Hadoop.
-For \sys, we consider the full pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
-
-The pipeline starts with a few stages in \sta  that generate URLs;
-  these URLs point to index files, each one of which contains pointers to compressed data files.
-The stream of index files is then analyzed (4 stages) to extract the URLs of the data files, passed again through the command to download the data.
-After download and un-compression, the stream is analyzed to extract the maximum temperature (another 4 stages).
-
-% The first interesting result 
-The effort required to write this pipeline is astonishingly low:\footnote{
-  Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.
-}
-  it amounts to 14 stages and, when expressed as a single line, only 240 characters.
-The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 117 lines of Java code---not accounting for code moving input and output in and out of HDFS.
-
-In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1:09 for moving the dataset to HDFS.
-For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
-
-% /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
-% hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
-
-% Dish p3 + p4 me platos pipeline 2 => real 4m30.554s (perilambanei download twn gz kai to processing)
-% Dish p3 + p4 me platos pipeline 10 => real 1m7.183s
-
-% Dish p3 me platos pipeline 2 => real	3m47.470s (perilambanei download twn gz kai save sto file system)
-% Dish p3 me platos pipeline 10 => real	1m38.945s
-
-A less obvious take-away from this case is that large, complex pipelines enable significant freedom in how to express their stages.
-Several stages in the original shell program (taken from the book) are expressed as a single \ttt{awk} script;
-  unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A simpler pipeline that builds on built-in \unix primitives is trivially distributable, 
-  but might not perform as well in the centralized case.
-% For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
-%   this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
-
-% few different ways to write some phases, 
-%   for example, 
-% A large pipeline resulted in a couple of 
-% An interesting lesson here was that there are multiple
-% First, longer pipelines can be written in multiple ways
-% It is worth noting
-% These are
-%
-
-\section{Related Work}
-\label{related}
-
-At a high level, techniques for re-writing programs to exploit distribution fall under a wide spectrum that ranges from
-  fully manual (but, ideally, optimal) to 
-  fully automated (but potentially sub-optimal) distribution   (Fig.~\ref{fig:spectrum}).
-A significant concern with approaches that require input from developers is correctness:
-  developers---often neither experts in distributed systems nor educated on the internals of the code they depend on---can easily break the semantics of centralized programs, especially when these depend on existing code (such as libraries or shell commands).
-
-\heading{Distributed Operating Systems and Shells}
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-hide many distribution problems, as long as a program uses the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
-% simplify many of the problems of distribution and
-
-Several packages expose commands for specifying parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}.
-%  or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Different from \sys, their effectiveness is predicated upon explicit and careful invocation and is limited to embarrassingly parallel (and short) programs.
-Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
-\sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently.
-
-Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target distribution.
-Here too, however, developers are expected to manually rewrite scripts to exploit these new primitives.
-
-Recently, the Smoosh authors have argued that certain shell features are useful for making concurrency explicit~\cite{smoosh:18}.
-Their argument is somewhat antithetical to \sys, % but rather that the shell language is itself an good DSL for concurrency;
-  which argues for automatically (and correcctly) parallelizing pipelines when users have not expressed parallelism explicitly---hence distribution-\emph{oblivious} shell scripting.
-
-\heading{Cluster Computing}
-Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
-These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
-By targeting specific classes of computation (\emph{viz.} \sys's distributability), they render distributed computation efficiently.
-
-% Lack of generality, no formal guaratees, their setup and is too tedious 
-\sys chooses a more general approach:
-  it does not require users to set up a new framework for every new class of computation they need to execute, nor to rewrite different parts of their computation using the specific abstractions provided by each framework they use.
-It also takes a more principled approach: by developing the semantic model outlined earlier, it can guarantee that transformations are correct with respect to the sequential program.
-
-% % Although still a research prototype, \sys's setup and deployment is simpler.
-% It also does not require these specialized frameworks, manually (re)write , and 
-% Different from \sys, however, 
-% Due to the specialization of these primitives, however, these systems suffer from a lack of generality ;
-% % and they still require developers to compose their work using the pipelines
-% Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
-
-\heading{Annotation-based Distribution}
-Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
-They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
-% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
-% Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
-Unfortunately, a challenge with this approach is that users composing applications by using third-party libraries do not understand their distributability details well-enough to annotate them.
-
-\sys occupies a different point in the design space, by removing any work from composers and adding a small load to developers.
-% More specifically, it is designed to be used automatically by composers,
-Its DSL is designed to be more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared towards the developers of these components---not the users composing them.
-
-% It also provides a more principled approach than both of these systems.
-% Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
-% The side-effects present in the shell dwarf those of functional languages
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Low-level Parallelization}
-Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
-%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
-Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
-
-% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
-% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
-% While they Moreover, these systems take advantage of functional purity;
-%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
-\caption{
-  \textbf{Manual--automated distribution spectrum.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-}
-\vspace{-15pt}
-\label{fig:spectrum}
-\end{figure}
-
-% \kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
-%   na ginoun formally proven claims gia to equivalence preservation twn
-%   optimizations etc.}
-
-\heading{Correct Distribution of Dataflow Graphs}
-%
-The dataflow graph is a prevalent model in several areas
-of data processing, as it is used by several batch-~\cite{mapreduce:08,
-  spark:12} and stream-processing systems~\cite{murray2013naiad,
-  carbone2015flink}. Despite its popularity, 
-most systems perform optimizations that do not preserve its semantics,
-introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
-  SHGW2015, MSAIT2019} attempts to address this issue by performing
-optimizations only in cases where correctness is
-preserved. \sys draws inspiration from these efforts, as we are
-also interested in optimizations that exploit parallelization but
-preserve the semantics of the underlying model. However, our proposed
-dataflow model is different from these other works, as nodes in our
-model read from their inputs \emph{in sequence}. Due to this
-intricacy, previous work is not straightforwardly applicable.
-
-%% \cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but
-%% their correctness is established using informal arguments based on
-%% operational intution.
-
-%% The paper \cite{MSAIT2019} proposes a denotational semantic framework
-%% for stream processing where the data streams are seen as partial
-%% orders, and establishes the soundness of some common parallelizing
-%% transformations on dataflow graphs.
-
-%% \km{Some very early references on the dataflow model of computation
-%% (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN,
-%% KMacQ1977}}
-
-\section{Discussion}
-\label{discussion}
-
-This section reflects on \sys's design and implementation, touches on future work, and concludes the paper.
-
-\heading{Limitations}
-Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
-
-\sys does not adequately handle failures or network partitions in the general sense, which would require re-scheduling passes on failed replicas.
-Prior work on fault-tolerant distributed stream processing can be of significant aid here.
-
-The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
-Prior work on operator placement of distributed dataflow graphs can be used to build a more sophisticated planner.
-
-\sys is conservative in the shell subsets that it handles, in order to
-avoid introducing unsafe behaviours. The point of this work was not to be
-able to handle a complete distributable subset of the shell, but
-rather to use significant part of it to demonstrate performance benefits.
-
-% Shell is highly dynamic, thus making it impossible to design sound
-% meaningful analyses to determine distributable regions and maximal
-% parallelization. In order to address this particularity, \sys could be
-% tighly integrated with a posix compliant shell
-% interpreter~\cite{smoosh:20} so that dynamic information can be
-% acquired selectively to aid the analysis and optimization process.
-
-\heading{Future Work}
-% \label{limitation}
-% \item No cycles (multiple commands writing and reading from the same file)
-There are several worthwhile directions for future work.  One
-direction would be an extension of the distributability analysis of
-\sys, in order to guide it using dynamic information. This could
-enable a formal study of the analysis~\sx{impl} to show that it
-returns regions that are safe to distribute.  Recent work on
-formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
-such an attempt possible.
-
-Another direction is to extend \sys to handle other classes outside \sta and \pur.
-These two classes enjoy a certain popularity with quick, one-off pipelines and are the easiest to distribute, but 
-it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).
-This direction would naturally introduce considerations about replication, consistency, and failure recovery.
-
-% There are two additional assumptions that have to be satisfied in
-% order for the distributed implementation that is generated by Dish to
-% have equivalent behaviour with the original implementation. First of
-% all, there should be no external signals to the commands while they
-% execute. At the moment Dish assumes no faults, and that communication
-% always succeeds. This assumption can be lifted by extending Dish with
-% a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
-% command reads and writes to arbitrary files that are not mentioned in
-% their arguments. This is a necessary assumption, since commands are
-% considered as black boxes and the only information that can be
-% inferred about them is through their arguments.
-
-\heading{Conclusion}
-This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
-\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
-To achieve its goal, \sys
-  decomposes primitives into distributability classes,
-  identifies high-distributability stages,
-  applies rewriting rules for largest possible subprograms,
-  and orchestrates the execution of the distributed program.
-% Its runtime component provides orchestration and planning support during the execution of the program.
-Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
-
+\paragraph{Conclusion}
 \begin{acks}

### 2020-01-03 osdi/draft.tex
Dump notes

@@
 
+\section{Introduction}
+\label{intro}
+
+Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
+Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
+Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
+  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
+   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
+% 
+To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
+
+Contributions:
+
+* 
+Extends beyond linear pipelines to full programs 
+that comply with the POSIX shell specification; while \sys does not optimize
+everything---it certainly cannot---, it does
+
+* Planner activations, the ability to call
+
+* A standard library of semantics-aware wrappers for
+
+* Augment data flows to allow performance optimizations and fault-tolerance
+
+\section{Background}
+\label{bg}
+
+To understand the limitations of pipelines and the opportunities for
+parallelization, consider the following pipeline
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+find . -name '*.sh' | xargs cat | wc -l > out.t
+\end{lstlisting}
+
+Unfortunately, this pipeline meshes the results from `cat` leading to a single
+count from all files. What the developer actually intended was the following:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+$(
+for f in $(find . -name '*.sh'); do
+ cat ${f} | wc -l
+done
+) > out.t
+\end{lstlisting}
+
+This remains trivial to parallelize, as none of the loop-internal steps depends
+on the input, so loop iterations can run in parallel.
+
+Similar for
+  * enclaves 
+  * sieve
+  * mapping max temperatures to years, etc.
+
+Different from prior systems, \sys's goal is to ensure correctness first and
+then improve performance.
+
+\section{Evaluation}
+
+\subsection{POSIX test-suite}
+
+\subsection{Pipelines and scripts}
+
+\subsection{Other pipelines}
+
+Rewrite pipelines expressed in other languages / scripts
+
 \paragraph{Conclusion}

### 2020-01-08 pldi/draft.tex
Teensy corrections to paper

@@
 
+\kk{After discussion on Wednesday Jan 8: Modify the narrative of this paper to talk about data processing in the Shell.
+Make this the main focus of the paper.
+This simplifies the narrative and allows us to not talk about late planner activation as well as environment variables.
+It also justifies why we only deal with pure and stateless.
+Note that this is a very expressive model, as shown by our benchmark suite.
+Note: Rememeber to clearly describe the subset of the shell that we handle.
+To be more precise the compositional operator of the shell that we can handle.}
+
 % 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
@@
 The DSL captures important information regarding the distributability of a command.
-\vspace{-20pt}
+\vspace{-10pt}
 }
@@
 \centering
-\includegraphics[width=0.70\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
+\includegraphics[width=0.40\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
 \caption{

### 2020-01-09 evaluation/tools/server.js
Update distribution scripts
(Related to #11)

@@ -6,71 +6,80 @@ const crypto = require('crypto');
 const HASH = "f4f8b9dc33fe805bcd0a854b19554f1daa596cf46ae3ef811a27b7d490dae42c";
 const PORT = 6001
 
-// nc -l 5000 | tr '[:lower:]' '[:upper:]' | nc -C 158.130.4.212 5555
-
 let jobs = [];
 
-// example object:
-// {dt: "passphrase", workerPort: 1234, serverIP: "192.168.0.1", serverPort: 4321, program: "..."  }
 let validateObj = (obj) => {
-  let keys = ["dt", "workerPort", "serverIP", "serverPort", "program"];
+  let keys = ["dt", "program"];
   for (var i = 0; i < keys.length; i++) {
     if (!obj[keys[i]]) {
-      return {"success": false, "msg": "object missing " + keys[i]}
+      return {success: false, msg: "object missing " + keys[i]}
     }
   };
 
   let h = crypto.createHash('sha256').update(obj.dt).digest('hex');
   if (h !== HASH) {
-    return {"success": false, "msg": "security token failed"}
+    return {success: false, msg: "security token failed"}
   }
+  return {success: true};
 }
 
 let validateURI = (uri) => {
   const parsed = parseInt(uri, 10);
   if (isNaN(parsed)) {
-    return {"success": false, "msg": "Did not call `/<int>`"};
+    return {success: false, msg: "Did not call `/<int>`"};
   } 
-  if ((parsed < 1) || (parsed > jobs.lenth - 1)) {
-    return {"success": false, "msg": "No such ID"};
+  if ((parsed < 1) || (parsed > jobs.length - 1)) {
+    return {success: false, msg: "No such ID"};
   }
-  reutn {"success": true, "msg": data[parsed]};
+  return {success: true, msg: data[parsed]};
 }
 
 let newJob = (obj) => {
-  const jobID = jobs.lenth;
-  const fileName = jobID + ".sh";
-  fs.writeFileSync(fileName, obj.program, "utf-8");
-  const pipeline = `
-  nc -l ${obj.workerPort} |
-        ./${fileName} |
-         nc -C ${serverIP} ${serverPort}`
-  let c = exec(pipeline, (error, stdout, stderr) => {
+  console.log(jobs, jobs.length, obj);
+  if (!validateObj(obj).success) {
+    return validateObj(obj)
+  }
+  const jobID = jobs.length;
+  jobs[jobID] = obj;
+  const filename = jobID + ".sh";
+  try {
[…]

### 2020-01-09 evaluation/tools/client.js
Update distribution scripts
(Related to #11)

@@ -1,11 +1,12 @@
 #!env node
+const http = require('http');
 
-const workers = [
-  {name: "local", ip: "127.0.0.1", port: 6000},
-  {name: "beta",  ip: "158.130.4.113", port: 6000},
-  {name: "gamma", ip: "158.130.4.114", port: 6000},
-  {name: "delta", ip: "158.130.4.120", port: 6000}
-];
+const workers = {
+  "local": {ip: "127.0.0.1", port: 6000},
+  "beta":  {ip: "158.130.4.113", port: 6000},
+  "gamma": {ip: "158.130.4.114", port: 6000},
+  "delta": {ip: "158.130.4.120", port: 6000}
+};
 
 // ./client.js            -- reports global status
 const help=`
@@ -13,11 +14,11 @@ const help=`
 ./client <w> <expr>    -- launches <expr> on <w>
 
 Workers currently available:
-  ${workers.map( (e) => e.name ).join('\n  ')}
+  ${Object.keys(workers).map( (e) => e ).join('\n  ')}
 `;
 
-let get = (url, cb) => {
-  http.get(url, (res) => {
+let get = (key, cb) => {
+  http.get(`http://${workers[key].ip}:${workers[key].port}/`, (res) => {
     res.setEncoding('utf8');
     let rawData = '';
     res.on('data', (chunk) => { rawData += chunk; });
@@ -27,8 +28,38 @@ let get = (url, cb) => {
   }).on('error', (e) => {
     cb(`Got error: ${e.message}`);
   });
+};
+
+let generateHeader = (key, data, op) => {
+  return {
+    hostname: workers[key].ip,
+    port: workers[key].port,
+    path: '/',
+    method: op || 'PUT',
+    headers: {
+      'Content-Type': 'application/json',
+      'Content-Length': data.length
+    }
+  }
 }
 
+let put = (key, json, cb) => {
+  const req = http.request(generateHeader(key, json), (res) => {
+    console.log(`statusCode: ${res.statusCode}`)
+
+    res.on('data', (d) => {
+      process.stdout.write(d)
+    })
+  })
+  
+  req.on('error', (error) => {
+    cb(error)
+  })
+  
+  req.end(json)
+  //req.end() // TODO combine
+}
+  
 const dtVar="DISH_TOKEN";
 
 let printAndExit = (msg) => {
@@ -40,6 +71,7 @@ let checkDishToken = () => {
   if (!process.env[dtVar]) {
     printAndExit(`Need to set ${dtVar} environment variable`);
[…]

### 2020-01-09 evaluation/tools/distributed.sh

@@ -1,15 +1,14 @@
 #!/bin/bash
 
 # Schematic for distributing computations
+# To try server client:
 
-A="158.130.4.119"  #TOBLIND
-B="158.130.4.113"  #TOBLIND
-C="158.130.4.114"  #TOBLIND
-D="158.130.4.120"  #TOBLIND
-DSTAR="158.130.4.212" #TOBLIND
-
-cat $IN | grep -v "onetwo" | tr '[:lower:]' '[:upper:]' > $OUT
-
+./client.js local 'ls'
+./client.js local | jq .
+./client.js local 'sleep 5; echo "yo"'
+./client.js local | jq .
+sleep 5
+./client.js local | jq .
 
 # The format is as follows:
 # cat FIFOs > $OUT &

### 2020-01-24 evaluation/tools/page-per-line.sh
Move into tools

@@ -1,10 +1,14 @@
 #!/bin/bash
 
+# Squash all HTML for each URL into a single line, streaming fashion
+# It also prefixes with the URL
+
 page_per_line () {
   curl -s $1 | tr -d "\n\r" | tr -d '\n' | sed "s/^/$0 /" | sed -e '/.$/a\'
 }
 
 export -f page_per_line
+
 # xargs:
 # add `-t` for debugging
 cat ./urls.txt | xargs -0 -d '\n' -n 1 bash -c 'page_per_line "$@"' _

### 2020-02-18 pldi/draft.tex
New intro sketch for OSDI

@@
 \title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
+%% \title{Extensible data processing in the shell}
+
 % \titlenote{with title note}             %% \titlenote is optional;
@@
 
-\kk{After discussion on Wednesday Jan 8: Modify the narrative of this paper to talk about data processing in the Shell.
-Make this the main focus of the paper.
-This simplifies the narrative and allows us to not talk about late planner activation as well as environment variables.
-It also justifies why we only deal with pure and stateless.
-Note that this is a very expressive model, as shown by our benchmark suite.
-Note: Rememeber to clearly describe the subset of the shell that we handle.
-To be more precise the compositional operator of the shell that we can handle.}
+\kk{After discussion on Wednesday Jan 8: Modify the narrative of this
+  paper to talk about data processing in the Shell. Make this the main
+  focus of the paper. This simplifies the narrative and allows us to
+  not talk about late planner activation as well as environment
+  variables. It also justifies why we only deal with pure and
+  stateless. Note that this is a very expressive model, as shown by
+  our benchmark suite. Note: Rememeber to clearly describe the subset
+  of the shell that we handle. To be more precise the compositional
+  operator of the shell that we can handle.}
+
+\kk{New intro sketch:}
+
+\kk{Note: A point that we need to establish and convince the readers
+  about is that the Shell IS used already for data-processing, and
+  that we are not just proposing it as an alternative to other
+  systems. We do not propose migrating to shell scripts if you already
+  have a parallel/efficient solution. I am not sure if that needs a
+  specific paragraph of its own at the start of the
+  introduction. Like: ``Consider the following bash script that does
+  some simple data processing... Examples like that exists in several
+  domains/applications (examples, cite).''}
+
+\begin{itemize}
+\item Why is parallelism important (especially in the context of data
+  processing). Significant performance benefits as most computer
+  systems today have multiple processors that are under utilized
+\item (Similarly to how it is now) unfortunately, it is quite hard
+  reaping these benefits in contrast to sequential data
+  processing. [Show Bash pipeline].
+\item To benefit from parallelism, the current options are
+  extreme. Either very fine-grained and low level (parallelism using
+  fine-grained C libraries) for maximum performance, or very coarse
+  grained and bulky where one has to use a proper big-data framework.
+\item The common thing is that one needs to rewrite big part of their
+  application, often requiring expertise and redesigning its
+  architecture, in order to get benefits that in the end might not
+  explain the cost.
+\item To address this issue, we propose Dish, a thin layer that
+  automatically! parallelizes a given shell script. Key insight is
+  that a large subset of the shell language can be encoded in a
+  dataflow graph model, a flexible representation that allows for
+  parallelization optimizations.
+\item Study of the shell primitives, identification of
+  families/classes of commands that share
+  distributability/parallelization properties, and a classification of
+  coreutils/POSIX commands in these distributability properties.
+\item Dataflow graph model that is suitable for the shell
+\item A set of optimizations on this model that can be thought of as
+  rewriting rules on the dataflow graph, that are sound and exploit
+  parallelism.
+\end{itemize}
+
+\kk{End of new intro sketch}
 

### 2020-02-21 pldi/draft.tex
Notes on section 4

@@
   fine-grained C libraries) for maximum performance, or very coarse
-  grained and bulky where one has to use a proper big-data framework.
+  grained and bulky where one has to use a proper big-data
+  framework. (Can we fit the shell ones here (gnu-parallel, etc))
 \item The common thing is that one needs to rewrite big part of their
@@
   parallelization optimizations.
+\item Contributions:
 \item Study of the shell primitives, identification of
@@
 
-The core of \sys is a dataflow graph model, inspired by the ones used in popular distributed stream processing systems~\sx{related}.
-% such as MapReduce \cite{mapreduce:08} and Apache Spark \cite{spark:12}).
-These systems introduce data parallelism by parallelizing nodes of the dataflow graph.
-Unfortunately, % their optimizations in many of these systems 
- they often do not preserve the model's semantics, blurring the lines between specification, optimization, and implementation.
-In contrast, \sys's dataflow model inherently supports the data parallelism found in shell pipelines.
-More specifically, it develops a set of parallelization-exposing optimizations represented as semantics-preserving graph transformations, effectively exposing data parallelism as task parallelism.
-The \sys optimizer then applies these transformations in an iterative way, optimizing the program to exploit the available computational resources.
-
-%% The standard dataflow graph exposes task parallelism. Different nodes
-%% in the graph can be executed concurrently, communicating through the
-%% edges.
-%% Task level parallelism is already exposed by the shell, by using
-%% pipes \ttt{|} and the background operator \ttt{\&}.
-%% Unfortunately, simple task parallelism leaves a lot to be desired,
-%% since scaling a computation requires manually introducing more
-%% operators.
-
-%% In order to adequately utilize the available computational resources,
-%% a distributed implementation should exploit data parallelism. Data
-%% parallelism is based on the observation that some operations produce
-%% the same result when performed independently on subsets of the input
-%% data, merging the results afterwards. Exploiting data parallelism is
-%% one of the major reasons that popular distributed stream processing
-%% systems (such as MapReduce \cite{mapreduce:08} and Apache Spark
-%% \cite{spark:12}) achieve significant performance.
-
-%% A very important consequence of this is that we can iteratively apply
-%% optimizations that expose data parallelism, or even revert them when
-%% the resulting dataflow graph is too wide. This flexibility unifies the
-%% unoptimized and optimized versions of a computation, enabling thinking
-%% about them with the same conceptual tools.
-
-In order to effectively apply optimizations and distribute a shell script, \sys first translates it to a manipulable representation.
-\sys first introduces a variant of the standard dataflow graph model as a candidate intermediate representation~\sx{graph-components}.
-The main benefit of this model is that it exposes both task and data parallelism.
-It then defines a set of optimizations that expose and exploit the data parallelism of commands in \sta and \pur as semantics-preserving graph transformations~\sx{ir:transformations}.
+The core of \sys is a dataflow graph model, inspired by the ones used
+in popular distributed stream processing systems~\sx{related}. The
+dataflow model can express a large subset of Shell programs and
+inherently supports the data parallelism found in shell pipelines by
+exposing it as task parallelism. More preciely, we have developed a
+set of parallelization-exposing optimizations represented as
+semantics-preserving graph transformations. These transformations can
+then be applied in an iterative way, optimizing the program to utilize
+available computational resources.
+
+%% In order to effectively apply optimizations and distribute a shell
+%% script, \sys first translates it to a manipulable representation.
+
+\kk{It might be beneficial to do a full pass over this section
+  describing the new model (where a node that has many sequential
+  inputs is always prepended with a cat node.}
 
@@
 
-As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of the
-shell are (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii) commands that communicate
-through these files. Based on that, \sys's graph model defines nodes to represent commands and edges to represent files.
+As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of
+the shell are (i) files (or pipes, \emph{viz.} correspondence between
+files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii)
+commands that communicate through these files. Based on that, \sys's
+graph model defines nodes to represent commands and edges to represent
+files.
 
@@
 For a set $D$, we write $D\kstar$ to denote the set of all finite
-words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$
-or $xy$ to denote their concatenation. We write $\eps$ for the empty
-word.
-We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-It is easy to see that $\leq$ is a partial order, often called the \emph{prefix order}.
-
-A node $f$ of the graph represents a function from one (possibly empty)
-input stream to an output stream $f : D\kstar \to D\kstar$, where $D$
-represents the basic data type of a line of characters. This
-representation captures the majority of shell commands. We require that
-the function $f$ is monotone w.r.t.\ the prefix order. This captures
-the idea that a node cannot retract output after it has been emitted.
-%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
+words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or
+$xy$ to denote their concatenation. We write $\eps$ for the empty
+word. We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq
+y$, if there is a word $z$ such that $y = xz$. It is easy to see that
+$\leq$ is a partial order, often called the \emph{prefix order}.
+
+A node $f$ of the graph represents a function from one (possibly
+empty) input stream to an output stream $f : D\kstar \to D\kstar$,
+where $D$ represents the basic data type of a line of characters
+\kk{We have to extend this to more general nodes. It has to be from
+  many inputs to many outputs. The commands that we optimize at the
+  moment are the ones that can be rewritten to take one stream input
+  to one stream output, and then some other inputs and outputs that
+  are considered static.}. This representation captures the majority
+of shell commands. We require that the function $f$ is monotone
+w.r.t.\ the prefix order. This captures the idea that a node cannot
+retract output after it has been emitted.
+%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
+%exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
 %concatenation and will be omitted when obvious.
@@
 
+\kk{I think that we should probably add the ``new'' split optimization
+  (and maybe any others that we develop until then) here.}
+
 In addition to node splits, we can also perform edge splits in order
@@
 the shell, the safety properties that are mentioned above cannot be
-soundly checked statically.
-For safety purposes, \sys takes a conservative approach where it does not parallelize nodes without complete information.
-For example, the translation pass cannot infer that two un-expand strings do not refer to the same file identifier, and as a result does not parallelize this sub-expression.
+soundly checked statically.  For safety purposes, \sys takes a
+conservative approach where it does not parallelize nodes without
+complete information.  For example, the translation pass cannot infer
+that two un-expand strings do not refer to the same file identifier,
+and as a result does not parallelize this sub-expression.
+
+\kk{Ideally we would want to talk here about the as-late-as possible
+  dish translation that allows us to be safe despite the Shell's
+  dynamic nature. I think that this is a great point about the
+  work. Maybe we should also focus on implementing it before the OSDI
+  deadline.}
 
@@
 
-At a high level, techniques for re-writing programs to exploit distribution fall under a wide spectrum that ranges from
-  fully manual (but, ideally, optimal) to 
-  fully automated (but potentially sub-optimal) distribution   (Fig.~\ref{fig:spectrum}).
-A significant concern with approaches that require input from developers is correctness:
-  developers---often neither experts in distributed systems nor educated on the internals of the code they depend on---can easily break the semantics of centralized programs, especially when these depend on existing code (such as libraries or shell commands).
+At a high level, techniques for re-writing programs to exploit
+distribution fall under a wide spectrum that ranges from fully manual
+(but, ideally, optimal) to fully automated (but potentially
+sub-optimal) distribution (Fig.~\ref{fig:spectrum}).  A significant
+concern with approaches that require input from developers is
+correctness: developers---often neither experts in distributed systems
+nor educated on the internals of the code they depend on---can easily
+break the semantics of centralized programs, especially when these
+depend on existing code (such as libraries or shell commands).
 
 \paragraph{Distributed Operating Systems and Shells}
-A plethora of systems assist in the construction of distributed software.
-Distributed operating systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
-hide many distribution problems, as long as a program uses the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
+A plethora of systems assist in the construction of distributed
+software.  Distributed operating systems~\cite{ousterhout1988sprite,
+  mullender1990amoeba, pike1990plan9, barak1998mosix} and programming
+languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide
+many distribution problems, as long as a program uses the provided
+abstractions---which are strongly coupled to the underlying operating
+or runtime system.
 % simplify many of the problems of distribution and
 
-Several packages expose commands for specifying parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}.
-%  or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Different from \sys, their effectiveness is predicated upon explicit and careful invocation and is limited to embarrassingly parallel (and short) programs.
-Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also includes other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
-\sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently.
-
-Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target distribution.
-Here too, however, developers are expected to manually rewrite scripts to exploit these new primitives.
-
-Recently, the Smoosh authors have argued that certain shell features are useful for making concurrency explicit~\cite{smoosh:18}.
-Their argument is somewhat antithetical to \sys, % but rather that the shell language is itself an good DSL for concurrency;
-  which argues for automatically (and correcctly) parallelizing pipelines when users have not expressed parallelism explicitly---hence distribution-\emph{oblivious} shell scripting.
+Several packages expose commands for specifying parallelism on modern
+\unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun},
+\textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU}
+\ttt{parallel}~\cite{Tange2011a}. % or any manual
+rewriting~\cite{mapreduce:08, ciel:11, spark:12}.  Different from
+\sys, their effectiveness is predicated upon explicit and careful
+invocation and is limited to embarrassingly parallel (and short)
+programs.  Often, these commands provide options to support an array
+of special sub-cases---a stark contradiction to the celebrated \unix
+philosophy.  For example, \ttt{parallel} contains flags such as
+\ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix
+user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also
+includes other programs with complex semantics, such as the ability
+transfer files between computers, separate text files, and parse CSV.
+\sys embraces the \unix philosophy and simply attempts to rewrite
+pipelines more efficiently.
+
+Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add
+primitives for non-linear pipe topologies---some of which target
+distribution.  Here too, however, developers are expected to manually
+rewrite scripts to exploit these new primitives.
+
+Recently, the Smoosh authors have argued that certain shell features
+are useful for making concurrency explicit~\cite{smoosh:18}.  Their
+argument is somewhat antithetical to \sys, % but rather that the shell
+language is itself an good DSL for concurrency; which argues for
+automatically (and correcctly) parallelizing pipelines when users have
+not expressed parallelism explicitly---hence
+distribution-\emph{oblivious} shell scripting.
 
 \paragraph{Cluster Computing}
-Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
-These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
-By targeting specific classes of computation (\emph{viz.} \sys's distributability), they render distributed computation efficiently.
+Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
+  naiad:13} and domain-specific languages~\cite{alvaro2011consistency,
+  distal:13, meiklejohn2015lasp} go a step further in terms of
+automation by providing a few carefully-designed primitives.  These
+primitives make strong assumptions about the nature of the
+computation---\eg strongly-eventual commutative functions that can
+proceed in parallel.  By targeting specific classes of computation
+(\emph{viz.} \sys's distributability), they render distributed
+computation efficiently.
 
 % Lack of generality, no formal guaratees, their setup and is too tedious 
-\sys chooses a more general approach:
-  it does not require users to set up a new framework for every new class of computation they need to execute, nor to rewrite different parts of their computation using the specific abstractions provided by each framework they use.
-It also takes a more principled approach: by developing the semantic model outlined earlier, it can guarantee that transformations are correct with respect to the sequential program.
+\sys chooses a more general approach: it does not require users to set
+up a new framework for every new class of computation they need to
+execute, nor to rewrite different parts of their computation using the
+specific abstractions provided by each framework they use.  It also
+takes a more principled approach: by developing the semantic model
+outlined earlier, it can guarantee that transformations are correct
+with respect to the sequential program.
 
@@
 \paragraph{Annotation-based Distribution}
-Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
-They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
+Recent systems on light-touch distribution, such as
+Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
+than cluster-computing frameworks.  They also lower (but not entirely
+eliminate) developer effort by automating most of the rewriting
+process.
 % Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
-% Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
-Unfortunately, a challenge with this approach is that users composing applications by using third-party libraries do not understand their distributability details well-enough to annotate them.
+For the critical points where user input is required, they provide a
+high-level declarative DSL with varying (but limited) degrees of
+expressive power.
+% Annotations are used by ---complicated by the use of third-party
+% libraries similar to \sys's extensions.
+Unfortunately, a challenge with this approach is that users composing
+applications by using third-party libraries do not understand their
+distributability details well-enough to annotate them.
 
@@
 \paragraph{Low-level Parallelization}
-Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require programmers to get involved in discovering and exposing parallelism.
+Instruction-level parallelization has a longer history, starting from
+explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2}
+and continuing with compilers that attempt to automatically extract
+parallelism~\cite{padua1993polaris,hall1996maximizing}.  These systems
+operate at a lower level than \sys (\eg that of instructions or loops
+instead of pipe boundaries) and typically do not exploit runtime
+information.
+
+More recent work focuses on extracting parallelism from
+domain-specific programming models~\cite{cilk5, streamIt, galois} and
+interactive parallelization tools~\cite{parascope, ipat}.  These tools
+simplify the expression of parallelism, but still require programmers
+to get involved in discovering and exposing parallelism.
 %  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
@@
 %
-The dataflow graph is a prevalent model in several areas
-of data processing, as it is used by several batch-~\cite{mapreduce:08,
+The dataflow graph is a prevalent model in several areas of data
+processing, as it is used by several batch-~\cite{mapreduce:08,
   spark:12} and stream-processing systems~\cite{murray2013naiad,
-  carbone2015flink}. Despite its popularity, 
-most systems perform optimizations that do not preserve its semantics,
-introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
-  SHGW2015, MSAIT2019} attempts to address this issue by performing
-optimizations only in cases where correctness is
-preserved. \sys draws inspiration from these efforts, as we are
-also interested in optimizations that exploit parallelization but
-preserve the semantics of the underlying model. However, our proposed
-dataflow model is different from these other works, as nodes in our
-model read from their inputs \emph{in sequence}. Due to this
-intricacy, previous work is not straightforwardly applicable.
+  carbone2015flink}. Despite its popularity, most systems perform
+optimizations that do not preserve its semantics, introducing subtle
+erroneous behaviours. Recent work~\cite{HSSGG2014, SHGW2015,
+  MSAIT2019} attempts to address this issue by performing
+optimizations only in cases where correctness is preserved. \sys draws
+inspiration from these efforts, as we are also interested in
+optimizations that exploit parallelization but preserve the semantics
+of the underlying model. However, our proposed dataflow model is
+different from these other works, as nodes in our model read from
+their inputs \emph{in sequence}. Due to this intricacy, previous work
+is not straightforwardly applicable.
+
+%% Possible points to add: These systems introduce data parallelism by
+%% parallelizing nodes of the dataflow graph.  Unfortunately, they
+%% often do not preserve the model's semantics, blurring the lines
+%% between specification, optimization, and implementation.  In
+%% contrast, \sys's dataflow model inherently supports the data
+%% parallelism found in shell pipelines.  More specifically, it
+%% develops a set of parallelization-exposing optimizations
+%% represented as semantics-preserving graph transformations,
+%% effectively exposing data parallelism as task parallelism.
 

### 2020-02-23 osdi/draft.tex
Notes on ICFP/OSDI

@@
 
-* 
+* new evaluation case
+* split / micro-batching optimizations
+* rewrite paper
+* DSL
+
 Extends beyond linear pipelines to full programs 

### 2020-02-23 compiler/json_ast.py
Fix a TODO in json_ast.py

@@ -1,10 +1,18 @@
 import json
+import os
 import re
 
 from subprocess import run, PIPE
 
-## TODO: Don't hardcode this
-JSON_TO_SHELL_APP="../parser/json_to_shell.native"
+GIT_TOP_CMD = [ 'git', 'rev-parse', '--show-toplevel', '--show-superproject-working-tree']
+if 'DISH_TOP' in os.environ:
+    DISH_TOP = os.environ['DISH_TOP']
+else:
+    DISH_TOP = subprocess.run(GIT_TOP_CMD, capture_output=True,
+            text=True).stdout.rstrip()
+
+SHELL_TO_JSON = os.path.join(DISH_TOP, "parser/parse_to_json.native")
+JSON_TO_SHELL= os.path.join(DISH_TOP, "parser/json_to_shell.native")
 
 ### --- From JSON --- ###
 
@@ -45,7 +53,7 @@ def serialize_ast_json(ast):
 
 
 def json_to_shell(json_string):
-    subproc = run([JSON_TO_SHELL_APP], stdout=PIPE, input=json_string, encoding='ascii')
+    subproc = run([JSON_TO_SHELL], stdout=PIPE, input=json_string, encoding='ascii')
     return subproc.stdout
 
 def ast_to_shell(ast):

### 2020-02-25 pldi/draft.tex
Change format to ACM small for ICFP
Notes on plan

--- 2020-02-18 pldi/draft.tex
@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
+% Limit: 25 pages
+% https://icfp20.sigplan.org/track/icfp-2020-papers#Call-for-Papers
+\documentclass[acmsmall,10pt,review,anonymous]{acmart}
 \settopmatter{printfolios=false,printccs=false,printacmref=false}
@@
 \label{intro}
+
+\nv{I don't think the intro needs any serious rewrite, apart from deleting any
+notion of distribution and highlighting parallel processing.
+
+Key tasks, in rough order of importance:
+  (i) change title and intro to avoid distribution
+  (ii) cleanup and rewrite sections 4, 5: introduce our high-level figure, talk
+  about things in order, add ordering in graphs, generally refine and beautify
+  as much as possible.
+  (iii) tighten section 2, everything in there has to be related to some part of
+  the story---make the connection explicit
+  (iv) Pump up evaluation: could we add 3 more micro-benchmarks? could we add
+  1--2 more macro-benchmarks?
+  (v) since this is IFCP, could we add some simple semantics and derivation
+  rules? we should certainly be able to describe some of the optimizations as
+  congruence rules.
+  (vi) complete the related work with some operating-systems literature.
+}
 

### 2020-02-29 pldi/draft.tex
Rewrite and refine of 4.1

--- 2020-02-21 pldi/draft.tex
@@
 The core of \sys is a dataflow graph model, inspired by the ones used
-in popular distributed stream processing systems~\sx{related}. The
-dataflow model can express a large subset of Shell programs and
-inherently supports the data parallelism found in shell pipelines by
-exposing it as task parallelism. More preciely, we have developed a
-set of parallelization-exposing optimizations represented as
+in popular distributed stream processing systems~\sx{related} \kk{Is
+  this reference here necessary?}. The dataflow model can express a
+large subset of Shell programs and inherently supports the data
+parallelism found in shell pipelines by exposing it as task
+parallelism. More preciely, we have developed a set of
+parallelization-exposing optimizations represented as
 semantics-preserving graph transformations. These transformations can
@@
 %% script, \sys first translates it to a manipulable representation.
-
-\kk{It might be beneficial to do a full pass over this section
-  describing the new model (where a node that has many sequential
-  inputs is always prepended with a cat node.}
 
@@
 
-\paragraph{Nodes---Commands}
-
-For a set $D$, we write $D\kstar$ to denote the set of all finite
-words over $D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or
-$xy$ to denote their concatenation. We write $\eps$ for the empty
-word. We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq
-y$, if there is a word $z$ such that $y = xz$. It is easy to see that
-$\leq$ is a partial order, often called the \emph{prefix order}.
-
-A node $f$ of the graph represents a function from one (possibly
-empty) input stream to an output stream $f : D\kstar \to D\kstar$,
-where $D$ represents the basic data type of a line of characters
-\kk{We have to extend this to more general nodes. It has to be from
-  many inputs to many outputs. The commands that we optimize at the
-  moment are the ones that can be rewritten to take one stream input
-  to one stream output, and then some other inputs and outputs that
-  are considered static.}. This representation captures the majority
-of shell commands. We require that the function $f$ is monotone
-w.r.t.\ the prefix order. This captures the idea that a node cannot
-retract output after it has been emitted.
-%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
-%exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
-%concatenation and will be omitted when obvious.
-An example of a node is the following command:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- grep ``foo''
-\end{lstlisting}
-
-\noindent
-This command takes a stream of lines from its standard input and
-returns a stream of lines on its standard output.
-
-An important
-observation is that commands might read their input
-stream from a sequence of different sources. An illustrative example
-is the command \ttt{cat}, which reads its input from a combination
-of files and its standard input:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat x - y
-\end{lstlisting}
-
-\noindent
-In the example above, \ttt{cat} reads file \ttt{x} until it encounters the EOF
-character, then reads from \ttt{stdin} until in encounters EOF, and finally reads from
-\ttt{y}.
-
-In order to handle this particularity, nodes in the
-dataflow graph can have many incoming edges, that are \emph{ordered} in a
-sequence. This is different from standard dataflow
-graph representations where the semantics of multiple incoming edges
-is either zipping the inputs into pairs, or interleaving them arbitrarily.
-
-%% \TODO{Is the following paragraph best suited here? It describes an
-%%   assumption that we make about commands, to guarantee the validity of
-%%   our transformations. The assumption is needed due to the limitation
-%%   of the programming model of one input one output per command.}
-
-Nodes may read from more files than just the files of their
-input stream (resp. write to more files than their output stream).
-However, these files are static (\eg used for configuration or
-logging) and do not represent streams of data, and therefore are not
-considered part of the dataflow graph.
-For now, we assume that these static
-files can only be accessed by one node, thus ensuring that they do not
-interfere with the distributed implementation.
-%% Because of the assumption in section Front End \ref{}, that each file
-%% exists only once in the dataflow graph, we can safely assume that
-%% reading and writing to these static files does not interfere with the
-%% distributed implementation, and doesn't alter the behaviour of the
-%% program.
-
 \paragraph{Edges---Files}
+
+Before describing the graph edges, we introduce basic notation that
+will be useful later on.  For a set $D$, we write $D\kstar$ to denote
+the set of all finite words over $D$. For words $x, y \in D\kstar$, we
+write $x \cdot y$ or $xy$ to denote their concatenation. We write
+$\eps$ for the empty word. We say that $x$ is a \emph{prefix} of $y$,
+and we write $x \leq y$, if there is a word $z$ such that $y = xz$. It
+is easy to see that $\leq$ is a partial order, often called the
+\emph{prefix order}.
 
@@
 $D\kstar$. They could either refer to a specific named file in the file
-system, or just be FIFO pipes used for interprocess communication
-%% \TODO{Are there any other file types, such as URLs, ...?}.
+system, or just be FIFO pipes used for interprocess communication.
 
@@
 
-In order to expose data parallelism from the output of a command, files that
-do not point to a resource can have
-\begin{wrapfigure}[3]{r}{0.20\columnwidth}
-\vspace{-7pt}
-  \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
-\vspace{-20pt}
-\end{wrapfigure}
-an upper bound on the lines that they can transfer. This helps utilize available
-parallelism after encountering a command that cannot be parallelized
-(\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.
+%% \kk{I think we should delete this since the node and not the edge
+%%   chooses how to write to its output files.}
+
+%% In order to expose data parallelism from the output of a command,
+%% files that do not point to a resource can have
+%% \begin{wrapfigure}[3]{r}{0.20\columnwidth}
+%% \vspace{-7pt}
+%%   \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
+%% \vspace{-20pt}
+%% \end{wrapfigure}
+%% an upper bound on the lines that they can transfer. This helps utilize available
+%% parallelism after encountering a command that cannot be parallelized
+%% (\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.
 
@@
 
-%% \kk{(Maybe) It might be beneficial to mention that many inputs do not
-%%   represent the same thing as many outputs. These two notions are not
-%%   symmetric.}
+\paragraph{Nodes---Commands}
+
+A node $f$ of the graph represents a function from a (possibly empty)
+list of input streams to a list of output streams $f : [D\kstar] \to
+[D\kstar]$, where $D$ represents the basic data type of a line of
+characters. This representation captures the majority of shell
+commands. We require that the function $f$ is monotone w.r.t.\ a
+lifting of the prefix order for a sequence of inputs \kk{Formalize
+this lifting if essential for later development}. This captures the
+  idea that a node cannot retract output after it has been emitted.
+%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
+%exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
+%concatenation and will be omitted when obvious.
+An example of a node with one input and one output stream is the
+following command:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ grep ``foo''
+\end{lstlisting}
+
+\noindent
+This command takes a stream of lines from its standard input and
+returns a stream of lines on its standard output.
+
+An important observation is that even though commands might read their
+input from several input streams, the order in which they read is not
+always arbitrary. An illustrative example is the command \ttt{cat},
+which consumes its input one stream at a time:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+ cat x - y
+\end{lstlisting}
+
+\noindent
+In the example above, \ttt{cat} reads file \ttt{x} until it encounters
+the EOF character, then reads from \ttt{stdin} until in encounters
+EOF, and finally reads from \ttt{y}. Other commands that read their
+inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
+order in which a node consumes its input is necessary to enable
+certain optimizations on the graph.
+
+%% \kk{Some could read in sequence (like cat), some in lockstep (like
+%%   zip), some read some files statically and the rest as streams (like
+%%   comm), some read arbitrarily.}
+
+\kk{I need to mention that there are nodes that we know more
+  properties of. For example cat, tee, zip, etc... and these are
+  handled specially because of their very important and special
+  properties.}
 
@@
 
+\kk{I have to mention that in order for the transformations to happen,
+  we transform all commands that read some of their inputs in sequence
+  to commands that read from one input but are preceded by a cat.}
+
+\kk{I have to mention that all stateless and pure commands fall into
+  the category of commands that have one input stream (and possibly
+  several input static files) and one output stream (and no(?!?)
+  output static files?). These commands are the bulk of
+  data-processing in the shell, and these are the ones that we handle
+  in our system.}
+
+\kk{Move this assumption in the transformation}
+Nodes may read from more files than just the files of their
+input stream (resp. write to more files than their output stream).
+However, these files are static (\eg used for configuration or
+logging) and do not represent streams of data, and therefore are not
+considered part of the dataflow graph.
+For now, we assume that these static
+files can only be accessed by one node, thus ensuring that they do not
+interfere with the distributed implementation.
+%% Because of the assumption in section Front End \ref{}, that each file
+%% exists only once in the dataflow graph, we can safely assume that
+%% reading and writing to these static files does not interfere with the
+%% distributed implementation, and doesn't alter the behaviour of the
+%% program.
+
 Before describing the different types of parallelization
@@
 %% \end{lemma}
+
+\kk{We should move FIgure 4 here}
 
@@
 \label{optimizer}
+
+\kk{This subsection has to only talk about the cool shell constructs
+  that are added and about optimizations that are not in the dataflow
+  graph (like file splitting), and the implementation of split, the
+  pure merges , etc...}
 

### 2020-02-29 pldi/draft.tex
Refine the first part of 4.2

@@
 
-\kk{I have to mention that all stateless and pure commands fall into
-  the category of commands that have one input stream (and possibly
-  several input static files) and one output stream (and no(?!?)
-  output static files?). These commands are the bulk of
-  data-processing in the shell, and these are the ones that we handle
-  in our system.}
-
-\kk{Move this assumption in the transformation}
-Nodes may read from more files than just the files of their
-input stream (resp. write to more files than their output stream).
-However, these files are static (\eg used for configuration or
-logging) and do not represent streams of data, and therefore are not
-considered part of the dataflow graph.
-For now, we assume that these static
-files can only be accessed by one node, thus ensuring that they do not
-interfere with the distributed implementation.
+\kk{Move this assumption in the transformation: Nodes may read from
+  more files than just the files of their input stream (resp. write to
+  more files than their output stream).  However, these files are
+  static (\eg used for configuration or logging) and do not represent
+  streams of data, and therefore are not considered part of the
+  dataflow graph.  For now, we assume that these static files can only
+  be accessed by one node, thus ensuring that they do not interfere
+  with the distributed implementation.
 %% Because of the assumption in section Front End \ref{}, that each file
@@
 %% program.
-
-Before describing the different types of parallelization
-optimizations, we formalize the intuition about stateless and pure
-commands that was described earlier~\sx{cmd}.
+}
+
+Having a dataflow graph representation allows us to define
+parallelization optimizations as graph transformations. Before
+describing the different types of optimizations, we formalize the
+intuition about stateless and pure commands that was described
+earlier~\sx{cmd}. We focus on parallelizing commands that produce one
+output and consume their inputs one at a time, with the possible
+exception of static input files that are used for configuration. More
+formally, we parallelize commands that can be represented with
+functions of type $f : D\kstar \times [D\kstar] \to D\kstar$, where
+the first argument represents the concatenation of the inputs that are
+consumed one at a time and the second argument represents the static
+inputs. Most stateless and pure commands fall in this category
+\kk{Maybe have an argument here that we show in the evaluation that
+  handling these commands is adequate for a big variety of data
+  processing scripts, from several microbenchmarks to some
+  large-realistic macrobenchmarks}.
 %% \nv{TODOs
@@
 %% }
-%
-
-As mentioned there, stateless commands such as \ttt{tr} operate independently on
-individual elements of the stream (characters, lines, or files)
-without maintaining any state.
-% Some examples include \ttt{grep}, \ttt{basename}, and \ttt{tr}.
-Formally, a command $f$ is stateless if
-it is ``linear'' with respect to concatenation, \ie 
-% the following equation holds:
-satisfying the following equation:
+
+As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr} operate
+independently on individual elements of the stream (characters, lines,
+or files) without maintaining any state. In this work, we consider
+lines as the data quantum, so commands that are file-stateless are not
+considered stateless. Formally, a command $f$ is stateless if it is
+``linear'' with respect to concatenation, \ie satisfying the following
+equation:
 
 \[
-\forall x, x', f(x.x') = f(x).f(x')
+\forall x, x', s, f(x.x', s) = f(x, s).f(x', s)
 \]
@@
 \[
-\forall x, x', f(x.x') = r(m(x).m(x'))
+\forall x, x', s, f(x.x', s) = r(m(x, s).m(x', s), s)
 \]

### 2020-02-29 pldi/draft.tex
Finish a pass on 4 only leaving some notes for additions

@@
 
-\kk{I have to mention that in order for the transformations to happen,
-  we transform all commands that read some of their inputs in sequence
-  to commands that read from one input but are preceded by a cat.}
-
-\kk{Move this assumption in the transformation: Nodes may read from
-  more files than just the files of their input stream (resp. write to
-  more files than their output stream).  However, these files are
-  static (\eg used for configuration or logging) and do not represent
-  streams of data, and therefore are not considered part of the
-  dataflow graph.  For now, we assume that these static files can only
-  be accessed by one node, thus ensuring that they do not interfere
-  with the distributed implementation.
-%% Because of the assumption in section Front End \ref{}, that each file
-%% exists only once in the dataflow graph, we can safely assume that
-%% reading and writing to these static files does not interfere with the
-%% distributed implementation, and doesn't alter the behaviour of the
-%% program.
-}
+%% \kk{Move this assumption in the transformation: Nodes may read from
+%%   more files than just the files of their input stream (resp. write to
+%%   more files than their output stream).  However, these files are
+%%   static (\eg used for configuration or logging) and do not represent
+%%   streams of data, and therefore are not considered part of the
+%%   dataflow graph.  For now, we assume that these static files can only
+%%   be accessed by one node, thus ensuring that they do not interfere
+%%   with the distributed implementation.
+%% %% Because of the assumption in section Front End \ref{}, that each file
+%% %% exists only once in the dataflow graph, we can safely assume that
+%% %% reading and writing to these static files does not interfere with the
+%% %% distributed implementation, and doesn't alter the behaviour of the
+%% %% program.
+%% }
 
@@
 consumed one at a time and the second argument represents the static
-inputs. Most stateless and pure commands fall in this category
-\kk{Maybe have an argument here that we show in the evaluation that
-  handling these commands is adequate for a big variety of data
-  processing scripts, from several microbenchmarks to some
-  large-realistic macrobenchmarks}.
+inputs. When translating a command that consumes its input streams one
+at a time, a \ttt{cat} node is introduced before it to concatenate the
+streams and its output is transferred to the command input. \kk{Maybe
+  have an argument here that we show in the evaluation that handling
+  these commands is adequate for a big variety of data processing
+  scripts (and most stateless and pure commands), from several
+  microbenchmarks to some large-realistic macrobenchmarks.}.
 %% \nv{TODOs
@@
 %
-Based on these equations, we can define a maximal \emph{node split
-  operation} on a stateless node $v$ with $n$ input ordered edges and
-an output edge to node $v'$, to replace $v$ with $n$ new nodes, giving
-each one of them one of the incoming edges of $v$, and routing them in
-order to point to $v'$. Since each incoming edge represents a stream
-of data $x_i : D^*$, and the only behaviour of a dataflow graph is its
-output, this optimization $ v(x_1.x_2...x_n) \Rightarrow
-v(x_1).v(x_2)...v(x_n)$ can be shown to preserve the behaviour of the
+Based on these equations, we can define a \emph{node parallelization
+  transformation} on a stateless node $v$ that is preceded by a
+\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
+transformation replaces $v$ with $n$ new nodes, routing each of the
+$n$ input streams to one of them, and commutes the \ttt{cat} node
+after them to concatenate their outputs and transfer them to
+$v'$. Since each incoming edge represents a stream of data $x_i :
+D^*$, and the only behaviour of a dataflow graph is its output, this
+optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
+s)...v(x_n, s)$ can be shown to preserve the behaviour of the
 graph. Furthermore, it can be straightforwardly extended to pure
 commands that can be implemented by a map-reduce pair $(m, r)$ as $
-v(x_1.x_2...x_n) \Rightarrow r(m(x_1).m(x_2)...m(x_n))$.
+v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
+s)$. A visual representation of this transformation can be seen in
+\Cref{fig:parallelization-transformation}.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.60\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
+\caption{
+  \textbf{Stateless parallelization transformation.}
+  The \texttt{cat} node is commuted with the stateless node
+  to utilize available data parallelism.
+  \TODO{Correct this diagram to show the cat node before
+    the node and then the cat node after the stateless
+    command after the optimization. Cat node should be
+    different than a circle. Maybe like an ``and'' node
+    in binary circuits?}
+}
+\vspace{-10pt}
+\label{fig:parallelization-transformation}
+\end{figure}
 
@@
 
-\kk{We should move FIgure 4 here}
-
-\kk{I think that we should probably add the ``new'' split optimization
-  (and maybe any others that we develop until then) here.}
+\kk{I think that we should probably rewrite this paragraph with the
+  ``new'' split optimization (and maybe any others that we develop
+  until then) here.}
 
@@
 can then be implemented by spawning a process for each node, and
-redirecting the outputs according to the graph edges. An example
-transformation can be seen in \Cref{fig:optimizer-transformation}.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.40\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
-\caption{
-  \textbf{Example graph transformation.}
-  The optimizer splits a stateless node to utilize
-  available data parallelism.
-}
-\vspace{-10pt}
-\label{fig:optimizer-transformation}
-\end{figure}
+redirecting the outputs according to the graph edges.
 

### 2020-03-03 pldi/draft.tex
Change format to OSDI

@@
+%% OSDI -- 12 pages limit
+\documentclass[letterpaper,twocolumn,10pt]{article}
+\usepackage{usenix2019_v3}
+
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
@@
 % https://icfp20.sigplan.org/track/icfp-2020-papers#Call-for-Papers
-\documentclass[acmsmall,10pt,review,anonymous]{acmart}
-\settopmatter{printfolios=false,printccs=false,printacmref=false}
+%% \documentclass[acmsmall,10pt,review,anonymous]{acmart}
+%% \settopmatter{printfolios=false,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-\captionsetup[figure]{font=footnotesize,name={Fig.},labelfont={bf, footnotesize}}
-\captionsetup[table]{font=footnotesize,name={Tab.},labelfont={bf, footnotesize}, skip=2pt, aboveskip=2pt}
-\captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
+%% \captionsetup[figure]{font=footnotesize,name={Fig.},labelfont={bf, footnotesize}}
+%% \captionsetup[table]{font=footnotesize,name={Tab.},labelfont={bf, footnotesize}, skip=2pt, aboveskip=2pt}
+%% \captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
 
@@
 %% use defaults for review submission.
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
-\acmYear{2018}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
+%% \acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+%% \acmYear{2018}
+%% \acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+%% \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+%% \startPage{1}
 
@@
 %% use 'none' for review submission.
-\setcopyright{none}
+%% \setcopyright{none}
 %\setcopyright{acmcopyright}
@@
 %% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{plain}
 %% Citation style
@@
 \author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
-}
-\email{first1.last1@inst1.edu}          %% \email is recommended
+%% \authornote{with author1 note}          %% \authornote is optional;
+%%                                         %% can be repeated if necessary
+%% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%% \affiliation{
+%%   \position{Position1}
+%%   \department{Department1}              %% \department is recommended
+%%   \institution{Institution1}            %% \institution is required
+%%   \streetaddress{Street1 Address1}
+%%   \city{City1}
+%%   \state{State1}
+%%   \postcode{Post-Code1}
+%%   \country{Country1}                    %% \country is recommended
+%% }
+%% \email{first1.last1@inst1.edu}          %% \email is recommended
 
@@
 \author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
+%% \authornote{with author2 note}          %% \authornote is optional;
+%%                                         %% can be repeated if necessary
+%% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%% \affiliation{
+%%   \position{Position2a}
+%%   \department{Department2a}             %% \department is recommended
+%%   \institution{Institution2a}           %% \institution is required
+%%   \streetaddress{Street2a Address2a}
+%%   \city{City2a}
+%%   \state{State2a}
+%%   \postcode{Post-Code2a}
+%%   \country{Country2a}                   %% \country is recommended
+%% }
+%% \email{first2.last2@inst2a.com}         %% \email is recommended
+%% \affiliation{
+%%   \position{Position2b}
+%%   \department{Department2b}             %% \department is recommended
+%%   \institution{Institution2b}           %% \institution is required
+%%   \streetaddress{Street3b Address2b}
+%%   \city{City2b}
+%%   \state{State2b}
+%%   \postcode{Post-Code2b}
+%%   \country{Country2b}                   %% \country is recommended
+%% }
+%% \email{first2.last2@inst2b.org}         %% \email is recommended
 
@@
 \setlist{noitemsep,leftmargin=10pt,topsep=2pt,parsep=2pt,partopsep=2pt}
+
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+%% \begin{CCSXML}
+%% <ccs2012>
+%% <concept>
+%% <concept_id>10011007.10011006.10011008</concept_id>
+%% <concept_desc>Software and its engineering~General programming languages</concept_desc>
+%% <concept_significance>500</concept_significance>
+%% </concept>
+%% <concept>
+%% <concept_id>10003456.10003457.10003521.10003525</concept_id>
+%% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
+%% <concept_significance>300</concept_significance>
+%% </concept>
+%% </ccs2012>
+%% \end{CCSXML}
+
+%% \ccsdesc[500]{Software and its engineering~General programming languages}
+%% \ccsdesc[300]{Social and professional topics~History of programming languages}
+%% End of generated code
+
+%% Keywords
+%% comma separated list
+% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+
+%% \maketitle
+%% Note: \maketitle command must come after title commands, author
+%% commands, abstract environment, Computing Classification System
+%% environment and commands, and keywords command.
+\maketitle
 
@@
 %
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
 
@@
 % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency,distal:13,meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
 More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
@@
 
-\begin{acks}
+\section*{Acknowledgments}
+
+%% \begin{acks}
   % Dumping people so that we don't forget
   % 
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
+  %% This material is based upon work supported by the
+  %% \grantsponsor{GS100000001}{National Science
+  %%   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  %% No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  %% No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  %% conclusions or recommendations expressed in this material are those
+  %% of the author and do not necessarily reflect the views of the
+  %% National Science Foundation.
+%% \end{acks}
 

### 2020-03-03 pldi/draft.tex
A pass on 5.1

@@
   \item
-    \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and orchestration of distributed shell programs.
+    \S\ref{impl} describes the \sys implementation and addresses
+    several challenges related to the translation and optimization of
+    shell programs.
     %% details other concerns, such as distributed state management and extensibility. 
@@
 
-This section describes challenges related to the implementation of \sys. 
-It starts by describing how the \sys front-end translates a shell script to the dataflow graph model defined in \S\ref{ir}, by applying transformation on distributable regions~\sx{front-end}.
-It closes with the \sys planner, which  optimizes, distributes, and orchestrates the execution of a given dataflow graph~\sx{optimizer}.
+This section describes challenges related to the implementation of
+\sys. It starts by describing how the \sys front-end translates a
+shell script to the dataflow graph model defined in \S\ref{ir}, by
+applying transformation on parallelizable regions~\sx{front-end}.  It
+closes with the \sys optimizer, which optimizes the dataflow graph
+exposing data-parallelism, and outputs an optimized
+parallel script~\sx{optimizer}.
 
@@
 
-As mentioned in \Cref{ir}, in order to generate the distributed implementation of a shell script, \sys translates it to a dataflow graph.
-However, there are several components in shell programs that cannot be arbitrarily parallelized without affecting the program behaviour---\eg commands connected with a \ttt{\&\&} operator~\sx{bg}.
-The fact that the shell is highly dynamic compounds the challenge---\eg by not knowing what files does an unexpanded command argument refer to.
+As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
+translates it to a dataflow graph.  However, there are several
+components in shell programs that cannot be arbitrarily parallelized
+without affecting the program behaviour---\eg commands connected with
+a \ttt{\&\&} operator~\sx{bg}. In addition, the shell's highly dynamic
+nature compounds the challenge---\eg by not knowing what files does an
+unexpanded command argument refer to.
 % of correctly identifying what parts can be distributed, e.g.
@@
 
-To address this issue, we introduce the notion of distributable program regions and a translation pass for converting them to a dataflow graph.
-The translation pass identifies distributable regions and translates each of them to a dataflow graph, provided the abstract syntax tree (AST) of a shell script.
-The AST is produced by \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
-
-\paragraph{Distributable Regions}
+To address the first issue, we introduce the notion of parallelizable
+program regions and a translation pass for converting them to a
+dataflow graph. To address the second issue, we limit dynamic
+components by deferring translation as late as possible using a
+wrapping scheme.
+
+%% The translation pass identifies distributable regions and
+%% translates each of them to a dataflow graph, provided the abstract
+%% syntax tree (AST) of a shell script.  The AST is produced by
+%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
+
+\paragraph{Parallelizable Regions}
 %
@@
 %
-Distributable regions are (ideally, maximal) program sub-expressions that can be parallelized safely with respect to their sequential execution.
-The structure of a shell program can be used to identify components that can be executed concurrently and barriers that are natural points for enforcing synchronization.
-Consider the following example:
+Parallelizable regions are program sub-expressions that can be
+parallelized safely without altering the total program behaviour.
+%% respect to their sequential execution.
+The search for these regions can be guided by the structure of shells
+programs, which inherently contains information about components that
+can be executed independently and barriers that are natural points for
+enforcing synchronization. Consider the following example:
 
@@
 \noindent
-The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would already execute concurrently in the standard shell, while \ttt{sort f3} would wait for their completion before executing.
-Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3} are distributable regions---maximal sub-expressions that cannot be extended beyond \ttt{\&\&}.
-
-Intuitively, distributed regions correspond to subsections of the program that would be allowed to execute concurrently by different processes in the POSIX standard~\cite{posix}.
-The pipe operator \ttt{|} and the parallel composition operator \ttt{\&} enable parallelization, and extend distributable regions.
-Conversely, the \ttt{;} sequential composition operator, the \ttt{\&\&} logical operator, and function definitions represent barrier constructs that do not allow distributable regions to permeate through.
+The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would execute as
+independent processes in the standard shell, while \ttt{sort f3} would
+wait for their completion before executing.  Both \ttt{cat f1 f2 |
+  grep ``foo'' > f3} and \ttt{sort f3} are therefore parallelizable
+regions, regions that cannot be extended beyond \ttt{\&\&}---even
+better, they are maximal.
+
+Intuitively, parallelizable regions correspond to sub-expressions of
+the program that would be allowed to execute independently by
+different processes in the POSIX standard~\cite{posix}. Larger
+parallelizable regions can be composed from smaller ones using the
+pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
+Conversely, the \ttt{;} sequential composition operator, the
+\ttt{\&\&} logical operator, and function definitions represent
+barrier constructs that do not allow parallelizable regions to
+permeate through.
 
@@
 \paragraph{Translation Pass}
-The \sys front-end performs a depth-first search pass on the AST of the given shell program.
-During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
-To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
-To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.
+
+\kk{It might be a good idea to change this subsection based on the new
+  as-late-as-possible architecture where everything is wrapped and
+  translated at the final moment. Ideally, this would become the main
+  point of this subsection, as the translation algorithm is kind of
+  redundant when having the distributable regions too.}
+
+The \sys front-end performs a depth-first search pass on the AST of
+the given shell program.  During this pass, it greedily extends the
+distributable regions from the bottom up, translating their
+independent components to dataflow graph nodes until a barrier
+construct is reached.  To identify possible opportunities for
+parallelism, the translation pass also extracts the category of each
+node together with its input and output pairs.  To perform this
+extraction, it uses the command name to index into a key-value map
+with values that are expressed in the DSL for distributability
+characterization~\sx{ext}.
 

### 2020-03-21 pldi/draft.tex
Note
Update
New introduction

@@
 %% Title information
-\title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
+%% \title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
+\title{DPSh: Automated Data-Parallel Processing in the Shell}         %% [Short Title] is optional;
 %% \title{Extensible data processing in the shell}
@@
 
+%% \kk{After discussion on Wednesday Jan 8: Modify the narrative of
+%%   this paper to talk about data processing in the Shell. Make this
+%%   the main focus of the paper. This simplifies the narrative and
+%%   allows us to not talk about late planner activation as well as
+%%   environment variables. It also justifies why we only deal with
+%%   pure and stateless. Note that this is a very expressive model, as
+%%   shown by our benchmark suite. Note: Rememeber to clearly describe
+%%   the subset of the shell that we handle. To be more precise the
+%%   compositional operator of the shell that we can handle.}
+
+%% \kk{Note: A point that we need to establish early on and convince
+%%   the readers about is that the Shell IS used already for
+%%   data-processing, and that we are not just proposing it as an
+%%   alternative to other systems. We do not propose migrating to
+%%   shell scripts if you already have a parallel/efficient
+%%   solution. I am not sure if that needs a specific paragraph of its
+%%   own at the start of the introduction. Like: ``Consider the
+%%   following bash script that does some simple data
+%%   processing... Examples like that exists in several
+%%   domains/applications (examples, cite).''}
+
+%% \kk{Decide if we will mainly focus on scripts or pipelines and
+%%   choose wordings appropriately.}
+%
+Consider the following data processing Shell script.
+%% \kk{We might want to have a more complex script here.}
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
+  sort | uniq -c | sort -rn | head 5 > out
+\end{lstlisting}
+
+\noindent
+Scripts like the one above are commonly used for orchestrating
+data-processing workflows; connecting system commands with custom
+tools to perform complex tasks.
+%% \kk{Do we need to mention domains where that happens?}
+Even though shell pipelines are designed to exploit task parallelism,
+they completely ignore available data-parallelism. For small input
+sizes this isn't a problem since scripts take less than a second to
+execute, but as data sizes increase, failing to utilize the underlying
+parallel computational resources is a significant bottleneck.
+
+To address this concern, a Shell \emph{user} could resort to several
+different solutions (e.g. gnuparallel, tsc, flags in individual
+commands etc...) that offer some form of parallelism. However these
+solutions are too ad-hoc (each having a disparate parallelism model),
+too limited (they only apply to specific use-cases), and too
+coarse-grained. \kk{Tone the previous claims down and back them up.}
+At the same time, a \emph{developer} of Shell commands could rewrite
+her commands using low level tools (e.g. Raftlib, openMP, etc) to
+improve utilization and performance. Manual rewriting is expensive and
+can introduce new bugs, cascading changes, and divergence from legacy
+functionality.
+
+To address this issue for both \emph{users} and \emph{developers}, we
+propose \sys, a tool that automatically exploits available
+data-parallelism in a given shell script. The key insight behind \sys
+is that a large subset of the Shell language can be encoded as a
+dataflow graph, a flexible representation that allows for
+parallelization optimizations. \sys builds on this insight with a
+careful study of the parallelizability properties of shell primitives
+and commands, a definition of a dataflow graph model that is suitable
+for the shell, a set of parallelization optimizations that preserve
+the behaviour of the graph, and a late-bound command-prefixing scheme
+for deferring optimizations as late-as-possible (\ie when runtime
+information is available). \kk{Maybe elaborate some more on the
+  contributions in this paragraph.}  A \sys-enabled $p_1$ will run
+\ttt{tr}s on parallel streams, use a mostly-parallel tree of
+\ttt{sort}s, run \ttt{uniq} mostly in parallel, execute \ttt{head} in
+one of the streams, and (trivially) merge streams in \ttt{out}. The
+only requirement to support a new command is that its developer adds a
+lightweight annotation to describe its parallelizability properties.
+
+The paper is structured as follows. It starts by introducing the
+necessary background on shell scripting and overviewing \sys~\sx{bg}.
+Sections \ref{distributability}--\ref{impl} highlight key
+contributions:
+\begin{itemize}
+  \item
+  \S\ref{distributability} studies the parallelizability properties of
+  a large class of Shell commands (coreutils) and introduces an
+  lightweight annotation language for commands that could be executed
+  in a data-parallel manner. \kk{Mention that we did this for all
+    coreutils.}
+
+  \item
+  \S\ref{ir} presents a dataflow graph model for encoding shell
+  scripts, and a set of parallelization transformations that preserve
+  the semantics of the sequential program.
+
+  \item
+  \S\ref{impl} describes the \sys implementation and addresses several
+  challenges related to the translation and optimization of shell
+  programs. \kk{Rewrite that when we know for sure what the section
+    contains.}
+
+\end{itemize}
+
+\noindent
+\sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$)
+for parallel execution using a combination of real pipelines.
+%micro-bench\-marks %---popular shell one-liners that highlight
+%certain features---and a multi-line macro-benchmark---a popular
+%weather analysis script.
+After a comparison with related prior work~\sx{related}, the paper
+closes with a discussion of limitations and possible future
+directions~\sx{discussion}.
+%% Additional material, in anonymized form, is available at
+%% \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+
 \nv{I don't think the intro needs any serious rewrite, apart from deleting any
@@
   (vi) complete the related work with some operating-systems literature.
+  --------
+%   Need to highlight the use of ad-hoc flags
+%   Need to highlight use cases / goals:
+%     still POSIX compliant, you can replace /bin/bash with /bin/dish and see
+%     \emph{some} performance improvements, even if we can't handle all cases
+%     today--that is, dish enables a platform for further research.
+%
+%  Talk about order-aware dataflow analysis?
 }
 
-\kk{After discussion on Wednesday Jan 8: Modify the narrative of this
-  paper to talk about data processing in the Shell. Make this the main
-  focus of the paper. This simplifies the narrative and allows us to
-  not talk about late planner activation as well as environment
-  variables. It also justifies why we only deal with pure and
-  stateless. Note that this is a very expressive model, as shown by
-  our benchmark suite. Note: Rememeber to clearly describe the subset
-  of the shell that we handle. To be more precise the compositional
-  operator of the shell that we can handle.}
-
-\kk{New intro sketch:}
-
-\kk{Note: A point that we need to establish and convince the readers
-  about is that the Shell IS used already for data-processing, and
-  that we are not just proposing it as an alternative to other
-  systems. We do not propose migrating to shell scripts if you already
-  have a parallel/efficient solution. I am not sure if that needs a
-  specific paragraph of its own at the start of the
-  introduction. Like: ``Consider the following bash script that does
-  some simple data processing... Examples like that exists in several
-  domains/applications (examples, cite).''}
-
-\begin{itemize}
-\item Why is parallelism important (especially in the context of data
-  processing). Significant performance benefits as most computer
-  systems today have multiple processors that are under utilized
-\item (Similarly to how it is now) unfortunately, it is quite hard
-  reaping these benefits in contrast to sequential data
-  processing. [Show Bash pipeline].
-\item To benefit from parallelism, the current options are
-  extreme. Either very fine-grained and low level (parallelism using
-  fine-grained C libraries) for maximum performance, or very coarse
-  grained and bulky where one has to use a proper big-data
-  framework. (Can we fit the shell ones here (gnu-parallel, etc))
-\item The common thing is that one needs to rewrite big part of their
-  application, often requiring expertise and redesigning its
-  architecture, in order to get benefits that in the end might not
-  explain the cost.
-\item To address this issue, we propose Dish, a thin layer that
-  automatically! parallelizes a given shell script. Key insight is
-  that a large subset of the shell language can be encoded in a
-  dataflow graph model, a flexible representation that allows for
-  parallelization optimizations.
-\item Contributions:
-\item Study of the shell primitives, identification of
-  families/classes of commands that share
-  distributability/parallelization properties, and a classification of
-  coreutils/POSIX commands in these distributability properties.
-\item Dataflow graph model that is suitable for the shell
-\item A set of optimizations on this model that can be thought of as
-  rewriting rules on the dataflow graph, that are sound and exploit
-  parallelism.
-\end{itemize}
-
-\kk{End of new intro sketch}
+%% \kk{Begin of new intro sketch}
+%% \begin{itemize}
+%% \item Why is parallelism important (especially in the context of data
+%%   processing). Significant performance benefits as most computer
+%%   systems today have multiple processors that are under utilized
+%% \item (Similarly to how it is now) unfortunately, it is quite hard
+%%   reaping these benefits in contrast to sequential data
+%%   processing. [Show Bash pipeline].
+%% \item To benefit from parallelism, the current options are
+%%   extreme. Either very fine-grained and low level (parallelism using
+%%   fine-grained C libraries) for maximum performance, or very coarse
+%%   grained and bulky where one has to use a proper big-data
+%%   framework. (Can we fit the shell ones here (gnu-parallel, etc))
+%% \item The common thing is that one needs to rewrite big part of their
+%%   application, often requiring expertise and redesigning its
+%%   architecture, in order to get benefits that in the end might not
+%%   explain the cost.
+%% \item To address this issue, we propose Dish, a thin layer that
+%%   automatically! parallelizes a given shell script. Key insight is
+%%   that a large subset of the shell language can be encoded in a
+%%   dataflow graph model, a flexible representation that allows for
+%%   parallelization optimizations.
+%% \end{itemize}
+%% Contributions:
+%% \begin{itemize}
+%% \item Study of the shell primitives, identification of
+%%   families/classes of commands that share
+%%   distributability/parallelization properties, and a classification of
+%%   coreutils/POSIX commands in these distributability properties.
+%% \item Dataflow graph model that is suitable for the shell
+%% \item A set of optimizations on this model that can be thought of as
+%%   rewriting rules on the dataflow graph, that are sound and exploit
+%%   parallelism.
+%% \end{itemize}
+%% \kk{End of new intro sketch}
 
@@
 
-Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
-% offer notable benefits over their centralized counterparts:
-%   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
-% Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
-% The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
-Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
-Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
-  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
-% 
-To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
-
-% \begin{lstlisting}[language=sh,float=h,numbers=none]
-% cat doc.ms |                   # print input file
-% groff -t -e -mandoc -Tascii |  # remove formatting
-% col -bx |                      # remove backspaces 
-% tr A-Z a-z |                   # convert to lower case
-% tr -d '[:punct:]' |            # remove punctuation
-% sort |                         # put words in alphabetical order
-% unique |                       # remove duplicate words
-% comm -13 /usr/share/dict -     # report words not in dictionary 
-% \end{lstlisting}
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
-  sort | uniq -c | sort -rn | head 5 > out
-\end{lstlisting}
-% \smallskip
-% \noindent
-Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
-Combining several features, \unix makes small tasks easy to express;
-  for a user with \emph{one} computer and a small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
-% composition allows the composition of general primitives
-
-% Key insight 
-% % The garden-hose philosophy is similar to Haskell 
-% This pipeline takes an input file (line 1), removes 
-% * do not need to use specialized frameworks for composition---instead, as long as they conform to the shell, individual primitives can be written in any programming language.
-% * do not need to rewrite ---
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-\caption{
-  \textbf{High-level schematic.}
-  \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
-  execution (right).
-}
-\vspace{-10pt}
-\label{fig:schematic}
-\end{figure}
-
-Unfortunately, developing $p_1$'s distributed equivalent requires significant effort.
-% For a user with many computers and larger inputs, 
-% The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency,distal:13,meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
+%% Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
+%% % offer notable benefits over their centralized counterparts:
+%% %   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
+%% % Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
+%% % The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
+%% Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
+%% Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
+%%   % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
+%%    scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
+%% % 
+%% To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
+
+%% % \begin{lstlisting}[language=sh,float=h,numbers=none]
+%% % cat doc.ms |                   # print input file
+%% % groff -t -e -mandoc -Tascii |  # remove formatting
+%% % col -bx |                      # remove backspaces 
+%% % tr A-Z a-z |                   # convert to lower case
+%% % tr -d '[:punct:]' |            # remove punctuation
+%% % sort |                         # put words in alphabetical order
+%% % unique |                       # remove duplicate words
+%% % comm -13 /usr/share/dict -     # report words not in dictionary 
+%% % \end{lstlisting}
+%% % \smallskip
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%% cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
+%%   sort | uniq -c | sort -rn | head 5 > out
+%% \end{lstlisting}
+%% % \smallskip
+%% % \noindent
+%% Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
+%% Combining several features, \unix makes small tasks easy to express;
+%%   for a user with \emph{one} computer and a small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
+%% % composition allows the composition of general primitives
+
+%% % Key insight 
+%% % % The garden-hose philosophy is similar to Haskell 
+%% % This pipeline takes an input file (line 1), removes 
+%% % * do not need to use specialized frameworks for composition---instead, as long as they conform to the shell, individual primitives can be written in any programming language.
+%% % * do not need to rewrite ---
+
+%% \begin{figure}[t]
+%% \centering
+%% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
+%% \caption{
+%%   \textbf{High-level schematic.}
+%%   \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
+%%   execution (right).
+%% }
+%% \vspace{-10pt}
+%% \label{fig:schematic}
+%% \end{figure}
+
+%% Unfortunately, developing $p_1$'s distributed equivalent requires significant effort.
+%% % For a user with many computers and larger inputs, 
+%% % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
+%% For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency,distal:13,meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
+%% More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
 %  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
-Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
-
-To answer this question we develop \sys, a shell variant that transforms traditional \unix pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-The key insight behind \sys is that a large part of the shell language can be encoded in a distributed dataflow graph model, providing structure to distribute computations.
-%% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
-% FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
+%% In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
+%% Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
+
+%% To answer this question we develop \sys, a shell variant that transforms traditional \unix pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
+%% The key insight behind \sys is that a large part of the shell language can be encoded in a distributed dataflow graph model, providing structure to distribute computations.
+%% %% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
+%% % FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
 Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
-  use a mostly-parallel tree of \ttt{sort}s,
-  run \ttt{uniq} mostly in parallel,
-  execute \ttt{head} in one of the streams,
-  and (trivially) merge streams in \ttt{out}.
-
-The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
-% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-Second, developers of new shell commands can use a lightweight domain-specific language to express their distributability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
-Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
-% Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
-
-% % Make Unix benefits explicit?
-% Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
-% Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
-% Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
-
-The paper is structured as follows.
-It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{impl} highlight key
-contributions:
-\begin{itemize}
-
-  \item
-  \S\ref{distributability} studies several distributability classes and introduces the distributability characterization language.
-
-  \item
-  \S\ref{ir} presents a dataflow graph model for encoding shell pipelines, and a set of parallelization transformations that preserve the semantics of the sequential program.
-
-  \item
-    \S\ref{impl} describes the \sys implementation and addresses
-    several challenges related to the translation and optimization of
-    shell programs.
-    %% details other concerns, such as distributed state management and extensibility. 
-\end{itemize}
-
-\noindent
-\sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$) for parallel execution using a combination of real pipelines.
+%% \sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+%% A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
+%%   use a mostly-parallel tree of \ttt{sort}s,
+%%   run \ttt{uniq} mostly in parallel,
+%%   execute \ttt{head} in one of the streams,
+%%   and (trivially) merge streams in \ttt{out}.
+
+%% % Developers of commands
+%% % Developers of pipelines
+%% % Language unifies both---no need for `-t` `-p`
+%% % POSIX:
+%%   % https://pubs.opengroup.org/onlinepubs/009695399/utilities/xcu_chap02.html
+
+%% The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
+%% % \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
+%% First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+%% Second, developers of new shell commands can use a lightweight domain-specific language to express their distributability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
+%% Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
+%% % Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
+
+%% % % Make Unix benefits explicit?
+%% % Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
+%% % Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
+%% % Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
+
+%% The paper is structured as follows.
+%% It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
+%% Sections \ref{distributability}--\ref{impl} highlight key
+%% contributions:
+%% \begin{itemize}
+
+%%   \item
+%%   \S\ref{distributability} studies several distributability classes and introduces the distributability characterization language.
+
+%%   \item
+%%   \S\ref{ir} presents a dataflow graph model for encoding shell pipelines, and a set of parallelization transformations that preserve the semantics of the sequential program.
+
+%%   \item
+%%     \S\ref{impl} describes the \sys implementation and addresses
+%%     several challenges related to the translation and optimization of
+%%     shell programs.
+%%     %% details other concerns, such as distributed state management and extensibility. 
+%% \end{itemize}
+
+%% \noindent
+%% \sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$) for parallel execution using a combination of real pipelines.
 %micro-bench\-marks %---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
-After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
-% \bigskip
-% \begin{quote}
-% \footnotesize
-Additional material, in anonymized form, is available at
-\href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
-% \end{quote}
-
-% \begin{figure*}[t]
-% \centering
-% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-% \caption{
-%   \textbf{Applying \sys to $p_1$.}
-% }
-% \label{fig:example}
-% \end{figure*}
-%
+%% After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
+%% % \bigskip
+%% % \begin{quote}
+%% % \footnotesize
+%% Additional material, in anonymized form, is available at
+%% \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+%% % \end{quote}
+
+%% % \begin{figure*}[t]
+%% % \centering
+%% % \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
+%% % \caption{
+%% %   \textbf{Applying \sys to $p_1$.}
+%% % }
+%% % \label{fig:example}
+%% % \end{figure*}
+%% %
 
@@
 
+\kk{Maybe guide all of this by showing an example of an output shell
+  script?}
+
 Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.

### 2020-03-21 pldi/draft.tex
Nit

@@
   \S\ref{ir} presents a dataflow graph model for encoding shell
-  scripts, and a set of parallelization transformations that preserve
-  the semantics of the sequential program.
+  scripts, and a set of safe parallelization transformations that
+  preserve the semantics of the sequential program.
 

### 2020-03-29 compiler/dish.py
Minor corrections

@@ -38,7 +38,7 @@ def main():
 
 def parse_shell(input_script_path):
     parser_output = subprocess.run([config.PARSER_BINARY, input_script_path], capture_output=True, text=True)
-    if(not parser_output.returncode == 0):
+    if (not parser_output.returncode == 0):
         print(parser_output.stderr)
         parser_output.check_returncode()
     return parser_output.stdout

### 2020-03-29 compiler/ast_to_ir.py
Minor corrections

@@ -75,7 +75,7 @@ def compile_asts(ast_objects, fileIdGen, config):
         print(ast_object)
 
         ## Compile subtrees of the AST to out intermediate representation
-        compiled_ast = compile_ast(ast_object, fileIdGen, config)
+        compiled_ast = compile_node(ast_object, fileIdGen, config)
 
         # print("Compiled AST:")
         # print(compiled_ast)
@@ -112,9 +112,6 @@ def compile_asts(ast_objects, fileIdGen, config):
 
     return compiled_asts
 
-
-def compile_ast(ast_object, fileIdGen, config):
-    return compile_node(ast_object, fileIdGen, config)
 
 def compile_node(ast_object, fileIdGen, config):
     global compile_cases

### 2020-03-29 compiler/json_ast.py

@@ -20,7 +20,6 @@ def parse_json_ast_string(json_string):
     ast_objects = [parse_json_line(line) for line in lines]
     return ast_objects
 
-
 ## Returns a list of AST objects
 def parse_json_ast(json_filename):
     with open(json_filename) as json_file:

### 2020-03-29 compiler/test.py
Add trivial test to show wrapper use

@@ -1,0 +1,11 @@
+import parse
+import json
+import wrapper
+import config
+
+print("Parse binary {}".format(config.PARSER_BINARY))
+
+ast = parse.shell_file_to_ir("./t1.sh")
+wrapper.rewrite_ast(ast)
+print("Wrapped {} commands".format(wrapper.get_results()[0]))
+parse.ir_to_shell(ast, "./t2.sh")

### 2020-03-30 pldi/draft.tex
Add some notes in the distributability DSL section

@@
 \paragraph{Distributability DSL}
-The DSL for distributability description (Fig.~\ref{fig:dsl}) is
-designed to classify commands into multiple categories
-given their options.
+The DSL for distributability description (Fig.~\ref{fig:dsl} \kk{I
+  think that the language definition should be in a technical report,
+  and here we should only give intuition and the \ttt{comm} example})
+is designed to classify commands into multiple categories given their
+options.
 
@@
 arbitrary commands in this category, \sys allows command developers to
-provide custom map and reduce steps for them, where the map
-is in \sta and the reduce in \pur, and their composition produces the
-same output with the initial command. \sys can then replace occurrences
-of these commands with their map and reduce parts, enabling their
-parallelization.
+provide custom map and reduce steps for them, where the map is in \sta
+and the reduce in \pur, and their composition produces the same output
+with the initial command. \sys can then replace occurrences of these
+commands with their map and reduce parts, enabling their
+parallelization. \TODO{Mention that we have done this for several pure
+  commands in coreutils, providing the foundation}
 
@@
 
+\TODO{Talk about the importance of eagerness and why we need to
+  enforce it to get benefits. Discuss the necessity of having files as
+  buffers and how we addressed the issue of cat's lazy consumption of
+  its inputs.}
+
 This section describes challenges related to the implementation of

### 2020-04-06 compiler/generate_unix50_scripts.py
Make script that generates unix50 pipelines from full script

@@ -1,6 +1,6 @@
 import sys
 import os
-from shutil import copyfile
+import re
 
 unix50_dir = sys.argv[1]
 intermediaries_dir = sys.argv[2]
@@ -19,7 +19,8 @@ with open(unix50_script) as file:
 ## Generate bigger inputs
 unix50_script_lines = unix50_script_data.split("\n")
 input_lines = [line for line in unix50_script_lines if line.startswith("IN")]
-input_file_names = [line.split("=")[1] for line in input_lines]
+input_file_name_assignments = {line.split("=")[0] : line.split("=")[1] for line in input_lines}
+input_file_names = input_file_name_assignments.values()
 
 for input_file in input_file_names:
     input_file_path = os.path.join(unix50_dir, input_file)
@@ -30,5 +31,26 @@ for input_file in input_file_names:
     with open(generated_input_file_path, "w") as file:
         file.write(input_file_data * input_size_increase)
 
+## Extract each pipeline of unix50
+unix50_pipelines = [line for line in unix50_script_lines if line.startswith("cat ")]
+p = re.compile('ab*')
 
-## TODO: Extract each pipeline of unix50
+for i, unix50_pipeline in enumerate(unix50_pipelines):
+
+    ## Find the input's name
+    input_name = re.search('IN[0-9]+', unix50_pipeline).group(0)
+    unix50_normalized_pipeline = re.sub('IN[0-9]+', 'IN', unix50_pipeline)
+
+    ## Generate the script
+    pipeline_file_path = os.path.join(intermediaries_dir, "unix50_pipeline_{}.sh".format(i))
+    with open(pipeline_file_path, "w") as file:
+        file.write(unix50_normalized_pipeline + "\n")
+
+    ## And its environment
+    pipeline_env_file_path = os.path.join(intermediaries_dir, "unix50_pipeline_{}_env.sh".format(i))
+
+    input_file_assignment_path = os.path.join(generated_inputs_dir, input_file_name_assignments[input_name])
+    environment_data = "IN={}\n".format(input_file_assignment_path)
+    print(environment_data)
+    with open(pipeline_env_file_path, "w") as file:
+        file.write(environment_data)

### 2020-04-06 evaluation/unix50/unix50.sh
Make script that generates unix50 pipelines from full script

@@ -17,15 +17,15 @@ IN5=5.txt
 IN6=6.txt
 IN7=7.txt
 IN8=8.txt
-IN9_1=9.1.txt
-IN9_2=9.2.txt
-IN9_3=9.3.txt
-IN9_4=9.4.txt
-IN9_5=9.5.txt
-IN9_6=9.6.txt
-IN9_7=9.7.txt
-IN9_8=9.8.txt
-IN9_9=9.9.txt
+IN91=9.1.txt
+IN92=9.2.txt
+IN93=9.3.txt
+IN94=9.4.txt
+IN95=9.5.txt
+IN96=9.6.txt
+IN97=9.7.txt
+IN98=9.8.txt
+IN99=9.9.txt
 IN10=10.txt
 IN11=11.txt
 IN12=12.txt
@@ -102,31 +102,31 @@ cat $IN8 | fmt -w1 | tr -c '[a-z][A-Z]' '\n' | sort | awk 'length >= 16'
 cat $IN8 > /dev/null
 
 # 9.1: extract the word PORT
-cat $IN9_1 | fmt -w1 | grep '[A-Z]' | tr '[a-z]' '\n' | grep '[A-Z]' | tr -d '\n' | cut -c 1-4
+cat $IN91 | fmt -w1 | grep '[A-Z]' | tr '[a-z]' '\n' | grep '[A-Z]' | tr -d '\n' | cut -c 1-4
 
 # 9.2: extract the word BELL
-cat $IN9_2 | cut -c 1-1 | tr -d '\n'
+cat $IN92 | cut -c 1-1 | tr -d '\n'
 
 # 9.3: animal that used to decorate the Unix room
-cat $IN9_3 | cut -c 1-2 | tr -d '\n'
+cat $IN93 | cut -c 1-2 | tr -d '\n'
 
 # 9.4: four corners with E centered, for an "X" configuration
-cat $IN9_4 | fmt -w1 | grep '"' | sed 4d | cut -d '"' -f 2 | tr -d '\n'
+cat $IN94 | fmt -w1 | grep '"' | sed 4d | cut -d '"' -f 2 | tr -d '\n'
 
 # 9.5: backwards running clock, in a backwards poem
-cat $IN9_5 > /dev/null
+cat $IN95 > /dev/null
 
 # 9.6: Follow the directions for grep
[…]

### 2020-04-06 pldi/draft.tex
Notes on intro

@@
 %% \title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
-\title{DPSh: Automated Data-Parallel Processing in the Shell}         %% [Short Title] is optional;
+\title{\sys: Automated Data-Parallel Shell Processing}         %% [Short Title] is optional;
 %% \title{Extensible data processing in the shell}
@@
 \newcommand{\sx}[1]{(\S\ref{#1})}
-\newcommand{\sys}{{\scshape Dish}\xspace}
+\newcommand{\sys}{{\scshape Pash}\xspace}
 \newcommand{\unix}{{\scshape Unix}\xspace}
@@
 
-\nv{I don't think the intro needs any serious rewrite, apart from deleting any
-notion of distribution and highlighting parallel processing.
-
-Key tasks, in rough order of importance:
-  (i) change title and intro to avoid distribution
-  (ii) cleanup and rewrite sections 4, 5: introduce our high-level figure, talk
-  about things in order, add ordering in graphs, generally refine and beautify
-  as much as possible.
-  (iii) tighten section 2, everything in there has to be related to some part of
-  the story---make the connection explicit
-  (iv) Pump up evaluation: could we add 3 more micro-benchmarks? could we add
-  1--2 more macro-benchmarks?
-  (v) since this is IFCP, could we add some simple semantics and derivation
-  rules? we should certainly be able to describe some of the optimizations as
-  congruence rules.
-  (vi) complete the related work with some operating-systems literature.
-  --------
+% TODO: Important to have a small script that highlights
+\nv{
+(i) different classes of commands
+(ii) challenge of shell expansion order
+(iii) segments that are parallelized by \sys (with provable correctness)
+(iv) segments that are \emph{not} parallelized by \sys (otherwise they can lead to
+incorrectness)
+}
+
+% \nv{I don't think the intro needs any serious rewrite, apart from deleting any
+% notion of distribution and highlighting parallel processing.
+% 
+% Key tasks, in rough order of importance:
+%   (i) change title and intro to avoid distribution
+%   (ii) cleanup and rewrite sections 4, 5: introduce our high-level figure, talk
+%   about things in order, add ordering in graphs, generally refine and beautify
+%   as much as possible.
+%   (iii) tighten section 2, everything in there has to be related to some part of
+%   the story---make the connection explicit
+%   (iv) Pump up evaluation: could we add 3 more micro-benchmarks? could we add
+%   1--2 more macro-benchmarks?
+%   (v) since this is IFCP, could we add some simple semantics and derivation
+%   rules? we should certainly be able to describe some of the optimizations as
+%   congruence rules.
+%   (vi) complete the related work with some operating-systems literature.
+%   --------
 %   Need to highlight the use of ad-hoc flags
@@
 %  Talk about order-aware dataflow analysis?
-}
+% }
 

### 2020-04-07 pldi/draft.tex
Notes on pldi draft

@@
 
+\subsection{Implementation Nuggets}
+
+\TODO{Problem: Cat (and other commands that have multiple inputs) is
+  not eager. Illustration: Show a parallel pipeline having a cat that
+  gathers input somewhere (and a sort merge maybe). The way pipes
+  work, this graph will essentially evaluate almost completely
+  sequentially. Maybe go through the steps that cat asks for the first
+  line of its first input until ithe first input is done and then goes
+  to the second. This leads to processor under-utilization. Solution:
+  We add eager nodes that eagerly consume all their input (even if
+  their output does not ask for it). Eager nodes are themselves part
+  of the dataflow graph (and model) and essentially perform an
+  identity function. This makes all processors as utilized as possible
+  and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
+  light) tasks it has some overhead. We evaluate eager in section X
+  (We should also say something about a CPU light task).}
+
+\TODO{(This problem is a nice followup for the one above) Problem:
+  Dataflow graph termination. Illustration: Show the example of head
+  deadlock and explain that this is inherent behaviour in the shell.
+  (Solution: emptying pipelines or terminating processes) In order for
+  things to terminate gracefully, dish wraps all commands in order to
+  always consume all of their input by sending it down to
+  /dev/null. Warning: This means that all commands in the pipeline
+  must terminate if all of their output is consumed. We evaluate the
+  overhead of emptying all outputs in section X.}
+
+\TODO{Problem: Pipeline width shrinks after parallel pure
+  commands. Solution: Split node that resplits its input to multiple
+  files in sequence (to preserve the model semantics). We should
+  probably evaluate that too.}
+
 \paragraph{Merge and Split Implementation}
@@
 \label{eval}
+
+\TODO{Notes on Unix50. We got the Unix50 solutions as is and we ran
+  them. These are the numbers. By changing fmt -w1 to tr (or by adding
+  it appropriately to the DSL) and by changing some occurences of awk
+  to ..., the speedup in these cases goes to...}
+
+\TODO{Remember to mention that deathstar has 128 threads but 64
+  cores. So it is reasonable for the speedup to not go further in cpu
+  heavy pipelines.}
 

### 2020-04-07 pldi/draft.tex
Take a stab on the eager figure

--- 2020-04-06 pldi/draft.tex
@@
 
+\begin{figure}[t]
+\centering
+% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
+\caption{
+  \textbf{Eager primitive.}
+  Describe three problems:
+    (i) FIFOs alone cannot solve it as they’re blocking;
+    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
+    (iii) writing to files + wait does not allow trask-based parallelism.
+  Eager is an elegant solution that solves the problem while remaining within
+  the model. Describe.
+}
+\vspace{-15pt}
+\label{fig:eager}
+\end{figure}
+
 \paragraph{Merge and Split Implementation}
@@
       (iii) +merge, adding a final merger step.
+      \nv{Candidate values, based on the HW we have 2, 4, 8, 16, 32, 64. Also, maybe it makes sense to show log x-axis?}
     }

### 2020-04-07 pldi/draft.tex

--- 2020-04-07 pldi/draft.tex
@@
 
+\begin{figure}[t]
+\centering
+% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
+\caption{
+  \textbf{Eager primitive.}
+  Describe three problems:
+    (i) FIFOs alone cannot solve it as they’re blocking;
+    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
+    (iii) writing to files + wait does not allow trask-based parallelism.
+  Eager is an elegant solution that solves the problem while remaining within
+  the model. Describe.
+}
+\vspace{-15pt}
+\label{fig:eager}
+\end{figure}
+
 \subsection{Implementation Nuggets}

### 2020-04-07 pldi/draft.tex

@@
   /dev/null. Warning: This means that all commands in the pipeline
-  must terminate if all of their output is consumed. We evaluate the
-  overhead of emptying all outputs in section X.}
+  must terminate gracefully if all of their output is consumed (if a
+  process dies unexpectedly, noone will consume its input. We leave
+  addressing that and other monitoring questions for future work). We
+  evaluate the overhead of emptying all outputs in section X.}
 

### 2020-04-07 pldi/draft.tex
Placeholder for the unix50 subsection

@@
 
-\TODO{Notes on Unix50. We got the Unix50 solutions as is and we ran
-  them. These are the numbers. By changing fmt -w1 to tr (or by adding
-  it appropriately to the DSL) and by changing some occurences of awk
-  to ..., the speedup in these cases goes to...}
-
 \TODO{Remember to mention that deathstar has 128 threads but 64
@@
 
+\subsection{Unix50}
+\label{unix50}
+
+The goal of this subsection is to evaluate \sys on an existing set of
+unix pipelines found in the wild. \todo{Motivate why unix50 is a nice
+  set of pipelines.} We took the solutions for the unix50 from
+\todo{Ref} and we executed them \emph{without any modification} with
+\sys's parallelism set to \todo{X}. The input given to the pipelines
+was the test input from unix50 multiplied \todo{10000} times. The
+speedup of each pipeline separately is shown in \todo{ref}. The
+average\todo{/median} speedup is \todo{X}.
+
+All pipelines that perform poorly (\todo{enumerate them}) contain
+non-parallelizable commands. An illustrative example is pipeline
+\todo{X}, which is shown here:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+  TODO
+\end{lstlisting}
+
+Notice in the above the use of (awk, fmt, or something else). We
+noticed that these occured in other pipelines too. By changing them to
+\todo{X}, the average speedup of these pipelines improved to \todo{x}
+and the new total average speedup is \todo{X}.
+
+\subsection{Comparison}
+
+\todo{Include comparison with Niko's speeduptest, with gnu-parallel,
+  with sort -parallel (and possibly other commands that have parallel
+  flags), and finally with Raftlib.}
+
 \subsection{Macro-benchmark: Weather Analysis}
@@
 For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
+\kk{I think it is fine keeping the Hadoop evaluation here, since this macrobenchmark was taken from a Hadoop book.}
 

### 2020-04-07 evaluation/unix50/unix50.sh
Comments and notes on unix50 pipelines that have no speedup

@@ -122,14 +122,18 @@ cat $IN95 > /dev/null
 # 9.6: Follow the directions for grep
 cat $IN96 | tr ' ' '\n' | grep '[A-Z]' | sed 1d | sed 3d | sed 3d | tr '[a-z]' '\n' | grep '[A-Z]' | sed 3d | tr -c '[A-Z]' '\n' | tr -d '\n'
 
-# 9.7: Four corners --- kk: Niko can we make this more general?
+# 9.7: Four corners
 cat $IN97 | sed 2d | sed 2d | tr -c '[A-Z]' '\n' | tr -d '\n'
 
 # 9.8: TELE-communications
 cat $IN98 | tr -c '[a-z][A-Z]' '\n' | grep '[A-Z]' | sed 1d | sed 2d | sed 3d | sed 4d | tr -c '[A-Z]' '\n' | tr -d '\n'
+# kk: A better way to implement this (since it want the final capital letter of each line)
+cat $IN98 | tr -cd '[A-Z]\n' | grep -o ".$" | tr -d '\n'
 
 # 9.9:
 cat $IN99 | tr -c '[a-z][A-Z]' '\n' | grep '[A-Z]' | sed 1d | sed 1d | sed 2d | sed 3d | sed 5d | tr -c '[A-Z]' '\n' | tr -d '\n'
+# kk: Similarly a better way to implement this (based on the above is
+cat $IN99 | tr -cd '[A-Z]\n' | grep -o "..$" | tr -d '\n'
 
 # 10.1: count Turing award recipients while working at Bell Labs
 cat $IN10 | sed 1d | grep 'Bell' | cut -f 2 | wc -l

### 2020-04-07 compiler/gather_results.py
New unix50 plots

@@ -187,9 +187,11 @@ def collect_unix50_scaleup_times(unix50_results_dir):
     ax.set_ylabel('Speedup')
     ax.set_xlabel('Pipeline')
     plt.bar(range(len(individual_results)), individual_results, align='center')
-    plt.yscale("log")
-    plt.yticks([0, 0.5, 1.0, 5, 10, 20])
-    plt.ylim((0, 20))
+    plt.hlines([1], -1, len(individual_results) + 1, linewidth=0.8)
+    plt.xlim(-1, len(individual_results) + 1)
+    # plt.yscale("log")
+    plt.yticks(range(1, 18, 2))
+    # plt.ylim((0.1, 20))
     # plt.legend(loc='lower right')
     plt.title("Unix50 Individual Speedups")
     plt.tight_layout()

### 2020-04-07 compiler/execute_evaluation_scripts.sh
Remove temporary eager files after running evaluation

@@ -3,6 +3,7 @@
 microbenchmarks_dir="../evaluation/microbenchmarks/"
 intermediary_dir="../evaluation/intermediary/"
 
+rm /tmp/eager*
 mkdir -p $intermediary_dir
 
 n_inputs=(

### 2020-04-07 pldi/draft.tex
Comments and notes on unix50 pipelines that have no speedup

@@
 \todo{Ref} and we executed them \emph{without any modification} with
-\sys's parallelism set to \todo{X}. The input given to the pipelines
+\sys's parallelism set to \todo{20}. The input given to the pipelines
 was the test input from unix50 multiplied \todo{10000} times. The
-speedup of each pipeline separately is shown in \todo{ref}. The
-average\todo{/median} speedup is \todo{X}.
+speedup of each pipeline separately is shown in
+\Cref{fig:unix50-individual}. The average\todo{/median} speedup is
+\todo{5.24}.
+
+%% The current average speedup includes unimplemented and small input pipelines.
+
+%% Bad result descriptions:
+%%  2. Contains just a head -n 2, and then a cut.
+%%  5. Very fast (small input). Sequential time is 0m0.103s (If we increase the input we will show speedup)
+%%  18. TODO: Implement
+%%  20. Four corners (not parallelizable). So performance is reasonable
+%%      (atm it doesn't return the correct result for multiplied input though)
+%%  21. Last capital character of each line. Their solution is not general (and it is not parallelizable).
+%%      Also, the input is very small. Changing it to a more general one and
+%%      increasing input size leads to speedup. TODO: Run with large input
+%%  22. Exactly the same problem as the above.
+%%  24. (Not sure: It seems that the computation is trivial)
+%%  32. TODO: Implement
+
+%% WARNING: Remember to fix `sed` and not have it be stateless in general.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_20.pdf}}
+  \caption{
+    \textbf{Speedup achieved by \sys for all pipelines with parallelism 20.}
+    \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}}
+  \vspace{-15pt}
+  \label{fig:unix50-individual}
+\end{figure}
 

### 2020-04-11 pldi/draft.tex
Update microbenchmarks table. It is now generated from a script

@@
 }
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllll}
-\toprule
-  Script                 ~&~ Structure                    & Input (MB)& Time (Seq)    & Script Size (LoC)             & Highlights                                        \\
-\midrule
-% https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
-  \tti{grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
-  \tti{sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ & 1000      &   2m 11.18s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-  \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ & 1000      &   6m 53.55s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
-  \tr{\tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\}
-% https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
-  \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ & 1000      &   8m 13.42s   &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
-% https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
-  \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
-  \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & extensive file-system operation   \\
-\bottomrule
-\end{tabular*}
+%% \begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllll}
+%% \toprule
+%%   Script                 ~&~ Structure                    & Input (MB)& Time (Seq)    & Script Size (LoC)             & Highlights                                        \\
+%% \midrule
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
+%%   \tti{grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
+%%   \tti{sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
+%%   \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ & 1000      &   2m 11.18s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
+%%   \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ & 1000      &   6m 53.55s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
+%%   \tr{\tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\}
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
+%%   \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ & 1000      &   8m 13.42s   &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
+%% % https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
+%%   \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
+%%   \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & extensive file-system operation   \\
+%% \bottomrule
+%% \end{tabular*}
+\input{microbenchmarks-table}
 \label{tab:eval}

### 2020-04-11 compiler/gather_results.py
Update microbenchmarks table. It is now generated from a script

@@ -19,6 +19,7 @@ plt.rcParams['mathtext.fontset'] = 'stix'
 plt.rcParams['font.family'] = 'STIXGeneral'
 
 
+MICROBENCHMARKS = "../evaluation/microbenchmarks/"
 RESULTS = "../evaluation/results/"
 UNIX50_RESULTS = "../evaluation/results/unix50/"
 
@@ -35,11 +36,31 @@ pretty_names = {"minimal_grep" : "grep",
                 "minimal_sort" : "sort",
                 "wf" : "wf",
                 "topn" : "top-n",
-                "grep" : "grep-old",
+                "grep" : "grep-light",
                 "bigram" : "bi-gram",
                 "spell" : "spell",
                 "shortest_scripts" : "shortest-scripts",
                 "diff" : "diff"}
+
+structures = {"minimal_grep" : "$3\\times$\\tsta",
+              "minimal_sort" : "$\\tsta, \\tpur$",
+              "wf" : "$3\\times\\tsta, 3\\times\\tpur$",
+              "topn" : "$2\\times\\tsta, 4\\times\\tpur$",
+              "grep" : "$3\\times$\\tsta",
+              "bigram" : "\\todo{TODO}",
+              "spell" : "$4\\times\\tsta, 3\\times\\tpur$",
+              "shortest_scripts" : "$5\\times\\tsta, 2\\times\\tpur$",
+              "diff" : "\\todo{TODO}"}
+
+highlights = {"minimal_grep" : "complex NFA regex",
+              "minimal_sort" : "\\tti{sort}ing",
+              "wf" : "double \\tti{sort}, \\tti{uniq} reduction",
+              "topn" : "double \\tti{sort}, \\tti{uniq} reduction",
+              "grep" : "$3\\times$\\tsta",
+              "bigram" : "stream shifting and merging",
+              "spell" : "comparisons (\\tti{comm})",
+              "shortest_scripts" : "\\todo{extensive file-system operation}",
+              "diff" : "non-parallelizable \\tti{diff}ing"}
 
 def get_experiment_files(experiment, results_dir):
     files = [f for f in os.listdir(results_dir) if f.startswith(experiment)]
@@ -145,6 +166,83 @@ def collect_scaleup_times(experiment, results_dir):
     plt.tight_layout()
     plt.savefig(os.path.join('../evaluation/plots', "{}_throughput_scaleup.pdf".format(experiment)))
 
+# \bottomrule
+# \end{tabular*}
+# \label{tab:eval}
+# \end{table*}
+
[…]

### 2020-04-12 evaluation/plots/microbenchmarks-table.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllll}
\toprule
Script ~&~ Structure & Input &Seq Time & Script Size(\todo{20, 100}) &Compile Time (\todo{20, 100}) & Highlights \\
\midrule
\tti{grep} ~&~ $3\times$\tsta & 1~GB & 95.45~s & \todo{\#Commands} & 63.84~ms\qquad 965.11~ms & complex NFA regex \\
\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 24.72~s & \todo{\#Commands} & 105.77~ms\qquad 1999.55~ms & \tti{sort}ing \\
\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 84.18~s & \todo{\#Commands} & 172.39~ms\qquad 3295.23~ms & double \tti{sort}, \tti{uniq} reduction \\
\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 25.75~s & \todo{\#Commands} & 172.83~ms\qquad 3294.47~ms & double \tti{sort}, \tti{uniq} reduction \\
\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 9.18~s & \todo{\#Commands} & 312.67~ms\qquad 6820.98~ms & comparisons (\tti{comm}) \\
\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 34.40~s & \todo{\#Commands} & 314.23~ms\qquad 6759.41~ms & \todo{extensive file-system operation} \\
\tti{diff} ~&~ \todo{TODO} & 10~GB & 30.69~s & \todo{\#Commands} & 225.73~ms\qquad 4336.29~ms & non-parallelizable \tti{diff}ing \\
\bottomrule
\end{tabular*}

### 2020-04-12 pldi/draft.tex
Add new figures and table in the paper

@@
     \caption{
-      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--200).} 
+      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
       Three different configurations per benchmark:
-      (i) distributed, focusing only on the distributed execution,
+      (i) parallel, focusing only on the parallel execution,
       (ii) +compile, adding \sys' pre-processing overheads, and
-      (iii) +merge, adding a final merger step.
+      (iii) \kk{Add with , without eager and split here}.
       \nv{Candidate values, based on the HW we have 2, 4, 8, 16, 32, 64. Also, maybe it makes sense to show log x-axis?}

### 2020-04-12 pldi/microbenchmarks-table.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllll}
\toprule
Script ~&~ Structure & Input &Seq Time & Script Size(\todo{20, 100}) &Compile Time (\todo{20, 100}) & Highlights \\
\midrule
\tti{grep} ~&~ $3\times$\tsta & 1~GB & 4809.61~s & \todo{\#Commands} & 63.37~ms\qquad 965.72~ms & complex NFA regex \\
\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 1429.04~s & \todo{\#Commands} & 201.12~ms\qquad 2005.70~ms & \tti{sort}ing \\
\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 4373.79~s & \todo{\#Commands} & 169.50~ms\qquad 3340.67~ms & double \tti{sort}, \tti{uniq} reduction \\
\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 1351.91~s & \todo{\#Commands} & 165.41~ms\qquad 3325.32~ms & double \tti{sort}, \tti{uniq} reduction \\
\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 481.13~s & \todo{\#Commands} & 311.45~ms\qquad 6655.26~ms & comparisons (\tti{comm}) \\
\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 1706.75~s & \todo{\#Commands} & 305.33~ms\qquad 6729.06~ms & \todo{extensive file-system operation} \\
\tti{diff} ~&~ \todo{TODO} & 10~GB & 1610.09~s & \todo{\#Commands} & 312.08~ms\qquad 4569.23~ms & non-parallelizable \tti{diff}ing \\
\bottomrule
\end{tabular*}

### 2020-04-12 pldi/draft.tex
A hypothesis that might explain our results

@@
 
+\kk{Hypothesis on results. Dish gives a great speedup if one step of
+  the pipeline is what takes the most time. If the pipeline is long
+  with many computation heavy tasks, task based parallelism gets you
+  part of the way there, so we can't expect Dish to give performance
+  benefits equal to the amount of parallelism that we give it.}
+
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.

### 2020-04-12 evaluation/plots/microbenchmarks-table.tex
Add optimized bi-grams plot

@@
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 95.45~s & \todo{\#Commands} & 63.84~ms\qquad 965.11~ms & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 24.72~s & \todo{\#Commands} & 105.77~ms\qquad 1999.55~ms & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 84.18~s & \todo{\#Commands} & 172.39~ms\qquad 3295.23~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 25.75~s & \todo{\#Commands} & 172.83~ms\qquad 3294.47~ms & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 4809.61~s & \todo{\#Commands} & 63.37~ms\qquad 965.72~ms & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 1429.04~s & \todo{\#Commands} & 201.12~ms\qquad 2005.70~ms & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 4373.79~s & \todo{\#Commands} & 169.50~ms\qquad 3340.67~ms & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 1351.91~s & \todo{\#Commands} & 165.41~ms\qquad 3325.32~ms & double \tti{sort}, \tti{uniq} reduction \\
 \tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 9.18~s & \todo{\#Commands} & 312.67~ms\qquad 6820.98~ms & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 34.40~s & \todo{\#Commands} & 314.23~ms\qquad 6759.41~ms & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 30.69~s & \todo{\#Commands} & 225.73~ms\qquad 4336.29~ms & non-parallelizable \tti{diff}ing \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 481.13~s & \todo{\#Commands} & 311.45~ms\qquad 6655.26~ms & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 1706.75~s & \todo{\#Commands} & 305.33~ms\qquad 6729.06~ms & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ \todo{TODO} & 10~GB & 1610.09~s & \todo{\#Commands} & 312.08~ms\qquad 4569.23~ms & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 7434.95~s & \todo{\#Commands} & 135.88~ms\qquad 2674.48~ms & optimized version of bigrams \\
 \bottomrule

### 2020-04-12 compiler/gather_results.py

@@ -30,14 +30,16 @@ experiments = ["minimal_grep",
                "grep",
                "spell",
                "shortest_scripts",
-               "diff"]
+               "diff",
+               "alt_bigrams"]
 
 pretty_names = {"minimal_grep" : "grep",
                 "minimal_sort" : "sort",
                 "wf" : "wf",
                 "topn" : "top-n",
                 "grep" : "grep-light",
-                "bigram" : "bi-gram",
+                "bigram" : "bi-grams",
+                "alt_bigrams" : "optimized bi-grams",
                 "spell" : "spell",
                 "shortest_scripts" : "shortest-scripts",
                 "diff" : "diff"}
@@ -48,6 +50,7 @@ structures = {"minimal_grep" : "$3\\times$\\tsta",
               "topn" : "$2\\times\\tsta, 4\\times\\tpur$",
               "grep" : "$3\\times$\\tsta",
               "bigram" : "\\todo{TODO}",
+              "alt_bigrams" : "$3\\times$\\tsta, \\tpur",
               "spell" : "$4\\times\\tsta, 3\\times\\tpur$",
               "shortest_scripts" : "$5\\times\\tsta, 2\\times\\tpur$",
               "diff" : "\\todo{TODO}"}
@@ -58,6 +61,7 @@ highlights = {"minimal_grep" : "complex NFA regex",
               "topn" : "double \\tti{sort}, \\tti{uniq} reduction",
               "grep" : "$3\\times$\\tsta",
               "bigram" : "stream shifting and merging",
+              "alt_bigrams" : "optimized version of bigrams",
               "spell" : "comparisons (\\tti{comm})",
               "shortest_scripts" : "\\todo{extensive file-system operation}",
               "diff" : "non-parallelizable \\tti{diff}ing"}

### 2020-04-12 compiler/gather_results.py
Update the formating of the microbenchmarks table

@@ -202,6 +202,26 @@ def collect_input_size(experiment):
     clean_name = input_file_name.split('/')[-1].split('.')[0]
     return clean_name
 
+def format_time_seconds(time_milliseconds):
+    time_seconds = time_milliseconds / 1000
+    time_minutes = int(time_seconds // 60)
+    time_only_seconds = time_seconds % 60
+    if(time_minutes > 0):
+        formatted_time = '{}m{:.3f}s'.format(time_minutes, time_only_seconds)
+    else:
+        formatted_time = '{:.3f}s'.format(time_seconds)
+    return formatted_time
+
+def format_time_milliseconds(time_milliseconds):
+    time_seconds = int(time_milliseconds // 1000)
+    time_only_milliseconds = time_milliseconds % 1000
+    if(time_seconds > 0):
+        formatted_time = '{}s{:.3f}ms'.format(time_seconds, time_only_milliseconds)
+    else:
+        formatted_time = '{:.3f}ms'.format(time_milliseconds)
+    return formatted_time
+
+
 def generate_table_header():
     header = []
     header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} lllllll}']
@@ -232,15 +252,16 @@ def generate_experiment_line(experiment):
     experiment_results_prefix = '{}/{}_'.format(RESULTS, experiment)
     seq_times, _, compile_times = collect_experiment_scaleup_times(experiment_results_prefix, scaleup_numbers)
     assert(len(seq_times) == 3)
-    seq_time_seconds = seq_times[0] / 1000
-    line += ['{:.2f}~s'.format(seq_time_seconds), '&']
+    seq_time_seconds = format_time_seconds(seq_times[0])
+    # seq_time_seconds = seq_times[0] / 1000
+    line += [seq_time_seconds, '&']
     line += ['\\todo{\\#Commands}', '&']
 
     ## Collect and output compile times
     compile_time_20_milliseconds = compile_times[1]
     compile_time_100_milliseconds = compile_times[2]
-    line += ['{:.2f}~ms\\qquad {:.2f}~ms'.format(compile_time_20_milliseconds,
-                                                 compile_time_100_milliseconds), '&']
+    line += ['{}\\qquad {}'.format(format_time_seconds(compile_time_20_milliseconds),
+                                   format_time_seconds(compile_time_100_milliseconds)), '&']
     line += [highlights[experiment], '\\\\']
     return " ".join(line)
 

### 2020-04-12 pldi/microbenchmarks-table.tex
Update the formating of the microbenchmarks table

@@
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 4809.61~s & \todo{\#Commands} & 63.37~ms\qquad 965.72~ms & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 1429.04~s & \todo{\#Commands} & 201.12~ms\qquad 2005.70~ms & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 4373.79~s & \todo{\#Commands} & 169.50~ms\qquad 3340.67~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 1351.91~s & \todo{\#Commands} & 165.41~ms\qquad 3325.32~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 481.13~s & \todo{\#Commands} & 311.45~ms\qquad 6655.26~ms & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 1706.75~s & \todo{\#Commands} & 305.33~ms\qquad 6729.06~ms & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 1610.09~s & \todo{\#Commands} & 312.08~ms\qquad 4569.23~ms & non-parallelizable \tti{diff}ing \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 80m9.607s & \todo{\#Commands} & 0.063s\qquad 0.966s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 23m49.044s & \todo{\#Commands} & 0.201s\qquad 2.006s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 72m53.786s & \todo{\#Commands} & 0.170s\qquad 3.341s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m31.907s & \todo{\#Commands} & 0.165s\qquad 3.325s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.035s\qquad 0.356s & $3\times$\tsta \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.131s & \todo{\#Commands} & 0.311s\qquad 6.655s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m26.746s & \todo{\#Commands} & 0.305s\qquad 6.729s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ \todo{TODO} & 10~GB & 26m50.085s & \todo{\#Commands} & 0.312s\qquad 4.569s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 123m54.946s & \todo{\#Commands} & 0.136s\qquad 2.674s & optimized version of bigrams \\
 \bottomrule

### 2020-04-13 compiler/execute_evaluation_scripts.sh
change microbenchmark experiments to run for 2-128

@@ -11,10 +11,16 @@ n_inputs=(
     # 1
     2
     4
-    10
-    20
-    50
-    100
+    8
+    16
+    32
+    64
+    96
+    128
+    # 10
+    # 20
+    # 50
+    # 100
 )
 
 microbenchmarks=(

### 2020-04-13 evaluation/plots/microbenchmarks-table.tex
New plots and table

--- 2020-04-12 evaluation/plots/microbenchmarks-table.tex
@@
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(\todo{20, 100}) &Compile Time (\todo{20, 100}) & Highlights \\
+Script ~&~ Structure & Input &Seq Time & Script Size(\todo{16, 128}) &Compile Time (\todo{16, 128}) & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 95.45~s & \todo{\#Commands} & 63.84~ms\qquad 965.11~ms & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 24.72~s & \todo{\#Commands} & 105.77~ms\qquad 1999.55~ms & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 84.18~s & \todo{\#Commands} & 172.39~ms\qquad 3295.23~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 25.75~s & \todo{\#Commands} & 172.83~ms\qquad 3294.47~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 9.18~s & \todo{\#Commands} & 312.67~ms\qquad 6820.98~ms & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 34.40~s & \todo{\#Commands} & 314.23~ms\qquad 6759.41~ms & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 30.69~s & \todo{\#Commands} & 225.73~ms\qquad 4336.29~ms & non-parallelizable \tti{diff}ing \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 80m9.607s & \todo{\#Commands} & 0.051s\qquad 1.551s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 23m49.044s & \todo{\#Commands} & 0.129s\qquad 3.362s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 72m53.786s & \todo{\#Commands} & 0.121s\qquad 5.346s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m31.907s & \todo{\#Commands} & 0.121s\qquad 5.378s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & $3\times$\tsta \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.131s & \todo{\#Commands} & 0.211s\qquad 10.907s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m26.746s & \todo{\#Commands} & 0.211s\qquad 10.755s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ \todo{TODO} & 10~GB & 26m50.085s & \todo{\#Commands} & 0.204s\qquad 7.222s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 123m54.946s & \todo{\#Commands} & 0.098s\qquad 4.410s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m34.784s & \todo{\#Commands} & 0.249s\qquad 12.590s & two pipelines merging to a \tti{comm} \\
 \bottomrule

### 2020-04-13 compiler/gather_results.py
New plots and table

@@ -154,9 +154,10 @@ def collect_experiment_speedups(prefix, scaleup_numbers):
 def collect_scaleup_times(experiment, results_dir):
     print(experiment)
 
-    all_scaleup_numbers = list(set(get_experiment_files(experiment, results_dir)))
-    all_scaleup_numbers.sort()
-    all_scaleup_numbers = [i for i in all_scaleup_numbers if i > 1]
+    # all_scaleup_numbers = list(set(get_experiment_files(experiment, results_dir)))
+    # all_scaleup_numbers.sort()
+    # all_scaleup_numbers = [i for i in all_scaleup_numbers if i > 1]
+    all_scaleup_numbers = [2, 4, 8, 16, 32, 64, 96, 128]
     prefix = '{}/{}_'.format(results_dir, experiment)
     distr_speedup, compile_distr_speedup = collect_experiment_speedups(prefix, all_scaleup_numbers)
 
@@ -231,8 +232,8 @@ def generate_table_header():
     header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} lllllll}']
     header += ['\\toprule']
     header += ['Script ~&~ Structure & Input &'
-               'Seq Time & Script Size(\\todo{20, 100}) &'
-               'Compile Time (\\todo{20, 100}) & Highlights \\\\']
+               'Seq Time & Script Size(\\todo{16, 128}) &'
+               'Compile Time (\\todo{16, 128}) & Highlights \\\\']
     header += ['\\midrule']
     return "\n".join(header)
 
@@ -252,7 +253,7 @@ def generate_experiment_line(experiment):
     line += [input_size, '&']
 
     ## Collect and output the sequential time for the experiment
-    scaleup_numbers = [2, 20, 100]
+    scaleup_numbers = [2, 16, 128]
     experiment_results_prefix = '{}/{}_'.format(RESULTS, experiment)
     seq_times, _, compile_times = collect_experiment_scaleup_times(experiment_results_prefix, scaleup_numbers)
     assert(len(seq_times) == 3)
@@ -262,10 +263,10 @@ def generate_experiment_line(experiment):
     line += ['\\todo{\\#Commands}', '&']
 
     ## Collect and output compile times
-    compile_time_20_milliseconds = compile_times[1]
-    compile_time_100_milliseconds = compile_times[2]
-    line += ['{}\\qquad {}'.format(format_time_seconds(compile_time_20_milliseconds),
[…]

### 2020-04-14 pldi/draft.tex
Add a comment in the PLDI draft

@@
   light) tasks it has some overhead. We evaluate eager in section X
-  (We should also say something about a CPU light task).}
+  (We should also say something about a CPU light task). A good thing
+  is that they work for all possible commands with almost no overhead,
+  without having to make each specific command eager.}
 
@@
     %% \caption{a}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.24\textwidth}
         \centering
@@
     ~
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.24\textwidth}
         \centering
@@
     \end{subfigure}%
-
-    \begin{subfigure}[b]{0.32\textwidth}
+    ~
+    \begin{subfigure}[b]{0.24\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{\detokenize{./figs/set-diff_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:minimal_sort}
+    \end{subfigure}%
+
+    \begin{subfigure}[b]{0.24\textwidth}
         \centering
@@
         %% \label{eval:spell}
+    \end{subfigure}%
+    ~
+    \begin{subfigure}[b]{0.24\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{\detokenize{./figs/alt_bigrams_throughput_scaleup.pdf}}
+        %% \caption{TODO}
+        %% \label{eval:minimal_sort}
     \end{subfigure}%

### 2020-04-16 pldi/draft.tex
Some notes on paper draft

@@
 
+\kk{Maximum scaleup for both diff and set-diff is around half because
+  they already have two sorts running in parallel (as diff, and comm
+  take two inputs).}
+
+\TODO{Explain the plots with eager/no-eager as follows. Say that we
+  initially decided to only make an eager version of cat (to allow for
+  parallelism). However, we saw that results are still not good, so we
+  resorted to making a proper generic eager node that can be put
+  anywhere in the graph and that eagerly consumes its input. The above
+  results show the improvement with it. Rerun them with eager only
+  before cat in the no eager cases to showcase the above
+  results. (Maybe) we need to mention that eager has disk usage
+  overhead that is linear to the size of the pipeline and the size of
+  the input.}
+
 At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.

### 2020-04-16 pldi/figs/microbenchmarks-table.tex
New figures with eager and no-eager results

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllll}
\toprule
Script ~&~ Structure & Input &Seq Time & Script Size(\todo{16, 128}) &Compile Time (\todo{16, 128}) & Highlights \\
\midrule
\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m49.082s & \todo{\#Commands} & 0.049s\qquad 1.542s & complex NFA regex \\
\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 23m25.333s & \todo{\#Commands} & 0.076s\qquad 3.213s & \tti{sort}ing \\
\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 74m39.671s & \todo{\#Commands} & 0.122s\qquad 5.390s & double \tti{sort}, \tti{uniq} reduction \\
\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.801s & \todo{\#Commands} & 0.120s\qquad 5.311s & double \tti{sort}, \tti{uniq} reduction \\
\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & $3\times$\tsta \\
\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m18.837s & \todo{\#Commands} & 0.211s\qquad 10.653s & comparisons (\tti{comm}) \\
\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m22.727s & \todo{\#Commands} & 0.210s\qquad 10.575s & \todo{extensive file-system operation} \\
\tti{diff} ~&~ \todo{TODO} & 10~GB & 27m10.031s & \todo{\#Commands} & 0.193s\qquad 7.238s & non-parallelizable \tti{diff}ing \\
\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 123m50.883s & \todo{\#Commands} & 0.095s\qquad 4.524s & optimized version of bigrams \\
\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m19.075s & \todo{\#Commands} & 0.248s\qquad 12.700s & two pipelines merging to a \tti{comm} \\
\bottomrule
\end{tabular*}

### 2020-04-18 evaluation/plots/microbenchmarks-table.tex
New great plots

@@
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 80m9.607s & \todo{\#Commands} & 0.051s\qquad 1.551s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 23m49.044s & \todo{\#Commands} & 0.129s\qquad 3.362s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 72m53.786s & \todo{\#Commands} & 0.121s\qquad 5.346s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m31.907s & \todo{\#Commands} & 0.121s\qquad 5.378s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.049s\qquad 1.542s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 1~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 3.213s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 73m28.098s & \todo{\#Commands} & 0.122s\qquad 5.390s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.120s\qquad 5.311s & double \tti{sort}, \tti{uniq} reduction \\
 \tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.131s & \todo{\#Commands} & 0.211s\qquad 10.907s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m26.746s & \todo{\#Commands} & 0.211s\qquad 10.755s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 26m50.085s & \todo{\#Commands} & 0.204s\qquad 7.222s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 123m54.946s & \todo{\#Commands} & 0.098s\qquad 4.410s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m34.784s & \todo{\#Commands} & 0.249s\qquad 12.590s & two pipelines merging to a \tti{comm} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.211s\qquad 10.653s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 10.575s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ \todo{TODO} & 10~GB & 25m10.092s & \todo{\#Commands} & 0.193s\qquad 7.238s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{\#Commands} & 0.095s\qquad 4.524s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.248s\qquad 12.700s & two pipelines merging to a \tti{comm} \\
 \bottomrule

### 2020-04-18 compiler/gather_results.py
New great plots

@@ -156,9 +156,13 @@ def collect_experiment_scaleup_times(prefix, scaleup_numbers):
 def collect_experiment_no_eager_times(prefix, scaleup_numbers):
     no_eager_distr_numbers = [read_distr_execution_time('{}{}_distr_no_eager.time'.format(prefix, n))
                               for n in scaleup_numbers]
-    no_eager_compile_numbers = [read_distr_total_compilation_time('{}{}_distr_no_eager.time'.format(prefix, n))
-                                for n in scaleup_numbers]
-    return (no_eager_distr_numbers, no_eager_compile_numbers)
+    # no_eager_compile_numbers = [read_distr_total_compilation_time('{}{}_distr_no_eager.time'.format(prefix, n))
+    #                             for n in scaleup_numbers]
+    no_task_par_eager_distr_numbers = [read_distr_execution_time('{}{}_distr_no_task_par_eager.time'.format(prefix, n))
+                                       for n in scaleup_numbers]
+    # no_eager_compile_numbers = [read_distr_total_compilation_time('{}{}_distr_no_eager.time'.format(prefix, n))
+    #                             for n in scaleup_numbers]
+    return (no_eager_distr_numbers, no_task_par_eager_distr_numbers)
 
 def collect_experiment_speedups(prefix, scaleup_numbers):
     seq_numbers, distr_numbers, compile_numbers = collect_experiment_scaleup_times(prefix, scaleup_numbers)
@@ -172,16 +176,23 @@ def collect_experiment_speedups(prefix, scaleup_numbers):
 
 def collect_experiment_no_eager_speedups(prefix, scaleup_numbers):
     seq_numbers, _, _ = collect_experiment_scaleup_times(prefix, scaleup_numbers)
-    no_eager_distr_numbers, no_eager_compile_numbers = collect_experiment_no_eager_times(prefix, scaleup_numbers)
-    no_eager_distr_speedup = [seq_numbers[i] / t for i, t in enumerate(no_eager_distr_numbers)]
-    no_eager_compile_distr_speedup = [seq_numbers[i] / (t + no_eager_compile_numbers[i])
-                                      for i, t in enumerate(no_eager_distr_numbers)]
-    return (no_eager_distr_speedup, no_eager_compile_distr_speedup)
+    no_eager_distr_numbers, no_task_par_eager_distr_numbers = collect_experiment_no_eager_times(prefix, scaleup_numbers)
[…]

### 2020-04-24 pldi/draft.tex
New figures with eager and no-eager results
Some notes on the paper draft

@@
 
-\tr{Remember to mention the assumptions that need to hold for the
+\kk{Remember to mention the assumptions that need to hold for the
   graph transformations to be valid in the Command categories
@@
   etc). However, these assumptions must already be checked when the
-  developer designates the categories.}
+  developer designates the categories. Commands are assumed to always
+  terminate gracefully. \sys changes the termination behaviour of
+  commands as a sequential command that would exit prematurely might
+  not exit completely if parallelized.}
 
@@
   results show the improvement with it. Rerun them with eager only
-  before cat in the no eager cases to showcase the above
-  results. (Maybe) we need to mention that eager has disk usage
-  overhead that is linear to the size of the pipeline and the size of
-  the input.}
+  before cat in the no eager cases so that the minimal-grep example is
+  the same with and without our eager node. (Maybe) we need to mention
+  that eager has disk usage overhead that is linear to the size of the
+  pipeline and the size of the input.}
 

### 2020-04-30 pldi/draft.tex
Updates (almost done) subsection about Unix50

@@
 \label{intro}
+
+\TODO{Drop the automated from the title maybe?}
 
@@
 \todo{Ref} and we executed them \emph{without any modification} with
-\sys's parallelism set to \todo{20}. The input given to the pipelines
-was the test input from unix50 multiplied \todo{10000} times. The
+\sys's parallelism set to \todo{16}. The input given to the pipelines
+was the test input from unix50 multiplied to reach \todo{10GBs}. The
 speedup of each pipeline separately is shown in
@@
 
-%% Bad result descriptions:
+%% (OLD) Bad result descriptions:
 %%  2. Contains just a head -n 2, and then a cut.
@@
   \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_20.pdf}}
-  \caption{
-    \textbf{Speedup achieved by \sys for all pipelines with parallelism 20.}
-    \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}}
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \caption{ \textbf{Speedup achieved by \sys for all pipelines with
+      parallelism 16. The red bars are the sequential execution time
+      of each pipeline.}
+    %% \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}
+  }
   \vspace{-15pt}
@@
 
-All pipelines that perform poorly (\todo{enumerate them}) contain
-non-parallelizable commands. An illustrative example is pipeline
-\todo{X}, which is shown here:
+\TODO{An important point that needs to be made clear is that these are
+  taken from the wild unmodified. And that Pash never leads to
+  slowdown (except for the trivial head cases, where the slowdown is
+  neligible).}
+
+\sys achieves speedup in all pipelines except for 2, 19, and 31. These
+pipelines include a \ttt{head} and only process one line of the
+input. This can be seen by their sequential execution time which is in
+the order of 10 milliseconds. The slowdown is due to constant costs of
+setting up the pipes and spawning many processes. In all three of
+these cases the execution time with \sys is in the order of \todo{X
+  ms}. For pipelines 13, 24, 25, 26, 29, and 30, \sys performs the
+same as the sequential implementation. This is because these pipelines
+contain non-parallelizable commands, such as \ttt{awk}, \ttt{sed} with
+option \ttt{d}, and \ttt{tr} with option \ttt{-d}. An illustrative
+example is pipeline 13, which is shown here:
 
 \begin{lstlisting}[language=sh, float=h, numbers=none]
-  TODO
+  cat $IN | awk "{print \$2, \$0}" | sort -nr | cut -d ' ' -f 2
 \end{lstlisting}
 
-Notice in the above the use of (awk, fmt, or something else). We
-noticed that these occured in other pipelines too. By changing them to
-\todo{X}, the average speedup of these pipelines improved to \todo{x}
-and the new total average speedup is \todo{X}.
+\kk{Rephrase that to sound more meaningful and surprising.}  Notice
+that \ttt{awk} is just used to permute two of the columns of the input
+to enable sorting. One could write this pipeline in a different way,
+by directly sorting on the second column:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none]
+  cat $IN6 | sort -nr -k 2 | cut -d ' ' -f 1
+\end{lstlisting}
+
+By executing this version of the pipeline with \sys, we achieve
+\todo{X} speedup. This indicates that just a little awaresess of
+command parallelizability can go a long way, and that the effort that
+a \sys user needs to spend is often negligible to achieve
+benefits. Finally, an important point that is reinforced from this
+experiment is that \sys doesn't lead to slowdown, even in cases where
+parallelism is not possible.
 

### 2020-05-01 compiler/gather_results.py
New unix50 plots and breakdown of all pipelines in paper draft

@@ -365,22 +365,8 @@ def aggregate_unix50_results(all_results, scaleup_numbers):
 
     return avg_distr_results
 
-def collect_unix50_scaleup_times(unix50_results_dir):
-    files = [f for f in os.listdir(unix50_results_dir)]
-    # print(files)
-    pipeline_numbers = sorted(list(set([f.split('_')[2] for f in files])))
-    # print(pipeline_numbers)
-
-    scaleup_numbers = [2, 16, 32]
-
-    all_results = [collect_unix50_pipeline_scaleup_times(pipeline_number,
-                                                         unix50_results_dir,
-                                                         scaleup_numbers)
-                   for pipeline_number in pipeline_numbers]
-    # print(all_results)
-
+def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
     ## Plot individual speedups
-    parallelism = 16
     individual_results = [distr_exec_speedup[scaleup_numbers.index(parallelism)]
                           for distr_exec_speedup, _ in all_results]
     absolute_seq_times_s = [absolute_seq[scaleup_numbers.index(parallelism)] / 1000
@@ -413,6 +399,24 @@ def collect_unix50_scaleup_times(unix50_results_dir):
     plt.title("Unix50 Individual Speedups")
     plt.tight_layout()
     plt.savefig(os.path.join('../evaluation/plots', "unix50_individual_speedups_{}.pdf".format(parallelism)))
+
+def collect_unix50_scaleup_times(unix50_results_dir):
+    files = [f for f in os.listdir(unix50_results_dir)]
+    # print(files)
+    pipeline_numbers = sorted(list(set([f.split('_')[2] for f in files])))
+    # print(pipeline_numbers)
+
+    scaleup_numbers = [2, 4, 8, 16]
+
+    all_results = [collect_unix50_pipeline_scaleup_times(pipeline_number,
+                                                         unix50_results_dir,
+                                                         scaleup_numbers)
+                   for pipeline_number in pipeline_numbers]
+    # print(all_results)
+
+    for parallelism in scaleup_numbers:
+        make_unix50_bar_chart(all_results, scaleup_numbers, parallelism)
+
 
     avg_results = aggregate_unix50_results(all_results, scaleup_numbers)
     print("Unix50 average speedup:", avg_results)

### 2020-05-01 pldi/draft.tex
New unix50 plots and breakdown of all pipelines in paper draft
Discussion of deadlock

--- 2020-04-24 pldi/draft.tex
@@
 
-\begin{lstlisting}[language=sh, float=h, numbers=none]
+\begin{lstlisting}[language=sh, numbers=none]
  cat $in1 $in2 ... > $out
@@
 
-\begin{lstlisting}[language=sh, float=h, numbers=none]
+\begin{lstlisting}[language=sh, numbers=none]
   tee >(head -n $N > $temp1;
@@
 arbitrarily many different pipes.
+
+\heading{Dangling FIFOs and Zombie Producers}
+% \nv{I think all of these titles should be the problem/challenge, not the solution}
+% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
+Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
+% \ttt{SIGPIPE} 
+However, when its consumer exits early, the producer needs to be notified so that it does not keep appending to the stream.
+% ---so that it avoids unnecessary work and buffering challenges in the kernel.
+
+In \unix, write errors are communicated differently from errors for other system calls.
+Whereas other system calls return an error, write errors are communicated through an out-of-band mechanism (signal handler).
+\unix is unusual in this regard from other systems,\footnote{
+  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
+}
+because pipes and pipelines are at the heart of its philosophy.
+Unfortunately, if the pipe hasn't been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
+Consider the following case:
+\begin{lstlisting}[language=sh, numbers=none]
+  cat file-chunk1.txt > fifo1 &
+  cat file-chunk2.txt > fifo2 &
+  cat fifo1 fifo2 | head -n 1 & wait
+\end{lstlisting}
+\noindent
+In the example above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
+As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer---after all, \ttt{fifo2} never even had a consumer.
+This, in turn, leaves to the second \ttt{cat} unable make process, as it cannot write (and thus process) its output but also unwire that its consumer has exited.
+Coupled with \ttt{wait} at the end, it causes the entire pipeline to deadlock.
+
+To solve this problem, \sys emits cleanup logic that operates from the end of the pipeline and towards its start.
+The emitted code first gathers the ID of the last process and passes it as a parameter to \ttt{wait}; 
+  this causes \ttt{wait} to block only on the last consumer of the pipeline.
+Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
+% need to define split and merge points; this should be "up to next merge point upstream"
+
+% not receiving an out-of-band signal about that.
+% never terminating, but 
+% never terminate while the entire pipeline is waiting for 
+% \ttt{\& echo \$!} and \ttt{kill -PIPE}
+
+% The generated program could also (i) have signal handlers (traps) for propagating signals
+% down the process tree; (ii) incorporate (a simple) cleanup logic for not leaving dangling processes and
+% intermediary files; (iii) generate a correct exit code, depending on its
+% execution. All of this can be easily generated by the compiler.
 

### 2020-05-01 pldi/draft.tex
Update deadlock

@@
 % \ttt{SIGPIPE} 
-However, when its consumer exits early, the producer needs to be notified so that it does not keep appending to the stream.
-% ---so that it avoids unnecessary work and buffering challenges in the kernel.
-
-In \unix, write errors are communicated differently from errors for other system calls.
-Whereas other system calls return an error, write errors are communicated through an out-of-band mechanism (signal handler).
-\unix is unusual in this regard from other systems,\footnote{
+However, if its consumer exits early, the producer needs to be notified so that it stops appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.
+
+In \unix, this is achieved by an out-of-band error mechanism:
+  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
+This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
   For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
 }
-because pipes and pipelines are at the heart of its philosophy.
-Unfortunately, if the pipe hasn't been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
-Consider the following case:
+primarily because pipes and pipelines are at the heart of its philosophy.
+% write errors are communicated differently 
+Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
+Consider the following emitted code from \sys's earlier versions:
 \begin{lstlisting}[language=sh, numbers=none]
@@
 \noindent
-In the example above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
-As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer---after all, \ttt{fifo2} never even had a consumer.
-This, in turn, leaves to the second \ttt{cat} unable make process, as it cannot write (and thus process) its output but also unwire that its consumer has exited.
-Coupled with \ttt{wait} at the end, it causes the entire pipeline to deadlock.
+In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
+As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
+This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
+Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
 
@@
 Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
-% need to define split and merge points; this should be "up to next merge point upstream"
+This epilogue is repeated for every \todo{merge/join} point in the transformed pipeline.
+% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"
 

### 2020-05-01 pldi/draft.tex

@@
 
-%% The current average speedup includes unimplemented and small input pipelines.
-
-%% Bad result descriptions:
+%% Breakdown of not ideal speedups in 8, 16 parallelism
+
+%% Head: 2, 19, 31
+%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
+%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
+%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14
+%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
+
+\kk{Out of all benchmarks, we have talked about the head and the
+  non-parallelizable ones in the paper. Maybe we also want to talk
+  about the long and the cheap ones. Also we might want to talk and
+  find out more about the sort bottleneck.}
+
+%% (OLD) Bad result descriptions:
 %%  2. Contains just a head -n 2, and then a cut.
@@
 
-%% WARNING: Remember to fix `sed` and not have it be stateless in general.
-
 \begin{figure}[t]
   \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_20.pdf}}
-  \caption{
-    \textbf{Speedup achieved by \sys for all pipelines with parallelism 20.}
-    \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}}
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \caption{ \textbf{Speedup achieved by \sys for all pipelines with
+      parallelism 16. The red bars are the sequential execution time
+      of each pipeline.}
+    %% \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}
+  }
   \vspace{-15pt}

### 2020-05-01 pldi/draft.tex
Update evaluation

@@
 Experiments were run on a %\todo{network of five workstations} connected by 1Gbps links: 
- machine with 512GB of memory and 128 $\times$ 2.1GHz Intel Xeon E5-2683 cores.
+  machine with 512GB of memory and 128 virtual (64 physical) $\times$ 2.1GHz Intel Xeon E5-2683 cores.
 %\todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
 For our software setup, we use Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
-Generally, no special configuration was made in hardware or software beyond disabling hyper-threading. % ---specifically, the network protocol stack was left unoptimized.
+Generally, no special configuration was made in hardware or software.
 All pipelines are set to (initially) read from and (finally) write to the file-system (except as otherwise stated).
 Whenever \ttt{curl} is part of a pipeline, it fetches data from a different physical host on the same network connected by 1Gbps links.
-More detailed setup and experiments, including the scripts, is included in the accompanying (anonymized) online repository.
-
-\subsection{Microbenchmarks: Shell One-liners}
+A more detailed description of the setup and experiments, including reproducible scripts, is included in the accompanying (anonymized) online repository.
+
+\subsection{Shell One-liners}
 \label{micro}
@@
 
-Interestingly, the overhead from \sys's analysis and generation phases remains collectively below 1s even for the high-scalability cases.
-We decided to push \sys's limits of these phases by synthesizing an artificial 1000-stage pipeline---a carefully-constructed, worst-case workload that is nowhere near the ones observed in practice.
-% (\S\ref{macro1}--\ref{macro3}).
-We find the resulting times quite acceptable: given two input arguments in the beginning of this pipeline, \sys takes 8.23s for parsing the script and 3.22s for analysis and optimization.
-Pushing the input file arguments to 100, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
-
-\subsection{Unix50}
+\subsection{Unix50 from Bell Labs}
 \label{unix50}
 
-The goal of this subsection is to evaluate \sys on an existing set of
-unix pipelines found in the wild. \todo{Motivate why unix50 is a nice
+\heading{Workload}
+The goal of this subsection is to evaluate \sys on an existing set of \unix pipelines found in the wild.
+In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of challenges
+
+solving challenges by
+
+ability to compose larger programs in versatile ways from smaller utility programs
+summarized by McIllroy's Unix philosophy memo.
+
+The ability to compose larger programs in versatile ways from smaller utility programs was a hallmark of the \unix design, and the game's levels are designed to stress that.
+
+This section reports on programs that were run unaltered---highlighting 
+These are pipelines that were run verbatim
+In cases where an obvious fix improves the pipeline's performance, we report on both \sys's original speedup and the one achieved by applying the fix.
+
+\todo{Motivate why unix50 is a nice
   set of pipelines.} We took the solutions for the unix50 from
 \todo{Ref} and we executed them \emph{without any modification} with
+
+\heading{Setup}
 \sys's parallelism set to \todo{16}. The input given to the pipelines
@@
 
-\subsection{Comparison}
-
-\todo{Include comparison with Niko's speeduptest, with gnu-parallel,
-  with sort -parallel (and possibly other commands that have parallel
-  flags), and finally with Raftlib.}
-
-\subsection{Macro-benchmark: Weather Analysis}
+\subsection{Use Case: NOAA Weather Analysis}
 \label{macro1}
@@
 % 
+\subsection{Use Case: Wikipedia Web Indexing}
+\label{macro1}
+
+\subsection{Use Case: Docker Container Orchestration}
+\label{macro1}
+
+\subsection{Further Micro-benchmarks}
+
+The previous several subsections show substantial speedups without any developer effort on a variety of smaller and larger shell programs.
+In this section, we are interested in zooming into \sys's limits and how they compare with perceived upper bounds.
+These upper bounds are calculated through a variety of means which, while requiring vastly more developer effort than \sys, they are expected to yield significantly better performance.
+
+\heading{Parallelization Command}
+We compare \sys to \ttt{parallel}~\cite{Tange2011a}, a GNU utility for running other commands in parallel.
+In the first experiment, we leverage \ttt{parallel}'s \ttt{--jobs} flag and redirect input from a file to 
+Sequential execution takes \todo{XX}s, and \sys's parallel execution takes \todo{XX}s (speedup: \todo{XX}\%).
+We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command, masking \sys's improvements in other parts of the pipeline.
+
+% % https://github.com/marcelm/cutadapt/issues/157
+% \begin{lstlisting}[language=sh, numbers=none]
+% 	uniq | sort | parallel cutadapt-fastqc
+% \end{lstlisting}
+% % -S server01,server02 
+% % -u -j24 --env PATH,PYTHONPATH,cutadapt_parallel
+% % --workdir $PWD cutadapt_parallel {}
+
+There are a few possible ways one might attempt to use GNU \ttt{parallel} to parallelize this program, which compared to \sys lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
+Assuming a user knows that the bottleneck comes from the \ttt{XX} stage, they could attempt to use \ttt{parallel} there leading to a runtime of \todo{XX}s (speedup: \todo{XX}\%)---worse than surprisingly worse than.
+
+Alternatively, a user could (incorrectly) sprinkle \ttt{parallel} across the entire program.
+This is a common strategy as GNU \ttt{parallel} allows feeding an entire script for parallelization via \ttt{stdin}. 
+Doing this leads to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
+\sys's conservative program transformations will not attempt parallelizing program sections that contain commands with unclear \todo{parallelizability} properties.
+	% or commands whose inputs depend on earlier outputs.
+GNU \ttt{parallel} is oblivious to command semantics, risking breaking program semantics as in this case.
+
+\heading{Manual Parallelization}
+In an attempt to highlight how a user might attempt to manually parallelize individual commands using the abstractions provided by the POSIX shell.
+The benefits depend on several factors, such as whether a command is I/O- or CPU-bound or whether but since
+
+Merging is provided as Python programs
+
+\heading{Streaming Framework}
+In a different attempt to identify how far \sys is from ideal performance, we start from programs already written in a high-performance parallel framework and rewrite as shell programs.
+Raftlib~\cite{} is a C++ library enabling streaming and dataflow parallelism without the fault-tolerance overheads of distributed systems.
+We rewrite three Raftlib benchmarks---\ttt{wordcount}, \ttt{XX}, and \ttt{XX}---as shell scripts.
+% These benchmarks were relatively easy found relatively easy to port and match many of the workloads one would expect to see in the shell as shell pipelines.
+
+\heading{Graph Transformations}
+In the benchmarks, the overhead from \sys's analysis and generation phases remains collectively below $1s$ even for the high-scalability cases.
+We decided to push \sys's limits by constructing an artificial 1000-stage pipeline---a worst-case workload that is nowhere near the ones observed in practice.
+We find the resulting times quite acceptable:
+  given two input arguments in the beginning of this pipeline, \sys takes $8.23s$ for parsing the script and $3.22s$ for analysis and optimization.
+Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
 

### 2020-05-02 pldi/draft.tex
Evaluation edits

@@
 \newcommand{\todo}[1]{\hl{#1}\xspace}
+\newcommand{\fixme}[1]{[{\color{red}#1}]}
 \newcommand{\nv}[1]{[{\color{cyan}#1 --- nv}]}
@@
 
-Micro-bench\-marks~\sx{micro} are one-off, simple one-liners that take a few seconds to write;
+Micro-bench\-marks~\sx{ours} are one-off, simple one-liners that take a few seconds to write;
   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
@@
 
-\subsection{Shell One-liners}
-\label{micro}
+\subsection{Classic \unix One-liners}
+\label{ours}
 
@@
 
-\kk{Hypothesis on results. Dish gives a great speedup if one step of
-  the pipeline is what takes the most time. If the pipeline is long
-  with many computation heavy tasks, task based parallelism gets you
-  part of the way there, so we can't expect Dish to give performance
-  benefits equal to the amount of parallelism that we give it.}
-
+\heading{Programs}
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
@@
 Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
-
 Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
@@
 
-In terms of benchmarks, \ttt{grep} is a short script centered around an expensive \sta command;
+In more detail, \ttt{grep} is a short script centered around an expensive \sta command;
 % a DFA-based regular expression matching phase;
@@
 
+\heading{Performance}
 % TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
@@
 
+\heading{Discussion}
+\kk{Hypothesis on results. Dish gives a great speedup if one step of
+  the pipeline is what takes the most time. If the pipeline is long
+  with many computation heavy tasks, task based parallelism gets you
+  part of the way there, so we can't expect Dish to give performance
+  benefits equal to the amount of parallelism that we give it.}
+
 \subsection{Unix50 from Bell Labs}
@@
 
-\heading{Workload}
 The goal of this subsection is to evaluate \sys on an existing set of \unix pipelines found in the wild.
+
+\heading{Programs}
 In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of challenges
@@
 
-\heading{Setup}
+\TODO{An important point that needs to be made clear is that these are
+  taken from the wild unmodified. And that Pash never leads to
+  slowdown (except for the trivial head cases, where the slowdown is
+  neligible).}
+
+the problems are well designed to highlight the \unix philosophy.
+pipelines were found out in the wild and are not necessarily written in a particularly
+
+\heading{Results}
 \sys's parallelism set to \todo{16}. The input given to the pipelines
@@
 \Cref{fig:unix50-individual}. The average\todo{/median} speedup is
-\todo{5.24}.
+\todo{$5.24\times$} and the \todo{geometric mean is $XX\times$}.
+
+\heading{Discussion}
+
+\heading{Take-away}
+\sys leads to significant performance improvements in completely unmodified \unix pipelines found out in the wild.
+Small tweaks can yield further improvements, showing that \sys-awareness and scripting expertise can improve results.
 
@@
   \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
-  \caption{ \textbf{Speedup achieved by \sys for all pipelines with
-      parallelism 16. The red bars are the sequential execution time
-      of each pipeline.}
+  \caption{
+	  \textbf{Speedup (left axis) and over sequential execution (right axis) for all Unix50 pipelines.}
+		Parallelism is \todo{16$\times$} on \todo{10GB} of input data~\cf{unix50}.
+		\nv{Can decrease height and increase width substantially?}
     %% \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}
@@
 \end{figure}
-
-\TODO{An important point that needs to be made clear is that these are
-  taken from the wild unmodified. And that Pash never leads to
-  slowdown (except for the trivial head cases, where the slowdown is
-  neligible).}
 
@@
 To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
-At a high level, the goal is to process multi-year weather data from NCDC to identify the maximum temperature over the past several years.
+
+\heading{Program}
+At a high level, the program processes multi-year weather data from NCDC to identify the maximum temperature over the past several years.
 The original task, as presented in the book, can be thought of comprising three sub-tasks:
@@
 
+\heading{Results}
 % The first interesting result 
@@
   it amounts to 14 stages and, when expressed as a single line, only 240 characters.
-The Hadoop program, expressed in Java and only capturing the 4 last stages, is about 137 lines of Java code---not accounting for code moving input and output in and out of HDFS.
-
-In terms of execution time, with two Hadoop datanodes running on Hadoop 3.2.1 and Openjdk 11.0.5 without any replication (\ie fault-tolerance), the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.
-For the same amount of parallelism, \sys takes 4m30.554s---but this includes downloading and uncompressing the data files.
-\kk{I think it is fine keeping the Hadoop evaluation here, since this macrobenchmark was taken from a Hadoop book.}
+The Hadoop program, expressed in Java and only capturing the \emph{four} last stages, amounts to 137 lines of Java code---not accounting for commands moving data in and out of HDFS.
+
+As this is benchmark is taken directly from the Hadoop book, it is worth reporting on its execution time.
+The comparison is not unreasonable as both systems
+  (i) favor message-passing rather than shared memory,
+	(ii) Hadoop has no fault-tolerance for master and worker nodes (\ie both systems do not replicate data),
+	(iii) the task at hand was presumably chosen by the book author as one that matches Hadoop's capabilities, and 
+	(iv) the Hadoop program is written as a single program tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs that each written to handle many other cases.
+\todo{The original pipeline executes in XXs}.
+With two Hadoop 3.2.1 (Openjdk 11.0.5) data-nodes, the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.
+For the same amount of parallelism, \sys takes 4m30.554s---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
 
@@
 
-A less obvious take-away from this case is that large, complex pipelines enable significant freedom in how to express their stages.
-Several stages in the original shell program (taken from the book) are expressed as a single \ttt{awk} script;
-  unfortunately, \ttt{awk} is too general to have a meaningful distributability signature.
-A simpler pipeline that builds on built-in \unix primitives might be outperformed by specialized programs on a single node, but turns out to be  trivially scalable to multiple nodes.
+\heading{Discussion}
+Similar to Unix50~\sx{unix50}, we found that large, complex pipelines enable significant freedom in terms of expressiveness.
+Several stages of the original program were expressed as a single \ttt{awk} script;
+  unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
+A simpler (but longer) pipeline that leverages on \unix built-ins might be outperformed by a specialized programs on a single node, but turns out to be trivially scalable to multiple nodes.
 % For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
 %   this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
+
+\heading{Take-away}
+Multi-line Java Program parallel program decomposed in two phases---a highly parallel map phase and a sequential reduce phase---corresponds to a much smaller shell script with
 
@@
 \subsection{Use Case: Wikipedia Web Indexing}
-\label{macro1}
-
-\subsection{Use Case: Docker Container Orchestration}
-\label{macro1}
+\label{macro2}
+
+As a second realistic workload, we use a large pipeline for downloading and processing the web, comprised of stages collected from various repositories.
+
+\heading{Program}
+At a high level, the pipeline is comprised of two distinct components that coarsely correspond to a front- and a back-end communicating via \unix pipes.
+The front-end downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
+  the back-end applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index the incoming pages.
+The pipeline contains a total of 52 stages (18 for the font-end and 34 for the back-end), with written in multiple languages---for example, its \ttt{url-extract}ion utility is written in JavaScript whereas its \ttt{word-stem}ming utility is in Python.
+\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be described by its DSL.
+
+\heading{Results}
+To simplify execution (and avoid hitting Wikipedia), we have saved a recent version of Wikipedia to a server within the network and to redirect all \ttt{curl} requests there.
+The complexity of this pipeline makes end-to-end measurements difficult, so we measure the improvements on the three components individually and then report the differences
+
+\heading{Discussion}
+
+\heading{Take-away}
+
+\subsection{Use Case: Container Orchestration}
+\label{macro3}
+
+As a third realistic workload, we use a script for orchestrating containers to test the latest built across various Linux distributions.
+Compatibility testing is a common task for large software projects, commonly automated through shell scripting.
+
+\heading{Program}
+The 30-line orchestration script is unusual among our set of benchmarks in that it is not structured as a linear pipeline.
+Rather, it makes extensive use of POSIX constructs common in the shell, including output redirection, conditionals, and here-documents.
+It also combines commands from other classes outside \sta and \pur---more prominently \dfs---which are not parallelizable by \sys.
+
+\heading{Results}
+
+\heading{Take-away}
 
 \subsection{Further Micro-benchmarks}
+\label{micro}
 
@@
 Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
+
+\heading{Take-away}
+\sys will choose not to improve performance rather than breaking correctness; as a result, it does not always improve performance.
+In highly-parallel benchmarks, it can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
 

### 2020-05-02 pldi/draft.tex
Comparison with sort --parallel and comment about scalability

--- 2020-05-01 pldi/draft.tex
@@
 
-\begin{lstlisting}[language=sh, numbers=none]
+\begin{lstlisting}[language=sh, float=h, numbers=none]
  cat $in1 $in2 ... > $out
@@
 
-\begin{lstlisting}[language=sh, numbers=none]
+\begin{lstlisting}[language=sh, float=h, numbers=none]
   tee >(head -n $N > $temp1;
@@
 
-\heading{Dangling FIFOs and Zombie Producers}
-% \nv{I think all of these titles should be the problem/challenge, not the solution}
-% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
-% \ttt{SIGPIPE} 
-However, if its consumer exits early, the producer needs to be notified so that it stops appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.
-
-In \unix, this is achieved by an out-of-band error mechanism:
-  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
-This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
-  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
-}
-primarily because pipes and pipelines are at the heart of its philosophy.
-% write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
-Consider the following emitted code from \sys's earlier versions:
-\begin{lstlisting}[language=sh, numbers=none]
-  cat file-chunk1.txt > fifo1 &
-  cat file-chunk2.txt > fifo2 &
-  cat fifo1 fifo2 | head -n 1 & wait
-\end{lstlisting}
-\noindent
-In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
-As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
-This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
-Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
-
-To solve this problem, \sys emits cleanup logic that operates from the end of the pipeline and towards its start.
-The emitted code first gathers the ID of the last process and passes it as a parameter to \ttt{wait}; 
-  this causes \ttt{wait} to block only on the last consumer of the pipeline.
-Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
-This epilogue is repeated for every \todo{merge/join} point in the transformed pipeline.
-% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"
-
-% not receiving an out-of-band signal about that.
-% never terminating, but 
-% never terminate while the entire pipeline is waiting for 
-% \ttt{\& echo \$!} and \ttt{kill -PIPE}
-
-% The generated program could also (i) have signal handlers (traps) for propagating signals
-% down the process tree; (ii) incorporate (a simple) cleanup logic for not leaving dangling processes and
-% intermediary files; (iii) generate a correct exit code, depending on its
-% execution. All of this can be easily generated by the compiler.
-
 \section{Evaluation}
@@
   \footnotesize{
-    \textbf{Summary of Micro-benchmarks}.
-    The micro-benchmarks are small pipelines (\ie ones developed on-the-fly) drawn from various sources and applied to large datasets.
+    \textbf{Summary of shell one-liners}.
+    The shell one-liners are small pipelines (\ie ones developed on-the-fly) drawn from various sources and applied to large datasets.
   }
@@
 
-\kk{Hypothesis on results. Dish gives a great speedup if one step of
-  the pipeline is what takes the most time. If the pipeline is long
-  with many computation heavy tasks, task based parallelism gets you
-  part of the way there, so we can't expect Dish to give performance
-  benefits equal to the amount of parallelism that we give it.}
+\kk{Below follows an important point about why scalability is more
+  than great in the beginning and then gets worse.}
+
+An interesting point is that \sys seems to achieve almost linear
+speedup with small parallelism configurations. The reason why this
+happens is not so straightforward though. \sys manages that because it
+exploits task-based parallelism by having different processes to do
+the \todo{map} and the \todo{reduce} phase. Due to task-based
+parallelism in the generated dataflow graphs, \sys essentially ``uses
+more parallelism'' than the given configuration. This can be seen in
+all pipelines that are not completely stateless. For example, in the
+\ttt{minimal\_sort} one-liner, \sys configured to have 8 parallelism
+spawns 37 nodes (8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \ttt{sort -m}
+nodes, and 14 \ttt{eager} nodes. This indicates that \sys achieves
+optimal performance (without unnecessary process spawns and context
+switches) in pipelines that contain pure commands by configuring it
+with a smaller parallelism than the capabilities of the underlying
+system -- in our case 16-32 for a 64 physical core system. This is
+also the case if the input pipeline is very long and so there is
+already task-based parallelism even in the sequential execution
+(several of the Unix50 pipelines fall in this category).
+
+\kk{Hypothesis on a scalability upper bound. Dish gives a great
+  speedup if one step of the pipeline is what takes the most time. If
+  the pipeline is long with many computation heavy tasks, task based
+  parallelism gets you part of the way there, so we can't expect Dish
+  to give performance benefits equal to the amount of parallelism that
+  we give it.}
 
@@
   flags), and finally with Raftlib.}
+
+\paragraph{\ttt{sort --parallel}}
+
+We compare a single \todo{GNU} \ttt{sort} optimized by \sys with the
+same \ttt{sort} with the \ttt{--parallel} flag set. The results are
+shown in \Cref{fig:sort-parallel-comparison}. There are a couple
+important points to note. First of all, \sys without the \ttt{eager}
+optimization enabled performs comparably with \ttt{sort
+  --parallel}. On the other hand, enabling the \ttt{eager}
+optimization allows \sys to perform much better (almost \todo{2}
+times) than \ttt{sort} with the \ttt{--parallel} flag. That is because
+\sys with the \ttt{eager} optimization enabled essentially uses
+intermediate buffers (on disk) between the merge phases. On the other
+hand, \ttt{sort --parallel} sorts the input trying to minimize
+additional memory (so doesn't allocate intermediate buffers). \kk{If
+  we want, we can add here numbers where we allow sort --parallel to
+  use additional ram for intermediate buffers (using the -S flag).} Of
+course, it would be possible to manually create a longer pipeline and
+use intermediate buffers to speed this up, but this defeats the
+purpose of having a simple solution.
+
+Finally, based on the comparison with \ttt{sort --parallel} it seems
+that the scaleup limit that \sys reaches with sort is \todo{inherent
+  and cannot be overpassed} \kk{We have to find a way to argue for the
+  previous statement}. Because of that, all pipelines and scripts that
+contain \ttt{sort} in our evaluation (several one-liners and unix50)
+have this upper bound on scalability. \kk{Maybe move this paragraph in
+  the one-liners section.}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
+  \caption{ \textbf{Comparison with \ttt{sort --parallel}.}
+  }
+  \vspace{-15pt}
+  \label{fig:sort-parallel-comparison}
+\end{figure}
+
+\kk{There is a weird thing in the results. That the sequential one
+  with --parallel=1 takes more than the \sys sequential one.}
 

### 2020-05-02 compiler/gather_results.py

@@ -274,7 +274,7 @@ def plot_sort_with_baseline(results_dir):
 
     plt.xticks(all_scaleup_numbers[1:])
     plt.legend(loc='lower right')
-    plt.title(pretty_names[experiment])
+    plt.title("Comparison with sort --parallel")
 
 
     plt.tight_layout()

### 2020-05-03 pldi/draft.tex
Teensy notes

--- 2020-05-02 pldi/draft.tex
@@
 \heading{Programs}
-In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of challenges
+In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of challenges~\cite{unix50}.
 
@@
 \heading{Parallelization Command}
-We compare \sys to \ttt{parallel}~\cite{Tange2011a}, a GNU utility for running other commands in parallel.
+We compare \sys to \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}.
 In the first experiment, we leverage \ttt{parallel}'s \ttt{--jobs} flag and redirect input from a file to 
@@
 % % https://github.com/marcelm/cutadapt/issues/157
+% % https://www.biostars.org/p/123237/
 % \begin{lstlisting}[language=sh, numbers=none]

### 2020-05-03 pldi/draft.tex

@@
 \heading{Discussion}
-\kk{Hypothesis on results. Dish gives a great speedup if one step of
-  the pipeline is what takes the most time. If the pipeline is long
-  with many computation heavy tasks, task based parallelism gets you
-  part of the way there, so we can't expect Dish to give performance
-  benefits equal to the amount of parallelism that we give it.}
+\kk{Below follows an important point about why scalability is more
+  than great in the beginning and then gets worse.}
+An interesting point is that \sys seems to achieve almost linear
+speedup with small parallelism configurations. The reason why this
+happens is not so straightforward though. \sys manages that because it
+exploits task-based parallelism by having different processes to do
+the \todo{map} and the \todo{reduce} phase. Due to task-based
+parallelism in the generated dataflow graphs, \sys essentially ``uses
+more parallelism'' than the given configuration. This can be seen in
+all pipelines that are not completely stateless. For example, in the
+\ttt{minimal\_sort} one-liner, \sys configured to have 8 parallelism
+spawns 37 nodes (8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \ttt{sort -m}
+nodes, and 14 \ttt{eager} nodes. This indicates that \sys achieves
+optimal performance (without unnecessary process spawns and context
+switches) in pipelines that contain pure commands by configuring it
+with a smaller parallelism than the capabilities of the underlying
+system -- in our case 16-32 for a 64 physical core system. This is
+also the case if the input pipeline is very long and so there is
+already task-based parallelism even in the sequential execution
+(several of the Unix50 pipelines fall in this category).
+
+\kk{Hypothesis on a scalability upper bound. Dish gives a great
+  speedup if one step of the pipeline is what takes the most time. If
+  the pipeline is long with many computation heavy tasks, task based
+  parallelism gets you part of the way there, so we can't expect Dish
+  to give performance benefits equal to the amount of parallelism that
+  we give it.}
 
@@
 
-\heading{Parallelization Command}
+\heading{\ttt{sort --parallel}}
+We compare a single \todo{GNU} \ttt{sort} optimized by \sys with the
+same \ttt{sort} with the \ttt{--parallel} flag set. The results are
+shown in \Cref{fig:sort-parallel-comparison}. There are a couple
+important points to note. First of all, \sys without the \ttt{eager}
+optimization enabled performs comparably with \ttt{sort
+  --parallel}. On the other hand, enabling the \ttt{eager}
+optimization allows \sys to perform much better (almost \todo{2}
+times) than \ttt{sort} with the \ttt{--parallel} flag. That is because
+\sys with the \ttt{eager} optimization enabled essentially uses
+intermediate buffers (on disk) between the merge phases. On the other
+hand, \ttt{sort --parallel} sorts the input trying to minimize
+additional memory (so doesn't allocate intermediate buffers). \kk{If
+  we want, we can add here numbers where we allow sort --parallel to
+  use additional ram for intermediate buffers (using the -S flag).} Of
+course, it would be possible to manually create a longer pipeline and
+use intermediate buffers to speed this up, but this defeats the
+purpose of having a simple solution.
+
+Finally, based on the comparison with \ttt{sort --parallel} it seems
+that the scaleup limit that \sys reaches with sort is \todo{inherent
+  and cannot be overpassed} \kk{We have to find a way to argue for the
+  previous statement}. Because of that, all pipelines and scripts that
+contain \ttt{sort} in our evaluation (several one-liners and unix50)
+have this upper bound on scalability. \kk{Maybe move this paragraph in
+  the one-liners section.}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
+  \caption{ \textbf{Comparison with \ttt{sort --parallel}.}
+  }
+  \vspace{-15pt}
+  \label{fig:sort-parallel-comparison}
+\end{figure}
+
+\kk{There is a weird thing in the results. That the sequential one
+  with --parallel=1 takes more than the \sys sequential one.}
+
+\heading{GNU Parallel}
 We compare \sys to \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}.

### 2020-05-03 pldi/draft.tex
Refactor implementation nuggets and leave comment

@@
 
+\subsection{Implementation Nuggets}
+
+\heading{Overcoming Laziness}
+
+\noindent
+\TODO{Problem: Cat (and other commands that have multiple inputs) is
+  not eager. Illustration: Show a parallel pipeline having a cat that
+  gathers input somewhere (and a sort merge maybe). The way pipes
+  work, this graph will essentially evaluate almost completely
+  sequentially. Maybe go through the steps that cat asks for the first
+  line of its first input until ithe first input is done and then goes
+  to the second. This leads to processor under-utilization. Solution:
+  We add eager nodes that eagerly consume all their input (even if
+  their output does not ask for it). Eager nodes are themselves part
+  of the dataflow graph (and model) and essentially perform an
+  identity function. This makes all processors as utilized as possible
+  and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
+  light) tasks it has some overhead. We evaluate eager in section X
+  (We should also say something about a CPU light task). A good thing
+  is that they work for all possible commands with almost no overhead,
+  without having to make each specific command eager.}
+
 \begin{figure}[t]
@@
 
-\subsection{Implementation Nuggets}
-
-\TODO{Problem: Cat (and other commands that have multiple inputs) is
-  not eager. Illustration: Show a parallel pipeline having a cat that
-  gathers input somewhere (and a sort merge maybe). The way pipes
-  work, this graph will essentially evaluate almost completely
-  sequentially. Maybe go through the steps that cat asks for the first
-  line of its first input until ithe first input is done and then goes
-  to the second. This leads to processor under-utilization. Solution:
-  We add eager nodes that eagerly consume all their input (even if
-  their output does not ask for it). Eager nodes are themselves part
-  of the dataflow graph (and model) and essentially perform an
-  identity function. This makes all processors as utilized as possible
-  and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
-  light) tasks it has some overhead. We evaluate eager in section X
-  (We should also say something about a CPU light task). A good thing
-  is that they work for all possible commands with almost no overhead,
-  without having to make each specific command eager.}
-
-\TODO{(This problem is a nice followup for the one above) Problem:
-  Dataflow graph termination. Illustration: Show the example of head
-  deadlock and explain that this is inherent behaviour in the shell.
-  (Solution: emptying pipelines or terminating processes) In order for
-  things to terminate gracefully, dish wraps all commands in order to
-  always consume all of their input by sending it down to
-  /dev/null. Warning: This means that all commands in the pipeline
-  must terminate gracefully if all of their output is consumed (if a
-  process dies unexpectedly, noone will consume its input. We leave
-  addressing that and other monitoring questions for future work). We
-  evaluate the overhead of emptying all outputs in section X.}
-
-\TODO{Problem: Pipeline width shrinks after parallel pure
-  commands. Solution: Split node that resplits its input to multiple
-  files in sequence (to preserve the model semantics). We should
-  probably evaluate that too.}
-
-\paragraph{Merge and Split Implementation}
-%
-It is common in the dataflow graph for a node to have many incoming
-edges or many outgoing edges (when all but the last outgoing edges
-are bounded). However some nodes, might only read their input from one
-source (e.g. stdin). In this case, the distributed implementation
-needs to concatenate input files (or conversely split output
-files).
+\heading{Readjusting Graph Width}
+
+\noindent
+\TODO{Problem: Pipeline width shrinks after pure commands. Solution:
+  Split node that resplits its input to multiple files in sequence (to
+  preserve the model semantics).}
+
+%% Old and obsolete
+
+%% \noindent
+%% It is common in the dataflow graph for a node to have many incoming
+%% edges or many outgoing edges (when all but the last outgoing edges
+%% are bounded). However some nodes, might only read their input from one
+%% source (e.g. stdin). In this case, the distributed implementation
+%% needs to concatenate input files (or conversely split output
+%% files).
 %
@@
 %
-We address this issue using primitive shell constructs for file
-manipulation. More concretely a merger can be implemented by means of a simple
-\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
-output file \ttt{out}, a merger can be implemented as:
-
-\begin{lstlisting}[language=sh, numbers=none]
- cat $in1 $in2 ... > $out
-\end{lstlisting}
-
-\noindent
-On the other hand, given an input file \ttt{in} that is a FIFO pipe,
-and two output files \ttt{out1, out2} the first of which is bounded
-to \ttt{|N} lines, a 2-splitter can be implemented as:
-
-\begin{lstlisting}[language=sh, numbers=none]
-  tee >(head -n $N > $temp1;
-        dd of=/dev/null > /dev/null 2>&1 &
-        cat $temp1 > $out1) |
-       (tail -n +${N+1} > $out2;
-        dd of=/dev/null > /dev/null 2>&1)
-\end{lstlisting}
+%% We address this issue using primitive shell constructs for file
+%% manipulation. More concretely a merger can be implemented by means of a simple
+%% \ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
+%% output file \ttt{out}, a merger can be implemented as:
+
+%% \begin{lstlisting}[language=sh, numbers=none]
+%%  cat $in1 $in2 ... > $out
+%% \end{lstlisting}
+
+%% \noindent
+%% On the other hand, given an input file \ttt{in} that is a FIFO pipe,
+%% and two output files \ttt{out1, out2} the first of which is bounded
+%% to \ttt{|N} lines, a 2-splitter can be implemented as:
+
+%% \begin{lstlisting}[language=sh, numbers=none]
+%%   tee >(head -n $N > $temp1;
+%%         dd of=/dev/null > /dev/null 2>&1 &
+%%         cat $temp1 > $out1) |
+%%        (tail -n +${N+1} > $out2;
+%%         dd of=/dev/null > /dev/null 2>&1)
+%% \end{lstlisting}
  %% head -n $N $in > $out1 ; cat $in > $out2
 
-\noindent
-Using the 2-splitter as a building block, a pipe can be split to
-arbitrarily many different pipes.
+%% \noindent
+%% Using the 2-splitter as a building block, a pipe can be split to
+%% arbitrarily many different pipes.
+
+\TODO{Talk about eagerness in split and how we needed to add eager
+  nodes to n-1 outputs.}
 
 \heading{Dangling FIFOs and Zombie Producers}
+
+%% \TODO{(This problem is a nice followup for the one above) Problem:
+%%   Dataflow graph termination. Illustration: Show the example of head
+%%   deadlock and explain that this is inherent behaviour in the shell.
+%%   (Solution: emptying pipelines or terminating processes) In order for
+%%   things to terminate gracefully, dish wraps all commands in order to
+%%   always consume all of their input by sending it down to
+%%   /dev/null. Warning: This means that all commands in the pipeline
+%%   must terminate gracefully if all of their output is consumed (if a
+%%   process dies unexpectedly, noone will consume its input. We leave
+%%   addressing that and other monitoring questions for future work). We
+%%   evaluate the overhead of emptying all outputs in section X.}
+
 % \nv{I think all of these titles should be the problem/challenge, not the solution}
@@
 
-To solve this problem, \sys emits cleanup logic that operates from the end of the pipeline and towards its start.
-The emitted code first gathers the ID of the last process and passes it as a parameter to \ttt{wait}; 
-  this causes \ttt{wait} to block only on the last consumer of the pipeline.
-Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
-This epilogue is repeated for every \todo{merge/join} point in the transformed pipeline.
+To solve this problem, \sys emits cleanup logic that operates from the
+end of the pipeline and towards its start.  The emitted code first
+gathers the IDs of the output processes and passes them as parameters
+to \ttt{wait}; this causes \ttt{wait} to block only on the output
+producers of the dataflow graph.  Right after \ttt{wait}, \sys inserts
+a routine that delivers \ttt{PIPE} signals to any remaining processes
+upstream.  This epilogue is repeated for every \todo{merge/join} point
+in the transformed pipeline \kk{I don't understand the final sentence}.
 % Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"

### 2020-05-03 pldi/draft.tex
Minor updates in eval

--- 2020-05-03 pldi/draft.tex
@@
 
-\subsection{Classic \unix One-liners}
+\subsection{Common \unix One-liners}
 \label{ours}
+
+The goal of this subsection is to evaluate \sys on popular, common, or classic \unix pipeline patterns.
 
@@
 \heading{Programs}
-In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of challenges~\cite{unix50}.
-
-solving challenges by
-
-ability to compose larger programs in versatile ways from smaller utility programs
-summarized by McIllroy's Unix philosophy memo.
-
-The ability to compose larger programs in versatile ways from smaller utility programs was a hallmark of the \unix design, and the game's levels are designed to stress that.
-
-This section reports on programs that were run unaltered---highlighting 
-These are pipelines that were run verbatim
+In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of \todo{37} challenges~\cite{unix50} solvable by means of composing \unix pipelines.
+By scanning GitHub, we were able to find a set of shell programs that solve all-but-one problems~\cite{}.
+We consider this a good set of benchmarks because 
+(i) the problems were designed to to highlight the \unix philosophy~\cite{} (rather than having a single command that solves the problem) and make extensive use of \unix built-ins with a plethora of flags, and (ii) the solutions were written by non-experts (contrary to the \S\ref{ours}) and were executed by \sys unmodified---stressing patterns that we, \sys's authors, did not necessarily anticipate.
 In cases where an obvious fix improves the pipeline's performance, we report on both \sys's original speedup and the one achieved by applying the fix.
-
-\todo{Motivate why unix50 is a nice
-  set of pipelines.} We took the solutions for the unix50 from
-\todo{Ref} and we executed them \emph{without any modification} with
+% TODO: talk about correct wrt developer intentions vs. correct wrt problem definition
+
+% The ability to compose larger programs in versatile ways from smaller utility programs was a hallmark of the \unix design, and the game's levels are designed to stress that.
+% These challenges were specifically designed to highlight the \unix's philosophy, its 
+% ability to compose larger programs in versatile ways from smaller utility programs summarized by McIllroy's Unix philosophy memo.
 
@@
   neligible).}
-
-the problems are well designed to highlight the \unix philosophy.
-pipelines were found out in the wild and are not necessarily written in a particularly
 
@@
 
-\heading{\ttt{sort --parallel}}
+\heading{Parallel Sort}
 We compare a single \todo{GNU} \ttt{sort} optimized by \sys with the
@@
 Sequential execution takes \todo{XX}s, and \sys's parallel execution takes \todo{XX}s (speedup: \todo{XX}\%).
-We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command, masking \sys's improvements in other parts of the pipeline.
+We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command---masking \sys's improvements in other parts of the pipeline.
 
@@
 
-There are a few possible ways one might attempt to use GNU \ttt{parallel} to parallelize this program, which compared to \sys lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
-Assuming a user knows that the bottleneck comes from the \ttt{XX} stage, they could attempt to use \ttt{parallel} there leading to a runtime of \todo{XX}s (speedup: \todo{XX}\%)---worse than surprisingly worse than.
-
-Alternatively, a user could (incorrectly) sprinkle \ttt{parallel} across the entire program.
-This is a common strategy as GNU \ttt{parallel} allows feeding an entire script for parallelization via \ttt{stdin}. 
+There are a few possible ways one might attempt to use GNU \ttt{parallel} to parallelize this program.
+Assuming a user knows that the bottleneck comes from the \ttt{XX} stage, they could attempt to use \ttt{parallel} there.
+This leads to a runtime of \todo{XX}s (speedup: \todo{XX}\%).
+
+Alternatively, a user could (incorrectly) sprinkle \ttt{parallel} across the entire program;
+  this is a common strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
 Doing this leads to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
@@
 
+% Parallel is a great program for its target audience.
+
 \heading{Manual Parallelization}
-In an attempt to highlight how a user might attempt to manually parallelize individual commands using the abstractions provided by the POSIX shell.
-The benefits depend on several factors, such as whether a command is I/O- or CPU-bound or whether but since
-
-Merging is provided as Python programs
+Alternatively, a user might attempt to manually parallelize individual commands by using the abstractions provided by the POSIX shell.
+The benefits depend on several factors, such as whether a command is I/O- or CPU-bound.
 
@@
 \heading{Take-away}
-\sys will choose not to improve performance rather than breaking correctness; as a result, it does not always improve performance.
-In highly-parallel benchmarks, it can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
+\sys does not improve performance when running the risk of breaking correctness. % This should go in the intro
+% Compared to \sys, \ttt{parallel} can lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
+In highly-parallel benchmarks, \sys can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
 

### 2020-05-03 evaluation/plots/microbenchmarks-table.tex
Update missing cells in one-liners table

@@
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.049s\qquad 1.542s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 1~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 3.213s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 73m28.098s & \todo{\#Commands} & 0.122s\qquad 5.390s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.120s\qquad 5.311s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.211s\qquad 10.653s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 10.575s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 25m10.092s & \todo{\#Commands} & 0.193s\qquad 7.238s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{\#Commands} & 0.095s\qquad 4.524s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.248s\qquad 12.700s & two pipelines merging to a \tti{comm} \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.050s\qquad 1.571s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 3.309s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 73m28.098s & \todo{\#Commands} & 0.119s\qquad 5.427s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.121s\qquad 5.365s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & 	odo{light computation} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.214s\qquad 10.659s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 10.545s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{\#Commands} & 0.254s\qquad 7.123s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{\#Commands} & 0.098s\qquad 4.476s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.250s\qquad 12.800s & two pipelines merging to a \tti{comm} \\
 \bottomrule

### 2020-05-03 compiler/gather_results.py

@@ -75,7 +75,8 @@ input_filename_sizes = {"1G": "1~GB",
                         "100G": "100~GB",
                         "1M": "1~MB",
                         "10M": "10~MB",
-                        "100M": "100~MB"}
+                        "100M": "100~MB",
+                        "all_cmds_x1000": "85~MB"}
 
 def safe_zero_div(a, b):
     if(b == 0):

### 2020-05-03 compiler/execute_baseline_sort.sh
sort --parallel results

@@ -21,6 +21,7 @@ n_inputs=(
 
 for n_in in "${n_inputs[@]}"; do
     experiment="baseline_sort_${n_in}"
+    echo "$experiment"
     exec_script="${intermediary_dir}sort_${n_in}_seq.sh"
     env_file="${intermediary_dir}sort_${n_in}_env.sh"
 
@@ -37,6 +38,7 @@ done
 
 for n_in in "${n_inputs[@]}"; do
     experiment="baseline_sort_opt_${n_in}"
+    echo "$experiment"
     exec_script="${intermediary_dir}sort-opt_${n_in}_seq.sh"
     env_file="${intermediary_dir}sort-opt_${n_in}_env.sh"
 

### 2020-05-03 pldi/draft.tex

@@
 
-\TODO{Talk about the importance of eagerness and why we need to
-  enforce it to get benefits. Discuss the necessity of having files as
-  buffers and how we addressed the issue of cat's lazy consumption of
-  its inputs.}
-
 This section describes challenges related to the implementation of
@@
 
-\kk{Ideally we would want to talk here about the as-late-as possible
-  dish translation that allows us to be safe despite the Shell's
-  dynamic nature. I think that this is a great point about the
-  work. Maybe we should also focus on implementing it before the OSDI
-  deadline.}
+%% \kk{Ideally we would want to talk here about the as-late-as possible
+%%   dish translation that allows us to be safe despite the Shell's
+%%   dynamic nature. I think that this is a great point about the
+%%   work. Maybe we should also focus on implementing it before the OSDI
+%%   deadline.}
 
@@
 
-\TODO{Explain the plots with eager/no-eager as follows. Say that we
-  initially decided to only make an eager version of cat (to allow for
-  parallelism). However, we saw that results are still not good, so we
-  resorted to making a proper generic eager node that can be put
-  anywhere in the graph and that eagerly consumes its input. The above
-  results show the improvement with it. Rerun them with eager only
-  before cat in the no eager cases so that the minimal-grep example is
-  the same with and without our eager node. (Maybe) we need to mention
-  that eager has disk usage overhead that is linear to the size of the
-  pipeline and the size of the input.}
+%% \TODO{Explain the plots with eager/no-eager as follows. Say that we
+%%   initially decided to only make an eager version of cat (to allow for
+%%   parallelism). However, we saw that results are still not good, so we
+%%   resorted to making a proper generic eager node that can be put
+%%   anywhere in the graph and that eagerly consumes its input. The above
+%%   results show the improvement with it. Rerun them with eager only
+%%   before cat in the no eager cases so that the minimal-grep example is
+%%   the same with and without our eager node.}
+
+\TODO{(Maybe) we need to mention that eager has disk usage overhead
+  that is linear to the size of the pipeline and the size of the
+  input.}
 

### 2020-05-03 pldi/draft.tex
Plot for comparison with sort --parallel

@@
 
+\kk{We have to remember to say that for this to work, all nodes in the
+  graph must always open their outputs! If not, then this technique
+  doesn't work, and we have to monitor if any node dies to do
+  something with it. At the moment we ensure that by doing that in our
+  nodes, and by assuming that all nodes that we handle do that.}
+
 % not receiving an out-of-band signal about that.

### 2020-05-03 compiler/gather_results.py
Plot for comparison with sort --parallel

@@ -182,10 +182,10 @@ def collect_experiment_speedups(prefix, scaleup_numbers):
     compile_distr_speedup = [safe_zero_div(seq_numbers[i], t + compile_numbers[i]) for i, t in enumerate(distr_numbers)]
     return (distr_speedup, compile_distr_speedup)
 
-def collect_baseline_experiment_speedups(prefix, scaleup_numbers):
+def collect_baseline_experiment_speedups(prefix, scaleup_numbers, base_seq):
     seq_numbers = [read_total_time('{}{}_seq.time'.format(prefix, n))
                    for n in scaleup_numbers]
-    speedup = [safe_zero_div(seq_numbers[0], t) for t in seq_numbers]
+    speedup = [safe_zero_div(base_seq, t) for t in seq_numbers]
     return speedup
 
 
@@ -259,9 +259,18 @@ def plot_sort_with_baseline(results_dir):
     all_scaleup_numbers = [2, 4, 8, 16, 32, 64, 96, 128]
     sort_prefix = '{}/sort_'.format(results_dir)
     baseline_sort_prefix = '{}/baseline_sort/baseline_sort_'.format(results_dir)
-    sort_distr_speedup, _ = collect_experiment_speedups(sort_prefix, all_scaleup_numbers)
+    baseline_sort_opt_prefix = '{}/baseline_sort/baseline_sort_opt_'.format(results_dir)
+
+    ## Collect all sort numbers
+    seq_numbers, distr_numbers, _ = collect_experiment_scaleup_times(sort_prefix, all_scaleup_numbers)
+    sort_distr_speedup = [safe_zero_div(seq_numbers[i], t) for i, t in enumerate(distr_numbers)]
+    # sort_distr_speedup, _ = collect_experiment_speedups(sort_prefix, all_scaleup_numbers)
     baseline_sort_distr_speedup = collect_baseline_experiment_speedups(baseline_sort_prefix,
-                                                                       [1] + all_scaleup_numbers)
+                                                                       [1] + all_scaleup_numbers,
+                                                                       seq_numbers[0])
+    baseline_sort_opt_distr_speedup = collect_baseline_experiment_speedups(baseline_sort_opt_prefix,
+                                                                           [1] + all_scaleup_numbers,
+                                                                           seq_numbers[0])
 
[…]

### 2020-05-05 pldi/draft.tex
Update structure

--- 2020-05-03 pldi/draft.tex
@@
 %% \title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
-\title{\sys: Automated Data-Parallel Shell Processing}         %% [Short Title] is optional;
+\title{\sys: Light-touch Data-Parallel Shell Processing}         %% [Short Title] is optional;
 %% \title{Extensible data processing in the shell}
@@
 %% Author with single affiliation.
-\author{First1 Last1}
+\author{
+  Anonymous Author(s)
+  \normalsize{Additional Anonymized Material: \href{https://github.com/u-3/dish}{https://github.com/u-3/dish}.}
+}
 %% \authornote{with author1 note}          %% \authornote is optional;
@@
 %% Author with two affiliations and emails.
-\author{First2 Last2}
+% \author{First2 Last2}
 %% \authornote{with author2 note}          %% \authornote is optional;
@@
 
-\TODO{Drop the automated from the title maybe?}
+% \TODO{Drop the automated from the title maybe?}
+% \nv{Updates}
 
@@
 
+We attribute the small drop above $96\times$ in \ttt{spell} \ttt{sort} \ttt{wf}
+and \ttt{bi-grams} to a combination of hyperthreading (128 virtual vs. 64
+physical CPUs) \todo{the fixed width on \ttt{sort}}.
+% I say "we attribute" because it's a way to say something without being certain
+% FIXME: We need to take linear benchmarks on deathstar and livestar to see
+% where it starts dropping
+% not explain scope the results appropriately
+% rinard
+
+\heading{Take-aways}
+
 \subsection{Unix50 from Bell Labs}
@@
 
-\heading{Take-away}
+\heading{Take-aways}
 \sys leads to significant performance improvements in completely unmodified \unix pipelines found out in the wild.
@@
 
-\heading{Take-away}
+\heading{Take-aways}
 Multi-line Java Program parallel program decomposed in two phases---a highly parallel map phase and a sequential reduce phase---corresponds to a much smaller shell script with
@@
 
-\heading{Take-away}
+\heading{Take-aways}
 
@@
 
-\heading{Take-away}
+\heading{Take-aways}
 \sys does not improve performance when running the risk of breaking correctness. % This should go in the intro

### 2020-05-05 pldi/draft.tex
Fix
Changes in the sort --parallel subsection

@@
 and \ttt{bi-grams} to a combination of hyperthreading (128 virtual vs. 64
-physical CPUs) \todo{the fixed width on \ttt{sort}}.
+physical CPUs) \todo{the fixed width on} \ttt{sort}.
 % I say "we attribute" because it's a way to say something without being certain
@@
 
+\kk{I think the scoping of this should change to be both
+  micro-benchmarks and comparison with existing work. Especially since
+  we will have some form of comparison with Raftlib and phoenix. I
+  guess this section is trying to establish \sys limits by comparing
+  it with similar (but not identical) work.}
+
 The previous several subsections show substantial speedups without any developer effort on a variety of smaller and larger shell programs.
@@
 \heading{Parallel Sort}
-We compare a single \todo{GNU} \ttt{sort} optimized by \sys with the
-same \ttt{sort} with the \ttt{--parallel} flag set. The results are
-shown in \Cref{fig:sort-parallel-comparison}. There are a couple
-important points to note. First of all, \sys without the \ttt{eager}
-optimization enabled performs comparably with \ttt{sort
-  --parallel}. On the other hand, enabling the \ttt{eager}
-optimization allows \sys to perform much better (almost \todo{2}
-times) than \ttt{sort} with the \ttt{--parallel} flag. That is because
-\sys with the \ttt{eager} optimization enabled essentially uses
-intermediate buffers (on disk) between the merge phases. On the other
-hand, \ttt{sort --parallel} sorts the input trying to minimize
-additional memory (so doesn't allocate intermediate buffers). \kk{If
-  we want, we can add here numbers where we allow sort --parallel to
-  use additional ram for intermediate buffers (using the -S flag).} Of
-course, it would be possible to manually create a longer pipeline and
-use intermediate buffers to speed this up, but this defeats the
-purpose of having a simple solution.
-
-Finally, based on the comparison with \ttt{sort --parallel} it seems
+%% \kk{Explain what is the purpose of the
+%%   microbenchmark. We want to establish that \sys scalability is as
+%%   good as it could be.}
+We first compare a single GNU \ttt{sort} optimized by \sys with the
+same \ttt{sort} with the \ttt{-{}-parallel} flag set. Even though the
+\ttt{-{}-parallel} flag is not a general parallelism solution, this
+comparison serves to establish a baseline for \sys's performance. The
+results are shown in \Cref{fig:sort-parallel-comparison}. Note that
+for \ttt{sort -{}-parallel} parallelism is set to double the number
+set of \sys (i.e. the final point is for \ttt{-{}-parallelism=128})
+since \sys also spawns separate processes for sorting and for
+merging. There are a couple important points to note.
+
+First of all, \sys without the \ttt{eager} optimization enabled
+performs comparably with \ttt{sort -{}-parallel}. On the other hand,
+enabling the \ttt{eager} optimization allows \sys to perform much
+better ($\sim 2$ times) than \ttt{sort} with the \ttt{-{}-parallel}
+flag. That is because \sys with the \ttt{eager} optimization enabled
+essentially uses intermediate buffers (on disk) between the merge
+phases.
+%% On the other hand, \ttt{sort --parallel} sorts the input trying to
+%% minimize additional memory (so doesn't allocate intermediate
+%% buffers).
+%% %% \kk{If we want, we can add here numbers where we allow sort --parallel
+%% %%   to use additional ram for intermediate buffers (using the -S flag).}
+%% Of course, it would be possible to manually create a longer pipeline
+%% and use intermediate buffers to speed this up, but this defeats the
+%% purpose of having a simple solution.
+
+Second, based on the comparison with \ttt{sort --parallel} it seems
 that the scaleup limit that \sys reaches with sort is \todo{inherent
@@
 
+Finally, this comparison showcases the benefit that \sys provides to
+command developers. Instead of implementing a custom flag that enables
+parallelism for a command, a simple specification of the command
+parallelizability allows \sys to scale it similarly to a handcrafted
+implementation with almost no effort.
+
 \begin{figure}[t]
@@
   \includegraphics[width=\columnwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
-  \caption{ \textbf{Comparison with \ttt{sort --parallel}.}
+  \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.}
   }
@@
 
-\kk{There is a weird thing in the results. That the sequential one
-  with --parallel=1 takes more than the \sys sequential one.}
+%% \kk{There is a weird thing in the results. That the sequential one
+%%   with --parallel=1 takes more than the \sys sequential one.}
 

### 2020-05-05 evaluation/plots/microbenchmarks-table.tex
New plots

--- 2020-04-18 evaluation/plots/microbenchmarks-table.tex
@@
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(\todo{16, 128}) &Compile Time (\todo{16, 128}) & Highlights \\
+Script ~&~ Structure & Input &Seq Time & Script Size(16, 64) &Compile Time (16, 64) & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.049s\qquad 1.542s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 1~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 3.213s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 73m28.098s & \todo{\#Commands} & 0.122s\qquad 5.390s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.120s\qquad 5.311s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.577s & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.211s\qquad 10.653s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 10.575s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 25m10.092s & \todo{\#Commands} & 0.193s\qquad 7.238s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{\#Commands} & 0.095s\qquad 4.524s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.248s\qquad 12.700s & two pipelines merging to a \tti{comm} \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.050s\qquad 0.417s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 0.839s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~MB & 73m28.098s & \todo{\#Commands} & 0.119s\qquad 1.407s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.121s\qquad 1.388s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.163s & 	odo{light computation} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.214s\qquad 2.800s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 2.762s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{\#Commands} & 0.254s\qquad 1.958s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~MB & 125m10.224s & \todo{\#Commands} & 0.098s\qquad 1.171s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.250s\qquad 3.316s & two pipelines merging to a \tti{comm} \\
 \bottomrule

### 2020-05-05 compiler/gather_results.py

@@ -215,7 +215,7 @@ def collect_scaleup_times(experiment, results_dir):
     # all_scaleup_numbers = list(set(get_experiment_files(experiment, results_dir)))
     # all_scaleup_numbers.sort()
     # all_scaleup_numbers = [i for i in all_scaleup_numbers if i > 1]
-    all_scaleup_numbers = [2, 4, 8, 16, 32, 64, 96, 128]
+    all_scaleup_numbers = [2, 4, 8, 16, 32, 64]
     prefix = '{}/{}_'.format(results_dir, experiment)
     distr_speedup, compile_distr_speedup = collect_experiment_speedups(prefix, all_scaleup_numbers)
 
@@ -250,13 +250,13 @@ def collect_scaleup_times(experiment, results_dir):
 
 
     plt.tight_layout()
-    plt.savefig(os.path.join('../evaluation/plots', "{}_throughput_scaleup.pdf".format(experiment)))
+    plt.savefig(os.path.join('../evaluation/plots', "{}_throughput_scaleup.pdf".format(experiment)),bbox_inches='tight')
 
     return output_diff
 
 def plot_sort_with_baseline(results_dir):
 
-    all_scaleup_numbers = [2, 4, 8, 16, 32, 64, 96, 128]
+    all_scaleup_numbers = [2, 4, 8, 16, 32, 64]
     sort_prefix = '{}/sort_'.format(results_dir)
     baseline_sort_prefix = '{}/baseline_sort/baseline_sort_'.format(results_dir)
     baseline_sort_opt_prefix = '{}/baseline_sort/baseline_sort_opt_'.format(results_dir)
@@ -266,11 +266,12 @@ def plot_sort_with_baseline(results_dir):
     sort_distr_speedup = [safe_zero_div(seq_numbers[i], t) for i, t in enumerate(distr_numbers)]
     # sort_distr_speedup, _ = collect_experiment_speedups(sort_prefix, all_scaleup_numbers)
     baseline_sort_distr_speedup = collect_baseline_experiment_speedups(baseline_sort_prefix,
-                                                                       [1] + all_scaleup_numbers,
+                                                                       [1] + [num*2
+                                                                              for num in all_scaleup_numbers],
                                                                        seq_numbers[0])
-    baseline_sort_opt_distr_speedup = collect_baseline_experiment_speedups(baseline_sort_opt_prefix,
[…]

### 2020-05-05 compiler/gather_results.py
Improve Unix50 section

@@ -2,6 +2,7 @@ import sys
 import os
 import numpy as np
 import matplotlib.pyplot as plt
+import statistics
 
 SMALL_SIZE = 16
 MEDIUM_SIZE = 18
@@ -428,6 +429,15 @@ def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
     absolute_seq_times_s = [absolute_seq[scaleup_numbers.index(parallelism)] / 1000
                             for _, absolute_seq in all_results]
     print("Unix50 individual speedups for {} parallelism:".format(parallelism), individual_results)
+    mean = sum(individual_results) / len(individual_results)
+    median = statistics.median(individual_results)
+    geo_mean = np.log(individual_results).sum() / len(individual_results)
+    weighted_res = [i*a for i, a in zip(individual_results, absolute_seq_times_s)]
+    weighted_avg = sum(weighted_res) / sum(absolute_seq_times_s)
+    print("  Mean:", mean)
+    print("  Median:", median)
+    print("  Geometric Mean:", geo_mean)
+    print("  Weighted Average:", weighted_avg)
 
     w = 0.2
     ind = np.arange(len(individual_results))

### 2020-05-05 pldi/draft.tex
Improve Unix50 section

@@
       \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
-      Three different configurations per benchmark:
-      (i) parallel, focusing only on the parallel execution,
-      (ii) +compile, adding \sys' pre-processing overheads, and
-      (iii) \kk{Add with , without eager and split here}.
-      \nv{Candidate values, based on the HW we have 2, 4, 8, 16, 32, 64. Also, maybe it makes sense to show log x-axis?}
+      Four different configurations per benchmark:
+      (i) \todo{Parallel}, showing the performance of \sys with both eager and split optimizations enabled,
+      (ii) \todo{No Split}, showing the performance of \sys with split disabled,
+      (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
+      (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
+      \nv{Maybe it makes sense to show log x-axis?}
     }
@@
 
-\heading{Programs}
-In a recent event celebrating \unix's 50-year legacy, Bell Labs created a set of \todo{37} challenges~\cite{unix50} solvable by means of composing \unix pipelines.
-By scanning GitHub, we were able to find a set of shell programs that solve all-but-one problems~\cite{}.
-We consider this a good set of benchmarks because 
-(i) the problems were designed to to highlight the \unix philosophy~\cite{} (rather than having a single command that solves the problem) and make extensive use of \unix built-ins with a plethora of flags, and (ii) the solutions were written by non-experts (contrary to the \S\ref{ours}) and were executed by \sys unmodified---stressing patterns that we, \sys's authors, did not necessarily anticipate.
-In cases where an obvious fix improves the pipeline's performance, we report on both \sys's original speedup and the one achieved by applying the fix.
+\heading{Programs} In a recent event celebrating \unix's 50-year
+legacy, Bell Labs created a set of \todo{36} challenges~\cite{unix50}
+solvable by means of composing \unix pipelines.  By scanning GitHub,
+we were able to find a set of shell programs that solve
+\todo{all-but-three} problems~\cite{}.  We consider this a good set of
+benchmarks because (i) the problems were designed to to highlight the
+\unix philosophy~\cite{} (rather than having a single command that
+solves the problem) and make extensive use of \unix built-ins with a
+plethora of flags, and (ii) the solutions were written by non-experts
+(contrary to the \S\ref{ours}) and were executed by \sys
+unmodified---stressing patterns that we, \sys's authors, did not
+necessarily anticipate \kk{I would change that to ``stressing a
+  variety of patterns'' since it doesn't really matter if we
+  anticipated them (especially since we can address any issues and
+  then re-run them).} .  In cases where an obvious fix improves the
+pipeline's performance, we report on both \sys's original speedup and
+the one achieved by applying the fix.
 % TODO: talk about correct wrt developer intentions vs. correct wrt problem definition
@@
 
-\TODO{An important point that needs to be made clear is that these are
-  taken from the wild unmodified. And that Pash never leads to
-  slowdown (except for the trivial head cases, where the slowdown is
-  neligible).}
-
-\heading{Results}
-\sys's parallelism set to \todo{16}. The input given to the pipelines
-was the test input from unix50 multiplied to reach \todo{10GBs}. The
-speedup of each pipeline separately is shown in
+%% \TODO{An important point that needs to be made clear is that these are
+%%   taken from the wild unmodified. And that Pash never leads to
+%%   slowdown (except for the trivial head cases, where the slowdown is
+%%   neligible).}
+
+\heading{Results} \sys's parallelism set to 16. The input given to the
+pipelines was the test input from unix50 multiplied to reach
+10GBs. The speedup of each pipeline separately is shown in
 \Cref{fig:unix50-individual}. The average\todo{/median} speedup is
-\todo{$5.24\times$} and the \todo{geometric mean is $XX\times$}.
+$5.49$/\todo{$6.07\times$}, the \todo{geometric mean is $1.08\times$},
+and the \todo{weighted average (with the absolute times as weights) is
+  $5.75\times$}.
+
+%% Mean: 5.499164686559543
+%% Median: 6.075774436154179
+%% Geometric Mean: 1.0798740409683736
+%% Weighted Average: 5.752406752503295
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \caption{
+	  \textbf{Speedup (left axis) and over sequential execution (right axis) for all Unix50 pipelines.}
+		Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
+  }
+  \vspace{-15pt}
+  \label{fig:unix50-individual}
+\end{figure}
 
 \heading{Discussion}
-
-\heading{Take-aways}
-\sys leads to significant performance improvements in completely unmodified \unix pipelines found out in the wild.
-Small tweaks can yield further improvements, showing that \sys-awareness and scripting expertise can improve results.
 
@@
 
-\kk{Out of all benchmarks, we have talked about the head and the
-  non-parallelizable ones in the paper. Maybe we also want to talk
-  about the long and the cheap ones. Also we might want to talk and
-  find out more about the sort bottleneck.}
+%% \kk{Out of all benchmarks, we have talked about the head and the
+%%   non-parallelizable ones in the paper. Maybe we also want to talk
+%%   about the long and the cheap ones. Also we might want to talk and
+%%   find out more about the sort bottleneck.}
 
@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
-  \caption{
-	  \textbf{Speedup (left axis) and over sequential execution (right axis) for all Unix50 pipelines.}
-		Parallelism is \todo{16$\times$} on \todo{10GB} of input data~\cf{unix50}.
-		\nv{Can decrease height and increase width substantially?}
-    %% \todo{Scale the 0-1 region of the yaxis as much as the 1-20.}
-  }
-  \vspace{-15pt}
-  \label{fig:unix50-individual}
-\end{figure}
-
 \sys achieves speedup in all pipelines except for 2, 19, and 31. These
@@
 the order of 10 milliseconds. The slowdown is due to constant costs of
-setting up the pipes and spawning many processes. In all three of
-these cases the execution time with \sys is in the order of \todo{X
-  ms}. For pipelines 13, 24, 25, 26, 29, and 30, \sys performs the
-same as the sequential implementation. This is because these pipelines
-contain non-parallelizable commands, such as \ttt{awk}, \ttt{sed} with
-option \ttt{d}, and \ttt{tr} with option \ttt{-d}. An illustrative
-example is pipeline 13, which is shown here:
+setting up the pipes and spawning processes. In all three of these
+cases the execution time with \sys is in the order of 1 second. For
+pipelines 13, 24, 25, 26, 29, and 30, \sys performs the same as the
+sequential implementation. This is because these pipelines contain
+commands that can not be parallelized in general, such as \ttt{awk},
+\ttt{sed} with option \ttt{d}, and \ttt{tr} with option \ttt{-d}. An
+illustrative example is pipeline 13, which is shown here:
 
 \begin{lstlisting}[language=sh, float=h, numbers=none]
-  cat $IN | awk "{print \$2, \$0}" | sort -nr | cut -d ' ' -f 2
+  cat $IN | awk "{print \$2, \$0}" |
+      sort -nr | cut -d ' ' -f 2
 \end{lstlisting}
@@
 a \sys user needs to spend is often negligible to achieve
-benefits. Finally, an important point that is reinforced from this
-experiment is that \sys doesn't lead to slowdown, even in cases where
-parallelism is not possible.
+benefits.
+
+Finally, for the rest of the pipelines \sys speedup is \todo{capped}
+because of a combination of the following reasons: (i) they contain
+pure commands that are parallelizable but don't scale linearly, such
+as \ttt{sort} (e.g. pipelines 0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
+(ii) they are deep pipelines that already exploit task-based
+parallelism (e.g. pipelines 7, 8, 9, 10, 11, 12, 14, 27, 28), and
+(iii) they are not cpu-intensive, leading to \todo{larger IO and
+  constant costs} (e.g. pipelines 4, 5, 6, 7, 8, 12, 14, 22, 23).
+
+%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
+%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
+%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14, 22, 23
+%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
+
+%% Finally, an important point that is reinforced from this
+%% experiment is that \sys doesn't lead to slowdown, even in cases where
+%% parallelism is not possible.
+
+\heading{Take-aways} \sys leads to significant performance
+improvements in completely unmodified \unix pipelines found out in the
+wild. Small tweaks can yield further improvements, showing that
+\sys-awareness and scripting expertise can improve
+results. Furthermore, it doesn't cause slowdown in any non-trivial
+computation, even in cases where parallelism is not possible.
 
@@
 \label{macro1}
+
+\kk{Maybe use phoenix for that one? Since it has a MapReduce interface
+  but it runs on multicore?}
 

### 2020-05-06 pldi/draft.tex
Checkpoint

--- 2020-05-05 pldi/draft.tex
@@
 \author{
-  Anonymous Author(s)
-  \normalsize{Additional Anonymized Material: \href{https://github.com/u-3/dish}{https://github.com/u-3/dish}.}
+  Anonymous Author(s)\\
+  \normalsize{Additional Anonymized Material: \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.}
 }
@@
 %
-Consider the following data processing Shell script.
-%% \kk{We might want to have a more complex script here.}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
-  sort | uniq -c | sort -rn | head 5 > out
-\end{lstlisting}
-
-\noindent
-Scripts like the one above are commonly used for orchestrating
-data-processing workflows; connecting system commands with custom
-tools to perform complex tasks.
-%% \kk{Do we need to mention domains where that happens?}
-Even though shell pipelines are designed to exploit task parallelism,
-they completely ignore available data-parallelism. For small input
-sizes this isn't a problem since scripts take less than a second to
-execute, but as data sizes increase, failing to utilize the underlying
-parallel computational resources is a significant bottleneck.
-
-To address this concern, a Shell \emph{user} could resort to several
-different solutions (e.g. gnuparallel, tsc, flags in individual
-commands etc...) that offer some form of parallelism. However these
-solutions are too ad-hoc (each having a disparate parallelism model),
-too limited (they only apply to specific use-cases), and too
-coarse-grained. \kk{Tone the previous claims down and back them up.}
-At the same time, a \emph{developer} of Shell commands could rewrite
-her commands using low level tools (e.g. Raftlib, openMP, etc) to
-improve utilization and performance. Manual rewriting is expensive and
-can introduce new bugs, cascading changes, and divergence from legacy
-functionality.
-
-To address this issue for both \emph{users} and \emph{developers}, we
-propose \sys, a tool that automatically exploits available
-data-parallelism in a given shell script. The key insight behind \sys
-is that a large subset of the Shell language can be encoded as a
-dataflow graph, a flexible representation that allows for
-parallelization optimizations. \sys builds on this insight with a
-careful study of the parallelizability properties of shell primitives
-and commands, a definition of a dataflow graph model that is suitable
-for the shell, a set of parallelization optimizations that preserve
-the behaviour of the graph, and a late-bound command-prefixing scheme
-for deferring optimizations as late-as-possible (\ie when runtime
-information is available). \kk{Maybe elaborate some more on the
-  contributions in this paragraph.}  A \sys-enabled $p_1$ will run
-\ttt{tr}s on parallel streams, use a mostly-parallel tree of
-\ttt{sort}s, run \ttt{uniq} mostly in parallel, execute \ttt{head} in
-one of the streams, and (trivially) merge streams in \ttt{out}. The
-only requirement to support a new command is that its developer adds a
-lightweight annotation to describe its parallelizability properties.
-
-The paper is structured as follows. It starts by introducing the
-necessary background on shell scripting and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{impl} highlight key
-contributions:
+% Consider the following data processing Shell script.
+% %% \kk{We might want to have a more complex script here.}
+% 
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+% cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
+%   sort | uniq -c | sort -rn | head 5 > out
+% \end{lstlisting}
+% 
+% \noindent
+% great sentence -- keep
+% Scripts like the one above are commonly used for orchestrating data-processing workflows; connecting system commands with custom tools to perform complex tasks.
+% %% \kk{Do we need to mention domains where that happens?}
+% Even though shell pipelines are designed to exploit task parallelism, they completely ignore available data-parallelism.
+% % Too low-level descriptions of the problem, maybe keep it high level?
+% For small input sizes this isn't a problem since scripts take less than a second to execute, but as data sizes increase, failing to utilize the underlying parallel computational resources is a significant bottleneck.
+
+The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
+This language agnosticism, coupled with \unix's toolbox philosophy~\cite{mcIlroymemo}, makes the shell the primary choice for specifying succinct and simple pipelines that target data processing, system orchestration, and other automation tasks.
+% Pipe-centric shell commands are parallel (and to some extent, functional) programs that can natively enjoy significant parallelism on a multicore machine.
+Unfortunately, parallelizing such pipelines requires significant effort and is shared unevenly between two different groups (or roles) of a programmer. % it's more like personas or "wearing hats"
+
+The first group is \emph{command developers} responsible for developing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
+These developers usually work in a single---often low-level---programming language and manually incorporate parallelism through means available in their chosen language, whenever this is possible.
+%  by the problem and language.
+% the language and abstractions they are using focused at a particular task 
+% also: target a small problem but aim at high reusability 
+As they have no visibility into the command's uses, they expose a plethora of ad-hoc command-specific flags such as \ttt{-t}~\cite{}, \ttt{--parallel}~\cite{}, \ttt{-p}, \ttt{CPUs}, and \ttt{NUM\_THREADS}.
+
+The second group is \emph{command users}, composing scripts out of individual commands.
+% Unfortunately, users do not operate in a single language
+As their programs combine a variety of commands written in many languages, they are left with only a few manual options for achieving parallelism:
+% use the composition abstractions of the \unix shell, 
+% and are often prototypical in nature---focusing on quickly prototyping a solution for the task at hand rather building an optimal solution to a recurring pattern.
+% short lived, changing quickly
+either coarse-grained tools such as \eg GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, or manual parallelism POSIX shell primitives and 
+% (ii) ad-hoc flags provided by command developers and mentioned above.
+
+Their last resort is 
+and writing the program from scratch
+Manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
+% The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
+Often, these users are domain experts lacking a deep understanding of parallel systems---in desperate need of automation when it comes to parallelization.
+
+This paper presents a solution targeting the second class of programmers, command users, 
+\sys can automate
+
+A system that is aware of the \unix philosophy, its key abstractions, and offers automated POSIX-aware parallelization as long as 
+These lightweight annotations can be expressed by developers 
+can be expressed 
+This is paired with parallelizability study that informed the design of the system and its annotation interface, and led to the parallelizability semantics for POSIX and GNU coreutils---a large class of commands that serves as the standard library.
+
+%% First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
+
+To address this challenge, we develop \sys, a system for transforming POSIX shell scripts to their parallel equivalents.
+\sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
+First, a careful study of the parallelizability properties of shell primitives and commands.
+\sys addresses the challenge by offering command developers a unified set of abstractions for describing the parallelizability characteristics of their commands.
+Using lightweight annotations akin to type signatures, \sys can automate the parallelization of entire POSIX scripts without any input from their user.
+hand-in-hand with the parallelizability study:
+  the study informed the annotation language and was used to annotate the entire
+  for all 
+---a set of annotations that \sys already offers for 
+for the standard library
+  akin to the standard library
+  the annotation language is informed by the study, but also the study is 
+XX to a type signature, these annotations describe
+
+Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
+(one sentence for the model)
+(one sentence for the optimizations)
+The result of the compilation is still a POSIX-compliant script that manages parallelism explicitly;
+  POSIX compliance means 
+it does not depend on a managed runtime
+
+Third, \sys packs a runtime component 
+A small library of highly-optimized data combiners allows 
+This is paired with a
+it comes with a few runtime primitives for eager data splitting and merging that address several practical issues ranging from deadlocks, performance bottlenecks;
+  these primitives were developed from uncovering several pathological situations, on many of which we report.
+
+We perform an extensive evaluation of \sys using 
+  (i) a series of benchmarks, ranging from classic scripts found in the \unix literature to modern data-processing pipelines;
+  (ii) three large and complex use cases, ranging from temperature analysis to web crawling and indexing; and
+  (iii) a series of micro-benchmarks inspired by (and comparing with) possible alternatives.
+The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
+Improvements on the performance of individual commands depend on several factors such as the script's execution time, the pipeline depth and the parallelizability class under which commands fall.
+
+%                                           a tool that automatically exploits available data-parallelism in a given shell script.
+
+The paper is structured as follows.
+It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
+Sections \ref{distributability}--\ref{impl} highlight key contributions:
 \begin{itemize}
   \item
-  \S\ref{distributability} studies the parallelizability properties of
-  a large class of Shell commands (coreutils) and introduces an
-  lightweight annotation language for commands that could be executed
-  in a data-parallel manner. \kk{Mention that we did this for all
-    coreutils.}
+  \S\ref{distributability}
+    studies the parallelizability properties of a large class of Shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
 
   \item
-  \S\ref{ir} presents a dataflow graph model for encoding shell
-  scripts, and a set of safe parallelization transformations that
-  preserve the semantics of the sequential program.
+  \S\ref{ir} presents a dataflow graph model for encoding shell scripts, and a set of parallelization transformations that focus on preserving the semantics of the sequential program.
 
   \item
-  \S\ref{impl} describes the \sys implementation and addresses several
-  challenges related to the translation and optimization of shell
-  programs. \kk{Rewrite that when we know for sure what the section
-    contains.}
+  \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and optimization of shell programs.
+% \kk{Rewrite that when we know for sure what the section contains.}
 
@@
 \noindent
-\sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$)
-for parallel execution using a combination of real pipelines.
+After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
 %micro-bench\-marks %---popular shell one-liners that highlight
@@
 %weather analysis script.
-After a comparison with related prior work~\sx{related}, the paper
-closes with a discussion of limitations and possible future
-directions~\sx{discussion}.
 %% Additional material, in anonymized form, is available at
@@
 
-% TODO: Important to have a small script that highlights
-\nv{
-(i) different classes of commands
-(ii) challenge of shell expansion order
-(iii) segments that are parallelized by \sys (with provable correctness)
-(iv) segments that are \emph{not} parallelized by \sys (otherwise they can lead to
-incorrectness)
-}
+% % TODO: Important to have a small script that highlights
+% \nv{
+% (i) different classes of commands
+% (ii) challenge of shell expansion order
+% (iii) segments that are parallelized by \sys (with provable correctness)
+% (iv) segments that are \emph{not} parallelized by \sys (otherwise they can lead to
+% incorrectness)
+% }
 
@@
 As this is benchmark is taken directly from the Hadoop book, it is worth reporting on its execution time.
-The comparison is not unreasonable as both systems
-  (i) favor message-passing rather than shared memory,
+While the setup is not typical of Hadoop, the comparison between the two systems is not unreasonable:  % we took several steps to make the comparison fair:
+  (i) both Hadoop and Spark favor message-passing rather than shared memory,
 	(ii) Hadoop has no fault-tolerance for master and worker nodes (\ie both systems do not replicate data),
-	(iii) the task at hand was presumably chosen by the book author as one that matches Hadoop's capabilities, and 
-	(iv) the Hadoop program is written as a single program tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs that each written to handle many other cases.
-\todo{The original pipeline executes in XXs}.
+	(iii) the task at hand is a good fit for  Hadoop,
+	(iv) the Hadoop program is written as a single program tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases.
+\todo{The original pipeline executes in XXs---long enough to amortize Hadoop's 0.5--0.9s startup costs.}
 With two Hadoop 3.2.1 (Openjdk 11.0.5) data-nodes, the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.

### 2020-05-07 pldi/draft.tex
Edit the web-index use case part of the paper

--- 2020-05-05 pldi/draft.tex
@@
 
-As a second realistic workload, we use a large pipeline for downloading and processing the web, comprised of stages collected from various repositories.
-
-\heading{Program}
-At a high level, the pipeline is comprised of two distinct components that coarsely correspond to a front- and a back-end communicating via \unix pipes.
-The front-end downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
-  the back-end applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index the incoming pages.
-The pipeline contains a total of 52 stages (18 for the font-end and 34 for the back-end), with written in multiple languages---for example, its \ttt{url-extract}ion utility is written in JavaScript whereas its \ttt{word-stem}ming utility is in Python.
-\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be described by its DSL.
-
-\heading{Results}
-To simplify execution (and avoid hitting Wikipedia), we have saved a recent version of Wikipedia to a server within the network and to redirect all \ttt{curl} requests there.
-The complexity of this pipeline makes end-to-end measurements difficult, so we measure the improvements on the three components individually and then report the differences
+As a second realistic workload, we use a large pipeline for
+downloading and processing the web, comprised of stages collected from
+various repositories.
+
+\heading{Program} At a high level, the pipeline is comprised of two
+distinct components that coarsely correspond to a front- and a
+back-end communicating via \unix pipes.  The front-end downloads pages
+and extracts their URLs from their source, feeding back \todo{to} the
+start of the pipeline; the back-end extracts the text from the html
+pages and then applies natural-language processing---\eg trigrams,
+character conversion, term frequencies---to index the incoming pages.
+The pipeline contains a total of 52 stages (18 for the font-end and 34
+for the back-end) written in multiple languages---for example,
+its \ttt{url-extract}ion utility is written in JavaScript whereas its
+\ttt{word-stem}ming utility is in Python.  \sys can still operate on
+them as their parallelizability properties---\sta for
+\ttt{url-extract} and \ttt{word-stem}---can be described by its DSL.
+
+\heading{Results} To simplify execution (and avoid hitting Wikipedia),
+we have saved a recent version of Wikipedia to a server within the
+network and to redirect all \ttt{curl} requests there.  The complexity
+of this pipeline makes end-to-end measurements difficult, so we
+measure the improvements on the two components components individually
+and then report the differences. With parallelism set to \todo{2/16}
+the execution time achieved by \sys is \todo{X} for the front-end and
+\todo{Y} for the back-end, while the original execution time is
+\todo{X} for the front-end and \todo{Y'} for the back-end. \kk{If the
+  front-end takes a lot of time (compared to the back-end), we might
+  better not report it with the premise that it is a feedback cycle
+  that cannot be parallelized by \sys.}
 
 \heading{Discussion}
+
+Since the front-end contains a feedback cycle, \sys \todo{doesn't
+  parallelize it}, and in both cases it takes \todo{X seconds} to
+complete. However, \sys parallelizes the back-end and achieves a
+speedup of \todo{X} with parallelism set to \todo{2/16}. Several
+stages of the pipeline are \sta (such as the \ttt{word-stem} script
+and the text extraction) and that allows \sys to achieve benefits by
+exposing data parallelism. Note that the back-end part of the script
+contains 34 stages executing in parallel, so the sequential version
+\todo{already benefits from task-based parallelism} \kk{not totally
+  true}.
 

### 2020-05-08 pldi/draft.tex
More intro

--- 2020-05-06 pldi/draft.tex
@@
 The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
-This language agnosticism, coupled with \unix's toolbox philosophy~\cite{mcIlroymemo}, makes the shell the primary choice for specifying succinct and simple pipelines that target data processing, system orchestration, and other automation tasks.
+This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple pipelines that target data processing, system orchestration, and other automation tasks.
 % Pipe-centric shell commands are parallel (and to some extent, functional) programs that can natively enjoy significant parallelism on a multicore machine.
-Unfortunately, parallelizing such pipelines requires significant effort and is shared unevenly between two different groups (or roles) of a programmer. % it's more like personas or "wearing hats"
-
-The first group is \emph{command developers} responsible for developing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
-These developers usually work in a single---often low-level---programming language and manually incorporate parallelism through means available in their chosen language, whenever this is possible.
-%  by the problem and language.
-% the language and abstractions they are using focused at a particular task 
-% also: target a small problem but aim at high reusability 
-As they have no visibility into the command's uses, they expose a plethora of ad-hoc command-specific flags such as \ttt{-t}~\cite{}, \ttt{--parallel}~\cite{}, \ttt{-p}, \ttt{CPUs}, and \ttt{NUM\_THREADS}.
-
-The second group is \emph{command users}, composing scripts out of individual commands.
-% Unfortunately, users do not operate in a single language
-As their programs combine a variety of commands written in many languages, they are left with only a few manual options for achieving parallelism:
+Unfortunately, parallelizing such pipelines requires significant effort shared between two different programmer groups. % it's more like roles personas or "wearing hats"
+
+The first group is \emph{command developers}, responsible for developing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
+These developers usually work in a single programming language, leveraging its abstractions to provide parallelism whenever possible.
+As they have no visibility into the command's uses, they expose a plethora of ad-hoc command-specific flags such as \ttt{-t}, \ttt{-}\ttt{-parallel}, \ttt{-p}, and \ttt{-j} % , \ttt{CPUs}, and \ttt{NUM\_THREADS}~\cite{}.
+\cite{pasetto2011comparative, mcilroy1993engineering, stallman1991gnu}.
+
+The second group is \emph{command users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}).
+% If versed into parallelism, 
+% Users versed in computing can apply 
+% these tools are command-unaware thus at risk of breaking program semantics,
+% require manual changes to the code to make parallelism explicit,
+% at times leverage command flags like the ones mentioned earlier---but each command 
+% oblivious? agnostic?  
+One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{tsc}~\cite{}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
+  these options are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus applying at the level of entire scripts rather than individual components.
+% Tools will attempt to parallelize anything 
 % use the composition abstractions of the \unix shell, 
@@
 % short lived, changing quickly
-either coarse-grained tools such as \eg GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, or manual parallelism POSIX shell primitives and 
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-
-Their last resort is 
-and writing the program from scratch
-Manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
+Another option is the shell's concurrency primitives (\eg \ttt{\&}, \ttt{wait}, \ttt{for}) % for explicit parallelism, 
+ that come at a cost of manual effort to carefully split inputs, manually rewrite scripts, and explicitly orchestrate execution---an expensive and error-prone process.
+% re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
 % The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
-Often, these users are domain experts lacking a deep understanding of parallel systems---in desperate need of automation when it comes to parallelization.
-
-This paper presents a solution targeting the second class of programmers, command users, 
-\sys can automate
-
-A system that is aware of the \unix philosophy, its key abstractions, and offers automated POSIX-aware parallelization as long as 
-These lightweight annotations can be expressed by developers 
-can be expressed 
-This is paired with parallelizability study that informed the design of the system and its annotation interface, and led to the parallelizability semantics for POSIX and GNU coreutils---a large class of commands that serves as the standard library.
+And all these options assume a good understanding of parallelism;
+  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options.% automation when it comes to parallelization.
+
+To address this challenge, we develop \sys, a shell transforming POSIX shell scripts to their parallel equivalents.
+% \sys offers both classes of users in the second group such much-needed automation.
+\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
+Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
+  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
+Command users are provided full automation:
+  \sys extracts latent parallelism from scripts as long as it understands their parallelizability properties.
+% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
+Its transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will not improve performance rather than risking breakage.
+
+To address cold-start issues, \sys comes with a library of parallelizability semantics for POSIX and GNU commands.
+These large classes of commands serve as the shell's standard library and are used pervasively. % in scripts found out in the wild.
+The study that led to the characterization of these commands also informed \sys's annotation language and transformation infrastructure.\footnote{
+  In fact, the scope of our study is somewhat broader as it also incorporates frequently used commands that fall outside GNU Coreutils and POSIX such as \ttt{curl} and \ttt{pandoc}.
+}
+
+The parallelization study and transformations are paired with \sys's runtime component aware of the \unix philosophy and its key abstractions.
+% \sys's runtime  component additional packs
+% Aside from these two components, \sys packs a runtime component that serves dual purpose:
+This component packs a small library of highly-optimized data combiners as well as high-performance primitives for eager data splitting and merging.
+It addresses several practical challenges such as task parallelism, deadlocks, and performance bottlenecks, and was developed by uncovering several pathological situations, on many of which we report.
 
@@
 
-To address this challenge, we develop \sys, a system for transforming POSIX shell scripts to their parallel equivalents.
-\sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
-First, a careful study of the parallelizability properties of shell primitives and commands.
-\sys addresses the challenge by offering command developers a unified set of abstractions for describing the parallelizability characteristics of their commands.
-Using lightweight annotations akin to type signatures, \sys can automate the parallelization of entire POSIX scripts without any input from their user.
-hand-in-hand with the parallelizability study:
-  the study informed the annotation language and was used to annotate the entire
-  for all 
----a set of annotations that \sys already offers for 
-for the standard library
-  akin to the standard library
-  the annotation language is informed by the study, but also the study is 
-XX to a type signature, these annotations describe
-
-Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-(one sentence for the model)
-(one sentence for the optimizations)
-The result of the compilation is still a POSIX-compliant script that manages parallelism explicitly;
-  POSIX compliance means 
-it does not depend on a managed runtime
-
-Third, \sys packs a runtime component 
-A small library of highly-optimized data combiners allows 
-This is paired with a
-it comes with a few runtime primitives for eager data splitting and merging that address several practical issues ranging from deadlocks, performance bottlenecks;
-  these primitives were developed from uncovering several pathological situations, on many of which we report.
-
-We perform an extensive evaluation of \sys using 
-  (i) a series of benchmarks, ranging from classic scripts found in the \unix literature to modern data-processing pipelines;
-  (ii) three large and complex use cases, ranging from temperature analysis to web crawling and indexing; and
-  (iii) a series of micro-benchmarks inspired by (and comparing with) possible alternatives.
+% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
+% First, a careful study of the parallelizability properties of shell primitives and commands.
+% \sys addresses the challenge by offering command developers a unified set of abstractions for describing the parallelizability characteristics of their commands.
+% Using lightweight annotations akin to type signatures, \sys can automate the parallelization of entire POSIX scripts without any input from their user.
+% hand-in-hand with the parallelizability study:
+%   the study informed the annotation language and was used to annotate the entire
+%   for all 
+% ---a set of annotations that \sys already offers for 
+% for the standard library
+%   akin to the standard library
+%   the annotation language is informed by the study, but also the study is 
+% XX to a type signature, these annotations describe 
+% 
+% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
+% (one sentence for the model)
+% (one sentence for the optimizations)
+% The result of the compilation is still a POSIX-compliant script that manages parallelism explicitly;
+%   POSIX compliance means 
+% it does not depend on a managed runtime 
+% 
+% Third, \sys packs a runtime component 
+% A small library of highly-optimized data combiners allows 
+% This is paired with a
+% it comes with a few runtime primitives for eager data splitting and merging that address several practical issues ranging from deadlocks, performance bottlenecks;
+%   these primitives were developed from uncovering several pathological situations, on many of which we report.
+
+We evaluate \sys using
+  (i) a series of benchmarks, ranging from classic one-liners found in the \unix literature to modern data-processing pipelines;
+  (ii) three large and complex use cases, covering temperature analysis, web crawling and indexing, and container orchestration; and
+  (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
 The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements on the performance of individual commands depend on several factors such as the script's execution time, the pipeline depth and the parallelizability class under which commands fall.
+Naturally, improvements on individual commands depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls.
 

### 2020-05-08 pldi/draft.tex
First draft of intro

@@
 \newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
-\newcommand{\ttt}[1]{\texttt{\small #1}}
+\newcommand{\ttt}[1]{\texttt{#1}}
 \newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
@@
 
-The second group is \emph{command users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}).
+The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}).
 % If versed into parallelism, 
@@
 % oblivious? agnostic?  
-One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{tsc}~\cite{}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
-  these options are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus applying at the level of entire scripts rather than individual components.
+One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
+  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus applying at the level of entire scripts rather than individual components.
 % Tools will attempt to parallelize anything 
@@
 
-To address this challenge, we develop \sys, a shell transforming POSIX shell scripts to their parallel equivalents.
+To address this challenge, we develop \sys, a shell parallelizing POSIX shell scripts. % to their parallel equivalents.
 % \sys offers both classes of users in the second group such much-needed automation.
@@
   rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
-Command users are provided full automation:
-  \sys extracts latent parallelism from scripts as long as it understands their parallelizability properties.
+Shell users, on the other hand, are provided full automation:
+  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
 % benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
-Its transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will not improve performance rather than risking breakage.
-
-To address cold-start issues, \sys comes with a library of parallelizability semantics for POSIX and GNU commands.
-These large classes of commands serve as the shell's standard library and are used pervasively. % in scripts found out in the wild.
-The study that led to the characterization of these commands also informed \sys's annotation language and transformation infrastructure.\footnote{
-  In fact, the scope of our study is somewhat broader as it also incorporates frequently used commands that fall outside GNU Coreutils and POSIX such as \ttt{curl} and \ttt{pandoc}.
+\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will simply not improve performance, rather than risking breakage.
+
+To address cold-start issues, \sys comes with a library of parallelizability semantics for commands in POSIX and GNU Coreutils.
+These large classes of commands serve as the shell's standard library, expected to be found pervasively. % in scripts found out in the wild.
+The study that led to their characterization also informed \sys's annotation and transformation components.\footnote{
+  In fact, the scope of the study is broader, incorporating also commands outside POSIX and GNU Coreutils such as \ttt{curl} and \ttt{pandoc}.
 }
 
-The parallelization study and transformations are paired with \sys's runtime component aware of the \unix philosophy and its key abstractions.
+These components are finally paired with \sys's runtime component.
 % \sys's runtime  component additional packs
 % Aside from these two components, \sys packs a runtime component that serves dual purpose:
-This component packs a small library of highly-optimized data combiners as well as high-performance primitives for eager data splitting and merging.
-It addresses several practical challenges such as task parallelism, deadlocks, and performance bottlenecks, and was developed by uncovering several pathological situations, on many of which we report.
+Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combiners as well as high-performance primitives for eager data splitting and merging.
+These address several practical challenges such as task parallelism, deadlocks, and performance bottlenecks, and were developed by uncovering several pathological situations, on many of which we report.
 
@@
   \S\ref{distributability}
-    studies the parallelizability properties of a large class of Shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
+    studies the parallelizability properties of a large class of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
 
@@
   \item
-  \S\ref{impl} describes the \sys implementation and addresses several challenges related to the translation and optimization of shell programs.
+  \S\ref{impl} describes \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
 % \kk{Rewrite that when we know for sure what the section contains.}
@@
 \noindent
-After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
+After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper closes with a discussion of \sys's limitations and possible future directions~\sx{discussion}.
 %micro-bench\-marks %---popular shell one-liners that highlight

### 2020-05-09 pldi/draft.tex
Teensy corrections

@@
 % benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
-\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will simply not improve performance, rather than risking breakage.
+\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.
 
@@
   but it runs on multicore?}
+\nv{Phoenix (i) is shared-memory, and (ii) would require re-writing the entire program in C++, which defeats the purpose of comparing with an existing program.
+Phoenix might be OK for micro-benchmarks however.}
 

### 2020-05-11 pldi/draft.tex
Pass on 3.2

@@
 
+\TODO{Change naming to parallelizability}
+
 To have any hope of distributing pipelines, \sys first needs to understand the distribution characteristics of their individual commands.
@@
 
-\subsection{Distributability Extensibility}
+\subsection{Extensibility}
 \label{ext}
@@
 % FIXME: \nv{TODO: remember to bubble up(me)}
-An important characteristic of the \unix shell is its language-agnostic extensibility~\sx{bg}.
-Users composing pipelines are free to install custom commands available from various sources.
-Being general programs, commands are developed in a variety of languages and are extended over long periods of time.
+An important characteristic of the \unix shell is its
+language-agnostic extensibility~\sx{bg}.  Users composing pipelines
+are free to install custom commands available from various sources.
+Being general programs, commands are developed in a variety of
+languages and are extended over long periods of time.
 %% \kk{Text point: This restricts the space of solutions for
 %%   distribution of shell scripts}
-As a result, any static characterization of shell commands would bs unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
-This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
+As a result, any static characterization of shell commands would bs
+unsatisfactory and would quickly become obsolete, supporting only a
+small subset of the commands needed in practice.  This poses a
+challenge to the developers of individual commands---how can they
+augment their commands to capture information critical to \sys without
+too much effort?
 %% \kk{Text point: The characteristics of a good solution}
 
-To address % additions and changes to the set of commands available to the shell,
-  this challenge, \sys arms command developers with a lightweight, domain-specific language (DSL).
-The DSL allows developers to express semantic information about a command's distributability that is critical to \sys.
-It can be used by the developers of new commands to succinctly express this information, such the class of a command and the sequence of sources from which it reads its input.
-It can also be used by maintainers extending or maintain existing commands, to express additions or changes to the command's implementation or interface.
+To address this challenge, \sys arms command developers with a
+lightweight, parallelizability annotation language (\todo{PAL}) that
+allows devevelopers to express semantic information about a command's
+parallelizability.  The PAL can be used by both developers of new
+commands, as well as maintainers of existing commands that can express
+additions or changes to the command's implmentation or
+interface. \todo{The two main challenges that PAL has to address is
+  supporting commands with mutiple inputs and outputs, and handle
+  command-line options.}
+
+\paragraph{Command Inputs and Outputs}
+
+Together with a command's parallelizability class (i.e. \sta, \pur,
+etc), \sys needs to know certain information about a command's inputs
+and outputs in order to construct a dataflow graph. More precisely,
+given a command and its arguments, \sys needs to know its inputs and
+outputs to connect commands in the dataflow graph with edges, as well
+as the order in which a command consumes its input; this is important
+for \sys to know how to order parallel instances of a command after it
+has performed any \todo{optimizations}. Consider the example of
+\ttt{grep "foo" f1 f2}, which can be parallelized as follows if we
+know that \ttt{grep} first reads from \ttt{f1} and then from \ttt{f2}.
+
+\begin{lstlisting}[float=h, numbers=none]
+  mkfifo t1 t2
+  grep "foo" f1 > t1 &
+  grep "foo" f2 > t2 &
+  cat t1 t2
+\end{lstlisting}
 
 \paragraph{Command-line Options}
-As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
-Options are a particularly popular way to control a command, 
-  directly affecting a command's distributability classification.
-That is, commands are assigned to default classes, and options can change the class in which they belong.
-
-As simple examples, consider \ttt{cat} and \ttt{sort}.
-By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
-Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta because it generates a static stream regardless of other input.
+As described earlier~\sx{bg}, com\-mand-line options and environment
+variables can affect the operation of a command.  Options are a
+particularly popular way to control a command, directly affecting a
+command's parallelizability classification.  That is, commands are
+assigned to default classes, and options can change the class in which
+they belong.  As a simple example, consider \ttt{cat}.  By default
+\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
+has to keep track of a counter and output the line number with each line.
+
+%% As simple examples, consider \ttt{cat} and \ttt{sort}.
+%% By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
+%% Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta because it generates a static stream regardless of other input.
 % TODO: closure?
+
+%% This is actually wrong. sort -h is not stateless since if called several times it produces the same output many times.
 
@@
 
+\tr{
 \begin{figure}
@@
 }
-
   \label{fig:dsl}
 \end{figure}
-
-\paragraph{Command Inputs and Outputs}
-A second concern captured by the DSL is the order of inputs and outputs.
-Commands generally read multiple input streams (and, at cases, write to multiple output streams).
-The invariants between inputs and outputs must be maintained over the distributed execution of the program.
-% but these maintain different invariants, enabling very different distributability characteristics.
-
-Consider the command \ttt{comm} that takes as input two files and
-performs a join operation. When invoked without any option, it
-produces a three-column output. The first column contains lines that
-are unique to the first file, the second column contains lines that are
-unique to the second file, and the last column contains lines that
-exist in both files. In the general case \ttt{comm} is in \pur, since
-it needs to wait until it reads both files completely before it can
-output its results.
-
-However, \ttt{comm} can also be invoked using some options that
-suppress the output of any combination of columns. More precisely, the
-option \ttt{-1} suppresses the first column of the output. Similarly,
-\ttt{-2} and \ttt{-3} supress the second and third column. Using both
-\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command jumps to \sta
-regarding its first (resp. the second) file argument.
+}
+
+%% \paragraph{Command Inputs and Outputs}
+%% A second concern captured by the AL is the order of inputs and outputs.
+%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
+%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
+%% % but these maintain different invariants, enabling very different distributability characteristics.
+
+%% Consider the command \ttt{comm} that takes as input two files and
+%% performs a join operation. When invoked without any option, it
+%% produces a three-column output. The first column contains lines that
+%% are unique to the first file, the second column contains lines that are
+%% unique to the second file, and the last column contains lines that
+%% exist in both files. In the general case \ttt{comm} is in \pur, since
+%% it needs to wait until it reads both files completely before it can
+%% output its results.
+
+%% However, \ttt{comm} can also be invoked using some options that
+%% suppress the output of any combination of columns. More precisely, the
+%% option \ttt{-1} suppresses the first column of the output. Similarly,
+%% \ttt{-2} and \ttt{-3} supress the second and third column. Using both
+%% \ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command jumps to \sta
+%% regarding its first (resp. the second) file argument.
 
@@
 
-\paragraph{Distributability DSL}
-The DSL for distributability description (Fig.~\ref{fig:dsl} \kk{I
-  think that the language definition should be in a technical report,
-  and here we should only give intuition and the \ttt{comm} example})
-is designed to classify commands into multiple categories given their
-options.
-
-It assumes that the arguments of a command can always be separated to
-options and file arguments (checking if they start with a
-\ttt{-}~\footnote{Note that the single character \ttt{-} is not an
-  option as it is most often used to designate that a command reads
-  from its standard input}). A file written in the DSL contains a
-record containing a mapping from predicates to assignments for each
-command. Predicates can be simple boolean formulas on the existence
-and the value of any option. Each assignment contains the command
-category, the sequence of its input sources, and its output
-source. Inputs to commands are a sequence of file identifiers and its
-standard input. Similarly, the output is either its standard output or
-a file identifier. An example record for \ttt{comm} can be seen below:
-
-\begin{lstlisting}[float=h, numbers=none]
+\paragraph{Parallelizability Annotation Language}
+
+To illustrate how the proposed PAL addresses the above challenges, we
+will follow the example of the annotation record for \ttt{comm}, a
+command that performs a ``join'' on its two inputs. When \ttt{comm} is
+invoked without any option, it produces a three-column output. The
+first column contains lines that are unique to the first file, the
+second column contains lines that are unique to the second file, and
+the last column contains lines that exist in both files. A user can
+invoke \ttt{comm} with options \ttt{-1}, \ttt{-2}, or \ttt{-3} to
+suppress some of its output columns.
+%
+\begin{lstlisting}[float=h, numbers=none,caption={PAL record for \ttt{comm}},label={lst:pal-record-comm}]
   comm {
     | -1 and -3 =>
-      (stateless, [args[1]], stdout)
+      (stateless, [args[1]], [stdout])
     | -2 and -3 =>
-      (stateless, [args[0]], stdout)
+      (stateless, [args[0]], [stdout])
     | otherwise =>
-      (pure, [args[0]], stdout)
+      (pure, [args[0], args[1]], [stdout])
   }
@@
 
-The semantics of the language is straightforward: given a command, the
-interpreter collects its options and then returns the assignment of
-the first predicate that is satisfied. The final predicate is always
-satisfied. For example the following two invocations of \ttt{comm}
-would be classified as follows:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- comm -13 f1 f2 # => (stateless, f2, stdout)
- comm f1 f2     # => (pure, f1, stdout)
-\end{lstlisting}
+The PAL annotation record for \ttt{comm} is shown in
+\cref{lst:pal-record-comm}. A record contains a set of clauses, each
+one identified by a predicate on the commands options. Each clause
+maps to an assignment for the command, indicating its class and the
+sequence of its inputs and outputs. \todo{PAL assumes} that arguments
+of a command can be separated to options and file arguments (checking
+if they start with a \ttt{-}~\footnote{Note that the single character
+  \ttt{-} is not an option as it is most often used to designate that
+  a command reads from its standard input}).
+
+In the record for \ttt{comm} the first two clauses describe \ttt{comm}
+invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}. In the
+first case, since the first and third columns are suppressed,
+\ttt{comm} can be considered \sta by regarding its first file argument
+a static ``configuration'' input. \kk{Does this need more
+  explanation?} The general case (that includes \ttt{comm} invocations
+without any options) is captured by the third clause. There,
+\ttt{comm} is classified as \pur, since it needs to wait until it
+reads both files completely before it can output its results. Its
+inputs are its first and second arguments (excluding the options), and
+its output is produced in \ttt{stdout}.
+
+%%  This case is described in the third clause of the annotation
+%% record, where \ttt{comm} is classified as \pur, since it needs to wait
+%% until it reads both files completely before it can output its
+%% results. Its inputs are its first and second arguments (excluding the
+%% options), and its output is produced in \ttt{stdout}.
+
+%% The first and second clauses describe the case when \ttt{comm} is
+%% invoked with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}. In this
+%% case, since the first and third columns are suppressed, \ttt{comm} can
+%% be considered \sta by regarding its first file argument a static
+%% ``configuration'' input. \kk{Does this need more explanation?}
+
+\tr{The complete syntax of the PAL is shown in (Fig.~\ref{fig:dsl})}
+%% \kk{I
+%%   think that the language definition should be in a technical report,
+%%   and here we should only give intuition and the \ttt{comm} example}
+%% is designed to classify commands into multiple categories given their
+%% options.
+
+%% It assumes that the arguments of a command can always be separated to
+%% options and file arguments (checking if they start with a
+%% \ttt{-}~\footnote{Note that the single character \ttt{-} is not an
+%%   option as it is most often used to designate that a command reads
+%%   from its standard input}). A file written in the DSL contains a
+%% record containing a mapping from predicates to assignments for each
+%% command. Predicates can be simple boolean formulas on the existence
+%% and the value of any option. Each assignment contains the command
+%% category, the sequence of its input sources, and its output
+%% source. Inputs to commands are a sequence of file identifiers and its
+%% standard input. Similarly, the output is either its standard output or
+%% a file identifier. An example record for \ttt{comm} can be seen below:
+
+%% \begin{lstlisting}[float=h, numbers=none]
+%%   comm {
+%%     | -1 and -3 =>
+%%       (stateless, [args[1]], stdout)
+%%     | -2 and -3 =>
+%%       (stateless, [args[0]], stdout)
+%%     | otherwise =>
+%%       (pure, [args[0]], stdout)
+%%   }
+%% \end{lstlisting}
+
+%% The semantics of the language is straightforward: given a command, the
+%% interpreter collects its options and then returns the assignment of
+%% the first predicate that is satisfied. The final predicate is always
+%% satisfied. For example the following two invocations of \ttt{comm}
+%% would be classified as follows:
+
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%%  comm -13 f1 f2 # => (stateless, f2, stdout)
+%%  comm f1 f2     # => (pure, f1, stdout)
+%% \end{lstlisting}
 
@@
 
-%% \paragraph{Pure Commands}
-%
-\noindent
+\paragraph{Parallelizable Pure Commands}
+
 In addition to the commands in \sta that can be parallelized in a
@@
 that can be parallelized using divide-and-conquer techniques (as
-mentioned in \cref{cmd}). In order to support the parallelization of
-arbitrary commands in this category, \sys allows command developers to
-provide custom map and reduce steps for them, where the map is in \sta
-and the reduce in \pur, and their composition produces the same output
-with the initial command. \sys can then replace occurrences of these
-commands with their map and reduce parts, enabling their
-parallelization. \TODO{Mention that we have done this for several pure
-  commands in coreutils, providing the foundation}
+mentioned in \cref{cmd}). \kk{Maybe show an example of map-reduce for
+  sort or wc. Either here or in \cref{cmd}.} In order to support the
+parallelization of arbitrary commands in this category, \sys allows
+command developers to provide custom map and reduce steps for them,
+where the map is in \sta and the reduce in \pur, and their composition
+produces the same output with the initial command. \sys can then
+replace occurrences of these commands with their map and reduce parts,
+enabling their parallelization. \todo{We have defined these primitives
+  for most of the parallelizable pure commands in coreutils, providing
+  a foundation for this effort.}
 

### 2020-05-11 pldi/draft.tex
Pass on 4.1

@@
 
-The core of \sys is a dataflow graph model, inspired by the ones used
-in popular distributed stream processing systems~\sx{related} \kk{Is
-  this reference here necessary?}. The dataflow model can express a
-large subset of Shell programs and inherently supports the data
-parallelism found in shell pipelines by exposing it as task
-parallelism. More preciely, we have developed a set of
-parallelization-exposing optimizations represented as
-semantics-preserving graph transformations. These transformations can
-then be applied in an iterative way, optimizing the program to utilize
-available computational resources.
+The core of \sys is a dataflow graph model that is expressive enough
+to support a large subset of shell programs. A fundamental difference
+with other dataflow graph models is that in \sys's model contains
+information about the order in which a node reads its inputs. This
+enables a set of graph transformations that expose data parallelism as
+task parallelism. These transformations can then be applied in an
+iterative way, optimizing the program to utilize available
+computational resources.
+
+%% The core of \sys is a dataflow graph model, inspired by the ones used
+%% in popular distributed stream processing systems~\sx{related} \kk{Is
+%%   this reference here necessary?}. The dataflow model can express a
+%% large subset of Shell programs and inherently supports the data
+%% parallelism found in shell pipelines by exposing it as task
+%% parallelism. More preciely, we have developed a set of
+%% parallelization-exposing optimizations represented as
+%% semantics-preserving graph transformations. These transformations can
+%% then be applied in an iterative way, optimizing the program to utilize
+%% available computational resources.
 
@@
 
-\subsection{Graph Components}
+\subsection{Definitions}
 \label{graph-components}
 
-As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of
-the shell are (i) files (or pipes, \emph{viz.} correspondence between
-files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii)
-commands that communicate through these files. Based on that, \sys's
-graph model defines nodes to represent commands and edges to represent
-files.
-
-\paragraph{Edges---Files}
-
-Before describing the graph edges, we introduce basic notation that
-will be useful later on.  For a set $D$, we write $D\kstar$ to denote
-the set of all finite words over $D$. For words $x, y \in D\kstar$, we
-write $x \cdot y$ or $xy$ to denote their concatenation. We write
-$\eps$ for the empty word. We say that $x$ is a \emph{prefix} of $y$,
-and we write $x \leq y$, if there is a word $z$ such that $y = xz$. It
-is easy to see that $\leq$ is a partial order, often called the
-\emph{prefix order}.
+The two main shell abstractions are (i) files (or pipes, \emph{viz.}
+correspondence between files and pipes in \S\ref{bg:pipelines}) that
+contain data, and (ii) commands that communicate through these
+files. Based on that, in \sys's graph model nodes represent commands
+and edges represent files.
+
+We first introduce basic notation that will be useful later on.  For a
+set $D$, we write $D\kstar$ to denote the set of all finite words over
+$D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to
+denote their concatenation. We write $\eps$ for the empty word. We say
+that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there
+is a word $z$ such that $y = xz$. The $\leq$ order is reflexive,
+antisymmetric, and transitive (i.e. is a partial order) and is often
+called the \emph{prefix order}.
+
+\paragraph{Edges -- Files}
 
@@
 graph. Edges are represented as possibly unbounded streams of type
-$D\kstar$. They could either refer to a specific named file in the file
-system, or just be FIFO pipes used for interprocess communication.
+$D\kstar$. Edges can either refer to named files or FIFO pipes used
+for interprocess communication. Edges that do not start from a node in
+the graph represent the graph input; edges that do not point to a node
+in the graph are its outputs.
 
@@
 
-%% \kk{I think we should delete this since the node and not the edge
-%%   chooses how to write to its output files.}
-
-%% In order to expose data parallelism from the output of a command,
-%% files that do not point to a resource can have
-%% \begin{wrapfigure}[3]{r}{0.20\columnwidth}
-%% \vspace{-7pt}
-%%   \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
-%% \vspace{-20pt}
-%% \end{wrapfigure}
-%% an upper bound on the lines that they can transfer. This helps utilize available
-%% parallelism after encountering a command that cannot be parallelized
-%% (\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.
-
-\tr{On the other hand, some forms of data parallelism can be exposed
-  when knowing the size of the input files. As mentioned in \ref{}
-  some pure commands (such as cat -n) only need line information to
-  become stateless, and knowing the size of a file could allow the
-  system to split it in different chunks that can be processed
-  independently. To account for that, edges that refer to an input
-  resource contain the number of lines of the file that they refer
-  to.}
-
-Finally, edges that do not refer to a resource and do not start
-from a node in the graph represent the graph input;
-edges that do not refer to a resource and do not point to a node in the graph are its outputs.
-
-\paragraph{Nodes---Commands}
+%% \tr{On the other hand, some forms of data parallelism can be exposed
+%%   when knowing the size of the input files. As mentioned in \ref{}
+%%   some pure commands (such as cat -n) only need line information to
+%%   become stateless, and knowing the size of a file could allow the
+%%   system to split it in different chunks that can be processed
+%%   independently. To account for that, edges that refer to an input
+%%   resource contain the number of lines of the file that they refer
+%%   to.}
+
+\paragraph{Nodes -- Commands}
 
@@
 lifting of the prefix order for a sequence of inputs \kk{Formalize
-this lifting if essential for later development}. This captures the
-  idea that a node cannot retract output after it has been emitted.
-%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
-%exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
-%concatenation and will be omitted when obvious.
-An example of a node with one input and one output stream is the
-following command:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- grep ``foo''
-\end{lstlisting}
-
-\noindent
-This command takes a stream of lines from its standard input and
-returns a stream of lines on its standard output.
-
-An important observation is that even though commands might read their
-input from several input streams, the order in which they read is not
-always arbitrary. An illustrative example is the command \ttt{cat},
-which consumes its input one stream at a time:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
- cat x - y
-\end{lstlisting}
-
-\noindent
-In the example above, \ttt{cat} reads file \ttt{x} until it encounters
-the EOF character, then reads from \ttt{stdin} until in encounters
-EOF, and finally reads from \ttt{y}. Other commands that read their
-inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
-order in which a node consumes its input is necessary to enable
-certain optimizations on the graph.
+  this lifting if essential for later development}.
+%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
+%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
+%% concatenation and will be omitted when obvious.
+This captures the idea that a node cannot retract output that it has
+already produced.
+
+\paragraph{\todo{One-input-output} Commands}
+
+\kk{Find a better name.} \todo{One-input-output} commands \todo{are
+  special commands} that consume their inputs one at a time, with the
+possible exception of static input files, and produce one output
+file. These commands can be represented as functions of type $f :
+D\kstar \times [D\kstar] \to D\kstar$, where the first argument
+represents the concatenation of the inputs that are consumed one at a
+time and the second argument represents the static inputs. The
+quintessential example of \todo{one-input-output} commands is
+\ttt{cat}, that consumes its inputs one at a time, producing their
+concatenation as output.
+%% A \todo{large subset} of shell commands fall in this category
+
+\kk{Maybe give an example of a script and its dataflow graph.}
+
+%% An example of a node with one input and one output stream is the
+%% following command:
+
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%%  grep ``foo''
+%% \end{lstlisting}
+
+%% \noindent
+%% This command takes a stream of lines from its standard input and
+%% returns a stream of lines on its standard output.
+
+%% An important observation is that even though commands might read their
+%% input from several input streams, the order in which they read is not
+%% always arbitrary. An illustrative example is the command \ttt{cat},
+%% which consumes its input one stream at a time:
+
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%%  cat x - y
+%% \end{lstlisting}
+
+%% \noindent
+%% In the example above, \ttt{cat} reads file \ttt{x} until it encounters
+%% the EOF character, then reads from \ttt{stdin} until in encounters
+%% EOF, and finally reads from \ttt{y}. Other commands that read their
+%% inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
+%% order in which a node consumes its input is necessary to enable
+%% certain optimizations on the graph.
 
@@
 
-\kk{I need to mention that there are nodes that we know more
-  properties of. For example cat, tee, zip, etc... and these are
-  handled specially because of their very important and special
-  properties.}
+%% \kk{I need to mention that there are nodes that we know more
+%%   properties of. For example cat, tee, zip, etc... and these are
+%%   handled specially because of their very important and special
+%%   properties.}
 
@@
 intuition about stateless and pure commands that was described
-earlier~\sx{cmd}. We focus on parallelizing commands that produce one
-output and consume their inputs one at a time, with the possible
-exception of static input files that are used for configuration. More
-formally, we parallelize commands that can be represented with
-functions of type $f : D\kstar \times [D\kstar] \to D\kstar$, where
-the first argument represents the concatenation of the inputs that are
-consumed one at a time and the second argument represents the static
-inputs. When translating a command that consumes its input streams one
-at a time, a \ttt{cat} node is introduced before it to concatenate the
-streams and its output is transferred to the command input. \kk{Maybe
+earlier~\sx{cmd}.
+%% \nv{TODOs
+%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
+%% }
+
+\kk{Maybe
   have an argument here that we show in the evaluation that handling
@@
   microbenchmarks to some large-realistic macrobenchmarks.}.
-%% \nv{TODOs
-%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
-%% }
 

### 2020-05-11 pldi/draft.tex
Pass over section 4 done

@@
 
+\kk{Do we want to get in more detail about requirements? Commands must
+  be deterministic, they must not do any other side effect (sending
+  signals, etc). These assumptions must be checked when the developer
+  designates the categories. Commands are assumed to always terminate
+  gracefully. \sys changes the termination behaviour of commands as a
+  sequential command that would exit prematurely might not exit
+  completely if parallelized.}
+
 \paragraph{Pure Commands}
@@
 Such constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $reduce$ function.
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $fold$ function.
 
@@
 that can be parallelized using divide-and-conquer techniques (as
-mentioned in \cref{cmd}). \kk{Maybe show an example of map-reduce for
+mentioned in \cref{cmd}). \kk{Maybe show an example of map-fold for
   sort or wc. Either here or in \cref{cmd}.} In order to support the
 parallelization of arbitrary commands in this category, \sys allows
-command developers to provide custom map and reduce steps for them,
-where the map is in \sta and the reduce in \pur, and their composition
+command developers to provide custom map and fold steps for them,
+where the map is in \sta and the fold in \pur, and their composition
 produces the same output with the initial command. \sys can then
-replace occurrences of these commands with their map and reduce parts,
+replace occurrences of these commands with their map and fold parts,
 enabling their parallelization. \todo{We have defined these primitives
@@
 \ttt{cat}, that consumes its inputs one at a time, producing their
-concatenation as output.
-%% A \todo{large subset} of shell commands fall in this category
+concatenation as output. We focus on these commands because a large
+subset of \sta and \pur commands fall in this category.
 
@@
 
-Having a dataflow graph representation allows us to define
-parallelization optimizations as graph transformations. Before
-describing the different types of optimizations, we formalize the
-intuition about stateless and pure commands that was described
+We have defined a set of semantics-preserving graph transformations
+that act as parallelization-exposing optimizations. Both the domain
+and range of these transformations is a graph in our dataflow model
+and therefore they can be composed arbitrarily and in any order.
+%% Having a dataflow graph representation allows us to define
+%% parallelization optimizations as graph transformations.
+Before describing the different types of optimizations, we formalize
+the intuition about stateless and pure commands that was described
 earlier~\sx{cmd}.
+
+%% We focus on \todo{one-input-output} commands (as defined in
+%% \cref{graph-components}) that constitute a large subset of shell
+%% commands.
+
 %% \nv{TODOs
@@
 
-\kk{Maybe
-  have an argument here that we show in the evaluation that handling
-  these commands is adequate for a big variety of data processing
-  scripts (and most stateless and pure commands), from several
-  microbenchmarks to some large-realistic macrobenchmarks.}.
-
-As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr} operate
-independently on individual elements of the stream (characters, lines,
-or files) without maintaining any state. In this work, we consider
-lines as the data quantum, so commands that are file-stateless are not
-considered stateless. Formally, a command $f$ is stateless if it is
-``linear'' with respect to concatenation, \ie satisfying the following
-equation:
+%% \kk{Maybe
+%%   have an argument here that we show in the evaluation that handling
+%%   these commands is adequate for a big variety of data processing
+%%   scripts (and most stateless and pure commands), from several
+%%   microbenchmarks to some large-realistic macrobenchmarks.}.
+
+\paragraph{Stateless and Parallelizable Pure Commands}
+
+As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
+operate independently on individual elements of the stream
+(characters, lines, or files) without maintaining any state. In this
+work, we consider lines as the data quantum, so only commands that are
+stateless regarding lines (or any finer granularity such as
+characters) are considered stateless. Formally, a
+\todo{one-input-output} command $f$ is stateless if it is ``linear''
+with respect to concatenation, \ie satisfying the following equation:
 
@@
 \noindent
-This means that one gets the same output by applying a command $f$ to a
-concatenation of two inputs $x, x'$ as by concatenating the result
-produced by the application of $f$ to these inputs
-separately.
-
-Similarly, it was mentioned that some pure commands (such
-as \ttt{sort} and \ttt{wc}) can be parallelized using
-divide-and-conquer techniques. More formally, these pure commands $f$
-can be implemented as a combination of functions map $m$ and reduce $r$, satisfying
-the following equation:
+This means that applying the command $f$ to a concatenation of two
+inputs $x, x'$ produces the same output as applying $f$ to each input
+$x, x'$ separately, and concatenating the outputs.
+
+Similarly, some pure commands (such as \ttt{sort} and \ttt{wc}) can be
+parallelized using divide-and-conquer techniques. More formally, these
+pure commands $f$ can be implemented as a combination of functions map
+$m$ and an associative fold $f'$, satisfying the following equation:
 
 \[
-\forall x, x', s, f(x.x', s) = r(m(x, s).m(x', s), s)
+\forall x, x', s, f(x.x', s) = f'(m(x, s).m(x', s), s)
 \]
@@
 \noindent
-Similarly to the earlier equation, this means that we get the same output by
-applying $f$ to a concatenation of two inputs $x, x'$, or by applying
-the reduce function $r$ to the concatenation of the outputs produced
-by applying map $m$ to each of these inputs.
+This means that we get the same output by applying $f$ to a
+concatenation of two inputs $x, x'$, or by applying the fold
+function $f'$ to the concatenation of the outputs produced by applying
+map $m$ to each of these inputs.
 
@@
 graph. Furthermore, it can be straightforwardly extended to pure
-commands that can be implemented by a map-reduce pair $(m, r)$ as $
+commands that can be implemented by a map-fold pair $(m, r)$ as $
 v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
@@
 \centering
-\includegraphics[width=0.60\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
+\includegraphics[width=\columnwidth]{\detokenize{./figs/transformation-sketch.png}}
 \caption{
@@
 
-\kk{Remember to mention the assumptions that need to hold for the
-  graph transformations to be valid in the Command categories
-  section. Commands must be deterministic, they must not do any other
-  side effect (such as writing to other files, sending signals,
-  etc). However, these assumptions must already be checked when the
-  developer designates the categories. Commands are assumed to always
-  terminate gracefully. \sys changes the termination behaviour of
-  commands as a sequential command that would exit prematurely might
-  not exit completely if parallelized.}
+%% \kk{Remember to mention the assumptions that need to hold for the
+%%   graph transformations to be valid in the Command categories
+%%   section. Commands must be deterministic, they must not do any other
+%%   side effect (such as writing to other files, sending signals,
+%%   etc). However, these assumptions must already be checked when the
+%%   developer designates the categories. Commands are assumed to always
+%%   terminate gracefully. \sys changes the termination behaviour of
+%%   commands as a sequential command that would exit prematurely might
+%%   not exit completely if parallelized.}
 
@@
 
-\kk{I think that we should probably rewrite this paragraph with the
-  ``new'' split optimization (and maybe any others that we develop
-  until then) here.}
-
-In addition to node splits, we can also perform edge splits in order
-to increase the number of incoming edges for some nodes to further expose
-possible data parallelism. In cases where the edge refers to a real file, the
-edge split represents splitting the real file in chunks. In
-cases where the edge connects two nodes, the split effectively refers to several edges
-with a bounded number of lines. Note that the final output edge of a
-valid node in the graph must always be unbounded if it does not refer
-to a file.
+\kk{Maybe say something about the rest of the transformations: (i)
+  changing a command with multiple inputs to a cat followed by the
+  command, (ii) Adding an identity node in the place of two edges, and
+  (iii) adding a split after a command.}
+
+%% \kk{I think that we should probably rewrite this paragraph with the
+%%   ``new'' split optimization (and maybe any others that we develop
+%%   until then) here.}
+
+%% In addition to node splits, we can also perform edge splits in order
+%% to increase the number of incoming edges for some nodes to further expose
+%% possible data parallelism. In cases where the edge refers to a real file, the
+%% edge split represents splitting the real file in chunks. In
+%% cases where the edge connects two nodes, the split effectively refers to several edges
+%% with a bounded number of lines. Note that the final output edge of a
+%% valid node in the graph must always be unbounded if it does not refer
+%% to a file.
 

### 2020-05-11 pldi/draft.tex
Add comments in intro. Overall I really like the structure and narrative. The separation between users and developers is convincing and brings the automation part in light. Nevertheless, I think it needs some polishing in the word/sentence level.

@@
 The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}).
+\kk{I don't really like the reference to such a later section.
+Can we remove it?}
 % If versed into parallelism, 
@@
 One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
-  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus applying at the level of entire scripts rather than individual components.
+  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained \kk{i.e.
+only being capable to exploit parallelism in the level of entire scripts} thus applying at the level of entire scripts rather than individual components.
 % Tools will attempt to parallelize anything 
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is the shell's concurrency primitives (\eg \ttt{\&}, \ttt{wait}, \ttt{for}) % for explicit parallelism, 
+Another option is using the shell's concurrency primitives (\eg \ttt{\&}, \ttt{wait}, \ttt{for}) \kk{Is \ttt{for} a concurrency primitive?} % for explicit parallelism, 
  that come at a cost of manual effort to carefully split inputs, manually rewrite scripts, and explicitly orchestrate execution---an expensive and error-prone process.
@@
 % The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
-And all these options assume a good understanding of parallelism;
-  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options.% automation when it comes to parallelization.
-
-To address this challenge, we develop \sys, a shell parallelizing POSIX shell scripts. % to their parallel equivalents.
+And\todo{/To top it off} all these options assume a good understanding of parallelism;
+  users whose domain of expertise falls outside computing---\todo{routinely} writing shell scripts to accomplish their tasks---are left without options.% automation when it comes to parallelization.
+
+To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts. % to their parallel equivalents.
 % \sys offers both classes of users in the second group such much-needed automation.
@@
 % Aside from these two components, \sys packs a runtime component that serves dual purpose:
-Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combiners as well as high-performance primitives for eager data splitting and merging.
-These address several practical challenges such as task parallelism, deadlocks, and performance bottlenecks, and were developed by uncovering several pathological situations, on many of which we report.
+Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combiners\todo{/combinators} as well as high-performance primitives for eager data splitting and merging.
+These address several practical challenges such as task parallelism, deadlocks, and performance bottlenecks\kk{I like the flow of this sentence but I am not sure it makes a lot of sense to name these challenges here.}, and were developed by uncovering several pathological situations, on many of which we report.
 
@@
 The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
+\kk{I would remove the ``aided further...'' part.}
 Naturally, improvements on individual commands depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls.
+\kk{Do we need to mention this last sentence here?
+Without context it is pretty difficult for anyone to get something out of it.}
 

### 2020-05-11 pldi/draft.tex
Comments on dangling fifos and zombie producers

@@
 
-To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts. % to their parallel equivalents.
+To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
+\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
 % \sys offers both classes of users in the second group such much-needed automation.
@@
 
+\kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}
+
 As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
@@
 
-\begin{lstlisting}[float=h, numbers=none]
+\begin{lstlisting}[language=sh, numbers=none]
   mkfifo t1 t2
@@
 
-\kk{Maybe say something about the rest of the transformations: (i)
-  changing a command with multiple inputs to a cat followed by the
-  command, (ii) Adding an identity node in the place of two edges, and
-  (iii) adding a split after a command.}
+\kk{I think we should say something about the rest of the
+  transformations: (i) changing a command with multiple inputs to a
+  cat followed by the command to enable the previous optimization,
+  (ii) adding a split node after a command (without saying anything
+  about its implementation for now, and (iii) Adding an identity node
+  in the place of an edge.}
 
@@
 applying transformation on parallelizable regions~\sx{front-end}.  It
-closes with the \sys optimizer, which optimizes the dataflow graph
-exposing data-parallelism, and outputs an optimized
-parallel script~\sx{optimizer}.
+closes with the \sys runtime\todo{/backend} and the technical
+challenges that were addressed to achieve performance
+benefits~\sx{optimizer}.
 
@@
 
-\subsection{Optimizer}
+\subsection{Runtime/Back End}
 \label{optimizer}
-
-\kk{This subsection has to only talk about the cool shell constructs
-  that are added and about optimizations that are not in the dataflow
-  graph (like file splitting), and the implementation of split, the
-  pure merges , etc...}
 
@@
 
-Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.
-More concretely, the
+Given a dataflow graph representing a shell program, the \sys
+optimizer outputs an optimized dataflow graph in which data
+parallelism opportunities have been exploited.  More concretely, the
 optimizer starts by splitting input files in several chunks, given the
@@
 nodes of the graph, iteratively performing the graph transformations
-that were described in \Cref{ir:transformations}. If the incoming
-edges of the node are less than the desirable graph width, it performs
-edge splits, and then it performs maximal node splits. The final graph
-can then be implemented by spawning a process for each node, and
-redirecting the outputs according to the graph edges.
-
-The optimizer also applies custom transformations on several commands in \pur
-to further improve their performance. Two notable examples are the
-transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
-that have more than one input are tranformed to \ttt{sort} commands on
-each input, followed by a \ttt{sort -m}, to produce the final sorted
-result. The transformation on occurences of \ttt{wc} that have more
-than input \ttt{i1, t2} is illustrated below as a shell script:
-% \nv{do not understand i1, t2}
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  paste -d '+' \
-   <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
-   <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
-   bc | tr -s '\n' ' ' |
-   sed 's/^/   /' | sed 's/$/\ /'
-\end{lstlisting}
-
-The above script takes the outputs produced by two \ttt{wc} commands
-and aggregates them accordingly.
+that were described in \Cref{ir:transformations}.
+%% If the incoming edges of the node are less than the desirable graph
+%% width, it performs edge splits, and then it performs maximal node
+%% splits.
+The final graph can then be implemented by spawning a process for each
+node, and redirecting the inputs and outputs according to the graph
+edges.
 
@@
 
-\subsection{Implementation Nuggets}
+%% \subsection{Implementation Nuggets}
 
@@
 % A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
+Under normal operation, a producer exits after it has produced and
+sent all of its output to a pipe\todo{/its output channel}.
 % \ttt{SIGPIPE} 
-However, if its consumer exits early, the producer needs to be notified so that it stops appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.
-
+However, if its consumer exits early, the producer needs to be
+notified so that it stops \todo{appending to the stream/trying to push
+  data to its output channel}.
+% ---so that it avoids unnecessary work and buffering challenges in the kernel.
 In \unix, this is achieved by an out-of-band error mechanism:
@@
 % write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
-Consider the following emitted code from \sys's earlier versions:
+Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g.
+\ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
+Consider the following emitted code from \sys's earlier versions:\kk{Proposed alternative: ``Consider the following shell script:''}
 \begin{lstlisting}[language=sh, numbers=none]
+  mkfifo fifo1 fifo2
   cat file-chunk1.txt > fifo1 &
@@
 % execution. All of this can be easily generated by the compiler.
+
+\paragraph{Pure Fold Function Implementations}
+
+\kk{Maybe say something about some of the pure fold functions that we
+  implemented. Interesting examples include sort and wc.}
+
+%% The optimizer also applies custom transformations on several commands in \pur
+%% to further improve their performance. Two notable examples are the
+%% transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
+%% that have more than one input are tranformed to \ttt{sort} commands on
+%% each input, followed by a \ttt{sort -m}, to produce the final sorted
+%% result. The transformation on occurences of \ttt{wc} that have more
+%% than input \ttt{i1, t2} is illustrated below as a shell script:
+%% % \nv{do not understand i1, t2}
+
+%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%%   paste -d '+' \
+%%    <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
+%%    <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
+%%    bc | tr -s '\n' ' ' |
+%%    sed 's/^/   /' | sed 's/$/\ /'
+%% \end{lstlisting}
+
+%% The above script takes the outputs produced by two \ttt{wc} commands
+%% and aggregates them accordingly.
 

### 2020-05-11 pldi/draft.tex
Pass on 5.1

@@
 \sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.
+\kk{Should we mention that we only show this empirically?}
 
@@
 \kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}
+
+\kk{Make sure for the running script to contain more than just a
+  pipeline, so that it is clear that \sys turns a script into the same
+  script with DFGs in place of ``parallelizable-regions'' (which are
+  usually pipelines).}
 
@@
 without affecting the program behaviour---\eg commands connected with
-a \ttt{\&\&} operator~\sx{bg}. In addition, the shell's highly dynamic
-nature compounds the challenge---\eg by not knowing what files does an
-unexpanded command argument refer to.
-% of correctly identifying what parts can be distributed, e.g.
+a \ttt{\&\&} operator~\sx{bg}.
+%% In addition, the shell's highly dynamic
+%% nature compounds the challenge---\eg by not knowing what files does an
+%% unexpanded command argument refer to.
+To address this issue, we introduce the notion of parallelizable
+program regions and a translation pass for converting them to dataflow
+dataflow graphs.
 
@@
 
-To address the first issue, we introduce the notion of parallelizable
-program regions and a translation pass for converting them to a
-dataflow graph. To address the second issue, we limit dynamic
-components by deferring translation as late as possible using a
-wrapping scheme.
+%% To address the first issue, we introduce the notion of parallelizable
+%% program regions and a translation pass for converting them to a
+%% dataflow graph. To address the second issue, we limit dynamic
+%% components by deferring translation as late as possible using a
+%% wrapping scheme.
 
@@
 \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat f1 f2 | grep ``foo'' > f3 && sort f3 
+  cat f1 f2 | grep "foo" > f3 && sort f3 
 \end{lstlisting}
@@
 \noindent
-The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would execute as
+The commands \ttt{cat f1 f2} and \ttt{grep "foo"} would execute as
 independent processes in the standard shell, while \ttt{sort f3} would
 wait for their completion before executing.  Both \ttt{cat f1 f2 |
-  grep ``foo'' > f3} and \ttt{sort f3} are therefore parallelizable
+  grep "foo" > f3} and \ttt{sort f3} are therefore parallelizable
 regions, regions that cannot be extended beyond \ttt{\&\&}---even
@@
 pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
-Conversely, the \ttt{;} sequential composition operator, the
-\ttt{\&\&} logical operator, and function definitions represent
+Conversely, all other operators, such as the \ttt{;} sequential
+composition operator, the \ttt{\&\&} logical operator represent
 barrier constructs that do not allow parallelizable regions to
@@
 
-\kk{It might be a good idea to change this subsection based on the new
-  as-late-as-possible architecture where everything is wrapped and
-  translated at the final moment. Ideally, this would become the main
-  point of this subsection, as the translation algorithm is kind of
-  redundant when having the distributable regions too.}
+%% \kk{It might be a good idea to change this subsection based on the new
+%%   as-late-as-possible architecture where everything is wrapped and
+%%   translated at the final moment. Ideally, this would become the main
+%%   point of this subsection, as the translation algorithm is kind of
+%%   redundant when having the distributable regions too.}
 
 The \sys front-end performs a depth-first search pass on the AST of
-the given shell program.  During this pass, it greedily extends the
-distributable regions from the bottom up, translating their
+the given shell program.  During this pass, it extends the
+parallelizable regions from the bottom up, translating their
 independent components to dataflow graph nodes until a barrier
-construct is reached.  To identify possible opportunities for
-parallelism, the translation pass also extracts the category of each
-node together with its input and output pairs.  To perform this
-extraction, it uses the command name to index into a key-value map
-with values that are expressed in the DSL for distributability
-characterization~\sx{ext}.
-
-A few illustrative cases of the analysis algorithm are shown in pseudocode below:
-
-\begin{lstlisting}[language=python, float=h]
-  def translate(node):
-    ...
-    elif(node.name == 'Pipe'):
-      return pipe_graphs([translate(child)
-        for child in node.children])
-    elif(node.name == 'And'):
-      node.children = [translate(child)
-        for child in node.children]
-      return node
-    elif(node.name == 'Command'):
-      if(not safe(node)):
-        return make_ir_node(node, 'unsafe')
-      else:
-        return make_ir_node(node,
-          find_category(node))
-    ...
-\end{lstlisting}
-
-Whenever the algorithm encounters a distributable node, it recursively
-translates its subcomponents.
-It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
-% what first and second? we didn't say anything about first and second
-Before the different dataflow graphs are connected, the
-algorithm checks that at most one node in the graph writes at every
-file, and that at most one node reads from a file that another node writes
-to. This is important, in order to avoid inconsistent behaviour after
-distributing nodes in the dataflow graph (\ie by introducing to
-concurrent reads and writes to the same file when
-parallelizing).
-
-Whenever the algorithm encounters a barrier node,
-it recursively translates its subcomponents, similarly to the distributable
-case.
-However, different from the distributable case, it does not merge them in a large dataflow graph.
-
-Finally, whenever the algorithm encounters a command, it first
-checks whether it is safe to parallelize it. A command is considered
-safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+construct is reached. All subtrees that are not translated to dataflow
+graphs are kept as they are. The output of the translation pass is the
+original AST where parallelizable regions have been replaced with
+dataflow graphs and a call to \sys's runtime.
+
+To identify possible opportunities for parallelism, the translation
+pass extracts the parallelizability class of each command together
+with its inputs and outputs. It achieves that by searching all
+available PAL~\sx{ext} records for each command and resorts to
+conservative defaults if a record for the command is not found. If the
+command is \sta or parallelizable \pur, the translation pass initiates
+a parallelizable region that is propagated up the tree.
+
+%% Finally, whenever the algorithm encounters a command, it first
+%% checks whether it is safe to parallelize it. A command is considered
+%% safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
 % why couldn't we replicate this?
-If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
+%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
+%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
+
+%% A few illustrative cases of the analysis algorithm are shown in pseudocode below:
+
+%% \begin{lstlisting}[language=python, float=h]
+%%   def translate(node):
+%%     ...
+%%     elif(node.name == 'Pipe'):
+%%       return pipe_graphs([translate(child)
+%%         for child in node.children])
+%%     elif(node.name == 'And'):
+%%       node.children = [translate(child)
+%%         for child in node.children]
+%%       return node
+%%     elif(node.name == 'Command'):
+%%       if(not safe(node)):
+%%         return make_ir_node(node, 'unsafe')
+%%       else:
+%%         return make_ir_node(node,
+%%           find_category(node))
+%%     ...
+%% \end{lstlisting}
+
+%% Whenever the algorithm encounters a distributable node, it recursively
+%% translates its subcomponents.
+%% It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
+%% % what first and second? we didn't say anything about first and second
+%% Before the different dataflow graphs are connected, the
+%% algorithm checks that at most one node in the graph writes at every
+%% file, and that at most one node reads from a file that another node writes
+%% to. This is important, in order to avoid inconsistent behaviour after
+%% distributing nodes in the dataflow graph (\ie by introducing to
+%% concurrent reads and writes to the same file when
+%% parallelizing).
+
+%% Whenever the algorithm encounters a barrier node,
+%% it recursively translates its subcomponents, similarly to the distributable
+%% case.
+%% However, different from the distributable case, it does not merge them in a large dataflow graph.
+
+%% Finally, whenever the algorithm encounters a command, it first
+%% checks whether it is safe to parallelize it. A command is considered
+%% safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+% why couldn't we replicate this?
+%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
+%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
 
 An important consideration is that due to the highly dynamic nature of
-the shell, the safety properties that are mentioned above cannot be
-soundly checked statically.  For safety purposes, \sys takes a
-conservative approach where it does not parallelize nodes without
-complete information.  For example, the translation pass cannot infer
-that two un-expand strings do not refer to the same file identifier,
+the shell, there is information (such as the values of environment
+variables) that are not known to \sys at the time of translation.
+%% to the the safety properties that are mentioned above cannot be
+%% soundly checked statically.
+For safety purposes, \sys takes a conservative approach where it does
+not parallelize nodes without complete information.  For example, the
+translation pass cannot infer that an environment variable that is
+given as an argument to a command is not a flag that changes its
+parallelizability class,
+%% two un-expand strings do not refer to the same file identifier,
 and as a result does not parallelize this sub-expression.
+
+\kk{Can we say anything more about the safety considerations? That (i)
+  the translation pass ``checks'' that no nodes write/read or
+  write/write to the same file (is that necessary?), that (ii) a
+  command should not read from a read-once file. (is that necessary?)}
 
@@
 
-\kk{Maybe guide all of this by showing an example of an output shell
-  script?}
+%% \kk{Maybe guide all of this by showing an example of an output shell
+%%   script?}
 
@@
   heavy pipelines.}
+
+\TODO{Should we mention that we empirically show that pash produces
+  correct code?}
 

### 2020-05-12 pldi/draft.tex
Turn all paragraph to heading

@@
 
-% \paragraph{Generation}
+% \heading{Generation}
 % Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
@@
 
-% \paragraph{Preliminaries}
-% \paragraph{POSIX, GNU Core-utils, and beyond}
-
-\paragraph{Stateless Commands}
+% \heading{Preliminaries}
+% \heading{POSIX, GNU Core-utils, and beyond}
+
+\heading{Stateless Commands}
 The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
@@
 
-\paragraph{Pure Commands}
+\heading{Pure Commands}
 The second class contains pure commands \pur.
@@
 
-\paragraph{File-system Commands}
+\heading{File-system Commands}
 The third class contains commands that access the file-system~\dfs.
@@
 
-\paragraph{Side-effectful Commands}
+\heading{Side-effectful Commands}
 Another large class of commands are ones that have side-effects to memory~\sid.
@@
 
-\paragraph{Irreversibly Side-effectful Commands}
+\heading{Irreversibly Side-effectful Commands}
 Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
@@
 
-\paragraph{Discussion}
+\heading{Discussion}
 There are a few technical details about \sys's design decisions worth noting.
@@
 
-\paragraph{Command Inputs and Outputs}
+\heading{Command Inputs and Outputs}
 
@@
 
-\paragraph{Command-line Options}
+\heading{Command-line Options}
 As described earlier~\sx{bg}, com\-mand-line options and environment
@@
 
-%% \paragraph{Command Inputs and Outputs}
+%% \heading{Command Inputs and Outputs}
 %% A second concern captured by the AL is the order of inputs and outputs.
@@
 
-\paragraph{Parallelizability Annotation Language}
+\heading{Parallelizability Annotation Language}
 
@@
 
-\paragraph{Parallelizable Pure Commands}
+\heading{Parallelizable Pure Commands}
 
@@
 
-\paragraph{Edges -- Files}
+\heading{Edges -- Files}
 
@@
 
-\paragraph{Nodes -- Commands}
+\heading{Nodes -- Commands}
 
@@
 
-\paragraph{\todo{One-input-output} Commands}
+\heading{\todo{One-input-output} Commands}
 
@@
 
-\paragraph{Stateless and Parallelizable Pure Commands}
+\heading{Stateless and Parallelizable Pure Commands}
 
@@
 
-\paragraph{Parallelization Transformations}
+\heading{Parallelization Transformations}
 %
@@
 
-\paragraph{Parallelizable Regions}
+\heading{Parallelizable Regions}
 %
@@
 
-\paragraph{Translation Pass}
+\heading{Translation Pass}
 
@@
 
-\paragraph{Pure Fold Function Implementations}
+\heading{Pure Fold Function Implementations}
 
@@
 
-\paragraph{Distributed Operating Systems and Shells}
+\heading{Distributed Operating Systems and Shells}
 A plethora of systems assist in the construction of distributed
@@
 
-\paragraph{Cluster Computing}
+\heading{Cluster Computing}
 Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
@@
 
-\paragraph{Annotation-based Distribution}
+\heading{Annotation-based Distribution}
 Recent systems on light-touch distribution, such as
@@
 % Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\paragraph{Low-level Parallelization}
+\heading{Low-level Parallelization}
 Instruction-level parallelization has a longer history, starting from
@@
 
-\paragraph{Correct Distribution of Dataflow Graphs}
+\heading{Correct Distribution of Dataflow Graphs}
 %
@@
 
-\paragraph{Limitations}
+\heading{Limitations}
 Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
@@
 
-\paragraph{Future Work}
+\heading{Future Work}
 % \label{limitation}
@@
 
-\paragraph{Conclusion}
+\heading{Conclusion}
 This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 

### 2020-05-14 compiler/gather_results.py
Update table and paper plots

@@ -138,6 +138,18 @@ def read_distr_total_compilation_time(filename):
         print("!! WARNING: Filename:", filename, "not found!!!")
         return 0
 
+def read_distr_command_number(filename):
+    try:
+        with open(filename) as f:
+            for line in f:
+                if(line.startswith("Total nodes after optimization:")):
+                    number_of_commands = int(line.split(": ")[1])
+                    return number_of_commands
+            return "\\todo{X}"
+    except:
+        print("!! WARNING: Filename:", filename, "not found!!!")
+        return "\\todo{X}"
+
 def check_output_diff_correctness_for_experiment(filename):
     try:
         with open(filename) as f:
@@ -165,20 +177,8 @@ def collect_experiment_scaleup_times(prefix, scaleup_numbers):
                        for n in scaleup_numbers]
     return (seq_numbers, distr_numbers, compile_numbers)
 
-# def collect_experiment_no_eager_times(prefix, scaleup_numbers):
-#     no_eager_distr_numbers = collect_distr_experiment_execution_times(prefix, 'distr_no_eager.time',
-#                                                                       scaleup_numbers)
-#     no_task_par_eager_distr_numbers = collect_distr_experiment_execution_times(prefix,
-#                                                                                'distr_no_task_par_eager.time',
-#                                                                                scaleup_numbers)
-#     return (no_eager_distr_numbers, no_task_par_eager_distr_numbers)
-
 def collect_experiment_speedups(prefix, scaleup_numbers):
     seq_numbers, distr_numbers, compile_numbers = collect_experiment_scaleup_times(prefix, scaleup_numbers)
-    # print(scaleup_numbers)
-    # print(seq_numbers)
-    # print(distr_numbers)
-    # print(compile_numbers)
     distr_speedup = [safe_zero_div(seq_numbers[i], t) for i, t in enumerate(distr_numbers)]
     compile_distr_speedup = [safe_zero_div(seq_numbers[i], t + compile_numbers[i]) for i, t in enumerate(distr_numbers)]
     return (distr_speedup, compile_distr_speedup)
@@ -189,23 +189,16 @@ def collect_baseline_experiment_speedups(prefix, scaleup_numbers, base_seq):
     speedup = [safe_zero_div(base_seq, t) for t in seq_numbers]
     return speedup
 
-
[…]

### 2020-05-14 pldi/microbenchmarks-table.tex
Add new plots and rows in paper
Update table and paper plots

--- 2020-04-12 pldi/microbenchmarks-table.tex
@@
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(\todo{20, 100}) &Compile Time (\todo{20, 100}) & Highlights \\
+Script ~&~ Structure & Input &Seq Time & \#Nodes(16, 64) &Compile Time (16, 64) & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 4809.61~s & \todo{\#Commands} & 63.37~ms\qquad 965.72~ms & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 1429.04~s & \todo{\#Commands} & 201.12~ms\qquad 2005.70~ms & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 4373.79~s & \todo{\#Commands} & 169.50~ms\qquad 3340.67~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 1351.91~s & \todo{\#Commands} & 165.41~ms\qquad 3325.32~ms & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 98.21~s & \todo{\#Commands} & 35.33~ms\qquad 355.90~ms & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 481.13~s & \todo{\#Commands} & 311.45~ms\qquad 6655.26~ms & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 1706.75~s & \todo{\#Commands} & 305.33~ms\qquad 6729.06~ms & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 1610.09~s & \todo{\#Commands} & 312.08~ms\qquad 4569.23~ms & non-parallelizable \tti{diff}ing \\
+\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m35.197s & 49\qquad 193 & 0.056s\qquad 0.523s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77\qquad 317 & 0.090s\qquad 1.083s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96\qquad 384 & 0.145s\qquad 1.790s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96\qquad 384 & 0.147s\qquad 1.809s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{X}\qquad \todo{X} & 0.031s\qquad 0.163s & 	odo{light computation} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.428s & 69\qquad 261 & 0.104s\qquad 1.060s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{X}\qquad \todo{X} & 0.210s\qquad 2.762s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{X}\qquad \todo{X} & 0.254s\qquad 1.958s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{X}\qquad \todo{X} & 0.098s\qquad 1.171s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{X}\qquad \todo{X} & 0.250s\qquad 3.316s & two pipelines merging to a \tti{comm} \\
 \bottomrule

### 2020-05-16 evaluation/plots/microbenchmarks-table.tex
New plots and table based on new results

@@
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(16, 64) &Compile Time (16, 64) & Highlights \\
+Script ~&~ Structure & Input &Seq Time & \#Nodes(16, 64) &Compile Time (16, 64) & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.050s\qquad 0.417s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 0.839s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~MB & 73m28.098s & \todo{\#Commands} & 0.119s\qquad 1.407s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.121s\qquad 1.388s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.163s & 	odo{light computation} \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.214s\qquad 2.800s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 2.762s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{\#Commands} & 0.254s\qquad 1.958s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~MB & 125m10.224s & \todo{\#Commands} & 0.098s\qquad 1.171s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.250s\qquad 3.316s & two pipelines merging to a \tti{comm} \\
+\tti{grep} ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49\qquad 193 & 0.056s\qquad 0.523s & complex NFA regex \\
+\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77\qquad 317 & 0.090s\qquad 1.083s & \tti{sort}ing \\
+\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96\qquad 384 & 0.145s\qquad 1.790s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96\qquad 384 & 0.147s\qquad 1.809s & double \tti{sort}, \tti{uniq} reduction \\
+\tti{grep-light} ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & \todo{X}\qquad \todo{X} & 0.031s\qquad 0.163s & \todo{light computation} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.428s & 69\qquad 261 & 0.104s\qquad 1.060s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142\qquad 574 & 0.328s\qquad 4.657s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125\qquad 509 & 0.186s\qquad 2.341s & non-parallelizable \tti{diff}ing \\
+\tti{optimized bi-grams} ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63\qquad 255 & 0.117s\qquad 1.482s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155\qquad 635 & 0.321s\qquad 4.358s & two pipelines merging to a \tti{comm} \\
+\tti{sort-sort} ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78\qquad 318 & 0.092s\qquad 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-05-16 pldi/draft.tex
Comment about PAL grammar
Minor note
New plots and table based on new results

@@
 %% ``configuration'' input. \kk{Does this need more explanation?}
+
+\kk{We could add the grammar of PAL in an appendix if we want}
 
@@
       Four different configurations per benchmark:
-      (i) \todo{Parallel}, showing the performance of \sys with both eager and split optimizations enabled,
-      (ii) \todo{No Split}, showing the performance of \sys with split disabled,
+      (i) \todo{Parallel + Split}, showing the performance of \sys with both eager and split optimizations enabled,
+      (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
       (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
@@
       \nv{Maybe it makes sense to show log x-axis?}
+      \kk{How can we also fit two more plots (sort-sort and diff) here?
+Maybe we can fit one by getting rid of sort (since it also shown in the sort--parallel one).
+Otherwise we could separate the ``split'' plots from the ones that don't have split.}
     }
@@
 \heading{Programs}
+
+\kk{How can we include a result for grep-light?  This is the IO
+  intensive CPU-light grep. \sys has no speedup but also no slowdown
+  (even if it IO heavy). Should we just mention that in text in
+  passing?}
+
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.

### 2020-05-16 compiler/gather_results.py
New plots and table based on new results

@@ -32,46 +32,54 @@ experiments = ["minimal_grep",
                "spell",
                "shortest_scripts",
                "diff",
+               "bigrams",
                "alt_bigrams",
-               "set-diff"]
+               "set-diff",
+               "double_sort"]
 
 pretty_names = {"minimal_grep" : "grep",
                 "minimal_sort" : "sort",
                 "wf" : "wf",
                 "topn" : "top-n",
                 "grep" : "grep-light",
-                "bigram" : "bi-grams",
+                "bigrams" : "bi-grams",
                 "alt_bigrams" : "optimized bi-grams",
                 "spell" : "spell",
                 "shortest_scripts" : "shortest-scripts",
                 "diff" : "diff",
-                "set-diff" : "set-diff"}
+                "set-diff" : "set-diff",
+                "double_sort" : "sort-sort"}
 
-structures = {"minimal_grep" : "$3\\times$\\tsta",
+structures = {"minimal_grep" : "$3\\times\\tsta$",
               "minimal_sort" : "$\\tsta, \\tpur$",
               "wf" : "$3\\times\\tsta, 3\\times\\tpur$",
               "topn" : "$2\\times\\tsta, 4\\times\\tpur$",
-              "grep" : "$3\\times$\\tsta",
-              "bigram" : "\\todo{TODO}",
-              "alt_bigrams" : "$3\\times$\\tsta, \\tpur",
+              "grep" : "$3\\times\\tsta$",
+              "bigrams" : "$3\\times\\tsta, 3\\times\\tpur$",
+              "alt_bigrams" : "$3\\times\\tsta, \\tpur$",
               "spell" : "$4\\times\\tsta, 3\\times\\tpur$",
               "shortest_scripts" : "$5\\times\\tsta, 2\\times\\tpur$",
               "diff" : "$2\\times\\tsta, 3\\times\\tpur$",
-              "set-diff" : "$5\\times\\tsta, 2\\times\\tpur$"}
+              "set-diff" : "$5\\times\\tsta, 2\\times\\tpur$",
+              "double_sort" : "$\\tsta, 2\\times\\tpur$"}
 
 highlights = {"minimal_grep" : "complex NFA regex",
               "minimal_sort" : "\\tti{sort}ing",
               "wf" : "double \\tti{sort}, \\tti{uniq} reduction",
               "topn" : "double \\tti{sort}, \\tti{uniq} reduction",
-              "grep" : "\todo{light computation}",
[…]

### 2020-05-16 pldi/draft.tex
New plots and results for one-liners

@@
 %% \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+
+\kk{Should we mention anywhere that we will open-source it or is it irrelevant?}
 
@@
       \centering
-      \includegraphics[width=\textwidth]{\detokenize{./figs/alt_bigrams_throughput_scaleup.pdf}}
+      \includegraphics[width=\textwidth]{\detokenize{./figs/bigrams_throughput_scaleup.pdf}}
         %% \caption{TODO}
@@
 
+\TODO{Briefly discuss the performance drop as we increase parallelism in some one-liners (sort-sort, bi-grams).The reason has to do with context switching.
+More precisely, sort-sort 32 has 314 nodes (and 54m sys time) and sort-sort 64 has 634 nodes (and 90m!
+sys time).
+On the other hand opt-bigrams 64 has 255 nodes and the sys time stays the same so I think that it might have to do with the fact that constant costs rise a lot (since opt-bigrams 32 takes just 1 minute).}
+
 We attribute the small drop above $96\times$ in \ttt{spell} \ttt{sort} \ttt{wf}

### 2020-05-16 compiler/gather_results.py

@@ -76,7 +76,10 @@ highlights = {"minimal_grep" : "complex NFA regex",
               "set-diff" : "two pipelines merging to a \\tti{comm}",
               "double_sort" : "parallelizable \\tpur after \\tpur"}
 
-custom_scaleup_plots = {"set-diff" : ["eager", "blocking-eager", "no-eager"]}
+custom_scaleup_plots = {"set-diff" : ["eager", "blocking-eager", "no-eager"],
+                        "spell" : ["split", "eager"],
+                        "bigrams" : ["split", "eager"]}
+
 
 input_filename_sizes = {"1G": "1~GB",
                         "3G": "3~GB",
@@ -234,6 +237,12 @@ def collect_scaleup_times(experiment, results_dir):
     ## Plot speedup
     ax.set_ylabel('Speedup')
     ax.set_xlabel('Level of Parallelism')
+
+    ## In the bigrams experiment we want to also have the alt_bigrams plot
+    if(experiment == "bigrams"):
+        opt_prefix = '{}/{}_'.format(results_dir, "alt_bigrams")
+        opt_distr_speedup, _ = collect_experiment_speedups(opt_prefix, all_scaleup_numbers)
+        ax.plot(all_scaleup_numbers, opt_distr_speedup, '-P', color='magenta', linewidth=0.5, label='Opt. Parallel')
 
     if(not (experiment in custom_scaleup_plots) or
        (experiment in custom_scaleup_plots and

### 2020-05-17 compiler/execute_compile_evaluation_script.sh
change of split sized for new inputs

@@ -69,13 +69,13 @@ if [ "$split_flag" -eq 1 ]; then
     distr_result_filename="${results}${experiment}_distr_split.time"
     sed -i "s#fan_out: [0-9]\+#fan_out: ${n_in}#" config.yaml
     if [ "${microbenchmark}" == "bigrams" ]; then
-        ## These are the total lines for 10G input
-        total_lines=2000000000
+        ## total_lines=2000000000 # 10G input
+        total_lines=600000000  # 3G input
         (( batch_size=total_lines / n_in ))
         sed -i "s#batch_size: [0-9]\+#batch_size: ${batch_size}#" config.yaml
     elif [ "${microbenchmark}" == "spell" ]; then
-        ## These are the total lines for 1G input
-        total_lines=200000000
+        ## total_lines=200000000 # 1G input
+        total_lines=600000000 # 3G input
         (( batch_size=total_lines / n_in ))
         sed -i "s#batch_size: [0-9]\+#batch_size: ${batch_size}#" config.yaml
     # At the moment set-diff cannot be split because of the issue

### 2020-05-17 pldi/microbenchmarks-table.tex

@@
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m35.197s & 49\qquad 193 & 0.056s\qquad 0.523s & complex NFA regex \\
+\tti{grep} ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49\qquad 193 & 0.056s\qquad 0.523s & complex NFA regex \\
 \tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77\qquad 317 & 0.090s\qquad 1.083s & \tti{sort}ing \\
@@
 \tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96\qquad 384 & 0.147s\qquad 1.809s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{X}\qquad \todo{X} & 0.031s\qquad 0.163s & 	odo{light computation} \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.428s & 69\qquad 261 & 0.104s\qquad 1.060s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{X}\qquad \todo{X} & 0.210s\qquad 2.762s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{X}\qquad \todo{X} & 0.254s\qquad 1.958s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 125m10.224s & \todo{X}\qquad \todo{X} & 0.098s\qquad 1.171s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{X}\qquad \todo{X} & 0.250s\qquad 3.316s & two pipelines merging to a \tti{comm} \\
+\tti{grep-light} ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & \todo{X}\qquad \todo{X} & 0.031s\qquad 0.163s & \todo{light computation} \\
+\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 69\qquad 261 & 0.104s\qquad 1.038s & comparisons (\tti{comm}) \\
+\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142\qquad 574 & 0.328s\qquad 4.657s & \todo{extensive file-system operation} \\
+\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125\qquad 509 & 0.186s\qquad 2.341s & non-parallelizable \tti{diff}ing \\
+\tti{bi-grams} ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 109\qquad 445 & 0.146s\qquad 1.716s & stream shifting and merging \\
+\tti{optimized bi-grams} ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63\qquad 255 & 0.117s\qquad 1.482s & optimized version of bigrams \\
+\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155\qquad 635 & 0.321s\qquad 4.358s & two pipelines merging to a \tti{comm} \\
+\tti{sort-sort} ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78\qquad 318 & 0.092s\qquad 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-05-17 pldi/draft.tex
Comments on the one-liners evaluation
Final one-liner results with auto split too

@@
       Four different configurations per benchmark:
-      (i) \todo{Parallel + Split}, showing the performance of \sys with both eager and split optimizations enabled,
+      (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
+      (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
       (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
@@
   passing?}
+
+\kk{Remember to mention that for spell and bigrams we dont show other
+  parallel because they are all close to 1 (no speedup).}
 
@@
 
-\TODO{Briefly discuss the performance drop as we increase parallelism in some one-liners (sort-sort, bi-grams).The reason has to do with context switching.
-More precisely, sort-sort 32 has 314 nodes (and 54m sys time) and sort-sort 64 has 634 nodes (and 90m!
-sys time).
-On the other hand opt-bigrams 64 has 255 nodes and the sys time stays the same so I think that it might have to do with the fact that constant costs rise a lot (since opt-bigrams 32 takes just 1 minute).}
+\TODO{Briefly discuss the performance drop as we increase parallelism
+  in some one-liners (sort-sort, bi-grams).The reason has to do with
+  context switching. More precisely, sort-sort 32 has 314 nodes (and
+  54m sys time) and sort-sort 64 has 634 nodes (and 90m! sys time). On
+  the other hand opt-bigrams 64 has 255 nodes and the sys time stays
+  the same so I think that it might have to do with the fact that
+  constant costs rise a lot (since opt-bigrams 32 takes just 1
+  minute).}
+
+\kk{It is visible that the automatic split is slightly worse than the
+  manual one in most cases (because it doesn't utilize task-based
+  parallelism).}
 

### 2020-05-17 compiler/gather_results.py

@@ -248,10 +248,14 @@ def collect_scaleup_times(experiment, results_dir):
        (experiment in custom_scaleup_plots and
         "split" in custom_scaleup_plots[experiment])):
         try:
+            auto_split_distr_speedup = collect_distr_experiment_speedup(prefix,
+                                                                        'distr_auto_split.time',
+                                                                        all_scaleup_numbers)
+            ax.plot(all_scaleup_numbers, auto_split_distr_speedup, '-D', color='tab:red', linewidth=0.5, label='Par + Split')
             split_distr_speedup = collect_distr_experiment_speedup(prefix,
                                                                    'distr_split.time',
                                                                    all_scaleup_numbers)
-            ax.plot(all_scaleup_numbers, split_distr_speedup, '-D', color='tab:red', linewidth=0.5, label='Parallel + Split')
+            ax.plot(all_scaleup_numbers, split_distr_speedup, '-h', color='brown', linewidth=0.5, label='Par + B. Split')
         except ValueError:
             pass
 

### 2020-05-17 pldi/draft.tex
Minor intro corrections

--- 2020-05-16 pldi/draft.tex
@@
 
+\nv{Add note about breadth of commands and the impossibility of our goal.}
+% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
+
 % \TODO{Drop the automated from the title maybe?}
@@
 
-The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism (see~\S\ref{related}).
-\kk{I don't really like the reference to such a later section.
-Can we remove it?}
+The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism.
+% \kk{I don't really like the reference to such a later section. Can we remove it?}
+% \nv{We have to say this is not just ``an opinion'', but that we give a scientific treatment/full coverage in the related work section.
+How far away should the sections be so that you're comfortable referencing to?}
 % If versed into parallelism, 
@@
 One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
-  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained \kk{i.e.
-only being capable to exploit parallelism in the level of entire scripts} thus applying at the level of entire scripts rather than individual components.
+  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus only being capable to exploit parallelism at the level of entire scripts rather than individual components.
 % Tools will attempt to parallelize anything 
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is using the shell's concurrency primitives (\eg \ttt{\&}, \ttt{wait}, \ttt{for}) \kk{Is \ttt{for} a concurrency primitive?} % for explicit parallelism, 
- that come at a cost of manual effort to carefully split inputs, manually rewrite scripts, and explicitly orchestrate execution---an expensive and error-prone process.
+Another option is to use shell primitives such as \ttt{\&}, \ttt{wait}, \ttt{for} to explicitly induce concurrency;
+ these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
 % re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
 % The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
-And\todo{/To top it off} all these options assume a good understanding of parallelism;
-  users whose domain of expertise falls outside computing---\todo{routinely} writing shell scripts to accomplish their tasks---are left without options.% automation when it comes to parallelization.
-
-To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
-\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
+To top it off, all these options assume a good understanding of parallelism;
+  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options (see~\S\ref{related} for a full discussion).
+% \nv{I still don't like ``for a full discussion''---raises too many expectations, and we don't have that much more to say}
+
+To address this challenge, we develop \sys, a shell that parallelizes POSIX shell scripts.
+% \kk{Mention that it outputs shell scripts.}
+% \nv{I think the ``how'' it solves it should come later, we're still trying to convince them we solve it (see next sentences in this paragraph)
 % \sys offers both classes of users in the second group such much-needed automation.
@@
 \sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.
-\kk{Should we mention that we only show this empirically?}
+% \kk{Should we mention that we only show this empirically?}
+% \nv{I don't think so; I don't think this is the right point to talk about our evaluation, we just state our intention/high-level principles.}
 
@@
 % Aside from these two components, \sys packs a runtime component that serves dual purpose:
-Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combiners\todo{/combinators} as well as high-performance primitives for eager data splitting and merging.
-These address several practical challenges such as task parallelism, deadlocks, and performance bottlenecks\kk{I like the flow of this sentence but I am not sure it makes a lot of sense to name these challenges here.}, and were developed by uncovering several pathological situations, on many of which we report.
+Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combinators as well as high-performance primitives for eager data splitting and merging.
+These address several practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
+  and were developed by uncovering several pathological situations, on many of which we report.
 
@@
 \kk{I would remove the ``aided further...'' part.}
-Naturally, improvements on individual commands depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls.
-\kk{Do we need to mention this last sentence here?
-Without context it is pretty difficult for anyone to get something out of it.}
+Naturally, these improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls. 
+%\kk{Do we need to mention this last sentence here? Without context it is pretty difficult for anyone to get something out of it.}
 
@@
   \S\ref{distributability}
-    studies the parallelizability properties of a large class of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
+    studies the parallelizability of a large set of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
 
@@
   \item
-  \S\ref{impl} describes \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
+  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
 % \kk{Rewrite that when we know for sure what the section contains.}
@@
 
-\heading{Command Inputs and Outputs}
+\paragraph{Command Inputs and Outputs}
 
@@
 
-\heading{Command-line Options}
+\paragraph{Command-line Options}
 As described earlier~\sx{bg}, com\-mand-line options and environment
@@
 
-%% \heading{Command Inputs and Outputs}
+%% \paragraph{Command Inputs and Outputs}
 %% A second concern captured by the AL is the order of inputs and outputs.
@@
 
-\heading{Parallelizability Annotation Language}
+\paragraph{Parallelizability Annotation Language}
 
@@
 %% ``configuration'' input. \kk{Does this need more explanation?}
-
-\kk{We could add the grammar of PAL in an appendix if we want}
 
@@
 
-\heading{Parallelizable Pure Commands}
+\paragraph{Parallelizable Pure Commands}
 
@@
 
-\heading{Edges -- Files}
+\paragraph{Edges -- Files}
 
@@
 
-\heading{Nodes -- Commands}
+\paragraph{Nodes -- Commands}
 
@@
 
-\heading{\todo{One-input-output} Commands}
+\paragraph{\todo{One-input-output} Commands}
 
@@
 
-\heading{Stateless and Parallelizable Pure Commands}
+\paragraph{Stateless and Parallelizable Pure Commands}
 
@@
 
-\heading{Parallelizable Regions}
+\paragraph{Parallelizable Regions}
 %
@@
 
-\heading{Translation Pass}
+\paragraph{Translation Pass}
 
@@
 
-\heading{Pure Fold Function Implementations}
+\paragraph{Pure Fold Function Implementations}
 
@@
       Four different configurations per benchmark:
-      (i) \todo{Parallel + Split}, showing the performance of \sys with both eager and split optimizations enabled,
-      (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
+      (i) \todo{Parallel}, showing the performance of \sys with both eager and split optimizations enabled,
+      (ii) \todo{No Split}, showing the performance of \sys with split disabled,
       (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
@@
       \nv{Maybe it makes sense to show log x-axis?}
-      \kk{How can we also fit two more plots (sort-sort and diff) here?
-Maybe we can fit one by getting rid of sort (since it also shown in the sort--parallel one).
-Otherwise we could separate the ``split'' plots from the ones that don't have split.}
     }
@@
 \heading{Programs}
-
-\kk{How can we include a result for grep-light?  This is the IO
-  intensive CPU-light grep. \sys has no speedup but also no slowdown
-  (even if it IO heavy). Should we just mention that in text in
-  passing?}
-
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
@@
 
-\heading{Distributed Operating Systems and Shells}
+\paragraph{Distributed Operating Systems and Shells}
 A plethora of systems assist in the construction of distributed
@@
 
-\heading{Cluster Computing}
+\paragraph{Cluster Computing}
 Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
@@
 
-\heading{Annotation-based Distribution}
+\paragraph{Annotation-based Distribution}
 Recent systems on light-touch distribution, such as
@@
 % Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Low-level Parallelization}
+\paragraph{Low-level Parallelization}
 Instruction-level parallelization has a longer history, starting from

### 2020-05-18 pldi/draft.tex
Rewrite example

@@
 \usepackage{syntax}
+\usepackage{caption}
 
@@
 
-%% \captionsetup[figure]{font=footnotesize,name={Fig.},labelfont={bf, footnotesize}}
-%% \captionsetup[table]{font=footnotesize,name={Tab.},labelfont={bf, footnotesize}, skip=2pt, aboveskip=2pt}
-%% \captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
+\captionsetup[figure]{font=footnotesize,name={Fig.},labelfont={bf, footnotesize}}
+\captionsetup[table]{font=footnotesize,name={Tab.},labelfont={bf, footnotesize}, skip=2pt, aboveskip=2pt}
+\captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
 
@@
 
-This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation~\sx{bg:overview}.
-
-\subsection{Unix Composition, Informally}
+This section first reviews shell scripting through a small example~\sx{bg:pipelines}, which it then uses to present an overview of \sys's design and implementation~\sx{bg:overview}.
+
+\subsection{Running Example: Weather Analysis}
 \label{bg:pipelines}
 
-% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
-\unix provides several ways for succinct program composition.
-Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
-These programs produce output and consume input concurrently and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
-Intermediate programs, termed \emph{commands}, can also be developed in different programming languages, as long as they conform to a common interface.
-%\begin{wrapfigure}{r}{0.06\textwidth}
-%  \vspace{-12pt}
-%  % \begin{center}
-%    % \includegraphics[width=0.48\textwidth]{gull}
-%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
-%  % \end{center}
-%  \vspace{-30pt}
-%\end{wrapfigure}
-This interface exposes a stream of contiguous character lines.
-A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
-As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example.pdf}}
+\caption{
+  \textbf{Calculating maximum temperatures per year.}
+  The script downloads daily temperatures recorded across the U.S. for the years 1920--2020 and extracts the maximum for every year.
+}
+\vspace{-15pt}
+\label{fig:example}
+\end{figure}
+
+% In the ongoing debate on climate change,
+Suppose a \todo{climate scientist}    % looking for a non-CS expert---journalist might be too low
+  wants to get a quick sense of the changes in maximum temperature over the past century.
+As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
+  answering this question is only a matter of a simple data-processing pipeline.
+
+The shell script in Fig.~\ref{fig:example} starts by pulling the yearly index files and filtering out URLs that are not part of the compressed dataset.
+It then downloads and decompresses each file in the remaining set, extracts the values that indicate the temperature, and filters out bogus inputs marked as \ttt{999}.
+It then calculates the maximum yearly temperature by sorting the values and picking the top element.
+Finally, it matches each maximum value with the appropriate year in order to print the result.
+% 
+The effort required to write this pipeline is astonishingly low:\footnote{
+  Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.
+}
+  its data-processing core amounts to 12 stages and, when expressed as a single line, only 165 characters long.
+To enable such a succinct program composition, \unix incorporates several features.
+
+\heading{\unix Features}
+An important \unix feature are pipes (\ttt{|}), introducing task-parallel program composition over character streams.
+These streams are contiguous character lines separated by newline character (\textsc{NL}), which delineate individual stream elements.
+For example, the first \ttt{grep} is given lines that contain file identifiers, of which it only outputs lines that contain \ttt{gz}, which are in turn consumed by \ttt{tr}.
+% As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
+A special end-of-file (\textsc{EOF}) condition marks the end of a stream.
+
+Another important feature is synchronization, as different pipeline stages process data concurrently and possibly at different rates.
+For example, the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
+The \unix kernel facilitates scheduling, communication, synchronization behind the scenes.
+
+Command flags, used pervasively in \unix, are configuration options that the command's developer has decided to expose to its users, to give command users significant freedom in their execution.
+For example, by omitting  \ttt{sort}'s \ttt{r} flag that enables reverse sorting, the user can easily get the minimum temperature.
+The shell does not have any visibility into these flags; 
+  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to commands.
+
+Finally, \unix provides composition environment for commands written in any language.
+Many the commands used in practice come built into the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
+The fact that commands are developed in a variety of languages---including shell scripts---provides users with a significant flexibility.
+For example, one could replace the call to \ttt{sort} and \ttt{head} with one to \ttt{./avg.py} to get an average or a standard deviation rather than a maximum.
+The pipeline would still work, as long \ttt{./avg.py} conforms to the interface outlined earlier.
+% Shell scripting allows incremental development through refinement
+
 % the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
@@
 
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls -l | wc -l                          ($$(p_2)$$)
-\end{lstlisting}
-% \smallskip
-
-Commands can be tuned by two mechanisms: com\-mand-line options and environment variables.
-Options control the execution of a program---\eg changing the \ttt{sort}ing order.
-The shell does not have any visibility into these options; 
-  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
-Environment variables carry more general information about the surrounding system---\eg search paths, system defaults, parsing switches \etc 
-They are organized as a map from names to values, both of which are strings, and can be evaluated at any point in the program---\eg as commands, flags, or inputs.
-% a shell feature termed parameter \emph{expansion}.
-
-% TODO: Talk about security primitives?
-Streams are a key abstraction in \unix, and go hand-in-hand with another \unix abstraction: \emph{files}.
-Files are persistent streams that can be manipulated by individual commands or pipelines:
-% \begin{wrapfigure}{r}{0.15\textwidth}
-%   \vspace{-15pt}
-%   % \begin{center}
-%     % \includegraphics[width=0.48\textwidth]{gull}
-%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-%   % \end{center}
-%   \vspace{-25pt}
-% \end{wrapfigure}
-  reading a (set of) file(s) generates a stream; 
-  conversely, redirecting to a file identifier generates a file.
-% streams can be trivially redirected to the file system---by appending with the file identifier.
-Files are named, corresponding to global identifiers---\eg \ttt{/x/y} corresponds to a persistent stream at the location \ttt{/y} in \ttt{/x}.
-\unix exposes a few different types of files---\eg directories, symbolic links, and named pipes (or FIFOs). % , and special-character devices.
-As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).
-
-% At a high level, POSIX defines several file types:
-% (-) regular files, storing lines of text or binary data,
-% (d) directories, grouping together multiple other files,
-% (l) symbolic links, pointing to other files,
-% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
-% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
-% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
-% \smallskip
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  ls | sed 's/1/2/' | > out.t 2> err.t    ($$(p_3)$$)
-\end{lstlisting}
-%\smallskip
-
-\noindent
-Special files are particularly interesting because they resolve to OS structures.
-One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
-  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
-% Here are a few examples:
-% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
-% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
-% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
-Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
-This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---enables the runtime wrapping and rebinding achieved during \sys's transformations.
-% TODO: compatibility
-
-A small but powerful set of features is concerned with stream manipulation.
-Some of these features are built into the shell;
-  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file---which itself can be a FIFO dubbing as input to another command.
-% \begin{wrapfigure}{r}{0.15\textwidth}
-%   \vspace{-15pt}
-%   % \begin{center}
-%     % \includegraphics[width=0.48\textwidth]{gull}
-%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
-%   % \end{center}
-%   \vspace{-25pt}
-% \end{wrapfigure}
-More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
-  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
-% TODO: talk about \ttt{xargs}?
-
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  cat * | tee >(wc -l > l.t) | zip > a.z   ($$(p_4)$$)
-\end{lstlisting}
-% \smallskip
-
-\noindent
-% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators 
-The combination of ephemeral file-system identifiers and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
-The most interesting feature of these pipelines is the presence of cycles:
-  a pipeline segment processes some input, part of which is extracting the next input to process.
-One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
-% These streams are created by the C library (glibc) at the start of program execution,
-% and new streams can be created to connect to files, sockets, pipes,
-
-% \smallskip
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-  read | { calc | square; } >/dev/fd/0     ($$(p_5)$$)
-\end{lstlisting}
-% \smallskip
-
-\noindent
-Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
-Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
-Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
-  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
-Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.
-
-To conclude, \unix pipelines are surprisingly powerful.
-This power is exploited directly by \sys's rewriting passes which re-target the shell rather than an implementation in Java or Go.
-
-% \begin{lstlisting}[
-%   language=es,
-%   mathescape,
-%   float=t,
-%   belowskip=-5mm,
-%   label=core-pl,
-%   upquote=true,
-%   caption={
-%     \textbf{The core language of the POSIX shell.}
-%     The language used by ; file identifiers and expansion are not shown.
-%   }]
-% $p$ ::= $s\in{}String$ | $n\in{}Number$ | $b\in{}Bool$ | $\emptyset$
-% $v$ ::= $p$ | ($x$,$\ldots$) => {e} | {$s$:$v$,$\ldots$} | [$v$,$\ldots$]
-% $e$ ::= x | $v$ | ($x$ = $e$) $e$ | $e$($e$) | $e$[$e$] = $e$
-% \end{lstlisting}
-
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_4$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%  cat /proc/ >                            (($$p_5$$))
-% \end{lstlisting}
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-% { echo 1; echo 2; } > out.txt
-% \end{lstlisting}
+% ---available publicly online via NCDC's NOAA~\cite{}---to obtain the maximum temperature per year.
+
+\heading{Parallelization Challenges}
+While these features aid developer economy through powerful program composition, they complicate shell script parallelization which even for simple scripts such as the one in Fig.~\ref{fig:example} face several challenges.
+% \footnote{
+%   Alternatives for automated parallelization were outlined in the introduction~\sx{intro} and are surveyed in the related work section~\sx{related}.
+% }
+
+One obvious challenge is due to the language of the shell, which has been developed through a long history of additions.
+Different parts of a script, exploiting different language constructs, are amenable to different levels of parallelization---if at all.
+% Some constructs, such as the parallel composition operator \ttt{\&} ; many others such as the sequential composition operator (\ttt{;}) are clearly sequential; yet others, such as the loop constructs are more difficult to categorize.
+For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), cannot be interleaved with the rest of the script.
+
+Another challenge is due to the multitude of available commands;
+  each command that is part of a script may require very different treatment.
+For example, \ttt{grep} can proceed in parallel on each stream without any coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination across streams.
+Worse even, the same command may be used with different flags---\eg % \ttt{-d" " -f9} for the first
+  Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+When a single flag can affect the functionality of a command in non-trivial ways, deciding how to parallelize becomes significantly more difficult.
+
+Yet another challenge is that different phases operate on different output sizes thus allowing different parallelization widths.
+Even if Fig.~\ref{fig:example}'s script was aiming on, say, only six years of data, the first \ttt{curl} would still output hundreds of lines for each year.
+Naive, coarse-grained parallelization can miss such opportunities.
+
+% \nv{Are these all the challenges we want to highlight?}
 
@@
   \textbf{\sys transformation overview.}
-  Given a shell pipeline, \sys leverages command distributability to identify
-  high-distributability stages, rewrite them, and orchestrate their execution.
+  Given a shell pipeline, \sys leverages command parallelizability to identify
+  high-parallelizability stages, rewrite them, and orchestrate their execution.
 }
@@
 
-\kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}
-
-\kk{Make sure for the running script to contain more than just a
-  pipeline, so that it is clear that \sys turns a script into the same
-  script with DFGs in place of ``parallelizable-regions'' (which are
-  usually pipelines).}
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs another shell script.
+The new script may incorporate parallelism, and handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
+To discover and enable parallel execution, \sys analyzes individual program fragments, addressing the aforementioned challenges.
 

### 2020-05-18 pldi/draft.tex
More example/overview

@@
 
-As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
-Intuitively,
- (i) the parallel-composition operator exposes task parallelism
-% FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
+The key insight behind the analysis is that, intuitively, the POSIX language amounts to a few constructs that indicate different levels of candidacy for pararllelization.
+For example,
+ (i) the parallel-composition operator exposes task parallelism,
  (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
  (iii) the pipe operator is a candidate for both task and data parallelism, and
-% \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
  (iv)  other composition operators can be viewed as barrier candidates.
-To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.
-
-% However, understanding the shell's primitives is not enough
-First, \sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
+
+Second, \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs~\sx{parallelizability}.
 This challenge can be broken down into two parts.
@@
 One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
-\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes~\sx{cmd}.
+\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood parallelizability classes~\sx{cmd}.
 This group is already useful-enough to allow the composition of practical pipelines~\sx{micro}.
 
+will have several fragments that explicitely execute in parallel, as long as \sys was able to identify and extract parallelism latent in the original script.
+
+By parsing the script of Fig.~\ref{fig:example},
+
+\sys will first identify a series of commands
+
+For now, different \ttt{for} branches can proceed in parallel as all the information is available.
+However, if the 
+In cases where \sys does not have enough information in place, it will not attempt parallelizing the script.
+
+\kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}
+
+\kk{Make sure for the running script to contain more than just a
+  pipeline, so that it is clear that \sys turns a script into the same
+  script with DFGs in place of ``parallelizable-regions'' (which are
+  usually pipelines).}
+
+As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
+
+% However, understanding the shell's primitives is not enough
 The second part is understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power.
 New commands are arbitrary programs, but still need to conform to the shell's pipeline interface.
-To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
-This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the distributability of the commands they are developing.
+To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its parallelizability.
+This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the parallelizability of the commands they are developing.
 % We haven't introduced the notion of flags
@@
 
-Given a pipeline of commands and their distributability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
+Given a pipeline of commands and their parallelizability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
 It addresses this challenge with a series of rewriting passes that perform dataflow analysis to identify distributable regions.
@@
 First, it needs to wrap single-node commands with corrective runtime interposition based on their classes.
-This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the distributability specification described earlier.
+This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the parallelizability specification described earlier.
 Thin wrappers avoid interference with legacy functionality, as \unix commands contain non-obvious logic and flags added over extended periods of time.
@@
 
-\kk{Maybe use phoenix for that one? Since it has a MapReduce interface
-  but it runs on multicore?}
-\nv{Phoenix (i) is shared-memory, and (ii) would require re-writing the entire program in C++, which defeats the purpose of comparing with an existing program.
-Phoenix might be OK for micro-benchmarks however.}
-
-To understand \sys better under a realistic workload, we use a large pipeline inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}.
+To understand \sys better under a realistic workload, we turn our attention to the Fig.~\ref{fig:example}'s script~\sx{bg}.
 
 \heading{Program}
-At a high level, the program processes multi-year weather data from NCDC to identify the maximum temperature over the past several years.
-The original task, as presented in the book, can be thought of comprising three sub-tasks:
+The full version of this program is inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it is used as an example of a realistic mid-scale analytics pipeline.
+
+The original task, as presented in the book, is comprised of three sub-tasks:
   download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
-Only the last of these tasks is the real focus of the chapter, demonstrating the benefits of using distributed analytics system such as Hadoop.
+Only the last of these tasks is the real focus of the book, demonstrating the benefits of using distributed framework such as Hadoop.
 For \sys, we consider the full pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
 
-The pipeline starts with a few stages in \sta  that generate URLs;
-  these URLs point to index files, each one of which contains pointers to compressed data files.
-The stream of index files is then analyzed (4 stages) to extract the URLs of the data files, passed again through the command to download the data.
-After download and un-compression, the stream is analyzed to extract the maximum temperature (another 4 stages).
+The full program used for our evaluation is slightly different than the one presented in Fig.~\ref{fig:example}, consisting of 14 stages and totalling 240 characters.
+We full program expresses the \ttt{for} loop as its first pipeline stage and uses \ttt{xargs} to 
+The program presented in Fig.~\ref{fig:example} is simplified only for clarity of exposition.
 
@@
 % The first interesting result 
-The effort required to write this pipeline is astonishingly low:\footnote{
-  Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.
-}
-  it amounts to 14 stages and, when expressed as a single line, only 240 characters.
+\TODO{Some results}
 The Hadoop program, expressed in Java and only capturing the \emph{four} last stages, amounts to 137 lines of Java code---not accounting for commands moving data in and out of HDFS.

### 2020-05-18 pldi/draft.tex
Distributability -> Parallelizability

@@
   To distribute them, \sys
-    decomposes primitives into distributability classes,
-    identifies high-distributability stages,
+    decomposes primitives into parallelizability classes,
+    identifies high-parallelizability stages,
     applies rewriting rules for largest possible subprograms,
@@
 %  with regards to the sequential program.
-  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known distributability classes.
+  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known parallelizability classes.
 % a series of techniques for automatically 
@@
 It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
-Sections \ref{distributability}--\ref{impl} highlight key contributions:
+Sections \ref{parallelizability}--\ref{impl} highlight key contributions:
 \begin{itemize}
   \item
-  \S\ref{distributability}
+  \S\ref{parallelizability}
     studies the parallelizability of a large set of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
@@
 %%   families/classes of commands that share
-%%   distributability/parallelization properties, and a classification of
-%%   coreutils/POSIX commands in these distributability properties.
+%%   parallelizability/parallelization properties, and a classification of
+%%   coreutils/POSIX commands in these parallelizability properties.
 %% \item Dataflow graph model that is suitable for the shell
@@
 %%   \textbf{High-level schematic.}
-%%   \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
+%%   \sys leverages parallelizability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
 %%   execution (right).
@@
 Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-%% \sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
+%% \sys builds on this insight with a careful study of the parallelizability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
 %% A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
@@
 %% First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-%% Second, developers of new shell commands can use a lightweight domain-specific language to express their distributability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
+%% Second, developers of new shell commands can use a lightweight domain-specific language to express their parallelizability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
 %% Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
@@
 %% It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
-%% Sections \ref{distributability}--\ref{impl} highlight key
+%% Sections \ref{parallelizability}--\ref{impl} highlight key
 %% contributions:
@@
 %%   \item
-%%   \S\ref{distributability} studies several distributability classes and introduces the distributability characterization language.
+%%   \S\ref{parallelizability} studies several parallelizability classes and introduces the parallelizability characterization language.
 
@@
 
-The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
+The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss \sys's design and implementation details.
 % They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
 
-\section{Distributability Classes}
-\label{distributability}
+\section{Parallelizability Classes}
+\label{parallelizability}
 
@@
 To have any hope of distributing pipelines, \sys first needs to understand the distribution characteristics of their individual commands.
-This understanding is encoded in a command's \emph{distributability} class.
-Informally, a distributability class represents the level of synchronization required by copies of a command that execute in parallel.
+This understanding is encoded in a command's \emph{parallelizability} class.
+Informally, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
 It focuses on capturing characteristics that are important for distribution, rather than the entire the observable behavior of a command.
@@
 
-This section starts by defining these classes, along with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+This section starts by defining these classes, along with a parallelizability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
 Building on this study, it develops a DSL that enables command classification by its developers~\sx{ext}.
@@
 
-\subsection{Distributability of Standard Libraries}
+\subsection{Parallelizability of Standard Libraries}
 \label{cmd}
@@
   \footnotesize{
-    \textbf{Distributability Classes}.
+    \textbf{Parallelizability Classes}.
     Broadly, \unix commands can be broken down into to five classes.
@@
 These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
-The details of this state and its propagation during piecewise element processing affect their distributability characteristics.
+The details of this state and its propagation during piecewise element processing affect their parallelizability characteristics.
 
@@
   \caption{
-\textbf{Distributability description language.}
-The DSL captures important information regarding the distributability of a command.
+\textbf{Parallelizability description language.}
+The DSL captures important information regarding the parallelizability of a command.
 \vspace{-10pt}
@@
 %% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
-%% % but these maintain different invariants, enabling very different distributability characteristics.
+%% % but these maintain different invariants, enabling very different parallelizability characteristics.
 
@@
 
-% As mentioned in distributability, pure commands can be broken down
+% As mentioned in parallelizability, pure commands can be broken down
 % into different categories. We might be able to get ideas about this
@@
 %% checks whether it is safe to parallelize it. A command is considered
-%% safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
 % why couldn't we replicate this?
@@
   \sys itself would never attempt a factor that is above the number of available compute nodes.
-The results show significant speedups, ranging between 4--70$\times$, depending on the distributability characteristics of individual pipelines.
+The results show significant speedups, ranging between 4--70$\times$, depending on the parallelizability characteristics of individual pipelines.
 
@@
 proceed in parallel.  By targeting specific classes of computation
-(\emph{viz.} \sys's distributability), they render distributed
+(\emph{viz.} \sys's parallelizability), they render distributed
 computation efficiently.
@@
 applications by using third-party libraries do not understand their
-distributability details well-enough to annotate them.
+parallelizability details well-enough to annotate them.
 
@@
 There are several worthwhile directions for future work.  One
-direction would be an extension of the distributability analysis of
+direction would be an extension of the parallelizability analysis of
 \sys, in order to guide it using dynamic information. This could
@@
 To achieve its goal, \sys
-  decomposes primitives into distributability classes,
-  identifies high-distributability stages,
+  decomposes primitives into parallelizability classes,
+  identifies high-parallelizability stages,
   applies rewriting rules for largest possible subprograms,
@@
 Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
-
-\section*{Acknowledgments}
 

### 2020-05-18 pldi/draft.tex
Finish rewrite

@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs another shell script.
-The new script may incorporate parallelism, and handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
-To discover and enable parallel execution, \sys analyzes individual program fragments, addressing the aforementioned challenges.
-
-The key insight behind the analysis is that, intuitively, the POSIX language amounts to a few constructs that indicate different levels of candidacy for pararllelization.
-For example,
- (i) the parallel-composition operator exposes task parallelism,
- (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator is a candidate for both task and data parallelism, and
- (iv)  other composition operators can be viewed as barrier candidates.
-
-Second, \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs~\sx{parallelizability}.
-This challenge can be broken down into two parts.
-First, understand standard commands part of the shell---that is, built-ins that come with any shell and support its core functions.
-One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
-\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood parallelizability classes~\sx{cmd}.
-This group is already useful-enough to allow the composition of practical pipelines~\sx{micro}.
-
-will have several fragments that explicitely execute in parallel, as long as \sys was able to identify and extract parallelism latent in the original script.
-
-By parsing the script of Fig.~\ref{fig:example},
-
-\sys will first identify a series of commands
-
-For now, different \ttt{for} branches can proceed in parallel as all the information is available.
-However, if the 
-In cases where \sys does not have enough information in place, it will not attempt parallelizing the script.
-
-\kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}
-
-\kk{Make sure for the running script to contain more than just a
-  pipeline, so that it is clear that \sys turns a script into the same
-  script with DFGs in place of ``parallelizable-regions'' (which are
-  usually pipelines).}
-
-As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
-
-% However, understanding the shell's primitives is not enough
-The second part is understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power.
-New commands are arbitrary programs, but still need to conform to the shell's pipeline interface.
-To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its parallelizability.
-This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the parallelizability of the commands they are developing.
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new POSIX shell script.
+The new script may incorporate parallelism, and is handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
+To discover and enable parallel execution, \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
+
+\sys first needs to understand the POSIX constructs used to compose larger programs, indicating indicate different levels of candidacy for pararllelization.
+To achieve this, \sys breaks POSIX constructs into a few broad classes:
+ (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
+ (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+ (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
+ (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
+
+Second, \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs---a two-fold challenge.
+To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but clear set of well-understood parallelizability classes;
+  these commands are already useful-enough to allow the composition of useful shell programs operating in practice.
+To understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power, \sys leverages this study to define an annotation language for describing a command's key properties with respect to its parallelizability.
+Such light annotations are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they are developing.
 % We haven't introduced the notion of flags
@@
 
-Given a pipeline of commands and their parallelizability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
-It addresses this challenge with a series of rewriting passes that perform dataflow analysis to identify distributable regions.
-It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
-These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc
-
-To schedule and execute the resulting distributed dataflow graph, \sys needs to solve several challenges~\sx{impl}.
-First, it needs to wrap single-node commands with corrective runtime interposition based on their classes.
-This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the parallelizability specification described earlier.
-Thin wrappers avoid interference with legacy functionality, as \unix commands contain non-obvious logic and flags added over extended periods of time.
-
-Second, \sys needs to fill-in planing details left blank by the rewriting pass.
-This is achieved by \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
-They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is empty, not affecting pipelines).
-They make clever use of shell features such as expansion to achieve the latest possible binding, when critical runtime information becomes available.
-% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
-
-% FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}
-
-The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss \sys's design and implementation details.
+% Given a pipeline of commands and their parallelizability characteristics, \sys's next challenge is 
+% \sys achieves this through a series of rewriting passes that perform dataflow analysis to identify distributable regions.
+% It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
+
+Provided information about the shell constructs and individual commands and their flags, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
+To achieve this, it converts the script into a data-flow graph (DFG) model.
+% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
+This is a flexible representation that allows a series of parallelization transformations to convert the graph into its parallel equivalent, in which DFGs indicate the parallelizable regions.
+Further transformations compile the DFG back to a parallel script that uses POSIX shell constructs to guide parallelism explicitly---while aiming at preserving the semantics of the sequential program.
+
+Finally, \sys needs to provide correct, high-performance combinators for certain classes commands, and be proactive about (addressing) several practical challenges related to quirks of the \unix environment.
+In terms of combiners, \sys offers a set of linkable components for stream splitting and merging;
+  these commands live in the \ttt{PATH} and are addressable by name, which means they can be used like (and by) any other commands.
+Practical challenges---related to task parallelism, deadlock prevention, and runtime performance---are addressed through a series of constructs that \sys provides.
+
+Returning to the script of Fig.~\ref{fig:example}...\nv{I think we should show fragments of the resulting script}
+
+The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
 % They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
@@
 \label{parallelizability}
-
-\TODO{Change naming to parallelizability}
 

### 2020-05-18 pldi/draft.tex

@@
       Four different configurations per benchmark:
-      (i) \todo{Parallel}, showing the performance of \sys with both eager and split optimizations enabled,
-      (ii) \todo{No Split}, showing the performance of \sys with split disabled,
+      (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
+      (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
+      (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
       (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
@@
 \heading{Programs}
+
+\kk{How can we include a result for grep-light?  This is the IO
+  intensive CPU-light grep. \sys has no speedup but also no slowdown
+  (even if it IO heavy). Should we just mention that in text in
+  passing?}
+
+\kk{Remember to mention that for spell and bigrams we dont show other
+  parallel because they are all close to 1 (no speedup).}
+
 Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
@@
 
+\TODO{Briefly discuss the performance drop as we increase parallelism
+  in some one-liners (sort-sort, bi-grams).The reason has to do with
+  context switching. More precisely, sort-sort 32 has 314 nodes (and
+  54m sys time) and sort-sort 64 has 634 nodes (and 90m! sys time). On
+  the other hand opt-bigrams 64 has 255 nodes and the sys time stays
+  the same so I think that it might have to do with the fact that
+  constant costs rise a lot (since opt-bigrams 32 takes just 1
+  minute).}
+
+\kk{It is visible that the automatic split is slightly worse than the
+  manual one in most cases (because it doesn't utilize task-based
+  parallelism).}
+
 We attribute the small drop above $96\times$ in \ttt{spell} \ttt{sort} \ttt{wf}

### 2020-05-18 pldi/draft.tex
Add fig, update classes

@@
 
-\nv{Add note about breadth of commands and the impossibility of our goal.}
-% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
-
 % \TODO{Drop the automated from the title maybe?}
@@
 After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper closes with a discussion of \sys's limitations and possible future directions~\sx{discussion}.
-%micro-bench\-marks %---popular shell one-liners that highlight
-%certain features---and a multi-line macro-benchmark---a popular
-%weather analysis script.
-%% Additional material, in anonymized form, is available at
-%% \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
-
-\kk{Should we mention anywhere that we will open-source it or is it irrelevant?}
+
+\sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
 
@@
 
-To have any hope of distributing pipelines, \sys first needs to understand the distribution characteristics of their individual commands.
-This understanding is encoded in a command's \emph{parallelizability} class.
-Informally, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
-It focuses on capturing characteristics that are important for distribution, rather than the entire the observable behavior of a command.
+To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they use.
+This information is encoded in a command's \emph{parallelizability class}.
+Intuitively, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
+It focuses on capturing characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 
-This section starts by defining these classes, along with a parallelizability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
-Building on this study, it develops a DSL that enables command classification by its developers~\sx{ext}.
-In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers, as presented in later sections.
+This section starts by defining these classes, along with a parallelizability study of the commands in {\sc POSIX} and GNU Coreutils~\sx{cmd}.
+Building on this study, it develops a lightweight annotation language that enables command classification by its developers~\sx{ext}.
+\sys in turn uses this language to annotate POSIX and GNU commands and generate their wrappers, as presented in later sections.
 
@@
 % nv TODO: this is a hierarchy
-Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
-These classes are ordered in ascending difficulty of distribution:
+Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on how they interact with state.
+These classes are ordered in ascending difficulty of parallelization:
   later classes require more effort.
@@
     \textbf{Parallelizability Classes}.
-    Broadly, \unix commands can be broken down into to five classes.
+    Broadly, \unix commands can be broken down into to \todo{three} classes according to their parallelizability properties.
   }
 }
-\begin{tabular}{l @{\extracolsep{\fill}} lll}
+\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
 \toprule
-Class                           &  Key    & Example Commands                            & Coreutils       \\ % & POSIX       \\
+Class            &  Key    & Example Commands                            & Coreutils       & POSIX       \\
 \midrule
-Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)    \\ %  &           \\  % 22
-Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)    \\ %  &           \\  % 21
-(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)    \\ %  &           \\  % 27
-Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)      \\ %  &           \\  % 25
-Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     \\ %  &           \\  % 5
+Stateless        & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)    & 22 (21.1\%)          \\  % 22
+Pure             & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)    & 21 (20.1\%)          \\  % 21
+% File-System~     & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)    & 27 (25.9\%)          \\  % 27
+Side-effectful   & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)      & 57 (58.8\%)            \\  % 25
+% Irreversible     & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     & 5  (0.4\%)           \\  % 5
 % \midrule
@@
 \bottomrule
-\end{tabular}
+\end{tabular*}
 \label{tab:classes}
@@
 \heading{Stateless Commands}
-The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
-These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
-At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
+The first class, ~\sta, contains commands that operate on individual elements of a data stream, without maintaining state across invocations.
+These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
+Stateless commands may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but when they receive empty input they return empty output.
 Workloads that use only stateless commands are trivial to parallelize:
-  they do not require any synchronization to maintain correctness, nor caution about where to be split.
+  they do not require any synchronization to maintain correctness, nor caution about where to split inputs.
  % map :: (a -> b) -> [a] -> [b]
 
-Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters within a line, one at a time).
-This property allows further parallelization, a feature that might seem of limited use as these commands are computationally inexpensive, precisely due to their narrow focus.
+Many of these commands (about 1/3 of \sta) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
+This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
 However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
@@
   completely if parallelized.}
+\nv{Can we please do this in 4? The difference in size between \S3 and \S4 is
+concerning.. }
 
 \heading{Pure Commands}
-The second class contains pure commands \pur.
-These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
-The details of this state and its propagation during piecewise element processing affect their parallelizability characteristics.
-
-Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
-Others, such as \ttt{sha1sum}, maintain more complex state that has to be updated in a sequential fashion.
+The second class,~ \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
+The details of this state and its propagation during element processing affect their parallelizability characteristics.
+Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc} simply maintains a counter.
+Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
+Commands such as \ttt{sha1sum} maintain even more complex state that has to be updated sequentially.
 
@@
 A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
-Such constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $fold$ function.
+\todo{Such constraints affect task parallelism, but not data parallelism}:
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by a $fold$ that merges the results.
 
@@
 The third class contains commands that access the file-system~\dfs.
-This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
-In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.
+This is the largest class, because the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
+In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
 
@@
 
-\heading{Discussion}
-There are a few technical details about \sys's design decisions worth noting.
-
-In terms of inputs, \sys supports commands with multiple input (resp. output) streams, by prepending (resp. appending) shell built-ins that merge input (resp. split output) streams~\sx{impl}.
-Thus, a command can be modelled as having one input and one output stream, a representation that is used throughout \sys including the extensibility DSL~\sx{ext}.
-
-The focus for the rest paper will be commands in \sta and \pur.
-\sys currently treats other commands as non-distributable~\sx{discussion}.
-In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
-Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
+% \heading{Discussion}
+% There are a few technical details about \sys's design decisions worth noting.
+
+% %\nv{I took this out because it tooks about \sys's design/implementation that goes far afield}
+% In terms of inputs, \sys supports commands with multiple input (resp. output) streams, by prepending (resp. appending) shell built-ins that merge input (resp. split output) streams~\sx{impl}.
+% Thus, a command can be modelled as having one input and one output stream, a representation that is used throughout \sys including the extensibility DSL~\sx{ext}.
+% 
+% %\nv{I took this out because there are no more classes now}
+% The focus for the rest paper will be commands in \sta and \pur.
+% \sys currently treats other commands as non-distributable~\sx{discussion}.
+% In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
+% Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
 
@@
 \label{ext}
+
+\nv{Add note about breadth of commands and the impossibility of our task.}
+% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
 

### 2020-05-18 pldi/draft.tex
Re-frame parallelizability discussion

@@
 
+\newcommand{\sta}{\cn{\textsc{S}}\xspace}
 \newcommand{\pur}{\cn{\textsc{P}}\xspace}
-\newcommand{\sta}{\cn{\textsc{S}}\xspace}
+\newcommand{\npu}{\cn{\textsc{N}}\xspace}
+\newcommand{\sid}{\cn{\textsc{E}}\xspace}
 \newcommand{\dfs}{\cn{\textsc{F}}\xspace}
-\newcommand{\sid}{\cn{\textsc{E}}\xspace}
 \newcommand{\irr}{\cn{\textsc{I}}\xspace}
 
+\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
 \newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
-\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
-\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
+\newcommand{\tnpu}{\tcn{\textsc{N}}\xspace}
 \newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
 \newcommand{\tirr}{\tcn{\textsc{I}}\xspace}
+\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
 
@@
 \toprule
-Class            &  Key    & Example Commands                            & Coreutils       & POSIX       \\
+Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
 \midrule
 Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 22 (21.1\%)          \\  % 22
-Pure             & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)    & 21 (20.1\%)          \\  % 21
-% File-System~     & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)    & 27 (25.9\%)          \\  % 27
+Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  \todo{21 (20.1\%)}    & 21 (20.1\%)          \\  % 21
+Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  \todo{27 (25.9\%)}    & 27 (25.9\%)          \\  % 27
 Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 57 (58.8\%)            \\  % 25
@@
 \heading{Stateless Commands}
-The first class, ~\sta, contains commands that operate on individual elements of a data stream, without maintaining state across invocations.
+The first class, \sta, contains commands that operate on individual elements of a data stream, without maintaining state across invocations.
 These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
@@
 
-\heading{Pure Commands}
-The second class,~ \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
+\heading{Parallelizable Pure Commands}
+The second class, \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
 The details of this state and its propagation during element processing affect their parallelizability characteristics.
@@
 Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
-Commands such as \ttt{sha1sum} maintain even more complex state that has to be updated sequentially.
+
+% \begin{figure}[t]
+% \centering
+% \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_classes.pdf}}
+% \caption{
+%   \textbf{\sys's parallelizability classes.}
+% }
+% \label{fig:example}
+% \end{figure}
 
@@
 
-\heading{File-system Commands}
-The third class contains commands that access the file-system~\dfs.
-This is the largest class, because the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
+\heading{Non-parallelizable Pure Commands}
+The third class, \npu, contains commands that, while purely functional, cannot be parallelized.
+This is because their internal state depends on prior state in the same pass in non-trivial ways. % should we say something about state machines?
+For example, hashing commands such as \ttt{sha1sum} maintain complex state that has to be updated sequentially.
+If parallelized on a single input, each \todo{stage} would need to repeat the work of all the previous stage foregoing any parallelism benefits.
+
+It is worth noting that while these commands are not parallelizable for a single input, they are still parallelizable across different inputs.
+For example, a web crawler involving hash generation for comparing individual pages would allow \ttt{sha1sum} to proceed in parallel for different pages.
+
+\heading{Side-effectful Commands}
+The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
+Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.
+
+This is the largest class because the file and file-system abstractions are central to the \unix design and philosophy~\cite{unix}.
 In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
 
-% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
-Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
-Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for all the commands in \dfs.
-The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---\eg \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
-% these operate by changing an inode
-Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
-Operations using different identifiers pose no problems;
-  operations sharing a single identifier might introduce problems---but these risks exist even in the non-distributed case.
-Such pathological cases are indeed possible, but are rare enough to not pose a concern---and have been, since \unix's inception~\cite[\S3.6]{unix}.
-
-% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.
-
-\heading{Side-effectful Commands}
-Another large class of commands are ones that have side-effects to memory~\sid.
-Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.
-
-A command in \sid will require some form of concurrency control to distribute property.
-However, because its side-effects stay within the distributed \unix environment, they can be manipulated using a transactional control mechanism---including optimistic and multi-versioning schemes that have the ability to roll-back side-effects.
-
-The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
+\nv{
+  A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
   For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
-The benefits of a scalable transactional scheme are even greater when the workload features only or mostly read accesses.
-
-\heading{Irreversibly Side-effectful Commands}
-Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
-Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.
+  Why aren't these parallelizable?
+}
 
@@
 
-\nv{Add note about breadth of commands and the impossibility of our task.}
+\nv{Note command breadth and the impossibility of our task.}
 % https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

### 2020-05-18 pldi/draft.tex
Update section 3

@@
 \newcommand{\fixme}[1]{[{\color{red}#1}]}
-\newcommand{\nv}[1]{[{\color{cyan}#1 --- nv}]}
-\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
-\newcommand{\km}[1]{[{\color{blue}#1 --- km}]}
+\newcommand{\nv}[1]{[{\color{cyan}nv: #1}]}
+\newcommand{\kk}[1]{[{\color{magenta}kk: #1}]}
+\newcommand{\km}[1]{[{\color{blue}km: #1}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
-  basicstyle=\small\ttfamily,  % the size of the fonts that are used for the code
+  basicstyle=\normalsize\ttfamily,  % the size of the fonts that are used for the code
   upquote=true,
@@
     \textbf{Parallelizability Classes}.
-    Broadly, \unix commands can be broken down into to \todo{three} classes according to their parallelizability properties.
+    Broadly, \unix commands can be broken down into to \todo{four} classes according to their parallelizability properties.
   }
@@
 
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_classes.pdf}}
-% \caption{
-%   \textbf{\sys's parallelizability classes.}
-% }
-% \label{fig:example}
-% \end{figure}
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
+\caption{
+  \textbf{Hierarchy of parallelizability classes.}
+  Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
+  \nv{The figure is incorrect---it's actually a lattice.}
+}
+\vspace{-15pt}
+\label{fig:example}
+\end{figure}
 
@@
   \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by a $fold$ that merges the results.
+
+\kk{Maybe show an example of map-fold for sort or wc}
 
@@
 
-This is the largest class because the file and file-system abstractions are central to the \unix design and philosophy~\cite{unix}.
+This is the largest class because it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
 In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
@@
 
-\subsection{Extensibility}
+\subsection{Extensibility Annotations}
 \label{ext}
 
-\nv{Note command breadth and the impossibility of our task.}
+\nv{Note command breadth/impossibility of our task.}
 % https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
@@
 
-% FIXME: \nv{TODO: remember to bubble up(me)}
-An important characteristic of the \unix shell is its
-language-agnostic extensibility~\sx{bg}.  Users composing pipelines
-are free to install custom commands available from various sources.
-Being general programs, commands are developed in a variety of
-languages and are extended over long periods of time.
-%% \kk{Text point: This restricts the space of solutions for
-%%   distribution of shell scripts}
-As a result, any static characterization of shell commands would bs
-unsatisfactory and would quickly become obsolete, supporting only a
-small subset of the commands needed in practice.  This poses a
-challenge to the developers of individual commands---how can they
-augment their commands to capture information critical to \sys without
-too much effort?
-%% \kk{Text point: The characteristics of a good solution}
-
-To address this challenge, \sys arms command developers with a
-lightweight, parallelizability annotation language (\todo{PAL}) that
-allows devevelopers to express semantic information about a command's
-parallelizability.  The PAL can be used by both developers of new
-commands, as well as maintainers of existing commands that can express
-additions or changes to the command's implmentation or
-interface. \todo{The two main challenges that PAL has to address is
-  supporting commands with mutiple inputs and outputs, and handle
-  command-line options.}
-
-\heading{Command Inputs and Outputs}
-
-Together with a command's parallelizability class (i.e. \sta, \pur,
-etc), \sys needs to know certain information about a command's inputs
-and outputs in order to construct a dataflow graph. More precisely,
-given a command and its arguments, \sys needs to know its inputs and
-outputs to connect commands in the dataflow graph with edges, as well
-as the order in which a command consumes its input; this is important
-for \sys to know how to order parallel instances of a command after it
-has performed any \todo{optimizations}. Consider the example of
-\ttt{grep "foo" f1 f2}, which can be parallelized as follows if we
-know that \ttt{grep} first reads from \ttt{f1} and then from \ttt{f2}.
+% % FIXME: \nv{TODO: remember to bubble up(me)}
+% An important characteristic of the \unix shell is its
+% language-agnostic extensibility~\sx{bg}.  Users composing pipelines
+% are free to install custom commands available from various sources.
+% Being general programs, commands are developed in a variety of
+% languages and are extended over long periods of time.
+% %% \kk{Text point: This restricts the space of solutions for
+% %%   distribution of shell scripts}
+% As a result, any static characterization of shell commands would bs
+% unsatisfactory and would quickly become obsolete, supporting only a
+% small subset of the commands needed in practice.  This poses a
+% challenge to the developers of individual commands---how can they
+% augment their commands to capture information critical to \sys without
+% too much effort?
+% %% \kk{Text point: The characteristics of a good solution}
+
+To address the challenge of language-agnostic extensibility~\sx{bg}, \sys allows communicating several key details about command parallelizability through lightweight annotations.
+These annotations can be used by both developers of new commands---which includes users developing their own scripts---as well as developers maintaining existing commands.
+The latter can express additions or changes to the command's implmentation or interface, important as commands are maintained or extended over long periods of time.
+
+\heading{Key Concerns}
+\sys's annotations focused on three crucial concerns:
+  (i) a command's parallelizability class,
+  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
+  (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
+As the first concern was discussed extensively in the previous section, we now focus on the latter two.\footnote{
+  \nv{Note to self: until now we have been talking only about parallelizability, and now we introduce two concerns that haven't been mentioned before.}
+}
+
+To be able to accurately construct the DFG representation, \sys needs to know certain information about a command's inputs and outputs.
+There are couple of reasons for this.
+One reason is due to \sys's DFG representation, which connects commands with each other;
+  to perform this connection correctly \sys needs to know the outputs of a command to correctly connect them with the inputs of the next.
+A second reason is due to ordering:
+  as some commands consume their inputs in a certain order, this order needs to be maintained in the parallel version of the program.
+For example, consider the parallel version of the expression \ttt{grep "foo" f1 f2}:
 
@@
 
-\heading{Command-line Options}
-As described earlier~\sx{bg}, com\-mand-line options and environment
-variables can affect the operation of a command.  Options are a
-particularly popular way to control a command, directly affecting a
-command's parallelizability classification.  That is, commands are
-assigned to default classes, and options can change the class in which
-they belong.  As a simple example, consider \ttt{cat}.  By default
-\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
-has to keep track of a counter and output the line number with each line.
+\noindent
+This is correct only because \sys knows that \ttt{grep} in the sequential program reads from \ttt{f1} first and from \ttt{f2} second.
+
+Command flags (options) are a particularly popular way to control a command's execution, directly affecting its parallelizability classification.
+Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
+For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.
+
+As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
+For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} that writes debugging output to a file, it ends up in \sta.
 
@@
 %% This is actually wrong. sort -h is not stateless since if called several times it produces the same output many times.
+%% nv: the above does not mean it is not stateless; if apply `map` multiple times on an input I will get combined results.
 
@@
 %% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
-
-\tr{
-\begin{figure}
-  \centering
-  \begin{grammar}
-    <option> ::= `-' <string>
-
-    <category> ::= `stateless' | `pure' | ...
-
-    <maybe-int> ::= | <int>
-
-    <arg> ::= `args[' <int> `]'
-
-    <args> ::= <arg>
-    \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-    <input> ::= `stdin' | <args>
-
-    <inputs> ::= <input>
-    \alt <input> `,' <inputs>
-
-    <output> ::= `stdout' | <arg>
-
-    <option-pred> ::= <option>
-    \alt `value' <option> = <string>
-    \alt `not' <option-pred>
-    \alt <option-pred> `or' <option-pred>
-    \alt <option-pred> `and' <option-pred>
-
-    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
-
-    <predicate> ::= <option-pred> `=>' <assignment>
-
-    <pred-list> ::= `|' <predicate> <pred-list>
-    \alt `|' `otherwise' `=>' <assignment>
-
-    <command> ::= <name> `\{' <pred-list> `\}'
-
-    <command-list> ::= <command>
-    \alt <command> <command-list>
-  \end{grammar}
-  \caption{
-\textbf{Parallelizability description language.}
-The DSL captures important information regarding the parallelizability of a command.
-\vspace{-10pt}
-}
-  \label{fig:dsl}
-\end{figure}
-}
 
@@
 
-\heading{Parallelizability Annotation Language}
-
-To illustrate how the proposed PAL addresses the above challenges, we
-will follow the example of the annotation record for \ttt{comm}, a
-command that performs a ``join'' on its two inputs. When \ttt{comm} is
-invoked without any option, it produces a three-column output. The
-first column contains lines that are unique to the first file, the
-second column contains lines that are unique to the second file, and
-the last column contains lines that exist in both files. A user can
-invoke \ttt{comm} with options \ttt{-1}, \ttt{-2}, or \ttt{-3} to
-suppress some of its output columns.
-%
-\begin{lstlisting}[float=h, numbers=none,caption={PAL record for \ttt{comm}},label={lst:pal-record-comm}]
+\heading{Example Annotations}
+A record contains a set of clauses, each one identified by a predicate on the commands options.
+Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
+Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
+% \footnote{
+%   Note that the single character \ttt{-} is not an option as it is most often used to designate that a command reads from its standard input.
+%   \nv{
+%     There are options of of the form --f "x; y", and BSD options do not take \ttt{-} (like \ttt{ps aux}).
+%     I am also not sure how important this last sentence is, as it did not answer a question I had---I expected arguments to be anything: if soneome wants to put ``pizza'' the DSL should not disallow it.
+%   }
+% }
+The grammar of the language for specifying annotations is shown in Appendix~\ref{alg}.
+
+For example, consider \ttt{comm}---a command that performs a join-like operation on its two inputs to identify common elements.
+When \ttt{comm} is invoked without any flags, it produces a three-column output:
+  the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
+A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output columns.
+Its annotation record is shown below:
+%
+\begin{lstlisting}[language=sh, numbers=none]
 comm {
-    | -1 and -3 =>
+  | -1 /\ -3 =>
     (stateless, [args[1]], [stdout])
-    | -2 and -3 =>
+  | -2 /\ -3 =>
     (stateless, [args[0]], [stdout])
-    | otherwise =>
+  | _  =>
     (pure, [args[0], args[1]], [stdout])
@@
 
-The PAL annotation record for \ttt{comm} is shown in
-\cref{lst:pal-record-comm}. A record contains a set of clauses, each
-one identified by a predicate on the commands options. Each clause
-maps to an assignment for the command, indicating its class and the
-sequence of its inputs and outputs. \todo{PAL assumes} that arguments
-of a command can be separated to options and file arguments (checking
-if they start with a \ttt{-}~\footnote{Note that the single character
-  \ttt{-} is not an option as it is most often used to designate that
-  a command reads from its standard input}).
-
-In the record for \ttt{comm} the first two clauses describe \ttt{comm}
-invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}. In the
-first case, since the first and third columns are suppressed,
-\ttt{comm} can be considered \sta by regarding its first file argument
-a static ``configuration'' input. \kk{Does this need more
-  explanation?} The general case (that includes \ttt{comm} invocations
-without any options) is captured by the third clause. There,
-\ttt{comm} is classified as \pur, since it needs to wait until it
-reads both files completely before it can output its results. Its
-inputs are its first and second arguments (excluding the options), and
-its output is produced in \ttt{stdout}.
+\noindent
+The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
+In the first case, since the first and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
+  this input accompanies every replicated instance of the command executing in parallel.
+The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
+There, \ttt{comm} is classified as \pur, since it needs to wait until it reads both files completely before it can output its results.
+Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.
 
@@
 
-\kk{We could add the grammar of PAL in an appendix if we want}
-
-\tr{The complete syntax of the PAL is shown in (Fig.~\ref{fig:dsl})}
 %% \kk{I
@@
 
-\heading{Parallelizable Pure Commands}
-
-In addition to the commands in \sta that can be parallelized in a
-straightforward way, there also exists a subclass of commands in \pur
-that can be parallelized using divide-and-conquer techniques (as
-mentioned in \cref{cmd}). \kk{Maybe show an example of map-fold for
-  sort or wc. Either here or in \cref{cmd}.} In order to support the
-parallelization of arbitrary commands in this category, \sys allows
-command developers to provide custom map and fold steps for them,
-where the map is in \sta and the fold in \pur, and their composition
-produces the same output with the initial command. \sys can then
-replace occurrences of these commands with their map and fold parts,
-enabling their parallelization. \todo{We have defined these primitives
-  for most of the parallelizable pure commands in coreutils, providing
-  a foundation for this effort.}
+\heading{Providing Combinators}
+For most commands, the annotations are enough to allow parallelization: 
+  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{cmd}.
+Unfortunately, there are commands with less trivial tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
+
+To support the parallelization of arbitrary commands in \pur, \sys allows command developers to provide custom $map$ and $fold$ functions.
+In line with \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
+  (i) $map$ in \sta and $fold$ is in \pur,
+  (ii) $map$ can consume (or extend) the output of the original command and  $fold$ can consume (and combine) the results of multiple $map$ invocations, and
+  (iii) their composition produces the same output with the original command.
+If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their $map$ and $fold$ equivalents, enabling their parallelization.
+
+\sys defines the combiners for most of the POSIX and Coreutils commands in \pur, which operate as both \sys's standard library and exemplar for community attempts to tackle other commands.
 
@@
 
+\appendix
+
+\newpage
+
+\section{Annotation Language Grammar}
+\label{alg}
+
+% \begin{figure}
+%   \centering
+%   \begin{grammar}
+%     <option> ::= `-' <string>
+% 
+%     <category> ::= `stateless' | `pure' | ...
+% 
+%     <maybe-int> ::= | <int>
+% 
+%     <arg> ::= `args[' <int> `]'
+% 
+%     <args> ::= <arg>
+%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
+% 
+%     <input> ::= `stdin' | <args>
+% 
+%     <inputs> ::= <input>
+%     \alt <input> `,' <inputs>
+% 
+%     <output> ::= `stdout' | <arg>
+% 
+%     <option-pred> ::= <option>
+%     \alt `value' <option> = <string>
+%     \alt `not' <option-pred>
+%     \alt <option-pred> `or' <option-pred>
+%     \alt <option-pred> `and' <option-pred>
+% 
+%     <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
+% 
+%     <predicate> ::= <option-pred> `=>' <assignment>
+% 
+%     <pred-list> ::= `|' <predicate> <pred-list>
+%     \alt `|' `otherwise' `=>' <assignment>
+% 
+%     <command> ::= <name> `\{' <pred-list> `\}'
+% 
+%     <command-list> ::= <command>
+%     \alt <command> <command-list>
+%   \end{grammar}
+%   \caption{
+%   \textbf{Parallelizability description language.}
+%     The DSL captures important information regarding the parallelizability of a command.
+%   }
+%   \label{fig:dsl}
+% \end{figure}
+
 %% %% Appendix

### 2020-05-18 pldi/draft.tex
Related work and conclusions

@@
   Anonymous Author(s)\\
-  \normalsize{Additional Anonymized Material: \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.}
+  \normalsize{Additional Anonymized Material: \href{https://git.io/JfEgr}{https://git.io/JfEgr}.}
 }
@@
 
-\sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
+\sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/JfEgr}{https://git.io/JfEgr}.
 
@@
 
-At a high level, techniques for re-writing programs to exploit
-distribution fall under a wide spectrum that ranges from fully manual
-(but, ideally, optimal) to fully automated (but potentially
-sub-optimal) distribution (Fig.~\ref{fig:spectrum}).  A significant
-concern with approaches that require input from developers is
-correctness: developers---often neither experts in distributed systems
-nor educated on the internals of the code they depend on---can easily
-break the semantics of centralized programs, especially when these
-depend on existing code (such as libraries or shell commands).
-
-\heading{Distributed Operating Systems and Shells}
-A plethora of systems assist in the construction of distributed
-software.  Distributed operating systems~\cite{ousterhout1988sprite,
-  mullender1990amoeba, pike1990plan9, barak1998mosix} and programming
-languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide
-many distribution problems, as long as a program uses the provided
-abstractions---which are strongly coupled to the underlying operating
-or runtime system.
+At a high level, techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
+A significant concern with approaches that require user input is correctness:
+   users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the semantics of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
+
+\heading{Parallel Shell Scripting}
+Several utilities expose commands for specifying parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}.
+Different from \sys, their effectiveness is predicated upon explicit and careful invocation and is limited to embarrassingly parallel (and short) programs.
+Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
+For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs};
+   it also introduces (and depends on) other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
+\sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently with minimal user effort---if any.
+
+Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target parallelism.
+Here too, however, users are expected to manually rewrite scripts to exploit these new primitives, contrary to \sys.
+
+Recently, the Smoosh authors have argued that certain shell features are useful for making concurrency explicit~\cite{smoosh:18}.
+Their argument is somewhat antithetical to \sys, which argues that users should experience mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
+
+% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
+\heading{Low-level Parallelization}
+Instruction-level parallelization has a long history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
+These systems operate at a lower level than \sys (\eg that of instructions or loops rather than the boundaries of programs that are part of a script), operating in a single target environment, and typically do not exploit high-level annotations.
+
+More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
+These tools simplify the expression of parallelism, but still require users to get involved in discovering and exposing parallelism.
+% Moreover, the insights behind these attempts are significantly different from \sys's, as they extract parallelism statically during compilation instead of dynamically during runtime.
+%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
+
+\heading{Correct Parallelization of Dataflow Graphs}
+The DFG is a prevalent model in several areas of data processing (including batch-~\cite{mapreduce:08, spark:12} and stream-processing ~\cite{murray2013naiad, carbone2015flink} systems).
+Despite its popularity, most systems perform optimizations that do not preserve its semantics, introducing subtle erroneous behaviours.
+Recent work~\cite{HSSGG2014, SHGW2015, MSAIT2019} attempts to address this issue by performing optimizations only in cases where correctness is preserved.
+\sys draws inspiration from these efforts, as it attempts transformations that maintain the program's correctness with respect to the sequential execution.
+
+\sys's proposed DFG model is diverges from prior works, as nodes it captures order.
+This is due to the intricacies of the \unix model---its streams, its argument processing, and concatenation operators all maintain ordering.
+% As a result, previous work is not directly applicable.
+
+\heading{Parallel Userspace Environments}
+By focusing on simplifying the development of distributed programs, a plethora of environments assist in the construction of parallel software.
+Such systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} or languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide many of the challenges of dealing with concurrency.
+They do so as long as a program leverages the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
+A notable example is Plan9's \ttt{rc} shell that provided an interface for the construction of general programs.
+Unfortunately these environments are not backward-compatible with the \unix shell, and often focus primarily on hiding the existence of a network---providing the illusion of a single-system image---rather than automatically accelerating scripts through parallel processing.
 % simplify many of the problems of distribution and
 
-Several packages expose commands for specifying parallelism on modern
-\unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun},
-\textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU}
-\ttt{parallel}~\cite{Tange2011a}. % or any manual
-rewriting~\cite{mapreduce:08, ciel:11, spark:12}.  Different from
-\sys, their effectiveness is predicated upon explicit and careful
-invocation and is limited to embarrassingly parallel (and short)
-programs.  Often, these commands provide options to support an array
-of special sub-cases---a stark contradiction to the celebrated \unix
-philosophy.  For example, \ttt{parallel} contains flags such as
-\ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix
-user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs}; it also
-includes other programs with complex semantics, such as the ability
-transfer files between computers, separate text files, and parse CSV.
-\sys embraces the \unix philosophy and simply attempts to rewrite
-pipelines more efficiently.
-
-Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add
-primitives for non-linear pipe topologies---some of which target
-distribution.  Here too, however, developers are expected to manually
-rewrite scripts to exploit these new primitives.
-
-Recently, the Smoosh authors have argued that certain shell features
-are useful for making concurrency explicit~\cite{smoosh:18}.  Their
-argument is somewhat antithetical to \sys, % but rather that the shell
-language is itself an good DSL for concurrency; which argues for
-automatically (and correcctly) parallelizing pipelines when users have
-not expressed parallelism explicitly---hence
-distribution-\emph{oblivious} shell scripting.
-
 \heading{Cluster Computing}
-Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
-  naiad:13} and domain-specific languages~\cite{alvaro2011consistency,
-  distal:13, meiklejohn2015lasp} go a step further in terms of
-automation by providing a few carefully-designed primitives.  These
-primitives make strong assumptions about the nature of the
-computation---\eg strongly-eventual commutative functions that can
-proceed in parallel.  By targeting specific classes of computation
-(\emph{viz.} \sys's parallelizability), they render distributed
-computation efficiently.
+Several frameworks such as Phoenix~\cite{} and Raftlib~\cite{} offer full automation as long as their use leverages special primitives---\eg map-reduce-style primitives for Phoenix.
+These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
+By targeting specific classes of computation (\emph{viz.} \sys's parallelizability), they can significantly optimize for their target domains.
 
 % Lack of generality, no formal guaratees, their setup and is too tedious 
-\sys chooses a more general approach: it does not require users to set
-up a new framework for every new class of computation they need to
-execute, nor to rewrite different parts of their computation using the
-specific abstractions provided by each framework they use.  It also
-takes a more principled approach: by developing the semantic model
-outlined earlier, it can guarantee that transformations are correct
-with respect to the sequential program.
+\sys chooses a more general approach: it does not require users to set up a new framework for every new class of computation they need to execute, nor to rewrite different parts of their computation using the specific abstractions provided by each framework they use.
+It also takes a more principled approach: by developing the semantic model outlined earlier, it can focus on transformations that are correct with respect to the sequential program.
 
@@
 
-\heading{Annotation-based Distribution}
-Recent systems on light-touch distribution, such as
-Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
-than cluster-computing frameworks.  They also lower (but not entirely
-eliminate) developer effort by automating most of the rewriting
-process.
-% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-For the critical points where user input is required, they provide a
-high-level declarative DSL with varying (but limited) degrees of
-expressive power.
-% Annotations are used by ---complicated by the use of third-party
-% libraries similar to \sys's extensions.
-Unfortunately, a challenge with this approach is that users composing
-applications by using third-party libraries do not understand their
-parallelizability details well-enough to annotate them.
-
-\sys occupies a different point in the design space, by removing any work from composers and adding a small load to developers.
-% More specifically, it is designed to be used automatically by composers,
-Its DSL is designed to be more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared towards the developers of these components---not the users composing them.
+% \heading{Annotation-based Transformations}
+% \sys inspiration from recent systems such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19} which, as long as a developer has provided a few high-level annotations, they take care of automating scale-out.
+% % Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
+% These systems works on single-language, single-runtime environments 
+% 
+% For the critical points where user input is required, they provide a
+% high-level declarative DSL with varying (but limited) degrees of
+% expressive power.
+% % Annotations are used by ---complicated by the use of third-party
+% % libraries similar to \sys's extensions.
+% Unfortunately, a challenge with this approach is that users composing
+% applications by using third-party libraries do not understand their
+% parallelizability details well-enough to annotate them.
+% 
+% \sys occupies a different point in the design space, by removing any work from composers and adding a small load to developers.
+% % More specifically, it is designed to be used automatically by composers,
+% Its DSL is designed to be more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared towards the developers of these components---not the users composing them.
 
@@
 % The side-effects present in the shell dwarf those of functional languages
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Low-level Parallelization}
-Instruction-level parallelization has a longer history, starting from
-explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2}
-and continuing with compilers that attempt to automatically extract
-parallelism~\cite{padua1993polaris,hall1996maximizing}.  These systems
-operate at a lower level than \sys (\eg that of instructions or loops
-instead of pipe boundaries) and typically do not exploit runtime
-information.
-
-More recent work focuses on extracting parallelism from
-domain-specific programming models~\cite{cilk5, streamIt, galois} and
-interactive parallelization tools~\cite{parascope, ipat}.  These tools
-simplify the expression of parallelism, but still require programmers
-to get involved in discovering and exposing parallelism.
-%  actively involve the programmer in the detection and mapping of application parallelism, but still demand great effort from the user. 
-Moreover, the insights behind these attempts are significantly different from ours, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
 
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
-\caption{
-  \textbf{Manual--automated distribution spectrum.}
-	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-}
-\vspace{-15pt}
-\label{fig:spectrum}
-\end{figure}
+% \begin{figure}[t]
+% \centering
+% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
+% \caption{
+%   \textbf{Manual--automated distribution spectrum.}
+% 	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
+% }
+% \vspace{-15pt}
+% \label{fig:spectrum}
+% \end{figure}
+%
 
@@
 %   optimizations etc.}
-
-\heading{Correct Distribution of Dataflow Graphs}
-%
-The dataflow graph is a prevalent model in several areas of data
-processing, as it is used by several batch-~\cite{mapreduce:08,
-  spark:12} and stream-processing systems~\cite{murray2013naiad,
-  carbone2015flink}. Despite its popularity, most systems perform
-optimizations that do not preserve its semantics, introducing subtle
-erroneous behaviours. Recent work~\cite{HSSGG2014, SHGW2015,
-  MSAIT2019} attempts to address this issue by performing
-optimizations only in cases where correctness is preserved. \sys draws
-inspiration from these efforts, as we are also interested in
-optimizations that exploit parallelization but preserve the semantics
-of the underlying model. However, our proposed dataflow model is
-different from these other works, as nodes in our model read from
-their inputs \emph{in sequence}. Due to this intricacy, previous work
-is not straightforwardly applicable.
 
@@
 
-\section{Discussion}
+\section{Discussion \& Conclusion}
 \label{discussion}
 
-This section reflects on \sys's design and implementation, touches on future work, and concludes the paper.
-
-\heading{Limitations}
-Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
-
-\sys does not adequately handle failures or network partitions in the general sense, which would require re-scheduling passes on failed replicas.
-Prior work on fault-tolerant distributed stream processing can be of significant aid here.
-
-The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
-Prior work on operator placement of distributed dataflow graphs can be used to build a more sophisticated planner.
-
-\sys is conservative in the shell subsets that it handles, in order to
-avoid introducing unsafe behaviours. The point of this work was not to be
-able to handle a complete distributable subset of the shell, but
-rather to use significant part of it to demonstrate performance benefits.
-
-% Shell is highly dynamic, thus making it impossible to design sound
-% meaningful analyses to determine distributable regions and maximal
-% parallelization. In order to address this particularity, \sys could be
-% tighly integrated with a posix compliant shell
-% interpreter~\cite{smoosh:20} so that dynamic information can be
-% acquired selectively to aid the analysis and optimization process.
-
-\heading{Future Work}
-% \label{limitation}
-% \item No cycles (multiple commands writing and reading from the same file)
-There are several worthwhile directions for future work.  One
-direction would be an extension of the parallelizability analysis of
-\sys, in order to guide it using dynamic information. This could
-enable a formal study of the analysis~\sx{impl} to show that it
-returns regions that are safe to distribute.  Recent work on
-formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
-such an attempt possible.
-
-Another direction is to extend \sys to handle other classes outside \sta and \pur.
-These two classes enjoy a certain popularity with quick, one-off pipelines and are the easiest to distribute, but 
-it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).
-This direction would naturally introduce considerations about replication, consistency, and failure recovery.
-
+This paper presented \sys, a shell variant that parallelizes shell programs mostly automatically.
+\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
+To achieve its goal, \sys
+  decomposes primitives into parallelizability classes,
+  identifies high-parallelizability stages,
+  applies a series of transformations according to a DFG model,
+  and orchestrates the execution of the parallel program.
+% Its runtime component provides orchestration and planning support during the execution of the program.
+
+\sys's current limitations revolve around shell expansion.
+Currently, \sys does not attempt to parallelize program fragments for which variables have not been expanded.
+Without substantial effort, \sys could be modified to at least suggest expansion of these fragments so that developers provide concrete values.
+More significant effort would be required however to fully exploit latent parallelism in fragments that have not been expanding.
+This effort would require tying \sys with the shell interpretation rather than having it as a compilation pass and is left for future work.
+
+Despite limitation related to expansions, \sys can lead to significant benefits for shell users.
+Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all minimal or zero developer effort.
+
+% \heading{Limitations}
+% Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
+% 
+% \sys does not adequately handle failures or network partitions in the general sense, which would require re-scheduling passes on failed replicas.
+% Prior work on fault-tolerant distributed stream processing can be of significant aid here.
+% 
+% The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
+% Prior work on operator placement of distributed dataflow graphs can be used to build a more sophisticated planner.
+% 
+% \sys is conservative in the shell subsets that it handles, in order to
+% avoid introducing unsafe behaviours. The point of this work was not to be
+% able to handle a complete distributable subset of the shell, but
+% rather to use significant part of it to demonstrate performance benefits.
+% 
+% % Shell is highly dynamic, thus making it impossible to design sound
+% % meaningful analyses to determine distributable regions and maximal
+% % parallelization. In order to address this particularity, \sys could be
+% % tighly integrated with a posix compliant shell
+% % interpreter~\cite{smoosh:20} so that dynamic information can be
+% % acquired selectively to aid the analysis and optimization process.
+% 
+% 
+% \heading{Future Work}
+% % \label{limitation}
+% % \item No cycles (multiple commands writing and reading from the same file)
+% There are several worthwhile directions for future work.  One
+% direction would be an extension of the parallelizability analysis of
+% \sys, in order to guide it using dynamic information. This could
+% enable a formal study of the analysis~\sx{impl} to show that it
+% returns regions that are safe to distribute.  Recent work on
+% formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
+% such an attempt possible.
+% 
+% Another direction is to extend \sys to handle other classes outside \sta and \pur.
+% These two classes enjoy a certain popularity with quick, one-off pipelines and are the easiest to distribute, but 
+% it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).
+% This direction would naturally introduce considerations about replication, consistency, and failure recovery.
+% 
 % There are two additional assumptions that have to be satisfied in
@@
 
-\heading{Conclusion}
-This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
-\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
-To achieve its goal, \sys
-  decomposes primitives into parallelizability classes,
-  identifies high-parallelizability stages,
-  applies rewriting rules for largest possible subprograms,
-  and orchestrates the execution of the distributed program.
-% Its runtime component provides orchestration and planning support during the execution of the program.
-Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
-
 %% \begin{acks}
@@
 
+\appendix
+
+\section{Annotation Language Grammar}
+\label{alg}
+
+\begin{figure}
+  \centering
+  \begin{grammar}
+    <option> ::= `-' <string>
+
+    <category> ::= `stateless' | `pure' | ...
+
+    <maybe-int> ::= | <int>
+
+    <arg> ::= `args[' <int> `]'
+
+    <args> ::= <arg>
+    \alt `args[' <maybe-int> `:' <maybe-int> `]'
+
+    <input> ::= `stdin' | <args>
+
+    <inputs> ::= <input>
+    \alt <input> `,' <inputs>
+
+    <output> ::= `stdout' | <arg>
+
+    <option-pred> ::= <option>
+    \alt `value' <option> = <string>
+    \alt `not' <option-pred>
+    \alt <option-pred> `or' <option-pred>
+    \alt <option-pred> `and' <option-pred>
+
+    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
+
+    <predicate> ::= <option-pred> `=>' <assignment>
+
+    <pred-list> ::= `|' <predicate> <pred-list>
+    \alt `|' `otherwise' `=>' <assignment>
+
+    <command> ::= <name> `\{' <pred-list> `\}'
+
+    <command-list> ::= <command>
+    \alt <command> <command-list>
+  \end{grammar}
+  \caption{
+  \textbf{Parallelizability description language.}
+    The DSL captures important information regarding the parallelizability of a command.
+  }
+  \label{fig:dsl}
+\end{figure}
+
 %% Bibliography
 \bibliography{./bib}
-
-\appendix
-
-\newpage
-
-\section{Annotation Language Grammar}
-\label{alg}
-
-% \begin{figure}
-%   \centering
-%   \begin{grammar}
-%     <option> ::= `-' <string>
-% 
-%     <category> ::= `stateless' | `pure' | ...
-% 
-%     <maybe-int> ::= | <int>
-% 
-%     <arg> ::= `args[' <int> `]'
-% 
-%     <args> ::= <arg>
-%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
-% 
-%     <input> ::= `stdin' | <args>
-% 
-%     <inputs> ::= <input>
-%     \alt <input> `,' <inputs>
-% 
-%     <output> ::= `stdout' | <arg>
-% 
-%     <option-pred> ::= <option>
-%     \alt `value' <option> = <string>
-%     \alt `not' <option-pred>
-%     \alt <option-pred> `or' <option-pred>
-%     \alt <option-pred> `and' <option-pred>
-% 
-%     <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
-% 
-%     <predicate> ::= <option-pred> `=>' <assignment>
-% 
-%     <pred-list> ::= `|' <predicate> <pred-list>
-%     \alt `|' `otherwise' `=>' <assignment>
-% 
-%     <command> ::= <name> `\{' <pred-list> `\}'
-% 
-%     <command-list> ::= <command>
-%     \alt <command> <command-list>
-%   \end{grammar}
-%   \caption{
-%   \textbf{Parallelizability description language.}
-%     The DSL captures important information regarding the parallelizability of a command.
-%   }
-%   \label{fig:dsl}
-% \end{figure}
 

### 2020-05-18 pldi/draft.tex
a couple of remarks

@@
 [D\kstar]$, where $D$ represents the basic data type of a line of
-characters. This representation captures the majority of shell
+characters.
+\km{Is the number of input/output streams variable?
+If $m$ is the number of input streams, is the number of output streams a function of $m$?
+If this the case, then it is not captured by this type.
+Maybe it would help to show a couple of example operators together with the types of their denotations.
+A graphical representation (as in Fig.~\ref{fig:parallelization-transformation} would also be nice.} This representation captures the majority of shell
 commands. We require that the function $f$ is monotone w.r.t.\ a
@@
 represents the concatenation of the inputs that are consumed one at a
-time and the second argument represents the static inputs. The
+time and the second argument represents the static inputs.
+\km{Again, a graphical presentation of these operators would be useful.
+Maybe with a way to distinguish between between streaming inputs and static inputs?
+Btw, this distinction is not completely clear to me at this point.
+Clearly, not all programs that we can write are streaming (monotone).} The
 quintessential example of \todo{one-input-output} commands is
@@
 \todo{one-input-output} command $f$ is stateless if it is ``linear''
-with respect to concatenation, \ie satisfying the following equation:
+with respect to concatenation \km{It commutes with the operation of concatenation.
+This is the definition of (semigroup) homomorphism.
+I am not sure that linearity is an appropriate term.}, \ie satisfying the following equation:
 
 \[
-\forall x, x', s, f(x.x', s) = f(x, s).f(x', s)
+\forall x, x', s, f(x \cdot x', s) = f(x, s) \cdot f(x', s)
 \]
@@
 \[
-\forall x, x', s, f(x.x', s) = f'(m(x, s).m(x', s), s)
+\forall x, x', s, f(x \cdot x', s) = f'(m(x, s) \cdot m(x', s), s)
 \]
@@
 map $m$ to each of these inputs.
+\km{We had also discussed here the case of operators that can be described as associative and commutative functions on the data (e.g., sum).
+In this case the parallelization has the flavor of data parallelism: one can split the input stream in an arbitrary way (e.g., randomly or using some other load-balancing scheme) and merge the outputs.
+For the more broad class of associative operations (e.g., sort), the formulation of this paragraph makes perfect sense.
+I just want to mention that this feels a lot more like \textbf{batching} than map-reduce-like data parallelization.}
 
@@
 D^*$, and the only behaviour of a dataflow graph is its output, this
-optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
-s)...v(x_n, s)$ can be shown to preserve the behaviour of the
-graph. Furthermore, it can be straightforwardly extended to pure
+optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the behavior of the graph.
+Furthermore, it can be straightforwardly extended to pure
 commands that can be implemented by a map-fold pair $(m, r)$ as $
-v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
-s)$. A visual representation of this transformation can be seen in
-\Cref{fig:parallelization-transformation}.
+v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n, s), s)$. A visual representation of this transformation can be seen in \Cref{fig:parallelization-transformation}.
 

### 2020-05-20 pldi/draft.tex
Comments and edits on OSDI draft

@@
 To address this challenge, we develop \sys, a shell that parallelizes POSIX shell scripts.
+\kk{We need to mention data-parallelism earlier.
+We don't focus on general parallelization but rather data-parallelism (performing operations on disjoint partitions of the input and then merging the results.
+WE need to give a brief idea about that early on so that we don't create false ideas.}
 % \kk{Mention that it outputs shell scripts.}
@@
   (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
-The results demonstrate significant advantages in correctness, manual effort, and performance---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-\kk{I would remove the ``aided further...'' part.}
+The results demonstrate significant advantages in correctness, manual effort, and performance
+ ---with \sys offering an order-of-maginute performance improvements due to its transformations and optimized runtime.
+%% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
 Naturally, these improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls. 
@@
 % In the ongoing debate on climate change,
-Suppose a \todo{climate scientist}    % looking for a non-CS expert---journalist might be too low
+Suppose a \todo{environmental scientist/engineer}    % looking for a non-CS expert---journalist might be too low
   wants to get a quick sense of the changes in maximum temperature over the past century.
@@
 \heading{\unix Features}
-An important \unix feature are pipes (\ttt{|}), introducing task-parallel program composition over character streams.
+Composition in \unix is primarily achieved through pipes (\ttt{|}), a
+construct that allows for task-parallel execution of two commands by
+connecting them with a character stream.
+%% An important \unix feature are pipes (\ttt{|}), introducing
+%% task-parallel program composition over character streams.
 These streams are contiguous character lines separated by newline character (\textsc{NL}), which delineate individual stream elements.
@@
 For example, the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
-The \unix kernel facilitates scheduling, communication, synchronization behind the scenes.
+The \unix kernel facilitates scheduling, communication, and synchronization behind the scenes.
 
@@
 
-Finally, \unix provides composition environment for commands written in any language.
-Many the commands used in practice come built into the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
+Finally, \unix provides an environment in which commands written in any language can be composed.
+Many of the commands used in practice come built into the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
 The fact that commands are developed in a variety of languages---including shell scripts---provides users with a significant flexibility.
@@
 Worse even, the same command may be used with different flags---\eg % \ttt{-d" " -f9} for the first
-  Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+  Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags\kk{I propose dropping the last sentence for this ``---making correct treatment even more difficult''}.
 When a single flag can affect the functionality of a command in non-trivial ways, deciding how to parallelize becomes significantly more difficult.
@@
 % \nv{Are these all the challenges we want to highlight?}
+
+\kk{There are two ways to go here: either mention the challenges that a user would face to manually parallelize a script (that have already been mentioned in the introducion), or mention the challenges that automatic parallelization---this paper---has to address.
+I think that we should focus on the second.}
+
+The first and greatest challenge is the gargantuan number of available
+commands that can be used to construct a shell script. In contrast to
+restricted programming frameworks that are amenable to parallelization
+by supporting a few carefully designed operators (see MapReduce,
+Spark, etc) shell scripts can contain commands that can have arbitrary
+behaviours.  In order to parallelize such a script, each of command
+requires special treatment as to not introduce non-determinism or
+erroneous behaviours. For example, \ttt{grep} can be performed on
+different partitions of its input in parallel without any
+coordination, \ttt{head} requires light coordination, and \ttt{sort}
+requires significant coordination.  To make matters worse, the same
+command may be used with different flags---\eg % \ttt{-d" " -f9} for
+the first Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two
+different sets of flags---that could potentially alter its behaviour,
+making correct treatment even more difficult. A good solution should
+handle several commands but should also be extensible, in the sense
+that newly added or modified commands don't require modifications in
+the solution itself. \kk{The solution to this challenge is already written in the third paragraph of the design overview.}
+
+Another issue is that all commands are black boxes. Therefore it is
+impossible to perform any kind of analysis on the code of the command
+to identify possible parallelization. Therefore a good solution should
+parallelize in a coarse-enough granularity that doesn't require
+opening up commands.
+
+Finally, shell scripts contain several constructs that enforce sequential execution.
+For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), has to be completed before the rest of the script starts executing.
+\kk{The solution is
+  written in the fourth paragraph of the design solution.}
 
@@
 The new script may incorporate parallelism, and is handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
-To discover and enable parallel execution, \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
+To \todo{discover and enable/expose and exploit} data-parallelism in a given shell script,
+%% parallel execution,
+\sys analyzes individual program fragments, addressing the challenges mentioned earlier.
 
@@
 To understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power, \sys leverages this study to define an annotation language for describing a command's key properties with respect to its parallelizability.
-Such light annotations are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they are developing.
+Such light annotations \kk{have to be written once per command (and not once per script) and} are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they are developing.
 % We haven't introduced the notion of flags
@@
 Provided information about the shell constructs and individual commands and their flags, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
-To achieve this, it converts the script into a data-flow graph (DFG) model.
+To achieve this, it converts sections of the script that are amenable to parallelization into a data-flow graph (DFG) representation.
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-This is a flexible representation that allows a series of parallelization transformations to convert the graph into its parallel equivalent, in which DFGs indicate the parallelizable regions.
+This is a flexible representation that enables a series of local transformations that expose data-parallelism, converting the graph into its parallel equivalent.
+%% , in which DFGs indicate the parallelizable regions.
 Further transformations compile the DFG back to a parallel script that uses POSIX shell constructs to guide parallelism explicitly---while aiming at preserving the semantics of the sequential program.
 
-Finally, \sys needs to provide correct, high-performance combinators for certain classes commands, and be proactive about (addressing) several practical challenges related to quirks of the \unix environment.
+Finally, \sys needs to provide correct, high-performance combiners for certain classes commands, and be proactive about (addressing) several practical challenges related to quirks of the \unix environment.
 In terms of combiners, \sys offers a set of linkable components for stream splitting and merging;
@@
 
-To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they use.
-This information is encoded in a command's \emph{parallelizability class}.
+To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
+As mentioned before \sys focuses on parallelizing data-parallel
+commands, i.e. commands that can process their input in parallel, and
+encodes their characteristics by assigning them to
+\emph{parallelizability classes}.
+%% This information is encoded in a command's \emph{parallelizability class}.
 Intuitively, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
@@
 
+\kk{Mention that commands change classes depending on their arguments and that we only write here (and in the table) the class of commands with no flags.
+Then the discussion about flags in the language part makes sense.}
+
 % nv TODO: this is a hierarchy
-Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on how they interact with state.
+Broadly, shell commands can be split into four major classes (summarized in Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
+%% how they interact with state.
 These classes are ordered in ascending difficulty of parallelization:
@@
 \heading{Stateless Commands}
-The first class, \sta, contains commands that operate on individual elements of a data stream, without maintaining state across invocations.
+The first class, \sta, contains commands that operate on individual \todo{lines/elements}~\footnote{\todo{the choice of line as the data-quantum is deliberate, since they strike a nice balance between very coarse grained separation (files) and very fine-grained ones (individual characters).
+We might need to mention that in the introduction of 3.1}} of their input, without maintaining state across invocations.
 These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
@@
 
-Many of these commands (about 1/3 of \sta) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
+\kk{Proposed insertion: The choice of where how to separate data elements would change the allocation of commands in parallelization classes---} Many of these commands (about 1/3 of \sta) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
 This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
 However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
-
-\kk{Do we want to get in more detail about requirements? Commands must
-  be deterministic, they must not do any other side effect (sending
-  signals, etc). These assumptions must be checked when the developer
-  designates the categories. Commands are assumed to always terminate
-  gracefully. \sys changes the termination behaviour of commands as a
-  sequential command that would exit prematurely might not exit
-  completely if parallelized.}
-\nv{Can we please do this in 4? The difference in size between \S3 and \S4 is
-concerning.. }
 
@@
 For example, hashing commands such as \ttt{sha1sum} maintain complex state that has to be updated sequentially.
-If parallelized on a single input, each \todo{stage} would need to repeat the work of all the previous stage foregoing any parallelism benefits.
+If parallelized on a single input, each \todo{stage} would need to wait on the results of all the previous stage foregoing any parallelism benefits.
 
@@
 The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
+\kk{Or commands that do not produce output consuming lines of their input---thus not being amenable to data-parallelism.}
 Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.
@@
 subset of \sta and \pur commands fall in this category.
+
+\kk{Do we want to get in more detail about requirements? Commands must
+  be deterministic, they must not do any other side effect (sending
+  signals, etc). These assumptions must be checked when the developer
+  designates the categories. Commands are assumed to always terminate
+  gracefully. \sys changes the termination behaviour of commands as a
+  sequential command that would exit prematurely might not exit
+  completely if parallelized.}
 

### 2020-05-20 pldi/draft.tex
Pass on Section 4

@@
 A node $f$ of the graph represents a function from a (possibly empty)
-list of input streams to a list of output streams $f : [D\kstar] \to
-[D\kstar]$, where $D$ represents the basic data type of a line of
-characters.
-\km{Is the number of input/output streams variable?
-If $m$ is the number of input streams, is the number of output streams a function of $m$?
-If this the case, then it is not captured by this type.
-Maybe it would help to show a couple of example operators together with the types of their denotations.
-A graphical representation (as in Fig.~\ref{fig:parallelization-transformation} would also be nice.} This representation captures the majority of shell
-commands. We require that the function $f$ is monotone w.r.t.\ a
-lifting of the prefix order for a sequence of inputs \kk{Formalize
-  this lifting if essential for later development}.
+list of inputs
+%% streams
+to a list of outputs
+%% streams
+$f : [D\kstar] \to [D\kstar]$, where $D$ represents the basic data
+type of a line of characters.
+%% \km{Is the number of input/output
+%%   streams variable? If $m$ is the number of input streams, is the
+%%   number of output streams a function of $m$? If this the case, then
+%%   it is not captured by this type. Maybe it would help to show a
+%%   couple of example operators together with the types of their
+%%   denotations. A graphical representation (as in
+%%   Fig.~\ref{fig:parallelization-transformation} would also be nice.}
+%% \kk{Is there a reason to restrict the number of outputs to be a
+%%   function of the number of inputs? I think that we don't need such a
+%%   restriction, therefore the above type should be fine.}
+This representation captures \todo{the majority/all} of commands in
+the \tsta, \tpur, \tnpu. Note that these functions should only produce
+output in the form of files and not perform any other side effect,
+such as sending signals.  We require that the function $f$ is monotone
+w.r.t.\ a lifting of the prefix order for a sequence of inputs
+%% \kk{Formalize
+%%   this lifting if essential for later development}.
 %% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
@@
 
-\heading{\todo{One-input-output} Commands}
-
-\kk{Find a better name.} \todo{One-input-output} commands \todo{are
-  special commands} that consume their inputs one at a time, with the
-possible exception of static input files, and produce one output
-file. These commands can be represented as functions of type $f :
-D\kstar \times [D\kstar] \to D\kstar$, where the first argument
-represents the concatenation of the inputs that are consumed one at a
-time and the second argument represents the static inputs.
-\km{Again, a graphical presentation of these operators would be useful.
-Maybe with a way to distinguish between between streaming inputs and static inputs?
-Btw, this distinction is not completely clear to me at this point.
-Clearly, not all programs that we can write are streaming (monotone).} The
-quintessential example of \todo{one-input-output} commands is
-\ttt{cat}, that consumes its inputs one at a time, producing their
-concatenation as output. We focus on these commands because a large
-subset of \sta and \pur commands fall in this category.
-
-\kk{Do we want to get in more detail about requirements? Commands must
-  be deterministic, they must not do any other side effect (sending
-  signals, etc). These assumptions must be checked when the developer
-  designates the categories. Commands are assumed to always terminate
-  gracefully. \sys changes the termination behaviour of commands as a
-  sequential command that would exit prematurely might not exit
-  completely if parallelized.}
-
-\kk{Maybe give an example of a script and its dataflow graph.}
+\heading{\todo{Streaming} Commands}
+
+\todo{Streaming} commands consume their inputs sequentially and one
+element at a time, with the possible exception of static input files,
+and produce one output file. The quintessential example of
+\todo{streaming} commands is \ttt{cat}, that consumes its inputs in
+order, producing their concatenation as output. A more interesting
+example is \ttt{comm -23 f1 f2}, which has \ttt{f2} as its static
+input (it needs to read it completely before consuming anything from
+\ttt{f1} and then consumes \ttt{f1} one element at a time. These
+commands can be represented as functions of type $f : D\kstar \times
+[D\kstar] \to D\kstar$, where the first argument represents the
+concatenation of the inputs that are consumed one at a time and the
+second argument represents the static inputs.
+%% \km{Again, a graphical
+%%   presentation of these operators would be useful. Maybe with a way to
+%%   distinguish between between streaming inputs and static inputs? Btw,
+%%   this distinction is not completely clear to me at this
+%%   point. Clearly, not all programs that we can write are streaming
+%%   (monotone).} \kk{Do you mean that there are shell scripts that are
+%%   non monotone? That is true but we are assuming monotone nodes. Do I
+%%   miss something?}
+We focus on these commands because a large subset of \sta and \pur
+commands fall in this category.
+
+%% \kk{Do we want to get in more detail about requirements? Commands must
+%%   be deterministic, they must not do any other side effect (sending
+%%   signals, etc). These assumptions must be checked when the developer
+%%   designates the categories. Commands are assumed to always terminate
+%%   gracefully. \sys changes the termination behaviour of commands as a
+%%   sequential command that would exit prematurely might not exit
+%%   completely if parallelized.}
+
+%% \kk{If we have time and space we could give an example of a script
+%%   together with its dataflow graph to give more intuition about nodes
+%%   and streaming commands.}
 
@@
 
-%% We focus on \todo{one-input-output} commands (as defined in
+%% We focus on \todo{streaming} commands (as defined in
 %% \cref{graph-components}) that constitute a large subset of shell
@@
 characters) are considered stateless. Formally, a
-\todo{one-input-output} command $f$ is stateless if it is ``linear''
-with respect to concatenation \km{It commutes with the operation of concatenation.
-This is the definition of (semigroup) homomorphism.
-I am not sure that linearity is an appropriate term.}, \ie satisfying the following equation:
+\todo{streaming} command $f$ is stateless if it commutes with
+the operation of concatenation, \ie it is a semigroup homomorphism:
+%% with respect to concatenation \km{It commutes with the operation of concatenation. This is the definition of (semigroup) homomorphism. I am not sure that linearity is an appropriate term.}
+%% , \ie satisfying the following equation:
 
@@
 This means that we get the same output by applying $f$ to a
-concatenation of two inputs $x, x'$, or by applying the fold
-function $f'$ to the concatenation of the outputs produced by applying
-map $m$ to each of these inputs.
-\km{We had also discussed here the case of operators that can be described as associative and commutative functions on the data (e.g., sum).
-In this case the parallelization has the flavor of data parallelism: one can split the input stream in an arbitrary way (e.g., randomly or using some other load-balancing scheme) and merge the outputs.
-For the more broad class of associative operations (e.g., sort), the formulation of this paragraph makes perfect sense.
-I just want to mention that this feels a lot more like \textbf{batching} than map-reduce-like data parallelization.}
+concatenation of two inputs $x, x'$, or by applying the fold function
+$f'$ to the concatenation of the outputs produced by applying map $m$
+to each of these inputs.
+%% \km{We had also discussed here the case of
+%%   operators that can be described as associative and commutative
+%%   functions on the data (e.g., sum). In this case the parallelization
+%%   has the flavor of data parallelism: one can split the input stream
+%%   in an arbitrary way (e.g., randomly or using some other
+%%   load-balancing scheme) and merge the outputs. For the more broad
+%%   class of associative operations (e.g., sort), the formulation of
+%%   this paragraph makes perfect sense. I just want to mention that this
+%%   feels a lot more like \textbf{batching} than map-reduce-like data
+%%   parallelization.} \kk{Thanks, I think you are right. Batching is a
+%%   better name for it. In the end we didn't handle
+%%   commutative-associative function (that are more like map-reduce) and
+%%   so we changes the name to map-fold to not imply commutativity.}
 
@@
 D^*$, and the only behaviour of a dataflow graph is its output, this
-optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the behavior of the graph.
-Furthermore, it can be straightforwardly extended to pure
-commands that can be implemented by a map-fold pair $(m, r)$ as $
-v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n, s), s)$. A visual representation of this transformation can be seen in \Cref{fig:parallelization-transformation}.
+optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s)
+\cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the
+behavior of the graph. Furthermore, it can be straightforwardly
+extended to pure commands that can be implemented by a map-fold pair
+$(m, r)$ as $ v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s)
+\cdot m(x_2, s) \cdots m(x_n, s), s)$. \todo{A visual representation
+  of this transformation can be seen in}
+\Cref{fig:parallelization-transformation}.
 
@@
 
+\heading{Auxiliary Transformations}
+
+In addition to the above, we also define a set of auxiliary
+transformations that are shown in
+\Cref{fig:auxiliary-transformations}.  The first two enable the
+parallelization transformations by inserting \ttt{cat} nodes, either
+by using them to concatenate many inputs of one command, or by first
+adding a \ttt{split} node that splits the input in batches. The final
+one inserts a relay that performs the identity transformation on its
+input. Relay nodes can be useful for monitoring and debugging, as well
+as for performance improvements, as we will see in \Cref{optimizer}.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{\detokenize{./figs/auxiliary-transformations.png}}
+\caption{
+  \textbf{Auxiliary transformations.}
+  \todo{Add all three of the auxiliary transformations.}
+}
+\vspace{-10pt}
+\label{fig:auxiliary-transformations}
+\end{figure}
+
 %% \kk{Remember to mention the assumptions that need to hold for the
@@
 %% \end{lemma}
-
-\kk{I think we should say something about the rest of the
-  transformations: (i) changing a command with multiple inputs to a
-  cat followed by the command to enable the previous optimization,
-  (ii) adding a split node after a command (without saying anything
-  about its implementation for now, and (iii) Adding an identity node
-  in the place of an edge.}
-
-%% \kk{I think that we should probably rewrite this paragraph with the
-%%   ``new'' split optimization (and maybe any others that we develop
-%%   until then) here.}
-
-%% In addition to node splits, we can also perform edge splits in order
-%% to increase the number of incoming edges for some nodes to further expose
-%% possible data parallelism. In cases where the edge refers to a real file, the
-%% edge split represents splitting the real file in chunks. In
-%% cases where the edge connects two nodes, the split effectively refers to several edges
-%% with a bounded number of lines. Note that the final output edge of a
-%% valid node in the graph must always be unbounded if it does not refer
-%% to a file.
 
@@
 
-In more detail, \ttt{grep} is a short script centered around an expensive \sta command;
+In more detail, \ttt{grep} is a short script centered around an expensive \tsta command;
 % a DFA-based regular expression matching phase;
@@
 
+%% Bibliography
+\bibliography{./bib}
+
 \appendix
@@
 
-%% Bibliography
-\bibliography{./bib}
-
 %% %% Appendix

### 2020-05-20 pldi/draft.tex
First attempt of abstract

--- 2020-05-20 pldi/draft.tex
@@
 \newcommand{\sx}[1]{(\S\ref{#1})}
-\newcommand{\sys}{{\scshape Pash}\xspace}
+\newcommand{\sys}{{\scshape PaSh}\xspace}
 \newcommand{\unix}{{\scshape Unix}\xspace}
@@
 \begin{abstract}
-  Distributed systems offer notable benefits over centralized ones.
-  Reaping these benefits, however, requires programming in a way that makes distribution explicit---\eg via new programming languages, distributed frameworks, or source-level annotations.
-  This paper presents \sys, a shell variant that automatically and correctly scales out \unix shell pipelines. 
-  Its insight is that pipelines already express stream computations that can be automatically distributed.
-  To distribute them, \sys
-    decomposes primitives into parallelizability classes,
-    identifies high-parallelizability stages,
-    applies rewriting rules for largest possible subprograms,
-    and orchestrates their execution.
-Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
+  This paper presents \sys, a shell aimed at parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
+  Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts it back to a script that uses POSIX constructs to explicitly guide parallelism.
+A set of lightweight annotations allow command developers to express key parallelizability properties of their commands.
+An accompanying pararllelizability study of commands in POSIX and GNU Coreutils, two large and commonly used groups, 
+informs \sys's annotation language and result in its equivalent of a data-parallel standard library.
+Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
+\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes—with order-of-maginute performance improvements stemming from the combination of its program transformations and runtime primitives.
 \end{abstract}
@@
   (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
-The results demonstrate significant advantages in correctness, manual effort, and performance
+The results demonstrate significant advantages in \todo{correctness}, \todo{manual effort}, and performance
  ---with \sys offering an order-of-maginute performance improvements due to its transformations and optimized runtime.

### 2020-05-20 pldi/draft.tex
Add parallel figure

@@
 %% purpose of having a simple solution.
-
+\begin{wrapfigure}{r}{0.25\textwidth}
+  \begin{center}
+    \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
+  \end{center}
+%   \vspace{-15pt}
+% \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
+\end{wrapfigure}
 Second, based on the comparison with \ttt{sort --parallel} it seems
@@
 implementation with almost no effort.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
-  \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.}
-  }
-  \vspace{-15pt}
-  \label{fig:sort-parallel-comparison}
-\end{figure}
 

### 2020-05-20 pldi/draft.tex
Stash-half pass on 5

@@
 
-Given a dataflow graph representing a shell program, the \sys
-optimizer outputs an optimized dataflow graph in which data
-parallelism opportunities have been exploited.  More concretely, the
-optimizer starts by splitting input files in several chunks, given the
-desirable output dataflow graph width. It then starts from the source
-nodes of the graph, iteratively performing the graph transformations
-that were described in \Cref{ir:transformations}.
-%% If the incoming edges of the node are less than the desirable graph
-%% width, it performs edge splits, and then it performs maximal node
-%% splits.
-The final graph can then be implemented by spawning a process for each
-node, and redirecting the inputs and outputs according to the graph
-edges.
+After translating the parallelizable regions of the input script to
+dataflow graphs and optimizing these graphs using the transformations
+described in \Cref{ir:transformations}, \sys implements the graph by
+translating it back to a shell script. In this section we describe how
+\sys achieves that, and several technical challenges that it addresses.
+
+%% Given a dataflow graph representing a shell program, the \sys
+%% optimizer outputs an optimized dataflow graph in which data
+%% parallelism opportunities have been exploited.  More concretely, the
+%% optimizer starts by splitting input files in several chunks, given the
+%% desirable output dataflow graph width. It then starts from the source
+%% nodes of the graph, iteratively performing the graph transformations
+%% that were described in \Cref{ir:transformations}.
+%% %% If the incoming edges of the node are less than the desirable graph
+%% %% width, it performs edge splits, and then it performs maximal node
+%% %% splits.
+%% The final graph can then be implemented by spawning a process for each
+%% node, and redirecting the inputs and outputs according to the graph
+%% edges. In this section, we describe several technical challenges that
+%% we addressed.
 
@@
 
-\noindent
-\TODO{Problem: Cat (and other commands that have multiple inputs) is
-  not eager. Illustration: Show a parallel pipeline having a cat that
-  gathers input somewhere (and a sort merge maybe). The way pipes
-  work, this graph will essentially evaluate almost completely
-  sequentially. Maybe go through the steps that cat asks for the first
-  line of its first input until ithe first input is done and then goes
-  to the second. This leads to processor under-utilization. Solution:
-  We add eager nodes that eagerly consume all their input (even if
-  their output does not ask for it). Eager nodes are themselves part
-  of the dataflow graph (and model) and essentially perform an
-  identity function. This makes all processors as utilized as possible
-  and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
-  light) tasks it has some overhead. We evaluate eager in section X
-  (We should also say something about a CPU light task). A good thing
-  is that they work for all possible commands with almost no overhead,
-  without having to make each specific command eager.}
+The shell is uniquely lazy, in the sense that most commands and pipes
+consume their inputs only when they are ready to process more. This
+allows the underlying system to be able to support concurrently
+executing many more commands than the processing threads of the
+underlying system, by scheduling them concurrently, allowing them to
+have progress one at a time. This design decision however, severely
+restricts parallelism since nodes are often blocked when their
+consumers are processing their input. Consider as an example the
+following script:
+
+\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+  mkfifo t1 t2
+  grep "foo" > t1 & grep "foo" > t2 &
+  cat t1 t2
+\end{lstlisting}
+
+The \ttt{cat} command will only try to consume input from \ttt{t2}
+when it is done reading from \ttt{t1}. This means that the second
+\ttt{grep} will block without processing any input until the first
+\ttt{grep} is done processing. Instead of addressing this issue for
+specific commands, \sys inserts and instants several eager \ttt{relay}
+nodes in critical points of the pipeline as seen in
+\Cref{fig:eager}. These nodes force their producers to produce output
+and not block, while also preserving the existing task-based
+parallelism---that would not be preserved say with the \todo{third}
+solution in \Cref{fig:eager}.
+
+%% \TODO{Problem: Cat (and other commands that have multiple inputs) is
+%%   not eager. Illustration: Show a parallel pipeline having a cat that
+%%   gathers input somewhere (and a sort merge maybe). The way pipes
+%%   work, this graph will essentially evaluate almost completely
+%%   sequentially. Maybe go through the steps that cat asks for the first
+%%   line of its first input until ithe first input is done and then goes
+%%   to the second. This leads to processor under-utilization. Solution:
+%%   We add eager nodes that eagerly consume all their input (even if
+%%   their output does not ask for it). Eager nodes are themselves part
+%%   of the dataflow graph (and model) and essentially perform an
+%%   identity function. This makes all processors as utilized as possible
+%%   and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
+%%   light) tasks it has some overhead. We evaluate eager in section X
+%%   (We should also say something about a CPU light task). A good thing
+%%   is that they work for all possible commands with almost no overhead,
+%%   without having to make each specific command eager.}
 
@@
   Eager is an elegant solution that solves the problem while remaining within
-  the model. Describe.
+  the model. Describe. \TODO{Combine this figure with the above code maybe.}
 }
@@
 
-\heading{Readjusting Graph Width}
+\heading{Split node implementation}
 

### 2020-05-20 pldi/draft.tex
Abstract typos

@@
 A set of lightweight annotations allow command developers to express key parallelizability properties of their commands.
-An accompanying pararllelizability study of commands in POSIX and GNU Coreutils, two large and commonly used groups, 
+An accompanying parallelizability study of commands in POSIX and GNU Coreutils, two large and commonly used groups, 
 informs \sys's annotation language and result in its equivalent of a data-parallel standard library.
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes—with order-of-maginute performance improvements stemming from the combination of its program transformations and runtime primitives.
+\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes—with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
 \end{abstract}

### 2020-05-20 pldi/draft.tex
First pass on section 5.2

@@
 
-\noindent
-\TODO{Problem: Pipeline width shrinks after pure commands. Solution:
-  Split node that resplits its input to multiple files in sequence (to
-  preserve the model semantics).}
+As described in \Cref{ir:transformations}, \sys transformations
+inserts \ttt{split} nodes to expose parallelism in cases where a
+parallelizable node only has one input. The \sys \todo{backend} offers
+two options for implementing this split node, to satisfy the inherent
+tension between task-based parallelism and knowledge of the number of
+lines in the \ttt{split} input. In the general case, the number of
+lines of a specific edge is not known a-priori (due to commands that
+remove or add new lines), and so \sys provides an automatic split that
+first counts the number of lines, and then splits its input to the
+desired number of outputs. If however, the user knows the number of
+lines at the split insertion edge, they can provide it to \sys, that
+then uses it to \ttt{split} without having to block and read its whole
+input first. \sys also inserts eager relay nodes after all but the
+last output of split in order to address the laziness issue that was
+described before.
+
+%% \noindent
+%% \TODO{Problem: Pipeline width shrinks after pure commands. Solution:
+%%   Split node that resplits its input to multiple files in sequence (to
+%%   preserve the model semantics).}
 

### 2020-05-20 pldi/draft.tex
Add tiled table

@@
     \centering
-    %% \subcaptionbox{\label{eval:minimal_grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:grep}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:minimal_sort}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:topn}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:wf}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-    %% }
-    %% \subcaptionbox{\label{eval:spell}}{
-    %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-    %% }
-    %% \caption{a}
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/spell_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:grep}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_sort}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{\detokenize{./figs/set-diff_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_sort}
-    \end{subfigure}%
-
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:topn}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:wf}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:spell}
-    \end{subfigure}%
-    ~
-    \begin{subfigure}[b]{0.24\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{\detokenize{./figs/bigrams_throughput_scaleup.pdf}}
-        %% \caption{TODO}
-        %% \label{eval:minimal_sort}
-    \end{subfigure}%
+    \includegraphics[width=\textwidth]{\detokenize{./figs/tiling_throughput_scaleup.pdf}}
     \caption{
@@
       (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
-      (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
-      (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
-      (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
+      (iii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
+      (iv) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
+      (v) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
+      (vi) \todo{Opt. Parallel}, showing the performance of an optimized version of \ttt{bigrams}.
       \nv{Maybe it makes sense to show log x-axis?}
@@
 \end{figure*}
+
+%% \begin{figure*}[t]
+%%     \centering
+%%     %% \subcaptionbox{\label{eval:minimal_grep}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \subcaptionbox{\label{eval:grep}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \subcaptionbox{\label{eval:minimal_sort}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \subcaptionbox{\label{eval:topn}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \subcaptionbox{\label{eval:wf}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \subcaptionbox{\label{eval:spell}}{
+%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+%%     %% }
+%%     %% \caption{a}
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:minimal_grep}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/spell_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:grep}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:minimal_sort}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%       \centering
+%%       \includegraphics[width=\textwidth]{\detokenize{./figs/set-diff_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:minimal_sort}
+%%     \end{subfigure}%
+
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:topn}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:wf}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%         \centering
+%%         \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:spell}
+%%     \end{subfigure}%
+%%     ~
+%%     \begin{subfigure}[b]{0.24\textwidth}
+%%       \centering
+%%       \includegraphics[width=\textwidth]{\detokenize{./figs/bigrams_throughput_scaleup.pdf}}
+%%         %% \caption{TODO}
+%%         %% \label{eval:minimal_sort}
+%%     \end{subfigure}%
+%%     \caption{
+%%       \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
+%%       Four different configurations per benchmark:
+%%       (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
+%%       (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
+%%       (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
+%%       (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
+%%       (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
+%%       \nv{Maybe it makes sense to show log x-axis?}
+%%       \kk{How can we also fit two more plots (sort-sort and diff) here?
+Maybe we can fit one by getting rid of sort (since it also shown in the sort--parallel one).
+Otherwise we could separate the ``split'' plots from the ones that don't have split.}
+%%     }
+%%     \vspace{-15pt}
+%%     \label{fig:microbenchmarks}
+%% \end{figure*}
 

### 2020-05-20 compiler/gather_results.py

@@ -220,29 +220,28 @@ def check_output_diff_correctness(prefix, scaleup_numbers):
                    if not check_output_diff_correctness_for_experiment('{}{}_distr_no_task_par_eager.time'.format(prefix, n))]
     return (wrong_diffs, wrong_diffs_no_eager, wrong_diffs_no_task_par_eager)
 
-def collect_scaleup_times(experiment, results_dir):
+
+def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, ax):
     print(experiment)
 
     # all_scaleup_numbers = list(set(get_experiment_files(experiment, results_dir)))
     # all_scaleup_numbers.sort()
     # all_scaleup_numbers = [i for i in all_scaleup_numbers if i > 1]
-    all_scaleup_numbers = [2, 4, 8, 16, 32, 64]
     prefix = '{}/{}_'.format(results_dir, experiment)
     distr_speedup, compile_distr_speedup = collect_experiment_speedups(prefix, all_scaleup_numbers)
 
     output_diff = check_output_diff_correctness(prefix, all_scaleup_numbers)
 
-    fig, ax = plt.subplots()
-
-    ## Plot speedup
-    ax.set_ylabel('Speedup')
-    ax.set_xlabel('Level of Parallelism')
+    maximum_y = 0
+    lines = []
 
     ## In the bigrams experiment we want to also have the alt_bigrams plot
     if(experiment == "bigrams"):
         opt_prefix = '{}/{}_'.format(results_dir, "alt_bigrams")
         opt_distr_speedup, _ = collect_experiment_speedups(opt_prefix, all_scaleup_numbers)
-        ax.plot(all_scaleup_numbers, opt_distr_speedup, '-P', color='magenta', linewidth=0.5, label='Opt. Parallel')
+        line, = ax.plot(all_scaleup_numbers, opt_distr_speedup, '-P', color='magenta', linewidth=0.5, label='Opt. Parallel')
+        maximum_y = max(maximum_y, max(opt_distr_speedup))
+        lines.append(line)
 
     if(not (experiment in custom_scaleup_plots) or
        (experiment in custom_scaleup_plots and
@@ -251,18 +250,24 @@ def collect_scaleup_times(experiment, results_dir):
             auto_split_distr_speedup = collect_distr_experiment_speedup(prefix,
                                                                         'distr_auto_split.time',
                                                                         all_scaleup_numbers)
[…]

### 2020-05-22 pldi/draft.tex
Pass on Section 5

@@
 
-\section{Implementation}
+\section{From scripts to graphs and back again}
 \label{impl}
+
+\kk{Less fun section title: Implementation}
 
@@
 \sys. It starts by describing how the \sys front-end translates a
-shell script to the dataflow graph model defined in \S\ref{ir}, by
-applying transformation on parallelizable regions~\sx{front-end}.  It
+shell script to the dataflow graph model defined in \S\ref{ir} by
+identifying and transforming parallelizable regions~\sx{front-end}. It
 closes with the \sys runtime\todo{/backend} and the technical
@@
 program regions and a translation pass for converting them to dataflow
-dataflow graphs.
+graphs.
 
@@
 programs, which inherently contains information about components that
-can be executed independently and barriers that are natural points for
+can be executed independently, and barriers that are natural points for
 enforcing synchronization. Consider the following example:
 
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
   cat f1 f2 | grep "foo" > f3 && sort f3 
@@
 Conversely, all other operators, such as the \ttt{;} sequential
-composition operator, the \ttt{\&\&} logical operator represent
+composition operator and the \ttt{\&\&} logical operator, represent
 barrier constructs that do not allow parallelizable regions to
@@
 \heading{Translation Pass}
-
+%
 %% \kk{It might be a good idea to change this subsection based on the new
@@
 %%   redundant when having the distributable regions too.}
-
+%
 The \sys front-end performs a depth-first search pass on the AST of
@@
 %% two un-expand strings do not refer to the same file identifier,
-and as a result does not parallelize this sub-expression.
-
-\kk{Can we say anything more about the safety considerations? That (i)
-  the translation pass ``checks'' that no nodes write/read or
-  write/write to the same file (is that necessary?), that (ii) a
-  command should not read from a read-once file. (is that necessary?)}
+and as a result does not parallelize such sub-expressions.
+
+%% \kk{Can we say anything more about the safety considerations? That (i)
+%%   the translation pass ``checks'' that no nodes write/read or
+%%   write/write to the same file (is that necessary?), that (ii) a
+%%   command should not read from a read-once file. (is that necessary?)}
 
@@
 dataflow graphs and optimizing these graphs using the transformations
-described in \Cref{ir:transformations}, \sys implements the graph by
-translating it back to a shell script. In this section we describe how
-\sys achieves that, and several technical challenges that it addresses.
+described in \Cref{ir:transformations}, \sys implements each graph by
+translating it back to a shell script. Nodes of the graph are
+instantiated with their respective commands, and edges are
+instantiated as named pipes. In this section, we describe several
+technical challenges that \sys addresses to efficiently perform this
+transformation, as well as the implementations of the custom \sys
+nodes that are added by the auxiliary transformations.
+%% it instantiates the \todo{custom} split and relay nodes as highly
+%% optimized commands to improve In this section we describe how \sys
+%% achieves that, and several technical challenges that it addresses.
 
@@
 \heading{Overcoming Laziness}
-
+%
+%
 The shell is uniquely lazy, in the sense that most commands and pipes
 consume their inputs only when they are ready to process more. This
-allows the underlying system to be able to support concurrently
-executing many more commands than the processing threads of the
-underlying system, by scheduling them concurrently, allowing them to
-have progress one at a time. This design decision however, severely
-restricts parallelism since nodes are often blocked when their
-consumers are processing their input. Consider as an example the
-following script:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+allows the underlying system to concurrently execute more commands
+than the available processing threads,
+%% support concurrently executing many more commands than the
+%% processing threads of the underlying system,
+%% by scheduling them concurrently, allowing them to
+%% have progress one at a time.
+by interleaving their executions. This design decision however,
+severely restricts parallelism since nodes are often blocked when
+their consumers are not requesting any input. Consider as an example
+the following script:
+
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
   mkfifo t1 t2
@@
 
+\noindent
 The \ttt{cat} command will only try to consume input from \ttt{t2}
@@
 \ttt{grep} is done processing. Instead of addressing this issue for
-specific commands, \sys inserts and instants several eager \ttt{relay}
-nodes in critical points of the pipeline as seen in
-\Cref{fig:eager}. These nodes force their producers to produce output
-and not block, while also preserving the existing task-based
-parallelism---that would not be preserved say with the \todo{third}
-solution in \Cref{fig:eager}.
+each command separately, \sys inserts and instantiates eager
+\ttt{relay} nodes in critical points of the pipeline as seen in
+\Cref{fig:eager}. These nodes eagerly consume their input, forcing
+their upstream nodes to produce output when possible,
+%% force their producers to
+%% produce output and not block,
+while also preserving the existing task-based parallelism---that would
+not be preserved say with the \todo{third} solution in
+\Cref{fig:eager}.
 
@@
 \heading{Split node implementation}
-
-As described in \Cref{ir:transformations}, \sys transformations
-inserts \ttt{split} nodes to expose parallelism in cases where a
-parallelizable node only has one input. The \sys \todo{backend} offers
-two options for implementing this split node, to satisfy the inherent
-tension between task-based parallelism and knowledge of the number of
-lines in the \ttt{split} input. In the general case, the number of
-lines of a specific edge is not known a-priori (due to commands that
-remove or add new lines), and so \sys provides an automatic split that
-first counts the number of lines, and then splits its input to the
+%
+As described in \Cref{ir:transformations}, \sys's optimizer inserts
+\ttt{split} nodes to expose parallelism in cases where a
+parallelizable node only has one input. In order for the split to be
+effective in exposing parallelism, it needs to split its input equally
+in its outputs, but the size of its input can not be known
+a-priori. To satisfy this inherent tension, \sys supports two split
+implementations, a general one that can be used if the script contains
+commands that could add or remove lines, and an optimized one that can
+be used if the size of the split input is known before the script
+executes.
+%% The \sys \todo{backend} support two split node implementations, to
+%% satisfy the inherent tension between the split efficiency and
+%% task-based parallelism.
+In the general case, since the number of lines of a specific edge is
+not known, \sys provides an automatic split that first consumes its
+complete input, counts its lines, and then splits it equally to the
 desired number of outputs. If however, the user knows the number of
-lines at the split insertion edge, they can provide it to \sys, that
-then uses it to \ttt{split} without having to block and read its whole
-input first. \sys also inserts eager relay nodes after all but the
-last output of split in order to address the laziness issue that was
-described before.
+lines at the split insertion edge, they can configure \sys to split
+without having to read its whole input, therefore exploiting available
+task-based parallelism. Finally, \sys also inserts eager relay nodes
+after all but the last output of split in order to address the
+laziness issue that was described before.
+
+%% \TODO{Talk about eagerness in split and how we needed to add eager
+%%   nodes to n-1 outputs.}
 
@@
 
-\TODO{Talk about eagerness in split and how we needed to add eager
-  nodes to n-1 outputs.}
-
 \heading{Dangling FIFOs and Zombie Producers}
-
+%
 %% \TODO{(This problem is a nice followup for the one above) Problem:
@@
 %%   evaluate the overhead of emptying all outputs in section X.}
-
+%
 % \nv{I think all of these titles should be the problem/challenge, not the solution}
 % A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a producer exits after it has produced and
-sent all of its output to a pipe\todo{/its output channel}.
+Under normal operation, a command exits after it has produced and
+sent all of its output to
+%% a pipe
+its output channel.
 % \ttt{SIGPIPE} 
-However, if its consumer exits early, the producer needs to be
-notified so that it stops \todo{appending to the stream/trying to push
-  data to its output channel}.
+However, if its output is a pipe and the pipe reader exits early, the
+command needs to be notified so that it stops writing to it.
+%% \todo{appending to the stream/trying to push data to its output
+%%   channel}.
 % ---so that it avoids unnecessary work and buffering challenges in the kernel.
-In \unix, this is achieved by an out-of-band error mechanism:
-  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
-This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
-  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
-}
-primarily because pipes and pipelines are at the heart of its philosophy.
+In \unix, this is achieved by an out-of-band error mechanism: the
+operating system will deliver a \ttt{PIPE} signal to the producer,
+notifying it that the pipe's consumer has exited.  This is different
+from errors for other system calls
+%% return an error
+and unusual compared to non-\unix systems\footnote{For example,
+  Windows indicate errors for \ttt{WriteFile} using its return
+  code---similar to \ttt{DeleteFile} and other Win32 functions.}
+primarily because pipes and pipelines are at the heart of its
+philosophy.
 % write errors are communicated differently 
@@
 \ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
-Consider the following emitted code from \sys's earlier versions:\kk{Proposed alternative: ``Consider the following shell script:''}
+Consider the following emitted code from \sys's earlier versions:
+%% \kk{Proposed alternative: ``Consider the following shell script:''}
 \begin{lstlisting}[language=sh, numbers=none]
@@
 \noindent
-In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
-As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
-This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
-Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
+In the code above, \ttt{head} exits early causing the last \ttt{cat}
+to exit before opening \ttt{fifo2}.  As a result, the second \ttt{cat}
+never receives a \ttt{PIPE} signal that its consumer exited---after
+all, \ttt{fifo2} never even had a consumer.  This, in turn, leaves the
+second \ttt{cat} unable to make progress, both (i) blocked unable to
+write output, and (ii) unaware that its consumer has exited.  Coupled
+with \ttt{wait} at the end, the entire snippet reaches a deadlock.
 
@@
 a routine that delivers \ttt{PIPE} signals to any remaining processes
-upstream.  This epilogue is repeated for every \todo{merge/join} point
-in the transformed pipeline \kk{I don't understand the final sentence}.
+upstream.
+%% This epilogue is repeated for every \todo{merge/join} point
+%% in the transformed pipeline \kk{I don't understand the final sentence}.
 % Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"
 
-\kk{We have to remember to say that for this to work, all nodes in the
-  graph must always open their outputs! If not, then this technique
-  doesn't work, and we have to monitor if any node dies to do
-  something with it. At the moment we ensure that by doing that in our
-  nodes, and by assuming that all nodes that we handle do that.}
+%% \kk{We have to remember to say that for this to work, all nodes in the
+%%   graph must always open their outputs! If not, then this technique
+%%   doesn't work, and we have to monitor if any node dies to do
+%%   something with it. At the moment we ensure that by doing that in our
+%%   nodes, and by assuming that all nodes that we handle do that. This
+%%   is not really a problem, because all of our commands have one
+%%   output, which if they dont open this would cause problems in the
+%%   sequential script too.}
 
@@
 
-\kk{Maybe say something about some of the pure fold functions that we
-  implemented. Interesting examples include sort and wc.}
+As mentioned in \Cref{parallelizability}, commands in \tpur can be
+parallelized using a map stage and a \todo{fold} stage. We have
+implemented these stages for several commands in \tpur to enable \sys
+to parallelize them. Some interesting examples are \ttt{sort}, which
+can be parallelized simply by using \ttt{sort} as a map and \ttt{sort
+  -m} as \todo{fold}, and \todo{wc} \TODO{Niko, can you complete
+  that with wc or any other interesting example?}
 

### 2020-05-22 compiler/gather_results.py
Update remaining empty cells in the table and fix a discrepancy in the one-liner tiling plot

@@ -602,12 +602,12 @@ def plot_one_liners_tiling(results_dir):
                    "double_sort",
                    "shortest_scripts"]
 
-    confs = ["Opt. Parallel",
-             "Par. + Split",
+    confs = ["Par. + Split",
              "Par. + B. Split",
              "Parallel",
              "Blocking Eager",
-             "No Eager"]
+             "No Eager",
+             "Opt. Parallel"]
 
     fig = plt.figure()
     gs = fig.add_gridspec(2, 5, hspace=0)
@@ -618,10 +618,10 @@ def plot_one_liners_tiling(results_dir):
     for i, experiment in enumerate(experiments):
         ax = fig.add_subplot(gs[i])
         _, lines = collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, ax)
-        if(experiment == "bigrams"):
-            total_lines += lines
-        elif(experiment == "shortest_scripts"):
-            total_lines += lines[1:]
+        if(experiment == "double_sort"):
+            total_lines = lines + total_lines
+        elif(experiment == "bigrams"):
+            total_lines += [lines[0]]
         ax.set_xticks(all_scaleup_numbers[1:])
         ax.text(.5,.91,pretty_names[experiment],
         horizontalalignment='center',

### 2020-05-22 pldi/draft.tex
Minor fix in the annotation language grammar
Tighten section 2

@@
   (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
-The results demonstrate significant advantages in \todo{correctness}, \todo{manual effort}, and performance
- ---with \sys offering an order-of-maginute performance improvements due to its transformations and optimized runtime.
+The results demonstrate significant advantages in \todo{correctness},
+\todo{manual effort}, and performance---with \sys offering an order-of-magnitude performance improvements due to its transformations and optimized runtime.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
@@
 
-\sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/JfEgr}{https://git.io/JfEgr}.
+% nv: took this out because it's wasting 6 lines of text (because of section titles hitting the column divide---classic problem of section 2)
+% \sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/JfEgr}{https://git.io/JfEgr}.
 
@@
 
-This section first reviews shell scripting through a small example~\sx{bg:pipelines}, which it then uses to present an overview of \sys's design and implementation~\sx{bg:overview}.
+This section reviews shell scripting through an example~\sx{bg:pipelines}, which it then uses to present 
+parallelization challenges~\sx{bg:challenges} and how they are addressed by \sys~\sx{bg:overview}.
 
@@
 % In the ongoing debate on climate change,
-Suppose a \todo{environmental scientist/engineer}    % looking for a non-CS expert---journalist might be too low
-  wants to get a quick sense of the changes in maximum temperature over the past century.
+Suppose a environmental scientist    % looking for a non-CS expert---journalist might be too low
+  wants to get a quick sense of the trends in the maximum temperature over the past century.
 As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
@@
 
-Finally, \unix provides an environment in which commands written in any language can be composed.
-Many of the commands used in practice come built into the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
+Finally, \unix provides an environment for composing commands written in any language.
+Many of the commands used in practice come with the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
 The fact that commands are developed in a variety of languages---including shell scripts---provides users with a significant flexibility.
@@
 
-\heading{Parallelization Challenges}
+\subsection{Parallelization Challenges}
+\label{bg:challenges}
+
 While these features aid developer economy through powerful program composition, they complicate shell script parallelization which even for simple scripts such as the one in Fig.~\ref{fig:example} face several challenges.
@@
 
-One obvious challenge is due to the language of the shell, which has been developed through a long history of additions.
-Different parts of a script, exploiting different language constructs, are amenable to different levels of parallelization---if at all.
-% Some constructs, such as the parallel composition operator \ttt{\&} ; many others such as the sequential composition operator (\ttt{;}) are clearly sequential; yet others, such as the loop constructs are more difficult to categorize.
-For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), cannot be interleaved with the rest of the script.
-
-Another challenge is due to the multitude of available commands;
-  each command that is part of a script may require very different treatment.
-For example, \ttt{grep} can proceed in parallel on each stream without any coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination across streams.
-Worse even, the same command may be used with different flags---\eg % \ttt{-d" " -f9} for the first
-  Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags\kk{I propose dropping the last sentence for this ``---making correct treatment even more difficult''}.
-When a single flag can affect the functionality of a command in non-trivial ways, deciding how to parallelize becomes significantly more difficult.
-
-Yet another challenge is that different phases operate on different output sizes thus allowing different parallelization widths.
-Even if Fig.~\ref{fig:example}'s script was aiming on, say, only six years of data, the first \ttt{curl} would still output hundreds of lines for each year.
-Naive, coarse-grained parallelization can miss such opportunities.
-
-% \nv{Are these all the challenges we want to highlight?}
-
-\kk{There are two ways to go here: either mention the challenges that a user would face to manually parallelize a script (that have already been mentioned in the introducion), or mention the challenges that automatic parallelization---this paper---has to address.
-I think that we should focus on the second.}
-
-The first and greatest challenge is the gargantuan number of available
-commands that can be used to construct a shell script. In contrast to
-restricted programming frameworks that are amenable to parallelization
-by supporting a few carefully designed operators (see MapReduce,
-Spark, etc) shell scripts can contain commands that can have arbitrary
-behaviours.  In order to parallelize such a script, each of command
-requires special treatment as to not introduce non-determinism or
-erroneous behaviours. For example, \ttt{grep} can be performed on
-different partitions of its input in parallel without any
-coordination, \ttt{head} requires light coordination, and \ttt{sort}
-requires significant coordination.  To make matters worse, the same
-command may be used with different flags---\eg % \ttt{-d" " -f9} for
-the first Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two
-different sets of flags---that could potentially alter its behaviour,
-making correct treatment even more difficult. A good solution should
-handle several commands but should also be extensible, in the sense
-that newly added or modified commands don't require modifications in
-the solution itself. \kk{The solution to this challenge is already written in the third paragraph of the design overview.}
-
-Another issue is that all commands are black boxes. Therefore it is
-impossible to perform any kind of analysis on the code of the command
-to identify possible parallelization. Therefore a good solution should
-parallelize in a coarse-enough granularity that doesn't require
-opening up commands.
-
-Finally, shell scripts contain several constructs that enforce sequential execution.
+\heading{Commands} % : Unbounded \& Mysterious}
+The first challenge is the nature and variety of shell commands.
+In contrast to restricted programming frameworks, amenable to parallelization by supporting a few carefully designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, shell scripts contain commands that can have arbitrary behaviors.
+For Fig.~\ref{fig:example}, each of command requires special treatment as to not introduce non-determinism or incorrectness:
+  \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
+These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
+Worse even, the same command may be used with different flags---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags---that could potentially alter its behaviour.
+% , making correct treatment even more difficult.
+% nv: we already said impossible so more difficult doesn't add much!
+
+% A good solution should handle several commands but should also be extensible, in the sense that newly added or modified commands don't require modifications in the solution itself.
+% TODO: Leave this for the solution
+
+\heading{Scripts} % 
+Another challenge is due to the language of the shell, which % has been developed through a long history of additions.
+  contains constructs with different parallelizability semantics. % that enforce sequential execution.
 For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), has to be completed before the rest of the script starts executing.
-\kk{The solution is
-  written in the fourth paragraph of the design solution.}
+Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
+For example, even if Fig.~\ref{fig:example}'s script was aiming on only six years of data, the first \ttt{curl} would still output hundreds of lines for each year;
+  naive, coarse-grained parallelization will miss such opportunities.
+Any attempt for automating parallelization needs to be aware of the shell semantics and induce as much parallelism as it can get but without breaking semantics.
+
+\heading{Implementation} % 
+A final challenge has to do with practical issues.
+Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
+Moreover, the \unix environment has its set of quirks % about opening streams, deadling with FIFOs,
+  and any attempt at parallelization will have to take care of these challenges---related to task parallelism, deadlock prevention, and runtime performance.
 
@@
   high-parallelizability stages, rewrite them, and orchestrate their execution.
+  \todo{reference in txt/decide}.
 }
@@
 The new script may incorporate parallelism, and is handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
-To \todo{discover and enable/expose and exploit} data-parallelism in a given shell script,
+To expose and exploit data-parallelism in a given shell script,
 %% parallel execution,
@@
 
-\sys first needs to understand the POSIX constructs used to compose larger programs, indicating indicate different levels of candidacy for pararllelization.
-To achieve this, \sys breaks POSIX constructs into a few broad classes:
- (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
- (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
- (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
-
-Second, \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs---a two-fold challenge.
+\heading{Commands: Parallelizability Classes}
+% \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs---a two-fold challenge.
 To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but clear set of well-understood parallelizability classes;
   these commands are already useful-enough to allow the composition of useful shell programs operating in practice.
-To understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power, \sys leverages this study to define an annotation language for describing a command's key properties with respect to its parallelizability.
-Such light annotations \kk{have to be written once per command (and not once per script) and} are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they are developing.
+To understand the characteristics of commands that fall outside this standard set and allow them to leverage its transformations,
+  \sys defines an annotation language for describing a command's % key properties with respect to its 
+  parallelizability.
+Such light annotations are expressed once per command (not once per script) and are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they develop.
 % We haven't introduced the notion of flags
@@
 
+\Cref{fig:dsl} shows the complete grammar of the parallelizability
+annotation language.
+
 \begin{figure}
@@
     <output> ::= `stdout' | <arg>
+
+    <outputs> ::= <output>
+    \alt <output> `,' <outputs>
 
@@
 
-    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
+    <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
 

### 2020-05-22 pldi/draft.tex
Update problem/solution

@@
 
-Provided information about the shell constructs and individual commands and their flags, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
+\heading{Scripts: Dataflow Graphs}
+% \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
+To assign a candidacy for parallelization, \sys breaks POSIX constructs into a few broad classes:
+ (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
+ (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+ (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
+ (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
+
+Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
 To achieve this, it converts sections of the script that are amenable to parallelization into a data-flow graph (DFG) representation.
@@
 
-Finally, \sys needs to provide correct, high-performance combiners for certain classes commands, and be proactive about (addressing) several practical challenges related to quirks of the \unix environment.
-In terms of combiners, \sys offers a set of linkable components for stream splitting and merging;
-  these commands live in the \ttt{PATH} and are addressable by name, which means they can be used like (and by) any other commands.
-Practical challenges---related to task parallelism, deadlock prevention, and runtime performance---are addressed through a series of constructs that \sys provides.
-
-Returning to the script of Fig.~\ref{fig:example}...\nv{I think we should show fragments of the resulting script}
-
+\heading{Implementation: Efficient Primitives}
+\sys addresses several practical challenges through a set of constructs it provides---\eg a set of linkable components for stream splitting and merging.
+For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
+All these commands live in the \ttt{PATH} and are addressable by name, which means they can be used like (and by) any other commands.
+
+Fig~\ref{fig:example2} shows a fragment of \sys's output for the script in Fig.~\ref{fig:example}.
 The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
 % They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
+\caption{
+  \textbf{\sys's $2\times$-parallel output for part of Fig.~\ref{fig:example}.}
+  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
+}
+\vspace{-15pt}
+\label{fig:example2}
+\end{figure}
 

### 2020-05-22 pldi/draft.tex
Pass over parallelizability

@@
 
-To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
-As mentioned before \sys focuses on parallelizing data-parallel
-commands, i.e. commands that can process their input in parallel, and
-encodes their characteristics by assigning them to
+% To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
+\sys aims at parallelizing data-parallel commands, \ie commands that can process their input in parallel, and encodes their characteristics by assigning them to
 \emph{parallelizability classes}.
 %% This information is encoded in a command's \emph{parallelizability class}.
-Intuitively, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
-It focuses on capturing characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
+% Intuitively, a parallelizability 
+A class represents the level of synchronization required by copies of a command executing in parallel.
+It captures characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 
-\kk{Mention that commands change classes depending on their arguments and that we only write here (and in the table) the class of commands with no flags.
-Then the discussion about flags in the language part makes sense.}
-
-% nv TODO: this is a hierarchy
-Broadly, shell commands can be split into four major classes (summarized in Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
+Broadly, shell commands can be split into four major classes (Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
 %% how they interact with state.
-These classes are ordered in ascending difficulty of parallelization:
-  later classes require more effort.
-In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).
+These classes are ordered in ascending difficulty (or impossibility) of parallelization.
+In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without gaining any performance improvements).
+Although commands can change classes depending on their arguments, we focus on no-flag defaults for now.
 
@@
 \heading{Stateless Commands}
-The first class, \sta, contains commands that operate on individual \todo{lines/elements}~\footnote{\todo{the choice of line as the data-quantum is deliberate, since they strike a nice balance between very coarse grained separation (files) and very fine-grained ones (individual characters).
-We might need to mention that in the introduction of 3.1}} of their input, without maintaining state across invocations.
+The first class, \sta, contains commands that operate on individual line elements of their input, without maintaining state across invocations.
 These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
@@
 
-\kk{Proposed insertion: The choice of where how to separate data elements would change the allocation of commands in parallelization classes---} Many of these commands (about 1/3 of \sta) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
+The choice of line as the data element strikes a convenient balance between coarse-grained (files) and fine-grained separation (characters) while staying in line with \unix's core abstractions.
+This choice can affect the allocation of commands in \sta, as many of its commands (about 1/3) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
 This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
-However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
+However, it turns out to be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
 
@@
 Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
-\caption{
-  \textbf{Hierarchy of parallelizability classes.}
-  Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
-  \nv{The figure is incorrect---it's actually a lattice.}
-}
-\vspace{-15pt}
-\label{fig:example}
-\end{figure}
+% 
+% \begin{figure}[t]
+% \centering
+% \includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
+% \caption{
+%   \textbf{Hierarchy of parallelizability classes.}
+%   Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
+%   \nv{The figure is incorrect---it's actually a lattice.}
+% }
+% \vspace{-15pt}
+% \label{fig:hierarchy}
+% \end{figure}
 
@@
 
-\kk{Maybe show an example of map-fold for sort or wc}
+% \kk{Maybe show an example of map-fold for sort or wc}
+% \nv{We agreed to put it end of §5}
 
@@
 The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
-\kk{Or commands that do not produce output consuming lines of their input---thus not being amenable to data-parallelism.}
 Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.
 
-This is the largest class because it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
+This is the largest class, for two reasons.
+First, it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
 In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
-
-\nv{
-  A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
-  For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
-  Why aren't these parallelizable?
-}
+Second, it contains commands that do not consume input or do not produce output---and thus are not amenable to data-parallelism.
+For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that
+  %A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
 
@@
 
-\nv{Note command breadth/impossibility of our task.}
+% TODO: \nv{Note command breadth/impossibility of our task.}
 % https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
@@
   (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
-As the first concern was discussed extensively in the previous section, we now focus on the latter two.\footnote{
-  \nv{Note to self: until now we have been talking only about parallelizability, and now we introduce two concerns that haven't been mentioned before.}
-}
+As the first concern was discussed extensively in the previous section, we now focus on the latter two.
 
@@
 \heading{Example Annotations}
-A record contains a set of clauses, each one identified by a predicate on the commands options.
+An annotation record contains a set of clauses, each one identified by a predicate on the command options.
 Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
@@
 % }
-The grammar of the language for specifying annotations is shown in Appendix~\ref{alg}.
-
-For example, consider \ttt{comm}---a command that performs a join-like operation on its two inputs to identify common elements.
+The annotation grammar is shown in Appendix~\ref{alg}.
+
+For example, consider \ttt{comm}, which performs a join-like operation on its two inputs to identify common elements.
 When \ttt{comm} is invoked without any flags, it produces a three-column output:
@@
 comm {
-  | -1 /\ -3 =>
-    (stateless, [args[1]], [stdout])
-  | -2 /\ -3 =>
-    (stateless, [args[0]], [stdout])
-  | _  =>
-    (pure, [args[0], args[1]], [stdout])
+| -1 /\ -3 => (S, [arg[1]], [stdout])
+| -2 /\ -3 => (S, [arg[0]], [stdout])
+| _  => (P, [arg[0], arg[1]], [stdout])
 }
@@
 The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
-There, \ttt{comm} is classified as \pur, since it needs to wait until it reads both files completely before it can output its results.
+There, \ttt{comm} is classified as \pur as it needs to % wait until it 
+  read both inputs completely before it can output its results.
 Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.
@@
 For most commands, the annotations are enough to allow parallelization: 
-  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{cmd}.
-Unfortunately, there are commands with less trivial tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
-
-To support the parallelization of arbitrary commands in \pur, \sys allows command developers to provide custom $map$ and $fold$ functions.
-In line with \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
-  (i) $map$ in \sta and $fold$ is in \pur,
-  (ii) $map$ can consume (or extend) the output of the original command and  $fold$ can consume (and combine) the results of multiple $map$ invocations, and
+  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
+Unfortunately, there are commands that require less obvious tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
+
+To support the parallelization of arbitrary commands in \pur, \sys allows supplying custom \emph{map} and \emph{fold} functions.
+In line with the \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
+  (i) \emph{map} is in \sta and \emph{fold} is in \pur,
+  (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{fold} can consume (and combine) the results of multiple \emph{map} invocations, and
   (iii) their composition produces the same output with the original command.
-If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their $map$ and $fold$ equivalents, enabling their parallelization.
+If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{fold} equivalents, enabling their parallelization.
 

### 2020-05-22 pldi/draft.tex
Pass over DFG

@@
 
-The core of \sys is a dataflow graph model that is expressive enough
-to support a large subset of shell programs. A fundamental difference
-with other dataflow graph models is that in \sys's model contains
-information about the order in which a node reads its inputs. This
-enables a set of graph transformations that expose data parallelism as
-task parallelism. These transformations can then be applied in an
-iterative way, optimizing the program to utilize available
-computational resources.
-
-%% The core of \sys is a dataflow graph model, inspired by the ones used
+The core of \sys is a dataflow graph (DFG) model that is expressive enough to support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
+A fundamental difference with other DFG models is that it contains information about the order in which a node reads its inputs.
+This enables a set of graph transformations that expose data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
+These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources.
+
+%% The core of \sys is a DFG model, inspired by the ones used
 %% in popular distributed stream processing systems~\sx{related} \kk{Is
@@
 
-The two main shell abstractions are (i) files (or pipes, \emph{viz.}
-correspondence between files and pipes in \S\ref{bg:pipelines}) that
-contain data, and (ii) commands that communicate through these
-files. Based on that, in \sys's graph model nodes represent commands
-and edges represent files.
-
-We first introduce basic notation that will be useful later on.  For a
-set $D$, we write $D\kstar$ to denote the set of all finite words over
-$D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to
-denote their concatenation. We write $\eps$ for the empty word. We say
-that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there
-is a word $z$ such that $y = xz$. The $\leq$ order is reflexive,
-antisymmetric, and transitive (i.e. is a partial order) and is often
-called the \emph{prefix order}.
-
-\heading{Edges -- Files}
-
-Edges in the dataflow graph represent files, the basic data
-abstraction of the shell. They are used as communication channels
-between nodes in the graph, and as the input or output of the whole
-graph. Edges are represented as possibly unbounded streams of type
-$D\kstar$. Edges can either refer to named files or FIFO pipes used
-for interprocess communication. Edges that do not start from a node in
-the graph represent the graph input; edges that do not point to a node
-in the graph are its outputs.
+The two main shell abstractions are
+  (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}\nv{Not in this version. Should we call them "streams"?}) that contain data, and
+  (ii) commands that communicate through these files.
+\nv{quick solution: change all files to streams and make the equivalence explicit here}
+\sys's DFG model represents commands as nodes and \todo{streams} as edges.
+
+We first introduce basic notation used throughput this section.
+For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
+For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
+We write $\eps$ for the empty word.
+We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
+The $\leq$ order is reflexive, antisymmetric, and transitive (\ie is a partial order) and is often called the \emph{prefix order}.
+
+\heading{Edges---Streams}
+Edges in the DFG represent streams, the basic data abstraction of the shell.
+They are used as communication channels between nodes in the graph, and as the input or output of the entire graph.
+Edges are represented as possibly unbounded streams of type $D\kstar$.
+Edges can either refer to named files or FIFO pipes used for interprocess communication.
+Edges that do not start from a node in the graph represent the graph input;
+  edges that do not point to a node in the graph are its outputs.
+\nv{OK This distinction is explained in this par; I would put this earlier}
 
@@
 
-\heading{Nodes -- Commands}
-
+\heading{Nodes---Commands}
 A node $f$ of the graph represents a function from a (possibly empty)
@@
 %%   restriction, therefore the above type should be fine.}
-This representation captures \todo{the majority/all} of commands in
-the \tsta, \tpur, \tnpu. Note that these functions should only produce
+This representation captures all the commands in
+the \tsta, \tpur, and \tnpu. Note that these functions should only produce
 output in the form of files and not perform any other side effect,
 such as sending signals.  We require that the function $f$ is monotone
-w.r.t.\ a lifting of the prefix order for a sequence of inputs
+with respect to a lifting of the prefix order for a sequence of inputs
 %% \kk{Formalize
@@
 \heading{\todo{Streaming} Commands}
-
-\todo{Streaming} commands consume their inputs sequentially and one
+These commands consume their inputs sequentially and one
 element at a time, with the possible exception of static input files,
@@
 commands fall in this category.
+\nv{Based on the intro of 4 or 4.1 I did not expect streaming.
+I understand edges and nodes, but how do streaming commands fit into the picture?
+the answer might be the last sentence which should be first---reviewer: why am I reading what I'm reading?:P}
 
@@
 %% \kk{If we have time and space we could give an example of a script
-%%   together with its dataflow graph to give more intuition about nodes
+%%   together with its DFG to give more intuition about nodes
 %%   and streaming commands.}
@@
 %%   streams of data, and therefore are not considered part of the
-%%   dataflow graph.  For now, we assume that these static files can only
+%%   DFG.  For now, we assume that these static files can only
 %%   be accessed by one node, thus ensuring that they do not interfere
@@
 %% %% Because of the assumption in section Front End \ref{}, that each file
-%% %% exists only once in the dataflow graph, we can safely assume that
+%% %% exists only once in the DFG, we can safely assume that
 %% %% reading and writing to these static files does not interfere with the
@@
 
-We have defined a set of semantics-preserving graph transformations
+\sys defines a set of semantics-preserving graph transformations
 that act as parallelization-exposing optimizations. Both the domain
-and range of these transformations is a graph in our dataflow model
+and range of these transformations is a graph in our DFG model
 and therefore they can be composed arbitrarily and in any order.
-%% Having a dataflow graph representation allows us to define
+%% Having a DFG representation allows us to define
 %% parallelization optimizations as graph transformations.
 Before describing the different types of optimizations, we formalize
-the intuition about stateless and pure commands that was described
+the intuition behind classes \sta and \pur described informally
 earlier~\sx{cmd}.
@@
 \heading{Stateless and Parallelizable Pure Commands}
-
 As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
@@
 parallelized using divide-and-conquer techniques. More formally, these
-pure commands $f$ can be implemented as a combination of functions map
+pure commands $f$ can be implemented as a combination of a function map
 $m$ and an associative fold $f'$, satisfying the following equation:
@@
 This means that we get the same output by applying $f$ to a
-concatenation of two inputs $x, x'$, or by applying the fold function
+concatenation of two inputs $x, x'$ as by applying the fold function
 $f'$ to the concatenation of the outputs produced by applying map $m$
-to each of these inputs.
+to each of $x$ and $x'$.
 %% \km{We had also discussed here the case of
@@
 \heading{Parallelization Transformations}
-%
 Based on these equations, we can define a \emph{node parallelization
   transformation} on a stateless node $v$ that is preceded by a
-\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
+\ttt{cat} with $n$ input streams and is followed by a node $v'$ (Fig.~\ref{fig:parallelization-transformation}). The
 transformation replaces $v$ with $n$ new nodes, routing each of the
@@
 $v'$. Since each incoming edge represents a stream of data $x_i :
-D^*$, and the only behaviour of a dataflow graph is its output, this
+D^*$, and the only behaviour of a DFG is its output, this
 optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s)
 \cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the
-behavior of the graph. Furthermore, it can be straightforwardly
-extended to pure commands that can be implemented by a map-fold pair
+behavior of the graph.
+
+This transformation can extended straightforwardly to pure commands implemented by a map-fold pair
 $(m, r)$ as $ v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s)
-\cdot m(x_2, s) \cdots m(x_n, s), s)$. \todo{A visual representation
-  of this transformation can be seen in}
-\Cref{fig:parallelization-transformation}.
+\cdot m(x_2, s) \cdots m(x_n, s), s)$.
+% \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.
 
@@
 \heading{Auxiliary Transformations}
-
-In addition to the above, we also define a set of auxiliary
-transformations that are shown in
-\Cref{fig:auxiliary-transformations}.  The first two enable the
-parallelization transformations by inserting \ttt{cat} nodes, either
+\todo{We} also define a set of auxiliary transformations that are shown in
+\Cref{fig:auxiliary-transformations}.  The first two \nv{Name them?} enable the
+parallelization transformations by inserting \ttt{cat} nodes,\footnote{\nv{This is minor, but for the model maybe name them concatenation nodes or ++? rather than \ttt{cat}}} either
 by using them to concatenate many inputs of one command, or by first
@@
 input. Relay nodes can be useful for monitoring and debugging, as well
-as for performance improvements, as we will see in \Cref{optimizer}.
+as for performance improvements~\sx{optimizer}.
 

### 2020-05-22 pldi/draft.tex
Pass over implementation

@@
 %% \begin{definition}
-%% Given a dataflow graph $G = (V, E, O)$, where $V$ is a set of nodes
+%% Given a DFG $G = (V, E, O)$, where $V$ is a set of nodes
 %% representing commands, $E$ is a set of edges representing files, and
@@
 %% edges. $v_{out}$ represents a concatenation of all the outputs of the
-%% dataflow graph. We represent the total order for a node $v$ as
+%% DFG. We represent the total order for a node $v$ as
 %% $<_v$. Given a node a node $v$
@@
 
-\section{From scripts to graphs and back again}
+\section{From Scripts to Graphs and Back Again}
 \label{impl}
 
-\kk{Less fun section title: Implementation}
-
-This section describes challenges related to the implementation of
-\sys. It starts by describing how the \sys front-end translates a
-shell script to the dataflow graph model defined in \S\ref{ir} by
-identifying and transforming parallelizable regions~\sx{front-end}. It
-closes with the \sys runtime\todo{/backend} and the technical
-challenges that were addressed to achieve performance
-benefits~\sx{optimizer}.
+This section describes \sys's implementation.
+It first details how \sys's front-end translates a shell script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
+It then describes \sys's runtime\todo{/backend} and the technical challenges addressed to achieve performance benefits~\sx{optimizer}.
 
@@
 
-As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
-translates it to a dataflow graph.  However, there are several
-components in shell programs that cannot be arbitrarily parallelized
-without affecting the program behaviour---\eg commands connected with
-a \ttt{\&\&} operator~\sx{bg}.
+To optimize a shell script, \sys translates it to a DFG~\sx{ir}.
+As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}~\S\ref{bg}---\sys introduce the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
 %% In addition, the shell's highly dynamic
@@
 %% unexpanded command argument refer to.
-To address this issue, we introduce the notion of parallelizable
-program regions and a translation pass for converting them to dataflow
-graphs.
 
@@
 %
-Parallelizable regions are program sub-expressions that can be
-parallelized safely without altering the total program behaviour.
+Parallelizable regions are program sub-expressions that can be parallelized safely, \ie without breaking the program.
 %% respect to their sequential execution.
-The search for these regions can be guided by the structure of shells
-programs, which inherently contains information about components that
-can be executed independently, and barriers that are natural points for
-enforcing synchronization. Consider the following example:
+The search for these regions is guided by the shell language and the structure of a particular program.
+These contain information about two types of components:
+  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
+Consider this script:
 
@@
 \noindent
-The commands \ttt{cat f1 f2} and \ttt{grep "foo"} would execute as
-independent processes in the standard shell, while \ttt{sort f3} would
-wait for their completion before executing.  Both \ttt{cat f1 f2 |
-  grep "foo" > f3} and \ttt{sort f3} are therefore parallelizable
-regions, regions that cannot be extended beyond \ttt{\&\&}---even
-better, they are maximal.
+The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---\ie more precisely, they are maximal.
 
@@
 parallelizable regions can be composed from smaller ones using the
-pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
-Conversely, all other operators, such as the \ttt{;} sequential
-composition operator and the \ttt{\&\&} logical operator, represent
+pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
+Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
 barrier constructs that do not allow parallelizable regions to
@@
 the given shell program.  During this pass, it extends the
-parallelizable regions from the bottom up, translating their
-independent components to dataflow graph nodes until a barrier
-construct is reached. All subtrees that are not translated to dataflow
-graphs are kept as they are. The output of the translation pass is the
+parallelizable regions bottom-up, translating their
+independent components to DFG nodes until a barrier
+construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
 original AST where parallelizable regions have been replaced with
-dataflow graphs and a call to \sys's runtime.
-
-To identify possible opportunities for parallelism, the translation
+DFGs and a call to \sys's runtime.
+
+To identify opportunities for parallelism, the translation
 pass extracts the parallelizability class of each command together
-with its inputs and outputs. It achieves that by searching all
-available PAL~\sx{ext} records for each command and resorts to
-conservative defaults if a record for the command is not found. If the
+with its inputs and outputs.  To achieve this for each command,
+it searches all available annotation~\sx{ext} and resorts to
+conservative defaults if no record is found. If the
 command is \sta or parallelizable \pur, the translation pass initiates
@@
 % A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a command exits after it has produced and
-sent all of its output to
-%% a pipe
-its output channel.
-% \ttt{SIGPIPE} 
-However, if its output is a pipe and the pipe reader exits early, the
-command needs to be notified so that it stops writing to it.
-%% \todo{appending to the stream/trying to push data to its output
-%%   channel}.
-% ---so that it avoids unnecessary work and buffering challenges in the kernel.
+Under normal operation, a command exits after it has produced and sent all of its output to its output channel.
+However, if its output is a pipe and the pipe reader exits early, the command is notified to stop writing early.
 In \unix, this is achieved by an out-of-band error mechanism: the
@@
 and unusual compared to non-\unix systems\footnote{For example,
-  Windows indicate errors for \ttt{WriteFile} using its return
+  Windows indicates errors for \ttt{WriteFile} using its return
   code---similar to \ttt{DeleteFile} and other Win32 functions.}
@@
 % write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g.
-\ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
-Consider the following emitted code from \sys's earlier versions:
+Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
+% \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
+% \nv{TODO! Will think about this later}
+Consider the following script:
+% emitted code from \sys's earlier versions:
 %% \kk{Proposed alternative: ``Consider the following shell script:''}
@@
 \noindent
-In the code above, \ttt{head} exits early causing the last \ttt{cat}
-to exit before opening \ttt{fifo2}.  As a result, the second \ttt{cat}
-never receives a \ttt{PIPE} signal that its consumer exited---after
-all, \ttt{fifo2} never even had a consumer.  This, in turn, leaves the
-second \ttt{cat} unable to make progress, both (i) blocked unable to
-write output, and (ii) unaware that its consumer has exited.  Coupled
-with \ttt{wait} at the end, the entire snippet reaches a deadlock.
+In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
+As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
+This, in turn, leaves it unable to make progress, as it is is both blocked and unaware of its consumer exiting.
+Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
 
@@
 \heading{Pure Fold Function Implementations}
-
-As mentioned in \Cref{parallelizability}, commands in \tpur can be
-parallelized using a map stage and a \todo{fold} stage. We have
-implemented these stages for several commands in \tpur to enable \sys
-to parallelize them. Some interesting examples are \ttt{sort}, which
-can be parallelized simply by using \ttt{sort} as a map and \ttt{sort
-  -m} as \todo{fold}, and \todo{wc} \TODO{Niko, can you complete
-  that with wc or any other interesting example?}
+Commands in \tpur can be parallelized using a \emph{map} and a \emph{fold} stage~\sx{parallelizability}.
+\sys implements \emph{fold} for several commands in \tpur to enable parallelization.
+A few interesting examples include \emph{fold} functions for
+  (i) \ttt{sort}, which amounts to the merge phase of a merge-sort (and on GNU systems is implemented as \ttt{sort -m}),
+  (ii) \ttt{uniq} and \ttt{uniq -c}, which need to check conditions at the boundary of their streams,
+  (iii) \ttt{tac}, which consumes stream descriptors in the reverse order, and 
+  (iv) \ttt{wc}, which adds inputs with an arbitrary number of elements (\eg \ttt{wc -lw} or \ttt{wc -lwc} \etc).
+The combiners iterate over the provided stream descriptors, \ie they work with more than two inputs, and apply pure functions at the boundaries of input streams (with the exception of \ttt{sort} that has to interleave inputs).
+% An important concern has been to maintain whitespace separation while
 

### 2020-05-22 pldi/draft.tex
Add dish cat
Add auxiliary figure

@@
 \centering
-\includegraphics[width=\columnwidth]{\detokenize{./figs/transformation-sketch.png}}
+\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_cat.pdf}}
+\vspace{-28pt}
 \caption{
   \textbf{Stateless parallelization transformation.}
-  The \texttt{cat} node is commuted with the stateless node
-  to utilize available data parallelism.
-  \TODO{Correct this diagram to show the cat node before
-    the node and then the cat node after the stateless
-    command after the optimization. Cat node should be
-    different than a circle. Maybe like an ``and'' node
-    in binary circuits?}
+  The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
+  \nv{Should graph be \emph{directed}?}
 }
-\vspace{-10pt}
+\vspace{-15pt}
 \label{fig:parallelization-transformation}
@@
 \centering
-\includegraphics[width=\columnwidth]{\detokenize{./figs/auxiliary-transformations.png}}
+\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_aux.pdf}}
 \caption{
   \textbf{Auxiliary transformations.}
-  \todo{Add all three of the auxiliary transformations.}
+  These augment the DFG with \ttt{cat}, \ttt{split}, and \ttt{relay} nodes.
 }

### 2020-05-22 pldi/draft.tex
Pass over 5

@@
 
-To identify opportunities for parallelism, the translation
-pass extracts the parallelizability class of each command together
-with its inputs and outputs.  To achieve this for each command,
-it searches all available annotation~\sx{ext} and resorts to
-conservative defaults if no record is found. If the
-command is \sta or parallelizable \pur, the translation pass initiates
-a parallelizable region that is propagated up the tree.
+To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
+To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
+If the command is in \sta or \pur, the translation pass initiates a parallelizable region that is propagated up the tree.
 
@@
 
-An important consideration is that due to the highly dynamic nature of
-the shell, there is information (such as the values of environment
-variables) that are not known to \sys at the time of translation.
+Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
+Examples of such information includes the values of environment variables, strings that have not been expanded, and sub-shell constructs.
 %% to the the safety properties that are mentioned above cannot be
 %% soundly checked statically.
-For safety purposes, \sys takes a conservative approach where it does
-not parallelize nodes without complete information.  For example, the
-translation pass cannot infer that an environment variable that is
-given as an argument to a command is not a flag that changes its
-parallelizability class,
-%% two un-expand strings do not refer to the same file identifier,
-and as a result does not parallelize such sub-expressions.
+For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
+It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.
 
@@
 
-After translating the parallelizable regions of the input script to
-dataflow graphs and optimizing these graphs using the transformations
-described in \Cref{ir:transformations}, \sys implements each graph by
-translating it back to a shell script. Nodes of the graph are
-instantiated with their respective commands, and edges are
-instantiated as named pipes. In this section, we describe several
-technical challenges that \sys addresses to efficiently perform this
-transformation, as well as the implementations of the custom \sys
-nodes that are added by the auxiliary transformations.
+After translating the parallelizable regions of the input script to DFGs and applying optimization passes using the transformations described earlier~\sx{ir:transformations},
+  \sys translated all DFGs back to a shell script.
+Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
+This section describes several technical challenges that \sys addresses to efficiently perform this transformation, as well as the implementations of the custom \sys nodes that are added by the auxiliary transformations.
 %% it instantiates the \todo{custom} split and relay nodes as highly
@@
 \heading{Overcoming Laziness}
-%
-%
-The shell is uniquely lazy, in the sense that most commands and pipes
-consume their inputs only when they are ready to process more. This
-allows the underlying system to concurrently execute more commands
-than the available processing threads,
+The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
+This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
 %% support concurrently executing many more commands than the
@@
 %% have progress one at a time.
-by interleaving their executions. This design decision however,
-severely restricts parallelism since nodes are often blocked when
-their consumers are not requesting any input. Consider as an example
-the following script:
+However, such laziness severely inhibits opportunities for parallelism as commands are often blocked when their consumers are not requesting any input.
+Consider the following script:
 
@@
 \noindent
-The \ttt{cat} command will only try to consume input from \ttt{t2}
-when it is done reading from \ttt{t1}. This means that the second
-\ttt{grep} will block without processing any input until the first
-\ttt{grep} is done processing. Instead of addressing this issue for
-each command separately, \sys inserts and instantiates eager
-\ttt{relay} nodes in critical points of the pipeline as seen in
-\Cref{fig:eager}. These nodes eagerly consume their input, forcing
-their upstream nodes to produce output when possible,
-%% force their producers to
-%% produce output and not block,
-while also preserving the existing task-based parallelism---that would
-not be preserved say with the \todo{third} solution in
-\Cref{fig:eager}.
+The \ttt{cat} command will only try to consume input from \ttt{t2} when it completes reading from \ttt{t1}.
+As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes. % processing.
+
+% replacing FIFOs with files is
+The closest solution \unix provides would be to replace FIFOs with files, as files are a central abstraction and can simulate pipes of arbitrary buffering.
+Aside from severe performance implications, naive replacement can lead to subtle race conditions as the consumer might reach EOF before the producer.
+Alternatively, a consumer could the producer to complete, but that would insert an artificial barrier inhibiting task-based parallelism and wasting disk resources.
+% \unix does not provide an easy way out.
+
+To address this challenge,
+% Instead of addressing this issue for each command separately, % nv: how could this be possible?
+  \sys inserts and instantiates \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}). % nv: not sure what critical meant
+These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
 
@@
   \textbf{Eager primitive.}
-  Describe three problems:
-    (i) FIFOs alone cannot solve it as they’re blocking;
-    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
-    (iii) writing to files + wait does not allow trask-based parallelism.
-  Eager is an elegant solution that solves the problem while remaining within
-  the model. Describe. \TODO{Combine this figure with the above code maybe.}
+  Addressing intermediary laziness is challenging:
+    (i) FIFOs are blocking;
+    (ii) files alone introduce race conditions between producer/consumer;
+    (iii) files \ttt{wait} inhibit task-based parallelism.
+Eager \ttt{relay} nodes address the challenge while remaining within the \sys model.
+% Describe. \TODO{Combine this figure with the above code maybe.}
 }
@@
 
-\heading{Split node implementation}
-%
-As described in \Cref{ir:transformations}, \sys's optimizer inserts
-\ttt{split} nodes to expose parallelism in cases where a
-parallelizable node only has one input. In order for the split to be
-effective in exposing parallelism, it needs to split its input equally
-in its outputs, but the size of its input can not be known
-a-priori. To satisfy this inherent tension, \sys supports two split
-implementations, a general one that can be used if the script contains
-commands that could add or remove lines, and an optimized one that can
-be used if the size of the split input is known before the script
-executes.
+\heading{Splitting Challenges}
+\sys's optimizer inserts \ttt{split} nodes to expose parallelism when parallelizable nodes only have one input~\sx{ir:transformations}.
+For \ttt{split} to be effective, it needs to disperse its input uniformly across its outputs.
+This requires the input size to be known beforehand, which is not always the case.
+
+To achieve this, % satisfy the tension between efficiency and \todo{knowledge},  % blanking on the right word here
+  \sys supports two \ttt{split} implementations:
+  (i) a general one that can be used with commands that could add or remove lines, and
+  (ii) an optimized one that can be used if the input size before its execution.
 %% The \sys \todo{backend} support two split node implementations, to
@@
 %% task-based parallelism.
-In the general case, since the number of lines of a specific edge is
-not known, \sys provides an automatic split that first consumes its
-complete input, counts its lines, and then splits it equally to the
-desired number of outputs. If however, the user knows the number of
-lines at the split insertion edge, they can configure \sys to split
-without having to read its whole input, therefore exploiting available
-task-based parallelism. Finally, \sys also inserts eager relay nodes
-after all but the last output of split in order to address the
-laziness issue that was described before.
+For the former, \ttt{split} first consumes its complete input, counts its lines, and then splits it equally to the desired number of outputs.
+For the latter, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
+
+\sys also inserts eager \ttt{relay} nodes after all \ttt{split} outputs (except the last one) to address laziness concerns.
 

### 2020-05-22 pldi/draft.tex
Quick resolution to make sure we can send a draft out

@@
 \newcommand{\todo}[1]{\hl{#1}\xspace}
-\newcommand{\fixme}[1]{[{\color{red}#1}]}
+\newcommand{\fixme}[1]{{\color{red}#1}}
 \newcommand{\nv}[1]{[{\color{cyan}nv: #1}]}
@@
 
-To address this challenge, we develop \sys, a shell that parallelizes POSIX shell scripts.
-\kk{We need to mention data-parallelism earlier.
+To address this challenge, we develop \sys, a shell that extracts data-parallelism from POSIX shell scripts.
+% \kk{We need to mention data-parallelism earlier.
 We don't focus on general parallelization but rather data-parallelism (performing operations on disjoint partitions of the input and then merging the results.
 WE need to give a brief idea about that early on so that we don't create false ideas.}
+% \nv{First attempt here. I am trying to get a copy out ASAP, we can refine this later}
 % \kk{Mention that it outputs shell scripts.}
@@
 % \sys offers both classes of users in the second group such much-needed automation.
-\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
+\sys benefits both programmer groups, % mentioned earlier, 
+  with a particular emphasis on users.
 Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:

### 2020-05-22 pldi/draft.tex
Half-way pass over eval

@@
 
+\begin{table*}[t]
+\center
+\footnotesize
+% \setlength\tabcolsep{3pt}
+\caption{
+  \footnotesize{
+    \textbf{Summary of shell one-liners}.
+    Small pipelines (\ie ones usually developed on-the-fly) drawn from various sources and applied to large datasets.
+    % Names correspond to the speedup plots (Fig.~\ref{fig:microbenchmarks}).
+    Structure summarizes the different classes of commands used in the script.
+    Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
+    Nodes and compile time report on \sys's resulting \todo{AST} size and compilation time for two indicative parallelization configurations.
+    % Highlights notes a few key features of each script.
+  }
+}
+\input{microbenchmarks-table}
+\label{tab:eval}
+\end{table*}
+
 \heading{Dangling FIFOs and Zombie Producers}
@@
 
-\TODO{Remember to mention that deathstar has 128 threads but 64
-  cores. So it is reasonable for the speedup to not go further in cpu
-  heavy pipelines.}
-
-\TODO{Should we mention that we empirically show that pash produces
-  correct code?}
-
+%> \TODO{Remember to mention that deathstar has 128 threads but 64
+%>   cores. So it is reasonable for the speedup to not go further in cpu
+%>   heavy pipelines.}
+%> 
+%> \TODO{Should we mention that we empirically show that pash produces
+%>   correct code?}
+%> 
 % FIXME
@@
 
-\begin{table*}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of shell one-liners}.
-    The shell one-liners are small pipelines (\ie ones developed on-the-fly) drawn from various sources and applied to large datasets.
-  }
-}
-%% \begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllll}
-%% \toprule
-%%   Script                 ~&~ Structure                    & Input (MB)& Time (Seq)    & Script Size (LoC)             & Highlights                                        \\
-%% \midrule
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/grep.sh
-%%   \tti{grep}         ~&~$3\times$\tsta                & 1000      &  78m 36.77s   &   30\qquad 134\qquad  1304    & complex NFA regex                                 \\
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/minimal5.sh
-%%   \tti{sort}         ~&~~$\tsta, \tpur$               & 10000     &  23m 19.85s   &   33\qquad 129\qquad  1209    & \tti{sort}ing                                     \\
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-%%   \tti{wf}               ~&~~$3\times\tsta, 3\times\tpur$ & 1000      &   2m 11.18s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/topn.sh
-%%   \tti{top-n}            ~&~~$2\times\tsta, 4\times\tpur$ & 1000      &   6m 53.55s   &   53\qquad 181\qquad  1621    & double \tti{sort}, \tti{uniq} reduction           \\
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/ngrams.sh
-%%   \tr{\tti{bi-gram}          ~&~~$2\times\tsta, 4\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & stream shifting and merging                       \\}
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/spell.sh
-%%   \tti{spell}            ~&~~$4\times\tsta, 3\times\tpur$ & 1000      &   8m 13.42s   &  37 \qquad 257\qquad  2417    & comparisons (\tti{comm})                          \\
-%% % https://github.com/andromeda/sdsh/blob/master/scripts/diff.sh
-%%   \tr{\tti{diff}             ~&~~$4\times\tsta, 3\times\tpur$ & \todo{X}  &               &   X \qquad  X \qquad          & shuffling and non-distributable \tti{diff}ing     \\}
-%%   \tti{shortest-scripts} ~&~~$5\times\tsta, 2\times\tpur$ & 8.6       &   2m 50.03s   &  61 \qquad 253\qquad  2413    & extensive file-system operation   \\
-%% \bottomrule
-%% \end{tabular*}
-\input{microbenchmarks-table}
-\label{tab:eval}
-\end{table*}
-
-\kk{Maximum scaleup for both diff and set-diff is around half because
-  they already have two sorts running in parallel (as diff, and comm
-  take two inputs).}
+%>\kk{Maximum scaleup for both diff and set-diff is around half because
+%>  they already have two sorts running in parallel (as diff, and comm
+%>  take two inputs).}
 
@@
 
-\TODO{(Maybe) we need to mention that eager has disk usage overhead
-  that is linear to the size of the pipeline and the size of the
-  input.}
-
-At a high level, we are interested in understanding whether \sys can indeed scale pipelines out automatically and correctly.
-To achieve this, we use six micro-benchmarks and one macro-benchmark, collected primarily from prior literature.
-
-Micro-bench\-marks~\sx{ours} are one-off, simple one-liners that take a few seconds to write;
-  such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-These pipelines are taken from real use cases~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked} and, precisely due to their small size, highlight stress on a handful of stages.
-
-The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
-It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-  as shown, if written in a conventional language it would correspond to a program on the order of hundreds of lines of code.
-
-Several results are worth highlighting.
-First, programs see a significant speedup, often more $10\times$ over sequential execution  while returning identical results (on very large datasets).
+%> \TODO{(Maybe) we need to mention that eager has disk usage overhead that is linear to the size of the pipeline and the size of the input.}
+
+At a high level, we are interested in understanding whether \sys can indeed scale shell scripts out automatically and correctly.
+To answer this question, we use several scripts collected out from the wild, along with a few micro-benchmarks for targeted comparisons.
+
+% Micro-bench\-marks~\sx{ours} are one-off, simple one-liners that take a few seconds to write;
+%   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
+% These pipelines are taken from real use cases~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked} and, precisely due to their small size, highlight stress on a handful of stages.
+% The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
+% It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
+%   as shown, if written in a conventional language it would correspond to a program on the order of hundreds of lines of code.
+
+\heading{Highlights}
+\fixme{
+Scripts see a significant speedup, often up to $10\times$ over sequential execution while returning identical results (on very large datasets).
 \sys's transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
@@
   the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
+}
   %% in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
 
-Experiments were run on a %\todo{network of five workstations} connected by 1Gbps links: 
-  machine with 512GB of memory and 128 virtual (64 physical) $\times$ 2.1GHz Intel Xeon E5-2683 cores.
-%\todo{four smaller machines (\wkq), each with 4GB of memory and two 3.33GHz Intel Core Duo E8600 processors}.
-For our software setup, we use Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
-Generally, no special configuration was made in hardware or software.
-All pipelines are set to (initially) read from and (finally) write to the file-system (except as otherwise stated).
-Whenever \ttt{curl} is part of a pipeline, it fetches data from a different physical host on the same network connected by 1Gbps links.
-A more detailed description of the setup and experiments, including reproducible scripts, is included in the accompanying (anonymized) online repository.
+\heading{Setup}
+Measurements were taken on a machine with 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores running Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
+No special configuration was made in hardware or software.
+Except as otherwise noted,
+  (i) all pipelines are set to (initially) read from and (finally) write to the file-system,
+  (ii) \ttt{curl} fetches data from a different physical host on the same network connected by 1Gbps links.
+% A more detailed description of the setup and experiments, including reproducible scripts, is included in the accompanying (anonymized) online repository.
 
@@
 
-The goal of this subsection is to evaluate \sys on popular, common, or classic \unix pipeline patterns.
+The goal of this subsection is to evaluate \sys on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
 
@@
 
-%% \begin{figure*}[t]
-%%     \centering
-%%     %% \subcaptionbox{\label{eval:minimal_grep}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \subcaptionbox{\label{eval:grep}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/grep_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \subcaptionbox{\label{eval:minimal_sort}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \subcaptionbox{\label{eval:topn}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \subcaptionbox{\label{eval:wf}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \subcaptionbox{\label{eval:spell}}{
-%%     %%     \includegraphics[width=0.32\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-%%     %% }
-%%     %% \caption{a}
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_grep_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:minimal_grep}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/spell_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:grep}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/minimal_sort_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:minimal_sort}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%       \centering
-%%       \includegraphics[width=\textwidth]{\detokenize{./figs/set-diff_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:minimal_sort}
-%%     \end{subfigure}%
-
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/topn_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:topn}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/wf_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:wf}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%         \centering
-%%         \includegraphics[width=\textwidth]{\detokenize{./figs/shortest_scripts_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:spell}
-%%     \end{subfigure}%
-%%     ~
-%%     \begin{subfigure}[b]{0.24\textwidth}
-%%       \centering
-%%       \includegraphics[width=\textwidth]{\detokenize{./figs/bigrams_throughput_scaleup.pdf}}
-%%         %% \caption{TODO}
-%%         %% \label{eval:minimal_sort}
-%%     \end{subfigure}%
-%%     \caption{
-%%       \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
-%%       Four different configurations per benchmark:
-%%       (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
-%%       (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
-%%       (ii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
-%%       (iii) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
-%%       (iv) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
-%%       \nv{Maybe it makes sense to show log x-axis?}
-%%       \kk{How can we also fit two more plots (sort-sort and diff) here?
-Maybe we can fit one by getting rid of sort (since it also shown in the sort--parallel one).
-Otherwise we could separate the ``split'' plots from the ones that don't have split.}
-%%     }
-%%     \vspace{-15pt}
-%%     \label{fig:microbenchmarks}
-%% \end{figure*}
-
 \heading{Programs}
-
-\kk{How can we include a result for grep-light?  This is the IO
-  intensive CPU-light grep. \sys has no speedup but also no slowdown
-  (even if it IO heavy). Should we just mention that in text in
-  passing?}
-
-\kk{Remember to mention that for spell and bigrams we dont show other
-  parallel because they are all close to 1 (no speedup).}
-
-Tab.~\ref{tab:eval} summarizes the collection of micro-benchmarks used to evaluate \sys.
-Benchmark names (col. 1) correspond to the names on the plots (Fig.~\ref{fig:microbenchmarks}), and structure (col. 2) summarizes the different classes of commands used in the script.
-% The centralized pipelines read files of various sizes from disk and write to disk.
-Input sizes and times (col. 3, 4) summarize the characteristics of the largest sequential executions. % (200 files in Fig.~\ref{fig:microbenchmarks}).
-Script sizes (col. 5) report on \sys's rewritten output for three different distribution sizes---$1\times$, 10$\times$, and $100\times$.
-Size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
-The first number alone is interesting, as it captures \sys's output for the sequential script.
-As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
-% this is the size of 
-This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-FIFOs are simply pipes made explicit;
-  the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
-
-In more detail, \ttt{grep} is a short script centered around an expensive \tsta command;
+%>\kk{How can we include a result for grep-light?  This is the IO
+%>  intensive CPU-light grep. \sys has no speedup but also no slowdown
+%>  (even if it IO heavy). Should we just mention that in text in
+%>  passing?}
+%>
+%>\kk{Remember to mention that for spell and bigrams we dont show other
+%>  parallel because they are all close to 1 (no speedup).}
+Tab.~\ref{tab:eval} summarizes the collection of programs shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
+Grep and Grep-light are in \sta, with the former centered around an expensive DFA-based backtracking expression.
 % a DFA-based regular expression matching phase;
-  while the \unix \ttt{grep} command defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
-The \ttt{sort} pipeline is equally short, but its central operation is a command in \pur.
-The next two benchmarks, \ttt{wf} and \ttt{top-n}, are based on McIlroy's now classic word-counting program~\cite{bentley1986literate};
+ % while the \unix \ttt{grep} command defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
+Sort is a short script centered around a \pur command.
+Wf and Top-n are based on McIlroy's classic word-counting program~\cite{bentley1986literate};
   they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
@@
 %   it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
-The next pipeline is based on the original \ttt{spell} program developed by Johnson~\cite{bentley1985spelling}---another \unix classic:
-  after some preprocessing, it makes clever use of \ttt{comm} to report words not in the dictionary.
-Finally, \ttt{shortest-scripts} is a pipeline~\cite[pg. 7]{taylor2004wicked}that identifies the 15 shortest shell scripts in the user's \ttt{PATH};
-   interestingly, it uses the \ttt{file} utility and a higher-order \ttt{wc} via \ttt{xargs}.
-
-\heading{Performance}
+Spell, based on the original \ttt{spell} developed by Johnson~\cite{bentley1985spelling}, is another \unix classic:
+  after some preprocessing, it makes clever use of \ttt{comm} to report words not in a dictionary.
+Shortest-scripts extracts the 15 shortest scripts in the user's \ttt{PATH}, using the \ttt{file} utility and a higher-order \ttt{wc} via \ttt{xargs}~\cite[pg. 7]{taylor2004wicked}.
+Diff and Set-diff compare streams via a \ttt{diff} (in \npu, non-parallelizable) and \ttt{comm} (in \pur), respectively.
+Sort-sort uses consecutive \pur commands without interleaving them with a \pur that has the effect of significantly condensing the input size (\eg \ttt{uniq}).
+The two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams over the input; the \ttt{optimized} version \todo{has a difference}.
+
+\heading{Results}
 % TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (1--200$\times$).
+Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (2--64$\times$).
 % we experiment with a $200\times$ level of parallelism to see potential overheads
-Each one of these plots reports on three different configurations:
-  (i) distributed, which captures only the runtime of the distributed computation,
-  (ii) +compile, which adds the \sys' analysis and generation overhead, and
-  (iii) +merge, which adds a final merger  gathering the results at the end of the pipeline.
-Parallelism of 200$\times$ was explicitly chosen to show the \sys-internal overheads on suboptimal configurations;
-  \sys itself would never attempt a factor that is above the number of available compute nodes.
-The results show significant speedups, ranging between 4--70$\times$, depending on the parallelizability characteristics of individual pipelines.
+
+In Tab.~\ref{tab:eval} contains two benchmarks that are not shown in Fig.~\ref{fig:microbenchmarks}, as they are artificial modifications intended to highlight \sys-internal characteristics and not lead to a speedup.
+  \todo{highlight 2--3 results on things that were not shown in Fig.7}
+%   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
+% The first number alone is interesting, as it captures \sys's output for the sequential script.
+% As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
+% % this is the size of 
+% This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
+% FIFOs are simply pipes made explicit;
+%   the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
 
@@
 \heading{Take-aways}
+\sys achieves significant speedups that range between 4--60$\times$ depending on the parallelizability characteristics of the commands used.
+\sys's runtime constructs offer significant improvements over the baseline speedup achieved by \sys's core transformations.
 
@@
   \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \includegraphics[width=0.80\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
   \caption{
-	  \textbf{Speedup (left axis) and over sequential execution (right axis) for all Unix50 pipelines.}
+	  \textbf{Unix50 scripts.}
+    Speedup (left axis) over sequential execution (right axis) for Unix50 scripts.
 		Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.

### 2020-05-22 pldi/draft.tex
More eval

@@
 
-At a high level, we are interested in understanding whether \sys can indeed scale shell scripts out automatically and correctly.
-To answer this question, we use several scripts collected out from the wild, along with a few micro-benchmarks for targeted comparisons.
+This section reports on whether \sys can indeed scale shell scripts out automatically and correctly, using several scripts collected out from the wild along with a few micro-benchmarks for targeted comparisons.
 
@@
 \heading{Setup}
-Measurements were taken on a machine with 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores running Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0.
-No special configuration was made in hardware or software.
+\sys was run on 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores, Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0---without any special configuration in hardware or software.
 Except as otherwise noted,
@@
 \heading{Discussion}
-\kk{Below follows an important point about why scalability is more
-  than great in the beginning and then gets worse.}
-An interesting point is that \sys seems to achieve almost linear
-speedup with small parallelism configurations. The reason why this
-happens is not so straightforward though. \sys manages that because it
-exploits task-based parallelism by having different processes to do
-the \todo{map} and the \todo{reduce} phase. Due to task-based
-parallelism in the generated dataflow graphs, \sys essentially ``uses
-more parallelism'' than the given configuration. This can be seen in
-all pipelines that are not completely stateless. For example, in the
-\ttt{minimal\_sort} one-liner, \sys configured to have 8 parallelism
-spawns 37 nodes (8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \ttt{sort -m}
-nodes, and 14 \ttt{eager} nodes. This indicates that \sys achieves
-optimal performance (without unnecessary process spawns and context
-switches) in pipelines that contain pure commands by configuring it
-with a smaller parallelism than the capabilities of the underlying
-system -- in our case 16-32 for a 64 physical core system. This is
-also the case if the input pipeline is very long and so there is
-already task-based parallelism even in the sequential execution
-(several of the Unix50 pipelines fall in this category).
-
-\kk{Hypothesis on a scalability upper bound. Dish gives a great
-  speedup if one step of the pipeline is what takes the most time. If
-  the pipeline is long with many computation heavy tasks, task based
-  parallelism gets you part of the way there, so we can't expect Dish
-  to give performance benefits equal to the amount of parallelism that
-  we give it.}
-
-\TODO{Briefly discuss the performance drop as we increase parallelism
-  in some one-liners (sort-sort, bi-grams).The reason has to do with
-  context switching. More precisely, sort-sort 32 has 314 nodes (and
-  54m sys time) and sort-sort 64 has 634 nodes (and 90m! sys time). On
-  the other hand opt-bigrams 64 has 255 nodes and the sys time stays
+\sys seems to achieve near-linear speedup on low-parallelism configurations.
+This is due to a high degree of task-based parallelism from distinct \emph{map} and \emph{fold} processes, which results in a longer tree of task-parallel stages.
+This is seen in all that include components in \pur.
+For example, Sort in $8\times$ spawns 37 nodes:
+  8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \ttt{sort -m} nodes, and 14 \ttt{relay} nodes.
+In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
+significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
+
+Thus, asking for parallelism beyond that point results in worse speedup, due to other overheads that start dominating as no further speedup is available.
+Sort-sort with $32\times$ has 314 nodes (and 54m sys time) and with $64\times$ has 634 nodes (and 90m! sys time).
+ On the other hand opt-bigrams 64 has 255 nodes and the sys time stays
   the same so I think that it might have to do with the fact that
   constant costs rise a lot (since opt-bigrams 32 takes just 1
-  minute).}
-
-\kk{It is visible that the automatic split is slightly worse than the
-  manual one in most cases (because it doesn't utilize task-based
-  parallelism).}
-
-We attribute the small drop above $96\times$ in \ttt{spell} \ttt{sort} \ttt{wf}
-and \ttt{bi-grams} to a combination of hyperthreading (128 virtual vs. 64
-physical CPUs) \todo{the fixed width on} \ttt{sort}.
-% I say "we attribute" because it's a way to say something without being certain
-% FIXME: We need to take linear benchmarks on deathstar and livestar to see
-% where it starts dropping
-% not explain scope the results appropriately
-% rinard
+  minute).
+
+%> \kk{It is visible that the automatic split is slightly worse than the
+%>   manual one in most cases (because it doesn't utilize task-based
+%>   parallelism).}
+
+%> This is also the case if the input pipeline is very long and so there is already task-based parallelism even in the sequential execution (several of the Unix50 pipelines fall in this category).
+%> \kk{Hypothesis on a scalability upper bound. Dish gives a great
+%>   speedup if one step of the pipeline is what takes the most time. If
+%>   the pipeline is long with many computation heavy tasks, task based
+%>   parallelism gets you part of the way there, so we can't expect Dish
+%>   to give performance benefits equal to the amount of parallelism that
+%>   we give it.}
 
@@
 
-\heading{Programs} In a recent event celebrating \unix's 50-year
-legacy, Bell Labs created a set of \todo{36} challenges~\cite{unix50}
-solvable by means of composing \unix pipelines.  By scanning GitHub,
-we were able to find a set of shell programs that solve
-\todo{all-but-three} problems~\cite{}.  We consider this a good set of
-benchmarks because (i) the problems were designed to to highlight the
-\unix philosophy~\cite{} (rather than having a single command that
-solves the problem) and make extensive use of \unix built-ins with a
-plethora of flags, and (ii) the solutions were written by non-experts
-(contrary to the \S\ref{ours}) and were executed by \sys
-unmodified---stressing patterns that we, \sys's authors, did not
-necessarily anticipate \kk{I would change that to ``stressing a
-  variety of patterns'' since it doesn't really matter if we
-  anticipated them (especially since we can address any issues and
-  then re-run them).} .  In cases where an obvious fix improves the
-pipeline's performance, we report on both \sys's original speedup and
-the one achieved by applying the fix.
+\heading{Programs}
+In a recent celebration of \unix's 50-year legacy, Bell Labs created 37 challenges~\cite{unix50} solvable by means of composing \unix pipelines.
+We found unofficial solutions to all-but-three problems on GitHub~\cite{unix50sol}, expressed as pipelines with 2--12 stages (avg.: 5.58).
+We consider this a good set of benchmarks because
+  (i) the problems were designed to highlight the \unix modular philosophy~\cite{mcilroy1978unix} and make extensive use of built-ins with a variety of flags, and
+  (ii) the solutions were written by non-experts (contrary to \S\ref{ours}).
+% ---stressing a patterns that we, \sys's authors, did not
+% necessarily anticipate \kk{I would change that to ``stressing a
+%   variety of patterns'' since it doesn't really matter if we
+%   anticipated them (especially since we can address any issues and
+%   then re-run them).} . 
+\sys executes each pipeline as-is; 
+  if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
 % TODO: talk about correct wrt developer intentions vs. correct wrt problem definition
@@
 %%   neligible).}
-
-\heading{Results} \sys's parallelism set to 16. The input given to the
-pipelines was the test input from unix50 multiplied to reach
-10GBs. The speedup of each pipeline separately is shown in
-\Cref{fig:unix50-individual}. The average\todo{/median} speedup is
-$5.49$/\todo{$6.07\times$}, the \todo{geometric mean is $1.08\times$},
-and the \todo{weighted average (with the absolute times as weights) is
-  $5.75\times$}.
+\heading{Results}
+\sys's parallelism was set to $16\times$, as many of these pipelines are relatively long (\S\ref{ours}, Discussion).
+Their input was generated by multiplying the original Unix50 input several times to reach 10GB.
+Fig~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
+The average speedup is $5.49$ ($6.07\times$), and the weighted average (with the absolute times as weights) is $5.75\times$.
+ % the \todo{geometric mean is $1.08\times$},
 
@@
 \heading{Discussion}
+Most pipelines see significant acceleration---except 13, 24, 25, 26, 29, and 30 that see no speedup and 2, 19, and 31 that see a slowdown.
+Ones in the first group contain stages that cannot be parallelized in general---\eg \ttt{awk}, \ttt{sed} \ttt{-d}, and \ttt{tr} \ttt{-d}.
+However, by applying trivial changes
+that change some of these commands with their equivalents in \sta---\eg, replacing a general \ttt{awk} with more specialized \ttt{cut} that achieves the same effect---\sys 
+can identify and exploit parallelism:
+  for, say, \todo{13} this leads  to \todo{XX}$\times$ speedup.
+This is a common pattern we witnessed elsewhere too~\sx{macro1}.
+
+Ones in the second group contain a \ttt{head}, thus always processing only one line and executing within $10ms$ regardless of the full input.
+\sys's slowdown is due to constant setup costs---\eg pipes, processes---but remains within $1s$.
+
+For pipelines  \sys performs the same as the sequential implementation.
+
+% illustrative example is pipeline 13, which is shown here:
+% 
+% \begin{lstlisting}[language=sh, float=h, numbers=none]
+%   cat $IN | awk "{print \$2, \$0}" |
+%       sort -nr | cut -d ' ' -f 2
+% \end{lstlisting}
+% 
+% \kk{Rephrase that to sound more meaningful and surprising.}  Notice
+% that \ttt{awk} is just used to permute two of the columns of the input
+% to enable sorting. One could write this pipeline in a different way,
+% by directly sorting on the second column:
+% 
+% \begin{lstlisting}[language=sh, float=h, numbers=none]
+%   cat $IN6 | sort -nr -k 2 | cut -d ' ' -f 1
+% \end{lstlisting}
+
+Finally, for the rest of the pipelines \sys speedup is \todo{capped}
+because of a combination of the following reasons: (i) they contain
+pure commands that are parallelizable but don't scale linearly, such
+as \ttt{sort} (e.g. pipelines 0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
+(ii) they are deep pipelines that already exploit task-based
+parallelism (e.g. pipelines 7, 8, 9, 10, 11, 12, 14, 27, 28), and
+(iii) they are not cpu-intensive, leading to \todo{larger IO and
+  constant costs} (e.g. pipelines 4, 5, 6, 7, 8, 12, 14, 22, 23).
+
+%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
+%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
+%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14, 22, 23
+%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
+
+%% Finally, an important point that is reinforced from this
+%% experiment is that \sys doesn't lead to slowdown, even in cases where
+%% parallelism is not possible.
 
 %% Breakdown of not ideal speedups in 8, 16 parallelism
-
 %% Head: 2, 19, 31
@@
 %%  32. TODO: Implement
-
-\sys achieves speedup in all pipelines except for 2, 19, and 31. These
-pipelines include a \ttt{head} and only process one line of the
-input. This can be seen by their sequential execution time which is in
-the order of 10 milliseconds. The slowdown is due to constant costs of
-setting up the pipes and spawning processes. In all three of these
-cases the execution time with \sys is in the order of 1 second. For
-pipelines 13, 24, 25, 26, 29, and 30, \sys performs the same as the
-sequential implementation. This is because these pipelines contain
-commands that can not be parallelized in general, such as \ttt{awk},
-\ttt{sed} with option \ttt{d}, and \ttt{tr} with option \ttt{-d}. An
-illustrative example is pipeline 13, which is shown here:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
-  cat $IN | awk "{print \$2, \$0}" |
-      sort -nr | cut -d ' ' -f 2
-\end{lstlisting}
-
-\kk{Rephrase that to sound more meaningful and surprising.}  Notice
-that \ttt{awk} is just used to permute two of the columns of the input
-to enable sorting. One could write this pipeline in a different way,
-by directly sorting on the second column:
-
-\begin{lstlisting}[language=sh, float=h, numbers=none]
-  cat $IN6 | sort -nr -k 2 | cut -d ' ' -f 1
-\end{lstlisting}
-
-By executing this version of the pipeline with \sys, we achieve
-\todo{X} speedup. This indicates that just a little awaresess of
-command parallelizability can go a long way, and that the effort that
-a \sys user needs to spend is often negligible to achieve
-benefits.
-
-Finally, for the rest of the pipelines \sys speedup is \todo{capped}
-because of a combination of the following reasons: (i) they contain
-pure commands that are parallelizable but don't scale linearly, such
-as \ttt{sort} (e.g. pipelines 0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
-(ii) they are deep pipelines that already exploit task-based
-parallelism (e.g. pipelines 7, 8, 9, 10, 11, 12, 14, 27, 28), and
-(iii) they are not cpu-intensive, leading to \todo{larger IO and
-  constant costs} (e.g. pipelines 4, 5, 6, 7, 8, 12, 14, 22, 23).
-
-%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
-%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
-%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14, 22, 23
-%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
-
-%% Finally, an important point that is reinforced from this
-%% experiment is that \sys doesn't lead to slowdown, even in cases where
-%% parallelism is not possible.
 

### 2020-05-22 pldi/draft.tex
Heavy pass on section 2

@@
   \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S. for the years 1920--2020 and extracts the maximum for every year.
+  The script downloads daily temperatures recorded across the U.S.
+for the years 1920--2020 and extracts the maximum for every year.
+\kk{Make this to be for 10 years only so that the point in the second challenge holds}
 }
@@
 
-While these features aid developer economy through powerful program composition, they complicate shell script parallelization which even for simple scripts such as the one in Fig.~\ref{fig:example} face several challenges.
+While these features aid developer economy through powerful program
+composition, they complicate shell script parallelization; which even
+for simple scripts such as the one in Fig.~\ref{fig:example} faces
+several challenges.
 % \footnote{
@@
 The first challenge is the nature and variety of shell commands.
-In contrast to restricted programming frameworks, amenable to parallelization by supporting a few carefully designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, shell scripts contain commands that can have arbitrary behaviors.
-For Fig.~\ref{fig:example}, each of command requires special treatment as to not introduce non-determinism or incorrectness:
-  \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
-These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
+In contrast to restricted programming frameworks that enable
+%% amenable to
+parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, shell scripts contain black-box commands that are written in a variety of programming languages and models and can have arbitrary behaviors.
+Therefore, each of them requires special treatment in order to be parallelized.
+For example, in Fig.~\ref{fig:example}
+%% each command requires special treatment as to not introduce non-determinism or incorrectness:
+\ttt{grep} can be parallelized simply by partitioning the lines of its input, while \ttt{sort} would require sorting each input partition and then merging the results.
+%% \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
+%% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
 Worse even, the same command may be used with different flags---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags---that could potentially alter its behaviour.
@@
 
+\kk{I tried to rephrase the above a little, but I can't get it just right.
+There are two parts to this argument.
+First, the fact that commands are arbitrary (and not known) requires that each of them has special treatment.
+However, the huge number of them, and the fact that flags change their behaviour even more makes an analysis of all commands a very difficult task.}
+
 % A good solution should handle several commands but should also be extensible, in the sense that newly added or modified commands don't require modifications in the solution itself.
@@
 
-\heading{Scripts} % 
-Another challenge is due to the language of the shell, which % has been developed through a long history of additions.
-  contains constructs with different parallelizability semantics. % that enforce sequential execution.
-For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), has to be completed before the rest of the script starts executing.
-Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
+\heading{Scripts} % Restrictive \& Coarse-grained}
+Another challenge is due to the language of the shell, which
+% has been developed through a long history of additions.
+is both restrictive and coarse-grained.
+First, it contains constructs that enforce sequential execution; in the example in \cref{fig:example}, the sequential composition operator \ttt{;} enforces that the assignment to \ttt{base} is completed before the rest of the script starts executing.
+Moreover, the structure of shell scripts only exposes very coarsed-grained forms of parallelism (usually in the form of task parallelism using the \ttt{\&} and \ttt{|} constructs).
+%%   contains constructs with different parallelizability semantics. % that enforce sequential execution.
+%% For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), has to be completed before the rest of the script starts executing.
+%% Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
 For example, even if Fig.~\ref{fig:example}'s script was aiming on only six years of data, the first \ttt{curl} would still output hundreds of lines for each year;
@@
 \heading{Implementation} %
-A final challenge has to do with practical issues.
-Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
-Moreover, the \unix environment has its set of quirks % about opening streams, deadling with FIFOs,
-  and any attempt at parallelization will have to take care of these challenges---related to task parallelism, deadlock prevention, and runtime performance.
+%
+A final challenge has to do with practical issues, as even manual
+parallelization of a given shell script is far from
+trivial. Parallelizing a script without reimplementing all of its
+commands, requires primitives to compose existing commands to achieve
+parallelization. Designing correct and efficient compositions is
+particularly challenging in the context of \unix, and requires
+addressing issues related to scheduling, synchronization, and deadlock
+prevention. Moreover, specific commands (such as \ttt{sort} and
+\ttt{uniq}) require specialized data aggregators to be correctly
+parallelized.
+%% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
+%% Moreover, the \unix environment has its set of quirks % about opening streams, deadling with FIFOs,
+%%   and any attempt at parallelization will have to take care of these challenges---related to task parallelism, deadlock prevention, and runtime performance.
 
@@
 % \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs---a two-fold challenge.
-To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but clear set of well-understood parallelizability classes;
-  these commands are already useful-enough to allow the composition of useful shell programs operating in practice.
-To understand the characteristics of commands that fall outside this standard set and allow them to leverage its transformations,
-  \sys defines an annotation language for describing a command's % key properties with respect to its 
-  parallelizability.
+To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but
+%% clear set of well-understood
+\todo{well-defined} set of
+parallelizability classes, which are \todo{useful-enough} to guide the parallelization of simple shell programs.
+%% classes that are useful on their own 
+%%   these classes are already useful-enough to allow the composition of useful shell programs operating in practice.
+To incorporate
+%% To understand the characteristics of
+commands that fall outside of this standard set and allow them to leverage \sys's transformations,
+  we define an annotation language for describing a command's % key properties with respect to its 
+  parallelizability characteristics.
 Such light annotations are expressed once per command (not once per script) and are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they develop.
@@
 \heading{Scripts: Dataflow Graphs}
+%
+To address the synchronization restrictions in shell scripts, \sys
+first analyzes a script to identify \emph{parallelizable regions} that
+can be safely executed in the context of parallelism. \sys uses the
+structure of shell scripts to identify parallelization candidates:
+constructs such as \ttt{\&} and \ttt{|} expose task parallelism, while
+\ttt{\&\&} and \ttt{||} enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
-To assign a candidacy for parallelization, \sys breaks POSIX constructs into a few broad classes:
- (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
- (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
- (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
- (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
-
-Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
-To achieve this, it converts sections of the script that are amenable to parallelization into a data-flow graph (DFG) representation.
+%% To assign a candidacy for parallelization, \sys breaks POSIX constructs into a few broad classes:
+%%  (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
+%%  (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
+%%  (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
+%%  (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
+Since the shell representation only exposes coarse grained parallelization opportunities, \sys then converts each parallelizable region to a dataflow graph.
+%% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
+%% To achieve this, it converts sections of the script that are amenable to parallelization into a data-flow graph (DFG) representation.
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
@@
 \heading{Implementation: Efficient Primitives}
-\sys addresses several practical challenges through a set of constructs it provides---\eg a set of linkable components for stream splitting and merging.
-For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
+\sys addresses several practical challenges through a set of constructs it provides---\eg a set of
+%% linkable
+independent components that can be used to compose individual commands.
+%% for stream splitting and merging.
+It also provides a small library of efficient data-aggregators for a large set of parallelizable commands.
+%% For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
+%% \kk{The terms pure, classes, and fold don't mean a lot to the reader at the moment, so I would prefer using more general terminology.}
 All these commands live in the \ttt{PATH} and are addressable by name, which means they can be used like (and by) any other commands.
@@
 The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
-% They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
 

### 2020-05-22 pldi/draft.tex
Section 3 mini-pass

@@
 
-\heading{Providing Combinators}
-For most commands, the annotations are enough to allow parallelization: 
-  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
-Unfortunately, there are commands that require less obvious tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
-
-To support the parallelization of arbitrary commands in \pur, \sys allows supplying custom \emph{map} and \emph{fold} functions.
+\heading{Providing \todo{Aggregators}}
+For commands in \sta, the annotations are enough to enable parallelization: they are applied to partitions of their input in parallel, and their outputs are simply concatenated.
+%% commands either fall under \sta, having straightforward aggregators, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
+\kk{pure commands are what we have to provide aggregators for, so this is not totally valid.}
+%% Unfortunately, there are commands that
+%% require less obvious tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
+\kk{I would propose deleting this last sentence.
+we have already separated commands into classes, so this sentence makes it seem like the classes don't help at all.}
+%
+To support the parallelization of arbitrary commands in \pur, \sys allows supplying custom \emph{map} and \emph{aggregate} functions.
 In line with the \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
-  (i) \emph{map} is in \sta and \emph{fold} is in \pur,
-  (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{fold} can consume (and combine) the results of multiple \emph{map} invocations, and
+  (i) \emph{map} is in \sta and \emph{aggregate} is in \pur,
+  (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{aggregate} can consume (and combine) the results of multiple \emph{map} invocations, and
   (iii) their composition produces the same output with the original command.
-If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{fold} equivalents, enabling their parallelization.
+If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their parallelization.
 
@@
 
-The core of \sys is a dataflow graph (DFG) model that is expressive enough to support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
-A fundamental difference with other DFG models is that it contains information about the order in which a node reads its inputs.
-This enables a set of graph transformations that expose data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
-These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources.
+The core of \sys is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
+%% representation that is expressive enough to model data parallelism in shell programs.
+%% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
+\sys first lifts sections of the input script to the DFG
+representation, it performs transformations to expose parallelism, and
+then instantiates each DFG back to a parallel shell script.  A
+fundamental difference with other DFG models is that it encodes the
+order in which a node reads its inputs, which in turn enables a set of
+graph transformations that can be iteratively applied to expose
+parallelization opportunities for \tsta and \tpur commands.
+%% data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
+%% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
+In \Cref{graph-components} we define the DFG model and describe the
+correspondence with shell scripts, and in \Cref{ir:transformations} we
+describe the graph transformations and \todo{give intuition about their
+correctness}.
 
@@
 The two main shell abstractions are
-  (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}\nv{Not in this version. Should we call them "streams"?}) that contain data, and
-  (ii) commands that communicate through these files.
-\nv{quick solution: change all files to streams and make the equivalence explicit here}
-\sys's DFG model represents commands as nodes and \todo{streams} as edges.
+  (i) streams, i.e. files or pipes, that contain data, and
+  (ii) commands that communicate through these streams.
+%% \nv{quick solution: change all files to streams and make the equivalence explicit here}
+\sys's DFG model represents commands as nodes and streams as edges.
 
@@
 pure commands $f$ can be implemented as a combination of a function map
-$m$ and an associative fold $f'$, satisfying the following equation:
+$m$ and an associative aggregate $agg$, satisfying the following equation:
 
 \[
-\forall x, x', s, f(x \cdot x', s) = f'(m(x, s) \cdot m(x', s), s)
+\forall x, x', s, f(x \cdot x', s) = agg(m(x, s) \cdot m(x', s), s)
 \]
@@
 This means that we get the same output by applying $f$ to a
-concatenation of two inputs $x, x'$ as by applying the fold function
-$f'$ to the concatenation of the outputs produced by applying map $m$
+concatenation of two inputs $x, x'$ as by applying the aggregation function
+$agg$ to the concatenation of the outputs produced by applying map $m$
 to each of $x$ and $x'$.
@@
 
-This transformation can extended straightforwardly to pure commands implemented by a map-fold pair
-$(m, r)$ as $ v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s)
-\cdot m(x_2, s) \cdots m(x_n, s), s)$.
+This transformation can extended straightforwardly to pure commands
+implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
+\cdots x_n, s) \Rightarrow agg(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n,
+s), s)$.
 % \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.

### 2020-05-22 pldi/draft.tex
Pass over related work

--- 2020-05-22 pldi/draft.tex
@@
 \heading{Discussion}
-\sys seems to achieve near-linear speedup on low-parallelism configurations.
-This is due to a high degree of task-based parallelism from distinct \emph{map} and \emph{fold} processes, which results in a longer tree of task-parallel stages.
-This is seen in all that include components in \pur.
+\sys achieves near-linear speedup on low-parallelism configurations.
+This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{fold}s and \ttt{relay}s---evident in all plots that contain \pur stages.
 For example, Sort in $8\times$ spawns 37 nodes:
@@
 significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-
-Thus, asking for parallelism beyond that point results in worse speedup, due to other overheads that start dominating as no further speedup is available.
-Sort-sort with $32\times$ has 314 nodes (and 54m sys time) and with $64\times$ has 634 nodes (and 90m! sys time).
- On the other hand opt-bigrams 64 has 255 nodes and the sys time stays
-  the same so I think that it might have to do with the fact that
-  constant costs rise a lot (since opt-bigrams 32 takes just 1
-  minute).
+Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available.
+Sort-sort with $32\times$ has 314 nodes (sys: 54m) and with $64\times$ has 634 nodes (sys: 90m).
+% 
+%  On the other hand opt-bigrams 64 has 255 nodes and the sys time stays
+%   the same so I think that it might have to do with the fact that
+%   constant costs rise a lot (since opt-bigrams 32 takes just 1
+%   minute).
 
@@
 \heading{Results}
-\sys's parallelism was set to $16\times$, as many of these pipelines are relatively long (\S\ref{ours}, Discussion).
-Their input was generated by multiplying the original Unix50 input several times to reach 10GB.
+\sys's parallelism is set to $16\times$, as many of these pipelines are relatively long (\S\ref{ours}, Discussion).
+Their input was generated by multiplying their original input several times, enough to reach 10GB.
 Fig~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
-  \caption{
-	  \textbf{Unix50 scripts.}
-    Speedup (left axis) over sequential execution (right axis) for Unix50 scripts.
-		Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-  }
-  \vspace{-15pt}
-  \label{fig:unix50-individual}
-\end{figure}
-
 \heading{Discussion}
-Most pipelines see significant acceleration---except 13, 24, 25, 26, 29, and 30 that see no speedup and 2, 19, and 31 that see a slowdown.
-Ones in the first group contain stages that cannot be parallelized in general---\eg \ttt{awk}, \ttt{sed} \ttt{-d}, and \ttt{tr} \ttt{-d}.
-However, by applying trivial changes
-that change some of these commands with their equivalents in \sta---\eg, replacing a general \ttt{awk} with more specialized \ttt{cut} that achieves the same effect---\sys 
-can identify and exploit parallelism:
-  for, say, \todo{13} this leads  to \todo{XX}$\times$ speedup.
-This is a common pattern we witnessed elsewhere too~\sx{macro1}.
-
-Ones in the second group contain a \ttt{head}, thus always processing only one line and executing within $10ms$ regardless of the full input.
-\sys's slowdown is due to constant setup costs---\eg pipes, processes---but remains within $1s$.
-
-For pipelines  \sys performs the same as the sequential implementation.
+Most pipelines see significant acceleration, except 13, 24, 25, 26, 29, 30 that see no speedup and 2, 19, 31 that see a slowdown.
+The first group contain commands which \sys cannot parallelize---\eg \ttt{awk}, \ttt{sed} \ttt{-d}, and \ttt{tr} \ttt{-d}.
+Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
+By applying only the replacements which to us experts felt more natural, we also give \sys the opportunity to identify and exploit parallelism---a common pattern we witnessed~\sx{macro1}.
+For example, after this change, \todo{13} sees a \todo{XX}$\times$ speedup.
+
+The second group contains \ttt{head}, thus practically processing only one line (regardless of the full input) and completing within $10ms$.
+\sys's slowdown is due to constant setup costs of pipes and processes but execution still remains under $1s$.
 
@@
 
-Finally, for the rest of the pipelines \sys speedup is \todo{capped}
-because of a combination of the following reasons: (i) they contain
-pure commands that are parallelizable but don't scale linearly, such
-as \ttt{sort} (e.g. pipelines 0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
-(ii) they are deep pipelines that already exploit task-based
-parallelism (e.g. pipelines 7, 8, 9, 10, 11, 12, 14, 27, 28), and
-(iii) they are not cpu-intensive, leading to \todo{larger IO and
-  constant costs} (e.g. pipelines 4, 5, 6, 7, 8, 12, 14, 22, 23).
+For the rest, \sys's speedup is \todo{capped} due to a combination of reasons:
+  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} (0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
+  (ii) some are deep pipelines that already exploit task-based parallelism (7, 8, 9, 10, 11, 12, 14, 27, 28,) and
+  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs (4, 5, 6, 7, 8, 12, 14, 22, 23).
 
@@
 
-\heading{Take-aways} \sys leads to significant performance
-improvements in completely unmodified \unix pipelines found out in the
-wild. Small tweaks can yield further improvements, showing that
-\sys-awareness and scripting expertise can improve
-results. Furthermore, it doesn't cause slowdown in any non-trivial
-computation, even in cases where parallelism is not possible.
+\heading{Take-aways} \sys accelerates unmodified pipelines found in the wild;
+  small tweaks can yield further improvements, showing that \sys-awareness and scripting expertise can improve results.
+Furthermore, \sys does not decelerate non-trivial computations, even when parallelism is impossible.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=0.80\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \caption{
+	  \textbf{Unix50 scripts.}
+    Speedup (left axis) over sequential execution (right axis) for Unix50 scripts.
+		Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
+  }
+  \vspace{-15pt}
+  \label{fig:unix50-individual}
+\end{figure}
 
@@
 \heading{Program}
-The full version of this program is inspired by the introductory chapter of Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it is used as an example of a realistic mid-scale analytics pipeline.
-
-The original task, as presented in the book, is comprised of three sub-tasks:
-  download weather data from NCDC (shell), convert them to a format that is Hadoop-friendly (shell), and process them to identify the maximum temperature (Hadoop).
-Only the last of these tasks is the real focus of the book, demonstrating the benefits of using distributed framework such as Hadoop.
-For \sys, we consider the full pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
-
-The full program used for our evaluation is slightly different than the one presented in Fig.~\ref{fig:example}, consisting of 14 stages and totalling 240 characters.
-We full program expresses the \ttt{for} loop as its first pipeline stage and uses \ttt{xargs} to 
-The program presented in Fig.~\ref{fig:example} is simplified only for clarity of exposition.
+This program is inspired by Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it exemplifies a realistic mid-scale analytics pipeline and comprises three sub-tasks:
+  download weather data from NCDC (shell), convert them to a Hadoop-friendly format (shell), and process them to get the maximum temperature (Hadoop).
+Only the last of these tasks is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
+  whereas we consider the end-to-end pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
+% The version used in the evaluation is slightly different than the one presented in Fig.~\ref{fig:example} (14 stages and totalling 240 characters), simplified only for clarity of exposition, in that it expresses the \ttt{for} loop as a first stage followed by \ttt{xargs}.
 
 \heading{Results}
-% The first interesting result 
-\TODO{Some results}
-The Hadoop program, expressed in Java and only capturing the \emph{four} last stages, amounts to 137 lines of Java code---not accounting for commands moving data in and out of HDFS.
-
-As this is benchmark is taken directly from the Hadoop book, it is worth reporting on its execution time.
-While the setup is not typical of Hadoop, the comparison between the two systems is not unreasonable:  % we took several steps to make the comparison fair:
-  (i) both Hadoop and Spark favor message-passing rather than shared memory,
-	(ii) Hadoop has no fault-tolerance for master and worker nodes (\ie both systems do not replicate data),
-	(iii) the task at hand is a good fit for  Hadoop,
-	(iv) the Hadoop program is written as a single program tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases.
-\todo{The original pipeline executes in XXs---long enough to amortize Hadoop's 0.5--0.9s startup costs.}
-With two Hadoop 3.2.1 (Openjdk 11.0.5) data-nodes, the Hadoop program takes 6m22s for data processing and 1m09s for moving the dataset to HDFS.
-For the same amount of parallelism, \sys takes 4m30.554s---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
+The original pipeline executes in \todo{XXs} on \todo{XX}GB of input data.
+Applying \sys with \todo{XX}$\times$ parallelism leads to (XX speedup).
+Naturally, different phases benefit differently from parallelization: \todo{part 1, 2, 3, 4}
+
+It is worth reporting Hadoop's (3.2.1, Oracle JDK1.8.0\_255) execution time for the curious reader (see Discussion for fairness), which takes \todo{6m22s} (sequential) and \todo{4m30.554s} ($\times$\todo{XX}-parallel).
+This amounts to \todo{XX}$\times$ speedup versus \todo{XX}$\times$ for \sys's equivalent (part 4).
+% ---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
+
+In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to \todo{XX} shell LoC for \sys's \todo{part 4}.
 
@@
 \heading{Discussion}
-Similar to Unix50~\sx{unix50}, we found that large, complex pipelines enable significant freedom in terms of expressiveness.
-Several stages of the original program were expressed as a single \ttt{awk} script;
-  unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
-A simpler (but longer) pipeline that leverages on \unix built-ins might be outperformed by a specialized programs on a single node, but turns out to be trivially scalable to multiple nodes.
+Similar to Unix50~\sx{unix50}, we found that large pipelines enable significant freedom in terms of expressiveness.
+A few stages of the original script were expressed in a single \ttt{awk};
+% unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
+  a simpler (but longer) pipeline that leverages on \unix built-ins % might be outperformed by a specialized programs on a single node, but
+  turns out to be trivially parallelizable.
 % For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
@@
 
+In terms of execution fairness, both systems are message-passing rather than shared-memory, and Hadoop's setup is atypical to improve fairness:
+  it operates on non-HDFS file-system and has replication disabled.
+Additionally, the Java program is written as a monolith tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases; and the task at hand is an ideal fit for MapReduce system such as Hadoop.
+Finally, execution time is long enough (\todo{XXh})to amortize Hadoop's startup costs.
+% FIXME OSDI mention tabulation vs. sorting
+
 \heading{Take-aways}
-Multi-line Java Program parallel program decomposed in two phases---a highly parallel map phase and a sequential reduce phase---corresponds to a much smaller shell script with
+A multi-line parallel program specialized for the task at hand corresponds to a much smaller shell script with comparable performance (on a comparable setup).
 
@@
 
-\heading{Program} At a high level, the pipeline is comprised of two
-distinct components that coarsely correspond to a front- and a
-back-end communicating via \unix pipes.  The front-end downloads pages
-and extracts their URLs from their source, feeding back \todo{to} the
-start of the pipeline; the back-end extracts the text from the html
-pages and then applies natural-language processing---\eg trigrams,
-character conversion, term frequencies---to index the incoming pages.
-The pipeline contains a total of 52 stages (18 for the font-end and 34
-for the back-end) written in multiple languages---for example,
-its \ttt{url-extract}ion utility is written in JavaScript whereas its
-\ttt{word-stem}ming utility is in Python.  \sys can still operate on
-them as their parallelizability properties---\sta for
-\ttt{url-extract} and \ttt{word-stem}---can be described by its DSL.
-
-\heading{Results} To simplify execution (and avoid hitting Wikipedia),
-we have saved a recent version of Wikipedia to a server within the
-network and to redirect all \ttt{curl} requests there.  The complexity
-of this pipeline makes end-to-end measurements difficult, so we
-measure the improvements on the two components components individually
-and then report the differences. With parallelism set to \todo{2/16}
-the execution time achieved by \sys is \todo{X} for the front-end and
-\todo{Y} for the back-end, while the original execution time is
-\todo{X} for the front-end and \todo{Y'} for the back-end. \kk{If the
-  front-end takes a lot of time (compared to the back-end), we might
-  better not report it with the premise that it is a feedback cycle
-  that cannot be parallelized by \sys.}
+\heading{Program}
+Coarsely, the pipeline is comprised of two distinct components % coarsely corresponding to a front- and a back-end
+  communicating via \unix pipes.
+The first downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
+ the second extracts the text from HTML and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
+Collectively, they contains a total of 52 stages (first: 18; second: 34) written in multiple languages---\eg
+its \ttt{url-extract}ion utility is written in JavaScript whereas its \ttt{word-stem}ming utility is in Python.
+\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be trivially described by annotations.
+
+\heading{Results}
+% To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
+\sys can operate on the entire program, but as its complexity makes end-to-end measurements difficult we report on the two components components individually.
+With parallelism set to \todo{2/16} the execution time achieved by \sys is \todo{X} for the front-end and \todo{Y} for the back-end, while the original execution time is \todo{X} for the front-end and \todo{Y'} for the back-end.
+
+%> \kk{If the front-end takes a lot of time (compared to the back-end), we might better not report it with the premise that it is a feedback cycle that cannot be parallelized by \sys.}
 
 \heading{Discussion}
-
 Since the front-end contains a feedback cycle, \sys \todo{doesn't
@@
 
-\subsection{Use Case: Container Orchestration}
-\label{macro3}
-
-As a third realistic workload, we use a script for orchestrating containers to test the latest built across various Linux distributions.
-Compatibility testing is a common task for large software projects, commonly automated through shell scripting.
-
-\heading{Program}
-The 30-line orchestration script is unusual among our set of benchmarks in that it is not structured as a linear pipeline.
-Rather, it makes extensive use of POSIX constructs common in the shell, including output redirection, conditionals, and here-documents.
-It also combines commands from other classes outside \sta and \pur---more prominently \dfs---which are not parallelizable by \sys.
-
-\heading{Results}
-
-\heading{Take-aways}
+% \subsection{Use Case: Container Orchestration}
+% \label{macro3}
+% 
+% As a third realistic workload, we use a script for orchestrating containers to test the latest built across various Linux distributions.
+% Compatibility testing is a common task for large software projects, commonly automated through shell scripting.
+% 
+% \heading{Program}
+% The 30-line orchestration script is unusual among our set of benchmarks in that it is not structured as a linear pipeline.
+% Rather, it makes extensive use of POSIX constructs common in the shell, including output redirection, conditionals, and here-documents.
+% It also combines commands from other classes outside \sta and \pur---more prominently \dfs---which are not parallelizable by \sys.
+% 
+% \heading{Results}
+% 
+% \heading{Take-aways}
 
@@
 
-\kk{I think the scoping of this should change to be both
-  micro-benchmarks and comparison with existing work. Especially since
-  we will have some form of comparison with Raftlib and phoenix. I
-  guess this section is trying to establish \sys limits by comparing
-  it with similar (but not identical) work.}
+%> \kk{I think the scoping of this should change to be both
+%>   micro-benchmarks and comparison with existing work. Especially since
+%>   we will have some form of comparison with Raftlib and phoenix. I
+%>   guess this section is trying to establish \sys limits by comparing
+%>   it with similar (but not identical) work.}
 
@@
 
-\heading{Manual Parallelization}
-Alternatively, a user might attempt to manually parallelize individual commands by using the abstractions provided by the POSIX shell.
-The benefits depend on several factors, such as whether a command is I/O- or CPU-bound.
-
-\heading{Streaming Framework}
-In a different attempt to identify how far \sys is from ideal performance, we start from programs already written in a high-performance parallel framework and rewrite as shell programs.
-Raftlib~\cite{} is a C++ library enabling streaming and dataflow parallelism without the fault-tolerance overheads of distributed systems.
-We rewrite three Raftlib benchmarks---\ttt{wordcount}, \ttt{XX}, and \ttt{XX}---as shell scripts.
-% These benchmarks were relatively easy found relatively easy to port and match many of the workloads one would expect to see in the shell as shell pipelines.
+%> \heading{Manual Parallelization}
+%> Alternatively, a user might attempt to manually parallelize individual commands by using the abstractions provided by the POSIX shell.
+%> The benefits depend on several factors, such as whether a command is I/O- or CPU-bound.
+%> 
+%> \heading{Streaming Framework}
+%> In a different attempt to identify how far \sys is from ideal performance, we start from programs already written in a high-performance parallel framework and rewrite as shell programs.
+%> Raftlib~\cite{} is a C++ library enabling streaming and dataflow parallelism without the fault-tolerance overheads of distributed systems.
+%> We rewrite three Raftlib benchmarks---\ttt{wordcount}, \ttt{XX}, and \ttt{XX}---as shell scripts.
+%> % These benchmarks were relatively easy found relatively easy to port and match many of the workloads one would expect to see in the shell as shell pipelines.
 

### 2020-05-23 pldi/draft.tex
Pass on section 4

@@
   edges that do not point to a node in the graph are its outputs.
-\nv{OK This distinction is explained in this par; I would put this earlier}
+%% \nv{OK This distinction is explained in this par; I would put this earlier}
 
@@
 
-\heading{\todo{Streaming} Commands}
-These commands consume their inputs sequentially and one
-element at a time, with the possible exception of static input files,
-and produce one output file. The quintessential example of
-\todo{streaming} commands is \ttt{cat}, that consumes its inputs in
-order, producing their concatenation as output. A more interesting
-example is \ttt{comm -23 f1 f2}, which has \ttt{f2} as its static
-input (it needs to read it completely before consuming anything from
-\ttt{f1} and then consumes \ttt{f1} one element at a time. These
-commands can be represented as functions of type $f : D\kstar \times
-[D\kstar] \to D\kstar$, where the first argument represents the
-concatenation of the inputs that are consumed one at a time and the
-second argument represents the static inputs.
+\heading{Streaming Commands}
+%
+A large subset of the parallelizable \tsta and \tpur classes falls
+into the special category of streaming commands. These commands
+consume their inputs sequentially and one element at a time, with the
+possible exception of static input files, and produce one output
+file. The quintessential example of streaming commands is \ttt{cat},
+that consumes its inputs in order, producing their concatenation as
+output. A more interesting example is \ttt{comm -23 f1 f2}, which has
+\ttt{f2} as its static input (it needs to read it completely before
+consuming anything from \ttt{f1} and then consumes \ttt{f1} one
+element at a time. These commands can be represented as functions of
+type $f : D\kstar \times [D\kstar] \to D\kstar$, where the first
+argument represents the concatenation of the inputs that are consumed
+one at a time and the second argument represents the static inputs.
 %% \km{Again, a graphical
@@
 %%   miss something?}
-We focus on these commands because a large subset of \sta and \pur
-commands fall in this category.
-\nv{Based on the intro of 4 or 4.1 I did not expect streaming.
+%% We focus on these commands because a large subset of \sta and \pur
+%% commands fall in this category.
+%% \nv{Based on the intro of 4 or 4.1 I did not expect streaming.
 I understand edges and nodes, but how do streaming commands fit into the picture?
@@
 characters) are considered stateless. Formally, a
-\todo{streaming} command $f$ is stateless if it commutes with
+streaming command $f$ is stateless if it commutes with
 the operation of concatenation, \ie it is a semigroup homomorphism:
@@
 
-\heading{Parallelization Transformations}
-Based on these equations, we can define a \emph{node parallelization
-  transformation} on a stateless node $v$ that is preceded by a
-\ttt{cat} with $n$ input streams and is followed by a node $v'$ (Fig.~\ref{fig:parallelization-transformation}). The
-transformation replaces $v$ with $n$ new nodes, routing each of the
-$n$ input streams to one of them, and commutes the \ttt{cat} node
-after them to concatenate their outputs and transfer them to
-$v'$. Since each incoming edge represents a stream of data $x_i :
-D^*$, and the only behaviour of a DFG is its output, this
-optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s)
-\cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the
-behavior of the graph.
-
-This transformation can extended straightforwardly to pure commands
-implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
-\cdots x_n, s) \Rightarrow agg(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n,
-s), s)$.
-% \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.
-
 \begin{figure}[t]
@@
 
+\heading{Parallelization Transformations}
+%
+Based on these equations, we can define a \emph{node parallelization
+  transformation} on a stateless node $v$ that is preceded by a
+concatenation, i.e. the command \ttt{cat}, of $n$ input streams and is
+followed by a node $v'$
+(Fig.~\ref{fig:parallelization-transformation}). The transformation
+replaces $v$ with $n$ new nodes, routing each of the $n$ input streams
+to one of them, and commutes the \ttt{cat} node after them to
+concatenate their outputs and transfer them to $v'$. Since each
+incoming edge represents a stream of data $x_i : D^*$, and the only
+behaviour of a DFG is its output, this optimization $v(x_1 \cdot x_2
+\cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$
+can be shown to preserve the behavior of the graph.
+
+This transformation can extended straightforwardly to pure commands
+implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
+\cdots x_n, s) \Rightarrow agg(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n,
+s), s)$.
+% \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.
+
 \heading{Auxiliary Transformations}
-\todo{We} also define a set of auxiliary transformations that are shown in
-\Cref{fig:auxiliary-transformations}.  The first two \nv{Name them?} enable the
-parallelization transformations by inserting \ttt{cat} nodes,\footnote{\nv{This is minor, but for the model maybe name them concatenation nodes or ++? rather than \ttt{cat}}} either
-by using them to concatenate many inputs of one command, or by first
-adding a \ttt{split} node that splits the input in batches. The final
-one inserts a relay that performs the identity transformation on its
-input. Relay nodes can be useful for monitoring and debugging, as well
-as for performance improvements~\sx{optimizer}.
+%
+\sys also performs a set of auxiliary transformations that are
+depicted in \Cref{fig:auxiliary-transformations}. If a node has many
+inputs, \sys concatenates these inputs by inserting a \ttt{cat} node
+to enable the parallelization transformations. In cases where a
+parallelizable node has one input and is not preceded by a
+concatenation, \sys inserts a \ttt{cat} node that is preceded by its
+inverse \ttt{split}, so that the concatenation can be commuted with
+the node. The third depicted transformation inserts a relay node that
+performs the identity transformation. Relay nodes can be useful for
+monitoring and debugging, as well as for performance
+improvements~\sx{optimizer}.
+
+%% The first two
+%% \nv{Name them?}  enable the parallelization transformations by
+%% inserting \ttt{cat} nodes,\footnote{\nv{This is minor, but for the
+%%     model maybe name them concatenation nodes or ++? rather than
+%%     \ttt{cat}}} either by using them to concatenate many inputs of one
+%% command, or by first adding a \ttt{split} node that splits the input
+%% in batches. The final one inserts a relay that performs the identity
+%% transformation on its input.
 

### 2020-05-23 pldi/draft.tex
Update plots
Update fig

@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
 \caption{
-  \textbf{\sys transformation overview.}
-  Given a shell pipeline, \sys leverages command parallelizability to identify
-  high-parallelizability stages, rewrite them, and orchestrate their execution.
-  \todo{reference in txt/decide}.
+  \textbf{\sys overview.}
+  A high-level schematic outline of \sys, and how this maps on the structure of this paper.
 }
@@
   The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
-  \nv{Should graph be \emph{directed}?}
 }
@@
 
-\section{Annotation Language Grammar}
-\label{alg}
-
-\Cref{fig:dsl} shows the complete grammar of the parallelizability
-annotation language.
-
-\begin{figure}
-  \centering
-  \begin{grammar}
-    <option> ::= `-' <string>
-
-    <category> ::= `stateless' | `pure' | ...
-
-    <maybe-int> ::= | <int>
-
-    <arg> ::= `args[' <int> `]'
-
-    <args> ::= <arg>
-    \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-    <input> ::= `stdin' | <args>
-
-    <inputs> ::= <input>
-    \alt <input> `,' <inputs>
-
-    <output> ::= `stdout' | <arg>
-
-    <outputs> ::= <output>
-    \alt <output> `,' <outputs>
-
-    <option-pred> ::= <option>
-    \alt `value' <option> = <string>
-    \alt `not' <option-pred>
-    \alt <option-pred> `or' <option-pred>
-    \alt <option-pred> `and' <option-pred>
-
-    <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
-
-    <predicate> ::= <option-pred> `=>' <assignment>
-
-    <pred-list> ::= `|' <predicate> <pred-list>
-    \alt `|' `otherwise' `=>' <assignment>
-
-    <command> ::= <name> `\{' <pred-list> `\}'
-
-    <command-list> ::= <command>
-    \alt <command> <command-list>
-  \end{grammar}
-  \caption{
-  \textbf{Parallelizability description language.}
-    The DSL captures important information regarding the parallelizability of a command.
-  }
-  \label{fig:dsl}
-\end{figure}
+% \section{Annotation Language Grammar}
+% \label{alg}
+% 
+% \Cref{fig:dsl} shows the complete grammar of the parallelizability
+% annotation language.
+% 
+% \begin{figure}
+%   \centering
+%   \begin{grammar}
+%     <option> ::= `-' <string>
+% 
+%     <category> ::= `stateless' | `pure' | ...
+% 
+%     <maybe-int> ::= | <int>
+% 
+%     <arg> ::= `args[' <int> `]'
+% 
+%     <args> ::= <arg>
+%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
+% 
+%     <input> ::= `stdin' | <args>
+% 
+%     <inputs> ::= <input>
+%     \alt <input> `,' <inputs>
+% 
+%     <output> ::= `stdout' | <arg>
+% 
+%     <outputs> ::= <output>
+%     \alt <output> `,' <outputs>
+% 
+%     <option-pred> ::= <option>
+%     \alt `value' <option> = <string>
+%     \alt `not' <option-pred>
+%     \alt <option-pred> `or' <option-pred>
+%     \alt <option-pred> `and' <option-pred>
+% 
+%     <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
+% 
+%     <predicate> ::= <option-pred> `=>' <assignment>
+% 
+%     <pred-list> ::= `|' <predicate> <pred-list>
+%     \alt `|' `otherwise' `=>' <assignment>
+% 
+%     <command> ::= <name> `\{' <pred-list> `\}'
+% 
+%     <command-list> ::= <command>
+%     \alt <command> <command-list>
+%   \end{grammar}
+%   \caption{
+%   \textbf{Parallelizability description language.}
+%     The DSL captures important information regarding the parallelizability of a command.
+%   }
+%   \label{fig:dsl}
+% \end{figure}
 

### 2020-05-23 pldi/draft.tex
Minor edits and adding results in evaluation

@@
     Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting \todo{AST} size and compilation time for two indicative parallelization configurations.
+    Nodes and compile time report on \sys's resulting DFG size and compilation time for two indicative parallelization configurations.
     % Highlights notes a few key features of each script.
@@
 Sort-sort uses consecutive \pur commands without interleaving them with a \pur that has the effect of significantly condensing the input size (\eg \ttt{uniq}).
-The two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams over the input; the \ttt{optimized} version \todo{has a difference}.
+Finally, the two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams over the input; \ttt{bi-grams-opt} is an optimized version where shift, \ttt{sort}, \ttt{uniq}, have been merged to one command to use a more efficient aggregator.
 
@@
 Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (2--64$\times$).
+\todo{The average speedups of all one-liners per level of parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, 13.47, and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
+
 % we experiment with a $200\times$ level of parallelism to see potential overheads
 
-In Tab.~\ref{tab:eval} contains two benchmarks that are not shown in Fig.~\ref{fig:microbenchmarks}, as they are artificial modifications intended to highlight \sys-internal characteristics and not lead to a speedup.
-  \todo{highlight 2--3 results on things that were not shown in Fig.7}
+Tab.~\ref{tab:eval} contains grep-light, a benchmarks that is not shown in Fig.~\ref{fig:microbenchmarks} since it was designed to show that \sys doesn't slow down IO-intensive scripts.
+On the contrary, \sys achieves a 1.5-2.5$\times$ speedup for grep-light.
+In addition, \Cref{tab:eval} shows that \sys compilation times are negligible ($<5s$) for all one-liners.
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available.
-Sort-sort with $32\times$ has 314 nodes (sys: 54m) and with $64\times$ has 634 nodes (sys: 90m).
+Sort-sort with $32\times$ has 314 nodes \todo{(sys: 54m)} and with $64\times$ has 634 nodes \todo{(sys: 90m)}.
 % 
@@
 By applying only the replacements which to us experts felt more natural, we also give \sys the opportunity to identify and exploit parallelism---a common pattern we witnessed~\sx{macro1}.
-For example, after this change, \todo{13} sees a \todo{XX}$\times$ speedup.
+For example, after this change, 13 sees a 8.1$\times$ speedup.
 
@@
   \centering
-  \includegraphics[width=0.80\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
   \caption{
@@
 \sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be trivially described by annotations.
+\kk{Mention the input.
+1/100 of the whole Wikipedia downloaded in a local directory.}
 
@@
 \sys can operate on the entire program, but as its complexity makes end-to-end measurements difficult we report on the two components components individually.
-With parallelism set to \todo{2/16} the execution time achieved by \sys is \todo{X} for the front-end and \todo{Y} for the back-end, while the original execution time is \todo{X} for the front-end and \todo{Y'} for the back-end.
+With parallelism set to 2/16 the execution time achieved by \sys is \todo{X} for the front-end and 97m/15m for the back-end, while the original execution time is \todo{X} for the front-end and 191m for the back-end.
 
@@
 complete. However, \sys parallelizes the back-end and achieves a
-speedup of \todo{X} with parallelism set to \todo{2/16}. Several
+speedup of 1.97/12.7x with parallelism set to 2/16. Several
 stages of the pipeline are \sta (such as the \ttt{word-stem} script
@@
 contains 34 stages executing in parallel, so the sequential version
-\todo{already benefits from task-based parallelism} \kk{not totally
-  true}.
+already benefits from task-based parallelism.
 

### 2020-05-23 compiler/gather_results.py

@@ -1,4 +1,5 @@
 import sys
+import math
 import os
 import numpy as np
 import matplotlib.pyplot as plt
@@ -43,7 +44,7 @@ pretty_names = {"minimal_grep" : "grep",
                 "topn" : "top-n",
                 "grep" : "grep-light",
                 "bigrams" : "bi-grams",
-                "alt_bigrams" : "optimized bi-grams",
+                "alt_bigrams" : "bi-grams-opt",
                 "spell" : "spell",
                 "shortest_scripts" : "shortest-scripts",
                 "diff" : "diff",
@@ -67,11 +68,11 @@ highlights = {"minimal_grep" : "complex NFA regex",
               "minimal_sort" : "\\tti{sort}ing",
               "wf" : "double \\tti{sort}, \\tti{uniq} reduction",
               "topn" : "double \\tti{sort}, \\tti{uniq} reduction",
-              "grep" : "\\todo{light computation}",
+              "grep" : "IO-intensive, computation-light",
               "bigrams" : "stream shifting and merging",
               "alt_bigrams" : "optimized version of bigrams",
               "spell" : "comparisons (\\tti{comm})",
-              "shortest_scripts" : "\\todo{extensive file-system operation}",
+              "shortest_scripts" : "long \\tsta pipeline ending with \\tpur",
               "diff" : "non-parallelizable \\tti{diff}ing",
               "set-diff" : "two pipelines merging to a \\tti{comm}",
               "double_sort" : "parallelizable \\tpur after \\tpur"}
@@ -234,6 +235,7 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, a
 
     maximum_y = 0
     lines = []
+    best_result = distr_speedup
 
     ## In the bigrams experiment we want to also have the alt_bigrams plot
     if(experiment == "bigrams"):
@@ -259,6 +261,7 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, a
             lines.append(line)
             maximum_y = max(maximum_y, max(auto_split_distr_speedup))
             maximum_y = max(maximum_y, max(split_distr_speedup))
+            best_result = split_distr_speedup
         except ValueError:
             pass
 
[…]

### 2020-05-23 pldi/microbenchmarks-table.tex
Half-way pass over eval

--- 2020-04-12 pldi/microbenchmarks-table.tex
@@
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllll}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllllll}
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(\todo{20, 100}) &Compile Time (\todo{20, 100}) & Highlights \\
+Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 80m9.607s & \todo{\#Commands} & 0.063s\qquad 0.966s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 23m49.044s & \todo{\#Commands} & 0.201s\qquad 2.006s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 72m53.786s & \todo{\#Commands} & 0.170s\qquad 3.341s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m31.907s & \todo{\#Commands} & 0.165s\qquad 3.325s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.035s\qquad 0.356s & $3\times$\tsta \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m1.131s & \todo{\#Commands} & 0.311s\qquad 6.655s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & \todo{UNKNOWN} & 28m26.746s & \todo{\#Commands} & 0.305s\qquad 6.729s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ \todo{TODO} & 10~GB & 26m50.085s & \todo{\#Commands} & 0.312s\qquad 4.569s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~GB & 123m54.946s & \todo{\#Commands} & 0.136s\qquad 2.674s & optimized version of bigrams \\
+grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
+sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
+wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
+grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
+spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 69 & 261 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
+shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
+diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
+bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 109 & 445 & 0.146s & 1.716s & stream shifting and merging \\
+bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
+set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
+sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78 & 318 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-05-23 evaluation/intermediary/max_temp_p123_10_seq.sh
max temp modifs and results

@@ -1,11 +1,12 @@
 #!/bin/bash
 
-seq 2005 2005 |
+seq 2000 2004 |
     sed 's;^;http://ndr.md/data/noaa/;' |
     sed 's;$;/;' |
     xargs -n 1 curl -s |
     grep gz |
     tr -s ' ' |
     cut -d ' ' -f9 |
+sed 's;^\(.*\)\(20[0-9][0-9]\).gz;\2/\1\2\.gz;' |
     sed 's;^;http://ndr.md/data/noaa/2005/;'
 

### 2020-05-23 evaluation/intermediary/p4_10_env.sh

@@ -1,1 +1,1 @@
-IN_DIR=/tmp/max-temp-output/
+IN_DIR=/home/nikos/max-temp-output/

### 2020-05-23 evaluation/scripts/max-temp/p3.sh

@@ -1,3 +1,4 @@
 #!/bin/bash
 
-sed 's;^;http://ndr.md/data/noaa/2005/;'
+sed 's;^\(.*\)\(20[0-9][0-9]\).gz;\2/\1\2\.gz;' |
+sed 's;^;http://ndr.md/data/noaa/;'

### 2020-05-23 compiler/execute_max_temp_dish_evaluation.sh
max temp modifs and results

@@ -8,7 +8,7 @@ p3="${directory}/p3.sh"
 p4="${directory}/p4.sh"
 p5="${directory}/p5.sh"
 
-output_dir="/tmp/max-temp-output/"
+output_dir="$HOME/max-temp-output/"
 p1_out="${output_dir}/p1.out"
 p2_out="${output_dir}/p2.out"
 p3_out="${output_dir}/p3.out"
@@ -17,17 +17,15 @@ p5_out="${output_dir}/p5.out"
 
 intermediary_dir="${eval_dir}/intermediary/"
 results_dir="${eval_dir}/results/"
-
+ 
 ## Make the temporary output dir
 mkdir -p $output_dir
 
 { time ( /bin/bash $p1 > $p1_out ) ; } 2> >(tee "${results_dir}/p1_seq.time" >&2)
 { time ( cat $p1_out | /bin/bash $p2 > $p2_out ) ; } 2> >(tee "${results_dir}/p2_seq.time" >&2)
-
-## TODO: p3 is currently only working for 2005
 { time ( cat $p2_out | /bin/bash $p3 > $p3_out ) ; } 2> >(tee "${results_dir}/p3_seq.time" >&2)
 { time ( cat $p3_out | /bin/bash $p4 > $p4_out ) ; } 2> >(tee "${results_dir}/p4_seq.time" >&2)
-{ time ( cat $p4_out | /bin/bash $p5 > $p5_out ) ; } 2> >(tee "${results_dir}/p5_seq.time" >&2)
+# { time ( cat $p4_out | /bin/bash $p5 > $p5_out ) ; } 2> >(tee "${results_dir}/p5_seq.time" >&2)
 
 echo "Sequential pipeline has been executed successfully."
 
@@ -38,7 +36,7 @@ split -n l/2 -d $p4_out ${p4_out}_2_
 split -n l/10 -d $p4_out ${p4_out}_10_
 
 echo "Intermediate files have been successfully produced."
-
+ 
 ## Setup the _env files of the experiments accordingly
 echo "IN_DIR=${output_dir}" > ${intermediary_dir}/p4_1_env.sh
 echo "IN_DIR=${output_dir}" > ${intermediary_dir}/p4_2_env.sh
@@ -48,10 +46,16 @@ echo "IN_DIR=${output_dir}" > ${intermediary_dir}/p5_2_env.sh
 echo "IN_DIR=${output_dir}" > ${intermediary_dir}/p5_10_env.sh
 
[…]

### 2020-05-23 evaluation/intermediary/max_temp_p123_2_seq.sh

@@ -1,11 +1,12 @@
 #!/bin/bash
 
-seq 2005 2005 |
+seq 2000 2004 |
     sed 's;^;http://ndr.md/data/noaa/;' |
     sed 's;$;/;' |
     xargs -n 1 curl -s |
     grep gz |
     tr -s ' ' |
     cut -d ' ' -f9 |
+    sed 's;^\(.*\)\(20[0-9][0-9]\).gz;\2/\1\2\.gz;' |
     sed 's;^;http://ndr.md/data/noaa/2005/;'
 

### 2020-05-23 evaluation/scripts/max-temp/p1.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-seq 2005 2005 |
+seq 2000 2004 |
     sed 's;^;http://ndr.md/data/noaa/;' |
     sed 's;$;/;' |
     xargs -n 1 curl -s

### 2020-05-23 evaluation/scripts/max-temp/p2.sh

@@ -3,5 +3,5 @@
 grep gz |
     # head -n 1 | # <-- remove this live for full scale
     tr -s ' ' |
-    cut -d ' ' -f9
+    cut -d ' ' -f9 
 

### 2020-05-23 pldi/draft.tex
Update eager fig
NOAA Pash results in paper

@@
 % https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
+\includegraphics[width=0.48\textwidth]{\detokenize{./figs/dish_eager.pdf}}
 \caption{
@@
 \heading{Results}
-The original pipeline executes in \todo{XXs} on \todo{XX}GB of input data.
-Applying \sys with \todo{XX}$\times$ parallelism leads to (XX speedup).
-Naturally, different phases benefit differently from parallelization: \todo{part 1, 2, 3, 4}
+The complete
+%% original
+shell pipeline executes in 44m2s for five years of weather data (which
+amounts to 82GB).  Applying \sys with 2/10$\times$ parallelism leads
+to (1.86/2.44$\times$ speedup).  Naturally, different phases benefit
+differently from parallelization: downloading, extracting, and
+preparing data benefits from 1.77/1.99$\times$ speedup (compared to
+original 33m58s), while computing the maximum temperature benefits
+from 2.30/10.79$\times$ speedup (original 10m4s). \todo{For an
+  explanation of why speedup is more than the amount of parallelism}
+  see Discussion-\sx{ours}.
 
@@
 
-In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to \todo{XX} shell LoC for \sys's \todo{part 4}.
+In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation.
 
@@
 % FIXME OSDI mention tabulation vs. sorting
+
+\kk{Should we mention that we serve the weather data in the local network}
 

### 2020-05-23 pldi/draft.tex
Remove third use case from intro

@@
   (i) a series of benchmarks, ranging from classic one-liners found in the \unix literature to modern data-processing pipelines;
-  (ii) three large and complex use cases, covering temperature analysis, web crawling and indexing, and container orchestration; and
-  (iii) a series of targeted micro-benchmarks inspired by (and comparing with) possible alternatives.
-The results demonstrate significant advantages in \todo{correctness},
-\todo{manual effort}, and performance---with \sys offering an order-of-magnitude performance improvements due to its transformations and optimized runtime.
+(ii) two large and complex use cases, covering temperature analysis, and web
+%% crawling and
+indexing;
+%% , and container orchestration;
+and (iii) a series of targeted micro-benchmarks inspired by (and
+comparing with) possible alternatives.  The results demonstrate
+significant advantages in \todo{correctness}, \todo{manual effort},
+and performance---with \sys offering an order-of-magnitude performance
+improvements due to its transformations and optimized runtime.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
@@
 correspondence with shell scripts, and in \Cref{ir:transformations} we
-describe the graph transformations and \todo{give intuition about their
-correctness}.
+describe the graph transformations that \sys performs.
 

### 2020-05-23 pldi/draft.tex
More work on eval

@@
       (v) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
-      (vi) \todo{Opt. Parallel}, showing the performance of an optimized version of \ttt{bigrams}.
-      \nv{Maybe it makes sense to show log x-axis?}
+      (vi) \todo{Opt. Parallel}, showing the performance of an optimized version of \ttt{bigrams}~\cf{ours}.
     }
@@
 Sort-sort uses consecutive \pur commands without interleaving them with a \pur that has the effect of significantly condensing the input size (\eg \ttt{uniq}).
-Finally, the two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams over the input; \ttt{bi-grams-opt} is an optimized version where shift, \ttt{sort}, \ttt{uniq}, have been merged to one command to use a more efficient aggregator.
+Finally, the two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams, but \ttt{-opt}'s shift, \ttt{sort}, and \ttt{uniq} are merged to one command to use a more efficient aggregator.
 
@@
 Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (2--64$\times$).
-\todo{The average speedups of all one-liners per level of parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, 13.47, and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
+(Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
+Average speedups per level of parallelism are 1.97 (for $2\times$), 3.5 ($4\times$), 5.78 ($8\times$), 8.83 ($16\times$), 10.96 ($32\times$), 13.47 ($64\times$).
+% and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
 
 % we experiment with a $200\times$ level of parallelism to see potential overheads
-
-Tab.~\ref{tab:eval} contains grep-light, a benchmarks that is not shown in Fig.~\ref{fig:microbenchmarks} since it was designed to show that \sys doesn't slow down IO-intensive scripts.
-On the contrary, \sys achieves a 1.5-2.5$\times$ speedup for grep-light.
-In addition, \Cref{tab:eval} shows that \sys compilation times are negligible ($<5s$) for all one-liners.
+Tab.~\ref{tab:eval} shows that \sys's analysis and compilation time is negligible % ($<5s$)
+  for all benchmarks.
+It also contains Grep-light (not shown in Fig.~\ref{fig:microbenchmarks}), designed to show that \sys does not slow down IO-intensive scripts;
+  on the contrary, \sys achieves a 1.5-2.5$\times$ speedup for grep-light. 
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available.
-Sort-sort with $32\times$ has 314 nodes \todo{(sys: 54m)} and with $64\times$ has 634 nodes \todo{(sys: 90m)}.
+Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available---as is the case, \eg  with Sort-sort for $32\times$ and $64\times$ which has 314 and 634 nodes, respectively.
 % 
@@
 Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-By applying only the replacements which to us experts felt more natural, we also give \sys the opportunity to identify and exploit parallelism---a common pattern we witnessed~\sx{macro1}.
+By applying only the replacements which to us experts felt the most natural, we also give \sys the opportunity to identify and exploit parallelism---a common pattern we witnessed~\sx{macro1}.
 For example, after this change, 13 sees a 8.1$\times$ speedup.
@@
 \heading{Results}
-The complete
-%% original
-shell pipeline executes in 44m2s for five years of weather data (which
-amounts to 82GB).  Applying \sys with 2/10$\times$ parallelism leads
-to (1.86/2.44$\times$ speedup).  Naturally, different phases benefit
-differently from parallelization: downloading, extracting, and
-preparing data benefits from 1.77/1.99$\times$ speedup (compared to
-original 33m58s), while computing the maximum temperature benefits
-from 2.30/10.79$\times$ speedup (original 10m4s). \todo{For an
-  explanation of why speedup is more than the amount of parallelism}
-  see Discussion-\sx{ours}.
-
-It is worth reporting Hadoop's (3.2.1, Oracle JDK1.8.0\_255) execution time for the curious reader (see Discussion for fairness), which takes \todo{6m22s} (sequential) and \todo{4m30.554s} ($\times$\todo{XX}-parallel).
-This amounts to \todo{XX}$\times$ speedup versus \todo{XX}$\times$ for \sys's equivalent (part 4).
-% ---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
+The complete shell pipeline executes in 44m2s for five years (82GB) of weather data.
+Applying \sys with 2/10$\times$ parallelism leads to 1.86/2.44$\times$ speedup. 
+Naturally, different phases see different benefits:
+  downloading, extracting, and preparing data sees 1.77/1.99$\times$ speedup (vs. 33m58s), and
+  computing the maximum temperature sees 2.30/10.79$\times$ speedup (vs. 10m4s).
+For a discussion of why speedup is higher than the amount of parallelism, see \S\ref{ours}'s Discussion.
+
+%> It is worth reporting Hadoop's (3.2.1, Oracle JDK1.8.0\_255) execution time for the curious reader (see below for fairness), which takes \todo{6m22s} (sequential) and \todo{4m30.554s} ($\times$\todo{XX}-parallel).
+%> This amounts to \todo{XX}$\times$ speedup versus \todo{XX}$\times$ for \sys's equivalent (part 4).
+%> % ---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
 
@@
 
-In terms of execution fairness, both systems are message-passing rather than shared-memory, and Hadoop's setup is atypical to improve fairness:
-  it operates on non-HDFS file-system and has replication disabled.
-Additionally, the Java program is written as a monolith tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases; and the task at hand is an ideal fit for MapReduce system such as Hadoop.
-Finally, execution time is long enough (\todo{XXh})to amortize Hadoop's startup costs.
-% FIXME OSDI mention tabulation vs. sorting
-
-\kk{Should we mention that we serve the weather data in the local network}
+%> In terms of execution fairness, both systems are message-passing rather than shared-memory, and Hadoop's setup is atypical to improve fairness:
+%>   it operates on non-HDFS file-system and has replication disabled.
+%> Additionally, the Java program is written as a monolith tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases; and the task at hand is an ideal fit for MapReduce system such as Hadoop.
+%> Finally, execution time is long enough (\todo{XXh})to amortize Hadoop's startup costs.
+%> % FIXME OSDI mention tabulation vs. sorting
+
+% \kk{Should we mention that we serve the weather data in the local network}
+% \nv{It's in the setup}
 
 \heading{Take-aways}
-A multi-line parallel program specialized for the task at hand corresponds to a much smaller shell script with comparable performance (on a comparable setup).
+\sys can be applied to \todo{larger} % real/realistic? implies that prior benchmarks were not  
+  programs with modern input sides to offer significant acceleration.
+\todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
 
@@
 
-As a second realistic workload, we use a large pipeline for
-downloading and processing the web, comprised of stages collected from
-various repositories.
+As a second realistic workload, we use a large web-indexing script composed from various \todo{sources}~\cite{}.
 
 \heading{Program}
-Coarsely, the pipeline is comprised of two distinct components % coarsely corresponding to a front- and a back-end
-  communicating via \unix pipes.
+Coarsely, the pipeline amounts to two distinct components communicating via \unix pipes.
 The first downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
  the second extracts the text from HTML and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
-Collectively, they contains a total of 52 stages (first: 18; second: 34) written in multiple languages---\eg
-its \ttt{url-extract}ion utility is written in JavaScript whereas its \ttt{word-stem}ming utility is in Python.
-\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be trivially described by annotations.
-\kk{Mention the input.
-1/100 of the whole Wikipedia downloaded in a local directory.}
+Collectively, they contain a total of 52 stages (first: 18; second: 34) written in multiple programming languages.
 
@@
 % To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
-\sys can operate on the entire program, but as its complexity makes end-to-end measurements difficult we report on the two components components individually.
-With parallelism set to 2/16 the execution time achieved by \sys is \todo{X} for the front-end and 97m/15m for the back-end, while the original execution time is \todo{X} for the front-end and 191m for the back-end.
+As the first component of the pipeline contains a cycle (URLs$\rightarrow$HTML$\rightarrow$URLs), \sys's conservative transformations do not attempt to parallelize it.
+The second component, applied to 1\% of Wikipedia (1.3GB), takes 191 minutes.
+With 2/16$\times$-parallelism, \sys brings it down to 97m/15m ($1.97/12.7\times$) with the majority of the speedup coming from the HTML-to-text conversion.
+% which uses \ttt{pandoc}, a utility written in Haskell.
 
@@
 \heading{Discussion}
-Since the front-end contains a feedback cycle, \sys \todo{doesn't
-  parallelize it}, and in both cases it takes \todo{X seconds} to
-complete. However, \sys parallelizes the back-end and achieves a
-speedup of 1.97/12.7x with parallelism set to 2/16. Several
-stages of the pipeline are \sta (such as the \ttt{word-stem} script
-and the text extraction) and that allows \sys to achieve benefits by
-exposing data parallelism. Note that the back-end part of the script
-contains 34 stages executing in parallel, so the sequential version
-already benefits from task-based parallelism.
+The original script contains 34 pipeline stages thus the sequential version already benefits from task-based parallelism.
+It also uses several utilities not part of the standard POSIX/GNU set---\eg its \ttt{url-extract}ion is written in JavaScript and its \ttt{word-stem}ming is in Python.
+% , and its HTML-to-ASCII is achieved via \ttt{pandoc}, a binary utility written in Haskell.
+\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be trivially described by annotations.
+Several other stages are in \sta allowing \sys to achieve benefits by exposing data parallelism.
 
 \heading{Take-aways}
+\sys operates programs with annotated commands outside the POSIX/GNU command subsets and leads to notable speedups even when the original program features significant task-based parallelism.
+\sys's conservative transformations do not risk breaking correctness, thus do not improve the first component.
 
@@
 
-The previous several subsections show substantial speedups without any developer effort on a variety of smaller and larger shell programs.
-In this section, we are interested in zooming into \sys's limits and how they compare with perceived upper bounds.
-These upper bounds are calculated through a variety of means which, while requiring vastly more developer effort than \sys, they are expected to yield significantly better performance.
+This section attempts to characterize \sys's limits by comparing it with similar alternatives.
+As there are no prior systems that directly comparable to \sys, we draw comparisons with a series of specialized ones that excel within smaller fragments of \sys's proposed domain.
 
@@
 %%   good as it could be.}
-We first compare a single GNU \ttt{sort} optimized by \sys with the
-same \ttt{sort} with the \ttt{-{}-parallel} flag set. Even though the
-\ttt{-{}-parallel} flag is not a general parallelism solution, this
-comparison serves to establish a baseline for \sys's performance. The
-results are shown in \Cref{fig:sort-parallel-comparison}. Note that
-for \ttt{sort -{}-parallel} parallelism is set to double the number
-set of \sys (i.e. the final point is for \ttt{-{}-parallelism=128})
-since \sys also spawns separate processes for sorting and for
-merging. There are a couple important points to note.
-
-First of all, \sys without the \ttt{eager} optimization enabled
-performs comparably with \ttt{sort -{}-parallel}. On the other hand,
-enabling the \ttt{eager} optimization allows \sys to perform much
-better ($\sim 2$ times) than \ttt{sort} with the \ttt{-{}-parallel}
-flag. That is because \sys with the \ttt{eager} optimization enabled
-essentially uses intermediate buffers (on disk) between the merge
-phases.
+We first compare a single GNU \ttt{sort} optimized by \sys ($S_{PaSh}$) versus the same \ttt{sort} with the 
+\begin{wrapfigure}{r}{0.25\textwidth}
+  \begin{center}
+    \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
+  \end{center}
+%   \vspace{-15pt}
+% \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
+\end{wrapfigure}
+\ttt{-{}-parallel} flag set ($S_{GNU}$), shown on the side.
+% The results are shown on the side-figure below.
+While \ttt{-{}-parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
+$S_{GNU}$'s parallelism is set to $2\times$ of $S_{PaSh}$'s (\ie the rightmost point is for \ttt{-{}-parallelism=128}), 
+to account for \sys's additional merge processes.
+% and as such we note that the measurements are harsh for \sys.
+
+A few points are worth noting.
+$S_{PaSh}$ without \ttt{eager} performs comparably to $S_{GNU}$ and \ttt{eager}  outperforms $S_{GNU}$ ($\sim2\times$).
+This is \ttt{eager} adds intermediate buffers %(on disk)
+  between merge phases.
 %% On the other hand, \ttt{sort --parallel} sorts the input trying to
@@
 %% purpose of having a simple solution.
-\begin{wrapfigure}{r}{0.25\textwidth}
-  \begin{center}
-    \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
-  \end{center}
-%   \vspace{-15pt}
-% \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
-\end{wrapfigure}
-Second, based on the comparison with \ttt{sort --parallel} it seems
-that the scaleup limit that \sys reaches with sort is \todo{inherent
-  and cannot be overpassed} \kk{We have to find a way to argue for the
-  previous statement}. Because of that, all pipelines and scripts that
-contain \ttt{sort} in our evaluation (several one-liners and unix50)
-have this upper bound on scalability. \kk{Maybe move this paragraph in
-  the one-liners section.}
-
-Finally, this comparison showcases the benefit that \sys provides to
-command developers. Instead of implementing a custom flag that enables
-parallelism for a command, a simple specification of the command
-parallelizability allows \sys to scale it similarly to a handcrafted
-implementation with almost no effort.
+Second, $S_{GNU}$ indicates that \sys's scalability limit is \todo{inherent and cannot be overpassed}.
+Because of this, all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) have this upper bound on scalability.
+Finally, the comparison also shows \sys's benefits to command developers:
+  rather than implementing a custom flag for parallelism, a simple  parallelizability annotation would allow \sys to scale similarly to or better than a handcrafted implementation with minimal effort.
 
@@
 \heading{GNU Parallel}
-We compare \sys to \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}.
-In the first experiment, we leverage \ttt{parallel}'s \ttt{--jobs} flag and redirect input from a file to 
-Sequential execution takes \todo{XX}s, and \sys's parallel execution takes \todo{XX}s (speedup: \todo{XX}\%).
+We compare \sys to GNU \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}, on a small bio-informatics script.
+% Using its \ttt{--jobs} flag, we  redirect input from a file to 
+Sequential execution takes \todo{XX}s \vs \sys's \todo{XX}s (\todo{XX}$\times$).
 We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command---masking \sys's improvements in other parts of the pipeline.
@@
 
-There are a few possible ways one might attempt to use GNU \ttt{parallel} to parallelize this program.
-Assuming a user knows that the bottleneck comes from the \ttt{XX} stage, they could attempt to use \ttt{parallel} there.
-This leads to a runtime of \todo{XX}s (speedup: \todo{XX}\%).
-
-Alternatively, a user could (incorrectly) sprinkle \ttt{parallel} across the entire program;
-  this is a common strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
-Doing this leads to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
+There are a few possible ways one might attempt to use GNU \ttt{parallel} on this program.
+They could apply \ttt{parallel} to \todo{XX}---assuming they know that it is the bottleneck---bringing execution time down to \todo{XX}s (speedup: \todo{XX}$\times$).
+Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program, a common strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
+This would lead to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
 \sys's conservative program transformations will not attempt parallelizing program sections that contain commands with unclear \todo{parallelizability} properties.
 	% or commands whose inputs depend on earlier outputs.
-GNU \ttt{parallel} is oblivious to command semantics, risking breaking program semantics as in this case.
+% GNU \ttt{parallel} is oblivious to command semantics, risking breaking program semantics as in this case.
 
@@
 \heading{Graph Transformations}
-In the benchmarks, the overhead from \sys's analysis and generation phases remains collectively below $1s$ even for the high-scalability cases.
+\sys's analysis and generation overheads remain collectively under $5s$ even for the high-scalability benchmarks.
 We decided to push \sys's limits by constructing an artificial 1000-stage pipeline---a worst-case workload that is nowhere near the ones observed in practice.
@@
 
-\heading{Take-aways}
-\sys does not improve performance when running the risk of breaking correctness. % This should go in the intro
-% Compared to \sys, \ttt{parallel} can lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
-In highly-parallel benchmarks, \sys can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
+% \heading{Take-aways}
+% \sys does not improve performance when running the risk of breaking correctness. % This should go in the intro
+% % Compared to \sys, \ttt{parallel} can lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
+% In highly-parallel benchmarks, \sys can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
 

### 2020-05-24 pldi/draft.tex
Pass over related work and conclusion

@@
 It first details how \sys's front-end translates a shell script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
-It then describes \sys's runtime\todo{/backend} and the technical challenges addressed to achieve performance benefits~\sx{optimizer}.
+It then describes \sys's back-end, responsible for generating the parallel script, and the technical challenges addressed to achieve performance benefits~\sx{optimizer}.
 
@@
 
-After translating the parallelizable regions of the input script to DFGs and applying optimization passes using the transformations described earlier~\sx{ir:transformations},
-  \sys translated all DFGs back to a shell script.
+After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
+  \sys translates all DFGs back into a shell script.
 Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
-This section describes several technical challenges that \sys addresses to efficiently perform this transformation, as well as the implementations of the custom \sys nodes that are added by the auxiliary transformations.
+This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.
 %% it instantiates the \todo{custom} split and relay nodes as highly
@@
 There are a few possible ways one might attempt to use GNU \ttt{parallel} on this program.
-They could apply \ttt{parallel} to \todo{XX}---assuming they know that it is the bottleneck---bringing execution time down to \todo{XX}s (speedup: \todo{XX}$\times$).
-Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program, a common strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
+They could use it on the \todo{XX} stage, assuming they know it is a bottleneck, bringing execution down to \todo{XX}s (\todo{XX}$\times$ speedup).
+Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
 This would lead to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
@@
 
-At a high level, techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
-A significant concern with approaches that require user input is correctness:
-   users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the semantics of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
+Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
+Following the distinction between \emph{users} and \emph{developers} established earlier~\sx{intro}, we note correctness trade-offs associated with approaches that require user input for the discussion below:
+   users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the correctness of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
 
 \heading{Parallel Shell Scripting}
-Several utilities expose commands for specifying parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}.
-Different from \sys, their effectiveness is predicated upon explicit and careful invocation and is limited to embarrassingly parallel (and short) programs.
-Often, these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs}, that a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs};
-   it also introduces (and depends on) other programs with complex semantics, such as the ability transfer files between computers, separate text files, and parse CSV.
-\sys embraces the \unix philosophy and simply attempts to rewrite pipelines more efficiently with minimal user effort---if any.
+Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are predicated upon explicit and careful user invocation and are limited to embarrassingly parallel (and short) programs.
+Often these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
+For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs} (which a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs})
+  and introduces (and depends on) other programs with complex semantics, such as ones for SQL querying and CSV parsing.
+\sys embraces the \unix philosophy, simply attempting to rewrite scripts more efficiently with minimal user effort.
 
@@
 
-Recently, the Smoosh authors have argued that certain shell features are useful for making concurrency explicit~\cite{smoosh:18}.
-Their argument is somewhat antithetical to \sys, which argues that users should experience mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
+Recently, Smoosh~\cite{smoosh:18} argued for making concurrency explicit via shell constructs.
+The argument is somewhat antithetical to \sys's, which argues for mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
 
@@
 Instruction-level parallelization has a long history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops rather than the boundaries of programs that are part of a script), operating in a single target environment, and typically do not exploit high-level annotations.
+These systems operate at a lower level than \sys (\eg that of instructions or loops rather than the boundaries of programs that are part of a script), within a single-language or single-target environments, and require source modifications.
 
 More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require users to get involved in discovering and exposing parallelism.
+These tools simplify the expression of parallelism, but still require significant user involvement in discovering and exposing parallelism.
 % Moreover, the insights behind these attempts are significantly different from \sys's, as they extract parallelism statically during compilation instead of dynamically during runtime.
@@
 \heading{Correct Parallelization of Dataflow Graphs}
-The DFG is a prevalent model in several areas of data processing (including batch-~\cite{mapreduce:08, spark:12} and stream-processing ~\cite{murray2013naiad, carbone2015flink} systems).
-Despite its popularity, most systems perform optimizations that do not preserve its semantics, introducing subtle erroneous behaviours.
+The DFG is a prevalent model in several areas of data processing (including batch-~\cite{mapreduce:08, spark:12} and stream-processing ~\cite{murray2013naiad, carbone2015flink}).
+Despite its popularity, most systems perform optimizations that do not preserve semantics, introducing subtle erroneous behaviours.
 Recent work~\cite{HSSGG2014, SHGW2015, MSAIT2019} attempts to address this issue by performing optimizations only in cases where correctness is preserved.
 \sys draws inspiration from these efforts, as it attempts transformations that maintain the program's correctness with respect to the sequential execution.
-
-\sys's proposed DFG model is diverges from prior works, as nodes it captures order.
-This is due to the intricacies of the \unix model---its streams, its argument processing, and concatenation operators all maintain ordering.
+Its  DFG model, however, is different and captures ordering constraints.
+This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.
 % As a result, previous work is not directly applicable.
@@
 \heading{Parallel Userspace Environments}
-By focusing on simplifying the development of distributed programs, a plethora of environments assist in the construction of parallel software.
-Such systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} or languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide many of the challenges of dealing with concurrency.
-They do so as long as a program leverages the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
-A notable example is Plan9's \ttt{rc} shell that provided an interface for the construction of general programs.
-Unfortunately these environments are not backward-compatible with the \unix shell, and often focus primarily on hiding the existence of a network---providing the illusion of a single-system image---rather than automatically accelerating scripts through parallel processing.
+By focusing on simplifying the development of distributed programs, a plethora of environments inadvertently assist in the construction of parallel software.
+Such systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} or languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide many of the challenges of dealing with concurrency as long as developers leverage the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
+% A notable example is Plan9's \ttt{rc} shell that provide an interface for the construction of general programs.
+Even shell-oriented efforts such as Plan9's \ttt{rc} are not backward-compatible with the \unix shell, and often focus primarily on hiding the existence of a network rather than automating parallel processing.
 % simplify many of the problems of distribution and
 
-\heading{Cluster Computing}
-Several frameworks such as Phoenix~\cite{} and Raftlib~\cite{} offer full automation as long as their use leverages special primitives---\eg map-reduce-style primitives for Phoenix.
+\heading{Parallel Frameworks}
+Several frameworks~\cite{streamit:02, brook:04, phoenix:11, raftlib:17} offer fully automated parallelism as long as special primitives are used---\eg map-reduce-style primitives for Phoenix~\cite{phoenix:11}.
 These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
-By targeting specific classes of computation (\emph{viz.} \sys's parallelizability), they can significantly optimize for their target domains.
-
+By targeting specific classes of computation (\emph{viz.} \sys's parallelizability), they are significantly optimized for their target domains.
 % Lack of generality, no formal guaratees, their setup and is too tedious 
-\sys chooses a more general approach: it does not require users to set up a new framework for every new class of computation they need to execute, nor to rewrite different parts of their computation using the specific abstractions provided by each framework they use.
-It also takes a more principled approach: by developing the semantic model outlined earlier, it can focus on transformations that are correct with respect to the sequential program.
+\sys chooses a more general approach:
+   it does not require setting up a new framework for each new class of computation used, nor rewriting different parts of the computation using a different set abstractions provided by each framework.
+It also takes a more principled approach:
+  by developing the semantic model outlined earlier, it can focus on transformations that are correct with respect to the sequential program.
 
@@
 
-\section{Discussion \& Conclusion}
+\section{Conclusion}
 \label{discussion}
@@
   applies a series of transformations according to a DFG model,
-  and orchestrates the execution of the parallel program.
+  and orchestrates the execution of the resulting parallel program.
+%
 % Its runtime component provides orchestration and planning support during the execution of the program.
-
-\sys's current limitations revolve around shell expansion.
-Currently, \sys does not attempt to parallelize program fragments for which variables have not been expanded.
-Without substantial effort, \sys could be modified to at least suggest expansion of these fragments so that developers provide concrete values.
-More significant effort would be required however to fully exploit latent parallelism in fragments that have not been expanding.
-This effort would require tying \sys with the shell interpretation rather than having it as a compilation pass and is left for future work.
-
-Despite limitation related to expansions, \sys can lead to significant benefits for shell users.
-Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all minimal or zero developer effort.
+% \sys's current limitations revolve around shell expansion.
+% Currently, \sys does not attempt to parallelize program fragments for which variables have not been expanded.
+% Without substantial effort, \sys could be modified to at least suggest expansion of these fragments so that developers provide concrete values.
+% More significant effort would be required however to fully exploit latent parallelism in fragments that have not been expanding.
+% This effort would require tying \sys with the shell interpretation rather than having it as a compilation pass and is left for future work.
+%
+% Despite limitation related to expansions,
+  \sys can lead to significant benefits for shell users.
+Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all with minimal or zero user effort.
 

### 2020-05-24 pldi/draft.tex
Minor fixes

@@
 We evaluate \sys using
-  (i) a series of benchmarks, ranging from classic one-liners found in the \unix literature to modern data-processing pipelines;
-(ii) two large and complex use cases, covering temperature analysis, and web
-%% crawling and
-indexing;
+  (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing pipelines;
+  (ii) two large and complex use cases for temperature analysis and web indexing;
 %% , and container orchestration;
-and (iii) a series of targeted micro-benchmarks inspired by (and
+and 
+  (iii) a series of targeted micro-benchmarks inspired by (and
 comparing with) possible alternatives.  The results demonstrate
@@
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Naturally, these improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability class under which a command falls. 
-%\kk{Do we need to mention this last sentence here? Without context it is pretty difficult for anyone to get something out of it.}
-
-%                                           a tool that automatically exploits available data-parallelism in a given shell script.
+Improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability classes in which its commands fall.
 
@@
   \item
-  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime phases.
+  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components.
 % \kk{Rewrite that when we know for sure what the section contains.}
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example.pdf}}
-\caption{
-  \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S.
-for the years 1920--2020 and extracts the maximum for every year.
-\kk{Make this to be for 10 years only so that the point in the second challenge holds}
-}
-\vspace{-15pt}
-\label{fig:example}
-\end{figure}
-
 % In the ongoing debate on climate change,
@@
 The \unix kernel facilitates scheduling, communication, and synchronization behind the scenes.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example.pdf}}
+\caption{
+  \textbf{Calculating maximum temperatures per year.}
+  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year.
+}
+\vspace{-15pt}
+\label{fig:example}
+\end{figure}
 
@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new POSIX shell script.
-The new script may incorporate parallelism, and is handed off to the user's original shell interpreter---\eg \ttt{bash} or \ttt{zsh}---for execution.
-To expose and exploit data-parallelism in a given shell script,
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new script.
+The new script may incorporate parallelism, and is handed off to the user's original shell interpreter % ---\eg \ttt{bash} or \ttt{zsh}---
+  for execution.
+To expose and exploit data-parallelism, % in a given shell script,
 %% parallel execution,
@@
 
-\heading{Commands: Parallelizability Classes}
-% \sys needs to understand the parallelization characteristics of individual commands used to compose larger programs---a two-fold challenge.
-To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but
-%% clear set of well-understood
-\todo{well-defined} set of
-parallelizability classes, which are \todo{useful-enough} to guide the parallelization of simple shell programs.
-%% classes that are useful on their own 
-%%   these classes are already useful-enough to allow the composition of useful shell programs operating in practice.
-To incorporate
-%% To understand the characteristics of
-commands that fall outside of this standard set and allow them to leverage \sys's transformations,
-  we define an annotation language for describing a command's % key properties with respect to its 
-  parallelizability characteristics.
-Such light annotations are expressed once per command (not once per script) and are aimed towards command developers (rather than users), who can quickly and easily capture the properties of the commands they develop.
+\heading{Commands}
+To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}, useful-enough for parallelizing practical programs.
+To allow commands that fall outside this standard set to leverage its transformations, \sys defines a light \emph{annotation language}.
+Annotations focus on describing a command's parallelizability characteristics rather than its full observable behavior, are expressed once per command rather than once per script, and are aimed towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
 % We haven't introduced the notion of flags
@@
 
-\heading{Scripts: Dataflow Graphs}
-%
-To address the synchronization restrictions in shell scripts, \sys
-first analyzes a script to identify \emph{parallelizable regions} that
-can be safely executed in the context of parallelism. \sys uses the
-structure of shell scripts to identify parallelization candidates:
-constructs such as \ttt{\&} and \ttt{|} expose task parallelism, while
-\ttt{\&\&} and \ttt{||} enforce synchronization.
+\heading{Scripts}
+To address the synchronization restrictions in shell scripts, \sys first analyzes a script to identify \emph{parallelizable regions} \todo{that can be safely executed in parallel}.
+\sys uses the structure of shell scripts to identify parallelization candidates:
+  some constructs \eg \ttt{\&} and \ttt{|} expose task parallelism; others \eg \ttt{\&\&} and \ttt{||} enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
@@
 %%  (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
-Since the shell representation only exposes coarse grained parallelization opportunities, \sys then converts each parallelizable region to a dataflow graph.
+% Since the shell representation only exposes coarse grained parallelization opportunities, 
+\sys then converts each parallelizable region to a \emph{dataflow graph} (DGF), 
 %% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
@@
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-This is a flexible representation that enables a series of local transformations that expose data-parallelism, converting the graph into its parallel equivalent.
+  a flexible representation that enables a series of local transformations that expose data parallelism, converting the graph into its parallel equivalent.
 %% , in which DFGs indicate the parallelizable regions.
-Further transformations compile the DFG back to a parallel script that uses POSIX shell constructs to guide parallelism explicitly---while aiming at preserving the semantics of the sequential program.
-
-\heading{Implementation: Efficient Primitives}
+Further transformations compile the DFG back to a parallel script that uses POSIX shell constructs to guide parallelism explicitly, while aiming at preserving the semantics of the sequential program.
+
+\heading{Implementation}
 \sys addresses several practical challenges through a set of constructs it provides---\eg a set of
 %% linkable
-independent components that can be used to compose individual commands.
+composable components used to \todo{wrap} individual stages.
 %% for stream splitting and merging.

### 2020-05-24 pldi/draft.tex
Mini-pass evaluation

@@
 These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
+We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
 
@@
 
-The goal of this subsection is to evaluate \sys on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
+The goal of this subsection is to evaluate \sys (and its different configurations) on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
 
@@
     \caption{
-      \textbf{Speedup achieved by \sys as a function of the level of parallelism (1--100).} 
-      Four different configurations per benchmark:
+      \textbf{Speedup achieved by \sys as a function of the level of parallelism (2--64).} 
+      Different configurations per one-liner:
       (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
@@
       (v) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
-      (vi) \todo{Opt. Parallel}, showing the performance of an optimized version of \ttt{bigrams}~\cf{ours}.
+      (vi) \todo{Opt.
+Parallel}, showing the performance of an optimized version of \ttt{bigrams}~\cf{ours}.
+Only the relevant configurations are shown for each one-liner (e.g.
+grep and sort don't benefit from split, and spell and bi-grams don't get any benefit without split).
     }
@@
 \sys achieves near-linear speedup on low-parallelism configurations.
-This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{fold}s and \ttt{relay}s---evident in all plots that contain \pur stages.
+This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{aggregations}s and \ttt{relay}s---evident in all plots that contain \pur stages.
 For example, Sort in $8\times$ spawns 37 nodes:
@@
 %   minute).
+An additional conclusion is that \sys's modular eager-relay significantly improves performance over the other alternatives (blocking eager or no eager, \S\ref{optimizer}).
 
@@
 Fig~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
-The average speedup is $5.49$ ($6.07\times$), and the weighted average (with the absolute times as weights) is $5.75\times$.
+The average speedup is $5.49$, \todo{the median $6.07\times$}, and the weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},

### 2020-05-24 pldi/draft.tex
Pass over section 5

@@
 \usepackage{caption}
+\usepackage{picins}
 
@@
 
+%% Very nice take-aways!
+
 \begin{figure}[t]
@@
 
-In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation.
+In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation. \kk{Is it worth leaving this sentence in here?}
+%% kk: I think now that we removed Hadoop, it is not so important. In
+%% the end of the day, we don't want to convince people that they
+%% should use shell for all their needs.
 
@@
 % hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
-
-% Dish p3 + p4 me platos pipeline 2 => real 4m30.554s (perilambanei download twn gz kai to processing)
-% Dish p3 + p4 me platos pipeline 10 => real 1m7.183s
-
-% Dish p3 me platos pipeline 2 => real	3m47.470s (perilambanei download twn gz kai save sto file system)
-% Dish p3 me platos pipeline 10 => real	1m38.945s
 
@@
 % unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
-  a simpler (but longer) pipeline that leverages on \unix built-ins % might be outperformed by a specialized programs on a single node, but
+  a simpler (but longer) pipeline that leverages \unix built-ins % might be outperformed by a specialized programs on a single node, but
   turns out to be trivially parallelizable.
@@
 \heading{Take-aways}
-\sys can be applied to \todo{larger} % real/realistic? implies that prior benchmarks were not  
-  programs with modern input sides to offer significant acceleration.
-\todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
+\sys can be applied to programs of significant size and input
+%% \todo{larger} % real/realistic? implies that prior benchmarks were not  
+%% programs
+%% with modern input sides
+to offer significant acceleration.
+A broader %% and more-ambitious
+take-away is that \sys is able to exploit the until now unharnessed parallelization opportunities of inter-program scripts that are used widely to download, extract, and clean data.
+%% \todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
+
+%% \kk{I feel that in order to make any ``bigger'' conclusion we need to at least mention some Hadoop results. Otherwise it feels that any argument doesn't stand. A reviewer could say that }
 
@@
 Coarsely, the pipeline amounts to two distinct components communicating via \unix pipes.
-The first downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
- the second extracts the text from HTML and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
+The \todo{first} downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
+ the second extracts the text from HTML and applies natural-language processing analyses---\eg trigrams, character conversion, term frequencies---to index it.
 Collectively, they contain a total of 52 stages (first: 18; second: 34) written in multiple programming languages.
@@
 % To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
-As the first component of the pipeline contains a cycle (URLs$\rightarrow$HTML$\rightarrow$URLs), \sys's conservative transformations do not attempt to parallelize it.
+As the \todo{first} component of the pipeline contains a cycle (URLs$\rightarrow$HTML$\rightarrow$URLs), \sys's conservative transformations do not attempt to parallelize it.
 The second component, applied to 1\% of Wikipedia (1.3GB), takes 191 minutes.
-With 2/16$\times$-parallelism, \sys brings it down to 97m/15m ($1.97/12.7\times$) with the majority of the speedup coming from the HTML-to-text conversion.
+With 2/16$\times$-parallelism, \sys brings it down to 97m/15m ($1.97/12.7\times$) with the majority of the speedup coming from parallelizing the HTML-to-text conversion.
 % which uses \ttt{pandoc}, a utility written in Haskell.
@@
 \heading{Take-aways}
-\sys operates programs with annotated commands outside the POSIX/GNU command subsets and leads to notable speedups even when the original program features significant task-based parallelism.
-\sys's conservative transformations do not risk breaking correctness, thus do not improve the first component.
+\sys operates on programs with annotated commands outside the POSIX/GNU command subsets and leads to notable speedups even when the original program features significant task-based parallelism.
+\sys's conservative transformations do not risk breaking correctness, \todo{thus do not improve the first component.}
 
@@
 This section attempts to characterize \sys's limits by comparing it with similar alternatives.
-As there are no prior systems that directly comparable to \sys, we draw comparisons with a series of specialized ones that excel within smaller fragments of \sys's proposed domain.
-
+As there are no prior systems directly comparable to \sys, we draw comparisons with a series of specialized ones that excel within smaller fragments of \sys's proposed domain.
+
+\pichskip{15pt}% Horizontal gap between picture and text
+\parpic[r][t]{%
+  \begin{minipage}{40mm}
+    \includegraphics[width=\linewidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}%
+    %% \captionof{figure}{This is a tiger.}
+  \end{minipage}
+}
 \heading{Parallel Sort}
@@
 We first compare a single GNU \ttt{sort} optimized by \sys ($S_{PaSh}$) versus the same \ttt{sort} with the 
-\begin{wrapfigure}{r}{0.25\textwidth}
-  \begin{center}
-    \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
-  \end{center}
-%   \vspace{-15pt}
-% \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
-\end{wrapfigure}
+%% \begin{wrapfigure}[10]{r}{0.25\textwidth}
+%%   \centering
+%%   %% \begin{center}
+%%     \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
+%%   %% \end{center}
+%% %   \vspace{-15pt}
+%% % \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
+%% \end{wrapfigure}
 \ttt{-{}-parallel} flag set ($S_{GNU}$), shown on the side.
@@
 $S_{PaSh}$ without \ttt{eager} performs comparably to $S_{GNU}$ and \ttt{eager}  outperforms $S_{GNU}$ ($\sim2\times$).
-This is \ttt{eager} adds intermediate buffers %(on disk)
+This is because \ttt{eager} adds intermediate buffers %(on disk)
   between merge phases.
@@
 Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
+\kk{I propose removing this subsection. 5 minutes is too much time for compilation and this happens atm because of a pathological scenario. The compilation should never take more than a second.}
 
@@
 
-Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
-Following the distinction between \emph{users} and \emph{developers} established earlier~\sx{intro}, we note correctness trade-offs associated with approaches that require user input for the discussion below:
-   users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the correctness of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
+Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they requiresignificantly more \emph{user} effort (see \sx{intro} for distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
+%% Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
+%% \kk{It is not very clear what the following sentence tries to achieve.}
+%% Following the distinction between \emph{users} and \emph{developers} established earlier~\sx{intro}, we note correctness trade-offs associated with approaches that require user input for the discussion below:
+%%    users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the correctness of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
 
 \heading{Parallel Shell Scripting}
-Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are predicated upon explicit and careful user invocation and are limited to embarrassingly parallel (and short) programs.
-Often these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs} (which a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs})
+Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are predicated upon explicit and careful user invocation: users have to navigate through a vast array of different configurations, flags, and modes of invocation to achieve parallelization without jeopardizing correctness.
+%% I like the part about the unix philosophy, but I want to really make it a side claim.
+The main claim, and the reason why we don't compare against all of these solutions, is that they require user effort and they cannot just be given a script to parallelize.
+%%
+%% and are limited to embarrassingly parallel (and short) programs.
+%% Often these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
+For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs},
+%% (which a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs})
+%% kk: I think that mentioning that they can be written using UNIX commands doesn't make the point stronger.
 and introduces (and depends on) other programs with complex semantics, such as ones for SQL querying and CSV parsing.
-\sys embraces the \unix philosophy, simply attempting to rewrite scripts more efficiently with minimal user effort.
+In contrast, \sys manages to parallelize large scripts correctly with minimal user effort.
+%% embraces the \unix philosophy, simply attempting to rewrite scripts more efficiently with minimal user effort.
 
@@
    it does not require setting up a new framework for each new class of computation used, nor rewriting different parts of the computation using a different set abstractions provided by each framework.
-It also takes a more principled approach:
-  by developing the semantic model outlined earlier, it can focus on transformations that are correct with respect to the sequential program.
+%% It also takes a more principled approach: by developing the
+%%   semantic model outlined earlier, it can focus on transformations
+%%   that are correct with respect to the sequential program.
+%% \kk{the other approaches are also principled and possibly event more correct.}
 

### 2020-05-24 pldi/draft.tex
Fixes

--- 2020-05-24 pldi/draft.tex
@@
 While these features aid developer economy through powerful program
-composition, they complicate shell script parallelization; which even
+composition, they complicate shell script parallelization, which even
 for simple scripts such as the one in Fig.~\ref{fig:example} faces
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
+\caption{
+  \textbf{\sys overview.}
+  A high-level schematic outline of \sys. %  and how this maps on the structure of this paper.
+}
+\vspace{-15pt}
+\label{fig:overview}
+\end{figure}
+
 \heading{Commands} % : Unbounded \& Mysterious}
-The first challenge is the nature and variety of shell commands.
-In contrast to restricted programming frameworks that enable
-%% amenable to
-parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, shell scripts contain black-box commands that are written in a variety of programming languages and models and can have arbitrary behaviors.
-Therefore, each of them requires special treatment in order to be parallelized.
-For example, in Fig.~\ref{fig:example}
-%% each command requires special treatment as to not introduce non-determinism or incorrectness:
-\ttt{grep} can be parallelized simply by partitioning the lines of its input, while \ttt{sort} would require sorting each input partition and then merging the results.
+In contrast to restricted programming frameworks that enable parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, the \unix shell provides a unprecedented number and variety of composable commands.
+% Due to their varying nature, 
+% As each command may exercise arbitrary behaviors,
+To be parallelized, each command may require special analysis and treatment---in Fig.~\ref{fig:example}, \ttt{grep}'s results can be combined by simple concatenation, while \ttt{sort}'s results would require careful merging.
+Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
 %% \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
 %% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
-Worse even, the same command may be used with different flags---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags---that could potentially alter its behaviour.
-% , making correct treatment even more difficult.
-% nv: we already said impossible so more difficult doesn't add much!
-
-\kk{I tried to rephrase the above a little, but I can't get it just right.
+Manual analysis is also challenging, due to the sheer number of commands and their many flags which affect their behaviour---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+ % , making correct treatment even more difficult. % nv: we already said impossible so more difficult doesn't add much!
+
+% \kk{I tried to rephrase the above a little, but I can't get it just right.
 There are two parts to this argument.
@@
 \heading{Implementation} %
-%
-A final challenge has to do with practical issues, as even manual
-parallelization of a given shell script is far from
-trivial. Parallelizing a script without reimplementing all of its
-commands, requires primitives to compose existing commands to achieve
-parallelization. Designing correct and efficient compositions is
-particularly challenging in the context of \unix, and requires
-addressing issues related to scheduling, synchronization, and deadlock
-prevention. Moreover, specific commands (such as \ttt{sort} and
-\ttt{uniq}) require specialized data aggregators to be correctly
-parallelized.
+Independently of the semantics of individual commands and the language of the shell, the \unix environment has its set of quirks.
+Thus, any manual attempt to orchestrate paralllel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
+Forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
+% A final challenge has to do with practical issues, as even manual parallelization of a given shell script is far from trivial.
+% Parallelizing a script without reimplementing all of its commands, requires primitives to compose existing commands to achieve parallelization.
+%  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
+Moreover, to be correctly parallelized, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators.
 %% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
@@
 %%   and any attempt at parallelization will have to take care of these challenges---related to task parallelism, deadlock prevention, and runtime performance.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-\caption{
-  \textbf{\sys overview.}
-  A high-level schematic outline of \sys, and how this maps on the structure of this paper.
-}
-\vspace{-15pt}
-\label{fig:overview}
-\end{figure}
 
@@
 \heading{Scripts}
-To address the synchronization restrictions in shell scripts, \sys first analyzes a script to identify \emph{parallelizable regions} \todo{that can be safely executed in parallel}.
+To address the synchronization restrictions in scripts, \sys first analyzes a script to identify \emph{parallelizable regions} that contain commands that are candidates for parallelization.
 \sys uses the structure of shell scripts to identify parallelization candidates:
@@
 %% , in which DFGs indicate the parallelizable regions.
-Further transformations compile the DFG back to a parallel script that uses POSIX shell constructs to guide parallelism explicitly, while aiming at preserving the semantics of the sequential program.
+Further transformations compile the DFG back to a parallel script, which uses POSIX shell constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program.
 
@@
 %% linkable
-composable components used to \todo{wrap} individual stages.
+modular components used to compose individual commands.
 %% for stream splitting and merging.

### 2020-05-24 pldi/draft.tex
Pass over bg

@@
 %% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
-Manual analysis is also challenging, due to the sheer number of commands and their many flags which affect their behaviour---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+Manual analysis is also challenging, due to the sheer number of commands, and the many flags that affect their behaviour---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
  % , making correct treatment even more difficult. % nv: we already said impossible so more difficult doesn't add much!
@@
 \heading{Scripts} % Restrictive \& Coarse-grained}
-Another challenge is due to the language of the shell, which
+Another challenge is the programming language of the POSIX shell. % of the shell, which is both restrictive and coarse-grained.
 % has been developed through a long history of additions.
-is both restrictive and coarse-grained.
-First, it contains constructs that enforce sequential execution; in the example in \cref{fig:example}, the sequential composition operator \ttt{;} enforces that the assignment to \ttt{base} is completed before the rest of the script starts executing.
-Moreover, the structure of shell scripts only exposes very coarsed-grained forms of parallelism (usually in the form of task parallelism using the \ttt{\&} and \ttt{|} constructs).
+First, it contains constructs that enforce sequential execution;
+   Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before the rest starts executing.
+Moreover, the language semantics only exposes limited task-based parallelism in the form of \ttt{\&}, \ttt{for}, and \ttt{|} constructs.
 %%   contains constructs with different parallelizability semantics. % that enforce sequential execution.
@@
 %% Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
-For example, even if Fig.~\ref{fig:example}'s script was aiming on only six years of data, the first \ttt{curl} would still output hundreds of lines for each year;
-  naive, coarse-grained parallelization will miss such opportunities.
-Any attempt for automating parallelization needs to be aware of the shell semantics and induce as much parallelism as it can get but without breaking semantics.
+For example, even though Fig.~\ref{fig:example}'s \ttt{for} focuses only on five years of data, \ttt{curl} still outputs thousands of lines per year;
+  naive parallelization of each loop iteration will miss such opportunities.
+Any attempt to automate parallelization should be aware of the shell semantics and induce as much parallelism as it can without breaking semantics.
 
@@
 Independently of the semantics of individual commands and the language of the shell, the \unix environment has its set of quirks.
-Thus, any manual attempt to orchestrate paralllel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
-Forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
+Thus, any attempt to orchestrate paralllel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
+For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
 % A final challenge has to do with practical issues, as even manual parallelization of a given shell script is far from trivial.
@@
 %  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
-Moreover, to be correctly parallelized, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators.
+Moreover, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators, to be correctly parallelized.
 %% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
@@
 \heading{Commands}
-To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}, useful-enough for parallelizing practical programs.
-To allow commands that fall outside this standard set to leverage its transformations, \sys defines a light \emph{annotation language}.
-Annotations focus on describing a command's parallelizability characteristics rather than its full observable behavior, are expressed once per command rather than once per script, and are aimed towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
+To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}.
+Rather than describing a command's full observable behavior, these classes focus on information that is important for data parallelism.
+To allow commands that fall outside of the standard set to leverage its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class.
+Annotations are expressed once per command rather than once per script and are aimed towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
 % We haven't introduced the notion of flags
@@
 \heading{Scripts}
-To address the synchronization restrictions in scripts, \sys first analyzes a script to identify \emph{parallelizable regions} that contain commands that are candidates for parallelization.
-\sys uses the structure of shell scripts to identify parallelization candidates:
-  some constructs \eg \ttt{\&} and \ttt{|} expose task parallelism; others \eg \ttt{\&\&} and \ttt{||} enforce synchronization.
+To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} that contain commands that are candidates for parallelization.
+This analysis is guided by the script structure: % to identify parallelization candidates:
+  some constructs (\eg \ttt{\&}, \ttt{|}) expose task parallelism; others (\eg \ttt{\&\&}, \ttt{||}) enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
@@
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-  a flexible representation that enables a series of local transformations that expose data parallelism, converting the graph into its parallel equivalent.
+  a flexible representation that enables a series of local transformations to expose data parallelism, converting the graph into its parallel equivalent.
 %% , in which DFGs indicate the parallelizable regions.
-Further transformations compile the DFG back to a parallel script, which uses POSIX shell constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program.
+Further transformations compile the DFG back to a shell script that uses POSIX constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program.
 
@@
 take-away is that \sys is able to exploit the until now unharnessed parallelization opportunities of inter-program scripts that are used widely to download, extract, and clean data.
+
+untapped potential 
+that is 
+exploitble
+
+does not fall within purely compute-intensive targets that are the focus of conventional parallelization techniques.
+
+% that falls outside the usual focus
+
 %% \todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
@@
 \heading{Parallel Shell Scripting}
-Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are predicated upon explicit and careful user invocation: users have to navigate through a vast array of different configurations, flags, and modes of invocation to achieve parallelization without jeopardizing correctness.
+Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are limited to embarrassingly parallel (and short) programs and are predicated upon explicit and careful user invocation:
+  users have to navigate through a vast array of different configurations, flags, and modes of invocation to achieve parallelization without jeopardizing correctness.
 %% I like the part about the unix philosophy, but I want to really make it a side claim.

### 2020-05-24 pldi/draft.tex
Evaluation highlights

@@
 \fixme{
-Scripts see a significant speedup, often up to $10\times$ over sequential execution while returning identical results (on very large datasets).
-\sys's transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
-As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
-Finally, \sys preserves the productivity of the shell:
-  the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
+Scripts from various sources and workloads see a significant speedup---almost linear in embarrassingly parallel cases---over sequential execution while returning identical results (on non-trivial datasets).
+The effort to reap this benefits is minimal; only sometimes requiring minimal annotations for \todo{custom} commands.
+\sys highly-optimized primitives (\sx{optimizer}) offer performance benefits even against manually implemented alternatives (such as GNU sort, and GNU parallel).
+\sys's compilation process takes $<$5s for all scripts; negligible time compared to any non trivial program execution.
+Finally, the \todo{two} larger use cases show that housekeeping tasks (such as data downloading.
+extracting, and preprocessing), that are often implemented using shell scripts and \todo{left without treatment} by specialized parallelization frameworks: (i) sometimes take comparable time to the execution of the ``main'' computation, and (ii) can be equally parallelized by \sys offering significant performance benefits.
+%% transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
+%% As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
+%% Finally, \sys preserves the productivity of the shell:
+%% the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
+\kk{What is that about the COST?}
 }

### 2020-05-24 pldi/draft.tex
Update takeaway

--- 2020-05-24 pldi/draft.tex
@@
 to offer significant acceleration.
-A broader %% and more-ambitious
-take-away is that \sys is able to exploit the until now unharnessed parallelization opportunities of inter-program scripts that are used widely to download, extract, and clean data.
-
-untapped potential 
-that is 
-exploitble
-
-does not fall within purely compute-intensive targets that are the focus of conventional parallelization techniques.
+A broader take-away is that \sys is able to extract parallelism from fragments that do not fall under purely compute-intensive regions that are the focus of conventional parallelization systems.
 

### 2020-05-24 pldi/draft.tex
End-to-end pass

@@
   Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts it back to a script that uses POSIX constructs to explicitly guide parallelism.
-A set of lightweight annotations allow command developers to express key parallelizability properties of their commands.
-An accompanying parallelizability study of commands in POSIX and GNU Coreutils, two large and commonly used groups, 
-informs \sys's annotation language and result in its equivalent of a data-parallel standard library.
+A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
+An accompanying parallelizability study of POSIX and GNU commands , two large and commonly used groups, 
+informs \sys's annotation language and results in its equivalent of a data-parallel standard library.
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is to use shell primitives such as \ttt{\&}, \ttt{wait}, \ttt{for} to explicitly induce concurrency;
+Another option is to use shell primitives such as \ttt{\&}, \ttt{wait}, \ttt{for} to explicitly induce parallelism;
  these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
@@
 To top it off, all these options assume a good understanding of parallelism;
-  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options (see~\S\ref{related} for a full discussion).
+  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options.
+%  (see~\S\ref{related} for a full discussion).
 % \nv{I still don't like ``for a full discussion''---raises too many expectations, and we don't have that much more to say}
@@
 
-These components are finally paired with \sys's runtime component.
+These two components are finally paired with \sys's runtime component.
 % \sys's runtime  component additional packs
 % Aside from these two components, \sys packs a runtime component that serves dual purpose:
-Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data combinators as well as high-performance primitives for eager data splitting and merging.
-These address several practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
-  and were developed by uncovering several pathological situations, on many of which we report.
+Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data aggregators as well as high-performance primitives for eager data splitting and merging.
+These address many practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
+  and were developed by uncovering several pathological situations, on a few of which we report.
 
@@
 We evaluate \sys using
-  (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing pipelines;
+  (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing scripts;
   (ii) two large and complex use cases for temperature analysis and web indexing;
@@
 and 
-  (iii) a series of targeted micro-benchmarks inspired by (and
-comparing with) possible alternatives.  The results demonstrate
-significant advantages in \todo{correctness}, \todo{manual effort},
-and performance---with \sys offering an order-of-magnitude performance
-improvements due to its transformations and optimized runtime.
+  (iii) a series of micro-benchmarks inspired by (and
+comparing with) possible alternatives.  The results show
+significant advantages in correctness, manual effort,
+and performance---with an order-of-magnitude performance
+improvements due to \sys's transformations and runtime primitives.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as the script's execution time, the pipeline depth, and the parallelizability classes in which its commands fall.
+Improvements depend on several factors such as the execution time, the pipeline depth, and the parallelizability classes in which its commands fall.
 
@@
   \item
-  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components.
+  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components and their solutions.
 % \kk{Rewrite that when we know for sure what the section contains.}
@@
 \noindent
-After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper closes with a discussion of \sys's limitations and possible future directions~\sx{discussion}.
+After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper concludes~\sx{discussion}.
 
@@
 
-This section reviews shell scripting through an example~\sx{bg:pipelines}, which it then uses to present 
+This section reviews \unix shell scripting through an example~\sx{bg:pipelines}, which it then uses to present 
 parallelization challenges~\sx{bg:challenges} and how they are addressed by \sys~\sx{bg:overview}.
@@
 Suppose a environmental scientist    % looking for a non-CS expert---journalist might be too low
-  wants to get a quick sense of the trends in the maximum temperature over the past century.
+  wants to get a quick sense of the trends in the maximum temperature across the U.S. over the past ten years.
 As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
@@
 % 
-The effort required to write this pipeline is astonishingly low:\footnote{
-  Some effort is required to understand NCDC's weather format, but this is true for any program processing this dataset.
+The effort required to write this script is astonishingly low:\footnote{
+  Some effort is required to understand NOAA's weather format, but this is true for any program processing this dataset.
 }
   its data-processing core amounts to 12 stages and, when expressed as a single line, only 165 characters long.
+Its succinctness is deceiving---this program is no toy:
+  a Java program implementing only the last four pipeline stages requires 137LoC~\cite[\S2.1]{hadoop:15}.
 To enable such a succinct program composition, \unix incorporates several features.
@@
 %% task-parallel program composition over character streams.
-These streams are contiguous character lines separated by newline character (\textsc{NL}), which delineate individual stream elements.
+These streams are contiguous character lines separated by newline characters (\textsc{NL}), which delineate individual stream elements.
 For example, the first \ttt{grep} is given lines that contain file identifiers, of which it only outputs lines that contain \ttt{gz}, which are in turn consumed by \ttt{tr}.
@@
 
-Another important feature is synchronization, as different pipeline stages process data concurrently and possibly at different rates.
-For example, the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
+% Another important feature is synchronization.
+Different pipeline stages process data concurrently and possibly at different rates---\eg the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
 The \unix kernel facilitates scheduling, communication, and synchronization behind the scenes.
@@
 
-Command flags, used pervasively in \unix, are configuration options that the command's developer has decided to expose to its users, to give command users significant freedom in their execution.
+Command flags, used pervasively in \unix, are configuration options that the command's developer has decided to expose to its users to improve the command's general applicability.
 For example, by omitting  \ttt{sort}'s \ttt{r} flag that enables reverse sorting, the user can easily get the minimum temperature.
 The shell does not have any visibility into these flags; 
-  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to commands.
+  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to individual commands.
 
 Finally, \unix provides an environment for composing commands written in any language.
-Many of the commands used in practice come with the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
-The fact that commands are developed in a variety of languages---including shell scripts---provides users with a significant flexibility.
-For example, one could replace the call to \ttt{sort} and \ttt{head} with one to \ttt{./avg.py} to get an average or a standard deviation rather than a maximum.
-The pipeline would still work, as long \ttt{./avg.py} conforms to the interface outlined earlier.
+Many of the commands used in practice come with the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils.
+Ohers are available as add-ons.
+The fact that commands are developed in a variety of languages---including shell scripts---provides users with significant flexibility.
+For example, one could replace the call to \ttt{sort} and \ttt{head} with one to \ttt{./avg.py} to get the average or standard deviation rather than the maximum.
+The pipeline will still work, as long \ttt{./avg.py} conforms to the interface outlined earlier.
 % Shell scripting allows incremental development through refinement
@@
 
-% ---available publicly online via NCDC's NOAA~\cite{}---to obtain the maximum temperature per year.
+% ---available publicly online via NOAA~\cite{}---to obtain the maximum temperature per year.
 
@@
 composition, they complicate shell script parallelization, which even
-for simple scripts such as the one in Fig.~\ref{fig:example} faces
+for simple scripts such as the one in Fig.~\ref{fig:example} create
 several challenges.
@@
 \heading{Commands} % : Unbounded \& Mysterious}
-In contrast to restricted programming frameworks that enable parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, the \unix shell provides a unprecedented number and variety of composable commands.
+In contrast to restricted programming frameworks that enable parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, the \unix shell provides an unprecedented number and variety of composable commands.
 % Due to their varying nature, 
@@
 \heading{Scripts} % Restrictive \& Coarse-grained}
-Another challenge is the programming language of the POSIX shell. % of the shell, which is both restrictive and coarse-grained.
+Another challenge is due to the language of the POSIX shell. % of the shell, which is both restrictive and coarse-grained.
 % has been developed through a long history of additions.
 First, it contains constructs that enforce sequential execution;
-   Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before the rest starts executing.
+   Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before everything else.
 Moreover, the language semantics only exposes limited task-based parallelism in the form of \ttt{\&}, \ttt{for}, and \ttt{|} constructs.
@@
 %% Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
-For example, even though Fig.~\ref{fig:example}'s \ttt{for} focuses only on five years of data, \ttt{curl} still outputs thousands of lines per year;
+Even though Fig.~\ref{fig:example}'s \ttt{for} focuses only on five years of data, \ttt{curl} still outputs thousands of lines per year;
   naive parallelization of each loop iteration will miss such opportunities.
-Any attempt to automate parallelization should be aware of the shell semantics and induce as much parallelism as it can without breaking semantics.
+Any attempt to automate parallelization should be aware of the POSIX shell language and induce as much parallelism as it can without breaking semantics.
 
 \heading{Implementation} %
-Independently of the semantics of individual commands and the language of the shell, the \unix environment has its set of quirks.
+Independently of the semantics of individual commands and the shell language, the \unix environment has its own set of quirks.
 Thus, any attempt to orchestrate paralllel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
@@
 %  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
-Moreover, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators, to be correctly parallelized.
+Moreover, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators so as to be correctly parallelized.
 %% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
@@
 %% linkable
-modular components used to compose individual commands.
+modular components for augmenting command composition.
 %% for stream splitting and merging.
-It also provides a small library of efficient data-aggregators for a large set of parallelizable commands.
+It also provides a small library of efficient data aggregators for a large set of parallelizable commands.
 %% For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
@@
 \caption{
-  \textbf{\sys's $2\times$-parallel output for part of Fig.~\ref{fig:example}.}
+  \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
   \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
@@
 % Intuitively, a parallelizability 
-A class represents the level of synchronization required by copies of a command executing in parallel.
-It captures characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
+A class represents the level of synchronization required by copies of a command executing in parallel, 
+  capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 
-Broadly, shell commands can be split into four major classes (Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
+Broadly, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they need when processing their input (Tab.\ref{tab:classes}).
 %% how they interact with state.
 These classes are ordered in ascending difficulty (or impossibility) of parallelization.
-In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without gaining any performance improvements).
-Although commands can change classes depending on their arguments, we focus on no-flag defaults for now.
+In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (but foregoing any performance improvements).
+Commands can change classes depending on their flags, we % focus on no-flag defaults for now and 
+  which are discussed later~\sx{ext}.
 
@@
 These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
-Stateless commands may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but when they receive empty input they return empty output.
+They may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but return empty output for empty input.
 Workloads that use only stateless commands are trivial to parallelize:
@@
 A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
-\todo{Such constraints affect task parallelism, but not data parallelism}:
+Such constraints affect task parallelism, but not data parallelism:
   \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by a $fold$ that merges the results.
@@
 For example, hashing commands such as \ttt{sha1sum} maintain complex state that has to be updated sequentially.
-If parallelized on a single input, each \todo{stage} would need to wait on the results of all the previous stage foregoing any parallelism benefits.
-
-It is worth noting that while these commands are not parallelizable for a single input, they are still parallelizable across different inputs.
-For example, a web crawler involving hash generation for comparing individual pages would allow \ttt{sha1sum} to proceed in parallel for different pages.
+If parallelized on a single input, each stage would need to wait on the results of all previous stages foregoing any parallelism benefits.
+
+It is worth noting that while these commands are not parallelizable at the granularity of a single input, they are still parallelizable across different inputs.
+For example, a web crawler involving hashing to compare individual pages would allow \ttt{sha1sum} to proceed in parallel for different pages.
 
 \heading{Side-effectful Commands}
-The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
+The last class, \sid, contains commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
 Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.
@@
 This is the largest class, for two reasons.
-First, it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
+First, it includes commands related to the file-system---a central abstraction of the \unix design and philosophy~\cite{unix}.
 In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
-Second, it contains commands that do not consume input or do not produce output---and thus are not amenable to data-parallelism.
-For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that
+Second, it contains commands that do not consume input or do not produce output---and thus not amenable to data parallelism.
+For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands interfacing with kernel- or hardware-generated information and do not consume any input from user programs.
   %A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
@@
 To address the challenge of language-agnostic extensibility~\sx{bg}, \sys allows communicating several key details about command parallelizability through lightweight annotations.
-These annotations can be used by both developers of new commands---which includes users developing their own scripts---as well as developers maintaining existing commands.
-The latter can express additions or changes to the command's implmentation or interface, important as commands are maintained or extended over long periods of time.
+These annotations can be used by both developers of new commands---including users developing their own scripts---as well as developers maintaining existing commands.
+The latter can express additions or changes to the command's implementation or interface, important as commands are maintained or extended over long periods of time.
 
 \heading{Key Concerns}
-\sys's annotations focused on three crucial concerns:
+\sys's annotations focus on three crucial concerns:
   (i) a command's parallelizability class,
   (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
+  (iii) for commands with multiple flags (options), the handling of competing sets of flags.
 As the first concern was discussed extensively in the previous section, we now focus on the latter two.
@@
 To be able to accurately construct the DFG representation, \sys needs to know certain information about a command's inputs and outputs.
-There are couple of reasons for this.
+There are a few reasons for this.
 One reason is due to \sys's DFG representation, which connects commands with each other;
@@
 \noindent
-This is correct only because \sys knows that \ttt{grep} in the sequential program reads from \ttt{f1} first and from \ttt{f2} second.
+This is correct only because \sys knows that \ttt{grep} in the sequential program reads first from \ttt{f1} and then from \ttt{f2}.
 
@@
 As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
-For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} that writes debugging output to a file, it ends up in \sta.
+For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
 
@@
   the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
-A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output columns.
+A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output column(s).
 Its annotation record is shown below:
@@
 The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
-In the first case, since the first and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
+As the first (second) and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
+% NV: I don't think this is correct, because the small input might be the other argument;  you could even have `comm - -`
   this input accompanies every replicated instance of the command executing in parallel.
-The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
+The third clause is the general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
 There, \ttt{comm} is classified as \pur as it needs to % wait until it 
@@
 
-\heading{Providing \todo{Aggregators}}
-For commands in \sta, the annotations are enough to enable parallelization: they are applied to partitions of their input in parallel, and their outputs are simply concatenated.
+\heading{Custom Aggregators}
+For commands in \sta, the annotations are enough to enable parallelization:
+  commands are applied to parts of their input in parallel, and their outputs are simply concatenated.
 %% commands either fall under \sta, having straightforward aggregators, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
@@
   (iii) their composition produces the same output with the original command.
-If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their parallelization.
-
-\sys defines the combiners for most of the POSIX and Coreutils commands in \pur, which operate as both \sys's standard library and exemplar for community attempts to tackle other commands.
+If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
+
+\sys defines aggregators many \pur POSIX and GNU commands, which operate as both \sys's standard library and an exemplar for community attempts to tackle other commands.
 
@@
 
-The core of \sys is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
+\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
 %% representation that is expressive enough to model data parallelism in shell programs.
@@
 then instantiates each DFG back to a parallel shell script.  A
-fundamental difference with other DFG models is that it encodes the
+fundamental difference with other DFG models is that \sys's DFG encodes the
 order in which a node reads its inputs, which in turn enables a set of
@@
 The two main shell abstractions are
-  (i) streams, i.e. files or pipes, that contain data, and
+  (i) streams, \ie files or pipes, that contain data, and
   (ii) commands that communicate through these streams.
@@
 
-We first introduce basic notation used throughput this section.
+We first introduce basic notation used in this section.
 For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
@@
 We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-The $\leq$ order is reflexive, antisymmetric, and transitive (\ie is a partial order) and is often called the \emph{prefix order}.
+The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.
 
@@
 Edges that do not start from a node in the graph represent the graph input;
-  edges that do not point to a node in the graph are its outputs.
+  edges that do not point to a node in the graph represent its outputs.
 %% \nv{OK This distinction is explained in this par; I would put this earlier}
@@
 such as sending signals.  We require that the function $f$ is monotone
-with respect to a lifting of the prefix order for a sequence of inputs
+with respect to a lifting of the prefix order for a sequence of inputs.
 %% \kk{Formalize
@@
 file. The quintessential example of streaming commands is \ttt{cat},
-that consumes its inputs in order, producing their concatenation as
+which consumes its inputs in order, producing their concatenation as
 output. A more interesting example is \ttt{comm -23 f1 f2}, which has
-\ttt{f2} as its static input (it needs to read it completely before
-consuming anything from \ttt{f1} and then consumes \ttt{f1} one
-element at a time. These commands can be represented as functions of
+\ttt{f2} as its static input (which it needs to read completely before
+consuming anything from \ttt{f1}, which in turn consumes one
+element at a time). These commands can be represented as functions of
 type $f : D\kstar \times [D\kstar] \to D\kstar$, where the first
@@
 that act as parallelization-exposing optimizations. Both the domain
-and range of these transformations is a graph in our DFG model
-and therefore they can be composed arbitrarily and in any order.
+and range of these transformations is a graph in \sys's DFG model
+and therefore transformations can be composed arbitrarily and in any order.
 %% Having a DFG representation allows us to define
 %% parallelization optimizations as graph transformations.
-Before describing the different types of optimizations, we formalize
+Before describing the different types of transformations, we formalize
 the intuition behind classes \sta and \pur described informally
@@
 \heading{Stateless and Parallelizable Pure Commands}
-As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
+As noted~\sx{cmd}, stateless commands such as \ttt{tr}
 operate independently on individual elements of the stream
 (characters, lines, or files) without maintaining any state. In this
-work, we consider lines as the data quantum, so only commands that are
-stateless regarding lines (or any finer granularity such as
-characters) are considered stateless. Formally, a
-streaming command $f$ is stateless if it commutes with
+work, we consider lines as the data quantum, thus we consider stateless only commands that are
+stateless with respect to lines (or any finer granularity such as
+characters). Formally, a streaming command $f$ is stateless if it commutes with
 the operation of concatenation, \ie it is a semigroup homomorphism:
@@
 %
-Based on these equations, we can define a \emph{node parallelization
-  transformation} on a stateless node $v$ that is preceded by a
-concatenation, i.e. the command \ttt{cat}, of $n$ input streams and is
+Based on these equations, we can define a node parallelization
+  transformation $T$ on a node $v \in$ \sta that is preceded by a
+concatenation, \ie the command \ttt{cat}, of $n$ input streams and is
 followed by a node $v'$
-(Fig.~\ref{fig:parallelization-transformation}). The transformation
+(Fig.~\ref{fig:parallelization-transformation}). $T$
 replaces $v$ with $n$ new nodes, routing each of the $n$ input streams
@@
 
-This transformation can extended straightforwardly to pure commands
+$T$ can be extended straightforwardly to nodes $v \in$ \pur,
 implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
@@
 s), s)$.
+As long as the pair $(m, agg)$ meets the three invariants outlined earlier~\sx{ext}, 
+$T$ can be shown to be behavior-preserving.
 % \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.
@@
 %
-\sys also performs a set of auxiliary transformations that are
+\sys also performs a set of auxiliary transformations $t_{1-3}$that are
 depicted in \Cref{fig:auxiliary-transformations}. If a node has many
-inputs, \sys concatenates these inputs by inserting a \ttt{cat} node
+inputs, $t_1$ concatenates these inputs by inserting a \ttt{cat} node
 to enable the parallelization transformations. In cases where a
 parallelizable node has one input and is not preceded by a
-concatenation, \sys inserts a \ttt{cat} node that is preceded by its
+concatenation, $t_2$ inserts a \ttt{cat} node that is preceded by its
 inverse \ttt{split}, so that the concatenation can be commuted with
-the node. The third depicted transformation inserts a relay node that
+the node. Transformation $t_3$ inserts a relay node that
 performs the identity transformation. Relay nodes can be useful for
@@
 
-This section describes \sys's implementation.
-It first details how \sys's front-end translates a shell script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
-It then describes \sys's back-end, responsible for generating the parallel script, and the technical challenges addressed to achieve performance benefits~\sx{optimizer}.
-
-\subsection{Front End}
+This section describes key points in the implementation of \sys.
+It first details how its front-end translates a script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
+It then describes its back-end, responsible for generating the parallel script~\sx{optimizer}.
+% , and the technical challenges addressed to achieve performance benefits
+
+\subsection{\sys Front-End}
 \label{front-end}
 
-To optimize a shell script, \sys translates it to a DFG~\sx{ir}.
-As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}~\S\ref{bg}---\sys introduce the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
+To optimize a script, \sys translates it to a DFG~\sx{ir}.
+As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
 %% In addition, the shell's highly dynamic
@@
   fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
-Consider this script:
+Consider this fragment:
 
@@
 The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---\ie more precisely, they are maximal.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---more precisely, they are maximal.
 
@@
 %
-The \sys front-end performs a depth-first search pass on the AST of
+\sys's front-end performs a depth-first search on the AST of
 the given shell program.  During this pass, it extends the
@@
 original AST where parallelizable regions have been replaced with
-DFGs and a call to \sys's runtime.
+DFGs and calls to \sys's runtime primitives~\sx{optimizer}.
 
@@
 
-\subsection{Runtime/Back End}
+\subsection{\sys Back End}
 \label{optimizer}
@@
 The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
-This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
+% This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
 %% support concurrently executing many more commands than the
@@
 %% have progress one at a time.
-However, such laziness severely inhibits opportunities for parallelism as commands are often blocked when their consumers are not requesting any input.
+Such laziness inhibits opportunities for parallelism, as commands are often blocked when their consumers are not requesting any input.
 Consider the following script:
@@
 \noindent
-The \ttt{cat} command will only try to consume input from \ttt{t2} when it completes reading from \ttt{t1}.
-As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes. % processing.
+The \ttt{cat} command will try to consume input from \ttt{t2} after it completes reading from \ttt{t1}.
+As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes (Fig. \ref{fig:eager}a). % processing.
 
 % replacing FIFOs with files is
-The closest solution \unix provides would be to replace FIFOs with files, as files are a central abstraction and can simulate pipes of arbitrary buffering.
-Aside from severe performance implications, naive replacement can lead to subtle race conditions as the consumer might reach EOF before the producer.
-Alternatively, a consumer could the producer to complete, but that would insert an artificial barrier inhibiting task-based parallelism and wasting disk resources.
+A solution would be to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
+Aside from severe performance implications, naive replacement can lead to subtle race conditions as a consumer might reach {\sc EOF} before a producer.
+Alternatively, consumers could wait for producers to complete before opening the file for reading (Fig. \ref{fig:eager}c);
+however, this would insert artificial barriers impeding task-based parallelism and wasting disk.
 % \unix does not provide an easy way out.
@@
 % Instead of addressing this issue for each command separately, % nv: how could this be possible?
-  \sys inserts and instantiates \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}). % nv: not sure what critical meant
+  \sys inserts and instantiates eager \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}d). % nv: not sure what critical meant
 These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
-We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
+% nv: Removed below because I don't think it adds much
+% We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
 
@@
   Addressing intermediary laziness is challenging:
-    (i) FIFOs are blocking;
-    (ii) files alone introduce race conditions between producer/consumer;
-    (iii) files \ttt{wait} inhibit task-based parallelism.
-Eager \ttt{relay} nodes address the challenge while remaining within the \sys model.
+    (a) FIFOs are blocking;
+    (b) files alone introduce race conditions between producer/consumer;
+    (c) files \ttt{wait} inhibit task-based parallelism.
+Eager \ttt{relay} nodes (d) address the challenge while remaining within the \sys model.
 % Describe. \TODO{Combine this figure with the above code maybe.}
@@
 
-To achieve this, % satisfy the tension between efficiency and \todo{knowledge},  % blanking on the right word here
+To solve this, % satisfy the tension between efficiency and \todo{knowledge},  % blanking on the right word here
   \sys supports two \ttt{split} implementations:
@@
 %% task-based parallelism.
-For the former, \ttt{split} first consumes its complete input, counts its lines, and then splits it equally to the desired number of outputs.
-For the latter, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
+In the former case, \ttt{split} first consumes its complete input, counts its lines, and then splits it equally to the desired number of outputs.
+In the latter case, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
 
@@
     \textbf{Summary of shell one-liners}.
-    Small pipelines (\ie ones usually developed on-the-fly) drawn from various sources and applied to large datasets.
+    % Small pipelines (\ie ones usually developed on-the-fly) drawn from various sources and applied to large datasets.
     % Names correspond to the speedup plots (Fig.~\ref{fig:microbenchmarks}).
@@
     Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting DFG size and compilation time for two indicative parallelization configurations.
+    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes aggregators, eager, and split nodes) and compilation time for two indicative parallelization configurations.
     % Highlights notes a few key features of each script.
@@
 % A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a command exits after it has produced and sent all of its output to its output channel.
-However, if its output is a pipe and the pipe reader exits early, the command is notified to stop writing early.
+Under normal operation, a command exits after it has produced and sent all its results to its output channel.
+If the channel is a pipe and its reader exits early, the command is notified to stop writing early.
 In \unix, this is achieved by an out-of-band error mechanism: the
-operating system will deliver a \ttt{PIPE} signal to the producer,
+operating system delivers a \ttt{PIPE} signal to the producer,
 notifying it that the pipe's consumer has exited.  This is different
@@
   code---similar to \ttt{DeleteFile} and other Win32 functions.}
-primarily because pipes and pipelines are at the heart of its
-philosophy.
+primarily because pipes and pipelines are at the heart of \unix.
 % write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
+Unfortunately though, if a pipe has not been opened for writing yet% ---possible for named pipes---
+  \unix cannot signal this condition.
 % \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
@@
 As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
-This, in turn, leaves it unable to make progress, as it is is both blocked and unaware of its consumer exiting.
+This, in turn, leaves it unable to make progress, as it is both blocked and unaware of its consumer exiting.
 Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
@@
 
-\heading{Pure Fold Function Implementations}
-Commands in \tpur can be parallelized using a \emph{map} and a \emph{fold} stage~\sx{parallelizability}.
-\sys implements \emph{fold} for several commands in \tpur to enable parallelization.
-A few interesting examples include \emph{fold} functions for
+\heading{Aggregator Implementations}
+Commands in \tpur can be parallelized using a \emph{map} and an \emph{aggregate} stage~\sx{parallelizability}.
+\sys implements \emph{aggregate} for several commands in \tpur to enable parallelization.
+A few interesting examples include \emph{aggregate} functions for
   (i) \ttt{sort}, which amounts to the merge phase of a merge-sort (and on GNU systems is implemented as \ttt{sort -m}),
-  (ii) \ttt{uniq} and \ttt{uniq -c}, which need to check conditions at the boundary of their streams,
-  (iii) \ttt{tac}, which consumes stream descriptors in the reverse order, and 
+  (ii) \ttt{uniq} and \ttt{uniq -c}, which need to check conditions at the boundary of their input streams,
+  (iii) \ttt{tac}, which consumes stream descriptors in reverse order, and 
   (iv) \ttt{wc}, which adds inputs with an arbitrary number of elements (\eg \ttt{wc -lw} or \ttt{wc -lwc} \etc).
@@
 %% and aggregates them accordingly.
+
+% OSDIFIX \nv{Should we add a small subsection about implementation details?}
 
@@
 \heading{Highlights}
-\fixme{
-Scripts from various sources and workloads see a significant speedup---almost linear in embarrassingly parallel cases---over sequential execution while returning identical results (on non-trivial datasets).
-The effort to reap this benefits is minimal; only sometimes requiring minimal annotations for \todo{custom} commands.
-\sys highly-optimized primitives (\sx{optimizer}) offer performance benefits even against manually implemented alternatives (such as GNU sort, and GNU parallel).
-\sys's compilation process takes $<$5s for all scripts; negligible time compared to any non trivial program execution.
-Finally, the \todo{two} larger use cases show that housekeeping tasks (such as data downloading.
-extracting, and preprocessing), that are often implemented using shell scripts and \todo{left without treatment} by specialized parallelization frameworks: (i) sometimes take comparable time to the execution of the ``main'' computation, and (ii) can be equally parallelized by \sys offering significant performance benefits.
+\sys accelerates ($1.92-61.1\times$) almost all scripts ($>50$, with $>200$ commands).
+From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
+\sys's optimized primitives offer benefits even against manually parallelized scripts. % (such as GNU sort, and GNU parallel).
+% \sys's compilation pass is negligible (avg. $<1s$). %  for non trivial program execution.
+In two large cases, tasks such as data download, extraction, and preprocessing---often implemented using shell scripts and outside the focus of specialized parallelization frameworks---take comparable time to the execution of the main computation, and can be significantly accelerated by \sys.
+% In terms of manual effort,
+
+The vast majority requires no effort other than invoking \sys;
+  the annotation for the remaining \todo{6} commands ($<3\%$) amounts to a single record.
+% 
+In terms of correctness, \sys's results over multi-GB inputs are identical to the sequential for all benchmarks.
+Scripts feature ample opportunities for break semantics, which \sys avoids.
+In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
+%in another, it preserves the semantics that another parallelization command breaks.
+
+% conservative program transformations will not attempt paral- lelizing program sections
+% In another comparison with \ttt{parallel}
+% and in another 
+% Does not break 
+% The effort to reap this benefits is minimal; 
+%  minimal annotations for \todo{custom} commands.
+
 %% transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
@@
 %% the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
-\kk{What is that about the COST?}
-}
+% \kk{What is that about the COST?}
   %% in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
+% In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation. \kk{Is it worth leaving this sentence in here?}
+%% kk: I think now that we removed Hadoop, it is not so important. In
+%% the end of the day, we don't want to convince people that they
+%% should use shell for all their needs.
 
@@
 
-The goal of this subsection is to evaluate \sys (and its different configurations) on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
+We evaluate \sys under multiple configurations on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
 
@@
     \caption{
-      \textbf{Speedup achieved by \sys as a function of the level of parallelism (2--64).} 
-      Different configurations per one-liner:
-      (i) \todo{Par + Split}, showing the performance of \sys with both eager and split optimizations enabled,
-      (ii) \todo{Par + B. Split}, showing the performance of \sys with eager and the \todo{batch-size annotated} split optimizations enabled,
-      (iii) \todo{Parallel}, showing the performance of \sys with eager enabled and split disabled,
-      (iv) \todo{No Eager}, showing the performance of \sys with split and eager disabled,
-      (v) \todo{Blocking Eager}, showing the performance of \sys with split disabled and the naive blocking eager enabled,
-      (vi) \todo{Opt.
-Parallel}, showing the performance of an optimized version of \ttt{bigrams}~\cf{ours}.
-Only the relevant configurations are shown for each one-liner (e.g.
-grep and sort don't benefit from split, and spell and bi-grams don't get any benefit without split).
+      \textbf{\sys's speedup for 2--6$\times$-parallelism.} 
+      Different configurations per benchmark:
+      (i) Par+Split: \ttt{eager} and general \ttt{split} enabled,
+      (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
+      (iii) Parallel: \ttt{eager} and \ttt{split} disabled,
+      (iv) No Eager: only \ttt{split} enabled,
+      (v) Blocking Eager, only \todo{naive blocking} \ttt{eager} enabled (no \ttt{split},
+      (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
+      Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
+      % TODO: WHY?
     }
@@
 Diff and Set-diff compare streams via a \ttt{diff} (in \npu, non-parallelizable) and \ttt{comm} (in \pur), respectively.
-Sort-sort uses consecutive \pur commands without interleaving them with a \pur that has the effect of significantly condensing the input size (\eg \ttt{uniq}).
-Finally, the two \ttt{bi-grams} replicate and shift a stream by one entry to calculate bigrams, but \ttt{-opt}'s shift, \ttt{sort}, and \ttt{uniq} are merged to one command to use a more efficient aggregator.
+Sort-sort uses consecutive \pur commands without interleaving them with commands that condense their input size (\eg \ttt{uniq}).
+Finally, the two Bi-grams replicate and shift a stream by one entry to calculate bigrams, but Bi-gram-opt's shift, \ttt{sort}, and \ttt{uniq} are merged to one command to use a more efficient aggregator.
 
@@
 % TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents the speedup gained by \sys as a function of the level of parallelism (2--64$\times$).
+Fig.~\ref{fig:microbenchmarks} presents \sys's speedup as a function of parallelism (2--64$\times$).
 (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups per level of parallelism are 1.97 (for $2\times$), 3.5 ($4\times$), 5.78 ($8\times$), 8.83 ($16\times$), 10.96 ($32\times$), 13.47 ($64\times$).
+Average speedups for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 % we experiment with a $200\times$ level of parallelism to see potential overheads
-Tab.~\ref{tab:eval} shows that \sys's analysis and compilation time is negligible % ($<5s$)
-  for all benchmarks.
-It also contains Grep-light (not shown in Fig.~\ref{fig:microbenchmarks}), designed to show that \sys does not slow down IO-intensive scripts;
-  on the contrary, \sys achieves a 1.5-2.5$\times$ speedup for grep-light. 
+Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
+It also contains Grep-light (not in Fig.~\ref{fig:microbenchmarks}), designed to show that \sys does not slow down IO-intensive scripts;
+  on the contrary, \sys achieves a 1.5-2.5$\times$ speedup for Grep-light. 
+\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 \sys achieves near-linear speedup on low-parallelism configurations.
-This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{aggregations}s and \ttt{relay}s---evident in all plots that contain \pur stages.
+This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
 For example, Sort in $8\times$ spawns 37 nodes:
-  8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \ttt{sort -m} nodes, and 14 \ttt{relay} nodes.
+  8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \emph{aggregation} nodes, and 14 \ttt{relay} nodes.
 In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
@@
 \heading{Take-aways}
-\sys achieves significant speedups that range between 4--60$\times$ depending on the parallelizability characteristics of the commands used.
-\sys's runtime constructs offer significant improvements over the baseline speedup achieved by \sys's core transformations.
+\sys accelerates scripts by 4--60$\times$, depending on the commands involved.
+Its runtime constructs improve over the baseline speedup achieved by its transformations.
 
@@
 
-The goal of this subsection is to evaluate \sys on an existing set of \unix pipelines found in the wild.
+We now evaluate \sys on an existing set of \unix pipelines found in the wild and written by non-experts.
 
@@
 We consider this a good set of benchmarks because
-  (i) the problems were designed to highlight the \unix modular philosophy~\cite{mcilroy1978unix} and make extensive use of built-ins with a variety of flags, and
+  (i) the problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix} and make extensive use of standard commands under a variety of flags, and
   (ii) the solutions were written by non-experts (contrary to \S\ref{ours}).
@@
 Their input was generated by multiplying their original input several times, enough to reach 10GB.
-Fig~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
-The average speedup is $5.49$, \todo{the median $6.07\times$}, and the weighted average (with the absolute times as weights) is $5.75\times$.
+Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
+Average speedup is $5.49\times$, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},
@@
 Most pipelines see significant acceleration, except 13, 24, 25, 26, 29, 30 that see no speedup and 2, 19, 31 that see a slowdown.
-The first group contain commands which \sys cannot parallelize---\eg \ttt{awk}, \ttt{sed} \ttt{-d}, and \ttt{tr} \ttt{-d}.
+The first group contains commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
 Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-By applying only the replacements which to us experts felt the most natural, we also give \sys the opportunity to identify and exploit parallelism---a common pattern we witnessed~\sx{macro1}.
-For example, after this change, 13 sees a 8.1$\times$ speedup.
+By attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
+% ---a common pattern we witnessed~\sx{macro1}.
 
@@
 
-For the rest, \sys's speedup is \todo{capped} due to a combination of reasons:
+For the rest, \sys's speedup is capped due to a combination of reasons:
   (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} (0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
@@
 
-To understand \sys better under a realistic workload, we turn our attention to the Fig.~\ref{fig:example}'s script~\sx{bg}.
+% For a realistic workload, 
+We now turn our attending to Fig.~\ref{fig:example}'s script~\sx{bg}.
 
 \heading{Program}
-This program is inspired by Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it exemplifies a realistic mid-scale analytics pipeline and comprises three sub-tasks:
-  download weather data from NCDC (shell), convert them to a Hadoop-friendly format (shell), and process them to get the maximum temperature (Hadoop).
-Only the last of these tasks is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
-  whereas we consider the end-to-end pipeline that implements all three sub-tasks---from fetching data to outputting the maximum temperature.
+This program is inspired by Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising three sub-tasks:
+  fetch data from NOAA (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
+Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
+  whereas we consider the entire pipeline. % ---from fetching data to outputting the maximum temperature.
 % The version used in the evaluation is slightly different than the one presented in Fig.~\ref{fig:example} (14 stages and totalling 240 characters), simplified only for clarity of exposition, in that it expresses the \ttt{for} loop as a first stage followed by \ttt{xargs}.
@@
 \heading{Results}
-The complete shell pipeline executes in 44m2s for five years (82GB) of weather data.
-Applying \sys with 2/10$\times$ parallelism leads to 1.86/2.44$\times$ speedup. 
-Naturally, different phases see different benefits:
-  downloading, extracting, and preparing data sees 1.77/1.99$\times$ speedup (vs. 33m58s), and
-  computing the maximum temperature sees 2.30/10.79$\times$ speedup (vs. 10m4s).
-For a discussion of why speedup is higher than the amount of parallelism, see \S\ref{ours}'s Discussion.
+The complete pipeline executes in 44m2s for five years (82GB) of data.
+\sys with 2/10$\times$ parallelism leads to 1.86/2.44$\times$ speedup, with different phases seeing different benefits:
+  1.77/1.99$\times$ (vs. 33m58s) for all the pre-processing and 
+  2.30/10.79$\times$ speedup (vs. 10m4s) for computing the max.
+(For why speedup is higher than parallelism, see \S\ref{ours}.) % 's Discussion.
 
@@
 %> % ---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
-
-In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation. \kk{Is it worth leaving this sentence in here?}
-%% kk: I think now that we removed Hadoop, it is not so important. In
-%% the end of the day, we don't want to convince people that they
-%% should use shell for all their needs.
 
@@
 \heading{Take-aways}
-\sys can be applied to programs of significant size and input
+\sys can be applied to programs of notable size and input
 %% \todo{larger} % real/realistic? implies that prior benchmarks were not  
@@
 to offer significant acceleration.
-A broader take-away is that \sys is able to extract parallelism from fragments that do not fall under purely compute-intensive regions that are the focus of conventional parallelization systems.
+A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, which are the focus of conventional parallelization systems.
 
@@
 
-As a second realistic workload, we use a large web-indexing script composed from various \todo{sources}~\cite{}.
+We now apply \sys to a large web-indexing script. % composed from various \todo{sources}~\cite{}.
 
 \heading{Program}
-Coarsely, the pipeline amounts to two distinct components communicating via \unix pipes.
+Coarsely, this script amounts to two distinct components communicating via \unix pipes.
 The \todo{first} downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
- the second extracts the text from HTML and applies natural-language processing analyses---\eg trigrams, character conversion, term frequencies---to index it.
+ the second extracts the text from HTML and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
 Collectively, they contain a total of 52 stages (first: 18; second: 34) written in multiple programming languages.
@@
 % To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
-As the \todo{first} component of the pipeline contains a cycle (URLs$\rightarrow$HTML$\rightarrow$URLs), \sys's conservative transformations do not attempt to parallelize it.
-The second component, applied to 1\% of Wikipedia (1.3GB), takes 191 minutes.
-With 2/16$\times$-parallelism, \sys brings it down to 97m/15m ($1.97/12.7\times$) with the majority of the speedup coming from parallelizing the HTML-to-text conversion.
+As the \todo{first} component contains a cycle (URLs $\rightarrow$ HTML $\rightarrow$ URLs), \sys's conservative transformations do not attempt to parallelize it.
+The latter, applied to 1\% of Wikipedia (1.3GB), takes 191min.
+With 2/16$\times$-parallelism, \sys brings it down to 97/15min ($1.97/12.7\times$), with the majority speedup coming from the HTML-to-text conversion.
 % which uses \ttt{pandoc}, a utility written in Haskell.
@@
 \heading{Discussion}
-The original script contains 34 pipeline stages thus the sequential version already benefits from task-based parallelism.
+The original script contains 34 pipeline stages, thus the sequential version already benefits from task-based parallelism.
 It also uses several utilities not part of the standard POSIX/GNU set---\eg its \ttt{url-extract}ion is written in JavaScript and its \ttt{word-stem}ming is in Python.
@@
 \heading{Take-aways}
-\sys operates on programs with annotated commands outside the POSIX/GNU command subsets and leads to notable speedups even when the original program features significant task-based parallelism.
-\sys's conservative transformations do not risk breaking correctness, \todo{thus do not improve the first component.}
+\sys operates on programs with annotated commands outside the POSIX/GNU subsets and leads to notable speedups even when the original program features significant task-based parallelism.
+\sys does not risk breaking semantics, thus does not improve the first component. % this will not change
 
@@
 %% \end{wrapfigure}
-\ttt{-{}-parallel} flag set ($S_{GNU}$), shown on the side.
+\ttt{-{}-parallel} flag set ($S_{GNU}$). % , shown on the side.
 % The results are shown on the side-figure below.
 While \ttt{-{}-parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{GNU}$'s parallelism is set to $2\times$ of $S_{PaSh}$'s (\ie the rightmost point is for \ttt{-{}-parallelism=128}), 
+$S_{GNU}$'s parallelism is $2\times$ $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{-{}-parallelism=128}), 
 to account for \sys's additional merge processes.
@@
 A few points are worth noting.
-$S_{PaSh}$ without \ttt{eager} performs comparably to $S_{GNU}$ and \ttt{eager}  outperforms $S_{GNU}$ ($\sim2\times$).
-This is because \ttt{eager} adds intermediate buffers %(on disk)
+$S_{PaSh}$ without \ttt{eager} performs comparably to $S_{GNU}$, and with \ttt{eager} it outperforms $S_{GNU}$ ($\sim2\times$);
+  this is because \ttt{eager} adds intermediate buffers %(on disk)
   between merge phases.
@@
 %% purpose of having a simple solution.
-Second, $S_{GNU}$ indicates that \sys's scalability limit is \todo{inherent and cannot be overpassed}.
-Because of this, all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) have this upper bound on scalability.
+$S_{GNU}$ indicates that \ttt{sort}'s scalability is inherently limited (\ie due to \ttt{sort} and not \sys);
+  this is why all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) do not go above $8\times$.
 Finally, the comparison also shows \sys's benefits to command developers:
-  rather than implementing a custom flag for parallelism, a simple  parallelizability annotation would allow \sys to scale similarly to or better than a handcrafted implementation with minimal effort.
+  a low-effort parallelizability annotation allows similar or better scalability than a custom flag added by developers.
 
@@
 
-\heading{Graph Transformations}
-\sys's analysis and generation overheads remain collectively under $5s$ even for the high-scalability benchmarks.
-We decided to push \sys's limits by constructing an artificial 1000-stage pipeline---a worst-case workload that is nowhere near the ones observed in practice.
-We find the resulting times quite acceptable:
-  given two input arguments in the beginning of this pipeline, \sys takes $8.23s$ for parsing the script and $3.22s$ for analysis and optimization.
-Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
-\kk{I propose removing this subsection. 5 minutes is too much time for compilation and this happens atm because of a pathological scenario. The compilation should never take more than a second.}
+% \heading{Graph Transformations}
+% \sys's analysis and generation overheads remain collectively under $5s$ even for the high-scalability benchmarks.
+% We decided to push \sys's limits by constructing an artificial 1000-stage parallelizable region---a worst-case workload that is nowhere near the ones observed in practice.
+% We find the resulting times quite acceptable:
+%   for a $2\times$-parallelization \sys takes $8.23s$ for parsing the script and $3.22s$ for analysis and optimization.
+% Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
+% \kk{I propose removing this subsection. 5 minutes is too much time for compilation and this happens atm because of a pathological scenario. The compilation should never take more than a second.}
 
@@
 and introduces (and depends on) other programs with complex semantics, such as ones for SQL querying and CSV parsing.
-In contrast, \sys manages to parallelize large scripts correctly with minimal user effort.
+In contrast, \sys manages to parallelize large scripts correctly with minimal-to-zero user effort.
 %% embraces the \unix philosophy, simply attempting to rewrite scripts more efficiently with minimal user effort.
@@
 
-% \section{Annotation Language Grammar}
-% \label{alg}
-% 
-% \Cref{fig:dsl} shows the complete grammar of the parallelizability
-% annotation language.
-% 
-% \begin{figure}
-%   \centering
-%   \begin{grammar}
-%     <option> ::= `-' <string>
-% 
-%     <category> ::= `stateless' | `pure' | ...
-% 
-%     <maybe-int> ::= | <int>
-% 
-%     <arg> ::= `args[' <int> `]'
-% 
-%     <args> ::= <arg>
-%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
-% 
-%     <input> ::= `stdin' | <args>
-% 
-%     <inputs> ::= <input>
-%     \alt <input> `,' <inputs>
-% 
-%     <output> ::= `stdout' | <arg>
-% 
-%     <outputs> ::= <output>
-%     \alt <output> `,' <outputs>
-% 
-%     <option-pred> ::= <option>
-%     \alt `value' <option> = <string>
-%     \alt `not' <option-pred>
-%     \alt <option-pred> `or' <option-pred>
-%     \alt <option-pred> `and' <option-pred>
-% 
-%     <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
-% 
-%     <predicate> ::= <option-pred> `=>' <assignment>
-% 
-%     <pred-list> ::= `|' <predicate> <pred-list>
-%     \alt `|' `otherwise' `=>' <assignment>
-% 
-%     <command> ::= <name> `\{' <pred-list> `\}'
-% 
-%     <command-list> ::= <command>
-%     \alt <command> <command-list>
-%   \end{grammar}
-%   \caption{
-%   \textbf{Parallelizability description language.}
-%     The DSL captures important information regarding the parallelizability of a command.
-%   }
-%   \label{fig:dsl}
-% \end{figure}
+\section{Annotation Language Grammar}
+\label{alg}
+
+\Cref{fig:dsl} shows the complete grammar of the parallelizability
+annotation language.
+
+\begin{figure}
+  \centering
+  \begin{grammar}
+    <option> ::= `-' <string>
+
+    <category> ::= `stateless' | `pure' | ...
+
+    <maybe-int> ::= | <int>
+
+    <arg> ::= `args[' <int> `]'
+
+    <args> ::= <arg>
+    \alt `args[' <maybe-int> `:' <maybe-int> `]'
+
+    <input> ::= `stdin' | <args>
+
+    <inputs> ::= <input>
+    \alt <input> `,' <inputs>
+
+    <output> ::= `stdout' | <arg>
+
+    <outputs> ::= <output>
+    \alt <output> `,' <outputs>
+
+    <option-pred> ::= <option>
+    \alt `value' <option> = <string>
+    \alt `not' <option-pred>
+    \alt <option-pred> `or' <option-pred>
+    \alt <option-pred> `and' <option-pred>
+
+    <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
+
+    <predicate> ::= <option-pred> `=>' <assignment>
+
+    <pred-list> ::= `|' <predicate> <pred-list>
+    \alt `|' `otherwise' `=>' <assignment>
+
+    <command> ::= <name> `\{' <pred-list> `\}'
+
+    <command-list> ::= <command>
+    \alt <command> <command-list>
+  \end{grammar}
+  \caption{
+  \textbf{Parallelizability description language.}
+    The DSL captures important information regarding the parallelizability of a command.
+  }
+  \label{fig:dsl}
+\end{figure}
 

### 2020-05-24 pldi/microbenchmarks-table.tex
End-to-end pass

@@
 \midrule
-grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
-sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
-top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
-wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
-grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
-spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 69 & 261 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
-shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
-diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
-bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 109 & 445 & 0.146s & 1.716s & stream shifting and merging \\
-bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
-set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
-sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78 & 318 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+Grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
+Sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+Top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
+Wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
+Grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
+Spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 69 & 261 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
+Shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
+Diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
+Bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 109 & 445 & 0.146s & 1.716s & stream shifting and merging \\
+Bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
+Set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
+Sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78 & 318 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-05-24 pldi/draft.tex
Minor modifications in Figure 7 caption

@@
       (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (iii) Parallel: \ttt{eager} and \ttt{split} disabled,
-      (iv) No Eager: only \ttt{split} enabled,
-      (v) Blocking Eager, only \todo{naive blocking} \ttt{eager} enabled (no \ttt{split},
+      (iii) Parallel: \ttt{eager} enabled (no \ttt{split}),
+      (v) Blocking Eager, only blocking \ttt{eager} enabled (no \ttt{split}),
+      (iv) No Eager: both \ttt{eager} and \ttt{split} disabled,
       (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.

### 2020-05-25 pldi/draft.tex
Fix typos, update wording

@@
   This paper presents \sys, a shell aimed at parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
-  Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts it back to a script that uses POSIX constructs to explicitly guide parallelism.
+  Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
 A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
@@
 % oblivious? agnostic?  
-One option is manual tools such GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
-  these tools are either command-unaware thus at risk of breaking program semantics, or too coarse-grained thus only being capable to exploit parallelism at the level of entire scripts rather than individual components.
+One option is manual tools such as GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
+  these tools are either command-unaware, and thus at risk of breaking program semantics, or too coarse-grained, and thus only capable of exploiting parallelism at the level of entire scripts rather than individual components.
 % Tools will attempt to parallelize anything 
@@
 and performance---with an order-of-magnitude performance
-improvements due to \sys's transformations and runtime primitives.
+improvement due to \sys's transformations and runtime primitives.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
@@
 % In the ongoing debate on climate change,
-Suppose a environmental scientist    % looking for a non-CS expert---journalist might be too low
+Suppose an environmental scientist    % looking for a non-CS expert---journalist might be too low
   wants to get a quick sense of the trends in the maximum temperature across the U.S. over the past ten years.
@@
 The effort required to write this script is astonishingly low:\footnote{
-  Some effort is required to understand NOAA's weather format, but this is true for any program processing this dataset.
+  Some effort is required to understand NOAA's weather format, but this is true for any program processing any dataset.
 }
-  its data-processing core amounts to 12 stages and, when expressed as a single line, only 165 characters long.
+  its data-processing core amounts to 12 stages and, when expressed as a single line, is only 165 characters long.
 Its succinctness is deceiving---this program is no toy:
@@
 Many of the commands used in practice come with the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils.
-Ohers are available as add-ons.
+Others are available as add-ons.
 The fact that commands are developed in a variety of languages---including shell scripts---provides users with significant flexibility.
 For example, one could replace the call to \ttt{sort} and \ttt{head} with one to \ttt{./avg.py} to get the average or standard deviation rather than the maximum.
-The pipeline will still work, as long \ttt{./avg.py} conforms to the interface outlined earlier.
+The pipeline will still work, as long as \ttt{./avg.py} conforms to the interface outlined earlier.
 % Shell scripting allows incremental development through refinement
@@
 
-While these features aid developer economy through powerful program
+While these features aid in development effort economy through powerful program
 composition, they complicate shell script parallelization, which even
@@
 %% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
-Manual analysis is also challenging, due to the sheer number of commands, and the many flags that affect their behaviour---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+Manual analysis is also challenging, due to the sheer number of commands, and the many flags that affect their behavior---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
  % , making correct treatment even more difficult. % nv: we already said impossible so more difficult doesn't add much!
@@
 
-Broadly, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they need when processing their input (Tab.\ref{tab:classes}).
+Broadly speaking, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they need when processing their input (Tab.\ref{tab:classes}).
 %% how they interact with state.
@@
 In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (but foregoing any performance improvements).
-Commands can change classes depending on their flags, we % focus on no-flag defaults for now and 
+Commands can change classes depending on their flags,% we focus on no-flag defaults for now and 
   which are discussed later~\sx{ext}.
@@
 
-The choice of line as the data element strikes a convenient balance between coarse-grained (files) and fine-grained separation (characters) while staying in line with \unix's core abstractions.
+The choice of line as the data element strikes a convenient balance between coarse-grained (files) and fine-grained (characters) separation while staying in line with \unix's core abstractions.
 This choice can affect the allocation of commands in \sta, as many of its commands (about 1/3) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
@@
 
-To be able to accurately construct the DFG representation, \sys needs to know certain information about a command's inputs and outputs.
+To be able to accurately construct the DFG representation, \sys needs to know certain details about a command's inputs and outputs.
 There are a few reasons for this.
@@
 
-\sys defines aggregators many \pur POSIX and GNU commands, which operate as both \sys's standard library and an exemplar for community attempts to tackle other commands.
+\sys defines aggregators for many \pur POSIX and GNU commands, which operate as both \sys's standard library and an exemplar for community attempts to tackle other commands.
 
@@
 \ttt{f2} as its static input (which it needs to read completely before
-consuming anything from \ttt{f1}, which in turn consumes one
+consuming anything from \ttt{f1}, which is in turn consumed one
 element at a time). These commands can be represented as functions of
@@
 \heading{Stateless and Parallelizable Pure Commands}
-As noted~\sx{cmd}, stateless commands such as \ttt{tr}
+As previously noted~\sx{cmd}, stateless commands such as \ttt{tr}
 operate independently on individual elements of the stream
@@
 incoming edge represents a stream of data $x_i : D^*$, and the only
-behaviour of a DFG is its output, this optimization $v(x_1 \cdot x_2
+behavior of a DFG is its output, this optimization $v(x_1 \cdot x_2
 \cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$
@@
   (i) a general one that can be used with commands that could add or remove lines, and
-  (ii) an optimized one that can be used if the input size before its execution.
+  (ii) an optimized one that can be used if the input size is known before its execution.
 %% The \sys \todo{backend} support two split node implementations, to
@@
 %% task-based parallelism.
-In the former case, \ttt{split} first consumes its complete input, counts its lines, and then splits it equally to the desired number of outputs.
+In the former case, \ttt{split} first consumes its complete input, counts its lines, and then splits it uniformly across the desired number of outputs.
 In the latter case, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
@@
 In terms of correctness, \sys's results over multi-GB inputs are identical to the sequential for all benchmarks.
-Scripts feature ample opportunities for break semantics, which \sys avoids.
+Scripts feature ample opportunities for breaking semantics, which \sys avoids.
 In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
@@
 (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47.
+Average speedups for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 % For a realistic workload, 
-We now turn our attending to Fig.~\ref{fig:example}'s script~\sx{bg}.
+We now turn our attention to Fig.~\ref{fig:example}'s script~\sx{bg}.
 
@@
 
-Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they requiresignificantly more \emph{user} effort (see \sx{intro} for distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
+Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they require significantly more \emph{user} effort (see \sx{intro} for distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
 %% Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
@@
 The DFG is a prevalent model in several areas of data processing (including batch-~\cite{mapreduce:08, spark:12} and stream-processing ~\cite{murray2013naiad, carbone2015flink}).
-Despite its popularity, most systems perform optimizations that do not preserve semantics, introducing subtle erroneous behaviours.
+Despite its popularity, most systems perform optimizations that do not preserve semantics, introducing subtle erroneous behaviors.
 Recent work~\cite{HSSGG2014, SHGW2015, MSAIT2019} attempts to address this issue by performing optimizations only in cases where correctness is preserved.
@@
 \sys chooses a more general approach:
-   it does not require setting up a new framework for each new class of computation used, nor rewriting different parts of the computation using a different set abstractions provided by each framework.
+   it does not require setting up a new framework for each new class of computation used, nor rewriting different parts of the computation using a different set of abstractions provided by each framework.
 %% It also takes a more principled approach: by developing the

### 2020-05-26 pldi/draft.tex
Typos and minor fixes

@@
 The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
-This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple pipelines that target data processing, system orchestration, and other automation tasks.
+This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple pipelines for data processing, system orchestration, and other automation tasks.
 % Pipe-centric shell commands are parallel (and to some extent, functional) programs that can natively enjoy significant parallelism on a multicore machine.
@@
 
-The first group is \emph{command developers}, responsible for developing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
+The first group is \emph{command developers}, responsible for implementing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
 These developers usually work in a single programming language, leveraging its abstractions to provide parallelism whenever possible.
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is to use shell primitives such as \ttt{\&}, \ttt{wait}, \ttt{for} to explicitly induce parallelism;
+Another option is to use shell primitives (such as \ttt{\&}, \ttt{wait}, \ttt{for}) to explicitly induce parallelism;
  these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
@@
 To address cold-start issues, \sys comes with a library of parallelizability semantics for commands in POSIX and GNU Coreutils.
-These large classes of commands serve as the shell's standard library, expected to be found pervasively. % in scripts found out in the wild.
+These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
 The study that led to their characterization also informed \sys's annotation and transformation components.\footnote{
@@
 Suppose an environmental scientist    % looking for a non-CS expert---journalist might be too low
-  wants to get a quick sense of the trends in the maximum temperature across the U.S. over the past ten years.
+  wants to get a quick sense of trends in the maximum temperature across the U.S. over the past ten years.
 As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
@@
 Its succinctness is deceiving---this program is no toy:
-  a Java program implementing only the last four pipeline stages requires 137LoC~\cite[\S2.1]{hadoop:15}.
+  a Java program implementing only the last four pipeline stages requires 137~LoC~\cite[\S2.1]{hadoop:15}.
 To enable such a succinct program composition, \unix incorporates several features.
@@
 Command flags, used pervasively in \unix, are configuration options that the command's developer has decided to expose to its users to improve the command's general applicability.
-For example, by omitting  \ttt{sort}'s \ttt{r} flag that enables reverse sorting, the user can easily get the minimum temperature.
+For example, by omitting  \ttt{sort}'s \ttt{-r} flag that enables reverse sorting, the user can easily get the minimum temperature.
 The shell does not have any visibility into these flags; 
@@
 Independently of the semantics of individual commands and the shell language, the \unix environment has its own set of quirks.
-Thus, any attempt to orchestrate paralllel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
+Thus, any attempt to orchestrate parallel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
 For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
@@
 %  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
-Moreover, specific commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators so as to be correctly parallelized.
+Moreover, several commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators so as to be correctly parallelized.
 %% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
@@
 % Since the shell representation only exposes coarse grained parallelization opportunities, 
-\sys then converts each parallelizable region to a \emph{dataflow graph} (DGF), 
+\sys then converts each parallelizable region to a \emph{dataflow graph} (DFG), 
 %% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
@@
   \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
-  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
+  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners. \TODO{Correct Dish!}
 }
@@
 Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc} simply maintains a counter.
-Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
+Other commands, such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
 % 
@@
 In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
-Second, it contains commands that do not consume input or do not produce output---and thus not amenable to data parallelism.
+Second, it contains commands that do not consume input or do not produce output---and thus are not amenable to data parallelism.
 For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands interfacing with kernel- or hardware-generated information and do not consume any input from user programs.
@@
 \heading{Key Concerns}
-\sys's annotations focus on three crucial concerns:
+\sys's annotations focus on four crucial concerns:
   (i) a command's parallelizability class,
+  (ii) a command's inputs and outputs,
   (ii) for commands that support multiple inputs, the characteristics of input consumption, and
   (iii) for commands with multiple flags (options), the handling of competing sets of flags.
-As the first concern was discussed extensively in the previous section, we now focus on the latter two.
+As the first concern was discussed extensively in the previous section, we now focus on the latter three.
 
@@
 \heading{Example Annotations}
-An annotation record contains a set of clauses, each one identified by a predicate on the command options.
+The parallelizability properties of each command are written in an annotation record that
+%% An annotation record
+contains a set of clauses, each one identified by a predicate on the command options.
 Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
@@
 % }
-The annotation grammar is shown in Appendix~\ref{alg}.
+The complete annotation grammar is shown in Appendix~\ref{alg}.
 
@@
   this input accompanies every replicated instance of the command executing in parallel.
-The third clause is the general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
+The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
 There, \ttt{comm} is classified as \pur as it needs to % wait until it 
@@
   (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{aggregate} can consume (and combine) the results of multiple \emph{map} invocations, and
-  (iii) their composition produces the same output with the original command.
+  (iii) their composition produces the same output as the original command.
 If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
@@
 % replacing FIFOs with files is
-A solution would be to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
+One might be tempted to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
 Aside from severe performance implications, naive replacement can lead to subtle race conditions as a consumer might reach {\sc EOF} before a producer.
@@
 % write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet% ---possible for named pipes---
+Unfortunately though, if a pipe has not been opened for writing yet
+% ---possible for named pipes---
   \unix cannot signal this condition.
@@
 Scripts feature ample opportunities for breaking semantics, which \sys avoids.
-In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
+%% In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
 %in another, it preserves the semantics that another parallelization command breaks.
@@
 \heading{Setup}
-\sys was run on 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores, Debian 4.9.144-3.1, GNU Core-utils 8.30-3, GNU Bash 5.0.3(1)-, Python 3.7.3, and OCaml 4.05.0---without any special configuration in hardware or software.
+\sys was run on 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores, Debian 4.9.144-3.1, GNU Coreutils 8.30-3, GNU Bash 5.0.3(1), and Python 3.7.3---without any special configuration in hardware or software.
+%% and OCaml 4.05.0 %% this was here for the parser that we don't mention atm
 Except as otherwise noted,
@@
 \sys achieves near-linear speedup on low-parallelism configurations.
-This is due to a high degree (deep trees) of task-based parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
+This is due to a high degree (deep trees) of task parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
+%% \kk{Ben's point: task parallelism is mentioned here for the first time. Since \sys exposes data parallelism as task parallelism, we might want to mention that somewhere earlier in the paper}
 For example, Sort in $8\times$ spawns 37 nodes:
@@
 \heading{Program}
-Coarsely, this script amounts to two distinct components communicating via \unix pipes.
-The \todo{first} downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
- the second extracts the text from HTML and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
-Collectively, they contain a total of 52 stages (first: 18; second: 34) written in multiple programming languages.
+This script reads a file containing Wikipedia URLs, downloads the pages,
+%% Coarsely, this script amounts to two distinct components communicating via \unix pipes.
+%% The \todo{first} downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
+%% the second
+extracts the text from HTML, and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
+In total, it contains 34 commands
+%% Collectively, they contain a total of 52 stages (first: 18; second: 34)
+written in multiple programming languages.
 
@@
 % To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
-As the \todo{first} component contains a cycle (URLs $\rightarrow$ HTML $\rightarrow$ URLs), \sys's conservative transformations do not attempt to parallelize it.
-The latter, applied to 1\% of Wikipedia (1.3GB), takes 191min.
+%% As the \todo{first} component contains a cycle (URLs $\rightarrow$ HTML $\rightarrow$ URLs), \sys's conservative transformations do not attempt to parallelize it.
+The original script takes 191min to execute on 1\% of Wikipedia (1.3GB).
 With 2/16$\times$-parallelism, \sys brings it down to 97/15min ($1.97/12.7\times$), with the majority speedup coming from the HTML-to-text conversion.
@@
 \sys operates on programs with annotated commands outside the POSIX/GNU subsets and leads to notable speedups even when the original program features significant task-based parallelism.
-\sys does not risk breaking semantics, thus does not improve the first component. % this will not change
+%% \sys does not risk breaking semantics, thus does not improve the first component. % this will not change
 
@@
 Recently, Smoosh~\cite{smoosh:18} argued for making concurrency explicit via shell constructs.
-The argument is somewhat antithetical to \sys's, which argues for mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
+The argument is dissimilar from \sys's, which argues for mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
 

### 2020-05-26 pldi/draft.tex
Checkpoint

--- 2020-05-24 pldi/draft.tex
@@
       (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (iii) Parallel: \ttt{eager} enabled (no \ttt{split}),
-      (v) Blocking Eager, only blocking \ttt{eager} enabled (no \ttt{split}),
-      (iv) No Eager: both \ttt{eager} and \ttt{split} disabled,
+      (iii) Parallel: \ttt{eager} and \ttt{split} disabled,
+      (iv) No Eager: only \ttt{split} enabled,
+      (v) Blocking Eager, only \todo{naive blocking} \ttt{eager} enabled (no \ttt{split},
       (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
@@
 
-Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they requiresignificantly more \emph{user} effort (see \sx{intro} for distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
+Many prior techniques for exploiting parallelism are not directly comparable with \sys, either because they require significantly more \emph{user} effort (see \S\ref{intro} for distinction between users and developers)
+  or because they are too specialized, targeting narrow domains or custom programming abstractions.
 %% Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).

### 2020-05-26 pldi/draft.tex
Minor corrections: D{a,i}sh -> Pash
Discussion about different configurations

--- 2020-05-26 pldi/draft.tex
@@
 %% extraction tools.
-
+https://git.io/JfEgr
 %% Author with single affiliation.
@@
   Anonymous Author(s)\\
-  \normalsize{Additional Anonymized Material: \href{https://git.io/JfEgr}{https://git.io/JfEgr}.}
+  \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
 }
@@
   \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
-  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners. \TODO{Correct Dish!}
+  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
 }
@@
     \caption{
-      \textbf{\sys's speedup for 2--6$\times$-parallelism.} 
+      \textbf{\sys's speedup for 2--64$\times$-parallelism.}
       Different configurations per benchmark:
@@
 (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
+Average speedups of the best \sys configuration, \ie \ttt{eager} enabled with \ttt{split} when it provides benefit, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 %   minute).
-An additional conclusion is that \sys's modular eager-relay significantly improves performance over the other alternatives (blocking eager or no eager, \S\ref{optimizer}).
+
+No Eager and Blocking Eager depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
+These configurations perform worse than \sys with \ttt{eager} in all cases (except for a negligible difference in shortest-scripts) for the reasons described in \Cref{optimizer}.
+Adding \ttt{split} on top of that (Par + Split, Par + B.
+Split) further improves speedup for some scripts that have \tpur or \tnpu commands, and for the rest it does not affect performance.
+Finally, the optimized bi-grams script achieves significantly more benefits from \sys than its unoptimized counterpart, showing that familiarity with \sys's parallelizability classes goes a long way in terms of performance.
 

### 2020-05-27 pldi/microbenchmarks-table.tex
updated table

@@
 Grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
-Spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 69 & 261 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
+Spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
 Shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
 Diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
-Bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 109 & 445 & 0.146s & 1.716s & stream shifting and merging \\
+Bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
 Bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
 Set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
-Sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 78 & 318 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+Sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-05-27 pldi/draft.tex
minor

@@
 %% extraction tools.
-https://git.io/JfEgr
+%https://git.io/JfEgr
 %% Author with single affiliation.
@@
   with a particular emphasis on users.
-Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
+Command developers are provided with a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
   rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
-Shell users, on the other hand, are provided full automation:
+Shell users, on the other hand, are provided with full automation:
   \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
@@
 %  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
-Moreover, several commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators so as to be correctly parallelized.
+Moreover, several commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators in order to be correctly parallelized.
 %% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
@@
   for execution.
-To expose and exploit data-parallelism, % in a given shell script,
+To expose and exploit data parallelism, % in a given shell script,
 %% parallel execution,

### 2020-05-27 pldi/draft.tex
Teensy fixes

--- 2020-05-26 pldi/draft.tex
@@
 %% extraction tools.
-https://git.io/JfEgr
+% https://git.io/JfEgr
 %% Author with single affiliation.
@@
 
+\heading{A Note on the Parallelism Factor}
+\nv{
+The discussion of the benchmarks often refers to the parallelism factor 
+This is not necessarily the resulting parallelism achieved by \sys, which typically is above the level configured by the user.
+This is because for each process (node in the graph), \sys creates multiple processes---aggregators, splitters, relays \etc
+Some times the results give the impression of superlinear parallelism, but this has a simple explanation.
+}
+
 \subsection{Common \unix One-liners}
@@
     \caption{
-      \textbf{\sys's speedup for 2--64$\times$-parallelism.}
+      \textbf{\sys's speedup for 2--6$\times$-parallelism.} 
       Different configurations per benchmark:
@@
 (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the best \sys configuration, \ie \ttt{eager} enabled with \ttt{split} when it provides benefit, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
+Average speedups for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 %   minute).
-
-No Eager and Blocking Eager depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-These configurations perform worse than \sys with \ttt{eager} in all cases (except for a negligible difference in shortest-scripts) for the reasons described in \Cref{optimizer}.
-Adding \ttt{split} on top of that (Par + Split, Par + B.
-Split) further improves speedup for some scripts that have \tpur or \tnpu commands, and for the rest it does not affect performance.
-Finally, the optimized bi-grams script achieves significantly more benefits from \sys than its unoptimized counterpart, showing that familiarity with \sys's parallelizability classes goes a long way in terms of performance.
+An additional conclusion is that \sys's modular eager-relay significantly improves performance over the other alternatives (blocking eager or no eager, \S\ref{optimizer}).
 
@@
 Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
+(\nv{Pipelines are referred by number, in both the plot and the discussion below.})
 Average speedup is $5.49\times$, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
@@
 While \ttt{-{}-parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{GNU}$'s parallelism is $2\times$ $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{-{}-parallelism=128}), 
+$S_{GNU}$'s parallelism is configured to $2\times$ that of $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{-{}-parallelism=127}), 
 to account for \sys's additional merge processes.

### 2020-05-27 pldi/draft.tex

--- 2020-05-27 pldi/draft.tex
@@
 %% extraction tools.
-%https://git.io/JfEgr
 %% Author with single affiliation.

### 2020-05-27 pldi/draft.tex
Small fixes in unix50 and the discussion of one-liners

@@
 No Eager and Blocking Eager depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-These configurations perform worse than \sys with \ttt{eager} in all cases (except for a negligible difference in shortest-scripts) for the reasons described in \Cref{optimizer}.
+As described in \Cref{optimizer}, these configurations perform worse than \sys with \ttt{eager} in all cases (except for a negligible difference in shortest-scripts).
 Adding \ttt{split} on top of that (Par + Split, Par + B.
 Split) further improves speedup for some scripts that have \tpur or \tnpu commands, and for the rest it does not affect performance.
-Finally, the optimized bi-grams script achieves significantly more benefits from \sys than its unoptimized counterpart, showing that familiarity with \sys's parallelizability classes goes a long way in terms of performance.
+Finally, the optimized bi-grams script achieves significantly more benefits from \sys than its unoptimized counterpart, showing that familiarity with \sys's parallelizability classes could go a long way in terms of performance.
 
@@
 Their input was generated by multiplying their original input several times, enough to reach 10GB.
-Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines.
-(\nv{Pipelines are referred by number, in both the plot and the discussion below.})
+Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines. We use the pipeline index to refer to them in both plots and discussion.
+%% (\nv{Pipelines are referred by number, in both the plot and the discussion below.})
 Average speedup is $5.49\times$, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
@@
 The first group contains commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
-Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-By attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
+Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind; a common example is the use of \ttt{awk} for reordering or removing columns of the input.
+%% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
+In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the 13th pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
+%% , attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
 % ---a common pattern we witnessed~\sx{macro1}.

### 2020-05-27 pldi/draft.tex
A note on parallelism

@@
 \heading{A Note on the Parallelism Factor}
-\nv{
-The discussion of the benchmarks often refers to the parallelism factor 
-This is not necessarily the resulting parallelism achieved by \sys, which typically is above the level configured by the user.
-This is because for each process (node in the graph), \sys creates multiple processes---aggregators, splitters, relays \etc
-Some times the results give the impression of superlinear parallelism, but this has a simple explanation.
-}
+%
+The discussion of the results often mentions the parallelism factor, which refers to the \sys \emph{parallelism configuration} rather than the resulting parallelization.
+In fact, for small parallelism configurations \sys often achieves higher parallelization due to the creation of multiple processes---aggregators, splitters, relays \etc
+%% \nv{
+
+%% This is not necessarily the resulting parallelism achieved by \sys, which typically is above the level configured by the user.
+%% This is because for each process (node in the graph), \sys creates multiple processes---aggregators, splitters, relays \etc
+%% Some times the results give the impression of superlinear parallelism, but this has a simple explanation.
+%% }
 

### 2020-05-27 pldi/draft.tex
changres

--- 2020-05-27 pldi/draft.tex
@@
     \textbf{Parallelizability Classes}.
-    Broadly, \unix commands can be broken down into to \todo{four} classes according to their parallelizability properties.
+    Broadly, \unix commands can be broken down into to four classes according to their parallelizability properties.
   }
@@
 \begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
-\toprule
+\toprule % 28 + 9 + 13 + 105
 Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
 \midrule
-Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 22 (21.1\%)          \\  % 22
-Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  \todo{21 (20.1\%)}    & 21 (20.1\%)          \\  % 21
-Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  \todo{27 (25.9\%)}    & 27 (25.9\%)          \\  % 27
-Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 57 (58.8\%)            \\  % 25
+Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 28 (18\%)          \\  % 22
+Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  8 (7.6\%)}            & 9  (5\%)          \\  % 21
+Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  13 (12.4\%)}          & 13 (8.3\%)          \\  % 27
+Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 105(67.8\%)            \\  % 25
 % Irreversible     & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     & 5  (0.4\%)           \\  % 5
@@
 % Using its \ttt{--jobs} flag, we  redirect input from a file to 
-Sequential execution takes \todo{XX}s \vs \sys's \todo{XX}s (\todo{XX}$\times$).
+% 554 / 330 
+Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$).
 We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command---masking \sys's improvements in other parts of the pipeline.
@@
 There are a few possible ways one might attempt to use GNU \ttt{parallel} on this program.
-They could use it on the \todo{XX} stage, assuming they know it is a bottleneck, bringing execution down to \todo{XX}s (\todo{XX}$\times$ speedup).
+They could use it on the 8th stage, assuming they know it is a bottleneck, bringing execution down to 304.4s ($1.8\times$ speedup).
 Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
-This would lead to \todo{XXX}\% performance improvements but severely incorrect results with respect to the sequential execution---a \ttt{diff} that averages about \todo{XX--XX} lines.
-\sys's conservative program transformations will not attempt parallelizing program sections that contain commands with unclear \todo{parallelizability} properties.
+This would lead to 3.2$\times$ performance improvements but severely incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
+\sys's conservative program transformations will not attempt parallelizing program sections that contain commands with unclear parallelizability properties.
 	% or commands whose inputs depend on earlier outputs.

### 2020-05-27 pldi/draft.tex
fix

@@
 \begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
-\toprule % 28 + 9 + 13 + 105
+\toprule
 Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
@@
 Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 28 (18\%)        \\  % 22
-Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  8 (7.6\%)}            & 9  (5\%)          \\  % 21
-Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  13 (12.4\%)}          & 13 (8.3\%)          \\  % 27
+Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  8 (7.6\%)            & 9  (5\%)          \\  % 21
+Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  13 (12.4\%)          & 13 (8.3\%)       \\  % 27
 Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 105(67.8\%)          \\  % 25

### 2020-07-10 evaluation/plots/microbenchmarks-table.tex
Update remaining empty cells in the table and fix a discrepancy in the one-liner tiling plot

--- 2020-05-05 evaluation/plots/microbenchmarks-table.tex
@@
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllll}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllllll}
 \toprule
-Script ~&~ Structure & Input &Seq Time & Script Size(16, 64) &Compile Time (16, 64) & Highlights \\
+Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
 \midrule
-\tti{grep} ~&~ $3\times$\tsta & 1~GB & 79m41.137s & \todo{\#Commands} & 0.050s\qquad 0.417s & complex NFA regex \\
-\tti{sort} ~&~ $\tsta, \tpur$ & 10~GB & 21m27.510s & \todo{\#Commands} & 0.076s\qquad 0.839s & \tti{sort}ing \\
-\tti{top-n} ~&~ $2\times\tsta, 4\times\tpur$ & 10~MB & 73m28.098s & \todo{\#Commands} & 0.119s\qquad 1.407s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{wf} ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m26.670s & \todo{\#Commands} & 0.121s\qquad 1.388s & double \tti{sort}, \tti{uniq} reduction \\
-\tti{grep-light} ~&~ $3\times$\tsta & 100~GB & 1m38.212s & \todo{\#Commands} & 0.031s\qquad 0.163s & 	odo{light computation} \\
-\tti{spell} ~&~ $4\times\tsta, 3\times\tpur$ & 1~GB & 8m9.370s & \todo{\#Commands} & 0.214s\qquad 2.800s & comparisons (\tti{comm}) \\
-\tti{shortest-scripts} ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m25.818s & \todo{\#Commands} & 0.210s\qquad 2.762s & \todo{extensive file-system operation} \\
-\tti{diff} ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m10.092s & \todo{\#Commands} & 0.254s\qquad 1.958s & non-parallelizable \tti{diff}ing \\
-\tti{optimized bi-grams} ~&~ $3\times$\tsta, \tpur & 10~MB & 125m10.224s & \todo{\#Commands} & 0.098s\qquad 1.171s & optimized version of bigrams \\
-\tti{set-diff} ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m22.068s & \todo{\#Commands} & 0.250s\qquad 3.316s & two pipelines merging to a \tti{comm} \\
+grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
+sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
+wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
+grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
+spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
+shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
+diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
+bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
+bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
+set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
+sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 \bottomrule

### 2020-07-16 pldi/draft.tex
Prepare arxiv version

@@
 
+\newcommand\blfootnote[1]{%
+  \begingroup
+  \renewcommand\thefootnote{}\footnote{#1}%
+  \addtocounter{footnote}{-1}%
+  \endgroup
+}
+
 \begin{document}
@@
 \author{
-  Anonymous Author(s)\\
-  \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
+% Anonymous Author(s)\\
+% \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
+  {\rm \normalsize Nikos Vasilakis$^{*}$\qquad Konstantinos Kallas$^{*\dagger}$\qquad  Konstantinos Mamouras$^{\circ}$ }\\
+  {\rm \normalsize Achilleas Benetopoulos$^{\diamond}$ \qquad Lazar Cvetkovich$^{\ddagger}$ }\\
+  {\small MIT, CSAIL \quad $^{\dagger}$University of Pennsylvania \quad $^{\circ}$Rice University \quad $^{\diamond}$Unaffiliated \quad $^{\ddagger}$University of Belgrade}
 }
-%% \authornote{with author1 note}          %% \authornote is optional;
+% \authornote{$^{*}$Equal contribution.}     %% \authornote is optional;
 %%                                         %% can be repeated if necessary
@@
 
+\blfootnote{$^{*}$Equal contribution.}
 The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
+\caption{
+  \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
+  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
+}
+\vspace{-15pt}
+\label{fig:example2}
+\end{figure}
+
 \heading{Scripts}
@@
 The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
-\caption{
-  \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
-  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
-}
-\vspace{-15pt}
-\label{fig:example2}
-\end{figure}
 
@@
 
-\sys defines aggregators for many \pur POSIX and GNU commands, which operate as both \sys's standard library and an exemplar for community attempts to tackle other commands.
+\sys defines aggregators for many POSIX and GNU commands in \pur, dubbing as both its standard library and an exemplar for community efforts tackling other commands.
 
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_cat.pdf}}
+\vspace{-28pt}
+\caption{
+  \textbf{Stateless parallelization transformation.}
+  The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
+}
+\vspace{-15pt}
+\label{fig:parallelization-transformation}
+\end{figure}
+
 \sys defines a set of semantics-preserving graph transformations
@@
 %%   so we changes the name to map-fold to not imply commutativity.}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_cat.pdf}}
-\vspace{-28pt}
-\caption{
-  \textbf{Stateless parallelization transformation.}
-  The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
-}
-\vspace{-15pt}
-\label{fig:parallelization-transformation}
-\end{figure}
 
@@
 The vast majority requires no effort other than invoking \sys;
-  the annotation for the remaining \todo{6} commands ($<3\%$) amounts to a single record.
+  the annotation for the remaining 6 commands ($<3\%$) amounts to a single record.
 % 
@@
 This would lead to 3.2$\times$ performance improvements but severely incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
-\sys's conservative program transformations will not attempt parallelizing program sections that contain commands with unclear parallelizability properties.
+\sys's conservative program transformations will not operate on fragments with unclear parallelizability properties.
 	% or commands whose inputs depend on earlier outputs.
@@
 
-%% \begin{acks}
+\subsection*{Acknowledgements}
 % Dumping people so that we don't forget
-  % 
-  %% This material is based upon work supported by the
+We want to thank 
+  Andr\'e DeHon,
+  Ben Karel,
+  Jean-S\'ebastien L\'egar\'e, and 
+  Nick Roessler.
+This material is based upon work supported by DARPA contract no. HR00112020013 (ALRSCC) and NSF award CCF 1763514.
 %% \grantsponsor{GS100000001}{National Science
@@
 %% National Science Foundation.
-%% \end{acks}
 
 %% Bibliography
+{\small
 \bibliography{./bib}
+}
 

### 2020-07-16 evaluation/plots/coarse-microbenchmarks-table.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllll}
\toprule
Script ~&~ Structure & Input &Seq. Time & Highlights \\
\midrule
grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & complex NFA regex \\
sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & \tti{sort}ing \\
top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & double \tti{sort}, \tti{uniq} reduction \\
wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & double \tti{sort}, \tti{uniq} reduction \\
spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & comparisons (\tti{comm}) \\
diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & non-parallelizable \tti{diff}ing \\
bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & stream shifting and merging \\
set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & two pipelines merging to a \tti{comm} \\
shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & long \tsta pipeline ending with \tpur \\
\bottomrule
\end{tabular*}

### 2020-07-17 pldi/draft.tex
Update based on arxiv feedback

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
+\pdfoutput=1 % for arxiv
+
 \usepackage{usenix2019_v3}
@@
 A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
-An accompanying parallelizability study of POSIX and GNU commands , two large and commonly used groups, 
+An accompanying parallelizability study of POSIX and GNU commands, two large and commonly used groups, 
 informs \sys's annotation language and results in its equivalent of a data-parallel standard library.
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes—with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
+\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes---with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
 \end{abstract}

### 2020-09-21 pldi/draft.tex
Update Unix50 figs to be sorted by speedup

@@
 \heading{Discussion}
-Most pipelines see significant acceleration, except 13, 24, 25, 26, 29, 30 that see no speedup and 2, 19, 31 that see a slowdown.
+Most pipelines see significant acceleration, except \todo{13, 24, 25, 26, 29, 30} that see no speedup and \todo{2, 19, 31} that see a slowdown.
 The first group contains commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
@@
 %% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the 13th pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
+In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the \todo{13th} pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
 %% , attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
@@
 For the rest, \sys's speedup is capped due to a combination of reasons:
-  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} (0, 1, 3, 15, 16, 18, 20, 27, 28, 33),
-  (ii) some are deep pipelines that already exploit task-based parallelism (7, 8, 9, 10, 11, 12, 14, 27, 28,) and
-  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs (4, 5, 6, 7, 8, 12, 14, 22, 23).
+  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
+  (ii) some are deep pipelines that already exploit task-based parallelism \todo{(7, 8, 9, 10, 11, 12, 14, 27, 28,)} and
+  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs \todo{(4, 5, 6, 7, 8, 12, 14, 22, 23)}.
 
@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
+    \todo{Pipelines are sorted in descending speedup order}.
   }

### 2020-09-21 compiler/gather_results.py
Sort unix50 pipelines by speedup:

@@ -587,6 +587,9 @@ def aggregate_unix50_results(all_results, scaleup_numbers):
     return avg_distr_results
 
 def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
+
+    ## TODO: Sort by speedup
+    all_results.sort(key=lambda x: x[0][scaleup_numbers.index(parallelism)], reverse=True)
     ## Plot individual speedups
     individual_results = [distr_exec_speedup[scaleup_numbers.index(parallelism)]
                           for distr_exec_speedup, _ in all_results]

### 2020-09-21 compiler/gather_results.py
Add average line in Unix50

@@ -611,23 +611,35 @@ def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
     speedup_color = 'tab:blue'
     seq_time_color = 'tab:red'
 
+
+
     fig, ax1 = plt.subplots()
+    ax2 = ax1.twinx()  # instantiate a second axes that shares the same x-axis
     # print(fig.get_size_inches())
     fig.set_size_inches(12, 6)
 
-    ## Plot speedup
-    ax1.set_ylabel('Speedup', color=speedup_color)
-    ax1.set_xlabel('Pipeline')
+    ## Plot line at 1
     plt.xlim(-1, len(individual_results))
-    plt.hlines([1], -1, len(individual_results) + 1, linewidth=0.8)
-    ax1.bar(ind-w, individual_results, width=2*w, align='center', color=speedup_color)
-    ax1.tick_params(axis='y', labelcolor=speedup_color)
 
-    ax2 = ax1.twinx()  # instantiate a second axes that shares the same x-axis
-    ax2.set_ylabel('Sequential Time (s)', color=seq_time_color)
-    ax2.set_yscale("log")
-    ax2.bar(ind+w, absolute_seq_times_s, width=2*w, align='center', color=seq_time_color)
-    ax2.tick_params(axis='y', labelcolor=seq_time_color)
+    ## Plot speedup
+    ax2.set_ylabel('Speedup', color=speedup_color)
+    ax2.set_xlabel('Pipeline')
+    ax2.bar(ind-w, individual_results, width=2*w, align='center', color=speedup_color)
+    ax2.tick_params(axis='y', labelcolor=speedup_color)
+
+    ax1.set_ylabel('Sequential Time (s)', color=seq_time_color)
+    ax1.set_yscale("log")
+    ax1.bar(ind+w, absolute_seq_times_s, width=2*w, align='center', color=seq_time_color)
+    ax1.tick_params(axis='y', labelcolor=seq_time_color)
+
+    ## Plot average line
+    ax2.hlines([1], -1, len(individual_results) + 1, linewidth=0.8, linestyles="dotted")
+    ax2.hlines([mean], -1, len(individual_results) + 1, linewidth=1.2)
+
+    ax1.yaxis.tick_right()
+    ax1.yaxis.set_label_position("right")
+    ax2.yaxis.tick_left()
+    ax2.yaxis.set_label_position("left")
     # plt.yscale("log")
     # plt.yticks(range(1, 18, 2))
     # plt.ylim((0.1, 20))

### 2020-09-21 evaluation/plots/coarse-microbenchmarks-table.tex
Change some one-liner names to not confuse reviewers that they are simple commands

@@
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllll}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
 \toprule
@@
 \midrule
-grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & complex NFA regex \\
+nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & complex NFA regex \\
 sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & \tti{sort}ing \\
@@
 spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & comparisons (\tti{comm}) \\
-diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & non-parallelizable \tti{diff}ing \\
+difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & non-parallelizable \tti{diff}ing \\
 bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & stream shifting and merging \\
-set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & two pipelines merging to a \tti{comm} \\
+set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & two pipelines merging to a \tti{comm} \\
 shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & long \tsta pipeline ending with \tpur \\

### 2020-09-21 compiler/gather_results.py

@@ -51,17 +51,17 @@ coarse_experiments = ["minimal_grep",
                       "shortest_scripts"]
 
 
-pretty_names = {"minimal_grep" : "grep",
+pretty_names = {"minimal_grep" : "nfa-regex",
                 "minimal_sort" : "sort",
                 "wf" : "wf",
                 "topn" : "top-n",
-                "grep" : "grep-light",
+                "grep" : "filter",
                 "bigrams" : "bi-grams",
                 "alt_bigrams" : "bi-grams-opt",
                 "spell" : "spell",
                 "shortest_scripts" : "shortest-scripts",
-                "diff" : "diff",
-                "set-diff" : "set-diff",
+                "diff" : "difference",
+                "set-diff" : "set-difference",
                 "double_sort" : "sort-sort"}
 
 structures = {"minimal_grep" : "$3\\times\\tsta$",
@@ -473,7 +473,7 @@ def format_time_milliseconds(time_milliseconds):
 
 def generate_table_header(full=True):
     header = []
-    header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} lllll}']
+    header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} llllllll}']
     header += ['\\toprule']
     if(not full):
         header += ['Script ~&~ Structure & Input &'

### 2020-09-25 paper/draft.tex
Rename to paper

--- 2020-09-21 pldi/draft.tex
@@
 %% OSDI -- 12 pages limit
-\documentclass[letterpaper,twocolumn,10pt]{article}
-\pdfoutput=1 % for arxiv
-
-\usepackage{usenix2019_v3}
+%% \documentclass[letterpaper,twocolumn,10pt]{article}
+%% \pdfoutput=1 % for arxiv
+%% \usepackage{usenix2019_v3}
+
+%% EuroSys 2021 -- 12 pages limit
+\documentclass[sigplan,review,anonymous]{acmart}
+\acmSubmissionID{TBD}
+\renewcommand\footnotetextcopyrightpermission[1]{}
+
+%% EuroSys rules: Submissions may have at most 12 pages of technical
+%% content, including all text, figures, tables, appendices,
+%% etc. Bibliographic references are not included in the 12-page
+%% limit. Use A4 or US letter paper size, with all text and figures
+%% fitting inside a 178 x 229 mm (7 x 9 in) block centered on the
+%% page, using two columns separated by ≥8 mm (0.33″) of
+%% whitespace. Use ≥10-point font (typeface Times Roman, Linux
+%% Libertine, etc.) on ≥12-point (single-spaced) leading. Graphs and
+%% figures should be readable when printed in grayscale, without
+%% magnification. All pages should be numbered. Authors are encouraged
+%% to hyperlink their references.
 
@@
 
-\begin{document}
-
 %% Title information
@@
 %% Author with single affiliation.
-\author{
-% Anonymous Author(s)\\
-% \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
-  {\rm \normalsize Nikos Vasilakis$^{*}$\qquad Konstantinos Kallas$^{*\dagger}$\qquad  Konstantinos Mamouras$^{\circ}$ }\\
-  {\rm \normalsize Achilleas Benetopoulos$^{\diamond}$ \qquad Lazar Cvetkovich$^{\ddagger}$ }\\
-  {\small MIT, CSAIL \quad $^{\dagger}$University of Pennsylvania \quad $^{\circ}$Rice University \quad $^{\diamond}$Unaffiliated \quad $^{\ddagger}$University of Belgrade}
-}
-% \authornote{$^{*}$Equal contribution.}     %% \authornote is optional;
+
+%% \author{
+%% % Anonymous Author(s)\\
+%% % \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
+%%   {\rm \normalsize Nikos Vasilakis$^{*}$\qquad Konstantinos Kallas$^{*\dagger}$\qquad  Konstantinos Mamouras$^{\circ}$ }\\
+%%   {\rm \normalsize Achilleas Benetopoulos$^{\diamond}$ \qquad Lazar Cvetkovich$^{\ddagger}$ }\\
+%%   {\small MIT, CSAIL \quad $^{\dagger}$University of Pennsylvania \quad $^{\circ}$Rice University \quad $^{\diamond}$Unaffiliated \quad $^{\ddagger}$University of Belgrade}
+%% }
+
+%% \author{Nikos Vasilakis}
+%% \authornote{The two marked authors contributed equally to the paper.}
+%% \affiliation{MIT}
+
+%% \author{Konstantinos Kallas}
+%% \authornotemark[1]
+%% \affiliation{University of Pennsylvania}
+
+%% \author{Konstantinos Mamouras}
+%% \affiliation{Rice University}
+
+%% \author{Achilleas Benetopoulos}
+%% \affiliation{Unaffiliated}
+
+%% \author{Lazar Cvetkovich}
+%% \affiliation{University of Belgrade}
+
+%% \authornote{$^{*}$Equal contribution.}     %% \authornote is optional;
+
 %%                                         %% can be repeated if necessary
@@
 
+\begin{document}
+
 %% 2012 ACM Computing Classification System (CSS) concepts
@@
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
-
 %% Abstract
@@
 
+%% \maketitle
+%% Note: \maketitle command must come after title commands, author
+%% commands, abstract environment, Computing Classification System
+%% environment and commands, and keywords command.
+\maketitle
+
 % Its runtime component provides orchestration and planning support during the execution of the program.
@@
 
-\blfootnote{$^{*}$Equal contribution.}
+%% \blfootnote{$^{*}$Equal contribution.}
 The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    \todo{Pipelines are sorted in descending speedup order}.
+    \todo{Pipelines are sorted in descending speedup order. The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.}
   }

### 2020-09-25 paper/draft.tex
Update paper

@@
 %% EuroSys 2021 -- 12 pages limit
-\documentclass[sigplan,review,anonymous]{acmart}
-\acmSubmissionID{TBD}
+\documentclass[sigplan, review, screen, anonymous]{acmart}
+\acmSubmissionID{49}
 \renewcommand\footnotetextcopyrightpermission[1]{}
+\settopmatter{printfolios=false,printccs=false,printacmref=false}
 
@@
 %% to hyperlink their references.
-
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-% Limit: 25 pages
-% https://icfp20.sigplan.org/track/icfp-2020-papers#Call-for-Papers
-%% \documentclass[acmsmall,10pt,review,anonymous]{acmart}
-%% \settopmatter{printfolios=false,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[sigplan]{acmart}\settopmatter{}
-
-% \acmSubmissionID{248}
 
@@
 
+\todo{Concurrent to our work, POSH is a framework that accelerates shell applications with I/O-heavy components. XXX}
+
 % Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
@@
 
-\subsection*{Acknowledgements}
-% Dumping people so that we don't forget
-We want to thank 
-  Andr\'e DeHon,
-  Ben Karel,
-  Jean-S\'ebastien L\'egar\'e, and 
-  Nick Roessler.
-This material is based upon work supported by DARPA contract no. HR00112020013 (ALRSCC) and NSF award CCF 1763514.
+% \subsection*{Acknowledgements}
+% % Dumping people so that we don't forget
+% We want to thank 
+%   Andr\'e DeHon,
+%   Ben Karel,
+%   Jean-S\'ebastien L\'egar\'e, and 
+%   Nick Roessler.
+% This material is based upon work supported by DARPA contract no. HR00112020013 (ALRSCC) and NSF award CCF 1763514.
+
 %% \grantsponsor{GS100000001}{National Science

### 2020-09-27 pldi/draft.tex
Add average line in Unix50
Update Figure 7 based on our last discussion

@@
       Different configurations per benchmark:
-      (i) Par+Split: \ttt{eager} and general \ttt{split} enabled,
-      (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (iii) Parallel: \ttt{eager} enabled (no \ttt{split}),
-      (v) Blocking Eager, only blocking \ttt{eager} enabled (no \ttt{split}),
-      (iv) No Eager: both \ttt{eager} and \ttt{split} disabled,
-      (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
+      (i) PaSh: the complete implementation with \ttt{eager} and \ttt{split} enabled,
+      % (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
+      (ii) PaSh w/o split: \ttt{eager} enabled (no \ttt{split}),
+      (iii) \todo{Blocking Eager}, only blocking \ttt{eager} enabled (no \ttt{split}),
+      (iv) \todo{No Eager}: both \ttt{eager} and \ttt{split} disabled.
+      % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
       Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
@@
 Fig.~\ref{fig:microbenchmarks} presents \sys's speedup as a function of parallelism (2--64$\times$).
-(Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the best \sys configuration, \ie \ttt{eager} enabled with \ttt{split} when it provides benefit, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are 1.97, 3.5, 5.78, 8.83, 10.96, and 13.47, respectively.
+% (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
+Average speedups of the best \sys configuration, \ie \ttt{eager} enabled with \ttt{split} when it provides benefit, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
+\kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
 
@@
 %   minute).
-
+\kk{Mention bigrams-opt result now that it is removed from the plot.}
 No Eager and Blocking Eager depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.

### 2020-09-27 compiler/gather_results.py

@@ -263,14 +263,16 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, c
     lines = []
     best_result = distr_speedup
 
-    ## In the bigrams experiment we want to also have the alt_bigrams plot
-    if(experiment == "bigrams" and not ("mini-split" in custom_scaleup_plots[experiment])):
-        opt_prefix = '{}/{}_'.format(results_dir, "alt_bigrams")
-        opt_distr_speedup, _ = collect_experiment_speedups(opt_prefix, all_scaleup_numbers)
-        line, = ax.plot(all_scaleup_numbers, opt_distr_speedup, '-P', color='magenta', linewidth=0.5, label='Opt. Parallel')
-        maximum_y = max(maximum_y, max(opt_distr_speedup))
-        lines.append(line)
+    ## Obsolete
+    # ## In the bigrams experiment we want to also have the alt_bigrams plot
+    # if(experiment == "bigrams" and not ("mini-split" in custom_scaleup_plots[experiment])):
+    #     opt_prefix = '{}/{}_'.format(results_dir, "alt_bigrams")
+    #     opt_distr_speedup, _ = collect_experiment_speedups(opt_prefix, all_scaleup_numbers)
+    #     line, = ax.plot(all_scaleup_numbers, opt_distr_speedup, '-P', color='magenta', linewidth=0.5, label='Opt. Parallel')
+    #     maximum_y = max(maximum_y, max(opt_distr_speedup))
+    #     lines.append(line)
 
+    split_exists = False
     if(not (experiment in custom_scaleup_plots) or
        (experiment in custom_scaleup_plots and
         "split" in custom_scaleup_plots[experiment])):
@@ -278,16 +280,17 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, c
             auto_split_distr_speedup = collect_distr_experiment_speedup(prefix,
                                                                         'distr_auto_split.time',
                                                                         all_scaleup_numbers)
-            line, = ax.plot(all_scaleup_numbers, auto_split_distr_speedup, '-D', color='tab:red', linewidth=0.5, label='Par + Split')
+            line, = ax.plot(all_scaleup_numbers, auto_split_distr_speedup, '-D', color='tab:blue', linewidth=0.5, label='Par + Split')
             lines.append(line)
-            split_distr_speedup = collect_distr_experiment_speedup(prefix,
-                                                                   'distr_split.time',
-                                                                   all_scaleup_numbers)
[…]

### 2020-09-27 pldi/draft.tex
Remove manual split from section 5

@@
 This requires the input size to be known beforehand, which is not always the case.
-
-To solve this, % satisfy the tension between efficiency and \todo{knowledge},  % blanking on the right word here
-  \sys supports two \ttt{split} implementations:
-  (i) a general one that can be used with commands that could add or remove lines, and
-  (ii) an optimized one that can be used if the input size is known before its execution.
+To address this, \sys provides a \ttt{split} implementations that first
+consumes its complete input, counts its lines, 
+and then splits it uniformly across the desired number of outputs. 
+% This \ttt{split} can
+% be used with any commands that add or remove lines in their output streams.
+  % (i) a general one that can be used with commands that could add or remove lines, and
+  % (ii) an optimized one that can be used if the input size is known before its execution.
 %% The \sys \todo{backend} support two split node implementations, to
@@
 %% task-based parallelism.
-In the former case, \ttt{split} first consumes its complete input, counts its lines, and then splits it uniformly across the desired number of outputs.
-In the latter case, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
-
+% In the latter case, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
 \sys also inserts eager \ttt{relay} nodes after all \ttt{split} outputs (except the last one) to address laziness concerns.

### 2020-09-27 paper/draft.tex
Minor note

@@
     \todo{Pipelines are sorted in descending speedup order. The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.}
+    \nv{Idea: scripts that only take over 0.01s?}
   }

### 2020-09-27 paper/microbenchmarks-table.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
\toprule
Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
\midrule
nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
filter ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
\bottomrule
\end{tabular*}

### 2020-09-27 paper/draft.tex
Remove < 0.1 unix50 pipelines

@@
 %% (\nv{Pipelines are referred by number, in both the plot and the discussion below.})
-Average speedup is $5.49\times$, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
+Average speedup is \todo{$5.49\times$}, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},
+\kk{We don't include three pipelines that took less than \todo{0.01} seconds to execute because they only modified the first line of their input using \ttt{head}}
 
@@
     \nv{Idea: scripts that only take over 0.01s?}
+    \kk{Now that we removed the small ones, sequential time does not add a lot to this plot. Maybe we can just show speedups?}
   }

### 2020-09-27 compiler/gather_results.py

@@ -595,8 +595,12 @@ def aggregate_unix50_results(all_results, scaleup_numbers):
 
 def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
 
-    ## TODO: Sort by speedup
+    ## Sort by speedup
     all_results.sort(key=lambda x: x[0][scaleup_numbers.index(parallelism)], reverse=True)
+    ## Filter small exec times.
+    all_results = [res for res in all_results
+                   if (res[1][scaleup_numbers.index(parallelism)] / 1000) > 0.1]
+
     ## Plot individual speedups
     individual_results = [distr_exec_speedup[scaleup_numbers.index(parallelism)]
                           for distr_exec_speedup, _ in all_results]
@@ -636,6 +640,7 @@ def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
 
     ax1.set_ylabel('Sequential Time (s)', color=seq_time_color)
     ax1.set_yscale("log")
+    ax1.set_ylim([5, 6000])
     ax1.bar(ind+w, absolute_seq_times_s, width=2*w, align='center', color=seq_time_color)
     ax1.tick_params(axis='y', labelcolor=seq_time_color)
 

### 2020-09-27 paper/draft.tex
Change text to reflect that unix50 scripts with seq time < 0.1 were cut

@@
  % the \todo{geometric mean is $1.08\times$},
-\kk{We don't include three pipelines that took less than \todo{0.01} seconds to execute because they only modified the first line of their input using \ttt{head}}
+\kk{We don't include three pipelines that took less than 0.1 seconds to execute because they only modified the first line of their input using \ttt{head}}
 

### 2020-10-01 paper/draft.tex
Updates on text

@@
 \begin{abstract}
-This paper presents \sys, a shell aimed at parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
-Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
+This paper presents \sys, a system aimed at parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
+% a source-to-source shell compiler and associated runtime library
+Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
 A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
@@
 
-% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
+% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a data-flow graph, enabling for parallelization optimizations.
 % First, a careful study of the parallelizability properties of shell primitives and commands.
@@
 % 
-% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
+% Second, \sys defines a data-flow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the data-flow graph.
 % (one sentence for the model)
@@
   \S\ref{parallelizability}
-    studies the parallelizability of a large set of shell commands and introduces a lightweight annotation language for commands that can be executed in a data-parallel manner.
+    studies the parallelizability of shell commands, and introduces a lightweight annotation language for commands that are executable in a data-parallel manner.
 
   \item
-  \S\ref{ir} presents a dataflow graph model for encoding shell scripts, and a set of parallelization transformations that focus on preserving the semantics of the sequential program.
+  \S\ref{ir} presents a dataflow model and associated parallelization transformations that focus on preserving the semantics of the sequential shell program.
 
@@
 Suppose an environmental scientist    % looking for a non-CS expert---journalist might be too low
-  wants to get a quick sense of trends in the maximum temperature across the U.S. over the past ten years.
+  wants to get a quick sense of trends in the maximum temperature across the U.S. over the past five years.
 As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
@@
 
-The shell script in Fig.~\ref{fig:example} starts by pulling the yearly index files and filtering out URLs that are not part of the compressed dataset.
+Fig.~\ref{fig:example}'s script starts by pulling the yearly index files and filtering out URLs that are not part of the compressed dataset.
 It then downloads and decompresses each file in the remaining set, extracts the values that indicate the temperature, and filters out bogus inputs marked as \ttt{999}.
@@
 % 
-The effort required to write this script is astonishingly low:\footnote{
+The effort expensed writing this script is low:\footnote{
   Some effort is required to understand NOAA's weather format, but this is true for any program processing any dataset.
@@
   its data-processing core amounts to 12 stages and, when expressed as a single line, is only 165 characters long.
-Its succinctness is deceiving---this program is no toy:
-  a Java program implementing only the last four pipeline stages requires 137~LoC~\cite[\S2.1]{hadoop:15}.
+% Its succinctness is deceiving---
+This program is no toy:
+  a Java program implementing only the last four stages takes 137~LoC~\cite[\S2.1]{hadoop:15}.
 To enable such a succinct program composition, \unix incorporates several features.
@@
 construct that allows for task-parallel execution of two commands by
-connecting them with a character stream.
+connecting them through a character stream.
 %% An important \unix feature are pipes (\ttt{|}), introducing
 %% task-parallel program composition over character streams.
-These streams are contiguous character lines separated by newline characters (\textsc{NL}), which delineate individual stream elements.
+This stream is comprised of contiguous character lines separated by newline characters (\textsc{NL}) delineating individual stream elements.
 For example, the first \ttt{grep} is given lines that contain file identifiers, of which it only outputs lines that contain \ttt{gz}, which are in turn consumed by \ttt{tr}.
@@
 Finally, \unix provides an environment for composing commands written in any language.
-Many of the commands used in practice come with the system---\eg commands defined by the POSIX standard or ones part of the GNU Coreutils.
-Others are available as add-ons.
+Many of these commands come with the system---\eg ones defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
 The fact that commands are developed in a variety of languages---including shell scripts---provides users with significant flexibility.
@@
 
-While these features aid in development effort economy through powerful program
-composition, they complicate shell script parallelization, which even
-for simple scripts such as the one in Fig.~\ref{fig:example} create
+While these features aid development-effort economy through powerful program composition, they complicate shell script parallelization, which even for simple scripts such as the one in Fig.~\ref{fig:example} create
 several challenges.
@@
 \caption{
-  \textbf{\sys overview.}
-  A high-level schematic outline of \sys. %  and how this maps on the structure of this paper.
+  \textbf{\sys overview and correspondence to sections.}
+  \sys identifies potentially parallelizable regions~\sx{bg}, 
+  converts them to a dataflow graph~\sx{graph-components}, applies iterative
+  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
+  of the commands used in these regions, and finally emits a parallel script
+  that uses custom primitives~\sx{optimizer}.
 }
@@
 % has been developed through a long history of additions.
-First, it contains constructs that enforce sequential execution;
+First, it contains constructs that enforce sequential execution:
    Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before everything else.
 Moreover, the language semantics only exposes limited task-based parallelism in the form of \ttt{\&}, \ttt{for}, and \ttt{|} constructs.
+% NV: is this really task parallelism? (to be clear: I'm only considering removing "task-based".
 %%   contains constructs with different parallelizability semantics. % that enforce sequential execution.

### 2020-10-03 paper/draft.tex
Move things around

@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new script.
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
 The new script may incorporate parallelism, and is handed off to the user's original shell interpreter % ---\eg \ttt{bash} or \ttt{zsh}---
@@
 
+\sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.
+% nv idea: let's call them may-par regions
+It then runs a series of iterative analyses and transformations within these regions that aim at pushing parallelism as far as possible without risking of breaking semantics.
+Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.
+% , and the technical challenges addressed to achieve performance benefits
+
 \heading{Commands}
@@
 Rather than describing a command's full observable behavior, these classes focus on information that is important for data parallelism.
-To allow commands that fall outside of the standard set to leverage its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class.
-Annotations are expressed once per command rather than once per script and are aimed towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
+To allow other commands to use its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class.
+Annotations are expressed once per command rather than once per script and aim towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
 % We haven't introduced the notion of flags
@@
 
-Fig~\ref{fig:example2} shows a fragment of \sys's output for the script in Fig.~\ref{fig:example}.
-The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
+% Fig~\ref{fig:example2} shows a fragment of \sys's output for the script in Fig.~\ref{fig:example}.
+% The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
 
@@
 describe the graph transformations that \sys performs.
+
+\subsection{Parallelizable Regions}
+The dataflow model and associated transformations are applied only to parallelizable regions.
+% 
+% To optimize a script, \sys translates it to a DFG~\sx{ir}.
+% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
+% 
+%% In addition, the shell's highly dynamic
+%% nature compounds the challenge---\eg by not knowing what files does an
+%% unexpanded command argument refer to.
+
+%% \kk{Give examples of writing to environment variables, producing
+%%   irreversible side effects, the ; and \&\& operators.}
+
+%% To address the first issue, we introduce the notion of parallelizable
+%% program regions and a translation pass for converting them to a
+%% dataflow graph. To address the second issue, we limit dynamic
+%% components by deferring translation as late as possible using a
+%% wrapping scheme.
+
+%% The translation pass identifies distributable regions and
+%% translates each of them to a dataflow graph, provided the abstract
+%% syntax tree (AST) of a shell script.  The AST is produced by
+%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
+
+%
+%% \kk{The goal of this subsection is to explain that Dish does a safe
+%%   but effective analysis to identify which parts of the scripts to
+%%   parallelize.}
+%
+Parallelizable regions are program sub-expressions that \sys deems may be parallelizable safely, \ie without breaking the program, based on the shell constructs they use.
+%% respect to their sequential execution.
+The search for these regions is guided by the shell language and the structure of a particular program.
+These contain information about two types of components:
+  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
+Consider this fragment:
+
+\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+ cat f1 f2 | grep "foo" > f3 && sort f3 
+\end{lstlisting}
+% # Replace sort with sth better
+
+\noindent
+The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---more precisely, they are maximal.
+
+Intuitively, parallelizable regions correspond to sub-expressions of
+the program that would be allowed to execute independently by
+different processes in the POSIX standard~\cite{posix}. Larger
+parallelizable regions can be composed from smaller ones using the
+pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
+Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
+barrier constructs that do not allow parallelizable regions to
+permeate through.
+
+% \tr{I don't know whether I should mention the following: While these
+%   control flow constructs can be parallelized and there has been
+%   research on it, it is orthogonal to our work, and can be
+%   incorporated as future work.}
+% 
+% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
+%   (tr ... f1 > g1 && tr ... f2 >> g1) &
+%   grep ``foo'' g1
+% \end{lstlisting}
+% 
+% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
+%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
+% the dataflow graph that corresponds to the whole program, has two
+% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
+%   g1}.
+%
+
+\heading{Translation Pass}
+%
+%% \kk{It might be a good idea to change this subsection based on the new
+%%   as-late-as-possible architecture where everything is wrapped and
+%%   translated at the final moment. Ideally, this would become the main
+%%   point of this subsection, as the translation algorithm is kind of
+%%   redundant when having the distributable regions too.}
+%
+\sys's front-end performs a depth-first search on the AST of
+the given shell program.  During this pass, it extends the
+parallelizable regions bottom-up, translating their
+independent components to DFG nodes until a barrier
+construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
+original AST where parallelizable regions have been replaced with
+DFGs and calls to \sys's runtime primitives~\sx{optimizer}.
+
+To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
+To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
+If the command is in \sta or \pur, the translation pass initiates a parallelizable region that is propagated up the tree.
+
+%% Finally, whenever the algorithm encounters a command, it first
+%% checks whether it is safe to parallelize it. A command is considered
+%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+% why couldn't we replicate this?
+%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
+%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
+
+%% A few illustrative cases of the analysis algorithm are shown in pseudocode below:
+
+%% \begin{lstlisting}[language=python, float=h]
+%%   def translate(node):
+%%     ...
+%%     elif(node.name == 'Pipe'):
+%%       return pipe_graphs([translate(child)
+%%         for child in node.children])
+%%     elif(node.name == 'And'):
+%%       node.children = [translate(child)
+%%         for child in node.children]
+%%       return node
+%%     elif(node.name == 'Command'):
+%%       if(not safe(node)):
+%%         return make_ir_node(node, 'unsafe')
+%%       else:
+%%         return make_ir_node(node,
+%%           find_category(node))
+%%     ...
+%% \end{lstlisting}
+
+%% Whenever the algorithm encounters a distributable node, it recursively
+%% translates its subcomponents.
+%% It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
+%% % what first and second? we didn't say anything about first and second
+%% Before the different dataflow graphs are connected, the
+%% algorithm checks that at most one node in the graph writes at every
+%% file, and that at most one node reads from a file that another node writes
+%% to. This is important, in order to avoid inconsistent behaviour after
+%% distributing nodes in the dataflow graph (\ie by introducing to
+%% concurrent reads and writes to the same file when
+%% parallelizing).
+
+%% Whenever the algorithm encounters a barrier node,
+%% it recursively translates its subcomponents, similarly to the distributable
+%% case.
+%% However, different from the distributable case, it does not merge them in a large dataflow graph.
+
+%% Finally, whenever the algorithm encounters a command, it first
+%% checks whether it is safe to parallelize it. A command is considered
+%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
+% why couldn't we replicate this?
+%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
+%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
+
+Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
+Examples of such information includes the values of environment variables, strings that have not been expanded, and sub-shell constructs.
+%% to the the safety properties that are mentioned above cannot be
+%% soundly checked statically.
+For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
+It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.
+
+%% \kk{Can we say anything more about the safety considerations? That (i)
+%%   the translation pass ``checks'' that no nodes write/read or
+%%   write/write to the same file (is that necessary?), that (ii) a
+%%   command should not read from a read-once file. (is that necessary?)}
+
+%% \kk{Ideally we would want to talk here about the as-late-as possible
+%%   dish translation that allows us to be safe despite the Shell's
+%%   dynamic nature. I think that this is a great point about the
+%%   work. Maybe we should also focus on implementing it before the OSDI
+%%   deadline.}
 
@@
 
-\section{From Scripts to Graphs and Back Again}
+\section{Runtime Library}
+% FIXME two labels here
 \label{impl}
-
-This section describes key points in the implementation of \sys.
-It first details how its front-end translates a script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
-It then describes its back-end, responsible for generating the parallel script~\sx{optimizer}.
-% , and the technical challenges addressed to achieve performance benefits
-
-\subsection{\sys Front-End}
-\label{front-end}
-
-To optimize a script, \sys translates it to a DFG~\sx{ir}.
-As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
-%% In addition, the shell's highly dynamic
-%% nature compounds the challenge---\eg by not knowing what files does an
-%% unexpanded command argument refer to.
-
-%% \kk{Give examples of writing to environment variables, producing
-%%   irreversible side effects, the ; and \&\& operators.}
-
-%% To address the first issue, we introduce the notion of parallelizable
-%% program regions and a translation pass for converting them to a
-%% dataflow graph. To address the second issue, we limit dynamic
-%% components by deferring translation as late as possible using a
-%% wrapping scheme.
-
-%% The translation pass identifies distributable regions and
-%% translates each of them to a dataflow graph, provided the abstract
-%% syntax tree (AST) of a shell script.  The AST is produced by
-%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
-
-\heading{Parallelizable Regions}
-%
-%% \kk{The goal of this subsection is to explain that Dish does a safe
-%%   but effective analysis to identify which parts of the scripts to
-%%   parallelize.}
-%
-Parallelizable regions are program sub-expressions that can be parallelized safely, \ie without breaking the program.
-%% respect to their sequential execution.
-The search for these regions is guided by the shell language and the structure of a particular program.
-These contain information about two types of components:
-  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
-Consider this fragment:
-
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
- cat f1 f2 | grep "foo" > f3 && sort f3 
-\end{lstlisting}
-% # Replace sort with sth better
-
-\noindent
-The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---more precisely, they are maximal.
-
-Intuitively, parallelizable regions correspond to sub-expressions of
-the program that would be allowed to execute independently by
-different processes in the POSIX standard~\cite{posix}. Larger
-parallelizable regions can be composed from smaller ones using the
-pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
-Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
-barrier constructs that do not allow parallelizable regions to
-permeate through.
-
-% \tr{I don't know whether I should mention the following: While these
-%   control flow constructs can be parallelized and there has been
-%   research on it, it is orthogonal to our work, and can be
-%   incorporated as future work.}
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%   (tr ... f1 > g1 && tr ... f2 >> g1) &
-%   grep ``foo'' g1
-% \end{lstlisting}
-% 
-% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
-%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
-% the dataflow graph that corresponds to the whole program, has two
-% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
-%   g1}.
-%
-
-\heading{Translation Pass}
-%
-%% \kk{It might be a good idea to change this subsection based on the new
-%%   as-late-as-possible architecture where everything is wrapped and
-%%   translated at the final moment. Ideally, this would become the main
-%%   point of this subsection, as the translation algorithm is kind of
-%%   redundant when having the distributable regions too.}
-%
-\sys's front-end performs a depth-first search on the AST of
-the given shell program.  During this pass, it extends the
-parallelizable regions bottom-up, translating their
-independent components to DFG nodes until a barrier
-construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
-original AST where parallelizable regions have been replaced with
-DFGs and calls to \sys's runtime primitives~\sx{optimizer}.
-
-To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
-To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
-If the command is in \sta or \pur, the translation pass initiates a parallelizable region that is propagated up the tree.
-
-%% Finally, whenever the algorithm encounters a command, it first
-%% checks whether it is safe to parallelize it. A command is considered
-%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
-% why couldn't we replicate this?
-%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
-
-%% A few illustrative cases of the analysis algorithm are shown in pseudocode below:
-
-%% \begin{lstlisting}[language=python, float=h]
-%%   def translate(node):
-%%     ...
-%%     elif(node.name == 'Pipe'):
-%%       return pipe_graphs([translate(child)
-%%         for child in node.children])
-%%     elif(node.name == 'And'):
-%%       node.children = [translate(child)
-%%         for child in node.children]
-%%       return node
-%%     elif(node.name == 'Command'):
-%%       if(not safe(node)):
-%%         return make_ir_node(node, 'unsafe')
-%%       else:
-%%         return make_ir_node(node,
-%%           find_category(node))
-%%     ...
-%% \end{lstlisting}
-
-%% Whenever the algorithm encounters a distributable node, it recursively
-%% translates its subcomponents.
-%% It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
-%% % what first and second? we didn't say anything about first and second
-%% Before the different dataflow graphs are connected, the
-%% algorithm checks that at most one node in the graph writes at every
-%% file, and that at most one node reads from a file that another node writes
-%% to. This is important, in order to avoid inconsistent behaviour after
-%% distributing nodes in the dataflow graph (\ie by introducing to
-%% concurrent reads and writes to the same file when
-%% parallelizing).
-
-%% Whenever the algorithm encounters a barrier node,
-%% it recursively translates its subcomponents, similarly to the distributable
-%% case.
-%% However, different from the distributable case, it does not merge them in a large dataflow graph.
-
-%% Finally, whenever the algorithm encounters a command, it first
-%% checks whether it is safe to parallelize it. A command is considered
-%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
-% why couldn't we replicate this?
-%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
-
-Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
-Examples of such information includes the values of environment variables, strings that have not been expanded, and sub-shell constructs.
-%% to the the safety properties that are mentioned above cannot be
-%% soundly checked statically.
-For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
-It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.
-
-%% \kk{Can we say anything more about the safety considerations? That (i)
-%%   the translation pass ``checks'' that no nodes write/read or
-%%   write/write to the same file (is that necessary?), that (ii) a
-%%   command should not read from a read-once file. (is that necessary?)}
-
-%% \kk{Ideally we would want to talk here about the as-late-as possible
-%%   dish translation that allows us to be safe despite the Shell's
-%%   dynamic nature. I think that this is a great point about the
-%%   work. Maybe we should also focus on implementing it before the OSDI
-%%   deadline.}
-
-\subsection{\sys Back End}
 \label{optimizer}

### 2020-10-04 paper/draft.tex
Some comments on the text
Checkpoint changes

@@
 The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---more precisely, they are maximal.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.
+% ---more precisely, they are maximal.
 
@@
 barrier constructs that do not allow parallelizable regions to
-permeate through.
+permeate through.\nv{I will make this more precise---this paragraph is almost the same as overview}
 
@@
 Such laziness inhibits opportunities for parallelism, as commands are often blocked when their consumers are not requesting any input.
-Consider the following script:
+Consider the following fragment:
 
@@
  mkfifo t1 t2
-  grep "foo" > t1 & grep "foo" > t2 &
+ grep "foo" in1 > t1 & 
+ grep "foo" in2 > t2 &
  cat t1 t2
@@
 \noindent
-The \ttt{cat} command will try to consume input from \ttt{t2} after it completes reading from \ttt{t1}.
+The \ttt{cat} command will consume input from \ttt{t2} only after it completes reading from \ttt{t1}.
 As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes (Fig. \ref{fig:eager}a). % processing.
@@
 % replacing FIFOs with files is
-One might be tempted to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
+To solve this, one might be tempted to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
 Aside from severe performance implications, naive replacement can lead to subtle race conditions as a consumer might reach {\sc EOF} before a producer.
 Alternatively, consumers could wait for producers to complete before opening the file for reading (Fig. \ref{fig:eager}c);
-however, this would insert artificial barriers impeding task-based parallelism and wasting disk.
+  however, this would insert artificial barriers impeding task-based parallelism and wasting disk resources---that is, this allows data parallelism at the expense of task parallelism.
 % \unix does not provide an easy way out.
@@
   \sys inserts and instantiates eager \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}d). % nv: not sure what critical meant
-These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
+These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push data to the output stream, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
 % nv: Removed below because I don't think it adds much
@@
 % https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.48\textwidth]{\detokenize{./figs/dish_eager.pdf}}
+\includegraphics[width=0.48\textwidth]{\detokenize{./figs/pash_eager.pdf}}
 \caption{
@@
     (b) files alone introduce race conditions between producer/consumer;
-    (c) files \ttt{wait} inhibit task-based parallelism.
+    (c) files + \ttt{wait} inhibit task parallelism.
 Eager \ttt{relay} nodes (d) address the challenge while remaining within the \sys model.
@@
 \sys implements \emph{aggregate} for several commands in \tpur to enable parallelization.
-A few interesting examples include \emph{aggregate} functions for
+A few interesting examples are \emph{aggregate} functions for
   (i) \ttt{sort}, which amounts to the merge phase of a merge-sort (and on GNU systems is implemented as \ttt{sort -m}),
@@
   (iv) \ttt{wc}, which adds inputs with an arbitrary number of elements (\eg \ttt{wc -lw} or \ttt{wc -lwc} \etc).
-The combiners iterate over the provided stream descriptors, \ie they work with more than two inputs, and apply pure functions at the boundaries of input streams (with the exception of \ttt{sort} that has to interleave inputs).
+The \emph{aggregate} functions iterate over the provided stream descriptors, \ie they work with more than two inputs, and apply pure functions at the boundaries of input streams (with the exception of \ttt{sort} that has to interleave inputs).
 % An important concern has been to maintain whitespace separation while
@@
 \label{eval}
+
+\begin{figure}[t]
+\centering
+% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
+\includegraphics[width=0.4\textwidth]{\detokenize{./figs/pash_lattice.pdf}}
+\caption{
+  \textbf{Lattice of evaluation setup, based on runtime primitives.}
+  \todo{Explain further}
+% Describe. \TODO{Combine this figure with the above code maybe.}
+}
+\vspace{-15pt}
+\label{fig:lattice}
+\end{figure}
 
@@
 % 
-In terms of correctness, \sys's results over multi-GB inputs are identical to the sequential for all benchmarks.
+In terms of correctness, \sys's results over multi-GB, \todo{XXX-line} inputs are identical to the sequential for all benchmarks.
 Scripts feature ample opportunities for breaking semantics, which \sys avoids.

### 2020-10-04 install.sh
Add some python3.8 dependencies

@@ -31,6 +31,10 @@ cd evaluation/tools/
 make
 cd ../../
 
+# Install python3.8 dependencies
+python3.8 -m pip install jsonpickle
+python3.8 -m pip install -U PyYAML
+
 # Generate inputs
 echo "Generating input files"
 echo "TODO: Running gen.sh currently doesn't work."

### 2020-10-04 paper/draft.tex
Comments and improvements in abstract

@@
 \newcommand{\tr}[1]{} %% Text and comments for technical report
+\newcommand{\str}{{\color{red}\textbf{\ttt{*}}}}
 
@@
 Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
+%% \kk{Sys converts a script into several data-flow graphs. Does precision matter here? Should we be super explicit about what it does?}
+A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
 An accompanying parallelizability study of POSIX and GNU commands, two large and commonly used groups, 
-informs \sys's annotation language and results in its equivalent of a data-parallel standard library.
+informs \sys's annotation language and guides the development of a data-parallel standard library.
+% \kk{Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
 \sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes---with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
+% \kk{What are the several axes? We only show speedup. I would propose renaming to significant benefits in execution time.}
 \end{abstract}
@@
 % }
-The complete annotation grammar is shown in Appendix~\ref{alg}.
+The complete annotation grammar is shown in Appendix~\ref{alg}. \str
+% \kk{EuroSys doesn't allow appendix. I think we should just remove the grammar.}
+% \kk{We should mention somewhere that we used the language to annotate N commands.} 
+% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
 
@@
 A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output column(s).
-Its annotation record is shown below:
+Its annotation record is shown below:\str
+% \kk{Update with an excerpt from `comm`.}
 %

### 2020-10-04 paper/draft.tex
Comments and minor modifications in intro

@@
 Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-%% \kk{Sys converts a script into several data-flow graphs. Does precision matter here? Should we be super explicit about what it does?}
+%% \kk{eurosys: Sys converts a script into several data-flow graphs. Does precision matter here? Should we be super explicit about what it does?}
 A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
@@
 informs \sys's annotation language and guides the development of a data-parallel standard library.
-% \kk{Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
+% \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
 \sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes---with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
-% \kk{What are the several axes? We only show speedup. I would propose renaming to significant benefits in execution time.}
+% \kk{eurosys: What are the several axes? We only show speedup. I would propose renaming to significant benefits in execution time.}
 \end{abstract}
@@
 Unfortunately, parallelizing such pipelines requires significant effort shared between two different programmer groups. % it's more like roles personas or "wearing hats"
+% \kk{eurosys: There is a small inconsistency between the abstract and this first paragraph, since we talked about scripts but now we talk about pipelines.
+I would suggest replacing "such pipelines" with "scripts with such pipelines", or just talking about scripts in the first paragraph (removing the word pipelines completely).}
 
@@
  these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
+%  \kk{eurosys: it is not immediately obvious how to use `for` to achieve parallelism. Maybe we can remove it and just leave & and `wait`.}
 % re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
@@
 To top it off, all these options assume a good understanding of parallelism;
-  users whose domain of expertise falls outside computing---routinely writing shell scripts to accomplish their tasks---are left without options.
+  users whose domain of expertise falls outside computing---from hobbyists to data analysts---are left without options.
+% \kk{eurosys: instead of routinely writing shell scripts to accomplish their tasks, I would prefer to focus on the fact that non-computing users are a large subset of all the shell users.}
 %  (see~\S\ref{related} for a full discussion).
@@
 
-To address this challenge, we develop \sys, a shell that extracts data-parallelism from POSIX shell scripts.
+To address this challenge, we develop \sys, a system that extracts data-parallelism from POSIX shell scripts.
 % \kk{We need to mention data-parallelism earlier.
@@
   with a particular emphasis on users.
-Command developers are provided with a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
+Command developers are armed with a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
   rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
 Shell users, on the other hand, are provided with full automation:
-  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
+\sys  analyzes their scripts and extracts latent parallelism.
+% \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
+% \kk{eurosys: I replaced the above to avoid saying that it analyzes parallelizability properties since it feels kinda redundant.}
 % benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
@@
 To address cold-start issues, \sys comes with a library of parallelizability semantics for commands in POSIX and GNU Coreutils.
+% \kk{eurosys: Suggestion: library of annotations, annotation library, annotation thesaurus for \several/ commands in POSIX and GNU... I find the term parallelizability semantics confusing.}
 These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
 The study that led to their characterization also informed \sys's annotation and transformation components.\footnote{
-  In fact, the scope of the study is broader, incorporating also commands outside POSIX and GNU Coreutils such as \ttt{curl} and \ttt{pandoc}.
+  In fact, the scope of the study is broader, also incorporating commands outside POSIX and GNU Coreutils such as \ttt{curl} and \ttt{pandoc}.
 }
@@
 These two components are finally paired with \sys's runtime component.
+% \kk{eurosys: It is not totally clear what the phrase: "These two components" refers to. Maybe we can say that everything is tied together by \sys's runtime component?}
 % \sys's runtime  component additional packs
@@
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as the execution time, the pipeline depth, and the parallelizability classes in which its commands fall.
+Improvements depend on several factors such as the initial script execution time, the pipeline depth, and the parallelizability properties of its commands.
+% \kk{eurosys: the pipeline depth seems to specific to pipelines.
+What we actually want to say is "how much parallelism did the initial script achieve (due to & and |).
+However I think that it is fine mentioning only two factors, "execution time, and the parallelizability properties of its commands."}
+% \kk{eurosys: I removed parallelizability classes since we haven't introduced them until now.}
 
@@
   \item
-  \S\ref{ir} presents a dataflow model and associated parallelization transformations that focus on preserving the semantics of the sequential shell program.
+  \S\ref{ir} presents a dataflow model and associated transformations that expose data parallelism while preserving the semantics of the sequential program.
+  % \kk{eurosys: I changed it since the transformations do no "focus" on preserving correctness.
+and associated transformations that expose data parallelism while preserving the semantics of the sequential program.
+}
 
@@
   \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components and their solutions.
+  % \kk{eurosys: suggestion: details \sys's runtime component, discussing several challenges related to correctness and performance as well as their solutions.
+The reason is that we don't talk about translation anymore and optimization is a bit too general.}
 % \kk{Rewrite that when we know for sure what the section contains.}

### 2020-10-05 paper/draft.tex
Comments and modifications in Section 2

@@
   \textbf{\sys overview and correspondence to sections.}
-  \sys identifies potentially parallelizable regions~\sx{bg}, 
+  \sys identifies potentially \todo{parallelizable regions}~\sx{bg}, 
   converts them to a dataflow graph~\sx{graph-components}, applies iterative
@@
   that uses custom primitives~\sx{optimizer}.
+  % \kk{eurosys: Can we put GNU under POSIX so that the DFG and transformations move towards the center to become more prominent?Also we need to remember to edit the term parallelizable regions and the section in which it appears}
 }
@@
 Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
+% \kk{eurosys: it is not clear why grep's results require concatenation if one hasn't first understood how we introduce parallelism.
+Maybe we can be a bit more verbose by saying "---e.g.
+exposing data parallelism when executing \ttt{sort} would require splitting its input, running sort on each part, and then merging the results---"}
 %% \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
@@
 Moreover, the language semantics only exposes limited task-based parallelism in the form of \ttt{\&}, \ttt{for}, and \ttt{|} constructs.
+% \kk{eurosys: I am not sure it is straightforward why `for` exposes task parallelism.}
 % NV: is this really task parallelism? (to be clear: I'm only considering removing "task-based".
@@
 Any attempt to automate parallelization should be aware of the POSIX shell language and induce as much parallelism as it can without breaking semantics.
+% \kk{eurosys: suggest replacing with: "...
+aware of the POSIX shell language; exposing latent data parallelism without modifying execution semantics." The reason is that I don't like "as much as it can" as a phrase.}
+% \kk{eurosys: I am not sure if I completely see the challenge in this paragraph.
+The only thing I see as challenging is the first sentence talking about ; and sequential composition.
+We should discuss to make sure we make any challenge here shine.}
 
@@
   for execution.
-To expose and exploit data parallelism, % in a given shell script,
-%% parallel execution,
-\sys analyzes individual program fragments, addressing the challenges mentioned earlier.
-
-\sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.
+% To expose and exploit data parallelism, % in a given shell script,
+% \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
+% \kk{eurosys: I deleted this sentence since it doesn't offer a lot now that we have the overview below.}
+An overview of \sys is shown in Fig.~\ref{fig:overview} and described below.
+First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{dataflow-regions} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.
 % nv idea: let's call them may-par regions
-It then runs a series of iterative analyses and transformations within these regions that aim at pushing parallelism as far as possible without risking of breaking semantics.
-Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.
+% \kk{eurosys: I suggest to replace the above sentence with the one below.
+There are a couple reasons.
+First I don't like the term source-to-source compiler because this sentence only describes the shell-graph compilation pass.
+Second, I would like to have some intuition about what there regions are (rather than deferring explanation for later).
+Third each region is tranformed to a DFG and not the whole script.}
+% \sys's \todo{compiler} first identifies sections of the script that lack any synchronization or scheduling constraints and transforms them to a dataflow graph (DFG) model~sx{dataflow-regions}. 
+It then performs a series of graph transformations that aim at exposing parallelism without breaking semantics~\sx{ir:transformations}.
+% \kk{eurosys: I changed this sentence because I didn't like the term analyses, and because it wasn't clear that the transformations are performed on the graph.}
+Finally, \sys converts these DFGs back to a shell script that it augments with its runtime primitives to improve performance~\sx{optimizer}.
+% \kk{eurosys: I changed the last sentence because I don't like the term runtime library.}
 % , and the technical challenges addressed to achieve performance benefits
@@
 \subsection{Parallelizable Regions}
+\label{dataflow-regions}
 The dataflow model and associated transformations are applied only to parallelizable regions.

### 2020-10-05 paper/draft.tex
Sections 1 and 2

--- 2020-10-04 paper/draft.tex
@@
 Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-%% \kk{Sys converts a script into several data-flow graphs. Does precision matter here? Should we be super explicit about what it does?}
-A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
+A set of lightweight annotations allow command developers to express key parallelizability properties about their commands.
 An accompanying parallelizability study of POSIX and GNU commands, two large and commonly used groups, 
-informs \sys's annotation language and guides the development of a data-parallel standard library.
-% \kk{Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
+informs \sys's annotation language and results in its equivalent of a data-parallel standard library.
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
 \sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes---with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
-% \kk{What are the several axes? We only show speedup. I would propose renaming to significant benefits in execution time.}
 \end{abstract}
@@
 
-To address this challenge, we develop \sys, a shell that extracts data-parallelism from POSIX shell scripts.
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/pash_overview.pdf}}
+\caption{
+  \textbf{\sys overview and correspondence to sections.}
+  \sys identifies potentially parallelizable regions~\sx{bg}, 
+  converts them to a dataflow graph~\sx{graph-components}, applies iterative
+  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
+  of the commands used in these regions, and finally emits a parallel script
+  that uses custom primitives~\sx{optimizer}.
+}
+\vspace{-15pt}
+\label{fig:overview}
+\end{figure}
+
+To address this challenge, we develop a system called \sys and outlined in Fig.~\ref{fig:overview} for parallelizing POSIX shell scripts.
 % \kk{We need to mention data-parallelism earlier.
@@
 \sys benefits both programmer groups, % mentioned earlier, 
-  with a particular emphasis on users.
-Command developers are provided with a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
-  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
+  with emphasis on users.
+Command developers are given a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
+  rather than expressing a command's full observable behavior, these annotations focus primarily on its interaction with state.
 Shell users, on the other hand, are provided with full automation:
-  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
+  \sys extracts parallelism latent in a script by analyzing its parallelizability properties. % of their fragments.
 % benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
@@
 These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
-The study that led to their characterization also informed \sys's annotation and transformation components.\footnote{
-  In fact, the scope of the study is broader, incorporating also commands outside POSIX and GNU Coreutils such as \ttt{curl} and \ttt{pandoc}.
-}
+The study that led to their characterization also informed \sys's annotation and transformation components.
+% \footnote{
+%   In fact, the scope of the study is broader, incorporating also commands outside POSIX and Coreutils such as \ttt{builtin}s, \ttt{curl} and \ttt{pandoc}.
+% }
 
@@
 significant advantages in correctness, manual effort,
-and performance---with an order-of-magnitude performance
-improvement due to \sys's transformations and runtime primitives.
+and performance---with 
+speedups $X-X\times$ due to \sys's transformations and $Y-Y\times$ due to \sys's runtime primitives.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as the execution time, the pipeline depth, and the parallelizability classes in which its commands fall.
+Improvements depend on several factors such as execution time, pipeline depth, and the parallelizability classes in which commands fall.
 
@@
   \item
-  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components and their solutions.
-% \kk{Rewrite that when we know for sure what the section contains.}
+  \S\ref{impl} details \sys's library of runtime primitives, discussing the correctness and performance challenges it addresses.
 
@@
 % 
-The effort expensed writing this script is low:\footnote{
-  Some effort is required to understand NOAA's weather format, but this is true for any program processing any dataset.
-}
+The effort expensed writing this script is low:
+% \footnote{
+%   Some effort is required to understand NOAA's weather format, but this is true for any program processing any dataset.
+% }
   its data-processing core amounts to 12 stages and, when expressed as a single line, is only 165 characters long.
@@
 \heading{\unix Features}
-Composition in \unix is primarily achieved through pipes (\ttt{|}), a
+Composition in \unix is primarily achieved with pipes (\ttt{|}), a
 construct that allows for task-parallel execution of two commands by
@@
 This stream is comprised of contiguous character lines separated by newline characters (\textsc{NL}) delineating individual stream elements.
-For example, the first \ttt{grep} is given lines that contain file identifiers, of which it only outputs lines that contain \ttt{gz}, which are in turn consumed by \ttt{tr}.
+For example, Fig~\ref{fig:example}'s first \ttt{grep} outputs (file-name) elements containing \ttt{gz}, which are then consumed by \ttt{tr}.
+% is given lines containing file identifiers and outputs ones that contain \ttt{gz}, 
 % As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
@@
 The fact that commands are developed in a variety of languages---including shell scripts---provides users with significant flexibility.
-For example, one could replace the call to \ttt{sort} and \ttt{head} with one to \ttt{./avg.py} to get the average or standard deviation rather than the maximum.
-The pipeline will still work, as long as \ttt{./avg.py} conforms to the interface outlined earlier.
+For example, one could replace \ttt{sort} and \ttt{head} with \ttt{./avg.py} to get the average rather than the maximum---the pipeline still works, as long as \ttt{./avg.py} conforms to the interface outlined earlier.
 % Shell scripting allows incremental development through refinement
@@
 % }
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-\caption{
-  \textbf{\sys overview and correspondence to sections.}
-  \sys identifies potentially parallelizable regions~\sx{bg}, 
-  converts them to a dataflow graph~\sx{graph-components}, applies iterative
-  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
-  of the commands used in these regions, and finally emits a parallel script
-  that uses custom primitives~\sx{optimizer}.
-}
-\vspace{-15pt}
-\label{fig:overview}
-\end{figure}
 
@@
 \heading{Implementation} %
-Independently of the semantics of individual commands and the shell language, the \unix environment has its own set of quirks.
-Thus, any attempt to orchestrate parallel execution will have to take care of several challenges related to task parallelism, deadlock prevention, and runtime performance.
+On top of command and shell semantics, the \unix environment has its own set of quirks.
+Any attempt to orchestrate parallel execution will hit challenges related to task parallelism, deadlock prevention, and runtime performance.
 For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
@@
 
-\sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.
+\sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} explained below.
 % nv idea: let's call them may-par regions
-It then runs a series of iterative analyses and transformations within these regions that aim at pushing parallelism as far as possible without risking of breaking semantics.
+It then runs a series of iterative analyses and transformations within these regions aimed at pushing parallelism as far as possible without risking breaking semantics.
 Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.

### 2020-10-05 paper/draft.tex

--- 2020-10-05 paper/draft.tex
@@
 
-To address this challenge, we develop \sys, a system that extracts data-parallelism from POSIX shell scripts.
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.49\textwidth]{\detokenize{./figs/pash_overview.pdf}}
+\caption{
+  \textbf{\sys overview and correspondence to sections.}
+  \sys identifies potentially parallelizable regions~\sx{bg}, 
+  converts them to a dataflow graph~\sx{graph-components}, applies iterative
+  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
+  of the commands used in these regions, and finally emits a parallel script
+  that uses custom primitives~\sx{optimizer}.
+  % \kk{eurosys: Can we put GNU under POSIX so that the DFG and transformations move towards the center to become more prominent?Also we need to remember to edit the term parallelizable regions and the section in which it appears}
+}
+\vspace{-15pt}
+\label{fig:overview}
+\end{figure}
+
+To address this challenge, we develop a system called \sys and outlined in Fig.~\ref{fig:overview} for parallelizing POSIX shell scripts.
 % \kk{We need to mention data-parallelism earlier.
@@
 \sys benefits both programmer groups, % mentioned earlier, 
-  with a particular emphasis on users.
-Command developers are armed with a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
-  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
+  with emphasis on users.
+Command developers are given a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
+  rather than expressing a command's full observable behavior, these annotations focus primarily on its interaction with state.
 Shell users, on the other hand, are provided with full automation:
   \sys  analyzes their scripts and extracts latent parallelism.
-% \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
 % \kk{eurosys: I replaced the above to avoid saying that it analyzes parallelizability properties since it feels kinda redundant.}
@@
 
-To address cold-start issues, \sys comes with a library of parallelizability semantics for commands in POSIX and GNU Coreutils.
+To address cold-start issues, \sys comes with a library of parallelizability annotations for commands in POSIX and GNU Coreutils.
 % \kk{eurosys: Suggestion: library of annotations, annotation library, annotation thesaurus for \several/ commands in POSIX and GNU... I find the term parallelizability semantics confusing.}
 These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
-The study that led to their characterization also informed \sys's annotation and transformation components.\footnote{
-  In fact, the scope of the study is broader, also incorporating commands outside POSIX and GNU Coreutils such as \ttt{curl} and \ttt{pandoc}.
-}
-
-These two components are finally paired with \sys's runtime component.
+The study that led to their characterization also informed \sys's annotation and transformation components.
+% \footnote{
+%   In fact, the scope of the study is broader, also incorporating commands outside POSIX and Coreutils such as \ttt{builtin}s, \ttt{curl} and \ttt{pandoc}.
+% }
+
+Everything is tied together with \sys's runtime component.
 % \kk{eurosys: It is not totally clear what the phrase: "These two components" refers to. Maybe we can say that everything is tied together by \sys's runtime component?}
@@
 significant advantages in correctness, manual effort,
-and performance---with an order-of-magnitude performance
-improvement due to \sys's transformations and runtime primitives.
+and performance---with 
+speedups $X-X\times$ due to \sys's transformations and $Y-Y\times$ due to \sys's runtime primitives.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as the initial script execution time, the pipeline depth, and the parallelizability properties of its commands.
+Improvements depend on several factors such as a script's initial execution time, its pipeline depth, and the parallelizability properties of its commands.
 % \kk{eurosys: the pipeline depth seems to specific to pipelines.
@@
   \item
-  \S\ref{impl} details \sys's implementation, discussing several challenges related to \sys's translation, optimization, and runtime components and their solutions.
+  \S\ref{impl} details \sys's library of runtime primitives, discussing the correctness and performance challenges it addresses.
   % \kk{eurosys: suggestion: details \sys's runtime component, discussing several challenges related to correctness and performance as well as their solutions.
@@
 % }
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-\caption{
-  \textbf{\sys overview and correspondence to sections.}
-  \sys identifies potentially \todo{parallelizable regions}~\sx{bg}, 
-  converts them to a dataflow graph~\sx{graph-components}, applies iterative
-  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
-  of the commands used in these regions, and finally emits a parallel script
-  that uses custom primitives~\sx{optimizer}.
-  % \kk{eurosys: Can we put GNU under POSIX so that the DFG and transformations move towards the center to become more prominent?Also we need to remember to edit the term parallelizable regions and the section in which it appears}
-}
-\vspace{-15pt}
-\label{fig:overview}
-\end{figure}
 
@@
   for execution.
+% concurrent version
 % To expose and exploit data parallelism, % in a given shell script,
+% %% parallel execution,
 % \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
+% 
+% \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} explained below.
+% nv idea: let's call them may-par regions
+% It then runs a series of iterative analyses and transformations within these regions aimed at pushing parallelism as far as possible without risking breaking semantics.
+% Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.
+To expose and exploit data parallelism, % in a given shell script,
+  \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
 % \kk{eurosys: I deleted this sentence since it doesn't offer a lot now that we have the overview below.}
-An overview of \sys is shown in Fig.~\ref{fig:overview} and described below.
+% An overview of \sys is shown in Fig.~\ref{fig:overview} and described below.
 First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{dataflow-regions} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.

### 2020-10-05 paper/draft.tex
Teensy fix

@@
 % To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
-\sys aims at parallelizing data-parallel commands, \ie commands that can process their input in parallel, and encodes their characteristics by assigning them to
+\sys aims at parallelizing data-parallel commands, \ie ones that can process their input in parallel, encoding their characteristics by assigning them to
 \emph{parallelizability classes}.

### 2020-10-05 paper/draft.tex
Resolve comments

@@
 Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-%% \kk{eurosys: Sys converts a script into several data-flow graphs. Does precision matter here? Should we be super explicit about what it does?}
 A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
@@
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation, combining several classic and modern Unix scripts, shows significant benefits across several axes---with order-of-magnitude performance improvements stemming from the combination of its program transformations and runtime primitives.
-% \kk{eurosys: What are the several axes? We only show speedup. I would propose renaming to significant benefits in execution time.}
+\sys's extensive evaluation, combining several classic and modern \unix scripts, shows significant \todo{speedups} stemming from the combination of its program transformations and runtime primitives.
 \end{abstract}
@@
 The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
-This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple pipelines for data processing, system orchestration, and other automation tasks.
+This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple scripts for data processing, system orchestration, and other automation tasks.
 % Pipe-centric shell commands are parallel (and to some extent, functional) programs that can natively enjoy significant parallelism on a multicore machine.
 Unfortunately, parallelizing such pipelines requires significant effort shared between two different programmer groups. % it's more like roles personas or "wearing hats"
-% \kk{eurosys: There is a small inconsistency between the abstract and this first paragraph, since we talked about scripts but now we talk about pipelines.
-I would suggest replacing "such pipelines" with "scripts with such pipelines", or just talking about scripts in the first paragraph (removing the word pipelines completely).}
 
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is to use shell primitives (such as \ttt{\&}, \ttt{wait}, \ttt{for}) to explicitly induce parallelism;
+Another option is to use shell primitives (such as \ttt{\&}, \ttt{wait}) to explicitly induce parallelism;
  these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
-%  \kk{eurosys: it is not immediately obvious how to use `for` to achieve parallelism. Maybe we can remove it and just leave & and `wait`.}
 % re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
@@
 To top it off, all these options assume a good understanding of parallelism;
-  users whose domain of expertise falls outside computing---from hobbyists to data analysts---are left without options.
-% \kk{eurosys: instead of routinely writing shell scripts to accomplish their tasks, I would prefer to focus on the fact that non-computing users are a large subset of all the shell users.}
+  users with domain of expertise outside computing---hobbyists to data analysts---are left without options.
 %  (see~\S\ref{related} for a full discussion).
@@
   that uses custom primitives~\sx{optimizer}.
-  % \kk{eurosys: Can we put GNU under POSIX so that the DFG and transformations move towards the center to become more prominent?Also we need to remember to edit the term parallelizable regions and the section in which it appears}
 }
@@
   \sys  analyzes their scripts and extracts latent parallelism.
-% \kk{eurosys: I replaced the above to avoid saying that it analyzes parallelizability properties since it feels kinda redundant.}
 % benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
@@
 To address cold-start issues, \sys comes with a library of parallelizability annotations for commands in POSIX and GNU Coreutils.
-% \kk{eurosys: Suggestion: library of annotations, annotation library, annotation thesaurus for \several/ commands in POSIX and GNU... I find the term parallelizability semantics confusing.}
 These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
@@
 
-Everything is tied together with \sys's runtime component.
-% \kk{eurosys: It is not totally clear what the phrase: "These two components" refers to. Maybe we can say that everything is tied together by \sys's runtime component?}
+The above are tied together with \sys's runtime component.
 % \sys's runtime  component additional packs
@@
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as a script's initial execution time, its pipeline depth, and the parallelizability properties of its commands.
-% \kk{eurosys: the pipeline depth seems to specific to pipelines.
-What we actually want to say is "how much parallelism did the initial script achieve (due to & and |).
-However I think that it is fine mentioning only two factors, "execution time, and the parallelizability properties of its commands."}
-% \kk{eurosys: I removed parallelizability classes since we haven't introduced them until now.}
+Improvements depend on several factors such as a script's initial execution time and the parallelizability properties of its commands.
 
@@
   \S\ref{ir} presents a dataflow model and associated transformations that expose data parallelism while preserving the semantics of the sequential program.
-  % \kk{eurosys: I changed it since the transformations do no "focus" on preserving correctness.
-and associated transformations that expose data parallelism while preserving the semantics of the sequential program.
-}
 
   \item
-  \S\ref{impl} details \sys's library of runtime primitives, discussing the correctness and performance challenges it addresses.
-  % \kk{eurosys: suggestion: details \sys's runtime component, discussing several challenges related to correctness and performance as well as their solutions.
-The reason is that we don't talk about translation anymore and optimization is a bit too general.}
-% \kk{Rewrite that when we know for sure what the section contains.}
+  \S\ref{impl} details \sys's runtime component, discussing the correctness and performance challenges it addresses.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example.pdf}}
+\includegraphics[width=0.47\textwidth]{\detokenize{./figs/dish_example.pdf}}
 \caption{
@@
 % As each command may exercise arbitrary behaviors,
-To be parallelized, each command may require special analysis and treatment---in Fig.~\ref{fig:example}, \ttt{grep}'s results can be combined by simple concatenation, while \ttt{sort}'s results would require careful merging.
+To be parallelized, each command may require special analysis and treatment---\eg exposing data parallelism in Fig.~\ref{fig:example}'s \ttt{sort} would require splitting its input, running \ttt{sort} on each part, and then merging the results.
 Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
-% \kk{eurosys: it is not clear why grep's results require concatenation if one hasn't first understood how we introduce parallelism.
-Maybe we can be a bit more verbose by saying "---e.g.
-exposing data parallelism when executing \ttt{sort} would require splitting its input, running sort on each part, and then merging the results---"}
 %% \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
 %% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
-Manual analysis is also challenging, due to the sheer number of commands, and the many flags that affect their behavior---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called twice with two different sets of flags.
+Manual analysis is also challenging, due to the sheer number of commands and the many flags that affect their behavior---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called with two flag sets.
  % , making correct treatment even more difficult. % nv: we already said impossible so more difficult doesn't add much!
@@
    Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before everything else.
-Moreover, the language semantics only exposes limited task-based parallelism in the form of \ttt{\&}, \ttt{for}, and \ttt{|} constructs.
-% \kk{eurosys: I am not sure it is straightforward why `for` exposes task parallelism.}
+Moreover, the language semantics only exposes limited task-based parallelism in the form of constructs such as \ttt{\&}. % and \ttt{|} 
 % NV: is this really task parallelism? (to be clear: I'm only considering removing "task-based".
@@
   naive parallelization of each loop iteration will miss such opportunities.
-Any attempt to automate parallelization should be aware of the POSIX shell language and induce as much parallelism as it can without breaking semantics.
-% \kk{eurosys: suggest replacing with: "...
-aware of the POSIX shell language; exposing latent data parallelism without modifying execution semantics." The reason is that I don't like "as much as it can" as a phrase.}
+Any attempt to automate parallelization should be aware of the POSIX shell language, exposing latent data parallelism without modifying execution semantics.
 % \kk{eurosys: I am not sure if I completely see the challenge in this paragraph.
@@
 At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
-The new script may incorporate parallelism, and is handed off to the user's original shell interpreter % ---\eg \ttt{bash} or \ttt{zsh}---
-  for execution.
 % concurrent version
@@
 % Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.
-To expose and exploit data parallelism, % in a given shell script,
-  \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
-% \kk{eurosys: I deleted this sentence since it doesn't offer a lot now that we have the overview below.}
 % An overview of \sys is shown in Fig.~\ref{fig:overview} and described below.
-First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{dataflow-regions} by identifying and transforming special regions termed \emph{parallelizable regions} and explained below.
-% nv idea: let's call them may-par regions
+% First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model % ~\sx{dataflow-regions}
+%   by identifying and transforming special regions termed \emph{parallelizable regions} and explained below. % nv idea: let's call them may-par regions
 % \kk{eurosys: I suggest to replace the above sentence with the one below.
@@
 Third each region is tranformed to a DFG and not the whole script.}
-% \sys's \todo{compiler} first identifies sections of the script that lack any synchronization or scheduling constraints and transforms them to a dataflow graph (DFG) model~sx{dataflow-regions}. 
-It then performs a series of graph transformations that aim at exposing parallelism without breaking semantics~\sx{ir:transformations}.
-% \kk{eurosys: I changed this sentence because I didn't like the term analyses, and because it wasn't clear that the transformations are performed on the graph.}
-Finally, \sys converts these DFGs back to a shell script that it augments with its runtime primitives to improve performance~\sx{optimizer}.
-% \kk{eurosys: I changed the last sentence because I don't like the term runtime library.}
-% , and the technical challenges addressed to achieve performance benefits
+\sys first identifies sections of the script that lack synchronization and scheduling constraints and converts them to dataflow graphs (DFGs).
+It then performs a series of DFG transformations aimed at exposing parallelism without breaking semantics. % ~\sx{ir:transformations}.
+Finally, \sys converts these DFGs back to a shell script augmented with \sys-provided commands. % ~\sx{optimizer}.
+The script is handed off to the user's original shell interpreter for execution.
+\sys addresses the aforementioned challenges~\sx{bg:challenges} as below.
 

### 2020-10-05 paper/draft.tex
Pass over 4

@@
 
+\TODO{Add pointers}
+
 \heading{Commands}
@@
 fundamental difference with other DFG models is that \sys's DFG encodes the
-order in which a node reads its inputs, which in turn enables a set of
+order in which a node reads its input streams (not just the order of input elements per stream), which in turn enables a set of
 graph transformations that can be iteratively applied to expose
@@
 %% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
-In \Cref{graph-components} we define the DFG model and describe the
-correspondence with shell scripts, and in \Cref{ir:transformations} we
-describe the graph transformations that \sys performs.
-
-\subsection{Parallelizable Regions}
+We start with the script-to-DFG tranlation~\sx{dataflow-regions}, 
+  then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
+  discuss the transformations that \sys performs~\sx{ir:transformations}, and close with
+  the DFG-to-script translation~\sx{backend}.
+
+\subsection{Font-end: From Shell Scripts to DFGs}
 \label{dataflow-regions}
-The dataflow model and associated transformations are applied only to parallelizable regions.
-% 
+
 % To optimize a script, \sys translates it to a DFG~\sx{ir}.
@@
 %% unexpanded command argument refer to.
-
+% 
 %% \kk{Give examples of writing to environment variables, producing
 %%   irreversible side effects, the ; and \&\& operators.}
-
 %% To address the first issue, we introduce the notion of parallelizable
@@
 %% wrapping scheme.
-
+% 
 %% The translation pass identifies distributable regions and
@@
 %% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
-
+% 
 %
@@
 %%   parallelize.}
-%
-Parallelizable regions are program sub-expressions that \sys deems may be parallelizable safely, \ie without breaking the program, based on the shell constructs they use.
+
+\heading{Parallelizable Reions}
+The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems \emph{may} (but not must) be safely parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
 %% respect to their sequential execution.
 The search for these regions is guided by the shell language and the structure of a particular program.
-These contain information about two types of components:
-  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
+These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
 Consider this fragment:
@@
 The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.\footnote{
+  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands?}
 % ---more precisely, they are maximal.
@@
 pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
-Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
-barrier constructs that do not allow parallelizable regions to
-permeate through.\nv{I will make this more precise---this paragraph is almost the same as overview}
+Conversely, all other operators, including sequential composition (\ttt{;}) and logical operators (\ttt{\&\&}, \ttt{||}), represent
+barrier constructs that do not allow parallelizable regions to expand beyond them.
 
@@
 independent components to DFG nodes until a barrier
-construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
+construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
 original AST where parallelizable regions have been replaced with
@@
 
-\subsection{Definitions}
+\subsection{DFG Definitions}
 \label{graph-components}
@@
 
-We first introduce basic notation used in this section.
+We first introduce the notation used in this section.
 For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
@@
 
+\subsection{Backend: From DFGs to Shell Scripts}
+\label{backend}
+
+After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
+  \sys translates all DFGs back into a shell script.
+Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
+A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
+The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
+
 %% The first two
@@
 
-After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
-  \sys translates all DFGs back into a shell script.
-Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
 This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.

### 2020-10-05 paper/draft.tex
Add minted in the paper code

@@
 \usepackage{picins}
+
+\usepackage{minted} % for code listings
 
@@
   \textbf{\sys overview and correspondence to sections.}
-  \sys identifies potentially parallelizable regions~\sx{bg}, 
+  \sys identifies dataflow regions~\sx{}, 
   converts them to a dataflow graph~\sx{graph-components}, applies iterative
@@
 \centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
+% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
+\begin{minted}[fontsize=\footnotesize]{bash}
+mkfifo $t{0,1...}
+curl $base/$y > $t0 & cat $t0 | split $t1 $t2 &
+cat $t1 | xargs -n1 curl -s >$t3 & 
+cat $t2 | xargs -n1 curl -s >$t4 &
+...
+cat $t9 | sort -rn > $t11 & cat $t10 | sort -rn > $t12 &
+cat $t11 | eager > $t13 & cat $t12 | eager > $t14 &
+sort -m -rn $t13 $t14 > $t15 &
+cat $t15 | head -n1 > $out1 &
+wait $! && ./dash_get_pids | xargs -n 1 kill -SIGPIPE
+\end{minted}
 \caption{
@@
 \emph{parallelizability classes}.
+% \kk{eurosys: I have two small concerns about this sentence but I don't know how to fix them.
+First, there is a redundancy "parallelizing data-parallel".
+Second and most important, I would like this sentence to say that \sys exploits latent data parallelism in commands, but not that it parallelizes data parallel commands.
+The reason is that I would like to make sure that no reviewer thinks that all we do is parallelizing each command on its own (as some OSDI reviewer thought).}
 %% This information is encoded in a command's \emph{parallelizability class}.
@@
   capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
+% \kk{eurosys: This sentence doesn't add much, especially given that we haven't talked about synchronization.}
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (but foregoing any performance improvements).
-Commands can change classes depending on their flags,% we focus on no-flag defaults for now and 
+Commands can change classes depending on their flags,
+% we focus on no-flag defaults for now and 
 which are discussed later~\sx{ext}.
@@
 The latter can express additions or changes to the command's implementation or interface, important as commands are maintained or extended over long periods of time.
+% \kk{eurosys: I have some concerns about the above sentence.
+First, I don't want to say that users developing their own scripts can use the annotation language because this goes against what we mentioned in the intro.
+Second, the last sentence is too prominent the way it is written now.
+I would prefer just saying that the annotations are aimed for command developers who can create an annotation when they develop a new command, or when a command's interface changes---which is important since commands are maintained and extended over long periods of time.}
 
@@
   as some commands consume their inputs in a certain order, this order needs to be maintained in the parallel version of the program.
-For example, consider the parallel version of the expression \ttt{grep "foo" f1 f2}:
-
-\begin{lstlisting}[language=sh, numbers=none]
+For example, consider the data-parallel version of \ttt{grep "foo" f1 f2}:
+
+% \begin{lstlisting}[language=sh, numbers=none]
+%   mkfifo t1 t2
+%   grep "foo" f1 > t1 &
+%   grep "foo" f2 > t2 &
+%   cat t1 t2
+% \end{lstlisting}
+\begin{minted}[fontsize=\footnotesize]{bash}
   mkfifo t1 t2
@@
   cat t1 t2
-\end{lstlisting}
+\end{minted}
 
@@
 
-As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
-For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
+% As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
+% For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
+% \kk{eurosys: I removed this paragraph because it is only a simplistic description of how flags are handled.}
 
@@
 \heading{Example Annotations}
-The parallelizability properties of each command are written in an annotation record that
+The parallelizability properties of each command are specified in an annotation record (a JSON object) that
 %% An annotation record
@@
 Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
-Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
+% Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
 % \footnote{
@@
 % }
-The complete annotation grammar is shown in Appendix~\ref{alg}. \str
+% The complete annotation grammar is shown in Appendix~\ref{alg}. \str
 % \kk{EuroSys doesn't allow appendix. I think we should just remove the grammar.}
-% \kk{We should mention somewhere that we used the language to annotate N commands.} 
-% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
 
@@
 A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output column(s).
-Its annotation record is shown below:\str
-% \kk{Update with an excerpt from `comm`.}
-%
-\begin{lstlisting}[language=sh, numbers=none]
-comm {
-| -1 /\ -3 => (S, [arg[1]], [stdout])
-| -2 /\ -3 => (S, [arg[0]], [stdout])
-| _  => (P, [arg[0], arg[1]], [stdout])
-}
-\end{lstlisting}
+The first clause from its annotation record is shown below:
+%
+\begin{minted}[fontsize=\footnotesize]{json-object}
+  "predicate":
+    { 
+      "operator": "all",
+      "operands": ["-1", "-3"] 
+    },    
+  "class": "stateless",
+  "inputs": ["args[1]"],
+  "outputs": ["stdout"],
+  "options": ["stdin-hyphen"]
+\end{minted}
+% \begin{lstlisting}[language=json, numbers=none]
+%   "predicate":
+%     { 
+%       "operator": "all",
+%       "operands": ["-1", "-3"] 
+%     },    
+%   "class": "stateless",
+%   "inputs": ["args[1]"],
+%   "outputs": ["stdout"],
+%   "options": ["stdin-hyphen"]
+% \end{lstlisting}
+% comm {
+% | -1 /\ -3 => (S, [arg[1]], [stdout])
+% | -2 /\ -3 => (S, [arg[0]], [stdout])
+% | _  => (P, [arg[0], arg[1]], [stdout])
+% }
 
 \noindent
-The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
-As the first (second) and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
+As the first and third columns are suppressed, \ttt{comm} can be categorized in \sta by regarding its first file argument as a static ``configuration'' input;
 % NV: I don't think this is correct, because the small input might be the other argument;  you could even have `comm - -`
   this input accompanies every replicated instance of the command executing in parallel.
-The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
-There, \ttt{comm} is classified as \pur as it needs to % wait until it 
-  read both inputs completely before it can output its results.
-Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.
+The second clause is symmetric to the first with respect to the second input, and the third clause, which is the general case classifies \ttt{comm} as \pur as it needs to read both inputs completely before outputing its results.
+The option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument.
+% The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
+% There, \ttt{comm} is classified as \pur as it needs to % wait until it 
+%   read both inputs completely before it can output its results.
+% Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.
 
@@
 
+\heading{Annotation Effort}
+%
 \sys defines aggregators for many POSIX and GNU commands in \pur, dubbing as both its standard library and an exemplar for community efforts tackling other commands.
+\str
+% \kk{We should mention somewhere that we used the language to annotate N commands.} 
+% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
 

### 2020-10-05 paper/draft.tex
Add minted in the rest of the code

@@
 
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+%  cat f1 f2 | grep "foo" > f3 && sort f3 
+% \end{lstlisting}
+\begin{minted}[fontsize=\footnotesize]{bash}
  cat f1 f2 | grep "foo" > f3 && sort f3 
-\end{lstlisting}
+\end{minted}
 % # Replace sort with sth better
@@
 
-\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+% \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
+%  mkfifo t1 t2
+%  grep "foo" in1 > t1 & 
+%  grep "foo" in2 > t2 &
+%  cat t1 t2
+% \end{lstlisting}
+\begin{minted}[fontsize=\footnotesize]{bash}
  mkfifo t1 t2
@@
  cat t1 t2
-\end{lstlisting}
+\end{minted}
 
@@
 %% \kk{Proposed alternative: ``Consider the following shell script:''}
-\begin{lstlisting}[language=sh, numbers=none]
+% \begin{lstlisting}[language=sh, numbers=none]
+%   mkfifo fifo1 fifo2
+%   cat file-chunk1.txt > fifo1 &
+%   cat file-chunk2.txt > fifo2 &
+%   cat fifo1 fifo2 | head -n 1 & wait
+% \end{lstlisting}
+\begin{minted}[fontsize=\footnotesize]{bash}
   mkfifo fifo1 fifo2
@@
   cat fifo1 fifo2 | head -n 1 & wait
-\end{lstlisting}
+\end{minted}
 \noindent

### 2020-10-05 paper/draft.tex
Add a paragraph on the complete annotation language and the annotation records that we have written and remove the appendix

@@
   mkfifo t1 t2
-  grep "foo" f1 > t1 &
-  grep "foo" f2 > t2 &
-  cat t1 t2
+  grep "foo" f1 > t1 & grep "foo" f2 > t2 & cat t1 t2
 \end{minted}
@@
 
+\heading{Complete Annotation Language}
+%
+The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands---an effort that took about \todo{N} hours.
+An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using combinations of flags.
+Our language is extensible with more operators (as long as the user defines an interpreter for them) and also supports writing arbitrary python code for commands whose properties are difficult to capture, such as the higher order \ttt{xargs} whose parallelizability class depends on the class of the command that it invokes.
+% \str
+% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
+
 \heading{Custom Aggregators}
@@
   (iii) their composition produces the same output as the original command.
-If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
-
-\heading{Annotation Effort}
-%
+\sys can use the \emph{map} and \emph{aggregate} functions in its graph transformations~\sx{ir} to further expose parallelism.
+% If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
 \sys defines aggregators for many POSIX and GNU commands in \pur, dubbing as both its standard library and an exemplar for community efforts tackling other commands.
-\str
-% \kk{We should mention somewhere that we used the language to annotate N commands.} 
-% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
 
@@
 
-\appendix
-
-\section{Annotation Language Grammar}
-\label{alg}
-
-\Cref{fig:dsl} shows the complete grammar of the parallelizability
-annotation language.
-
-\begin{figure}
-  \centering
-  \begin{grammar}
-    <option> ::= `-' <string>
-
-    <category> ::= `stateless' | `pure' | ...
-
-    <maybe-int> ::= | <int>
-
-    <arg> ::= `args[' <int> `]'
-
-    <args> ::= <arg>
-    \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-    <input> ::= `stdin' | <args>
-
-    <inputs> ::= <input>
-    \alt <input> `,' <inputs>
-
-    <output> ::= `stdout' | <arg>
-
-    <outputs> ::= <output>
-    \alt <output> `,' <outputs>
-
-    <option-pred> ::= <option>
-    \alt `value' <option> = <string>
-    \alt `not' <option-pred>
-    \alt <option-pred> `or' <option-pred>
-    \alt <option-pred> `and' <option-pred>
-
-    <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
-
-    <predicate> ::= <option-pred> `=>' <assignment>
-
-    <pred-list> ::= `|' <predicate> <pred-list>
-    \alt `|' `otherwise' `=>' <assignment>
-
-    <command> ::= <name> `\{' <pred-list> `\}'
-
-    <command-list> ::= <command>
-    \alt <command> <command-list>
-  \end{grammar}
-  \caption{
-  \textbf{Parallelizability description language.}
-    The DSL captures important information regarding the parallelizability of a command.
-  }
-  \label{fig:dsl}
-\end{figure}
+% \appendix
+
+% \section{Annotation Language Grammar}
+% \label{alg}
+
+% \Cref{fig:dsl} shows the complete grammar of the parallelizability
+% annotation language.
+
+% \begin{figure}
+%   \centering
+%   \begin{grammar}
+%     <option> ::= `-' <string>
+
+%     <category> ::= `stateless' | `pure' | ...
+
+%     <maybe-int> ::= | <int>
+
+%     <arg> ::= `args[' <int> `]'
+
+%     <args> ::= <arg>
+%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
+
+%     <input> ::= `stdin' | <args>
+
+%     <inputs> ::= <input>
+%     \alt <input> `,' <inputs>
+
+%     <output> ::= `stdout' | <arg>
+
+%     <outputs> ::= <output>
+%     \alt <output> `,' <outputs>
+
+%     <option-pred> ::= <option>
+%     \alt `value' <option> = <string>
+%     \alt `not' <option-pred>
+%     \alt <option-pred> `or' <option-pred>
+%     \alt <option-pred> `and' <option-pred>
+
+%     <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
+
+%     <predicate> ::= <option-pred> `=>' <assignment>
+
+%     <pred-list> ::= `|' <predicate> <pred-list>
+%     \alt `|' `otherwise' `=>' <assignment>
+
+%     <command> ::= <name> `\{' <pred-list> `\}'
+
+%     <command-list> ::= <command>
+%     \alt <command> <command-list>
+%   \end{grammar}
+%   \caption{
+%   \textbf{Parallelizability description language.}
+%     The DSL captures important information regarding the parallelizability of a command.
+%   }
+%   \label{fig:dsl}
+% \end{figure}
 

### 2020-10-05 paper/draft.tex
Change ttt to be mintinline

@@
 \newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
-\newcommand{\ttt}[1]{\texttt{#1}}
+% \newcommand{\ttt}[1]{\texttt{#1}}
+\newcommand{\ttt}[1]{\mintinline{bash}{#1}}
 \newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
@@
 % (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is to use shell primitives (such as \ttt{\&}, \ttt{wait}) to explicitly induce parallelism;
+Another option is to use shell primitives (such as \ttt{&}, \ttt{wait}) to explicitly induce parallelism;
  these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
@@
   \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year.
+  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year. \kk{Make minted.}
 }
@@
    Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before everything else.
-Moreover, the language semantics only exposes limited task-based parallelism in the form of constructs such as \ttt{\&}. % and \ttt{|} 
+Moreover, the language semantics only exposes limited task-based parallelism in the form of constructs such as \ttt{&}. % and \ttt{|} 
 % NV: is this really task parallelism? (to be clear: I'm only considering removing "task-based".
@@
 This analysis is guided by the script structure: % to identify parallelization candidates:
-  some constructs (\eg \ttt{\&}, \ttt{|}) expose task parallelism; others (\eg \ttt{\&\&}, \ttt{||}) enforce synchronization.
+  some constructs (\eg \ttt{&}, \ttt{|}) expose task parallelism; others (\eg \ttt{&&}, \ttt{||}) enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
@@
 
-\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs optimizations.
+\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs parallelization-exposing transformations.
 %% representation that is expressive enough to model data parallelism in shell programs.
@@
 
-\subsection{Font-end: From Shell Scripts to DFGs}
+\subsection{Front-end: From Shell Scripts to DFGs}
 \label{dataflow-regions}
@@
 
-\heading{Parallelizable Reions}
-The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems \emph{may} (but not must) be safely parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
+\heading{Dataflow Regions}
+The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
 %% respect to their sequential execution.
@@
 The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}.\footnote{
-  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands?}
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus dataflow regions, each not extending beyond \ttt{&&}.\footnote{
+  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands?
+\kk{Commands do matter a bit though.
+If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}}
 % ---more precisely, they are maximal.
 
-Intuitively, parallelizable regions correspond to sub-expressions of
+Intuitively, dataflow regions correspond to sub-expressions of
 the program that would be allowed to execute independently by
 different processes in the POSIX standard~\cite{posix}. Larger
-parallelizable regions can be composed from smaller ones using the
-pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
-Conversely, all other operators, including sequential composition (\ttt{;}) and logical operators (\ttt{\&\&}, \ttt{||}), represent
-barrier constructs that do not allow parallelizable regions to expand beyond them.
+dataflow regions can be composed from smaller ones using the
+pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{&}).
+Conversely, all other operators, including sequential composition (\ttt{;}) and logical operators (\ttt{&&}, \ttt{||}), represent
+barrier constructs that do not allow dataflow regions to expand beyond them.
 
@@
 the given shell program.  During this pass, it extends the
-parallelizable regions bottom-up, translating their
+dataflow regions bottom-up, translating their
 independent components to DFG nodes until a barrier
 construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
-original AST where parallelizable regions have been replaced with
+original AST where dataflow regions have been replaced with
 DFGs and calls to \sys's runtime primitives~\sx{optimizer}.
@@
 To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
-If the command is in \sta or \pur, the translation pass initiates a parallelizable region that is propagated up the tree.
+If the command is in \sta or \pur, the translation pass initiates a dataflow region that is propagated up the tree.
 
@@
 
-After translating the parallelizable regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
+After translating the dataflow regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
   \sys translates all DFGs back into a shell script.
@@
 %% \end{wrapfigure}
-\ttt{-{}-parallel} flag set ($S_{GNU}$). % , shown on the side.
+\ttt{--parallel} flag set ($S_{GNU}$). % , shown on the side.
 % The results are shown on the side-figure below.
-While \ttt{-{}-parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{GNU}$'s parallelism is configured to $2\times$ that of $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{-{}-parallelism=127}), 
+While \ttt{--parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
+$S_{GNU}$'s parallelism is configured to $2\times$ that of $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{--parallelism=127}), 
 to account for \sys's additional merge processes.

### 2020-10-05 paper/draft.tex
Update the unix50 plot

@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    \todo{Pipelines are sorted in descending speedup order. The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.}
-    \nv{Idea: scripts that only take over 0.01s?}
-    \kk{Now that we removed the small ones, sequential time does not add a lot to this plot. Maybe we can just show speedups?}
+    \todo{Pipelines are sorted in descending speedup order.} 
+    % The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.
+    % \nv{Idea: scripts that only take over 0.01s?}
+    % \kk{Now that we removed the small ones, sequential time does not add a lot to this plot. Maybe we can just show speedups?}
   }

### 2020-10-05 compiler/gather_results.py

@@ -627,35 +627,41 @@ def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
     fig, ax1 = plt.subplots()
     ax2 = ax1.twinx()  # instantiate a second axes that shares the same x-axis
     # print(fig.get_size_inches())
-    fig.set_size_inches(12, 6)
+    fig.set_size_inches(14, 6)
 
     ## Plot line at 1
     plt.xlim(-1, len(individual_results))
 
     ## Plot speedup
     ax2.set_ylabel('Speedup', color=speedup_color)
-    ax2.set_xlabel('Pipeline')
     ax2.bar(ind-w, individual_results, width=2*w, align='center', color=speedup_color)
     ax2.tick_params(axis='y', labelcolor=speedup_color)
 
     ax1.set_ylabel('Sequential Time (s)', color=seq_time_color)
     ax1.set_yscale("log")
     ax1.set_ylim([5, 6000])
+    ax1.set_xlabel('Script Index')
     ax1.bar(ind+w, absolute_seq_times_s, width=2*w, align='center', color=seq_time_color)
     ax1.tick_params(axis='y', labelcolor=seq_time_color)
 
     ## Plot average line
     ax2.hlines([1], -1, len(individual_results) + 1, linewidth=0.8, linestyles="dotted")
-    ax2.hlines([mean], -1, len(individual_results) + 1, linewidth=1.2)
+    # ax2.hlines([mean], -1, len(individual_results) + 1, linewidth=1.2)
+    mean_xs = range(-1, len(individual_results) + 1)
+    ax2.plot(mean_xs, [mean for _ in mean_xs], linewidth=1.2, label='Mean Speedup')
 
     ax1.yaxis.tick_right()
     ax1.yaxis.set_label_position("right")
     ax2.yaxis.tick_left()
     ax2.yaxis.set_label_position("left")
+    old_ticks = ax2.get_yticks()
+    # print(old_ticks)
+    ax2.set_yticks([1] + old_ticks[:-1])
     # plt.yscale("log")
     # plt.yticks(range(1, 18, 2))
     # plt.ylim((0.1, 20))
-    # plt.legend(loc='lower right')
+    ax2.legend()
+    # ax2.grid()
     # plt.title("Unix50 Individual Speedups")
     plt.tight_layout()
     plt.savefig(os.path.join('../evaluation/plots', "unix50_individual_speedups_{}.pdf".format(parallelism)),bbox_inches='tight')

### 2020-10-05 paper/draft.tex
Change Figure 2 to minted

@@
 \centering
-\includegraphics[width=0.47\textwidth]{\detokenize{./figs/dish_example.pdf}}
+\begin{minted}[fontsize=\footnotesize]{bash}
+base="ftp://ftp.ncdc.noaa.gov/pub/data/noaa";
+for y in {2015..2020}; do
+ curl $base/$y | grep gz | tr -s" " | cut -d" " -f9 |
+ sed "s;^;$base/$y/;" | xargs -n 1 curl -s | gunzip |
+ cut -c 89-92 | grep -iv 999 | sort -rn | head -n 1 |
+ sed "s/^/Maximum temperature for $y is: /"
+done
+\end{minted}
+% \includegraphics[width=0.47\textwidth]{\detokenize{./figs/dish_example.pdf}}
 \caption{

### 2020-10-06 paper/draft.tex
Modify key concerns for annotation language and make them more prominent in our solution

@@
 Such constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by a $fold$ that merges the results.
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by an $aggregate$ that merges the results.
 
@@
 \heading{Key Concerns}
-\sys's annotations focus on four crucial concerns:
+\sys's annotations focus on three crucial concerns:
   (i) a command's parallelizability class,
-  (ii) a command's inputs and outputs,
-  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (iii) for commands with multiple flags (options), the handling of competing sets of flags.
-As the first concern was discussed extensively in the previous section, we now focus on the latter three.
+  (ii) a command's inputs and outputs, and the characteristics of input consumption, and
+  % (ii) for commands that support multiple inputs, the characteristics of input consumption, and
+  (iii) how do flags affect a command's class, inputs, and outputs.
+As the first concern was discussed extensively in the previous section, we now focus on the latter two.
 
@@
 Command flags (options) are a particularly popular way to control a command's execution, directly affecting its parallelizability classification.
-Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
+Commands are thus assigned a default parallelizability class, which should then be refined by the set of flags the command uses.
 For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.
@@
 \heading{Example Annotations}
-The parallelizability properties of each command are specified in an annotation record (a JSON object) that
-%% An annotation record
-contains a set of clauses, each one identified by a predicate on the command options.
-Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
+The parallelizability properties of each command are specified in an annotation record that contains several components, each addressing one of the key concerns.
+On the top level, each annotation record contains a set of clauses, each one identified by a predicate on the command arguments (Concern (iii)). 
+Each clause contains an assignment for the command, indicating its class (Concern (i)), and the sequence of its inputs and outputs (Concern (ii)).
+%  (a JSON object) that
 % Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
@@
 %
-The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands---an effort that took about \todo{N} hours.
-An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using combinations of flags.
+The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands totalling \todo{Y} lines of JSON---an effort that took about \todo{N} hours.
+An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
 Our language is extensible with more operators (as long as the user defines an interpreter for them) and also supports writing arbitrary python code for commands whose properties are difficult to capture, such as the higher order \ttt{xargs} whose parallelizability class depends on the class of the command that it invokes.

### 2020-10-06 paper/draft.tex
Pass over 4

@@
 \heading{Dataflow Regions}
-The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
+In order to apply the graph transformations that expose data parallelism, \sys first has to identify program sub-expressions that can be safely transformed to a dataflow graph, \ie sub-expressions that do not impose any scheduling or synchronization (e.g.
+by using \ttt{;}) and take a set of files as inputs and produce a set of files as outputs.
+% The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
 %% respect to their sequential execution.
@@
   Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands? 
-\kk{Commands do matter a bit though.
-If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}}
+  % \kk{Commands do matter a bit though.
+If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}
+  \kk{As we discussed we can add a small figure that shows the script, the AST, and the DFG.}}
 % ---more precisely, they are maximal.
@@
 original AST where dataflow regions have been replaced with
-DFGs and calls to \sys's runtime primitives~\sx{optimizer}.
+DFGs. 
+% and calls to \sys's runtime primitives~\sx{optimizer}.
+% \kk{eurosys: I removed this since it is slightly inaccurate and readers don't know what the primitives are at that point.}
 
@@
 To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
-If the command is in \sta or \pur, the translation pass initiates a dataflow region that is propagated up the tree.
+If the command is in \sta, \pur, or \npu, the translation pass initiates a dataflow region that is propagated up the tree.
 
@@
 \sys's DFG model represents commands as nodes and streams as edges.
-
-We first introduce the notation used in this section.
+We now introduce the notation used in throughout this section.
 For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
@@
 
-\subsection{Backend: From DFGs to Shell Scripts}
+\subsection{Backend: From DFGs to a Parallel Shell Script}
 \label{backend}
 
-After translating the dataflow regions of the input script to DFGs and applying optimization passes~\sx{ir:transformations},
+After translating the dataflow regions of the input script to DFGs and applying the transformations~\sx{ir:transformations},
   \sys translates all DFGs back into a shell script.
@@
 A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
-The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
+% The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
+% \kk{I removed this, since we describe them just below.}
 
@@
 %%   not exit completely if parallelized.}
-
-\tr{If there is time I can work out a formal definition and a proof
-  sketch why this transformation preserves the output of the dataflow
-  graph.}
 

### 2020-10-06 paper/draft.tex
Pass over section 5 and comments

@@
 A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
+% \kk{What does this "trap" refer to?}
 % The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
@@
 
-\section{Runtime Library}
+\section{Runtime}
 % FIXME two labels here
@@
 
-This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.
+This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's runtime and custom primitives.
 %% it instantiates the \todo{custom} split and relay nodes as highly
@@
 \heading{Overcoming Laziness}
+% \kk{eurosys: Maybe we can call this section pull vs push, and rephrase the terminology to explain the problem as "execution is bounded by the speed in which commands pull their inputs, leading to underutilization".
+Using eager we get rid of this bound prefetch the inputs of the next command.}
 The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
@@
 %% have progress one at a time.
-Such laziness inhibits opportunities for parallelism, as commands are often blocked when their consumers are not requesting any input.
+Such laziness leads to CPU underutilization, as commands are often blocked when their consumers are not requesting any input.
 Consider the following fragment:
@@
  mkfifo t1 t2
- grep "foo" in1 > t1 & 
- grep "foo" in2 > t2 &
- cat t1 t2
+ grep "foo" in1 > t1 & grep "foo" in2 > t2 & cat t1 t2
 \end{minted}
@@
   mkfifo fifo1 fifo2
-  cat file-chunk1.txt > fifo1 &
-  cat file-chunk2.txt > fifo2 &
+  cat in1 > fifo1 & cat in2 > fifo2 &
   cat fifo1 fifo2 | head -n 1 & wait

### 2020-10-07 paper/draft.tex
Checkpoint

--- 2020-10-06 paper/draft.tex
@@
 cat $t11 | eager > $t13 & cat $t12 | eager > $t14 &
-sort -m -rn $t13 $t14 > $t15 &
+sort-merge -rn $t13 $t14 > $t15 &
 cat $t15 | head -n1 > $out1 &
-wait $! && ./dash_get_pids | xargs -n 1 kill -SIGPIPE
+wait $! && ./pash_get_pids | xargs -n 1 kill -SIGPIPE
 \end{minted}
@@
   \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
+  \nv{Added \ttt{sort-merge}, can we sprinke more \sys primitives?}
 }
@@
 Such constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by an $aggregate$ that merges the results.
+  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by a $fold$ that merges the results.
 
@@
 \heading{Key Concerns}
-\sys's annotations focus on three crucial concerns:
+\sys's annotations focus on four crucial concerns:
   (i) a command's parallelizability class,
-  (ii) a command's inputs and outputs, and the characteristics of input consumption, and
-  % (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (iii) how do flags affect a command's class, inputs, and outputs.
-As the first concern was discussed extensively in the previous section, we now focus on the latter two.
+  (ii) a command's inputs and outputs,
+  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
+  (iii) for commands with multiple flags (options), the handling of competing sets of flags.
+As the first concern was discussed extensively in the previous section, we now focus on the latter three.
 
@@
 Command flags (options) are a particularly popular way to control a command's execution, directly affecting its parallelizability classification.
-Commands are thus assigned a default parallelizability class, which should then be refined by the set of flags the command uses.
+Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
 For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.
@@
 \heading{Example Annotations}
-The parallelizability properties of each command are specified in an annotation record that contains several components, each addressing one of the key concerns.
-On the top level, each annotation record contains a set of clauses, each one identified by a predicate on the command arguments (Concern (iii)). 
-Each clause contains an assignment for the command, indicating its class (Concern (i)), and the sequence of its inputs and outputs (Concern (ii)).
-%  (a JSON object) that
+The parallelizability properties of each command are specified in an annotation record (a JSON object) that
+%% An annotation record
+contains a set of clauses, each one identified by a predicate on the command options.
+Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
 % Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
@@
 %
-The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands totalling \todo{Y} lines of JSON---an effort that took about \todo{N} hours.
-An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
+The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands---an effort that took about \todo{N} hours.
+An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using combinations of flags.
 Our language is extensible with more operators (as long as the user defines an interpreter for them) and also supports writing arbitrary python code for commands whose properties are difficult to capture, such as the higher order \ttt{xargs} whose parallelizability class depends on the class of the command that it invokes.

### 2020-10-07 paper/draft.tex
Pass over 6, 6.1

--- 2020-10-06 paper/draft.tex
@@
 % 
-In terms of correctness, \sys's results over multi-GB, \todo{XXX-line} inputs are identical to the sequential for all benchmarks.
+In terms of correctness, \sys's results over multi-GB, \todo{XXX-line}, inputs are identical to the sequential for all benchmarks.
 Scripts feature ample opportunities for breaking semantics, which \sys avoids.
@@
 
-\heading{A Note on the Parallelism Factor}
-%
-The discussion of the results often mentions the parallelism factor, which refers to the \sys \emph{parallelism configuration} rather than the resulting parallelization.
+\heading{Parallelism}
+%
+All of the experiments are run on the same machine and we configure the level of parallelism using \sys's \ttt{--parallel=N} flag that determines how many times will \sys apply its transformations---determining the ``width'' of the parallel script.
+The value of this flag, \ttt{N} is what is shown in the x-axis of \Cref{fig:microbenchmarks} and what we refer to as \emph{parallelism} throughout this section.
+Note that we cannot control the initial parallelism of a script (e.g.
+a command could spawn 10 processes), and therefore parallel scripts produced by \sys often reach maximum parallelization benefits with a value of \ttt{N} that is smaller than the physical cores of the system (in our case 64).
+
+% \heading{A Note on the Parallelism Factor}
+% %
+% The discussion of the results often mentions the parallelism factor, which refers to the \sys \emph{parallelism configuration} rather than the resulting parallelization.
 In fact, for small parallelism configurations \sys often achieves higher parallelization due to the creation of multiple processes---aggregators, splitters, relays \etc
@@
     \caption{
-      \textbf{\sys's speedup for 2--64$\times$-parallelism.}
+      \textbf{\sys's speedup for 2--64$\times$-parallelism (\S\ref{eval}, Parallelism).}
       Different configurations per benchmark:
@@
       % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
-      Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
+      % Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
       % TODO: WHY?
@@
 % (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the best \sys configuration, \ie \ttt{eager} enabled with \ttt{split} when it provides benefit, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
+Average speedups of the optimized \sys, \ie \ttt{eager} and \ttt{split} enabled, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
-\kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
-
+% \kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
+%
 % we experiment with a $200\times$ level of parallelism to see potential overheads
 Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
-It also contains Grep-light (not in Fig.~\ref{fig:microbenchmarks}), designed to show that \sys does not slow down IO-intensive scripts;
-  on the contrary, \sys achieves a 1.5-2.5$\times$ speedup for Grep-light. 
-\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
+It also contains filter and bigrams-opt (not in Fig.~\ref{fig:microbenchmarks}).
+The first (filter) is used to evaluate \sys's impact on IO-intensive scripts (\sys achieves a 1.5-2.5$\times$ speedup).
+The second (bigrams-opt) is an optimized version of bigrams where we grouped 5 commands into one \tpur command thas has a more efficient $map$ and $aggregate$ implementation.
+\sys achieves up to x30 speedup for this script (significantly higher than what is achieved for the standard bigrams Fig.\ref{fig:microbenchmarks}).
+\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.\str
+% \kk{Do we want to talk about cost without mentioning it further?}
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 \heading{Discussion}
-\sys achieves near-linear speedup on low-parallelism configurations.
-This is due to a high degree (deep trees) of task parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
+%
+As discussed in (\S\ref{eval}, Parallelism) \sys does not have any control over the input script parallelism, and therefore it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
+This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.\str
+% \kk{Can we somehow turn this point around and make it seems good/fine.
+That here we report this parallelism, but in the real system, it tries to do a best effort guess of parallelism based on the size of the input script.}
+% \sys achieves near-linear speedup on low-parallelism configurations.
+% This is due to a high degree (deep trees) of task parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
 %% \kk{Ben's point: task parallelism is mentioned here for the first time. Since \sys exposes data parallelism as task parallelism, we might want to mention that somewhere earlier in the paper}
@@
   8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \emph{aggregation} nodes, and 14 \ttt{relay} nodes.
-In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
-significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available---as is the case, \eg  with Sort-sort for $32\times$ and $64\times$ which has 314 and 634 nodes, respectively.
+% In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
+% significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
+% Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available---as is the case, \eg  with Sort-sort for $32\times$ and $64\times$ which has 314 and 634 nodes, respectively.
 % 
@@
 %   minute).
-\kk{Mention bigrams-opt result now that it is removed from the plot.}
-No Eager and Blocking Eager depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-As described in \Cref{optimizer}, these configurations perform worse than \sys with \ttt{eager} in all cases (except for a negligible difference in shortest-scripts).
-Adding \ttt{split} on top of that (Par + Split, Par + B.
-Split) further improves speedup for some scripts that have \tpur or \tnpu commands, and for the rest it does not affect performance.
-Finally, the optimized bi-grams script achieves significantly more benefits from \sys than its unoptimized counterpart, showing that familiarity with \sys's parallelizability classes could go a long way in terms of performance.
+% \kk{Mention bigrams-opt result now that it is removed from the plot.}
+\todo{No Eager} and \todo{Blocking Eager} depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
+As described in \Cref{optimizer}, these eager implementations perform worse than \sys's optimized relay \ttt{eager} in all cases (except for a negligible difference in shortest-scripts).
+The benefits of \ttt{split} are also eveident for scripts that have \tpur or \tnpu commands, and for the rest it does not negatively affect performance.
 

### 2020-10-07 paper/draft.tex
A minor fix in intro

@@
   \textbf{\sys overview and correspondence to sections.}
-  \sys identifies dataflow regions~\sx{}, 
+  \sys identifies dataflow regions~\sx{dataflow-regions}, 
   converts them to a dataflow graph~\sx{graph-components}, applies iterative
@@
   \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year. \kk{Make minted.}
+  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year.
 }

### 2020-10-07 paper/draft.tex
Update pipelines in Unix50

@@
 \heading{Results}
-\sys's parallelism is set to $16\times$, as many of these pipelines are relatively long (\S\ref{ours}, Discussion).
+\sys's parallelism is set to $16\times$, as many of these pipelines are relatively long (\S\ref{eval}, Parallelism).
 Their input was generated by multiplying their original input several times, enough to reach 10GB.
@@
 %% (\nv{Pipelines are referred by number, in both the plot and the discussion below.})
-Average speedup is \todo{$5.49\times$}, median is $6.07\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
+Average speedup is \todo{$6.02\times$}, median is $6.48\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},
-\kk{We don't include three pipelines that took less than 0.1 seconds to execute because they only modified the first line of their input using \ttt{head}}
-
-%% Mean: 5.499164686559543
-%% Median: 6.075774436154179
-%% Geometric Mean: 1.0798740409683736
-%% Weighted Average: 5.752406752503295
+We do not include three pipelines that take less than 0.1 seconds to execute because they only output the first line of their input using \ttt{head}.
+
+  % Mean: 6.028795730618471
+  % Median: 6.48837000168888
+  % Geometric Mean: 4.765602271688225
+  % Weighted Average: 5.752417060139914
 
 \heading{Discussion}
-Most pipelines see significant acceleration, except \todo{13, 24, 25, 26, 29, 30} that see no speedup and \todo{2, 19, 31} that see a slowdown.
-The first group contains commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
+Most pipelines see significant acceleration, except 25-30
+  % \todo{13, 24, 25, 26, 29, 30} 
+  that see no speedup. 
+  % and \todo{2, 19, 31} that see a slowdown.
+These pipelines contain commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
 Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind; a common example is the use of \ttt{awk} for reordering or removing columns of the input.
 %% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the \todo{13th} pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
+In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the 
+  % \todo{13th} 
+  26th pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
 %% , attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
@@
 
-The second group contains \ttt{head}, thus practically processing only one line (regardless of the full input) and completing within $10ms$.
-\sys's slowdown is due to constant setup costs of pipes and processes but execution still remains under $1s$.
+% The second group contains \ttt{head}, thus practically processing only one line (regardless of the full input) and completing within $10ms$. 
+% \sys's slowdown is due to constant setup costs of pipes and processes but execution still remains under $1s$.
 
@@
 For the rest, \sys's speedup is capped due to a combination of reasons:
-  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
-  (ii) some are deep pipelines that already exploit task-based parallelism \todo{(7, 8, 9, 10, 11, 12, 14, 27, 28,)} and
-  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs \todo{(4, 5, 6, 7, 8, 12, 14, 22, 23)}.
+  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} 
+  5, 6, 7, 8, 9, 19, 20, 21, 23, 24,
+  % 24, 9, 7, 6, 8, 23, 20, 19, 21, 5
+  % \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
+  (ii) some are deep pipelines that already exploit task-based parallelism 
+  4, 10, 11, 13, 15, 17, 19, 21, 22,
+  % 17, 4, 13, 15, 10, 22, 11, 19, 21,
+  % \todo{(7, 8, 9, 10, 11, 12, 14, 27, 28,)} 
+  and
+  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs 
+  3, 4, 11, 12, 14, 16, 17, 18, 22.
+  % 14, 3, 16, 17, 4, 22, 11, 12, 18
+  % \todo{(4, 5, 6, 7, 8, 12, 14, 22, 23)}.
 
@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    \todo{Pipelines are sorted in descending speedup order.} 
+    Pipelines are sorted in descending speedup order. 
     % The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.

### 2020-10-07 compiler/gather_results.py
Update plotting script to output the indices of the Unix50 scripts

@@ -596,16 +596,22 @@ def aggregate_unix50_results(all_results, scaleup_numbers):
 def make_unix50_bar_chart(all_results, scaleup_numbers, parallelism):
 
     ## Sort by speedup
-    all_results.sort(key=lambda x: x[0][scaleup_numbers.index(parallelism)], reverse=True)
+    sorted_indices = sorted(range(len(all_results)), key=lambda x:all_results[x][0][scaleup_numbers.index(parallelism)], reverse=True)
+    sorted_all_results = sorted(all_results, key=lambda x: x[0][scaleup_numbers.index(parallelism)], reverse=True)
+    print("Unix50 Par: {} Sorted Indices:".format(parallelism))
+    print("|------------------------") 
+    indices_map = sorted([(v,i) for i, v in enumerate(sorted_indices)], key=lambda x: x[0])
+    print("\n".join(["Old: {}, New: {}".format(v,i) for v, i in indices_map]))
+    print("|------------------------") 
     ## Filter small exec times.
-    all_results = [res for res in all_results
-                   if (res[1][scaleup_numbers.index(parallelism)] / 1000) > 0.1]
+    sorted_all_results = [res for res in sorted_all_results
+                          if (res[1][scaleup_numbers.index(parallelism)] / 1000) > 0.1]
 
     ## Plot individual speedups
     individual_results = [distr_exec_speedup[scaleup_numbers.index(parallelism)]
-                          for distr_exec_speedup, _ in all_results]
+                          for distr_exec_speedup, _ in sorted_all_results]
     absolute_seq_times_s = [absolute_seq[scaleup_numbers.index(parallelism)] / 1000
-                            for _, absolute_seq in all_results]
+                            for _, absolute_seq in sorted_all_results]
     print("Unix50 individual speedups for {} parallelism:".format(parallelism), individual_results)
     mean = sum(individual_results) / len(individual_results)
     median = statistics.median(individual_results)

### 2020-10-07 paper/draft.tex
Complete pash on evaluation

@@
 We consider this a good set of benchmarks because
-  (i) the problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix} and make extensive use of standard commands under a variety of flags, and
-  (ii) the solutions were written by non-experts (contrary to \S\ref{ours}).
+  (i) the problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix}, 
+  (ii) they make extensive use of standard commands under a variety of flags, and
+  (iii) the solutions were written by non-experts (contrary to \S\ref{ours}).
 % ---stressing a patterns that we, \sys's authors, did not
@@
 %   then re-run them).} . 
-\sys executes each pipeline as-is; 
-  if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
-% TODO: talk about correct wrt developer intentions vs. correct wrt problem definition
+\sys executes each pipeline as-is, without any modification.
+  % if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
 
@@
 
-%% \TODO{An important point that needs to be made clear is that these are
-%%   taken from the wild unmodified. And that Pash never leads to
-%%   slowdown (except for the trivial head cases, where the slowdown is
-%%   neligible).}
 \heading{Results}
@@
   2.30/10.79$\times$ speedup (vs. 10m4s) for computing the max.
-(For why speedup is higher than parallelism, see \S\ref{ours}.) % 's Discussion.
+The reason why speedup is higher than parallelism is also also explained in the Discussion of \S\ref{ours}.
 
@@
 to offer significant acceleration.
-A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, which are the focus of conventional parallelization systems.
+A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, \ie the usual focus of conventional parallelization systems.
 
@@
 The original script takes 191min to execute on 1\% of Wikipedia (1.3GB).
-With 2/16$\times$-parallelism, \sys brings it down to 97/15min ($1.97/12.7\times$), with the majority speedup coming from the HTML-to-text conversion.
+With 2/16$\times$-parallelism, \sys brings it down to 97/15min ($1.97/12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion.
 % which uses \ttt{pandoc}, a utility written in Haskell.
@@
 %%   good as it could be.}
-We first compare a single GNU \ttt{sort} optimized by \sys ($S_{PaSh}$) versus the same \ttt{sort} with the 
+We first compare a single GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) versus the same \ttt{sort} with the 
 %% \begin{wrapfigure}[10]{r}{0.25\textwidth}
@@
 $S_{GNU}$'s parallelism is configured to $2\times$ that of $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{--parallelism=127}), 
-to account for \sys's additional merge processes.
+to account for \sys's additional runtime processes.
 % and as such we note that the measurements are harsh for \sys.
@@
   this is because \ttt{eager} adds intermediate buffers %(on disk)
-  between merge phases.
+  that ensure CPU utilization is high.
 %% On the other hand, \ttt{sort --parallel} sorts the input trying to
@@
 Finally, the comparison also shows \sys's benefits to command developers:
-  a low-effort parallelizability annotation allows similar or better scalability than a custom flag added by developers.
+  a low-effort parallelizability annotation achieves better scalability than a custom flag (and underlying parallel implementation) added by developers.
 
@@
 This would lead to 3.2$\times$ performance improvements but severely incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
-\sys's conservative program transformations will not operate on fragments with unclear parallelizability properties.
+\sys's conservative program transformations are not applied in program fragments with unclear parallelizability properties.
 	% or commands whose inputs depend on earlier outputs.

### 2020-10-07 paper/draft.tex
Minor fix for a tilde character

@@
 The shell does not have any visibility into these flags; 
-  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to individual commands.
+  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to individual commands.
 

### 2020-10-07 paper/draft.tex
Add a comment to add separate speedups for compilation + runtime primitives
Fix some issues with minted

@@
 \usepackage{minted} % for code listings
+% \usemintedstyle{xcode}
+% ----- For red boxes ------
+\usepackage{etoolbox}
+\makeatletter
+\AtBeginEnvironment{minted}{\dontdofcolorbox}
+\def\dontdofcolorbox{\renewcommand\fcolorbox[4][]{##4}}
+\makeatother
+% --------------------------
 
@@
 % \newcommand{\ttt}[1]{\texttt{#1}}
-\newcommand{\ttt}[1]{\mintinline{bash}{#1}}
+\newcommand{\ttt}[1]{\mintinline[fontsize=\small]{bash}{#1}}
 \newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
@@
 \heading{Highlights}
-\sys accelerates ($1.92-61.1\times$) almost all scripts ($>50$, with $>200$ commands).
-From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
+\sys accelerates ($1.92-61.1\times$) all scripts ($>50$, with $>200$ commands).
+\kk{Add speedup separated: (i) transformations, (ii) runtime primitives.}
+% From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
 \sys's optimized primitives offer benefits even against manually parallelized scripts. % (such as GNU sort, and GNU parallel).

### 2020-10-07 compiler/gather_results.py
Add no-eager-averages for one-liners

@@ -332,6 +332,7 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, c
             maximum_y = max(maximum_y, max(no_task_par_eager_distr_speedup))
         except ValueError:
             pass
+    no_eager_distr_speedup = None
     if(not (experiment in custom_scaleup_plots) or
        (experiment in custom_scaleup_plots and
         "no-eager" in custom_scaleup_plots[experiment])):
@@ -351,7 +352,10 @@ def collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, c
     ## Set the ylim
     ax.set_ylim(top=maximum_y*1.15)
 
-    return output_diff, lines, best_result
+    # Return the no-eager speedup
+    if(no_eager_distr_speedup is None):
+        no_eager_distr_speedup = distr_speedup
+    return output_diff, lines, best_result, no_eager_distr_speedup
 
 
 def collect_scaleup_times(experiment, results_dir):
@@ -366,7 +370,7 @@ def collect_scaleup_times(experiment, results_dir):
     # ax.plot(all_scaleup_numbers, total_distr_speedup, '-^', linewidth=0.5, label='+ Merge')
     # ax.plot(all_scaleup_numbers, all_scaleup_numbers, '-', color='tab:gray', linewidth=0.5, label='Ideal')
     all_scaleup_numbers = [2, 4, 8, 16, 32, 64]
-    output_diff, _, _ = collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, custom_scaleup_plots, ax)
+    output_diff, _, _, _ = collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, custom_scaleup_plots, ax)
 
     # plt.yscale("log")
     plt.xticks(all_scaleup_numbers[1:])
@@ -816,10 +820,11 @@ def plot_one_liners_tiling(results_dir):
 
     total_lines = []
     averages = [[] for _ in all_scaleup_numbers]
+    no_eager_averages = [[] for _ in all_scaleup_numbers]
     ## Plot microbenchmarks
     for i, experiment in enumerate(experiments):
         ax = fig.add_subplot(gs[i])
-        _, lines, best_result = collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, custom_scaleup_plots, ax)
+        _, lines, best_result, no_eager_result = collect_scaleup_times_common(experiment, all_scaleup_numbers, results_dir, custom_scaleup_plots, ax)
[…]

### 2020-10-08 paper/draft.tex
Resolve comments

--- 2020-10-07 paper/draft.tex
@@
 % Intuitively, a parallelizability 
-A class represents the level of synchronization required by copies of a command executing in parallel, 
-  capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
-% \kk{eurosys: This sentence doesn't add much, especially given that we haven't talked about synchronization.}
+% A class represents the level of synchronization required by copies of a command executing in parallel, 
+%   capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 % For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
-% \kk{eurosys: I removed this paragraph because it is only a simplistic description of how flags are handled.}
 
@@
 % The complete annotation grammar is shown in Appendix~\ref{alg}. \str
-% \kk{EuroSys doesn't allow appendix. I think we should just remove the grammar.}
 
@@
 % and calls to \sys's runtime primitives~\sx{optimizer}.
-% \kk{eurosys: I removed this since it is slightly inaccurate and readers don't know what the primitives are at that point.}
 
@@
 \kk{Add speedup separated: (i) transformations, (ii) runtime primitives.}
+% Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
+% No Eager Averages: [1.6327834370824543, 2.5481043878162444, 3.8623841529643173, 5.931851620007484, 7.466566647963421, 9.35752032439642]
 % From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
@@
 \sys achieves up to x30 speedup for this script (significantly higher than what is achieved for the standard bigrams Fig.\ref{fig:microbenchmarks}).
-\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.\str
-% \kk{Do we want to talk about cost without mentioning it further?}
+\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 As discussed in (\S\ref{eval}, Parallelism) \sys does not have any control over the input script parallelism, and therefore it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.\str
-% \kk{Can we somehow turn this point around and make it seems good/fine.
+This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.
+% \kk{eurosys: Can we somehow turn this point around and make it seems good/fine.
 That here we report this parallelism, but in the real system, it tries to do a best effort guess of parallelism based on the size of the input script.}

### 2020-10-08 paper/draft.tex
Add \ttiny, reword

@@
 \newcommand{\ttt}[1]{\mintinline[fontsize=\small]{bash}{#1}}
-\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
+\newcommand{\ttiny}[1]{\mintinline[fontsize=\footnotesize]{bash}{#1}}
 \newcommand{\tti}[1]{\texttt{\scriptsize #1}}
@@
   \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
-  \nv{Added \ttt{sort-merge}, can we sprinke more \sys primitives?}
 }
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/pash_dfg.pdf}}
+  \vspace{-25pt}
+  \caption{
+    \textbf{From Shell script AST to DFGs.}
+    The AST on the left has two parallelizable regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
+  }
+  \vspace{-15pt}
+  \label{fig:ast-dfg}
+\end{figure}
+
 \subsection{Front-end: From Shell Scripts to DFGs}
@@
 \noindent
-The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
-Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus dataflow regions, each not extending beyond \ttt{&&}.\footnote{
-  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands? 
-  % \kk{Commands do matter a bit though.
-If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}
-  \kk{As we discussed we can add a small figure that shows the script, the AST, and the DFG.}}
-% ---more precisely, they are maximal.
-
+The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently)
+in the standard shell, but \ttt{sort} waits for their completion prior to start
+(Fig.~\ref{fig:ast-dfg}).
+% 
 Intuitively, dataflow regions correspond to sub-expressions of
@@
 For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
-It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.
+It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, \eg  an environment variable passed as an argument to a command does not change its parallelizability class.
 
@@
   (i) streams, \ie files or pipes, that contain data, and
-  (ii) commands that communicate through these streams.
+  (ii) commands, communicating through streams.
 %% \nv{quick solution: change all files to streams and make the equivalence explicit here}
@@
   \textbf{Stateless parallelization transformation.}
-  The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
+  The \ttiny{cat} node is commuted with the stateless node to utilize available data parallelism.
 }
@@
   \textbf{Auxiliary transformations.}
-  These augment the DFG with \ttt{cat}, \ttt{split}, and \ttt{relay} nodes.
+  These augment the DFG with \ttiny{cat}, \ttiny{split}, and \ttiny{relay} nodes.
 }
@@
     (b) files alone introduce race conditions between producer/consumer;
-    (c) files + \ttt{wait} inhibit task parallelism.
-Eager \ttt{relay} nodes (d) address the challenge while remaining within the \sys model.
+    (c) files + \ttiny{wait} inhibit task parallelism.
+Eager \ttiny{relay} nodes (d) address the challenge while remaining within the \sys model.
 % Describe. \TODO{Combine this figure with the above code maybe.}
@@
     Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes aggregators, eager, and split nodes) and compilation time for two indicative parallelization configurations.
+    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes \ttiny{agg}regators, \ttiny{eager}, and \ttiny{split} nodes) and compilation time for two indicative parallelization configurations.
     % Highlights notes a few key features of each script.
@@
 \caption{
-  \textbf{Lattice of evaluation setup, based on runtime primitives.}
-  \todo{Explain further}
+  \textbf{Runtime setup lattice.}
+  Parallel No Eager and Blocking Eager improve over sequential, but are not directly comparable.
+  Eager adds \sys's optimized \ttiny{eager} relay, and \sys uses all primitives in \S\ref{optimizer} (Fig.~\ref{fig:microbenchmarks}).
 % Describe. \TODO{Combine this figure with the above code maybe.}
@@
       Different configurations per benchmark:
-      (i) PaSh: the complete implementation with \ttt{eager} and \ttt{split} enabled,
+      (i) PaSh: the complete implementation with \ttiny{eager} and \ttiny{split} enabled,
       % (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (ii) PaSh w/o split: \ttt{eager} enabled (no \ttt{split}),
-      (iii) \todo{Blocking Eager}, only blocking \ttt{eager} enabled (no \ttt{split}),
-      (iv) \todo{No Eager}: both \ttt{eager} and \ttt{split} disabled.
+      (ii) PaSh w/o split: \ttiny{eager} enabled (no \ttiny{split}),
+      (iii) \todo{Blocking Eager}, only blocking \ttiny{eager} enabled (no \ttiny{split}),
+      (iv) \todo{No Eager}: both \ttiny{eager} and \ttiny{split} disabled.
       % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.

### 2020-10-08 paper/draft.tex
Partial pass over eval

@@
 \heading{Key Concerns}
-\sys's annotations focus on three crucial concerns:
-  (i) a command's parallelizability class,
-  (ii) a command's inputs and outputs, and the characteristics of input consumption, and
+\sys's annotations focus on three crucial concerns about a command:
+  (i) its parallelizability class,
+  (ii) its inputs and outputs, and the characteristics of its input consumption, and
   % (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (iii) how do flags affect a command's class, inputs, and outputs.
-As the first concern was discussed extensively in the previous section, we now focus on the latter two.
+  (iii) how flags affect its class, inputs, and outputs.
+The first concern was discussed extensively in the previous section; we now turn to the latter two.
 
@@
 \heading{Complete Annotation Language}
-%
-The complete annotation language contains 6 operators (one of which supports regural expressions), and we used it to specify the annotation records for \todo{X} commands totalling \todo{Y} lines of JSON---an effort that took about \todo{N} hours.
-An important point is that annotation records can be conservative by default to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
-Our language is extensible with more operators (as long as the user defines an interpreter for them) and also supports writing arbitrary python code for commands whose properties are difficult to capture, such as the higher order \ttt{xargs} whose parallelizability class depends on the class of the command that it invokes.
+The complete annotation language contains 6 operators, one of which supports regural expressions.
+It was used to annotate 47 commands, totalling 708 lines of JSON---an effort that took about 2--3 hours.
+Annotation records are by default conservative so as to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
+The language is extensible with more operators (as long as the \todo{user} defines an \todo{interpreter} for them);
+  it also supports writing arbitrary python code for commands whose properties are difficult to capture---\eg higher order \ttt{xargs}, whose parallelizability class depends on the class of the first-order command that it invokes.
 % \str
@@
 \heading{Stateless and Parallelizable Pure Commands}
-As previously noted~\sx{cmd}, stateless commands such as \ttt{tr}
+Stateless commands such as \ttt{tr}
 operate independently on individual elements of the stream
-(characters, lines, or files) without maintaining any state. In this
+(characters, lines, or files) without maintaining any state~\sx{cmd}. In this
 work, we consider lines as the data quantum, thus we consider stateless only commands that are
@@
 
-After translating the dataflow regions of the input script to DFGs and applying the transformations~\sx{ir:transformations},
-  \sys translates all DFGs back into a shell script.
+After % translating the dataflow regions of the input script to DFGs and 
+  applying transformations~\sx{ir:transformations}, \sys translates all DFGs back into a shell script.
 Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
-A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
+A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning up when the script aborts.
 % \kk{What does this "trap" refer to?}
@@
 % \kk{I removed this, since we describe them just below.}
+% \nv{Eurosys: Exactly, I had the last sentence as a segway connecting the two sections. There is not connection in the first paragraph of 5!}
 
@@
 \sys accelerates ($1.92-61.1\times$) all scripts ($>50$, with $>200$ commands).
-\kk{Add speedup separated: (i) transformations, (ii) runtime primitives.}
+For some of our benchmarks~\sx{ours}, we decouple the performance benefits stemming from \sys's compiler alone from the ones stemming from \sys's runtime primitives:
+  the compiler's average speedup seen is $9.3\times$, jumping to $13.2\times$ when runtime optimizations are used.
+In a different set of $30$ benchmarks parallelized $16\times$\footnote{
+  \nv{Immediate question: why use $64\times$ in one and $16\times$ in another?}
+}
+ see an average speedup of $6\times$, depending on the class of piecewise commands~\sx{unix50}.
+Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see $2.44\times$ and $12.7\times$ speedup on $16\times$-parallel configuration.
+\sys outperforms \ttt{sort --parallel} by a significant margin, and offers small performance benefits over GNU \ttt{parallel}, which returns incorrect results in one configuration~\sx{micro}.
 % Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
@@
 % From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
-\sys's optimized primitives offer benefits even against manually parallelized scripts. % (such as GNU sort, and GNU parallel).
+% \sys's optimized primitives offer benefits even against manually parallelized scripts. % (such as GNU sort, and GNU parallel).
 % \sys's compilation pass is negligible (avg. $<1s$). %  for non trivial program execution.
-In two large cases, tasks such as data download, extraction, and preprocessing---often implemented using shell scripts and outside the focus of specialized parallelization frameworks---take comparable time to the execution of the main computation, and can be significantly accelerated by \sys.
+% In two large cases, tasks such as data download, extraction, and preprocessing---often implemented using shell scripts and outside the focus of specialized parallelization frameworks---take comparable time to the execution of the main computation, and can be significantly accelerated by \sys.
 % In terms of manual effort,
 
-The vast majority requires no effort other than invoking \sys;
-  the annotation for the remaining 6 commands ($<3\%$) amounts to a single record.
-% 
-In terms of correctness, \sys's results over multi-GB, \todo{XXX-line}, inputs are identical to the sequential for all benchmarks.
+The vast majority of programs require no effort to parallelize other than invoking \sys;
+  only 6 commands ($<3\%$) needed a single-record annotation.
+% 
+In terms of correctness, \sys's results over multi-GB  inputs are identical to the sequential for all benchmarks.
 Scripts feature ample opportunities for breaking semantics, which \sys avoids.
@@
 %
-All of the experiments are run on the same machine and we configure the level of parallelism using \sys's \ttt{--parallel=N} flag that determines how many times will \sys apply its transformations---determining the ``width'' of the parallel script.
-The value of this flag, \ttt{N} is what is shown in the x-axis of \Cref{fig:microbenchmarks} and what we refer to as \emph{parallelism} throughout this section.
-Note that we cannot control the initial parallelism of a script (e.g.
-a command could spawn 10 processes), and therefore parallel scripts produced by \sys often reach maximum parallelization benefits with a value of \ttt{N} that is smaller than the physical cores of the system (in our case 64).
+% All experiments are run on the same machine and 
+% \nv{Eurosys: This /\ is described above}
+The level of parallelism is configured using \sys's \ttt{--parallel=N}, a flag that determines the number of times that \sys applies its transformations---determining the ``width'' of the parallel script.
+The value  of \ttt{N} is shown as the x-axis of Fig.\ref{fig:microbenchmarks}, called \emph{parallelism} throughout this section.
+We cannot control the initial parallelism of a script (\eg a command could spawn 10 processes), and thus parallel scripts produced by \sys often reach maximum parallelization benefits with a value of \ttt{N} that is smaller than the physical cores of the hardware setup (in our case 64).
 
@@
 %>  parallel because they are all close to 1 (no speedup).}
-Tab.~\ref{tab:eval} summarizes the collection of programs shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
+Tab.~\ref{tab:eval} summarizes the first collection of programs.
+%  shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
 Grep and Grep-light are in \sta, with the former centered around an expensive DFA-based backtracking expression.
@@
 In a recent celebration of \unix's 50-year legacy, Bell Labs created 37 challenges~\cite{unix50} solvable by means of composing \unix pipelines.
+The problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix}.
 We found unofficial solutions to all-but-three problems on GitHub~\cite{unix50sol}, expressed as pipelines with 2--12 stages (avg.: 5.58).
-We consider this a good set of benchmarks because
-  (i) the problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix}, 
-  (ii) they make extensive use of standard commands under a variety of flags, and
-  (iii) the solutions were written by non-experts (contrary to \S\ref{ours}).
+They make extensive use of standard commands under a variety of flags, and appear to be written by non-experts (contrary to \S\ref{ours}, they often use sub-optimal or non-\unix-y constructs).
 % ---stressing a patterns that we, \sys's authors, did not
@@
 Their input was generated by multiplying their original input several times, enough to reach 10GB.
-Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all 34 pipelines. We use the pipeline index to refer to them in both plots and discussion.
+Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all \todo{30} pipelines. We use the pipeline index to refer to them in both plots and discussion.
 %% (\nv{Pipelines are referred by number, in both the plot and the discussion below.})
@@
   2.30/10.79$\times$ speedup (vs. 10m4s) for computing the max.
-The reason why speedup is higher than parallelism is also also explained in the Discussion of \S\ref{ours}.
+The reason why speedup is higher than parallelism is explained at the beginning of \S\ref{eval} (Parallelism).\footnote{
+  Can we avoid this by not showing the $2\times$? Why show $2\times$?
+}
 
@@
 extracts the text from HTML, and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
-In total, it contains 34 commands
+It totals 34 commands written in multiple programming languages.
 %% Collectively, they contain a total of 52 stages (first: 18; second: 34)
-written in multiple programming languages.
 
@@
 \heading{GNU Parallel}
-We compare \sys to GNU \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}, on a small bio-informatics script.
+We compare \sys to \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}, on a small bio-informatics script.
 % Using its \ttt{--jobs} flag, we  redirect input from a file to 
@@
 \heading{Parallel Shell Scripting}
-Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}---are limited to embarrassingly parallel (and short) programs and are predicated upon explicit and careful user invocation:
+Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, % \textsc{GNU} 
+  \ttt{parallel}~\cite{Tange2011a}---are limited to embarrassingly parallel (and short) programs and are predicated upon explicit and careful user invocation:
   users have to navigate through a vast array of different configurations, flags, and modes of invocation to achieve parallelization without jeopardizing correctness.
@@
 
-\todo{Concurrent to our work, POSH is a framework that accelerates shell applications with I/O-heavy components. XXX}
+Concurrent to our work, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
+POSH focuses on colocating distributed operation, and introduces a novel scheduling algorithm. 
+\sys instead attempts non-distributed parallelism, is location-agnostic, and focuses on correctness through its dataflow transformations.
 
@@
 
-This paper presented \sys, a shell variant that parallelizes shell programs mostly automatically.
-\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
-To achieve its goal, \sys
-  decomposes primitives into parallelizability classes,
-  identifies high-parallelizability stages,
-  applies a series of transformations according to a DFG model,
-  and orchestrates the execution of the resulting parallel program.
-%
+This paper presented \sys, a system allowing shell users to parallelize shell programs mostly automatically.
+% a source-to-source shell compiler and associated runtime library
+\sys can be thought as a parallelizing source-to-source compiler coupled with a runtime component.
+The compiler operates by converting parallelizable regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
+Characteristics are known by the  use of classification and annotations (for certain classes), which \sys offers for commands in POSIX and Coreutils used frequently.
+\sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
+Applying \sys to classic and modern \unix scripts shows significant speedups---in one set of benchmarks, the compiler's average speedup is $9.3\times$, jumping to $13.2\times$ due to runtime optimizations.
+\sys is open source and available for download at \ttt{blinded}.
+
 % Its runtime component provides orchestration and planning support during the execution of the program.
@@
 % Despite limitation related to expansions,
-  \sys can lead to significant benefits for shell users.
-Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all with minimal or zero user effort.
+%   \sys can lead to significant benefits for shell users.
+% Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all with minimal or zero user effort.
 

### 2020-10-08 paper/draft.tex
Corrections

@@
-%% OSDI -- 12 pages limit
+% TODO Add Ignis and Mozart
 %% \documentclass[letterpaper,twocolumn,10pt]{article}
@@
   classified) can be represented with one, two clauses etc. If we have
-  a different solution (like python function that given a command and
+  a different solution (like Python function that given a command and
   its arguments returns the category) then we could talk about how
@@
   should be a backup mechanism, when the language is not expressive
-  enough, like a python function that categorizes the command, and
+  enough, like a Python function that categorizes the command, and
   returns its input argument if it is stateless.}}
@@
   language, and why don't we just allow someone to write a function in
-  python for each command, that given the command and its arguments,
+  Python for each command, that given the command and its arguments,
   returns whether the command category. Possible arguments include,
@@
   should be shareable among users, it would be bad to execute
-  arbitrary python code, so using this language, expressivity is
+  arbitrary Python code, so using this language, expressivity is
   limited. All of these arguments are a little bit weak though. Niko,
@@
 The language is extensible with more operators (as long as the \todo{user} defines an \todo{interpreter} for them);
-  it also supports writing arbitrary python code for commands whose properties are difficult to capture---\eg higher order \ttt{xargs}, whose parallelizability class depends on the class of the first-order command that it invokes.
+  it also supports writing arbitrary Python code for commands whose properties are difficult to capture---\eg higher order \ttt{xargs}, whose parallelizability class depends on the class of the first-order command that it invokes.
 % \str
@@
   could then go to the implementation? Or maybe say that one can write
-  a python function that given a node of the graph, transforms it into
+  a Python function that given a node of the graph, transforms it into
   many. I am not sure what is best...}}
@@
 \heading{Highlights}
-\sys accelerates ($1.92-61.1\times$) all scripts ($>50$, with $>200$ commands).
+\sys accelerates ($1.92-61.1\times$) all scripts.\footnote{
+  \nv{But this is not true; ``the vast majority''? maybe put 51/54 scripts?}.
+}
 For some of our benchmarks~\sx{ours}, we decouple the performance benefits stemming from \sys's compiler alone from the ones stemming from \sys's runtime primitives:
@@
 
-The vast majority of programs require no effort to parallelize other than invoking \sys;
+The vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
   only 6 commands ($<3\%$) needed a single-record annotation.
@@
 We evaluate \sys under multiple configurations on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
+% \nv{Eurosys: goal of experiment: decouple compiler from runtime, and talk about all overheads}
 
@@
 % \kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
-%
+
 % we experiment with a $200\times$ level of parallelism to see potential overheads
 Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
-It also contains filter and bigrams-opt (not in Fig.~\ref{fig:microbenchmarks}).
-The first (filter) is used to evaluate \sys's impact on IO-intensive scripts (\sys achieves a 1.5-2.5$\times$ speedup).
-The second (bigrams-opt) is an optimized version of bigrams where we grouped 5 commands into one \tpur command thas has a more efficient $map$ and $aggregate$ implementation.
-\sys achieves up to x30 speedup for this script (significantly higher than what is achieved for the standard bigrams Fig.\ref{fig:microbenchmarks}).
+It also contains Filter and Bigrams-opt (not in Fig.~\ref{fig:microbenchmarks}).
+Filter is used to evaluate \sys's impact on IO-intensive scripts (\sys achieves a 1.5--2.5$\times$ speedup).
+Bigrams-opt is an optimized version of Bigrams that groups 5 commands into one \pur command and thus has a more efficient $map$ and $aggregate$ implementation.
+\sys achieves up to $30\times$ speedup for this script (significantly higher than what is achieved for the standard Bigrams Fig.\ref{fig:microbenchmarks}).
+% \nv{Eurosys: I think this should be part of a "do custom aggregators help?" question, not the main results}.
 \sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
@@
 %
-As discussed in (\S\ref{eval}, Parallelism) \sys does not have any control over the input script parallelism, and therefore it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
+\todo{These results explain a detail before getting the main points across: A is better than B, C is worse except G}
+As discussed earlier (\S\ref{eval}, Parallelism), \sys does not have control over the input script's parallelism.
+Thus, it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
 This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.
@@
 \todo{No Eager} and \todo{Blocking Eager} depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-As described in \Cref{optimizer}, these eager implementations perform worse than \sys's optimized relay \ttt{eager} in all cases (except for a negligible difference in shortest-scripts).
-The benefits of \ttt{split} are also eveident for scripts that have \tpur or \tnpu commands, and for the rest it does not negatively affect performance.
+As described in Fig.\ref{optimizer}, these eager implementations perform worse than \sys's optimized \ttt{eager} relay in all cases (except for a negligible difference in Shortest-Scripts).
+A \ttt{split} benefits scripts with \tpur or \tnpu commands, but does not negatively affect the rest.
 
@@
 The reason why speedup is higher than parallelism is explained at the beginning of \S\ref{eval} (Parallelism).\footnote{
-  Can we avoid this by not showing the $2\times$? Why show $2\times$?
+  \nv{Can we avoid this by not showing the $2\times$? Why show $2\times$?} % <--- Eurosys
 }

### 2020-10-08 paper/draft.tex
Add a note for a future version of PaSh

--- 2020-10-08 paper/draft.tex
@@
 \newcommand{\ttt}[1]{\mintinline[fontsize=\small]{bash}{#1}}
-\newcommand{\ttiny}[1]{\mintinline[fontsize=\footnotesize]{bash}{#1}}
+\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
 \newcommand{\tti}[1]{\texttt{\scriptsize #1}}
@@
 % Intuitively, a parallelizability 
-% A class represents the level of synchronization required by copies of a command executing in parallel, 
-%   capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
+A class represents the level of synchronization required by copies of a command executing in parallel, 
+  capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
+% \kk{eurosys: This sentence doesn't add much, especially given that we haven't talked about synchronization.}
 \sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
@@
 % For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
+% \kk{eurosys: I removed this paragraph because it is only a simplistic description of how flags are handled.}
 
@@
 % The complete annotation grammar is shown in Appendix~\ref{alg}. \str
+% \kk{EuroSys doesn't allow appendix. I think we should just remove the grammar.}
 
@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/pash_dfg.pdf}}
-  \vspace{-25pt}
-  \caption{
-    \textbf{From Shell script AST to DFGs.}
-    The AST on the left has two parallelizable regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
-  }
-  \vspace{-15pt}
-  \label{fig:ast-dfg}
-\end{figure}
-
 \subsection{Front-end: From Shell Scripts to DFGs}
@@
 \noindent
-The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently)
-in the standard shell, but \ttt{sort} waits for their completion prior to start
-(Fig.~\ref{fig:ast-dfg}).
-% 
+The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
+Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus dataflow regions, each not extending beyond \ttt{&&}.\footnote{
+  Can we make the commands \ttt{cmd1}, \ttt{cmd2}, \etc to ensure the readers understand this is driven by structure and not commands? 
+  % \kk{Commands do matter a bit though.
+If a command is sideeffectful we don't know how to connect it in the dataflow graph and thus \sys would not parallelize it (as is written later in the translation pass).}
+  \kk{As we discussed we can add a small figure that shows the script, the AST, and the DFG.}}
+% ---more precisely, they are maximal.
+
 Intuitively, dataflow regions correspond to sub-expressions of
@@
 % and calls to \sys's runtime primitives~\sx{optimizer}.
+% \kk{eurosys: I removed this since it is slightly inaccurate and readers don't know what the primitives are at that point.}
 
@@
 For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
-It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, \eg  an environment variable passed as an argument to a command does not change its parallelizability class.
+It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.
 
@@
   (i) streams, \ie files or pipes, that contain data, and
-  (ii) commands, communicating through streams.
+  (ii) commands that communicate through these streams.
 %% \nv{quick solution: change all files to streams and make the equivalence explicit here}
@@
   \textbf{Stateless parallelization transformation.}
-  The \ttiny{cat} node is commuted with the stateless node to utilize available data parallelism.
+  The \texttt{cat} node is commuted with the stateless node to utilize available data parallelism.
 }
@@
   \textbf{Auxiliary transformations.}
-  These augment the DFG with \ttiny{cat}, \ttiny{split}, and \ttiny{relay} nodes.
+  These augment the DFG with \ttt{cat}, \ttt{split}, and \ttt{relay} nodes.
 }
@@
     (b) files alone introduce race conditions between producer/consumer;
-    (c) files + \ttiny{wait} inhibit task parallelism.
-Eager \ttiny{relay} nodes (d) address the challenge while remaining within the \sys model.
+    (c) files + \ttt{wait} inhibit task parallelism.
+Eager \ttt{relay} nodes (d) address the challenge while remaining within the \sys model.
 % Describe. \TODO{Combine this figure with the above code maybe.}
@@
     Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes \ttiny{agg}regators, \ttiny{eager}, and \ttiny{split} nodes) and compilation time for two indicative parallelization configurations.
+    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes aggregators, eager, and split nodes) and compilation time for two indicative parallelization configurations.
     % Highlights notes a few key features of each script.
@@
 \caption{
-  \textbf{Runtime setup lattice.}
-  Parallel No Eager and Blocking Eager improve over sequential, but are not directly comparable.
-  Eager adds \sys's optimized \ttiny{eager} relay, and \sys uses all primitives in \S\ref{optimizer} (Fig.~\ref{fig:microbenchmarks}).
+  \textbf{Lattice of evaluation setup, based on runtime primitives.}
+  \todo{Explain further}
 % Describe. \TODO{Combine this figure with the above code maybe.}
@@
       Different configurations per benchmark:
-      (i) PaSh: the complete implementation with \ttiny{eager} and \ttiny{split} enabled,
+      (i) PaSh: the complete implementation with \ttt{eager} and \ttt{split} enabled,
       % (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (ii) PaSh w/o split: \ttiny{eager} enabled (no \ttiny{split}),
-      (iii) \todo{Blocking Eager}, only blocking \ttiny{eager} enabled (no \ttiny{split}),
-      (iv) \todo{No Eager}: both \ttiny{eager} and \ttiny{split} disabled.
+      (ii) PaSh w/o split: \ttt{eager} enabled (no \ttt{split}),
+      (iii) \todo{Blocking Eager}, only blocking \ttt{eager} enabled (no \ttt{split}),
+      (iv) \todo{No Eager}: both \ttt{eager} and \ttt{split} disabled.
       % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
@@
 \sys achieves up to x30 speedup for this script (significantly higher than what is achieved for the standard bigrams Fig.\ref{fig:microbenchmarks}).
-\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
+\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.\str
+% \kk{Do we want to talk about cost without mentioning it further?}
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 As discussed in (\S\ref{eval}, Parallelism) \sys does not have any control over the input script parallelism, and therefore it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.
-% \kk{eurosys: Can we somehow turn this point around and make it seems good/fine.
+This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.\str
+% \kk{Can we somehow turn this point around and make it seems good/fine.
 That here we report this parallelism, but in the real system, it tries to do a best effort guess of parallelism based on the size of the input script.}
@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    Pipelines are sorted in descending speedup order. 
+    Pipelines are sorted in descending speedup order.\str
+    % \kk{Add speedup by primitives and transformations in bar-chart.}
     % The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.

### 2020-10-08 paper/draft.tex
Minor fix

--- 2020-10-08 paper/draft.tex
@@
     \textbf{From Shell script AST to DFGs.}
-    The AST on the left has two parallelizable regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
+    The AST on the left has two dataflow regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
   }

### 2020-10-08 paper/draft.tex
Add comments from OSDI reviews

@@
 
-We now evaluate \sys on an existing set of \unix pipelines found in the wild and written by non-experts.
+We now evaluate \sys on an existing set of \unix pipelines found in the wild and \todo{written by non-experts}.
+% \kk{OSDI reviewer asked how do you know if they are non-experts.}
 
@@
 \sys's parallelism is set to $16\times$, as many of these pipelines are relatively long (\S\ref{eval}, Parallelism).
-Their input was generated by multiplying their original input several times, enough to reach 10GB.
+Their input was generated by \todo{multiplying their original input several times, enough to reach 10GB.}
+% \kk{eurosys: One OSDI reviewer didn't like how this is phrased. Can we change that?}
 Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all \todo{30} pipelines. We use the pipeline index to refer to them in both plots and discussion.
@@
 
+\kk{Mention something about Nebula (as an OSDI reviewer asked for.)}
+
 Concurrent to our work, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.

### 2020-10-08 paper/draft.tex
Fix some small issues in the evaluation

@@
 The complete pipeline executes in 44m2s for five years (82GB) of data.
-\sys with 2/10$\times$ parallelism leads to 1.86/2.44$\times$ speedup, with different phases seeing different benefits:
-  1.77/1.99$\times$ (vs. 33m58s) for all the pre-processing and 
-  2.30/10.79$\times$ speedup (vs. 10m4s) for computing the max.
-The reason why speedup is higher than parallelism is explained at the beginning of \S\ref{eval} (Parallelism).\footnote{
+\sys with \todo{10$\times$ parallelism} leads to \todo{2.44$\times$} speedup, with different phases seeing different benefits:
+  \todo{1.99$\times$} (vs. 33m58s) for all the pre-processing and 
+  \todo{10.79$\times$} speedup (vs. 10m4s) for computing the max.
+\todo{The reason why speedup is higher than parallelism} is explained at the beginning of \S\ref{eval} (Parallelism).\footnote{
   \nv{Can we avoid this by not showing the $2\times$? Why show $2\times$?} % <--- Eurosys
+  \kk{This will probably not be the case if we run for x16}
 }
@@
 % unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
-  a simpler (but longer) pipeline that leverages \unix built-ins % might be outperformed by a specialized programs on a single node, but
-  turns out to be trivially parallelizable.
-% For example, the \ttt{sort} command at the end is an expensive version of extracting the maximum element;
-%   this could have been expressed with a simple \ttt{awk} script that only keeps track of the max.
+  a simpler (but longer) pipeline that leverages \unix built-ins turns out to be trivially parallelizable.
 
@@
 
-% \kk{Should we mention that we serve the weather data in the local network}
-% \nv{It's in the setup}
-
 \heading{Take-aways}
-\sys can be applied to programs of notable size and input
-%% \todo{larger} % real/realistic? implies that prior benchmarks were not  
-%% programs
-%% with modern input sides
-to offer significant acceleration.
+\sys can be applied to programs of notable size and input to offer significant acceleration.
 A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, \ie the usual focus of conventional parallelization systems.
 
-% that falls outside the usual focus
-
 %% \todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
@@
 
-% few different ways to write some phases, 
-%   for example, 
-% A large pipeline resulted in a couple of 
-% An interesting lesson here was that there are multiple
-% First, longer pipelines can be written in multiple ways
-% It is worth noting
-% These are
-% 
 \subsection{Use Case: Wikipedia Web Indexing}
@@
 The original script takes 191min to execute on 1\% of Wikipedia (1.3GB).
-With 2/16$\times$-parallelism, \sys brings it down to 97/15min ($1.97/12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion.
+With 16$\times$\todo{-parallelism}, \sys brings it down to 15min ($12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion.
 % which uses \ttt{pandoc}, a utility written in Haskell.
@@
 \sys operates on programs with annotated commands outside the POSIX/GNU subsets and leads to notable speedups even when the original program features significant task-based parallelism.
-%% \sys does not risk breaking semantics, thus does not improve the first component. % this will not change
 
@@
 While \ttt{--parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{GNU}$'s parallelism is configured to $2\times$ that of $S_{PaSh}$'s (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{--parallelism=127}), 
+$S_{GNU}$'s parallelism is configured to $2\times$ that of \todo{$S_{PaSh}$'s} \ttt{--width} (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{--parallelism=128}), 
 to account for \sys's additional runtime processes.
@@
 Finally, the comparison also shows \sys's benefits to command developers:
-  a low-effort parallelizability annotation achieves better scalability than a custom flag (and underlying parallel implementation) added by developers.
+  a low-effort parallelizability annotation achieves better scalability than a custom flag (and underlying parallel implementation) manually added by developers.
 
@@
 There are a few possible ways one might attempt to use GNU \ttt{parallel} on this program.
-They could use it on the 8th stage, assuming they know it is a bottleneck, bringing execution down to 304.4s ($1.8\times$ speedup).
+They could use it on the bottleneck stage, assuming they are can deduce that, bringing execution down to 304.4s ($1.8\times$ speedup).
 Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.

### 2020-10-08 paper/draft.tex
Corrections

@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and \todo{a parameter} \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
 % concurrent version
@@
 \sys first identifies sections of the script that lack synchronization and scheduling constraints and converts them to dataflow graphs (DFGs).
-It then performs a series of DFG transformations aimed at exposing parallelism without breaking semantics. % ~\sx{ir:transformations}.
+It then performs a series of DFG transformations aimed at exposing parallelism without breaking semantics, \todo{up to the desired graph} \ttt{width}.
+. % ~\sx{ir:transformations}.
 Finally, \sys converts these DFGs back to a shell script augmented with \sys-provided commands. % ~\sx{optimizer}.
@@
 \heading{Scripts}
-To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} that contain commands that are candidates for parallelization.
+To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} containing commands that are candidates for parallelization.
 This analysis is guided by the script structure: % to identify parallelization candidates:
-  some constructs (\eg \ttt{&}, \ttt{|}) expose task parallelism; others (\eg \ttt{&&}, \ttt{||}) enforce synchronization.
+  some constructs expose task parallelism; others enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
@@
     The AST on the left has two dataflow regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
+    \nv{DFG2 reads from f2, grep writes to f3}
   }
@@
 improvements~\sx{optimizer}.
+The additional relay nodes inserted by the auxiliary transformations affect the resulting % (task) 
+  parallelism.
 
@@
 \heading{Highlights}
-\sys accelerates ($1.92-61.1\times$) all scripts.\footnote{
-  \nv{But this is not true; ``the vast majority''? maybe put 51/54 scripts?}.
-}
-For some of our benchmarks~\sx{ours}, we decouple the performance benefits stemming from \sys's compiler alone from the ones stemming from \sys's runtime primitives:
-  the compiler's average speedup seen is $9.3\times$, jumping to $13.2\times$ when runtime optimizations are used.
-In a different set of $30$ benchmarks parallelized $16\times$\footnote{
-  \nv{Immediate question: why use $64\times$ in one and $16\times$ in another?}
-}
- see an average speedup of $6\times$, depending on the class of piecewise commands~\sx{unix50}.
-Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see $2.44\times$ and $12.7\times$ speedup on $16\times$-parallel configuration.
+\sys accelerates ($1.92-61.1\times$) \todo{49/54} scripts, whereas for the rest 5 the parallel execution time is close to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
+In this paragraph we summarize results for \ttt{--width=16}, but the rest of our evaluation section reports on varying \ttt{width}s.
+For a subset of our benchmarks~\sx{ours}, we decouple the performance benefits stemming from \sys's DFG transformations alone from the ones stemming from its runtime primitives:
+  the compiler's average speedup is \todo{$9.3\times$}, jumping to \todo{$13.2\times$} when runtime optimizations are used.
+A different subset ($30$ benchmarks) see an average speedup of $6.02\times$~\sx{unix50}.
+% , depending on the classed of commands involved
+Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see speedups of $2.44\times$ and $12.7\times$.
 \sys outperforms \ttt{sort --parallel} by a significant margin, and offers small performance benefits over GNU \ttt{parallel}, which returns incorrect results in one configuration~\sx{micro}.
@@
 
-The vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
-  only 6 commands ($<3\%$) needed a single-record annotation.
-% 
-In terms of correctness, \sys's results over multi-GB  inputs are identical to the sequential for all benchmarks.
+Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
+  only 6 commands outside these classes ($<3\%$) needed a single-record annotation.
+
+In terms of correctness, \sys's results on multi-GB inputs are always identical to the sequential.
 Scripts feature ample opportunities for breaking semantics, which \sys avoids.

### 2020-10-08 paper/draft.tex
Changes to eval

@@
 Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
-  only 6 commands outside these classes ($<3\%$) needed a single-record annotation.
+  only 6  ($<3\%$) commands, outside this library, needed a single-record annotation.
 
@@
 \heading{Parallelism}
-%
-% All experiments are run on the same machine and 
-% \nv{Eurosys: This /\ is described above}
-The level of parallelism is configured using \sys's \ttt{--parallel=N}, a flag that determines the number of times that \sys applies its transformations---determining the ``width'' of the parallel script.
-The value  of \ttt{N} is shown as the x-axis of Fig.\ref{fig:microbenchmarks}, called \emph{parallelism} throughout this section.
+The level of parallelism is configured using \sys's \ttt{--width=N}, a flag that determines the number of times that \sys applies its transformations and the resulting width of the data-parallel DFG.
+% The value  of \ttt{N} is shown as the x-axis of Fig.\ref{fig:microbenchmarks}. 
 We cannot control the initial parallelism of a script (\eg a command could spawn 10 processes), and thus parallel scripts produced by \sys often reach maximum parallelization benefits with a value of \ttt{N} that is smaller than the physical cores of the hardware setup (in our case 64).
@@
 
-We evaluate \sys under multiple configurations on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
-% \nv{Eurosys: goal of experiment: decouple compiler from runtime, and talk about all overheads}
+We evaluate \sys on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
+The goal is to evaluate
+  (i) the performance benefits stemming from \sys's  DFG transformations alone, including how \ttt{--width} affects speedup, 
+  (ii) performance benefits stemming from \sys's runtime primitives, showing results for several configurations of the runtime lattice (Fig.~\ref{fig:lattice}).
 
@@
     \caption{
-      \textbf{\sys's speedup for 2--64$\times$-parallelism (\S\ref{eval}, Parallelism).}
+      \textbf{\sys's speedup for \ttt{width=} 2--64.}
       Different configurations per benchmark:
@@
       (iv) \todo{No Eager}: both \ttiny{eager} and \ttiny{split} disabled.
+  For some pairs of configurations, \sys produces identical parallel scripts and thus only one is shown.
+%   Two types of pairs where x y z coincide
+%   - \sys adds no split, whenever whenever the DFG width is constant for the entire pipeline. In these cases the blue line is identical to the red line, which not shown.
+%    Conversely, whenever both blue and red lines are shown, split nodes are needed due to changes in the pipeline width and result in performance improvements.
+% 
+%   - without split, the script is sequential (spell, bigrams)
+% Early on they have a pure or no parallelizable pure
+% and thus introduce an aggregator resulting in width of 1 right after, and requires a split.
+% Thus, any configuration that does not has \ttt{split} is practically sequential beyond that point---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup.
+% 
+% Blocking eager: data parallelism, but no task parallelism
+% No eager: blocking FIFO does not allow pushing -> underutilize CPU
+% 
+% The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
+% 
+% Sort-sort possibly shows the full spectrum:
+%   (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
+%   (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
+
+  % Configurations where the parallel script produced is identical to an existing configuration are not shown.
+
       % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
@@
 %  shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
-Grep and Grep-light are in \sta, with the former centered around an expensive DFA-based backtracking expression.
+NFA-Regex and Filter are in \sta, with the former centered around an expensive \todo{NFA-based} backtracking expression and the latter being an IO-intensive script.
 % a DFA-based regular expression matching phase;
@@
 Sort-sort uses consecutive \pur commands without interleaving them with commands that condense their input size (\eg \ttt{uniq}).
-Finally, the two Bi-grams replicate and shift a stream by one entry to calculate bigrams, but Bi-gram-opt's shift, \ttt{sort}, and \ttt{uniq} are merged to one command to use a more efficient aggregator.
+Finally, the two Bi-grams replicate and shift a stream by one entry to calculate bigrams, but Bi-gram-opt's groups 5 commands into one \pur command that has a more efficient $map$ and $aggregate$ implementation.
 
@@
 % TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents \sys's speedup as a function of parallelism (2--64$\times$).
+Fig.~\ref{fig:microbenchmarks} presents \sys's speedup as a function of \ttt{width=}2--64.
 % (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the optimized \sys, \ie \ttt{eager} and \ttt{split} enabled, for \{2, 4, 8, 16, 32, 64\}$\times$-parallelism are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
+Average speedups of the optimized \sys, \ie \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
+Average speedups for No Eager, \ie only \sys' transformations without its runtime support, are 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
 % \kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
-
 % we experiment with a $200\times$ level of parallelism to see potential overheads
 Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
-It also contains Filter and Bigrams-opt (not in Fig.~\ref{fig:microbenchmarks}).
-Filter is used to evaluate \sys's impact on IO-intensive scripts (\sys achieves a 1.5--2.5$\times$ speedup).
-Bigrams-opt is an optimized version of Bigrams that groups 5 commands into one \pur command and thus has a more efficient $map$ and $aggregate$ implementation.
-\sys achieves up to $30\times$ speedup for this script (significantly higher than what is achieved for the standard Bigrams Fig.\ref{fig:microbenchmarks}).
-% \nv{Eurosys: I think this should be part of a "do custom aggregators help?" question, not the main results}.
+Filter and Bigrams-opt are not in Fig.~\ref{fig:microbenchmarks}, for which \sys achieves a speedup 1.5--2.5$\times$ and $30\times$.
 \sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
@@
 \heading{Discussion}
-%
+% TODO: Add
+% * for scripts only with commands in \sta, we get linear speedup.
+% * eager is better than no eager and blocking eager for all scripts
+% * A \ttt{split} benefits scripts with \tpur or \tnpu commands, but does not negatively affect the rest.
+% * \ttt{width}
+% * why blocking is less than no eager
+% ( standard vs non-standard Bigrams Fig.\ref{fig:microbenchmarks}).
 \todo{These results explain a detail before getting the main points across: A is better than B, C is worse except G}
@@
 As described in Fig.\ref{optimizer}, these eager implementations perform worse than \sys's optimized \ttt{eager} relay in all cases (except for a negligible difference in Shortest-Scripts).
-A \ttt{split} benefits scripts with \tpur or \tnpu commands, but does not negatively affect the rest.
 

### 2020-10-08 paper/draft.tex
Corrections

@@
 %   these primitives were developed from uncovering several pathological situations, on many of which we report.
+
+% TODO: keep only points necessary for the punchline 
+% In the vast majority of scripts (49/54), \sys achieves significant speedups \todo{(average for width=16)}, and for the rest 5 slowdown is insignificant (worst case 0.89x).
+% For the temperature analysis script (which is taken from a Hadoop case study), \sys not only 
+% contains pre-processing that takes X time and Y main processing (Hadoop focus).
+% \sys accelerates the entire pipeline by , including the fragment that is outside the focus of conventional parallelization systems.
+% accelerates the CPU-intensive part that is parallelized by Hadoop, but also for the preprocessing part which takes even more time than what the other part takes.
+% XX--YY speedups in the vast majority of scripts, ones taken from the wild 
+% 1. Sentence about transformations vs runtime primitives
+% 2. Unix 50? 
+% 3. point about Hadoop expressing a fraction of the pipeline and speedup
+% \sys achieves higher speedup than \ttt{sort --parallel} that involved hand-tuned parallel implementation by command developers.
+% correctness of GNU parallel
+% PaSh can be applied to programs of notable
+% size and input to offer significant acceleration. A broader
+% take-away is that PaSh is also able to extract parallelism from
+% fragments that are not purely compute-intensive regions, i.e.,
+% the usual focus of conventional parallelization systems.
 
@@
 Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see speedups of $2.44\times$ and $12.7\times$.
-\sys outperforms \ttt{sort --parallel} by a significant margin, and offers small performance benefits over GNU \ttt{parallel}, which returns incorrect results in one configuration~\sx{micro}.
+\sys outperforms \ttt{sort --parallel} \todo{by a significant margin}\footnote{
+\nv{Nikos add point about experts}
+}, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
 % Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
@@
   For some pairs of configurations, \sys produces identical parallel scripts and thus only one is shown.
+}
 %   Two types of pairs where x y z coincide
@@
       % TODO: WHY?
-    }
     \vspace{-15pt}
@@
 
-We now evaluate \sys on an existing set of \unix pipelines found in the wild and \todo{written by non-experts}.
-% \kk{OSDI reviewer asked how do you know if they are non-experts.}
+We now evaluate \sys on an existing set of \unix pipelines found in the wild.
 
 \heading{Programs}
-In a recent celebration of \unix's 50-year legacy, Bell Labs created 37 challenges~\cite{unix50} solvable by means of composing \unix pipelines.
+In a recent celebration of \unix's 50-year legacy, Bell Labs created 37 challenges~\cite{unix50} solvable by \unix pipelines.
 The problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix}.
@@
 \heading{Results}
-\sys's parallelism is set to $16\times$, as many of these pipelines are relatively long (\S\ref{eval}, Parallelism).
-Their input was generated by \todo{multiplying their original input several times, enough to reach 10GB.}
-% \kk{eurosys: One OSDI reviewer didn't like how this is phrased. Can we change that?}
-Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for all \todo{30} pipelines. We use the pipeline index to refer to them in both plots and discussion.
-%% (\nv{Pipelines are referred by number, in both the plot and the discussion below.})
+\sys's \ttt{width} is set to 16 and pipelines are given 10GB inputs.
+Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for 31 pipelines.
+We do not include three pipelines that take less than 0.1 seconds to execute because they only output the first line of their input using \ttt{head}.
+We use the pipeline index to refer to them in both plots and discussion.
 Average speedup is \todo{$6.02\times$}, median is $6.48\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},
-We do not include three pipelines that take less than 0.1 seconds to execute because they only output the first line of their input using \ttt{head}.
 
@@
   % and \todo{2, 19, 31} that see a slowdown.
-These pipelines contain commands which \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
-Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind; a common example is the use of \ttt{awk} for reordering or removing columns of the input.
+These pipelines often contain general commands that \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
+% Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind;
+%   a common example is the use of \ttt{awk} for reordering or removing columns of the input.
 %% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-In particular, replacing an \ttt{awk} that was just used to sort on the second field \ttt{awk "{print \$2, \$0}" | sort -nr} with a single \ttt{sort -nr -k 2} in the 
+A \unix expert would notice that some of these uses could be replaced with \unix-specific commands---\eg replacing \ttt{awk "{print \$2, \$0}" | sort -nr} just used to sort on the second field with a single \ttt{sort -nr -k 2} in the 26th pipeline.
   % \todo{13th} 
-  26th pipeline allows \sys to achieve 8.1$\times$ speedup (in contrast to the original 1.01$\times$).
+The limited expressiveness of these commands can be exploited by \sys---in this specific case, achieving 8.1$\times$ speedup (\vs the original 1.01$\times$).
 %% , attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
@@
   (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} 
-  5, 6, 7, 8, 9, 19, 20, 21, 23, 24,
+  (\# 5, 6, 7, 8, 9, 19, 20, 21, 23, 24), 
   % 24, 9, 7, 6, 8, 23, 20, 19, 21, 5
   % \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
-  (ii) some are deep pipelines that already exploit task-based parallelism 
-  4, 10, 11, 13, 15, 17, 19, 21, 22,
+  (ii) they are deep pipelines that already exploit task-based parallelism
+  (\# 4, 10, 11, 13, 15, 17, 19, 21, 22),
   % 17, 4, 13, 15, 10, 22, 11, 19, 21,
   % \todo{(7, 8, 9, 10, 11, 12, 14, 27, 28,)} 
-  and
+or 
   (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs 
-  3, 4, 11, 12, 14, 16, 17, 18, 22.
+  (\# 3, 4, 11, 12, 14, 16, 17, 18, 22).
   % 14, 3, 16, 17, 4, 22, 11, 12, 18
@@
   fetch data from NOAA (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
-Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
-  whereas we consider the entire pipeline. % ---from fetching data to outputting the maximum temperature.
+While the book focuses only on the last one, \sys parallelizes the entire pipeline.
+% Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
+%   whereas we consider the entire pipeline. % ---from fetching data to outputting the maximum temperature.
 % The version used in the evaluation is slightly different than the one presented in Fig.~\ref{fig:example} (14 stages and totalling 240 characters), simplified only for clarity of exposition, in that it expresses the \ttt{for} loop as a first stage followed by \ttt{xargs}.
@@
 The complete pipeline executes in 44m2s for five years (82GB) of data.
-\sys with \todo{10$\times$ parallelism} leads to \todo{2.44$\times$} speedup, with different phases seeing different benefits:
+\sys with \ttt{width=}16 leads to \todo{2.44$\times$} speedup, with different phases seeing different benefits:
   \todo{1.99$\times$} (vs. 33m58s) for all the pre-processing and 
   \todo{10.79$\times$} speedup (vs. 10m4s) for computing the max.
-\todo{The reason why speedup is higher than parallelism} is explained at the beginning of \S\ref{eval} (Parallelism).\footnote{
-  \nv{Can we avoid this by not showing the $2\times$? Why show $2\times$?} % <--- Eurosys
-  \kk{This will probably not be the case if we run for x16}
-}
 
@@
 \heading{Take-aways}
-\sys can be applied to programs of notable size and input to offer significant acceleration.
+\sys can be applied to programs of notable size \todo{and complexity} to offer significant acceleration.
 A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, \ie the usual focus of conventional parallelization systems.
@@
 The original script takes 191min to execute on 1\% of Wikipedia (1.3GB).
-With 16$\times$\todo{-parallelism}, \sys brings it down to 15min ($12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion.
-% which uses \ttt{pandoc}, a utility written in Haskell.
+With \ttt{width=}16, \sys brings it down to 15min ($12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion. % which uses \ttt{pandoc}, a utility written in Haskell.
 
@@
 \heading{Take-aways}
-\sys operates on programs with annotated commands outside the POSIX/GNU subsets and leads to notable speedups even when the original program features significant task-based parallelism.
+\sys operates on programs with (annotated) commands outside the POSIX/GNU subsets and leads to notable speedups, even when the original program features significant task-based parallelism.
 
@@
 
-This section attempts to characterize \sys's limits by comparing it with similar alternatives.
-As there are no prior systems directly comparable to \sys, we draw comparisons with a series of specialized ones that excel within smaller fragments of \sys's proposed domain.
+% This section attempts to compare \sys with best-of-breed alternatives for particular taks.
+As there are no prior systems directly comparable to \sys, this section draws comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
 
@@
 %%   good as it could be.}
-We first compare a single GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) versus the same \ttt{sort} with the 
+We first compare a single GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
 %% \begin{wrapfigure}[10]{r}{0.25\textwidth}
@@
 $S_{GNU}$ indicates that \ttt{sort}'s scalability is inherently limited (\ie due to \ttt{sort} and not \sys);
-  this is why all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) do not go above $8\times$.
+  this is why all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) are capped at $8\times$ speedup.
 Finally, the comparison also shows \sys's benefits to command developers:
@@
 % 554 / 330 
-Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$).
-We note that this pipeline is harsh for \sys, in that most of the overhead comes from a single command---masking \sys's improvements in other parts of the pipeline.
+Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from \todo{a single command}.
 
@@
 
-There are a few possible ways one might attempt to use GNU \ttt{parallel} on this program.
-They could use it on the bottleneck stage, assuming they are can deduce that, bringing execution down to 304.4s ($1.8\times$ speedup).
-Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
-This would lead to 3.2$\times$ performance improvements but severely incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
+There are a few possible ways users might attempt to use GNU \ttt{parallel} on this program.
+They could use it on the bottleneck stage, assuming they can deduce it, bringing execution down to 304.4s ($1.8\times$ speedup).
+Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program.
+% ---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
+This would lead to 3.2$\times$ performance improvements but incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
 \sys's conservative program transformations are not applied in program fragments with unclear parallelizability properties.

### 2020-10-08 paper/draft.tex
Checkpoint eval overview

@@
 
-% \TODO{Drop the automated from the title maybe?}
-% \nv{Updates}
-
 %% \kk{After discussion on Wednesday Jan 8: Modify the narrative of
@@
 % For small input sizes this isn't a problem since scripts take less than a second to execute, but as data sizes increase, failing to utilize the underlying parallel computational resources is a significant bottleneck.
+
+\nv{Make sure we use \ttt{width} everywhere}
+\nv{Decide on parallelizable regions}
 
@@
   Eager adds \sys's optimized \ttiny{eager} relay, and \sys uses all primitives in \S\ref{optimizer} (Fig.~\ref{fig:microbenchmarks}).
-% Describe. \TODO{Combine this figure with the above code maybe.}
 }
@@
 
-This section reports on whether \sys can indeed scale shell scripts out automatically and correctly, using several scripts collected out from the wild along with a few micro-benchmarks for targeted comparisons.
+This section reports on whether \sys can indeed offer performance benefits automatically and correctly using several scripts collected out from the wild along with a few micro-benchmarks for targeted comparisons.
 
@@
 \heading{Highlights}
-\sys accelerates ($1.92-61.1\times$) \todo{49/54} scripts, whereas for the rest 5 the parallel execution time is close to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
-In this paragraph we summarize results for \ttt{--width=16}, but the rest of our evaluation section reports on varying \ttt{width}s.
-For a subset of our benchmarks~\sx{ours}, we decouple the performance benefits stemming from \sys's DFG transformations alone from the ones stemming from its runtime primitives:
-  the compiler's average speedup is \todo{$9.3\times$}, jumping to \todo{$13.2\times$} when runtime optimizations are used.
-A different subset ($30$ benchmarks) see an average speedup of $6.02\times$~\sx{unix50}.
+In this paragraph we summarize results for \ttt{--width=16}, but \sys's evaluation reports on varying \ttt{width}s.
+Overall, \sys accelerates 49/54 scripts by \todo{1.92--61.1}$\times$; % Eurosys: results for 16!!
+  for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
+For a subset of our benchmarks~\sx{ours}, we decouple the speedup due \sys's DFG transformations alone from the one due to its runtime primitives:
+  the former averages \todo{$9.3\times$}, jumping to \todo{$13.2\times$} due to the latter.
+A different subset ($30$ benchmarks) sees average speedups of $6.02\times$~\sx{unix50}.
 % , depending on the classed of commands involved
 Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see speedups of $2.44\times$ and $12.7\times$.
-\sys outperforms \ttt{sort --parallel} \todo{by a significant margin}\footnote{
-\nv{Nikos add point about experts}
-}, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
+\sys significantly outperforms \ttt{sort --parallel}, which involves hand-tuning by its command developers, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
 % Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
@@
 \heading{Parallelism}
-The level of parallelism is configured using \sys's \ttt{--width=N}, a flag that determines the number of times that \sys applies its transformations and the resulting width of the data-parallel DFG.
+The level of parallelism is configured using \sys's \ttt{--width=W}, a flag that determines the number of times that \sys applies its transformations and the resulting width of the data-parallel DFG.
 % The value  of \ttt{N} is shown as the x-axis of Fig.\ref{fig:microbenchmarks}. 
-We cannot control the initial parallelism of a script (\eg a command could spawn 10 processes), and thus parallel scripts produced by \sys often reach maximum parallelization benefits with a value of \ttt{N} that is smaller than the physical cores of the hardware setup (in our case 64).
+\sys does not control a script's initial parallelism (\eg a command could spawn 10 processes), and thus the resulting scripts often reach maximum parallelization benefits with a value of \ttt{W} smaller than the physical cores of our hardware setup (in our case 64).
 
@@
 We evaluate \sys on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
-The goal is to evaluate
-  (i) the performance benefits stemming from \sys's  DFG transformations alone, including how \ttt{--width} affects speedup, 
-  (ii) performance benefits stemming from \sys's runtime primitives, showing results for several configurations of the runtime lattice (Fig.~\ref{fig:lattice}).
+The goal is to evaluate performance benefits due to \sys's  
+  (i) DFG transformations alone, including how \ttt{--width} affects speedup, 
+  (ii) runtime primitives, showing results for several configurations of the runtime lattice (Fig.~\ref{fig:lattice}).
 
@@
 }
-%   Two types of pairs where x y z coincide
-%   - \sys adds no split, whenever whenever the DFG width is constant for the entire pipeline. In these cases the blue line is identical to the red line, which not shown.
-%    Conversely, whenever both blue and red lines are shown, split nodes are needed due to changes in the pipeline width and result in performance improvements.
-% 
-%   - without split, the script is sequential (spell, bigrams)
-% Early on they have a pure or no parallelizable pure
-% and thus introduce an aggregator resulting in width of 1 right after, and requires a split.
-% Thus, any configuration that does not has \ttt{split} is practically sequential beyond that point---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup.
-% 
-% Blocking eager: data parallelism, but no task parallelism
-% No eager: blocking FIFO does not allow pushing -> underutilize CPU
-% 
-% The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
-% 
-% Sort-sort possibly shows the full spectrum:
-%   (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
-%   (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
-
-  % Configurations where the parallel script produced is identical to an existing configuration are not shown.
-
-      % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
-      % Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
-      % TODO: WHY?
     \vspace{-15pt}
@@
 %   the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
+
+%   Two types of pairs where x y z coincide
+%   - \sys adds no split, whenever whenever the DFG width is constant for the entire pipeline. In these cases the blue line is identical to the red line, which not shown.
+%    Conversely, whenever both blue and red lines are shown, split nodes are needed due to changes in the pipeline width and result in performance improvements.
+% 
+%   - without split, the script is sequential (spell, bigrams)
+% Early on they have a pure or no parallelizable pure
+% and thus introduce an aggregator resulting in width of 1 right after, and requires a split.
+% Thus, any configuration that does not has \ttt{split} is practically sequential beyond that point---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup.
+% 
+% Blocking eager: data parallelism, but no task parallelism
+% No eager: blocking FIFO does not allow pushing -> underutilize CPU
+% 
+% The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
+% 
+% Sort-sort possibly shows the full spectrum:
+%   (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
+%   (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
+
+  % Configurations where the parallel script produced is identical to an existing configuration are not shown.
+
+      % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
+      % Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
+      % TODO: WHY?
 

### 2020-10-08 compiler/gather_results.py
Change parallelism to width in plots

@@ -365,7 +365,7 @@ def collect_scaleup_times(experiment, results_dir):
 
     ## Plot speedup
     ax.set_ylabel('Speedup')
-    ax.set_xlabel('Level of Parallelism')
+    ax.set_xlabel('--width')
 
     # ax.plot(all_scaleup_numbers, total_distr_speedup, '-^', linewidth=0.5, label='+ Merge')
     # ax.plot(all_scaleup_numbers, all_scaleup_numbers, '-', color='tab:gray', linewidth=0.5, label='Ideal')
@@ -408,7 +408,7 @@ def plot_sort_with_baseline(results_dir):
 
     ## Plot speedup
     ax.set_ylabel('Speedup')
-    ax.set_xlabel('Level of Parallelism')
+    ax.set_xlabel('--width')
     ax.plot(all_scaleup_numbers, sort_distr_speedup, '-o', linewidth=0.5, label='Pash')
     ## Add the no eager times if they exist
     # try:
@@ -750,7 +750,7 @@ def collect_unix50_scaleup_times(unix50_results_dir):
 
     ## Plot speedup
     ax.set_ylabel('Speedup')
-    ax.set_xlabel('Level of Parallelism')
+    ax.set_xlabel('--width')
     ax.plot(scaleup_numbers, avg_results, '-o', linewidth=0.5, label='Parallel')
     # plt.yscale("log")
     plt.xticks(scaleup_numbers)
@@ -785,7 +785,7 @@ def collect_unix50_coarse_scaleup_times(unix50_results_dir):
 
     ## Plot speedup
     ax.set_ylabel('Speedup')
-    ax.set_xlabel('Level of Parallelism')
+    ax.set_xlabel('--width')
     ax.plot(scaleup_numbers, avg_results, '-o', linewidth=0.5, label='Parallel')
     # plt.yscale("log")
     plt.xticks(scaleup_numbers)
@@ -848,7 +848,7 @@ def plot_one_liners_tiling(results_dir):
         if(ax.is_first_col()):
             ax.set_ylabel('Speedup')
         if(ax.is_last_row()):
-            ax.set_xlabel('Level of Parallelism')
+            ax.set_xlabel('--width')
         if(ax.is_first_row()):
             ax.set_xticklabels([])
         # ax.label_outer()
@@ -916,7 +916,7 @@ def plot_less_one_liners_tiling(results_dir):
         if(ax.is_first_col()):
             ax.set_ylabel('Speedup')
[…]

### 2020-10-08 paper/draft.tex
Some minor fixes

@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and \todo{a parameter} \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
 % concurrent version
@@
 \sys first identifies sections of the script that lack synchronization and scheduling constraints and converts them to dataflow graphs (DFGs).
-It then performs a series of DFG transformations aimed at exposing parallelism without breaking semantics, \todo{up to the desired graph} \ttt{width}.
+It then performs a series of DFG transformations that expose parallelism without breaking semantics, by expanding the DFG to the desired \ttt{width}.
 . % ~\sx{ir:transformations}.
@@
 \sys first lifts sections of the input script to the DFG
-representation, it performs transformations to expose parallelism, and
+representation, it performs transformations to expose parallelism (up to the desired \ttt{--width}), and
 then instantiates each DFG back to a parallel shell script.  A
@@
     Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes \ttiny{agg}regators, \ttiny{eager}, and \ttiny{split} nodes) and compilation time for two indicative parallelization configurations.
+    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes \ttiny{agg}regators, \ttiny{eager}, and \ttiny{split} nodes) and compilation time for two indicative \ttiny{--width}s.
     % Highlights notes a few key features of each script.

### 2020-10-08 paper/draft.tex
Checkpoint

--- 2020-10-08 paper/draft.tex
@@
 % (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the optimized \sys, \ie \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \todo{1.97, 3.5, 5.78, 8.83, 10.96, and 13.47}, respectively.
-Average speedups for No Eager, \ie only \sys' transformations without its runtime support, are 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
+Average speedups of the optimized \sys, \ie with \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \{\todo{1.97, 3.5, 5.78, 8.83, 10.96, 13.47}\}$\times$, respectively.
+For No-Eager, \ie for only \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 % we experiment with a $200\times$ level of parallelism to see potential overheads
+
+Configurations where the parallel script produced is identical to an existing configuration are not shown.
+There are two types of such cases (that appear in \todo{four} plots).
+In the first, blue and red (not shown) lines are identical when \sys adds no \ttt{split} because the DFG width is constant for the entire pipeline;
+  conversely, if both blue and red lines are shown, \ttt{split}s are needed due to changes in the pipeline width and result in performance improvements.
+In the second type, the script features a command in \pur that introduces an aggregator resulting in width of 1 right after, needing a \ttt{split} to regain width.
+Non-\ttt{split} configurations are sequential beyond that aggregator---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup and are identical (\todo{spell, bigrams}).
+
 Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
-Filter and Bigrams-opt are not in Fig.~\ref{fig:microbenchmarks}, for which \sys achieves a speedup 1.5--2.5$\times$ and $30\times$.
+Filter and Bigrams-opt are not in Fig.~\ref{fig:microbenchmarks}, for which \sys achieves an average speedup 1.5--2.5$\times$ and $30\times$.
 \sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
@@
 %   the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
-
-%   Two types of pairs where x y z coincide
-%   - \sys adds no split, whenever whenever the DFG width is constant for the entire pipeline. In these cases the blue line is identical to the red line, which not shown.
-%    Conversely, whenever both blue and red lines are shown, split nodes are needed due to changes in the pipeline width and result in performance improvements.
-% 
-%   - without split, the script is sequential (spell, bigrams)
-% Early on they have a pure or no parallelizable pure
-% and thus introduce an aggregator resulting in width of 1 right after, and requires a split.
-% Thus, any configuration that does not has \ttt{split} is practically sequential beyond that point---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup.
-% 
-% Blocking eager: data parallelism, but no task parallelism
-% No eager: blocking FIFO does not allow pushing -> underutilize CPU
-% 
-% The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
-% 
-% Sort-sort possibly shows the full spectrum:
-%   (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
-%   (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
-
-  % Configurations where the parallel script produced is identical to an existing configuration are not shown.
 
@@
 %>   parallelism).}
+
+Blocking eager: data parallelism, but no task parallelism
+No eager: blocking FIFO does not allow pushing -> underutilize CPU
+
+The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
+
+Sort-sort possibly shows the full spectrum:
+  (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
+  (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
 

### 2020-10-08 paper/draft.tex
Add Dryad and improve paragraph about PoSh

--- 2020-10-08 paper/draft.tex
@@
 
-\kk{Mention something about Nebula (as an OSDI reviewer asked for.)}
-
 Concurrent to our work, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
 POSH focuses on colocating distributed operation, and introduces a novel scheduling algorithm. 
-\sys instead attempts non-distributed parallelism, is location-agnostic, and focuses on correctness through its dataflow transformations.
+\sys instead focuses on the parallelization of CPU-intensive scripts, by transforming them to DFGs, applying transformations, and then transforming them back to parallel schell scripts augmented with \sys's runtime primitives to be executed on a standard shell.
+%  non-distributed parallelism, is location-agnostic, and focuses on correctness through its dataflow transformations.
 
@@
 Its  DFG model, however, is different and captures ordering constraints.
-This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.
+This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.\str
+% \kk{eurosys: I could see this being deleted in need of space.}
+
 % As a result, previous work is not directly applicable.
@@
 
+% \kk{Mention something about Nebula (as an OSDI reviewer asked for.)}
+
+Dryad~\cite{isard2007dryad} is a distributed system for dataflow graphs.
+Dryad offers a scripting language, Nebula, that allows using shell commands such as \ttt{grep} or \ttt{sed} in place of individual dataflow nodes.
+The main difference with \sys is that in Dryad the programmer needs to explicitly express the data flow graph, which is then executed in a distributed fashion, whereas Pash automatically parallelizes a given shell script by producing a parallel script that runs on an unmodified shell of choice.
+
 % % Although still a research prototype, \sys's setup and deployment is simpler.

### 2020-10-08 paper/draft.tex
Finish 6.1

@@
 Average speedups of the optimized \sys, \ie with \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \{\todo{1.97, 3.5, 5.78, 8.83, 10.96, 13.47}\}$\times$, respectively.
-For No-Eager, \ie for only \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
+For \emph{No-Eager}, \ie \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 
-Configurations where the parallel script produced is identical to an existing configuration are not shown.
-There are two types of such cases (that appear in \todo{four} plots).
-In the first, blue and red (not shown) lines are identical when \sys adds no \ttt{split} because the DFG width is constant for the entire pipeline;
-  conversely, if both blue and red lines are shown, \ttt{split}s are needed due to changes in the pipeline width and result in performance improvements.
-In the second type, the script features a command in \pur that introduces an aggregator resulting in width of 1 right after, needing a \ttt{split} to regain width.
-Non-\ttt{split} configurations are sequential beyond that aggregator---\ie No Eager, Blocking Eager, and \sys w/o Split see no speedup and are identical (\todo{spell, bigrams}).
-
-Tab.~\ref{tab:eval} shows that \sys's compilation time is negligible. % ($<5s$)
-Filter and Bigrams-opt are not in Fig.~\ref{fig:microbenchmarks}, for which \sys achieves an average speedup 1.5--2.5$\times$ and $30\times$.
-\sys achieves a COST~\cite{mcsherryscalability} of 2 for all benchmarks.
+Plots hide lines for configurations that lead to identical parallel programs.
+There are two types of such cases (that appear in \todo{five} plots).
+In the first type, \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical when \sys adds no \ttt{split} because the DFG width remains constant for the whole program;
+  conversely, if both lines are shown, \ttt{split}s are needed due to changes in parallelism width and result in performance improvements.
+In the second type, No Eager (green), Blocking Eager (orange), and \emph{Pash w/o Split} (red) are identical 
+  because the program features a command in \pur relatively early.
+This command requires an \ttt{agg}regator, whose output is of width 1, and beyond which non-\ttt{split} configurations are sequential (and see no speedup).
+Conversely, all three lines exist when the width remains above 1 for the full program.
+% needing a \ttt{split} to regain width.
+
+Tab.~\ref{tab:eval} shows that \sys's transformation time is negligible, and its COST~\cite{mcsherryscalability} is 2. %  for all benchmarks.
+Filter and Bigrams-opt, not in Fig.~\ref{fig:microbenchmarks}, achieve speedups of 1.5--2.5$\times$ and 2--30$\times$.
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 % TODO: Add
-% * for scripts only with commands in \sta, we get linear speedup.
-% * eager is better than no eager and blocking eager for all scripts
-% * A \ttt{split} benefits scripts with \tpur or \tnpu commands, but does not negatively affect the rest.
 % * \ttt{width}
 % * why blocking is less than no eager
-% ( standard vs non-standard Bigrams Fig.\ref{fig:microbenchmarks}).
-\todo{These results explain a detail before getting the main points across: A is better than B, C is worse except G}
-As discussed earlier (\S\ref{eval}, Parallelism), \sys does not have control over the input script's parallelism.
-Thus, it often achieves the maximum possible speedup, \ie without unnecessary process spawns and context switches, earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-This is also due to the fact that \sys also adds its runtime primitives (\ttt{split} and \ttt{eager}), further increasing process spawns.
+As expected, scripts with commands only in \sta get linear speedup.
+\sys's \ttt{split} benefits scripts with \tpur or \tnpu commands, without negative effects on the rest.
+Its \ttt{eager} primitive improves over Eager and Blocking-eager for all scripts.
+% Blocking eager: data parallelism, but no task parallelism; No eager: blocking FIFO does not allow pushing -> underutilize CPU
+\emph{Blocking Eager} is usually slower than \emph{No Eager} due to \ttt{agg}regators that read multiple input streams at the same time---\eg \ttt{sort -m}.
+Sort-sort illustrates the full spectrum of primitives:
+  (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
+  (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
+
+% \todo{No Eager} and \todo{Blocking Eager} depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
+% As described in Fig.\ref{optimizer}, these eager implementations perform worse than \sys's optimized \ttt{eager} relay in all cases (except for a negligible difference in Shortest-Scripts).
+
+As described earlier, \sys often achieves the maximum possible speedup % \ie without unnecessary process spawns and context switches, 
+  in \ttt{width} that is lower than the number of available cores---\ie \ttt{width=}16--32 for a 64-core system.
+This is also because \sys's runtime primitives spawn new processes---\eg Sort with \ttt{width=}8 spawns 37 processes:
+  8 \ttt{tr}, 8 \ttt{sort}, 7 \emph{aggregation}, and 14 \ttt{relay} processes.
+
+\todo{standard vs non-standard Bigrams}
+
 % \kk{eurosys: Can we somehow turn this point around and make it seems good/fine.
@@
 %% \kk{Ben's point: task parallelism is mentioned here for the first time. Since \sys exposes data parallelism as task parallelism, we might want to mention that somewhere earlier in the paper}
-For example, Sort in $8\times$ spawns 37 nodes:
-  8 \ttt{tr} nodes, 8 \ttt{sort} nodes, 7 \emph{aggregation} nodes, and 14 \ttt{relay} nodes.
 % In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
@@
 % \kk{Mention bigrams-opt result now that it is removed from the plot.}
-\todo{No Eager} and \todo{Blocking Eager} depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-As described in Fig.\ref{optimizer}, these eager implementations perform worse than \sys's optimized \ttt{eager} relay in all cases (except for a negligible difference in Shortest-Scripts).
 
@@
 %>   parallelism).}
-
-Blocking eager: data parallelism, but no task parallelism
-No eager: blocking FIFO does not allow pushing -> underutilize CPU
-
-The reason why Blocking Eager is usually slower than No Eager is due to \ttt{agg}regators that read multiple input streams at the same time---such \ttt{sort -m}.
-
-Sort-sort possibly shows the full spectrum:
-  (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
-  (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
 

### 2020-10-08 paper/draft.tex
Update 6.2

@@
 
-We now evaluate \sys on an existing set of \unix pipelines found in the wild.
+We now turn to a set of \unix pipelines found out in the wild.
 
@@
 %   then re-run them).} . 
-\sys executes each pipeline as-is, without any modification.
+\sys executes each pipeline as-is, without any modification.\footnote{
+  \nv{Do we need this?}
+}
   % if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
@@
 \heading{Results}
-\sys's \ttt{width} is set to 16 and pipelines are given 10GB inputs.
-Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for 31 pipelines.
-We do not include three pipelines that take less than 0.1 seconds to execute because they only output the first line of their input using \ttt{head}.
-We use the pipeline index to refer to them in both plots and discussion.
-Average speedup is \todo{$6.02\times$}, median is $6.48\times$, and weighted average (with the absolute times as weights) is $5.75\times$.
+Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for 31 pipelines, with \ttt{width=}16 and 10GB inputs.
+It does not include 3 pipelines that execute in under $0.1s$ by using \ttt{head} to output the first line of input.
+The script index refers to the challenge level, in both the plots and the discussion.
+Average speedup is \todo{$6.02\times$}, % median is $6.48\times$, 
+  and weighted average (with the absolute times as weights) is $5.75\times$.
  % the \todo{geometric mean is $1.08\times$},
@@
 \heading{Discussion}
-Most pipelines see significant acceleration, except 25-30
+Most pipelines see significant speedup, except \#25-30
   % \todo{13, 24, 25, 26, 29, 30} 
-  that see no speedup. 
+  that see no speedup because they contain general commands that \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
   % and \todo{2, 19, 31} that see a slowdown.
-These pipelines often contain general commands that \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
 % Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind;
@@
 %% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-A \unix expert would notice that some of these uses could be replaced with \unix-specific commands---\eg replacing \ttt{awk "{print \$2, \$0}" | sort -nr} just used to sort on the second field with a single \ttt{sort -nr -k 2} in the 26th pipeline.
+A \unix expert would notice that some of them can be replaced with \unix-specific commands---\eg \ttt{awk "{print \$2, \$0}" | sort -nr}, used to sort on the second field can be replaced with a single \ttt{sort -nr -k 2} (\#26).
   % \todo{13th} 
@@
   (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} 
-  (\# 5, 6, 7, 8, 9, 19, 20, 21, 23, 24), 
+  (\#5, 6, 7, 8, 9, 19, 20, 21, 23, 24), 
   % 24, 9, 7, 6, 8, 23, 20, 19, 21, 5
   % \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
-  (ii) they are deep pipelines that already exploit task-based parallelism
-  (\# 4, 10, 11, 13, 15, 17, 19, 21, 22),
+  (ii) they are deep pipelines that already exploit task parallelism
+  (\#4, 10, 11, 13, 15, 17, 19, 21, 22),
   % 17, 4, 13, 15, 10, 22, 11, 19, 21,
@@
 or 
-  (iii) they are not CPU-intensive, resulting in pronounced IO and constant costs 
-  (\# 3, 4, 11, 12, 14, 16, 17, 18, 22).
+  (iii) they are not CPU-intensive, resulting in pronounced I/O and constant costs 
+  (\#3, 4, 11, 12, 14, 16, 17, 18, 22).
   % 14, 3, 16, 17, 4, 22, 11, 12, 18

### 2020-10-08 paper/draft.tex
Complete eval

@@
 Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
-  only 6  ($<3\%$) commands, outside this library, needed a single-record annotation.
-
-In terms of correctness, \sys's results on multi-GB inputs are always identical to the sequential.
-Scripts feature ample opportunities for breaking semantics, which \sys avoids.
+  only 6 ($<3\%$) commands, outside this library, needed a single-record annotation~\sx{macro2}.
+
+In terms of correctness, \sys's results on multi-GB inputs are identical to the sequential ones.
+Scripts feature ample opportunities for breaking semantics~\sx{micro}, which \sys avoids.
 %% In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
@@
 \heading{Program}
-This program is inspired by Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising three sub-tasks:
-  fetch data from NOAA (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
-While the book focuses only on the last one, \sys parallelizes the entire pipeline.
+This program is inspired by Hadoop's Definitive Guide~\cite[\S2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising 3 stages:
+  fetch NOAA data (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
+While the book focuses only on the last stage, \sys parallelizes the entire pipeline.
 % Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
@@
 \sys can be applied to programs of notable size \todo{and complexity} to offer significant acceleration.
-A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, \ie the usual focus of conventional parallelization systems.
+A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive, \ie the usual focus of conventional parallelization systems.
 
@@
 % This section attempts to compare \sys with best-of-breed alternatives for particular taks.
-As there are no prior systems directly comparable to \sys, this section draws comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
+As there are no prior systems directly comparable to \sys, we now draw comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
 
@@
 %%   good as it could be.}
-We first compare a single GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
+We first compare a GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
 %% \begin{wrapfigure}[10]{r}{0.25\textwidth}
@@
 % 554 / 330 
-Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from \todo{a single command}.
+Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from a single command---\ttt{cutadapt}.
 

### 2020-10-08 paper/draft.tex
Update DFG figure

@@
     The AST on the left has two dataflow regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
-    \nv{DFG2 reads from f2, grep writes to f3}
   }
@@
     Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    Pipelines are sorted in descending speedup order.\str
-    % \kk{Add speedup by primitives and transformations in bar-chart.}
+    Pipelines are sorted in descending speedup order. % \str
+    % \kk{FUTURE: Add speedup by primitives and transformations in bar-chart.}
     % The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.
@@
 Concurrent to our work, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
-POSH focuses on colocating distributed operation, and introduces a novel scheduling algorithm. 
+POSH focuses on co-locating distributed operation, and introduces a novel scheduling algorithm. 
 \sys instead focuses on the parallelization of CPU-intensive scripts, by transforming them to DFGs, applying transformations, and then transforming them back to parallel schell scripts augmented with \sys's runtime primitives to be executed on a standard shell.
@@
 Dryad offers a scripting language, Nebula, that allows using shell commands such as \ttt{grep} or \ttt{sed} in place of individual dataflow nodes.
-The main difference with \sys is that in Dryad the programmer needs to explicitly express the data flow graph, which is then executed in a distributed fashion, whereas Pash automatically parallelizes a given shell script by producing a parallel script that runs on an unmodified shell of choice.
+The main difference with \sys is that in Dryad the programmer needs to explicitly express the dataflow graph, which is then executed in a distributed fashion, whereas Pash automatically parallelizes a given shell script by producing a parallel script that runs on an unmodified shell of choice.
 

### 2020-10-08 paper/draft.tex
Update annotation

@@
 This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
-However, it turns out to be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3
+However, it turns out to be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics.
 
@@
 A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output column(s).
-The first clause from its annotation record is shown below:
-%
-\begin{minted}[fontsize=\footnotesize]{json-object}
-  "predicate":
-    { 
-      "operator": "all",
-      "operands": ["-1", "-3"] 
-    },    
-  "class": "stateless",
-  "inputs": ["args[1]"],
-  "outputs": ["stdout"],
-  "options": ["stdin-hyphen"]
-\end{minted}
+% The first clause from its annotation record is shown below:
+Its annotation record is shown on the left:
+%
+% \begin{minted}[fontsize=\footnotesize]{json-object}
+%   "predicate":
+%     { 
+%       "operator": "all",
+%       "operands": ["-1", "-3"] 
+%     },    
+%   "class": "stateless",
+%   "inputs": ["args[1]"],
+%   "outputs": ["stdout"],
+%   "options": ["stdin-hyphen"]
+% \end{minted}
+
+\hspace*{-0.3cm}\includegraphics[width=\columnwidth]{\detokenize{./figs/pash_annotation.pdf}}
+
 % \begin{lstlisting}[language=json, numbers=none]
@@
 Annotation records are by default conservative so as to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
-The language is extensible with more operators (as long as the \todo{user} defines an \todo{interpreter} for them);
+The language is extensible with more operators (as long as the developer defines their semantics);
   it also supports writing arbitrary Python code for commands whose properties are difficult to capture---\eg higher order \ttt{xargs}, whose parallelizability class depends on the class of the first-order command that it invokes.

### 2020-10-09 paper/draft.tex
Pass over 6, 6.1

--- 2020-10-08 paper/draft.tex
@@
 In this paragraph we summarize results for \ttt{--width=16}, but \sys's evaluation reports on varying \ttt{width}s.
-Overall, \sys accelerates 49/54 scripts by \todo{1.92--61.1}$\times$; % Eurosys: results for 16!!
+Overall, \sys accelerates 47/52 scripts by 1.92--17.42$\times$; % Eurosys: results for 16!! \kk{I added them}
   for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
-For a subset of our benchmarks~\sx{ours}, we decouple the speedup due \sys's DFG transformations alone from the one due to its runtime primitives:
-  the former averages \todo{$9.3\times$}, jumping to \todo{$13.2\times$} due to the latter.
+  % \kk{eurosys: I have an issue that I am not sure how to address.
+I would like to have the 61.1x here (which is due to 64 width)  but I don't see a good way of including it without sounding suspicious.}
+For a subset of our benchmarks~\sx{ours}, we decouple the speedup due to \sys's DFG transformations alone from the one achieved when including its runtime primitives:
+  the former averages $5.9\times$, jumping to $8.6\times$ due to the latter.
 A different subset ($30$ benchmarks) sees average speedups of $6.02\times$~\sx{unix50}.
 % , depending on the classed of commands involved
-Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands in multiple languages see speedups of $2.44\times$ and $12.7\times$.
-\sys significantly outperforms \ttt{sort --parallel}, which involves hand-tuning by its command developers, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
+Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands written in multiple languages see speedups of $2.44\times$ and $12.7\times$.
+\sys significantly outperforms \ttt{sort --parallel}, a hand-tuned parallel implementation, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
 % Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
@@
 Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
-  only 6 ($<3\%$) commands, outside this library, needed a single-record annotation~\sx{macro2}.
-
-In terms of correctness, \sys's results on multi-GB inputs are identical to the sequential ones.
-Scripts feature ample opportunities for breaking semantics~\sx{micro}, which \sys avoids.
+  only 6  ($<3\%$) commands, outside this library, needed a single-record annotation.
+
+In terms of correctness, \sys's results on multi-GB inputs are always identical to the sequential.
+Scripts feature ample opportunities for breaking semantics, which \sys avoids.
 %% In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
@@
 
-We evaluate \sys on popular, common, or classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
+We first evaluate \sys on a set of popular, common, and classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
 The goal is to evaluate performance benefits due to \sys's  
-  (i) DFG transformations alone, including how \ttt{--width} affects speedup, 
+  (i) DFG transformations alone, including how \ttt{--width} affects speedup, and
   (ii) runtime primitives, showing results for several configurations of the runtime lattice (Fig.~\ref{fig:lattice}).
@@
     \caption{
-      \textbf{\sys's speedup for \ttt{width=} 2--64.}
+      \textbf{\sys's speedup for \ttiny{width=} 2--64.}
       Different configurations per benchmark:
@@
 %  shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
-NFA-Regex and Filter are in \sta, with the former centered around an expensive \todo{NFA-based} backtracking expression and the latter being an IO-intensive script.
+NFA-Regex is centered around an expensive \todo{NFA-based} backtracking expression and all of its commands are in \sta.
+% and the latter being an IO-intensive script.
 % a DFA-based regular expression matching phase;
@@
 Sort-sort uses consecutive \pur commands without interleaving them with commands that condense their input size (\eg \ttt{uniq}).
-Finally, the two Bi-grams replicate and shift a stream by one entry to calculate bigrams, but Bi-gram-opt's groups 5 commands into one \pur command that has a more efficient $map$ and $aggregate$ implementation.
+Finally, Bi-grams replicates and shifts a stream by one entry to calculate bigrams.
+% , but Bi-gram-opt's groups 5 commands into one \pur command that has a more efficient $map$ and $aggregate$ implementation.
 
@@
 
-Plots hide lines for configurations that lead to identical parallel programs.
-There are two types of such cases (that appear in \todo{five} plots).
-In the first type, \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical when \sys adds no \ttt{split} because the DFG width remains constant for the whole program;
-  conversely, if both lines are shown, \ttt{split}s are needed due to changes in parallelism width and result in performance improvements.
-In the second type, No Eager (green), Blocking Eager (orange), and \emph{Pash w/o Split} (red) are identical 
-  because the program features a command in \pur relatively early.
+Plots do not include lines for configurations that lead to identical parallel programs.
+There are two types of such cases. %  (that appear in \todo{five} plots).
+In the first, \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical for scripts that \sys does not add any \ttt{split} since the width of the DFG is constant;
+%  when \sys adds no \ttt{split} because the DFG width remains constant for the whole program;
+  conversely (for spell, bi-grams, and sort), when both lines are shown, \sys has added \ttt{split}s due to changes in the DFG width (e.g. due to a \npu command).
+In the second type, \emph{Pash w/o Split} (red) is identical to No Eager (green, hidden) and Blocking Eager (orange, hidden) because the input script features a command in \pur or \npu relatively early.
 This command requires an \ttt{agg}regator, whose output is of width 1, and beyond which non-\ttt{split} configurations are sequential (and see no speedup).
-Conversely, all three lines exist when the width remains above 1 for the full program.
+% Conversely, all three lines exist when the width remains above 1 for the full program.
 % needing a \ttt{split} to regain width.
 
-Tab.~\ref{tab:eval} shows that \sys's transformation time is negligible, and its COST~\cite{mcsherryscalability} is 2. %  for all benchmarks.
-Filter and Bigrams-opt, not in Fig.~\ref{fig:microbenchmarks}, achieve speedups of 1.5--2.5$\times$ and 2--30$\times$.
+Finally, Tab.~\ref{tab:eval} shows that \sys's transformation time is negligible, and its COST~\cite{mcsherryscalability} is 2. %  for all benchmarks.
+% Filter and Bigrams-opt, not in Fig.~\ref{fig:microbenchmarks}, achieve speedups of 1.5--2.5$\times$ and 2--30$\times$.
 %   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
@@
 % Blocking eager: data parallelism, but no task parallelism; No eager: blocking FIFO does not allow pushing -> underutilize CPU
-\emph{Blocking Eager} is usually slower than \emph{No Eager} due to \ttt{agg}regators that read multiple input streams at the same time---\eg \ttt{sort -m}.
+\emph{No Eager} is usually faster than \emph{Blocking Eager} 
+% due to \ttt{agg}regators that read multiple input streams at the same time---\eg \ttt{sort -m}.
+% \kk{eurosys: while the above is technically true, it is a bit difficult to explain here, so I prefer the sentence below.}
+since it allows its producer and consumer to execute in parallel.
 Sort-sort illustrates the full spectrum of primitives:
-  (iii) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} while maintaining task parallelism
-  (iv) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
+  (i) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} 
+  % while maintaining task parallelism
+  (ii) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
 
@@
 As described earlier, \sys often achieves the maximum possible speedup % \ie without unnecessary process spawns and context switches, 
-  in \ttt{width} that is lower than the number of available cores---\ie \ttt{width=}16--32 for a 64-core system.
+  for a \ttt{width} that is lower than the number of available cores---\ie \ttt{width=}16--32 for a 64-core system.
 This is also because \sys's runtime primitives spawn new processes---\eg Sort with \ttt{width=}8 spawns 37 processes:
@@
 
-\todo{standard vs non-standard Bigrams}
+% \todo{standard vs non-standard Bigrams}
 
@@
 \heading{Take-aways}
-\sys accelerates scripts by 4--60$\times$, depending on the commands involved.
+\sys accelerates scripts by up to 60$\times$, depending on the commands involved.
 Its runtime constructs improve over the baseline speedup achieved by its transformations.
@@
 \heading{Program}
-This program is inspired by Hadoop's Definitive Guide~\cite[\S2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising 3 stages:
-  fetch NOAA data (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
-While the book focuses only on the last stage, \sys parallelizes the entire pipeline.
+This program is inspired by Hadoop's Definitive Guide~\cite[Chapter 2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising three sub-tasks:
+  fetch data from NOAA (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
+While the book focuses only on the last one, \sys parallelizes the entire pipeline.
 % Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
@@
 \sys can be applied to programs of notable size \todo{and complexity} to offer significant acceleration.
-A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive, \ie the usual focus of conventional parallelization systems.
+A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive regions, \ie the usual focus of conventional parallelization systems.
 
@@
 % This section attempts to compare \sys with best-of-breed alternatives for particular taks.
-As there are no prior systems directly comparable to \sys, we now draw comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
+As there are no prior systems directly comparable to \sys, this section draws comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
 
@@
 %%   good as it could be.}
-We first compare a GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
+We first compare a single GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
 %% \begin{wrapfigure}[10]{r}{0.25\textwidth}
@@
 % 554 / 330 
-Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from a single command---\ttt{cutadapt}.
+Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from \todo{a single command}.
 

### 2020-10-09 compiler/gather_results.py
Table now doesnt contain filter and bigrams-opt

@@ -31,12 +31,12 @@ experiments = ["minimal_grep",
                "minimal_sort",
                "topn",
                "wf",
-               "grep",
+            #    "grep",
                "spell",
                "shortest_scripts",
                "diff",
                "bigrams",
-               "alt_bigrams",
+            #    "alt_bigrams",
                "set-diff",
                "double_sort"]
 

### 2020-10-09 paper/draft.tex
Corrections

@@
 
-\TODO{Add pointers}
-
 \heading{Commands}
-To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}.
+To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}~\sx{cmd}.
 Rather than describing a command's full observable behavior, these classes focus on information that is important for data parallelism.
-To allow other commands to use its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class.
+To allow other commands to use its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class~\sx{ext}.
 Annotations are expressed once per command rather than once per script and aim towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
@@
 \heading{Scripts}
-To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} containing commands that are candidates for parallelization.
+To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} containing commands that are candidates for parallelization~\sx{dataflow-regions}.
 This analysis is guided by the script structure: % to identify parallelization candidates:
@@
 % Since the shell representation only exposes coarse grained parallelization opportunities, 
-\sys then converts each parallelizable region to a \emph{dataflow graph} (DFG), 
+\sys then converts each parallelizable region to a \emph{dataflow graph} (DFG)~\sx{graph-components}, 
 %% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
@@
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-  a flexible representation that enables a series of local transformations to expose data parallelism, converting the graph into its parallel equivalent.
+  a flexible representation that enables a series of local transformations to expose data parallelism, converting the graph into its parallel equivalent~\sx{ir:transformations}.
 %% , in which DFGs indicate the parallelizable regions.
-Further transformations compile the DFG back to a shell script that uses POSIX constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program.
+Further transformations compile the DFG back to a shell script that uses POSIX constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program~\sx{backend}.
 
 \heading{Implementation}
-\sys addresses several practical challenges through a set of constructs it provides---\eg a set of
+\sys addresses several practical challenges through a set of constructs it provides---\ie 
 %% linkable
-modular components for augmenting command composition.
+modular components for augmenting command composition~\sx{impl}.
 %% for stream splitting and merging.
-It also provides a small library of efficient data aggregators for a large set of parallelizable commands.
+It also provides a small and efficient \emph{aggregator library} targeting a large set of parallelizable commands.
 %% For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
@@
 % The first clause from its annotation record is shown below:
-Its annotation record is shown on the left:
+Below is its annotation record (left) and the implementation of record's first clause (right):
 %
@@
   this input accompanies every replicated instance of the command executing in parallel.
-The second clause is symmetric to the first with respect to the second input, and the third clause, which is the general case classifies \ttt{comm} as \pur as it needs to read both inputs completely before outputing its results.
-The option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument.
+The second clause is symmetric to the first with respect to the second input. 
+The third clause is the general case, classifiying \ttt{comm} in \pur as it needs to read both inputs completely before outputing its results.
+Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument.
 % The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
@@
 
-\subsection{DFG Definitions}
+\subsection{Dataflow Model Definitions}
 \label{graph-components}
@@
     \includegraphics[width=\textwidth]{\detokenize{./figs/tiling_throughput_scaleup.pdf}}
+    \vspace{-20pt}
     \caption{
-      \textbf{\sys's speedup for \ttiny{width=} 2--64.}
+      \textbf{\sys's speedup for \ttiny{width=}2--64.}
       Different configurations per benchmark:
@@
 }
-    \vspace{-15pt}
+    \vspace{-14pt}
     \label{fig:microbenchmarks}
@@
   \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
+  \vspace{-15pt}
   \caption{

### 2020-10-09 paper/draft.tex
end-to-end pass

@@
 % TODO Add Ignis and Mozart
-%% \documentclass[letterpaper,twocolumn,10pt]{article}
-%% \pdfoutput=1 % for arxiv
-%% \usepackage{usenix2019_v3}
+% \nv{Make sure we use \ttt{width} everywhere}
+% \nv{Decide on parallelizable regions}
 
@@
 % a source-to-source shell compiler and associated runtime library
-Given a script, \sys converts it to a data-flow graph, performs a series of analyses that aim at exposing parallelism while preserving its sequential semantics, and then converts the data-flow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
+Given a script, \sys converts it to a dataflow graph, performs a series of analyses and transformations that aim at exposing parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
 A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
-An accompanying parallelizability study of POSIX and GNU commands, two large and commonly used groups, 
-informs \sys's annotation language and guides the development of a data-parallel standard library.
+An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---informs \sys's annotation language and guides the development of a data-parallel standard library.
 % \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
@@
 
-\nv{Make sure we use \ttt{width} everywhere}
-\nv{Decide on parallelizable regions}
-
 %% \blfootnote{$^{*}$Equal contribution.}
@@
 % oblivious? agnostic?  
-One option is manual tools such as GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
+One option is to use manual tools such as GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
   these tools are either command-unaware, and thus at risk of breaking program semantics, or too coarse-grained, and thus only capable of exploiting parallelism at the level of entire scripts rather than individual components.
@@
 To top it off, all these options assume a good understanding of parallelism;
-  users with domain of expertise outside computing---hobbyists to data analysts---are left without options.
+  users with domain of expertise outside computing---from hobbyists to data analysts---are left without options.
 %  (see~\S\ref{related} for a full discussion).
@@
 }
-\vspace{-15pt}
+\vspace{-25pt}
 \label{fig:overview}
@@
 % Aside from these two components, \sys packs a runtime component that serves dual purpose:
-Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data aggregators as well as high-performance primitives for eager data splitting and merging.
+Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data aggregators as well as high-performance primitives such as ones for eager data splitting and merging.
 These address many practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
@@
 
-% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a data-flow graph, enabling for parallelization optimizations.
+% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
 % First, a careful study of the parallelizability properties of shell primitives and commands.
@@
 % 
-% Second, \sys defines a data-flow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the data-flow graph.
+% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
 % (one sentence for the model)
@@
 and 
-  (iii) a series of micro-benchmarks inspired by (and
-comparing with) possible alternatives.  The results show
-significant advantages in correctness, manual effort,
-and performance---with 
-speedups $X-X\times$ due to \sys's transformations and $Y-Y\times$ due to \sys's runtime primitives.
+  (iii) a series of micro-benchmarks inspired by (and comparing with) possible alternatives.
+The vast majority (47/52) of scripts---all unmodified---see significant speedups.
+Indicatively, applied to 10 programs with parallelism \ttt{--width=64}, \sys achieves 9.35$\times$ due to program transformations alone and 13.24$\times$ when coupled with its runtime primitives.
+% \sys performs better than \ttt{sort --parallel}, a hand-tuned parallel implementation, and GNU parallel, which returns incorrect results if used without care.
+In another case of \ttt{--width=16}, \sys successfully accelerates a large program by 2.44$\times$---extracting 10.79$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
+
+% For the temperature analysis script (which is taken from a Hadoop case study), \sys not only 
+% contains pre-processing that takes X time and Y main processing (Hadoop focus).
+% \sys accelerates the entire pipeline by , including the fragment that is outside the focus of conventional parallelization systems.
+% accelerates the CPU-intensive part that is parallelized by Hadoop, but also for the preprocessing part which takes even more time than what the other part takes.
+% XX--YY speedups in the vast majority of scripts, ones taken from the wild 
+% 1. Sentence about transformations vs runtime primitives
+% 2. Unix 50? 
+% 3. point about Hadoop expressing a fraction of the pipeline and speedup
+% \sys achieves higher speedup than \ttt{sort --parallel} that involved hand-tuned parallel implementation by command developers.
+
+%   The results show
+% significant advantages in correctness, manual effort,
+% and performance---with 
+% speedups $X-X\times$ due to \sys's transformations and $Y-Y\times$ due to \sys's runtime primitives.
 %% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-Improvements depend on several factors such as a script's initial execution time and the parallelizability properties of its commands.
+% Improvements depend on several factors such as a script's initial execution time and the parallelizability properties of its commands.
 
@@
 \heading{Implementation} %
-On top of command and shell semantics, the \unix environment has its own set of quirks.
+On top of command and shell semantics, the broader \unix environment has its own set of quirks.
 Any attempt to orchestrate parallel execution will hit challenges related to task parallelism, deadlock prevention, and runtime performance.
-For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} might not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
+For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} may not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
 % A final challenge has to do with practical issues, as even manual parallelization of a given shell script is far from trivial.
@@
 % 
-% \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{parallelizable regions} explained below.
+% \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{dataflow regions} explained below.
 % nv idea: let's call them may-par regions
@@
 % First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model % ~\sx{dataflow-regions}
-%   by identifying and transforming special regions termed \emph{parallelizable regions} and explained below. % nv idea: let's call them may-par regions
+%   by identifying and transforming special regions termed \emph{dataflow regions} and explained below. % nv idea: let's call them may-par regions
 % \kk{eurosys: I suggest to replace the above sentence with the one below.
@@
 % \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
-\begin{minted}[fontsize=\footnotesize]{bash}
+\begin{minted}[fontsize=\footnotesize,escapeinside=()]{bash}
 mkfifo $t{0,1...}
@@
 cat $t15 | head -n1 > $out1 &
-wait $! && ./pash_get_pids | xargs -n 1 kill -SIGPIPE
+wait $! && get-pids | xargs -n 1 kill -SIGPIPE
 \end{minted}
 \caption{
-  \textbf{\sys's $2\times$-parallel output for fragment of Fig.~\ref{fig:example}.}
-  \sys orchestrates the script's parallel execution through named pipes, parallel operators and careful placement of combiners.
+  \textbf{Output of \ttiny{pash --width=2} for Fig.~\ref{fig:example} (fragment).}
+  \sys orchestrates parallel execution through named pipes, parallel operators, and custom runtime primitives---\eg \ttiny{eager}, \ttiny{sort-merge}, and \ttiny{get-pids}.
 }
@@
 \heading{Scripts}
-To maintain sequential semantics, \sys first analyzes a script to identify \emph{parallelizable regions} containing commands that are candidates for parallelization~\sx{dataflow-regions}.
+To maintain sequential semantics, \sys first analyzes a script to identify \emph{dataflow regions} containing commands that are candidates for parallelization~\sx{dataflow-regions}.
 This analysis is guided by the script structure: % to identify parallelization candidates:
-  some constructs expose task parallelism; others enforce synchronization.
+  some constructs expose parallelism; others enforce synchronization.
 % \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
@@
 % Since the shell representation only exposes coarse grained parallelization opportunities, 
-\sys then converts each parallelizable region to a \emph{dataflow graph} (DFG)~\sx{graph-components}, 
+\sys then converts each dataflow region to a \emph{dataflow graph} (DFG)~\sx{graph-components}, 
 %% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
-%% To achieve this, it converts sections of the script that are amenable to parallelization into a data-flow graph (DFG) representation.
+%% To achieve this, it converts sections of the script that are amenable to parallelization into a dataflow graph (DFG) representation.
 % Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
   a flexible representation that enables a series of local transformations to expose data parallelism, converting the graph into its parallel equivalent~\sx{ir:transformations}.
-%% , in which DFGs indicate the parallelizable regions.
+%% , in which DFGs indicate the dataflow regions.
 Further transformations compile the DFG back to a shell script that uses POSIX constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program~\sx{backend}.
@@
 
-Broadly speaking, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they need when processing their input (Tab.\ref{tab:classes}).
+Broadly speaking, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they mutate when processing their input (Tab.\ref{tab:classes}).
 %% how they interact with state.
@@
 These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
-They may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but return empty output for empty input.
+They may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but always return empty output for empty input.
 Workloads that use only stateless commands are trivial to parallelize:
@@
 This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
-However, it turns out to be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics.
+However, it may be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics.
 
@@
 \sys's annotations focus on three crucial concerns about a command:
-  (i) its parallelizability class,
-  (ii) its inputs and outputs, and the characteristics of its input consumption, and
+  (C1) its parallelizability class,
+  (C2) its inputs and outputs, and the characteristics of its input consumption, and
   % (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (iii) how flags affect its class, inputs, and outputs.
+  (C3) how flags affect its class, inputs, and outputs.
 The first concern was discussed extensively in the previous section; we now turn to the latter two.
@@
 There are a few reasons for this.
-One reason is due to \sys's DFG representation, which connects commands with each other;
-  to perform this connection correctly \sys needs to know the outputs of a command to correctly connect them with the inputs of the next.
+One reason is due to \sys's DFG representation, which connects commands with each other.
+  To perform this connection correctly, \sys needs to know the outputs of a command. % to correctly connect them with the inputs of the next.
 A second reason is due to ordering:
@@
 
-Command flags (options) are a particularly popular way to control a command's execution, directly affecting its parallelizability classification.
-Commands are thus assigned a default parallelizability class, which should then be refined by the set of flags the command uses.
+Command flags (options) are a particularly popular way of controlling a command's execution, directly affecting its parallelizability classification.
+Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
 For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.
@@
 \heading{Example Annotations}
-The parallelizability properties of each command are specified in an annotation record that contains several components, each addressing one of the key concerns.
-On the top level, each annotation record contains a set of clauses, each one identified by a predicate on the command arguments (Concern (iii)). 
-Each clause contains an assignment for the command, indicating its class (Concern (i)), and the sequence of its inputs and outputs (Concern (ii)).
+The parallelizability properties of each command are specified in an annotation record that contains several components, each addressing one of the key concerns C1--3.
+At the top level, each annotation record contains a set of clauses, each one identified by a predicate on the command arguments (C3). 
+Each clause contains an assignment for the command, indicating its class (C1), and the sequence of its inputs and outputs (C2).
 %  (a JSON object) that
@@
 % The first clause from its annotation record is shown below:
-Below is its annotation record (left) and the implementation of record's first clause (right):
+Below is its annotation record (left) and the implementation of the record's first clause (right):
 %
@@
 The second clause is symmetric to the first with respect to the second input. 
-The third clause is the general case, classifiying \ttt{comm} in \pur as it needs to read both inputs completely before outputing its results.
+The third clause is the general case, classifiying \ttt{comm} in \pur, as \ttt{comm} needs to read both inputs completely before outputing its results.
 Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument.
@@
 %% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
-\sys first lifts sections of the input script to the DFG
-representation, it performs transformations to expose parallelism (up to the desired \ttt{--width}), and
-then instantiates each DFG back to a parallel shell script.  A
+\sys first lifts sections of the input script to the DFG representation, 
+  then performs transformations to expose parallelism (up to the desired \ttt{--width}), 
+  and finally instantiates each DFG back to a parallel shell script.  A
 fundamental difference with other DFG models is that \sys's DFG encodes the
@@
   \caption{
-    \textbf{From Shell script AST to DFGs.}
+    \textbf{From a script AST to DFGs.}
     The AST on the left has two dataflow regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
@@
 
-\subsection{Front-end: From Shell Scripts to DFGs}
+\subsection{Frontend: From a Sequential Script to DFGs}
 \label{dataflow-regions}
@@
 % To optimize a script, \sys translates it to a DFG~\sx{ir}.
-% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
+% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{dataflow program regions} and a translation pass for converting them to DFGs.
 % 
@@
 %%   irreversible side effects, the ; and \&\& operators.}
-%% To address the first issue, we introduce the notion of parallelizable
+%% To address the first issue, we introduce the notion of dataflow
 %% program regions and a translation pass for converting them to a
@@
 \heading{Dataflow Regions}
-In order to apply the graph transformations that expose data parallelism, \sys first has to identify program sub-expressions that can be safely transformed to a dataflow graph, \ie sub-expressions that do not impose any scheduling or synchronization (e.g.
-by using \ttt{;}) and take a set of files as inputs and produce a set of files as outputs.
+In order to apply the graph transformations that expose data parallelism, \sys first has to identify program sub-expressions that can be safely transformed to a dataflow graph, 
+  \ie sub-expressions that (i) do not impose any scheduling or synchronization constraints (\eg by using \ttt{;}), and (ii) take a set of files as inputs and produce a set of files as outputs.
 % The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
@@
 These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
-Consider this fragment:
+Consider this code fragment:
 
@@
 Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
-Examples of such information includes the values of environment variables, strings that have not been expanded, and sub-shell constructs.
+Examples of such information include the values of environment variables, strings that have not been expanded, and sub-shell constructs.
 %% to the the safety properties that are mentioned above cannot be
 %% soundly checked statically.
-For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
+For safety purposes, \sys takes a conservative approach, avoiding parallelization of such nodes with incomplete information.
 It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, \eg  an environment variable passed as an argument to a command does not change its parallelizability class.
@@
 
-This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's runtime and custom primitives.
+This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom runtime primitives.
 %% it instantiates the \todo{custom} split and relay nodes as highly
@@
  mkfifo t1 t2
- grep "foo" in1 > t1 & grep "foo" in2 > t2 & cat t1 t2
+ grep "foo" f1 > t1 & grep "foo" f2 > t2 & cat t1 t2
 \end{minted}
@@
 Alternatively, consumers could wait for producers to complete before opening the file for reading (Fig. \ref{fig:eager}c);
-  however, this would insert artificial barriers impeding task-based parallelism and wasting disk resources---that is, this allows data parallelism at the expense of task parallelism.
+  however, this would insert artificial barriers impeding task-based parallelism and wasting disk resources---that is, this allows for data parallelism at the expense of task parallelism.
 % \unix does not provide an easy way out.
@@
 % We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
+
+In \sys's evaluation~\sx{eval}, these primitives have the names presented in Fig.~\ref{fig:eager}.
 
@@
 % write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet
+Unfortunately though, if a pipe has not been opened for writing yet,
 % ---possible for named pipes---
@@
 \heading{Highlights}
-In this paragraph we summarize results for \ttt{--width=16}, but \sys's evaluation reports on varying \ttt{width}s.
-Overall, \sys accelerates 47/52 scripts by 1.92--17.42$\times$; % Eurosys: results for 16!! \kk{I added them}
+This paragraph highlights results for \ttt{width=16}, but \sys's evaluation reports on varying \ttt{width}s (2--64).
+Overall, \sys accelerates 47/52 scripts---all unmodified---by 1.92--17.42$\times$; 
   for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
@@
 %  shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
-NFA-Regex is centered around an expensive \todo{NFA-based} backtracking expression and all of its commands are in \sta.
+NFA-Regex is centered around an expensive NFA-based backtracking expression and all of its commands are in \sta.
 % and the latter being an IO-intensive script.
@@
 % (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the optimized \sys, \ie with \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \{\todo{1.97, 3.5, 5.78, 8.83, 10.96, 13.47}\}$\times$, respectively.
-For \emph{No-Eager}, \ie \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93 7.46 9.35$\times$.
+Average speedups of the optimized \sys, \ie with \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \{1.97, 3.5, 5.78, 8.83, 10.96, 13.47\}$\times$, respectively.
+For \emph{No-Eager}, \ie \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93, 7.46, 9.35$\times$.
 % and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
@@
 There are two types of such cases. %  (that appear in \todo{five} plots).
-In the first, \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical for scripts that \sys does not add any \ttt{split} since the width of the DFG is constant;
+In the first, the \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical for scripts that \sys does not add \ttt{split} since the width of the DFG is constant;
 %  when \sys adds no \ttt{split} because the DFG width remains constant for the whole program;
-  conversely (for spell, bi-grams, and sort), when both lines are shown, \sys has added \ttt{split}s due to changes in the DFG width (e.g. due to a \npu command).
-In the second type, \emph{Pash w/o Split} (red) is identical to No Eager (green, hidden) and Blocking Eager (orange, hidden) because the input script features a command in \pur or \npu relatively early.
+  conversely, when both lines are shown (\eg Spell, Bi-grams, and Sort), \sys has added \ttt{split}s due to changes in the DFG width (e.g. due to a \npu command).
+In the second type, \emph{Pash w/o Split} (red) is identical to \emph{No-Eager} (green, hidden) and \emph{Blocking Eager} (orange, hidden) because the input script features a command in \pur or \npu relatively early.
 This command requires an \ttt{agg}regator, whose output is of width 1, and beyond which non-\ttt{split} configurations are sequential (and see no speedup).
@@
 % * why blocking is less than no eager
-As expected, scripts with commands only in \sta get linear speedup.
+As expected, scripts with commands only in \sta see linear speedup.
 \sys's \ttt{split} benefits scripts with \tpur or \tnpu commands, without negative effects on the rest.
@@
 Sort-sort illustrates the full spectrum of primitives:
-  (i) \sys's \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} 
+  (i) \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} 
   % while maintaining task parallelism
-  (ii) \sys gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
+  (ii) \emph{Pash} gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
 
@@
 %   then re-run them).} . 
-\sys executes each pipeline as-is, without any modification.\footnote{
-  \nv{Do we need this?}
-}
+\sys executes each pipeline as-is, without any modification.
   % if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
@@
 % \sys's analysis and generation overheads remain collectively under $5s$ even for the high-scalability benchmarks.
-% We decided to push \sys's limits by constructing an artificial 1000-stage parallelizable region---a worst-case workload that is nowhere near the ones observed in practice.
+% We decided to push \sys's limits by constructing an artificial 1000-stage dataflow region---a worst-case workload that is nowhere near the ones observed in practice.
 % We find the resulting times quite acceptable:
@@
 
-Concurrent to our work, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
+Concurrent to \sys, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
 POSH focuses on co-locating distributed operation, and introduces a novel scheduling algorithm. 
@@
 Its  DFG model, however, is different and captures ordering constraints.
-This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.\str
+This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.
 % \kk{eurosys: I could see this being deleted in need of space.}
@@
 \sys can be thought as a parallelizing source-to-source compiler coupled with a runtime component.
-The compiler operates by converting parallelizable regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
+The compiler operates by converting dataflow regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
 Characteristics are known by the  use of classification and annotations (for certain classes), which \sys offers for commands in POSIX and Coreutils used frequently.

### 2020-10-09 evaluation/plots/microbenchmarks-table.tex

@@
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllllllll}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
 \toprule
@@
 \midrule
-grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
+nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
 sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
@@
 wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
-grep-light ~&~ $3\times\tsta$ & 100~GB & 1m38.212s & 49 & 193 & 0.031s & 0.163s & IO-intensive, computation-light \\
 spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
 shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
-diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
+difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
 bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
-bi-grams-opt ~&~ $3\times\tsta, \tpur$ & 3~GB & 38m21.501s & 63 & 255 & 0.117s & 1.482s & optimized version of bigrams \\
-set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
+set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
 sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\

### 2020-10-09 paper/draft.tex
Minor modifications in the evaluation

@@
   For some pairs of configurations, \sys produces identical parallel scripts and thus only one is shown.
+  % \kk{FUTURE: Run and show all plots (with and without split for all experiments.)}
 }
@@
 % needing a \ttt{split} to regain width. 
-
 Finally, Tab.~\ref{tab:eval} shows that \sys's transformation time is negligible, and its COST~\cite{mcsherryscalability} is 2. %  for all benchmarks.
@@
 Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for 31 pipelines, with \ttt{width=}16 and 10GB inputs.
-It does not include 3 pipelines that execute in under $0.1s$ by using \ttt{head} to output the first line of input.
-The script index refers to the challenge level, in both the plots and the discussion.
+It does not include 3 pipelines that use \ttt{head} fairly early thereby finishing execution in under $0.1$ seconds.
+We refer to each pipeline using its x-axis index in Fig.~\ref{fig:unix50-individual}.
+% The script index refers to the challenge level, in both the plots and the discussion.
+% \kk{The above is not exactly true since we sorted them. If someone goes to take a look at the solution source they will be confused.}
 Average speedup is \todo{$6.02\times$}, % median is $6.48\times$, 
@@
 
-%% \kk{Out of all benchmarks, we have talked about the head and the
-%%   non-parallelizable ones in the paper. Maybe we also want to talk
-%%   about the long and the cheap ones. Also we might want to talk and
-%%   find out more about the sort bottleneck.}
-
-%% (OLD) Bad result descriptions:
-%%  2. Contains just a head -n 2, and then a cut.
-%%  5. Very fast (small input). Sequential time is 0m0.103s (If we increase the input we will show speedup)
-%%  18. TODO: Implement
-%%  20. Four corners (not parallelizable). So performance is reasonable
-%%      (atm it doesn't return the correct result for multiplied input though)
-%%  21. Last capital character of each line. Their solution is not general (and it is not parallelizable).
-%%      Also, the input is very small. Changing it to a more general one and
-%%      increasing input size leads to speedup. TODO: Run with large input
-%%  22. Exactly the same problem as the above.
-%%  24. (Not sure: It seems that the computation is trivial)
-%%  32. TODO: Implement
-
 \heading{Take-aways} \sys accelerates unmodified pipelines found in the wild;
   small tweaks can yield further improvements, showing that \sys-awareness and scripting expertise can improve results.
-Furthermore, \sys does not decelerate non-trivial computations, even when parallelism is impossible.
+Furthermore, \sys does not significantly decelerate non-parallelizable scripts.
+% \kk{eurosys: we cant say it doesnt decelarate since it does slow down some of them. If we dont like this sentence we can also delete it.}
+% non-trivial computations, even when parallelism is impossible.
 
@@
 The complete pipeline executes in 44m2s for five years (82GB) of data.
-\sys with \ttt{width=}16 leads to \todo{2.44$\times$} speedup, with different phases seeing different benefits:
-  \todo{1.99$\times$} (vs. 33m58s) for all the pre-processing and 
-  \todo{10.79$\times$} speedup (vs. 10m4s) for computing the max.
+\sys with \ttt{width=}16 leads to 2.52$\times$ speedup, with different phases seeing different benefits:
+  2.04$\times$ (vs. 33m58s) for all the pre-processing and 
+  12.31$\times$ speedup (vs. 10m4s) for computing the max.
 

### 2020-10-09 paper/draft.tex
Minor tweaks in evaluation paragraph in intro

@@
 The vast majority (47/52) of scripts---all unmodified---see significant speedups.
-Indicatively, applied to 10 programs with parallelism \ttt{--width=64}, \sys achieves 9.35$\times$ due to program transformations alone and 13.24$\times$ when coupled with its runtime primitives.
+Indicatively, applied to 10 representative \unix one-liners,
+%  with \ttt{--width=64}, 
+ \sys achieves 9.35$\times$ due to program transformations alone and 13.24$\times$ when coupled with its runtime primitives.
+% \kk{eurosys: Readers don't know what --width is here so it might make sense to not mention it.}
 % \sys performs better than \ttt{sort --parallel}, a hand-tuned parallel implementation, and GNU parallel, which returns incorrect results if used without care.
-In another case of \ttt{--width=16}, \sys successfully accelerates a large program by 2.44$\times$---extracting 10.79$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
+When applied to a large program used for temperature analysis, 
+  \sys successfully accelerates it by 2.52$\times$---
+  parallelizing both the computation as well as the preprocessing fragment (downloading, extracting, and cleaning the data) which traditionally falls outside of the focus of conventional parallelization systems even though it takes $75\%$ of the whole program's execution time.
+  % extracting 12.31$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
+  % \kk{\sys parallelizes the computation by 12.31 and not the preprocessing. It parallelizes the preprocessing by 2.}
 

### 2020-10-09 paper/draft.tex
merge paragraphs

@@
 % We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
-
 In \sys's evaluation~\sx{eval}, these primitives have the names presented in Fig.~\ref{fig:eager}.
@@
 
-Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they require significantly more \emph{user} effort (see \sx{intro} for distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
+Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they require significantly more \emph{user} effort (see \sx{intro} for the distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
 %% Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).

### 2020-10-09 paper/draft.tex
Remove some space above equations
Add updated program numbers in eval, slightly improve abstract, address Caleb's comment for PoSh

@@
 \begin{abstract}
-This paper presents \sys, a system aimed at parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
+This paper presents \sys, a system for parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
 % a source-to-source shell compiler and associated runtime library
-Given a script, \sys converts it to a dataflow graph, performs a series of analyses and transformations that aim at exposing parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-A lightweight annotation language allow command developers to express key parallelizability properties about their commands.
-An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---informs \sys's annotation language and guides the development of a data-parallel standard library.
+Given a script, \sys converts it to a dataflow graph, performs a series of transformations that expose parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
+A lightweight annotation language allows command developers to express key parallelizability properties about their commands.
+An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---informs said annotation language and guides the development of an optimized aggregator library that \sys uses.
 % \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation, combining several classic and modern \unix scripts, shows significant \todo{speedups} stemming from the combination of its program transformations and runtime primitives.
+\sys's extensive evaluation, combining several classic and modern \unix scripts, shows significant speedups (total average over 44 scripts is $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
+% The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
 \end{abstract}
@@
   (iii) a series of micro-benchmarks inspired by (and comparing with) possible alternatives.
-The vast majority (47/52) of scripts---all unmodified---see significant speedups.
+The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
 Indicatively, applied to 10 representative \unix one-liners,
@@
 %% , \ie satisfying the following equation:
-
+%
 \[
@@
 \]
-
+%
 %% This is a nice characterization, which I think is true in most
@@
 %% the equation above holds.
-
+%
 \noindent
@@
 $m$ and an associative aggregate $agg$, satisfying the following equation:
-
+%
 \[
@@
 \]
-
+%
 \noindent
@@
 This paragraph highlights results for \ttt{width=16}, but \sys's evaluation reports on varying \ttt{width}s (2--64).
-Overall, \sys accelerates 47/52 scripts---all unmodified---by 1.92--17.42$\times$; 
+% The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
+Overall, \sys accelerates 39/44 scripts---all unmodified---by 1.92--17.42$\times$; 
   for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
@@
 
-Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>50$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
+Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>40$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
   only 6 ($<3\%$) commands, outside this library, needed a single-record annotation~\sx{macro2}.
@@
 \sys with \ttt{width=}16 leads to 2.52$\times$ speedup, with different phases seeing different benefits:
-  2.04$\times$ (vs. 33m58s) for all the pre-processing and 
+  2.04$\times$ (vs. 33m58s) for all the pre-processing ($75\%$ of the total running time) and 
   12.31$\times$ speedup (vs. 10m4s) for computing the max.
@@
 
-Concurrent to \sys, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
+Developed independently and at the same time with \sys, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
 POSH focuses on co-locating distributed operation, and introduces a novel scheduling algorithm. 
@@
 \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-Applying \sys to classic and modern \unix scripts shows significant speedups---in one set of benchmarks, the compiler's average speedup is $9.3\times$, jumping to $13.2\times$ due to runtime optimizations.
-\sys is open source and available for download at \ttt{blinded}.
+Our evaluation of \sys on a wide variety of \emph{unmodified} scripts shows that the vast majority (39/44) see significant speedups (average over all is $6.7\times$, up to $61\times$).
+% Applying \sys to classic and modern \unix scripts shows significant speedups---in one set of benchmarks, the compiler's average speedup is $9.3\times$, jumping to $13.2\times$ due to runtime optimizations.
+\sys is open source and available for download online~[Blind].
 

### 2020-10-09 paper/draft.tex
Address another one of Caleb's comments

@@
 
-We evaluate \sys using
+We evaluate \sys on a set of 44 scripts including:
   (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing scripts;

### 2020-10-09 paper/draft.tex
Add text for short-combined
Thin down hightlights as Ben mentioned

@@
 The third clause is the general case, classifiying \ttt{comm} in \pur, as \ttt{comm} needs to read both inputs completely before outputing its results.
-Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument.
+Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument, and option \ttt{short-combined} indicates that flags can be combined, as in \ttt{comm -13}.
 % The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
@@
 % The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
-Overall, \sys accelerates 39/44 scripts---all unmodified---by 1.92--17.42$\times$; 
-  for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
+Overall, applying \sys to all 44 unmodified scripts accelerates 39 of them by 1.92--17.42$\times$; for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
+The total average speedup over all 44 benchmarks is $6.7\times$.
+\sys's runtime primitives offer significant benefits---for the 10 scripts that we measured with and without the runtime primitives they bump the average speedup from $5.9\times$ to $8.6\times$.
+% Overall, \sys accelerates 39/44 scripts---all unmodified---by 1.92--17.42$\times$; 
+  % for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
   % \kk{eurosys: I have an issue that I am not sure how to address.
 I would like to have the 61.1x here (which is due to 64 width)  but I don't see a good way of including it without sounding suspicious.}
-For a subset of our benchmarks~\sx{ours}, we decouple the speedup due to \sys's DFG transformations alone from the one achieved when including its runtime primitives:
-  the former averages $5.9\times$, jumping to $8.6\times$ due to the latter.
-A different subset ($30$ benchmarks) sees average speedups of $6.02\times$~\sx{unix50}.
-% , depending on the classed of commands involved
-Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands written in multiple languages see speedups of $2.44\times$ and $12.7\times$.
+% For a subset of our benchmarks~\sx{ours}, we decouple the speedup due to \sys's DFG transformations alone from the one achieved when including its runtime primitives:
+  % the former averages $5.9\times$, jumping to $8.6\times$ due to the latter.
+% A different subset ($30$ benchmarks) sees average speedups of $6.02\times$~\sx{unix50}.
+% Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands written in multiple languages see speedups of $2.44\times$ and $12.7\times$.
 \sys significantly outperforms \ttt{sort --parallel}, a hand-tuned parallel implementation, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.

### 2020-10-09 paper/draft.tex
Address some of Ben's nits

@@
 % 
-The effort expensed writing this script is low:
+The effort expended writing this script is low:
 % \footnote{
@@
 It then performs a series of DFG transformations that expose parallelism without breaking semantics, by expanding the DFG to the desired \ttt{width}.
-. % ~\sx{ir:transformations}.
+% ~\sx{ir:transformations}.
 Finally, \sys converts these DFGs back to a shell script augmented with \sys-provided commands. % ~\sx{optimizer}.
@@
 \heading{Complete Annotation Language}
-The complete annotation language contains 6 operators, one of which supports regural expressions.
-It was used to annotate 47 commands, totalling 708 lines of JSON---an effort that took about 2--3 hours.
+The complete annotation language contains 6 operators, one of which supports regular expressions.
+It was used to annotate 47 commands, totaling 708 lines of JSON---an effort that took about 2--3 hours.
 Annotation records are by default conservative so as to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
@@
 % If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
-\sys defines aggregators for many POSIX and GNU commands in \pur, dubbing as both its standard library and an exemplar for community efforts tackling other commands.
+\sys defines aggregators for many POSIX and GNU commands in \pur, doubling as both its standard library and an exemplar for community efforts tackling other commands.
 
@@
 %% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
-We start with the script-to-DFG tranlation~\sx{dataflow-regions}, 
+We start with the script-to-DFG translation~\sx{dataflow-regions}, 
   then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
@@
 % The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
-Overall, applying \sys to all 44 unmodified scripts accelerates 39 of them by 1.92--17.42$\times$; for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
+Overall, applying \sys to all 44 unmodified scripts accelerates 39 of them by 1.92--17.42$\times$; for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 0.99, 1.01$\times$).
 The total average speedup over all 44 benchmarks is $6.7\times$.

### 2020-10-09 paper/draft.tex
Three spelling fixes

@@
 The second clause is symmetric to the first with respect to the second input. 
-The third clause is the general case, classifiying \ttt{comm} in \pur, as \ttt{comm} needs to read both inputs completely before outputing its results.
+The third clause is the general case, classifying \ttt{comm} in \pur, as \ttt{comm} needs to read both inputs completely before outputting its results.
 Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument, and option \ttt{short-combined} indicates that flags can be combined, as in \ttt{comm -13}.
@@
 POSH focuses on co-locating distributed operation, and introduces a novel scheduling algorithm. 
-\sys instead focuses on the parallelization of CPU-intensive scripts, by transforming them to DFGs, applying transformations, and then transforming them back to parallel schell scripts augmented with \sys's runtime primitives to be executed on a standard shell.
+\sys instead focuses on the parallelization of CPU-intensive scripts, by transforming them to DFGs, applying transformations, and then transforming them back to parallel shell scripts augmented with \sys's runtime primitives to be executed on a standard shell.
 %  non-distributed parallelism, is location-agnostic, and focuses on correctness through its dataflow transformations.

### 2020-10-09 paper/draft.tex
cdot in 4.3 and nits

--- 2020-10-09 paper/draft.tex
@@
 
-% Another important feature is synchronization.
+% Another important feature is agg(synchronization.
 Different pipeline stages process data concurrently and possibly at different rates---\eg the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
@@
 base="ftp://ftp.ncdc.noaa.gov/pub/data/noaa";
-for y in {2015..2020}; do
+for y in {2015..2019}; do
  curl $base/$y | grep gz | tr -s" " | cut -d" " -f9 |
@@
   \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S. for the years 2015--2020 and extracts the maximum for every year.
+  The script downloads daily temperatures recorded across the U.S. for the years 2015--2019 and extracts the maximum for every year.
 }
@@
 cat $t11 | eager > $t13 & cat $t12 | eager > $t14 &
-sort-merge -rn $t13 $t14 > $t15 &
+sort -mrn $t13 $t14 > $t15 &
 cat $t15 | head -n1 > $out1 &
@@
   \textbf{Output of \ttiny{pash --width=2} for Fig.~\ref{fig:example} (fragment).}
-  \sys orchestrates parallel execution through named pipes, parallel operators, and custom runtime primitives---\eg \ttiny{eager}, \ttiny{sort-merge}, and \ttiny{get-pids}.
+  \sys orchestrates parallel execution through named pipes, parallel operators, and custom runtime primitives---\eg \ttiny{eager}, \ttiny{split}, and \ttiny{get-pids}.
 }
@@
 The first class, \sta, contains commands that operate on individual line elements of their input, without maintaining state across invocations.
-These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
+These are commands that can be expressed as a purely functional \emph{map} or \emph{filter}---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
 They may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but always return empty output for empty input.
@@
 For example, hashing commands such as \ttt{sha1sum} maintain complex state that has to be updated sequentially.
-If parallelized on a single input, each stage would need to wait on the results of all previous stages foregoing any parallelism benefits.
+If parallelized on a single input, each stage would need to wait on the results of all previous stages, foregoing any parallelism benefits.
 
@@
 \[
-\forall x, x', s, f(x \cdot x', s) = agg(m(x, s) \cdot m(x', s), s)
+\forall x, x', s, f(x \cdot x', s) = agg(m(x, s), m(x', s), s)
 \]
@@
   concatenation of two inputs $x, x'$ as by applying the aggregation function
-$agg$ to the concatenation of the outputs produced by applying map $m$
-to each of $x$ and $x'$.
+  $agg$ to the outputs produced by applying map $m$ to each of $x$ and $x'$.
 %% \km{We had also discussed here the case of
@@
 implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
-\cdots x_n, s) \Rightarrow agg(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n,
+\cdots x_n, s) \Rightarrow agg(m(x_1, s), m(x_2, s), \ldots m(x_n,
 s), s)$.
@@
 %
-\sys also performs a set of auxiliary transformations $t_{1-3}$that are
+\sys also performs a set of auxiliary transformations $t_{1-3}$ that are
 depicted in \Cref{fig:auxiliary-transformations}. If a node has many
@@
 %   Ben Karel,
+%   Caleb Stanford,
+%   Thurston Dang,
 %   Jean-S\'ebastien L\'egar\'e, and 

### 2020-10-09 paper/draft.tex
NOAA in eval

@@
 \heading{Discussion}
+% 
+The speedup of the preprocessing phase of the pipeline is bound by the network and IO costs since \ttt{curl} downloads 82GB of data.
+However, in the processing phase (which is CPU-bound) the speedup is $12.31\times$, much higher than what would be achieved by parallelizing per loop iteration.
 Similar to Unix50~\sx{unix50}, we found that large pipelines enable significant freedom in terms of expressiveness.
-A few stages of the original script were expressed in a single \ttt{awk};
+% A few stages of the original script were expressed in a single \ttt{awk};
 % unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
-  a simpler (but longer) pipeline that leverages \unix built-ins turns out to be trivially parallelizable.
+  % a simpler (but longer) pipeline that leverages \unix built-ins turns out to be trivially parallelizable.
 

### 2020-10-09 paper/draft.tex
Corrections

@@
 % As each command may exercise arbitrary behaviors,
-To be parallelized, each command may require special analysis and treatment---\eg exposing data parallelism in Fig.~\ref{fig:example}'s \ttt{sort} would require splitting its input, running \ttt{sort} on each part, and then merging the results.
+To be parallelized, each command may require special analysis and treatment---\eg exposing data parallelism in Fig.~\ref{fig:example}'s \ttt{tr} or \ttt{sort} would require splitting their inputs, running them on each partial input, and then merging the partial results.\footnote{
+  As explained in the introduction~\sx{intro}, commands such as \ttiny{sort} may have \emph{ad hoc} flags like \ttiny{--parallel}, which do not compose across commands and may risk breaking correctness and unexploiting performance potential~\sx{micro}.
+}
 Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}.
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2} for \ttt{--width=2} (\ie 2$\times$ \{\ttt{xargs}, \ttt{sort}, \etc\}).
 % concurrent version
@@
 and unusual compared to non-\unix systems\footnote{For example,
-  Windows indicates errors for \ttt{WriteFile} using its return
-  code---similar to \ttt{DeleteFile} and other Win32 functions.}
+  Windows indicates errors for \ttiny{WriteFile} using its return
+  code---similar to \ttiny{DeleteFile} and other Win32 functions.}
 primarily because pipes and pipelines are at the heart of \unix.

### 2020-10-09 paper/draft.tex
Minor corrections

@@
 A lightweight annotation language allows command developers to express key parallelizability properties about their commands.
-An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---informs said annotation language and guides the development of an optimized aggregator library that \sys uses.
+An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---guides the annotation language and optimized aggregator library that \sys uses.
 % \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation, combining several classic and modern \unix scripts, shows significant speedups (total average over 44 scripts is $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
+\sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.98$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
 % The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
@@
 
-We evaluate \sys on a set of 44 scripts including:
+We evaluate \sys 44 unmodified scripts including:
   (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing scripts;
@@
 and 
-  (iii) a series of micro-benchmarks inspired by (and comparing with) possible alternatives.
-The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
-Indicatively, applied to 10 representative \unix one-liners,
+  (iii) two micro-benchmarks inspired by (and comparing with) possible alternatives.
+Speedups range between 0.98--61.1$\times$ (avg: $6.7\times$), with the 39 out of 44 scripts seeing non-trivial speedups. 
+\sys's runtime primitives add to the base speedup extracted by \sys's program transformations---\eg 13.24$\times$ over a base 9.35$\times$ average for 10 representative \unix one-liners.
 %  with \ttt{--width=64}, 
- \sys achieves 9.35$\times$ due to program transformations alone and 13.24$\times$ when coupled with its runtime primitives.
 % \kk{eurosys: Readers don't know what --width is here so it might make sense to not mention it.}
 % \sys performs better than \ttt{sort --parallel}, a hand-tuned parallel implementation, and GNU parallel, which returns incorrect results if used without care.
-When applied to a large program used for temperature analysis, 
-  \sys successfully accelerates it by 2.52$\times$---
-  parallelizing both the computation as well as the preprocessing fragment (downloading, extracting, and cleaning the data) which traditionally falls outside of the focus of conventional parallelization systems even though it takes $75\%$ of the whole program's execution time.
+\sys accelerates a large program for temperature analysis by 2.52$\times$, parallelizing both the computation and the preprocessing fragment (\ie data download, extraction, and cleanup), which traditionally falls outside of the focus of conventional parallelization systems---even though it takes $75\%$ of the total execution time.
   % extracting 12.31$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
@@
 }
+% Key point: ...analyses and treatment---\eg combining partial results for \ttt{tr} is different than for \ttt{sort}.
 Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
@@
 %%   good as it could be.}
-We first compare a GNU \ttt{sort} parallelized by \sys ($S_{PaSh}$) \vs the same \ttt{sort} with the 
+We first compare a GNU \ttt{sort} parallelized by \sys ($S_{p}$) \vs the same \ttt{sort} with the 
 %% \begin{wrapfigure}[10]{r}{0.25\textwidth}
@@
 %% \end{wrapfigure}
-\ttt{--parallel} flag set ($S_{GNU}$). % , shown on the side.
+\ttt{--parallel} flag set ($S_{g}$).\footnote{
+  Both \ttiny{sort}s use the same buffer size internally~\cite{sortp:15}.
+}
 % The results are shown on the side-figure below.
 While \ttt{--parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{GNU}$'s parallelism is configured to $2\times$ that of \todo{$S_{PaSh}$'s} \ttt{--width} (\ie the rightmost plot point for $S_{GNU}$ is for \ttt{--parallelism=128}), 
+$S_{g}$'s parallelism is configured to $2\times$ that of $S_{p}$'s \ttt{--width} (\ie the rightmost plot point for $S_{g}$ is for \ttt{--parallelism=128}), 
 to account for \sys's additional runtime processes.
@@
 A few points are worth noting.
-$S_{PaSh}$ without \ttt{eager} performs comparably to $S_{GNU}$, and with \ttt{eager} it outperforms $S_{GNU}$ ($\sim2\times$);
+$S_{p}$ without \ttt{eager} performs comparably to $S_{g}$, and with \ttt{eager} it outperforms $S_{g}$ ($\sim2\times$);
   this is because \ttt{eager} adds intermediate buffers %(on disk)
@@
 %% purpose of having a simple solution.
-$S_{GNU}$ indicates that \ttt{sort}'s scalability is inherently limited (\ie due to \ttt{sort} and not \sys);
+$S_{g}$ indicates that \ttt{sort}'s scalability is inherently limited (\ie due to \ttt{sort}, not \sys);
   this is why all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) are capped at $8\times$ speedup.
-Finally, the comparison also shows \sys's benefits to command developers:
+The comparison also shows \sys's benefits to command developers:
   a low-effort parallelizability annotation achieves better scalability than a custom flag (and underlying parallel implementation) manually added by developers.

### 2020-10-09 paper/draft.tex
Fix todos

@@
 \caption{
-  \textbf{\sys overview and correspondence to sections.}
+  \textbf{\sys overview.}
   \sys identifies dataflow regions~\sx{dataflow-regions}, 
-  converts them to a dataflow graph~\sx{graph-components}, applies iterative
-  transformations~\sx{ir:transformations} based on the parallelizability properties (\S\ref{cmd}, \S\ref{ext}) 
-  of the commands used in these regions, and finally emits a parallel script
-  that uses custom primitives~\sx{optimizer}.
+  converts them to dataflow graphs~\sx{graph-components},
+  applies transformations~\sx{ir:transformations} based on the parallelizability properties  of the commands in these regions(\S\ref{cmd}, \S\ref{ext}),
+  and emits a parallel script that uses custom primitives~\sx{optimizer}.
 }
-\vspace{-25pt}
+\vspace{-18pt}
 \label{fig:overview}
@@
       (ii) PaSh w/o split: \ttiny{eager} enabled (no \ttiny{split}),
-      (iii) \todo{Blocking Eager}, only blocking \ttiny{eager} enabled (no \ttiny{split}),
-      (iv) \todo{No Eager}: both \ttiny{eager} and \ttiny{split} disabled.
+      (iii) Blocking Eager: only blocking \ttiny{eager} enabled (no \ttiny{split}),
+      (iv) No Eager: both \ttiny{eager} and \ttiny{split} disabled.
   For some pairs of configurations, \sys produces identical parallel scripts and thus only one is shown.
@@
 % \kk{The above is not exactly true since we sorted them. If someone goes to take a look at the solution source they will be confused.}
-Average speedup is \todo{$6.02\times$}, % median is $6.48\times$, 
+Average speedup is $6.02\times$, % median is $6.48\times$, 
   and weighted average (with the absolute times as weights) is $5.75\times$.
@@
 \heading{Take-aways}
-\sys can be applied to programs of notable size \todo{and complexity} to offer significant acceleration.
+\sys can be applied to programs of notable size and complexity to offer significant acceleration.
 A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive, \ie the usual focus of conventional parallelization systems.

### 2020-10-09 paper/draft.tex
phone nits

@@
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.98$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
+\sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.89$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
 % The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
@@
 
-We evaluate \sys 44 unmodified scripts including:
+We evaluate \sys on 44 unmodified scripts including:
   (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing scripts;
@@
   (iii) two micro-benchmarks inspired by (and comparing with) possible alternatives.
-Speedups range between 0.98--61.1$\times$ (avg: $6.7\times$), with the 39 out of 44 scripts seeing non-trivial speedups. 
-\sys's runtime primitives add to the base speedup extracted by \sys's program transformations---\eg 13.24$\times$ over a base 9.35$\times$ average for 10 representative \unix one-liners.
+Speedups range between 0.89--61.1$\times$ (avg: $6.7\times$), with the 39 out of 44 scripts seeing non-trivial speedups. 
+\sys's runtime primitives add to the base speedup extracted by \sys's program transformations---\eg 8.83$\times$ over a base 5.93$\times$ average for 10 representative \unix one-liners.
 %  with \ttt{--width=64}, 
@@
 % \sys performs better than \ttt{sort --parallel}, a hand-tuned parallel implementation, and GNU parallel, which returns incorrect results if used without care.
-\sys accelerates a large program for temperature analysis by 2.52$\times$, parallelizing both the computation and the preprocessing fragment (\ie data download, extraction, and cleanup), which traditionally falls outside of the focus of conventional parallelization systems---even though it takes $75\%$ of the total execution time.
+\sys accelerates a large program for temperature analysis by 2.52$\times$, parallelizing both the computation (12.31$\times$) and the preprocessing (2.04$\times$) fragment (\ie data download, extraction, and cleanup), which traditionally falls outside of the focus of conventional parallelization systems---even though it takes $75\%$ of the total execution time.
   % extracting 12.31$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
@@
 curl $base/$y > $t0 & cat $t0 | split $t1 $t2 &
-cat $t1 | xargs -n1 curl -s >$t3 & 
-cat $t2 | xargs -n1 curl -s >$t4 &
+cat $t1 | grep gz > $t3 & 
+cat $t2 | grep gz > $t4 &
 ...

### 2020-10-09 paper/draft.tex
Fixes

--- 2020-10-09 paper/draft.tex
@@
 
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width} and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2} for \ttt{--width=2} (\ie 2$\times$ \{\ttt{xargs}, \ttt{sort}, \etc\}).
+At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width}, and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}---for \ttt{--width=2}, \ie 2$\times$ \{\ttt{grep}, \ttt{tr}, \ttt{cut}, \etc\}.
 % concurrent version
@@
 \bottomrule
+\vspace{-18pt}
 \end{tabular*}
@@
 
-\sys's core is an abstract dataflow graph (DFG) model that is used as the intermediate representation on which \sys performs parallelization-exposing transformations.
+\sys's core is an abstract dataflow graph (DFG) model~\sx{graph-components} that is used as the intermediate representation on which \sys performs parallelization-exposing transformations~\sx{ir:transformations}.
 %% representation that is expressive enough to model data parallelism in shell programs.
 %% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
-\sys first lifts sections of the input script to the DFG representation, 
-  then performs transformations to expose parallelism (up to the desired \ttt{--width}), 
-  and finally instantiates each DFG back to a parallel shell script.  A
+\sys first lifts sections of the input script to the DFG representation~\sx{dataflow-regions}, 
+  then performs transformations to expose parallelism (up to the desired \ttt{--width})~\sx{ir:transformations}, 
+  and finally instantiates each DFG back to a parallel shell script~\sx{backend}.  A
 fundamental difference with other DFG models is that \sys's DFG encodes the
@@
 %% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
-We start with the script-to-DFG translation~\sx{dataflow-regions}, 
-  then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
-  discuss the transformations that \sys performs~\sx{ir:transformations}, and close with
-  the DFG-to-script translation~\sx{backend}.
+% The section script-to-DFG translation~\sx{dataflow-regions}, 
+%   then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
+%   discuss the transformations that \sys performs~\sx{ir:transformations}, and close with
+%   the DFG-to-script translation~\sx{backend}.
 
@@
 % https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.48\textwidth]{\detokenize{./figs/pash_eager.pdf}}
+\includegraphics[width=0.42\textwidth]{\detokenize{./figs/pash_eager.pdf}}
 \caption{
@@
 % https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.4\textwidth]{\detokenize{./figs/pash_lattice.pdf}}
+\includegraphics[width=0.37\textwidth]{\detokenize{./figs/pash_lattice.pdf}}
 \caption{

### 2020-10-09 paper/draft.tex
Stab at conclusion

@@
 % a source-to-source shell compiler and associated runtime library
-\sys can be thought as a parallelizing source-to-source compiler coupled with a runtime component.
-The compiler operates by converting dataflow regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
-Characteristics are known by the  use of classification and annotations (for certain classes), which \sys offers for commands in POSIX and Coreutils used frequently.
-\sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-Our evaluation of \sys on a wide variety of \emph{unmodified} scripts shows that the vast majority (39/44) see significant speedups (average over all is $6.7\times$, up to $61\times$).
+\sys consists of (i) a source-to-source compiler, which transforms scripts to DFGs, parallelizes them, and transforms them back to scripts; and (ii) a runtime component that addresses several practical challenges related to performance and correctness.
+Our evaluation on a wide variety of \emph{unmodified} scripts shows that \sys achieves significant speedups due to parallelism (total avg: $6.7\times$) in the vast majority of them (39/44), including scripts that perform mundane tasks such as downloading, extracting, and cleaning data, which are usually not the focus of parallelization systems.
+\sys is open source and available for download online~[Blind].
+
+% The compiler operates by converting dataflow regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
+% Characteristics are known by the  use of classification and annotations (for certain classes), which \sys offers for commands in POSIX and Coreutils used frequently.
+% \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
+% Our evaluation of \sys on a wide variety of \emph{unmodified} scripts shows that the vast majority (39/44) see significant speedups (average over all is $6.7\times$, up to $61\times$).
 % Applying \sys to classic and modern \unix scripts shows significant speedups---in one set of benchmarks, the compiler's average speedup is $9.3\times$, jumping to $13.2\times$ due to runtime optimizations.
-\sys is open source and available for download online~[Blind].
 

### 2020-10-09 paper/draft.tex
Update abstract
Conclusion

@@
 This paper presents \sys, a system for parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
-% a source-to-source shell compiler and associated runtime library
 Given a script, \sys converts it to a dataflow graph, performs a series of transformations that expose parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
@@
 An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---guides the annotation language and optimized aggregator library that \sys uses.
-% \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
 \sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.89$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
+\end{abstract}
+
+% a source-to-source shell compiler and associated runtime library
+% \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
 % The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
-\end{abstract}
 
@@
 
-This paper presented \sys, a system allowing shell users to parallelize shell programs mostly automatically.
-% a source-to-source shell compiler and associated runtime library
-\sys consists of (i) a source-to-source compiler, which transforms scripts to DFGs, parallelizes them, and transforms them back to scripts; and (ii) a runtime component that addresses several practical challenges related to performance and correctness.
-Our evaluation on a wide variety of \emph{unmodified} scripts shows that \sys achieves significant speedups due to parallelism (total avg: $6.7\times$) in the vast majority of them (39/44), including scripts that perform mundane tasks such as downloading, extracting, and cleaning data, which are usually not the focus of parallelization systems.
-\sys is open source and available for download online~[Blind].
+Shell programs are ubiquitous, utilize programs from a plethora of programming languages, and spend a significant fraction of their time interacting with the broader environment to download, extract, and process data---falling outside the focus of conventional parallelization systems.
+This paper presents \sys, a system that allows shell users to parallelize shell programs mostly automatically.
+\sys can be viewed as (i) a source-to-source compiler that transforms scripts to DFGs, parallelizes them, and transforms them back to scripts; coupled with (ii) a  runtime component that addresses several practical challenges related to performance and correctness.
+\sys's extensive evaluation over 44 unmodified Unix scripts demonstrates non-trivial speedups (0.89--61.1$\times$, avg: 6.7$\times$).
+\sys is open source and available for download~[Blind].
 

### 2020-10-11 paper/draft.tex
Small update for arxiv

@@
 %% EuroSys 2021 -- 12 pages limit
-\documentclass[sigplan, review, screen, anonymous]{acmart}
+\documentclass[sigplan, screen]{acmart}
 \acmSubmissionID{49}
@@
 
-%% \author{Nikos Vasilakis}
-%% \authornote{The two marked authors contributed equally to the paper.}
-%% \affiliation{MIT}
-
-%% \author{Konstantinos Kallas}
-%% \authornotemark[1]
-%% \affiliation{University of Pennsylvania}
-
-%% \author{Konstantinos Mamouras}
-%% \affiliation{Rice University}
-
-%% \author{Achilleas Benetopoulos}
-%% \affiliation{Unaffiliated}
-
-%% \author{Lazar Cvetkovich}
-%% \affiliation{University of Belgrade}
+\author{Nikos Vasilakis}
+\authornote{The two marked authors contributed equally to the paper.}
+\affiliation{MIT}
+
+\author{Konstantinos Kallas}
+\authornotemark[1]
+\affiliation{University of Pennsylvania}
+
+\author{Konstantinos Mamouras}
+\affiliation{Rice University}
+
+\author{Achilleas Benetopoulos}
+\affiliation{Unaffiliated}
+
+\author{Lazar Cvetkovich}
+\affiliation{Unaffiliated}
 
@@
 %% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+%% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
 
@@
 \begin{abstract}
-This paper presents \sys, a system for parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
-Given a script, \sys converts it to a dataflow graph, performs a series of transformations that expose parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
+This paper presents \sys, a system for parallelizing POSIX shell scripts. %  through a combination of program transformations and runtime primitives.
+Given a script, \sys converts it to a dataflow graph, performs a series of semantics-preserving program transformations that expose parallelism, and then converts the dataflow graph back into a script---one that adds POSIX constructs to explicitly guide parallelism coupled with \sys-provided \unix-aware runtime primitives for addressing per\-for\-mance- and correctness-related issues.
 A lightweight annotation language allows command developers to express key parallelizability properties about their commands.
 An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---guides the annotation language and optimized aggregator library that \sys uses.
-Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
+Finally, \sys's 
 \sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.89$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
@@
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
@@
 % The first clause from its annotation record is shown below:
-Below is its annotation record (left) and the implementation of the record's first clause (right):
+
+Fig.~\ref{annotation} presents its annotation record (left) and the implementation of the record's first clause (right).
 %
@@
 
-\hspace*{-0.3cm}\includegraphics[width=\columnwidth]{\detokenize{./figs/pash_annotation.pdf}}
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{\detokenize{./figs/pash_annotation.pdf}}
+  \vspace{-15pt}
+  \caption{
+    \textbf{Parallelizability annotation.}
+    Annotation record for \ttiny{comm} (left) and the implementation of the record's first clause (right).
+  }
+  \vspace{-15pt}
+  \label{fig:annotation}
+\end{figure}
 
@@
 
-\noindent
 As the first and third columns are suppressed, \ttt{comm} can be categorized in \sta by regarding its first file argument as a static ``configuration'' input;
@@
 \sys's extensive evaluation over 44 unmodified Unix scripts demonstrates non-trivial speedups (0.89--61.1$\times$, avg: 6.7$\times$).
-\sys is open source and available for download~[Blind].
+
+\sys's implementation, as well as all the example code and benchmarks presented in this paper, are all open source and available for download:
+\href{https://github.com/andromeda/pash}{github.com/andromeda/pash}.
 
@@
 % \subsection*{Acknowledgements}
-% % Dumping people so that we don't forget
-% We want to thank 
-%   Andr\'e DeHon,
-%   Ben Karel,
-%   Caleb Stanford,
-%   Thurston Dang,
-%   Jean-S\'ebastien L\'egar\'e, and 
-%   Nick Roessler.
-% This material is based upon work supported by DARPA contract no. HR00112020013 (ALRSCC) and NSF award CCF 1763514.
+\begin{acks}
+% Dumping people so that we don't forget
+We want to thank 
+  Andr\'e DeHon,
+  Ben Karel,
+  Caleb Stanford,
+  Thurston Dang,
+  Jean-S\'ebastien L\'egar\'e, and 
+  Nick Roessler.
+This material is based upon work supported by DARPA contract no. HR00112020013 and NSF award CCF 1763514.
+\end{acks}
 

### 2020-10-31 compiler/gather_results.py
Refactor and improve the plotting script

@@ -114,12 +114,90 @@ input_filename_sizes = {"1G": "1~GB",
                         "100M": "100~MB",
                         "all_cmds_x1000": "85~MB"}
 
+suffix_to_runtime_config = {"distr": "eager",
+                            "distr_auto_split": "split",
+                            "distr_no_task_par_eager": "blocking-eager",
+                            "distr_no_eager": "no-eager"}
+
+
+
+class Config:
+    def __init__(self, pash, width=None, runtime=None):
+        if(pash):
+            assert(Config.is_runtime_valid(runtime))
+            assert(isinstance(width, int))
+            self.pash = True
+            self.width = width
+            self.runtime = runtime
+        else:
+            self.pash = False
+    
+    def __repr__(self):
+        if(self.pash):
+            return("PaSh Config(width={}, runtime={})".format(self.width, self.runtime))
+        else:
+            return("No PaSh")
+
+    def is_runtime_valid(runtime):
+        possible_runtimes = ["no-eager",
+                             "blocking-eager",
+                             "eager",
+                             "mini-split",
+                             "split"]
+        return runtime in possible_runtimes
+
+
+class Result:
+    def __init__(self, script, config, value, description=""):
+        self.script = script
+        self.config = config
+        self.value = value
+        self.description = description
+    
+    def __repr__(self):
+        return("Result(description={}, script={}, value={}, config={})".format(self.description,
+                                                                               self.script,
+                                                                               self.value,
+                                                                               self.config))
+
+    def __truediv__(self, other):
+        if not isinstance(other, (int, float, Result)):
+            return NotImplemented
+
+        ## TODO: Change that to Result too
+        return self.value / other
+
+    def __rtruediv__(self, other):
+        if not isinstance(other, (int, float, Result)):
+            return NotImplemented
+
+        ## TODO: Change that to Result too
+        return other / self.value
+
+    def __add__(self, other):
+        if not isinstance(other, (int, float, Result)):
[…]

### 2020-11-02 evaluation/plots/microbenchmarks-table.tex
Refactor plotting scripts

@@
 spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
-shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
 difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
@@
 sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
 \bottomrule

### 2020-11-02 compiler/gather_results.py

@@ -28,29 +28,18 @@ UNIX50_RESULTS = "../evaluation/results/unix50/"
 COARSE_UNIX50_RESULTS = "../evaluation/results/unix50-naive/"
 
 
-experiments = ["minimal_grep",
-               "minimal_sort",
-               "topn",
-               "wf",
-            #    "grep",
-               "spell",
-               "shortest_scripts",
-               "diff",
-               "bigrams",
-            #    "alt_bigrams",
-               "set-diff",
-               "double_sort"]
-
-coarse_experiments = ["minimal_grep",
-                      "minimal_sort",
-                      "topn",
-                      "wf",
-                      "spell",
-                      "diff",
-                      "bigrams",
-                      "set-diff",
-                      "shortest_scripts"]
-
+all_experiments = ["minimal_grep",
+                   "minimal_sort",
+                   "topn",
+                   "wf",
+                   "grep",
+                   "spell",
+                   "shortest_scripts",
+                   "diff",
+                   "bigrams",
+                   "alt_bigrams",
+                   "set-diff",
+                   "double_sort"]
 
 pretty_names = {"minimal_grep" : "nfa-regex",
                 "minimal_sort" : "sort",
@@ -505,19 +494,15 @@ def plot_scaleup_lines(experiment, all_scaleup_numbers, all_speedup_results, cus
 
 ## TODO: Rename this. At the moment it is used to produce a pdf plot for each, 
 ##       but maybe we should produce a report of all of them together instead.
-def collect_scaleup_times(experiment, results_dir):
-
-
+## TODO: Once add the xlabels in the resultvector obhect we can remove the second
+##       argument here.
+def plot_scaleup_times(experiment, all_scaleup_numbers, all_speedup_results):
     fig, ax = plt.subplots()
 
     ## Plot speedup
     ax.set_ylabel('Speedup')
     ax.set_xlabel('--width')
 
-    # ax.plot(all_scaleup_numbers, total_distr_speedup, '-^', linewidth=0.5, label='+ Merge')
-    # ax.plot(all_scaleup_numbers, all_scaleup_numbers, '-', color='tab:gray', linewidth=0.5, label='Ideal')
-    all_scaleup_numbers = [2, 4, 8, 16, 32, 64]
-    all_speedup_results, output_diff = collect_scaleup_line_speedups(experiment, all_scaleup_numbers, results_dir)
[…]

### 2020-11-11 compiler/gather_results.py
Add an aggregate coarse plot

@@ -4,6 +4,8 @@ import math
 import os
 import numpy as np
 import matplotlib.pyplot as plt
+import matplotlib.lines as pltlines
+import matplotlib.ticker as plticker
 import statistics
 
 SMALL_SIZE = 16
@@ -117,6 +119,25 @@ class LinePlotConfig:
     
     def plot(self, xvalues, yvalues, ax):
         return ax.plot(xvalues, yvalues, self.linestyle, linewidth=self.linewidth, color=self.color, label=self.label)
+
+    ## The following two methods are needed since the linestyle argument in plot
+    ## does not correspond to the one in lines2D
+    def get_ls(self):
+        linestyle = self.linestyle
+        if(len(linestyle) == 2 and linestyle[0] == '-' and not linestyle[1] == '-'):
+            return '-'
+        return linestyle
+
+    def get_marker(self):
+        linestyle = self.linestyle
+        if(len(linestyle) == 2 and linestyle[0] == '-' and not linestyle[1] == '-'):
+            return linestyle[1]
+        return None
+
+    def get_artist(self):
+        marker = self.get_marker()
+        linestyle = self.get_ls()
+        return pltlines.Line2D([], [], marker=marker, linestyle=linestyle, linewidth=self.linewidth, color=self.color, label=self.label)
 
 default_line_plot_configs = {'eager': LinePlotConfig('-D', 'tab:red', 'Parallel w/o split'),
                              'split': LinePlotConfig('-o', 'tab:blue', 'Parallel'),
@@ -895,6 +916,18 @@ def collect_unix50_coarse_scaleup_times(unix50_results_dir):
     plt.savefig(os.path.join('../evaluation/plots', "unix50_coarse_throughput_scaleup.pdf"),bbox_inches='tight')
 
 
+def get_statistics_from_lines(lines):
+    minimums = lines[0].get_ydata().copy()
+    maximums = lines[0].get_ydata().copy()
+    sums = [0 for _ in lines[0].get_ydata()]
+    for line in lines:
+        for i, y in enumerate(line.get_ydata()):
+            minimums[i] = min(minimums[i], y)
+            maximums[i] = max(maximums[i], y)
+            sums[i] += y
+    avgs = [float(s)/len(lines) for s in sums]
+    return minimums, maximums, avgs
+
 def set_tiling_axes_labels_ticks(fig):
     axs = fig.get_axes()
     for ax in axs:
@@ -918,8 +951,8 @@ def plot_tiling_experiments(fig, gs, experiments, all_experiment_results, all_sc
[…]

### 2020-11-11 evaluation/plots/microbenchmarks-table.tex
Add an aggregate coarse plot

@@
 set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
-sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~MB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\

### 2020-11-12 compiler/gather_results.py
Add a table that has the scripts in it

@@ -584,6 +584,22 @@ def collect_input_size(experiment):
     clean_name = input_file_name.split('/')[-1].split('.')[0]
     return clean_name
 
+def collect_script(experiment):
+    # print(experiment)
+    script_file = os.path.join(MICROBENCHMARKS, '{}.sh'.format(experiment))
+    with open(script_file) as file:
+        script_lines = file.readlines()
+        clean_script_lines = [line.lstrip() for line in script_lines
+                              if not line.lstrip().startswith('#')]
+        clean_script_lines = [line.split('#')[0].rstrip(' ') for line in clean_script_lines]
+        script = "".join(clean_script_lines).rstrip()
+        script = script.replace("|\n", "| ")
+        script = script.replace("&\n", "& ")
+        script = script.replace("\n", "; ")
+        print(script)
+    wrapped_script = '\lstinline[columns=fixed,basicstyle=\\footnotesize\\ttfamily]!{}!'.format(script)
+    return wrapped_script
+
 def format_time_seconds(time_milliseconds):
     time_seconds = time_milliseconds / 1000
     time_minutes = int(time_seconds // 60)
@@ -606,12 +622,16 @@ def format_time_milliseconds(time_milliseconds):
 
 def generate_table_header(full=True):
     header = []
-    header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} llllllll}']
-    header += ['\\toprule']
     if(not full):
-        header += ['Script ~&~ Structure & Input &'
-                   'Seq. Time & Highlights \\\\']
+        # header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} llll}']
+        header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} l}']
+        header += ['\\toprule']
+        # header += ['Script ~&~ Structure & Input &'
+        #            'Seq. Time & Highlights \\\\']
+        header += ['Name ~&~ Script \\\\']
     else:
+        header += ['\\begin{tabular*}{\\textwidth}{l @{\\extracolsep{\\fill}} llllllll}']
+        header += ['\\toprule']
         header += ['Script ~&~ Structure & Input &'
                    'Seq. Time & \\multicolumn{2}{l}{\\#Nodes(16, 64)} &'
                    '\\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\\\']
@@ -627,8 +647,13 @@ def generate_table_footer(full=True):
[…]

### 2020-11-12 evaluation/plots/coarse-microbenchmarks-table.tex
Add a table that has the scripts in it

--- 2020-07-16 evaluation/plots/coarse-microbenchmarks-table.tex
@@
-\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} lllll}
+\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} l}
 \toprule
-Script ~&~ Structure & Input &Seq. Time & Highlights \\
+Name ~&~ Script \\
 \midrule
-grep ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & complex NFA regex \\
-sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & \tti{sort}ing \\
-top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & double \tti{sort}, \tti{uniq} reduction \\
-wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & double \tti{sort}, \tti{uniq} reduction \\
-spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & comparisons (\tti{comm}) \\
-diff ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & non-parallelizable \tti{diff}ing \\
-bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & stream shifting and merging \\
-set-diff ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & two pipelines merging to a \tti{comm} \\
-shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & long \tsta pipeline ending with \tpur \\
+nfa-regex ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr A-Z a-z | grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4'! \\
+sort ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr A-Z a-z | sort! \\
+top-n ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z '\n' | tr A-Z a-z | sort | uniq -c | sort -rn | sed ${N}q! \\
+wf ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z'\n' | tr A-Z a-z | sort | uniq -c | sort -rn | sed ${1}q! \\
+spell ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | col -bx |tr -cs A-Za-z '\n' | tr A-Z a-z |tr -d '[:punct:]' |sort |uniq |comm -23 - $dict! \\
+difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!mkfifo s1 s2; cat $IN | sort | tr [:lower:] [:upper:] > s1 & cat $IN | sort | tr [:upper:] [:lower:] > s2 & diff -B s1 s2; rm s1 s2!
+\\
+bi-grams ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z '\n' | tr A-Z a-z | bigrams_aux | sort | uniq! \\
+set-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!mkfifo s1 s2; cat $IN | cut -d ' ' -f 1 | sort > s1 & cat $IN | cut -d ' ' -f 1 | sort > s2 & comm -23 s1 s2; rm s1 s2!
+\\
+shortest-scripts ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | xargs file | grep "shell script" | cut -d: -f1 | xargs -L 1 wc -l | grep -v '^0$' | sort -n | head -15! \\
 \bottomrule

### 2020-11-12 compiler/gather_results.py
Remove mkfifo from table

@@ -592,6 +592,10 @@ def collect_script(experiment):
         clean_script_lines = [line.lstrip() for line in script_lines
                               if not line.lstrip().startswith('#')]
         clean_script_lines = [line.split('#')[0].rstrip(' ') for line in clean_script_lines]
+        ## Remove mkfifo, cat, rm
+        clean_script_lines = [line for line in clean_script_lines
+                              if not line.startswith("mkfifo") and
+                                 not line.startswith("rm")]
         script = "".join(clean_script_lines).rstrip()
         script = script.replace("|\n", "| ")
         script = script.replace("&\n", "& ")

### 2020-11-12 evaluation/plots/coarse-microbenchmarks-table.tex
Remove mkfifo from table

@@
 spell ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | col -bx |tr -cs A-Za-z '\n' | tr A-Z a-z |tr -d '[:punct:]' |sort |uniq |comm -23 - $dict! \\
-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!mkfifo s1 s2; cat $IN | sort | tr [:lower:] [:upper:] > s1 & cat $IN | sort | tr [:upper:] [:lower:] > s2 & diff -B s1 s2; rm s1 s2!
-\\
+difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | sort | tr [:lower:] [:upper:] > s1 & cat $IN | sort | tr [:upper:] [:lower:] > s2 & diff -B s1 s2! \\
 bi-grams ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z '\n' | tr A-Z a-z | bigrams_aux | sort | uniq! \\
-set-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!mkfifo s1 s2; cat $IN | cut -d ' ' -f 1 | sort > s1 & cat $IN | cut -d ' ' -f 1 | sort > s2 & comm -23 s1 s2; rm s1 s2!
-\\
+set-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | cut -d ' ' -f 1 | sort > s1 & cat $IN | cut -d ' ' -f 1 | sort > s2 & comm -23 s1 s2! \\
 shortest-scripts ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | xargs file | grep "shell script" | cut -d: -f1 | xargs -L 1 wc -l | grep -v '^0$' | sort -n | head -15! \\

### 2020-11-14 evaluation/plots/coarse-microbenchmarks-table.tex
Correctly format scripts in table

@@
 wf ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z'\n' | tr A-Z a-z | sort | uniq -c | sort -rn | sed ${1}q! \\
-spell ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | col -bx |tr -cs A-Za-z '\n' | tr A-Z a-z |tr -d '[:punct:]' |sort |uniq |comm -23 - $dict! \\
-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | sort | tr [:lower:] [:upper:] > s1 & cat $IN | sort | tr [:upper:] [:lower:] > s2 & diff -B s1 s2! \\
+spell ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | col -bx |tr -cs A-Za-z '\n' | tr A-Z a-z |tr -d '[:punct:]' |sort |uniq |!\\ ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!
+comm -23 - $dict!
+\\
+difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | sort | tr [:lower:] [:upper:] > s1 & cat $IN | sort |!\\ ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!
+tr [:upper:] [:lower:] > s2 & diff -B s1 s2!
+\\
 bi-grams ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | tr -cs A-Za-z '\n' | tr A-Z a-z | bigrams_aux | sort | uniq! \\
-set-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | cut -d ' ' -f 1 | sort > s1 & cat $IN | cut -d ' ' -f 1 | sort > s2 & comm -23 s1 s2! \\
-shortest-scripts ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | xargs file | grep "shell script" | cut -d: -f1 | xargs -L 1 wc -l | grep -v '^0$' | sort -n | head -15! \\
+set-difference ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | cut -d ' ' -f 1 | sort > s1 & cat $IN | cut -d ' ' -f 1 | sort > s2 &!\\ ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!
+comm -23 s1 s2!
+\\
+shortest-scripts ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!cat $IN | xargs file | grep "shell script" | cut -d: -f1 | xargs -L 1 wc -l |!\\ ~&~ \lstinline[columns=fixed,basicstyle=\footnotesize\ttfamily]!
+grep -v '^0$' | sort -n | head -15!
+\\
 \bottomrule

### 2020-11-14 compiler/gather_results.py

@@ -2,6 +2,7 @@ import copy
 import sys
 import math
 import os
+import re
 import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib.lines as pltlines
@@ -584,24 +585,46 @@ def collect_input_size(experiment):
     clean_name = input_file_name.split('/')[-1].split('.')[0]
     return clean_name
 
+def separate_into_lines(script):
+    indent_len = 4
+    lines = [script]
+    while (len(lines[-1]) > 90):
+        last_line = lines[-1]
+        last_break_index = -1
+        for i in range(90):
+            if(last_line[i] in ["&", "|"]):
+                last_break_index = i
+
+        if(last_break_index == -1):
+            break
+        new_pre_last_line = last_line[:(last_break_index+1)]
+        new_last_line = " " * indent_len + last_line[(last_break_index+1):].lstrip()
+        lines[-1] = new_pre_last_line
+        lines.append(new_last_line)
+    return lines
+
+
 def collect_script(experiment):
     # print(experiment)
     script_file = os.path.join(MICROBENCHMARKS, '{}.sh'.format(experiment))
     with open(script_file) as file:
         script_lines = file.readlines()
-        clean_script_lines = [line.lstrip() for line in script_lines
-                              if not line.lstrip().startswith('#')]
-        clean_script_lines = [line.split('#')[0].rstrip(' ') for line in clean_script_lines]
-        ## Remove mkfifo, cat, rm
-        clean_script_lines = [line for line in clean_script_lines
-                              if not line.startswith("mkfifo") and
-                                 not line.startswith("rm")]
-        script = "".join(clean_script_lines).rstrip()
-        script = script.replace("|\n", "| ")
-        script = script.replace("&\n", "& ")
-        script = script.replace("\n", "; ")
-        print(script)
-    wrapped_script = '\lstinline[columns=fixed,basicstyle=\\footnotesize\\ttfamily]!{}!'.format(script)
+    clean_script_lines = [line.lstrip() for line in script_lines
+                            if not line.lstrip().startswith('#')]
+    clean_script_lines = [line.split('#')[0].rstrip(' ') for line in clean_script_lines]
+    ## Remove mkfifo, cat, rm
+    clean_script_lines = [line for line in clean_script_lines
+                            if not line.startswith("mkfifo") and
+                                not line.startswith("rm")]
+    script = "".join(clean_script_lines).rstrip()
[…]

### 2020-11-14 evaluation/plots/microbenchmarks-table.tex
Correctly format scripts in table

@@
 nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
-sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+sort ~&~ $\tsta, \tpur$ & 10~MB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
 top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
 wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
-spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
+spell ~&~ $4\times\tsta, 3\times\tpur$ & 10~MB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
 difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\

### 2020-11-26 compiler/ast_to_ir.py
A comment regarding the correctness of expansion

@@ -354,6 +354,9 @@ def naive_expand(arg_char, config):
 
 ## This function expands an arg_char. 
 ## At the moment it is pretty inefficient as it serves as a prototype.
+##
+## TODO: At the moment this has the issue that a command that has the words which we want to expand 
+##       might have assignments of its own, therefore requiring that we use them to properly expand.
 def expand(arg_char, config):
     return naive_expand(arg_char, config)
 

### 2020-11-29 papers/SOSP21/draft.tex
Add paper structure

--- 2020-10-09 paper/draft.tex
@@
-% TODO Add Ignis and Mozart
-% \nv{Make sure we use \ttt{width} everywhere}
-% \nv{Decide on parallelizable regions}
-
-%% EuroSys 2021 -- 12 pages limit
-\documentclass[sigplan, review, screen, anonymous]{acmart}
-\acmSubmissionID{49}
+\documentclass[sigplan, screen, review, anonymous]{acmart}
+\acmSubmissionID{XXX}
 \renewcommand\footnotetextcopyrightpermission[1]{}
@@
 
-%% EuroSys rules: Submissions may have at most 12 pages of technical
-%% content, including all text, figures, tables, appendices,
-%% etc. Bibliographic references are not included in the 12-page
-%% limit. Use A4 or US letter paper size, with all text and figures
-%% fitting inside a 178 x 229 mm (7 x 9 in) block centered on the
-%% page, using two columns separated by ≥8 mm (0.33″) of
-%% whitespace. Use ≥10-point font (typeface Times Roman, Linux
-%% Libertine, etc.) on ≥12-point (single-spaced) leading. Graphs and
-%% figures should be readable when printed in grayscale, without
-%% magnification. All pages should be numbered. Authors are encouraged
-%% to hyperlink their references.
-
+% \usepackage{amssymb}
+% \usepackage{cleveref}
 \usepackage{enumitem}
 \usepackage{booktabs}
-\usepackage{amssymb}
 \usepackage{soul}
@@
 \usepackage{amsmath}
-\usepackage{cleveref}
 \usepackage{wrapfig}
@@
 \usepackage{caption}
-\usepackage{picins}
-
-\usepackage{minted} % for code listings
-% \usemintedstyle{xcode}
-% ----- For red boxes ------
-\usepackage{etoolbox}
-\makeatletter
-\AtBeginEnvironment{minted}{\dontdofcolorbox}
-\def\dontdofcolorbox{\renewcommand\fcolorbox[4][]{##4}}
-\makeatother
-% --------------------------
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,shapes.misc,shapes.geometric,positioning}
 
@@
 \newcommand{\vs}{{\em vs.} }
-\newcommand{\cmpn}{compartmentalization} 
 \newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
 % \newcommand{\ttt}[1]{\texttt{#1}}
-\newcommand{\ttt}[1]{\mintinline[fontsize=\small]{bash}{#1}}
-\newcommand{\ttiny}[1]{\mintinline[fontsize=\footnotesize]{bash}{#1}}
+\newcommand{\ttt}[1]{\texttt{#1}}
+\newcommand{\ttiny}[1]{\texttt{#1}}
 \newcommand{\tti}[1]{\texttt{\scriptsize #1}}
 \newcommand{\spol}[1]{\scriptsize{\sc#1}}
-\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
-\newcommand{\rf}[1]{\ref{#1}}
-\newcommand{\wka}{\ttt{a\textsubscript{1}}}
-\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}
-
-\newcommand{\cn}[1]{\mbox{\textcircled{\footnotesize #1}}}
-\newcommand{\tcn}[1]{\mbox{\textcircled{\scriptsize #1}}}
-
-\newcommand{\sta}{\cn{\textsc{S}}\xspace}
-\newcommand{\pur}{\cn{\textsc{P}}\xspace}
-\newcommand{\npu}{\cn{\textsc{N}}\xspace}
-\newcommand{\sid}{\cn{\textsc{E}}\xspace}
-\newcommand{\dfs}{\cn{\textsc{F}}\xspace}
-\newcommand{\irr}{\cn{\textsc{I}}\xspace}
-
-\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
-\newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
-\newcommand{\tnpu}{\tcn{\textsc{N}}\xspace}
-\newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
-\newcommand{\tirr}{\tcn{\textsc{I}}\xspace}
-\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
+\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
+\newcommand{\sx}[1]{(\S\ref{#1})}
+\newcommand{\unix}{{\scshape Unix}\xspace}
+\newcommand{\sys}{{\scshape Opaque}\xspace}
 
@@
 \newcommand{\kk}[1]{[{\color{magenta}kk: #1}]}
-\newcommand{\km}[1]{[{\color{blue}km: #1}]}
 \newcommand{\review}[1]{{\color{red}#1}}
@@
 \newcommand{\str}{{\color{red}\textbf{\ttt{*}}}}
-
-\newcommand{\kstar}{^{\textstyle *}}
-\newcommand{\eps}{\varepsilon}
-
-\definecolor{editorGray}{rgb}{0.95, 0.95, 0.95}
-\definecolor{editorOcher}{rgb}{1, 0.5, 0} % #FF7F00 -> rgb(239, 169, 0)
-\definecolor{editorGreen}{rgb}{0, 0.5, 0} % #007C00 -> rgb(0, 124, 0)
-
-\definecolor{cdb}{rgb}{0.37, 0.62, 0.63} % cadet blue
-
-\lstdefinelanguage{sh}{
-  morekeywords={for, in, do, done, \|},
-  keywordstyle=\color{purple}\ttfamily,
-  % ndkeywords={curl, grep, wget, awk, xargs, find, nc, mdc, gunzip, cut, sort, head, join},
-  ndkeywordstyle=\color{black}\ttfamily\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{\#},
-  commentstyle=\color{lightgray},
-% morecomment=[s]{/\\*\\*, \\*/},
-  stringstyle=\color{darkgray}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-% numbersep=1pt,
-% numberstyle=\footnotesize\bf\color{gray},   % the style that is used for the line-numbers
-  abovecaptionskip=0pt,
-  aboveskip=0pt,
-  belowcaptionskip=0pt,
-  belowskip=0pt,
-  frame=none                     % adds a frame around the code
-% moredelim=[s][\color{gray}]{c:}{>},
-% moredelim=[s][\color{orange}]{/*}{/}
-}
-
-\lstset{ %
-  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
-  basicstyle=\normalsize\ttfamily,  % the size of the fonts that are used for the code
-  upquote=true,
-  captionpos=b,                    % sets the caption-position to bottom
-% frame=B,                    % adds a frame around the code
-  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
-  numbersep=2pt,                   % how far the line-numbers are from the code
-  numberstyle=\tiny\color{gray},   % the style that is used for the line-numbers
-  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
-  framerule=0pt,
-	xleftmargin=0pt,
-	xrightmargin=0pt,
-	breakindent=0pt,
-  aboveskip=0pt,
-  framesep=0pt,
-  abovecaptionskip=0pt,
-  aboveskip=0pt,
-  belowcaptionskip=0pt,
-  belowskip=0pt,
-  frame=none,
-  framexbottommargin=0pt,
-  resetmargins=true
-}
-
-%% Conference information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-%% \acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
-%% \acmYear{2018}
-%% \acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-%% \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-%% \startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-%% \setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{plain}
-%% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
-%\citestyle{acmnumeric}     %% For numeric citations
-%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
-                            %% sorting of references within a single citation;
-                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
-                            %% rendered as [14,5,2] rather than [2,5,14].
-%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
-                            %% compression of sequential references within a
-                            %% single citation;
-                            %% e.g., \cite{Baker12,Baker14,Baker16}
-                            %% rendered as [2,3,4] rather than [2-4].
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from traditional SIGPLAN
-%% proceedings format to PACMPL format must update the
-%% '\documentclass' and topmatter commands above; see
-%% 'acmart-pacmpl-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-\newcommand\blfootnote[1]{%
-  \begingroup
-  \renewcommand\thefootnote{}\footnote{#1}%
-  \addtocounter{footnote}{-1}%
-  \endgroup
-}
-
-%% Title information
-%% \title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
-\title{\sys: Light-touch Data-Parallel Shell Processing}         %% [Short Title] is optional;
-%% \title{Extensible data processing in the shell}
-
-% \titlenote{with title note}             %% \titlenote is optional;
-%                                         %% can be repeated if necessary;
-%                                         %% contents suppressed with 'anonymous'
-% \subtitle{Subtitle}                     %% \subtitle is optional
-% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-%% Author with single affiliation.
-
-%% \author{
-%% % Anonymous Author(s)\\
-%% % \normalsize{Additional Anonymized Material: \href{https://git.io/JfwSq}{https://git.io/JfwSq}.}
-%%   {\rm \normalsize Nikos Vasilakis$^{*}$\qquad Konstantinos Kallas$^{*\dagger}$\qquad  Konstantinos Mamouras$^{\circ}$ }\\
-%%   {\rm \normalsize Achilleas Benetopoulos$^{\diamond}$ \qquad Lazar Cvetkovich$^{\ddagger}$ }\\
-%%   {\small MIT, CSAIL \quad $^{\dagger}$University of Pennsylvania \quad $^{\circ}$Rice University \quad $^{\diamond}$Unaffiliated \quad $^{\ddagger}$University of Belgrade}
-%% }
-
-%% \author{Nikos Vasilakis}
-%% \authornote{The two marked authors contributed equally to the paper.}
-%% \affiliation{MIT}
-
-%% \author{Konstantinos Kallas}
-%% \authornotemark[1]
-%% \affiliation{University of Pennsylvania}
-
-%% \author{Konstantinos Mamouras}
-%% \affiliation{Rice University}
-
-%% \author{Achilleas Benetopoulos}
-%% \affiliation{Unaffiliated}
-
-%% \author{Lazar Cvetkovich}
-%% \affiliation{University of Belgrade}
-
-%% \authornote{$^{*}$Equal contribution.}     %% \authornote is optional;
-
-%%                                         %% can be repeated if necessary
-%% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-%% \affiliation{
-%%   \position{Position1}
-%%   \department{Department1}              %% \department is recommended
-%%   \institution{Institution1}            %% \institution is required
-%%   \streetaddress{Street1 Address1}
-%%   \city{City1}
-%%   \state{State1}
-%%   \postcode{Post-Code1}
-%%   \country{Country1}                    %% \country is recommended
-%% }
-%% \email{first1.last1@inst1.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-% \author{First2 Last2}
-%% \authornote{with author2 note}          %% \authornote is optional;
-%%                                         %% can be repeated if necessary
-%% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-%% \affiliation{
-%%   \position{Position2a}
-%%   \department{Department2a}             %% \department is recommended
-%%   \institution{Institution2a}           %% \institution is required
-%%   \streetaddress{Street2a Address2a}
-%%   \city{City2a}
-%%   \state{State2a}
-%%   \postcode{Post-Code2a}
-%%   \country{Country2a}                   %% \country is recommended
-%% }
-%% \email{first2.last2@inst2a.com}         %% \email is recommended
-%% \affiliation{
-%%   \position{Position2b}
-%%   \department{Department2b}             %% \department is recommended
-%%   \institution{Institution2b}           %% \institution is required
-%%   \streetaddress{Street3b Address2b}
-%%   \city{City2b}
-%%   \state{State2b}
-%%   \postcode{Post-Code2b}
-%%   \country{Country2b}                   %% \country is recommended
-%% }
-%% \email{first2.last2@inst2b.org}         %% \email is recommended
-
-\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
-\newcommand{\sx}[1]{(\S\ref{#1})}
-\newcommand{\sys}{{\scshape PaSh}\xspace}
-\newcommand{\unix}{{\scshape Unix}\xspace}
 
@@
 
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-%% \begin{CCSXML}
-%% <ccs2012>
-%% <concept>
-%% <concept_id>10011007.10011006.10011008</concept_id>
-%% <concept_desc>Software and its engineering~General programming languages</concept_desc>
-%% <concept_significance>500</concept_significance>
-%% </concept>
-%% <concept>
-%% <concept_id>10003456.10003457.10003521.10003525</concept_id>
-%% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
-%% <concept_significance>300</concept_significance>
-%% </concept>
-%% </ccs2012>
-%% \end{CCSXML}
+\title{Something Something}
 
-%% \ccsdesc[500]{Software and its engineering~General programming languages}
-%% \ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-% \keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
 \begin{abstract}
-This paper presents \sys, a system for parallelizing POSIX shell scripts through a combination of program transformations and runtime primitives.
-% a source-to-source shell compiler and associated runtime library
-Given a script, \sys converts it to a dataflow graph, performs a series of transformations that expose parallelism while preserving its sequential semantics, and then converts the dataflow graph back into a script that uses POSIX constructs to explicitly guide parallelism.
-A lightweight annotation language allows command developers to express key parallelizability properties about their commands.
-An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---guides the annotation language and optimized aggregator library that \sys uses.
-% \kk{eurosys: Can we extend this to say that PaSh automatically uses constructs from this library? Otherwise it sounds like the user has to use it to make their scripts parallel.}
-Finally, \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-\sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.89$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
-% The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
 \end{abstract}
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
-
-% Its runtime component provides orchestration and planning support during the execution of the program.
-% These are from the abstract:
-%  with regards to the sequential program.
-  % leveraging the insight that such programs already express stream computations, with their stages falling under a few, known parallelizability classes.
-% a series of techniques for automatically 
-% 
-% Key challenges include 
-% surprisingly expressive, 
-% maximally distributable subprograms
-%
 
@@
 \label{intro}
-
-%% \kk{After discussion on Wednesday Jan 8: Modify the narrative of
-%%   this paper to talk about data processing in the Shell. Make this
-%%   the main focus of the paper. This simplifies the narrative and
-%%   allows us to not talk about late planner activation as well as
-%%   environment variables. It also justifies why we only deal with
-%%   pure and stateless. Note that this is a very expressive model, as
-%%   shown by our benchmark suite. Note: Rememeber to clearly describe
-%%   the subset of the shell that we handle. To be more precise the
-%%   compositional operator of the shell that we can handle.}
-
-%% \kk{Note: A point that we need to establish early on and convince
-%%   the readers about is that the Shell IS used already for
-%%   data-processing, and that we are not just proposing it as an
-%%   alternative to other systems. We do not propose migrating to
-%%   shell scripts if you already have a parallel/efficient
-%%   solution. I am not sure if that needs a specific paragraph of its
-%%   own at the start of the introduction. Like: ``Consider the
-%%   following bash script that does some simple data
-%%   processing... Examples like that exists in several
-%%   domains/applications (examples, cite).''}
-
-%% \kk{Decide if we will mainly focus on scripts or pipelines and
-%%   choose wordings appropriately.}
-%
-% Consider the following data processing Shell script.
-% %% \kk{We might want to have a more complex script here.}
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-% cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
-%   sort | uniq -c | sort -rn | head 5 > out
-% \end{lstlisting}
-% 
-% \noindent
-% great sentence -- keep
-% Scripts like the one above are commonly used for orchestrating data-processing workflows; connecting system commands with custom tools to perform complex tasks.
-% %% \kk{Do we need to mention domains where that happens?}
-% Even though shell pipelines are designed to exploit task parallelism, they completely ignore available data-parallelism.
-% % Too low-level descriptions of the problem, maybe keep it high level?
-% For small input sizes this isn't a problem since scripts take less than a second to execute, but as data sizes increase, failing to utilize the underlying parallel computational resources is a significant bottleneck.
-
-%% \blfootnote{$^{*}$Equal contribution.}
-The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
-This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple scripts for data processing, system orchestration, and other automation tasks.
-% Pipe-centric shell commands are parallel (and to some extent, functional) programs that can natively enjoy significant parallelism on a multicore machine.
-Unfortunately, parallelizing such pipelines requires significant effort shared between two different programmer groups. % it's more like roles personas or "wearing hats"
-
-The first group is \emph{command developers}, responsible for implementing individual commands such as \ttt{sort}, \ttt{uniq}, and \ttt{jq}.
-These developers usually work in a single programming language, leveraging its abstractions to provide parallelism whenever possible.
-As they have no visibility into the command's uses, they expose a plethora of ad-hoc command-specific flags such as \ttt{-t}, \ttt{-}\ttt{-parallel}, \ttt{-p}, and \ttt{-j} % , \ttt{CPUs}, and \ttt{NUM\_THREADS}~\cite{}.
-\cite{pasetto2011comparative, mcilroy1993engineering, stallman1991gnu}.
-
-The second group is \emph{shell users}, who use POSIX shell constructs to combine multiple such commands from many languages into their scripts and are thus left with only a few options for incorporating parallelism.
-% \kk{I don't really like the reference to such a later section. Can we remove it?}
-% \nv{We have to say this is not just ``an opinion'', but that we give a scientific treatment/full coverage in the related work section.
-How far away should the sections be so that you're comfortable referencing to?}
-% If versed into parallelism, 
-% Users versed in computing can apply 
-% these tools are command-unaware thus at risk of breaking program semantics,
-% require manual changes to the code to make parallelism explicit,
-% at times leverage command flags like the ones mentioned earlier---but each command 
-% oblivious? agnostic?  
-One option is to use manual tools such as GNU \ttt{parallel}~\cite{Tange2011a}, \ttt{ts}~\cite{tsp}, \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm};
-  these tools are either command-unaware, and thus at risk of breaking program semantics, or too coarse-grained, and thus only capable of exploiting parallelism at the level of entire scripts rather than individual components.
-% Tools will attempt to parallelize anything 
-% use the composition abstractions of the \unix shell, 
-% and are often prototypical in nature---focusing on quickly prototyping a solution for the task at hand rather building an optimal solution to a recurring pattern.
-% short lived, changing quickly
-% (ii) ad-hoc flags provided by command developers and mentioned above.
-Another option is to use shell primitives (such as \ttt{&}, \ttt{wait}) to explicitly induce parallelism;
- these come at a cost of manual effort to split inputs, rewrite scripts, and orchestrate execution---an expensive and error-prone process.
-% re-writing of the entire script from scratch; unfortunately, this process is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality---but also foregoes all the productivity benefits of the original script.
-% The fact that a programmer often belongs to both groups at the same time---as composed scripts can themselves form a command used in other scrips---compounds the challenge.
-To top it off, all these options assume a good understanding of parallelism;
-  users with domain of expertise outside computing---from hobbyists to data analysts---are left without options.
-%  (see~\S\ref{related} for a full discussion).
-% \nv{I still don't like ``for a full discussion''---raises too many expectations, and we don't have that much more to say}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.49\textwidth]{\detokenize{./figs/pash_overview.pdf}}
-\caption{
-  \textbf{\sys overview.}
-  \sys identifies dataflow regions~\sx{dataflow-regions}, 
-  converts them to dataflow graphs~\sx{graph-components},
-  applies transformations~\sx{ir:transformations} based on the parallelizability properties  of the commands in these regions(\S\ref{cmd}, \S\ref{ext}),
-  and emits a parallel script that uses custom primitives~\sx{optimizer}.
-}
-\vspace{-18pt}
-\label{fig:overview}
-\end{figure}
-
-To address this challenge, we develop a system called \sys and outlined in Fig.~\ref{fig:overview} for parallelizing POSIX shell scripts.
-% \kk{We need to mention data-parallelism earlier.
-We don't focus on general parallelization but rather data-parallelism (performing operations on disjoint partitions of the input and then merging the results.
-WE need to give a brief idea about that early on so that we don't create false ideas.}
-% \nv{First attempt here. I am trying to get a copy out ASAP, we can refine this later}
-% \kk{Mention that it outputs shell scripts.}
-% \nv{I think the ``how'' it solves it should come later, we're still trying to convince them we solve it (see next sentences in this paragraph)
-% \sys offers both classes of users in the second group such much-needed automation.
-\sys benefits both programmer groups, % mentioned earlier, 
-  with emphasis on users.
-Command developers are given a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
-  rather than expressing a command's full observable behavior, these annotations focus primarily on its interaction with state.
-Shell users, on the other hand, are provided with full automation:
-  \sys  analyzes their scripts and extracts latent parallelism.
-% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
-\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.
-% \kk{Should we mention that we only show this empirically?}
-% \nv{I don't think so; I don't think this is the right point to talk about our evaluation, we just state our intention/high-level principles.}
-
-To address cold-start issues, \sys comes with a library of parallelizability annotations for commands in POSIX and GNU Coreutils.
-These large classes of commands serve as the shell's standard library, expected to be used pervasively. % in scripts found out in the wild.
-The study that led to their characterization also informed \sys's annotation and transformation components.
-% \footnote{
-%   In fact, the scope of the study is broader, also incorporating commands outside POSIX and Coreutils such as \ttt{builtin}s, \ttt{curl} and \ttt{pandoc}.
-% }
-
-The above are tied together with \sys's runtime component.
-% \sys's runtime  component additional packs
-% Aside from these two components, \sys packs a runtime component that serves dual purpose:
-Aware of the \unix philosophy and abstractions, it packs a small library of highly-optimized data aggregators as well as high-performance primitives such as ones for eager data splitting and merging.
-These address many practical challenges % such as task parallelism, deadlocks, and performance bottlenecks,
-  and were developed by uncovering several pathological situations, on a few of which we report.
-
-%% First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-
-% \sys packs several ideas around the insight that a large subset of the Shell language can be represented as a dataflow graph, enabling for parallelization optimizations.
-% First, a careful study of the parallelizability properties of shell primitives and commands.
-% \sys addresses the challenge by offering command developers a unified set of abstractions for describing the parallelizability characteristics of their commands.
-% Using lightweight annotations akin to type signatures, \sys can automate the parallelization of entire POSIX scripts without any input from their user.
-% hand-in-hand with the parallelizability study:
-%   the study informed the annotation language and was used to annotate the entire
-%   for all 
-% ---a set of annotations that \sys already offers for 
-% for the standard library
-%   akin to the standard library
-%   the annotation language is informed by the study, but also the study is 
-% XX to a type signature, these annotations describe 
-% 
-% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-% (one sentence for the model)
-% (one sentence for the optimizations)
-% The result of the compilation is still a POSIX-compliant script that manages parallelism explicitly;
-%   POSIX compliance means 
-% it does not depend on a managed runtime 
-% 
-% Third, \sys packs a runtime component 
-% A small library of highly-optimized data combiners allows 
-% This is paired with a
-% it comes with a few runtime primitives for eager data splitting and merging that address several practical issues ranging from deadlocks, performance bottlenecks;
-%   these primitives were developed from uncovering several pathological situations, on many of which we report.
-
-% TODO: keep only points necessary for the punchline 
-% In the vast majority of scripts (49/54), \sys achieves significant speedups \todo{(average for width=16)}, and for the rest 5 slowdown is insignificant (worst case 0.89x).
-% For the temperature analysis script (which is taken from a Hadoop case study), \sys not only 
-% contains pre-processing that takes X time and Y main processing (Hadoop focus).
-% \sys accelerates the entire pipeline by , including the fragment that is outside the focus of conventional parallelization systems.
-% accelerates the CPU-intensive part that is parallelized by Hadoop, but also for the preprocessing part which takes even more time than what the other part takes.
-% XX--YY speedups in the vast majority of scripts, ones taken from the wild 
-% 1. Sentence about transformations vs runtime primitives
-% 2. Unix 50? 
-% 3. point about Hadoop expressing a fraction of the pipeline and speedup
-% \sys achieves higher speedup than \ttt{sort --parallel} that involved hand-tuned parallel implementation by command developers.
-% correctness of GNU parallel
-% PaSh can be applied to programs of notable
-% size and input to offer significant acceleration. A broader
-% take-away is that PaSh is also able to extract parallelism from
-% fragments that are not purely compute-intensive regions, i.e.,
-% the usual focus of conventional parallelization systems.
-
-We evaluate \sys on 44 unmodified scripts including:
-  (i) a series of benchmarks, ranging from classic \unix one-liners to modern data-processing scripts;
-  (ii) two large and complex use cases for temperature analysis and web indexing;
-%% , and container orchestration;
-and 
-  (iii) two micro-benchmarks inspired by (and comparing with) possible alternatives.
-Speedups range between 0.89--61.1$\times$ (avg: $6.7\times$), with the 39 out of 44 scripts seeing non-trivial speedups. 
-\sys's runtime primitives add to the base speedup extracted by \sys's program transformations---\eg 8.83$\times$ over a base 5.93$\times$ average for 10 representative \unix one-liners.
-%  with \ttt{--width=64}, 
-% \kk{eurosys: Readers don't know what --width is here so it might make sense to not mention it.}
-% \sys performs better than \ttt{sort --parallel}, a hand-tuned parallel implementation, and GNU parallel, which returns incorrect results if used without care.
-\sys accelerates a large program for temperature analysis by 2.52$\times$, parallelizing both the computation (12.31$\times$) and the preprocessing (2.04$\times$) fragment (\ie data download, extraction, and cleanup), which traditionally falls outside of the focus of conventional parallelization systems---even though it takes $75\%$ of the total execution time.
-  % extracting 12.31$\times$ from the portion of the program that traditionally falls outside the focus of conventional parallelization systems.
-  % \kk{\sys parallelizes the computation by 12.31 and not the preprocessing. It parallelizes the preprocessing by 2.}
-
-% For the temperature analysis script (which is taken from a Hadoop case study), \sys not only 
-% contains pre-processing that takes X time and Y main processing (Hadoop focus).
-% \sys accelerates the entire pipeline by , including the fragment that is outside the focus of conventional parallelization systems.
-% accelerates the CPU-intensive part that is parallelized by Hadoop, but also for the preprocessing part which takes even more time than what the other part takes.
-% XX--YY speedups in the vast majority of scripts, ones taken from the wild 
-% 1. Sentence about transformations vs runtime primitives
-% 2. Unix 50? 
-% 3. point about Hadoop expressing a fraction of the pipeline and speedup
-% \sys achieves higher speedup than \ttt{sort --parallel} that involved hand-tuned parallel implementation by command developers.
-
-%   The results show
-% significant advantages in correctness, manual effort,
-% and performance---with 
-% speedups $X-X\times$ due to \sys's transformations and $Y-Y\times$ due to \sys's runtime primitives.
-%% ---with \sys's transformations offering an order-of-magnitude performance improvements aided further by \sys's optimized runtime components.
-% Improvements depend on several factors such as a script's initial execution time and the parallelizability properties of its commands.
-
-The paper is structured as follows.
-It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
-Sections \ref{parallelizability}--\ref{impl} highlight key contributions:
-\begin{itemize}
-  \item
-  \S\ref{parallelizability}
-    studies the parallelizability of shell commands, and introduces a lightweight annotation language for commands that are executable in a data-parallel manner.
-
-  \item
-  \S\ref{ir} presents a dataflow model and associated transformations that expose data parallelism while preserving the semantics of the sequential program.
-
-  \item
-  \S\ref{impl} details \sys's runtime component, discussing the correctness and performance challenges it addresses.
-
-\end{itemize}
-
-\noindent
-After \sys's evaluation~\sx{eval} and  comparison with related work~\sx{related}, the paper concludes~\sx{discussion}.
-
-% nv: took this out because it's wasting 6 lines of text (because of section titles hitting the column divide---classic problem of section 2)
-% \sys's source code will be open-sourced with the camera-ready version of the paper, and is provided in anonymized form at \href{https://git.io/JfEgr}{https://git.io/JfEgr}.
-
-% % TODO: Important to have a small script that highlights
-% \nv{
-% (i) different classes of commands
-% (ii) challenge of shell expansion order
-% (iii) segments that are parallelized by \sys (with provable correctness)
-% (iv) segments that are \emph{not} parallelized by \sys (otherwise they can lead to
-% incorrectness)
-% }
-
-% \nv{I don't think the intro needs any serious rewrite, apart from deleting any
-% notion of distribution and highlighting parallel processing.
-% 
-% Key tasks, in rough order of importance:
-%   (i) change title and intro to avoid distribution
-%   (ii) cleanup and rewrite sections 4, 5: introduce our high-level figure, talk
-%   about things in order, add ordering in graphs, generally refine and beautify
-%   as much as possible.
-%   (iii) tighten section 2, everything in there has to be related to some part of
-%   the story---make the connection explicit
-%   (iv) Pump up evaluation: could we add 3 more micro-benchmarks? could we add
-%   1--2 more macro-benchmarks?
-%   (v) since this is IFCP, could we add some simple semantics and derivation
-%   rules? we should certainly be able to describe some of the optimizations as
-%   congruence rules.
-%   (vi) complete the related work with some operating-systems literature.
-%   --------
-%   Need to highlight the use of ad-hoc flags
-%   Need to highlight use cases / goals:
-%     still POSIX compliant, you can replace /bin/bash with /bin/dish and see
-%     \emph{some} performance improvements, even if we can't handle all cases
-%     today--that is, dish enables a platform for further research.
-%
-%  Talk about order-aware dataflow analysis?
-% }
-
-%% \kk{Begin of new intro sketch}
-%% \begin{itemize}
-%% \item Why is parallelism important (especially in the context of data
-%%   processing). Significant performance benefits as most computer
-%%   systems today have multiple processors that are under utilized
-%% \item (Similarly to how it is now) unfortunately, it is quite hard
-%%   reaping these benefits in contrast to sequential data
-%%   processing. [Show Bash pipeline].
-%% \item To benefit from parallelism, the current options are
-%%   extreme. Either very fine-grained and low level (parallelism using
-%%   fine-grained C libraries) for maximum performance, or very coarse
-%%   grained and bulky where one has to use a proper big-data
-%%   framework. (Can we fit the shell ones here (gnu-parallel, etc))
-%% \item The common thing is that one needs to rewrite big part of their
-%%   application, often requiring expertise and redesigning its
-%%   architecture, in order to get benefits that in the end might not
-%%   explain the cost.
-%% \item To address this issue, we propose Dish, a thin layer that
-%%   automatically! parallelizes a given shell script. Key insight is
-%%   that a large subset of the shell language can be encoded in a
-%%   dataflow graph model, a flexible representation that allows for
-%%   parallelization optimizations.
-%% \end{itemize}
-%% Contributions:
-%% \begin{itemize}
-%% \item Study of the shell primitives, identification of
-%%   families/classes of commands that share
-%%   parallelizability/parallelization properties, and a classification of
-%%   coreutils/POSIX commands in these parallelizability properties.
-%% \item Dataflow graph model that is suitable for the shell
-%% \item A set of optimizations on this model that can be thought of as
-%%   rewriting rules on the dataflow graph, that are sound and exploit
-%%   parallelism.
-%% \end{itemize}
-%% \kk{End of new intro sketch}
-
-% 1. Shell scripting; pipelines, in particular, are a common abstraction for expressing filters
-% They are an easy way to , because they combine a set of assumptions that work well with Unix
-% They work great on a single machine, but are difficult to scale out
-% Could we fully automate distribution? 
-% The key insight is that pipelines already express a domain-specific
-% computation that is easily amenable to distribution.
-%
-
-%% Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
-%% % offer notable benefits over their centralized counterparts:
-%% %   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
-%% % Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
-%% % The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
-%% Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
-%% Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
-%%   % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
-%%    scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.
-%% % 
-%% To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:
-
-%% % \begin{lstlisting}[language=sh,float=h,numbers=none]
-%% % cat doc.ms |                   # print input file
-%% % groff -t -e -mandoc -Tascii |  # remove formatting
-%% % col -bx |                      # remove backspaces 
-%% % tr A-Z a-z |                   # convert to lower case
-%% % tr -d '[:punct:]' |            # remove punctuation
-%% % sort |                         # put words in alphabetical order
-%% % unique |                       # remove duplicate words
-%% % comm -13 /usr/share/dict -     # report words not in dictionary 
-%% % \end{lstlisting}
-%% % \smallskip
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%% cat * | tr -cs A-Za-z\n | tr A-Z a-z |    ($$(p_1)$$)
-%%   sort | uniq -c | sort -rn | head 5 > out
-%% \end{lstlisting}
-%% % \smallskip
-%% % \noindent
-%% Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
-%% Combining several features, \unix makes small tasks easy to express;
-%%   for a user with \emph{one} computer and a small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
-%% % composition allows the composition of general primitives
-
-%% % Key insight 
-%% % % The garden-hose philosophy is similar to Haskell 
-%% % This pipeline takes an input file (line 1), removes 
-%% % * do not need to use specialized frameworks for composition---instead, as long as they conform to the shell, individual primitives can be written in any programming language.
-%% % * do not need to rewrite ---
-
-%% \begin{figure}[t]
-%% \centering
-%% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-%% \caption{
-%%   \textbf{High-level schematic.}
-%%   \sys leverages parallelizability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
-%%   execution (right).
-%% }
-%% \vspace{-10pt}
-%% \label{fig:schematic}
-%% \end{figure}
-
-%% Unfortunately, developing $p_1$'s distributed equivalent requires significant effort.
-%% % For a user with many computers and larger inputs, 
-%% % The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
-%% For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency,distal:13,meiklejohn2015lasp}---an unjustifiable cost for small, one-off pipelines that take a few minutes to compose (but are applied to large datasets).
-%% More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}.
-%  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
-%% In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
-%% Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?
-
-%% To answer this question we develop \sys, a shell variant that transforms traditional \unix pipelines into their distributed equivalents, enabling \emph{distribution-oblivious programming}.
-%% The key insight behind \sys is that a large part of the shell language can be encoded in a distributed dataflow graph model, providing structure to distribute computations.
-%% %% \kk{Replaced that: the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation. }
-%% % FIXME \kk{I am not sure whether stream processing provides information for distributing a computation on its own.
-Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
-%% \sys builds on this insight with a careful study of the parallelizability properties of shell primitives and commands, a definition of a dataflow graph model that is suitable for the shell, a set of parallelization optimizations that preserve the behaviour of the graph, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
-%% A \sys-enabled $p_1$ will run \ttt{tr}s on parallel streams,
-%%   use a mostly-parallel tree of \ttt{sort}s,
-%%   run \ttt{uniq} mostly in parallel,
-%%   execute \ttt{head} in one of the streams,
-%%   and (trivially) merge streams in \ttt{out}.
-
-%% % Developers of commands
-%% % Developers of pipelines
-%% % Language unifies both---no need for `-t` `-p`
-%% % POSIX:
-%%   % https://pubs.opengroup.org/onlinepubs/009695399/utilities/xcu_chap02.html
-
-%% The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
-%% % \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
-%% First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, calls to \textsc{GNU} \ttt{parallel}~\cite{Tange2011a}, or any manual rewriting~\cite{mapreduce:08, ciel:11, spark:12}.
-%% Second, developers of new shell commands can use a lightweight domain-specific language to express their parallelizability properties rather than ad-hoc, command-specific flags such as {\tt -t},  {\tt NUM\_THREADS}, \ttt{-t}, \ttt{-p} \etc
-%% Most importantly, \sys provides an architectural lesson for system designers---namely, that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} would have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.
-%% % Nice critique on parallel: https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=597050#75
-
-%% % % Make Unix benefits explicit?
-%% % Indeed, the primary reason behind $p_1$'s succinctness is that the pipeline is a domain-specific language for describing operations over streams.
-%% % Key elements of \unix are the ability to compose programs written in different languages, the abstraction of a file system as a set of resident streams, a small but extensible library of commands, and the ability to resolve names within a global context.
-%% % Under the hood, the \unix kernel buffers results, synchronizes processing stages, and generally orchestrates the computation.
-
-%% The paper is structured as follows.
-%% It starts by introducing the necessary background on shell scripting and overviewing \sys~\sx{bg}.
-%% Sections \ref{parallelizability}--\ref{impl} highlight key
-%% contributions:
-%% \begin{itemize}
-
-%%   \item
-%%   \S\ref{parallelizability} studies several parallelizability classes and introduces the parallelizability characterization language.
-
-%%   \item
-%%   \S\ref{ir} presents a dataflow graph model for encoding shell pipelines, and a set of parallelization transformations that preserve the semantics of the sequential program.
-
-%%   \item
-%%     \S\ref{impl} describes the \sys implementation and addresses
-%%     several challenges related to the translation and optimization of
-%%     shell programs.
-%%     %% details other concerns, such as distributed state management and extensibility. 
-%% \end{itemize}
-
-%% \noindent
-%% \sys's evaluation~\sx{eval} shows significant speedups (4--10$\times$) for parallel execution using a combination of real pipelines.
-%micro-bench\-marks %---popular shell one-liners that highlight certain features---and a multi-line macro-benchmark---a popular weather analysis script.
-%% After a comparison with related prior work~\sx{related}, the paper closes with a discussion of limitations and possible future directions~\sx{discussion}.
-%% % \bigskip
-%% % \begin{quote}
-%% % \footnotesize
-%% Additional material, in anonymized form, is available at
-%% \href{https://git.io/Je6Nk}{https://git.io/Je6Nk}.
-%% % \end{quote}
-
-%% % \begin{figure*}[t]
-%% % \centering
-%% % \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
-%% % \caption{
-%% %   \textbf{Applying \sys to $p_1$.}
-%% % }
-%% % \label{fig:example}
-%% % \end{figure*}
-%% %
-
-\section{Background and Overview}
-\label{bg}
-
-This section reviews \unix shell scripting through an example~\sx{bg:pipelines}, which it then uses to present 
-parallelization challenges~\sx{bg:challenges} and how they are addressed by \sys~\sx{bg:overview}.
-
-\subsection{Running Example: Weather Analysis}
-\label{bg:pipelines}
-
-% In the ongoing debate on climate change,
-Suppose an environmental scientist    % looking for a non-CS expert---journalist might be too low
-  wants to get a quick sense of trends in the maximum temperature across the U.S. over the past five years.
-As the National Oceanic and Atmospheric Administration (NOAA) has made historic temperature data publicly available~\cite{noaa}, 
-  answering this question is only a matter of a simple data-processing pipeline.
-
-Fig.~\ref{fig:example}'s script starts by pulling the yearly index files and filtering out URLs that are not part of the compressed dataset.
-It then downloads and decompresses each file in the remaining set, extracts the values that indicate the temperature, and filters out bogus inputs marked as \ttt{999}.
-It then calculates the maximum yearly temperature by sorting the values and picking the top element.
-Finally, it matches each maximum value with the appropriate year in order to print the result.
-% 
-The effort expended writing this script is low:
-% \footnote{
-%   Some effort is required to understand NOAA's weather format, but this is true for any program processing any dataset.
-% }
-  its data-processing core amounts to 12 stages and, when expressed as a single line, is only 165 characters long.
-% Its succinctness is deceiving---
-This program is no toy:
-  a Java program implementing only the last four stages takes 137~LoC~\cite[\S2.1]{hadoop:15}.
-To enable such a succinct program composition, \unix incorporates several features.
-
-\heading{\unix Features}
-Composition in \unix is primarily achieved with pipes (\ttt{|}), a
-construct that allows for task-parallel execution of two commands by
-connecting them through a character stream.
-%% An important \unix feature are pipes (\ttt{|}), introducing
-%% task-parallel program composition over character streams.
-This stream is comprised of contiguous character lines separated by newline characters (\textsc{NL}) delineating individual stream elements.
-For example, Fig~\ref{fig:example}'s first \ttt{grep} outputs (file-name) elements containing \ttt{gz}, which are then consumed by \ttt{tr}.
-% is given lines containing file identifiers and outputs ones that contain \ttt{gz}, 
-% As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
-A special end-of-file (\textsc{EOF}) condition marks the end of a stream.
-
-% Another important feature is agg(synchronization.
-Different pipeline stages process data concurrently and possibly at different rates---\eg the second \ttt{curl} produces output at a significantly slower pace than the \ttt{grep} commands before and after it.
-The \unix kernel facilitates scheduling, communication, and synchronization behind the scenes.
-
-\begin{figure}[t]
-\centering
-\begin{minted}[fontsize=\footnotesize]{bash}
-base="ftp://ftp.ncdc.noaa.gov/pub/data/noaa";
-for y in {2015..2019}; do
- curl $base/$y | grep gz | tr -s" " | cut -d" " -f9 |
- sed "s;^;$base/$y/;" | xargs -n 1 curl -s | gunzip |
- cut -c 89-92 | grep -iv 999 | sort -rn | head -n 1 |
- sed "s/^/Maximum temperature for $y is: /"
-done
-\end{minted}
-% \includegraphics[width=0.47\textwidth]{\detokenize{./figs/dish_example.pdf}}
-\caption{
-  \textbf{Calculating maximum temperatures per year.}
-  The script downloads daily temperatures recorded across the U.S. for the years 2015--2019 and extracts the maximum for every year.
-}
-\vspace{-15pt}
-\label{fig:example}
-\end{figure}
-
-Command flags, used pervasively in \unix, are configuration options that the command's developer has decided to expose to its users to improve the command's general applicability.
-For example, by omitting  \ttt{sort}'s \ttt{-r} flag that enables reverse sorting, the user can easily get the minimum temperature.
-The shell does not have any visibility into these flags; 
-  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to individual commands.
-
-Finally, \unix provides an environment for composing commands written in any language.
-Many of these commands come with the system---\eg ones defined by the POSIX standard or ones part of the GNU Coreutils---whereas others are available as add-ons.
-The fact that commands are developed in a variety of languages---including shell scripts---provides users with significant flexibility.
-For example, one could replace \ttt{sort} and \ttt{head} with \ttt{./avg.py} to get the average rather than the maximum---the pipeline still works, as long as \ttt{./avg.py} conforms to the interface outlined earlier.
-% Shell scripting allows incremental development through refinement
-
-% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
-% generating errors or diagnostics to the standard 
-% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
-% \smallskip
-
-% ---available publicly online via NOAA~\cite{}---to obtain the maximum temperature per year.
-
-\subsection{Parallelization Challenges}
-\label{bg:challenges}
-
-While these features aid development-effort economy through powerful program composition, they complicate shell script parallelization, which even for simple scripts such as the one in Fig.~\ref{fig:example} create
-several challenges.
-% \footnote{
-%   Alternatives for automated parallelization were outlined in the introduction~\sx{intro} and are surveyed in the related work section~\sx{related}.
-% }
-
-\heading{Commands} % : Unbounded \& Mysterious}
-In contrast to restricted programming frameworks that enable parallelization by supporting a few carefully-designed primitives~\cite{streamit:02, brook:04, mapreduce:08, spark:12}, the \unix shell provides an unprecedented number and variety of composable commands.
-% Due to their varying nature, 
-% As each command may exercise arbitrary behaviors,
-To be parallelized, each command may require special analysis and treatment---\eg exposing data parallelism in Fig.~\ref{fig:example}'s \ttt{tr} or \ttt{sort} would require splitting their inputs, running them on each partial input, and then merging the partial results.\footnote{
-  As explained in the introduction~\sx{intro}, commands such as \ttiny{sort} may have \emph{ad hoc} flags like \ttiny{--parallel}, which do not compose across commands and may risk breaking correctness and unexploiting performance potential~\sx{micro}.
-}
-% Key point: ...analyses and treatment---\eg combining partial results for \ttt{tr} is different than for \ttt{sort}.
-Automating such an analysis is infeasible, as individual commands are black boxes written in a variety of programming languages and models.
-%% \ttt{grep} requires no coordination, \ttt{head} requires light coordination, and \ttt{sort} requires significant coordination.
-%% These behaviors are further concealed by the fact that commands are black boxes implemented in a variety of programming languages and models.
-Manual analysis is also challenging, due to the sheer number of commands and the many flags that affect their behavior---\eg Fig.~\ref{fig:example}'s \ttt{cut} is called with two flag sets.
- % , making correct treatment even more difficult. % nv: we already said impossible so more difficult doesn't add much!
-
-% \kk{I tried to rephrase the above a little, but I can't get it just right.
-There are two parts to this argument.
-First, the fact that commands are arbitrary (and not known) requires that each of them has special treatment.
-However, the huge number of them, and the fact that flags change their behaviour even more makes an analysis of all commands a very difficult task.}
-
-% A good solution should handle several commands but should also be extensible, in the sense that newly added or modified commands don't require modifications in the solution itself.
-% TODO: Leave this for the solution
-
-\heading{Scripts} % Restrictive \& Coarse-grained}
-Another challenge is due to the language of the POSIX shell. % of the shell, which is both restrictive and coarse-grained.
-% has been developed through a long history of additions.
-First, it contains constructs that enforce sequential execution:
-   Fig.\ref{fig:example}'s sequential composition operator (\ttt{;}) means that the assignment to \ttt{base} is completed before everything else.
-Moreover, the language semantics only exposes limited task-based parallelism in the form of constructs such as \ttt{&}. % and \ttt{|} 
-% NV: is this really task parallelism? (to be clear: I'm only considering removing "task-based".
-%%   contains constructs with different parallelizability semantics. % that enforce sequential execution.
-%% For example, Fig.~\ref{fig:example} assignment to \ttt{base}, connected with the rest of the script by the sequential composition operator (\ttt{;}), has to be completed before the rest of the script starts executing.
-%% Moreover, depending on the construct, different fragments operate on different output sizes thus allowing different parallelization widths.
-Even though Fig.~\ref{fig:example}'s \ttt{for} focuses only on five years of data, \ttt{curl} still outputs thousands of lines per year;
-  naive parallelization of each loop iteration will miss such opportunities.
-Any attempt to automate parallelization should be aware of the POSIX shell language, exposing latent data parallelism without modifying execution semantics.
-% \kk{eurosys: I am not sure if I completely see the challenge in this paragraph.
-The only thing I see as challenging is the first sentence talking about ; and sequential composition.
-We should discuss to make sure we make any challenge here shine.}
-
-\heading{Implementation} %
-On top of command and shell semantics, the broader \unix environment has its own set of quirks.
-Any attempt to orchestrate parallel execution will hit challenges related to task parallelism, deadlock prevention, and runtime performance.
-For example, forked processes piping their combined results to Fig.~\ref{fig:example}'s \ttt{head} may not receive a \ttt{PIPE} signal if \ttt{head} exits prior to opening all pipes.
-% A final challenge has to do with practical issues, as even manual parallelization of a given shell script is far from trivial.
-% Parallelizing a script without reimplementing all of its commands, requires primitives to compose existing commands to achieve parallelization.
-%  Designing correct and efficient compositions is particularly challenging in the context of \unix, and requires addressing issues related to scheduling, synchronization, and deadlock prevention. 
-Moreover, several commands such as \ttt{sort} and \ttt{uniq} require specialized data aggregators in order to be correctly parallelized.
-%% Developing correct, high-performance combiners for certain classes of commands such as \ttt{sort} and \ttt{uniq} is non trivial.
-%% Moreover, the \unix environment has its set of quirks % about opening streams, deadling with FIFOs,
-%%   and any attempt at parallelization will have to take care of these challenges---related to task parallelism, deadlock prevention, and runtime performance.
-
-% \heading{Generation}
-% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
-% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams
-
-\subsection{\sys Design Overview}
-\label{bg:overview}
-
-At a high level, \sys takes as input a POSIX shell script like the one in Fig.~\ref{fig:example} and a parameter \ttt{--width}, and outputs a new script, a fragment of which is shown in Fig.~\ref{fig:example2}---for \ttt{--width=2}, \ie 2$\times$ \{\ttt{grep}, \ttt{tr}, \ttt{cut}, \etc\}.
-% concurrent version
-% To expose and exploit data parallelism, % in a given shell script,
-% %% parallel execution,
-% \sys analyzes individual program fragments, addressing the challenges mentioned earlier.
-% 
-% \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model~\sx{ir} by identifying and transforming special regions termed \emph{dataflow regions} explained below.
-% nv idea: let's call them may-par regions
-% It then runs a series of iterative analyses and transformations within these regions aimed at pushing parallelism as far as possible without risking breaking semantics.
-% Finally, the compiler converts the DFG back into a shell script, replacing nodes with instantiated commands or primitives taken from its runtime library~\sx{optimizer}.
-% An overview of \sys is shown in Fig.~\ref{fig:overview} and described below.
-% First, \sys's source-to-source compiler translates the script to a dataflow graph (DFG) model % ~\sx{dataflow-regions}
-%   by identifying and transforming special regions termed \emph{dataflow regions} and explained below. % nv idea: let's call them may-par regions
-% \kk{eurosys: I suggest to replace the above sentence with the one below.
-There are a couple reasons.
-First I don't like the term source-to-source compiler because this sentence only describes the shell-graph compilation pass.
-Second, I would like to have some intuition about what there regions are (rather than deferring explanation for later).
-Third each region is tranformed to a DFG and not the whole script.}
-\sys first identifies sections of the script that lack synchronization and scheduling constraints and converts them to dataflow graphs (DFGs).
-It then performs a series of DFG transformations that expose parallelism without breaking semantics, by expanding the DFG to the desired \ttt{width}.
-% ~\sx{ir:transformations}.
-Finally, \sys converts these DFGs back to a shell script augmented with \sys-provided commands. % ~\sx{optimizer}.
-The script is handed off to the user's original shell interpreter for execution.
-\sys addresses the aforementioned challenges~\sx{bg:challenges} as below.
-
-\heading{Commands}
-To understand standard commands available in any shell, \sys groups POSIX and GNU commands into a small but well-defined set of \emph{parallelizability classes}~\sx{cmd}.
-Rather than describing a command's full observable behavior, these classes focus on information that is important for data parallelism.
-To allow other commands to use its transformations, \sys defines a light \emph{annotation language} for describing a command's parallelizability class~\sx{ext}.
-Annotations are expressed once per command rather than once per script and aim towards command developers rather than its users, so that they can quickly and easily capture the characteristics of the commands they develop.
-% We haven't introduced the notion of flags
-% A command developer can only annotate the default behavior;
-%   if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.
-
-% Given a pipeline of commands and their parallelizability characteristics, \sys's next challenge is 
-% \sys achieves this through a series of rewriting passes that perform dataflow analysis to identify distributable regions.
-% It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
-
-\begin{figure}[t]
-\centering
-% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_example2.pdf}}
-\begin{minted}[fontsize=\footnotesize,escapeinside=()]{bash}
-mkfifo $t{0,1...}
-curl $base/$y > $t0 & cat $t0 | split $t1 $t2 &
-cat $t1 | grep gz > $t3 & 
-cat $t2 | grep gz > $t4 &
-...
-cat $t9 | sort -rn > $t11 & cat $t10 | sort -rn > $t12 &
-cat $t11 | eager > $t13 & cat $t12 | eager > $t14 &
-sort -mrn $t13 $t14 > $t15 &
-cat $t15 | head -n1 > $out1 &
-wait $! && get-pids | xargs -n 1 kill -SIGPIPE
-\end{minted}
-\caption{
-  \textbf{Output of \ttiny{pash --width=2} for Fig.~\ref{fig:example} (fragment).}
-  \sys orchestrates parallel execution through named pipes, parallel operators, and custom runtime primitives---\eg \ttiny{eager}, \ttiny{split}, and \ttiny{get-pids}.
-}
-\vspace{-15pt}
-\label{fig:example2}
-\end{figure}
-
-\heading{Scripts}
-To maintain sequential semantics, \sys first analyzes a script to identify \emph{dataflow regions} containing commands that are candidates for parallelization~\sx{dataflow-regions}.
-This analysis is guided by the script structure: % to identify parallelization candidates:
-  some constructs expose parallelism; others enforce synchronization.
-% \sys first needs to understand the POSIX constructs used to compose larger programs, indicating different levels of candidacy for pararllelization.
-%% To assign a candidacy for parallelization, \sys breaks POSIX constructs into a few broad classes:
-%%  (i) the parallel-composition operator (\ttt{\&}) exposes task parallelism,
-%%  (ii) special commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
-%%  (iii) the pipe operator (\ttt{|}) is a candidate for both task and data parallelism, and
-%%  (iv)  other composition operators such as \ttt{\&\&}, \ttt{||}, and \ttt{;} are barrier candidates.
-% Since the shell representation only exposes coarse grained parallelization opportunities, 
-\sys then converts each dataflow region to a \emph{dataflow graph} (DFG)~\sx{graph-components}, 
-%% Provided information commands and POSIX constructs, \sys needs to analyze the given script and identify program fragments that are candidates for parallelization.
-%% To achieve this, it converts sections of the script that are amenable to parallelization into a dataflow graph (DFG) representation.
-% Second, \sys defines a dataflow graph model that is suitable for the semantics of the POSIX shell and proposes a set of parallelization optimizations that preserve the behavior of the dataflow graph.
-  a flexible representation that enables a series of local transformations to expose data parallelism, converting the graph into its parallel equivalent~\sx{ir:transformations}.
-%% , in which DFGs indicate the dataflow regions.
-Further transformations compile the DFG back to a shell script that uses POSIX constructs to guide parallelism explicitly while aiming at preserving the semantics of the sequential program~\sx{backend}.
-
-\heading{Implementation}
-\sys addresses several practical challenges through a set of constructs it provides---\ie 
-%% linkable
-modular components for augmenting command composition~\sx{impl}.
-%% for stream splitting and merging.
-It also provides a small and efficient \emph{aggregator library} targeting a large set of parallelizable commands.
-%% For the vast majority of commands in one of the parallelizable classes, it also provides  a small library of pure fold function implementations.
-%% \kk{The terms pure, classes, and fold don't mean a lot to the reader at the moment, so I would prefer using more general terminology.}
-All these commands live in the \ttt{PATH} and are addressable by name, which means they can be used like (and by) any other commands.
-
-% Fig~\ref{fig:example2} shows a fragment of \sys's output for the script in Fig.~\ref{fig:example}.
-% The next few sections (\S\ref{parallelizability}--\ref{impl}) discuss the details.
-
-\section{Parallelizability Classes}
-\label{parallelizability}
-
-% To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
-\sys aims at parallelizing data-parallel commands, \ie ones that can process their input in parallel, encoding their characteristics by assigning them to
-\emph{parallelizability classes}.
-% \kk{eurosys: I have two small concerns about this sentence but I don't know how to fix them.
-First, there is a redundancy "parallelizing data-parallel".
-Second and most important, I would like this sentence to say that \sys exploits latent data parallelism in commands, but not that it parallelizes data parallel commands.
-The reason is that I would like to make sure that no reviewer thinks that all we do is parallelizing each command on its own (as some OSDI reviewer thought).}
-%% This information is encoded in a command's \emph{parallelizability class}.
-% Intuitively, a parallelizability 
-% A class represents the level of synchronization required by copies of a command executing in parallel, 
-%   capturing only characteristics that are important for its parallel execution. % , rather than the command's full observable behavior.
-\sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
-% Make the connections that this is like a type, that doesn't capture everything, but is good-enough
-
-This section starts by defining these classes, along with a parallelizability study of the commands in {\sc POSIX} and GNU Coreutils~\sx{cmd}.
-Building on this study, it develops a lightweight annotation language that enables command classification by its developers~\sx{ext}.
-\sys in turn uses this language to annotate POSIX and GNU commands and generate their wrappers, as presented in later sections.
-
-\subsection{Parallelizability of Standard Libraries}
-\label{cmd}
-
-Broadly speaking, shell commands can be split into four major classes with respect to their parallelization characteristics, depending on what kind of state they mutate when processing their input (Tab.\ref{tab:classes}).
-%% how they interact with state.
-These classes are ordered in ascending difficulty (or impossibility) of parallelization.
-In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (but foregoing any performance improvements).
-Commands can change classes depending on their flags,
-% we focus on no-flag defaults for now and 
-which are discussed later~\sx{ext}.
-
-\begin{table}[t]
-\center
-\footnotesize
-\setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Parallelizability Classes}.
-    Broadly, \unix commands can be broken down into to four classes according to their parallelizability properties.
-  }
-}
-\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
-\toprule
-Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
-\midrule
-Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 28 (18\%)        \\  % 22
-Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  8 (7.6\%)            & 9  (5\%)          \\  % 21
-Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  13 (12.4\%)          & 13 (8.3\%)       \\  % 27
-Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 105(67.8\%)          \\  % 25
-% Irreversible     & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     & 5  (0.4\%)           \\  % 5
-% \midrule
-% Shell PL Constructs             &                         &                  \\
-% \etc                            &                         &                  \\
-\bottomrule
-\vspace{-18pt}
-\end{tabular*}
-\label{tab:classes}
-\end{table}
-
-% \heading{Preliminaries}
-% \heading{POSIX, GNU Core-utils, and beyond}
-
-\heading{Stateless Commands}
-The first class, \sta, contains commands that operate on individual line elements of their input, without maintaining state across invocations.
-These are commands that can be expressed as a purely functional \emph{map} or \emph{filter}---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
-They may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but always return empty output for empty input.
-Workloads that use only stateless commands are trivial to parallelize:
-  they do not require any synchronization to maintain correctness, nor caution about where to split inputs.
- % map :: (a -> b) -> [a] -> [b]
-
-The choice of line as the data element strikes a convenient balance between coarse-grained (files) and fine-grained (characters) separation while staying in line with \unix's core abstractions.
-This choice can affect the allocation of commands in \sta, as many of its commands (about 1/3) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
-This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
-However, it may be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics.
-
-\heading{Parallelizable Pure Commands}
-The second class, \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
-The details of this state and its propagation during element processing affect their parallelizability characteristics.
-Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc} simply maintains a counter.
-Other commands, such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
-% 
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
-% \caption{
-%   \textbf{Hierarchy of parallelizability classes.}
-%   Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
-%   \nv{The figure is incorrect---it's actually a lattice.}
-% }
-% \vspace{-15pt}
-% \label{fig:hierarchy}
-% \end{figure}
-
-Often these commands do not operate in an online fashion, but need to block until the end of a stream.
-A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
-Such constraints affect task parallelism, but not data parallelism:
-  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---\ie by encoding it as a group of (parallel) $map$ functions followed by an $aggregate$ that merges the results.
-
-% \kk{Maybe show an example of map-fold for sort or wc}
-% \nv{We agreed to put it end of §5}
-
-\heading{Non-parallelizable Pure Commands}
-The third class, \npu, contains commands that, while purely functional, cannot be parallelized.
-This is because their internal state depends on prior state in the same pass in non-trivial ways. % should we say something about state machines?
-For example, hashing commands such as \ttt{sha1sum} maintain complex state that has to be updated sequentially.
-If parallelized on a single input, each stage would need to wait on the results of all previous stages, foregoing any parallelism benefits.
-
-It is worth noting that while these commands are not parallelizable at the granularity of a single input, they are still parallelizable across different inputs.
-For example, a web crawler involving hashing to compare individual pages would allow \ttt{sha1sum} to proceed in parallel for different pages.
-
-\heading{Side-effectful Commands}
-The last class, \sid, contains commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
-Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.
-
-This is the largest class, for two reasons.
-First, it includes commands related to the file-system---a central abstraction of the \unix design and philosophy~\cite{unix}.
-In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
-Second, it contains commands that do not consume input or do not produce output---and thus are not amenable to data parallelism.
-For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands interfacing with kernel- or hardware-generated information and do not consume any input from user programs.
-  %A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
-
-% \heading{Discussion}
-% There are a few technical details about \sys's design decisions worth noting.
-
-% %\nv{I took this out because it tooks about \sys's design/implementation that goes far afield}
-% In terms of inputs, \sys supports commands with multiple input (resp. output) streams, by prepending (resp. appending) shell built-ins that merge input (resp. split output) streams~\sx{impl}.
-% Thus, a command can be modelled as having one input and one output stream, a representation that is used throughout \sys including the extensibility DSL~\sx{ext}.
-% 
-% %\nv{I took this out because there are no more classes now}
-% The focus for the rest paper will be commands in \sta and \pur.
-% \sys currently treats other commands as non-distributable~\sx{discussion}.
-% In future work, by building infrastructure to handle other classes, \sys will be able to (i) extract more parallelism and (ii) handle larger programs.
-% Even with only these two classes, however, \sys distributes highly useful pipelines that are strictly more expressive than programs written in popular frameworks such as Hadoop~\cite{mapreduce:08} and Spark~\cite{spark:12}~\sx{eval}.
-
-\subsection{Extensibility Annotations}
-\label{ext}
-
-% TODO: \nv{Note command breadth/impossibility of our task.}
-% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html
-
-%% Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
-%% % ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
-%% %    | -h :: ( Sless ) => { Out }                              -- if needed
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%% (bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
-%%   |-h :: (Sless)($$\Rightarrow$$){Out}                              
-%% \end{lstlisting}
-
-%% It states that the command defaults to the Pure class.
-%% It's operation can be thought as a transformation from input streams to output streams.
-%% It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
-%% % TODO: check actual man page for how to specify files
-%%   it writes to the output and error stream rather than a file.
-%% The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
-%%   its output is a constant function.
-
-% % FIXME: \nv{TODO: remember to bubble up(me)}
-% An important characteristic of the \unix shell is its
-% language-agnostic extensibility~\sx{bg}.  Users composing pipelines
-% are free to install custom commands available from various sources.
-% Being general programs, commands are developed in a variety of
-% languages and are extended over long periods of time.
-% %% \kk{Text point: This restricts the space of solutions for
-% %%   distribution of shell scripts}
-% As a result, any static characterization of shell commands would bs
-% unsatisfactory and would quickly become obsolete, supporting only a
-% small subset of the commands needed in practice.  This poses a
-% challenge to the developers of individual commands---how can they
-% augment their commands to capture information critical to \sys without
-% too much effort?
-% %% \kk{Text point: The characteristics of a good solution}
-
-To address the challenge of language-agnostic extensibility~\sx{bg}, \sys allows communicating several key details about command parallelizability through lightweight annotations.
-These annotations can be used by both developers of new commands---including users developing their own scripts---as well as developers maintaining existing commands.
-The latter can express additions or changes to the command's implementation or interface, important as commands are maintained or extended over long periods of time.
-% \kk{eurosys: I have some concerns about the above sentence.
-First, I don't want to say that users developing their own scripts can use the annotation language because this goes against what we mentioned in the intro.
-Second, the last sentence is too prominent the way it is written now.
-I would prefer just saying that the annotations are aimed for command developers who can create an annotation when they develop a new command, or when a command's interface changes---which is important since commands are maintained and extended over long periods of time.}
-
-\heading{Key Concerns}
-\sys's annotations focus on three crucial concerns about a command:
-  (C1) its parallelizability class,
-  (C2) its inputs and outputs, and the characteristics of its input consumption, and
-  % (ii) for commands that support multiple inputs, the characteristics of input consumption, and
-  (C3) how flags affect its class, inputs, and outputs.
-The first concern was discussed extensively in the previous section; we now turn to the latter two.
-
-To be able to accurately construct the DFG representation, \sys needs to know certain details about a command's inputs and outputs.
-There are a few reasons for this.
-One reason is due to \sys's DFG representation, which connects commands with each other.
-  To perform this connection correctly, \sys needs to know the outputs of a command. % to correctly connect them with the inputs of the next.
-A second reason is due to ordering:
-  as some commands consume their inputs in a certain order, this order needs to be maintained in the parallel version of the program.
-For example, consider the data-parallel version of \ttt{grep "foo" f1 f2}:
-
-% \begin{lstlisting}[language=sh, numbers=none]
-%   mkfifo t1 t2
-%   grep "foo" f1 > t1 &
-%   grep "foo" f2 > t2 &
-%   cat t1 t2
-% \end{lstlisting}
-\begin{minted}[fontsize=\footnotesize]{bash}
-  mkfifo t1 t2
-  grep "foo" f1 > t1 & grep "foo" f2 > t2 & cat t1 t2
-\end{minted}
-
-\noindent
-This is correct only because \sys knows that \ttt{grep} in the sequential program reads first from \ttt{f1} and then from \ttt{f2}.
-
-Command flags (options) are a particularly popular way of controlling a command's execution, directly affecting its parallelizability classification.
-Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
-For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.
-
-% As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
-% For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} in \sid that writes debugging output to a file, it ends up in \sid.
-
-%% As simple examples, consider \ttt{cat} and \ttt{sort}.
-%% By default \ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter.
-%% Conversely, \ttt{sort} defaults to \pur, but \ttt{with} \ttt{-h} it jumps into \sta because it generates a static stream regardless of other input.
-% TODO: closure?
-
-%% This is actually wrong. sort -h is not stateless since if called several times it produces the same output many times.
-%% nv: the above does not mean it is not stateless; if apply `map` multiple times on an input I will get combined results.
-
-%% Can be tweaked according to this:
-%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars
-
-%% \heading{Command Inputs and Outputs}
-%% A second concern captured by the AL is the order of inputs and outputs.
-%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
-%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
-%% % but these maintain different invariants, enabling very different parallelizability characteristics.
-
-%% Consider the command \ttt{comm} that takes as input two files and
-%% performs a join operation. When invoked without any option, it
-%% produces a three-column output. The first column contains lines that
-%% are unique to the first file, the second column contains lines that are
-%% unique to the second file, and the last column contains lines that
-%% exist in both files. In the general case \ttt{comm} is in \pur, since
-%% it needs to wait until it reads both files completely before it can
-%% output its results.
-
-%% However, \ttt{comm} can also be invoked using some options that
-%% suppress the output of any combination of columns. More precisely, the
-%% option \ttt{-1} suppresses the first column of the output. Similarly,
-%% \ttt{-2} and \ttt{-3} supress the second and third column. Using both
-%% \ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command jumps to \sta
-%% regarding its first (resp. the second) file argument.
-
-% \ttt{comm} is in \pur, but with \ttt{/input1} it jumps into \sta---as it is now partially evaluated, it becomes stateless with respect to the second argument.
-
-\heading{Example Annotations}
-The parallelizability properties of each command are specified in an annotation record that contains several components, each addressing one of the key concerns C1--3.
-At the top level, each annotation record contains a set of clauses, each one identified by a predicate on the command arguments (C3). 
-Each clause contains an assignment for the command, indicating its class (C1), and the sequence of its inputs and outputs (C2).
-%  (a JSON object) that
-% Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
-% \footnote{
-%   Note that the single character \ttt{-} is not an option as it is most often used to designate that a command reads from its standard input.
-%   \nv{
-%     There are options of of the form --f "x; y", and BSD options do not take \ttt{-} (like \ttt{ps aux}).
-%     I am also not sure how important this last sentence is, as it did not answer a question I had---I expected arguments to be anything: if soneome wants to put ``pizza'' the DSL should not disallow it.
-%   }
-% }
-% The complete annotation grammar is shown in Appendix~\ref{alg}. \str
-
-For example, consider \ttt{comm}, which performs a join-like operation on its two inputs to identify common elements.
-When \ttt{comm} is invoked without any flags, it produces a three-column output:
-  the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
-A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output column(s).
-% The first clause from its annotation record is shown below:
-Below is its annotation record (left) and the implementation of the record's first clause (right):
-%
-% \begin{minted}[fontsize=\footnotesize]{json-object}
-%   "predicate":
-%     { 
-%       "operator": "all",
-%       "operands": ["-1", "-3"] 
-%     },    
-%   "class": "stateless",
-%   "inputs": ["args[1]"],
-%   "outputs": ["stdout"],
-%   "options": ["stdin-hyphen"]
-% \end{minted}
-
-\hspace*{-0.3cm}\includegraphics[width=\columnwidth]{\detokenize{./figs/pash_annotation.pdf}}
-
-% \begin{lstlisting}[language=json, numbers=none]
-%   "predicate":
-%     { 
-%       "operator": "all",
-%       "operands": ["-1", "-3"] 
-%     },    
-%   "class": "stateless",
-%   "inputs": ["args[1]"],
-%   "outputs": ["stdout"],
-%   "options": ["stdin-hyphen"]
-% \end{lstlisting}
-% comm {
-% | -1 /\ -3 => (S, [arg[1]], [stdout])
-% | -2 /\ -3 => (S, [arg[0]], [stdout])
-% | _  => (P, [arg[0], arg[1]], [stdout])
-% }
-
-\noindent
-As the first and third columns are suppressed, \ttt{comm} can be categorized in \sta by regarding its first file argument as a static ``configuration'' input;
-% NV: I don't think this is correct, because the small input might be the other argument;  you could even have `comm - -`
-  this input accompanies every replicated instance of the command executing in parallel.
-The second clause is symmetric to the first with respect to the second input. 
-The third clause is the general case, classifying \ttt{comm} in \pur, as \ttt{comm} needs to read both inputs completely before outputting its results.
-Option \ttt{stdin-hyphen} indicates that \ttt{-} represents standard input if given as a file argument, and option \ttt{short-combined} indicates that flags can be combined, as in \ttt{comm -13}.
-% The third clause is the general case, which includes \ttt{comm} invocations without any options, and is captured by the third clause.
-% There, \ttt{comm} is classified as \pur as it needs to % wait until it 
-%   read both inputs completely before it can output its results.
-% Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.
-
-%%  This case is described in the third clause of the annotation
-%% record, where \ttt{comm} is classified as \pur, since it needs to wait
-%% until it reads both files completely before it can output its
-%% results. Its inputs are its first and second arguments (excluding the
-%% options), and its output is produced in \ttt{stdout}.
-
-%% The first and second clauses describe the case when \ttt{comm} is
-%% invoked with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}. In this
-%% case, since the first and third columns are suppressed, \ttt{comm} can
-%% be considered \sta by regarding its first file argument a static
-%% ``configuration'' input. \kk{Does this need more explanation?}
-
-%% \kk{I
-%%   think that the language definition should be in a technical report,
-%%   and here we should only give intuition and the \ttt{comm} example}
-%% is designed to classify commands into multiple categories given their
-%% options.
-
-%% It assumes that the arguments of a command can always be separated to
-%% options and file arguments (checking if they start with a
-%% \ttt{-}~\footnote{Note that the single character \ttt{-} is not an
-%%   option as it is most often used to designate that a command reads
-%%   from its standard input}). A file written in the DSL contains a
-%% record containing a mapping from predicates to assignments for each
-%% command. Predicates can be simple boolean formulas on the existence
-%% and the value of any option. Each assignment contains the command
-%% category, the sequence of its input sources, and its output
-%% source. Inputs to commands are a sequence of file identifiers and its
-%% standard input. Similarly, the output is either its standard output or
-%% a file identifier. An example record for \ttt{comm} can be seen below:
-
-%% \begin{lstlisting}[float=h, numbers=none]
-%%   comm {
-%%     | -1 and -3 =>
-%%       (stateless, [args[1]], stdout)
-%%     | -2 and -3 =>
-%%       (stateless, [args[0]], stdout)
-%%     | otherwise =>
-%%       (pure, [args[0]], stdout)
-%%   }
-%% \end{lstlisting}
-
-%% The semantics of the language is straightforward: given a command, the
-%% interpreter collects its options and then returns the assignment of
-%% the first predicate that is satisfied. The final predicate is always
-%% satisfied. For example the following two invocations of \ttt{comm}
-%% would be classified as follows:
-
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%%  comm -13 f1 f2 # => (stateless, f2, stdout)
-%%  comm f1 f2     # => (pure, f1, stdout)
-%% \end{lstlisting}
-
-\tr{\kk{We don't need to designate the stderror, because we assume that it
-  is never the main output of a command and it will never be used by
-  the input of another command in the pipeline. If someone indeed
-  wants to do this, they can just use a redirect I think to get around
-  it.}}
-
-\tr{\kk{If we talk about the language it would be good to give some
-  statistics about how many commands (out of the ones that we have
-  classified) can be represented with one, two clauses etc. If we have
-  a different solution (like Python function that given a command and
-  its arguments returns the category) then we could talk about how
-  many lines of code it took to categorize all the commands that we
-  did.}}
-
-\tr{\kk{Note: Since a categorization language will probably not be
-  complete (especially the one showing the input arguments) there
-  should be a backup mechanism, when the language is not expressive
-  enough, like a Python function that categorizes the command, and
-  returns its input argument if it is stateless.}}
-
-\tr{\kk{(Maybe) We should have a crisp point about why we have this categorization
-  language, and why don't we just allow someone to write a function in
-  Python for each command, that given the command and its arguments,
-  returns whether the command category. Possible arguments include,
-  the fact that this language is simpler to use, especially by non
-  experts that just run some script but have installed some commands
-  that are not ``supported''. Another possible argument is that we
-  could be able to reason about the constructs of the categorization
-  lagnauge, and that they will be easier to read and
-  understand. Another argument is that since these categorizations
-  should be shareable among users, it would be bad to execute
-  arbitrary Python code, so using this language, expressivity is
-  limited. All of these arguments are a little bit weak though. Niko,
-  what do you think?}}
-
-\heading{Complete Annotation Language}
-The complete annotation language contains 6 operators, one of which supports regular expressions.
-It was used to annotate 47 commands, totaling 708 lines of JSON---an effort that took about 2--3 hours.
-Annotation records are by default conservative so as to not jeopardize correctness, but can be incrementally refined to capture parallelizability when using increasingly complex combinations of flags.
-The language is extensible with more operators (as long as the developer defines their semantics);
-  it also supports writing arbitrary Python code for commands whose properties are difficult to capture---\eg higher order \ttt{xargs}, whose parallelizability class depends on the class of the first-order command that it invokes.
-% \str
-% \kk{Mention that some commands, such as xargs are handled internally due to their particularities.}
-
-\heading{Custom Aggregators}
-For commands in \sta, the annotations are enough to enable parallelization:
-  commands are applied to parts of their input in parallel, and their outputs are simply concatenated.
-%% commands either fall under \sta, having straightforward aggregators, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
-\kk{pure commands are what we have to provide aggregators for, so this is not totally valid.}
-%% Unfortunately, there are commands that
-%% require less obvious tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.
-\kk{I would propose deleting this last sentence.
-we have already separated commands into classes, so this sentence makes it seem like the classes don't help at all.}
-%
-To support the parallelization of arbitrary commands in \pur, \sys allows supplying custom \emph{map} and \emph{aggregate} functions.
-In line with the \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
-  (i) \emph{map} is in \sta and \emph{aggregate} is in \pur,
-  (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{aggregate} can consume (and combine) the results of multiple \emph{map} invocations, and
-  (iii) their composition produces the same output as the original command.
-\sys can use the \emph{map} and \emph{aggregate} functions in its graph transformations~\sx{ir} to further expose parallelism.
-% If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{aggregate} equivalents, enabling their full parallelization.
-\sys defines aggregators for many POSIX and GNU commands in \pur, doubling as both its standard library and an exemplar for community efforts tackling other commands.
-
-\tr{\kk{I am not sure a general interface is so easy to design. It needs
-  more though. It might be beneficial to just talk about sort and wc
-  here and how we implemented them and nothing more. Or maybe this
-  could then go to the implementation? Or maybe say that one can write
-  a Python function that given a node of the graph, transforms it into
-  many. I am not sure what is best...}}
-
-% As mentioned in parallelizability, pure commands can be broken down
-% into different categories. We might be able to get ideas about this
-% from this paper:
-% \url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
-% continuation in PLDI 2019.
-% FIXME: \nv{Cite this paper?}
-\tr{Can we find a solution for the commands in coreutils?}
-
-\section{Dataflow Graph Model}
-\label{ir}
-
-\sys's core is an abstract dataflow graph (DFG) model~\sx{graph-components} that is used as the intermediate representation on which \sys performs parallelization-exposing transformations~\sx{ir:transformations}.
-%% representation that is expressive enough to model data parallelism in shell programs.
-%% support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
-\sys first lifts sections of the input script to the DFG representation~\sx{dataflow-regions}, 
-  then performs transformations to expose parallelism (up to the desired \ttt{--width})~\sx{ir:transformations}, 
-  and finally instantiates each DFG back to a parallel shell script~\sx{backend}.  A
-fundamental difference with other DFG models is that \sys's DFG encodes the
-order in which a node reads its input streams (not just the order of input elements per stream), which in turn enables a set of
-graph transformations that can be iteratively applied to expose
-parallelization opportunities for \tsta and \tpur commands.
-%% data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
-%% These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources. 
-% The section script-to-DFG translation~\sx{dataflow-regions}, 
-%   then define the DFG model and describe the correspondence with shell scripts~\sx{graph-components},
-%   discuss the transformations that \sys performs~\sx{ir:transformations}, and close with
-%   the DFG-to-script translation~\sx{backend}.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/pash_dfg.pdf}}
-  \vspace{-25pt}
-  \caption{
-    \textbf{From a script AST to DFGs.}
-    The AST on the left has two dataflow regions, each not extending beyond \ttiny{&&}~\cf{dataflow-regions}.
-  }
-  \vspace{-15pt}
-  \label{fig:ast-dfg}
-\end{figure}
-
-\subsection{Frontend: From a Sequential Script to DFGs}
-\label{dataflow-regions}
-
-% To optimize a script, \sys translates it to a DFG~\sx{ir}.
-% As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}, ~\S\ref{bg}---\sys introduces the notion of \emph{dataflow program regions} and a translation pass for converting them to DFGs.
-% 
-%% In addition, the shell's highly dynamic
-%% nature compounds the challenge---\eg by not knowing what files does an
-%% unexpanded command argument refer to.
-% 
-%% \kk{Give examples of writing to environment variables, producing
-%%   irreversible side effects, the ; and \&\& operators.}
-%% To address the first issue, we introduce the notion of dataflow
-%% program regions and a translation pass for converting them to a
-%% dataflow graph. To address the second issue, we limit dynamic
-%% components by deferring translation as late as possible using a
-%% wrapping scheme.
-% 
-%% The translation pass identifies distributable regions and
-%% translates each of them to a dataflow graph, provided the abstract
-%% syntax tree (AST) of a shell script.  The AST is produced by
-%% \ttt{LibDash}, a POSIX compliant parser~\cite{libdash}.
-% 
-%
-%% \kk{The goal of this subsection is to explain that Dish does a safe
-%%   but effective analysis to identify which parts of the scripts to
-%%   parallelize.}
-
-\heading{Dataflow Regions}
-In order to apply the graph transformations that expose data parallelism, \sys first has to identify program sub-expressions that can be safely transformed to a dataflow graph, 
-  \ie sub-expressions that (i) do not impose any scheduling or synchronization constraints (\eg by using \ttt{;}), and (ii) take a set of files as inputs and produce a set of files as outputs.
-% The DFG model and associated transformations are applied only to program sub-ex\-pres\-sions that \sys deems would be safe to parallelizable---\ie ones that would not break sequential semantics, as long as the commands involved in these regions fall under the right classes.
-%% respect to their sequential execution.
-The search for these regions is guided by the shell language and the structure of a particular program.
-These contain information about (i) fragments that can be executed independently, and (ii) barriers that are natural synchronization points. % for enforcing synchronization.
-Consider this code fragment:
-
-% \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-%  cat f1 f2 | grep "foo" > f3 && sort f3 
-% \end{lstlisting}
-\begin{minted}[fontsize=\footnotesize]{bash}
- cat f1 f2 | grep "foo" > f3 && sort f3 
-\end{minted}
-% # Replace sort with sth better
-
-\noindent
-The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently)
-in the standard shell, but \ttt{sort} waits for their completion prior to start
-(Fig.~\ref{fig:ast-dfg}).
-% 
-Intuitively, dataflow regions correspond to sub-expressions of
-the program that would be allowed to execute independently by
-different processes in the POSIX standard~\cite{posix}. Larger
-dataflow regions can be composed from smaller ones using the
-pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{&}).
-Conversely, all other operators, including sequential composition (\ttt{;}) and logical operators (\ttt{&&}, \ttt{||}), represent
-barrier constructs that do not allow dataflow regions to expand beyond them.
-
-% \tr{I don't know whether I should mention the following: While these
-%   control flow constructs can be parallelized and there has been
-%   research on it, it is orthogonal to our work, and can be
-%   incorporated as future work.}
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%   (tr ... f1 > g1 && tr ... f2 >> g1) &
-%   grep ``foo'' g1
-% \end{lstlisting}
-% 
-% In the above example, while \ttt{tr ... f1 > g1}, \ttt{tr ... f2 >>
-%   g1}, and \ttt{grep ``foo'' g1} are all in different dataflow graphs,
-% the dataflow graph that corresponds to the whole program, has two
-% nodes, \ttt{tr ... f1 > g1 \&\& tr ... f2 >> g1} and \ttt{grep ``foo''
-%   g1}.
-%
-
-\heading{Translation Pass}
-%
-%% \kk{It might be a good idea to change this subsection based on the new
-%%   as-late-as-possible architecture where everything is wrapped and
-%%   translated at the final moment. Ideally, this would become the main
-%%   point of this subsection, as the translation algorithm is kind of
-%%   redundant when having the distributable regions too.}
-%
-\sys's front-end performs a depth-first search on the AST of
-the given shell program.  During this pass, it extends the
-dataflow regions bottom-up, translating their
-independent components to DFG nodes until a barrier
-construct is reached. All AST subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
-original AST where dataflow regions have been replaced with
-DFGs. 
-% and calls to \sys's runtime primitives~\sx{optimizer}.
-
-To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
-To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
-If the command is in \sta, \pur, or \npu, the translation pass initiates a dataflow region that is propagated up the tree.
-
-%% Finally, whenever the algorithm encounters a command, it first
-%% checks whether it is safe to parallelize it. A command is considered
-%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
-% why couldn't we replicate this?
-%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
-
-%% A few illustrative cases of the analysis algorithm are shown in pseudocode below:
-
-%% \begin{lstlisting}[language=python, float=h]
-%%   def translate(node):
-%%     ...
-%%     elif(node.name == 'Pipe'):
-%%       return pipe_graphs([translate(child)
-%%         for child in node.children])
-%%     elif(node.name == 'And'):
-%%       node.children = [translate(child)
-%%         for child in node.children]
-%%       return node
-%%     elif(node.name == 'Command'):
-%%       if(not safe(node)):
-%%         return make_ir_node(node, 'unsafe')
-%%       else:
-%%         return make_ir_node(node,
-%%           find_category(node))
-%%     ...
-%% \end{lstlisting}
-
-%% Whenever the algorithm encounters a distributable node, it recursively
-%% translates its subcomponents.
-%% It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
-%% % what first and second? we didn't say anything about first and second
-%% Before the different dataflow graphs are connected, the
-%% algorithm checks that at most one node in the graph writes at every
-%% file, and that at most one node reads from a file that another node writes
-%% to. This is important, in order to avoid inconsistent behaviour after
-%% distributing nodes in the dataflow graph (\ie by introducing to
-%% concurrent reads and writes to the same file when
-%% parallelizing).
-
-%% Whenever the algorithm encounters a barrier node,
-%% it recursively translates its subcomponents, similarly to the distributable
-%% case.
-%% However, different from the distributable case, it does not merge them in a large dataflow graph.
-
-%% Finally, whenever the algorithm encounters a command, it first
-%% checks whether it is safe to parallelize it. A command is considered
-%% safe for parallelization if (i) it belongs to the \sta or \pur parallelizability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this.
-% why couldn't we replicate this?
-%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
-%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.
-
-Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
-Examples of such information include the values of environment variables, strings that have not been expanded, and sub-shell constructs.
-%% to the the safety properties that are mentioned above cannot be
-%% soundly checked statically.
-For safety purposes, \sys takes a conservative approach, avoiding parallelization of such nodes with incomplete information.
-It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, \eg  an environment variable passed as an argument to a command does not change its parallelizability class.
-
-%% \kk{Can we say anything more about the safety considerations? That (i)
-%%   the translation pass ``checks'' that no nodes write/read or
-%%   write/write to the same file (is that necessary?), that (ii) a
-%%   command should not read from a read-once file. (is that necessary?)}
-
-%% \kk{Ideally we would want to talk here about the as-late-as possible
-%%   dish translation that allows us to be safe despite the Shell's
-%%   dynamic nature. I think that this is a great point about the
-%%   work. Maybe we should also focus on implementing it before the OSDI
-%%   deadline.}
-
-%% The core of \sys is a DFG model, inspired by the ones used
-%% in popular distributed stream processing systems~\sx{related} \kk{Is
-%%   this reference here necessary?}. The dataflow model can express a
-%% large subset of Shell programs and inherently supports the data
-%% parallelism found in shell pipelines by exposing it as task
-%% parallelism. More preciely, we have developed a set of
-%% parallelization-exposing optimizations represented as
-%% semantics-preserving graph transformations. These transformations can
-%% then be applied in an iterative way, optimizing the program to utilize
-%% available computational resources.
-
-%% In order to effectively apply optimizations and distribute a shell
-%% script, \sys first translates it to a manipulable representation.
-
-\subsection{Dataflow Model Definitions}
-\label{graph-components}
-
-The two main shell abstractions are
-  (i) streams, \ie files or pipes, that contain data, and
-  (ii) commands, communicating through streams.
-%% \nv{quick solution: change all files to streams and make the equivalence explicit here}
-\sys's DFG model represents commands as nodes and streams as edges.
-We now introduce the notation used in throughout this section.
-For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
-For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
-We write $\eps$ for the empty word.
-We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
-The $\leq$ order is reflexive, antisymmetric, and transitive (\ie it is a partial order), and is often called the \emph{prefix order}.
-
-\heading{Edges---Streams}
-Edges in the DFG represent streams, the basic data abstraction of the shell.
-They are used as communication channels between nodes in the graph, and as the input or output of the entire graph.
-Edges are represented as possibly unbounded streams of type $D\kstar$.
-Edges can either refer to named files or FIFO pipes used for interprocess communication.
-Edges that do not start from a node in the graph represent the graph input;
-  edges that do not point to a node in the graph represent its outputs.
-%% \nv{OK This distinction is explained in this par; I would put this earlier}
-
-%% \TODO{Show an example pipe and its graph}.
-
-%% \begin{center}
-%% \small
-%% \begin{tikzpicture}[node distance=2.25cm, ->]
-%% \node (S) {};
-%% \node (A) [draw, right of=S, node distance=1.75cm] {node 1};
-%% \node (B) [draw, right of=A] {node 2};
-%% \node (C) [draw, right of=B] {node 3};
-%% \node (E) [right of=C, node distance=1.75cm] {};
-%% \path (S) edge node[above] {input} (A);
-%% \path (A) edge node[above] {stream} (B);
-%% \path (B) edge node[above] {stream} (C);
-%% \path (C) edge node[above] {output} (E);
-%% \end{tikzpicture}
-%% \end{center}
-
-%% \tr{On the other hand, some forms of data parallelism can be exposed
-%%   when knowing the size of the input files. As mentioned in \ref{}
-%%   some pure commands (such as cat -n) only need line information to
-%%   become stateless, and knowing the size of a file could allow the
-%%   system to split it in different chunks that can be processed
-%%   independently. To account for that, edges that refer to an input
-%%   resource contain the number of lines of the file that they refer
-%%   to.}
-
-\heading{Nodes---Commands}
-A node $f$ of the graph represents a function from a (possibly empty)
-list of inputs
-%% streams
-to a list of outputs
-%% streams
-$f : [D\kstar] \to [D\kstar]$, where $D$ represents the basic data
-type of a line of characters.
-%% \km{Is the number of input/output
-%%   streams variable? If $m$ is the number of input streams, is the
-%%   number of output streams a function of $m$? If this the case, then
-%%   it is not captured by this type. Maybe it would help to show a
-%%   couple of example operators together with the types of their
-%%   denotations. A graphical representation (as in
-%%   Fig.~\ref{fig:parallelization-transformation} would also be nice.}
-%% \kk{Is there a reason to restrict the number of outputs to be a
-%%   function of the number of inputs? I think that we don't need such a
-%%   restriction, therefore the above type should be fine.}
-This representation captures all the commands in
-the \tsta, \tpur, and \tnpu. Note that these functions should only produce
-output in the form of files and not perform any other side effect,
-such as sending signals.  We require that the function $f$ is monotone
-with respect to a lifting of the prefix order for a sequence of inputs.
-%% \kk{Formalize
-%%   this lifting if essential for later development}.
-%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
-%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
-%% concatenation and will be omitted when obvious.
-This captures the idea that a node cannot retract output that it has
-already produced.
-
-\heading{Streaming Commands}
-%
-A large subset of the parallelizable \tsta and \tpur classes falls
-into the special category of streaming commands. These commands
-consume their inputs sequentially and one element at a time, with the
-possible exception of static input files, and produce one output
-file. The quintessential example of streaming commands is \ttt{cat},
-which consumes its inputs in order, producing their concatenation as
-output. A more interesting example is \ttt{comm -23 f1 f2}, which has
-\ttt{f2} as its static input (which it needs to read completely before
-consuming anything from \ttt{f1}, which is in turn consumed one
-element at a time). These commands can be represented as functions of
-type $f : D\kstar \times [D\kstar] \to D\kstar$, where the first
-argument represents the concatenation of the inputs that are consumed
-one at a time and the second argument represents the static inputs.
-%% \km{Again, a graphical
-%%   presentation of these operators would be useful. Maybe with a way to
-%%   distinguish between between streaming inputs and static inputs? Btw,
-%%   this distinction is not completely clear to me at this
-%%   point. Clearly, not all programs that we can write are streaming
-%%   (monotone).} \kk{Do you mean that there are shell scripts that are
-%%   non monotone? That is true but we are assuming monotone nodes. Do I
-%%   miss something?}
-%% We focus on these commands because a large subset of \sta and \pur
-%% commands fall in this category.
-%% \nv{Based on the intro of 4 or 4.1 I did not expect streaming.
-I understand edges and nodes, but how do streaming commands fit into the picture?
-the answer might be the last sentence which should be first---reviewer: why am I reading what I'm reading?:P}
-
-%% \kk{Do we want to get in more detail about requirements? Commands must
-%%   be deterministic, they must not do any other side effect (sending
-%%   signals, etc). These assumptions must be checked when the developer
-%%   designates the categories. Commands are assumed to always terminate
-%%   gracefully. \sys changes the termination behaviour of commands as a
-%%   sequential command that would exit prematurely might not exit
-%%   completely if parallelized.}
-
-%% \kk{If we have time and space we could give an example of a script
-%%   together with its DFG to give more intuition about nodes
-%%   and streaming commands.}
-
-%% An example of a node with one input and one output stream is the
-%% following command:
-
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%%  grep ``foo''
-%% \end{lstlisting}
-
-%% \noindent
-%% This command takes a stream of lines from its standard input and
-%% returns a stream of lines on its standard output.
-
-%% An important observation is that even though commands might read their
-%% input from several input streams, the order in which they read is not
-%% always arbitrary. An illustrative example is the command \ttt{cat},
-%% which consumes its input one stream at a time:
-
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%%  cat x - y
-%% \end{lstlisting}
-
-%% \noindent
-%% In the example above, \ttt{cat} reads file \ttt{x} until it encounters
-%% the EOF character, then reads from \ttt{stdin} until in encounters
-%% EOF, and finally reads from \ttt{y}. Other commands that read their
-%% inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
-%% order in which a node consumes its input is necessary to enable
-%% certain optimizations on the graph.
-
-%% \kk{Some could read in sequence (like cat), some in lockstep (like
-%%   zip), some read some files statically and the rest as streams (like
-%%   comm), some read arbitrarily.}
-
-%% \kk{I need to mention that there are nodes that we know more
-%%   properties of. For example cat, tee, zip, etc... and these are
-%%   handled specially because of their very important and special
-%%   properties.}
-
-%% \begin{figure*}
-%% \centering
-%% \begin{tikzpicture}[->, >=to, auto, node distance=1cm, semithick, transform shape, inner sep=2pt]
-%% %
-%% \small
-%% %
-%% \node (M1) {};
-%% \node (M2) [draw, right of=M1, node distance=2cm] {Map};
-%% \node (M3) [draw, right of=M2, node distance=3cm] {GroupBy(Reduce)};
-%% \node (M4) [right of=M3, node distance=3cm] {};
-%% %
-%% \path (M1) edge node {input} (M2);
-%% \path (M2) edge node {stream} (M3);
-%% \path (M3) edge node {output} (M4);
-%% \end{tikzpicture}
-%% \end{figure*}
-
-%% \begin{figure*}
-%% \newcommand{\Split}{\text{Split}}
-%% \newcommand{\Merge}{\text{Merge}}
-%% \newcommand{\Map}{\text{Map}}
-%% \newcommand{\GroupBy}{\text{GroupBy}}
-%% \newcommand{\Reduce}{\text{Reduce}}
-%% \centering
-%% \begin{tikzpicture}[->, >=to, auto, node distance=0.6cm, semithick, transform shape, inner sep=1.75pt]
-%% %
-%% \footnotesize
-%% %
-%% \node (M1) {};
-%% \node (M2) [draw, right of=M1, node distance=1.3cm] {$\Split$};
-%% \node (M3) [right of=M2, node distance=1.5cm] {};
-%% \node (M4) [right of=M3, node distance=2.5cm] {};
-%% \node (M5) [draw, right of=M4, node distance=2.2cm] {$\Merge$};
-%% \node (M6) [draw, right of=M5, node distance=3.8cm] {$\GroupBy(K_2,\Reduce)$};
-%% \node (M7) [draw, right of=M6, node distance=3cm] {$\Merge_K$};
-%% \node (M8) [right of=M7, node distance=2cm] {};
-%% %
-%% \node (T3) [draw, above of=M3] {$\Map$};
-%% \node (T4) [draw, above of=M4] {$\Split_K$};
-%% \node (T5) [draw, above of=M5] {$\Merge$};
-%% \node (T6) [draw, above of=M6] {$\GroupBy(K_1,\Reduce)$};
-%% %
-%% \node (B3) [draw, below of=M3] {$\Map$};
-%% \node (B4) [draw, below of=M4] {$\Split_K$};
-%% \node (B5) [draw, below of=M5] {$\Merge$};
-%% \node (B6) [draw, below of=M6] {$\GroupBy(K_3,\Reduce)$};
-%% %
-%% \path (M1) edge node {$A$} (M2);
-%% \path (M2) edge[bend left=10] node {$A$} (T3);
-%% \path (M2) edge[bend right=10] node[swap] {$A$} (B3);
-%% \path (M5) edge node {$K_2 \times B$} (M6);
-%% \path (M6) edge (M7);
-%% \path (M7) edge node {$K \times B$} (M8);
-%% %
-%% \path (T3) edge node {$K \times B$} (T4);
-%% \path (T4) edge (T5);
-%% \path (T4) edge (M5);
-%% \path (T4.330) edge[bend right=10] (B5.170);
-%% \path (T5) edge node {$K_1 \times B$} (T6);
-%% \path (T6) edge[bend left=10] node {$K_1 \times B$} (M7);
-%% %
-%% \path (B3) edge node {$K \times B$} (B4);
-%% \path (B4) edge (B5);
-%% \path (B4) edge (M5);
-%% \path (B4.30) edge[bend left=10] (T5.190);
-%% \path (B5) edge node {$K_3 \times B$} (B6);
-%% \path (B6) edge[bend right=10] node[swap] {$K_3 \times B$} (M7);
-%% %
-%% \end{tikzpicture}
-%% \caption{Distributed implementation of the map-reduce pipeline $\Map \gg \GroupBy(K,\Reduce)$.}
-%% \label{fig:mapReduce}
-%% \end{figure*}
-
-\subsection{Graph Transformations}
-\label{ir:transformations}
-
-%% \kk{Move this assumption in the transformation: Nodes may read from
-%%   more files than just the files of their input stream (resp. write to
-%%   more files than their output stream).  However, these files are
-%%   static (\eg used for configuration or logging) and do not represent
-%%   streams of data, and therefore are not considered part of the
-%%   DFG.  For now, we assume that these static files can only
-%%   be accessed by one node, thus ensuring that they do not interfere
-%%   with the distributed implementation.
-%% %% Because of the assumption in section Front End \ref{}, that each file
-%% %% exists only once in the DFG, we can safely assume that
-%% %% reading and writing to these static files does not interfere with the
-%% %% distributed implementation, and doesn't alter the behaviour of the
-%% %% program.
-%% }
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_cat.pdf}}
-\vspace{-28pt}
-\caption{
-  \textbf{Stateless parallelization transformation.}
-  The \ttiny{cat} node is commuted with the stateless node to utilize available data parallelism.
-}
-\vspace{-15pt}
-\label{fig:parallelization-transformation}
-\end{figure}
-
-\sys defines a set of semantics-preserving graph transformations
-that act as parallelization-exposing optimizations. Both the domain
-and range of these transformations is a graph in \sys's DFG model
-and therefore transformations can be composed arbitrarily and in any order.
-%% Having a DFG representation allows us to define
-%% parallelization optimizations as graph transformations.
-Before describing the different types of transformations, we formalize
-the intuition behind classes \sta and \pur described informally
-earlier~\sx{cmd}.
-
-%% We focus on \todo{streaming} commands (as defined in
-%% \cref{graph-components}) that constitute a large subset of shell
-%% commands.
-
-%% \nv{TODOs
-%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
-%% }
-
-%% \kk{Maybe
-%%   have an argument here that we show in the evaluation that handling
-%%   these commands is adequate for a big variety of data processing
-%%   scripts (and most stateless and pure commands), from several
-%%   microbenchmarks to some large-realistic macrobenchmarks.}.
-
-\heading{Stateless and Parallelizable Pure Commands}
-Stateless commands such as \ttt{tr}
-operate independently on individual elements of the stream
-(characters, lines, or files) without maintaining any state~\sx{cmd}. In this
-work, we consider lines as the data quantum, thus we consider stateless only commands that are
-stateless with respect to lines (or any finer granularity such as
-characters). Formally, a streaming command $f$ is stateless if it commutes with
-the operation of concatenation, \ie it is a semigroup homomorphism:
-%% with respect to concatenation \km{It commutes with the operation of concatenation. This is the definition of (semigroup) homomorphism. I am not sure that linearity is an appropriate term.}
-%% , \ie satisfying the following equation:
-%
-\[
-\forall x, x', s, f(x \cdot x', s) = f(x, s) \cdot f(x', s)
-\]
-%
-%% This is a nice characterization, which I think is true in most
-%% settings. There is a caveat though: You can have stateless programs
-%% that emit output when they start (before they consume the stream)
-%% and/or when they consume the EOF symbol. In this case I am not sure
-%% the equation above holds.
-%
-\noindent
-This means that applying the command $f$ to a concatenation of two
-inputs $x, x'$ produces the same output as applying $f$ to each input
-$x, x'$ separately, and concatenating the outputs.
-
-Similarly, some pure commands (such as \ttt{sort} and \ttt{wc}) can be
-parallelized using divide-and-conquer techniques. More formally, these
-pure commands $f$ can be implemented as a combination of a function map
-$m$ and an associative aggregate $agg$, satisfying the following equation:
-%
-\[
-\forall x, x', s, f(x \cdot x', s) = agg(m(x, s), m(x', s), s)
-\]
-%
-\noindent
-This means that we get the same output by applying $f$ to a
-  concatenation of two inputs $x, x'$ as by applying the aggregation function
-  $agg$ to the outputs produced by applying map $m$ to each of $x$ and $x'$.
-%% \km{We had also discussed here the case of
-%%   operators that can be described as associative and commutative
-%%   functions on the data (e.g., sum). In this case the parallelization
-%%   has the flavor of data parallelism: one can split the input stream
-%%   in an arbitrary way (e.g., randomly or using some other
-%%   load-balancing scheme) and merge the outputs. For the more broad
-%%   class of associative operations (e.g., sort), the formulation of
-%%   this paragraph makes perfect sense. I just want to mention that this
-%%   feels a lot more like \textbf{batching} than map-reduce-like data
-%%   parallelization.} \kk{Thanks, I think you are right. Batching is a
-%%   better name for it. In the end we didn't handle
-%%   commutative-associative function (that are more like map-reduce) and
-%%   so we changes the name to map-fold to not imply commutativity.}
-
-\heading{Parallelization Transformations}
-%
-Based on these equations, we can define a node parallelization
-  transformation $T$ on a node $v \in$ \sta that is preceded by a
-concatenation, \ie the command \ttt{cat}, of $n$ input streams and is
-followed by a node $v'$
-(Fig.~\ref{fig:parallelization-transformation}). $T$
-replaces $v$ with $n$ new nodes, routing each of the $n$ input streams
-to one of them, and commutes the \ttt{cat} node after them to
-concatenate their outputs and transfer them to $v'$. Since each
-incoming edge represents a stream of data $x_i : D^*$, and the only
-behavior of a DFG is its output, this optimization $v(x_1 \cdot x_2
-\cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$
-can be shown to preserve the behavior of the graph.
-
-$T$ can be extended straightforwardly to nodes $v \in$ \pur,
-implemented by a map-aggregate pair $(m, agg)$ as $ v(x_1 \cdot x_2
-\cdots x_n, s) \Rightarrow agg(m(x_1, s), m(x_2, s), \ldots m(x_n,
-s), s)$.
-As long as the pair $(m, agg)$ meets the three invariants outlined earlier~\sx{ext}, 
-$T$ can be shown to be behavior-preserving.
-% \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.
-
-\heading{Auxiliary Transformations}
-%
-\sys also performs a set of auxiliary transformations $t_{1-3}$ that are
-depicted in \Cref{fig:auxiliary-transformations}. If a node has many
-inputs, $t_1$ concatenates these inputs by inserting a \ttt{cat} node
-to enable the parallelization transformations. In cases where a
-parallelizable node has one input and is not preceded by a
-concatenation, $t_2$ inserts a \ttt{cat} node that is preceded by its
-inverse \ttt{split}, so that the concatenation can be commuted with
-the node. Transformation $t_3$ inserts a relay node that
-performs the identity transformation. Relay nodes can be useful for
-monitoring and debugging, as well as for performance
-improvements~\sx{optimizer}.
-The additional relay nodes inserted by the auxiliary transformations affect the resulting % (task) 
-  parallelism.
-
-\subsection{Backend: From DFGs to a Parallel Shell Script}
-\label{backend}
-
-After % translating the dataflow regions of the input script to DFGs and 
-  applying transformations~\sx{ir:transformations}, \sys translates all DFGs back into a shell script.
-Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
-A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning up when the script aborts.
-% \kk{What does this "trap" refer to?}
-% The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
-% \kk{I removed this, since we describe them just below.}
-% \nv{Eurosys: Exactly, I had the last sentence as a segway connecting the two sections. There is not connection in the first paragraph of 5!}
-
-%% The first two
-%% \nv{Name them?}  enable the parallelization transformations by
-%% inserting \ttt{cat} nodes,\footnote{\nv{This is minor, but for the
-%%     model maybe name them concatenation nodes or ++? rather than
-%%     \ttt{cat}}} either by using them to concatenate many inputs of one
-%% command, or by first adding a \ttt{split} node that splits the input
-%% in batches. The final one inserts a relay that performs the identity
-%% transformation on its input.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{\detokenize{./figs/dish_aux.pdf}}
-\caption{
-  \textbf{Auxiliary transformations.}
-  These augment the DFG with \ttiny{cat}, \ttiny{split}, and \ttiny{relay} nodes.
-}
-\vspace{-10pt}
-\label{fig:auxiliary-transformations}
-\end{figure}
-
-%% \kk{Remember to mention the assumptions that need to hold for the
-%%   graph transformations to be valid in the Command categories
-%%   section. Commands must be deterministic, they must not do any other
-%%   side effect (such as writing to other files, sending signals,
-%%   etc). However, these assumptions must already be checked when the
-%%   developer designates the categories. Commands are assumed to always
-%%   terminate gracefully. \sys changes the termination behaviour of
-%%   commands as a sequential command that would exit prematurely might
-%%   not exit completely if parallelized.}
-
-%% \begin{definition}
-%% Given a DFG $G = (V, E, O)$, where $V$ is a set of nodes
-%% representing commands, $E$ is a set of edges representing files, and
-%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
-%% edges. $v_{out}$ represents a concatenation of all the outputs of the
-%% DFG. We represent the total order for a node $v$ as
-%% $<_v$. Given a node a node $v$
-%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
-%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
-%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
-%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
-%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
-%% \end{definition}
-
-%% \begin{lemma}
-%% tofill
-%% \end{lemma}
-
-\section{Runtime}
-% FIXME two labels here
-\label{impl}
-\label{optimizer}
-
-%% \kk{Maybe guide all of this by showing an example of an output shell
-%%   script?}
-
-This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom runtime primitives.
-%% it instantiates the \todo{custom} split and relay nodes as highly
-%% optimized commands to improve In this section we describe how \sys
-%% achieves that, and several technical challenges that it addresses.
-
-%% Given a dataflow graph representing a shell program, the \sys
-%% optimizer outputs an optimized dataflow graph in which data
-%% parallelism opportunities have been exploited.  More concretely, the
-%% optimizer starts by splitting input files in several chunks, given the
-%% desirable output dataflow graph width. It then starts from the source
-%% nodes of the graph, iteratively performing the graph transformations
-%% that were described in \Cref{ir:transformations}.
-%% %% If the incoming edges of the node are less than the desirable graph
-%% %% width, it performs edge splits, and then it performs maximal node
-%% %% splits.
-%% The final graph can then be implemented by spawning a process for each
-%% node, and redirecting the inputs and outputs according to the graph
-%% edges. In this section, we describe several technical challenges that
-%% we addressed.
-
-%% \subsection{Mapping operators to nodes}
-
-%% \kk{I don't think we should mention this.}
-
-%% Explain the simple algorithm that we used to minimize intra node
-%% communication, that tries to map contiguous parts of the graph to the
-%% same node. We could do this by having an algorithm that minimizes the
-%% number of cuts or something.
-
-%% We don't try to make a proper planner, as there is a lot of work on
-%% operator placement etc. that we can borrow from.
-
-% \subsection{Planning Activators and Just-in-Time Planning}
-% 
-% \nv{Nikos working on this and next}
-% 
-% % \kk{This whole section for just in itme planning, is also relevant for
-% %   the front end. If the front end is executed just in time, it would
-% %   have more information, and maybe it could make a sound analysis.}
-% 
-% Important point: Shell is extremely dynamic. Because of this, a
-% distributor cannot decide how to distribute a subexpression
-% statically, by just reading the script, (as in many other systems like
-% MapReduce, Spark, etc), since most of the information is not there
-% before the script starts executing. Environment variables,
-% unexpanded(unevaluated) strings \kk{Make sure that terminology is
-%   consistent with Greenberg}, could all contain information that is
-% valuable to make the distribution plan \kk{give an example}.
-% 
-% 
-% Future work on this: calling the shell shtepper to only partially
-% evaluate strings, just expand arguments, and not do any significant
-% computation. Even better, we could be calling the shtepper just after
-% having parsed the ast, and decide on maximal distributable subtrees as
-% late as possible in the process. This would be awesome.
-%
-
-%% \subsection{Implementation Nuggets}
-
-\heading{Overcoming Laziness}
-% \kk{eurosys: Maybe we can call this section pull vs push, and rephrase the terminology to explain the problem as "execution is bounded by the speed in which commands pull their inputs, leading to underutilization".
-Using eager we get rid of this bound prefetch the inputs of the next command.}
-The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
-% This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
-%% support concurrently executing many more commands than the
-%% processing threads of the underlying system,
-%% by scheduling them concurrently, allowing them to
-%% have progress one at a time.
-Such laziness leads to CPU underutilization, as commands are often blocked when their consumers are not requesting any input.
-Consider the following fragment:
-
-% \begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
-%  mkfifo t1 t2
-%  grep "foo" in1 > t1 & 
-%  grep "foo" in2 > t2 &
-%  cat t1 t2
-% \end{lstlisting}
-\begin{minted}[fontsize=\footnotesize]{bash}
- mkfifo t1 t2
- grep "foo" f1 > t1 & grep "foo" f2 > t2 & cat t1 t2
-\end{minted}
-
-\noindent
-The \ttt{cat} command will consume input from \ttt{t2} only after it completes reading from \ttt{t1}.
-As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes (Fig. \ref{fig:eager}a). % processing.
-
-% replacing FIFOs with files is
-To solve this, one might be tempted to replace FIFOs with files, a central \unix abstraction, simulating pipes of arbitrary buffering (Fig. \ref{fig:eager}b).
-Aside from severe performance implications, naive replacement can lead to subtle race conditions as a consumer might reach {\sc EOF} before a producer.
-Alternatively, consumers could wait for producers to complete before opening the file for reading (Fig. \ref{fig:eager}c);
-  however, this would insert artificial barriers impeding task-based parallelism and wasting disk resources---that is, this allows for data parallelism at the expense of task parallelism.
-% \unix does not provide an easy way out.
-
-To address this challenge,
-% Instead of addressing this issue for each command separately, % nv: how could this be possible?
-  \sys inserts and instantiates eager \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}d). % nv: not sure what critical meant
-These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push data to the output stream, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.
-% nv: Removed below because I don't think it adds much
-% We compare \sys's eager implementation with the alternatives from \cref{fig:eager} in \Cref{ours}.
-In \sys's evaluation~\sx{eval}, these primitives have the names presented in Fig.~\ref{fig:eager}.
-
-%% \TODO{Problem: Cat (and other commands that have multiple inputs) is
-%%   not eager. Illustration: Show a parallel pipeline having a cat that
-%%   gathers input somewhere (and a sort merge maybe). The way pipes
-%%   work, this graph will essentially evaluate almost completely
-%%   sequentially. Maybe go through the steps that cat asks for the first
-%%   line of its first input until ithe first input is done and then goes
-%%   to the second. This leads to processor under-utilization. Solution:
-%%   We add eager nodes that eagerly consume all their input (even if
-%%   their output does not ask for it). Eager nodes are themselves part
-%%   of the dataflow graph (and model) and essentially perform an
-%%   identity function. This makes all processors as utilized as possible
-%%   and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
-%%   light) tasks it has some overhead. We evaluate eager in section X
-%%   (We should also say something about a CPU light task). A good thing
-%%   is that they work for all possible commands with almost no overhead,
-%%   without having to make each specific command eager.}
-
-\begin{figure}[t]
-\centering
-% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.42\textwidth]{\detokenize{./figs/pash_eager.pdf}}
-\caption{
-  \textbf{Eager primitive.}
-  Addressing intermediary laziness is challenging:
-    (a) FIFOs are blocking;
-    (b) files alone introduce race conditions between producer/consumer;
-    (c) files + \ttiny{wait} inhibit task parallelism.
-Eager \ttiny{relay} nodes (d) address the challenge while remaining within the \sys model.
-% Describe. \TODO{Combine this figure with the above code maybe.}
-}
-\vspace{-15pt}
-\label{fig:eager}
-\end{figure}
-
-\heading{Splitting Challenges}
-\sys's optimizer inserts \ttt{split} nodes to expose parallelism when parallelizable nodes only have one input~\sx{ir:transformations}.
-For \ttt{split} to be effective, it needs to disperse its input uniformly across its outputs.
-This requires the input size to be known beforehand, which is not always the case.
-To address this, \sys provides a \ttt{split} implementations that first
-consumes its complete input, counts its lines, 
-and then splits it uniformly across the desired number of outputs. 
-% This \ttt{split} can
-% be used with any commands that add or remove lines in their output streams.
-  % (i) a general one that can be used with commands that could add or remove lines, and
-  % (ii) an optimized one that can be used if the input size is known before its execution.
-%% The \sys \todo{backend} support two split node implementations, to
-%% satisfy the inherent tension between the split efficiency and
-%% task-based parallelism.
-% In the latter case, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.
-\sys also inserts eager \ttt{relay} nodes after all \ttt{split} outputs (except the last one) to address laziness concerns.
-
-%% \TODO{Talk about eagerness in split and how we needed to add eager
-%%   nodes to n-1 outputs.}
-
-%% \noindent
-%% \TODO{Problem: Pipeline width shrinks after pure commands. Solution:
-%%   Split node that resplits its input to multiple files in sequence (to
-%%   preserve the model semantics).}
-
-%% Old and obsolete
-
-%% \noindent
-%% It is common in the dataflow graph for a node to have many incoming
-%% edges or many outgoing edges (when all but the last outgoing edges
-%% are bounded). However some nodes, might only read their input from one
-%% source (e.g. stdin). In this case, the distributed implementation
-%% needs to concatenate input files (or conversely split output
-%% files).
-%
-%% \tr{Show an example graph! of cat | tr | sort. The command in the
-%%   middle must only take input from one file (maybe just stdin).}
-%
-%% We address this issue using primitive shell constructs for file
-%% manipulation. More concretely a merger can be implemented by means of a simple
-%% \ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
-%% output file \ttt{out}, a merger can be implemented as:
-
-%% \begin{lstlisting}[language=sh, numbers=none]
-%%  cat $in1 $in2 ... > $out
-%% \end{lstlisting}
-
-%% \noindent
-%% On the other hand, given an input file \ttt{in} that is a FIFO pipe,
-%% and two output files \ttt{out1, out2} the first of which is bounded
-%% to \ttt{|N} lines, a 2-splitter can be implemented as:
-
-%% \begin{lstlisting}[language=sh, numbers=none]
-%%   tee >(head -n $N > $temp1;
-%%         dd of=/dev/null > /dev/null 2>&1 &
-%%         cat $temp1 > $out1) |
-%%        (tail -n +${N+1} > $out2;
-%%         dd of=/dev/null > /dev/null 2>&1)
-%% \end{lstlisting}
- %% head -n $N $in > $out1 ; cat $in > $out2
-
-%% \noindent
-%% Using the 2-splitter as a building block, a pipe can be split to
-%% arbitrarily many different pipes.
-
-\begin{table*}[t]
-\center
-\footnotesize
-% \setlength\tabcolsep{3pt}
-\caption{
-  \footnotesize{
-    \textbf{Summary of shell one-liners}.
-    % Small pipelines (\ie ones usually developed on-the-fly) drawn from various sources and applied to large datasets.
-    % Names correspond to the speedup plots (Fig.~\ref{fig:microbenchmarks}).
-    Structure summarizes the different classes of commands used in the script.
-    Input and seq. time report on the input size fed to the script and the timing of its sequential execution.
-    Nodes and compile time report on \sys's resulting DFG size (which is equal to the number of resulting processes and includes \ttiny{agg}regators, \ttiny{eager}, and \ttiny{split} nodes) and compilation time for two indicative \ttiny{--width}s.
-    % Highlights notes a few key features of each script.
-  }
-}
-\input{microbenchmarks-table}
-\label{tab:eval}
-\end{table*}
-
-\heading{Dangling FIFOs and Zombie Producers}
-%
-%% \TODO{(This problem is a nice followup for the one above) Problem:
-%%   Dataflow graph termination. Illustration: Show the example of head
-%%   deadlock and explain that this is inherent behaviour in the shell.
-%%   (Solution: emptying pipelines or terminating processes) In order for
-%%   things to terminate gracefully, dish wraps all commands in order to
-%%   always consume all of their input by sending it down to
-%%   /dev/null. Warning: This means that all commands in the pipeline
-%%   must terminate gracefully if all of their output is consumed (if a
-%%   process dies unexpectedly, noone will consume its input. We leave
-%%   addressing that and other monitoring questions for future work). We
-%%   evaluate the overhead of emptying all outputs in section X.}
-%
-% \nv{I think all of these titles should be the problem/challenge, not the solution}
-% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
-Under normal operation, a command exits after it has produced and sent all its results to its output channel.
-If the channel is a pipe and its reader exits early, the command is notified to stop writing early.
-In \unix, this is achieved by an out-of-band error mechanism: the
-operating system delivers a \ttt{PIPE} signal to the producer,
-notifying it that the pipe's consumer has exited.  This is different
-from errors for other system calls
-%% return an error
-and unusual compared to non-\unix systems\footnote{For example,
-  Windows indicates errors for \ttiny{WriteFile} using its return
-  code---similar to \ttiny{DeleteFile} and other Win32 functions.}
-primarily because pipes and pipelines are at the heart of \unix.
-% write errors are communicated differently 
-Unfortunately though, if a pipe has not been opened for writing yet,
-% ---possible for named pipes---
-  \unix cannot signal this condition.
-% \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
-% \nv{TODO! Will think about this later}
-Consider the following script:
-% emitted code from \sys's earlier versions:
-%% \kk{Proposed alternative: ``Consider the following shell script:''}
-% \begin{lstlisting}[language=sh, numbers=none]
-%   mkfifo fifo1 fifo2
-%   cat file-chunk1.txt > fifo1 &
-%   cat file-chunk2.txt > fifo2 &
-%   cat fifo1 fifo2 | head -n 1 & wait
-% \end{lstlisting}
-\begin{minted}[fontsize=\footnotesize]{bash}
-  mkfifo fifo1 fifo2
-  cat in1 > fifo1 & cat in2 > fifo2 &
-  cat fifo1 fifo2 | head -n 1 & wait
-\end{minted}
-\noindent
-In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
-As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
-This, in turn, leaves it unable to make progress, as it is both blocked and unaware of its consumer exiting.
-Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
-
-To solve this problem, \sys emits cleanup logic that operates from the
-end of the pipeline and towards its start.  The emitted code first
-gathers the IDs of the output processes and passes them as parameters
-to \ttt{wait}; this causes \ttt{wait} to block only on the output
-producers of the dataflow graph.  Right after \ttt{wait}, \sys inserts
-a routine that delivers \ttt{PIPE} signals to any remaining processes
-upstream.
-%% This epilogue is repeated for every \todo{merge/join} point
-%% in the transformed pipeline \kk{I don't understand the final sentence}.
-% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"
-
-%% \kk{We have to remember to say that for this to work, all nodes in the
-%%   graph must always open their outputs! If not, then this technique
-%%   doesn't work, and we have to monitor if any node dies to do
-%%   something with it. At the moment we ensure that by doing that in our
-%%   nodes, and by assuming that all nodes that we handle do that. This
-%%   is not really a problem, because all of our commands have one
-%%   output, which if they dont open this would cause problems in the
-%%   sequential script too.}
-
-% not receiving an out-of-band signal about that.
-% never terminating, but 
-% never terminate while the entire pipeline is waiting for 
-% \ttt{\& echo \$!} and \ttt{kill -PIPE}
-
-% The generated program could also (i) have signal handlers (traps) for propagating signals
-% down the process tree; (ii) incorporate (a simple) cleanup logic for not leaving dangling processes and
-% intermediary files; (iii) generate a correct exit code, depending on its
-% execution. All of this can be easily generated by the compiler.
-
-\heading{Aggregator Implementations}
-Commands in \tpur can be parallelized using a \emph{map} and an \emph{aggregate} stage~\sx{parallelizability}.
-\sys implements \emph{aggregate} for several commands in \tpur to enable parallelization.
-A few interesting examples are \emph{aggregate} functions for
-  (i) \ttt{sort}, which amounts to the merge phase of a merge-sort (and on GNU systems is implemented as \ttt{sort -m}),
-  (ii) \ttt{uniq} and \ttt{uniq -c}, which need to check conditions at the boundary of their input streams,
-  (iii) \ttt{tac}, which consumes stream descriptors in reverse order, and 
-  (iv) \ttt{wc}, which adds inputs with an arbitrary number of elements (\eg \ttt{wc -lw} or \ttt{wc -lwc} \etc).
-The \emph{aggregate} functions iterate over the provided stream descriptors, \ie they work with more than two inputs, and apply pure functions at the boundaries of input streams (with the exception of \ttt{sort} that has to interleave inputs).
-% An important concern has been to maintain whitespace separation while
-
-%% The optimizer also applies custom transformations on several commands in \pur
-%% to further improve their performance. Two notable examples are the
-%% transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
-%% that have more than one input are tranformed to \ttt{sort} commands on
-%% each input, followed by a \ttt{sort -m}, to produce the final sorted
-%% result. The transformation on occurences of \ttt{wc} that have more
-%% than input \ttt{i1, t2} is illustrated below as a shell script:
-%% % \nv{do not understand i1, t2}
-
-%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
-%%   paste -d '+' \
-%%    <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
-%%    <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
-%%    bc | tr -s '\n' ' ' |
-%%    sed 's/^/   /' | sed 's/$/\ /'
-%% \end{lstlisting}
-
-%% The above script takes the outputs produced by two \ttt{wc} commands
-%% and aggregates them accordingly.
-
-% OSDIFIX \nv{Should we add a small subsection about implementation details?}
+\section{Example \& Overview}
+\label{overview}
 
@@
 
-\begin{figure}[t]
-\centering
-% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
-\includegraphics[width=0.37\textwidth]{\detokenize{./figs/pash_lattice.pdf}}
-\caption{
-  \textbf{Runtime setup lattice.}
-  Parallel No Eager and Blocking Eager improve over sequential, but are not directly comparable.
-  Eager adds \sys's optimized \ttiny{eager} relay, and \sys uses all primitives in \S\ref{optimizer} (Fig.~\ref{fig:microbenchmarks}).
-}
-\vspace{-15pt}
-\label{fig:lattice}
-\end{figure}
-
-%> \TODO{Remember to mention that deathstar has 128 threads but 64
-%>   cores. So it is reasonable for the speedup to not go further in cpu
-%>   heavy pipelines.}
-%> 
-%> \TODO{Should we mention that we empirically show that pash produces
-%>   correct code?}
-%> 
-% FIXME
-% \kk{We should make sure to measure the time that it takes for the
-%   whole process to run (parsing, translating, optimizing, planning)
-%   together with the scripts. This will probably be negligible, but we
-%   still have to mention it. }
-% 
-% \kk{Also, we have to have one-two handcrafted examples that have more
-%   than one pipeline, to show the execution time end to end, as this
-%   cannot be shown with the one-pipeline examples since they only have
-%   one graph, so no back and forth between shell and dish.}
-% 
-% \kk{Furthermore it would be great to have one distributed example that
-%   runs in more than one node (if we have time)}
-% 
-% \kk{Also we have to make sure to measure with and without the final
-%   cat of output file, so that we show that this is also negligible.}
-% 
-% \kk{We have to make sure that we repeat that the results are the same
-%   in all tests that we run. Corectness, blah blah ...}
-% 
-% \kk{We should have at least some pipelines that have a bad pure
-%   command in them.}
-
-%>\kk{Maximum scaleup for both diff and set-diff is around half because
-%>  they already have two sorts running in parallel (as diff, and comm
-%>  take two inputs).}
-
-%% \TODO{Explain the plots with eager/no-eager as follows. Say that we
-%%   initially decided to only make an eager version of cat (to allow for
-%%   parallelism). However, we saw that results are still not good, so we
-%%   resorted to making a proper generic eager node that can be put
-%%   anywhere in the graph and that eagerly consumes its input. The above
-%%   results show the improvement with it. Rerun them with eager only
-%%   before cat in the no eager cases so that the minimal-grep example is
-%%   the same with and without our eager node.}
-
-%> \TODO{(Maybe) we need to mention that eager has disk usage overhead that is linear to the size of the pipeline and the size of the input.}
-
-This section reports on whether \sys can indeed offer performance benefits automatically and correctly using several scripts collected out from the wild along with a few micro-benchmarks for targeted comparisons.
-
-% Micro-bench\-marks~\sx{ours} are one-off, simple one-liners that take a few seconds to write;
-%   such pipelines are usually composed interactively to solve a task at hand, test a hypothesis, or ``smoke out bugs''~\cite{bentley1986literate}---but, with \sys, applied to very large data sets.
-% These pipelines are taken from real use cases~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked} and, precisely due to their small size, highlight stress on a handful of stages.
-% The complex macro-benchmark (\S\ref{macro1}) is designed to handle realistic workloads by today's standards.
-% It is about an order of magnitude larger than the micro-benchmarks, but its perceived lack of complexity is deceiving:
-%   as shown, if written in a conventional language it would correspond to a program on the order of hundreds of lines of code.
-
-\heading{Highlights}
-This paragraph highlights results for \ttt{width=16}, but \sys's evaluation reports on varying \ttt{width}s (2--64).
-% The vast majority (39/44) of scripts---all unmodified---see significant speedups (average over all is $6.7\times$, up to $61\times$).
-Overall, applying \sys to all 44 unmodified scripts accelerates 39 of them by 1.92--17.42$\times$; for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 0.99, 1.01$\times$).
-The total average speedup over all 44 benchmarks is $6.7\times$.
-\sys's runtime primitives offer significant benefits---for the 10 scripts that we measured with and without the runtime primitives they bump the average speedup from $5.9\times$ to $8.6\times$.
-% Overall, \sys accelerates 39/44 scripts---all unmodified---by 1.92--17.42$\times$; 
-  % for the rest, the parallel performance is comparable to the sequential (0.89, 0.91, 0.94, 099, 1.01$\times$).
-  % \kk{eurosys: I have an issue that I am not sure how to address.
-I would like to have the 61.1x here (which is due to 64 width)  but I don't see a good way of including it without sounding suspicious.}
-% For a subset of our benchmarks~\sx{ours}, we decouple the speedup due to \sys's DFG transformations alone from the one achieved when including its runtime primitives:
-  % the former averages $5.9\times$, jumping to $8.6\times$ due to the latter.
-% A different subset ($30$ benchmarks) sees average speedups of $6.02\times$~\sx{unix50}.
-% Two complex scripts with 15~\sx{macro1} and 34~\sx{macro2} commands written in multiple languages see speedups of $2.44\times$ and $12.7\times$.
-\sys significantly outperforms \ttt{sort --parallel}, a hand-tuned parallel implementation, and performs better than GNU \ttt{parallel}, which returns incorrect results if used without care~\sx{micro}.
-% Averages: [1.955223776839714, 3.478591807121036, 5.721328556478535, 8.604380509595906, 10.666921423438112, 13.246864076708423]
-% No Eager Averages: [1.6327834370824543, 2.5481043878162444, 3.8623841529643173, 5.931851620007484, 7.466566647963421, 9.35752032439642]
-% From the rest, only two see a slowdown, only because \sys's setup ($\sim1s$) is higher than their runtime ($\sim0.4s$).
-% \sys's optimized primitives offer benefits even against manually parallelized scripts. % (such as GNU sort, and GNU parallel).
-% \sys's compilation pass is negligible (avg. $<1s$). %  for non trivial program execution.
-% In two large cases, tasks such as data download, extraction, and preprocessing---often implemented using shell scripts and outside the focus of specialized parallelization frameworks---take comparable time to the execution of the main computation, and can be significantly accelerated by \sys.
-% In terms of manual effort,
-
-Using \sys's standard library of annotations for POSIX and GNU commands~\sx{parallelizability}, the vast majority of programs ($>40$, with $>200$ commands) require no effort to parallelize other than invoking \sys;
-  only 6 ($<3\%$) commands, outside this library, needed a single-record annotation~\sx{macro2}.
-
-In terms of correctness, \sys's results on multi-GB inputs are identical to the sequential ones.
-Scripts feature ample opportunities for breaking semantics~\sx{micro}, which \sys avoids.
-%% In one case, \sys avoids transforming a cycle (\ie a fragment whose input and output identifiers are the same).
-%in another, it preserves the semantics that another parallelization command breaks.
-
-% conservative program transformations will not attempt paral- lelizing program sections
-% In another comparison with \ttt{parallel}
-% and in another 
-% Does not break 
-% The effort to reap this benefits is minimal; 
-%  minimal annotations for \todo{custom} commands.
-
-%% transformation and planning phase take around 50--300ms, negligible ($<$2\%) for short-running pipelines and virtually non-existent for long-running ones.
-%% As \sys rewrites to highly-optimized shell primitives rather than using a managed language runtime, it always has a COST~\cite{mcsherryscalability} of 2.
-%% Finally, \sys preserves the productivity of the shell:
-%% the equivalent program for a small part of the macro-benchmark pipeline requires over a 150 lines of code in combined Java and Bash to run distributed on top of Hadoop.
-% \kk{What is that about the COST?}
-  %% in another, the program expressed by the pipeline was the subject of a semester project in a graduate course on distributed systems~\cite{blinded} where student implementations ranged between 1--3\textbf{K} lines of code.
-% In terms of program size, the Hadoop program amounts to 137 Java LoC as opposed to a pipeline of \todo{5} shell commands for the same computation. \kk{Is it worth leaving this sentence in here?}
-%% kk: I think now that we removed Hadoop, it is not so important. In
-%% the end of the day, we don't want to convince people that they
-%% should use shell for all their needs.
-
-\heading{Setup}
-\sys was run on 512GB of memory and 64 physical $\times$ 2.1GHz Intel Xeon E5-2683 cores, Debian 4.9.144-3.1, GNU Coreutils 8.30-3, GNU Bash 5.0.3(1), and Python 3.7.3---without any special configuration in hardware or software.
-%% and OCaml 4.05.0 %% this was here for the parser that we don't mention atm
-Except as otherwise noted,
-  (i) all pipelines are set to (initially) read from and (finally) write to the file-system,
-  (ii) \ttt{curl} fetches data from a different physical host on the same network connected by 1Gbps links.
-% A more detailed description of the setup and experiments, including reproducible scripts, is included in the accompanying (anonymized) online repository.
-
-\heading{Parallelism}
-The level of parallelism is configured using \sys's \ttt{--width=W}, a flag that determines the number of times that \sys applies its transformations and the resulting width of the data-parallel DFG.
-% The value  of \ttt{N} is shown as the x-axis of Fig.\ref{fig:microbenchmarks}. 
-\sys does not control a script's initial parallelism (\eg a command could spawn 10 processes), and thus the resulting scripts often reach maximum parallelization benefits with a value of \ttt{W} smaller than the physical cores of our hardware setup (in our case 64).
-
-% \heading{A Note on the Parallelism Factor}
-% %
-% The discussion of the results often mentions the parallelism factor, which refers to the \sys \emph{parallelism configuration} rather than the resulting parallelization.
-In fact, for small parallelism configurations \sys often achieves higher parallelization due to the creation of multiple processes---aggregators, splitters, relays \etc
-%% \nv{
-
-%% This is not necessarily the resulting parallelism achieved by \sys, which typically is above the level configured by the user.
-%% This is because for each process (node in the graph), \sys creates multiple processes---aggregators, splitters, relays \etc
-%% Some times the results give the impression of superlinear parallelism, but this has a simple explanation.
-%% }
-
-\subsection{Common \unix One-liners}
-\label{ours}
-
-We first evaluate \sys on a set of popular, common, and classic \unix pipeline patterns~\cite{bentley1985spelling, bentley1986literate, taylor2004wicked}.
-The goal is to evaluate performance benefits due to \sys's  
-  (i) DFG transformations alone, including how \ttt{--width} affects speedup, and
-  (ii) runtime primitives, showing results for several configurations of the runtime lattice (Fig.~\ref{fig:lattice}).
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{\detokenize{./figs/tiling_throughput_scaleup.pdf}}
-    \vspace{-20pt}
-    \caption{
-      \textbf{\sys's speedup for \ttiny{width=}2--64.}
-      Different configurations per benchmark:
-      (i) PaSh: the complete implementation with \ttiny{eager} and \ttiny{split} enabled,
-      % (ii) Par+B.Split: \ttt{eager} and input-aware \ttt{split} enabled,
-      (ii) PaSh w/o split: \ttiny{eager} enabled (no \ttiny{split}),
-      (iii) Blocking Eager: only blocking \ttiny{eager} enabled (no \ttiny{split}),
-      (iv) No Eager: both \ttiny{eager} and \ttiny{split} disabled.
-  For some pairs of configurations, \sys produces identical parallel scripts and thus only one is shown.
-  % \kk{FUTURE: Run and show all plots (with and without split for all experiments.)}
-}
-    \vspace{-14pt}
-    \label{fig:microbenchmarks}
-\end{figure*}
-
-\heading{Programs}
-%>\kk{How can we include a result for grep-light?  This is the IO
-%>  intensive CPU-light grep. \sys has no speedup but also no slowdown
-%>  (even if it IO heavy). Should we just mention that in text in
-%>  passing?}
-%>
-%>\kk{Remember to mention that for spell and bigrams we dont show other
-%>  parallel because they are all close to 1 (no speedup).}
-Tab.~\ref{tab:eval} summarizes the first collection of programs.
-%  shown in Fig.~\ref{fig:microbenchmarks}, which in turn reports on their performance characteristics.
-NFA-Regex is centered around an expensive NFA-based backtracking expression and all of its commands are in \sta.
-% and the latter being an IO-intensive script.
-% a DFA-based regular expression matching phase;
- % while the \unix \ttt{grep} command defaults to a Thompson NFA, this particular expression makes use of DFA-based backtracking patterns resulting in high runtime overhead.
-Sort is a short script centered around a \pur command.
-Wf and Top-n are based on McIlroy's classic word-counting program~\cite{bentley1986literate};
-  they use sorting, rather than tabulation, to identify high-frequency terms in a corpus.
-% \tr{The \ttt{bi-gram} pipeline calculates n-grams of base two;
-%   it makes clever use of \ttt{tail} to shift a stream by one word and \ttt{paste} to fuse (zip) two streams together.}
-Spell, based on the original \ttt{spell} developed by Johnson~\cite{bentley1985spelling}, is another \unix classic:
-  after some preprocessing, it makes clever use of \ttt{comm} to report words not in a dictionary.
-Shortest-scripts extracts the 15 shortest scripts in the user's \ttt{PATH}, using the \ttt{file} utility and a higher-order \ttt{wc} via \ttt{xargs}~\cite[pg. 7]{taylor2004wicked}.
-Diff and Set-diff compare streams via a \ttt{diff} (in \npu, non-parallelizable) and \ttt{comm} (in \pur), respectively.
-Sort-sort uses consecutive \pur commands without interleaving them with commands that condense their input size (\eg \ttt{uniq}).
-Finally, Bi-grams replicates and shifts a stream by one entry to calculate bigrams.
-% , but Bi-gram-opt's groups 5 commands into one \pur command that has a more efficient $map$ and $aggregate$ implementation.
-
-\heading{Results}
-% TODO \nv{Need to change terminology in plots---files should be (attempted) parallelism, name benchmark, cat should be merge.}
-Fig.~\ref{fig:microbenchmarks} presents \sys's speedup as a function of \ttt{width=}2--64.
-% (Bi-grams-opt is shown as the \ttt{opt} line in Fig.~\ref{fig:microbenchmarks}'s Bi-grams.)
-Average speedups of the optimized \sys, \ie with \ttt{eager} and \ttt{split} enabled, for \ttt{width=}\{2, 4, 8, 16, 32, 64\} are \{1.97, 3.5, 5.78, 8.83, 10.96, 13.47\}$\times$, respectively.
-For \emph{No-Eager}, \ie \sys' transformations without its runtime support, speedups drop to 1.63, 2.54, 3.86, 5.93, 7.46, 9.35$\times$.
-% and the geometric means are 1.92, 3.39, 5.48, 7.94, 8.86, 8.51.}
-% \kk{For the one-liners where split doesn't provide any benefit (or overhead) we don't show the PaSh w/o split line.}
-% we experiment with a $200\times$ level of parallelism to see potential overheads
-
-Plots do not include lines for configurations that lead to identical parallel programs.
-There are two types of such cases. %  (that appear in \todo{five} plots).
-In the first, the \emph{Pash} (blue) and \emph{Pash-w/o-split} (red, hidden) lines are identical for scripts that \sys does not add \ttt{split} since the width of the DFG is constant;
-%  when \sys adds no \ttt{split} because the DFG width remains constant for the whole program;
-  conversely, when both lines are shown (\eg Spell, Bi-grams, and Sort), \sys has added \ttt{split}s due to changes in the DFG width (e.g. due to a \npu command).
-In the second type, \emph{Pash w/o Split} (red) is identical to \emph{No-Eager} (green, hidden) and \emph{Blocking Eager} (orange, hidden) because the input script features a command in \pur or \npu relatively early.
-This command requires an \ttt{agg}regator, whose output is of width 1, and beyond which non-\ttt{split} configurations are sequential (and see no speedup).
-% Conversely, all three lines exist when the width remains above 1 for the full program.
-% needing a \ttt{split} to regain width. 
-Finally, Tab.~\ref{tab:eval} shows that \sys's transformation time is negligible, and its COST~\cite{mcsherryscalability} is 2. %  for all benchmarks.
-% Filter and Bigrams-opt, not in Fig.~\ref{fig:microbenchmarks}, achieve speedups of 1.5--2.5$\times$ and 2--30$\times$.
-%   size grows due to the resulting complexity of coordinating among parallel executions---multiple FIFOs per pipeline stage, merging and splitting, encoding of divide-and-conquer synchronization \etc
-% The first number alone is interesting, as it captures \sys's output for the sequential script.
-% As there is no parallelism involved, it highlights the initial (bare minimum) cost corresponding to using \sys.
-% % this is the size of 
-% This change in size does not necessarily translate to observed runtime, as most of the size increase comes from the setup and teardown of FIFOs responsible for interprocess (IPC) communication.
-% FIFOs are simply pipes made explicit;
-%   the performance characteristics, including in-kernel buffering and synchronization mechanisms, are identical between the two.
-
-      % (vi) Opt. Parallel, an optimized version of \ttt{bigrams}~\cf{ours}.
-      % Only relevant configurations are shown---\eg \ttt{grep} and \ttt{sort} do not benefit from \ttt{split}; \ttt{spell} and \ttt{bi-grams} do not see benefits without \ttt{split}.
-      % TODO: WHY?
-
-\heading{Discussion}
-% TODO: Add
-% * \ttt{width}
-% * why blocking is less than no eager
-As expected, scripts with commands only in \sta see linear speedup.
-\sys's \ttt{split} benefits scripts with \tpur or \tnpu commands, without negative effects on the rest.
-Its \ttt{eager} primitive improves over Eager and Blocking-eager for all scripts.
-% Blocking eager: data parallelism, but no task parallelism; No eager: blocking FIFO does not allow pushing -> underutilize CPU
-\emph{No Eager} is usually faster than \emph{Blocking Eager} 
-% due to \ttt{agg}regators that read multiple input streams at the same time---\eg \ttt{sort -m}.
-% \kk{eurosys: while the above is technically true, it is a bit difficult to explain here, so I prefer the sentence below.}
-since it allows its producer and consumer to execute in parallel.
-Sort-sort illustrates the full spectrum of primitives:
-  (i) \ttt{eager} offers benefits despite the lack of \ttt{split} because it fully parallelizes the first \ttt{sort} 
-  % while maintaining task parallelism
-  (ii) \emph{Pash} gets full benefits because \ttt{split}ting allows parallelizing the second \ttt{sort} too.
-
-% \todo{No Eager} and \todo{Blocking Eager} depict configurations without \sys's \ttt{eager} optimization and without \ttt{split}.
-% As described in Fig.\ref{optimizer}, these eager implementations perform worse than \sys's optimized \ttt{eager} relay in all cases (except for a negligible difference in Shortest-Scripts).
-
-As described earlier, \sys often achieves the maximum possible speedup % \ie without unnecessary process spawns and context switches, 
-  for a \ttt{width} that is lower than the number of available cores---\ie \ttt{width=}16--32 for a 64-core system.
-This is also because \sys's runtime primitives spawn new processes---\eg Sort with \ttt{width=}8 spawns 37 processes:
-  8 \ttt{tr}, 8 \ttt{sort}, 7 \emph{aggregation}, and 14 \ttt{relay} processes.
-
-% \todo{standard vs non-standard Bigrams}
-
-% \kk{eurosys: Can we somehow turn this point around and make it seems good/fine.
-That here we report this parallelism, but in the real system, it tries to do a best effort guess of parallelism based on the size of the input script.}
-% \sys achieves near-linear speedup on low-parallelism configurations.
-% This is due to a high degree (deep trees) of task parallelism from \sys's multiple constructs---\eg \emph{aggregations} and \ttt{relay}s---evident in all plots that contain \pur stages.
-%% \kk{Ben's point: task parallelism is mentioned here for the first time. Since \sys exposes data parallelism as task parallelism, we might want to mention that somewhere earlier in the paper}
-% In effect, \sys hits the maximum number of parallel processes (and thus achieves optimal performance, \ie one without unnecessary process spawns and context switches) 
-% significantly earlier than the purported parallelism---$16-32\times$ for a 64-core system.
-% Asking for parallelism beyond that point degrades speedup, as overheads still increase but no further parallelism is available---as is the case, \eg  with Sort-sort for $32\times$ and $64\times$ which has 314 and 634 nodes, respectively.
-% 
-%  On the other hand opt-bigrams 64 has 255 nodes and the sys time stays
-%   the same so I think that it might have to do with the fact that
-%   constant costs rise a lot (since opt-bigrams 32 takes just 1
-%   minute).
-% \kk{Mention bigrams-opt result now that it is removed from the plot.}
-
-%> \kk{It is visible that the automatic split is slightly worse than the
-%>   manual one in most cases (because it doesn't utilize task-based
-%>   parallelism).}
-
-%> This is also the case if the input pipeline is very long and so there is already task-based parallelism even in the sequential execution (several of the Unix50 pipelines fall in this category).
-%> \kk{Hypothesis on a scalability upper bound. Dish gives a great
-%>   speedup if one step of the pipeline is what takes the most time. If
-%>   the pipeline is long with many computation heavy tasks, task based
-%>   parallelism gets you part of the way there, so we can't expect Dish
-%>   to give performance benefits equal to the amount of parallelism that
-%>   we give it.}
-
-\heading{Take-aways}
-\sys accelerates scripts by up to 60$\times$, depending on the commands involved.
-Its runtime constructs improve over the baseline speedup achieved by its transformations.
-
-\subsection{Unix50 from Bell Labs}
-\label{unix50}
-
-We now turn to a set of \unix pipelines found out in the wild.
-
-\heading{Programs}
-In a recent celebration of \unix's 50-year legacy, Bell Labs created 37 challenges~\cite{unix50} solvable by \unix pipelines.
-The problems were designed to highlight \unix's modular philosophy~\cite{mcilroy1978unix}.
-We found unofficial solutions to all-but-three problems on GitHub~\cite{unix50sol}, expressed as pipelines with 2--12 stages (avg.: 5.58).
-They make extensive use of standard commands under a variety of flags, and appear to be written by non-experts (contrary to \S\ref{ours}, they often use sub-optimal or non-\unix-y constructs).
-% ---stressing a patterns that we, \sys's authors, did not
-% necessarily anticipate \kk{I would change that to ``stressing a
-%   variety of patterns'' since it doesn't really matter if we
-%   anticipated them (especially since we can address any issues and
-%   then re-run them).} . 
-\sys executes each pipeline as-is, without any modification.
-  % if an obvious fix improves performance, we report the speedup of both versions. %  original and modified. %  pipeline.
-
-% The ability to compose larger programs in versatile ways from smaller utility programs was a hallmark of the \unix design, and the game's levels are designed to stress that.
-% These challenges were specifically designed to highlight the \unix's philosophy, its 
-% ability to compose larger programs in versatile ways from smaller utility programs summarized by McIllroy's Unix philosophy memo.
-
-\heading{Results}
-Fig.~\ref{fig:unix50-individual} shows the speedup (left) over the sequential runtime (right) for 31 pipelines, with \ttt{width=}16 and 10GB inputs.
-It does not include 3 pipelines that use \ttt{head} fairly early thereby finishing execution in under $0.1$ seconds.
-We refer to each pipeline using its x-axis index in Fig.~\ref{fig:unix50-individual}.
-% The script index refers to the challenge level, in both the plots and the discussion.
-% \kk{The above is not exactly true since we sorted them. If someone goes to take a look at the solution source they will be confused.}
-Average speedup is $6.02\times$, % median is $6.48\times$, 
-  and weighted average (with the absolute times as weights) is $5.75\times$.
- % the \todo{geometric mean is $1.08\times$},
-
-  % Mean: 6.028795730618471
-  % Median: 6.48837000168888
-  % Geometric Mean: 4.765602271688225
-  % Weighted Average: 5.752417060139914
-
-\heading{Discussion}
-Most pipelines see significant speedup, except \#25-30
-  % \todo{13, 24, 25, 26, 29, 30} 
-  that see no speedup because they contain general commands that \sys cannot parallelize without risking breakage---\eg \ttt{awk} and \ttt{sed} \ttt{-d}. % and \ttt{tr} \ttt{-d}.
-  % and \todo{2, 19, 31} that see a slowdown.
-% Note however, that some of the pipelines contain more general commands that cannot be parallelized since they were not written with that goal in mind;
-%   a common example is the use of \ttt{awk} for reordering or removing columns of the input.
-%% Some of them use general interpreters when they could be using a specialized equivalent tailored to the task at hand, the typical example being \ttt{awk} vs \ttt{cut}.
-A \unix expert would notice that some of them can be replaced with \unix-specific commands---\eg \ttt{awk "{print \$2, \$0}" | sort -nr}, used to sort on the second field can be replaced with a single \ttt{sort -nr -k 2} (\#26).
-  % \todo{13th} 
-The limited expressiveness of these commands can be exploited by \sys---in this specific case, achieving 8.1$\times$ speedup (\vs the original 1.01$\times$).
-%% , attempting only such obvious (to us, experts) replacements, \sys extracts further parallelism---\eg 13 sees a 8.1$\times$ speedup.
-% ---a common pattern we witnessed~\sx{macro1}.
-
-% The second group contains \ttt{head}, thus practically processing only one line (regardless of the full input) and completing within $10ms$. 
-% \sys's slowdown is due to constant setup costs of pipes and processes but execution still remains under $1s$.
-
-% illustrative example is pipeline 13, which is shown here:
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none]
-%   cat $IN | awk "{print \$2, \$0}" |
-%       sort -nr | cut -d ' ' -f 2
-% \end{lstlisting}
-% 
-% \kk{Rephrase that to sound more meaningful and surprising.}  Notice
-% that \ttt{awk} is just used to permute two of the columns of the input
-% to enable sorting. One could write this pipeline in a different way,
-% by directly sorting on the second column:
-% 
-% \begin{lstlisting}[language=sh, float=h, numbers=none]
-%   cat $IN6 | sort -nr -k 2 | cut -d ' ' -f 1
-% \end{lstlisting}
-
-For the rest, \sys's speedup is capped due to a combination of reasons:
-  (i) they contain pure commands that are parallelizable but don't scale linearly, such as \ttt{sort} 
-  (\#5, 6, 7, 8, 9, 19, 20, 21, 23, 24), 
-  % 24, 9, 7, 6, 8, 23, 20, 19, 21, 5
-  % \todo{(0, 1, 3, 15, 16, 18, 20, 27, 28, 33)},
-  (ii) they are deep pipelines that already exploit task parallelism
-  (\#4, 10, 11, 13, 15, 17, 19, 21, 22),
-  % 17, 4, 13, 15, 10, 22, 11, 19, 21,
-  % \todo{(7, 8, 9, 10, 11, 12, 14, 27, 28,)} 
-or 
-  (iii) they are not CPU-intensive, resulting in pronounced I/O and constant costs 
-  (\#3, 4, 11, 12, 14, 16, 17, 18, 22).
-  % 14, 3, 16, 17, 4, 22, 11, 12, 18
-  % \todo{(4, 5, 6, 7, 8, 12, 14, 22, 23)}.
-
-%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
-%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
-%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14, 22, 23
-%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
-
-%% Finally, an important point that is reinforced from this
-%% experiment is that \sys doesn't lead to slowdown, even in cases where
-%% parallelism is not possible.
-
-%% Breakdown of not ideal speedups in 8, 16 parallelism
-%% Head: 2, 19, 31
-%% Sort-bottleneck: 1, 3, 15, 16, 20, 33
-%% Non-parallelizable completely: 13 (awk-solved), 18 (awk), 24, 25, 26, 29, 30
-%% Cheap (non cpu intensive -- constant costs grow): 0, 4, 5, 6, 7, 8, 12, 14
-%% Long (task-based parallelism in sequential and context switches): 7, 8, 9, 10, 11, 12, 14, 21, 27, 28
-
-\heading{Take-aways} \sys accelerates unmodified pipelines found in the wild;
-  small tweaks can yield further improvements, showing that \sys-awareness and scripting expertise can improve results.
-Furthermore, \sys does not significantly decelerate non-parallelizable scripts.
-% \kk{eurosys: we cant say it doesnt decelarate since it does slow down some of them. If we dont like this sentence we can also delete it.}
-% non-trivial computations, even when parallelism is impossible.
-
-%% Very nice take-aways!
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{\detokenize{./figs/unix50_individual_speedups_16.pdf}}
-  \vspace{-15pt}
-  \caption{
-	  \textbf{Unix50 scripts.}
-    Speedup (left axis) over sequential execution (right axis) for Unix50 scripts.
-    Parallelism is 16$\times$ on 10GB of input data~\cf{unix50}.
-    Pipelines are sorted in descending speedup order. % \str
-    % \kk{FUTURE: Add speedup by primitives and transformations in bar-chart.}
-    % The solid line indicates the average speedup $\times5.49$ and the dotted line indicates 1.
-    % \nv{Idea: scripts that only take over 0.01s?}
-    % \kk{Now that we removed the small ones, sequential time does not add a lot to this plot. Maybe we can just show speedups?}
-  }
-  \vspace{-15pt}
-  \label{fig:unix50-individual}
-\end{figure}
-
-\subsection{Use Case: NOAA Weather Analysis}
-\label{macro1}
-
-% For a realistic workload, 
-We now turn our attention to Fig.~\ref{fig:example}'s script~\sx{bg}.
-
-\heading{Program}
-This program is inspired by Hadoop's Definitive Guide~\cite[\S2]{hadoop:15}, where it exemplifies a realistic analytics pipeline comprising 3 stages:
-  fetch NOAA data (shell), convert them to a Hadoop-friendly format (shell), and calculate the maximum temperature (Hadoop).
-While the book focuses only on the last stage, \sys parallelizes the entire pipeline.
-% Only the last one is the real focus of the book, % demonstrating the benefits of using distributed framework such as Hadoop.
-%   whereas we consider the entire pipeline. % ---from fetching data to outputting the maximum temperature.
-% The version used in the evaluation is slightly different than the one presented in Fig.~\ref{fig:example} (14 stages and totalling 240 characters), simplified only for clarity of exposition, in that it expresses the \ttt{for} loop as a first stage followed by \ttt{xargs}.
-
-\heading{Results}
-The complete pipeline executes in 44m2s for five years (82GB) of data.
-\sys with \ttt{width=}16 leads to 2.52$\times$ speedup, with different phases seeing different benefits:
-  2.04$\times$ (vs. 33m58s) for all the pre-processing ($75\%$ of the total running time) and 
-  12.31$\times$ speedup (vs. 10m4s) for computing the max.
-
-%> It is worth reporting Hadoop's (3.2.1, Oracle JDK1.8.0\_255) execution time for the curious reader (see below for fairness), which takes \todo{6m22s} (sequential) and \todo{4m30.554s} ($\times$\todo{XX}-parallel).
-%> This amounts to \todo{XX}$\times$ speedup versus \todo{XX}$\times$ for \sys's equivalent (part 4).
-%> % ---this includes downloading and uncompressing the data files which, as described earlier, is not expressible in the Hadoop equivalent.
-
-% /home/nikos/hadoop-3.2.1/bin/hdfs dfs -put /home/nikos/dish/scripts/max-temp 77.24s user 50.79s system 185% cpu 1:09.14 total
-% hadoop jar $HADOOP_HOME/share/hadoop/tools/lib/hadoop-streaming-*.jar -input 719.26s user 63.88s system 155% cpu 8:22.47 total
-
-\heading{Discussion}
-% 
-The speedup of the preprocessing phase of the pipeline is bound by the network and IO costs since \ttt{curl} downloads 82GB of data.
-However, in the processing phase (which is CPU-bound) the speedup is $12.31\times$, much higher than what would be achieved by parallelizing per loop iteration.
-Similar to Unix50~\sx{unix50}, we found that large pipelines enable significant freedom in terms of expressiveness.
-% A few stages of the original script were expressed in a single \ttt{awk};
-% unfortunately, \ttt{awk} is too general to have a single meaningful \todo{parallelizability} signature.
-  % a simpler (but longer) pipeline that leverages \unix built-ins turns out to be trivially parallelizable.
-
-%> In terms of execution fairness, both systems are message-passing rather than shared-memory, and Hadoop's setup is atypical to improve fairness:
-%>   it operates on non-HDFS file-system and has replication disabled.
-%> Additionally, the Java program is written as a monolith tailored for this task---with ample opportunity for optimization from the compiler and runtime system---rather than a series of loosely coupled programs each written to handle many other cases; and the task at hand is an ideal fit for MapReduce system such as Hadoop.
-%> Finally, execution time is long enough (\todo{XXh})to amortize Hadoop's startup costs.
-%> % FIXME OSDI mention tabulation vs. sorting
-
-\heading{Take-aways}
-\sys can be applied to programs of notable size and complexity to offer significant acceleration.
-A broader take-away is that \sys is also able to extract parallelism from fragments that are not purely compute-intensive, \ie the usual focus of conventional parallelization systems.
-
-%% \todo{These programs correspond to much larger programs in other languages, a small fraction of which is automatically parallelizable with conventional systems.}
-
-%% \kk{I feel that in order to make any ``bigger'' conclusion we need to at least mention some Hadoop results. Otherwise it feels that any argument doesn't stand. A reviewer could say that }
-
-\subsection{Use Case: Wikipedia Web Indexing}
-\label{macro2}
-
-We now apply \sys to a large web-indexing script. % composed from various \todo{sources}~\cite{}.
-
-\heading{Program}
-This script reads a file containing Wikipedia URLs, downloads the pages,
-%% Coarsely, this script amounts to two distinct components communicating via \unix pipes.
-%% The \todo{first} downloads pages and extracts their URLs from their source, feeding back \todo{to} the start of the pipeline;
-%% the second
-extracts the text from HTML, and applies natural-language processing---\eg trigrams, character conversion, term frequencies---to index it.
-It totals 34 commands written in multiple programming languages.
-%% Collectively, they contain a total of 52 stages (first: 18; second: 34)
-
-\heading{Results}
-% To simplify execution (and avoid Wikipedia), we have saved a recent Wikipedia version on a server within the network and to redirect all \ttt{curl} requests there.
-%% As the \todo{first} component contains a cycle (URLs $\rightarrow$ HTML $\rightarrow$ URLs), \sys's conservative transformations do not attempt to parallelize it.
-The original script takes 191min to execute on 1\% of Wikipedia (1.3GB).
-With \ttt{width=}16, \sys brings it down to 15min ($12.7\times$), with the majority of the speedup coming from the HTML-to-text conversion. % which uses \ttt{pandoc}, a utility written in Haskell.
-
-%> \kk{If the front-end takes a lot of time (compared to the back-end), we might better not report it with the premise that it is a feedback cycle that cannot be parallelized by \sys.}
-
-\heading{Discussion}
-The original script contains 34 pipeline stages, thus the sequential version already benefits from task-based parallelism.
-It also uses several utilities not part of the standard POSIX/GNU set---\eg its \ttt{url-extract}ion is written in JavaScript and its \ttt{word-stem}ming is in Python.
-% , and its HTML-to-ASCII is achieved via \ttt{pandoc}, a binary utility written in Haskell.
-\sys can still operate on them as their parallelizability properties---\sta for \ttt{url-extract} and \ttt{word-stem}---can be trivially described by annotations.
-Several other stages are in \sta allowing \sys to achieve benefits by exposing data parallelism.
-
-\heading{Take-aways}
-\sys operates on programs with (annotated) commands outside the POSIX/GNU subsets and leads to notable speedups, even when the original program features significant task-based parallelism.
-
-% \subsection{Use Case: Container Orchestration}
-% \label{macro3}
-% 
-% As a third realistic workload, we use a script for orchestrating containers to test the latest built across various Linux distributions.
-% Compatibility testing is a common task for large software projects, commonly automated through shell scripting.
-% 
-% \heading{Program}
-% The 30-line orchestration script is unusual among our set of benchmarks in that it is not structured as a linear pipeline.
-% Rather, it makes extensive use of POSIX constructs common in the shell, including output redirection, conditionals, and here-documents.
-% It also combines commands from other classes outside \sta and \pur---more prominently \dfs---which are not parallelizable by \sys.
-% 
-% \heading{Results}
-% 
-% \heading{Take-aways}
-
-\subsection{Further Micro-benchmarks}
-\label{micro}
-
-%> \kk{I think the scoping of this should change to be both
-%>   micro-benchmarks and comparison with existing work. Especially since
-%>   we will have some form of comparison with Raftlib and phoenix. I
-%>   guess this section is trying to establish \sys limits by comparing
-%>   it with similar (but not identical) work.}
-
-% This section attempts to compare \sys with best-of-breed alternatives for particular taks.
-As there are no prior systems directly comparable to \sys, we now draw comparisons with two specialized cases that excel within smaller fragments of \sys's proposed domain.
-
-\pichskip{15pt}% Horizontal gap between picture and text
-\parpic[r][t]{%
-  \begin{minipage}{40mm}
-    \includegraphics[width=\linewidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}%
-    %% \captionof{figure}{This is a tiger.}
-  \end{minipage}
-}
-\heading{Parallel Sort}
-%% \kk{Explain what is the purpose of the
-%%   microbenchmark. We want to establish that \sys scalability is as
-%%   good as it could be.}
-We first compare a GNU \ttt{sort} parallelized by \sys ($S_{p}$) \vs the same \ttt{sort} with the 
-%% \begin{wrapfigure}[10]{r}{0.25\textwidth}
-%%   \centering
-%%   %% \begin{center}
-%%     \includegraphics[width=0.25\textwidth]{\detokenize{./figs/sort_baseline_comparison_scaleup.pdf}}
-%%   %% \end{center}
-%% %   \vspace{-15pt}
-%% % \caption{ \textbf{Comparison with \ttt{sort -{}-parallel}.} }
-%% \end{wrapfigure}
-\ttt{--parallel} flag set ($S_{g}$).\footnote{
-  Both \ttiny{sort}s use the same buffer size internally~\cite{sortp:15}.
-}
-% The results are shown on the side-figure below.
-While \ttt{--parallel} is not a general solution, the comparison serves to establish a baseline for \sys.
-$S_{g}$'s parallelism is configured to $2\times$ that of $S_{p}$'s \ttt{--width} (\ie the rightmost plot point for $S_{g}$ is for \ttt{--parallelism=128}), 
-to account for \sys's additional runtime processes.
-% and as such we note that the measurements are harsh for \sys.
-
-A few points are worth noting.
-$S_{p}$ without \ttt{eager} performs comparably to $S_{g}$, and with \ttt{eager} it outperforms $S_{g}$ ($\sim2\times$);
-  this is because \ttt{eager} adds intermediate buffers %(on disk)
-  that ensure CPU utilization is high.
-%% On the other hand, \ttt{sort --parallel} sorts the input trying to
-%% minimize additional memory (so doesn't allocate intermediate
-%% buffers).
-%% %% \kk{If we want, we can add here numbers where we allow sort --parallel
-%% %%   to use additional ram for intermediate buffers (using the -S flag).}
-%% Of course, it would be possible to manually create a longer pipeline
-%% and use intermediate buffers to speed this up, but this defeats the
-%% purpose of having a simple solution.
-$S_{g}$ indicates that \ttt{sort}'s scalability is inherently limited (\ie due to \ttt{sort}, not \sys);
-  this is why all scripts that contain \ttt{sort} (\eg \S\ref{ours}--\ref{macro2}) are capped at $8\times$ speedup.
-The comparison also shows \sys's benefits to command developers:
-  a low-effort parallelizability annotation achieves better scalability than a custom flag (and underlying parallel implementation) manually added by developers.
-
-%% \kk{There is a weird thing in the results. That the sequential one
-%%   with --parallel=1 takes more than the \sys sequential one.}
-
-\heading{GNU Parallel}
-We compare \sys to \ttt{parallel} (v.20160422), a GNU utility for running other commands in parallel~\cite{Tange2011a}, on a small bio-informatics script.
-% Using its \ttt{--jobs} flag, we  redirect input from a file to 
-% 554 / 330 
-Sequential execution takes 554.8s \vs \sys's 128.5s (4.3$\times$), with most of the overhead coming from a single command---\ttt{cutadapt}.
-
-% % https://github.com/marcelm/cutadapt/issues/157
-% % https://www.biostars.org/p/123237/
-% \begin{lstlisting}[language=sh, numbers=none]
-% 	uniq | sort | parallel cutadapt-fastqc
-% \end{lstlisting}
-% % -S server01,server02 
-% % -u -j24 --env PATH,PYTHONPATH,cutadapt_parallel
-% % --workdir $PWD cutadapt_parallel {}
-
-There are a few possible ways users might attempt to use GNU \ttt{parallel} on this program.
-They could use it on the bottleneck stage, assuming they can deduce it, bringing execution down to 304.4s ($1.8\times$ speedup).
-Alternatively, they could (incorrectly) sprinkle \ttt{parallel} across the entire program.
-% ---a strategy simplified by \ttt{parallel}'s \ttt{stdin} redirection feature.
-This would lead to 3.2$\times$ performance improvements but incorrect results with respect to the sequential execution---with 92\% of the output showing a difference between sequential and parallel execution.
-\sys's conservative program transformations are not applied in program fragments with unclear parallelizability properties.
-	% or commands whose inputs depend on earlier outputs.
-% GNU \ttt{parallel} is oblivious to command semantics, risking breaking program semantics as in this case.
-
-% Parallel is a great program for its target audience.
-
-%> \heading{Manual Parallelization}
-%> Alternatively, a user might attempt to manually parallelize individual commands by using the abstractions provided by the POSIX shell.
-%> The benefits depend on several factors, such as whether a command is I/O- or CPU-bound.
-%> 
-%> \heading{Streaming Framework}
-%> In a different attempt to identify how far \sys is from ideal performance, we start from programs already written in a high-performance parallel framework and rewrite as shell programs.
-%> Raftlib~\cite{} is a C++ library enabling streaming and dataflow parallelism without the fault-tolerance overheads of distributed systems.
-%> We rewrite three Raftlib benchmarks---\ttt{wordcount}, \ttt{XX}, and \ttt{XX}---as shell scripts.
-%> % These benchmarks were relatively easy found relatively easy to port and match many of the workloads one would expect to see in the shell as shell pipelines.
-
-% \heading{Graph Transformations}
-% \sys's analysis and generation overheads remain collectively under $5s$ even for the high-scalability benchmarks.
-% We decided to push \sys's limits by constructing an artificial 1000-stage dataflow region---a worst-case workload that is nowhere near the ones observed in practice.
-% We find the resulting times quite acceptable:
-%   for a $2\times$-parallelization \sys takes $8.23s$ for parsing the script and $3.22s$ for analysis and optimization.
-% Pushing parallelization factor to $100\times$, \sys's runtime jumps to 5 minutes and 8 seconds to perform the analysis, producing an output dataflow graph that has more than 100 thousand edges in a file of about 11MB.
-% \kk{I propose removing this subsection. 5 minutes is too much time for compilation and this happens atm because of a pathological scenario. The compilation should never take more than a second.}
-
-% \heading{Take-aways}
-% \sys does not improve performance when running the risk of breaking correctness. % This should go in the intro
-% % Compared to \sys, \ttt{parallel} can lead to a combination of smaller performance improvements, incorrect execution with respect to the sequential program, and higher rewriting effort.
-% In highly-parallel benchmarks, \sys can lead to performance that is comparable to highly specialized streaming frameworks---at a fraction of the cost in terms of manual development.
-
-\section{Related Work}
-\label{related}
-
-Existing techniques for exploiting parallelism are not directly comparable to \sys, either because they require significantly more \emph{user} effort (see \sx{intro} for the distinction between users and developers); or are too specialized, targeting narrow domains or custom programming abstractions.
-%% Techniques for re-writing programs to exploit parallelism fall under a wide spectrum that ranges from fully manual (but, ideally, optimal or more general) to fully automated (but potentially sub-optimal or specialized).
-%% \kk{It is not very clear what the following sentence tries to achieve.}
-%% Following the distinction between \emph{users} and \emph{developers} established earlier~\sx{intro}, we note correctness trade-offs associated with approaches that require user input for the discussion below:
-%%    users---often neither experts in parallelism nor educated on the internals of the code they depend on---can easily break the correctness of sequential programs, especially when these depend on existing code (such as is the case with shell commands).
-
-\heading{Parallel Shell Scripting}
-Utilities exposing parallelism on modern \unix{}es---\eg \ttt{qsub}~\cite{gentzsch2001sun}, \textsc{SLURM}~\cite{yoo2003slurm}, % \textsc{GNU} 
-  \ttt{parallel}~\cite{Tange2011a}---are limited to embarrassingly parallel (and short) programs and are predicated upon explicit and careful user invocation:
-  users have to navigate through a vast array of different configurations, flags, and modes of invocation to achieve parallelization without jeopardizing correctness.
-%% I like the part about the unix philosophy, but I want to really make it a side claim.
-The main claim, and the reason why we don't compare against all of these solutions, is that they require user effort and they cannot just be given a script to parallelize.
-%%
-%% and are limited to embarrassingly parallel (and short) programs.
-%% Often these commands provide options to support an array of special sub-cases---a stark contradiction to the celebrated \unix philosophy.
-For example, \ttt{parallel} contains flags such as \ttt{--skip-first-line}, \ttt{-trim}, and \ttt{--xargs},
-%% (which a \unix user can achieve using \ttt{head}, \ttt{sed}, and \ttt{xargs})
-%% kk: I think that mentioning that they can be written using UNIX commands doesn't make the point stronger.
-and introduces (and depends on) other programs with complex semantics, such as ones for SQL querying and CSV parsing.
-In contrast, \sys manages to parallelize large scripts correctly with minimal-to-zero user effort.
-%% embraces the \unix philosophy, simply attempting to rewrite scripts more efficiently with minimal user effort.
-
-Several shells~\cite{duff1990rc, mcdonald1988support, dagsh:17} add primitives for non-linear pipe topologies---some of which target parallelism.
-Here too, however, users are expected to manually rewrite scripts to exploit these new primitives, contrary to \sys.
-
-Recently, Smoosh~\cite{smoosh:18} argued for making concurrency explicit via shell constructs.
-The argument is dissimilar from \sys's, which argues for mostly automated (and correct) parallelization---hence \emph{light-touch} parallel scripting.
-
-Developed independently and at the same time with \sys, POSH~\cite{raghavan2020posh} is a framework for running shell scripts over remote storage by offloading the I/O-intensive portions to proxy servers closer to the data.
-POSH focuses on co-locating distributed operation, and introduces a novel scheduling algorithm. 
-\sys instead focuses on the parallelization of CPU-intensive scripts, by transforming them to DFGs, applying transformations, and then transforming them back to parallel shell scripts augmented with \sys's runtime primitives to be executed on a standard shell.
-%  non-distributed parallelism, is location-agnostic, and focuses on correctness through its dataflow transformations.
-
-% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
-\heading{Low-level Parallelization}
-Instruction-level parallelization has a long history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
-These systems operate at a lower level than \sys (\eg that of instructions or loops rather than the boundaries of programs that are part of a script), within a single-language or single-target environments, and require source modifications.
-
-More recent work focuses on extracting parallelism from domain-specific programming models~\cite{cilk5, streamIt, galois} and interactive parallelization tools~\cite{parascope, ipat}.
-These tools simplify the expression of parallelism, but still require significant user involvement in discovering and exposing parallelism.
-% Moreover, the insights behind these attempts are significantly different from \sys's, as they extract parallelism statically during compilation instead of dynamically during runtime.
-%Despite significant progress~\cite{}, it remains due to its need for complex program analysis and the unknown factors (such as input data range) during compilation.
-
-\heading{Correct Parallelization of Dataflow Graphs}
-The DFG is a prevalent model in several areas of data processing (including batch-~\cite{mapreduce:08, spark:12} and stream-processing ~\cite{murray2013naiad, carbone2015flink}).
-Despite its popularity, most systems perform optimizations that do not preserve semantics, introducing subtle erroneous behaviors.
-Recent work~\cite{HSSGG2014, SHGW2015, MSAIT2019} attempts to address this issue by performing optimizations only in cases where correctness is preserved.
-\sys draws inspiration from these efforts, as it attempts transformations that maintain the program's correctness with respect to the sequential execution.
-Its  DFG model, however, is different and captures ordering constraints.
-This is due to the intricacies of the \unix model---\eg streams, argument processing, and concatenation operators.
-% \kk{eurosys: I could see this being deleted in need of space.}
-
-% As a result, previous work is not directly applicable.
-
-\heading{Parallel Userspace Environments}
-By focusing on simplifying the development of distributed programs, a plethora of environments inadvertently assist in the construction of parallel software.
-Such systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} or languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide many of the challenges of dealing with concurrency as long as developers leverage the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
-% A notable example is Plan9's \ttt{rc} shell that provide an interface for the construction of general programs.
-Even shell-oriented efforts such as Plan9's \ttt{rc} are not backward-compatible with the \unix shell, and often focus primarily on hiding the existence of a network rather than automating parallel processing.
-% simplify many of the problems of distribution and
-
-\heading{Parallel Frameworks}
-Several frameworks~\cite{streamit:02, brook:04, phoenix:11, raftlib:17} offer fully automated parallelism as long as special primitives are used---\eg map-reduce-style primitives for Phoenix~\cite{phoenix:11}.
-These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
-By targeting specific classes of computation (\emph{viz.} \sys's parallelizability), they are significantly optimized for their target domains.
-% Lack of generality, no formal guaratees, their setup and is too tedious 
-\sys chooses a more general approach:
-   it does not require setting up a new framework for each new class of computation used, nor rewriting different parts of the computation using a different set of abstractions provided by each framework.
-%% It also takes a more principled approach: by developing the
-%%   semantic model outlined earlier, it can focus on transformations
-%%   that are correct with respect to the sequential program.
-%% \kk{the other approaches are also principled and possibly event more correct.}
-
-% \kk{Mention something about Nebula (as an OSDI reviewer asked for.)}
-
-Dryad~\cite{isard2007dryad} is a distributed system for dataflow graphs.
-Dryad offers a scripting language, Nebula, that allows using shell commands such as \ttt{grep} or \ttt{sed} in place of individual dataflow nodes.
-The main difference with \sys is that in Dryad the programmer needs to explicitly express the dataflow graph, which is then executed in a distributed fashion, whereas Pash automatically parallelizes a given shell script by producing a parallel script that runs on an unmodified shell of choice.
-
-% % Although still a research prototype, \sys's setup and deployment is simpler.
-% It also does not require these specialized frameworks, manually (re)write , and 
-% Different from \sys, however, 
-% Due to the specialization of these primitives, however, these systems suffer from a lack of generality ;
-% % and they still require developers to compose their work using the pipelines
-% Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
-
-% \heading{Annotation-based Transformations}
-% \sys inspiration from recent systems such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19} which, as long as a developer has provided a few high-level annotations, they take care of automating scale-out.
-% % Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
-% These systems works on single-language, single-runtime environments 
-% 
-% For the critical points where user input is required, they provide a
-% high-level declarative DSL with varying (but limited) degrees of
-% expressive power.
-% % Annotations are used by ---complicated by the use of third-party
-% % libraries similar to \sys's extensions.
-% Unfortunately, a challenge with this approach is that users composing
-% applications by using third-party libraries do not understand their
-% parallelizability details well-enough to annotate them.
-% 
-% \sys occupies a different point in the design space, by removing any work from composers and adding a small load to developers.
-% % More specifically, it is designed to be used automatically by composers,
-% Its DSL is designed to be more heavy-weight than Ignis' declarative recipes or Mozart's split annotations, but geared towards the developers of these components---not the users composing them.
-
-% It also provides a more principled approach than both of these systems.
-% Their annotations language is not as principled as \sys's---and one that is critical in ensuring that the dataflow analyses preserves the semantics expected by the developer
-% The side-effects present in the shell dwarf those of functional languages
-
-% https://ecommons.cornell.edu/bitstream/handle/1813/6508/85-668.pdf?sequence=1
-% The growth of the web led to specialized frameworks for massively distributed computation~\cite{mapreduce:08, spark:10, naiad:13}.
-% While they Moreover, these systems take advantage of functional purity;
-%   for programs that are not data-intensive processing pipelines (\eg web servers), purely functional code is generally responsible for only a small fraction of the program runtime.
-
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
-% \caption{
-%   \textbf{Manual--automated distribution spectrum.}
-% 	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness.
-% }
-% \vspace{-15pt}
-% \label{fig:spectrum}
-% \end{figure}
-%
-
-% \kk{exoume ena semantic model to opoio mporei na xrhsimopoiithei gia
-%   na ginoun formally proven claims gia to equivalence preservation twn
-%   optimizations etc.}
-
-%% Possible points to add: These systems introduce data parallelism by
-%% parallelizing nodes of the dataflow graph.  Unfortunately, they
-%% often do not preserve the model's semantics, blurring the lines
-%% between specification, optimization, and implementation.  In
-%% contrast, \sys's dataflow model inherently supports the data
-%% parallelism found in shell pipelines.  More specifically, it
-%% develops a set of parallelization-exposing optimizations
-%% represented as semantics-preserving graph transformations,
-%% effectively exposing data parallelism as task parallelism.
-
-%% \cite{HSSGG2014, SHGW2015} discuss optimizing transformations, but
-%% their correctness is established using informal arguments based on
-%% operational intution.
-
-%% The paper \cite{MSAIT2019} proposes a denotational semantic framework
-%% for stream processing where the data streams are seen as partial
-%% orders, and establishes the soundness of some common parallelizing
-%% transformations on dataflow graphs.
-
-%% \km{Some very early references on the dataflow model of computation
-%% (maybe there are relevant?): \cite{KM1966, D1974Dataflow, K1974KPN,
-%% KMacQ1977}}
-
-\section{Conclusion}
-\label{discussion}
-
-This paper presented \sys, a system allowing shell users to parallelize shell programs mostly automatically.
-% a source-to-source shell compiler and associated runtime library
-\sys consists of (i) a source-to-source compiler, which transforms scripts to DFGs, parallelizes them, and transforms them back to scripts; and (ii) a runtime component that addresses several practical challenges related to performance and correctness.
-Our evaluation on a wide variety of \emph{unmodified} scripts shows that \sys achieves significant speedups due to parallelism (total avg: $6.7\times$) in the vast majority of them (39/44), including scripts that perform mundane tasks such as downloading, extracting, and cleaning data, which are usually not the focus of parallelization systems.
-\sys is open source and available for download online~[Blind].
-
-% The compiler operates by converting dataflow regions into dataflow graphs and then applying iterative transformations based on the characteristics of commands in these regions.
-% Characteristics are known by the  use of classification and annotations (for certain classes), which \sys offers for commands in POSIX and Coreutils used frequently.
-% \sys's \unix-aware runtime primitives address several practical issues related to performance and correctness.
-% Our evaluation of \sys on a wide variety of \emph{unmodified} scripts shows that the vast majority (39/44) see significant speedups (average over all is $6.7\times$, up to $61\times$).
-% Applying \sys to classic and modern \unix scripts shows significant speedups---in one set of benchmarks, the compiler's average speedup is $9.3\times$, jumping to $13.2\times$ due to runtime optimizations.
-
-% Its runtime component provides orchestration and planning support during the execution of the program.
-% \sys's current limitations revolve around shell expansion.
-% Currently, \sys does not attempt to parallelize program fragments for which variables have not been expanded.
-% Without substantial effort, \sys could be modified to at least suggest expansion of these fragments so that developers provide concrete values.
-% More significant effort would be required however to fully exploit latent parallelism in fragments that have not been expanding.
-% This effort would require tying \sys with the shell interpretation rather than having it as a compilation pass and is left for future work.
-%
-% Despite limitation related to expansions,
-%   \sys can lead to significant benefits for shell users.
-% Experiments with real programs show substantial speedups and the ability to operate on large input datasets, all with minimal or zero user effort.
-
-% \heading{Limitations}
-% Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.
-% 
-% \sys does not adequately handle failures or network partitions in the general sense, which would require re-scheduling passes on failed replicas.
-% Prior work on fault-tolerant distributed stream processing can be of significant aid here.
-% 
-% The logic of the planner is quite simplified, attempting only straightforward placement of tasks to nodes (and using a hard-coded, homogeneous node structure).
-% Prior work on operator placement of distributed dataflow graphs can be used to build a more sophisticated planner.
-% 
-% \sys is conservative in the shell subsets that it handles, in order to
-% avoid introducing unsafe behaviours. The point of this work was not to be
-% able to handle a complete distributable subset of the shell, but
-% rather to use significant part of it to demonstrate performance benefits.
-% 
-% % Shell is highly dynamic, thus making it impossible to design sound
-% % meaningful analyses to determine distributable regions and maximal
-% % parallelization. In order to address this particularity, \sys could be
-% % tighly integrated with a posix compliant shell
-% % interpreter~\cite{smoosh:20} so that dynamic information can be
-% % acquired selectively to aid the analysis and optimization process.
-% 
-% 
-% \heading{Future Work}
-% % \label{limitation}
-% % \item No cycles (multiple commands writing and reading from the same file)
-% There are several worthwhile directions for future work.  One
-% direction would be an extension of the parallelizability analysis of
-% \sys, in order to guide it using dynamic information. This could
-% enable a formal study of the analysis~\sx{impl} to show that it
-% returns regions that are safe to distribute.  Recent work on
-% formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
-% such an attempt possible.
-% 
-% Another direction is to extend \sys to handle other classes outside \sta and \pur.
-% These two classes enjoy a certain popularity with quick, one-off pipelines and are the easiest to distribute, but 
-% it would be interesting to expand to \dfs (with the use of a distributed file-system) and \sid (with the use of transactional protocols).
-% This direction would naturally introduce considerations about replication, consistency, and failure recovery.
-% 
-% There are two additional assumptions that have to be satisfied in
-% order for the distributed implementation that is generated by Dish to
-% have equivalent behaviour with the original implementation. First of
-% all, there should be no external signals to the commands while they
-% execute. At the moment Dish assumes no faults, and that communication
-% always succeeds. This assumption can be lifted by extending Dish with
-% a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
-% command reads and writes to arbitrary files that are not mentioned in
-% their arguments. This is a necessary assumption, since commands are
-% considered as black boxes and the only information that can be
-% inferred about them is through their arguments.
-
-% \subsection*{Acknowledgements}
-% % Dumping people so that we don't forget
-% We want to thank 
-%   Andr\'e DeHon,
-%   Ben Karel,
-%   Caleb Stanford,
-%   Thurston Dang,
-%   Jean-S\'ebastien L\'egar\'e, and 
-%   Nick Roessler.
-% This material is based upon work supported by DARPA contract no. HR00112020013 (ALRSCC) and NSF award CCF 1763514.
-
-%% \grantsponsor{GS100000001}{National Science
-%%   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%% No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%% No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%% conclusions or recommendations expressed in this material are those
-%% of the author and do not necessarily reflect the views of the
-%% National Science Foundation.
-
-%% Bibliography
+\bibliographystyle{ACM-Reference-Format}
 {\small
-\bibliography{./bib}
+  \bibliography{../EuroSys21/bib}
 }
 
-% \appendix
-
-% \section{Annotation Language Grammar}
-% \label{alg}
-
-% \Cref{fig:dsl} shows the complete grammar of the parallelizability
-% annotation language.
-
-% \begin{figure}
-%   \centering
-%   \begin{grammar}
-%     <option> ::= `-' <string>
-
-%     <category> ::= `stateless' | `pure' | ...
-
-%     <maybe-int> ::= | <int>
-
-%     <arg> ::= `args[' <int> `]'
-
-%     <args> ::= <arg>
-%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
-
-%     <input> ::= `stdin' | <args>
-
-%     <inputs> ::= <input>
-%     \alt <input> `,' <inputs>
-
-%     <output> ::= `stdout' | <arg>
-
-%     <outputs> ::= <output>
-%     \alt <output> `,' <outputs>
-
-%     <option-pred> ::= <option>
-%     \alt `value' <option> = <string>
-%     \alt `not' <option-pred>
-%     \alt <option-pred> `or' <option-pred>
-%     \alt <option-pred> `and' <option-pred>
-
-%     <assignment> ::= `(' <category>, `[' <inputs> `]' `,' `[' <output> `]' `)'
-
-%     <predicate> ::= <option-pred> `=>' <assignment>
-
-%     <pred-list> ::= `|' <predicate> <pred-list>
-%     \alt `|' `otherwise' `=>' <assignment>
-
-%     <command> ::= <name> `\{' <pred-list> `\}'
-
-%     <command-list> ::= <command>
-%     \alt <command> <command-list>
-%   \end{grammar}
-%   \caption{
-%   \textbf{Parallelizability description language.}
-%     The DSL captures important information regarding the parallelizability of a command.
-%   }
-%   \label{fig:dsl}
-% \end{figure}
-
-%% %% Appendix
-%% \appendix
-%% \section{Scripts used in the evaluation}
-
-%% This appendix contains the source code of the scripts used in the evaluation of
-%% the \sys. They are part of the codebase (released as open source with the camera
-%% ready), and are provided here only to aid the reviewers.
-
 \end{document}

### 2020-11-30 compiler/execute_evaluation_scripts.sh
Add new experimentation script in Deleting eager intermediate files... Executing: test_script Generating input and intermediary scripts... be patient... Flag: -a and comment out the configurations that we dont want to run Note that I only added the -a configuration after the semicolon since this is the 'complete' PaSh configuration that adds both eager and split

@@ -11,35 +11,36 @@ gnu_parallel_flag=0
 
 n_inputs=(
     2
-    4
-    8
-    16
-    32
-    64
+    # 4
+    # 8
+    # 16
+    # 32
+    # 64
 )
 
 microbenchmarks=(
-    'grep;-n;-e'           # PLDI
-    'minimal_sort;;-n;-e'         # PLDI
-    'minimal_grep;-n;-e'         # PLDI
-    'topn;;-n;-e'                 # PLDI
-    'wf;;-n;-e'                   # PLDI
-    'spell;-e;-a'                # PLDI
-    'shortest_scripts;;-n;-e'     # PLDI
-    # micro_1000           # PLDI
+    # 'grep;-n;-e'           # PLDI
+    # 'minimal_sort;;-n;-e'         # PLDI
+    # 'minimal_grep;-n;-e'         # PLDI
+    # 'topn;;-n;-e'                 # PLDI
+    # 'wf;;-n;-e'                   # PLDI
+    # 'spell;-e;-a'                # PLDI
+    # 'shortest_scripts;;-n;-e'     # PLDI
+    # # micro_1000           # PLDI
 
-    'bigrams;-e;-a'              # TODO: Fix bug. Run with good split.
-    'alt_bigrams;;-n;-e'          # Optimized version of Bigrams
-    'diff;;-n;-e'                 # TODO: Optimize diff
-    'set-diff;;-n;-e'             # TODO: Handle redirection after reduce
+    # 'bigrams;-e;-a'              # TODO: Fix bug. Run with good split.
+    # 'alt_bigrams;;-n;-e'          # Optimized version of Bigrams
+    # 'diff;;-n;-e'                 # TODO: Optimize diff
+    # 'set-diff;;-n;-e'             # TODO: Handle redirection after reduce
 
     # sort                 # For comparison with sort --parallel
 
     ## Tests
     # deadlock_test          # Test to check deadlock prevention using drain_stream
-    'double_sort;;-n;-e;-a'            # Checks maximum peformance gains from split
+    # 'double_sort;;-n;-e;-a'            # Checks maximum peformance gains from split
 
     ## TODO: Add some more for OSDI
+    'test_script;-a'
     # wc                   # Extremely simple
     # page-count           # TODO: Change it so that it is one pipeline and it has many files
     # sieve                # Nice to show posix compliance

### 2020-11-30 evaluation/plots/microbenchmarks-table.tex
Modify  and add the new updates plots

--- 2020-11-11 evaluation/plots/microbenchmarks-table.tex
@@
 nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
-sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+sort ~&~ $\tsta, \tpur$ & 10~MB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
 top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
@@
 bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
-set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 51m32.313s & 155 & 635 & 0.321s & 4.358s & two pipelines merging to a \tti{comm} \\
-sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~MB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~MB & 51m32.313s & 155 & 635 & 0.316s & 4.358s & two pipelines merging to a \tti{comm} \\
+sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~MB & 0.687s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
 shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\

### 2020-11-30 compiler/gather_results.py

@@ -42,7 +42,8 @@ all_experiments = ["minimal_grep",
                    "bigrams",
                    "alt_bigrams",
                    "set-diff",
-                   "double_sort"]
+                   "double_sort",
+                   "test_script"]
 
 pretty_names = {"minimal_grep" : "nfa-regex",
                 "minimal_sort" : "sort",
@@ -55,7 +56,8 @@ pretty_names = {"minimal_grep" : "nfa-regex",
                 "shortest_scripts" : "shortest-scripts",
                 "diff" : "difference",
                 "set-diff" : "set-difference",
-                "double_sort" : "sort-sort"}
+                "double_sort" : "sort-sort",
+                "test_script" : "temp"}
 
 structures = {"minimal_grep" : "$3\\times\\tsta$",
               "minimal_sort" : "$\\tsta, \\tpur$",

### 2020-12-05 papers/SOSP21/draft.tex
Update paper notes

@@
 \begin{abstract}
+This paper presents \sys, a system for parallelizing \unix shell scripts. 
+To address performance and correctness concerns 
+\sys introduces an ahead-of-time/just-in-time compilation hybrid to address key runtime challenges specific to the \unix shell---\eg parameter and variable expansion, command substitution, and word splitting.
+A set of high-performance primitives such as a windowed \ttt{drain}er and token-based \ttt{split}ter offer further  performance and correctness improvements.
+A resource-aware superoptimization component explores the trade-off space to identify configurations 
+We evaluate \sys across a variety of hardware and software configurations using an extensive suite of more than 3K scripts including all 418 locale-independent tests from the POSIX suite.
+% (1) all 418 locale-independent tests from the POSIX suite,
+% (2) 112 scripts from the literature,
+% (3) 
+\sys's execution produces results that are correct with respect to sequential execution, significantly speeding up parallelizable scripts with sub-second elasticity characteristics.
+% important for interactive use.
 \end{abstract}
@@
 \label{intro}
+
+The \unix shell is an environment---often interactive---for composing scripts written in a plethora of programming languages.
+This language-agnosticism, coupled with \unix's toolbox philosophy~\cite{mcilroy1978unix}, makes the shell the primary choice for specifying succinct and simple scripts for data processing, system orchestration, and other automation tasks.
+
 \section{Example \& Overview}

### 2020-12-23 papers/SOSP21/draft.tex
Update notes

@@
 This paper presents \sys, a system for parallelizing \unix shell scripts. 
-To address performance and correctness concerns 
 \sys introduces an ahead-of-time/just-in-time compilation hybrid to address key runtime challenges specific to the \unix shell---\eg parameter and variable expansion, command substitution, and word splitting.
-A set of high-performance primitives such as a windowed \ttt{drain}er and token-based \ttt{split}ter offer further  performance and correctness improvements.
+A set of high-performance streaming primitives such as a windowed \ttt{drain}er and token-based \ttt{split}ter offer further  performance and correctness improvements.
 A resource-aware superoptimization component explores the trade-off space to identify configurations 
-We evaluate \sys across a variety of hardware and software configurations using an extensive suite of more than 3K scripts including all 418 locale-independent tests from the POSIX suite.
+We evaluate \sys across a variety of hardware and software configurations using an extensive suite of more than 1K scripts including all 418 locale-independent tests from the POSIX suite.
 % (1) all 418 locale-independent tests from the POSIX suite,
@@
 
+\heading{Methodology}
+
+\section{Related Work}
+
+\heading{Staged Compilation}
+(i) staged compilation, (ii) partial evaluation
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-12-31 papers/EuroSys21/draft.tex
Move papers into their own directory

--- 2020-10-11 paper/draft.tex
@@
-% TODO Add Ignis and Mozart
-% \nv{Make sure we use \ttt{width} everywhere}
-% \nv{Decide on parallelizable regions}
+% TODO Add Ignis, m31, and Mozart
+
+% arxiv: https://arxiv.org/abs/2007.09436
 
@@
 
+% \usepackage{amssymb}
+% \usepackage{cleveref}
 \usepackage{enumitem}
 \usepackage{booktabs}
-\usepackage{amssymb}
 \usepackage{soul}
@@
 
-\usepackage{minted} % for code listings
+% arxiv: use these options:
+% finalizecache: to generate cache before uploading to arxiv
+% frozencache: to load it, for when uploading to arxiv
+% https://tex.stackexchange.com/questions/280590/work-around-for-minted-code-highlighting-in-arxiv
+% https://github.com/gpoore/minted/issues/113#issuecomment-223451550
+\usepackage{minted}
 % \usemintedstyle{xcode}
@@
 %% Bibliography style
-\bibliographystyle{plain}
+\bibliographystyle{ACM-Reference-Format}
 %% Citation style
@@
 \affiliation{Unaffiliated}
+
+\renewcommand{\shortauthors}{N. Vasilakis*, K. Kallas*, K. Mamouras, A.  Benetopoulos, L. Cvetkovich}
 
@@
 An accompanying parallelizability study of POSIX and GNU commands---two large and commonly used groups---guides the annotation language and optimized aggregator library that \sys uses.
-Finally, \sys's 
 \sys's extensive evaluation over 44 unmodified \unix scripts shows significant speedups ($0.89$--$61.1\times$, avg: $6.7\times$) stemming from the combination of its program transformations and runtime primitives.
@@
 \maketitle
+% arxiv: use plan to remove header/footer
+\pagestyle{plain}
 
@@
 
+\kk{We need a way to instantiate the correspondence (bidirectional transformations) from commands to DFG nodes and back.
+This could be done with arbitrary python code in the general setting that checks arbitrary things for each command.
+We should allow that.
+For a big subset of commands though (how big) we come up with an annotation language that allows one to easily instantiate this correspondence.}
+
+\kk{Maybe rename this whole thing a framework for instantiating the node $\leftrightarrow$ command correspondence.}
+
 % TODO: \nv{Note command breadth/impossibility of our task.}
@@
 \label{ir}
+
+\TODO{Remember to make this more informal.}
 
@@
 \sys also performs a set of auxiliary transformations $t_{1-3}$ that are
-depicted in \Cref{fig:auxiliary-transformations}. If a node has many
+depicted in Fig.~\ref{fig:auxiliary-transformations}. If a node has many
 inputs, $t_1$ concatenates these inputs by inserting a \ttt{cat} node

### 2021-01-04 papers/EuroSys21/draft.tex

@@
-% TODO Add Ignis, m31, and Mozart
+% TODO CAMERA READY
+% * Add Ignis, m31, Mozart, ODFG, kq
+% * Make model  more informal.
+% * Talk annotations as a special case of the node<->command corr (pair of) function(s)
+
+% Detailed comments for camera ready
+% \kk{We need a way to instantiate the correspondence (bidirectional transformations) from commands to DFG nodes and back.
+This could be done with arbitrary python code in the general setting that checks arbitrary things for each command.
+We should allow that.
+For a big subset of commands though (how big) we come up with an annotation language that allows one to easily instantiate this correspondence.}
+% 
+% \kk{Maybe rename this whole thing a framework for instantiating the node $\leftrightarrow$ command correspondence.}
 
@@
 
-\author{Lazar Cvetkovich}
-\affiliation{Unaffiliated}
+\author{Lazar Cvetkovi\'{c}}
+\affiliation{University of Belgrade}
 
@@
 
-\kk{We need a way to instantiate the correspondence (bidirectional transformations) from commands to DFG nodes and back.
-This could be done with arbitrary python code in the general setting that checks arbitrary things for each command.
-We should allow that.
-For a big subset of commands though (how big) we come up with an annotation language that allows one to easily instantiate this correspondence.}
-
-\kk{Maybe rename this whole thing a framework for instantiating the node $\leftrightarrow$ command correspondence.}
-
 % TODO: \nv{Note command breadth/impossibility of our task.}
@@
 \label{ir}
-
-\TODO{Remember to make this more informal.}
 

### 2021-03-16 evaluation/plots/microbenchmarks-table-small.tex
Small fixes to some plots for final version

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
\toprule
Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
\midrule
nfa-regex ~&~ $3\times\tsta$ & 1~GB & 7m59.843s & 49 & \todo{X} & 0.305s & 0.000s & complex NFA regex \\
sort ~&~ $\tsta, \tpur$ & 10~GB & 2m6.756s & 77 & \todo{X} & 0.311s & 0.000s & \tti{sort}ing \\
top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 7m6.963s & 174 & \todo{X} & 0.404s & 0.000s & double \tti{sort}, \tti{uniq} reduction \\
wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 2m10.746s & 174 & \todo{X} & 0.385s & 0.000s & double \tti{sort}, \tti{uniq} reduction \\
spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 46.520s & 195 & \todo{X} & 0.412s & 0.000s & comparisons (\tti{comm}) \\
difference ~&~ $2\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 2m35.132s & 157 & \todo{X} & 0.661s & 0.000s & non-parallelizable \tti{diff}ing \\
bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 1m11.240s & 284 & \todo{X} & 0.452s & 0.000s & stream shifting and merging \\
set-difference ~&~ $5\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 5m2.111s & 173 & \todo{X} & 0.658s & 0.000s & two pipelines merging to a \tti{comm} \\
sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 3m8.657s & 154 & \todo{X} & 0.357s & 0.000s & parallelizable \tpur after \tpur \\
shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 3m4.233s & 142 & \todo{X} & 0.348s & 0.000s & long \tsta pipeline ending with \tpur \\
\bottomrule
\end{tabular*}

### 2021-03-16 evaluation/plots/microbenchmarks-table.tex
Small fixes to some plots for final version

@@
 nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
-sort ~&~ $\tsta, \tpur$ & 10~MB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
+sort ~&~ $\tsta, \tpur$ & 10~GB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
 top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
 wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
-spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
-difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
-bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
-set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~MB & 51m32.313s & 155 & 635 & 0.316s & 4.358s & two pipelines merging to a \tti{comm} \\
-sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~MB & 0.687s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
+spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 25m7.560s & 193 & 769 & 0.335s & 4.560s & comparisons (\tti{comm}) \\
+difference ~&~ $2\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
+bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.313s & 4.310s & stream shifting and merging \\
+set-difference ~&~ $5\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 51m32.313s & 155 & 635 & 0.316s & 4.358s & two pipelines merging to a \tti{comm} \\
+sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 31m26.147s & 154 & 634 & 0.293s & 3.255s & parallelizable \tpur after \tpur \\
 shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\

### 2021-03-16 compiler/gather_results.py

@@ -88,8 +88,8 @@ structures = {"minimal_grep" : "$3\\times\\tsta$",
               "alt_bigrams" : "$3\\times\\tsta, \\tpur$",
               "spell" : "$4\\times\\tsta, 3\\times\\tpur$",
               "shortest_scripts" : "$5\\times\\tsta, 2\\times\\tpur$",
-              "diff" : "$2\\times\\tsta, 3\\times\\tpur$",
-              "set-diff" : "$5\\times\\tsta, 2\\times\\tpur$",
+              "diff" : "$2\\times\\tsta, 2\\times\\tpur, \\tnpu$",
+              "set-diff" : "$5\\times\\tsta, 2\\times\\tpur, \\tnpu$",
               "double_sort" : "$\\tsta, 2\\times\\tpur$"}
 
 highlights = {"minimal_grep" : "complex NFA regex",
@@ -603,11 +603,12 @@ def plot_sort_with_baseline(results_dir):
     ax.plot(all_scaleup_numbers, baseline_sort_distr_speedup[1:], '-p', linewidth=0.5, label='sort --parallel')
     # ax.plot(all_scaleup_numbers, baseline_sort_opt_distr_speedup[1:], '-', linewidth=0.5, label='sort --parallel -S 30%')
 
+    plt.ylim(1, 8.6)
 
     plt.xticks(all_scaleup_numbers[1:])
     plt.legend(loc='lower right')
     # plt.title("Comparison with sort --parallel")
-
+    fig.set_size_inches(8, 5.5)
 
     plt.tight_layout()
     plt.savefig(os.path.join('../evaluation/plots', "sort_baseline_comparison_scaleup.pdf"),bbox_inches='tight')

### 2021-03-17 evaluation/plots/microbenchmarks-table-small.tex

@@
 spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 46.520s & 195 & \todo{X} & 0.412s & 0.000s & comparisons (\tti{comm}) \\
-difference ~&~ $2\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 2m35.132s & 157 & \todo{X} & 0.661s & 0.000s & non-parallelizable \tti{diff}ing \\
+difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 2m35.132s & 157 & \todo{X} & 0.661s & 0.000s & non-parallelizable \tti{diff}ing \\
 bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 1m11.240s & 284 & \todo{X} & 0.452s & 0.000s & stream shifting and merging \\
-set-difference ~&~ $5\times\tsta, 2\times\tpur, \tnpu$ & 10~GB & 5m2.111s & 173 & \todo{X} & 0.658s & 0.000s & two pipelines merging to a \tti{comm} \\
+set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 5m2.111s & 173 & \todo{X} & 0.658s & 0.000s & two pipelines merging to a \tti{comm} \\
 sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 3m8.657s & 154 & \todo{X} & 0.357s & 0.000s & parallelizable \tpur after \tpur \\

### 2021-03-17 evaluation/round_split/r-bell_grep.sh
compatibility changes to tests

@@ -10,7 +10,7 @@ file8=8.out
 file9=9.out
 rm -f *.out
 
-testFile="/home/ubuntu/pash/evaluation/scripts/input/1G.txt"
+testFile="/home/tamlu/pash/evaluation/scripts/input/1G.txt"
 batchSize=10000000
 if [ "$#" -gt "0" ]
  then
@@ -24,7 +24,8 @@ mkfifo $file1
 mkfifo $file2
 mkfifo $file3
 mkfifo $file4
-# mkfifo $file5
+mkfifo $file5
+mkfifo $file6
 
 # mkfifo $file7
 # mkfifo $file8
@@ -33,11 +34,14 @@ mkfifo $file4
 
 ../../runtime/r_split $testFile $batchSize $file1 $file2 &
 
-../../runtime/r_wrap grep 'Bell' < $file1 > $file3 &
-../../runtime/r_wrap grep 'Bell' < $file2 > $file4 &
+../../runtime/r_wrap tr A-Z a-z < $file1 > $file5 &
+../../runtime/r_wrap tr A-Z a-z < $file2 > $file6 &
+
+# ../../runtime/r_wrap grep 'Bell' < $file1 > $file5 &
+# ../../runtime/r_wrap grep 'Bell' < $file2 > $file6 &
 # ../r_wrap grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4' < $file7 > $file8 &
 
-../../runtime/r_merge $file3 $file4
+../../runtime/r_merge $file5 $file6
 
 # cat $testFile | grep '\(.\).*\1\(.\).*\2\(.\).*\3\(.\).*\4' > $file6
 # if cmp -s "$file6" "$file5"; then

### 2021-03-17 runtime/r_wrap.c
small changes to r_wrap.c

@@ -17,10 +17,12 @@ void processCmd(char* args[]) {
         int fdOut[2];
         if (pipe(fdIn) < 0) {
             perror("pipe failed");
+            exit(1);
         }
 
         if (pipe(fdOut) < 0) {
             perror("pipe failed");
+            exit(1);
         }
         
         int pid = fork();
@@ -31,15 +33,17 @@ void processCmd(char* args[]) {
             close(fdOut[WRITE_END]);
             close(fdIn[READ_END]);
             close(fdIn[WRITE_END]);
-            free(buffer);
+            // free(buffer);
             execvp(args[0], args);
             //shouldn't get here
             perror("Exec failed");
+            exit(1);
         } else {
             close(fdIn[READ_END]);
             close(fdOut[WRITE_END]);
             
             FILE* execOutFile = fdopen(fdOut[READ_END], "rb");
+            FILE* execInFile = fdopen(fdIn[WRITE_END], "wb");
             fcntl(fdOut[READ_END], F_SETFL, O_NONBLOCK);
 
             size_t outBufLen = 0, len = 0, currLen = 0;
@@ -49,19 +53,20 @@ void processCmd(char* args[]) {
             size_t tot_read = 0, readSize = 0;
             while (tot_read < blockSize) {
                 readSize = MIN(bufLen, blockSize-tot_read);
-                // fprintf(stderr, "reading stdin\n");
+                fprintf(stderr, "reading stdin\n");
                 if (fread(buffer, 1, readSize, stdin) != readSize) {
                     fprintf(stderr, "r_wrap: There is a problem with reading the block\n");
                     exit(1);
                 }
                 
-                // fprintf(stderr, "writing to fork\n");
+                fprintf(stderr, "writing to fork\n");
                 //Write to forked process
-                if (write(fdIn[WRITE_END], buffer, readSize) != readSize) {
-                    fprintf(stderr, "Failed wrtiting to pipe\n");
-                    exit(1);
-                }
-                // fprintf(stderr, "wrote to fork and reading\n");
+                safeWrite(buffer, 1 , readSize, execInFile);
+                // if (write(fdIn[WRITE_END], buffer, readSize) != readSize) {
+                //     fprintf(stderr, "Failed wrtiting to pipe\n");
+                //     exit(1);
+                // }
+                fprintf(stderr, "wrote to fork and reading\n");
[…]

### 2021-03-17 evaluation/round_split/r-minimal_grep.sh
improved performance of split -r (split without headers)

@@ -10,7 +10,7 @@ file8=8.out
 file9=9.out
 rm -f *.out
 
-testFile="/home/ubuntu/pash/evaluation/scripts/input/10M.txt"
+testFile="$PASH_TOP/evaluation/scripts/input/100M.txt"
 batchSize=100000
 if [ "$#" -gt "0" ]
  then

### 2021-03-17 runtime/r_split.c

@@ -1,28 +1,31 @@
 #include "r_split.h"
 
-void SplitByBytes(FILE* inputFile, int batchSize, FILE* outputFiles[], unsigned int numOutputFiles) {
+void SplitByBytes(FILE *inputFile, int batchSize, FILE *outputFiles[], unsigned int numOutputFiles)
+{
   int current = 0;
   int64_t id = 0;
   size_t len = 0;
-  FILE* outputFile = outputFiles[current];
-  
-  char* buffer = malloc(batchSize + 1);
+  FILE *outputFile = outputFiles[current];
+
+  char *buffer = malloc(batchSize + 1);
 
   // Do round robin copying of the input file to the output files
   // Each block has a header of "ID blockSize\n"
-  while ((len = fread(buffer, 1, batchSize, inputFile)) > 0) {
+  while ((len = fread(buffer, 1, batchSize, inputFile)) > 0)
+  {
     //write header
     writeHeader(outputFile, id, len);
 
     //write blocks
     fwrite(buffer, 1, len, outputFile);
-    
+
     current = (current + 1) % numOutputFiles;
     outputFile = outputFiles[current];
-    id+=1;
+    id += 1;
   }
 
-  if (len < 0) {
+  if (len < 0)
+  {
     perror(LOC);
     exit(1);
   }
@@ -31,23 +34,26 @@ void SplitByBytes(FILE* inputFile, int batchSize, FILE* outputFiles[], unsigned 
   free(buffer);
 }
 
-
-void SplitByLines(FILE* inputFile, int batchSize, FILE* outputFiles[], unsigned int numOutputFiles) {
+void SplitByLines(FILE *inputFile, int batchSize, FILE *outputFiles[], unsigned int numOutputFiles)
+{
   int current = 0;
   int64_t id = 0;
   size_t len = 0, headSize = 0, restSize = 0, prevRestSize = 0, blockSize = 0, bufLen = 0;
-  FILE* outputFile = outputFiles[current];
-  
-  char* buffer = malloc(batchSize + 1);
-  char* incompleteLine = malloc(batchSize + 1);
-  char* newLineBuffer = NULL; //only used when a newline character is not found in chunk
+  FILE *outputFile = outputFiles[current];
+
+  char *buffer = malloc(batchSize + 1);
+  char *incompleteLine = malloc(batchSize + 1);
+  char *newLineBuffer = NULL; //only used when a newline character is not found in chunk
 
[…]

### 2021-03-18 evaluation/results_archive/plots/microbenchmarks-tablesmall.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
\toprule
Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
\midrule
nfa-regex ~&~ $3\times\tsta$ & 1~GB & 7m59.843s & \todo{X} & \todo{X} & 0.000s & 0.000s & complex NFA regex \\
sort ~&~ $\tsta, \tpur$ & 10~GB & 2m6.756s & \todo{X} & \todo{X} & 0.000s & 0.000s & \tti{sort}ing \\
top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 7m6.963s & \todo{X} & \todo{X} & 0.000s & 0.000s & double \tti{sort}, \tti{uniq} reduction \\
wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 2m10.746s & \todo{X} & \todo{X} & 0.000s & 0.000s & double \tti{sort}, \tti{uniq} reduction \\
spell ~&~ $4\times\tsta, 3\times\tpur$ & 3~GB & 46.520s & 195 & \todo{X} & 0.324s & 0.000s & comparisons (\tti{comm}) \\
difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 2m35.132s & \todo{X} & \todo{X} & 0.000s & 0.000s & non-parallelizable \tti{diff}ing \\
bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 1m11.240s & 284 & \todo{X} & 0.301s & 0.000s & stream shifting and merging \\
set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~GB & 5m2.111s & \todo{X} & \todo{X} & 0.000s & 0.000s & two pipelines merging to a \tti{comm} \\
sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~GB & 3m8.657s & 154 & \todo{X} & 0.327s & 0.000s & parallelizable \tpur after \tpur \\
shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 3m4.233s & \todo{X} & \todo{X} & 0.000s & 0.000s & long \tsta pipeline ending with \tpur \\
\bottomrule
\end{tabular*}

### 2021-03-27 evaluation/results/plots/microbenchmarks-table.tex

\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} llllllll}
\toprule
Script ~&~ Structure & Input &Seq. Time & \multicolumn{2}{l}{\#Nodes(16, 64)} &\multicolumn{2}{l}{Compile Time (16, 64)} & Highlights \\
\midrule
nfa-regex ~&~ $3\times\tsta$ & 1~GB & 79m35.197s & 49 & 193 & 0.056s & 0.523s & complex NFA regex \\
sort ~&~ $\tsta, \tpur$ & 10~MB & 21m46.807s & 77 & 317 & 0.090s & 1.083s & \tti{sort}ing \\
top-n ~&~ $2\times\tsta, 4\times\tpur$ & 10~GB & 78m45.872s & 96 & 384 & 0.145s & 1.790s & double \tti{sort}, \tti{uniq} reduction \\
wf ~&~ $3\times\tsta, 3\times\tpur$ & 10~GB & 22m30.048s & 96 & 384 & 0.147s & 1.809s & double \tti{sort}, \tti{uniq} reduction \\
spell ~&~ $4\times\tsta, 3\times\tpur$ & 10~MB & 25m7.560s & 193 & 769 & 0.104s & 1.038s & comparisons (\tti{comm}) \\
difference ~&~ $2\times\tsta, 3\times\tpur$ & 10~GB & 25m49.097s & 125 & 509 & 0.186s & 2.341s & non-parallelizable \tti{diff}ing \\
bi-grams ~&~ $3\times\tsta, 3\times\tpur$ & 3~GB & 38m9.922s & 185 & 761 & 0.146s & 1.716s & stream shifting and merging \\
set-difference ~&~ $5\times\tsta, 2\times\tpur$ & 10~MB & 51m32.313s & 155 & 635 & 0.316s & 4.358s & two pipelines merging to a \tti{comm} \\
sort-sort ~&~ $\tsta, 2\times\tpur$ & 10~MB & 31m26.147s & 154 & 634 & 0.092s & 1.077s & parallelizable \tpur after \tpur \\
shortest-scripts ~&~ $5\times\tsta, 2\times\tpur$ & 85~MB & 28m45.900s & 142 & 574 & 0.328s & 4.657s & long \tsta pipeline ending with \tpur \\
\bottomrule
\end{tabular*}
